### 2024-03-13 analyses/simulate-sce/notebooks/splatter-ScPCA.Rmd
add dependencies and initial nb

---
title: "Simulating Single Cell experiments from ScPCA data"
output: html_notebook
---

```{r setup}
suppressMessages({
  library(SingleCellExperiment)
})
```

```{r}
sample_id <- "SCPCS000430"
library_id <- "SCPCL000750"
suffixes <- c( "_processed.rds", "_filtered.rds", "_unfiltered.rds")

infiles <- file.path("data", sample_id, paste0(library_id, suffixes))

sce <- readRDS(infiles[1])
```

```{r}
ncells <- 100
sce_subset <- sce[, sample.int(ncol(sce), ncells)]
sce_sim <- sce_subset
# splatter doesn't know new matrix types?
sim_params <- splatter::splatEstimate(as.matrix(counts(sce_subset)))
counts(sce_sim, withDimnames = FALSE) <- counts(splatter::splatSimulate(sim_params))

sce_sim <- sce_sim |>
  scater::logNormCounts()|>
  scater::runPCA(ncomponents = 10) |> # we don't need all the PCA components
  scater::runUMAP()
```

Are the same genes zero counted as before? That would be ideal, but maybe isn't the case.

```{r}
sce_sim
```

```{r}
metadata(sce_sim)$miQC_model <- NULL
readr::write_rds(sce_sim, paste0(infiles[1], "_sim_gz.rds"), compress = "gz")
readr::write_rds(sce_sim, paste0(infiles[1], "_sim_bz.rds"), compress = "bz2")
```

```{r}
sce2 <- readRDS(infiles[2])
ncells <- 100
sce2_subset <- sce2[, sample.int(ncol(sce), ncells)]
sce2_sim <- sce2_subset
# splatter doesn't know new matrix types?
sim_params <- splatter::splatEstimate(as.matrix(counts(sce_subset)))
counts(sce2_sim, withDimnames = FALSE) <- counts(splatter::splatSimulate(sim_params))
```

```{r}
metadata(sce2_sim)$miQC_model <- NULL
readr::write_rds(sce2_sim, paste0(infiles[2], "_sim_gz.rds"), compress = "gz")
readr::write_rds(sce2_sim, paste0(infiles[2], "_sim_bz.rds"), compress = "bz2")
```

### 2024-03-13 analyses/simulate-sce/notebooks/splatter-ScPCA.Rmd
Notebook version of sim complete

@@
 
+The data used in this notebook was previously downloaded with the following command:
+
+```
+./download-data.py --include "filtered,unfiltered,processed" --format "sce,anndata"
+```
+
+Note that we will only use the SCE data, but having the AnnData is useful for comparison in the final output.
+
 ```{r setup}
@@
 })
+
+# get the current data directory (data/current), tracing from to the repo root
+data_dir <- file.path(rprojroot::find_root(rprojroot::is_git_root), "data", "current")
+results_dir <- here::here("results", "simulated") # starts from this modules root, since it is an r project.
+```
+
+For now, we will pick a single library to work with.
+```{r}
+project_id <- "SCPCP000015"
+sample_id <- "SCPCS000490"
+library_id <- "SCPCL000822"
+
+suffixes <- c( "_processed.rds", "_filtered.rds", "_unfiltered.rds")
+
+infiles <- file.path(data_dir, project_id, sample_id, paste0(library_id, suffixes))
+outfiles <- file.path(results_dir, project_id, sample_id, paste0(library_id, suffixes))
+
+sce <- readRDS(infiles[1]) # start with the processed file
+```
+
+Start by creating a subset of the original data set and adjusting the metadata.
+
+```{r}
+ncells <- 100
+
+# reduce the data set to a smaller number of cells
+cell_subset <- sample.int(ncol(sce), ncells)
+sce_sim <- sce[, cell_subset]
+
+# remove participant id, but leave it as a string
+metadata(sce_sim)$sample_metadata$participant_id <- ""
+
+# remove any miQC model
+metadata(sce_sim)$miQC_model <- NULL
+
+# reduce the cell type data matrices, if present
+if (!is.null(metadata(sce_sim)$singler_results)) {
+  metadata(sce_sim)$singler_results <- metadata(sce_sim)$singler_results[cell_subset,]
+}
+if (!is.null(metadata(sce_sim)$cellassign_predictions)) {
+  metadata(sce_sim)$cellassign_predictions <- metadata(sce_sim)$cellassign_predictions[cell_subset,]
+}
+```
+
+Create a function to randomly select labels from a set, ensuring each label is included at least once.
+
+```{r}
+random_label<- function(label_set, n){
+  # randomly select n labels from the label set, ensuring each label is included at least once
+  # ensure labels are unique
+  label_set <- unique(label_set)
+  if (length(label_set) > n) {
+    stop("The number of labels must not be greater than the number requested")
+  }
+  r_labels <- sample(label_set, n, replace = TRUE)
+  # add each label at least once
+  idx <- sample.int(n, length(label_set))
+  r_labels[idx] <- label_set
+  return(r_labels)
+}
+
+```
+
+Modify the cluster and cell type labels in the subest SCE to ensure they are all represented
+
+```{r}
+# Adjust cluster/cell type labels, using original label sets
+
+# get label sets
+cluster_set <- unique(colData(sce)$cluster)
+cellassign_set <- unique(colData(sce)$cellassign_celltype_annotation)
+
+# create a mapping for singler ontology and annotation
+singler_df <- colData(sce)|>
+  data.frame()|>
+  dplyr::select(singler_celltype_ontology, singler_celltype_annotation) |>
+  dplyr::distinct()
+
+singler_dict <- setNames(
+  singler_df$singler_celltype_annotation,
+  singler_df$singler_celltype_ontology
+)
+
+colData(sce_sim)$cluster <- random_label(cluster_set, ncells)
+colData(sce_sim)$singler_celltype_ontology <- random_label(names(singler_dict), ncells)
+colData(sce_sim)$singler_celltype_annotation <- unname(
+  singler_dict[colData(sce_sim)$singler_celltype_ontology]
+)
+colData(sce_sim)$cellassign_celltype_annotation <- random_label(cellassign_set, ncells)
+```
+
+Now we can do the simulation.
+We will use the `splatter` package to simulate the data.
+Not sure whether we want to do simulations with groups, but we probably don't need to.
+
+```{r}
+# for simulation, get probabilities of each cluster
+# cluster_probs <- c(table(colData(sce)$cluster)/ncol(sce))
+sim_params <- splatter::splatEstimate(as.matrix(counts(sce_sim)))
+
+# sim_params <- splatter::setParams(sim_params, group.prob = cluster_probs)
+
+counts(sce_sim, withDimnames = FALSE) <- splatter::splatSimulate(sim_params)|>
+  counts()
+# recalculate dimension reduction
+sce_sim <- sce_sim |>
+  scater::logNormCounts()|>
+  scater::runPCA(10) |> # we don't need all the PCA components
+  scater::runUMAP(dimred = "PCA")
+```
+
+Replace and update column stats
+
+```{r}
+# store coldata column names to restore order later
+colData_names <- names(colData(sce_sim))
+
+mito_detected_factor <- colData(sce_sim)$subsets_mito_detected / colData(sce_sim)$detected
+
+# remove stats that will be recalculated
+remove_stats <- c(
+  "sum",
+  "detected",
+  "total",
+  "subsets_mito_sum",
+  "subsets_mito_detected",
+  names(colData(sce_sim))[grep("altExp_.*_(sum|detected|percent)", names(colData(sce_sim)))]
+)
+colData(sce_sim)[,remove_stats] <- NULL
+
+sce_sim <- scuttle::addPerCellQC(sce_sim)
+colData(sce_sim)$subsets_mito_sum <- colData(sce_sim)$sum * colData(sce_sim)$subsets_mito_percent / 100
+colData(sce_sim)$subsets_mito_detected <- round(colData(sce_sim)$detected * mito_detected_factor)
+
+# restore column order
+colData(sce_sim) <- colData(sce_sim)[,colData_names]
+
 ```
@@
 ```{r}
-sample_id <- "SCPCS000430"
-library_id <- "SCPCL000750"
-suffixes <- c( "_processed.rds", "_filtered.rds", "_unfiltered.rds")
-
-infiles <- file.path("data", sample_id, paste0(library_id, suffixes))
-
-sce <- readRDS(infiles[1])
+fs::dir_create(dirname(outfiles[1]))
+readr::write_rds(sce_sim, paste0(outfiles[1]), compress = "bz2")
 ```
-
-```{r}
-ncells <- 100
-sce_subset <- sce[, sample.int(ncol(sce), ncells)]
-sce_sim <- sce_subset
-# splatter doesn't know new matrix types?
-sim_params <- splatter::splatEstimate(as.matrix(counts(sce_subset)))
-counts(sce_sim, withDimnames = FALSE) <- counts(splatter::splatSimulate(sim_params))
-
-sce_sim <- sce_sim |>
-  scater::logNormCounts()|>
-  scater::runPCA(ncomponents = 10) |> # we don't need all the PCA components
-  scater::runUMAP()
-```
-
-Are the same genes zero counted as before? That would be ideal, but maybe isn't the case.
-
-```{r}
-sce_sim
-```
-
-```{r}
-metadata(sce_sim)$miQC_model <- NULL
-readr::write_rds(sce_sim, paste0(infiles[1], "_sim_gz.rds"), compress = "gz")
-readr::write_rds(sce_sim, paste0(infiles[1], "_sim_bz.rds"), compress = "bz2")
-```
-
-```{r}
-sce2 <- readRDS(infiles[2])
-ncells <- 100
-sce2_subset <- sce2[, sample.int(ncol(sce), ncells)]
-sce2_sim <- sce2_subset
-# splatter doesn't know new matrix types?
-sim_params <- splatter::splatEstimate(as.matrix(counts(sce_subset)))
-counts(sce2_sim, withDimnames = FALSE) <- counts(splatter::splatSimulate(sim_params))
-```
-
-```{r}
-metadata(sce2_sim)$miQC_model <- NULL
-readr::write_rds(sce2_sim, paste0(infiles[2], "_sim_gz.rds"), compress = "gz")
-readr::write_rds(sce2_sim, paste0(infiles[2], "_sim_bz.rds"), compress = "bz2")
-```

### 2024-03-14 analyses/simulate-sce/notebooks/splatter-ScPCA.Rmd
Spelling updates
Flesh out readme and update notebook intro a

@@
 ---
+
+This notebook was used as exploration for the `simulate-sce.R` script.
+The code here is not fully current with that script, but has been left here for reference about the development process and for future use.
 
@@
 
-Modify the cluster and cell type labels in the subest SCE to ensure they are all represented
+Modify the cluster and cell type labels in the subset SCE to ensure they are all represented
 

### 2024-03-15 analyses/hello-R/notebooks/hello.Rmd
Add notebook

---
title: "Hello OpenScPCA"
output: html_notebook
---

## Introduction

This notebook presents a simple example of an analysis in the OpenScPCA project, using R/Bioconductor.

We will go through the processed SCE objects in the project `data` library and and count the number of cells in each object, plotting the results.

### Setup

Load libraries and set initial paths.

```{r setup}
# load required libraries
suppressPackageStartupMessages({
  library("SingleCellExperiment")
  library("ggplot2")
})

# Find the repository root
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

# set current data dir
data_dir = file.path(repo_root, "data", "current")

# set results dir (using the analysis project file to find root)
results_dir <- here::here("results")
```

## Functions

First, we will define a function to load and count the cells in a SCE object, returning a one line data frame with the project ID, sample ID, library ID, and number of cells.

```{r}
# function to count cells in an sce file
# returns a data frame with ids and cell count
count_sce <- function(sce_file) {
  # load the SCE object
  sce <- readRDS(sce_file)

  # get stats and count the cells
  result_df <- data.frame(
      project_id = metadata(sce)$project_id,
      sample_id = metadata(sce)$sample_id,
      library_id = metadata(sce)$library_id,
      n_cells = ncol(sce)
    )
  return(result_df)
}
```

I think that is it! Now we can get the list of SCE objects and count the cells.

## Count cells and plot

First we will list all of the processed RDS files in the data directory.
We will use `list.files()` to do this all in one shot, using the option to search recursively.

```{r}
# get the list of processed SCE files
sce_files <- list.files(
  data_dir,
  pattern = "_processed.rds$",
  full.names = TRUE,
  recursive = TRUE
)
```

Now we will use the `purrr` package to apply our counting function to each of the files, and then combine all of the results into a single data frame.

```{r}
# count the cells in each file
cell_counts <- sce_files |>
  purrr::map(count_sce) |>
  purrr::list_rbind() # combine results into a single data frame
```

Let's look at the table of results.

```{r}
cell_counts
```

Finally, we will plot the results, faceted and colored by project ID.

```{r}
ggplot(cell_counts, aes(x = n_cells, fill = project_id)) +
  geom_histogram(bins = 20) +
  labs(x = "Number of cells") +
  facet_wrap(facets = vars(project_id)) +
  guides(fill = "none") + # don't show the legend
  theme_bw()

```

Excellent!

### 2024-03-15 analyses/hello-R/notebooks/hello.Rmd
Add outputs and wrapper script

@@
 
-# set results dir (using the analysis project file to find root)
+# set results and plots directories (using the analysis project file to find root)
 results_dir <- here::here("results")
+plots_dir <- here::here("plots")
+```
+
+We will be creating two output files: a table and a plot.
+Defining their locations here will make it easy for later users to know what to expect.
+
+```{r}
+count_file <- file.path(results_dir, "cell_counts.csv")
+plot_file <- file.path(plots_dir, "cell_counts.pdf")
 ```
@@
 
-Finally, we will plot the results, faceted and colored by project ID.
+And save it to our results directory.
+
+```{r}
+readr::write_csv(cell_counts, count_file)
+```
+
+Finally, we will plot the results as a histogram, faceted and colored by project ID.
 
@@
 Excellent!
+
+Now let's save the plot.
+```{r}
+ggsave(plot_file, width = 8, height = 6)
+```

### 2024-03-15 analyses/hello-R/hello-R.sh

@@ -1,0 +1,11 @@
+#!/bin/bash
+set -euo pipefail
+
+# Set the working directory to the directory of this file
+cd "$(dirname "${BASH_SOURCE[0]}")"
+
+# Render the R notebook
+
+notebook="notebooks/hello.Rmd"
+
+Rscript -e "rmarkdown::render('${notebook}')"

### 2024-03-15 analyses/hello-R/notebooks/hello.Rmd
Responses to review and linting

@@
 
-We will go through the processed SCE objects in the project `data` library and and count the number of cells in each object, plotting the results.
+We will go through the processed SCE objects in the project `data` library and count the number of cells in each object, plotting the results.
 
@@
 # set current data dir
-data_dir = file.path(repo_root, "data", "current")
+data_dir <- file.path(repo_root, "data", "current")
 
@@
 ```{r}
-ggplot(cell_counts, aes(x = n_cells, fill = project_id)) +
+count_plot <- ggplot(cell_counts, aes(x = n_cells, fill = project_id)) +
   geom_histogram(bins = 20) +
@@
   theme_bw()
-
+count_plot # show the plot
 ```
@@
 ```{r}
-ggsave(plot_file, width = 8, height = 6)
+ggsave(count_plot, plot_file, width = 8, height = 6)
 ```

### 2024-03-15 analyses/hello-R/notebooks/hello.Rmd
More responses to review

@@
 We will be creating two output files: a table and a plot.
-Defining their locations here will make it easy for later users to know what to expect.
+Defining their locations in the `Setup` section will make it easy for future notebook users to know what to expect.
 
@@
 
-Finally, we will plot the results as a histogram, faceted and colored by project ID.
+Finally, we will plot the results as a faceted histogram, colored by project ID.
 
@@
 Now let's save the plot.
+
 ```{r}
-ggsave(count_plot, plot_file, width = 8, height = 6)
+ggsave(plot_file, count_plot, width = 8, height = 6)
 ```

### 2024-03-16 analyses/hello-R/notebooks/hello.Rmd
start to respond to review

@@
 ---
-title: "Hello OpenScPCA"
+title: Hello OpenScPCA
+author:
+  - Joshua A. Shapiro, Childhood Cancer Data Lab, ALSF
+date: "`r Sys.Date()`"
 output: html_notebook
@@
 
-### Setup
+## Setup
 
@@
 
-```{r setup}
+```{r libraries}
 # load required libraries
@@
 })
+```
 
-# Find the repository root
+Set the paths to the data, results, and plots directories.
+
+```{r}
+# Find the repository and module root directories
 repo_root <- rprojroot::find_root(rprojroot::is_git_root)
+module_root <- rprojroot::find_root(rprojroot::is_renv_project)
 
@@
 # set results and plots directories (using the analysis project file to find root)
-results_dir <- here::here("results")
-plots_dir <- here::here("plots")
+results_dir <- file.path(module_root, "results")
+plots_dir <- file.path(module_root, "plots")
 ```
@@
 ```
+
+## Session Info
+
+```{r}
+sessioninfo::session_info()
+```

### 2024-03-16 analyses/hello-R/hello-R.sh

@@ -6,6 +6,4 @@ cd "$(dirname "${BASH_SOURCE[0]}")"
 
 # Render the R notebook
 
-notebook="notebooks/hello.Rmd"
-
-Rscript -e "rmarkdown::render('${notebook}')"
+Rscript -e "rmarkdown::render('notebooks/hello.Rmd')"

### 2024-03-18 analyses/hello-R/hello.Rmd
Move notebook out of subdirectory

--- 2024-03-16 analyses/hello-R/notebooks/hello.Rmd
@@
 Set the paths to the data, results, and plots directories.
+Here we use the `rprojroot` package to find:
+
+- the root of the `OpenScPCA-analysis` repository by looking for the (hidden) `.git` directory.
+- the root of this module by looking for the `renv.lock` file.
+
+Then we define the directories for input data and output results based on those root directories.
 
@@
 We will be creating two output files: a table and a plot.
-Defining their locations in the `Setup` section will make it easy for future notebook users to know what to expect.
+Defining their locations in the `Setup` section will make it easy for future users (including us in the future) to know what to expect as output from this notebook.
 
@@
 ```{r}
+count_sce <- function(sce_file) {
   # function to count cells in an sce file
-# returns a data frame with ids and cell count
-count_sce <- function(sce_file) {
+  # returns a data frame with the following columns:
+  # - project_id
+  # - sample_id
+  # - library_id
+  # - n_cells: number of cells in the library
+
   # load the SCE object
   sce <- readRDS(sce_file)
+
+  # input check
+  stopifnot("`sce_file` must contain a SingleCellExperiment object" = is(sce, "SingleCellExperiment"))
 

### 2024-03-18 analyses/hello-R/hello-R.sh

@@ -6,4 +6,4 @@ cd "$(dirname "${BASH_SOURCE[0]}")"
 
 # Render the R notebook
 
-Rscript -e "rmarkdown::render('notebooks/hello.Rmd')"
+Rscript -e "rmarkdown::render('hello.Rmd')"

### 2024-03-18 templates/analysis-module-R/notebook-template.Rmd
Add R notebook template

---
title: Analysis Title
author:
  - Your name here
date: "`r Sys.Date()`"
output: html_notebook
---

## Introduction

This a template for an analysis notebook using R Markdown.

It is a good idea to start with a brief introduction to the analysis, including the purpose of the analysis, the data used, and the methods applied.
Replace this text with your own introduction, and be sure to update the `title` and `author` fields at the top of the document.

## Setup

### Packages

Load required packages in the following chunk, if needed.
Do not install packages here, only load them with the `library()` function.

```{r packages}
# load required packages
# e.g. library("tidyverse")
```

You may also want to use the `package::function()` syntax within your code for better tracking of which functions come from which packages, and to avoid conflicts between functions with the same name in different packages.

### Paths

Setting paths to directories and files for input and output at the start of a notebook can be helpful for organization and reproducibility.
You may not know all of the paths you need when you start a notebook, but once you do, keeping them all in one place can be helpful, both for yourself in the future and for other users of the notebook.

### Base directories

In the chunk below, we included some common directories that you might need, finding their locations using the very useful `rprojroot` package.

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current")

# This module's base path, found by the presence of an `renv` directory
module_base <- rprojroot::find_root(rprojroot::is_renv_project)
# If you are not using `renv`, you might use something like the following instead:
# module_base <- file.path(repository_base, "analyses", "{module_name}")
```

### Input and output files

Set paths to input and output directories and files in the chunk below

```{r paths}
# Input files

# Output files
```

## Analysis content

Organize the remainder of your content into sections and subsections as appropriate for your analysis.

Add a new code chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I* on Windows or Linux, *Cmd+Option+I* on a Mac.
Be sure to comment your code and explain your steps as you go.

When you save the notebook, an HTML file containing the code and output will be saved alongside it.
To preview the HTML file, click the *Preview* button or press *Ctrl+Shift+K* on Windows or Linux, *Cmd+Shift+K* on a Mac.

The preview shows you a rendered HTML copy of the contents of the editor.
Consequently, *Preview* does not run any R code chunks.
Instead, the output of the chunk when it was last run in the editor is displayed.

It is usually a good idea to rerun the entire notebook before sharing any output, to ensure that the notebook and output files are fully up to date.
You can do this by clicking on the *Run* menu in the toolbar and clicking on *Restart R and Run All Chunks*.

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-03-18 templates/analysis-module-R/notebook-template.Rmd
Get fancy with handlebars

@@
 ---
-title: Analysis Title
+title: "{{module}} Analysis"
 author:
@@
 
-# This module's base path, found by the presence of an `renv` directory
-module_base <- rprojroot::find_root(rprojroot::is_renv_project)
-# If you are not using `renv`, you might use something like the following instead:
-# module_base <- file.path(repository_base, "analyses", "{module_name}")
+# The path to this module
+module_base <- file.path(repository_base, "analyses", "{{module}}")
+# If you are using `renv`, you might use something like the following instead:
+# module_base <- rprojroot::find_root(rprojroot::is_renv_project)
 ```

### 2024-03-18 create-analysis-module.py

@@ -185,13 +185,24 @@ def main() -> None:
         print(f"\nInitialized new renv environment in `{module_dir}`.")
 
     # Add template files
+
+    # find the {{module}} tag in template files
+    module_tag = re.compile(r"{{\s*module\s*}}")
+
     if args.use_r:
-        # add an R Markdown notebook template
-        shutil.copy(
-            base_dir / "templates" / "analysis-module-R" / "notebook-template.Rmd",
-            module_dir / "notebook-template.Rmd",
+        template_rmd = (
+            base_dir / "templates" / "analysis-module-R" / "notebook-template.Rmd"
         )
-        print(f"\nAdded an R Markdown notebook template in `{module_dir}`.")
+        with open(template_rmd, "r") as f:
+            template_content = f.readlines()
+        # replace the module tag with the module name
+        template_content = (
+            module_tag.sub(args.name, line) for line in template_content
+        )
+        with open(module_dir / "notebook-template.Rmd", "w") as f:
+            f.writelines(template_content)
+
+        print(f"Added an R Markdown notebook template in `{module_dir}`.")
 
 
 if __name__ == "__main__":

### 2024-03-18 templates/analysis-module-R/notebook-template.Rmd
Apply suggestions from code review
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 It is a good idea to start with a brief introduction to the analysis, including the purpose of the analysis, the data used, and the methods applied.
-Replace this text with your own introduction, and be sure to update the `title` and `author` fields at the top of the document.
+Replace this text with your own introduction, and be sure to update the `title:` and `author:` fields in the YAML header at the top of the document.
+Don't forget to rename this files as well!
 
@@
 Load required packages in the following chunk, if needed.
-Do not install packages here, only load them with the `library()` function.
+Do not install packages here; only load them with the `library()` function.
 
@@
 # load required packages
-# e.g. library("tidyverse")
+# e.g. library(tidyverse)
 ```
@@
 Setting paths to directories and files for input and output at the start of a notebook can be helpful for organization and reproducibility.
-You may not know all of the paths you need when you start a notebook, but once you do, keeping them all in one place can be helpful, both for yourself in the future and for other users of the notebook.
+You may not know all of the paths you need when you start a notebook, but once you do, keeping them all in one place is helpful, both for yourself in the future and for other users of the notebook.
 
@@
 
-In the chunk below, we included some common directories that you might need, finding their locations using the very useful `rprojroot` package.
+In the chunk below, we include some common directories that you might need, finding their locations using the very useful [`rprojroot` package](https://rprojroot.r-lib.org/).
 
@@
 
-Set paths to input and output directories and files in the chunk below
+Set paths to input and output directories and files in the chunk below.
 
@@
 # Output files
+
 ```

### 2024-03-18 templates/analysis-module-R/notebook-template.Rmd
remove rprojroot renv option

@@
 module_base <- file.path(repository_base, "analyses", "{{module}}")
-# If you are using `renv`, you might use something like the following instead:
-# module_base <- rprojroot::find_root(rprojroot::is_renv_project)
 ```

### 2024-03-21 templates/rmarkdown/notebook-template.Rmd

--- 2024-03-18 templates/analysis-module-R/notebook-template.Rmd
@@
 ---
-title: "{{module}} Analysis"
+title: "{{openscpca_module}} Analysis"
 author:
@@
 Replace this text with your own introduction, and be sure to update the `title:` and `author:` fields in the YAML header at the top of the document.
-Don't forget to rename this files as well!
+
+Don't forget to rename this file as well!
 
@@
 
-### Base directories
+#### Base directories
 
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "{{module}}")
+module_base <- file.path(repository_base, "analyses", "{{openscpca_module}}")
 ```
 
-### Input and output files
+#### Input and output files
 
@@
 # Output files
-
 ```

### 2024-04-02 analyses/hello-R/hello.Rmd

@@
 count_sce <- function(sce_file) {
-  # function to count cells in an sce file
-  # returns a data frame with the following columns:
+  # Count cells in an sce file
+  # Args:
+  # - sce_file: path to the SingleCellExperiment file
+  # Returns a data frame with the following columns:
   # - project_id
@@
 
-## Session Info
+## Session info
 

### 2024-04-05 analyses/simulate-sce/notebooks/splatter-ScPCA.Rmd

@@
 
-suffixes <- c( "_processed.rds", "_filtered.rds", "_unfiltered.rds")
+suffixes <- c("_processed.rds", "_filtered.rds", "_unfiltered.rds")
 
@@
 if (!is.null(metadata(sce_sim)$singler_results)) {
-  metadata(sce_sim)$singler_results <- metadata(sce_sim)$singler_results[cell_subset,]
+  metadata(sce_sim)$singler_results <- metadata(sce_sim)$singler_results[cell_subset, ]
 }
 if (!is.null(metadata(sce_sim)$cellassign_predictions)) {
-  metadata(sce_sim)$cellassign_predictions <- metadata(sce_sim)$cellassign_predictions[cell_subset,]
+  metadata(sce_sim)$cellassign_predictions <- metadata(sce_sim)$cellassign_predictions[cell_subset, ]
 }
@@
 ```{r}
-random_label<- function(label_set, n){
+random_label <- function(label_set, n) {
   # randomly select n labels from the label set, ensuring each label is included at least once
@@
 }
-
 ```
@@
 # create a mapping for singler ontology and annotation
-singler_df <- colData(sce)|>
-  data.frame()|>
+singler_df <- colData(sce) |>
+  data.frame() |>
   dplyr::select(singler_celltype_ontology, singler_celltype_annotation) |>
@@
 
-counts(sce_sim, withDimnames = FALSE) <- splatter::splatSimulate(sim_params)|>
+counts(sce_sim, withDimnames = FALSE) <- splatter::splatSimulate(sim_params) |>
   counts()
@@
 sce_sim <- sce_sim |>
-  scater::logNormCounts()|>
+  scater::logNormCounts() |>
   scater::runPCA(10) |> # we don't need all the PCA components
@@
 )
-colData(sce_sim)[,remove_stats] <- NULL
+colData(sce_sim)[, remove_stats] <- NULL
 
@@
 # restore column order
-colData(sce_sim) <- colData(sce_sim)[,colData_names]
-
+colData(sce_sim) <- colData(sce_sim)[, colData_names]
 ```

### 2024-04-23 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md
started docs on responding to review; img and doc in progress

# Responding to code reviews

After you file your PR, the Data Lab team will assign a reviewer to your pull request.
We aim to begin the process of code review within three business days of you filing the PR.
You can expect that each PR will have undergo several rounds of code review and response to review

Your reviewer will evaluate the code and/or documentation in your PR for the following considerations:

- **Context**
    - Has enough information been provided for the reviewer to fully understand the scope and context of what they are reviewing?
    If not, your reviewer's initial code review will likely request additional information to be able to perform review.
- **Clarity**
    - Is the code readable, reasonably efficient, and well-commented?
- **Reproducibility**
    - Can the code be re-run to successfully generate the same results?
- **Documentation**
    - Are the code and results (if applicable) clearly documented?
    - Are the steps to set up the code environment and run the code clearly documented?
- **Code checks**
    - We have set up several automated checks using GitHub actions for code quality control.
    If any of these automated checks fail, your reviewer may include feedback about what changes you may need to make for code checks to pass.

## Responding to review

Responding to code reviews establishes a conversation between you and your reviewer.
To ensure that this process goes smoothly and with fewer opportunities for miscommunication, we strongly recommend that read through all of your reviewer's comments before diving into review:

- You'll gain more context about why the reviewer left certain comments by reading the whole review first.
    - Several comments may also be related to each other, and taking in the full review will help you identify similar concepts or areas the reviewer noted.
- It's also possible that your reviewer misunderstood, or differently interpreted, some changes you made.
    - Reading the full review will help you understand if there are areas that need clarification before you can dive into responding.

### The pull request page

On the PR page, you will see several tabs:

- **Conversation**
    - This tab shows the main PR page.
    Here, you can find all comments left on the PR, both reviewer comments and any comments you leave in response.
    - You can also find any [resolved comments](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/commenting-on-a-pull-request#resolving-conversations) here, if you need to refer to them again.
- **Commits**
    - This tab shows a list of all commits added to the branch that is being reviewed.
    Your reviewer may use this page to explore changes in individual commits.
- **Checks**
    - This tab shows results from certain types of code checks; you can ignore this tab.
- **Files Changed**
    - This tab shows a line-by-line overview of content you have added and removed in your branch, relative to the base `AlexsLemonade/OpenScPCA-analysis` repository's `main` branch that your PR is targetting.
    - Referring to this tab is a great way to quickly see all your changes, as well as work with _reviewer suggestions_, [which are described below](#inline-suggestions).

### Types of reviewer comments

When reviewing your pull request, there are several types of comments your reviewer may leave for you to respond to.
<!-- Please refer to [our documentation on example reviews](STUB_LINK example reviews). -->

- Comments with overall feedback
    - In most circumstances, your reviewer will leave an overall high-level comment explaining their overall impressions from reviewing the code and synthesizing the types of changes they are requesting you make.
- File-level comments
    - Your reviewer can leave file-specific comments that provide high-level feedback about the overall contents of a given file.
- Inline comments
    - Your reviewer can also leave inline comments with feedback about specific lines of code.
    - These comments can also include _suggestions_ for code changes, that you can incorporate into your PR directly via on GitHub.

#### Inline suggestions

https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request

## Re-requesting review

Once you feel that you have addressed your reviewer's comments, you can [re-request review by clicking the cycle icon next to their handle on the right side of the PR page](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request#re-requesting-a-review).

<figure markdown="span">
    ![Re-request review.](../../img/responding-to-review-x.png){width="300"}
</figure>

### 2024-04-24 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md
First final draft of respond to review page with images that are still a bit TBD

@@
     - Has enough information been provided for the reviewer to fully understand the scope and context of what they are reviewing?
-    If not, your reviewer's initial code review will likely request additional information to be able to perform review.
+    If not, your reviewer's initial code review will likely request additional information they need to be able to perform review.
 - **Clarity**
     - Is the code readable, reasonably efficient, and well-commented?
-- **Reproducibility**
-    - Can the code be re-run to successfully generate the same results?
 - **Documentation**
@@
     - Are the steps to set up the code environment and run the code clearly documented?
+- **Reproducibility**
+    - Can the code be re-run to successfully generate the same results?
 - **Code checks**
@@
 
-## Responding to review
-
+!!! tip "Read the full review first"
     Responding to code reviews establishes a conversation between you and your reviewer.
-To ensure that this process goes smoothly and with fewer opportunities for miscommunication, we strongly recommend that read through all of your reviewer's comments before diving into review:
+    To ensure that this process goes smoothly and with fewer opportunities for miscommunication, we strongly recommend reading through all of your reviewer's comments before diving into review:
 
@@
 
-### The pull request page
+## The pull request page
 
-On the PR page, you will see several tabs:
+On the top of the PR page, you will see four tabs:
 
-- **Conversation**
+<figure markdown="span">
+    ![Pull request tabs.](../../img/respond-to-review-tabs.png){width="800"}
+</figure>
+
+<div class="grid" markdown>
+
+- **`Conversation`**
     - This tab shows the main PR page.
@@
     - You can also find any [resolved comments](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/commenting-on-a-pull-request#resolving-conversations) here, if you need to refer to them again.
-- **Commits**
+- **`Commits`**
     - This tab shows a list of all commits added to the branch that is being reviewed.
     Your reviewer may use this page to explore changes in individual commits.
-- **Checks**
+
+<!-- comment to force two grid columns -->
+
+- **`Checks`**
     - This tab shows results from certain types of code checks; you can ignore this tab.
-- **Files Changed**
-    - This tab shows a line-by-line overview of content you have added and removed in your branch, relative to the base `AlexsLemonade/OpenScPCA-analysis` repository's `main` branch that your PR is targetting.
-    - Referring to this tab is a great way to quickly see all your changes, as well as work with _reviewer suggestions_, [which are described below](#inline-suggestions).
+- **`Files changed`**
+    - This tab shows a line-by-line overview of content you have added and removed in your branch, relative to the base `AlexsLemonade/OpenScPCA-analysis` repository's `main` branch that your PR is targeting.
+    - Referring to this tab is a great way to quickly see all your changes, as well as work with [reviewer suggestions comments](#inline-comments).
 
-### Types of reviewer comments
+</div>
+
+## Types of reviewer comments
 
@@
 
-- Comments with overall feedback
-    - In most circumstances, your reviewer will leave an overall high-level comment explaining their overall impressions from reviewing the code and synthesizing the types of changes they are requesting you make.
-- File-level comments
-    - Your reviewer can leave file-specific comments that provide high-level feedback about the overall contents of a given file.
-- Inline comments
-    - Your reviewer can also leave inline comments with feedback about specific lines of code.
-    - These comments can also include _suggestions_ for code changes, that you can incorporate into your PR directly via on GitHub.
+!!! tip "Navigating review comments"
+    Unresolved file-level and inline review comments will appear both in the `Conversation` and in the `Files Changed` tabs.
 
-#### Inline suggestions
+    - In the `Conversation` tab, comments are listed in chronological order based on when the reviewer wrote the comment, which might not match the order of the code itself.
+    - In the `Files Changed` tab, comments appear right at the file or line of code they are about.
 
-https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request
+### Comments with overall feedback
+
+In most circumstances, your reviewer will leave an overall high-level comment explaining their overall impressions from reviewing the code and synthesizing the types of changes they are requesting you make.
+
+### File-level comments
+
+Your reviewer can leave file-specific comments that provide high-level feedback about the overall contents of a given file.
+
+For example, this file-level review comment requests that the code author comment and document a script:
+
+<figure markdown="span">
+    ![Re-request review.](../../img/respond-to-review-filecomment.png){width="600"}
+</figure>
+
+### Inline comments
+
+Your reviewer can also leave inline comments with feedback about specific lines of code.
+These comments can also include _suggestions_ for code changes, that you can incorporate into your PR directly via on GitHub.
+
+For example, below you can see three inline comments left on an R script, where the first and third comment are suggestions, as seen in the `Files changed` tab:
+
+<div class="grid" markdown>
+
+<!-- Note that I've used <br> to make the bullets line up with the associated comment in the image -->
+
+- In the first comment, the reviewer suggested that the code author add a new line to load the `readr` library.<br><br><br><br>
+- In the second comment, the reviewer did not formally suggest a code change, but left a comment recommending to use `file.path()`.
+(Learn more about [structuring your scripts](../analysis-modules/script-structure.md)!)<br><br>
+- In the third comment, the reviewer suggested that the code author remove a line that loads the `dplyr` library.
+
+<figure markdown="span">
+    ![Re-request review.](../../img/respond-to-review-inline.png){width="600"}
+</figure>
+
+</div>
+
+#### Working with suggestion comments
+
+Suggestion comments have two associated buttons: `Commit suggestion` and `Add suggestion to batch`.
+
+- You can directly commit a _single_ suggestion by clicking `Commit suggestion` button, and then clicking `Commit changes` in the dropdown box:
+<figure markdown="span">
+    ![Re-request review.](../../img/respond-to-review-commit-1-file.png){width="600"}
+</figure>
+
+- Alternatively, if there are multiple suggestions to accept, you can click the `Add suggestion to batch` button on each one.
+_The `Add suggestion to batch` option is only available via the `Files changed` tab!_
+    - Now, both suggestions are `Pending in batch` to be committed as a group:
+<figure markdown="span">
+    ![Re-request review.](../../img/respond-to-review-pending-batch.png){width="600"}
+</figure>
+
+You can commit the batched suggestions at once with the `Commit suggestions` button that will appear at the top of the `Files changed` tab:
+
+<figure markdown="span">
+    ![Re-request review.](../../img/respond-to-review-commitbatch.png){width="600"}
+</figure>
+
+!!! warning
+    After committing suggestions, be sure to _pull your changes locally_ before making more code changes in response to review!
+    This will help you avoid [merge conflicts](../creating-pull-requests/resolve-merge-conflicts.md).
+
+Read this GitHub documentation for more about [applying suggested changes while responding to review](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request).
 
@@
 <figure markdown="span">
-    ![Re-request review.](../../img/responding-to-review-x.png){width="300"}
+    ![Re-request review.](../../img/respond-to-review-x.png){width="400"}
 </figure>

### 2024-04-24 analyses/ewings-cell-type/notebook-template.Rmd

--- 2024-03-21 templates/rmarkdown/notebook-template.Rmd
@@
 ---
-title: "{{openscpca_module}} Analysis"
+title: "ewings-cell-type Analysis"
 author:
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "{{openscpca_module}}")
+module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
 ```

### 2024-04-24 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md
Apply suggestions from code review
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

@@
 We aim to begin the process of code review within three business days of you filing the PR.
-You can expect that each PR will have undergo several rounds of code review and response to review
+You can expect that each PR will undergo several rounds of code review and response to review
 
@@
 - **`Conversation`**
-    - This tab shows the main PR page.
-    Here, you can find all comments left on the PR, both reviewer comments and any comments you leave in response.
+    - This tab shows the main page of the PR.
+    Here, you can find the original description of the PR along with any comments left on the PR, both reviewer comments and any comments you leave in response.
     - You can also find any [resolved comments](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/commenting-on-a-pull-request#resolving-conversations) here, if you need to refer to them again.
@@
 - **`Checks`**
-    - This tab shows results from certain types of code checks; you can ignore this tab.
+    - This tab shows results from certain types of code checks that have been automatically set up by the Data Lab; you can ignore this tab.
 - **`Files changed`**
-    - This tab shows a line-by-line overview of content you have added and removed in your branch, relative to the base `AlexsLemonade/OpenScPCA-analysis` repository's `main` branch that your PR is targeting.
-    - Referring to this tab is a great way to quickly see all your changes, as well as work with [reviewer suggestions comments](#inline-comments).
+    - This tab shows a line-by-line overview of content you have added and removed in your branch.
+    Changes shown here should reflect the difference between files in your branch and the base branch your PR is targeting. 
+This should be the `main` branch of the `AlexsLemonade/OpenScPCA-analysis` repository. 
+    - Referring to this tab is a great way to quickly see all your changes, as well as respond to [reviewer suggestions or comments](#inline-comments).
 

### 2024-04-24 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md
add section on responding to review, which includes pushing

@@
 
-After you file your PR, the Data Lab team will assign a reviewer to your pull request.
-We aim to begin the process of code review within three business days of you filing the PR.
-You can expect that each PR will undergo several rounds of code review and response to review
-
-Your reviewer will evaluate the code and/or documentation in your PR for the following considerations:
-
-- **Context**
-    - Has enough information been provided for the reviewer to fully understand the scope and context of what they are reviewing?
-    If not, your reviewer's initial code review will likely request additional information they need to be able to perform review.
-- **Clarity**
-    - Is the code readable, reasonably efficient, and well-commented?
-- **Documentation**
-    - Are the code and results (if applicable) clearly documented?
-    - Are the steps to set up the code environment and run the code clearly documented?
-- **Reproducibility**
-    - Can the code be re-run to successfully generate the same results?
-- **Code checks**
-    - We have set up several automated checks using GitHub actions for code quality control.
-    If any of these automated checks fail, your reviewer may include feedback about what changes you may need to make for code checks to pass.
-
-!!! tip "Read the full review first"
-    Responding to code reviews establishes a conversation between you and your reviewer.
+Code review is essentially a conversation between you and your reviewer.
 To ensure that this process goes smoothly and with fewer opportunities for miscommunication, we strongly recommend reading through all of your reviewer's comments before diving into review:
@@
 
-## The pull request page
+## Navigating the pull request page
 
@@
 
-!!! tip "Navigating review comments"
+!!! tip "Reading review comments"
     Unresolved file-level and inline review comments will appear both in the `Conversation` and in the `Files Changed` tabs.
@@
 
-!!! warning
-    After committing suggestions, be sure to _pull your changes locally_ before making more code changes in response to review!
-    This will help you avoid [merge conflicts](../creating-pull-requests/resolve-merge-conflicts.md).
+Read this GitHub documentation for more about [applying suggested changes while responding to review](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request).
 
-Read this GitHub documentation for more about [applying suggested changes while responding to review](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request).
+## Responding to review comments
+
+When making code changes in response to review, you need to make sure your local repository is staying in-sync with any changes made directly on GitHub.
+Follow these steps when responding to review to avoid [merge conflicts](../creating-pull-requests/resolve-merge-conflicts.md):
+
+1. Before making any code changes, _pull_ your branch from GitHub.
+This will ensure you are up-to-date with any code changes made on GitHub, including committed review suggestions or updates from merging the upstream `main` branch into your pull request branch.
+    <figure markdown="span">
+        ![Pull in GitKraken.](../../img/respond-to-review-pull.png){width="600"}
+    </figure>
+
+1. Make changes, commit, and push as you normally would.
+Any changes you push to the branch with an open PR will automatically be reflected in the PR.
 

### 2024-04-24 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md
one review per commit
explain how to reply to a comment; may be moved to new doc

@@
 
-#### Working with suggestion comments
+## Replying to comments
+
+You can directly reply to review comments on GitHub.
+For example, you might ask your reviewer to clarify a particular comment, or ask for more guidance about how to implement the review.
+Or, you might just want to leave your reviewer a helpful note of some kind!
+
+The way to responding to an existing comment differs depending on which PR tab you are leaving the comment from:
+
+- If you are in the **`Files changed`** tab, use the button `Add single comment` to reply (_do not use `Start a review`_):
+    <figure markdown="span">
+        ![Reply to a review comment from the `Files changed` tab.](../../img/respond-to-review-comment-filed-changed.png){width="600"}
+    </figure>
+
+- If you are in the **`Conversation`** tab, use the button `Comment`.
+    <figure markdown="span">
+        ![Reply to a review comment from the `Conversation` tab.](../../img/respond-to-review-comment-conversation.png){width="600"}
+    </figure>
+
+When leaving a comment, you can either click `Add single comment` or `Start a review`.
+You should click `Add single comment`:
+
+## Working with suggestion comments
 
@@
 Any changes you push to the branch with an open PR will automatically be reflected in the PR.
+    - When committing changes in response to review, the smaller the better!
+    Review will go more smoothly if each commit addresses only one review comment.
 
@@
 </figure>
+
+You can also post a comment in your PR summarizing your response to review, which will help your reviewer navigate your code changes:
+
+- Which review comments did you address?
+    - Is there more information you can tell your reviewer that will help them understand your code changes?
+- Which review comments, if any, did you not address?
+    - Was it because the review comment itself was confusing, or because you weren't sure how to address it?
+    Communicating this with your reviewer will help you both out!

### 2024-04-24 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md
more content and reorg

@@
 
-## Types of reviewer comments
-
-When reviewing your pull request, there are several types of comments your reviewer may leave for you to respond to.
-<!-- Please refer to [our documentation on example reviews](STUB_LINK example reviews). -->
-
-!!! tip "Reading review comments"
-    Unresolved file-level and inline review comments will appear both in the `Conversation` and in the `Files Changed` tabs.
-
-    - In the `Conversation` tab, comments are listed in chronological order based on when the reviewer wrote the comment, which might not match the order of the code itself.
-    - In the `Files Changed` tab, comments appear right at the file or line of code they are about.
-
-### Comments with overall feedback
-
-In most circumstances, your reviewer will leave an overall high-level comment explaining their overall impressions from reviewing the code and synthesizing the types of changes they are requesting you make.
-
-### File-level comments
-
-Your reviewer can leave file-specific comments that provide high-level feedback about the overall contents of a given file.
-
-For example, this file-level review comment requests that the code author comment and document a script:
-
-<figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-filecomment.png){width="600"}
-</figure>
-
-### Inline comments
-
-Your reviewer can also leave inline comments with feedback about specific lines of code.
-These comments can also include _suggestions_ for code changes, that you can incorporate into your PR directly via on GitHub.
-
-For example, below you can see three inline comments left on an R script, where the first and third comment are suggestions, as seen in the `Files changed` tab:
-
-<div class="grid" markdown>
-
-<!-- Note that I've used <br> to make the bullets line up with the associated comment in the image -->
-
-- In the first comment, the reviewer suggested that the code author add a new line to load the `readr` library.<br><br><br><br>
-- In the second comment, the reviewer did not formally suggest a code change, but left a comment recommending to use `file.path()`.
-(Learn more about [structuring your scripts](../analysis-modules/script-structure.md)!)<br><br>
-- In the third comment, the reviewer suggested that the code author remove a line that loads the `dplyr` library.
-
-<figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-inline.png){width="600"}
-</figure>
-
-</div>
-
-## Replying to comments
-
-You can directly reply to review comments on GitHub.
-For example, you might ask your reviewer to clarify a particular comment, or ask for more guidance about how to implement the review.
-Or, you might just want to leave your reviewer a helpful note of some kind!
-
-The way to responding to an existing comment differs depending on which PR tab you are leaving the comment from:
-
-- If you are in the **`Files changed`** tab, use the button `Add single comment` to reply (_do not use `Start a review`_):
-    <figure markdown="span">
-        ![Reply to a review comment from the `Files changed` tab.](../../img/respond-to-review-comment-filed-changed.png){width="600"}
-    </figure>
-
-- If you are in the **`Conversation`** tab, use the button `Comment`.
-    <figure markdown="span">
-        ![Reply to a review comment from the `Conversation` tab.](../../img/respond-to-review-comment-conversation.png){width="600"}
-    </figure>
-
-When leaving a comment, you can either click `Add single comment` or `Start a review`.
-You should click `Add single comment`:
-
-## Working with suggestion comments
-
-Suggestion comments have two associated buttons: `Commit suggestion` and `Add suggestion to batch`.
-
-- You can directly commit a _single_ suggestion by clicking `Commit suggestion` button, and then clicking `Commit changes` in the dropdown box:
-<figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-commit-1-file.png){width="600"}
-</figure>
-
-- Alternatively, if there are multiple suggestions to accept, you can click the `Add suggestion to batch` button on each one.
-_The `Add suggestion to batch` option is only available via the `Files changed` tab!_
-    - Now, both suggestions are `Pending in batch` to be committed as a group:
-<figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-pending-batch.png){width="600"}
-</figure>
-
-You can commit the batched suggestions at once with the `Commit suggestions` button that will appear at the top of the `Files changed` tab:
-
-<figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-commitbatch.png){width="600"}
-</figure>
-
-Read this GitHub documentation for more about [applying suggested changes while responding to review](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request).
-
 ## Responding to review comments
@@
     Communicating this with your reviewer will help you both out!
+
+## Types of reviewer comments
+
+When reviewing your pull request, there are several types of comments your reviewer may leave for you to respond to.
+<!-- Please refer to [our documentation on example reviews](STUB_LINK example reviews). -->
+
+!!! tip "Reading review comments"
+    Unresolved file-level and inline review comments will appear both in the `Conversation` and in the `Files Changed` tabs.
+
+    - In the `Conversation` tab, comments are listed in chronological order based on when the reviewer wrote the comment, which might not match the order of the code itself.
+    - In the `Files Changed` tab, comments appear right at the file or line of code they are about.
+
+### Comments with overall feedback
+
+In most circumstances, your reviewer will leave an overall high-level comment explaining their overall impressions from reviewing the code and synthesizing the types of changes they are requesting you make.
+
+### File-level comments
+
+Your reviewer can leave file-specific comments that provide high-level feedback about the overall contents of a given file.
+
+For example, this file-level review comment requests that the code author comment and document a script:
+
+<figure markdown="span">
+    ![Re-request review.](../../img/respond-to-review-filecomment.png){width="600"}
+</figure>
+
+### Inline comments
+
+Your reviewer can also leave inline comments with feedback about specific lines of code:
+
+<figure markdown="span">
+    ![Re-request review.](../../img/respond-to-review-inline.png){width="600"}
+</figure>
+
+You can directly reply to these review comments on GitHub.
+For example, you might ask your reviewer to clarify a particular comment, or ask for more guidance about how to implement the review.
+Or, you might just want to leave your reviewer a helpful note of some kind!
+
+The way to responding to an existing comment differs depending on which PR tab you are leaving the comment from:
+
+- If you are in the **`Files changed`** tab, use the button `Add single comment` to reply (_do not use `Start a review`_):
+    <figure markdown="span">
+        ![Reply to a review comment from the `Files changed` tab.](../../img/respond-to-review-comment-filed-changed.png){width="600"}
+    </figure>
+
+- If you are in the **`Conversation`** tab, use the button `Comment`.
+    <figure markdown="span">
+        ![Reply to a review comment from the `Conversation` tab.](../../img/respond-to-review-comment-conversation.png){width="600"}
+    </figure>
+
+(Learn more about [structuring your scripts](../analysis-modules/script-structure.md)!)
+
+### Suggestion comments
+
+These comments can also include _suggestions_ for code changes, that you can incorporate into your PR directly via on GitHub.
+
+For example, below you are two suggestions review comments on an R script:
+
+- In the first comment, the reviewer suggested that the code author add a new line to load the `readr` library.
+- In the second comment, the reviewer suggested that the code author remove a line that loads the `dplyr` library.
+
+<figure markdown="span">
+    ![Re-request review.](../../img/respond-to-review-suggestions.png){width="600"}
+</figure>
+
+Suggestion comments have two associated buttons boxed in the image above: `Commit suggestion` and `Add suggestion to batch`.
+
+- You can directly commit a _single_ suggestion by clicking `Commit suggestion` button, and then clicking `Commit changes` in the dropdown box:
+<figure markdown="span">
+    ![Re-request review.](../../img/respond-to-review-commit-1-file.png){width="600"}
+</figure>
+
+- Alternatively, if there are multiple suggestions to accept, you can click the `Add suggestion to batch` button on each one.
+_The `Add suggestion to batch` option is only available via the `Files changed` tab!_
+    - Now, both suggestions are `Pending in batch` to be committed as a group:
+<figure markdown="span">
+    ![Re-request review.](../../img/respond-to-review-pending-batch.png){width="600"}
+</figure>
+
+You can commit the batched suggestions at once with the `Commit suggestions` button that will appear at the top of the `Files changed` tab:
+
+<figure markdown="span">
+    ![Re-request review.](../../img/respond-to-review-commitbatch.png){width="600"}
+</figure>
+
+Read this GitHub documentation for more about [applying suggested changes while responding to review](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request).

### 2024-04-25 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md
Apply suggestions from code review
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

@@
 
-Code review is essentially a conversation between you and your reviewer.
+Code review is essentially a conversation between you and your reviewer about your proposed changes. 
+This document includes some example review comments and guidance on how to respond to review comments and suggestions. 
 To ensure that this process goes smoothly and with fewer opportunities for miscommunication, we strongly recommend reading through all of your reviewer's comments before diving into review:
@@
 
-(Learn more about [structuring your scripts](../analysis-modules/script-structure.md)!)
-
 ### Suggestion comments
 
-These comments can also include _suggestions_ for code changes, that you can incorporate into your PR directly via on GitHub.
+Reviewer comments can also include _suggestions_ for code changes, that you can incorporate into your PR directly on GitHub.
 
-For example, below you are two suggestions review comments on an R script:
+For example, below are two suggestions left as comments on an R script:
 

### 2024-04-25 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md
reorganization and rewording in response to review, and update links in index

@@
 This document includes some example review comments and guidance on how to respond to review comments and suggestions.
-To ensure that this process goes smoothly and with fewer opportunities for miscommunication, we strongly recommend reading through all of your reviewer's comments before diving into review:
+
+To ensure that code review goes smoothly and with fewer opportunities for miscommunication, we strongly recommend reading through all of your reviewer's comments before diving into review:
 
@@
 This should be the `main` branch of the `AlexsLemonade/OpenScPCA-analysis` repository.
-    - Referring to this tab is a great way to quickly see all your changes, as well as respond to [reviewer suggestions or comments](#inline-comments).
+    - Referring to this tab is a great way to quickly see all your changes, as well as respond to [reviewer suggestions or comments](#in-line-comments).
 
 </div>
-
-## Responding to review comments
-
-When making code changes in response to review, you need to make sure your local repository is staying in-sync with any changes made directly on GitHub.
-Follow these steps when responding to review to avoid [merge conflicts](../creating-pull-requests/resolve-merge-conflicts.md):
-
-1. Before making any code changes, _pull_ your branch from GitHub.
-This will ensure you are up-to-date with any code changes made on GitHub, including committed review suggestions or updates from merging the upstream `main` branch into your pull request branch.
-    <figure markdown="span">
-        ![Pull in GitKraken.](../../img/respond-to-review-pull.png){width="600"}
-    </figure>
-
-1. Make changes, commit, and push as you normally would.
-Any changes you push to the branch with an open PR will automatically be reflected in the PR.
-    - When committing changes in response to review, the smaller the better!
-    Review will go more smoothly if each commit addresses only one review comment.
-
-## Re-requesting review
-
-Once you feel that you have addressed your reviewer's comments, you can [re-request review by clicking the cycle icon next to their handle on the right side of the PR page](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request#re-requesting-a-review).
-
-<figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-x.png){width="400"}
-</figure>
-
-You can also post a comment in your PR summarizing your response to review, which will help your reviewer navigate your code changes:
-
-- Which review comments did you address?
-    - Is there more information you can tell your reviewer that will help them understand your code changes?
-- Which review comments, if any, did you not address?
-    - Was it because the review comment itself was confusing, or because you weren't sure how to address it?
-    Communicating this with your reviewer will help you both out!
 
@@
 When reviewing your pull request, there are several types of comments your reviewer may leave for you to respond to.
-<!-- Please refer to [our documentation on example reviews](STUB_LINK example reviews). -->
 
 !!! tip "Reading review comments"
-    Unresolved file-level and inline review comments will appear both in the `Conversation` and in the `Files Changed` tabs.
+    Unresolved file-level and in-line review comments will appear both in the `Conversation` and in the `Files Changed` tabs.
 
@@
 
-### Inline comments
+### In-line comments
 
-Your reviewer can also leave inline comments with feedback about specific lines of code:
+Your reviewer can also leave in-line comments with feedback about specific lines of code:
 
@@
 
-Read this GitHub documentation for more about [applying suggested changes while responding to review](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request).
+Read this GitHub documentation for more about [applying suggested changes while responding to review](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request#applying-suggested-changes).
+
+## Responding to review comments
+
+In some cases a reviewer will request you make changes without providing in-line suggestions.
+When this happens, you will need to make the code changes locally.
+If you committed suggestions directly on GitHub before working on additional changes, be sure to sync your local branch with changes made on GitHub before changing any local code.
+
+When making code changes in response to review, you need to make sure your local repository is staying in-sync with any changes made directly on GitHub.
+Follow these steps when responding to review to avoid [merge conflicts](../creating-pull-requests/resolve-merge-conflicts.md):
+
+1. Before making any code changes, _pull_ your branch from GitHub.
+This will ensure you are up-to-date with any code changes made on GitHub, including committed review suggestions or updates from merging the upstream `main` branch into your pull request branch.
+    <figure markdown="span">
+        ![Pull in GitKraken.](../../img/respond-to-review-pull.png){width="600"}
+    </figure>
+
+1. Make changes, commit, and push as you normally would.
+Any changes you push to the branch with an open PR will automatically be reflected in the PR.
+    - When committing changes in response to review, the smaller the better!
+    Review will go more smoothly if each commit addresses only one review comment.
+
+## Re-requesting review
+
+Once you feel that you have addressed your reviewer's comments, you can [re-request review by clicking the cycle icon next to their handle on the right side of the PR page](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request#re-requesting-a-review).
+
+<figure markdown="span">
+    ![Re-request review.](../../img/respond-to-review-x.png){width="400"}
+</figure>
+
+You can also post a comment in your PR summarizing your response to review, which will help your reviewer navigate your code changes:
+
+- Which review comments did you address?
+    - Is there more information you can tell your reviewer that will help them understand your code changes?
+- Which review comments, if any, did you not address?
+    - Was it because the review comment itself was confusing, or because you weren't sure how to address it?
+    Communicating this with your reviewer will help you both out!

### 2024-04-25 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md
final image names

@@
 <figure markdown="span">
-    ![Pull request tabs.](../../img/respond-to-review-tabs.png){width="800"}
+    ![Pull request tabs.](../../img/respond-to-review-01.png){width="800"}
 </figure>
@@
 <figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-filecomment.png){width="600"}
+    ![Re-request review.](../../img/respond-to-review-02.png){width="600"}
 </figure>
@@
 <figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-inline.png){width="600"}
+    ![Re-request review.](../../img/respond-to-review-03.png){width="600"}
 </figure>
@@
     <figure markdown="span">
-        ![Reply to a review comment from the `Files changed` tab.](../../img/respond-to-review-comment-filed-changed.png){width="600"}
+        ![Reply to a review comment from the `Files changed` tab.](../../img/respond-to-review-04.png){width="600"}
     </figure>
@@
     <figure markdown="span">
-        ![Reply to a review comment from the `Conversation` tab.](../../img/respond-to-review-comment-conversation.png){width="600"}
+        ![Reply to a review comment from the `Conversation` tab.](../../img/respond-to-review-05.png){width="600"}
     </figure>
@@
 <figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-suggestions.png){width="600"}
+    ![Re-request review.](../../img/respond-to-review-06.png){width="600"}
 </figure>
@@
 <figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-commit-1-file.png){width="600"}
+    ![Re-request review.](../../img/respond-to-review-07.png){width="600"}
 </figure>
@@
 <figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-pending-batch.png){width="600"}
+    ![Re-request review.](../../img/respond-to-review-08.png){width="600"}
 </figure>
@@
 <figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-commitbatch.png){width="600"}
+    ![Re-request review.](../../img/respond-to-review-09.png){width="600"}
 </figure>
@@
     <figure markdown="span">
-        ![Pull in GitKraken.](../../img/respond-to-review-pull.png){width="600"}
+        ![Pull in GitKraken.](../../img/respond-to-review-10.png){width="600"}
     </figure>
@@
 <figure markdown="span">
-    ![Re-request review.](../../img/respond-to-review-x.png){width="400"}
+    ![Re-request review.](../../img/respond-to-review-11.png){width="400"}
 </figure>

### 2024-04-25 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md
Apply suggestions from code review
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

@@
 
-The way to responding to an existing comment differs depending on which PR tab you are leaving the comment from:
+The way to respond to an existing comment differs depending on which PR tab you are leaving the comment from:
 
@@
 
+## Responding to review comments
+
+### Responding to suggestion comments 
 Suggestion comments have two associated buttons boxed in the image above: `Commit suggestion` and `Add suggestion to batch`.
@@
 
-## Responding to review comments
+## Making additional code changes in response to comments
 
-In some cases a reviewer will request you make changes without providing in-line suggestions.
+In some cases a reviewer will request you make changes to the code without providing in-line suggestions.
 When this happens, you will need to make the code changes locally.

### 2024-04-25 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md
update doc: header levels and indicate that you can use a custom commit message

@@
 
-- You can directly commit a _single_ suggestion by clicking `Commit suggestion` button, and then clicking `Commit changes` in the dropdown box:
+- You can directly commit a _single_ suggestion by clicking `Commit suggestion` button.
+Type a commit message, and then click `Commit changes` in the dropdown box:
 <figure markdown="span">
@@
 
-## Making additional code changes in response to comments
+### Making additional code changes in response to comments
 

### 2024-04-25 analyses/ewings-cell-type/01-tumor-cell-classification.Rmd
initial exploration notebook for marker genes

---
title: "Explor tumor cell classification - SCPCS000490"
author: Ally Hawkins
date: "`r Sys.Date()`"
output: html_notebook
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

## Introduction

This notebook looks at expression of marker genes for tumor cells in a Ewing sarcoma sample, `r {params$sample_id}`. 
Additionally, we see if we can use those marker genes to classify tumor and normal cells using `CellAssign`.

- First we look at expression of each of the marker genes across all cells. 
- Then we plot the mean expression of all marker genes across all cells. 
- Next, we perform clustering and see if we can assign each cluster as tumor or normal cells based on mean expression of marker genes. 
- Finally, we use `CellAssign` with both the full list of marker genes and filtered list of marker genes. 
- The results from `CellAssign` are then compared to manual annotation of cells.

## Setup

```{r packages}
suppressMessages(
  # load required packages
  library(SingleCellExperiment),
  library(ggplot2)
)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current")
sample_dir <- file.path(data_dir, "SCPCP000015", params$sample_id)

# The path to this module
module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
```

```{r paths}
# Input files
sce_filename <- glue::glue("{params$library_id}_processed.rds")
sce_file <- file.path(sample_dir, sce_filename)

anndata_filename <- glue::glue("{params$library_id}_processed_rna.hdf5")
anndata_file <- file.path(sample_dir, anndata_filename)

marker_genes <- file.path(module_base, "references", "tumor_marker_genes.tsv")

# Output files
tumor_markers_mtx_file <- file.path(module_base, "references", "tumor-marker-cellassign.tsv")
filtered_markers_mtx_file <- file.path(module_base, "references", "filtered-tumor-marker-cellassign.tsv")
cellassign_predictions_file <- file.path(module_base, "results", "cellassign-tumor-predictions.tsv")
filtered_predictions_file <- file.path(module_base, "results", "filtered-cellassign-tumor-predictions.tsv")
```

```{r}
# read in processed sce
sce <- readr::read_rds(sce_file)

# read in marker genes table 
marker_genes_df <- readr::read_tsv(marker_genes) |> 
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
  dplyr::distinct()

marker_genes <- marker_genes_df |> 
  dplyr::filter(cell_type == "tumor") |> 
  dplyr::pull(ensembl_gene_id)
```

## Analysis content

### Explore marker gene expression

The first thing we do here is just create a faceted UMAP showing the expression of each marker gene for tumor cells.

```{r}
# get the gene expression counts for all marker genes
marker_gene_exp <- logcounts(sce[marker_genes, ]) |>
    as.matrix() |>
    t() |>
    as.data.frame() |>
    tibble::rownames_to_column("barcodes")

# pull out the UMAP coordinates and make a data frame to use for plotting
umap_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
        starts_with("UMAP")
      )

marker_gene_plot_df <- umap_df |> 
  # add in marker gene expression to dataframe 
  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
  # combine all genes into a single column for easy faceting
  tidyr::pivot_longer(
    cols = starts_with("ENSG"),
    names_to = "ensembl_gene_id",
    values_to = "gene_expression"
  ) |> 
  # join with marker gene df to get gene symbols for plotting 
  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression, clusters)
```

```{r}
# faceted umap showing a umap panel for each marker gene 
ggplot(marker_gene_plot_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
    geom_point(alpha = 0.1, size = 0.4) +
    facet_wrap(vars(gene_symbol)) +
    scale_color_viridis_c() +
    labs(
      color = "Log-normalized gene expression"
    ) +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    theme(
      aspect.ratio = 1,
      legend.position = "bottom",
      axis.title = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
  theme_bw()
```

Looking at this plot, there is definitely some variation in the expression of marker genes. 
It looks like the highest gene expression is seen in the larger group of cells on the bottom right of the UMAP.

Below, we can get the mean gene expression of all marker genes and plot in a single UMAP.

```{r}
# calculate mean gene expression across all marker genes in list 
marker_mean_exp <- umap_df |> 
  dplyr::group_by(barcodes) |> 
  dplyr::mutate(mean_exp = mean(gene_expression)) |> 
  dplyr::select(barcodes, UMAP1, UMAP2, mean_exp) |> 
  dplyr::distinct()

# plot mean gene expression 
ggplot(marker_mean_exp, aes(x = UMAP1, y = UMAP2, color = mean_exp)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_viridis_c() +
  theme_bw()
```

Similar to the individual plots, it looks like there is one group of cells on the bottom right that has the highest mean gene expression. 
We would anticipate that these are most likely to be the tumor cells.

### Clustering

```{r}
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = clusters)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

```{r}
# classify cells as tumor/normal 
classification_df <- marker_mean_exp |> 
  dplyr::mutate(tumor_normal_classification = dplyr::if_else(mean_exp > 0.5, "Tumor", "Normal"))

ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = tumor_normal_classification)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

### Run CellAssign with marker genes

Create marker genes reference

```{r}
binary_mtx <- marker_genes_df |>
  dplyr::select(cell_type, ensembl_gene_id) |> 
  unique() |> 
  tidyr::pivot_wider(
    id_cols = ensembl_gene_id,
    names_from = cell_type,
    values_from = cell_type,
    values_fn = length,
    values_fill = 0
  ) |>
  # add a column with no marker genes
  # cell assign will assign cells to "other" when no other cell types are appropriate
  dplyr::mutate(other = 0)

# export matrix 
readr::write_tsv(binary_mtx, tumor_markers_mtx_file)
```

Add script for running CellAssign 
```{r, eval=FALSE}
cellassign_call <- glue::glue("
python scripts/run-cellassign.py \
  --anndata_file '{anndata_file}' \
  --output_predictions '{cellassign_predictions_file}' \
  --reference '{tumor_markers_mtx}'
")

system(cellassign_call)
```

Read in CellAssign marker genes
```{r}
predictions <- readr::read_tsv(cellassign_predictions_file)

celltype_assignments <- predictions |>
      tidyr::pivot_longer(
        !barcode,
        names_to = "celltype",
        values_to = "prediction"
      ) |>
      dplyr::group_by(barcode) |>
      dplyr::slice_max(prediction, n = 1) |>
      dplyr::ungroup()

table(celltype_assignments$celltype)
```

What if we restrict our list of marker genes based on our dataset? So we want something that has a mean gene expression of 1.

```{r}
filtered_markers <- rowData(sce) |> 
  as.data.frame() |> 
  dplyr::filter(gene_symbol %in% marker_genes_df$gene_symbol,
                mean > 1) |> 
  dplyr::pull(gene_ids)

filtered_binary_mtx <- binary_mtx |> 
  dplyr::filter(ensembl_gene_id %in% filtered_markers)

readr::write_tsv(filtered_binary_mtx, filtered_markers_mtx_file)
```

```{r, eval=FALSE}
cellassign_call <- glue::glue("
python scripts/run-cellassign.py --anndata_file '{anndata_file}' --output_predictions '{filtered_predictions_file}' --reference '{filtered_markers_mtx_file}'
")

system(cellassign_call)
```

```{r}
predictions <- readr::read_tsv(filtered_predictions_file)

celltype_assignments <- predictions |>
      tidyr::pivot_longer(
        !barcode,
        names_to = "celltype",
        values_to = "prediction"
      ) |>
      dplyr::group_by(barcode) |>
      dplyr::slice_max(prediction, n = 1) |>
      dplyr::ungroup()

table(celltype_assignments$celltype)
```

```{r}
celltype_df <- marker_mean_exp |> 
  dplyr::left_join(celltype_assignments, by = c("barcodes" = "barcode"))

ggplot(celltype_df, aes(x = UMAP1, UMAP2, color = celltype)) +
  geom_point(size = 0.5, alpha = 0.5) + 
  theme_bw()
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-04-25 analyses/ewings-cell-type/01-marker-gene-classification-cellassign.Rmd
update notebook name

---
title: "Explore tumor cell classification with marker genes and CellAssign - SCPCS000490"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

## Introduction

This notebook looks at expression of marker genes for tumor cells in a Ewing sarcoma sample, `r {params$sample_id}`. 
Additionally, we see if we can use those marker genes to classify tumor and normal cells using both manual annotation and `CellAssign`.

The first section focuses on assigning tumor and normal cells by looking at marker gene expression alone:

- First we look at expression of each of the marker genes across all cells. 
- Then we plot the mean expression of all marker genes across all cells. 
- Next, we perform clustering and see if we can assign each cluster as tumor or normal cells based on mean expression of marker genes.

The next section uses `CellAssign` to assign tumor cells with the same set of marker genes:

- We use `CellAssign` with both the full list of marker genes and filtered list of marker genes. 
- We then compare the average gene expression of all marker genes between tumor and other cells.

## Setup

```{r packages}
suppressMessages(
  # load required packages
  library(SingleCellExperiment),
  library(ggplot2)
)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current")
sample_dir <- file.path(data_dir, "SCPCP000015", params$sample_id)

# The path to this module
module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
```

```{r paths}
# Input files
sce_filename <- glue::glue("{params$library_id}_processed.rds")
sce_file <- file.path(sample_dir, sce_filename)

anndata_filename <- glue::glue("{params$library_id}_processed_rna.hdf5")
anndata_file <- file.path(sample_dir, anndata_filename)

marker_genes <- file.path(module_base, "references", "tumor_marker_genes.tsv")

# Output files
tumor_markers_mtx_file <- file.path(module_base, "references", "cellassign_refs", "tumor-marker-cellassign.tsv")
filtered_markers_mtx_file <- file.path(module_base, "references", "cellassign_refs", "filtered-tumor-marker-cellassign.tsv")
cellassign_predictions_file <- file.path(module_base, "results", "cellassign-tumor-predictions.tsv")
filtered_predictions_file <- file.path(module_base, "results", "filtered-cellassign-tumor-predictions.tsv")
```

```{r}
# read in processed sce
sce <- readr::read_rds(sce_file)

# read in marker genes table 
marker_genes_df <- readr::read_tsv(marker_genes) |> 
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
  dplyr::distinct()

marker_genes <- marker_genes_df |> 
  dplyr::filter(cell_type == "tumor") |> 
  dplyr::pull(ensembl_gene_id)
```

## Analysis content

### Explore marker gene expression

The first thing we do here is just create a faceted UMAP showing the expression of each marker gene for tumor cells.

```{r}
# get the gene expression counts for all marker genes
marker_gene_exp <- logcounts(sce[marker_genes, ]) |>
    as.matrix() |>
    t() |>
    as.data.frame() |>
    tibble::rownames_to_column("barcodes")

# pull out the UMAP coordinates and make a data frame to use for plotting
umap_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
        starts_with("UMAP")
      ) |> 
  # add in marker gene expression to dataframe 
  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
  # combine all genes into a single column for easy faceting
  tidyr::pivot_longer(
    cols = starts_with("ENSG"),
    names_to = "ensembl_gene_id",
    values_to = "gene_expression"
  ) |> 
  # join with marker gene df to get gene symbols for plotting 
  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression, clusters)
```

```{r}
# faceted umap showing a umap panel for each marker gene 
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
    geom_point(alpha = 0.1, size = 0.2) +
    facet_wrap(vars(gene_symbol)) +
    scale_color_viridis_c() +
    labs(
      color = "Log-normalized gene expression"
    ) +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    theme(
      aspect.ratio = 1,
      legend.position = "bottom",
      axis.title = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
  theme_bw()
```

Looking at this plot, there is definitely some variation in the expression of marker genes. 
It looks like the highest gene expression is seen in the larger group of cells on the bottom right of the UMAP.

Below, we can get the mean gene expression of all marker genes and plot in a single UMAP.

```{r}
# calculate mean gene expression across all marker genes in list 
marker_mean_exp <- umap_df |> 
  dplyr::group_by(barcodes) |> 
  dplyr::mutate(mean_exp = mean(gene_expression)) |> 
  dplyr::select(barcodes, UMAP1, UMAP2, mean_exp, clusters) |> 
  dplyr::distinct()

# plot mean gene expression 
ggplot(marker_mean_exp, aes(x = UMAP1, y = UMAP2, color = mean_exp)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_viridis_c() +
  theme_bw()
```

Similar to the individual plots, it looks like there is one group of cells on the bottom right that has the highest mean gene expression. 
We would anticipate that these are most likely to be the tumor cells.

### Using clusters and marker gene expression to annotate tumor cells

We anticipate that the tumor cells and normal cells would cluster separately. 
First, let's just plot the cluster assignments alone.

```{r}
ggplot(marker_mean_exp, aes(x = UMAP1, y = UMAP2, color = clusters)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

It looks like our large group of cells on the bottom right is comprised of a few clusters while most other groups of cells are a single cluster. 
Let's see if there's any difference in mean expression of the marker genes across the clusters. 
Based on the gene expression I would expect most cells in the bottom right to be considered tumor and all other cells to be normal.

```{r}
# plot marker gene expression across all clusters 
ggplot(marker_mean_exp, aes(x = mean_exp, fill = clusters)) +
  geom_density() +
  facet_grid(rows = vars(clusters)) +
  theme_bw()
```

Looking at this there are definitely distinct groups of cells with higher mean expression of the marker genes than other cells. 
Before doing anything fancy, let's use this to manually assign tumor and normal cells. 
One approach is to assign based on the cluster, so if the cluster has a median expression > 0.75 then all cells in that cluster would get assigned as tumor cells. 
We could also pick a mean expression cutoff and use that to assign each cell individually, but there are some tails that overlap.

```{r}
# classify clusters as tumor/normal 
cluster_classification_df <- marker_mean_exp |> 
  dplyr::group_by(clusters) |> 
  # calculate the median mean gene expression across all cells in a cluster
  dplyr::summarise(median_cluster_exp = median(mean_exp)) |> 
  # use 0.75 as the cutoff for tumor/normal 
  dplyr::mutate(tumor_normal_classification = dplyr::if_else(median_cluster_exp > 0.75, "Tumor", "Normal"))

# get individual cell classifications
celltype_classification_df <- marker_mean_exp |> 
  dplyr::left_join(cluster_classification_df)

# umap showing tumor and normal cells 
ggplot(celltype_classification_df, aes(x = UMAP1, y = UMAP2, color = tumor_normal_classification)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

### Run CellAssign with marker genes

Now let's see if we can use an automated marker gene-based method to identify tumor cells. 
To do this we will use `CellAssign` with a reference matrix created from the marker gene list.

First, we need to create a binary matrix with two columns, "tumor" and "other". 
The "other" will refer to any normal cells.

```{r}
binary_mtx <- marker_genes_df |>
  dplyr::select(cell_type, ensembl_gene_id) |> 
  unique() |> 
  tidyr::pivot_wider(
    id_cols = ensembl_gene_id,
    names_from = cell_type,
    values_from = cell_type,
    values_fn = length,
    values_fill = 0
  ) |>
  # add a column with no marker genes
  # cell assign will assign cells to "other" when no other cell types are appropriate
  dplyr::mutate(other = 0)

# export matrix 
readr::write_tsv(binary_mtx, tumor_markers_mtx_file)
```

Now run `CellAssign` with the full marker gene list as the reference.

```{r, eval=FALSE}
cellassign_call <- glue::glue("
python scripts/run-cellassign.py \
  --anndata_file '{anndata_file}' \
  --output_predictions '{cellassign_predictions_file}' \
  --reference '{tumor_markers_mtx_file}'
")

system(cellassign_call)
```

We can then read in the predictions file and find the predicted cell type for each cell, either "tumor" or "other".

```{r}
# read in predictions
predictions <- readr::read_tsv(cellassign_predictions_file)

# get individual cell type assignments
# those with the max prediction 
celltype_assignments <- predictions |>
      tidyr::pivot_longer(
        !barcode,
        names_to = "celltype",
        values_to = "prediction"
      ) |>
      dplyr::group_by(barcode) |>
      dplyr::slice_max(prediction, n = 1) |>
      dplyr::ungroup()
# create a table of assignments 
table(celltype_assignments$celltype)
```

Looking at this, it appears that `CellAssign` is classifying all cells as "other"? 
I find this hard to believe since this sample was taken directly from a Ewing sarcoma tumor, so there are bound to be tumor cells.

In looking at the original UMAP there are quite a few genes that have very low expression across all cells, so perhaps they are adding noise to the model. 
Although this is a bit circular, we could restrict our list of marker genes to be only those that have a high mean gene expression in our dataset? 
I would expect that when we do that, we are able to identify more tumor cells that line up with what we would expect based on gene expression patterns.

```{r}
# grab the mean gene expression for all marker genes 
rowdata_df <- rowData(sce) |> 
  as.data.frame() |> 
  dplyr::filter(gene_symbol %in% marker_genes_df$gene_symbol)
rowdata_df
```

```{r}
# what's the mean gene expression of all genes in this sample for reference
mean(rowData(sce)$mean)
```

There are a few genes that have low gene expression < 1. 
Let's try and filter out those genes and then use `CellAssign`.

```{r}
# get list of marker genes to keep
filtered_markers <- rowdata_df |> 
  dplyr::filter(mean > 1) |> 
  dplyr::pull(gene_ids)

# filter binary mtx 
filtered_binary_mtx <- binary_mtx |> 
  dplyr::filter(ensembl_gene_id %in% filtered_markers)

readr::write_tsv(filtered_binary_mtx, filtered_markers_mtx_file)
```

```{r, eval=FALSE}
cellassign_call <- glue::glue("
python scripts/run-cellassign.py --anndata_file '{anndata_file}' --output_predictions '{filtered_predictions_file}' --reference '{filtered_markers_mtx_file}'
")

system(cellassign_call)
```

Now we can read in the predictions for the filtered marker gene list.

```{r}
predictions <- readr::read_tsv(filtered_predictions_file)

celltype_assignments <- predictions |>
      tidyr::pivot_longer(
        !barcode,
        names_to = "celltype",
        values_to = "prediction"
      ) |>
      dplyr::group_by(barcode) |>
      dplyr::slice_max(prediction, n = 1) |>
      dplyr::ungroup()

table(celltype_assignments$celltype)
```

Now we are seeing a few more tumor cells, let's see where they are?

```{r}
celltype_df <- marker_mean_exp |> 
  dplyr::left_join(celltype_assignments, by = c("barcodes" = "barcode"))

ggplot(celltype_df, aes(x = UMAP1, UMAP2, color = celltype)) +
  geom_point(size = 0.5, alpha = 0.5) + 
  theme_bw()
```

Not quite where I would expect to see tumor cells. 
Let's look at the mean expression of marker genes between tumor and other cells as classified by `CellAssign`.

```{r}
ggplot(celltype_df, aes(x = mean_exp, fill = celltype)) +
  geom_density() +
  facet_grid(rows = vars(celltype)) +
  theme_bw()
```

Unlike what I would expect, the "other" cells appear to have higher expression of marker genes? 
Maybe using `CellAssign` with only two options and a small set of marker genes doesn't work in the way we would expect.

## Conclusions

- We do see variation in marker gene expression, with a large group of cells expressing multiple marker genes at a higher level than all other cells. 
- This same group of cells has the highest mean gene expression of all marker genes. 
- We see a clear separation in marker gene expression between clusters. 
- Assigning tumor cells by cluster identifies a distinct group of tumor cells. 
- Using `CellAssign` with the same list of marker genes doesn't identify any tumor cells. 
- Using `CellAssign` with a restricted list of marker genes identfies tumor cells that have lower marker gene expression than other cells. 
This is opposite of the assignment we obtained using manual annotation.

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-04-30 analyses/ewings-cell-type/01-marker-gene-tumor-classification.Rmd
add individual marker genes and comparison to automated annotations

---
title: "Manual classification of tumor cells"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

## Introduction

This notebook looks at expression of marker genes for tumor cells in a Ewing sarcoma sample, `r {params$sample_id}`. 
We then see if we can use expression of marker genes to manually classify tumor and normal cells.
The main goal of this notebook is only to identify tumor cells, identification and labelling of the other cells is a separate question that we do not answer here.

- First we look at expression of each of the marker genes across all cells. 
- Then we use a z-transform prior to summing expression of all marker genes. 
Cells with a z-score for any marker gene > 0 are classified as tumor cells. 
- We look at marker genes for these tumor cells using `scran::findMarkers()` and compare to the tumor markers.
- Marker gene expression is examined across clusters and tumor and normal cells are classified using clusters rather than individual cells. 
- Finally, we compare the cells classified as tumor/normal manually to assignments from `SingleR` and `CellAssign`.

## Setup

```{r packages}
suppressMessages(
  # load required packages
  library(SingleCellExperiment),
  library(ggplot2)
)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current")
sample_dir <- file.path(data_dir, "SCPCP000015", params$sample_id)

# The path to this module
module_base <- file.path(repository_base, "analyses", "ewings-cell-type")

# source in helper functions
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)
```

```{r paths}
# Input files
sce_filename <- glue::glue("{params$library_id}_processed_annotated.rds")
sce_file <- file.path(sample_dir, sce_filename)

marker_genes <- file.path(module_base, "references", "tumor-marker-genes.tsv")
```

```{r}
# read in processed sce
sce <- readr::read_rds(sce_file)

# read in marker genes table 
marker_genes_df <- readr::read_tsv(marker_genes) |> 
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
  dplyr::distinct()

marker_genes <- marker_genes_df |> 
  dplyr::filter(cell_type == "tumor") |> 
  dplyr::pull(ensembl_gene_id)
```

## Analysis content

### Explore marker gene expression

The first thing we do here is just create a faceted UMAP showing the expression of each marker gene for tumor cells.

```{r}
# get the gene expression counts for all marker genes
marker_gene_exp <- logcounts(sce[marker_genes, ]) |>
  as.matrix() |>
  t() |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes")

# pull out the UMAP coordinates and make a data frame to use for plotting
umap_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
        starts_with("UMAP")
      ) |> 
  # add in marker gene expression to dataframe 
  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
  # combine all genes into a single column for easy faceting
  tidyr::pivot_longer(
    cols = starts_with("ENSG"),
    names_to = "ensembl_gene_id",
    values_to = "gene_expression"
  ) |> 
  # join with marker gene df to get gene symbols for plotting 
  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression, cluster)
```

```{r}
# faceted umap showing a umap panel for each marker gene 
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
    geom_point(alpha = 0.1, size = 0.2) +
    facet_wrap(vars(gene_symbol)) +
    scale_color_viridis_c() +
    labs(
      color = "Log-normalized gene expression"
    ) +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    theme(
      aspect.ratio = 1,
      legend.position = "bottom",
      axis.title = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
  theme_bw()
```

Looking at this plot, there is definitely some variation in the expression of marker genes. 
It looks like the highest gene expression is seen in the larger group of cells on the bottom right of the UMAP.

```{r}
ggplot(umap_df, aes(x = gene_expression, fill = gene_symbol)) +
  geom_density() +
  theme_bw() +
  facet_wrap(vars(gene_symbol))
```

Now we will transform each of the gene expression vectors by generating z-scores. 
Then we might be able to find a cut off we can use across samples for if marker genes are present in a cell or not.

```{r}
# get the z-scores for all genes, transforming across each column that contains expression for a given gene 
transformed_exp_df <- umap_df |> 
  dplyr::select(barcodes, gene_symbol, gene_expression) |> 
  tidyr::pivot_wider(
    names_from = "gene_symbol",
    values_from = "gene_expression") |> 
  dplyr::mutate(across(!barcodes, ~ scale(.x))) |> 
  tidyr::pivot_longer(
    cols = - barcodes,
    names_to = "gene_symbol",
    values_to = "transformed_gene_expression"
  )

# add the transformed gene expression back to the main df used for plotting
umap_df <- umap_df |> 
  dplyr::left_join(transformed_exp_df, by = c("barcodes", "gene_symbol"))
```

Now we can create the same density plot but looking at z-scores.

```{r}
ggplot(umap_df, aes(x = transformed_gene_expression, fill = gene_symbol)) +
  geom_density() +
  theme_bw() +
  facet_wrap(vars(gene_symbol))
```

It looks like some marker genes have distinct groups of cells with z-score > 0, while other marker genes may not be as informative. 
Keep in mind this could vary somewhat across patients, so I'm wary to remove marker genes without looking across multiple references.

### Classify tumor cells using marker genes only

Let's try and use the marker gene expression to classify tumor cells. 
It looks like we could use a cutoff of z-score > 0 to count that cell as a tumor cell.

We could either count any cell that expresses at least one marker gene > 0 as a tumor cell, or look at the combined expression. 
Let's start with classifying tumor cells as tumor if any marker gene is present (z-score > 0). 
This also means the sum of all z-scores > 0.

Below, we can get the sum of the transformed gene expression of all marker genes and plot in a single UMAP.

```{r}
# calculate sum gene expression across all marker genes in list 
marker_sum_exp <- umap_df |> 
  dplyr::group_by(barcodes) |> 
  dplyr::mutate(sum_exp = sum(transformed_gene_expression)) |> 
  dplyr::select(barcodes, UMAP1, UMAP2, sum_exp, cluster) |> 
  dplyr::distinct()

# plot mean gene expression 
ggplot(marker_sum_exp, aes(x = UMAP1, y = UMAP2, color = sum_exp)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_viridis_c() +
  theme_bw()
```
Similar to the individual plots, it looks like there is one group of cells on the bottom right that has the highest marker gene expression. 
We would anticipate that these are most likely to be the tumor cells.

Now let's classify any cell that has a sum of marker genes > 0 (after z-transformation) as tumor cells.

```{r}
# classify tumor cells based on presence of any marker genes 
marker_sum_exp <- marker_sum_exp |> 
  dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))

ggplot(marker_sum_exp, aes(x = UMAP1, y = UMAP2, color = sum_classification)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

This gives us a rough idea of cells that may be classified as tumor cells. 
Let's look at the expression of each marker gene in tumor vs. normal cells and see if we see distinct separation between all of them or only a subset.

```{r}
# add cell types from sum classification to plotting df 
celltype_df <- umap_df |> 
  dplyr::left_join(marker_sum_exp)

# make plot with gene on y-axis, expression on x-axis and then color density by tumor/normal 
# do all marker genes so differences between tumor/normal or only a subset? 
ggplot(celltype_df, aes(x=transformed_gene_expression, color = sum_classification)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) +
  theme_bw()
```

It looks like ADGRG2, HES1, NROB1, and STEAP1 don't show a lot of difference between cells that might be tumor vs. normal cells and are generally lowly expressed.
CD99 appears to be expressed in both normal and tumor at similar levels. 
All other genes show differences between the two, although at varying degrees. 
Again, we should make sure this is consistent across multiple patients before removing these genes as marker genes.

### Marker genes using `findMarkers`

Below, we will use `scran::findMarkers()` to find any markers between tumor and normal cells. 
Since we used marker genes to classify tumor cells, I would expect that at least some of the top markers would be from our list of tumor markers.

```{r}
# add in tumor/normal classification to colData of SCE 
celltypes_only_df <- marker_sum_exp |> 
  dplyr::select(barcodes, sum_classification)

coldata_df <- as.data.frame(colData(sce)) |> 
  dplyr::left_join(celltypes_only_df)

colData(sce) <- DataFrame(coldata_df, row.names = rownames(coldata_df))

# find markers between tumor and normal 
markers <- scran::findMarkers(sce,
                              groups = sce$sum_classification,
                              pval.type = "all") 
```

```{r}
# pull out top markers for tumor cells 
tumor_markers <- markers$Tumor |> 
  as.data.frame() |> 
  tibble::rownames_to_column("ensembl_gene_id") |> 
  dplyr::left_join(marker_genes_df) |>
  # just take those with FDR < 0.05 and logFC > 1.5
  dplyr::filter(FDR < 0.05, summary.logFC > 1.5) |> 
  dplyr::arrange(FDR)

# get any marker genes that are in both findMarkers and our curated list
marker_gene_intersect <- tumor_markers |> 
  dplyr::filter(cell_type == "tumor")

# print out genes that are marker genes for both 
marker_gene_intersect
```

### Using clusters and marker gene expression to annotate tumor cells

We anticipate that the tumor cells and normal cells would cluster separately. 
First, let's just plot the cluster assignments alone.

These are clusters calculated in `scpca-nf` and its probable that these aren't the best clusters to use. 
Separately, we will evaluate clustering for these samples, but for now we will look at the clusters we have.

```{r}
# plot clusters on UMAP
ggplot(marker_sum_exp, aes(x = UMAP1, y = UMAP2, color = cluster)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

It looks like our large group of cells on the bottom right is comprised of a few clusters while most other groups of cells are a single cluster.

We can then look at marker gene expression across clusters.
Below we visualize the total number of cells in each cluster that are classified as tumor or normal using the sum of marker gene expression. 
Then for each set of cells, how many of those actually express each marker gene.

We expect to see two things:

- Tumor cells and normal cells tend to separate by cluster. 
- Good marker genes will be expressed in tumor cells but not normal cells.

```{r, fig.height=10, fig.width=10}
cluster_plot_df <- celltype_df |> 
  dplyr::mutate(gene_expresed = dplyr::if_else(transformed_gene_expression > 0, TRUE, FALSE)) |> 
  dplyr::select(gene_symbol, gene_expresed, sum_classification, cluster)

ggplot(cluster_plot_df, aes(x = sum_classification, fill = gene_expresed)) +
  geom_bar(stat = "count") +
  facet_grid(rows = vars(gene_symbol),
             cols = vars(cluster)) +
  scale_fill_manual(values = c("grey", "blue")) +
  theme_classic()

```

Looking at this, most clusters are made up mostly of either normal or tumor. 
Also, genes like ADGRG2 and HES1 do not seem specific to tumor cells. There are normal and tumor cells that express the gene and not very many tumor cells actually have expression of ADGRG2. 
STEAP1 and NROB1 although not expressed in very many tumor cells, look to be specific to tumor cells which is promising.

CD99 and CCND1 are expressed in a high proportion of tumor cells but are also expressed in quite a few groups of normal cells. 
This is consistent with the literature for CD99 that it is expressed in normal cells, just at a lower level than in Ewing's cells.

Almost all genes have at least some expression in normal cells. 
Perhaps those cells should also be categorized as tumor cells, or they do have some baseline expression in normal cells which is possible.

Now we can classify the cells based on clustering and marker gene expression.

```{r}
# plot marker gene expression across all clusters 
ggplot(marker_sum_exp, aes(x = sum_exp, fill = cluster)) +
  geom_density() +
  facet_grid(rows = vars(cluster)) +
  theme_bw()
```

Looking at this there are definitely clusters with higher sum expression of the marker genes than other cells. 
If the cluster has a median expression > 0 then all cells in that cluster will get assigned as tumor cells.

```{r}
# classify clusters as tumor/normal 
cluster_classification_df <- marker_sum_exp |> 
  dplyr::group_by(cluster) |> 
  # calculate the median mean gene expression across all cells in a cluster
  dplyr::summarise(median_cluster_exp = median(sum_exp)) |> 
  # use 0.75 as the cutoff for tumor/normal 
  dplyr::mutate(cluster_classification = dplyr::if_else(median_cluster_exp > 0, "Tumor", "Normal"))

# get individual cell classifications
celltype_classification_df <- marker_sum_exp |> 
  dplyr::left_join(cluster_classification_df)

# umap showing tumor and normal cells 
ggplot(celltype_classification_df, aes(x = UMAP1, y = UMAP2, color = cluster_classification)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

### Compare with SingleR and CellAssign classifictions

The last thing we will do here is compare our manual annotations with the annotations using `SingleR` and `CellAssign`. 
Looking at the annotations and UMAPs in the supplemental report, it looks like our group of "Tumor" cells are classified mostly as "Muscle cells" by both SingleR and CellAssign. 
This makes sense since Ewing's sarcoma cells tend to resemble muscle cells.

I would expect the Endothelial cells, Fibroblasts, and Macrophages assigned by `CellAssign` and `SingleR` to line up with the normal cells in this sample.

Let's create a heatmap looking at the Jaccard similarity index between each of the assigned cell types and the tumor/normal classification.

```{r}
# get all assigned cell types in one dataframe 
celltypes_df <- as.data.frame(colData(sce)) |> 
  dplyr::select(barcodes, singler_celltype_annotation, cellassign_celltype_annotation) |> 
  dplyr::left_join(marker_sum_exp)

celltype_columns <- c(
  "singler_celltype_annotation",
  "cellassign_celltype_annotation"
)

# create jaccard matrices for SingleR and CellAssign compared to tumor/normal 
jaccard_matrices <- celltype_columns |>
  purrr::map(\(name) {
    make_jaccard_matrix(
      celltypes_df,
      "sum_classification",
      name
    )
  }) |> 
  purrr::set_names("SingleR", "CellAssign")
```

```{r, fig.height=10, fig.width=7}
# Set heatmap padding option
heatmap_padding <- 0.2
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))

# list of heatmaps looking at SingleR/ CellAssign vs tumor/normal 
heatmap <- jaccard_matrices |>
  purrr::imap(
    \(celltype_mat, celltype_method) {
      ComplexHeatmap::Heatmap(
        t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
        border = TRUE,
        ## Row parameters
        cluster_rows = TRUE,
        row_title = celltype_method,
        row_title_gp = grid::gpar(fontsize = 12),
        row_title_side = "left",
        row_names_side = "left",
        row_dend_side = "right",
        row_names_gp = grid::gpar(fontsize = 10),
        ## Column parameters
        cluster_columns = FALSE,
        column_title = "",
        column_title_gp = grid::gpar(fontsize = 12),
        column_names_side = "bottom",
        column_names_gp = grid::gpar(fontsize = 10),
        column_names_rot = 90,
        ## Legend parameters
        heatmap_legend_param = list(
          title = "Jaccard index",
          direction = "vertical",
          legend_width = unit(1.5, "in")
        ),
        show_heatmap_legend = celltype_method == "SingleR",
      )
    }) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )
```

When looking at `SingleR`, we see that the tumor cells line up with smooth muscle cells, chondrocytes, neurons, and fibroblasts. 
Ewing's cells have a neuronal like phenotype and express high levels of ECM genes, which may explain the overlap with chondrocytes and fibroblasts. 
We do see that macrophages and endothelial cells tend to line up with the normal cells, which is expected.

For CellAssign, all tumor cells are pulmonary vascular smooth muscle cells. 
We also see tha normal cells contain some smooth muscle cells, but mostly are fibroblasts and endothelial cells.

Looking at the heatmap in the supplemental report, the endothelial cells called by SingleR and CellAssign have a lot of overlap. 
Let's look at expression of marker genes in these endothlial cells compared to cells that are categorized as smooth muscle cells.

```{r}
select_celltype_df <- celltypes_df |> 
  dplyr::mutate(
    automated_annotation = dplyr::case_when(
      singler_celltype_annotation == "smooth muscle cell" & cellassign_celltype_annotation == "Pulmonary vascular smooth muscle cells" ~ "Muscle cell",
      singler_celltype_annotation == "endothelial cell" & cellassign_celltype_annotation == "Endothelial cells" ~ "Endothelial cell"
    )
  ) |> 
  dplyr::select(barcodes, automated_annotation)

annotation_plot_df <- umap_df |> 
  dplyr::left_join(select_celltype_df) |>
  dplyr::filter(!is.na(automated_annotation))

ggplot(annotation_plot_df, aes(x = transformed_gene_expression, color = automated_annotation)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) +
  theme_bw()
```

Consistent with previous findings, ADGRG2, HES1, NROB1, and STEAP1 don't seem to show any expression in either groups of cells. 
CCND1 and CD99 appear to have similar expression across both groups of cells, while MAPT and PRKCB seem to have the most distinct expression profiles between muscle and endothelial cells.

## Conclusions

- We do see variation in marker gene expression across cells, suggesting that there are both tumor and normal cells here. - Using a z-score cut-off identifies a group of cells that appear to be tumor. 
- This mostly lines up with the default clustering that has been performed, where tumor and normal cells have been separated.
- The top marker genes for the "tumor" cells contain PRKCB and MAPT, two of the marker genes in our gene list. 
- It's likely that ADGRG2, HES1, NROB1, and STEAP1 will not be informative in assigning tumor cells. 
- We may be weary of genes like CD99 too as expression is more consistent across cells rather than specific to a group of cells. 
- Endothelial cells as identified by SingleR and CellAssign generally have low expression of these marker genes and muscle cells have higher expression of marker genes.

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-05-01 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd
initiate cellassign notebook

--- 2024-04-25 analyses/ewings-cell-type/01-marker-gene-classification-cellassign.Rmd
@@
 ---
-title: "Explore tumor cell classification with marker genes and CellAssign - SCPCS000490"
+title: "Explore tumor cell classification with CellAssign"
 author: Ally Hawkins
@@
 
-This notebook looks at expression of marker genes for tumor cells in a Ewing sarcoma sample, `r {params$sample_id}`. 
-Additionally, we see if we can use those marker genes to classify tumor and normal cells using both manual annotation and `CellAssign`.
+This notebook looks at classifying tumor cells with `CellAssign` in a Ewing sarcoma sample, `r {params$sample_id}`.
 
-The first section focuses on assigning tumor and normal cells by looking at marker gene expression alone:
-
-- First we look at expression of each of the marker genes across all cells. 
-- Then we plot the mean expression of all marker genes across all cells. 
-- Next, we perform clustering and see if we can assign each cluster as tumor or normal cells based on mean expression of marker genes.
-
-The next section uses `CellAssign` to assign tumor cells with the same set of marker genes:
-
-- We use `CellAssign` with both the full list of marker genes and filtered list of marker genes. 
+- We use `CellAssign` with both a full list of marker genes and filtered list of marker genes. 
 - We then compare the average gene expression of all marker genes between tumor and other cells. 
+- Finally we use `CellAssign` with marker genes from [Visser et al.](https://doi.org/10.1158/2767-9764.CRC-23-0027) specified in the supplemental methods.
 
@@
 
-marker_genes <- file.path(module_base, "references", "tumor_marker_genes.tsv")
+# tumor marker genes
+tumor_markers_file <- file.path(module_base, "references", "tumor-marker-genes.tsv")
+# all markers 
+all_markers_file <- file.path(module_base, "references", "all-marker-genes.tsv")
 
-# Output files
-tumor_markers_mtx_file <- file.path(module_base, "references", "cellassign_refs", "tumor-marker-cellassign.tsv")
-filtered_markers_mtx_file <- file.path(module_base, "references", "cellassign_refs", "filtered-tumor-marker-cellassign.tsv")
-cellassign_predictions_file <- file.path(module_base, "results", "cellassign-tumor-predictions.tsv")
-filtered_predictions_file <- file.path(module_base, "results", "filtered-cellassign-tumor-predictions.tsv")
+# Output reference matrix files
+ref_dir <- file.path(module_base, "references", "cellassign_refs")
+
+tumor_markers_mtx_file <- file.path(ref_dir, "tumor-marker-cellassign.tsv")
+filtered_markers_mtx_file <- file.path(ref_dir, "filtered-tumor-marker-cellassign.tsv")
+visser_markers_mtx_file <- file.path(ref_dir, "visser-all-marker-cellassign.tsv")
+
+# output predictions from cellassign 
+results_dir <- file.path(module_base, "results")
+cellassign_predictions_file <- file.path(results_dir, "cellassign-tumor-predictions.tsv")
+filtered_predictions_file <- file.path(results_dir, "filtered-cellassign-tumor-predictions.tsv")
+visser_predictions_file <- file.path(results_dir, "visser-cellassign-predictions.tsv")
+
 ```
@@
 
-# read in marker genes table 
-marker_genes_df <- readr::read_tsv(marker_genes) |> 
+# read in tumor marker genes table 
+tumor_markers_df <- readr::read_tsv(tumor_markers_file) |> 
   # account for genes being from multiple sources
@@
 
-marker_genes <- marker_genes_df |> 
-  dplyr::filter(cell_type == "tumor") |> 
-  dplyr::pull(ensembl_gene_id)
+# read in all maker genes 
+all_markers_df <- readr::read_tsv(all_markers_file) |> 
+  # account for genes being from multiple sources
+  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
+  dplyr::distinct()
 ```
 
-## Analysis content
-
-### Explore marker gene expression
-
-The first thing we do here is just create a faceted UMAP showing the expression of each marker gene for tumor cells.
+Define some helper functions for creating a reference matrix and obtaining cell type annotations from `CellAssign` predictions.
 
 ```{r}
-# get the gene expression counts for all marker genes
-marker_gene_exp <- logcounts(sce[marker_genes, ]) |>
-    as.matrix() |>
-    t() |>
-    as.data.frame() |>
-    tibble::rownames_to_column("barcodes")
+# function for creating a binary matrix with cell types as columns and marker genes as rows 
+build_binary_mtx <- function(marker_genes_df,
+                             gene_col, 
+                             cell_type_col) {
 
-# pull out the UMAP coordinates and make a data frame to use for plotting
-umap_df <- sce |> 
-  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
-  # replace UMAP.1 with UMAP1
-  dplyr::rename_with(
-        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
-        starts_with("UMAP")
-      ) |> 
-  # add in marker gene expression to dataframe 
-  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
-  # combine all genes into a single column for easy faceting
-  tidyr::pivot_longer(
-    cols = starts_with("ENSG"),
-    names_to = "ensembl_gene_id",
-    values_to = "gene_expression"
-  ) |> 
-  # join with marker gene df to get gene symbols for plotting 
-  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
-  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression, clusters)
-```
-
-```{r}
-# faceted umap showing a umap panel for each marker gene 
-ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
-    geom_point(alpha = 0.1, size = 0.2) +
-    facet_wrap(vars(gene_symbol)) +
-    scale_color_viridis_c() +
-    labs(
-      color = "Log-normalized gene expression"
-    ) +
-    # remove axis numbers and background grid
-    scale_x_continuous(labels = NULL, breaks = NULL) +
-    scale_y_continuous(labels = NULL, breaks = NULL) +
-    theme(
-      aspect.ratio = 1,
-      legend.position = "bottom",
-      axis.title = element_text(size = 9, color = "black"),
-      strip.text = element_text(size = 8),
-      legend.title = element_text(size = 9),
-      legend.text = element_text(size = 8)
-    ) +
-    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
-  theme_bw()
-```
-
-Looking at this plot, there is definitely some variation in the expression of marker genes. 
-It looks like the highest gene expression is seen in the larger group of cells on the bottom right of the UMAP.
-
-Below, we can get the mean gene expression of all marker genes and plot in a single UMAP.
-
-```{r}
-# calculate mean gene expression across all marker genes in list 
-marker_mean_exp <- umap_df |> 
-  dplyr::group_by(barcodes) |> 
-  dplyr::mutate(mean_exp = mean(gene_expression)) |> 
-  dplyr::select(barcodes, UMAP1, UMAP2, mean_exp, clusters) |> 
-  dplyr::distinct()
-
-# plot mean gene expression 
-ggplot(marker_mean_exp, aes(x = UMAP1, y = UMAP2, color = mean_exp)) +
-  geom_point(size = 0.5, alpha = 0.5) +
-  scale_color_viridis_c() +
-  theme_bw()
-```
-
-Similar to the individual plots, it looks like there is one group of cells on the bottom right that has the highest mean gene expression. 
-We would anticipate that these are most likely to be the tumor cells.
-
-### Using clusters and marker gene expression to annotate tumor cells
-
-We anticipate that the tumor cells and normal cells would cluster separately. 
-First, let's just plot the cluster assignments alone.
-
-```{r}
-ggplot(marker_mean_exp, aes(x = UMAP1, y = UMAP2, color = clusters)) +
-  geom_point(size = 0.5, alpha = 0.5) +
-  theme_bw()
-```
-
-It looks like our large group of cells on the bottom right is comprised of a few clusters while most other groups of cells are a single cluster. 
-Let's see if there's any difference in mean expression of the marker genes across the clusters. 
-Based on the gene expression I would expect most cells in the bottom right to be considered tumor and all other cells to be normal.
-
-```{r}
-# plot marker gene expression across all clusters 
-ggplot(marker_mean_exp, aes(x = mean_exp, fill = clusters)) +
-  geom_density() +
-  facet_grid(rows = vars(clusters)) +
-  theme_bw()
-```
-
-Looking at this there are definitely distinct groups of cells with higher mean expression of the marker genes than other cells. 
-Before doing anything fancy, let's use this to manually assign tumor and normal cells. 
-One approach is to assign based on the cluster, so if the cluster has a median expression > 0.75 then all cells in that cluster would get assigned as tumor cells. 
-We could also pick a mean expression cutoff and use that to assign each cell individually, but there are some tails that overlap.
-
-```{r}
-# classify clusters as tumor/normal 
-cluster_classification_df <- marker_mean_exp |> 
-  dplyr::group_by(clusters) |> 
-  # calculate the median mean gene expression across all cells in a cluster
-  dplyr::summarise(median_cluster_exp = median(mean_exp)) |> 
-  # use 0.75 as the cutoff for tumor/normal 
-  dplyr::mutate(tumor_normal_classification = dplyr::if_else(median_cluster_exp > 0.75, "Tumor", "Normal"))
-
-# get individual cell classifications
-celltype_classification_df <- marker_mean_exp |> 
-  dplyr::left_join(cluster_classification_df)
-
-# umap showing tumor and normal cells 
-ggplot(celltype_classification_df, aes(x = UMAP1, y = UMAP2, color = tumor_normal_classification)) +
-  geom_point(size = 0.5, alpha = 0.5) +
-  theme_bw()
-```
-
-### Run CellAssign with marker genes
-
-Now let's see if we can use an automated marker gene-based method to identify tumor cells. 
-To do this we will use `CellAssign` with a reference matrix created from the marker gene list.
-
-First, we need to create a binary matrix with two columns, "tumor" and "other". 
-The "other" will refer to any normal cells.
-
-```{r}
   binary_mtx <- marker_genes_df |>
-  dplyr::select(cell_type, ensembl_gene_id) |> 
+  dplyr::select({{cell_type_col}}, {{gene_col}}) |> 
   unique() |> 
   tidyr::pivot_wider(
-    id_cols = ensembl_gene_id,
-    names_from = cell_type,
-    values_from = cell_type,
+    id_cols = {{gene_col}},
+    names_from = {{cell_type_col}},
+    values_from = {{cell_type_col}},
     values_fn = length,
@@
 
-# export matrix 
-readr::write_tsv(binary_mtx, tumor_markers_mtx_file)
-```
+  return(binary_mtx)
+}
 
-Now run `CellAssign` with the full marker gene list as the reference.
-
-```{r, eval=FALSE}
-cellassign_call <- glue::glue("
-python scripts/run-cellassign.py \
-  --anndata_file '{anndata_file}' \
-  --output_predictions '{cellassign_predictions_file}' \
-  --reference '{tumor_markers_mtx_file}'
-")
-
-system(cellassign_call)
-```
-
-We can then read in the predictions file and find the predicted cell type for each cell, either "tumor" or "other".
-
-```{r}
-# read in predictions
-predictions <- readr::read_tsv(cellassign_predictions_file)
-
+# function to get assigned cell type based on max prediction returned by CellAssign 
+get_celltype_assignment <- function(predictions){
   # get individual cell type assignments
@@
     dplyr::ungroup()
+
+  return(celltype_assignments)
+}
+```
+
+## Analysis content
+
+### Run CellAssign with only tumor marker genes
+
+First let's see if we can use `CellAssign` to classify cells as tumor or normal using just the list of tumor marker genes.
+
+First, we need to create a binary matrix with two columns, "tumor" and "other". 
+The "other" will refer to any normal cells.
+
+```{r}
+tumor_ref_mtx <- build_binary_mtx(tumor_markers_df, 
+                                  gene_col = "ensembl_gene_id",
+                                  cell_type_col = "cell_type")
+
+# export matrix 
+readr::write_tsv(tumor_ref_mtx, tumor_markers_mtx_file)
+```
+
+Now run `CellAssign` with the full marker gene list as the reference.
+
+```{r, eval=FALSE}
+cellassign_call <- glue::glue(
+  "python scripts/run-cellassign.py",
+  "--anndata_file '{anndata_file}'",
+  "--output_predictions '{cellassign_predictions_file}'",
+  "--reference '{tumor_markers_mtx_file}'"
+)
+
+system(cellassign_call)
+```
+
+We can then read in the predictions file and find the predicted cell type for each cell, either "tumor" or "other".
+
+```{r}
+# read in predictions file
+tumor_predictions <- readr::read_tsv(cellassign_predictions_file)
+
+# get cell type assignments 
+tumor_celltype_assignments <- get_celltype_assignment(tumor_predictions)
+
 # create a table of assignments 
-table(celltype_assignments$celltype)
+table(tumor_celltype_assignments$celltype)
 ```
@@
 I find this hard to believe since this sample was taken directly from a Ewing sarcoma tumor, so there are bound to be tumor cells.
+
+### Run CellAssign with a filtered set of tumor marker genes
 
@@
 # filter binary mtx 
-filtered_binary_mtx <- binary_mtx |> 
+filtered_ref_mtx <- tumor_ref_mtx |> 
   dplyr::filter(ensembl_gene_id %in% filtered_markers)
 
-readr::write_tsv(filtered_binary_mtx, filtered_markers_mtx_file)
+readr::write_tsv(filtered_ref_mtx, filtered_markers_mtx_file)
 ```
@@
 ```{r, eval=FALSE}
-cellassign_call <- glue::glue("
-python scripts/run-cellassign.py --anndata_file '{anndata_file}' --output_predictions '{filtered_predictions_file}' --reference '{filtered_markers_mtx_file}'
-")
+cellassign_call <- glue::glue(
+  "python scripts/run-cellassign.py",
+  "--anndata_file '{anndata_file}'",
+  "--output_predictions '{filtered_predictions_file}'",
+  "--reference '{filtered_markers_mtx_file}'"
+)
 
@@
 ```{r}
-predictions <- readr::read_tsv(filtered_predictions_file)
+filtered_predictions <- readr::read_tsv(filtered_predictions_file)
 
-celltype_assignments <- predictions |>
-      tidyr::pivot_longer(
-        !barcode,
-        names_to = "celltype",
-        values_to = "prediction"
-      ) |>
-      dplyr::group_by(barcode) |>
-      dplyr::slice_max(prediction, n = 1) |>
-      dplyr::ungroup()
+filtered_celltype_assignments <- get_celltype_assignment(filtered_predictions)
 
-table(celltype_assignments$celltype)
+table(filtered_celltype_assignments$celltype)
 ```
@@
 
+### Run CellAssign with tumor markers and all marker genes from Visser et al.
+
+```{r}
+visser_ref_mtx <- build_binary_mtx(all_markers_df, 
+                                   gene_col = "ensembl_gene_id",
+                                   cell_type_col = "cell_type")
+
+# export matrix 
+readr::write_tsv(visser_ref_mtx, visser_markers_mtx_file)
+```
+
+Now run `CellAssign` with the full marker gene list as the reference.
+
+```{r, eval=FALSE}
+cellassign_call <- glue::glue(
+  "python scripts/run-cellassign.py ",
+  "--anndata_file '{anndata_file}' ",
+  "--output_predictions '{visser_predictions_file}' ",
+  "--reference '{visser_markers_mtx_file}'"
+)
+
+system(cellassign_call)
+```
+
+```{r}
+visser_predictions <- readr::read_tsv(visser_predictions_file)
+
+visser_celltype_assignments <- get_celltype_assignment(visser_predictions)
+
+table(visser_celltype_assignments$celltype)
+```
+
 ## Conclusions
 
-- We do see variation in marker gene expression, with a large group of cells expressing multiple marker genes at a higher level than all other cells. 
-- This same group of cells has the highest mean gene expression of all marker genes. 
-- We see a clear separation in marker gene expression between clusters. 
-- Assigning tumor cells by cluster identifies a distinct group of tumor cells. 
 - Using `CellAssign` with the same list of marker genes doesn't identify any tumor cells. 
@@
 This is opposite of the assignment we obtained using manual annotation.
+
+```{r}
+
+```
 

### 2024-05-01 analyses/ewings-cell-type/01-marker-gene-tumor-classification.Rmd
save tumor normal classifications

@@
 # Input files
-sce_filename <- glue::glue("{params$library_id}_processed_annotated.rds")
+sce_filename <- glue::glue("{params$library_id}_processed.rds")
 sce_file <- file.path(sample_dir, sce_filename)
@@
 marker_genes <- file.path(module_base, "references", "tumor-marker-genes.tsv")
+
+# output tumor/normal classifications
+results_dir <- file.path(module_base, "results", "marker_gene_analysis")
+fs::dir_create(results_dir)
+
+classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
+output_classifications_file <- file.path(results_dir, classifications_filename)
 ```
@@
 
+## Save outputs
+
+```{r}
+# get a dataframe with barcodes, sum_classification and cluster_classification 
+celltype_classification_df <- celltype_classification_df |> 
+  dplyr::select(
+    barcodes,
+    marker_gene_classification = sum_classification, 
+    cluster_classification
+  )
+
+readr::write_tsv(celltype_classification_df, output_classifications_file)
+```
+
 ## Session Info

### 2024-05-01 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd
add visser marker gene analysis

@@
 module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
+
+# source in helper functions
+jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
+source(jaccard_functions)
 ```
@@
 
+# tumor normal classifications 
+classifications_results_dir <- file.path(module_base, "results", "marker_gene_analysis")
+classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
+classifications_file <- file.path(classifications_results_dir, classifications_filename)
+
 # tumor marker genes
@@
 all_markers_file <- file.path(module_base, "references", "all-marker-genes.tsv")
+# panglao markers
+panglao_markers_file <- file.path(module_base, "references", "PanglaoDB_markers_2020-03-27.tsv")
 
@@
 visser_markers_mtx_file <- file.path(ref_dir, "visser-all-marker-cellassign.tsv")
+panglao_markers_mtx_file <- file.path(ref_dir, "panglao-endo-fibro-cellassign.tsv")
 
 # output predictions from cellassign 
-results_dir <- file.path(module_base, "results")
-cellassign_predictions_file <- file.path(results_dir, "cellassign-tumor-predictions.tsv")
-filtered_predictions_file <- file.path(results_dir, "filtered-cellassign-tumor-predictions.tsv")
-visser_predictions_file <- file.path(results_dir, "visser-cellassign-predictions.tsv")
+preds_results_dir <- file.path(module_base, "results", "cellassign_predictions")
+fs::dir_create(preds_results_dir)
 
+cellassign_predictions_file <- file.path(preds_results_dir, "cellassign-tumor-predictions.tsv")
+filtered_predictions_file <- file.path(preds_results_dir, "filtered-cellassign-tumor-predictions.tsv")
+visser_predictions_file <- file.path(preds_results_dir, "visser-cellassign-predictions.tsv")
+panglao_predictions_file <- file.path(preds_results_dir, "panglao-endo-fibro-predictions.tsv")
 ```
@@
 
-In looking at the original UMAP there are quite a few genes that have very low expression across all cells, so perhaps they are adding noise to the model. 
+In looking at the UMAPs of all marker genes (see `01-marker-gene-tumor-classification.Rmd`) there are quite a few genes that have very low expression across all cells, so perhaps they are adding noise to the model.
 Although this is a bit circular, we could restrict our list of marker genes to be only those that have a high mean gene expression in our dataset? 
@@
   as.data.frame() |> 
-  dplyr::filter(gene_symbol %in% marker_genes_df$gene_symbol)
+  dplyr::filter(gene_symbol %in% tumor_markers_df$gene_symbol)
 rowdata_df
@@
 
-Now we are seeing a few more tumor cells, let's see where they are?
+Now we are seeing a few more tumor cells, let's see where they are on the UMAP. 
+We can also compare with the tumor normal classifications we made using marker genes alone.
 
 ```{r}
-celltype_df <- marker_mean_exp |> 
-  dplyr::left_join(celltype_assignments, by = c("barcodes" = "barcode"))
+# first read in tumor normal classifications 
+classifications_df <- readr::read_tsv(classifications_file)
 
-ggplot(celltype_df, aes(x = UMAP1, UMAP2, color = celltype)) +
+filtered_celltype_assignments <- filtered_celltype_assignments |> 
+  dplyr::rename(filtered_cellassign_annotations = celltype)
+
+celltype_df <- sce |> 
+  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
+  # replace UMAP.1 with UMAP1
+  dplyr::rename_with(
+        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
+        starts_with("UMAP")
+      ) |> 
+  # join with tumor normal
+  dplyr::left_join(classifications_df) |> 
+  # join with filtered cell type assignments from cellassign 
+  dplyr::left_join(filtered_celltype_assignments, by = c("barcodes" = "barcode"))
+
+ggplot(celltype_df, aes(x = UMAP1, UMAP2, color = filtered_cellassign_annotations)) +
   geom_point(size = 0.5, alpha = 0.5) + 
@@
 Not quite where I would expect to see tumor cells. 
+Let's see how many tumor cells overlap with our manual classifications using the Jaccard similarity index.
+
+```{r}
+jaccard_mtx <- make_jaccard_matrix(celltype_df, "filtered_cellassign_annotations", "marker_gene_classification")
+
+ComplexHeatmap::Heatmap(
+  t(jaccard_mtx), # transpose because matrix rows are in common & we want a vertical arrangement
+  col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
+  border = TRUE,
+  ## Row parameters
+  cluster_rows = FALSE,
+  row_title = "Marker gene",
+  row_title_gp = grid::gpar(fontsize = 12),
+  row_title_side = "left",
+  row_names_side = "left",
+  row_dend_side = "right",
+  row_names_gp = grid::gpar(fontsize = 10),
+  ## Column parameters
+  cluster_columns = FALSE,
+  column_title = "CellAssign",
+  column_title_gp = grid::gpar(fontsize = 12),
+  column_names_side = "bottom",
+  column_names_gp = grid::gpar(fontsize = 10),
+  column_names_rot = 90,
+  ## Legend parameters
+  heatmap_legend_param = list(
+    title = "Jaccard index",
+    direction = "vertical",
+    legend_width = unit(1.5, "in")
+  ),
+  show_heatmap_legend = TRUE,
+)
+```
+As expected, it looks like CellAssign is assigning very few cells to tumor that overlap with the tumor cells annotated with marker gene analysis. 
+It looks like most tumor cells are being classified as normal with the marker gene analysis.
+
 Let's look at the mean expression of marker genes between tumor and other cells as classified by `CellAssign`.
@@
 ```{r}
-ggplot(celltype_df, aes(x = mean_exp, fill = celltype)) +
+# get marker gene expression 
+marker_gene_exp <- logcounts(sce[filtered_markers, ]) |>
+  as.matrix() |>
+  t() |>
+  as.data.frame() |>
+  tibble::rownames_to_column("barcodes")
+
+marker_gene_exp_df <- celltype_df |> 
+  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
+  tidyr::pivot_longer(
+    cols = starts_with("ENSG"),
+    names_to = "ensembl_gene_id",
+    values_to = "transformed_gene_expression"
+  ) |> 
+  dplyr::group_by(barcodes) |> 
+  dplyr::mutate(sum_exp = sum(transformed_gene_expression))
+
+ggplot(marker_gene_exp_df, aes(x = sum_exp, fill = filtered_cellassign_annotations)) +
   geom_density() +
-  facet_grid(rows = vars(celltype)) +
+  facet_grid(rows = vars(filtered_cellassign_annotations)) +
   theme_bw()
@@
 ### Run CellAssign with tumor markers and all marker genes from Visser et al.
+
+Perhaps we need to add a few more cell types, rather than just have the two options. 
+Below we will use CellAssign but with all tumor marker genes and all marker genes in Visser et al. for non-tumor cells.
 
@@
 ```
+Here we see that there are still only a few hundred tumor cells and most cells are endothelial cells. 
+Generally, I expect that there to be a large portion of tumor cells with much fewer cells from normal cell types.
+
+Below we will compare assignments between CellAssign and manual marker gene assignment.
+
+```{r}
+# first rename column to combine with cell type df 
+visser_celltype_assignments <- visser_celltype_assignments |> 
+  dplyr::rename(visser_cellassign_annotations = celltype)
+
+celltype_df <- celltype_df |> 
+  dplyr::left_join(visser_celltype_assignments, by = c("barcodes" = "barcode"))
+
+ggplot(celltype_df, aes(x = UMAP1, UMAP2, color = visser_cellassign_annotations)) +
+  geom_point(size = 0.5, alpha = 0.5) + 
+  theme_bw()
+```
+
+```{r}
+jaccard_mtx <- make_jaccard_matrix(celltype_df, "visser_cellassign_annotations", "marker_gene_classification")
+
+ComplexHeatmap::Heatmap(
+  t(jaccard_mtx), # transpose because matrix rows are in common & we want a vertical arrangement
+  col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
+  border = TRUE,
+  ## Row parameters
+  cluster_rows = FALSE,
+  row_title = "Marker gene",
+  row_title_gp = grid::gpar(fontsize = 12),
+  row_title_side = "left",
+  row_names_side = "left",
+  row_dend_side = "right",
+  row_names_gp = grid::gpar(fontsize = 10),
+  ## Column parameters
+  cluster_columns = FALSE,
+  column_title = "CellAssign - Visser",
+  column_title_gp = grid::gpar(fontsize = 12),
+  column_names_side = "bottom",
+  column_names_gp = grid::gpar(fontsize = 10),
+  column_names_rot = 90,
+  ## Legend parameters
+  heatmap_legend_param = list(
+    title = "Jaccard index",
+    direction = "vertical",
+    legend_width = unit(1.5, "in")
+  ),
+  show_heatmap_legend = TRUE,
+)
+```
+
+```{r}
+all_markers <- all_markers_df |> 
+  dplyr::pull(ensembl_gene_id)
+
+marker_gene_exp <- logcounts(sce[all_markers, ]) |>
+  as.matrix() |>
+  t() |>
+  as.data.frame() |>
+  tibble::rownames_to_column("barcodes")
+
+marker_gene_exp_df <- celltype_df |> 
+  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
+  tidyr::pivot_longer(
+    cols = starts_with("ENSG"),
+    names_to = "ensembl_gene_id",
+    values_to = "gene_expression"
+  ) |> 
+  dplyr::left_join(all_markers_df)
+
+celltypes <- unique(all_markers_df$cell_type)
+
+celltypes |> 
+  purrr::map(\(celltype){
+    plot_df <- marker_gene_exp_df |> 
+      dplyr::filter(cell_type == celltype)
+
+    # faceted umap showing a umap panel for each marker gene 
+    ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
+      geom_point(alpha = 0.1, size = 0.2) +
+      facet_wrap(vars(gene_symbol)) +
+      scale_color_viridis_c() +
+      labs(
+        title = celltype,
+        color = "Log-normalized gene expression"
+      ) +
+      # remove axis numbers and background grid
+      scale_x_continuous(labels = NULL, breaks = NULL) +
+      scale_y_continuous(labels = NULL, breaks = NULL) +
+      theme(
+        aspect.ratio = 1,
+        legend.position = "bottom",
+        axis.title = element_text(size = 9, color = "black"),
+        strip.text = element_text(size = 8),
+        legend.title = element_text(size = 9),
+        legend.text = element_text(size = 8)
+      ) +
+      guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
+      theme_bw()
+
+  })
+```
+
+Then we will look at the gene expression by cell type for each marker gene. 
+We would expect that the specified marker genes for each cell type are highly expressed in those cells compared to other cells.
+
+```{r}
+transformed_exp_df <- marker_gene_exp_df |> 
+  dplyr::select(barcodes, gene_symbol, gene_expression) |> 
+  tidyr::pivot_wider(
+    names_from = "gene_symbol",
+    values_from = "gene_expression") |> 
+  dplyr::mutate(across(!barcodes, ~ scale(.x))) |> 
+  tidyr::pivot_longer(
+    cols = - barcodes,
+    names_to = "gene_symbol",
+    values_to = "transformed_gene_expression"
+  )
+
+# add the transformed gene expression back to the main df used for plotting
+grouped_celltype_df <- marker_gene_exp_df |> 
+  dplyr::left_join(transformed_exp_df, by = c("barcodes", "gene_symbol")) |> 
+  dplyr::group_by(barcodes, cell_type) |> 
+  dplyr::mutate(sum_exp = sum(transformed_gene_expression))
+```
+
+```{r}
+ggplot(grouped_celltype_df, aes(x = sum_exp, color = cell_type)) +
+  geom_density() +
+  facet_wrap(vars(visser_cellassign_annotations)) +
+  theme_bw() +
+  labs(
+    color = "Marker gene group"
+  )
+```
+
+### Use markers from PanglaoDB
+
+```{r}
+panglao_markers_df <- readr::read_tsv(panglao_markers_file) |> 
+  dplyr::filter(`cell type` %in% c("Endothelial cells", "Fibroblasts")) |> 
+  dplyr::select(`official gene symbol`, `cell type`)
+
+panglao_tumor_markers_df <- rowData(sce) |> 
+  as.data.frame() |> 
+  dplyr::filter(gene_symbol %in% panglao_markers_df$`official gene symbol`) |> 
+  dplyr::left_join(panglao_markers_df, by = c("gene_symbol" = "official gene symbol")) |> 
+  dplyr::select(gene_symbol, ensembl_gene_id = gene_ids, cell_type = `cell type`) |> 
+  dplyr::bind_rows(tumor_markers_df)
+
+```
+
+```{r}
+panglao_tumor_ref_mtx <- build_binary_mtx(panglao_tumor_markers_df, 
+                                          gene_col = "ensembl_gene_id",
+                                          cell_type_col = "cell_type")
+
+# export matrix 
+readr::write_tsv(panglao_tumor_ref_mtx, panglao_markers_mtx_file)
+```
+
+Now run `CellAssign` with the full marker gene list as the reference.
+
+```{r, eval=FALSE}
+cellassign_call <- glue::glue(
+  "python scripts/run-cellassign.py ",
+  "--anndata_file '{anndata_file}' ",
+  "--output_predictions '{panglao_predictions_file}' ",
+  "--reference '{panglao_markers_mtx_file}' "
+)
+
+system(cellassign_call)
+```
+
+We can then read in the predictions file and find the predicted cell type for each cell, either "tumor" or "other".
+
+```{r}
+# read in predictions file
+panglao_predictions <- readr::read_tsv(panglao_markers_mtx_file)
+
+# get cell type assignments 
+panglao_celltype_assignments <- get_celltype_assignment(panglao_predictions)
+
+# create a table of assignments 
+table(panglao_celltype_assignments$celltype)
+```
 

### 2024-05-01 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd
clean up comments and text

@@
 
-- We use `CellAssign` with both a full list of marker genes and filtered list of marker genes. 
-- We then compare the average gene expression of all marker genes between tumor and other cells. 
-- Finally we use `CellAssign` with marker genes from [Visser et al.](https://doi.org/10.1158/2767-9764.CRC-23-0027) specified in the supplemental methods.
+We use `CellAssign` with a variety of references: 
+- A full list of tumor marker genes only. 
+- A filtered list of tumor marker genes only. 
+- Tumor marker genes and non-tumor marker genes from [Visser et al.](https://doi.org/10.1158/2767-9764.CRC-23-0027) specified in the supplemental methods. 
+- Tumor marker genes and marker genes from endothelial cells and fibroblast marker genes in PanglaoDB.
+
+We then compare the results from running `CellAssign` with each reference to the tumor/normal classification obtained in `01-marker-gene-tumor-classification.Rmd`. 
+We also look at marker gene expression in the assigned cells.
 
@@
 
-anndata_filename <- glue::glue("{params$library_id}_processed_rna.hdf5")
+anndata_filename <- glue::glue("{params$library_id}_processed_rna.h5ad")
 anndata_file <- file.path(sample_dir, anndata_filename)
@@
 
-First, we need to create a binary matrix with two columns, "tumor" and "other". 
+We need to create a binary matrix with two columns, "tumor" and "other". 
 The "other" will refer to any normal cells.
@@
 Now we are seeing a few more tumor cells, let's see where they are on the UMAP.
-We can also compare with the tumor normal classifications we made using marker genes alone.
 
@@
     names_to = "ensembl_gene_id",
-    values_to = "transformed_gene_expression"
+    values_to = "gene_expression"
   ) |> 
   dplyr::group_by(barcodes) |> 
-  dplyr::mutate(sum_exp = sum(transformed_gene_expression))
+  dplyr::mutate(mean_exp = mean(gene_expression))
 
-ggplot(marker_gene_exp_df, aes(x = sum_exp, fill = filtered_cellassign_annotations)) +
+ggplot(marker_gene_exp_df, aes(x = mean_exp, fill = filtered_cellassign_annotations)) +
   geom_density() +
@@
 ```{r}
+# build the reference matrix 
 visser_ref_mtx <- build_binary_mtx(all_markers_df, 
@@
 ```
-
-Now run `CellAssign` with the full marker gene list as the reference.
 
@@
 ```{r}
+# read in predictions and get cell type assignments 
 visser_predictions <- readr::read_tsv(visser_predictions_file)
@@
 
-Below we will compare assignments between CellAssign and manual marker gene assignment.
+Below we will look at which cells are assigned to which cell type and then compare assignments between CellAssign and manual marker gene assignment.
 
@@
 
+# umap showing cell type annotations from using Visser reference 
 ggplot(celltype_df, aes(x = UMAP1, UMAP2, color = visser_cellassign_annotations)) +
@@
 ```{r}
+# get jaccard similarity index 
 jaccard_mtx <- make_jaccard_matrix(celltype_df, "visser_cellassign_annotations", "marker_gene_classification")
@@
 
+Again, we don't see much overlap of cells that we are classifying as tumor manually to the tumor cells being classified by CellAssign. 
+We do see that the majority of the immune and mesenchymal-like cells fall into the "normal" category, but this isn't the case for endothelial cells.
+
+Below, we look at the individual expression of each marker genes in the cells. 
+These plots are grouped to show all marker genes for a given cell type in a single panel.
+
 ```{r}
+# get all markers 
 all_markers <- all_markers_df |> 
@@
 
+# get expression for all markers 
 marker_gene_exp <- logcounts(sce[all_markers, ]) |>
@@
 
+# add in gene expression to cell type df for plotting
 marker_gene_exp_df <- celltype_df |> 
@@
   dplyr::left_join(all_markers_df)
+```
 
+```{r}
+# get a list of all celltypes in the reference
 celltypes <- unique(all_markers_df$cell_type)
 
+# for each cell type, plot each individual marker as a UMAP 
 celltypes |> 
-  purrr::map(\(celltype){
+  purrr::walk(\(celltype){
     plot_df <- marker_gene_exp_df |> 
@@
 
-Then we will look at the gene expression by cell type for each marker gene. 
-We would expect that the specified marker genes for each cell type are highly expressed in those cells compared to other cells.
+Interestingly, the endothelial and MSC like markers seem to be localized to a small group of cells, where as CellAssign assigns most cells to endothelial cells.
+
+Below we look at the expression of marker genes for each cell type across all cells identified to be each type by CellAssign. 
+We would expect to see expression of the marker genes for tumor cells to be higher in the cells that are identified to be tumor cells compared to other cell types.
 
 ```{r}
-transformed_exp_df <- marker_gene_exp_df |> 
-  dplyr::select(barcodes, gene_symbol, gene_expression) |> 
-  tidyr::pivot_wider(
-    names_from = "gene_symbol",
-    values_from = "gene_expression") |> 
-  dplyr::mutate(across(!barcodes, ~ scale(.x))) |> 
-  tidyr::pivot_longer(
-    cols = - barcodes,
-    names_to = "gene_symbol",
-    values_to = "transformed_gene_expression"
-  )
+grouped_celltype_df <- marker_gene_exp_df |> 
+  # group by barcode and cell type to get the sum 
+  dplyr::group_by(barcodes, cell_type) |> 
+  dplyr::mutate(mean_exp = mean(gene_expression))
 
-# add the transformed gene expression back to the main df used for plotting
-grouped_celltype_df <- marker_gene_exp_df |> 
-  dplyr::left_join(transformed_exp_df, by = c("barcodes", "gene_symbol")) |> 
-  dplyr::group_by(barcodes, cell_type) |> 
-  dplyr::mutate(sum_exp = sum(transformed_gene_expression))
-```
-
-```{r}
-ggplot(grouped_celltype_df, aes(x = sum_exp, color = cell_type)) +
+ggplot(grouped_celltype_df, aes(x = mean_exp, color = cell_type)) +
   geom_density() +
@@
 
-### Use markers from PanglaoDB
+It looks like the only real difference between groups is that the endothelial cells have higher expression of the tumor markers then other cells? 
+However the tumor cells do not show increased expression of the tumor marker genes. 
+This really is not what I would expect... maybe we need a larger reference...
+
+### Run CellAssign with markers from PanglaoDB
+
+The last thing I'm trying here is to run CellAssign using some of the same reference that we used when running CellAssign as part of scpca-nf. 
+In particular, the majority of non-muscle cells (which we believe to be tumor cells) were identified as either Endothelial cells or Fibroblasts.
+So we will create a reference that takes the markers for Endothelial cells and Fibroblasts from PanglaoDB and combines with the tumor markers.
 
 ```{r}
+# get just the genes for endothelial and fibroblasts from panglao 
 panglao_markers_df <- readr::read_tsv(panglao_markers_file) |> 
@@
 
+# get the ensembl id from rowdata for each panglao marker 
 panglao_tumor_markers_df <- rowData(sce) |> 
@@
   dplyr::select(gene_symbol, ensembl_gene_id = gene_ids, cell_type = `cell type`) |> 
+  # add in the tumor markers 
   dplyr::bind_rows(tumor_markers_df)
@@
 ```{r}
+# build the reference matrix 
 panglao_tumor_ref_mtx <- build_binary_mtx(panglao_tumor_markers_df, 
@@
 
+# remove any genes that may be present in multiple cell types 
+rowsums <- panglao_tumor_ref_mtx |> 
+  dplyr::select(-ensembl_gene_id) |> 
+  rowSums()
+
+panglao_tumor_ref_mtx <- panglao_tumor_ref_mtx[which(rowsums > 1),]
+
 # export matrix 
@@
 
-We can then read in the predictions file and find the predicted cell type for each cell, either "tumor" or "other".
-
 ```{r}
 # read in predictions file
-panglao_predictions <- readr::read_tsv(panglao_markers_mtx_file)
+panglao_predictions <- readr::read_tsv(panglao_predictions_file)
 
@@
 
+This is definitely not what I expected... 
+It looks like most cells are categorized as "other" with a few categorized as tumor. 
+I would have thought that at least some cells would be endothelial or fibroblast.
+
+```{r}
+# first rename column to combine with cell type df 
+panglao_celltype_assignments <- panglao_celltype_assignments |> 
+  dplyr::rename(panglao_cellassign_annotations = celltype)
+
+celltype_df <- celltype_df |> 
+  dplyr::left_join(panglao_celltype_assignments, by = c("barcodes" = "barcode"))
+
+# umap showing cell type annotations from using Visser reference 
+ggplot(celltype_df, aes(x = UMAP1, UMAP2, color = panglao_cellassign_annotations)) +
+  geom_point(size = 0.5, alpha = 0.5) + 
+  theme_bw()
+
+```
+
+Again, it looks like cells that were classified as normal are considered tumor here. 
+I think if we want to use CellAssign, we will need to spend more time carefully curating a marker gene list that includes all possible cell types for these samples. 
+Another alternative is to use the CellAssign annotations that we already have to help guide which cells may be tumor or normal. 
+For example, we expect that muscle cells will be tumor cells.
+
 ## Conclusions
@@
 This is opposite of the assignment we obtained using manual annotation. 
-
-```{r}
-
-```
+- Using `CellAssign` with an expanded list of marker genes from Visser et al., identifies all cell types, but the majority are endothelial cells. 
+This doesn't match up with what's expected based on marker gene expression in the sample. 
+- Using `CellAssign` with the Panglao markers and tumor markers identifies mostly normal cells and very few tumor cells. 
+Again, the cells labeled as tumor cells are opposite of the manual annotation.
 

### 2024-05-01 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd
render notebook

@@
 
-panglao_tumor_ref_mtx <- panglao_tumor_ref_mtx[which(rowsums > 1),]
+panglao_tumor_ref_mtx <- panglao_tumor_ref_mtx[-which(rowsums > 1),]
 
@@
 This is definitely not what I expected... 
-It looks like most cells are categorized as "other" with a few categorized as tumor. 
+It looks like most cells are categorized as "other" with a few categorized as tumor and only 1 endothlial cell. 
 I would have thought that at least some cells would be endothelial or fibroblast.
@@
 Again, it looks like cells that were classified as normal are considered tumor here. 
-I think if we want to use CellAssign, we will need to spend more time carefully curating a marker gene list that includes all possible cell types for these samples. 
-Another alternative is to use the CellAssign annotations that we already have to help guide which cells may be tumor or normal. 
+I think if we want to use `CellAssign`, we will need to spend more time carefully curating a marker gene list that includes all possible cell types for these samples. 
+Another alternative is to use the `CellAssign` annotations that we already have to help guide which cells may be tumor or normal. 
 For example, we expect that muscle cells will be tumor cells.

### 2024-05-02 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd
formatting fixes

@@
 We use `CellAssign` with a variety of references:
+
 - A full list of tumor marker genes only. 
@@
 ```{r packages}
-suppressMessages(
+suppressPackageStartupMessages({
   # load required packages
-  library(SingleCellExperiment),
+  library(SingleCellExperiment)
   library(ggplot2)
-)
+})
 ```
@@
 ```
+
 As expected, it looks like CellAssign is assigning very few cells to tumor that overlap with the tumor cells annotated with marker gene analysis. 
@@
 celltypes |> 
-  purrr::walk(\(celltype){
+  purrr::map(\(celltype){
     plot_df <- marker_gene_exp_df |> 

### 2024-05-03 analyses/ewings-cell-type/03-cnv-inference.Rmd
add initial notebook

---
title: "Copy number inference with Ewing sarcoma samples"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

## Introduction

## Setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
  library(copykat)
})
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current")
sample_dir <- file.path(data_dir, "SCPCP000015", params$sample_id)

# The path to this module
module_base <- file.path(repository_base, "analyses", "ewings-cell-type")

# source in helper functions
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)
```

```{r paths}
# Input files
sce_filename <- glue::glue("{params$library_id}_processed.rds")
sce_file <- file.path(sample_dir, sce_filename)

# tumor/normal classifications 
classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
classifications_file <- file.path(module_base, "results", "marker_gene_analysis", classifications_filename)

# output files 
```

```{r}
# read in processed sce
sce <- readr::read_rds(sce_file)
```

## Analysis content

### CopyKAT

```{r}
ck_results <- copykat::copykat(rawmat = as.matrix(counts(sce)),
                               id.type = "E", 
                               n.cores = 4)
```

Save these results so we don't need to re-run! 
```{r}
table(ck_results$prediction$copykat.pred)
# temp save ck_results 
readr::write_rds(ck_results, "results/SCPCL000490_copykat_results.rds")
```

```{r}
umap_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
        starts_with("UMAP")
      )

ck_preds <- ck_results$prediction |> 
  dplyr::rename(
    barcodes = cell.names
  )

cnv_df <- umap_df |> 
  dplyr::left_join(ck_preds)
```

```{r}
ggplot(cnv_df, aes(x = UMAP1, y = UMAP2, color = copykat.pred)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw()
```

```{r}
# read in tumor normal classifications 
manual_classifications_df <- readr::read_tsv(classifications_file)

cnv_df <- cnv_df |> 
  dplyr::left_join(manual_classifications_df)
```

```{r}
caret_df <- cnv_df |> 
  dplyr::filter(copykat.pred != "not.defined") |> 
  dplyr::mutate(copykat = ifelse(
    # use str_detect for test data
    copykat.pred == "diploid", "Normal", "Tumor"
  ))

caret::confusionMatrix(table(caret_df$marker_gene_classification, 
      caret_df$copykat))
```

Look at how many cells are aneuploid vs. diploid
Compare tumor vs. normal from manual to copykat using confusion matrix

Try and repeat but with providing a baseline reference

### 2024-05-03 analyses/ewings-cell-type/01-marker-gene-tumor-classification.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
   theme_bw() +
-  facet_wrap(vars(gene_symbol))
+  facet_wrap(vars(gene_symbol)) +
+  theme(legend.position = "none")
 ```
@@
 
-```{r}
-# get the z-scores for all genes, transforming across each column that contains expression for a given gene 
-transformed_exp_df <- umap_df |> 
-  dplyr::select(barcodes, gene_symbol, gene_expression) |> 
-  tidyr::pivot_wider(
-    names_from = "gene_symbol",
-    values_from = "gene_expression") |> 
-  dplyr::mutate(across(!barcodes, ~ scale(.x))) |> 
-  tidyr::pivot_longer(
-    cols = - barcodes,
-    names_to = "gene_symbol",
-    values_to = "transformed_gene_expression"
-  )
-
-# add the transformed gene expression back to the main df used for plotting
 umap_df <- umap_df |> 
-  dplyr::left_join(transformed_exp_df, by = c("barcodes", "gene_symbol"))
-```
+  dplyr::group_by(gene_symbol) |>
+  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1] |>
+  dplyr::ungroup()
 

### 2024-05-03 analyses/ewings-cell-type/01-marker-gene-tumor-classification.Rmd

@@
 - We look at marker genes for these tumor cells using `scran::findMarkers()` and compare to the tumor markers.
-- Marker gene expression is examined across clusters and tumor and normal cells are classified using clusters rather than individual cells. 
+- Marker gene expression is examined across tumor and normal cells. 
 - Finally, we compare the cells classified as tumor/normal manually to assignments from `SingleR` and `CellAssign`.
@@
 ```{r packages}
-suppressMessages(
+suppressPackageStartupMessages({
   # load required packages
-  library(SingleCellExperiment),
+  library(SingleCellExperiment)
   library(ggplot2)
+})
+
+# Set default ggplot theme
+theme_set(
+  theme_bw()
 )
@@
 module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
+```
 
-# source in helper functions
+```{r}
+# source in helper functions: make_jaccard_matrix and jaccard 
 jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
@@
     ) +
-    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
-  theme_bw()
+    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) 
 ```
@@
 
+We can also look at the distributions for each marker gene. 
+I would expect to see some sort of bimodal distribution separating cells that do and do not have expression of the marker gene.
+
 ```{r}
@@
   geom_density() +
-  theme_bw() +
   facet_wrap(vars(gene_symbol)) +
@@
 
+Although we see some slight semblance of a bimodal distribution for most marker genes, it is hard to see and we cannot directly compare gene expression values across genes. 
+This would make it hard to identify a cut off to categorize cells as expression or not expressing the marker gene.
+
 Now we will transform each of the gene expression vectors by generating z-scores. 
@@
 
+```{r}
 umap_df <- umap_df |> 
   dplyr::group_by(gene_symbol) |>
-  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1] |>
+  # get z-scores for each gene
+  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |>
   dplyr::ungroup()
+```
 
@@
   geom_point(size = 0.5, alpha = 0.5) +
-  scale_color_viridis_c() +
-  theme_bw()
+  scale_color_viridis_c()
 ```
@@
 ggplot(marker_sum_exp, aes(x = UMAP1, y = UMAP2, color = sum_classification)) +
-  geom_point(size = 0.5, alpha = 0.5) +
-  theme_bw()
+  geom_point(size = 0.5, alpha = 0.5)
 ```
@@
   geom_density() +
-  facet_wrap(vars(gene_symbol)) +
-  theme_bw()
+  facet_wrap(vars(gene_symbol)) 
 ```
@@
   tibble::rownames_to_column("ensembl_gene_id") |> 
+  # join to get gene symbols to print out later
   dplyr::left_join(marker_genes_df) |>
@@
 
-### Using clusters and marker gene expression to annotate tumor cells
+### Looking at marker genes across tumor cells
 
-We anticipate that the tumor cells and normal cells would cluster separately. 
-First, let's just plot the cluster assignments alone.
+We would expect that cells we are labeling as tumor cells have expression of the tumor marker genes. 
+Cells labeled as normal should not have expression of the markers.
 
-These are clusters calculated in `scpca-nf` and its probable that these aren't the best clusters to use. 
-Separately, we will evaluate clustering for these samples, but for now we will look at the clusters we have.
-
-```{r}
-# plot clusters on UMAP
-ggplot(marker_sum_exp, aes(x = UMAP1, y = UMAP2, color = cluster)) +
-  geom_point(size = 0.5, alpha = 0.5) +
-  theme_bw()
-```
-
-It looks like our large group of cells on the bottom right is comprised of a few clusters while most other groups of cells are a single cluster.
-
-We can then look at marker gene expression across clusters.
-Below we visualize the total number of cells in each cluster that are classified as tumor or normal using the sum of marker gene expression. 
+Below we visualize the total number of cells that are classified as tumor or normal using the sum of marker gene expression. 
 Then for each set of cells, how many of those actually express each marker gene.
 
-We expect to see two things:
-
-- Tumor cells and normal cells tend to separate by cluster. 
-- Good marker genes will be expressed in tumor cells but not normal cells.
+We expect that good marker genes will be expressed in tumor cells but not normal cells.
 
 ```{r, fig.height=10, fig.width=10}
-cluster_plot_df <- celltype_df |> 
+marker_plot_df <- celltype_df |> 
   dplyr::mutate(gene_expresed = dplyr::if_else(transformed_gene_expression > 0, TRUE, FALSE)) |> 
-  dplyr::select(gene_symbol, gene_expresed, sum_classification, cluster)
+  dplyr::select(gene_symbol, gene_expresed, sum_classification)
 
-ggplot(cluster_plot_df, aes(x = sum_classification, fill = gene_expresed)) +
+ggplot(marker_plot_df, aes(x = sum_classification, fill = gene_expresed)) +
   geom_bar(stat = "count") +
-  facet_grid(rows = vars(gene_symbol),
-             cols = vars(cluster)) +
+  facet_grid(rows = vars(gene_symbol)) +
   scale_fill_manual(values = c("grey", "blue")) +
@@
 
-Looking at this, most clusters are made up mostly of either normal or tumor. 
-Also, genes like ADGRG2 and HES1 do not seem specific to tumor cells. There are normal and tumor cells that express the gene and not very many tumor cells actually have expression of ADGRG2. 
+Looking at this might help give us an idea of genes that are more specifically expressed in tumor cells. 
+Genes like ADGRG2 and HES1 do not seem specific to tumor cells. 
+There are normal and tumor cells that express the gene and not very many tumor cells actually have expression of ADGRG2. 
 STEAP1 and NROB1 although not expressed in very many tumor cells, look to be specific to tumor cells which is promising.
 
-CD99 and CCND1 are expressed in a high proportion of tumor cells but are also expressed in quite a few groups of normal cells. 
+CD99, CAV1, and CCND1 are expressed in a high proportion of tumor cells but are also expressed in quite a few normal cells. 
 This is consistent with the literature for CD99 that it is expressed in normal cells, just at a lower level than in Ewing's cells.
 
+KDSR, MAPT, NKX2-2, PAPPA, and PRKCB all seem to have a high proportion of tumor cells expressing the gene and a low proportion of normal cells that express the gene. 
+PRKCB and MAPT are also 2 of the top genes identified using the marker analysis.
+
 Almost all genes have at least some expression in normal cells. 
-Perhaps those cells should also be categorized as tumor cells, or they do have some baseline expression in normal cells which is possible.
-
-Now we can classify the cells based on clustering and marker gene expression.
-
-```{r}
-# plot marker gene expression across all clusters 
-ggplot(marker_sum_exp, aes(x = sum_exp, fill = cluster)) +
-  geom_density() +
-  facet_grid(rows = vars(cluster)) +
-  theme_bw()
-```
-
-Looking at this there are definitely clusters with higher sum expression of the marker genes than other cells. 
-If the cluster has a median expression > 0 then all cells in that cluster will get assigned as tumor cells.
-
-```{r}
-# classify clusters as tumor/normal 
-cluster_classification_df <- marker_sum_exp |> 
-  dplyr::group_by(cluster) |> 
-  # calculate the median mean gene expression across all cells in a cluster
-  dplyr::summarise(median_cluster_exp = median(sum_exp)) |> 
-  # use 0.75 as the cutoff for tumor/normal 
-  dplyr::mutate(cluster_classification = dplyr::if_else(median_cluster_exp > 0, "Tumor", "Normal"))
-
-# get individual cell classifications
-celltype_classification_df <- marker_sum_exp |> 
-  dplyr::left_join(cluster_classification_df)
-
-# umap showing tumor and normal cells 
-ggplot(celltype_classification_df, aes(x = UMAP1, y = UMAP2, color = cluster_classification)) +
-  geom_point(size = 0.5, alpha = 0.5) +
-  theme_bw()
-```
+Perhaps those cells should also be categorized as tumor cells, or they just have some baseline expression in normal cells which is possible.
 
@@
 Looking at the heatmap in the supplemental report, the endothelial cells called by SingleR and CellAssign have a lot of overlap. 
-Let's look at expression of marker genes in these endothlial cells compared to cells that are categorized as smooth muscle cells.
+Let's look at expression of marker genes in these endothelial cells compared to cells that are categorized as smooth muscle cells.
 
@@
 - Using a z-score cut-off identifies a group of cells that appear to be tumor.
-- This mostly lines up with the default clustering that has been performed, where tumor and normal cells have been separated.
+.
 - The top marker genes for the "tumor" cells contain PRKCB and MAPT, two of the marker genes in our gene list. 
@@
 # get a dataframe with barcodes, sum_classification and cluster_classification 
-celltype_classification_df <- celltype_classification_df |> 
+celltypes_df <- celltypes_df |> 
   dplyr::select(
     barcodes,
-    marker_gene_classification = sum_classification, 
-    cluster_classification
+    marker_gene_classification = sum_classification
   )
 
-readr::write_tsv(celltype_classification_df, output_classifications_file)
+readr::write_tsv(celltypes_df, output_classifications_file)
 ```

### 2024-05-03 analyses/ewings-cell-type/03-cnv-inference.Rmd
compare ref to no ref

@@
 module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
+```
 
-# source in helper functions
+```{r}
+# source in helper functions for make_jaccard_matrix() and jaccard()
 jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
@@
 
-# output files 
+# copykat results 
+results_dir <- file.path(module_base, "results", "copykat", params$library_id)
+results_filenames <- c(
+  no_ref = glue::glue("{params$library_id}_no_normal_copykat_prediction.txt"), 
+  with_ref = glue::glue("{params$library_id}_with_normal_copykat_prediction.txt")
+) 
+ck_results_files <- file.path(results_dir, results_filenames) |> 
+  purrr::set_names(names(results_filenames))
 ```
@@
 sce <- readr::read_rds(sce_file)
+
+# read in tumor normal classifications 
+manual_classifications_df <- readr::read_tsv(classifications_file)
+
+# read in ck results 
+ck_results_df <- ck_results_files |> 
+  purrr::map(readr::read_tsv) |> 
+  dplyr::bind_rows(.id = "reference_used")
 ```
 
-## Analysis content
-
-### CopyKAT
-
-```{r}
-ck_results <- copykat::copykat(rawmat = as.matrix(counts(sce)),
-                               id.type = "E", 
-                               n.cores = 4)
-```
-
-Save these results so we don't need to re-run! 
-```{r}
-table(ck_results$prediction$copykat.pred)
-# temp save ck_results 
-readr::write_rds(ck_results, "results/SCPCL000490_copykat_results.rds")
-```
+## CopyKAT results
 
@@
   dplyr::rename_with(
-        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
-        starts_with("UMAP")
-      )
-
-ck_preds <- ck_results$prediction |> 
-  dplyr::rename(
-    barcodes = cell.names
+        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
       )
@@
 cnv_df <- umap_df |> 
-  dplyr::left_join(ck_preds)
+  # first add manual annotations
+  dplyr::left_join(manual_classifications_df) |> 
+  # now add copykat results
+  dplyr::left_join(ck_results_df, by = c("barcodes" = "cell.names"))
 ```
@@
   geom_point(alpha = 0.5, size = 0.5) +
-  theme_bw()
-```
-
-```{r}
-# read in tumor normal classifications 
-manual_classifications_df <- readr::read_tsv(classifications_file)
-
-cnv_df <- cnv_df |> 
-  dplyr::left_join(manual_classifications_df)
+  theme_bw() +
+  facet_wrap(vars(reference_used))
 ```
@@
     copykat.pred == "diploid", "Normal", "Tumor"
-  ))
+  )) |> 
+  split(cnv_df$reference_used)
 
-caret::confusionMatrix(table(caret_df$marker_gene_classification, 
-      caret_df$copykat))
+caret_df |> 
+  purrr::imap(\(df, ref_type){
+    glue::glue("\n\n================== {ref_type} ====================")
+    caret::confusionMatrix(
+      table(
+        df$marker_gene_classification, 
+        df$copykat)
+    )
+  })
 ```
 
-Look at how many cells are aneuploid vs. diploid
-Compare tumor vs. normal from manual to copykat using confusion matrix
+```{r}
+# calculate Jaccard similarity index for each reference type
+ref_types <- c("no_ref", "with_ref")
+jaccard_matrices <- ref_types |>
+  purrr::map(\(ref_type) {
+    jaccard_df <- cnv_df |> 
+      dplyr::filter(reference_used == ref_type)
 
-Try and repeat but with providing a baseline reference
+    make_jaccard_matrix(
+      cnv_df,
+      "marker_gene_classification",
+      "copykat.pred"
+    )
+  }) |> 
+  purrr::set_names(ref_types)
+```
+
+```{r}
+# Set heatmap padding option
+heatmap_padding <- 0.2
+ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))
+
+# heatmaps comparing tumor/normal annotations manually vs. copyKAT
+heatmap <- jaccard_matrices |>
+  purrr::imap(
+    \(jaccard_mtx, ref_type) {
+      ComplexHeatmap::Heatmap(
+        t(jaccard_mtx), # transpose because matrix rows are in common & we want a vertical arrangement
+        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
+        border = TRUE,
+        ## Row parameters
+        cluster_rows = TRUE,
+        row_title = ref_type,
+        row_title_gp = grid::gpar(fontsize = 12),
+        row_title_side = "left",
+        row_names_side = "left",
+        row_dend_side = "right",
+        row_names_gp = grid::gpar(fontsize = 10),
+        ## Column parameters
+        cluster_columns = FALSE,
+        column_title = "",
+        column_title_gp = grid::gpar(fontsize = 12),
+        column_names_side = "bottom",
+        column_names_gp = grid::gpar(fontsize = 10),
+        column_names_rot = 90,
+        ## Legend parameters
+        heatmap_legend_param = list(
+          title = "Jaccard index",
+          direction = "vertical",
+          legend_width = unit(1.5, "in")
+        ),
+        show_heatmap_legend = ref_type == "no_ref",
+      )
+    }) |>
+  # concatenate vertically into HeatmapList object
+  purrr::reduce(ComplexHeatmap::`%v%`) |>
+  ComplexHeatmap::draw(
+    heatmap_legend_side = "right",
+    # add a margin to the heatmap so labels don't get cut off
+    padding = unit(c(2, 20, 2, 2), "mm")
+  )
+```
+
+Now compare to CellAssign and SingleR annotations.
+
+```{r}
+celltype_columns <- c(
+  "singler_celltype_annotation",
+  "cellassign_celltype_annotation"
+)
+
+# create jaccard matrices for SingleR and CellAssign compared to tumor/normal 
+jaccard_matrices <- celltype_columns |>
+  purrr::map(\(name) {
+    make_jaccard_matrix(
+      cnv_df,
+      "copykat.pred",
+      name
+    )
+  }) |> 
+  purrr::set_names("SingleR", "CellAssign")
+```
+
+```{r, fig.height=10, fig.width=7}
+# Set heatmap padding option
+heatmap_padding <- 0.2
+ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))
+
+# list of heatmaps looking at SingleR/ CellAssign vs tumor/normal 
+heatmap <- jaccard_matrices |>
+  purrr::imap(
+    \(celltype_mat, celltype_method) {
+      ComplexHeatmap::Heatmap(
+        t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
+        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
+        border = TRUE,
+        ## Row parameters
+        cluster_rows = TRUE,
+        row_title = celltype_method,
+        row_title_gp = grid::gpar(fontsize = 12),
+        row_title_side = "left",
+        row_names_side = "left",
+        row_dend_side = "right",
+        row_names_gp = grid::gpar(fontsize = 10),
+        ## Column parameters
+        cluster_columns = FALSE,
+        column_title = "",
+        column_title_gp = grid::gpar(fontsize = 12),
+        column_names_side = "bottom",
+        column_names_gp = grid::gpar(fontsize = 10),
+        column_names_rot = 90,
+        ## Legend parameters
+        heatmap_legend_param = list(
+          title = "Jaccard index",
+          direction = "vertical",
+          legend_width = unit(1.5, "in")
+        ),
+        show_heatmap_legend = celltype_method == "SingleR",
+      )
+    }) |>
+  # concatenate vertically into HeatmapList object
+  purrr::reduce(ComplexHeatmap::`%v%`) |>
+  ComplexHeatmap::draw(
+    heatmap_legend_side = "right",
+    # add a margin to the heatmap so labels don't get cut off
+    padding = unit(c(2, 20, 2, 2), "mm")
+  )
+```
+
+## Session Info
+
+```{r session info}
+# record the versions of the packages used in this analysis and other environment information
+sessionInfo()
+```

### 2024-05-06 analyses/ewings-cell-type/03-copykat.Rmd
rename notebook and add comments

--- 2024-05-03 analyses/ewings-cell-type/03-cnv-inference.Rmd
@@
 ---
-title: "Copy number inference with Ewing sarcoma samples"
+title: "Using CopyKAT to idenfity tumor cells in Ewing sarcoma samples"
 author: Ally Hawkins
@@
 ## Introduction
+
+This notebook explores using [`CopyKAT`](https://github.com/navinlabcode/copykat) to estimate tumor and normal cells in `r params$sample_id` from SCPCP000015.
+
+`CopyKAT` was run using the `run-copykat.R` script and specifying endothelial cells from both `SingleR` and `CellAssign` as the normal reference. 
+This produced results from `CopyKAT` both with and without a normal reference. 
+These results are read into this notebook and used to:
+
+- Visualize diploid and anueploid cells on the UMAP. 
+- Calculate the confusion matrix comparing manual annotations of tumor cells to using `CopyKAT` to annotate tumor cells. 
+- Compare the annotations from `CopyKAT` to cell type annotations from `SingleR` and `CellAssign`.
 
@@
 
-# read in ck results 
+# read in ck results from both reference types (no_normal and with_normal)
 ck_results_df <- ck_results_files |> 
@@
 ## CopyKAT results
+
+### UMAP
+
+Below we prepare and plot a UMAP that shows which cells are classified as diploid, aneuploid, and not defined by `CopyKAT`. 
+We show a side by side UMAP with results from running `CopyKAT` both with and without a reference of normal cells.
 
@@
 
+If we look at the UMAPs shown in `01-marker-gene-tumor-classification.Rmd`, the cells that are labeled aneuploid here are consistent with where we saw marker gene expession for tumor cells. 
+It also looks like there is not a visible difference between using a reference of normal cells and not using a reference.
+
+### Confusion Matrix
+
+Below we directly compare the annotations obtained using manual classification of tumor and normal cells to annotating cells with `CopyKAT`. 
+To do this, we will calculate the confusion matrix using `caret::confusionMatrix()`.
+
 ```{r}
@@
   )) |> 
+  # make tumor the positive class
+  dplyr::mutate(
+    copykat = forcats::fct_relevel(copykat, "Tumor"),
+    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
+  ) |> 
   split(cnv_df$reference_used)
@@
 ```
+
+Looking at these results, we see that cells classified as tumor cells using marker genes are mostly overlapping with cells that are aneuploid in CopyKAT, which is what we expect! 
+We also see that the Kappa values range are close to 1, indicating almost perfect agreement.
+
+It also looks like there isn't a difference between using the reference and not using the reference. 
+The reference Kappa is .1 higher, but I don't think that's enough to argue that you need to use a reference. 
+The one thing to note is that using a reference cut the run time for `CopyKAT` in half, so that is one benefit.
+
+We can also calculate the Jaccard similarity index to visualize the amount of cells that have overlapping annotations. 
+Given the confusion matrix, I would expect to see good agreement between marker gene and `CopyKAT` annotations.
 
@@
 
-Now compare to CellAssign and SingleR annotations.
+As expected, it looks like cells classified as tumor cells using marker gene annotation are classified as `aneuploid` with `CopyKAT`.
+
+### Compare CopyKAT to SingleR and CellAssign
+
+Lastly, we will compare the annotations from `CopyKAT` to those obtained using `SingleR` and `CellAssign` by calculating the Jaccard similarity index. 
+For this comparison we will use just the annotations from `CopyKAT` with no reference.
 
@@
 
-# create jaccard matrices for SingleR and CellAssign compared to tumor/normal 
+# filter to only get annotations from no ref
+no_ref_only <- cnv_df |> 
+  dplyr::filter(reference_used == "no_ref")
+
+# create jaccard matrices for SingleR and CellAssign compared to aneuploid/diploid 
 jaccard_matrices <- celltype_columns |>
@@
     make_jaccard_matrix(
-      cnv_df,
+      no_ref_only,
       "copykat.pred",
@@
 
+We see that aneuploid cells are mostly muscle cells in both `CellAssign` and `SingleR`. 
+In `SingleR` there are some other cell types that were assigned that have a few anueploid cells (chondrocytes, neurons, and fibroblasts).
+
+For diploid cells, we see the most overlap with fibroblasts, endothelial cells, and macrophages. 
+These are consistent with the cell types that we see normal cells line up with in `01-marker-gene-tumor-classification.Rmd`.
+
+## Conclusions
+
+- `CopyKAT` seems to do a good job in this Ewing sarcoma sample of identifying aneuploid and diploid cells. 
+- The annotations from `CopyKAT` mostly line up with annotations obtained using marker gene annotation. 
+- Using a reference of normal cells doesn't change the overall results but does reduce the run time for `CopyKAT`.
+
 ## Session Info

### 2024-05-07 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 
-# read in all maker genes 
+# read in all marker genes 
 all_markers_df <- readr::read_tsv(all_markers_file) |> 
@@
       names_to = "celltype",
-      values_to = "prediction"
+      values_to = "score"
     ) |>
     dplyr::group_by(barcode) |>
-    dplyr::slice_max(prediction, n = 1) |>
+    dplyr::slice_max(score, n = 1) |>
     dplyr::ungroup()
@@
   "python scripts/run-cellassign.py",
-  "--anndata_file '{anndata_file}'",
-  "--output_predictions '{cellassign_predictions_file}'",
-  "--reference '{tumor_markers_mtx_file}'"
+  " --anndata_file '{anndata_file}'",
+  " --output_predictions '{cellassign_predictions_file}'",
+  " --reference '{tumor_markers_mtx_file}'"
 )
@@
   "python scripts/run-cellassign.py",
-  "--anndata_file '{anndata_file}'",
-  "--output_predictions '{filtered_predictions_file}'",
-  "--reference '{filtered_markers_mtx_file}'"
+  " --anndata_file '{anndata_file}'",
+  " --output_predictions '{filtered_predictions_file}'",
+  " --reference '{filtered_markers_mtx_file}'"
 )
@@
 
-ggplot(celltype_df, aes(x = UMAP1, UMAP2, color = filtered_cellassign_annotations)) +
+ggplot(celltype_df, aes(x = UMAP1, y = UMAP2, color = filtered_cellassign_annotations)) +
   geom_point(size = 0.5, alpha = 0.5) + 
@@
 cellassign_call <- glue::glue(
-  "python scripts/run-cellassign.py ",
-  "--anndata_file '{anndata_file}' ",
-  "--output_predictions '{visser_predictions_file}' ",
-  "--reference '{visser_markers_mtx_file}'"
+  "python scripts/run-cellassign.py",
+  " --anndata_file '{anndata_file}'",
+  " --output_predictions '{visser_predictions_file}'",
+  " --reference '{visser_markers_mtx_file}'"
 )
@@
 cellassign_call <- glue::glue(
-  "python scripts/run-cellassign.py ",
-  "--anndata_file '{anndata_file}' ",
-  "--output_predictions '{panglao_predictions_file}' ",
-  "--reference '{panglao_markers_mtx_file}' "
+  "python scripts/run-cellassign.py",
+  " --anndata_file '{anndata_file}'",
+  " --output_predictions '{panglao_predictions_file}'",
+  " --reference '{panglao_markers_mtx_file}'"
 )

### 2024-05-07 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd
add heatmaps to notebook

@@
 
-```{r}
-# read in processed sce
-sce <- readr::read_rds(sce_file)
-
-# read in tumor marker genes table 
-tumor_markers_df <- readr::read_tsv(tumor_markers_file) |> 
-  # account for genes being from multiple sources
-  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
-  dplyr::distinct()
-
-# read in all marker genes 
-all_markers_df <- readr::read_tsv(all_markers_file) |> 
-  # account for genes being from multiple sources
-  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
-  dplyr::distinct()
-```
-
 Define some helper functions for creating a reference matrix and obtaining cell type annotations from `CellAssign` predictions.
@@
 # function for creating a binary matrix with cell types as columns and marker genes as rows 
-build_binary_mtx <- function(marker_genes_df,
+build_binary_df <- function(marker_genes_df,
                             gene_col, 
@@
 
-  binary_mtx <- marker_genes_df |>
+  binary_df <- marker_genes_df |>
     dplyr::select({{cell_type_col}}, {{gene_col}}) |> 
@@
 
-  return(binary_mtx)
+  return(binary_df)
 }
@@
 
+```{r}
+# read in processed sce
+sce <- readr::read_rds(sce_file)
+
+# read in tumor marker genes table 
+tumor_markers_df <- readr::read_tsv(tumor_markers_file) |> 
+  # account for genes being from multiple sources
+  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
+  dplyr::distinct()
+
+# read in all marker genes 
+all_markers_df <- readr::read_tsv(all_markers_file) |> 
+  # account for genes being from multiple sources
+  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
+  dplyr::distinct()
+```
+
 ## Analysis content
@@
 ```{r}
-tumor_ref_mtx <- build_binary_mtx(tumor_markers_df, 
+tumor_ref_mtx <- build_binary_df(tumor_markers_df, 
                                   gene_col = "ensembl_gene_id",
@@
 table(tumor_celltype_assignments$celltype)
+```
+
+```{r}
+# heatmap of prediction scores 
+tumor_predictions |>
+  dplyr::select(-barcode) |> 
+  as.matrix() |> 
+  pheatmap::pheatmap(show_rownames = FALSE)
 ```
@@
 
+```{r}
+# heatmap of prediction scores 
+filtered_predictions |>
+  dplyr::select(-barcode) |> 
+  as.matrix() |> 
+  pheatmap::pheatmap(show_rownames = FALSE)
+```
+
 Now we are seeing a few more tumor cells, let's see where they are on the UMAP.
@@
 # first read in tumor normal classifications 
+# obtained from manual classification in `01-marker-gene-tumor-classifications.Rmd`
 classifications_df <- readr::read_tsv(classifications_file)
@@
   dplyr::rename_with(
-        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
-        starts_with("UMAP")
+        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
       ) |> 
   # join with tumor normal
-  dplyr::left_join(classifications_df) |> 
+  dplyr::left_join(classifications_df, by = "barcodes") |> 
   # join with filtered cell type assignments from cellassign 
@@
 # build the reference matrix 
-visser_ref_mtx <- build_binary_mtx(all_markers_df, 
+visser_ref_mtx <- build_binary_df(all_markers_df, 
                                    gene_col = "ensembl_gene_id",
@@
 ```
+
+```{r}
+# heatmap of prediction scores 
+visser_predictions |>
+  dplyr::select(-barcode) |> 
+  as.matrix() |> 
+  pheatmap::pheatmap(show_rownames = FALSE)
+```
+
 Here we see that there are still only a few hundred tumor cells and most cells are endothelial cells. 
@@
 # build the reference matrix 
-panglao_tumor_ref_mtx <- build_binary_mtx(panglao_tumor_markers_df, 
+panglao_tumor_ref_mtx <- build_binary_df(panglao_tumor_markers_df, 
                                           gene_col = "ensembl_gene_id",
@@
 table(panglao_celltype_assignments$celltype)
+```
+
+```{r}
+# heatmap of prediction scores 
+panglao_predictions |>
+  dplyr::select(-barcode) |> 
+  as.matrix() |> 
+  pheatmap::pheatmap(show_rownames = FALSE)
 ```

### 2024-05-07 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd

@@
 ```{r}
+# first read in tumor normal classifications 
+# obtained from manual classification in `01-marker-gene-tumor-classifications.Rmd`
+classifications_df <- readr::read_tsv(classifications_file)
+
+# get annotations for heatmaps
+annotation_df <- data.frame(
+  marker_gene_classification = classifications_df$marker_gene_classification,
+  row.names = classifications_df$barcodes
+)
+
 # heatmap of prediction scores 
 tumor_predictions |>
-  dplyr::select(-barcode) |> 
+  tibble::column_to_rownames("barcode") |> 
   as.matrix() |> 
-  pheatmap::pheatmap(show_rownames = FALSE)
+  pheatmap::pheatmap(show_rownames = FALSE,
+                     annotation_row = annotation_df)
 ```
@@
 filtered_predictions |>
-  dplyr::select(-barcode) |> 
+  tibble::column_to_rownames("barcode") |> 
   as.matrix() |> 
-  pheatmap::pheatmap(show_rownames = FALSE)
+  pheatmap::pheatmap(show_rownames = FALSE,
+                     annotation_row = annotation_df)
 ```
@@
 ```{r}
-# first read in tumor normal classifications 
-# obtained from manual classification in `01-marker-gene-tumor-classifications.Rmd`
-classifications_df <- readr::read_tsv(classifications_file)
-
 filtered_celltype_assignments <- filtered_celltype_assignments |> 
@@
 visser_predictions |>
-  dplyr::select(-barcode) |> 
+  tibble::column_to_rownames("barcode") |> 
   as.matrix() |> 
-  pheatmap::pheatmap(show_rownames = FALSE)
+  pheatmap::pheatmap(show_rownames = FALSE,
+                     annotation_row = annotation_df)
 ```
@@
 panglao_predictions |>
-  dplyr::select(-barcode) |> 
+  tibble::column_to_rownames("barcode") |> 
   as.matrix() |> 
-  pheatmap::pheatmap(show_rownames = FALSE)
+  pheatmap::pheatmap(show_rownames = FALSE,
+                     annotation_row = annotation_df)
 ```

### 2024-05-08 analyses/ewings-cell-type/04-infercnv.Rmd
initial notebook

--- 2024-05-07 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd
@@
 ---
-title: "Explore tumor cell classification with CellAssign"
+title: "Copy number inference with InferCNV"
 author: Ally Hawkins
@@
 ## Introduction
-
-This notebook looks at classifying tumor cells with `CellAssign` in a Ewing sarcoma sample, `r {params$sample_id}`.
-
-We use `CellAssign` with a variety of references:
-
-- A full list of tumor marker genes only. 
-- A filtered list of tumor marker genes only. 
-- Tumor marker genes and non-tumor marker genes from [Visser et al.](https://doi.org/10.1158/2767-9764.CRC-23-0027) specified in the supplemental methods. 
-- Tumor marker genes and marker genes from endothelial cells and fibroblast marker genes in PanglaoDB.
-
-We then compare the results from running `CellAssign` with each reference to the tumor/normal classification obtained in `01-marker-gene-tumor-classification.Rmd`. 
-We also look at marker gene expression in the assigned cells.
 
@@
 module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
+```
 
+```{r base paths}
 # source in helper functions
@@
 
-anndata_filename <- glue::glue("{params$library_id}_processed_rna.h5ad")
-anndata_file <- file.path(sample_dir, anndata_filename)
-
 # tumor normal classifications 
@@
 
-# tumor marker genes
-tumor_markers_file <- file.path(module_base, "references", "tumor-marker-genes.tsv")
-# all markers 
-all_markers_file <- file.path(module_base, "references", "all-marker-genes.tsv")
-# panglao markers
-panglao_markers_file <- file.path(module_base, "references", "PanglaoDB_markers_2020-03-27.tsv")
-
-# Output reference matrix files
-ref_dir <- file.path(module_base, "references", "cellassign_refs")
-
-tumor_markers_mtx_file <- file.path(ref_dir, "tumor-marker-cellassign.tsv")
-filtered_markers_mtx_file <- file.path(ref_dir, "filtered-tumor-marker-cellassign.tsv")
-visser_markers_mtx_file <- file.path(ref_dir, "visser-all-marker-cellassign.tsv")
-panglao_markers_mtx_file <- file.path(ref_dir, "panglao-endo-fibro-cellassign.tsv")
-
-# output predictions from cellassign 
-preds_results_dir <- file.path(module_base, "results", "cellassign_predictions")
-fs::dir_create(preds_results_dir)
-
-cellassign_predictions_file <- file.path(preds_results_dir, "cellassign-tumor-predictions.tsv")
-filtered_predictions_file <- file.path(preds_results_dir, "filtered-cellassign-tumor-predictions.tsv")
-visser_predictions_file <- file.path(preds_results_dir, "visser-cellassign-predictions.tsv")
-panglao_predictions_file <- file.path(preds_results_dir, "panglao-endo-fibro-predictions.tsv")
+# output from running infercnv 
 ```
@@
 ```{r}
-# read in processed sce
 sce <- readr::read_rds(sce_file)
-
-# read in tumor marker genes table 
-tumor_markers_df <- readr::read_tsv(tumor_markers_file) |> 
-  # account for genes being from multiple sources
-  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
-  dplyr::distinct()
-
-# read in all marker genes 
-all_markers_df <- readr::read_tsv(all_markers_file) |> 
-  # account for genes being from multiple sources
-  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
-  dplyr::distinct()
-```
-
-Define some helper functions for creating a reference matrix and obtaining cell type annotations from `CellAssign` predictions.
-
-```{r}
-# function for creating a binary matrix with cell types as columns and marker genes as rows 
-build_binary_mtx <- function(marker_genes_df,
-                             gene_col, 
-                             cell_type_col) {
-
-  binary_mtx <- marker_genes_df |>
-  dplyr::select({{cell_type_col}}, {{gene_col}}) |> 
-  unique() |> 
-  tidyr::pivot_wider(
-    id_cols = {{gene_col}},
-    names_from = {{cell_type_col}},
-    values_from = {{cell_type_col}},
-    values_fn = length,
-    values_fill = 0
-  ) |>
-  # add a column with no marker genes
-  # cell assign will assign cells to "other" when no other cell types are appropriate
-  dplyr::mutate(other = 0)
-
-  return(binary_mtx)
-}
-
-# function to get assigned cell type based on max prediction returned by CellAssign 
-get_celltype_assignment <- function(predictions){
-  # get individual cell type assignments
-  # those with the max prediction 
-  celltype_assignments <- predictions |>
-    tidyr::pivot_longer(
-      !barcode,
-      names_to = "celltype",
-      values_to = "score"
-    ) |>
-    dplyr::group_by(barcode) |>
-    dplyr::slice_max(score, n = 1) |>
-    dplyr::ungroup()
-
-  return(celltype_assignments)
-}
-```
-
-## Analysis content
-
-### Run CellAssign with only tumor marker genes
-
-First let's see if we can use `CellAssign` to classify cells as tumor or normal using just the list of tumor marker genes.
-
-We need to create a binary matrix with two columns, "tumor" and "other". 
-The "other" will refer to any normal cells.
-
-```{r}
-tumor_ref_mtx <- build_binary_mtx(tumor_markers_df, 
-                                  gene_col = "ensembl_gene_id",
-                                  cell_type_col = "cell_type")
-
-# export matrix 
-readr::write_tsv(tumor_ref_mtx, tumor_markers_mtx_file)
-```
-
-Now run `CellAssign` with the full marker gene list as the reference.
-
-```{r, eval=FALSE}
-cellassign_call <- glue::glue(
-  "python scripts/run-cellassign.py",
-  " --anndata_file '{anndata_file}'",
-  " --output_predictions '{cellassign_predictions_file}'",
-  " --reference '{tumor_markers_mtx_file}'"
-)
-
-system(cellassign_call)
-```
-
-We can then read in the predictions file and find the predicted cell type for each cell, either "tumor" or "other".
-
-```{r}
-# read in predictions file
-tumor_predictions <- readr::read_tsv(cellassign_predictions_file)
-
-# get cell type assignments 
-tumor_celltype_assignments <- get_celltype_assignment(tumor_predictions)
-
-# create a table of assignments 
-table(tumor_celltype_assignments$celltype)
-```
-
-Looking at this, it appears that `CellAssign` is classifying all cells as "other"? 
-I find this hard to believe since this sample was taken directly from a Ewing sarcoma tumor, so there are bound to be tumor cells.
-
-### Run CellAssign with a filtered set of tumor marker genes
-
-In looking at the UMAPs of all marker genes (see `01-marker-gene-tumor-classification.Rmd`) there are quite a few genes that have very low expression across all cells, so perhaps they are adding noise to the model.
-Although this is a bit circular, we could restrict our list of marker genes to be only those that have a high mean gene expression in our dataset? 
-I would expect that when we do that, we are able to identify more tumor cells that line up with what we would expect based on gene expression patterns.
-
-```{r}
-# grab the mean gene expression for all marker genes 
-rowdata_df <- rowData(sce) |> 
-  as.data.frame() |> 
-  dplyr::filter(gene_symbol %in% tumor_markers_df$gene_symbol)
-rowdata_df
 ```
@@
 ```{r}
-# what's the mean gene expression of all genes in this sample for reference
-mean(rowData(sce)$mean)
+# create the infercnv object
+infercnv_obj <- infercnv::CreateInfercnvObject(raw_counts_matrix=counts(sce),
+                                               annotations_file=annotations_file,
+                                               delim="\t",
+                                               gene_order_file=gene_order_file,
+                                               ref_group_names=c("endothelial"))
+
 ```
-
-There are a few genes that have low gene expression < 1. 
-Let's try and filter out those genes and then use `CellAssign`.
-
 ```{r}
-# get list of marker genes to keep
-filtered_markers <- rowdata_df |> 
-  dplyr::filter(mean > 1) |> 
-  dplyr::pull(gene_ids)
-
-# filter binary mtx 
-filtered_ref_mtx <- tumor_ref_mtx |> 
-  dplyr::filter(ensembl_gene_id %in% filtered_markers)
-
-readr::write_tsv(filtered_ref_mtx, filtered_markers_mtx_file)
-```
-
-```{r, eval=FALSE}
-cellassign_call <- glue::glue(
-  "python scripts/run-cellassign.py",
-  " --anndata_file '{anndata_file}'",
-  " --output_predictions '{filtered_predictions_file}'",
-  " --reference '{filtered_markers_mtx_file}'"
-)
-
-system(cellassign_call)
-```
-
-Now we can read in the predictions for the filtered marker gene list.
-
-```{r}
-filtered_predictions <- readr::read_tsv(filtered_predictions_file)
-
-filtered_celltype_assignments <- get_celltype_assignment(filtered_predictions)
-
-table(filtered_celltype_assignments$celltype)
-```
-
-Now we are seeing a few more tumor cells, let's see where they are on the UMAP.
-
-```{r}
-# first read in tumor normal classifications 
-classifications_df <- readr::read_tsv(classifications_file)
-
-filtered_celltype_assignments <- filtered_celltype_assignments |> 
-  dplyr::rename(filtered_cellassign_annotations = celltype)
-
-celltype_df <- sce |> 
-  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
-  # replace UMAP.1 with UMAP1
-  dplyr::rename_with(
-        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
-        starts_with("UMAP")
-      ) |> 
-  # join with tumor normal
-  dplyr::left_join(classifications_df) |> 
-  # join with filtered cell type assignments from cellassign 
-  dplyr::left_join(filtered_celltype_assignments, by = c("barcodes" = "barcode"))
-
-ggplot(celltype_df, aes(x = UMAP1, y = UMAP2, color = filtered_cellassign_annotations)) +
-  geom_point(size = 0.5, alpha = 0.5) + 
-  theme_bw()
-```
-
-Not quite where I would expect to see tumor cells. 
-Let's see how many tumor cells overlap with our manual classifications using the Jaccard similarity index.
-
-```{r}
-jaccard_mtx <- make_jaccard_matrix(celltype_df, "filtered_cellassign_annotations", "marker_gene_classification")
-
-ComplexHeatmap::Heatmap(
-  t(jaccard_mtx), # transpose because matrix rows are in common & we want a vertical arrangement
-  col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
-  border = TRUE,
-  ## Row parameters
-  cluster_rows = FALSE,
-  row_title = "Marker gene",
-  row_title_gp = grid::gpar(fontsize = 12),
-  row_title_side = "left",
-  row_names_side = "left",
-  row_dend_side = "right",
-  row_names_gp = grid::gpar(fontsize = 10),
-  ## Column parameters
-  cluster_columns = FALSE,
-  column_title = "CellAssign",
-  column_title_gp = grid::gpar(fontsize = 12),
-  column_names_side = "bottom",
-  column_names_gp = grid::gpar(fontsize = 10),
-  column_names_rot = 90,
-  ## Legend parameters
-  heatmap_legend_param = list(
-    title = "Jaccard index",
-    direction = "vertical",
-    legend_width = unit(1.5, "in")
-  ),
-  show_heatmap_legend = TRUE,
+# perform infercnv operations to reveal cnv signal
+infercnv_obj <- infercnv::run(infercnv_obj,
+                              cutoff=0.1,  # use 1 for smart-seq, 0.1 for 10x-genomics
+                              out_dir=infer_results_dir,  # dir is auto-created for storing outputs
+                              denoise=T,
+                              HMM=T ,
+                              report_by = "cell"
 )
@@
 
-As expected, it looks like CellAssign is assigning very few cells to tumor that overlap with the tumor cells annotated with marker gene analysis. 
-It looks like most tumor cells are being classified as normal with the marker gene analysis.
+```{r}
+infercnv_obj@observation_grouped_cell_indices
 
-Let's look at the mean expression of marker genes between tumor and other cells as classified by `CellAssign`.
-
-```{r}
-# get marker gene expression 
-marker_gene_exp <- logcounts(sce[filtered_markers, ]) |>
-  as.matrix() |>
-  t() |>
-  as.data.frame() |>
-  tibble::rownames_to_column("barcodes")
-
-marker_gene_exp_df <- celltype_df |> 
-  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
-  tidyr::pivot_longer(
-    cols = starts_with("ENSG"),
-    names_to = "ensembl_gene_id",
-    values_to = "gene_expression"
-  ) |> 
-  dplyr::group_by(barcodes) |> 
-  dplyr::mutate(mean_exp = mean(gene_expression))
-
-ggplot(marker_gene_exp_df, aes(x = mean_exp, fill = filtered_cellassign_annotations)) +
-  geom_density() +
-  facet_grid(rows = vars(filtered_cellassign_annotations)) +
-  theme_bw()
+infercnv_obj@expr.data[1:5, 1:5]
 ```
-
-Unlike what I would expect, the "other" cells appear to have higher expression of marker genes? 
-Maybe using `CellAssign` with only two options and a small set of marker genes doesn't work in the way we would expect.
-
-### Run CellAssign with tumor markers and all marker genes from Visser et al.
-
-Perhaps we need to add a few more cell types, rather than just have the two options. 
-Below we will use CellAssign but with all tumor marker genes and all marker genes in Visser et al. for non-tumor cells.
-
-```{r}
-# build the reference matrix 
-visser_ref_mtx <- build_binary_mtx(all_markers_df, 
-                                   gene_col = "ensembl_gene_id",
-                                   cell_type_col = "cell_type")
-
-# export matrix 
-readr::write_tsv(visser_ref_mtx, visser_markers_mtx_file)
-```
-
-```{r, eval=FALSE}
-cellassign_call <- glue::glue(
-  "python scripts/run-cellassign.py",
-  " --anndata_file '{anndata_file}'",
-  " --output_predictions '{visser_predictions_file}'",
-  " --reference '{visser_markers_mtx_file}'"
-)
-
-system(cellassign_call)
-```
-
-```{r}
-# read in predictions and get cell type assignments 
-visser_predictions <- readr::read_tsv(visser_predictions_file)
-
-visser_celltype_assignments <- get_celltype_assignment(visser_predictions)
-
-table(visser_celltype_assignments$celltype)
-```
-Here we see that there are still only a few hundred tumor cells and most cells are endothelial cells. 
-Generally, I expect that there to be a large portion of tumor cells with much fewer cells from normal cell types.
-
-Below we will look at which cells are assigned to which cell type and then compare assignments between CellAssign and manual marker gene assignment.
-
-```{r}
-# first rename column to combine with cell type df 
-visser_celltype_assignments <- visser_celltype_assignments |> 
-  dplyr::rename(visser_cellassign_annotations = celltype)
-
-celltype_df <- celltype_df |> 
-  dplyr::left_join(visser_celltype_assignments, by = c("barcodes" = "barcode"))
-
-# umap showing cell type annotations from using Visser reference 
-ggplot(celltype_df, aes(x = UMAP1, UMAP2, color = visser_cellassign_annotations)) +
-  geom_point(size = 0.5, alpha = 0.5) + 
-  theme_bw()
-```
-
-```{r}
-# get jaccard similarity index 
-jaccard_mtx <- make_jaccard_matrix(celltype_df, "visser_cellassign_annotations", "marker_gene_classification")
-
-ComplexHeatmap::Heatmap(
-  t(jaccard_mtx), # transpose because matrix rows are in common & we want a vertical arrangement
-  col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
-  border = TRUE,
-  ## Row parameters
-  cluster_rows = FALSE,
-  row_title = "Marker gene",
-  row_title_gp = grid::gpar(fontsize = 12),
-  row_title_side = "left",
-  row_names_side = "left",
-  row_dend_side = "right",
-  row_names_gp = grid::gpar(fontsize = 10),
-  ## Column parameters
-  cluster_columns = FALSE,
-  column_title = "CellAssign - Visser",
-  column_title_gp = grid::gpar(fontsize = 12),
-  column_names_side = "bottom",
-  column_names_gp = grid::gpar(fontsize = 10),
-  column_names_rot = 90,
-  ## Legend parameters
-  heatmap_legend_param = list(
-    title = "Jaccard index",
-    direction = "vertical",
-    legend_width = unit(1.5, "in")
-  ),
-  show_heatmap_legend = TRUE,
-)
-```
-
-Again, we don't see much overlap of cells that we are classifying as tumor manually to the tumor cells being classified by CellAssign. 
-We do see that the majority of the immune and mesenchymal-like cells fall into the "normal" category, but this isn't the case for endothelial cells.
-
-Below, we look at the individual expression of each marker genes in the cells. 
-These plots are grouped to show all marker genes for a given cell type in a single panel.
-
-```{r}
-# get all markers 
-all_markers <- all_markers_df |> 
-  dplyr::pull(ensembl_gene_id)
-
-# get expression for all markers 
-marker_gene_exp <- logcounts(sce[all_markers, ]) |>
-  as.matrix() |>
-  t() |>
-  as.data.frame() |>
-  tibble::rownames_to_column("barcodes")
-
-# add in gene expression to cell type df for plotting
-marker_gene_exp_df <- celltype_df |> 
-  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
-  tidyr::pivot_longer(
-    cols = starts_with("ENSG"),
-    names_to = "ensembl_gene_id",
-    values_to = "gene_expression"
-  ) |> 
-  dplyr::left_join(all_markers_df)
-```
-
-```{r}
-# get a list of all celltypes in the reference
-celltypes <- unique(all_markers_df$cell_type)
-
-# for each cell type, plot each individual marker as a UMAP 
-celltypes |> 
-  purrr::map(\(celltype){
-    plot_df <- marker_gene_exp_df |> 
-      dplyr::filter(cell_type == celltype)
-
-    # faceted umap showing a umap panel for each marker gene 
-    ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
-      geom_point(alpha = 0.1, size = 0.2) +
-      facet_wrap(vars(gene_symbol)) +
-      scale_color_viridis_c() +
-      labs(
-        title = celltype,
-        color = "Log-normalized gene expression"
-      ) +
-      # remove axis numbers and background grid
-      scale_x_continuous(labels = NULL, breaks = NULL) +
-      scale_y_continuous(labels = NULL, breaks = NULL) +
-      theme(
-        aspect.ratio = 1,
-        legend.position = "bottom",
-        axis.title = element_text(size = 9, color = "black"),
-        strip.text = element_text(size = 8),
-        legend.title = element_text(size = 9),
-        legend.text = element_text(size = 8)
-      ) +
-      guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
-      theme_bw()
-
-  })
-```
-
-Interestingly, the endothelial and MSC like markers seem to be localized to a small group of cells, where as CellAssign assigns most cells to endothelial cells.
-
-Below we look at the expression of marker genes for each cell type across all cells identified to be each type by CellAssign. 
-We would expect to see expression of the marker genes for tumor cells to be higher in the cells that are identified to be tumor cells compared to other cell types.
-
-```{r}
-grouped_celltype_df <- marker_gene_exp_df |> 
-  # group by barcode and cell type to get the sum 
-  dplyr::group_by(barcodes, cell_type) |> 
-  dplyr::mutate(mean_exp = mean(gene_expression))
-
-ggplot(grouped_celltype_df, aes(x = mean_exp, color = cell_type)) +
-  geom_density() +
-  facet_wrap(vars(visser_cellassign_annotations)) +
-  theme_bw() +
-  labs(
-    color = "Marker gene group"
-  )
-```
-
-It looks like the only real difference between groups is that the endothelial cells have higher expression of the tumor markers then other cells? 
-However the tumor cells do not show increased expression of the tumor marker genes. 
-This really is not what I would expect... maybe we need a larger reference...
-
-### Run CellAssign with markers from PanglaoDB
-
-The last thing I'm trying here is to run CellAssign using some of the same reference that we used when running CellAssign as part of scpca-nf. 
-In particular, the majority of non-muscle cells (which we believe to be tumor cells) were identified as either Endothelial cells or Fibroblasts.
-So we will create a reference that takes the markers for Endothelial cells and Fibroblasts from PanglaoDB and combines with the tumor markers.
-
-```{r}
-# get just the genes for endothelial and fibroblasts from panglao 
-panglao_markers_df <- readr::read_tsv(panglao_markers_file) |> 
-  dplyr::filter(`cell type` %in% c("Endothelial cells", "Fibroblasts")) |> 
-  dplyr::select(`official gene symbol`, `cell type`)
-
-# get the ensembl id from rowdata for each panglao marker 
-panglao_tumor_markers_df <- rowData(sce) |> 
-  as.data.frame() |> 
-  dplyr::filter(gene_symbol %in% panglao_markers_df$`official gene symbol`) |> 
-  dplyr::left_join(panglao_markers_df, by = c("gene_symbol" = "official gene symbol")) |> 
-  dplyr::select(gene_symbol, ensembl_gene_id = gene_ids, cell_type = `cell type`) |> 
-  # add in the tumor markers 
-  dplyr::bind_rows(tumor_markers_df)
-
-```
-
-```{r}
-# build the reference matrix 
-panglao_tumor_ref_mtx <- build_binary_mtx(panglao_tumor_markers_df, 
-                                          gene_col = "ensembl_gene_id",
-                                          cell_type_col = "cell_type")
-
-# remove any genes that may be present in multiple cell types 
-rowsums <- panglao_tumor_ref_mtx |> 
-  dplyr::select(-ensembl_gene_id) |> 
-  rowSums()
-
-panglao_tumor_ref_mtx <- panglao_tumor_ref_mtx[-which(rowsums > 1),]
-
-# export matrix 
-readr::write_tsv(panglao_tumor_ref_mtx, panglao_markers_mtx_file)
-```
-
-```{r, eval=FALSE}
-cellassign_call <- glue::glue(
-  "python scripts/run-cellassign.py",
-  " --anndata_file '{anndata_file}'",
-  " --output_predictions '{panglao_predictions_file}'",
-  " --reference '{panglao_markers_mtx_file}'"
-)
-
-system(cellassign_call)
-```
-
-```{r}
-# read in predictions file
-panglao_predictions <- readr::read_tsv(panglao_predictions_file)
-
-# get cell type assignments 
-panglao_celltype_assignments <- get_celltype_assignment(panglao_predictions)
-
-# create a table of assignments 
-table(panglao_celltype_assignments$celltype)
-```
-
-This is definitely not what I expected... 
-It looks like most cells are categorized as "other" with a few categorized as tumor and only 1 endothlial cell. 
-I would have thought that at least some cells would be endothelial or fibroblast.
-
-```{r}
-# first rename column to combine with cell type df 
-panglao_celltype_assignments <- panglao_celltype_assignments |> 
-  dplyr::rename(panglao_cellassign_annotations = celltype)
-
-celltype_df <- celltype_df |> 
-  dplyr::left_join(panglao_celltype_assignments, by = c("barcodes" = "barcode"))
-
-# umap showing cell type annotations from using Visser reference 
-ggplot(celltype_df, aes(x = UMAP1, UMAP2, color = panglao_cellassign_annotations)) +
-  geom_point(size = 0.5, alpha = 0.5) + 
-  theme_bw()
-
-```
-
-Again, it looks like cells that were classified as normal are considered tumor here. 
-I think if we want to use `CellAssign`, we will need to spend more time carefully curating a marker gene list that includes all possible cell types for these samples. 
-Another alternative is to use the `CellAssign` annotations that we already have to help guide which cells may be tumor or normal. 
-For example, we expect that muscle cells will be tumor cells.
-
-## Conclusions
-
-- Using `CellAssign` with the same list of marker genes doesn't identify any tumor cells. 
-- Using `CellAssign` with a restricted list of marker genes identfies tumor cells that have lower marker gene expression than other cells. 
-This is opposite of the assignment we obtained using manual annotation. 
-- Using `CellAssign` with an expanded list of marker genes from Visser et al., identifies all cell types, but the majority are endothelial cells. 
-This doesn't match up with what's expected based on marker gene expression in the sample. 
-- Using `CellAssign` with the Panglao markers and tumor markers identifies mostly normal cells and very few tumor cells. 
-Again, the cells labeled as tumor cells are opposite of the manual annotation.
-
-## Session Info
-
-```{r session info}
-# record the versions of the packages used in this analysis and other environment information
-sessionInfo()
-```

### 2024-05-09 analyses/ewings-cell-type/04-infercnv.Rmd
look at cnvs for each cell

@@
 
+Here we look at using [`InferCNV`](https://github.com/broadinstitute/infercnv) to identify tumor and normal cells in `r params$sample_id`.
+
+Before running this notebook, `InferCNV` is run using the `run-infercnv.R` script. 
+This script runs `InferCNV` once with Endothelial cells as the reference cells and once without any cells as the reference.
+
+Here, we look the copy number variations found reported by `InferCNV` and use that copy number information to call cells as tumor or normal cells. 
+We then compare these classifications to manual classifications and annotations from `SingleR` and `CellAssign`.
+
 ## Setup
@@
   library(ggplot2)
+  library(dplyr)
 })
@@
 
-```{r base paths}
+```{r}
 # source in helper functions
@@
 # output from running infercnv 
+infer_cnv_dir <- file.path(module_base, "results", "infercnv", "SCPCL000822")
+
+# add code to read in both ref and no ref? 
 ```
@@
 ```{r}
+# read in sce object
 sce <- readr::read_rds(sce_file)
+
+# read in tumor normal classifications 
+manual_classifications_df <- readr::read_tsv(classifications_file)
+```
+
+## InferCNV Heatmap
+
+Below is the heatmap produced by `InferCNV` showing the CNV's for each cell. 
+Each row is a cell and each column is a gene. 
+The top group contains all cells that were used as the reference group and the bottom is all other cells. 
+The color indicates the relative expression intensity of each gene with the darker colors either indicating a gain of expression or loss of expression compared to the reference cells.
+
+![InferCNV Heatmap - Endothelial cells as reference](results/infercnv/SCPCL000822/infercnv.png)
+From this heatmap we see that there is a distinct group of cells within the bottom portion of the plot that appear to have a lot more CNVs than the rest of the cells. 
+It's probable that these correspond to the tumor cells, but `InferCNV` does not provide any classification of cells as tumor or normal, it just provides information on copy number variations.
+
+## Using CNVs to annotate tumor cells
+
+`InferCNV` uses an HMM-based model to predict whether or not a copy number variation (either loss or gain) is present for each chromosome in each cell. 
+For each predicted CNV region, a posterior probability is calculated for that CNV for each cell.
+
+From the [`InferCNV` documentation](https://github.com/broadinstitute/inferCNV/wiki/inferCNV-HMM-based-CNV-Prediction-Methods):
+
+> CNV regions with mean posterior probabilities of being normal (no CNV) that are above a maximum threshold are removed as likely false positive predictions.
+
+The predictions information lies in a few files that live in the output directory. 
+We want to know which cells had CNVs and how many CNVs were found. 
+To map the information about CNVs to individual cells, they recommend using the `infercnv::add_to_seurat()` function which takes the output directory from running `InferCNV` and reads in all necessary files and adds the per cell CNV information to the `meta.data`.
+So we will start by converting our processed SCE object to a Seurat object, incorporate the CNV information, and then add the CNV information to the SCE object.
+
+```{r}
+# first create a Seurat object from our SCE object
+seurat_obj <- scpcaTools::sce_to_seurat(sce)
+
+# add in the CNV information 
+# this adds new columns to the meta.data indicating presence of CNVs for each chromosome
+seurat_obj <- infercnv::add_to_seurat(seurat_obj = seurat_obj,
+                                      infercnv_output_path = infer_cnv_dir)
+
+# grab metadata from Seurat and add back to SCE 
+seurat_metadata <- seurat_obj@meta.data |> 
+  dplyr::select(-c(orig.ident, nCount_RNA, nFeature_RNA))
+
+colData(sce) <- DataFrame(seurat_metadata, row.names = seurat_metadata$barcodes)
+```
+
+The metadata that we added includes the following columns for each chromosome:
+
+- `has_cnv_chr1`
+- `has_loss_chr1`
+- `has_dupli_chr1`
+- `proportion_cnv_chr1`
+- `proportion_loss_chr1`
+- `proportion_dupli_chr1`
+- `proportion_scaled_cnv_chr1`
+- `proportion_scaled_loss_chr1`
+- `proportion_scaled_dupli_chr1`
+
+The `has` columns are boolean columns indicating if that cell has presence of a CNV, loss, or duplication event. 
+If `has_cnv` is TRUE if either `has_loss` or `has_dupli` is TRUE.
+
+See [the comments in this issue](https://github.com/broadinstitute/infercnv/issues/208) explaining what the contents of the columns are.
+
+```{r}
+coldata_df <- sce |> 
+  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
+  # replace UMAP.1 with UMAP1
+  dplyr::rename_with(
+        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
+      )
+
+# remove MT and GL chr 
+chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL000219.1"))
+cnv_df <- coldata_df[, chr_columns_to_keep]
+```
+
+One approach to classifying tumor cells is to look for cells that have a lot of CNVs. 
+Let's start by just summing up the total CNVs/cell and plotting that.
+
+```{r}
+# grab any columns that have has_cnv 
+has_cnv_cols <- names(cnv_df)[startsWith(colnames(cnv_df), "has_cnv")]
+
+cnv_df <- cnv_df |>
+  # get total number of cnvs per cell and then scale that number 
+  dplyr::mutate(has_cnv = rowSums(across(has_cnv_cols)),
+                scaled_cnv = scale(has_cnv))
 ```
@@
 ```{r}
-# create the infercnv object
-infercnv_obj <- infercnv::CreateInfercnvObject(raw_counts_matrix=counts(sce),
-                                               annotations_file=annotations_file,
-                                               delim="\t",
-                                               gene_order_file=gene_order_file,
-                                               ref_group_names=c("endothelial"))
+# plot the total number of CNVs
+ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = has_cnv)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  theme_bw() +
+  scale_color_viridis_c()
+```
+This shows that the total CNVs are highest in the group of cells that correspond to the same cells that are manually annotated as tumor cells, which is promising. 
+If we were to use total CNVs we would need some sort of cut off for the number. 
+Let's look at the distribution of total CNVs and also the scaled total CNVs to see if we can identify a cut off.
 
+```{r}
+ggplot(cnv_df, aes(x = has_cnv)) +
+  geom_density() +
+  theme_bw() +
+  labs(x = "Total number of CNVs")
+
+ggplot(cnv_df, aes(x= scaled_cnv)) + 
+  geom_density() +
+  theme_bw() +
+  labs(x = "Scaled number of CNVS")
 ```
+
+Looking at the scaled number of CNVs it does look like we could use 0 as the cutoff and call any cell that has > 0 CNVs (after scaling) as tumor and all other cells as normal.
+
 ```{r}
-# perform infercnv operations to reveal cnv signal
-infercnv_obj <- infercnv::run(infercnv_obj,
-                              cutoff=0.1,  # use 1 for smart-seq, 0.1 for 10x-genomics
-                              out_dir=infer_results_dir,  # dir is auto-created for storing outputs
-                              denoise=T,
-                              HMM=T ,
-                              report_by = "cell"
+# add cnv classification 
+cnv_df <- cnv_df |> 
+  dplyr::mutate(cnv_classification = dplyr::if_else(scaled_cnv > 0, "Tumor", "Normal"))
+
+ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = cnv_classification)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  theme_bw() 
+```
+
+Another option is to use the scaled proportion of CNVs which tells us the proportion of genes on that chromosome that are part of the CNV.
+
+```{r}
+proportion_cols <- names(cnv_df)[startsWith(colnames(cnv_df), "proportion_scaled_cnv")]
+
+# get mean proportion of cnvs per cell and then scale that number 
+cnv_df <- cnv_df |> 
+  dplyr::mutate(mean_proportion = rowMeans(across(proportion_cols)),
+                scaled_mean_proportion = scale(mean_proportion))
+```
+
+```{r}
+ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = mean_proportion)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  theme_bw() +
+  scale_color_viridis_c()
+
+ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = scaled_mean_proportion)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  theme_bw() +
+  scale_color_viridis_c()
+```
+
+```{r}
+ggplot(cnv_df, aes(x = scaled_mean_proportion)) +
+  geom_density() +
+  theme_bw()
+```
+It looks like maybe we could again use 0 as the cutoff here and anything above 0 is a tumor cell. 
+Let's see how this matches up with the tumor/normal classifications we defined just using total CNVs.
+
+```{r}
+ggplot(cnv_df, aes(x = scaled_mean_proportion, color = cnv_classification)) +
+  geom_density() +
+  theme_bw()
+```
+
+Interestingly, the cells we labeled as tumor have a higher mean proportion value then those that we labeled as normal. 
+I'm not really sure which is "better" to use the mean proportion or just the detection of CNVs to call tumor cells.
+
+## Compare InferCNV annotations to manual annotations
+
+```{r}
+all_classifications_df <- cnv_df |> 
+  dplyr::left_join(manual_classifications_df) |>
+  # relevel classifications for confusion matrix 
+  dplyr::mutate(
+    cnv_classification = forcats::fct_relevel(cnv_classification, "Tumor"),
+    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
+  )
+
+caret::confusionMatrix(
+  table(
+    all_classifications_df$cnv_classification,
+    all_classifications_df$marker_gene_classification
+  )
 )
@@
 ```{r}
-infercnv_obj@observation_grouped_cell_indices
+celltype_columns <- c(
+  "singler_celltype_annotation",
+  "cellassign_celltype_annotation"
+)
 
-infercnv_obj@expr.data[1:5, 1:5]
+# create jaccard matrices for SingleR and CellAssign compared to aneuploid/diploid 
+jaccard_matrices <- celltype_columns |>
+  purrr::map(\(name) {
+    make_jaccard_matrix(
+      cnv_df,
+      "cnv_classification",
+      name
+    )
+  }) |> 
+  purrr::set_names("SingleR", "CellAssign")
 ```
+
+```{r, fig.height=10, fig.width=7}
+# Set heatmap padding option
+heatmap_padding <- 0.2
+ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))
+# list of heatmaps looking at SingleR/ CellAssign vs tumor/normal 
+heatmap <- jaccard_matrices |>
+  purrr::imap(
+    \(celltype_mat, celltype_method) {
+      ComplexHeatmap::Heatmap(
+        t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
+        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
+        border = TRUE,
+        ## Row parameters
+        cluster_rows = TRUE,
+        row_title = celltype_method,
+        row_title_gp = grid::gpar(fontsize = 12),
+        row_title_side = "left",
+        row_names_side = "left",
+        row_dend_side = "right",
+        row_names_gp = grid::gpar(fontsize = 10),
+        ## Column parameters
+        cluster_columns = FALSE,
+        column_title = "",
+        column_title_gp = grid::gpar(fontsize = 12),
+        column_names_side = "bottom",
+        column_names_gp = grid::gpar(fontsize = 10),
+        column_names_rot = 90,
+        ## Legend parameters
+        heatmap_legend_param = list(
+          title = "Jaccard index",
+          direction = "vertical",
+          legend_width = unit(1.5, "in")
+        ),
+        show_heatmap_legend = celltype_method == "SingleR",
+      )
+    }) |>
+  # concatenate vertically into HeatmapList object
+  purrr::reduce(ComplexHeatmap::`%v%`) |>
+  ComplexHeatmap::draw(
+    heatmap_legend_side = "right",
+    # add a margin to the heatmap so labels don't get cut off
+    padding = unit(c(2, 20, 2, 2), "mm")
+  )
+```
+
+## Check for known CNVs in Ewing's
+
+There are a few known copy number variations in Ewing's sarcoma:
+
+- Gain of Chr8
+- Gain of Chr12
+- Gain of Chr1p
+- Loss of Chr16q
+
+Although these are the most frequent, there are patients who do not have any of these alterations and patients that only have some of these alterations. 
+See [Tirode et al.,](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4264969/) and [Crompton et al.](https://doi.org/10.1158/2159-8290.CD-13-1037).
+
+```{r}
+# umaps showing where cells are with gains/losses 
+ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = has_dupli_chr8)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  labs(title = "Chr8 gain") +
+  theme_bw()
+
+ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = has_dupli_chr12)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  labs(title = "Chr12 gain") +
+  theme_bw()
+
+ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = has_dupli_chr1)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  labs(title = "Chr1 gain") +
+  theme_bw()
+
+ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = has_loss_chr16)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  labs(title = "Chr16 loss") +
+  theme_bw()
+```
+
+From this we see that the group of tumor cells at the bottom have chr8 and chr1 gains with some having gains of chr12. 
+It doesn't look like any cells have loss of chr16. 
+The other interesting thing is that there is a group of cells that have gains of chr8, chr1, and chr12 that we did not classify as tumor cells. Should they be...
+
+```{r}
+ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = proportion_scaled_dupli_chr8)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  labs(title = "Chr8 gain") +
+  theme_bw() +
+  scale_color_viridis_c()
+
+ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = proportion_scaled_dupli_chr12)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  labs(title = "Chr12 gain") +
+  theme_bw() +
+  scale_color_viridis_c()
+
+ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = proportion_scaled_dupli_chr1)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  labs(title = "Chr1 gain") +
+  theme_bw() +
+  scale_color_viridis_c()
+```
+Now when we look at the proportion, we see that only the group of tumor cells on the bottom seem to have a higher proportion of genes in chr8 that have a gain in expression. 
+Based on the papers, we should see a gain in the whole chr8 and chr12, not just one arm or specific genes, so we expect a high value for the proportion. 
+We see this for chr8 but not in chr12.
+
+I think with the combination of marker gene expression and CNVs, the bottom group of cells represent the tumor cells.
+
+## Do the same thing but without a reference set of cells?
+
+## Session Info
+
+```{r session info}
+# record the versions of the packages used in this analysis and other environment information
+sessionInfo()
+```

### 2024-05-09 analyses/ewings-cell-type/04-infercnv.Rmd
small txt adjustments

@@
 ```
+
 Now when we look at the proportion, we see that only the group of tumor cells on the bottom seem to have a higher proportion of genes in chr8 that have a gain in expression. 
@@
 
+## Conclusions
+
 ## Session Info

### 2024-05-10 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd
initial changes to notebook

@@
                      annotation_row = annotation_df)
+```
+
+```{r}
+tumor_celltype_assignments <- tumor_celltype_assignments |> 
+  dplyr::rename(cellassign_annotations = celltype)
+
+celltype_df <- sce |> 
+  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
+  # replace UMAP.1 with UMAP1
+  dplyr::rename_with(
+        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
+      ) |> 
+  # join with tumor normal
+  dplyr::left_join(classifications_df, by = "barcodes") |> 
+  # join with filtered cell type assignments from cellassign 
+  dplyr::left_join(tumor_celltype_assignments, by = c("barcodes" = "barcode"))
+
+ggplot(celltype_df, aes(x = UMAP1, y = UMAP2, color = cellassign_annotations)) +
+  geom_point(size = 0.5, alpha = 0.5) + 
+  theme_bw()
 ```

### 2024-05-10 analyses/ewings-cell-type/03-copykat.Rmd
remove old notebook
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 ```{r}
-caret_df <- cnv_df |> 
+filtered_cnv_df <- cnv_df |> 
   dplyr::filter(copykat.pred != "not.defined") |> 
+caret_df_list <- filtered_cnv_df |> 
   dplyr::mutate(copykat = ifelse(
@@
 Looking at these results, we see that cells classified as tumor cells using marker genes are mostly overlapping with cells that are aneuploid in CopyKAT, which is what we expect! 
-We also see that the Kappa values range are close to 1, indicating almost perfect agreement.
+We also see that the Kappa values are close to 1, indicating almost perfect agreement.
 

### 2024-05-10 analyses/ewings-cell-type/scripts/run-copykat.R

@@ -49,9 +49,7 @@ opt <- parse_args(OptionParser(option_list = option_list))
 # Set up -----------------------------------------------------------------------
 
 # make sure path to sce file exists
-if(!file.exists(opt$sce_file)){
-  stop("--sce_file does not exist.")
-}
+stopifnot("sce_file does not exist" = file.exists(opt$sce_file))
 
 # read in sce file
 sce <- readr::read_rds(opt$sce_file)

### 2024-05-10 analyses/ewings-cell-type/03-copykat.Rmd
add common CNAs to analysis

@@
 - Visualize diploid and anueploid cells on the UMAP. 
+- Evaluate common copy number gains and losses in Ewing sarcoma. 
 - Calculate the confusion matrix comparing manual annotations of tumor cells to using `CopyKAT` to annotate tumor cells. 
@@
 results_dir <- file.path(module_base, "results", "copykat", params$library_id)
-results_filenames <- c(
-  no_ref = glue::glue("{params$library_id}_no_normal_copykat_prediction.txt"), 
-  with_ref = glue::glue("{params$library_id}_with_normal_copykat_prediction.txt")
+
+# predictions files 
+predictions_file <- glue::glue("{params$library_id}_copykat_prediction.txt")
+
+predictions_paths <- c(
+  no_ref = file.path(results_dir, "no_normal", predictions_file),
+  with_ref = file.path(results_dir, "endothelial_normal", predictions_file)
 )
-ck_results_files <- file.path(results_dir, results_filenames) |> 
-  purrr::set_names(names(results_filenames))
+
+# full results gene by cell 
+full_ck_result_file <- glue::glue("{params$library_id}_copykat_CNA_raw_results_gene_by_cell.txt")
+
+full_ck_result_paths <- c(
+  no_ref = file.path(results_dir, "no_normal", full_ck_result_file),
+  with_ref = file.path(results_dir, "endothelial_normal", full_ck_result_file)
+)
+
 ```
@@
 
-# read in ck results from both reference types (no_normal and with_normal)
-ck_results_df <- ck_results_files |> 
+# read in ck predictions from both reference types (no_normal and with_normal)
+ck_results_df <- predictions_paths |> 
+  purrr::map(readr::read_tsv) |> 
+  dplyr::bind_rows(.id = "reference_used")
+
+# read in full gene by cell copy number detection results 
+full_ck_results_df <- full_ck_result_paths |> 
   purrr::map(readr::read_tsv) |> 
@@
 ## CopyKAT results
+
+Below we look at the heatmaps produced by `CopyKAT`.
+
+### Heatmap without reference
+
+![](`r file.path(results_dir, "no_normal", "SCPCL000822_copykat_heatmap.jpeg")`)
+
+### Heatmap with endothelial cells as reference
+
+![](`r file.path(results_dir, "endothelial_normal", "SCPCL000822_copykat_heatmap.jpeg")`)
+
+Both of these heatmaps appear to show some increase in observed CNVs in aneuploid vs. diploid which is to be expected.
 
@@
 
+## Validate common CNAs found in Ewing sarcoma
+
+To validate some of these annotations, we can also look at some commonly found copy number variations found in Ewing sarcoma patients. 
+There are a few known copy number variations in Ewing's sarcoma:
+
+- Gain of Chr8
+- Gain of Chr12
+- Gain of Chr1p
+- Loss of Chr16q
+
+Although these are the most frequent, there are patients who do not have any of these alterations and patients that only have some of these alterations. 
+See [Tirode et al.,](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4264969/) and [Crompton et al.](https://doi.org/10.1158/2159-8290.CD-13-1037).
+
+`CopyKAT` outputs a matrix that contains the estimated copy numbers for each gene in each cell. 
+We can read that in and look at the mean estimated copy numbers for each chromosome across each cell. 
+We might expect that tumor cells would show an increased estimated copy number in Chr8, Chr12, and/or Chf 1 and a loss of Chr16.
+
+```{r}
+# for every cell, calculate the mean detection level across all genes in a given chromosome
+full_cnv_df <- full_ck_results_df |> 
+  tidyr::pivot_longer(cols = -c(
+    reference_used,
+    abspos, 
+    chromosome_name, 
+    start_position, 
+    end_position, 
+    ensembl_gene_id, 
+    hgnc_symbol, 
+    band),
+                      names_to = "barcodes",
+                      values_to = "cnv_detection") |> 
+  dplyr::group_by(chromosome_name, barcodes, reference_used) |> 
+  dplyr::summarise(mean_cnv_detection = mean(cnv_detection))
+
+# filter to only have cnvs found in Ewings to keep things a bit smaller when joining
+ewings_chrs <- c(1, 8, 12, 16)
+
+ewings_cnv_df <- full_cnv_df |> 
+  dplyr::filter(chromosome_name %in% ewings_chrs)
+
+# join with 
+cnv_df <- cnv_df |> 
+  dplyr::left_join(ewings_cnv_df, by = c("barcodes", "reference_used"))
+```
+
+```{r}
+# create faceted UMAPs showing estimation of CNV detection across each chr of interest 
+ewings_chrs |> 
+  purrr::map(\(chr){
+    plot_df <- cnv_df |> 
+      dplyr::filter(chromosome_name == chr)
+
+    ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = mean_cnv_detection)) +
+      geom_point(alpha = 0.5, size = 0.5) +
+      theme_bw() +
+      scale_color_viridis_c() +
+      facet_wrap(vars(reference_used)) + 
+      labs(title = glue::glue("chr{chr}"))
+  })
+```
+
+Looking at the UMAPs, I don't see any really obvious differences between CNV estimation and groups of cells. 
+Let's look at the distribution of CNV estimation in cells that are called aneuploid and diploid by `CopyKAT`.
+
+```{r}
+# create faceted density plots showing estimation of CNV detection across each chr of interest 
+# colored by aneuploid/diploid estimation 
+ewings_chrs |> 
+  purrr::map(\(chr){
+    plot_df <- cnv_df |> 
+      dplyr::filter(chromosome_name == chr)
+
+    ggplot(plot_df, aes(x = mean_cnv_detection, color = copykat.pred)) +
+      geom_density() +
+      theme_bw() +
+      facet_wrap(vars(reference_used)) + 
+      labs(title = glue::glue("chr{chr}"))
+  })
+```
+
+It looks like there is a slight increase in chr8 and chr12 indicating maybe a copy number gain for those chromosomes in aneuploid cells, but nothing striking.
+
 ### Confusion Matrix
@@
 filtered_cnv_df <- cnv_df |> 
-  dplyr::filter(copykat.pred != "not.defined") |> 
+  dplyr::filter(copykat.pred != "not.defined")
+
 caret_df_list <- filtered_cnv_df |> 
@@
 
-caret_df |> 
+caret_df_list |> 
   purrr::imap(\(df, ref_type){
-    glue::glue("\n\n================== {ref_type} ====================")
     caret::confusionMatrix(
@@
 # calculate Jaccard similarity index for each reference type
-ref_types <- c("no_ref", "with_ref")
-jaccard_matrices <- ref_types |>
-  purrr::map(\(ref_type) {
-    jaccard_df <- cnv_df |> 
-      dplyr::filter(reference_used == ref_type)
+jaccard_matrices <- caret_df_list |>
+  purrr::map(\(df) {
 
     make_jaccard_matrix(
-      cnv_df,
+      df,
       "marker_gene_classification",
@@
     )
-  }) |> 
-  purrr::set_names(ref_types)
+  })
 ```
@@
 - `CopyKAT` seems to do a good job in this Ewing sarcoma sample of identifying aneuploid and diploid cells. 
+- There is some small evidence of copy number gains in Chr8 and Chr12 which is consistent with findings in Ewing sarcoma patients. 
 - The annotations from `CopyKAT` mostly line up with annotations obtained using marker gene annotation. 

### 2024-05-13 analyses/doublet-detection/01_compare-doublet-methods.Rmd

--- 2024-04-24 analyses/ewings-cell-type/notebook-template.Rmd
@@
 ---
-title: "ewings-cell-type Analysis"
+title: "Compare doublet detection methods"
 author:
-  - Your name here
+  - Stephanie J. Spielman
 date: "`r Sys.Date()`"
@@
 
-This a template for an analysis notebook using R Markdown.
-
-It is a good idea to start with a brief introduction to the analysis, including the purpose of the analysis, the data used, and the methods applied.
-Replace this text with your own introduction, and be sure to update the `title:` and `author:` fields in the YAML header at the top of the document.
-
-Don't forget to rename this file as well!
+This notebook will compare doublet detection results from several methods across a set of ScPCA libraries (_not multiplexed ones_) with varying size and diagnoses. 
+The methods include [`scDblFinder`](https://bioconductor.org/packages/3.18/scDblFinder) and three methods from [`scds`](https://bioconductor.org/packages/devel/bioc/vignettes/scds/inst/doc/scds.html): `cxds`, `bsds`, and `hybrid` (a combination of the previous two methods).
 
@@
 
-Load required packages in the following chunk, if needed.
-Do not install packages here; only load them with the `library()` function.
-
 ```{r packages}
 # load required packages
-# e.g. library(tidyverse)
+library(SingleCellExperiment)
+library(scDblFinder)
+library(scds)
 ```
-
-You may also want to use the `package::function()` syntax within your code for better tracking of which functions come from which packages, and to avoid conflicts between functions with the same name in different packages.
 
@@
 
-Setting paths to directories and files for input and output at the start of a notebook can be helpful for organization and reproducibility.
-You may not know all of the paths you need when you start a notebook, but once you do, keeping them all in one place is helpful, both for yourself in the future and for other users of the notebook.
-
 #### Base directories
-
-In the chunk below, we include some common directories that you might need, finding their locations using the very useful [`rprojroot` package](https://rprojroot.r-lib.org/).
 
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
+module_base <- file.path(repository_base, "analyses", "doublet-detection")
 ```
@@
 
-Organize the remainder of your content into sections and subsections as appropriate for your analysis.
-
-Add a new code chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I* on Windows or Linux, *Cmd+Option+I* on a Mac.
-Be sure to comment your code and explain your steps as you go.
-
-When you save the notebook, an HTML file containing the code and output will be saved alongside it.
-To preview the HTML file, click the *Preview* button or press *Ctrl+Shift+K* on Windows or Linux, *Cmd+Shift+K* on a Mac.
-
-The preview shows you a rendered HTML copy of the contents of the editor.
-Consequently, *Preview* does not run any R code chunks.
-Instead, the output of the chunk when it was last run in the editor is displayed.
-
-It is usually a good idea to rerun the entire notebook before sharing any output, to ensure that the notebook and output files are fully up to date.
-You can do this by clicking on the *Run* menu in the toolbar and clicking on *Restart R and Run All Chunks*.
-
 ## Session Info

### 2024-05-13 analyses/ewings-cell-type/03-copykat.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 
-## Validate common CNAs found in Ewing sarcoma
+### Validate common CNAs found in Ewing sarcoma
 
@@
 We can read that in and look at the mean estimated copy numbers for each chromosome across each cell. 
-We might expect that tumor cells would show an increased estimated copy number in Chr8, Chr12, and/or Chf 1 and a loss of Chr16.
+We might expect that tumor cells would show an increased estimated copy number in Chr8, Chr12, and/or Chr 1 and a loss of Chr16.
 

### 2024-05-13 analyses/ewings-cell-type/scripts/identify-normal-cells.R

@@ -41,7 +41,7 @@ option_list <- list(
   make_option(
     opt_str = c("--output_filename"),
     type = "character",
-    help = "filename to use for saved list of normal cells. Will be saved inside `--output_dir/library_id`.
+    help = "filename to use for saved list of normal cells. Will be saved inside `<output_dir>/library_id`.
       Must end in `.txt`"
   )
 )

### 2024-05-13 analyses/ewings-cell-type/03-copykat.Rmd
plot each chr and remove UMAPs

@@
 
-# filter to only have cnvs found in Ewings to keep things a bit smaller when joining
-ewings_chrs <- c(1, 8, 12, 16)
-
-ewings_cnv_df <- full_cnv_df |> 
-  dplyr::filter(chromosome_name %in% ewings_chrs)
-
-# join with 
+# join with cnv info
 cnv_df <- cnv_df |> 
-  dplyr::left_join(ewings_cnv_df, by = c("barcodes", "reference_used"))
+  dplyr::left_join(full_cnv_df, by = c("barcodes", "reference_used"))
 ```
 
-```{r}
-# create faceted UMAPs showing estimation of CNV detection across each chr of interest 
-ewings_chrs |> 
-  purrr::map(\(chr){
-    plot_df <- cnv_df |> 
-      dplyr::filter(chromosome_name == chr)
+Let's look at the distribution of CNV estimation in cells that are called aneuploid and diploid by `CopyKAT`.
 
-    ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = mean_cnv_detection)) +
-      geom_point(alpha = 0.5, size = 0.5) +
+```{r, fig.height=15, fig.width=10}
+# create faceted density plots showing estimation of CNV detection across each chr of interest 
+# colored by aneuploid/diploid estimation 
+ggplot(cnv_df, aes(x = mean_cnv_detection, color = copykat.pred)) +
+      geom_density() +
       theme_bw() +
-      scale_color_viridis_c() +
-      facet_wrap(vars(reference_used)) + 
-      labs(title = glue::glue("chr{chr}"))
-  })
+      facet_grid(rows = vars(chromosome_name), 
+                 cols = vars(reference_used))
 ```
 
-Looking at the UMAPs, I don't see any really obvious differences between CNV estimation and groups of cells. 
-Let's look at the distribution of CNV estimation in cells that are called aneuploid and diploid by `CopyKAT`.
+It looks like the largest increases are in chr8 and chr21. 
+Chr8 is one of the more common CNA's in Ewing's so this is a good sign.
 
-```{r}
-# create faceted density plots showing estimation of CNV detection across each chr of interest 
-# colored by aneuploid/diploid estimation 
-ewings_chrs |> 
-  purrr::map(\(chr){
-    plot_df <- cnv_df |> 
-      dplyr::filter(chromosome_name == chr)
-
-    ggplot(plot_df, aes(x = mean_cnv_detection, color = copykat.pred)) +
-      geom_density() +
-      theme_bw() +
-      facet_wrap(vars(reference_used)) + 
-      labs(title = glue::glue("chr{chr}"))
-  })
-```
-
-It looks like there is a slight increase in chr8 and chr12 indicating maybe a copy number gain for those chromosomes in aneuploid cells, but nothing striking.
+Chr12 may show some increase, but Chr1 and Chr16 don't show any real differences.
 

### 2024-05-13 analyses/cell-type-ewings/02-cellassign-tumor-cell-classification.Rmd
change folder name

--- 2024-05-07 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd
@@
 - A full list of tumor marker genes only. 
-- A filtered list of tumor marker genes only. 
+- A filtered list of marker genes. 
 - Tumor marker genes and non-tumor marker genes from [Visser et al.](https://doi.org/10.1158/2767-9764.CRC-23-0027) specified in the supplemental methods. 
@@
 module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
+```
 
-# source in helper functions
+```{r}
+# source in helper functions make_jaccard_matrix() and jaccard()
 jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
@@
   dplyr::distinct()
+
+# read in tumor normal classifications 
+# obtained from manual classification in `01-marker-gene-tumor-classifications.Rmd`
+classifications_df <- readr::read_tsv(classifications_file)
 ```
@@
 We can then read in the predictions file and find the predicted cell type for each cell, either "tumor" or "other". 
+Then we'll show a heatmap of the scores for each cell and cell type. 
+Included in that heatmap will be annotation of which cells were classified as tumor or normal using the manual classification.
 
@@
 ```{r}
-# first read in tumor normal classifications 
-# obtained from manual classification in `01-marker-gene-tumor-classifications.Rmd`
-classifications_df <- readr::read_tsv(classifications_file)
-
 # get annotations for heatmaps
@@
 
-Looking at this, it appears that `CellAssign` is classifying all cells as "other"? 
-I find this hard to believe since this sample was taken directly from a Ewing sarcoma tumor, so there are bound to be tumor cells.
+It looks like we might have pretty good agreement between the annotations from `CellAssign` and manual annotations. 
+Let's also look at the cell annotations on a UMAP and then look at the confusion matrix.
+
+```{r}
+# rename cell type column 
+tumor_celltype_assignments <- tumor_celltype_assignments |> 
+  dplyr::rename(cellassign_annotations = celltype)
+
+# create a dataframe to use for plotting with UMAP and cell information 
+celltype_df <- sce |> 
+  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
+  # replace UMAP.1 with UMAP1
+  dplyr::rename_with(
+        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
+      ) |> 
+  # join with tumor normal
+  dplyr::left_join(classifications_df, by = "barcodes") |> 
+  # join with filtered cell type assignments from cellassign 
+  dplyr::left_join(tumor_celltype_assignments, by = c("barcodes" = "barcode"))
+```
+
+```{r}
+ggplot(celltype_df, aes(x = UMAP1, y = UMAP2, color = cellassign_annotations)) +
+  geom_point(size = 0.5, alpha = 0.5) + 
+  theme_bw()
+```
+
+```{r}
+# prep annotations for confusion matrix
+celltype_df <- celltype_df |> 
+  dplyr::mutate(cellassign_annotations = ifelse(
+    # make sure annotations match marker gene annotations
+    cellassign_annotations == "tumor", "Tumor", "Normal"
+  )) |> 
+  # make tumor the positive class
+  dplyr::mutate(
+    cellassign_annotations = forcats::fct_relevel(cellassign_annotations, "Tumor"),
+    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
+  )
+
+caret::confusionMatrix(
+  table(
+    celltype_df$marker_gene_classification, 
+    celltype_df$cellassign_annotations)
+)
+```
+
+This looks pretty good! 
+There is quite a bit of agreement between classifying cells manually with marker genes vs. classifying cells with `CellAssign` and using the marker genes.
+
+To confirm that the cells with tumor cells have higher expression of marker genes, let's look at the mean expression of marker genes between tumor and other cells as classified by `CellAssign`.
+
+```{r}
+# get marker gene expression 
+marker_gene_exp <- logcounts(sce[tumor_markers_df$ensembl_gene_id, ]) |>
+  as.matrix() |>
+  t() |>
+  as.data.frame() |>
+  tibble::rownames_to_column("barcodes")
+
+marker_gene_exp_df <- celltype_df |> 
+  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
+  tidyr::pivot_longer(
+    cols = starts_with("ENSG"),
+    names_to = "ensembl_gene_id",
+    values_to = "gene_expression"
+  ) |> 
+  dplyr::group_by(barcodes) |> 
+  dplyr::mutate(mean_exp = mean(gene_expression))
+
+ggplot(marker_gene_exp_df, aes(x = mean_exp, fill = cellassign_annotations)) +
+  geom_density() +
+  facet_grid(rows = vars(cellassign_annotations)) +
+  theme_bw()
+```
+
+As expected, we see that cells classified as tumor cells have higher expression of the tumor marker genes.
 
@@
 
-Now we are seeing a few more tumor cells, let's see where they are on the UMAP.
-
-```{r}
-filtered_celltype_assignments <- filtered_celltype_assignments |> 
-  dplyr::rename(filtered_cellassign_annotations = celltype)
-
-celltype_df <- sce |> 
-  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
-  # replace UMAP.1 with UMAP1
-  dplyr::rename_with(
-        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
-      ) |> 
-  # join with tumor normal
-  dplyr::left_join(classifications_df, by = "barcodes") |> 
-  # join with filtered cell type assignments from cellassign 
-  dplyr::left_join(filtered_celltype_assignments, by = c("barcodes" = "barcode"))
-
-ggplot(celltype_df, aes(x = UMAP1, y = UMAP2, color = filtered_cellassign_annotations)) +
-  geom_point(size = 0.5, alpha = 0.5) + 
-  theme_bw()
-```
-
-Not quite where I would expect to see tumor cells. 
-Let's see how many tumor cells overlap with our manual classifications using the Jaccard similarity index.
-
-```{r}
-jaccard_mtx <- make_jaccard_matrix(celltype_df, "filtered_cellassign_annotations", "marker_gene_classification")
-
-ComplexHeatmap::Heatmap(
-  t(jaccard_mtx), # transpose because matrix rows are in common & we want a vertical arrangement
-  col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
-  border = TRUE,
-  ## Row parameters
-  cluster_rows = FALSE,
-  row_title = "Marker gene",
-  row_title_gp = grid::gpar(fontsize = 12),
-  row_title_side = "left",
-  row_names_side = "left",
-  row_dend_side = "right",
-  row_names_gp = grid::gpar(fontsize = 10),
-  ## Column parameters
-  cluster_columns = FALSE,
-  column_title = "CellAssign",
-  column_title_gp = grid::gpar(fontsize = 12),
-  column_names_side = "bottom",
-  column_names_gp = grid::gpar(fontsize = 10),
-  column_names_rot = 90,
-  ## Legend parameters
-  heatmap_legend_param = list(
-    title = "Jaccard index",
-    direction = "vertical",
-    legend_width = unit(1.5, "in")
-  ),
-  show_heatmap_legend = TRUE,
-)
-```
-
-As expected, it looks like CellAssign is assigning very few cells to tumor that overlap with the tumor cells annotated with marker gene analysis. 
-It looks like most tumor cells are being classified as normal with the marker gene analysis.
-
-Let's look at the mean expression of marker genes between tumor and other cells as classified by `CellAssign`.
-
-```{r}
-# get marker gene expression 
-marker_gene_exp <- logcounts(sce[filtered_markers, ]) |>
-  as.matrix() |>
-  t() |>
-  as.data.frame() |>
-  tibble::rownames_to_column("barcodes")
-
-marker_gene_exp_df <- celltype_df |> 
-  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
-  tidyr::pivot_longer(
-    cols = starts_with("ENSG"),
-    names_to = "ensembl_gene_id",
-    values_to = "gene_expression"
-  ) |> 
-  dplyr::group_by(barcodes) |> 
-  dplyr::mutate(mean_exp = mean(gene_expression))
-
-ggplot(marker_gene_exp_df, aes(x = mean_exp, fill = filtered_cellassign_annotations)) +
-  geom_density() +
-  facet_grid(rows = vars(filtered_cellassign_annotations)) +
-  theme_bw()
-```
-
-Unlike what I would expect, the "other" cells appear to have higher expression of marker genes? 
-Maybe using `CellAssign` with only two options and a small set of marker genes doesn't work in the way we would expect.
+Interestingly, `CellAssign` doesn't seem to like using such a limited list of marker genes and assigns all cells as "other".
 
@@
 
-Perhaps we need to add a few more cell types, rather than just have the two options. 
-Below we will use CellAssign but with all tumor marker genes and all marker genes in Visser et al. for non-tumor cells.
+We can also test to see if classifying tumor cells works with adding a few more cell types, rather than just have the two options. 
+Below we will use `CellAssign` but with all tumor marker genes and all marker genes in Visser et al. for non-tumor cells.
 
@@
 
-Here we see that there are still only a few hundred tumor cells and most cells are endothelial cells. 
-Generally, I expect that there to be a large portion of tumor cells with much fewer cells from normal cell types.
+Interestingly when we use this reference we actually see that most cells are classified as other, with some cells being classified into the different normal cell types.
 
-Below we will look at which cells are assigned to which cell type and then compare assignments between CellAssign and manual marker gene assignment.
+Below we will look at which cells are assigned to which cell type and then compare assignments between `CellAssign` and manual marker gene assignment.
 
@@
 Again, we don't see much overlap of cells that we are classifying as tumor manually to the tumor cells being classified by CellAssign. 
-We do see that the majority of the immune and mesenchymal-like cells fall into the "normal" category, but this isn't the case for endothelial cells.
+We do see that the majority of the immune and mesenchymal-like cells fall into the "normal" category.
+
+The one cell type that is identified here and in the `CellAssign` annotations with the reference from `scpca-nf` is Endothelial cells. 
+Let's compare the classifications of Endothelial cells between the two references.
+
+```{r}
+# prep annotations for confusion matrix
+celltype_df <- celltype_df |> 
+  dplyr::mutate(
+    caret_visser_annotations = ifelse(
+      # make sure annotations match marker gene annotations
+      visser_cellassign_annotations == "endothelial cells", "Endothelial cells", "Other"
+    ),
+    caret_cellassign_annotations = ifelse(
+      cellassign_celltype_annotation == "Endothelial cells", "Endothelial cells", "Other"
+    ),  
+    # make tumor the positive class
+    caret_visser_annotations = forcats::fct_relevel(caret_visser_annotations, "Endothelial cells"),
+    caret_cellassign_annotations = forcats::fct_relevel(caret_cellassign_annotations, "Endothelial cells")
+  )
+
+caret::confusionMatrix(
+  table(
+    celltype_df$caret_cellassign_annotations, 
+    celltype_df$caret_visser_annotations)
+)
+```
+
+The good news is although tumor cells aren't getting classified, the same group of cells is getting classified as Endothelial cells, even with the smaller list of marker genes.
 
@@
 
-Interestingly, the endothelial and MSC like markers seem to be localized to a small group of cells, where as CellAssign assigns most cells to endothelial cells.
+In comparing these UMAPs to the UMAP showing the cell type annotations with the Visser reference, it does appear that Endothelial cells are expressing the marker genes and that the MSC-like cells are expressing the expected marker genes.
 
@@
 
-It looks like the only real difference between groups is that the endothelial cells have higher expression of the tumor markers then other cells? 
-However the tumor cells do not show increased expression of the tumor marker genes. 
-This really is not what I would expect... maybe we need a larger reference...
+It looks like the "other" cells actually have high expression of the tumor marker genes and instead should be classified as tumor cells.
 
@@
 
-This is definitely not what I expected... 
-It looks like most cells are categorized as "other" with a few categorized as tumor and only 1 endothlial cell. 
-I would have thought that at least some cells would be endothelial or fibroblast.
+Although only 2 cells are classified as Fibroblasts, we do see a group of Endothelial cells and then tumor cells that line up with our manual annotations of tumor/normal.
 
@@
 
-Again, it looks like cells that were classified as normal are considered tumor here. 
-I think if we want to use `CellAssign`, we will need to spend more time carefully curating a marker gene list that includes all possible cell types for these samples. 
-Another alternative is to use the `CellAssign` annotations that we already have to help guide which cells may be tumor or normal. 
-For example, we expect that muscle cells will be tumor cells.
+The UMAP tells me that we are classifying the same group of endothelial cells as we did in `scpca-nf` with `CellAssign` and that the tumor cells line up with the manual annotations of tumor cells (muscle cells in `CellAssign`).
+
+Let's directly compare these annotations to the `CellAssign` annotations obtained from `scpca-nf`.
+
+```{r}
+# get jaccard similarity index 
+jaccard_mtx <- make_jaccard_matrix(celltype_df, "cellassign_celltype_annotation", "panglao_cellassign_annotations")
+
+ComplexHeatmap::Heatmap(
+  t(jaccard_mtx), # transpose because matrix rows are in common & we want a vertical arrangement
+  col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
+  border = TRUE,
+  ## Row parameters
+  cluster_rows = FALSE,
+  row_title = "Panglao + marker genes",
+  row_title_gp = grid::gpar(fontsize = 12),
+  row_title_side = "left",
+  row_names_side = "left",
+  row_dend_side = "right",
+  row_names_gp = grid::gpar(fontsize = 10),
+  ## Column parameters
+  cluster_columns = FALSE,
+  column_title = "scpca-nf CellAssign",
+  column_title_gp = grid::gpar(fontsize = 12),
+  column_names_side = "bottom",
+  column_names_gp = grid::gpar(fontsize = 10),
+  column_names_rot = 90,
+  ## Legend parameters
+  heatmap_legend_param = list(
+    title = "Jaccard index",
+    direction = "vertical",
+    legend_width = unit(1.5, "in")
+  ),
+  show_heatmap_legend = TRUE,
+)
+```
+
+It looks like combining some of the marker genes from PanglaoDB with marker genes does a good job of capturing the endothelial cells and tumor cells. 
+It doesn't capture the Fibroblasts, but instead classifies those as other.
+
+If we want to combine annotation of both tumor cells and normal cells, we may consider using the full PanglaoDB marker gene references we have and adding tumor marker genes to that list. 
+However, this will be more computationally expensive since those references are quite large.
 
@@
 
-- Using `CellAssign` with the same list of marker genes doesn't identify any tumor cells. 
-- Using `CellAssign` with a restricted list of marker genes identfies tumor cells that have lower marker gene expression than other cells. 
-This is opposite of the assignment we obtained using manual annotation. 
-- Using `CellAssign` with an expanded list of marker genes from Visser et al., identifies all cell types, but the majority are endothelial cells. 
+- Using `CellAssign` with a list of tumor markers identifies tumor cells similar to manual annotations. 
+- Using `CellAssign` with a restricted list of marker genes doesn't identify any tumor cells. 
+- Using `CellAssign` with an expanded list of marker genes from Visser et al., identifies some normal cell types, but struggles to identify any tumor cells. 
 This doesn't match up with what's expected based on marker gene expression in the sample. 
-- Using `CellAssign` with the Panglao markers and tumor markers identifies mostly normal cells and very few tumor cells. 
-Again, the cells labeled as tumor cells are opposite of the manual annotation.
+- Using `CellAssign` with the Panglao markers and tumor markers identifies endothelial cells and tumor cells similar to what we would expect.
 

### 2024-05-13 analyses/cell-type-ewings/03-copykat.Rmd
change folder name

--- 2024-05-13 analyses/ewings-cell-type/03-copykat.Rmd
@@
 # full results gene by cell 
-full_ck_result_file <- glue::glue("{params$library_id}_copykat_CNA_raw_results_gene_by_cell.txt")
+full_ck_result_file <- glue::glue("{params$library_id}_copykat_CNA_results.txt")
 
@@
     reference_used,
-    abspos, 
-    chromosome_name, 
-    start_position, 
-    end_position, 
-    ensembl_gene_id, 
-    hgnc_symbol, 
-    band),
+    chrom
+    ),
     names_to = "barcodes",
     values_to = "cnv_detection") |> 
-  dplyr::group_by(chromosome_name, barcodes, reference_used) |> 
+  dplyr::group_by(chrom, barcodes, reference_used) |> 
   dplyr::summarise(mean_cnv_detection = mean(cnv_detection))
@@
       theme_bw() +
-      facet_grid(rows = vars(chromosome_name), 
+      facet_grid(rows = vars(chrom), 
                  cols = vars(reference_used))

### 2024-05-13 analyses/ewings-cell-type/scripts/run-cellassign.py

@@ -1,127 +1,0 @@
-#!/usr/bin/env python3
-
-# The contents of this script were taken from https://github.com/AlexsLemonade/scpca-nf/blob/main/bin/predict_cellassign.py
-# This script takes an AnnData object and a reference marker gene matrix
-# where genes are rows and columns are cell types
-# CellAssign is run and the predictions matrix is returned
-
-import argparse
-import os
-import re
-
-import anndata as adata
-import numpy as np
-import pandas as pd
-import scvi
-from scvi.external import CellAssign
-
-parser = argparse.ArgumentParser()
-parser.add_argument(
-    "-i",
-    "--anndata_file",
-    dest="anndata_file",
-    required=True,
-    help="Path to H5AD file with processed AnnData object to annotate",
-)
-parser.add_argument(
-    "-o",
-    "--output_predictions",
-    dest="output_predictions",
-    required=True,
-    help="Path to file to save the predictions, should end in tsv",
-)
-parser.add_argument(
-    "-r",
-    "--reference",
-    dest="reference",
-    required=True,
-    help="Path to marker by cell type reference file, should end in tsv",
-)
-parser.add_argument(
-    "-s",
-    "--seed",
-    dest="seed",
-    type=int,
-    default=None,
-    help="Random seed to set for scvi.",
-)
-parser.add_argument(
-    "-t",
-    "--threads",
-    dest="threads",
-    default=1,
-    type=int,
-    help="Number of threads to use when training the CellAssign model",
-)
-
-args = parser.parse_args()
-
-# set seed
-scvi.settings.seed = args.seed
-
-# define the number of threads to use
-scvi.settings.num_threads = args.threads
-
-# compile extension regex
-file_ext = re.compile(r"\.h5ad$|\.hdf5$|.h5$", re.IGNORECASE)
-
-# check that input file exists, if it does exist, make sure it's an h5 file
-if not os.path.exists(args.anndata_file):
-    raise FileExistsError("--anndata_file file not found.")
-elif not file_ext.search(args.anndata_file):
-    raise ValueError(
-        "--anndata_file must end in .h5ad, .hdf5, or .h5 and contain a processed AnnData object."
-    )
-
-# check that marker file exists and make sure its a tsv
-if not os.path.exists(args.reference):
-    raise FileExistsError("--reference file not found.")
-
-# make sure output file path is tsv file
-if not args.output_predictions.endswith(".tsv"):
[…]

### 2024-05-13 analyses/cell-type-ewings/01-marker-gene-tumor-classification.Rmd
move files

--- 2024-05-03 analyses/ewings-cell-type/01-marker-gene-tumor-classification.Rmd
@@
 ```{r}
-# source in helper functions: make_jaccard_matrix and jaccard 
+# source in helper functions: make_jaccard_matrix() and jaccard() 
 jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
@@
   dplyr::rename_with(
-        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
-        starts_with("UMAP")
+        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
       ) |> 

### 2024-05-13 analyses/cell-type-ewings/04-infercnv.Rmd
move files

--- 2024-05-09 analyses/ewings-cell-type/04-infercnv.Rmd
@@
 # output from running infercnv 
-infer_cnv_dir <- file.path(module_base, "results", "infercnv", "SCPCL000822")
+results_dir <- file.path(module_base, "results", "infercnv", params$library_id)
 
-# add code to read in both ref and no ref? 
+all_results_dir_list <- c(
+  no_ref = file.path(results_dir, "no_reference"),
+  with_ref = file.path(results_dir, "endothelial_reference")
+)
 ```
@@
 
-Below is the heatmap produced by `InferCNV` showing the CNV's for each cell. 
+### Endothelial cells as reference
+
+Below is the heatmap produced by `InferCNV` showing the CNV's for each cell using endothelial cells as a reference. 
 Each row is a cell and each column is a gene. 
@@
 
-![InferCNV Heatmap - Endothelial cells as reference](results/infercnv/SCPCL000822/infercnv.png)
+![InferCNV Heatmap - Endothelial cells as reference](`r file.path(all_results_dir_list["with_ref"], "infercnv.png")`)
+
 From this heatmap we see that there is a distinct group of cells within the bottom portion of the plot that appear to have a lot more CNVs than the rest of the cells. 
 It's probable that these correspond to the tumor cells, but `InferCNV` does not provide any classification of cells as tumor or normal, it just provides information on copy number variations.
+
+### No reference
+
+Below is the heatmap produced by `InferCNV` showing the CNV's for each cell without using a reference.
+
+From the [`InferCNV` documentation](https://github.com/broadinstitute/inferCNV/wiki/Running-InferCNV):
+
+> Note, if you do not have reference cells, you can set ref_group_names=NULL, in which case the average signal across all cells will be used to define the baseline.
+This can work well when there are sufficient differences among the cells included (ie.
+they do not all show a chromosomal deletion at the same place).
+
+Given this information, I wouldn't expect the CNV inference to work well because we see that all cells that do have prominent CNVs have the same CNVs. 
+There is not a lot of variation.
+
+![InferCNV Heatmap - No reference](`r file.path(all_results_dir_list["no_ref"], "infercnv.png")`)
+
+Looking at this we see a small group of cells that may have some variations, but they are different from the variations seen above. 
+I don't know that using this without a reference is the best call since it's unlikely that it will identify distinct groups of cells from tumors with quiet genomes with no baseline. 
+The rest of the analysis will use the results from running `InferCNV` with endothelial cells as the reference.
 
@@
 ```{r}
+infer_cnv_dir <- all_results_dir_list["with_ref"]
+
 # first create a Seurat object from our SCE object
@@
 
+One approach to classifying tumor cells is to look for cells that have a lot of CNVs. 
+Let's start by just summing up the total CNVs/cell and plotting that.
+
 ```{r}
+# make df for plotting
 coldata_df <- sce |> 
@@
 
-# remove MT and GL chr 
+# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
 chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL000219.1"))
@@
 ```
-
-One approach to classifying tumor cells is to look for cells that have a lot of CNVs. 
-Let's start by just summing up the total CNVs/cell and plotting that.
 
@@
 
+For the rest of the notebook, we will use the classifications of tumor cells based on the sum of the total CNV occurrences.
+
 ## Compare InferCNV annotations to manual annotations
+
+Below we can compare these annotations from InferCNV to classifying tumor cells manually with marker genes.
 
@@
 ```
+
+This looks pretty consistent with what we see when we annotate cells manually.
+
+We can also compare to the existing annotations obtained using `SingleR` and `CellAssign` as part of `scpca-nf`.
 
@@
 
+This looks as we expect with mostly muscle cells lining up with tumor cells.
+
 ## Check for known CNVs in Ewing's
@@
 See [Tirode et al.,](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4264969/) and [Crompton et al.](https://doi.org/10.1158/2159-8290.CD-13-1037).
+
+Below we will plot whether or not the gain or loss is present in each cell according to `InferCNV`.
 
@@
 
+We can also look at the proportion of the chromosome with the gain.
+
 ```{r}
@@
 We see this for chr8 but not in chr12. 
+However, when you look at the heatmap you see that most of Chr8 is included in the gain, but only a small part of Chr12 has a gain, so these plots are not surprising.
 
-I think with the combination of marker gene expression and CNVs, the bottom group of cells represent the tumor cells.
-
-## Do the same thing but without a reference set of cells?
+I think with using the combination of marker gene expression and CNVs, the bottom group of cells represent the tumor cells.
 
 ## Conclusions
+
+- If we have a known group of normal cells, `InferCNV` can be helpful in identifying cells that could be tumor cells, e.g., cells that have a high number of CNVs. 
+- Using `InferCNV` without a group of normal cells does not seem feasible for this particular sample since all tumor cells appear to have the same set of CNVs with little variation. 
+- We can define tumor cells by setting a cutoff for the number of CNVs that must be present in a cell, but I'm not sure how transferrable this is to other samples. 
+- `InferCNV` can be helpful in looking at specific CNVs that are expected in a tumor type, something that is harder to look at in a tool like `CopyKAT`.
 

### 2024-05-13 analyses/ewings-cell-type/components/dependencies.R

@@ -1,2 +1,0 @@
-# R dependencies not captured by `renv`
-# library("missing_package")

### 2024-05-14 analyses/cell-type-ewings/template_notebooks/01-marker-genes.Rmd
add template notebooks

--- 2024-05-13 analyses/cell-type-ewings/01-marker-gene-tumor-classification.Rmd
@@
     toc_depth: 3
+    code_folding: "hide"
 params:
@@
   library_id: SCPCL000822
+  results_dir: "../results/annotate_tumor_cells_output"
+  reference_cell_file: "../references/cell_lists/SCPCS000490/SCPCL000822_reference-cells.tsv"
 ---
@@
 ## Setup
+
+```{r}
+project_root <- here::here()
+renv::load(project_root)
+```
 
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
+module_base <- file.path(repository_base, "analyses", "cell-type-ewings")
 ```
@@
 # output tumor/normal classifications
-results_dir <- file.path(module_base, "results", "marker_gene_analysis")
-fs::dir_create(results_dir)
+fs::dir_create(params$results_dir)
 
 classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
-output_classifications_file <- file.path(results_dir, classifications_filename)
+output_classifications_file <- file.path(params$results_dir, classifications_filename)
 ```
@@
   dplyr::pull(ensembl_gene_id)
+```
+
+```{r}
+# read in normal and tumor cells for later comparisons 
+if(file.exists(params$reference_cell_file)){
+  ref_cells_df <- readr::read_tsv(params$reference_cell_file)
+} else {
+  ref_cells_df <- NULL 
+}
 ```
@@
 
-Looking at this plot, there is definitely some variation in the expression of marker genes. 
-It looks like the highest gene expression is seen in the larger group of cells on the bottom right of the UMAP.
-
 We can also look at the distributions for each marker gene. 
-I would expect to see some sort of bimodal distribution separating cells that do and do not have expression of the marker gene.
+We would expect to see some sort of bimodal distribution separating cells that do and do not have expression of the marker gene.
 
@@
 
-Although we see some slight semblance of a bimodal distribution for most marker genes, it is hard to see and we cannot directly compare gene expression values across genes. 
-This would make it hard to identify a cut off to categorize cells as expression or not expressing the marker gene.
-
-Now we will transform each of the gene expression vectors by generating z-scores. 
-Then we might be able to find a cut off we can use across samples for if marker genes are present in a cell or not.
+Now we will transform each of the gene expression vectors by generating z-scores and use a cutoff of z-score > 0 to call tumor cells.
 
@@
 
-Now we can create the same density plot but looking at z-scores.
-
 ```{r}
@@
 
-It looks like some marker genes have distinct groups of cells with z-score > 0, while other marker genes may not be as informative. 
-Keep in mind this could vary somewhat across patients, so I'm wary to remove marker genes without looking across multiple references.
-
 ### Classify tumor cells using marker genes only
 
-Let's try and use the marker gene expression to classify tumor cells. 
-It looks like we could use a cutoff of z-score > 0 to count that cell as a tumor cell.
-
-We could either count any cell that expresses at least one marker gene > 0 as a tumor cell, or look at the combined expression. 
-Let's start with classifying tumor cells as tumor if any marker gene is present (z-score > 0). 
+To classify tumor cells, we use a cutoff of z-score > 0 to count that cell as a tumor cell. 
+We will call cells as tumor if any marker gene is present (z-score > 0). 
 This also means the sum of all z-scores > 0.
@@
 ```
-Similar to the individual plots, it looks like there is one group of cells on the bottom right that has the highest marker gene expression. 
-We would anticipate that these are most likely to be the tumor cells.
 
@@
 
-It looks like ADGRG2, HES1, NROB1, and STEAP1 don't show a lot of difference between cells that might be tumor vs. normal cells and are generally lowly expressed.
-CD99 appears to be expressed in both normal and tumor at similar levels. 
-All other genes show differences between the two, although at varying degrees. 
-Again, we should make sure this is consistent across multiple patients before removing these genes as marker genes.
-
 ### Marker genes using `findMarkers`
@@
 Below, we will use `scran::findMarkers()` to find any markers between tumor and normal cells. 
-Since we used marker genes to classify tumor cells, I would expect that at least some of the top markers would be from our list of tumor markers.
+Since we used marker genes to classify tumor cells, we expect that at least some of the top markers would be from our list of tumor markers.
 
@@
 
-We would expect that cells we are labeling as tumor cells have expression of the tumor marker genes. 
+We expect that cells we are labeling as tumor cells have expression of the tumor marker genes. 
 Cells labeled as normal should not have expression of the markers.
@@
 
-Looking at this might help give us an idea of genes that are more specifically expressed in tumor cells. 
-Genes like ADGRG2 and HES1 do not seem specific to tumor cells. 
-There are normal and tumor cells that express the gene and not very many tumor cells actually have expression of ADGRG2. 
-STEAP1 and NROB1 although not expressed in very many tumor cells, look to be specific to tumor cells which is promising.
-
-CD99, CAV1, and CCND1 are expressed in a high proportion of tumor cells but are also expressed in quite a few normal cells. 
-This is consistent with the literature for CD99 that it is expressed in normal cells, just at a lower level than in Ewing's cells.
-
-KDSR, MAPT, NKX2-2, PAPPA, and PRKCB all seem to have a high proportion of tumor cells expressing the gene and a low proportion of normal cells that express the gene. 
-PRKCB and MAPT are also 2 of the top genes identified using the marker analysis.
-
-Almost all genes have at least some expression in normal cells. 
-Perhaps those cells should also be categorized as tumor cells, or they just have some baseline expression in normal cells which is possible.
+Looking at this might help give us an idea of genes that are specifically expressed in only tumor cells.
 
@@
 
-The last thing we will do here is compare our manual annotations with the annotations using `SingleR` and `CellAssign`. 
-Looking at the annotations and UMAPs in the supplemental report, it looks like our group of "Tumor" cells are classified mostly as "Muscle cells" by both SingleR and CellAssign. 
-This makes sense since Ewing's sarcoma cells tend to resemble muscle cells.
-
-I would expect the Endothelial cells, Fibroblasts, and Macrophages assigned by `CellAssign` and `SingleR` to line up with the normal cells in this sample.
+The last thing we will do here is compare the manual annotations with the annotations from `SingleR` and `CellAssign`.
 
@@
 
-When looking at `SingleR`, we see that the tumor cells line up with smooth muscle cells, chondrocytes, neurons, and fibroblasts. 
-Ewing's cells have a neuronal like phenotype and express high levels of ECM genes, which may explain the overlap with chondrocytes and fibroblasts. 
-We do see that macrophages and endothelial cells tend to line up with the normal cells, which is expected.
+```{r}
+# plot comparing marker gene expression in cells annotated as "normal" and "tumor" by SingleR and CellAssign 
+# use already identified normal/tumor cell types 
+if(!is.null(ref_cells_df)){
 
-For CellAssign, all tumor cells are pulmonary vascular smooth muscle cells. 
-We also see tha normal cells contain some smooth muscle cells, but mostly are fibroblasts and endothelial cells.
+  # get list of normal cells 
+  normal_cells <- ref_cells_df |> 
+    dplyr::filter(reference_cell_class == "Normal") |> 
+    dplyr::pull(barcodes) |> 
+    unique()
 
-Looking at the heatmap in the supplemental report, the endothelial cells called by SingleR and CellAssign have a lot of overlap. 
-Let's look at expression of marker genes in these endothelial cells compared to cells that are categorized as smooth muscle cells.
+  # get list of tumor cells 
+  tumor_cells <- ref_cells_df |> 
+    dplyr::filter(reference_cell_class == "Tumor") |> 
+    dplyr::pull(barcodes) |> 
+    unique()
 
-```{r}
+  # only create this plot if normal and tumor cells are provided 
+  if(length(normal_cells) > 0 && length(tumor_cells) > 0){
+
+    # indicate cells that are part of normal and tumor lists
     select_celltype_df <- celltypes_df |> 
@@
         automated_annotation = dplyr::case_when(
-      singler_celltype_annotation == "smooth muscle cell" & cellassign_celltype_annotation == "Pulmonary vascular smooth muscle cells" ~ "Muscle cell",
-      singler_celltype_annotation == "endothelial cell" & cellassign_celltype_annotation == "Endothelial cells" ~ "Endothelial cell"
+          barcodes %in% normal_cells ~ "Normal cell",
+          barcodes %in% tumor_cells ~ "Tumor cell"
         )
@@
 
+    # join with umap df
     annotation_plot_df <- umap_df |> 
@@
       facet_wrap(vars(gene_symbol)) 
+  } 
+} 
 ```
-
-Consistent with previous findings, ADGRG2, HES1, NROB1, and STEAP1 don't seem to show any expression in either groups of cells. 
-CCND1 and CD99 appear to have similar expression across both groups of cells, while MAPT and PRKCB seem to have the most distinct expression profiles between muscle and endothelial cells.
-
-## Conclusions
-
-- We do see variation in marker gene expression across cells, suggesting that there are both tumor and normal cells here.
-- Using a z-score cut-off identifies a group of cells that appear to be tumor.
-.
-- The top marker genes for the "tumor" cells contain PRKCB and MAPT, two of the marker genes in our gene list. 
-- It's likely that ADGRG2, HES1, NROB1, and STEAP1 will not be informative in assigning tumor cells. 
-- We may be weary of genes like CD99 too as expression is more consistent across cells rather than specific to a group of cells. 
-- Endothelial cells as identified by SingleR and CellAssign generally have low expression of these marker genes and muscle cells have higher expression of marker genes.
 

### 2024-05-15 analyses/cell-type-ewings/04-infercnv.Rmd
update notebook to read in cnv metadata

@@
 )
+
+# with ref cnv metadata file 
+cnv_metadata_file <- file.path(results_dir, "endothelial_reference", glue::glue("{params$library_id}_cnv-metadata.tsv"))
 ```
@@
 sce <- readr::read_rds(sce_file)
+
+# read in cnv metadata 
+cnv_metadata_df <- readr::read_tsv(cnv_metadata_file)
 
@@
 
-The predictions information lies in a few files that live in the output directory. 
-We want to know which cells had CNVs and how many CNVs were found. 
-To map the information about CNVs to individual cells, they recommend using the `infercnv::add_to_seurat()` function which takes the output directory from running `InferCNV` and reads in all necessary files and adds the per cell CNV information to the `meta.data`.
-So we will start by converting our processed SCE object to a Seurat object, incorporate the CNV information, and then add the CNV information to the SCE object.
+Before we can make any plots we need to add some metadata that contains CNV information on a per cell level to the cell metadata.
 
 ```{r}
-infer_cnv_dir <- all_results_dir_list["with_ref"]
+# make df for plotting
+coldata_df <- sce |> 
+  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
+  # replace UMAP.1 with UMAP1
+  dplyr::rename_with(
+        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
+      ) |> 
+  # add in per cell cnv information 
+  dplyr::left_join(cnv_metadata_df, by = c("barcodes"))
 
-# first create a Seurat object from our SCE object
-seurat_obj <- scpcaTools::sce_to_seurat(sce)
-
-# add in the CNV information 
-# this adds new columns to the meta.data indicating presence of CNVs for each chromosome
-seurat_obj <- infercnv::add_to_seurat(seurat_obj = seurat_obj,
-                                      infercnv_output_path = infer_cnv_dir)
-
-# grab metadata from Seurat and add back to SCE 
-seurat_metadata <- seurat_obj@meta.data |> 
-  dplyr::select(-c(orig.ident, nCount_RNA, nFeature_RNA))
-
-colData(sce) <- DataFrame(seurat_metadata, row.names = seurat_metadata$barcodes)
+# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
+chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL000219.1"))
+cnv_df <- coldata_df[, chr_columns_to_keep]
 ```
@@
 Let's start by just summing up the total CNVs/cell and plotting that.
-
-```{r}
-# make df for plotting
-coldata_df <- sce |> 
-  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
-  # replace UMAP.1 with UMAP1
-  dplyr::rename_with(
-        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
-      )
-
-# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
-chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL000219.1"))
-cnv_df <- coldata_df[, chr_columns_to_keep]
-```
 

### 2024-05-15 analyses/cell-type-ewings/04-infercnv.Rmd
update png paths

@@
 
+# png ref files 
+png_files <- file.path(all_results_dir_list, glue::glue("{params$library_id}_infercnv.png")) |> 
+  purrr::set_names(names(all_results_dir_list))
+
 # with ref cnv metadata file 
@@
 
-![InferCNV Heatmap - Endothelial cells as reference](`r file.path(all_results_dir_list["with_ref"], "infercnv.png")`)
+![InferCNV Heatmap - Endothelial cells as reference](`r png_files["with_ref"]`)
 
@@
 
-![InferCNV Heatmap - No reference](`r file.path(all_results_dir_list["no_ref"], "infercnv.png")`)
+![InferCNV Heatmap - No reference](`r png_files["no_ref"]`)
 

### 2024-05-16 analyses/cell-type-ewings/template_notebooks/01-marker-genes.Rmd
update documentation

@@
 
-classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
+classifications_filename <- glue::glue("{params$library_id}_tumor-normal-classifications.tsv")
 output_classifications_file <- file.path(params$results_dir, classifications_filename)
@@
 
-### Compare with SingleR and CellAssign classifictions
+### Compare with SingleR and CellAssign classifications
 

### 2024-05-16 analyses/cell-type-ewings/annotate-tumor-cells-workflow.sh

@@ -12,26 +12,18 @@ Tumor cell annotations are obtained by:
 Before running this workflow, you must download both the processed `SingleCellExperiment` objects and `AnnData` objects with `download-data.py`.
 Do this for any samples you would like to run through this workflow.
 
-Then create a tsv file with the following columns for each sample that you would like to run through the worfklow:
+The following arguments are optional and can be used to run this workflow on additional samples (Default sample is "SCPCS000490"):
 
 sample_id: Unique sample ID (name of folder containing libray data)
-library_id: Unique library ID (prefix of data files)
 normal_celltypes: Comma separated list of cell types annotated with either `SingleR` or `CellAssign` to use as a reference list of normal cells.
 tumor_celltypes: Comma separated list of cell typs annotated with either `SingleR` or `CellAssign` that are expected to align with tumor cells.
 Any cell types used here will be used for comparing to tumor cells annotated in this workflow.
-If you do not wish to provide any tumor cell types, leave blank.
 
-The path to this file should be provided when running the workflow using the --sample_metadata argument.
+A set of reports summarizing the tumor cell annotations and a TSV file containing annotations from all used methods will be saved to `results/annotate_tumor_cells_output`.
 
-Optionally, you can include the `data_dir` argument. Use this if you want to use a different release other than `current` after downloading data.
+Example of running the workflow with a different sample:
 
-A set of reports summarizing the tumor cell annotations and a TSV file containing annotations from all used methods will be saved to the `results_dir`.
-The default `results_dir` is `results/annotate_tumor_cells_output`.
-
-Example of running the workflow:
-
-./annotate-tumor-cells-workflow.sh \
-  --sample_metadata "sample_metadata.tsv"
+sample_id="SCPCS000491" ./annotate-tumor-cells-workflow.sh
 
 '
 ####
@@ -67,6 +59,7 @@ for sce in $data_dir/$sample_id/*_processed.rds; do
 
     # define output directory for ref file
     cell_lists_dir="$ref_dir/cell_lists/$sample_id"
+    mkdir -p $cell_lists_dir
     # define output reference file
     reference_cell_file="$cell_lists_dir/${library_id}_reference-cells.tsv"
 

### 2024-05-16 analyses/cell-type-ewings/scripts/select-cell-types.R

@@ -56,10 +56,6 @@ stopifnot(
 # read in sce file
 sce <- readr::read_rds(opt$sce_file)
 
-# create output directory if not already present
-output_dir <- dir(opt$output_filename)
-fs::dir_create(output_dir)
-
 # Create table of reference cells ----------------------------------------------
 
 # possible annotation columns 

### 2024-05-17 analyses/cell-type-ewings/template_notebooks/03-copykat.Rmd

--- 2024-05-13 analyses/ewings-cell-type/03-copykat.Rmd
@@
 cnv_df <- cnv_df |> 
-  dplyr::left_join(full_cnv_df, by = c("barcodes", "reference_used"))
+  dplyr::left_join(full_cnv_df, by = c("barcodes", "reference_used")) |> 
+  dplyr::filter(!is.na(chrom))
 ```

### 2024-05-17 analyses/cell-type-ewings/template_notebooks/02-cellassign.Rmd
clean up notebook to remove conclusions and add params

--- 2024-05-13 analyses/cell-type-ewings/02-cellassign-tumor-cell-classification.Rmd
@@
   library_id: SCPCL000822
+  tumor_markers_file: "../references/tumor-marker-genes.tsv"
+  visser_markers_file: "../references/visser-all-marker-genes.tsv"
+  marker_gene_classification: "../results/annotate_tumor_cells_output/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
+  tumor_marker_predictions: "../results/annotate_tumor_cells_output/SCPCS000490/cellassign/SCPCL000822_tumor-only-predictions.tsv"
+  visser_marker_predictions: "../results/annotate_tumor_cells_output/SCPCS000490/cellassign/SCPCL000822_visser-predictions.tsv"
+  panglao_predictions: "../results/annotate_tumor_cells_output/SCPCS000490/cellassign/SCPCL000822_panglao-predictions.tsv"
+  results_dir: "../results/annotate_tumor_cells_output/SCPCS000490"
 ---
@@
 - A full list of tumor marker genes only. 
-- A filtered list of marker genes. 
 - Tumor marker genes and non-tumor marker genes from [Visser et al.](https://doi.org/10.1158/2767-9764.CRC-23-0027) specified in the supplemental methods.
-- Tumor marker genes and marker genes from endothelial cells and fibroblast marker genes in PanglaoDB.
 
-We then compare the results from running `CellAssign` with each reference to the tumor/normal classification obtained in `01-marker-gene-tumor-classification.Rmd`. 
+We then compare the results from running `CellAssign` with each reference to the tumor/normal classifications obtained from looking only at marker genes. 
 We also look at marker gene expression in the assigned cells.
@@
 
-anndata_filename <- glue::glue("{params$library_id}_processed_rna.h5ad")
-anndata_file <- file.path(sample_dir, anndata_filename)
-
-# tumor normal classifications 
-classifications_results_dir <- file.path(module_base, "results", "marker_gene_analysis")
-classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
-classifications_file <- file.path(classifications_results_dir, classifications_filename)
-
-# tumor marker genes
-tumor_markers_file <- file.path(module_base, "references", "tumor-marker-genes.tsv")
-# all markers 
-all_markers_file <- file.path(module_base, "references", "all-marker-genes.tsv")
-# panglao markers
-panglao_markers_file <- file.path(module_base, "references", "PanglaoDB_markers_2020-03-27.tsv")
-
-# Output reference matrix files
-ref_dir <- file.path(module_base, "references", "cellassign_refs")
-
-tumor_markers_mtx_file <- file.path(ref_dir, "tumor-marker-cellassign.tsv")
-filtered_markers_mtx_file <- file.path(ref_dir, "filtered-tumor-marker-cellassign.tsv")
-visser_markers_mtx_file <- file.path(ref_dir, "visser-all-marker-cellassign.tsv")
-panglao_markers_mtx_file <- file.path(ref_dir, "panglao-endo-fibro-cellassign.tsv")
-
-# output predictions from cellassign 
-preds_results_dir <- file.path(module_base, "results", "cellassign_predictions")
-fs::dir_create(preds_results_dir)
-
-cellassign_predictions_file <- file.path(preds_results_dir, "cellassign-tumor-predictions.tsv")
-filtered_predictions_file <- file.path(preds_results_dir, "filtered-cellassign-tumor-predictions.tsv")
-visser_predictions_file <- file.path(preds_results_dir, "visser-cellassign-predictions.tsv")
-panglao_predictions_file <- file.path(preds_results_dir, "panglao-endo-fibro-predictions.tsv")
+# output annotations
+cellassign_annotations_file <- file.path(params$results_dir, glue::glue("{params$library_id}_cellassign-annotations.tsv"))
 ```
@@
 ```{r}
-# function for creating a binary matrix with cell types as columns and marker genes as rows 
-build_binary_df <- function(marker_genes_df,
-                            gene_col, 
-                            cell_type_col) {
-
-  binary_df <- marker_genes_df |>
-    dplyr::select({{cell_type_col}}, {{gene_col}}) |> 
-    unique() |> 
-    tidyr::pivot_wider(
-      id_cols = {{gene_col}},
-      names_from = {{cell_type_col}},
-      values_from = {{cell_type_col}},
-      values_fn = length,
-      values_fill = 0
-    ) |>
-    # add a column with no marker genes
-    # cell assign will assign cells to "other" when no other cell types are appropriate
-    dplyr::mutate(other = 0)
-
-  return(binary_df)
-}
-
 # function to get assigned cell type based on max prediction returned by CellAssign 
@@
 # read in tumor marker genes table 
-tumor_markers_df <- readr::read_tsv(tumor_markers_file) |> 
+tumor_markers_df <- readr::read_tsv(params$tumor_markers_file) |> 
   # account for genes being from multiple sources
@@
 # read in all marker genes 
-all_markers_df <- readr::read_tsv(all_markers_file) |> 
+visser_markers_df <- readr::read_tsv(params$visser_markers_file) |> 
   # account for genes being from multiple sources
@@
 # obtained from manual classification in `01-marker-gene-tumor-classifications.Rmd`
-classifications_df <- readr::read_tsv(classifications_file)
+classifications_df <- readr::read_tsv(params$marker_gene_classification)
+tumor_predictions <- readr::read_tsv(params$tumor_marker_predictions)
+visser_predictions <- readr::read_tsv(params$visser_marker_predictions)
 ```
@@
 
-We need to create a binary matrix with two columns, "tumor" and "other". 
-The "other" will refer to any normal cells.
-
-```{r}
-tumor_ref_mtx <- build_binary_df(tumor_markers_df, 
-                                  gene_col = "ensembl_gene_id",
-                                  cell_type_col = "cell_type")
-
-# export matrix 
-readr::write_tsv(tumor_ref_mtx, tumor_markers_mtx_file)
-```
-
-Now run `CellAssign` with the full marker gene list as the reference.
-
-```{r, eval=FALSE}
-cellassign_call <- glue::glue(
-  "python scripts/run-cellassign.py",
-  " --anndata_file '{anndata_file}'",
-  " --output_predictions '{cellassign_predictions_file}'",
-  " --reference '{tumor_markers_mtx_file}'"
-)
-
-system(cellassign_call)
-```
-
-We can then read in the predictions file and find the predicted cell type for each cell, either "tumor" or "other". 
+We will read in the predictions file and find the predicted cell type for each cell, either "tumor" or "other". 
 Then we'll show a heatmap of the scores for each cell and cell type. 
@@
 ```{r}
-# read in predictions file
-tumor_predictions <- readr::read_tsv(cellassign_predictions_file)
-
 # get cell type assignments 
@@
 
-It looks like we might have pretty good agreement between the annotations from `CellAssign` and manual annotations. 
 Let's also look at the cell annotations on a UMAP and then look at the confusion matrix.
@@
 
-This looks pretty good! 
-There is quite a bit of agreement between classifying cells manually with marker genes vs. classifying cells with `CellAssign` and using the marker genes.
-
 To confirm that the cells with tumor cells have higher expression of marker genes, let's look at the mean expression of marker genes between tumor and other cells as classified by `CellAssign`.
@@
 
-As expected, we see that cells classified as tumor cells have higher expression of the tumor marker genes.
-
-### Run CellAssign with a filtered set of tumor marker genes
-
-In looking at the UMAPs of all marker genes (see `01-marker-gene-tumor-classification.Rmd`) there are quite a few genes that have very low expression across all cells, so perhaps they are adding noise to the model.
-Although this is a bit circular, we could restrict our list of marker genes to be only those that have a high mean gene expression in our dataset? 
-I would expect that when we do that, we are able to identify more tumor cells that line up with what we would expect based on gene expression patterns.
-
-```{r}
-# grab the mean gene expression for all marker genes 
-rowdata_df <- rowData(sce) |> 
-  as.data.frame() |> 
-  dplyr::filter(gene_symbol %in% tumor_markers_df$gene_symbol)
-rowdata_df
-```
-
-```{r}
-# what's the mean gene expression of all genes in this sample for reference
-mean(rowData(sce)$mean)
-```
-
-There are a few genes that have low gene expression < 1. 
-Let's try and filter out those genes and then use `CellAssign`.
-
-```{r}
-# get list of marker genes to keep
-filtered_markers <- rowdata_df |> 
-  dplyr::filter(mean > 1) |> 
-  dplyr::pull(gene_ids)
-
-# filter binary mtx 
-filtered_ref_mtx <- tumor_ref_mtx |> 
-  dplyr::filter(ensembl_gene_id %in% filtered_markers)
-
-readr::write_tsv(filtered_ref_mtx, filtered_markers_mtx_file)
-```
-
-```{r, eval=FALSE}
-cellassign_call <- glue::glue(
-  "python scripts/run-cellassign.py",
-  " --anndata_file '{anndata_file}'",
-  " --output_predictions '{filtered_predictions_file}'",
-  " --reference '{filtered_markers_mtx_file}'"
-)
-
-system(cellassign_call)
-```
-
-Now we can read in the predictions for the filtered marker gene list.
-
-```{r}
-filtered_predictions <- readr::read_tsv(filtered_predictions_file)
-
-filtered_celltype_assignments <- get_celltype_assignment(filtered_predictions)
-
-table(filtered_celltype_assignments$celltype)
-```
-
-```{r}
-# heatmap of prediction scores 
-filtered_predictions |>
-  tibble::column_to_rownames("barcode") |> 
-  as.matrix() |> 
-  pheatmap::pheatmap(show_rownames = FALSE,
-                     annotation_row = annotation_df)
-```
-
-Interestingly, `CellAssign` doesn't seem to like using such a limited list of marker genes and assigns all cells as "other".
-
 ### Run CellAssign with tumor markers and all marker genes from Visser et al.
@@
 Below we will use `CellAssign` but with all tumor marker genes and all marker genes in Visser et al. for non-tumor cells.
-
-```{r}
-# build the reference matrix 
-visser_ref_mtx <- build_binary_df(all_markers_df, 
-                                   gene_col = "ensembl_gene_id",
-                                   cell_type_col = "cell_type")
-
-# export matrix 
-readr::write_tsv(visser_ref_mtx, visser_markers_mtx_file)
-```
-
-```{r, eval=FALSE}
-cellassign_call <- glue::glue(
-  "python scripts/run-cellassign.py",
-  " --anndata_file '{anndata_file}'",
-  " --output_predictions '{visser_predictions_file}'",
-  " --reference '{visser_markers_mtx_file}'"
-)
-
-system(cellassign_call)
-```
 
@@
 ```
-
-Interestingly when we use this reference we actually see that most cells are classified as other, with some cells being classified into the different normal cell types.
 
@@
 
-Again, we don't see much overlap of cells that we are classifying as tumor manually to the tumor cells being classified by CellAssign. 
-We do see that the majority of the immune and mesenchymal-like cells fall into the "normal" category.
-
 The one cell type that is identified here and in the `CellAssign` annotations with the reference from `scpca-nf` is Endothelial cells. 
@@
 ```{r}
+# first make sure cells have been classified as endothelial 
+has_endothelial <- "Endothelial cells" %in% unique(celltype_df$cellassign_celltype_annotation) &&
+  "endothelial cells" %in% unique(celltype_df$visser_cellassign_annotations)
+
+# if present, compare across both references 
+if(has_endothelial){
   # prep annotations for confusion matrix
@@
   ) 
+} else {
+  message("No endothelial cells were classified by `CellAssign`.")
+}
 ```
-
-The good news is although tumor cells aren't getting classified, the same group of cells is getting classified as Endothelial cells, even with the smaller list of marker genes.
 
@@
 # get all markers 
-all_markers <- all_markers_df |> 
+all_markers <- visser_markers_df |> 
   dplyr::pull(ensembl_gene_id)
@@
 
-In comparing these UMAPs to the UMAP showing the cell type annotations with the Visser reference, it does appear that Endothelial cells are expressing the marker genes and that the MSC-like cells are expressing the expected marker genes.
-
-Below we look at the expression of marker genes for each cell type across all cells identified to be each type by CellAssign. 
+Below we look at the expression of marker genes for each cell type across all cells identified to be each type by `CellAssign.` 
 We would expect to see expression of the marker genes for tumor cells to be higher in the cells that are identified to be tumor cells compared to other cell types.
@@
 
-It looks like the "other" cells actually have high expression of the tumor marker genes and instead should be classified as tumor cells.
-
 ### Run CellAssign with markers from PanglaoDB
 
-The last thing I'm trying here is to run CellAssign using some of the same reference that we used when running CellAssign as part of scpca-nf. 
+The last thing we will do here is run CellAssign using some of the same reference that we used when running CellAssign as part of scpca-nf. 
 In particular, the majority of non-muscle cells (which we believe to be tumor cells) were identified as either Endothelial cells or Fibroblasts.
-So we will create a reference that takes the markers for Endothelial cells and Fibroblasts from PanglaoDB and combines with the tumor markers.
-
-```{r}
-# get just the genes for endothelial and fibroblasts from panglao 
-panglao_markers_df <- readr::read_tsv(panglao_markers_file) |> 
-  dplyr::filter(`cell type` %in% c("Endothelial cells", "Fibroblasts")) |> 
-  dplyr::select(`official gene symbol`, `cell type`)
-
-# get the ensembl id from rowdata for each panglao marker 
-panglao_tumor_markers_df <- rowData(sce) |> 
-  as.data.frame() |> 
-  dplyr::filter(gene_symbol %in% panglao_markers_df$`official gene symbol`) |> 
-  dplyr::left_join(panglao_markers_df, by = c("gene_symbol" = "official gene symbol")) |> 
-  dplyr::select(gene_symbol, ensembl_gene_id = gene_ids, cell_type = `cell type`) |> 
-  # add in the tumor markers 
-  dplyr::bind_rows(tumor_markers_df)
-
-```
-
-```{r}
-# build the reference matrix 
-panglao_tumor_ref_mtx <- build_binary_df(panglao_tumor_markers_df, 
-                                          gene_col = "ensembl_gene_id",
-                                          cell_type_col = "cell_type")
-
-# remove any genes that may be present in multiple cell types 
-rowsums <- panglao_tumor_ref_mtx |> 
-  dplyr::select(-ensembl_gene_id) |> 
-  rowSums()
-
-panglao_tumor_ref_mtx <- panglao_tumor_ref_mtx[-which(rowsums > 1),]
-
-# export matrix 
-readr::write_tsv(panglao_tumor_ref_mtx, panglao_markers_mtx_file)
-```
-
-```{r, eval=FALSE}
-cellassign_call <- glue::glue(
-  "python scripts/run-cellassign.py",
-  " --anndata_file '{anndata_file}'",
-  " --output_predictions '{panglao_predictions_file}'",
-  " --reference '{panglao_markers_mtx_file}'"
-)
-
-system(cellassign_call)
-```
+So we will use a reference that takes the markers for Endothelial cells and Fibroblasts from PanglaoDB and combines with the tumor markers.
 
@@
 # read in predictions file
-panglao_predictions <- readr::read_tsv(panglao_predictions_file)
+panglao_predictions <- readr::read_tsv(params$panglao_predictions)
 
@@
 
-Although only 2 cells are classified as Fibroblasts, we do see a group of Endothelial cells and then tumor cells that line up with our manual annotations of tumor/normal.
-
 ```{r}
@@
 ```
-
-The UMAP tells me that we are classifying the same group of endothelial cells as we did in `scpca-nf` with `CellAssign` and that the tumor cells line up with the manual annotations of tumor cells (muscle cells in `CellAssign`).
 
@@
 
-It looks like combining some of the marker genes from PanglaoDB with marker genes does a good job of capturing the endothelial cells and tumor cells. 
-It doesn't capture the Fibroblasts, but instead classifies those as other.
-
-If we want to combine annotation of both tumor cells and normal cells, we may consider using the full PanglaoDB marker gene references we have and adding tumor marker genes to that list. 
-However, this will be more computationally expensive since those references are quite large.
-
-## Conclusions
-
-- Using `CellAssign` with a list of tumor markers identifies tumor cells similar to manual annotations. 
-- Using `CellAssign` with a restricted list of marker genes doesn't identify any tumor cells. 
-- Using `CellAssign` with an expanded list of marker genes from Visser et al., identifies some normal cell types, but struggles to identify any tumor cells. 
-This doesn't match up with what's expected based on marker gene expression in the sample. 
-- Using `CellAssign` with the Panglao markers and tumor markers identifies endothelial cells and tumor cells similar to what we would expect.
-
 ## Session Info

### 2024-05-20 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd
move notebook to exploratory analysis

--- 2024-05-15 analyses/cell-type-ewings/04-infercnv.Rmd
@@
 
-all_results_dir_list <- c(
+all_results_dir_list <- list(
   no_ref = file.path(results_dir, "no_reference"),
@@
 cnv_df <- cnv_df |>
-  # get total number of cnvs per cell and then scale that number 
+  # get total number of chromosomes with cnvs per cell and then scale that number 
   dplyr::mutate(has_cnv = rowSums(across(has_cnv_cols)),
@@
   theme_bw() +
-  labs(x = "Scaled number of CNVS")
+  labs(x = "z-scaled chromosome CNV count")
 ```

### 2024-05-20 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd
scale proportion by number of genes

@@
 # png ref files 
-png_files <- file.path(all_results_dir_list, glue::glue("{params$library_id}_infercnv.png")) |> 
-  purrr::set_names(names(all_results_dir_list))
+png_files <- all_results_dir_list |>
+  purrr::map_chr(\(dir) file.path(dir, glue::glue("{params$library_id}_infercnv.png"))) |> 
+  # make sure png files exists
+  purrr::map(\(file)
+    stopifnot("Missing PNG file" = file.exists(file))
+  )
 
@@
 cnv_metadata_file <- file.path(results_dir, "endothelial_reference", glue::glue("{params$library_id}_cnv-metadata.tsv"))
+
+# full infercnv object with information on genes/ chr to use later 
+final_cnv_obj_file <- file.path(results_dir, "endothelial_reference", glue::glue("{params$library_id}_cnv-obj.rds"))
 ```
@@
 # read in cnv metadata 
-cnv_metadata_df <- readr::read_tsv(cnv_metadata_file)
+# use read.delim instead of readr because we want the rownames to be read in properly 
+cnv_metadata_df <- read.delim(cnv_metadata_file) |> 
+  tibble::rownames_to_column("barcodes")
 
@@
 manual_classifications_df <- readr::read_tsv(classifications_file)
+
+# read in cnv object 
+final_cnv_obj <- readr::read_rds(final_cnv_obj_file)
 ```
@@
 The color indicates the relative expression intensity of each gene with the darker colors either indicating a gain of expression or loss of expression compared to the reference cells.
+
+From the [`InferCNV` documentation](https://github.com/broadinstitute/inferCNV/wiki/Interpreting-the-figure):
+
+> The normal cells in the top heatmap define baseline expression for genes in normal cells.
+This baseline distribution of normal gene expression is subtracted from both the normal cells from which was defined as well as the tumor cells.
+Afterwards, the normal cell expression heatmap should be largely devoid of signal, with the exception of certain outlier gene expression values in certain cells.
+Removing this baseline of normal expression signal from the tumor cells should reveal those chromosomal regions that have significantly more or less expression than the normal cells, highlighting likely amplified or deleted whole chromosomes or large chromosomal regions.
+
+The dendogram on the left side is hierarchical clustering of all non-reference cells. 
+The color annotation next to it indicates subdivisions of the dendogram by cutting it in `k_obs_groups` groups, with the default `k_obs_groups` = 1.
 
@@
 # remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
-chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL000219.1"))
-cnv_df <- coldata_df[, chr_columns_to_keep]
+chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL000219.1|chrGL000194.1"))
+all_cnv_df <- coldata_df[, chr_columns_to_keep]
 ```
@@
 One approach to classifying tumor cells is to look for cells that have a lot of CNVs. 
-Let's start by just summing up the total CNVs/cell and plotting that.
+Let's start by just summing up the total number of chromosomes with CNVs per cell and plotting that.
 
@@
 # grab any columns that have has_cnv 
-has_cnv_cols <- names(cnv_df)[startsWith(colnames(cnv_df), "has_cnv")]
+has_cnv_cols <- names(all_cnv_df)[startsWith(colnames(all_cnv_df), "has_cnv")]
 
-cnv_df <- cnv_df |>
+reference_cells_only <- all_cnv_df |> 
+  dplyr::filter(stringr::str_detect(subcluster, "reference")) |> 
+  dplyr::mutate(has_cnv = NA,
+                z_scaled_cnv = NA)
+
+unknown_cells_only <- all_cnv_df |>
+  dplyr::filter(stringr::str_detect(subcluster, "unknown")) |> 
   # get total number of chromosomes with cnvs per cell and then scale that number 
   dplyr::mutate(has_cnv = rowSums(across(has_cnv_cols)),
-                scaled_cnv = scale(has_cnv))
+                z_scaled_cnv = scale(has_cnv))
+
+# combine back together for plotting 
+cnv_df <- bind_rows(reference_cells_only, unknown_cells_only)
 ```
@@
 This shows that the total CNVs are highest in the group of cells that correspond to the same cells that are manually annotated as tumor cells, which is promising. 
+Those cells in grey (bottom left) are cells used as a reference. 
 If we were to use total CNVs we would need some sort of cut off for the number. 
@@
 
-ggplot(cnv_df, aes(x= scaled_cnv)) + 
+ggplot(cnv_df, aes(x= z_scaled_cnv)) + 
   geom_density() +
@@
 cnv_df <- cnv_df |> 
-  dplyr::mutate(cnv_classification = dplyr::if_else(scaled_cnv > 0, "Tumor", "Normal"))
+  dplyr::mutate(cnv_classification = dplyr::if_else(z_scaled_cnv > 0, "Tumor", "Normal"))
 
@@
 
-# get mean proportion of cnvs per cell and then scale that number 
+# get total number of genes per chromosome to use in weighted mean 
+chr_weights <- final_cnv_obj@gene_order |> 
+  as.data.frame() |> 
+  dplyr::count(chr) |> 
+  # only keep chr 1-22, drops MT and GL chr
+  dplyr::filter(chr %in% glue::glue("chr{seq(1,22)}")) |> 
+  dplyr::pull(n)
+
+# get mean proportion of cnvs per cell weighted by chromsome size (number of genes)
 cnv_df <- cnv_df |> 
-  dplyr::mutate(mean_proportion = rowMeans(across(proportion_cols)),
-                scaled_mean_proportion = scale(mean_proportion))
+  dplyr::rowwise() |> 
+  # calculate weighted mean, excluding normal cells 
+  dplyr::mutate(mean_proportion = ifelse(!is.na(has_cnv), weighted.mean(across(proportion_cols), num_genes_df$n), NA)) |> 
+  dplyr::ungroup()
 ```
@@
   scale_color_viridis_c()
-
-ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = scaled_mean_proportion)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  theme_bw() +
-  scale_color_viridis_c()
 ```
@@
 ```{r}
-ggplot(cnv_df, aes(x = scaled_mean_proportion)) +
+ggplot(cnv_df, aes(x = mean_proportion)) +
   geom_density() +
@@
 ```
-It looks like maybe we could again use 0 as the cutoff here and anything above 0 is a tumor cell. 
+It looks like we could use a cutoff of 0.1 here to classify tumor/normal cells. 
 Let's see how this matches up with the tumor/normal classifications we defined just using total CNVs.
@@
 ```{r}
-ggplot(cnv_df, aes(x = scaled_mean_proportion, color = cnv_classification)) +
+ggplot(cnv_df, aes(x = mean_proportion, color = cnv_classification)) +
   geom_density() +
@@
 
-Interestingly, the cells we labeled as tumor have a higher mean proportion value then those that we labeled as normal. 
+As expected, the cells we labeled as tumor have a higher mean proportion value then those that we labeled as normal. 
 I'm not really sure which is "better" to use the mean proportion or just the detection of CNVs to call tumor cells.
 
-For the rest of the notebook, we will use the classifications of tumor cells based on the sum of the total CNV occurrences.
+```{r}
+# add cnv classification 
+cnv_df <- cnv_df |> 
+  dplyr::mutate(cnv_mean_classification = dplyr::if_else(mean_proportion > 0.1, "Tumor", "Normal"))
+
+ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = cnv_mean_classification)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  theme_bw() 
+```
+
+For the rest of the notebook, we will use the classifications of tumor cells based on the mean proportion of total CNV occurrences.
 
@@
   dplyr::mutate(
-    cnv_classification = forcats::fct_relevel(cnv_classification, "Tumor"),
+    cnv_classification = forcats::fct_relevel(cnv_mean_classification, "Tumor"),
     marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
@@
       cnv_df,
-      "cnv_classification",
+      "cnv_mean_classification",
       name

### 2024-05-20 analyses/cell-type-ewings/scripts/run-infercnv.R

@@ -86,6 +86,10 @@ fs::dir_create(scratch_dir)
 # define output metadata file 
 cnv_metadata_file <- file.path(output_dir, glue::glue("{library_id}_cnv-metadata.tsv"))
 
+# define output infercnv obj
+scratch_obj_file <- file.;path(scratch_dir, "run.final.infercnv_obj")
+output_obj_file <- file.path(output_dir, glue::glue("{library_id}_cnv-obj.rds"))
+
 # png file to save 
 scratch_png <- file.path(scratch_dir, "infercnv.png")
 output_png <- file.path(output_dir, glue::glue("{library_id}_infercnv.png"))

### 2024-05-20 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd
add some more comments to main exploratory notebook

@@
 
+# don't include the normal cells that were provided as reference in scaling 
 reference_cells_only <- all_cnv_df |> 
@@
 Looking at the scaled number of CNVs it does look like we could use 0 as the cutoff and call any cell that has > 0 CNVs (after scaling) as tumor and all other cells as normal.
+
+In the below plot we label all non-reference cells as either tumor or normal. 
+Cells that are NA are those that were already classified as reference cells.
 

### 2024-05-20 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd
make simplifed copy of infercnv notebook

--- 2024-05-20 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd
@@
   library_id: SCPCL000822
+  marker_genes_classification_file: "../results/annotate_tumor_cells_output/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
+  infercnv_dir: "../results/annotate_tumor_cells_output/SCPCS000490/infercnv/endothelial_reference"
+  results_dir: "../results/annotate_tumor_cells_output/SCPCS000490"
 ---
@@
 
-# tumor normal classifications 
-classifications_results_dir <- file.path(module_base, "results", "marker_gene_analysis")
-classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
-classifications_file <- file.path(classifications_results_dir, classifications_filename)
-
-# output from running infercnv 
-results_dir <- file.path(module_base, "results", "infercnv", params$library_id)
-
-all_results_dir_list <- list(
-  no_ref = file.path(results_dir, "no_reference"),
-  with_ref = file.path(results_dir, "endothelial_reference")
-)
-
 # png ref files 
-png_files <- all_results_dir_list |>
-  purrr::map_chr(\(dir) file.path(dir, glue::glue("{params$library_id}_infercnv.png"))) |> 
-  # make sure png files exists
-  purrr::map(\(file)
-    stopifnot("Missing PNG file" = file.exists(file))
-  )
+png_file <-file.path(params$infercnv_dir, glue::glue("{params$library_id}_infercnv.png"))
 
 # with ref cnv metadata file 
-cnv_metadata_file <- file.path(results_dir, "endothelial_reference", glue::glue("{params$library_id}_cnv-metadata.tsv"))
+cnv_metadata_file <- file.path(params$infercnv_dir, "endothelial_reference", glue::glue("{params$library_id}_cnv-metadata.tsv"))
 
 # full infercnv object with information on genes/ chr to use later 
-final_cnv_obj_file <- file.path(results_dir, "endothelial_reference", glue::glue("{params$library_id}_cnv-obj.rds"))
+final_cnv_obj_file <- file.path(params$infercnv_dir, "endothelial_reference", glue::glue("{params$library_id}_cnv-obj.rds"))
+
+# output classifications file 
+infercnv_results_file <- file.path(params$results_dir, glue::glue("{params$library_id}_infercnv-classifications.tsv"))
+```
+
+```{r}
+# make sure files exist
+stopifnot(
+  "SCE file is missing" = file.exists(sce_file),
+  "PNG file with heatmap is missing" = file.exists(png_file),
+  "CNV metadata file is missing" = file.exists(cnv_metadata_file),
+  "Final InferCNV object is missing" = file.exists(final_cnv_obj_file),
+  "Marker gene classifications are missing" = file.exists(params$marker_genes_classification_file)
+)
 ```
@@
 # read in tumor normal classifications 
-manual_classifications_df <- readr::read_tsv(classifications_file)
+manual_classifications_df <- readr::read_tsv(params$marker_genes_classification_file)
 
@@
 
-Below is the heatmap produced by `InferCNV` showing the CNV's for each cell using endothelial cells as a reference. 
+Below is the heatmap produced by `InferCNV` showing the CNV's for each cell using normal cells as a reference. 
 Each row is a cell and each column is a gene. 
@@
 
-![InferCNV Heatmap - Endothelial cells as reference](`r png_files["with_ref"]`)
-
-From this heatmap we see that there is a distinct group of cells within the bottom portion of the plot that appear to have a lot more CNVs than the rest of the cells. 
-It's probable that these correspond to the tumor cells, but `InferCNV` does not provide any classification of cells as tumor or normal, it just provides information on copy number variations.
-
-### No reference
-
-Below is the heatmap produced by `InferCNV` showing the CNV's for each cell without using a reference.
-
-From the [`InferCNV` documentation](https://github.com/broadinstitute/inferCNV/wiki/Running-InferCNV):
-
-> Note, if you do not have reference cells, you can set ref_group_names=NULL, in which case the average signal across all cells will be used to define the baseline.
-This can work well when there are sufficient differences among the cells included (ie.
-they do not all show a chromosomal deletion at the same place).
-
-Given this information, I wouldn't expect the CNV inference to work well because we see that all cells that do have prominent CNVs have the same CNVs. 
-There is not a lot of variation.
-
-![InferCNV Heatmap - No reference](`r png_files["no_ref"]`)
-
-Looking at this we see a small group of cells that may have some variations, but they are different from the variations seen above. 
-I don't know that using this without a reference is the best call since it's unlikely that it will identify distinct groups of cells from tumors with quiet genomes with no baseline. 
-The rest of the analysis will use the results from running `InferCNV` with endothelial cells as the reference.
+![InferCNV Heatmap - Endothelial cells as reference](`r png_file`)
 
@@
 ```
-This shows that the total CNVs are highest in the group of cells that correspond to the same cells that are manually annotated as tumor cells, which is promising. 
+
 Those cells in grey (bottom left) are cells used as a reference. 
@@
 
-Looking at the scaled number of CNVs it does look like we could use 0 as the cutoff and call any cell that has > 0 CNVs (after scaling) as tumor and all other cells as normal.
+Looking at the scaled number of CNVs we will use 0 as the cutoff and call any cell that has > 0 CNVs (after scaling) as tumor and all other cells as normal.
 
@@
 ```
-It looks like we could use a cutoff of 0.1 here to classify tumor/normal cells. 
+
+We will use a cutoff of 0.1 here to classify tumor/normal cells. 
 Let's see how this matches up with the tumor/normal classifications we defined just using total CNVs.
+
+In the below plot we label all non-reference cells as either tumor or normal. 
+Cells that are NA are those that were already classified as reference cells.
 
@@
 ```
-
-As expected, the cells we labeled as tumor have a higher mean proportion value then those that we labeled as normal. 
-I'm not really sure which is "better" to use the mean proportion or just the detection of CNVs to call tumor cells.
 
@@
 ```
-
-This looks pretty consistent with what we see when we annotate cells manually.
 
@@
 
-This looks as we expect with mostly muscle cells lining up with tumor cells.
-
 ## Check for known CNVs in Ewing's
@@
 
-From this we see that the group of tumor cells at the bottom have chr8 and chr1 gains with some having gains of chr12. 
-It doesn't look like any cells have loss of chr16. 
-The other interesting thing is that there is a group of cells that have gains of chr8, chr1, and chr12 that we did not classify as tumor cells. Should they be...
-
 We can also look at the proportion of the chromosome with the gain.
@@
 
-Now when we look at the proportion, we see that only the group of tumor cells on the bottom seem to have a higher proportion of genes in chr8 that have a gain in expression. 
-Based on the papers, we should see a gain in the whole chr8 and chr12, not just one arm or specific genes, so we expect a high value for the proportion. 
-We see this for chr8 but not in chr12. 
-However, when you look at the heatmap you see that most of Chr8 is included in the gain, but only a small part of Chr12 has a gain, so these plots are not surprising.
+## Export results
 
-I think with using the combination of marker gene expression and CNVs, the bottom group of cells represent the tumor cells.
+We will save the tumor/normal classifications using both the sum and proportion.
 
-## Conclusions
+```{r}
+# get a dataframe with barcodes, sum_classification and cluster_classification 
+celltypes_df <- cnv_df |> 
+  dplyr::select(
+    barcodes,
+    cnv_proportion_classification = cnv_mean_classification,
+    cnv_sum_classification = cnv_classification
+  )
 
-- If we have a known group of normal cells, `InferCNV` can be helpful in identifying cells that could be tumor cells, e.g., cells that have a high number of CNVs. 
-- Using `InferCNV` without a group of normal cells does not seem feasible for this particular sample since all tumor cells appear to have the same set of CNVs with little variation. 
-- We can define tumor cells by setting a cutoff for the number of CNVs that must be present in a cell, but I'm not sure how transferrable this is to other samples. 
-- `InferCNV` can be helpful in looking at specific CNVs that are expected in a tumor type, something that is harder to look at in a tool like `CopyKAT`.
+readr::write_tsv(celltypes_df, infercnv_results_file)
+```
 

### 2024-05-20 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd
don't create null pngs

@@
 png_files <- all_results_dir_list |>
-  purrr::map_chr(\(dir) file.path(dir, glue::glue("{params$library_id}_infercnv.png"))) |> 
+  purrr::map_chr(\(dir) file.path(dir, glue::glue("{params$library_id}_infercnv.png")))
+
+png_files |> 
   # make sure png files exists

### 2024-05-21 analyses/cell-type-ewings/template_notebooks/02-cellassign.Rmd
fix path errors in cellassign notebook

@@
   library_id: SCPCL000822
-  tumor_markers_file: "../references/tumor-marker-genes.tsv"
-  visser_markers_file: "../references/visser-all-marker-genes.tsv"
   marker_gene_classification: "../results/annotate_tumor_cells_output/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
@@
 
+tumor_markers_file <- file.path(module_base, "references", "tumor-marker-genes.tsv")
+visser_markers_file <- file.path(module_base, "references", "visser-all-marker-genes.tsv")
+
 # output annotations
@@
 # read in tumor marker genes table 
-tumor_markers_df <- readr::read_tsv(params$tumor_markers_file) |> 
+tumor_markers_df <- readr::read_tsv(tumor_markers_file) |> 
   # account for genes being from multiple sources
@@
 # read in all marker genes 
-visser_markers_df <- readr::read_tsv(params$visser_markers_file) |> 
+visser_markers_df <- readr::read_tsv(visser_markers_file) |> 
   # account for genes being from multiple sources
@@
 visser_predictions <- readr::read_tsv(params$visser_marker_predictions)
+panglao_predictions <- readr::read_tsv(params$panglao_predictions)
 ```
@@
 ```{r}
-# read in predictions and get cell type assignments 
-visser_predictions <- readr::read_tsv(visser_predictions_file)
-
+# get cell type assignments 
 visser_celltype_assignments <- get_celltype_assignment(visser_predictions)
@@
   ) |> 
-  dplyr::left_join(all_markers_df)
+  dplyr::left_join(visser_markers_df)
 ```
@@
 # get a list of all celltypes in the reference
-celltypes <- unique(all_markers_df$cell_type)
+celltypes <- unique(visser_markers_df$cell_type)
 
@@
 ```{r}
-# read in predictions file
-panglao_predictions <- readr::read_tsv(params$panglao_predictions)
-
 # get cell type assignments 

### 2024-05-21 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
   # make sure png files exists
-  purrr::map(\(file)
+  purrr::walk(\(file)
     stopifnot("Missing PNG file" = file.exists(file))
@@
 # don't include the normal cells that were provided as reference in scaling 
-reference_cells_only <- all_cnv_df |> 
-  dplyr::filter(stringr::str_detect(subcluster, "reference")) |> 
-  dplyr::mutate(has_cnv = NA,
-                z_scaled_cnv = NA)
-
-unknown_cells_only <- all_cnv_df |>
-  dplyr::filter(stringr::str_detect(subcluster, "unknown")) |> 
+cnv_df <- all_cnv_df |>
   # get total number of chromosomes with cnvs per cell and then scale that number 
-  dplyr::mutate(has_cnv = rowSums(across(has_cnv_cols)),
+  dplyr::mutate(
+    has_cnv = ifelse(
+      stringr::str_detect(subcluster, "reference"), 
+      NA_real_, 
+      rowSums(across(has_cnv_cols))
+    ),
     z_scaled_cnv = scale(has_cnv))
@@
 ```
+
 It looks like we could use a cutoff of 0.1 here to classify tumor/normal cells. 

### 2024-05-21 analyses/cell-type-ewings/scripts/run-infercnv.R

@@ -144,7 +144,7 @@ infercnv_obj <- infercnv::run(infercnv_obj,
 
 # create table with barcodes and CNVs for each chromosome
 infercnv::add_to_seurat(seurat_obj = NULL,
-                                      infercnv_output_path = scratch_dir)
+                        infercnv_output_path = scratch_dir)
 
 # define path to output cnv metadata file saved in scratch dir
 scratch_metadata_file <- file.path(scratch_dir, "map_metadata_from_infercnv.txt")

### 2024-05-21 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd
create some faceted density and umaps

@@
     z_scaled_cnv = scale(has_cnv))
-
-# combine back together for plotting 
-cnv_df <- bind_rows(reference_cells_only, unknown_cells_only)
 ```
@@
   # calculate weighted mean, excluding normal cells 
-  dplyr::mutate(mean_proportion = ifelse(!is.na(has_cnv), weighted.mean(across(proportion_cols), num_genes_df$n), NA)) |> 
+  dplyr::mutate(
+    mean_proportion = ifelse(
+      !is.na(has_cnv), 
+      weighted.mean(across(proportion_cols), chr_weights), 
+      NA)) |> 
   dplyr::ungroup()
@@
 
-Below we will plot whether or not the gain or loss is present in each cell according to `InferCNV`.
+First we will compare the proportions across each chromosome between cells we have labeled as tumor or normal. 
+Then we will look specifically at the chromosomes commonly affected in Ewing sarcoma.
 
 ```{r}
-# umaps showing where cells are with gains/losses 
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = has_dupli_chr8)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr8 gain") +
-  theme_bw()
+# prepare a data frame with one row per chromosome per cell for plotting
+proportion_df <- cnv_df |> 
+  # select only relevant columns for plotting
+  dplyr::select(barcodes, UMAP1, UMAP2, cnv_mean_classification, all_of(proportion_cols)) |> 
+  tidyr::pivot_longer(cols = all_of(proportion_cols),
+    names_to = "chromosome",
+    values_to = "proportion") |> 
+  # remove extra string in front of chr 
+  dplyr::mutate(
+    chromosome = stringr::str_remove(chromosome, "proportion_scaled_cnv_chr"),
+    # turn into numeric for easy ordering in plot 
+    chromosome = as.numeric(chromosome)
+  )
 
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = has_dupli_chr12)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr12 gain") +
-  theme_bw()
-
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = has_dupli_chr1)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr1 gain") +
-  theme_bw()
-
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = has_loss_chr16)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr16 loss") +
-  theme_bw()
 ```
 
-From this we see that the group of tumor cells at the bottom have chr8 and chr1 gains with some having gains of chr12. 
-It doesn't look like any cells have loss of chr16. 
-The other interesting thing is that there is a group of cells that have gains of chr8, chr1, and chr12 that we did not classify as tumor cells. Should they be...
+```{r, fig.height=10, fig.width=10}
+ggplot(proportion_df, aes(x = proportion, color = cnv_mean_classification)) +
+  geom_density() +
+  theme_bw() +
+  facet_wrap(vars(chromosome), ncol = 3)
+```
 
-We can also look at the proportion of the chromosome with the gain.
+This shows us that Chr1, Chr8, and Chr12 all appear to have a gain in the proportion of the chromosome with a CNV between tumor and normal cells. 
+There are a few other chromosomes that also have alterations (Chr3, Chr4, etc.) with another set of chromosomes that appear to be unaffected. 
+This includes Chr16.
+
+Below we plot only Chr1, Chr8, Chr12, and Chr16 and look at the proportion on a UMAP.
 
 ```{r}
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = proportion_scaled_dupli_chr8)) +
+ewings_chr_df <- proportion_df |> 
+  dplyr::filter(chromosome %in% c(1,8,12,16))
+
+ggplot(ewings_chr_df, aes(x = UMAP1, y=UMAP2 , color = proportion)) +
   geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr8 gain") +
   theme_bw() +
+  facet_wrap(vars(chromosome)) +
   scale_color_viridis_c()
 
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = proportion_scaled_dupli_chr12)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr12 gain") +
-  theme_bw() +
-  scale_color_viridis_c()
-
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = proportion_scaled_dupli_chr1)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr1 gain") +
-  theme_bw() +
-  scale_color_viridis_c()
 ```

### 2024-05-21 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd
remove has_cnv umaps from template

@@
 
-reference_cells_only <- all_cnv_df |> 
-  dplyr::filter(stringr::str_detect(subcluster, "reference")) |> 
-  dplyr::mutate(has_cnv = NA,
-                z_scaled_cnv = NA)
-
-unknown_cells_only <- all_cnv_df |>
-  dplyr::filter(stringr::str_detect(subcluster, "unknown")) |> 
-  # get total number of chromosomes with cnvs per cell and then scale that number 
-  dplyr::mutate(has_cnv = rowSums(across(has_cnv_cols)),
+cnv_df <- all_cnv_df |> 
+  dplyr::mutate(
+    has_cnv = ifelse(
+      stringr::str_detect(subcluster, "reference"), 
+      NA_real_, 
+      rowSums(across(has_cnv_cols))
+    ),
     z_scaled_cnv = scale(has_cnv))
-
-# combine back together for plotting 
-cnv_df <- bind_rows(reference_cells_only, unknown_cells_only)
 ```
@@
 
-Below we will plot whether or not the gain or loss is present in each cell according to `InferCNV`.
+First we will compare the proportions across each chromosome between cells we have labeled as tumor or normal. 
+Then we will look specifically at the chromosomes commonly affected in Ewing sarcoma.
 
 ```{r}
-# umaps showing where cells are with gains/losses 
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = has_dupli_chr8)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr8 gain") +
-  theme_bw()
+# prepare a data frame with one row per chromosome per cell for plotting
+proportion_df <- cnv_df |> 
+  # select only relevant columns for plotting
+  dplyr::select(barcodes, UMAP1, UMAP2, cnv_mean_classification, all_of(proportion_cols)) |> 
+  tidyr::pivot_longer(cols = all_of(proportion_cols),
+    names_to = "chromosome",
+    values_to = "proportion") |> 
+  # remove extra string in front of chr 
+  dplyr::mutate(
+    chromosome = stringr::str_remove(chromosome, "proportion_scaled_cnv_chr"),
+    # turn into numeric for easy ordering in plot 
+    chromosome = as.numeric(chromosome)
+  )
 
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = has_dupli_chr12)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr12 gain") +
-  theme_bw()
-
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = has_dupli_chr1)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr1 gain") +
-  theme_bw()
-
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = has_loss_chr16)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr16 loss") +
-  theme_bw()
 ```
 
-We can also look at the proportion of the chromosome with the gain.
+```{r, fig.height=10, fig.width=10}
+ggplot(proportion_df, aes(x = proportion, color = cnv_mean_classification)) +
+  geom_density() +
+  theme_bw() +
+  facet_wrap(vars(chromosome), ncol = 3)
+```
+
+Below we plot only Chr1, Chr8, Chr12, and Chr16 and look at the proportion on a UMAP.
 
 ```{r}
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = proportion_scaled_dupli_chr8)) +
+ewings_chr_df <- proportion_df |> 
+  dplyr::filter(chromosome %in% c(1,8,12,16))
+
+ggplot(ewings_chr_df, aes(x = UMAP1, y=UMAP2 , color = proportion)) +
   geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr8 gain") +
   theme_bw() +
+  facet_wrap(vars(chromosome)) +
   scale_color_viridis_c()
 
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = proportion_scaled_dupli_chr12)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr12 gain") +
-  theme_bw() +
-  scale_color_viridis_c()
-
-ggplot(cnv_df, aes(x= UMAP1, y = UMAP2, color = proportion_scaled_dupli_chr1)) +
-  geom_point(alpha = 0.5, size = 0.5) +
-  labs(title = "Chr1 gain") +
-  theme_bw() +
-  scale_color_viridis_c()
 ```

### 2024-05-21 analyses/cell-type-ewings/template_notebooks/03-copykat.Rmd
update template notebook to take in new params

@@
   library_id: SCPCL000822
+  marker_gene_classification: "../results/annotate_tumor_cells_output/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
+  no_ref_copykat_results: "../results/annotate_tumor_cells_output/SCPCS000490/no_reference"
+  with_ref_copykat_results: "../results/annotate_tumor_cells_output/SCPCS000490/with_reference"
 ---
@@
 
-`CopyKAT` was run using the `run-copykat.R` script and specifying endothelial cells from both `SingleR` and `CellAssign` as the normal reference. 
-This produced results from `CopyKAT` both with and without a normal reference. 
+`CopyKAT` was run using the `run-copykat.R` script with and without a normal reference. 
 These results are read into this notebook and used to:
@@
 
-# tumor/normal classifications 
-classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
-classifications_file <- file.path(module_base, "results", "marker_gene_analysis", classifications_filename)
+obj_file <- glue::glue("{params$library_id}_final-copykat.rds")
+png_file <- glue::glue("{params$libary_id}_heatmap-copykat.png")
 
-# copykat results 
-results_dir <- file.path(module_base, "results", "copykat", params$library_id)
+copykat_objs <- c(
+  no_ref = file.path(params$no_ref_copykat_results, obj_file),
+  with_ref = file.path(params$with_ref_coypkat_results, obj_file)
+) |> 
+  purrr::map(readr::read_rds)
 
-# predictions files 
-predictions_file <- glue::glue("{params$library_id}_copykat_prediction.txt")
-
-predictions_paths <- c(
-  no_ref = file.path(results_dir, "no_normal", predictions_file),
-  with_ref = file.path(results_dir, "endothelial_normal", predictions_file)
+png_list <- c(
+  no_ref = file.path(params$no_ref_copykat_results, png_file),
+  with_ref = file.path(params$with_ref_copykat_results, png_file)
 )
-
-# full results gene by cell 
-full_ck_result_file <- glue::glue("{params$library_id}_copykat_CNA_results.txt")
-
-full_ck_result_paths <- c(
-  no_ref = file.path(results_dir, "no_normal", full_ck_result_file),
-  with_ref = file.path(results_dir, "endothelial_normal", full_ck_result_file)
-)
-
 ```
@@
 # read in tumor normal classifications 
-manual_classifications_df <- readr::read_tsv(classifications_file)
+manual_classifications_df <- readr::read_tsv(params$marker_gene_classification)
 
 # read in ck predictions from both reference types (no_normal and with_normal)
-ck_results_df <- predictions_paths |> 
-  purrr::map(readr::read_tsv) |> 
+ck_results_df <- copykat_objs |> 
+  purrr::map(\(obj){
+    obj$prediction |> 
+      as.data.frame()
+  }) |> 
   dplyr::bind_rows(.id = "reference_used")
@@
 # read in full gene by cell copy number detection results 
-full_ck_results_df <- full_ck_result_paths |> 
-  purrr::map(readr::read_tsv) |> 
+full_ck_results_df <- copykat_objs |> 
+  purrr::map(\(obj){
+    obj$CNAmat |> 
+      as.data.frame()
+  }) |> 
   dplyr::bind_rows(.id = "reference_used")
@@
 
-![](`r file.path(results_dir, "no_normal", "SCPCL000822_copykat_heatmap.jpeg")`)
+![](`r png_list["no_ref"]`)
 
@@
 
-![](`r file.path(results_dir, "endothelial_normal", "SCPCL000822_copykat_heatmap.jpeg")`)
-
-Both of these heatmaps appear to show some increase in observed CNVs in aneuploid vs. diploid which is to be expected.
+![](`r png_list["with_ref"]`)9
 
@@
 ```
-
-If we look at the UMAPs shown in `01-marker-gene-tumor-classification.Rmd`, the cells that are labeled aneuploid here are consistent with where we saw marker gene expession for tumor cells. 
-It also looks like there is not a visible difference between using a reference of normal cells and not using a reference.
 
@@
 
-It looks like the largest increases are in chr8 and chr21. 
-Chr8 is one of the more common CNA's in Ewing's so this is a good sign.
-
-Chr12 may show some increase, but Chr1 and Chr16 don't show any real differences.
-
 ### Confusion Matrix
@@
 
-Looking at these results, we see that cells classified as tumor cells using marker genes are mostly overlapping with cells that are aneuploid in CopyKAT, which is what we expect! 
-We also see that the Kappa values are close to 1, indicating almost perfect agreement.
-
-It also looks like there isn't a difference between using the reference and not using the reference. 
-The reference Kappa is .1 higher, but I don't think that's enough to argue that you need to use a reference. 
-The one thing to note is that using a reference cut the run time for `CopyKAT` in half, so that is one benefit.
-
 We can also calculate the Jaccard similarity index to visualize the amount of cells that have overlapping annotations.
-Given the confusion matrix, I would expect to see good agreement between marker gene and `CopyKAT` annotations.
 
@@
 ```
-
-As expected, it looks like cells classified as tumor cells using marker gene annotation are classified as `aneuploid` with `CopyKAT`.
 
@@
 
-We see that aneuploid cells are mostly muscle cells in both `CellAssign` and `SingleR`. 
-In `SingleR` there are some other cell types that were assigned that have a few anueploid cells (chondrocytes, neurons, and fibroblasts).
-
-For diploid cells, we see the most overlap with fibroblasts, endothelial cells, and macrophages. 
-These are consistent with the cell types that we see normal cells line up with in `01-marker-gene-tumor-classification.Rmd`.
-
-## Conclusions
-
-- `CopyKAT` seems to do a good job in this Ewing sarcoma sample of identifying aneuploid and diploid cells. 
-- There is some small evidence of copy number gains in Chr8 and Chr12 which is consistent with findings in Ewing sarcoma patients. 
-- The annotations from `CopyKAT` mostly line up with annotations obtained using marker gene annotation. 
-- Using a reference of normal cells doesn't change the overall results but does reduce the run time for `CopyKAT`.
-
 ## Session Info

### 2024-05-22 analyses/cell-type-ewings/scripts/run-copykat.R
fix path to heatmap

@@ -69,10 +69,10 @@ scratch_dir <- file.path(opt$scratch_dir, library_id)
 # path to output copykat object 
 copykat_output_obj <- file.path(opt$results_dir, glue::glue("{library_id}_final-copykat.rds"))
 
-# path to scratch and final png file to copy over 
-png_file <- glue::glue("{library_id}_copykat_heatmap.png")
-scratch_png <- file.path(opt$scratch_dir, png_file)
-output_png <- file.path(opt$results_dir, png_file)
+# path to scratch and final heatmap file to copy over 
+jpeg_file <- glue::glue("{library_id}_copykat_heatmap.jpeg")
+scratch_jpeg <- file.path(opt$scratch_dir, jpeg_file)
+output_jpeg <- file.path(opt$results_dir, jpeg_file)
 
 # change working directory of the script to the scratch directory
 # this ensures copykat files get saved to the right location
@@ -114,5 +114,5 @@ copykat_result <- copykat(
 readr::write_rds(copykat_result, copykat_output_obj)
 
 # copy over png file 
-fs::file_copy(scratch_png, output_png, overwrite = TRUE)
+fs::file_copy(scratch_jpeg, output_jpeg, overwrite = TRUE)
 

### 2024-05-22 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 ggplot(cnv_df, aes(x = has_cnv)) +
-  geom_density() +
+  geom_freqpoly(binwidth = 1, center = 0) +
   theme_bw() +
@@
 cnv_df <- cnv_df |> 
-  dplyr::mutate(cnv_classification = dplyr::if_else(z_scaled_cnv > 0, "Tumor", "Normal"))
+  dplyr::mutate(cnv_classification = dplyr::if_else(has_cnv > mean(has_cnv), "Tumor", "Normal"))
 
@@
 ggplot(proportion_df, aes(x = proportion, color = cnv_mean_classification)) +
-  geom_density() +
+  geom_density(bounds = c(0,1), bw = 0.05) +
   theme_bw() +

### 2024-05-22 analyses/cell-type-ewings/scripts/make-gene-order-file.R

@@ -27,7 +27,7 @@ opt <- parse_args(OptionParser(option_list = option_list))
 fs::dir_create(opt$local_ref_dir)
 
 # sync gtf file to local directory 
-gtf_filename <- stringr::word(opt$gtf_file, -1, sep = "/")
+gtf_filename <- basename(opt$gtf_file)
 
 local_gtf_file <- file.path(opt$local_ref_dir, gtf_filename)
 if(!file.exists(local_gtf_file)){

### 2024-05-22 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd
use mean cutoff and improve density plots

@@
       rowSums(across(has_cnv_cols))
-    ),
-    z_scaled_cnv = scale(has_cnv))
+    ))
 ```
@@
 If we were to use total CNVs we would need some sort of cut off for the number. 
-Let's look at the distribution of total CNVs and also the scaled total CNVs to see if we can identify a cut off.
+Let's look at the distribution of total CNVs to see if we can identify a cut off.
 
@@
   labs(x = "Total number of CNVs")
-
-ggplot(cnv_df, aes(x= z_scaled_cnv)) + 
-  geom_density() +
-  theme_bw() +
-  labs(x = "z-scaled chromosome CNV count")
 ```
 
-Looking at the scaled number of CNVs it does look like we could use 0 as the cutoff and call any cell that has > 0 CNVs (after scaling) as tumor and all other cells as normal.
+We will call any tumor cell as those with a total number of cnvs greater than the mean(number of cnvs).
 
@@
 ```{r}
+# get mean value for has_cnv, excluding any NAs
+mean_cnv <- cnv_df |> 
+  dplyr::filter(!is.na(has_cnv)) |> 
+  dplyr::pull(has_cnv) |> 
+  mean()
+
 # add cnv classification 
 cnv_df <- cnv_df |> 
-  dplyr::mutate(cnv_classification = dplyr::if_else(has_cnv > mean(has_cnv), "Tumor", "Normal"))
+  dplyr::mutate(cnv_classification = dplyr::case_when(is.na(has_cnv) ~ "Reference",
+                                                      has_cnv > mean_cnv ~ "Tumor",
+                                                      has_cnv <= mean_cnv ~ "Normal"))
 
@@
 
-It looks like we could use a cutoff of 0.1 here to classify tumor/normal cells. 
 Let's see how this matches up with the tumor/normal classifications we defined just using total CNVs.
@@
 
+Below we will call tumor cells based on those that have a mean scaled proportion higher than the mean of all proportions.
+
 ```{r}
+# get mean value for mean_proportion, excluding any NAs
+all_mean_proportion <- cnv_df |> 
+  dplyr::filter(!is.na(mean_proportion)) |> 
+  dplyr::pull(mean_proportion) |> 
+  mean()
+
 # add cnv classification 
 cnv_df <- cnv_df |> 
-  dplyr::mutate(cnv_mean_classification = dplyr::if_else(mean_proportion > 0.1, "Tumor", "Normal"))
+  dplyr::mutate(cnv_mean_classification = dplyr::case_when(is.na(mean_proportion) ~ "Reference",
+                                                           mean_proportion > all_mean_proportion ~ "Tumor",
+                                                           mean_proportion <= all_mean_proportion ~ "Normal"))
 
@@
   # relevel classifications for confusion matrix 
+  dplyr::filter(cnv_mean_classification != "Reference") |> 
   dplyr::mutate(
@@
 There are a few other chromosomes that also have alterations (Chr3, Chr4, etc.) with another set of chromosomes that appear to be unaffected.
-This includes Chr16.
 

### 2024-05-22 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd
use mean cutoff and improve density plots

@@
 
+# don't include the normal cells that were provided as reference in scaling 
 cnv_df <- all_cnv_df |>
+  # get total number of chromosomes with cnvs per cell and then scale that number 
   dplyr::mutate(
@@
 ggplot(cnv_df, aes(x = has_cnv)) +
-  geom_density() +
+  geom_freqpoly(binwidth = 1, center = 0) +
   theme_bw() +
   labs(x = "Total number of CNVs")
-
-ggplot(cnv_df, aes(x= z_scaled_cnv)) + 
-  geom_density() +
-  theme_bw() +
-  labs(x = "z-scaled chromosome CNV count")
 ```
 
-Looking at the scaled number of CNVs we will use 0 as the cutoff and call any cell that has > 0 CNVs (after scaling) as tumor and all other cells as normal.
+We will call any tumor cell as those with a total number of cnvs greater than the mean(number of cnvs).
 
 ```{r}
+# get mean value for has_cnv, excluding any NAs
+mean_cnv <- cnv_df |> 
+  dplyr::filter(!is.na(has_cnv)) |> 
+  dplyr::pull(has_cnv) |> 
+  mean()
+
 # add cnv classification 
 cnv_df <- cnv_df |> 
-  dplyr::mutate(cnv_classification = dplyr::if_else(z_scaled_cnv > 0, "Tumor", "Normal"))
+  dplyr::mutate(cnv_classification = dplyr::case_when(is.na(has_cnv) ~ "Reference",
+                                                      has_cnv > mean_cnv ~ "Tumor",
+                                                      has_cnv <= mean_cnv ~ "Normal"))
 
@@
 
-We will use a cutoff of 0.1 here to classify tumor/normal cells. 
 Let's see how this matches up with the tumor/normal classifications we defined just using total CNVs.

### 2024-05-22 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd
match template

@@
 # get mean value for has_cnv, excluding any NAs
-mean_cnv <- cnv_df |> 
-  dplyr::filter(!is.na(has_cnv)) |> 
-  dplyr::pull(has_cnv) |> 
-  mean()
+mean_cnv <- mean(cnv_df$has_cnv, na.rm=TRUE)
 
@@
 # get mean value for mean_proportion, excluding any NAs
-all_mean_proportion <- cnv_df |> 
-  dplyr::filter(!is.na(mean_proportion)) |> 
-  dplyr::pull(mean_proportion) |> 
-  mean()
+all_mean_proportion <- mean(cnv_df$mean_proportion, na.rm = TRUE)
 

### 2024-05-22 analyses/cell-type-ewings/template_notebooks/02-cellassign.Rmd
save cellassign output

@@
 # output annotations
-cellassign_annotations_file <- file.path(params$results_dir, glue::glue("{params$library_id}_cellassign-annotations.tsv"))
+cellassign_annotations_file <- file.path(params$results_dir, glue::glue("{params$library_id}_cellassign-classifications.tsv"))
 ```
@@
 
+## Save annotations
+
+```{r}
+celltype_df <- celltype_df |> 
+  dplyr::select(barcodes, 
+                marker_gene_cellassign_annotations = cellassign_annotations, 
+                visser_cellassign_annotations, 
+                panglao_cellassign_annotations)
+
+readr::write_tsv(celltype_df, cellassign_annotations_file)
+```
+
 ## Session Info

### 2024-05-22 analyses/cell-type-ewings/template_notebooks/03-copykat.Rmd
fix path to heatmap
save copykat output

@@
   marker_gene_classification: "../results/annotate_tumor_cells_output/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
-  no_ref_copykat_results: "../results/annotate_tumor_cells_output/SCPCS000490/no_reference"
-  with_ref_copykat_results: "../results/annotate_tumor_cells_output/SCPCS000490/with_reference"
+  no_ref_copykat_results: "../results/annotate_tumor_cells_output/SCPCS000490/copykat/no_reference"
+  with_ref_copykat_results: "../results/annotate_tumor_cells_output/SCPCS000490/copykat/with_reference"
+  reference_cell_file: "../results/annotate_tumor_cells_output/SCPCS000490/annotations/SCPCL000822_reference-cells.tsv"
+  results_dir: "../results/annotate_tumor_cells_output/SCPCS000490"
 ---
@@
 obj_file <- glue::glue("{params$library_id}_final-copykat.rds")
-png_file <- glue::glue("{params$libary_id}_heatmap-copykat.png")
+png_file <- glue::glue("{params$libary_id}_copykat_heatmap.jpeg")
 
@@
   no_ref = file.path(params$no_ref_copykat_results, obj_file),
-  with_ref = file.path(params$with_ref_coypkat_results, obj_file)
+  with_ref = file.path(params$with_ref_copykat_results, obj_file)
 ) |> 
@@
 )
+
+# output classifications file 
+copykat_results_file <- file.path(params$results_dir, glue::glue("{params$library_id}_copykat-classifications.tsv"))
 ```
@@
 manual_classifications_df <- readr::read_tsv(params$marker_gene_classification)
+
+# read in ref cells 
+normal_cells <- readr::read_tsv(params$reference_cell_file) |> 
+  dplyr::filter(reference_cell_class == "Normal") |> 
+  dplyr::pull(barcodes)
 
@@
   # now add copykat results
-  dplyr::left_join(ck_results_df, by = c("barcodes" = "cell.names"))
+  dplyr::left_join(ck_results_df, by = c("barcodes" = "cell.names")) |> 
+  dplyr::mutate(
+    copykat.pred = dplyr::if_else(
+      # label any reference cells that were used 
+      (barcodes %in% normal_cells) & (reference_used == "with_ref"), 
+      "reference", 
+      copykat.pred)
+  )
 ```
@@
 filtered_cnv_df <- cnv_df |> 
-  dplyr::filter(copykat.pred != "not.defined")
+  dplyr::filter(!(copykat.pred %in% c("not.defined", "reference")))
 
@@
 
+## Save annotations
+
+```{r}
+celltype_df <- cnv_df |> 
+  dplyr::select(barcodes, 
+                reference_used,
+                copykat.pred) |> 
+  dplyr::distinct() |> 
+  tidyr::pivot_wider(
+    names_from = reference_used,
+    values_from = copykat.pred
+  )
+
+readr::write_tsv(celltype_df, copykat_results_file)
+```
+
 ## Session Info

### 2024-05-22 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd

@@
   library_id: SCPCL000822
-  marker_genes_classification_file: "../results/annotate_tumor_cells_output/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
-  infercnv_dir: "../results/annotate_tumor_cells_output/SCPCS000490/infercnv/endothelial_reference"
+  marker_gene_classification: "../results/annotate_tumor_cells_output/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
+  infercnv_dir: "../results/annotate_tumor_cells_output/SCPCS000490/infercnv"
   results_dir: "../results/annotate_tumor_cells_output/SCPCS000490"
@@
 # with ref cnv metadata file 
-cnv_metadata_file <- file.path(params$infercnv_dir, "endothelial_reference", glue::glue("{params$library_id}_cnv-metadata.tsv"))
+cnv_metadata_file <- file.path(params$infercnv_dir, glue::glue("{params$library_id}_cnv-metadata.tsv"))
 
 # full infercnv object with information on genes/ chr to use later 
-final_cnv_obj_file <- file.path(params$infercnv_dir, "endothelial_reference", glue::glue("{params$library_id}_cnv-obj.rds"))
+final_cnv_obj_file <- file.path(params$infercnv_dir, glue::glue("{params$library_id}_cnv-obj.rds"))
 
@@
   "Final InferCNV object is missing" = file.exists(final_cnv_obj_file),
-  "Marker gene classifications are missing" = file.exists(params$marker_genes_classification_file)
+  "Marker gene classifications are missing" = file.exists(params$marker_gene_classification)
 )
@@
 # read in tumor normal classifications 
-manual_classifications_df <- readr::read_tsv(params$marker_genes_classification_file)
+manual_classifications_df <- readr::read_tsv(params$marker_gene_classification)
 

### 2024-05-22 analyses/cell-type-ewings/template_notebooks/02-cellassign.Rmd
account for missing tumor markers and no cell assign results

@@
 get_celltype_assignment <- function(predictions){
+
+  # if no predictions, assign every cell to NA
+  if(length(colnames(predictions)) == 1){
+    celltype_assignments <- data.frame(
+      barcode = predictions$barcode,
+      celltype = NA
+    )
+  } else {
     # get individual cell type assignments
@@
       dplyr::ungroup() 
+  }
 
@@
 
+```{r}
+# check if marker gene annotations are present 
+if(all(is.na(classifications_df$marker_gene_classification))){
+  has_marker_gene <- FALSE
+} else {
+  has_marker_gene <- TRUE 
+}
+```
+
 ## Analysis content
@@
 
+if(all(is.na(tumor_celltype_assignments$celltype))){
+  has_tumor_assignments <- FALSE
+  message("No CellAssign assignments found for tumor marker genes.")
+} else {
  # create a table of assignments 
   table(tumor_celltype_assignments$celltype) 
+  has_tumor_assignments <- TRUE
+}
+
 ```
@@
 ```{r}
+# define annotations to use in all heatmaps
+if(has_marker_gene){
   # get annotations for heatmaps
@@
   ) 
+} else {
+  annotation_df <- NULL
+}
+```
 
+```{r, eval=has_tumor_assignments}
 # heatmap of prediction scores 
@@
 
-```{r}
+```{r, eval=has_tumor_assignments}
 ggplot(celltype_df, aes(x = UMAP1, y = UMAP2, color = cellassign_annotations)) +
@@
 
-```{r}
+```{r, eval=has_tumor_assignments}
 # prep annotations for confusion matrix
@@
 
-```{r}
+```{r, eval=has_marker_gene}
 # get marker gene expression 
@@
 
+if(all(is.na(visser_celltype_assignments$celltype))){
+  has_visser_assignments <- FALSE
+  message("No CellAssign assignments found for tumor marker genes.")
+} else {
+ # create a table of assignments 
   table(visser_celltype_assignments$celltype) 
+  has_visser_assignments <- TRUE
+}
 ```
 
-```{r}
+```{r, eval=has_visser_assignments}
 # heatmap of prediction scores 
@@
 
+if(all(is.na(panglao_celltype_assignments$celltype))){
+  has_panglao_assignments <- FALSE
+  message("No CellAssign assignments found for tumor marker genes.")
+} else {
  # create a table of assignments 
   table(panglao_celltype_assignments$celltype) 
+  has_panglao_assignments <- TRUE
+}
 ```
 
-```{r}
+```{r, eval=has_panglao_assignments}
 # heatmap of prediction scores 

### 2024-05-22 analyses/cell-type-ewings/template_notebooks/01-marker-genes.Rmd
account for no marker gene expression

@@
 ```{r}
+if(sum(umap_df$gene_expression) == 0){
+  has_marker_gene_exp <- FALSE
+} else {
+  has_marker_gene_exp <- TRUE
+}
+```
+
+```{r, eval=has_marker_gene_exp}
 # faceted umap showing a umap panel for each marker gene 
@@
 
-```{r}
+```{r, eval=has_marker_gene_exp}
 ggplot(umap_df, aes(x = gene_expression, fill = gene_symbol)) +
@@
 
-```{r}
+```{r, eval=has_marker_gene_exp}
 umap_df <- umap_df |> 
@@
 
-```{r}
+```{r, eval=has_marker_gene_exp}
 ggplot(umap_df, aes(x = transformed_gene_expression, fill = gene_symbol)) +
@@
 
-```{r}
+```{r, eval=has_marker_gene_exp}
 # calculate sum gene expression across all marker genes in list 
@@
 
-```{r}
+```{r, eval=has_marker_gene_exp}
 # classify tumor cells based on presence of any marker genes 
@@
 
-```{r}
+```{r, eval=has_marker_gene_exp}
 # add cell types from sum classification to plotting df 
@@
 
-```{r}
+```{r, eval=has_marker_gene_exp}
 # add in tumor/normal classification to colData of SCE 
@@
 
-```{r}
+```{r, eval=has_marker_gene_exp}
 # pull out top markers for tumor cells 
@@
 
-```{r, fig.height=10, fig.width=10}
+```{r, fig.height=10, fig.width=10, eval=has_marker_gene_exp}
 marker_plot_df <- celltype_df |> 
@@
 
-```{r}
+```{r, eval=has_marker_gene_exp}
 # get all assigned cell types in one dataframe 
@@
 
-```{r, fig.height=10, fig.width=7}
+```{r, fig.height=10, fig.width=7, eval=has_marker_gene_exp}
 # Set heatmap padding option
@@
 
-```{r}
+```{r, eval=has_marker_gene_exp}
 # plot comparing marker gene expression in cells annotated as "normal" and "tumor" by SingleR and CellAssign 
@@
 ```{r}
+if(has_marker_gene_exp){
   # get a dataframe with barcodes, sum_classification and cluster_classification 
@@
   readr::write_tsv(celltypes_df, output_classifications_file)
+} else{
+  # save marker gene classification as NA if no expression 
+  celltypes_df <- data.frame(
+    barcodes = umap_df$barcodes,
+    marker_gene_classification = NA
+  )
+
+  readr::write_tsv(celltypes_df, output_classifications_file)
+}
 ```

### 2024-05-23 analyses/cell-type-ewings/template_notebooks/03-copykat.Rmd
account for no marker genes

@@
 manual_classifications_df <- readr::read_tsv(params$marker_gene_classification)
+
+# check if marker gene annotations are present 
+if(all(is.na(manual_classifications_df$marker_gene_classification))){
+  has_marker_gene <- FALSE
+} else {
+  has_marker_gene <- TRUE 
+}
 
@@
 
-```{r}
+```{r, eval=has_marker_gene}
 filtered_cnv_df <- cnv_df |> 
-  dplyr::filter(!(copykat.pred %in% c("not.defined", "reference")))
+  dplyr::filter(!(copykat.pred %in% c("not.defined", "reference")),
+                # filter any low confidence calls
+                !stringr::str_detect(copykat.pred, "low.conf"))
 
@@
 
-```{r}
+```{r, eval=has_marker_gene}
 # calculate Jaccard similarity index for each reference type
@@
 
-```{r}
+```{r, eval=has_marker_gene}
 # Set heatmap padding option

### 2024-05-23 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd
account for no marker genes

@@
 
+# check if marker gene annotations are present 
+if(all(is.na(manual_classifications_df$marker_gene_classification))){
+  has_marker_gene <- FALSE
+} else {
+  has_marker_gene <- TRUE 
+}
+
 # read in cnv object 
@@
 # remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
-chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL000219.1|chrGL000194.1"))
+chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL"))
 all_cnv_df <- coldata_df[, chr_columns_to_keep]
@@
 
-```{r}
+```{r, eval=has_marker_gene}
 all_classifications_df <- cnv_df |> 

### 2024-05-23 analyses/cell-type-ewings/template_notebooks/01-marker-genes.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>
add some messages about skipped plots

@@
 ```{r}
+# test if _any_ marker genes have expression
 if(sum(umap_df$gene_expression) == 0){
   has_marker_gene_exp <- FALSE
+  message("No marker gene expression detected, so all plots of marker genes will be skipped.")
 } else {

### 2024-05-23 analyses/cell-type-ewings/template_notebooks/02-cellassign.Rmd
add some messages about skipped plots

@@
   has_marker_gene <- FALSE
+  message("No annotations were available using only marker gene expression. 
+          Any plots comparing CellAssign to marker gene annotation will be skipped.")
 } else {

### 2024-05-23 analyses/cell-type-ewings/template_notebooks/03-copykat.Rmd

@@
   has_marker_gene <- FALSE
+  message("No annotations were available using only marker gene expression. 
+          Any plots comparing CopyKAT to marker gene annotation will be skipped.")
 } else {

### 2024-05-23 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd

@@
   has_marker_gene <- FALSE
+  message("No annotations were available using only marker gene expression. 
+          Any plots comparing InferCNV to marker gene annotation will be skipped.")
 } else {

### 2024-05-28 analyses/cell-type-ewings/template_notebooks/02-cellassign.Rmd
set eval based on existing annotations

@@
   library_id: SCPCL000822
-  marker_gene_classification: "../results/annotate_tumor_cells_output/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
-  tumor_marker_predictions: "../results/annotate_tumor_cells_output/SCPCS000490/cellassign/SCPCL000822_tumor-only-predictions.tsv"
-  visser_marker_predictions: "../results/annotate_tumor_cells_output/SCPCS000490/cellassign/SCPCL000822_visser-predictions.tsv"
-  panglao_predictions: "../results/annotate_tumor_cells_output/SCPCS000490/cellassign/SCPCL000822_panglao-predictions.tsv"
-  results_dir: "../results/annotate_tumor_cells_output/SCPCS000490"
+  marker_gene_classification: "../results/sim_annotate_tumor/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
+  tumor_marker_predictions: "../results/sim_annotate_tumor/SCPCS000490/cellassign/SCPCL000822_tumor-marker_predictions.tsv"
+  visser_marker_predictions: "../results/sim_annotate_tumor/SCPCS000490/cellassign/SCPCL000822_visser-all-marker_predictions.tsv"
+  panglao_predictions: "../results/sim_annotate_tumor/SCPCS000490/cellassign/SCPCL000822_panglao-endo-fibro_predictions.tsv"
+  results_dir: "../results/sim_annotate_tumor/SCPCS000490"
 ---
@@
 
-```{r}
+```{r, eval=has_visser_assignments}
 # first rename column to combine with cell type df 
@@
 
-```{r}
+```{r, eval=has_visser_assignments}
 # get jaccard similarity index 
@@
 
-```{r}
+```{r, eval=has_visser_assignments}
 # first make sure cells have been classified as endothelial 
@@
 
-```{r}
+```{r, eval=has_visser_assignments}
 # get all markers 
@@
 
-```{r}
+```{r, eval=has_visser_assignments}
 # get a list of all celltypes in the reference
@@
 
-```{r}
+```{r, eval=has_visser_assignments}
 grouped_celltype_df <- marker_gene_exp_df |> 
@@
 
-```{r}
+```{r, eval=has_panglao_assignments}
 # first rename column to combine with cell type df 
@@
 
-```{r}
+```{r, eval=has_panglao_assignments}
 # get jaccard similarity index 

### 2024-05-28 analyses/cell-type-ewings/template_notebooks/03-copykat.Rmd
fix spelling in notebook
simplify template notebook defaults

@@
   library_id: SCPCL000822
-  marker_gene_classification: "../results/annotate_tumor_cells_output/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
-  no_ref_copykat_results: "../results/annotate_tumor_cells_output/SCPCS000490/copykat/no_reference"
-  with_ref_copykat_results: "../results/annotate_tumor_cells_output/SCPCS000490/copykat/with_reference"
-  reference_cell_file: "../results/annotate_tumor_cells_output/SCPCS000490/annotations/SCPCL000822_reference-cells.tsv"
-  results_dir: "../results/annotate_tumor_cells_output/SCPCS000490"
+  marker_gene_classification: "tumor-normal-classifications.tsv"
+  no_ref_copykat_results: "copykat/no_reference"
+  with_ref_copykat_results: "copykat/with_reference"
+  reference_cell_file: ".reference-cells.tsv"
+  results_dir: "annotate_tumor_cells_output"
 ---
@@
 obj_file <- glue::glue("{params$library_id}_final-copykat.rds")
-png_file <- glue::glue("{params$libary_id}_copykat_heatmap.jpeg")
+png_file <- glue::glue("{params$library_id}_copykat_heatmap.jpeg")
 

### 2024-05-28 analyses/doublet-detection/exploratory-analyses/02_explore-benchmark-results.Rmd
notebook to explore individual benchmarking results on each benchmark dataset, and associated environment updates

---
params: 
  dataset: "hm-6k"
title: "`r glue::glue('Doublet benchmarking results: {params$dataset}')`"
author: Stephanie J. Spielman
date: "`r Sys.Date()`"
output: html_notebook
---

The goal of this notebook is to explore doublet inferences run on ground-truth benchmarking data. 
These three doublet detection methods are explored:

- `scDblFinder`
- `scrublet`
- A version of `cxds` calculated by `scDblFinder`.
Unlike the other two methods, this one reports only a score and does not attempt to make single/doublet calls, so we use a `0.5` threshold to make `cxds` calls.

This notebook explores inferences made on a single dataset at a time (here, **`r params$dataset`**) with the goal of examining the relationship between scores and accuracy of calls (if made) compared to the ground-truth values.
Note that ground-truth calls are available only for heterotypic doublets, which these methods are more tuned for in the first place (vs homotypic).

## Setup

### Packages

```{r packages}
library(SingleCellExperiment)
library(readr)
library(ggplot2)
library(caret)

theme_set(theme_bw())
```

### Paths

#### Base directories

```{r base paths}
module_base <- rprojroot::find_root(rprojroot::is_renv_project)
data_dir <- file.path(module_base, "scratch", "benchmark_datasets", params$dataset)
result_dir <- file.path(module_base, "results", "benchmark_results")
```

#### Input and output files

Set paths to input and output directories and files in the chunk below.

```{r paths}
# Input files
scdbl_tsv <- file.path(
  result_dir, 
  glue::glue("{params$dataset}_scdblfinder.tsv")
)
scrub_tsv <- file.path(
  result_dir, 
  glue::glue("{params$dataset}_scrublet.tsv")
)
truth_tsv <- file.path(
  data_dir, 
  glue::glue("{params$dataset}_ground-truth.tsv")
)

# Output files
```

### Read and prepare input data

First, we'll read in and combine TSV files with doublet results.

```{r}
scdbl_df <- read_tsv(scdbl_tsv) |>
  dplyr::select(
    barcodes,
    cxds_score, 
    scdbl_score = score, 
    scdbl_prediction  = class
  ) |>
  # add cxds calls at 0.5 threshold
  dplyr::mutate(
    cxds_prediction = dplyr::if_else(
      cxds_score >= 0.5,
      "doublet",
      "singlet"
    )
  )

scrub_df <- read_tsv(scrub_tsv)

true_df <- read_tsv(truth_tsv)

doublet_df <- dplyr::inner_join(
  scdbl_df,
  scrub_df, 
  by = "barcodes"
) |>
  dplyr::inner_join(
    true_df, 
    by = "barcodes"
  )
```

## Explore results: Score distributions

Both `scDblFinder` and `scrublet` make doublet calls based on an internally-determined threshold, which may not be optimal.

These sections explore the score distributions and how they relate to calls, including a confusion matrix and associated performance metrics. 
We're particularly interested in "Balanced accuracy," as there is a much higher proportion of singlets vs doublets.

### scDblFinder results

```{r}
ggplot(doublet_df) +
  aes(x = ground_truth,
      y = scdbl_score, 
      color = scdbl_prediction) + 
  geom_jitter(
    alpha = 0.2, 
    width = 0.1
  ) + 
  labs(
    x = "Ground truth call",
    y = "scDblFinder score",
    color = "scDblFinder prediction"
  )

confusion <-  caret::confusionMatrix(
    # truth should be first
    table(
      "Truth" = doublet_df$ground_truth,
      "Prediction" = doublet_df$scdbl_prediction
    ), 
    positive = "doublet"
)

print(confusion)
```

### scrublet results

```{r}
ggplot(doublet_df) +
  aes(x = ground_truth,
      y = scrublet_score, 
      color = scrublet_prediction) + 
  geom_jitter(
    alpha = 0.2, 
    width = 0.1
  ) + 
  labs(
    x = "Ground truth call",
    y = "scrublet score",
    color = "scrublet prediction"
  )

confusion <-  caret::confusionMatrix(
    # truth should be first
    table(
      "Truth" = doublet_df$ground_truth,
      "Prediction" = doublet_df$scrublet_prediction
    ), 
    positive = "doublet"
)

print(confusion)
```

### cxds results

> This method does not call droplets as singlet/doublet.
> Instead, performance is assessed below at a threshold of 0.5.

```{r}
ggplot(doublet_df) +
  aes(x = ground_truth,
      y = cxds_score, 
      color = cxds_prediction) + 
  geom_jitter(
    alpha = 0.2, 
    width = 0.1
  ) + 
  labs(
    x = "Ground truth call",
    y = "cxds score",
    color = "cxds prediction"
  )

confusion <-  caret::confusionMatrix(
    # truth should be first
    table(
      "Truth" = doublet_df$ground_truth,
      "Prediction" = doublet_df$cxds_prediction
    ), 
    positive = "doublet"
)

print(confusion)
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-05-29 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd

@@
   library_id: SCPCL000822
-  marker_gene_classification: "../results/annotate_tumor_cells_output/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
-  infercnv_dir: "../results/annotate_tumor_cells_output/SCPCS000490/infercnv"
-  results_dir: "../results/annotate_tumor_cells_output/SCPCS000490"
+  marker_gene_classification: "tumor-normal-classifications.tsv"
+  infercnv_dir: "infercnv"
+  results_dir: "annotate_tumor_cells_output"
 ---

### 2024-05-29 analyses/cell-type-ewings/template_notebooks/01-marker-genes.Rmd

@@
   library_id: SCPCL000822
-  results_dir: "../results/annotate_tumor_cells_output"
-  reference_cell_file: "../references/cell_lists/SCPCS000490/SCPCL000822_reference-cells.tsv"
+  results_dir: "annotate_tumor_cells_output"
+  reference_cell_file: "reference-cells.tsv"
 ---

### 2024-05-29 analyses/cell-type-ewings/template_notebooks/02-cellassign.Rmd

@@
   library_id: SCPCL000822
-  marker_gene_classification: "../results/sim_annotate_tumor/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
-  tumor_marker_predictions: "../results/sim_annotate_tumor/SCPCS000490/cellassign/SCPCL000822_tumor-marker_predictions.tsv"
-  visser_marker_predictions: "../results/sim_annotate_tumor/SCPCS000490/cellassign/SCPCL000822_visser-all-marker_predictions.tsv"
-  panglao_predictions: "../results/sim_annotate_tumor/SCPCS000490/cellassign/SCPCL000822_panglao-endo-fibro_predictions.tsv"
-  results_dir: "../results/sim_annotate_tumor/SCPCS000490"
+  marker_gene_classification: "tumor-normal-classifications.tsv"
+  tumor_marker_predictions: "tumor-marker_predictions.tsv"
+  visser_marker_predictions: "visser-all-marker_predictions.tsv"
+  panglao_predictions: "panglao-endo-fibro_predictions.tsv"
+  results_dir: "annotate_tumor_cells_output"
 ---

### 2024-05-29 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd
rename notebook directory and document it a bit

---
params: 
  dataset: "hm-6k"
title: "`r glue::glue('Doublet benchmarking results: {params$dataset}')`"
author: Stephanie J. Spielman
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: true
---

The goal of this notebook is to explore doublet inferences run on ground-truth benchmarking data. 
These three doublet detection methods are explored:

- `scDblFinder`
- `scrublet`
- A version of `cxds` calculated by `scDblFinder`.
Unlike the other two methods, this one reports only a score and does not attempt to make single/doublet calls, so we use thresholds of 0.5, 0.75, and 0.9 threshold to make `cxds` calls.

This notebook explores inferences made on a single dataset at a time (here, **`r params$dataset`**) with the goal of examining the relationship between scores and accuracy of calls (if made) compared to the ground-truth values.
Note that ground-truth calls are available only for heterotypic doublets, which these methods are more tuned for in the first place (vs homotypic).

## Setup

### Packages

```{r packages}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(readr)
  library(ggplot2)
  library(caret)
})

theme_set(theme_bw())
```

### Paths

#### Base directories

```{r base paths}
module_base <- rprojroot::find_root(rprojroot::is_renv_project)
data_dir <- file.path(module_base, "scratch", "benchmark-datasets", params$dataset)
result_dir <- file.path(module_base, "results", "benchmark-results")
```

#### Input and output files

Set paths to input and output directories and files in the chunk below.

```{r paths}
# Input files
scdbl_tsv <- file.path(
  result_dir, 
  glue::glue("{params$dataset}_scdblfinder.tsv")
)
scrub_tsv <- file.path(
  result_dir, 
  glue::glue("{params$dataset}_scrublet.tsv")
)
truth_tsv <- file.path(
  data_dir, 
  glue::glue("{params$dataset}_ground-truth.tsv")
)
```

### Functions

This chunk defines functions to explore doublet detection results

```{r functions}
plot_doublets <- function(df, score_col, pred_col, method_name){
  # df is expected to contain columns `ground_truth`, `score_col`, and `pred_col`
  # The column arguments should _not_ be provided as strings
  ggplot(df) +
    aes(x = ground_truth,
        y = {{score_col}}, 
        color = {{pred_col}}) + 
    geom_jitter(
      alpha = 0.2, 
      width = 0.1
    ) + 
    labs(
      x = "Ground truth call",
      y = glue::glue("{method_name} score"),
      color = glue::glue("{method_name} prediction")
    )
}

assess_doublets <- function(df, pred_col) {
  # df is expected to contain columns `ground_truth` and `pred_col`
  # The `pred_col` argument _should_ be provided as a strings

  caret::confusionMatrix(
    # truth should be first
    table(
      "Truth" = doublet_df$ground_truth,
      "Prediction" = doublet_df[[pred_col]]
    ), 
    positive = "doublet"
  )

}

```

### Read and prepare input data

First, we'll read in and combine TSV files with doublet results.

```{r}
scdbl_df <- read_tsv(scdbl_tsv) |>
  dplyr::select(
    barcodes,
    cxds_score, 
    scdbl_score = score, 
    scdbl_prediction  = class
  ) |>
  # add cxds calls at three thresholds:
  # 0.5, 0.75, 0.9
  dplyr::mutate(
    cxds_prediction_0.5 = dplyr::if_else(
      cxds_score >= 0.5,
      "doublet",
      "singlet"
    ),
    cxds_prediction_0.75 = dplyr::if_else(
      cxds_score >= 0.75,
      "doublet",
      "singlet"
    ),
    cxds_prediction_0.9 = dplyr::if_else(
      cxds_score >= 0.9,
      "doublet",
      "singlet"
    )
  )

scrub_df <- read_tsv(scrub_tsv)

true_df <- read_tsv(truth_tsv)

doublet_df <- dplyr::inner_join(
  scdbl_df,
  scrub_df, 
  by = "barcodes"
) |>
  dplyr::inner_join(
    true_df, 
    by = "barcodes"
  )
```

## Explore results: Score distributions

Both `scDblFinder` and `scrublet` make doublet calls based on an internally-determined threshold, which may not be optimal.

These sections explore the score distributions and how they relate to calls, including a confusion matrix and associated performance metrics. 
We're particularly interested in "Balanced accuracy," as there is a much higher proportion of singlets vs doublets.

### scDblFinder results

```{r}
plot_doublets(
  doublet_df,
  scdbl_score,
  scdbl_prediction,
  "scDblFinder"
)

assess_doublets(
  doublet_df,
  "scdbl_prediction"
) |>
  print()
```

### scrublet results

```{r}
plot_doublets(
  doublet_df,
  scrublet_score,
  scrublet_prediction,
  "scrublet"
)

assess_doublets(
  doublet_df,
  "scrublet_prediction"
) |>
  print()
```

### cxds results

> This method does not call droplets as singlet/doublet.
> Instead, performance is assessed below at several thresholds:

#### cxds: 0.5 threshold

```{r}
plot_doublets(
  doublet_df,
  cxds_score,
  cxds_prediction_0.5,
  "cxds, 0.5"
)

assess_doublets(
  doublet_df,
  "cxds_prediction_0.5"
) |>
  print()
```

#### cxds: 0.75 threshold

```{r}
plot_doublets(
  doublet_df,
  cxds_score,
  cxds_prediction_0.75,
  "cxds, 0.75"
)

assess_doublets(
  doublet_df,
  "cxds_prediction_0.75"
) |>
  print()
```

#### cxds: 0.9 threshold

```{r}
plot_doublets(
  doublet_df,
  cxds_score,
  cxds_prediction_0.9,
  "cxds, 0.9"
)

assess_doublets(
  doublet_df,
  "cxds_prediction_0.9"
) |>
  print()
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-05-29 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd
Update notebook to include PCAs and tables

@@
     toc: true
+    toc_depth: 4
+    code_folding: hide
 ---
@@
 The goal of this notebook is to explore doublet inferences run on ground-truth benchmarking data. 
-These three doublet detection methods are explored:
+These three doublet detection methods are explored.
 
@@
 - `scrublet`
-- A version of `cxds` calculated by `scDblFinder`.
-Unlike the other two methods, this one reports only a score and does not attempt to make single/doublet calls, so we use thresholds of 0.5, 0.75, and 0.9 threshold to make `cxds` calls.
+- A version of `cxds` [calculated by `scDblFinder`](https://github.com/plger/scDblFinder/blob/893c958322df5a4b3d364f28ed818c740de16316/R/misc.R#L268-L274).
+Unlike the other two methods, this method reports only a score and does not attempt to make single/doublet calls, so we use thresholds of 0.5, 0.75, and 0.9 threshold to make `cxds` calls.
 
@@
 Note that ground-truth calls are available only for heterotypic doublets, which these methods are more tuned for in the first place (vs homotypic).
+
+## Dataset description
+
+The following dataset description is adapted from [Xi and Li (2021)](https://doi.org/10.1016/j.cels.2020.11.008).
+
+```{r results = "asis"}
+if (params$dataset == "hm-6k") {
+  data_desc <- "The `hm-6k` dataset is a mixture of human HEK293T and mouse NIH3T3 cells with 6806 droplets."
+} else if (params$dataset == "HMEC-orig-MULTI") {
+  data_desc <- "The `HMEC-orig-MULTI` dataset contains human primary mammary epithelial cells (HMECs) with 26426 droplets."
+} else if (params$dataset == "pdx-MULTI") {
+  data_desc <- "The `pdx-MULTI` dataset is a mixture of human breast cancer cells and mouse immune cells from a PDX mouse model."
+} else if (params$dataset == "pbmc-1B-dm") {
+  data_desc <- "The `pbmc-1B-dm` dataset is a PMBC sample from a patient with systemic lupus erythematosus."
+} else {
+  stop("Unknown dataset provided.")
+}
+
+data_desc |>
+  glue::glue() |>
+  print()
+```
 
@@
   library(ggplot2)
+  library(patchwork)
   library(caret)
@@
 ```{r paths}
-# Input files
+# doublet calls
 scdbl_tsv <- file.path(
@@
 )
-truth_tsv <- file.path(
+
+# SCE, with ground truth included
+sce_file <- file.path(
   data_dir, 
-  glue::glue("{params$dataset}_ground-truth.tsv")
+  glue::glue("{params$dataset}_sce.rds")
 )
+
 ```
@@
 
-This chunk defines functions to explore doublet detection results
+This chunk defines functions to explore doublet detection results.
 
 ```{r functions}
-plot_doublets <- function(df, score_col, pred_col, method_name){
+plot_pca_d <- function(df, color_column, color_lab) {
+  # Plot PC with a discrete color variable, showing doublets on top
+  # df is expected to contain columns PC1, PC2, and `color_column`
+  # The `color_column` argument should _not_ be provided as a string
+  ggplot(doublet_df) + 
+    aes(x = PC1, 
+        y = PC2, 
+        color = {{color_column}}) +
+  geom_point(alpha = 0.5) + 
+  scale_color_manual(values = c("black", "lightblue")) + 
+  # ensure doublets are on top for visibility
+  geom_point(
+    data = dplyr::filter(doublet_df, {{color_column}} == "doublet"), 
+    color = "black"
+  ) + 
+  labs(color = color_lab) + 
+  theme(legend.position = "bottom")
+}
+
+plot_pca_c <- function(df, color_column, pred_column, color_lab) {
+  # Plot PC with a continuous color variable, showing doublets on top
+  # df is expected to contain columns PC1, PC2, `color_column`, and `pred_column`
+  # The `color_column` and `pred_column` arguments should _not_ be provided as strings
+  ggplot(doublet_df) + 
+    aes(x = PC1, 
+        y = PC2, 
+        color = {{color_column}}) +
+  geom_point(alpha = 0.5) + 
+  scale_color_viridis_c(direction = -1) + 
+  # ensure doublets are on top for visibility
+  geom_point(
+    data = dplyr::filter(doublet_df, {{pred_column}} == "doublet"), 
+    aes(color = {{color_column}})
+  ) + 
+  labs(color = color_lab) + 
+  theme(legend.position = "bottom")
+}
+
+plot_doublet_scores <- function(df, score_col, pred_col, method_name){
   # df is expected to contain columns `ground_truth`, `score_col`, and `pred_col`
@@
       color = glue::glue("{method_name} prediction")
-    )
+    ) + 
+  theme(legend.position = "bottom")
 }
@@
 
-true_df <- read_tsv(truth_tsv)
+# grab ground truth and PCA coordinates
+sce <- read_rds(sce_file)
+sce_df <- scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
+  tibble::rownames_to_column(var = "barcodes") |>
+  dplyr::select(barcodes,
+                ground_truth = ground_truth_doublets, 
+                PC1 = PCA.1, 
+                PC2 = PCA.2)
 
@@
   dplyr::inner_join(
-    true_df, 
+    sce_df, 
     by = "barcodes"
@@
 
-## Explore results: Score distributions
+## Ground truth
+
+First, we assess the ground truth doublet distributions to contextualize the doublet calls.
+
+```{r}
+dub_table <- table(doublet_df$ground_truth)
+
+dub_percent <- round(dub_table[["doublet"]]/sum(dub_table), 2) * 100
+dub_table
+```
+`r dub_percent`% of droplets in this dataset are doublets.
+
+```{r}
+plot_pca_d(doublet_df, ground_truth, "Ground truth doublets")
+```
+
+## Doublet predictions and score distributions
 
 Both `scDblFinder` and `scrublet` make doublet calls based on an internally-determined threshold, which may not be optimal. 
+These sections explore the score distributions and how they relate to calls, including:
 
-These sections explore the score distributions and how they relate to calls, including a confusion matrix and associated performance metrics. 
-We're particularly interested in "Balanced accuracy," as there is a much higher proportion of singlets vs doublets.
+- PCAs of predicted doublet vs. singlets
+  - The first PCA is colored by doublet scores, where scores closer to 1 are more likely to be doublets
+  - The second PCA is colored by doublet predictions made by the given method and/or threshold.
+- Jitter plots of doublet scores
+- A confusion with associated performance metrics
+  - We're particularly interested in "Balanced accuracy," as there is a much higher proportion of singlets vs doublets.
 
@@
 ```{r}
-plot_doublets(
+table(doublet_df$scdl_prediction)
+```
+
+#### PCA plots
+
+```{r fig.width = 8}
+d_plot <- plot_pca_d(
+  doublet_df, 
+  scdbl_prediction,
+  glue::glue("scDblFinder prediction")
+)
+
+c_plot <- plot_pca_c(
   doublet_df, 
@@
   scdbl_prediction,
-  "scDblFinder"
+  glue::glue("scDblFinder score")
 )
 
+d_plot + c_plot
+```
+
+#### Doublet score distributions
+
+```{r}
+plot_doublet_scores(
+  doublet_df,
+  scdbl_score,
+  scdbl_prediction,
+  "scDblFinder prediction"
+)
+```
+
+#### Confusion matrix and performance metrics
+
+```{r}
 assess_doublets(
@@
 
-```{r}
-plot_doublets(
+#### PCA plots
+
+```{r fig.width = 8}
+d_plot <- plot_pca_d(
+  doublet_df, 
+  scrublet_prediction,
+  glue::glue("scrublet prediction")
+)
+
+c_plot <- plot_pca_c(
   doublet_df, 
@@
   scrublet_prediction,
-  "scrublet"
+  glue::glue("scrublet score")
 )
 
+d_plot + c_plot
+```
+
+#### Doublet score distributions
+
+```{r}
+plot_doublet_scores(
+  doublet_df,
+  scrublet_score,
+  scrublet_prediction,
+  "scrublet prediction"
+)
+```
+
+#### Confusion matrix and performance metrics
+
+```{r}
 assess_doublets(
@@
 > This method does not call droplets as singlet/doublet.
-> Instead, performance is assessed below at several thresholds:
+> Instead, performance is assessed below at several thresholds.
+
+#### PCA plots
+
+```{r fig.width = 10}
+
+cxds_cols <- names(doublet_df)[grep("^cxds_prediction_", names(doublet_df))]
+
+prediction_pcas <- cxds_cols |>
+  purrr::map(
+    \(col_name) {
+      threshold <- stringr::str_remove(col_name, "cxds_prediction_")
+
+      plot_pca_d(
+        doublet_df, 
+        !!as.symbol(col_name),
+        glue::glue("cxds prediction at {threshold}")
+      ) +
+      labs(title = glue::glue("cxds prediction at {threshold}"))
+    }
+)
+wrap_plots(prediction_pcas)
+```
+
+```{r}
+# just need a single score pca for this case
+# we'll provide the least stringent predictions; this only affects point layering in the plot
+pca_c <- plot_pca_c(
+  doublet_df, 
+  cxds_score, 
+  cxds_prediction_0.5,
+  glue::glue("cxds scores")
+)
+wrap_plots(pca_c)
+```
 
@@
 
+##### Doublet score distributions
+
 ```{r}
-plot_doublets(
+plot_doublet_scores(
   doublet_df,
@@
   cxds_prediction_0.5,
-  "cxds, 0.5"
+  "cxds prediction at 0.5"
 )
+```
 
+##### Confusion matrix and performance metrics
+
+```{r}
 assess_doublets(
@@
   cxds_prediction_0.75,
-  "cxds, 0.75"
+  "cxds prediction at 0.75"
 )
+```
 
+##### Confusion matrix and performance metrics
+
+```{r}
 assess_doublets(
@@
   cxds_prediction_0.9,
-  "cxds, 0.9"
+  "cxds prediction at 0.9"
 )
+```
 
+##### Confusion matrix and performance metrics
+
+```{r}
 assess_doublets(

### 2024-05-29 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd
left_join, density plots, and other styling

@@
   labs(color = color_lab) + 
-  theme(legend.position = "bottom")
+  theme(
+    legend.title.position = "top",
+    legend.position = "bottom"
+  )
 }
@@
   labs(color = color_lab) + 
-  theme(legend.position = "bottom")
+  theme(
+    legend.title.position = "top",
+    legend.position = "bottom"
+  )
 }
 
-plot_doublet_scores <- function(df, score_col, pred_col, method_name){
+plot_jitter_scores <- function(df, score_col, pred_col, method_name){
+  # Plot jitter plot of score distributions and predictions, colored by ground truth
   # df is expected to contain columns `ground_truth`, `score_col`, and `pred_col`
@@
     ) + 
-  theme(legend.position = "bottom")
+  theme(
+    legend.title.position = "top",
+    legend.position = "bottom"
+  )
+}
+
+plot_density_scores <- function(df, score_col, method_name){
+  # Plot density plot of score distributions, colored by ground truth
+  # df is expected to contain columns `ground_truth` and `score_col`
+  # The column arguments should _not_ be provided as strings
+  ggplot(df) +
+    aes(x = {{score_col}}, 
+        fill = ground_truth) + 
+    geom_density(alpha = 0.7) + 
+    labs(
+      x = glue::glue("{method_name} score"),
+      fill = "Ground truth call"
+    ) + 
+  theme(
+    legend.title.position = "top",
+    legend.position = "bottom"
+  )
 }
@@
 
-doublet_df <- dplyr::inner_join(
+# we can now remove the sce
+rm(sce)
+
+doublet_df <- dplyr::left_join(
   scdbl_df,
@@
 ) |>
-  dplyr::inner_join(
+  dplyr::left_join(
     sce_df, 
@@
 ```{r}
-table(doublet_df$scdl_prediction)
+table(doublet_df$scdbl_prediction)
+```
+
+#### Score distributions relative to ground truth
+
+```{r}
+plot_density_scores(
+  doublet_df, 
+  scdbl_score, 
+  "scDblFinder"
+)
 ```
@@
   scdbl_prediction,
-  glue::glue("scDblFinder prediction")
+  "scDblFinder prediction"
 )
@@
   scdbl_prediction,
-  glue::glue("scDblFinder score")
+  "scDblFinder score"
 )
@@
 
-#### Doublet score distributions
+#### Jitter plot
 
 ```{r}
-plot_doublet_scores(
+plot_jitter_scores(
   doublet_df,
@@
 
+#### Score distributions relative to ground truth
+
+```{r}
+plot_density_scores(
+  doublet_df, 
+  scrublet_score, 
+  "scrublet"
+)
+```
+
 #### PCA plots
@@
   scrublet_prediction,
-  glue::glue("scrublet prediction")
+  "scrublet prediction"
 )
@@
   scrublet_prediction,
-  glue::glue("scrublet score")
+  "scrublet score"
 )
@@
 d_plot + c_plot
-```
-
-#### Doublet score distributions
-
-```{r}
-plot_doublet_scores(
-  doublet_df,
-  scrublet_score,
-  scrublet_prediction,
-  "scrublet prediction"
-)
 ```
@@
 
-> This method does not call droplets as singlet/doublet.
-> Instead, performance is assessed below at several thresholds.
+Since this method does not call droplets as singlet/doublet, performance is assessed below at several thresholds: 0.5, 0.75, and 0.9.
+
+#### Score distributions relative to ground truth
+
+```{r}
+plot_density_scores(
+  doublet_df, 
+  cxds_score, 
+  "cxds"
+)
+```
 
@@
 
-##### Doublet score distributions
+##### Jitter plot
 
 ```{r}
-plot_doublet_scores(
+plot_jitter_scores(
   doublet_df,

### 2024-05-29 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd
table printing

@@
 ```{r functions}
+print_doublet_table <- function(df, pred_col, is_ground_truth = FALSE) {
+  # Function to print a table of doublet counts and a sentence stating the percentage of doublets
+  # df is expected to contain `pred_col`, should should be provided as a string
+  dub_table <- table(df[[pred_col]])
+  dub_percent <- round(dub_table[["doublet"]]/sum(dub_table), 4) * 100
+
+  print(dub_table)
+
+  if (is_ground_truth) {
+    output <- glue::glue("{dub_percent}% of droplets in this dataset are doublets.") 
+  } else {
+    output <- glue::glue("{dub_percent}% of droplets in this dataset were predicted to be doublets.")
+  }
+
+  return(output)
+}
+
 plot_pca_d <- function(df, color_column, color_lab) {
@@
     labs(
-      x = "Ground truth call",
+      x = "Ground truth",
       y = glue::glue("{method_name} score"),
@@
 ```{r}
-dub_table <- table(doublet_df$ground_truth)
+percent <- print_doublet_table(doublet_df, "ground_truth", is_ground_truth = TRUE)
+```
 
-dub_percent <- round(dub_table[["doublet"]]/sum(dub_table), 2) * 100
-dub_table
+```{r results = "asis"}
+percent
 ```
-`r dub_percent`% of droplets in this dataset are doublets.
 
@@
 ```{r}
-table(doublet_df$scdbl_prediction)
+percent <- print_doublet_table(doublet_df, "scdbl_prediction")
+```
+
+```{r results = "asis"}
+percent
 ```
@@
 ### scrublet results
+
+```{r}
+percent <- print_doublet_table(doublet_df, "scrublet_prediction")
+```
+
+```{r results = "asis"}
+percent
+```
 
@@
 
+```{r}
+percent <- print_doublet_table(doublet_df, "cxds_prediction_0.5")
+```
+
+```{r results = "asis"}
+percent
+```
+
 ##### Jitter plot
@@
 
+```{r}
+percent <- print_doublet_table(doublet_df, "cxds_prediction_0.75")
+```
+
+```{r results = "asis"}
+percent
+```
+
 ##### Jitter plot
@@
 #### cxds: 0.9 threshold
+
+```{r}
+percent <- print_doublet_table(doublet_df, "cxds_prediction_0.9")
+```
+
+```{r results = "asis"}
+percent
+```
 

### 2024-05-30 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd
fix table function usage to avoid duplication, and consolidate pca function

@@
   # df is expected to contain `pred_col`, should should be provided as a string
-  dub_table <- table(df[[pred_col]])
-  dub_percent <- round(dub_table[["doublet"]]/sum(dub_table), 4) * 100
-
-  print(dub_table)
+  dub_tab <- table(df[[pred_col]])
+  dub_percent <- round(dub_tab[["doublet"]]/sum(dub_tab), 4) * 100
 
@@
 
-  return(output)
+  return(
+    list(
+      "table" = dub_tab, 
+      "percent" = output
+    )
+  )
 }
 
-plot_pca_d <- function(df, color_column, color_lab) {
-  # Plot PC with a discrete color variable, showing doublets on top
-  # df is expected to contain columns PC1, PC2, and `color_column`
-  # The `color_column` argument should _not_ be provided as a string
-  ggplot(doublet_df) + 
+plot_pca <- function(df, 
+                     color_column, 
+                     pred_column = NULL,
+                     color_type = "d",
+                     color_lab) {
+  # Plot PCs colored by either discrete or continuous color variable, showing doublets on top
+  # df is expected to contain columns PC1, PC2, `color_column`, and if color_type == "c" also `pred_column`. These should _not_ be provided as strings
+  # By default, this function assumes a discrete color variable (`color_type = "d"`) with two levels. Use `color_type = "c"` for continuous.
+
+  p <- ggplot(doublet_df) + 
     aes(x = PC1, 
@@
   geom_point(alpha = 0.5) + 
-  scale_color_manual(values = c("black", "lightblue")) + 
-  # ensure doublets are on top for visibility
-  geom_point(
-    data = dplyr::filter(doublet_df, {{color_column}} == "doublet"), 
-    color = "black"
-  ) + 
   labs(color = color_lab) + 
@@
   )
-}
 
-plot_pca_c <- function(df, color_column, pred_column, color_lab) {
-  # Plot PC with a continuous color variable, showing doublets on top
-  # df is expected to contain columns PC1, PC2, `color_column`, and `pred_column`
-  # The `color_column` and `pred_column` arguments should _not_ be provided as strings
-  ggplot(doublet_df) + 
-    aes(x = PC1, 
-        y = PC2, 
-        color = {{color_column}}) +
-  geom_point(alpha = 0.5) + 
+  # Set the palette, and ensure doublets are on top for visibility
+  if (color_type == "d") {
+    p <- p + 
+      scale_color_manual(values = c("black", "lightblue")) + 
+    geom_point(
+      data = dplyr::filter(doublet_df, {{color_column}} == "doublet"), 
+      color = "black"
+    )
+  } else if (color_type == "c") {
+    p <- p + 
       scale_color_viridis_c(direction = -1) + 
-  # ensure doublets are on top for visibility
     geom_point(
@@
       aes(color = {{color_column}})
-  ) + 
-  labs(color = color_lab) + 
-  theme(
-    legend.title.position = "top",
-    legend.position = "bottom"
     )
+  }
+
+  return(p)
 }
@@
 ```{r}
-percent <- print_doublet_table(doublet_df, "ground_truth", is_ground_truth = TRUE)
+dub_tab <- print_doublet_table(doublet_df, "ground_truth", is_ground_truth = TRUE)
+dub_tab$table
 ```
@@
 ```{r results = "asis"}
-percent
+dub_tab$percent
 ```
@@
 ```{r}
-plot_pca_d(doublet_df, ground_truth, "Ground truth doublets")
+plot_pca(doublet_df, ground_truth, color_lab = "Ground truth doublets")
 ```
@@
 ```{r}
-percent <- print_doublet_table(doublet_df, "scdbl_prediction")
+dub_tab <- print_doublet_table(doublet_df, "scdbl_prediction")
+dub_tab$table
 ```
@@
 ```{r fig.width = 8}
-d_plot <- plot_pca_d(
+
+d_plot <- plot_pca(
   doublet_df, 
-  scdbl_prediction,
-  "scDblFinder prediction"
+  color_column = scdbl_prediction,
+  color_lab = "scDblFinder prediction"
 )
 
-c_plot <- plot_pca_c(
+c_plot <- plot_pca(
   doublet_df, 
-  scdbl_score,
-  scdbl_prediction,
-  "scDblFinder score"
+  color_column = scdbl_score,
+  pred_column = scdbl_prediction,
+  color_type = "c",
+  color_lab = "scDblFinder score"
 )
@@
 ```{r}
-percent <- print_doublet_table(doublet_df, "scrublet_prediction")
+dub_tab <- print_doublet_table(doublet_df, "scrublet_prediction")
+dub_tab$table
 ```
@@
 ```{r fig.width = 8}
-d_plot <- plot_pca_d(
+d_plot <- plot_pca(
   doublet_df, 
-  scrublet_prediction,
-  "scrublet prediction"
+  color_column = scrublet_prediction,
+  color_lab = "scrublet prediction"
 )
 
-c_plot <- plot_pca_c(
+c_plot <- plot_pca(
   doublet_df, 
-  scrublet_score,
-  scrublet_prediction,
-  "scrublet score"
+  color_column = scrublet_score,
+  pred_column = scrublet_prediction,
+  color_type = "c",
+  color_lab = "scrublet score"
 )
@@
 
-      plot_pca_d(
+      plot_pca(
         doublet_df, 
-        !!as.symbol(col_name),
-        glue::glue("cxds prediction at {threshold}")
+        color_column = !!as.symbol(col_name),
+        color_lab = glue::glue("cxds prediction at {threshold}")
       ) +
@@
 # we'll provide the least stringent predictions; this only affects point layering in the plot
-pca_c <- plot_pca_c(
+plot_pca(
   doublet_df, 
-  cxds_score, 
-  cxds_prediction_0.5,
-  glue::glue("cxds scores")
+  color_column = cxds_score, 
+  pred_column = cxds_prediction_0.5,
+  color_type = "c",
+  color_lab = glue::glue("cxds scores")
 )
-wrap_plots(pca_c)
 ```
@@
 ```{r}
-percent <- print_doublet_table(doublet_df, "cxds_prediction_0.5")
+dub_tab <- print_doublet_table(doublet_df, "cxds_prediction_0.5")
+dub_tab$table
 ```
@@
 ```{r}
-percent <- print_doublet_table(doublet_df, "cxds_prediction_0.75")
+dub_tab <- print_doublet_table(doublet_df, "cxds_prediction_0.75")
+dub_tab$table
 ```
@@
 ```{r}
-percent <- print_doublet_table(doublet_df, "cxds_prediction_0.9")
+dub_tab <- print_doublet_table(doublet_df, "cxds_prediction_0.9")
+dub_tab$table
 ```

### 2024-05-30 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd
add missing jitter plot
ground truth context

@@
 The goal of this notebook is to explore doublet inferences run on ground-truth benchmarking data. 
-These three doublet detection methods are explored.
+Importantly, the ground-truth doublets in these benchmarking datasets were themselves called with computational approaches in the context of multiplexed experiments.
+This means the true values are not necessarily entirely accurate.
+
+These three doublet detection methods are explored:
 
@@
 
+#### Jitter plot
+
+```{r}
+plot_jitter_scores(
+  doublet_df,
+  scrublet_score,
+  scrublet_prediction,
+  "scrublet prediction"
+)
+```
+
 #### Confusion matrix and performance metrics

### 2024-05-30 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd
bug fix - wrong variable used in function

@@
 
-  p <- ggplot(doublet_df) + 
+  p <- ggplot(df) + 
     aes(x = PC1, 
@@
     geom_point(
-      data = dplyr::filter(doublet_df, {{color_column}} == "doublet"), 
+      data = dplyr::filter(df, {{color_column}} == "doublet"), 
       color = "black"
@@
     geom_point(
-      data = dplyr::filter(doublet_df, {{pred_column}} == "doublet"), 
+      data = dplyr::filter(df, {{pred_column}} == "doublet"), 
       aes(color = {{color_column}})

### 2024-05-30 analyses/cell-type-ewings/template_notebooks/02-cellassign.Rmd
account for only one cell type and skip confusion matrix

@@
 ```{r, eval=has_tumor_assignments}
+# make sure that we only run if both classes are found in both cell types 
+if(all(c("other", "tumor") %in% celltype_df$cellassign_annotations) &&
+   all(c("Tumor", "Normal") %in% celltype_df$marker_gene_classification)){
   # prep annotations for confusion matrix
@@
   ) 
+}
 ```

### 2024-05-31 analyses/doublet-detection/template-notebooks/03_compare-benchmark-results.Rmd
extremely WIP comparison notebook

---
title: "Comparison among doublet predictions on benchmarking datasets"
author: Stephanie J. Spielman
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: true
    toc_depth: 4
    code_folding: hide
---

## Setup

### Packages

```{r packages}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(ggplot2)
  library(patchwork)
  library(caret)
  library(UpSetR)
})

theme_set(theme_bw())
```

### Paths

```{r base paths}
module_base <- rprojroot::find_root(rprojroot::is_renv_project)
data_dir <- file.path(module_base, "scratch", "benchmark-datasets")
result_dir <- file.path(module_base, "results", "benchmark-results")
```

### Functions

Import shared functions:
```{r}

plot_pca <- function(df, 
                     color_column, 
                     pred_column = NULL,
                     color_type = "d",
                     color_lab) {
  # Plot PCs colored by either discrete or continuous color variable, showing doublets on top
  # df is expected to contain columns PC1, PC2, `color_column`, and if color_type == "c" also `pred_column`. These should _not_ be provided as strings
  # By default, this function assumes a discrete color variable (`color_type = "d"`) with two levels. Use `color_type = "c"` for continuous.

  p <- ggplot(df) + 
    aes(x = PC1, 
        y = PC2, 
        color = {{color_column}}) +
  geom_point(alpha = 0.5) + 
  labs(color = color_lab) + 
  theme(
    legend.title.position = "top",
    legend.position = "bottom"
  )

  # Set the palette, and ensure doublets are on top for visibility
  if (color_type == "d") {
    p <- p + 
      scale_color_manual(values = c("black", "lightblue")) + 
    geom_point(
      data = dplyr::filter(df, {{color_column}} == "doublet"), 
      color = "black"
    )
  } else if (color_type == "c") {
    p <- p + 
      scale_color_viridis_c(direction = -1) + 
    geom_point(
      data = dplyr::filter(df, {{pred_column}} == "doublet"), 
      aes(color = {{color_column}})
    )
  }

  return(p)
}
```

### Read and prepare input data

First, we'll read in and combine doublet results into a list of data frames for each dataset.
We'll also two columns for each dataset:

- `consensus_call`, which will be "doublet" if _all_ methods predict "doublet," and "singlet" otherwise
- `call_type`, which will classify the consensus call as one of "tp", "tn", "fp", or "fn" (true/false positive/negative)

```{r paths}
# find all dataset names to process:
dataset_names <- list.files(result_dir, pattern = "*_scrublet.tsv") |>
  stringr::str_remove("_scrublet.tsv")
```

```{r read_data}
# Read in and data for analysis
doublet_df_list <- dataset_names |>
  purrr::map(
    \(dataset) {

      scdbl_tsv <- file.path(result_dir, glue::glue("{dataset}_scdblfinder.tsv"))
      scrub_tsv <- file.path(result_dir, glue::glue("{dataset}_scrublet.tsv"))
      sce_file <- file.path(data_dir, dataset, glue::glue("{dataset}_sce.rds"))

      scdbl_df <- scdbl_tsv |>
        readr::read_tsv(show_col_types = FALSE) |>
        dplyr::select(
          barcodes,
          cxds_score, 
          scdbl_score = score, 
          scdbl_prediction  = class
        ) |>
        # add cxds calls at 0.75 threshold
        dplyr::mutate(
          cxds_prediction = dplyr::if_else(
            cxds_score >= 0.75,
            "doublet",
            "singlet"
          )
        )

      scrub_df <- readr::read_tsv(scrub_tsv, show_col_types = FALSE)

      # grab ground truth and PCA coordinates
      sce <- readr::read_rds(sce_file)
      sce_df <- scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
        tibble::rownames_to_column(var = "barcodes") |>
        dplyr::select(barcodes,
                      ground_truth = ground_truth_doublets, 
                      PC1 = PCA.1, 
                      PC2 = PCA.2)
      rm(sce)

      dataset_df <- scdbl_df |>
        dplyr::left_join(
          scrub_df, 
          by = "barcodes"
        ) |>
        dplyr::left_join(
          sce_df, 
          by = "barcodes"
        )

      # Add a consensus call column
      dataset_df <- dataset_df |>
        dplyr::rowwise() |>
        dplyr::mutate(consensus_call = dplyr::if_else(
          all(
            c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"
          ),
          "doublet", 
          "singlet"
        )) |>
        dplyr::mutate(
          call_type = dplyr::case_when(
            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
          ), 
          # set associated plotting colors
          call_type_color = dplyr::case_when(
            call_type == "tp" ~ "blue4",
            call_type == "tn" ~ "blue",
            call_type == "fp" ~ "brown1",
            call_type == "fn" ~ "darkorange1"
          )
          )

      return(dataset_df)
    }
  )
names(doublet_df_list) <- dataset_names
```

## Viz: PCA

```{r, fig.width = 10}
doublet_df_list |>
  purrr::iwalk(
    \(df, dataset) {

      # First, ground truth
      p1 <- plot_pca(
        df, 
        color_column = ground_truth, 
        color_lab = "Ground truth"
      )

      # Second, consensus call
      p2 <- plot_pca(
        df, 
        color_column = consensus_call, 
        color_lab = "Consensus call"
      )

      # Third, call type
      p3 <- ggplot(df) + 
        aes(x = PC1, y = PC2, color = call_type_color) + 
        geom_point(alpha = 0.6, size = 0.25) + 
        scale_color_identity()

      # combine and plot
      plot( p1 + p2 + p3 + plot_annotation(glue::glue("PCA for {dataset}")) )
    }
  )
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-06-03 analyses/doublet-detection/template-notebooks/03_compare-benchmark-results.Rmd
PCA section for comparison notebook

@@
 
-Import shared functions:
 ```{r}
-
-plot_pca <- function(df, 
+plot_pca_calls <- function(df, 
                            color_column, 
-                     pred_column = NULL,
-                     color_type = "d",
+                           pred_column,
                            color_lab) {
-  # Plot PCs colored by either discrete or continuous color variable, showing doublets on top
-  # df is expected to contain columns PC1, PC2, `color_column`, and if color_type == "c" also `pred_column`. These should _not_ be provided as strings
-  # By default, this function assumes a discrete color variable (`color_type = "d"`) with two levels. Use `color_type = "c"` for continuous.
-
-  p <- ggplot(df) + 
+  # Plot PCs colored by singlet/doublet, showing doublets on top
+  # df is expected to contain columns PC1, PC2, `color_column`, and `pred_column`. These should _not_ be provided as strings
+  ggplot(df) + 
     aes(x = PC1, 
@@
         color = {{color_column}}) +
-  geom_point(alpha = 0.5) + 
-  labs(color = color_lab) + 
+  geom_point(
+    size = 0.75, 
+    alpha = 0.6
+  ) +
+  scale_color_manual(name = color_lab, values = c("black", "lightblue")) + 
+  geom_point(
+    data = dplyr::filter(df, {{color_column}} == "doublet"), 
+    color = "black",
+    size = 0.75
+  ) +
   theme(
@@
   )
-
-  # Set the palette, and ensure doublets are on top for visibility
-  if (color_type == "d") {
-    p <- p + 
-      scale_color_manual(values = c("black", "lightblue")) + 
-    geom_point(
-      data = dplyr::filter(df, {{color_column}} == "doublet"), 
-      color = "black"
-    )
-  } else if (color_type == "c") {
-    p <- p + 
-      scale_color_viridis_c(direction = -1) + 
-    geom_point(
-      data = dplyr::filter(df, {{pred_column}} == "doublet"), 
-      aes(color = {{color_column}})
-    )
 }
 
-  return(p)
+plot_pca_metrics <- function(df, color_column, false_colors) {
+  # Plot PCs colored by performance metric, showing false calls on top
+  # false_colors is a vector of colors used for fn and fp points
+  # df is expected to contain columns PC1, PC2, and `color_column`. This should _not_ be provided as a string.
+  ggplot(df) + 
+    aes(x = PC1, 
+        y = PC2, 
+        color = {{color_column}}) +
+  geom_point(
+    size = 0.75, 
+    alpha = 0.6
+  ) + 
+  geom_point(
+    data = dplyr::filter(df, {{color_column}} %in% false_colors), 
+    size = 0.75
+  ) +
+  scale_color_identity() +
+  theme(legend.position = "none")
 }
@@
 ```{r read_data}
+confusion_colors <- c(
+  "tp" = "lightblue",
+  "tn" = "pink",
+  "fp" = "blue",
+  "fn" = "firebrick2"
+)
+
 # Read in and data for analysis
@@
           call_type_color = dplyr::case_when(
-            call_type == "tp" ~ "blue4",
-            call_type == "tn" ~ "blue",
-            call_type == "fp" ~ "brown1",
-            call_type == "fn" ~ "darkorange1"
+            call_type == "tp" ~ unname(confusion_colors["tp"]),
+            call_type == "tn" ~ unname(confusion_colors["tn"]),
+            call_type == "fp" ~ unname(confusion_colors["fp"]),
+            call_type == "fn" ~ unname(confusion_colors["fn"])
           )
@@
 
-## Viz: PCA
+## Visualizations
 
-```{r, fig.width = 10}
+### PCA
+
+This section plots the PCA for each dataset, with three color schemes from left to right:
+- Ground truth doublets are shown in black
+- Consensus doublets are shown in black
+- Points are colored as tp, tn, fp, or fn based on comparing the consensus call to the ground truth
+
+```{r, fig.width = 12, fig.height = 6}
+# Make a legend for the confusion-colored PCA
+legend_plot <- data.frame(
+  x = factor(names(confusion_colors), levels = names(confusion_colors)), y = 1:4
+) |>
+ ggplot(aes(x = x, y = y, color = x)) + 
+  geom_point(size = 3) + 
+  scale_color_manual(name = "Metric", values = confusion_colors) 
+confusion_legend <- ggpubr::get_legend(legend_plot) |> ggpubr::as_ggplot()
+
 doublet_df_list |>
@@
       # First, ground truth
-      p1 <- plot_pca(
+      p1 <- plot_pca_calls(
         df, 
@@
       # Second, consensus call
-      p2 <- plot_pca(
+      p2 <- plot_pca_calls(
         df, 
@@
       # Third, call type
-      p3 <- ggplot(df) + 
-        aes(x = PC1, y = PC2, color = call_type_color) + 
-        geom_point(alpha = 0.6, size = 0.25) + 
-        scale_color_identity()
+      p3 <- plot_pca_metrics(
+        df,
+        call_type_color,
+        false_colors = unname(c(confusion_colors["fn"], confusion_colors["fp"]))
+      )
 
       # combine and plot
-      plot( p1 + p2 + p3 + plot_annotation(glue::glue("PCA for {dataset}")) )
+      plot( p1 + p2 + p3 + confusion_legend + plot_annotation(glue::glue("PCA for {dataset}")) + plot_layout(ncol=4, widths = c(1,1,1,0.25)) )
     }

### 2024-06-03 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
add section with metrics

--- 2024-06-03 analyses/doublet-detection/template-notebooks/03_compare-benchmark-results.Rmd
@@
 
+## Performance metrics
+
+This section shows performance metrics for the consensus calls for each dataset.
+
+```{r}
+doublet_df_list |>
+  purrr::iwalk( 
+    \(df, dataset) {
+        print(glue::glue("======================== {dataset} ========================"))
+
+        cat("Table of consensus calls:")
+        print(table(df$consensus_call))
+
+        cat("\n\n")
+
+        caret::confusionMatrix(
+          # truth should be first
+          table(
+            "Truth" = df$ground_truth,
+            "Consensus prediction" = df$consensus_call
+          ), 
+          positive = "doublet"
+        ) |>
+        print()
+    }
+  )
+```
+
 ## Visualizations

### 2024-06-03 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
analysis portion of notebook largely completed, and next it needs interpretation

@@
 ```{r read_data}
+# used in PCA plots
 confusion_colors <- c(
@@
 
+### Upset plots
+
+This section shows upset plots for overlap among methods for each dataset.
+
+```{r}
+pull_barcodes <- function(df, pred_var) {
+  # Helper function to pull out barcodes for doublet calls
+  df$barcodes[df[[pred_var]] == "doublet"]
+}
+
+upset_list <- doublet_df_list |>
+  purrr::iwalk(
+    \(df, dataset) {
+
+      doublet_barcodes <- list(
+        "scDblFinder" = pull_barcodes(df, "scdbl_prediction"),
+        "scrublet"    = pull_barcodes(df, "scrublet_prediction"),
+        "cxds"        = pull_barcodes(df, "cxds_prediction")
+      )
+
+      UpSetR::upset(fromList(doublet_barcodes), order.by = "freq") |> print()
+      grid::grid.text( # plot title
+        dataset,
+        x = 0.65, 
+        y = 0.95, 
+        gp = grid::gpar(fontsize=16)
+      )
+
+    }
+  )
+
+```
+
 ## Session Info

### 2024-06-03 analyses/cell-type-ewings/exploratory_analysis/03-copykat.Rmd

--- 2024-05-13 analyses/ewings-cell-type/03-copykat.Rmd
@@
 ---
-title: "Using CopyKAT to idenfity tumor cells in Ewing sarcoma samples"
+title: "Using CopyKAT to identify tumor cells in Ewing sarcoma samples"
 author: Ally Hawkins
@@
 
-- Visualize diploid and anueploid cells on the UMAP. 
+- Visualize diploid and aneuploid cells on the UMAP. 
 - Evaluate common copy number gains and losses in Ewing sarcoma. 
@@
 # The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "current")
+data_dir <- file.path(repository_base, "data", "2024-05-01")
 sample_dir <- file.path(data_dir, "SCPCP000015", params$sample_id)
@@
 
-If we look at the UMAPs shown in `01-marker-gene-tumor-classification.Rmd`, the cells that are labeled aneuploid here are consistent with where we saw marker gene expession for tumor cells. 
+If we look at the UMAPs shown in `01-marker-gene-tumor-classification.Rmd`, the cells that are labeled aneuploid here are consistent with where we saw marker gene expression for tumor cells. 
 It also looks like there is not a visible difference between using a reference of normal cells and not using a reference.
@@
 It looks like the largest increases are in chr8 and chr21. 
-Chr8 is one of the more common CNA's in Ewing's so this is a good sign.
+Chr8 is one of the more common CNAs in Ewing's so this is a good sign.
 
@@
 We see that aneuploid cells are mostly muscle cells in both `CellAssign` and `SingleR`. 
-In `SingleR` there are some other cell types that were assigned that have a few anueploid cells (chondrocytes, neurons, and fibroblasts).
+In `SingleR` there are some other cell types that were assigned that have a few aneuploid cells (chondrocytes, neurons, and fibroblasts).
 

### 2024-06-03 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd

@@
 
-Below is the heatmap produced by `InferCNV` showing the CNV's for each cell using endothelial cells as a reference. 
+Below is the heatmap produced by `InferCNV` showing the CNVs for each cell using endothelial cells as a reference. 
 Each row is a cell and each column is a gene. 
@@
 
-Below is the heatmap produced by `InferCNV` showing the CNV's for each cell without using a reference.
+Below is the heatmap produced by `InferCNV` showing the CNVs for each cell without using a reference.
 
@@
 This shows that the total CNVs are highest in the group of cells that correspond to the same cells that are manually annotated as tumor cells, which is promising. 
-Those cells in grey (bottom left) are cells used as a reference. 
+Those cells in gray (bottom left) are cells used as a reference. 
 If we were to use total CNVs we would need some sort of cut off for the number. 

### 2024-06-03 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd
show images as jpeg

@@
 })
+
+knitr::opts_chunk$set(
+  dev = "jpeg"
+)
 ```
@@
 
-Below is the heatmap produced by `InferCNV` showing the CNV's for each cell using normal cells as a reference. 
+Below is the heatmap produced by `InferCNV` showing the CNVs for each cell using normal cells as a reference. 
 Each row is a cell and each column is a gene. 

### 2024-06-04 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
Some reorg for logical flow, intro and conclusions section

@@
 ---
+
+This notebook compares doublet results calculated on benchmarking datasets to one another, with the primary goal of addressing these questions:
+
+1. Do methods tend to predict overlapping or distinct sets of doublets on the same dataset?
+2. Is the consensus doublet call across methods predictive of the true doublet status?
+Here, the "consensus doublets" are those droplets which all methods identify as doublets.
+
+There are several items to bear in mind when interpreting these results:
+
+- The ground truth calls themselves to which we are comparing consensus calls may not be entirely accurate, since they were also computationally identified generally with demultiplexing algorithms. 
+- `cxds`, as we are using it, does not have a specific threshold for calling droplets. 
+By contrast, both `scublet` and `scDblFinder` identify a threshold based on the given dataset they are processing. 
+This notebook used a doublet threshold of `>=0.75` for `cxds`, which may not be universally suitable (though choosing a universally suitable threshold is not easy either!).
 
@@
 
-### Read and prepare input data
+## Read and prepare input data
 
 First, we'll read in and combine doublet results into a list of data frames for each dataset.
-We'll also two columns for each dataset:
+We'll also create new columns for each dataset:
 
@@
 
-## Performance metrics
+## Upset plots
 
-This section shows performance metrics for the consensus calls for each dataset.
+This section contains upset plots that show overlap across doublet calls from each method, displayed for each dataset.
 
 ```{r}
-doublet_df_list |>
+pull_barcodes <- function(df, pred_var) {
+  # Helper function to pull out barcodes for doublet calls
+  df$barcodes[df[[pred_var]] == "doublet"]
+}
+
+upset_list <- doublet_df_list |>
   purrr::iwalk(
     \(df, dataset) {
-        print(glue::glue("======================== {dataset} ========================"))
 
-        cat("Table of consensus calls:")
-        print(table(df$consensus_call))
+      doublet_barcodes <- list(
+        "scDblFinder" = pull_barcodes(df, "scdbl_prediction"),
+        "scrublet"    = pull_barcodes(df, "scrublet_prediction"),
+        "cxds"        = pull_barcodes(df, "cxds_prediction")
+      )
 
-        cat("\n\n")
+      UpSetR::upset(fromList(doublet_barcodes), order.by = "freq") |> print()
+      grid::grid.text( # plot title
+        dataset,
+        x = 0.65, 
+        y = 0.95, 
+        gp = grid::gpar(fontsize=16)
+      )
 
-        caret::confusionMatrix(
-          # truth should be first
-          table(
-            "Truth" = df$ground_truth,
-            "Consensus prediction" = df$consensus_call
-          ), 
-          positive = "doublet"
-        ) |>
-        print()
     }
   )
+
 ```
 
-## Visualizations
+## Evaluating consensus performance
+
+This section visualizes and evaluates the consensus doublet calls across each dataset.
 
@@
 This section plots the PCA for each dataset, with three color schemes from left to right:
-- Ground truth doublets are shown in black
-- Consensus doublets are shown in black
-- Points are colored as tp, tn, fp, or fn based on comparing the consensus call to the ground truth
+
+1. Ground truth doublets are shown in black and singlets in blue
+2. Consensus doublets are shown in black and all remaining droplets in blue
+3. Points are colored based on comparing the consensus call to the ground truth as one of:
+  - true positive (`tp`), true negative (`tn`), false positive (`fp`), false negative (`fn`)
 
@@
 
-### Upset plots
+### Performance metrics
 
-This section shows upset plots for overlap among methods for each dataset.
+This section calculates a confusion matrix and associated statistics on the consensus calls.
 
 ```{r}
-pull_barcodes <- function(df, pred_var) {
-  # Helper function to pull out barcodes for doublet calls
-  df$barcodes[df[[pred_var]] == "doublet"]
-}
+metric_df <- doublet_df_list |>
+  purrr::imap( 
+    \(df, dataset) {
+        print(glue::glue("======================== {dataset} ========================"))
 
-upset_list <- doublet_df_list |>
-  purrr::iwalk(
-    \(df, dataset) {
+        cat("Table of consensus calls:")
+        print(table(df$consensus_call))
 
-      doublet_barcodes <- list(
-        "scDblFinder" = pull_barcodes(df, "scdbl_prediction"),
-        "scrublet"    = pull_barcodes(df, "scrublet_prediction"),
-        "cxds"        = pull_barcodes(df, "cxds_prediction")
+        cat("\n\n")
+
+        confusion_result <- caret::confusionMatrix(
+          # truth should be first
+          table(
+            "Truth" = df$ground_truth,
+            "Consensus prediction" = df$consensus_call
+          ), 
+          positive = "doublet"
         )
 
-      UpSetR::upset(fromList(doublet_barcodes), order.by = "freq") |> print()
-      grid::grid.text( # plot title
-        dataset,
-        x = 0.65, 
-        y = 0.95, 
-        gp = grid::gpar(fontsize=16)
+        print(confusion_result)
+
+        # Extract information we want to present later in a table
+        tibble::tibble(
+          "Dataset name" = dataset,
+          "Kappa" = round(confusion_result$overall["Kappa"], 3), 
+          "Balanced accuracy" = round(confusion_result$byClass["Balanced Accuracy"], 3)
         )
+    }
+  ) |>
+  dplyr::bind_rows()
+```
 
-    }
-  )
+## Conclusions
 
+Overall, methods do not have substantial overlap with out another. 
+They each tend to detect different sets of doublets, leading to fairly small sets of consensus doublets. 
+Further, the consensus doublets called by all three methods have some, but not substantial, overlap with the ground truth.
+
+For three out of four datasets, `scDblFinder` predicts a much larger number of doublets compared to other methods. 
+For `pdx-MULTI`, however, `cxds` predicts a much larger number of droplets.
+
+The table below summarizes performance of the "consensus caller". 
+Note that, in the [benchmarking paper these datasets were originally analyzed in](https://doi.org/10.1016/j.cels.2020.11.008), `hm-6k` was observed to be one of the "easiest" datasets to classify across methods.
+Consistent with that observation, it has the highest `kappa` value here, although it is still fairly low - though not as low as the other methods, which are very close to 0.
+
+```{r}
+metric_df
 ```

### 2024-06-04 analyses/cell-type-ewings/exploratory_analysis/01-marker-gene-tumor-classification.Rmd

--- 2024-05-13 analyses/cell-type-ewings/01-marker-gene-tumor-classification.Rmd
@@
 We then see if we can use expression of marker genes to manually classify tumor and normal cells.
-The main goal of this notebook is only to identify tumor cells, identification and labelling of the other cells is a separate question that we do not answer here.
+The main goal of this notebook is only to identify tumor cells, identification and labeling of the other cells is a separate question that we do not answer here.
 
@@
 
-It looks like ADGRG2, HES1, NROB1, and STEAP1 don't show a lot of difference between cells that might be tumor vs. normal cells and are generally lowly expressed.
+It looks like `ADGRG2`, `HES1`, `NROB1`, and `STEAP1` don't show a lot of difference between cells that might be tumor vs. normal cells and are generally lowly expressed.
 CD99 appears to be expressed in both normal and tumor at similar levels. 
@@
 Looking at this might help give us an idea of genes that are more specifically expressed in tumor cells. 
-Genes like ADGRG2 and HES1 do not seem specific to tumor cells. 
-There are normal and tumor cells that express the gene and not very many tumor cells actually have expression of ADGRG2. 
-STEAP1 and NROB1 although not expressed in very many tumor cells, look to be specific to tumor cells which is promising.
+Genes like `ADGRG2` and `HES1` do not seem specific to tumor cells. 
+There are normal and tumor cells that express the gene and not very many tumor cells actually have expression of `ADGRG2.` 
+`STEAP1` and `NROB1` although not expressed in very many tumor cells, look to be specific to tumor cells which is promising.
 
-CD99, CAV1, and CCND1 are expressed in a high proportion of tumor cells but are also expressed in quite a few normal cells. 
-This is consistent with the literature for CD99 that it is expressed in normal cells, just at a lower level than in Ewing's cells.
+`CD99`, `CAV1`, and `CCND1` are expressed in a high proportion of tumor cells but are also expressed in quite a few normal cells. 
+This is consistent with the literature for `CD99` that it is expressed in normal cells, just at a lower level than in Ewing's cells.
 
-KDSR, MAPT, NKX2-2, PAPPA, and PRKCB all seem to have a high proportion of tumor cells expressing the gene and a low proportion of normal cells that express the gene. 
-PRKCB and MAPT are also 2 of the top genes identified using the marker analysis.
+`KDSR`, `MAPT`, `NKX2-2`, `PAPPA`, and `PRKCB` all seem to have a high proportion of tumor cells expressing the gene and a low proportion of normal cells that express the gene. 
+`PRKCB` and `MAPT` are also 2 of the top genes identified using the marker analysis.
 
@@
 For CellAssign, all tumor cells are pulmonary vascular smooth muscle cells. 
-We also see tha normal cells contain some smooth muscle cells, but mostly are fibroblasts and endothelial cells.
+We also see that normal cells contain some smooth muscle cells, but mostly are fibroblasts and endothelial cells.
 
@@
 
-Consistent with previous findings, ADGRG2, HES1, NROB1, and STEAP1 don't seem to show any expression in either groups of cells. 
-CCND1 and CD99 appear to have similar expression across both groups of cells, while MAPT and PRKCB seem to have the most distinct expression profiles between muscle and endothelial cells.
+Consistent with previous findings, `ADGRG2`, `HES1`, `NROB1`, and `STEAP1` don't seem to show any expression in either groups of cells. 
+`CCND1` and `CD99` appear to have similar expression across both groups of cells, while `MAPT` and `PRKCB` seem to have the most distinct expression profiles between muscle and endothelial cells.
 
@@
 - The top marker genes for the "tumor" cells contain PRKCB and MAPT, two of the marker genes in our gene list. 
-- It's likely that ADGRG2, HES1, NROB1, and STEAP1 will not be informative in assigning tumor cells. 
-- We may be weary of genes like CD99 too as expression is more consistent across cells rather than specific to a group of cells. 
+- It's likely that `ADGRG2`, `HES1`, `NROB1`, and `STEAP1` will not be informative in assigning tumor cells. 
+- We may be weary of genes like `CD99` too as expression is more consistent across cells rather than specific to a group of cells. 
 - Endothelial cells as identified by SingleR and CellAssign generally have low expression of these marker genes and muscle cells have higher expression of marker genes.

### 2024-06-04 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
notebook validating tumor cells

---
title: "SCPCL000822 - Validation of tumor cells"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
---

## Introduction

In this notebook we compare tumor cell annotations obtained from marker genes, `CopyKAT`, and `InferCNV` to identify a group of tumor cells we are confident in.

- First we look at the cells that are annotated as tumor cells by each method. 
Using this information, we call any cell that is annotated as a tumor cell by both `CopyKAT` and `InferCNV` as a tumor cell. 
- We then validate that these are in fact tumor cells by doing the following: 
  - Looking at marker gene expression. 
  - Looking at expression of EWS-FLI1 target gene sets. 
  - Confirming presence of known CNVs in Ewing sarcoma

## Setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_bw()
)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "2024-05-01")
sample_dir <- file.path(data_dir, "SCPCP000015", "SCPCS000490")

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-ewings")
```

```{r}
# Input files
sce_file <- file.path(sample_dir, "SCPCL000822_processed.rds")
marker_genes <- file.path(module_base, "references", "tumor-marker-genes.tsv")

# results from annotation workflow 
results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000490")

marker_gene_results_file <- file.path(results_dir, "SCPCL000822_tumor-normal-classifications.tsv")
copykat_predictions_file <- file.path(results_dir, "SCPCL000822_copykat-classifications.tsv")

infercnv_predictions_file <- file.path(results_dir, "SCPCL000822_infercnv-classifications.tsv")
infercnv_metadata_file <- file.path(results_dir, "infercnv", "SCPCL000822_cnv-metadata.tsv")

geneset_scores_file <- file.path(results_dir, "SCPCL000822_gene-set-scores.tsv")

# output files 
final_annotations_dir <- file.path(module_base, "cell_type_annotations", "SCPCS000490")
fs::dir_create(final_annotations_dir)
final_annotations_file <- file.path(final_annotations_dir, "SCPCL000822_tumor-classifications.tsv")
```

```{r}
# read in all files 
sce <- readr::read_rds(sce_file)

# read in marker genes table 
marker_genes_df <- readr::read_tsv(marker_genes) |> 
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
  dplyr::distinct()

marker_gene_results_df <- readr::read_tsv(marker_gene_results_file)

# copykat output
copykat_results_df <- readr::read_tsv(copykat_predictions_file) |> 
  # only select no_ref classification and rename to be consistent with naming of other classifications
  dplyr::select(barcodes, copykat_classification = no_ref) |> 
  dplyr::mutate(
    copykat_classification = dplyr::case_when(
      copykat_classification == "aneuploid" ~ "Tumor",
      copykat_classification == "diploid" ~ "Normal",
      TRUE ~ copykat_classification
    )
  )

# infercnv output 
infercnv_results_df <- readr::read_tsv(infercnv_predictions_file)
infercnv_metadata_df <- read.delim(infercnv_metadata_file) |> 
  tibble::rownames_to_column("barcodes")

# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
chr_columns_to_keep <- which(!stringr::str_detect(names(infercnv_metadata_df), "chrMT|chrGL"))
infercnv_metadata_df <- infercnv_metadata_df[, chr_columns_to_keep]

# gene set scores 
geneset_scores_df <- readr::read_tsv(geneset_scores_file)
```

```{r}
# grab all coldata and add in all classifications and results to use for plotting throughout notebook 
# get list of marker genes to use 
marker_genes <- marker_genes_df |> 
  dplyr::filter(cell_type == "tumor") |> 
  dplyr::pull(ensembl_gene_id)

# get the gene expression counts for all marker genes
marker_gene_exp <- logcounts(sce[marker_genes, ]) |>
  as.matrix() |>
  t() |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes")

# pull out the UMAP coordinates and make a data frame to use for plotting
coldata_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
      ) |> 
  # add in marker gene expression to dataframe 
  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
  # combine all genes into a single column for easy faceting
  tidyr::pivot_longer(
    cols = starts_with("ENSG"),
    names_to = "ensembl_gene_id",
    values_to = "gene_expression"
  ) |> 
  # join with marker gene df to get gene symbols for plotting 
  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression) |> 
  # add in marker gene classifications
  dplyr::left_join(marker_gene_results_df) |> 
  # add in copykat results 
  dplyr::left_join(copykat_results_df) |> 
  # add in infercnv results and metadata
  dplyr::left_join(infercnv_results_df) |> 
  dplyr::left_join(infercnv_metadata_df) |> 
  # gene set scores 
  dplyr::left_join(geneset_scores_df)

```

## Identify tumor cell population

Below is a UMAP showing which cells are labeled as tumor cells using each of the various classification methods.

```{r}
# plot tumor cells for each classification method 
tumor_umap_df <- coldata_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, marker_gene_classification, copykat_classification, cnv_proportion_classification) |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  )

ggplot(tumor_umap_df, aes(x = UMAP1, UMAP2, color = classification)) +
  geom_point(size = 0.5, alpha = 0.5) +
  facet_wrap(vars(method))
```

This plot shows us that there is a pretty clear consensus that most of the cells in the bottom left group are tumor cells. 
From the original marker gene analysis we know that not all marker genes are expressed and that there is quite a bit of variance in expression. 
Because of that and the fact that both CNV inference methods show similar annotations, we will classify any cells that are labeled as tumor by both `CopyKAT` and `InferCNV` as tumor cells.

```{r}
# create a combined classification between inferCNV and copyKAT
# cells that are classified as tumor by both are kept as tumor, all other are normal 
coldata_df <- coldata_df |> 
  dplyr::mutate(
    combined_cnv_classification = dplyr::if_else(
      cnv_proportion_classification == "Tumor" & copykat_classification == "Tumor",
      "Tumor",
      "Normal"
    ),
    # there are cells classified as NA by copykat, we want those to be normal 
    combined_cnv_classification = dplyr::if_else(is.na(copykat_classification) & is.na(combined_cnv_classification), "Normal", combined_cnv_classification)
  )
```

```{r}
# look at which cells are tumor/normal
# be sure to remove any duplicates 
coldata_df |> 
  dplyr::select(UMAP1, UMAP2, barcodes, combined_cnv_classification) |>
  unique() |> 
  ggplot(aes(x = UMAP1, y = UMAP2, color = combined_cnv_classification)) +
  geom_point(alpha = 0.5, size = 0.5)
```

The remainder of the notebook will focus on validating that these in fact are tumor cells and all other cells are classified as normal.

## Validate tumor cells

### Marker gene expression

First, we will look at the expression of the tumor marker genes. 
We expect that these should be expressed higher in the tumor cells than in the normal cells.

In this first plot, we show a density plot of marker gene expression coloring by the cell classification (tumor or normal). 
For this plot, we will include all classifications, including the new combined CNV classification.

```{r}
# calculate sum of the scaled gene expression values for plotting 
marker_density_df <- coldata_df |> 
  dplyr::select(barcodes, gene_symbol, gene_expression, ends_with("classification")) |> 
  unique() |> 
  dplyr::group_by(gene_symbol) |>
  # get z-scores for each gene
  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |> 
  dplyr::ungroup() |> 
  dplyr::group_by(barcodes) |> 
  # add all the scores together 
  dplyr::mutate(sum_exp = sum(transformed_gene_expression)) |>
  # expand classifications column for easy faceting 
  dplyr::ungroup() |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  )

ggplot(marker_density_df, aes(x = sum_exp, color = classification)) +
  geom_density() +
  facet_wrap(vars(method))
```

This plot shows us that overall the tumor cells in each method have higher marker gene expression. 
Although the `InferCNV` alone and combined CNV classifications appear to have the most distinct separation in marker gene expression.

Next, we will plot the expression of each gene in each cell using a heatmap. 
Here the rows are the marker genes and the columns are the cells with the color indicating the expression of that gene in that cell.

```{r}
# create matrix with marker genes as rows and barcodes as columns
marker_gene_heatmap <- coldata_df |> 
  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
  tidyr::pivot_wider(values_from = gene_expression, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("gene_symbol") |> 
  as.matrix()

# create annotation df, make sure that dimensions match up by only getting unique barcodes 
annotation_df <- coldata_df |> 
  dplyr::select(barcodes, combined_cnv_classification) |> 
  unique()

annotation <- ComplexHeatmap::columnAnnotation(
  classification = annotation_df$combined_cnv_classification,
  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05")))
```

```{r}
# plot heatmap of marker genes 
ComplexHeatmap::Heatmap(
  marker_gene_heatmap,
  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
  border = TRUE,
  ## Row parameters
  cluster_rows = TRUE,
  row_title = "Marker gene symbol",
  row_title_gp = grid::gpar(fontsize = 12),
  row_title_side = "left",
  row_names_side = "left",
  row_dend_side = "right",
  row_names_gp = grid::gpar(fontsize = 10),
  ## Column parameters
  cluster_columns = TRUE,
  show_column_names = FALSE,
  bottom_annotation = annotation
)
```
As expected, the tumor cells appear to have more marker genes that are expressed than normal cells.

### Gene set scores

Next, we will look at the gene set scores for three different EWS-FLI1 gene sets. 
The below plots will all use the `mean-GENESET` columns.

First we will just look at the gene set scores for each cell on the UMAP.

```{r}
# plot tumor cells for each classification method 
geneset_plot_df <- coldata_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean")) |> 
  tidyr::pivot_longer(
    cols = starts_with("mean"),
    names_to = "geneset",
    values_to = "mean_score"
  ) |> 
  dplyr::mutate(
    geneset = stringr::word(geneset, -1, sep = "-")
  )

ggplot(geneset_plot_df, aes(x = UMAP1, UMAP2, color = mean_score)) +
  geom_point(size = 0.5, alpha = 0.5) +
  facet_wrap(vars(geneset)) +
  scale_color_viridis_c()
```

Generally, the higher gene set scores appear to be in the group of cells that are tumor cells with some slight variations between gene sets.

Below we look at gene set scores for cells classified as tumor cells vs normal cells for all classification methods.

```{r, fig.height=10}
geneset_plot_df |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  ggplot(aes(x = mean_score, color = classification)) +
  geom_density(bounds = c(0,1), bw = 0.05) +
  facet_grid(rows = vars(method), 
             cols = vars(geneset))
```

For the most part the gene set scores are higher in tumor cells regardless of classification.

Below we make a heatmap of the gene set scores for just the combined CNV classification. 
Each row is a gene set and each column is a barcode. 
The color indicates the gene set score for that cell.

```{r}
# make a matrix of gene set by barcode 
geneset_heatmap <- geneset_plot_df |> 
  dplyr::select(mean_score, geneset, barcodes) |> 
  unique() |> 
  tidyr::pivot_wider(values_from = mean_score, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("geneset") |> 
  as.matrix()

ComplexHeatmap::Heatmap(
  geneset_heatmap,
  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
  border = TRUE,
  ## Row parameters
  cluster_rows = TRUE,
  row_title = "",
  row_title_gp = grid::gpar(fontsize = 12),
  row_title_side = "left",
  row_names_side = "left",
  row_dend_side = "right",
  row_names_gp = grid::gpar(fontsize = 10),
  ## Column parameters
  cluster_columns = TRUE,
  show_column_names = FALSE,
  bottom_annotation = annotation
)
```

For the most part we see that cells with higher gene set scores are classified as tumor cells. 
One thing to note is that the `SILIGAN` gene set appears to have some expression in normal cells. 
However, if you look at the spots where all three gene sets have higher scores, these apply to tumor cells.

### Validate based on CNAs

The last thing we will do here is look to see if tumor cells have any of the expected CNAs found in Ewing sarcoma:

- Gain of Chr8
- Gain of Chr12
- Gain of Chr1p
- Loss of Chr16q

Although these are the most frequent, there are patients who do not have any of these alterations and patients that only have some of these alterations. 
See [Tirode et al.,](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4264969/) and [Crompton et al.](https://doi.org/10.1158/2159-8290.CD-13-1037).

We will look specifically at gains and losses in each chromosome across tumor and normal cells. 
To do this, we will use the output from `InferCNV` since that output separates out gains and losses in a way that makes it easy for plotting.

The first heatmap looks at the proportion of each chromosome with a CNV that corresponds to an amplification or gain. 
Here each row is a cell and each column is a chromosome with the color corresponding to the proportion of the chromosome with a gain detected.

```{r}
# get matrix of barcodes by chromosome
gain_cnv_df <- coldata_df |> 
  dplyr::select(barcodes, starts_with("proportion_scaled_dupli")) |> 
  unique() |> 
  tibble::column_to_rownames("barcodes") |> 
  as.matrix()

# remove extra characters in colnames, proportion_scaled_dupli
colnames(gain_cnv_df) <- stringr::word(colnames(gain_cnv_df), -1, sep = "_")

# need to make a new annotation, since we want to annotate the rows here 
annotation <- ComplexHeatmap::rowAnnotation(classification = annotation_df$combined_cnv_classification,
                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05")))
```

```{r}
ComplexHeatmap::Heatmap(
  gain_cnv_df,
  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
  border = TRUE,
  ## Row parameters
  cluster_rows = TRUE,
  row_title = "",
  row_title_gp = grid::gpar(fontsize = 12),
  row_title_side = "left",
  show_row_names = FALSE,
  row_dend_side = "right",
  row_names_gp = grid::gpar(fontsize = 10),
  ## Column parameters
  cluster_columns = FALSE,
  show_column_names = TRUE,
  right_annotation = annotation
)
```

Here we see a gain in Chr8 and Chr1 in tumor cells compared to normal cells. 
We also see other chromosomes with amplifications in tumor cells compared to normal cells and that the tumor and normal cells are nicely separated in the heirarchical clustering.

Below we make the same plot but with only the losses or deletions.

```{r}
# barcode by chromosome mtx with losses 
loss_cnv_df <- coldata_df |> 
  dplyr::select(barcodes, starts_with("proportion_scaled_loss")) |> 
  unique() |> 
  tibble::column_to_rownames("barcodes") |> 
  as.matrix()

colnames(loss_cnv_df) <- stringr::word(colnames(loss_cnv_df), -1, sep = "_")
```

```{r}
ComplexHeatmap::Heatmap(
  loss_cnv_df,
  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
  border = TRUE,
  ## Row parameters
  cluster_rows = TRUE,
  row_title = "",
  row_title_gp = grid::gpar(fontsize = 12),
  row_title_side = "left",
  show_row_names = FALSE,
  row_dend_side = "right",
  row_names_gp = grid::gpar(fontsize = 10),
  ## Column parameters
  cluster_columns = FALSE,
  show_column_names = TRUE,
  right_annotation = annotation
)
```

We do not see the expected loss in Chr16, but this is not present in every patient. 
Additionally, we had already noted the gain in Chr8 in the tumor cells, so there is still at least one expected CNV present. 
Again, there is a clear separation of tumor and normal cells in terms of presence of CNVs.

## Save final tumor annotations

After validating that the cells classified as tumor by both `CopyKAT` and `InferCNV` show the expected expression patterns and CNV patterns as tumor cells, we can confidently label those cells as tumor cells. 
We will export these labels to the final annotations file for later use.

```{r}
# export final TSV with annotations 
final_classifications <- coldata_df |> 
  dplyr::select(
    cell_barcode = barcodes,
    tumor_cell_classification = combined_cnv_classification
  ) |> 
  unique()

readr::write_tsv(final_classifications, final_annotations_file)
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-06-05 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd
remaining spelling
use CNVs

@@
 
-Those cells in grey (bottom left) are cells used as a reference. 
+Those cells in gray (bottom left) are cells used as a reference. 
 If we were to use total CNVs we would need some sort of cut off for the number. 
@@
 
-We will call any tumor cell as those with a total number of cnvs greater than the mean(number of cnvs).
+We will call any tumor cell as those with a total number of CNVs greater than the mean(number of CNVs).
 

### 2024-06-05 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
       copykat_classification == "diploid" ~ "Normal",
-      TRUE ~ copykat_classification
+      .default = copykat_classification
     )
@@
   # join with marker gene df to get gene symbols for plotting 
-  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
+  dplyr::left_join(marker_genes_df, by = "ensembl_gene_id") |>
   # get rid of excess columns
@@
   # add in marker gene classifications
-  dplyr::left_join(marker_gene_results_df) |> 
+  dplyr::left_join(marker_gene_results_df, by = "barcodes") |> 
   # add in copykat results 
-  dplyr::left_join(copykat_results_df) |> 
+  dplyr::left_join(copykat_results_df, by = "barcodes") |> 
   # add in infercnv results and metadata
-  dplyr::left_join(infercnv_results_df) |> 
-  dplyr::left_join(infercnv_metadata_df) |> 
+  dplyr::left_join(infercnv_results_df, by = "barcodes") |> 
+  dplyr::left_join(infercnv_metadata_df, by = "barcodes") |> 
   # gene set scores 
-  dplyr::left_join(geneset_scores_df)
+  dplyr::left_join(geneset_scores_df, by = "barcodes")
 
@@
   dplyr::select(UMAP1, UMAP2, barcodes, combined_cnv_classification) |>
-  unique() |> 
+  dplyr::distinct() |> 
   ggplot(aes(x = UMAP1, y = UMAP2, color = combined_cnv_classification)) +
@@
 ```
+
 As expected, the tumor cells appear to have more marker genes that are expressed than normal cells.
@@
   ggplot(aes(x = mean_score, color = classification)) +
-  geom_density(bounds = c(0,1), bw = 0.05) +
+  geom_density(bw = 0.05) +
   facet_grid(rows = vars(method), 

### 2024-06-05 analyses/cell-type-ewings/template_notebooks/01-marker-genes.Rmd

@@
 renv::load(project_root)
+
+knitr::opts_chunk$set(
+  dev = "jpeg"
+)
 ```

### 2024-06-05 analyses/cell-type-ewings/template_notebooks/02-cellassign.Rmd
fix cellassign errors

@@
       # make sure annotations match marker gene annotations
-      cellassign_annotations == "tumor", "Tumor", "Normal"
+      cellassign_annotations == "tumor", 
+      "Tumor", 
+      "Normal"
     )) |> 
@@
 
+  if(all("Endothelial cells", "Other") %in% celltype_df$visser_cellassign_annotations &&
+     all("Endothelial cells", "Other") %in% celltype_df$cellassign_celltype_annotation){
     caret::confusionMatrix(
@@
     )  
+  }
 } else {
@@
 
-The last thing we will do here is run CellAssign using some of the same reference that we used when running CellAssign as part of scpca-nf. 
+The last thing we will do here is run CellAssign using some of the same reference that we used when running CellAssign as part of `scpca-nf`. 
 In particular, the majority of non-muscle cells (which we believe to be tumor cells) were identified as either Endothelial cells or Fibroblasts.

### 2024-06-11 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
Apply suggestions from code review
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

@@
 - `cxds`, as we are using it, does not have a specific threshold for calling droplets. 
-By contrast, both `scublet` and `scDblFinder` identify a threshold based on the given dataset they are processing. 
+By contrast, both `scrublet` and `scDblFinder` identify a threshold based on the given dataset they are processing. 
 This notebook used a doublet threshold of `>=0.75` for `cxds`, which may not be universally suitable (though choosing a universally suitable threshold is not easy either!).
@@
 
-# Read in and data for analysis
+# Read in data for analysis
 doublet_df_list <- dataset_names |>
@@
                       PC1 = PCA.1, 
-                      PC2 = PCA.2)
-      rm(sce)
-
-      dataset_df <- scdbl_df |>
+                      PC2 = PCA.2) |> 
         dplyr::left_join(
@@
         dplyr::left_join(
-          sce_df, 
+          scdbl_df, 
           by = "barcodes"
@@
     }
-  )
-names(doublet_df_list) <- dataset_names
+  ) |> 
+  purrr::set_names(dataset_names)
 ```
@@
 
-Overall, methods do not have substantial overlap with out another. 
+Overall, methods do not have substantial overlap with each other. 
 They each tend to detect different sets of doublets, leading to fairly small sets of consensus doublets. 

### 2024-06-12 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
break out heatmap functions and explore the unknown

@@
 
+## Define functions
+
+```{r}
+# create a heatmap where rows are marker genes or gene sets and columns are cells 
+# this function adds a column annotation since cells are columns 
+plot_gene_heatmap <- function(df, 
+                              row_title = "",
+                              legend_title = "", 
+                              annotation = NULL){
+
+  # plot heatmap of marker genes 
+  heatmap <- ComplexHeatmap::Heatmap(
+    df,
+    # set the color scale based on min and max values
+    col = circlize::colorRamp2(seq(min(df), max(df), length = 2), colors = c("white", "#00274C")),
+    border = TRUE,
+    ## Row parameters
+    cluster_rows = TRUE,
+    row_title = row_title,
+    row_title_gp = grid::gpar(fontsize = 12),
+    row_title_side = "left",
+    row_names_side = "left",
+    row_dend_side = "right",
+    row_names_gp = grid::gpar(fontsize = 10),
+    ## Column parameters
+    cluster_columns = TRUE,
+    show_column_names = FALSE,
+    bottom_annotation = annotation,
+    heatmap_legend_param = list(
+      title = legend_title
+    )
+  )
+
+  return(heatmap)
+}
+
+# create a heatmap where rows are cells and columns are chromosomes 
+# this function adds a row annotation since cells are rows 
+plot_cnv_heatmap <- function(df,
+                             cnv_col, 
+                             annotation = NULL,
+                             legend_title = ""){
+
+  # barcode by chromosome df 
+  cnv_df <- df |> 
+    dplyr::select(barcodes, starts_with(cnv_col)) |> 
+    unique() |> 
+    tibble::column_to_rownames("barcodes")
+
+  # just get the name of the chromosome as column name 
+  colnames(cnv_df) <- stringr::word(colnames(cnv_df), -1, sep = "_")
+
+  # do a little bit of column reordering before converting to a matrix
+  cnv_df <- cnv_df |> 
+    dplyr::relocate(chr10, .before = chr11) |> 
+    dplyr::relocate(chr19, .before = chr20) |> 
+    dplyr::relocate(chr21, .before = chr22) |> 
+    as.matrix()
+
+  heatmap <- ComplexHeatmap::Heatmap(
+    cnv_df,
+    col = circlize::colorRamp2(seq(min(cnv_df), max(cnv_df), length = 2), colors = c("white", "#00274C")),
+    border = TRUE,
+    ## Row parameters
+    cluster_rows = TRUE,
+    row_title = "",
+    row_title_gp = grid::gpar(fontsize = 12),
+    row_title_side = "left",
+    show_row_names = FALSE,
+    row_dend_side = "right",
+    row_names_gp = grid::gpar(fontsize = 10),
+    ## Column parameters
+    cluster_columns = FALSE,
+    show_column_names = TRUE,
+    right_annotation = annotation,
+    heatmap_legend_param = list(
+      title = legend_title
+    )
+  )
+  return(heatmap)
+}
+
+```
+
 ## Setup
@@
 )
+
+# use jpegs for images 
+knitr::opts_chunk$set(
+  dev = "jpeg"
+)
+
+# quiet messages
+options(readr.show_col_types = FALSE)
+ComplexHeatmap::ht_opt(message = FALSE)
 ```
@@
 
-# pull out the UMAP coordinates and make a data frame to use for plotting
-coldata_df <- sce |> 
+# pull out the UMAP coordinates, add in marker genes and classifications 
+# and make a data frame to use for plotting
+classification_df <- sce |> 
   scuttle::makePerCellDF(use.dimred = "UMAP") |> 
@@
 
+- Marker genes only: For each cell, the expression of all marker genes was transformed using a z-transformation and summed together. 
+Any cells with a total marker gene expression value > 0 were called as tumor cells. 
+- CopyKAT - no reference: [`CopyKAT`](https://github.com/navinlabcode/copykat) was run without indicating any reference cells and cells were classified as aneuploid (Tumor) or diploid (Normal). 
+- InferCNV - [`InferCNV`](https://github.com/broadinstitute/inferCNV/wiki) was used to call copy number variations using endothelial and fibroblasts assigned by `CellAssign` as the normal reference group.
+This returns a proportion of each chromosome with a CNV detected.
+We then calculate the genomic CNV proportion for each cell across all chromosomes, weighted by the number of genes in a chromosome.
+Cells with a genomic CNV proportion greater than the mean  CNV proportion across all cells are called as tumor cells.
+
 ```{r}
 # plot tumor cells for each classification method 
-tumor_umap_df <- coldata_df |> 
+tumor_umap_df <- classification_df |> 
   dplyr::select(barcodes, UMAP1, UMAP2, marker_gene_classification, copykat_classification, cnv_proportion_classification) |> 
@@
     values_to = "classification"
+  ) |> 
+  dplyr::mutate(
+    method = dplyr::case_when(
+      method == "marker_gene_classification" ~ "Marker genes only",
+      method == "copykat_classification" ~ "CopyKAT - no reference",
+      method == "cnv_proportion_classification" ~ "InferCNV proportion"
+    )
   )
@@
 
-This plot shows us that there is a pretty clear consensus that most of the cells in the bottom left group are tumor cells. 
+This plot shows us that there is a pretty clear consensus that most of the cells in the bottom right group are tumor cells. 
 From the original marker gene analysis we know that not all marker genes are expressed and that there is quite a bit of variance in expression. 
@@
 # create a combined classification between inferCNV and copyKAT
-# cells that are classified as tumor by both are kept as tumor, all other are normal 
-coldata_df <- coldata_df |> 
+# cells that are classified as tumor by both are kept as tumor
+# classified by both as normal are normal 
+# in between are classified as unknown
+classification_df <- classification_df |> 
   dplyr::mutate(
-    combined_cnv_classification = dplyr::if_else(
-      cnv_proportion_classification == "Tumor" & copykat_classification == "Tumor",
-      "Tumor",
-      "Normal"
-    ),
-    # there are cells classified as NA by copykat, we want those to be normal 
-    combined_cnv_classification = dplyr::if_else(is.na(copykat_classification) & is.na(combined_cnv_classification), "Normal", combined_cnv_classification)
-  )
+    combined_cnv_classification = dplyr::case_when(
+      # only tumor if both call tumor 
+      cnv_proportion_classification == "Tumor" & copykat_classification == "Tumor" ~ "Tumor",
+      # only normal if both call normal and/or reference
+      cnv_proportion_classification == "Normal" & copykat_classification == "Normal" ~ "Normal",
+      cnv_proportion_classification == "Reference" & copykat_classification == "Normal" ~ "Normal",
+      # if normal/tumor mismatch set to unknown 
+      cnv_proportion_classification == "Normal" & copykat_classification == "Tumor" ~ "Unknown",
+      cnv_proportion_classification == "Tumor" & copykat_classification == "Normal" ~ "Unknown",
+      cnv_proportion_classification == "Reference" &  copykat_classification == "Tumor" ~ "Unknown", 
+      !is.na(cnv_proportion_classification) & is.na(copykat_classification) ~ "Unknown", 
+      is.na(cnv_proportion_classification) & !is.na(copykat_classification) ~ "Unknown",
+      is.na(cnv_proportion_classification) & is.na(copykat_classification) ~ "Unknown"
+    ))
 ```
@@
 # be sure to remove any duplicates 
-coldata_df |> 
+classification_df |> 
   dplyr::select(UMAP1, UMAP2, barcodes, combined_cnv_classification) |>
@@
 The remainder of the notebook will focus on validating that these in fact are tumor cells and all other cells are classified as normal. 
+We will also look at where the unknown cells fall in relation to tumor vs. normal.
 
@@
 
+Just a note that the `marker_gene_classification` is a circular since these same marker genes were used to classify those cells.
+
 ```{r}
 # calculate sum of the scaled gene expression values for plotting 
-marker_density_df <- coldata_df |> 
-  dplyr::select(barcodes, gene_symbol, gene_expression, ends_with("classification")) |> 
+marker_density_df <- classification_df |> 
+  dplyr::select(barcodes, gene_symbol, gene_expression, copykat_classification, cnv_proportion_classification, marker_gene_classification, combined_cnv_classification) |> 
   unique() |> 
@@
     values_to = "classification"
+  ) |> 
+  dplyr::mutate(
+    method = dplyr::case_when(
+      method == "marker_gene_classification" ~ "Marker genes only",
+      method == "copykat_classification" ~ "CopyKAT - no reference",
+      method == "cnv_proportion_classification" ~ "InferCNV proportion",
+      method == "combined_cnv_classification" ~ "CopyKAT and InferCNV"
+    )
   )
@@
 
+Another note is that the "Unknown" class appears to have marker gene expression more similar to the normal cells.
+
 Next, we will plot the expression of each gene in each cell using a heatmap. 
@@
 # create matrix with marker genes as rows and barcodes as columns
-marker_gene_heatmap <- coldata_df |> 
+marker_gene_heatmap <- classification_df |> 
   dplyr::select(gene_expression, gene_symbol, barcodes) |> 
@@
 # create annotation df, make sure that dimensions match up by only getting unique barcodes 
-annotation_df <- coldata_df |> 
+annotation_df <- classification_df |> 
   dplyr::select(barcodes, combined_cnv_classification) |> 
@@
   classification = annotation_df$combined_cnv_classification,
-  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05")))
+  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))
 ```
@@
 # plot heatmap of marker genes 
-ComplexHeatmap::Heatmap(
-  marker_gene_heatmap,
-  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
-  border = TRUE,
-  ## Row parameters
-  cluster_rows = TRUE,
+plot_gene_heatmap(marker_gene_heatmap, 
                   row_title = "Marker gene symbol",
-  row_title_gp = grid::gpar(fontsize = 12),
-  row_title_side = "left",
-  row_names_side = "left",
-  row_dend_side = "right",
-  row_names_gp = grid::gpar(fontsize = 10),
-  ## Column parameters
-  cluster_columns = TRUE,
-  show_column_names = FALSE,
-  bottom_annotation = annotation
-)
+                  legend_title = "Marker gene \nexpression",
+                  annotation = annotation)
 ```
@@
 As expected, the tumor cells appear to have more marker genes that are expressed than normal cells. 
+The "Unknown" cells appear to have a small group of cells that cluster with the tumor cells on the right hand side of the plot.
 
@@
 # plot tumor cells for each classification method 
-geneset_plot_df <- coldata_df |> 
+geneset_plot_df <- classification_df |> 
   dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean")) |> 
@@
 
+We also see that the "Unknown" cells identified in the combined classification mostly mirror the gene set expression scores for normal cells with maybe a small group of cells having more similar gene set scores to tumor cells.
+
 Below we make a heatmap of the gene set scores for just the combined CNV classification. 
@@
 
-ComplexHeatmap::Heatmap(
-  geneset_heatmap,
-  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
-  border = TRUE,
-  ## Row parameters
-  cluster_rows = TRUE,
-  row_title = "",
-  row_title_gp = grid::gpar(fontsize = 12),
-  row_title_side = "left",
-  row_names_side = "left",
-  row_dend_side = "right",
-  row_names_gp = grid::gpar(fontsize = 10),
-  ## Column parameters
-  cluster_columns = TRUE,
-  show_column_names = FALSE,
-  bottom_annotation = annotation
-)
+# plot heatmap of gene set score 
+plot_gene_heatmap(geneset_heatmap,
+                  annotation = annotation,
+                  legend_title = "Gene set \nscore")
 ```
@@
 However, if you look at the spots where all three gene sets have higher scores, these apply to tumor cells.
+
+Again, most of the "Unknown" tumor cells tend to align with normal cells, with a few grouping in the area where cells transition from tumor to normal.
 
@@
 ```{r}
-# get matrix of barcodes by chromosome
-gain_cnv_df <- coldata_df |> 
-  dplyr::select(barcodes, starts_with("proportion_scaled_dupli")) |> 
-  unique() |> 
-  tibble::column_to_rownames("barcodes") |> 
-  as.matrix()
-
-# remove extra characters in colnames, proportion_scaled_dupli
-colnames(gain_cnv_df) <- stringr::word(colnames(gain_cnv_df), -1, sep = "_")
-
 # need to make a new annotation, since we want to annotate the rows here 
 annotation <- ComplexHeatmap::rowAnnotation(classification = annotation_df$combined_cnv_classification,
-                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05")))
-```
+                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))
 
-```{r}
-ComplexHeatmap::Heatmap(
-  gain_cnv_df,
-  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
-  border = TRUE,
-  ## Row parameters
-  cluster_rows = TRUE,
-  row_title = "",
-  row_title_gp = grid::gpar(fontsize = 12),
-  row_title_side = "left",
-  show_row_names = FALSE,
-  row_dend_side = "right",
-  row_names_gp = grid::gpar(fontsize = 10),
-  ## Column parameters
-  cluster_columns = FALSE,
-  show_column_names = TRUE,
-  right_annotation = annotation
-)
+# plot gains in cnv
+plot_cnv_heatmap(classification_df,
+                 cnv_col = "proportion_scaled_dupli",
+                 annotation = annotation,
+                 legend_title = "Copy number \ngain")
 ```
@@
 ```{r}
-# barcode by chromosome mtx with losses 
-loss_cnv_df <- coldata_df |> 
-  dplyr::select(barcodes, starts_with("proportion_scaled_loss")) |> 
-  unique() |> 
-  tibble::column_to_rownames("barcodes") |> 
-  as.matrix()
-
-colnames(loss_cnv_df) <- stringr::word(colnames(loss_cnv_df), -1, sep = "_")
-```
-
-```{r}
-ComplexHeatmap::Heatmap(
-  loss_cnv_df,
-  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
-  border = TRUE,
-  ## Row parameters
-  cluster_rows = TRUE,
-  row_title = "",
-  row_title_gp = grid::gpar(fontsize = 12),
-  row_title_side = "left",
-  show_row_names = FALSE,
-  row_dend_side = "right",
-  row_names_gp = grid::gpar(fontsize = 10),
-  ## Column parameters
-  cluster_columns = FALSE,
-  show_column_names = TRUE,
-  right_annotation = annotation
-)
+# plot loss in cnv
+plot_cnv_heatmap(classification_df, 
+                 cnv_col = "proportion_scaled_loss", 
+                 annotation = annotation, 
+                 legend_title = "Copy number \nloss")
 ```
@@
 
-## Save final tumor annotations
+In both of these plots, the majority of "Unknown" cells cluster between tumor and normal or with the normal cells.
+
+## Create final tumor annotations
+
+In looking at the plots, it looks like there might be some of the "Unknown" cells that should actually be classified as tumor cells. 
+In particular, we expect those that have a gain in Chr8 are tumor cells, so let's just look at Chr8 gains and see if we can find a cut off to use to categorize our "Unknown" cells.
+
+```{r}
+#density plot of chr8 gains
+ggplot(classification_df, aes(x = proportion_scaled_dupli_chr8, fill = combined_cnv_classification)) +
+  geom_density() +
+  facet_grid(rows = vars(combined_cnv_classification))
+```
+
+It looks like we have a pretty clear distinction between the `proportion_scaled_dupli_chr8` for normal and tumor cells and we could call any cell with a value > 0.25 as tumor cells and those that are less than 0.25 as normal cells.
+
+```{r}
+new_classification_df <- classification_df |> 
+  # remove duplicat
+  dplyr::select(-c(gene_symbol, ensembl_gene_id, gene_expression)) |> 
+  dplyr::distinct() |> 
+  dplyr::mutate(
+    combined_cnv_classification = dplyr::case_when(
+      combined_cnv_classification == "Unknown" & proportion_scaled_dupli_chr8 >= 0.25 ~ "Tumor",
+      combined_cnv_classification == "Unknown" & proportion_scaled_dupli_chr8 < 0.25 ~ "Normal",
+      .default = combined_cnv_classification
+    )
+  )
+```
+
+```{r}
+# account for new annotations 
+annotation <- ComplexHeatmap::rowAnnotation(classification = new_classification_df$combined_cnv_classification,
+                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))
+
+# recreate the cnv heatmap 
+plot_cnv_heatmap(new_classification_df,
+                 cnv_col = "proportion_scaled_dupli",
+                 annotation = annotation,
+                 legend_title = "Copy number \ngain")
+```
+
+It looks like we now have a prety good separation between tumor and normal cells based mostly on Chr8 gain. 
+Before we finish, we'll make a UMAP showing where the final annotations are.
+
+```{r}
+# look at which cells are tumor/normal
+new_classification_df |> 
+  ggplot(aes(x = UMAP1, y = UMAP2, color = combined_cnv_classification)) +
+  geom_point(alpha = 0.5, size = 0.5)
+```
 
@@
 # export final TSV with annotations 
-final_classifications <- coldata_df |> 
+final_classifications <- classification_df |> 
   dplyr::select(

### 2024-06-12 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
output the updated annotations

@@
 # export final TSV with annotations 
-final_classifications <- classification_df |> 
+final_classifications <- new_classification_df |> 
   dplyr::select(

### 2024-06-12 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
simplify case when statement

@@
       # if normal/tumor mismatch set to unknown 
-      cnv_proportion_classification == "Normal" & copykat_classification == "Tumor" ~ "Unknown",
-      cnv_proportion_classification == "Tumor" & copykat_classification == "Normal" ~ "Unknown",
-      cnv_proportion_classification == "Reference" &  copykat_classification == "Tumor" ~ "Unknown", 
-      !is.na(cnv_proportion_classification) & is.na(copykat_classification) ~ "Unknown", 
-      is.na(cnv_proportion_classification) & !is.na(copykat_classification) ~ "Unknown",
-      is.na(cnv_proportion_classification) & is.na(copykat_classification) ~ "Unknown"
+      .default = "Unknown"
     ))

### 2024-06-12 analyses/cell-type-ewings/template_notebooks/03-copykat.Rmd
make sure copykat results are saved to proper location

@@
 # output classifications file 
-copykat_results_file <- file.path(params$results_dir, glue::glue("{params$library_id}_copykat-classifications.tsv"))
+copykat_results_file <- file.path(params$results_dir, params$sample_id, glue::glue("{params$library_id}_copykat-classifications.tsv"))
 ```

### 2024-06-12 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_tumor-cell-validation.Rmd
initial notebook

--- 2024-06-12 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
@@
 ---
-title: "SCPCL000822 - Validation of tumor cells"
+title: "SCPCL000824 - Validation of tumor cells"
 author: Ally Hawkins
@@
 
-## Define functions
-
-```{r}
-# create a heatmap where rows are marker genes or gene sets and columns are cells 
-# this function adds a column annotation since cells are columns 
-plot_gene_heatmap <- function(df, 
-                              row_title = "",
-                              legend_title = "", 
-                              annotation = NULL){
-
-  # plot heatmap of marker genes 
-  heatmap <- ComplexHeatmap::Heatmap(
-    df,
-    # set the color scale based on min and max values
-    col = circlize::colorRamp2(seq(min(df), max(df), length = 2), colors = c("white", "#00274C")),
-    border = TRUE,
-    ## Row parameters
-    cluster_rows = TRUE,
-    row_title = row_title,
-    row_title_gp = grid::gpar(fontsize = 12),
-    row_title_side = "left",
-    row_names_side = "left",
-    row_dend_side = "right",
-    row_names_gp = grid::gpar(fontsize = 10),
-    ## Column parameters
-    cluster_columns = TRUE,
-    show_column_names = FALSE,
-    bottom_annotation = annotation,
-    heatmap_legend_param = list(
-      title = legend_title
-    )
-  )
-
-  return(heatmap)
-}
-
-# create a heatmap where rows are cells and columns are chromosomes 
-# this function adds a row annotation since cells are rows 
-plot_cnv_heatmap <- function(df,
-                             cnv_col, 
-                             annotation = NULL,
-                             legend_title = ""){
-
-  # barcode by chromosome df 
-  cnv_df <- df |> 
-    dplyr::select(barcodes, starts_with(cnv_col)) |> 
-    unique() |> 
-    tibble::column_to_rownames("barcodes")
-
-  # just get the name of the chromosome as column name 
-  colnames(cnv_df) <- stringr::word(colnames(cnv_df), -1, sep = "_")
-
-  # do a little bit of column reordering before converting to a matrix
-  cnv_df <- cnv_df |> 
-    dplyr::relocate(chr10, .before = chr11) |> 
-    dplyr::relocate(chr19, .before = chr20) |> 
-    dplyr::relocate(chr21, .before = chr22) |> 
-    as.matrix()
-
-  heatmap <- ComplexHeatmap::Heatmap(
-    cnv_df,
-    col = circlize::colorRamp2(seq(min(cnv_df), max(cnv_df), length = 2), colors = c("white", "#00274C")),
-    border = TRUE,
-    ## Row parameters
-    cluster_rows = TRUE,
-    row_title = "",
-    row_title_gp = grid::gpar(fontsize = 12),
-    row_title_side = "left",
-    show_row_names = FALSE,
-    row_dend_side = "right",
-    row_names_gp = grid::gpar(fontsize = 10),
-    ## Column parameters
-    cluster_columns = FALSE,
-    show_column_names = TRUE,
-    right_annotation = annotation,
-    heatmap_legend_param = list(
-      title = legend_title
-    )
-  )
-  return(heatmap)
-}
-
-```
-
 ## Setup
@@
 )
-
-# use jpegs for images 
-knitr::opts_chunk$set(
-  dev = "jpeg"
-)
-
-# quiet messages
-options(readr.show_col_types = FALSE)
-ComplexHeatmap::ht_opt(message = FALSE)
 ```
@@
 data_dir <- file.path(repository_base, "data", "2024-05-01")
-sample_dir <- file.path(data_dir, "SCPCP000015", "SCPCS000490")
+sample_dir <- file.path(data_dir, "SCPCP000015", "SCPCS000492")
 
@@
 # Input files
-sce_file <- file.path(sample_dir, "SCPCL000822_processed.rds")
+sce_file <- file.path(sample_dir, "SCPCL000824_processed.rds")
 marker_genes <- file.path(module_base, "references", "tumor-marker-genes.tsv")
@@
 # results from annotation workflow 
-results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000490")
+results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000492")
 
-marker_gene_results_file <- file.path(results_dir, "SCPCL000822_tumor-normal-classifications.tsv")
-copykat_predictions_file <- file.path(results_dir, "SCPCL000822_copykat-classifications.tsv")
+marker_gene_results_file <- file.path(results_dir, "SCPCL000824_tumor-normal-classifications.tsv")
+copykat_predictions_file <- file.path(results_dir, "SCPCL000824_copykat-classifications.tsv")
 
-infercnv_predictions_file <- file.path(results_dir, "SCPCL000822_infercnv-classifications.tsv")
-infercnv_metadata_file <- file.path(results_dir, "infercnv", "SCPCL000822_cnv-metadata.tsv")
+infercnv_predictions_file <- file.path(results_dir, "SCPCL000824_infercnv-classifications.tsv")
+infercnv_metadata_file <- file.path(results_dir, "infercnv", "SCPCL000824_cnv-metadata.tsv")
 
-geneset_scores_file <- file.path(results_dir, "SCPCL000822_gene-set-scores.tsv")
+geneset_scores_file <- file.path(results_dir, "SCPCL000824_gene-set-scores.tsv")
 
 # output files 
-final_annotations_dir <- file.path(module_base, "cell_type_annotations", "SCPCS000490")
+final_annotations_dir <- file.path(module_base, "cell_type_annotations", "SCPCS000492")
 fs::dir_create(final_annotations_dir)
-final_annotations_file <- file.path(final_annotations_dir, "SCPCL000822_tumor-classifications.tsv")
+final_annotations_file <- file.path(final_annotations_dir, "SCPCL000824_tumor-classifications.tsv")
 ```
@@
       copykat_classification == "diploid" ~ "Normal",
-      .default = copykat_classification
+      TRUE ~ copykat_classification
     )
@@
 
-# pull out the UMAP coordinates, add in marker genes and classifications 
-# and make a data frame to use for plotting
+# pull out the UMAP coordinates and make a data frame to use for plotting
 classification_df <- sce |> 
@@
   # join with marker gene df to get gene symbols for plotting 
-  dplyr::left_join(marker_genes_df, by = "ensembl_gene_id") |>
+  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
   # get rid of excess columns
@@
   # add in marker gene classifications
-  dplyr::left_join(marker_gene_results_df, by = "barcodes") |> 
+  dplyr::left_join(marker_gene_results_df) |> 
   # add in copykat results 
-  dplyr::left_join(copykat_results_df, by = "barcodes") |> 
+  dplyr::left_join(copykat_results_df) |> 
   # add in infercnv results and metadata
-  dplyr::left_join(infercnv_results_df, by = "barcodes") |> 
-  dplyr::left_join(infercnv_metadata_df, by = "barcodes") |> 
+  dplyr::left_join(infercnv_results_df) |> 
+  dplyr::left_join(infercnv_metadata_df) |> 
   # gene set scores 
-  dplyr::left_join(geneset_scores_df, by = "barcodes")
-
+  dplyr::left_join(geneset_scores_df)
 ```
@@
 
-- Marker genes only: For each cell, the expression of all marker genes was transformed using a z-transformation and summed together. 
-Any cells with a total marker gene expression value > 0 were called as tumor cells. 
-- CopyKAT - no reference: [`CopyKAT`](https://github.com/navinlabcode/copykat) was run without indicating any reference cells and cells were classified as aneuploid (Tumor) or diploid (Normal). 
-- InferCNV - [`InferCNV`](https://github.com/broadinstitute/inferCNV/wiki) was used to call copy number variations using endothelial and fibroblasts assigned by `CellAssign` as the normal reference group.
-This returns a proportion of each chromosome with a CNV detected.
-We then calculate the genomic CNV proportion for each cell across all chromosomes, weighted by the number of genes in a chromosome.
-Cells with a genomic CNV proportion greater than the mean  CNV proportion across all cells are called as tumor cells.
-
 ```{r}
 # plot tumor cells for each classification method 
-tumor_umap_df <- classification_df |> 
+tumor_umap_df <- coldata_df |> 
   dplyr::select(barcodes, UMAP1, UMAP2, marker_gene_classification, copykat_classification, cnv_proportion_classification) |> 
@@
     values_to = "classification"
-  ) |> 
-  dplyr::mutate(
-    method = dplyr::case_when(
-      method == "marker_gene_classification" ~ "Marker genes only",
-      method == "copykat_classification" ~ "CopyKAT - no reference",
-      method == "cnv_proportion_classification" ~ "InferCNV proportion"
-    )
   )
@@
 ```
-
-This plot shows us that there is a pretty clear consensus that most of the cells in the bottom right group are tumor cells. 
-From the original marker gene analysis we know that not all marker genes are expressed and that there is quite a bit of variance in expression. 
-Because of that and the fact that both CNV inference methods show similar annotations, we will classify any cells that are labeled as tumor by both `CopyKAT` and `InferCNV` as tumor cells.
-
-```{r}
-# create a combined classification between inferCNV and copyKAT
-# cells that are classified as tumor by both are kept as tumor
-# classified by both as normal are normal 
-# in between are classified as unknown
-classification_df <- classification_df |> 
-  dplyr::mutate(
-    combined_cnv_classification = dplyr::case_when(
-      # only tumor if both call tumor 
-      cnv_proportion_classification == "Tumor" & copykat_classification == "Tumor" ~ "Tumor",
-      # only normal if both call normal and/or reference
-      cnv_proportion_classification == "Normal" & copykat_classification == "Normal" ~ "Normal",
-      cnv_proportion_classification == "Reference" & copykat_classification == "Normal" ~ "Normal",
-      # if normal/tumor mismatch set to unknown 
-      .default = "Unknown"
-    ))
-```
-
-```{r}
-# look at which cells are tumor/normal
-# be sure to remove any duplicates 
-classification_df |> 
-  dplyr::select(UMAP1, UMAP2, barcodes, combined_cnv_classification) |>
-  dplyr::distinct() |> 
-  ggplot(aes(x = UMAP1, y = UMAP2, color = combined_cnv_classification)) +
-  geom_point(alpha = 0.5, size = 0.5)
-```
-
-The remainder of the notebook will focus on validating that these in fact are tumor cells and all other cells are classified as normal. 
-We will also look at where the unknown cells fall in relation to tumor vs. normal.
-
-## Validate tumor cells
-
-### Marker gene expression
-
-First, we will look at the expression of the tumor marker genes. 
-We expect that these should be expressed higher in the tumor cells than in the normal cells.
-
-In this first plot, we show a density plot of marker gene expression coloring by the cell classification (tumor or normal). 
-For this plot, we will include all classifications, including the new combined CNV classification.
-
-Just a note that the `marker_gene_classification` is a circular since these same marker genes were used to classify those cells.
-
-```{r}
-# calculate sum of the scaled gene expression values for plotting 
-marker_density_df <- classification_df |> 
-  dplyr::select(barcodes, gene_symbol, gene_expression, copykat_classification, cnv_proportion_classification, marker_gene_classification, combined_cnv_classification) |> 
-  unique() |> 
-  dplyr::group_by(gene_symbol) |>
-  # get z-scores for each gene
-  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |> 
-  dplyr::ungroup() |> 
-  dplyr::group_by(barcodes) |> 
-  # add all the scores together 
-  dplyr::mutate(sum_exp = sum(transformed_gene_expression)) |>
-  # expand classifications column for easy faceting 
-  dplyr::ungroup() |> 
-  tidyr::pivot_longer(
-    cols = ends_with("classification"),
-    names_to = "method",
-    values_to = "classification"
-  ) |> 
-  dplyr::mutate(
-    method = dplyr::case_when(
-      method == "marker_gene_classification" ~ "Marker genes only",
-      method == "copykat_classification" ~ "CopyKAT - no reference",
-      method == "cnv_proportion_classification" ~ "InferCNV proportion",
-      method == "combined_cnv_classification" ~ "CopyKAT and InferCNV"
-    )
-  )
-
-ggplot(marker_density_df, aes(x = sum_exp, color = classification)) +
-  geom_density() +
-  facet_wrap(vars(method))
-```
-
-This plot shows us that overall the tumor cells in each method have higher marker gene expression. 
-Although the `InferCNV` alone and combined CNV classifications appear to have the most distinct separation in marker gene expression.
-
-Another note is that the "Unknown" class appears to have marker gene expression more similar to the normal cells.
-
-Next, we will plot the expression of each gene in each cell using a heatmap. 
-Here the rows are the marker genes and the columns are the cells with the color indicating the expression of that gene in that cell.
-
-```{r}
-# create matrix with marker genes as rows and barcodes as columns
-marker_gene_heatmap <- classification_df |> 
-  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
-  tidyr::pivot_wider(values_from = gene_expression, 
-                     names_from = barcodes) |> 
-  tibble::column_to_rownames("gene_symbol") |> 
-  as.matrix()
-
-# create annotation df, make sure that dimensions match up by only getting unique barcodes 
-annotation_df <- classification_df |> 
-  dplyr::select(barcodes, combined_cnv_classification) |> 
-  unique()
-
-annotation <- ComplexHeatmap::columnAnnotation(
-  classification = annotation_df$combined_cnv_classification,
-  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))
-```
-
-```{r}
-# plot heatmap of marker genes 
-plot_gene_heatmap(marker_gene_heatmap, 
-                  row_title = "Marker gene symbol",
-                  legend_title = "Marker gene \nexpression",
-                  annotation = annotation)
-```
-
-As expected, the tumor cells appear to have more marker genes that are expressed than normal cells. 
-The "Unknown" cells appear to have a small group of cells that cluster with the tumor cells on the right hand side of the plot.
-
-### Gene set scores
-
-Next, we will look at the gene set scores for three different EWS-FLI1 gene sets. 
-The below plots will all use the `mean-GENESET` columns.
-
-First we will just look at the gene set scores for each cell on the UMAP.
-
-```{r}
-# plot tumor cells for each classification method 
-geneset_plot_df <- classification_df |> 
-  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean")) |> 
-  tidyr::pivot_longer(
-    cols = starts_with("mean"),
-    names_to = "geneset",
-    values_to = "mean_score"
-  ) |> 
-  dplyr::mutate(
-    geneset = stringr::word(geneset, -1, sep = "-")
-  )
-
-ggplot(geneset_plot_df, aes(x = UMAP1, UMAP2, color = mean_score)) +
-  geom_point(size = 0.5, alpha = 0.5) +
-  facet_wrap(vars(geneset)) +
-  scale_color_viridis_c()
-```
-
-Generally, the higher gene set scores appear to be in the group of cells that are tumor cells with some slight variations between gene sets.
-
-Below we look at gene set scores for cells classified as tumor cells vs normal cells for all classification methods.
-
-```{r, fig.height=10}
-geneset_plot_df |> 
-  tidyr::pivot_longer(
-    cols = ends_with("classification"),
-    names_to = "method",
-    values_to = "classification"
-  ) |> 
-  ggplot(aes(x = mean_score, color = classification)) +
-  geom_density(bw = 0.05) +
-  facet_grid(rows = vars(method), 
-             cols = vars(geneset))
-```
-
-For the most part the gene set scores are higher in tumor cells regardless of classification.
-
-We also see that the "Unknown" cells identified in the combined classification mostly mirror the gene set expression scores for normal cells with maybe a small group of cells having more similar gene set scores to tumor cells.
-
-Below we make a heatmap of the gene set scores for just the combined CNV classification. 
-Each row is a gene set and each column is a barcode. 
-The color indicates the gene set score for that cell.
-
-```{r}
-# make a matrix of gene set by barcode 
-geneset_heatmap <- geneset_plot_df |> 
-  dplyr::select(mean_score, geneset, barcodes) |> 
-  unique() |> 
-  tidyr::pivot_wider(values_from = mean_score, 
-                     names_from = barcodes) |> 
-  tibble::column_to_rownames("geneset") |> 
-  as.matrix()
-
-# plot heatmap of gene set score 
-plot_gene_heatmap(geneset_heatmap,
-                  annotation = annotation,
-                  legend_title = "Gene set \nscore")
-```
-
-For the most part we see that cells with higher gene set scores are classified as tumor cells. 
-One thing to note is that the `SILIGAN` gene set appears to have some expression in normal cells. 
-However, if you look at the spots where all three gene sets have higher scores, these apply to tumor cells.
-
-Again, most of the "Unknown" tumor cells tend to align with normal cells, with a few grouping in the area where cells transition from tumor to normal.
-
-### Validate based on CNAs
-
-The last thing we will do here is look to see if tumor cells have any of the expected CNAs found in Ewing sarcoma:
-
-- Gain of Chr8
-- Gain of Chr12
-- Gain of Chr1p
-- Loss of Chr16q
-
-Although these are the most frequent, there are patients who do not have any of these alterations and patients that only have some of these alterations. 
-See [Tirode et al.,](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4264969/) and [Crompton et al.](https://doi.org/10.1158/2159-8290.CD-13-1037).
-
-We will look specifically at gains and losses in each chromosome across tumor and normal cells. 
-To do this, we will use the output from `InferCNV` since that output separates out gains and losses in a way that makes it easy for plotting.
-
-The first heatmap looks at the proportion of each chromosome with a CNV that corresponds to an amplification or gain. 
-Here each row is a cell and each column is a chromosome with the color corresponding to the proportion of the chromosome with a gain detected.
-
-```{r}
-# need to make a new annotation, since we want to annotate the rows here 
-annotation <- ComplexHeatmap::rowAnnotation(classification = annotation_df$combined_cnv_classification,
-                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))
-
-# plot gains in cnv
-plot_cnv_heatmap(classification_df,
-                 cnv_col = "proportion_scaled_dupli",
-                 annotation = annotation,
-                 legend_title = "Copy number \ngain")
-```
-
-Here we see a gain in Chr8 and Chr1 in tumor cells compared to normal cells. 
-We also see other chromosomes with amplifications in tumor cells compared to normal cells and that the tumor and normal cells are nicely separated in the heirarchical clustering.
-
-Below we make the same plot but with only the losses or deletions.
-
-```{r}
-# plot loss in cnv
-plot_cnv_heatmap(classification_df, 
-                 cnv_col = "proportion_scaled_loss", 
-                 annotation = annotation, 
-                 legend_title = "Copy number \nloss")
-```
-
-We do not see the expected loss in Chr16, but this is not present in every patient. 
-Additionally, we had already noted the gain in Chr8 in the tumor cells, so there is still at least one expected CNV present. 
-Again, there is a clear separation of tumor and normal cells in terms of presence of CNVs.
-
-In both of these plots, the majority of "Unknown" cells cluster between tumor and normal or with the normal cells.
-
-## Create final tumor annotations
-
-In looking at the plots, it looks like there might be some of the "Unknown" cells that should actually be classified as tumor cells. 
-In particular, we expect those that have a gain in Chr8 are tumor cells, so let's just look at Chr8 gains and see if we can find a cut off to use to categorize our "Unknown" cells.
-
-```{r}
-#density plot of chr8 gains
-ggplot(classification_df, aes(x = proportion_scaled_dupli_chr8, fill = combined_cnv_classification)) +
-  geom_density() +
-  facet_grid(rows = vars(combined_cnv_classification))
-```
-
-It looks like we have a pretty clear distinction between the `proportion_scaled_dupli_chr8` for normal and tumor cells and we could call any cell with a value > 0.25 as tumor cells and those that are less than 0.25 as normal cells.
-
-```{r}
-new_classification_df <- classification_df |> 
-  # remove duplicat
-  dplyr::select(-c(gene_symbol, ensembl_gene_id, gene_expression)) |> 
-  dplyr::distinct() |> 
-  dplyr::mutate(
-    combined_cnv_classification = dplyr::case_when(
-      combined_cnv_classification == "Unknown" & proportion_scaled_dupli_chr8 >= 0.25 ~ "Tumor",
-      combined_cnv_classification == "Unknown" & proportion_scaled_dupli_chr8 < 0.25 ~ "Normal",
-      .default = combined_cnv_classification
-    )
-  )
-```
-
-```{r}
-# account for new annotations 
-annotation <- ComplexHeatmap::rowAnnotation(classification = new_classification_df$combined_cnv_classification,
-                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))
-
-# recreate the cnv heatmap 
-plot_cnv_heatmap(new_classification_df,
-                 cnv_col = "proportion_scaled_dupli",
-                 annotation = annotation,
-                 legend_title = "Copy number \ngain")
-```
-
-It looks like we now have a prety good separation between tumor and normal cells based mostly on Chr8 gain. 
-Before we finish, we'll make a UMAP showing where the final annotations are.
-
-```{r}
-# look at which cells are tumor/normal
-new_classification_df |> 
-  ggplot(aes(x = UMAP1, y = UMAP2, color = combined_cnv_classification)) +
-  geom_point(alpha = 0.5, size = 0.5)
-```
-
-After validating that the cells classified as tumor by both `CopyKAT` and `InferCNV` show the expected expression patterns and CNV patterns as tumor cells, we can confidently label those cells as tumor cells.
-We will export these labels to the final annotations file for later use.
-
-```{r}
-# export final TSV with annotations 
-final_classifications <- new_classification_df |> 
-  dplyr::select(
-    cell_barcode = barcodes,
-    tumor_cell_classification = combined_cnv_classification
-  ) |> 
-  unique()
-
-readr::write_tsv(final_classifications, final_annotations_file)
-```
-
-## Session Info
-
-```{r session info}
-# record the versions of the packages used in this analysis and other environment information
-sessionInfo()
-```

### 2024-06-14 analyses/cell-type-ewings/template_notebooks/03-copykat.Rmd
add mean cnv detection to copykat output

@@
 
-readr::write_tsv(celltype_df, copykat_results_file)
+all_info_df <- cnv_df |> 
+  dplyr::select(barcodes,
+                mean_cnv_detection) |> 
+  dplyr::group_by(barcodes) |>
+  dplyr::mutate(mean_cnv_detection = mean(mean_cnv_detection)) |> 
+  dplyr::distinct() |> 
+  dplyr::left_join(celltype_df)
+
+readr::write_tsv(all_info_df, copykat_results_file)
 ```

### 2024-06-14 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_tumor-cell-validation.Rmd
some more exploratory plots

@@
   - Confirming presence of known CNVs in Ewing sarcoma
+
+## Define functions
+
+```{r}
+# create a heatmap where rows are marker genes or gene sets and columns are cells 
+# this function adds a column annotation since cells are columns 
+plot_gene_heatmap <- function(df, 
+                              row_title = "",
+                              legend_title = "", 
+                              annotation = NULL){
+
+  # plot heatmap of marker genes 
+  heatmap <- ComplexHeatmap::Heatmap(
+    df,
+    # set the color scale based on min and max values
+    col = circlize::colorRamp2(seq(min(df), max(df), length = 2), colors = c("white", "#00274C")),
+    border = TRUE,
+    ## Row parameters
+    cluster_rows = TRUE,
+    row_title = row_title,
+    row_title_gp = grid::gpar(fontsize = 12),
+    row_title_side = "left",
+    row_names_side = "left",
+    row_dend_side = "right",
+    row_names_gp = grid::gpar(fontsize = 10),
+    ## Column parameters
+    cluster_columns = TRUE,
+    show_column_names = FALSE,
+    bottom_annotation = annotation,
+    heatmap_legend_param = list(
+      title = legend_title
+    )
+  )
+
+  return(heatmap)
+}
+
+# create a heatmap where rows are cells and columns are chromosomes 
+# this function adds a row annotation since cells are rows 
+plot_cnv_heatmap <- function(df,
+                             cnv_col, 
+                             annotation = NULL,
+                             legend_title = ""){
+
+  # barcode by chromosome df 
+  cnv_df <- df |> 
+    dplyr::select(barcodes, starts_with(cnv_col)) |> 
+    unique() |> 
+    tibble::column_to_rownames("barcodes")
+
+  # just get the name of the chromosome as column name 
+  colnames(cnv_df) <- stringr::word(colnames(cnv_df), -1, sep = "_")
+
+  # do a little bit of column reordering before converting to a matrix
+  cnv_df <- cnv_df |> 
+    dplyr::relocate(chr10, .before = chr11) |> 
+    dplyr::relocate(chr19, .before = chr20) |> 
+    dplyr::relocate(chr21, .before = chr22) |> 
+    as.matrix()
+
+  heatmap <- ComplexHeatmap::Heatmap(
+    cnv_df,
+    col = circlize::colorRamp2(seq(min(cnv_df), max(cnv_df), length = 2), colors = c("white", "#00274C")),
+    border = TRUE,
+    ## Row parameters
+    cluster_rows = TRUE,
+    row_title = "",
+    row_title_gp = grid::gpar(fontsize = 12),
+    row_title_side = "left",
+    show_row_names = FALSE,
+    row_dend_side = "right",
+    row_names_gp = grid::gpar(fontsize = 10),
+    ## Column parameters
+    cluster_columns = FALSE,
+    show_column_names = TRUE,
+    right_annotation = annotation,
+    heatmap_legend_param = list(
+      title = legend_title
+    )
+  )
+  return(heatmap)
+}
+
+```
 
@@
   # only select no_ref classification and rename to be consistent with naming of other classifications
-  dplyr::select(barcodes, copykat_classification = no_ref) |> 
+  dplyr::select(barcodes, copykat_classification = no_ref, mean_cnv_detection) |> 
+  dplyr::distinct() |> 
   dplyr::mutate(
@@
       ) |>
-  # add in marker gene expression to dataframe 
-  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
-  # combine all genes into a single column for easy faceting
-  tidyr::pivot_longer(
-    cols = starts_with("ENSG"),
-    names_to = "ensembl_gene_id",
-    values_to = "gene_expression"
-  ) |> 
-  # join with marker gene df to get gene symbols for plotting 
-  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
   # get rid of excess columns
-  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression) |> 
+  dplyr::select(barcodes, UMAP1, UMAP2) |> 
   # add in marker gene classifications
@@
 # plot tumor cells for each classification method 
-tumor_umap_df <- coldata_df |> 
+tumor_umap_df <- classification_df |> 
   dplyr::select(barcodes, UMAP1, UMAP2, marker_gene_classification, copykat_classification, cnv_proportion_classification) |> 
@@
 ```
+
+```{r}
+classification_df <- classification_df |>
+  dplyr::mutate(
+    geneset_classification = dplyr::if_else(
+      `scaled_mean-RIGGI` > 0 | `scaled_mean-ZHANG` > 0 | `scaled_mean-SILIGAN` > 0,
+      "Tumor",
+      "Normal"
+    ),
+    combined_classification = dplyr::case_when(
+      marker_gene_classification == "Normal" & geneset_classification == "Normal" ~ "Normal",
+      marker_gene_classification == "Tumor" & geneset_classification == "Tumor" ~ "Tumor",
+      .default = "Unknown"
+    )
+    # combined_classification = dplyr::case_when(
+    #   copykat_classification == "Normal" & cnv_proportion_classification == "Normal" & marker_gene_classification == "Normal" ~ "Normal",
+    #   copykat_classification == "Normal" & cnv_proportion_classification == "Reference" & marker_gene_classification == "Normal" ~ "Normal",
+    #   copykat_classification == "Tumor" & cnv_proportion_classification == "Tumor" & marker_gene_classification == "Tumor" ~ "Tumor",
+    #   .default = "Unknown"
+    # )
+  )
+```
+
+```{r}
+ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = geneset_classification)) +
+  geom_point(size = 0.5, alpha = 0.5)
+```
+
+```{r}
+# calculate sum of the scaled gene expression values for plotting 
+marker_density_df <- classification_df |> 
+  # add in marker gene expression to dataframe 
+  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
+  # combine all genes into a single column for easy faceting
+  tidyr::pivot_longer(
+    cols = starts_with("ENSG"),
+    names_to = "ensembl_gene_id",
+    values_to = "gene_expression"
+  ) |> 
+  # join with marker gene df to get gene symbols for plotting 
+  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |> 
+  dplyr::select(barcodes, gene_symbol, gene_expression, copykat_classification, cnv_proportion_classification, marker_gene_classification, combined_classification, geneset_classification) |> 
+  unique() |> 
+  dplyr::group_by(gene_symbol) |>
+  # get z-scores for each gene
+  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |> 
+  dplyr::ungroup() |> 
+  dplyr::group_by(barcodes) |> 
+  # add all the scores together 
+  dplyr::mutate(sum_exp = sum(transformed_gene_expression)) |>
+  # expand classifications column for easy faceting 
+  dplyr::ungroup() |> 
+  tidyr::pivot_longer(
+    cols = ends_with("classification"),
+    names_to = "method",
+    values_to = "classification"
+  ) |> 
+  dplyr::mutate(
+    method = dplyr::case_when(
+      method == "marker_gene_classification" ~ "Marker genes only",
+      method == "copykat_classification" ~ "CopyKAT - no reference",
+      method == "cnv_proportion_classification" ~ "InferCNV proportion",
+      method == "combined_classification" ~ "Combined",
+      method == "geneset_classification" ~ "EWS-FLI1 Gene set"
+    )
+  )
+
+ggplot(marker_density_df, aes(x = sum_exp, color = classification)) +
+  geom_density() +
+  facet_wrap(vars(method))
+```
+
+```{r}
+# plot tumor cells for each classification method 
+geneset_plot_df <- classification_df |> 
+  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("scaled_mean")) |> 
+  tidyr::pivot_longer(
+    cols = starts_with("scaled_mean"),
+    names_to = "geneset",
+    values_to = "mean_score"
+  ) |> 
+  dplyr::mutate(
+    geneset = stringr::word(geneset, -1, sep = "-")
+  )
+
+ggplot(geneset_plot_df, aes(x = UMAP1, UMAP2, color = mean_score)) +
+  geom_point(size = 0.5, alpha = 0.5) +
+  facet_wrap(vars(geneset)) +
+  scale_color_viridis_c()
+```
+
+```{r}
+geneset_plot_df |> 
+  tidyr::pivot_longer(
+    cols = ends_with("classification"),
+    names_to = "method",
+    values_to = "classification"
+  ) |> 
+  ggplot(aes(x = mean_score, color = classification)) +
+  geom_density(bw = 0.05) +
+  facet_grid(rows = vars(method), 
+             cols = vars(geneset))
+```
+
+CNV heatmap annotated by copykat and marker gene prediction, two annotation columns
+
+```{r}
+annotation_df <- classification_df |> 
+  dplyr::select(barcodes, ends_with("classification")) |> 
+  unique()
+
+# need to make a new annotation, since we want to annotate the rows here 
+annotation <- ComplexHeatmap::rowAnnotation(classification = annotation_df$combined_classification,
+                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))
+
+# plot gains in cnv
+plot_cnv_heatmap(classification_df,
+                 cnv_col = "proportion_scaled_dupli",
+                 annotation = annotation,
+                 legend_title = "Copy number \ngain")
+```
+
+```{r}
+# plot gains in cnv
+plot_cnv_heatmap(classification_df,
+                 cnv_col = "proportion_scaled_loss",
+                 annotation = annotation,
+                 legend_title = "Copy number \nloss")
+```
+
+Density plot with mean proportion CNV from InferCNV on x-axis, colored by prediction, faceted by classification method
+
+```{r}
+classification_df |> 
+  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), mean_cnv_proportion) |> 
+  tidyr::pivot_longer(
+    cols = ends_with("classification"),
+    names_to = "method",
+    values_to = "classification"
+  ) |> 
+  ggplot(aes(x = mean_cnv_proportion, color = classification)) +
+  geom_density(bw = 0.05) +
+  facet_grid(rows = vars(method))
+```
+
+Mean InferCNV proportion on x-axis, mean copyKAT proportion on y-axis, color by scaled sum of marker gene expression
+
+```{r}
+cnv_prop_df <- classification_df |> 
+  dplyr::select(barcodes, ends_with("classification"), starts_with("scaled_mean"), mean_cnv_proportion, mean_cnv_detection) |> 
+  tidyr::pivot_longer(
+    cols = starts_with("scaled_mean"),
+    names_to = "geneset",
+    values_to = "mean_score"
+  ) |> 
+  dplyr::mutate(
+    geneset = stringr::word(geneset, -1, sep = "-")
+  )
+
+ggplot(cnv_prop_df, aes(x = mean_cnv_proportion, y = abs(mean_cnv_detection), color = mean_score)) +
+  geom_point(size = 0.5, alpha = 0.5) +
+  facet_wrap(vars(geneset)) +
+  scale_color_viridis_c()
+```
+
+```{r}
+ggplot(cnv_prop_df, aes(x = mean_score, y = abs(mean_cnv_detection))) +
+  geom_point(size = 0.5, alpha = 0.5) +
+  facet_wrap(vars(geneset)) +
+  scale_color_viridis_c()
+```
+
+faceted density plot with marker gene expression on x axis, colored by copykat and infercnv prediction and faceted by marker gene
+
+```{r}
+geneset_scores_df |> 
+  dplyr::pivot
+
+ggplot(geneset_scores_df, aes(x = mean_score, color = ))
+```
+
+use SingleR to classify as tumor/normal using SCPCS000490
+
+correlation of geneset score and marker gene score 
+```{r}
+geneset_plot_df |>
+
+  dplyr::left_join(marker_density_df) |> 
+  ggplot(aes(x = sum_exp, y = mean_score)) +
+    facet_wrap()
+```

### 2024-06-14 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
rename unknown to ambiguous and no extra classification

@@
   - Confirming presence of known CNVs in Ewing sarcoma
+
+## Setup
+
+```{r packages}
+suppressPackageStartupMessages({
+  # load required packages
+  library(SingleCellExperiment)
+  library(ggplot2)
+})
+
+# Set default ggplot theme
+theme_set(
+  theme_bw()
+)
+
+# use jpegs for images 
+knitr::opts_chunk$set(
+  dev = "jpeg"
+)
+
+# quiet messages
+options(readr.show_col_types = FALSE)
+ComplexHeatmap::ht_opt(message = FALSE)
+```
+
+```{r base paths}
+# The base path for the OpenScPCA repository, found by its (hidden) .git directory
+repository_base <- rprojroot::find_root(rprojroot::is_git_root)
+
+# The current data directory, found within the repository base directory
+data_dir <- file.path(repository_base, "data", "2024-05-01")
+sample_dir <- file.path(data_dir, "SCPCP000015", "SCPCS000490")
+
+# The path to this module
+module_base <- file.path(repository_base, "analyses", "cell-type-ewings")
+```
+
+```{r}
+# Input files
+sce_file <- file.path(sample_dir, "SCPCL000822_processed.rds")
+marker_genes <- file.path(module_base, "references", "tumor-marker-genes.tsv")
+
+# results from annotation workflow 
+results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000490")
+
+marker_gene_results_file <- file.path(results_dir, "SCPCL000822_tumor-normal-classifications.tsv")
+copykat_predictions_file <- file.path(results_dir, "SCPCL000822_copykat-classifications.tsv")
+
+infercnv_predictions_file <- file.path(results_dir, "SCPCL000822_infercnv-classifications.tsv")
+infercnv_metadata_file <- file.path(results_dir, "infercnv", "SCPCL000822_cnv-metadata.tsv")
+
+geneset_scores_file <- file.path(results_dir, "SCPCL000822_gene-set-scores.tsv")
+
+# output files 
+final_annotations_dir <- file.path(module_base, "cell_type_annotations", "SCPCS000490")
+fs::dir_create(final_annotations_dir)
+final_annotations_file <- file.path(final_annotations_dir, "SCPCL000822_tumor-classifications.tsv")
+```
+
+### Process input files
+
+```{r}
+# read in all files 
+sce <- readr::read_rds(sce_file)
+
+# read in marker genes table 
+marker_genes_df <- readr::read_tsv(marker_genes) |> 
+  # account for genes being from multiple sources
+  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
+  dplyr::distinct()
+
+marker_gene_results_df <- readr::read_tsv(marker_gene_results_file)
+
+# copykat output
+copykat_results_df <- readr::read_tsv(copykat_predictions_file) |> 
+  # only select no_ref classification and rename to be consistent with naming of other classifications
+  dplyr::select(barcodes, copykat_classification = no_ref) |> 
+  dplyr::mutate(
+    copykat_classification = dplyr::case_when(
+      copykat_classification == "aneuploid" ~ "Tumor",
+      copykat_classification == "diploid" ~ "Normal",
+      .default = copykat_classification
+    )
+  )
+
+# infercnv output 
+infercnv_results_df <- readr::read_tsv(infercnv_predictions_file)
+infercnv_metadata_df <- read.delim(infercnv_metadata_file) |> 
+  tibble::rownames_to_column("barcodes")
+
+# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
+chr_columns_to_keep <- which(!stringr::str_detect(names(infercnv_metadata_df), "chrMT|chrGL"))
+infercnv_metadata_df <- infercnv_metadata_df[, chr_columns_to_keep]
+
+# gene set scores 
+geneset_scores_df <- readr::read_tsv(geneset_scores_file)
+```
+
+```{r}
+# grab all coldata and add in all classifications and results to use for plotting throughout notebook 
+# get list of marker genes to use 
+marker_genes <- marker_genes_df |> 
+  dplyr::filter(cell_type == "tumor") |> 
+  dplyr::pull(ensembl_gene_id)
+
+# get the gene expression counts for all marker genes
+marker_gene_exp <- logcounts(sce[marker_genes, ]) |>
+  as.matrix() |>
+  t() |>
+  as.data.frame() |>
+  tibble::rownames_to_column("barcodes")
+
+# pull out the UMAP coordinates, add in marker genes and classifications 
+# and make a data frame to use for plotting
+classification_df <- sce |> 
+  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
+  # replace UMAP.1 with UMAP1
+  dplyr::rename_with(
+        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
+      ) |> 
+  # add in marker gene expression to dataframe 
+  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
+  # combine all genes into a single column for easy faceting
+  tidyr::pivot_longer(
+    cols = starts_with("ENSG"),
+    names_to = "ensembl_gene_id",
+    values_to = "gene_expression"
+  ) |> 
+  # join with marker gene df to get gene symbols for plotting 
+  dplyr::left_join(marker_genes_df, by = "ensembl_gene_id") |>
+  # get rid of excess columns
+  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression) |> 
+  # add in marker gene classifications
+  dplyr::left_join(marker_gene_results_df, by = "barcodes") |> 
+  # add in copykat results 
+  dplyr::left_join(copykat_results_df, by = "barcodes") |> 
+  # add in infercnv results and metadata
+  dplyr::left_join(infercnv_results_df, by = "barcodes") |> 
+  dplyr::left_join(infercnv_metadata_df, by = "barcodes") |> 
+  # gene set scores 
+  dplyr::left_join(geneset_scores_df, by = "barcodes")
+
+```
 
@@
 
-## Setup
-
-```{r packages}
-suppressPackageStartupMessages({
-  # load required packages
-  library(SingleCellExperiment)
-  library(ggplot2)
-})
-
-# Set default ggplot theme
-theme_set(
-  theme_bw()
-)
-
-# use jpegs for images 
-knitr::opts_chunk$set(
-  dev = "jpeg"
-)
-
-# quiet messages
-options(readr.show_col_types = FALSE)
-ComplexHeatmap::ht_opt(message = FALSE)
-```
-
-```{r base paths}
-# The base path for the OpenScPCA repository, found by its (hidden) .git directory
-repository_base <- rprojroot::find_root(rprojroot::is_git_root)
-
-# The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "2024-05-01")
-sample_dir <- file.path(data_dir, "SCPCP000015", "SCPCS000490")
-
-# The path to this module
-module_base <- file.path(repository_base, "analyses", "cell-type-ewings")
-```
-
-```{r}
-# Input files
-sce_file <- file.path(sample_dir, "SCPCL000822_processed.rds")
-marker_genes <- file.path(module_base, "references", "tumor-marker-genes.tsv")
-
-# results from annotation workflow 
-results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000490")
-
-marker_gene_results_file <- file.path(results_dir, "SCPCL000822_tumor-normal-classifications.tsv")
-copykat_predictions_file <- file.path(results_dir, "SCPCL000822_copykat-classifications.tsv")
-
-infercnv_predictions_file <- file.path(results_dir, "SCPCL000822_infercnv-classifications.tsv")
-infercnv_metadata_file <- file.path(results_dir, "infercnv", "SCPCL000822_cnv-metadata.tsv")
-
-geneset_scores_file <- file.path(results_dir, "SCPCL000822_gene-set-scores.tsv")
-
-# output files 
-final_annotations_dir <- file.path(module_base, "cell_type_annotations", "SCPCS000490")
-fs::dir_create(final_annotations_dir)
-final_annotations_file <- file.path(final_annotations_dir, "SCPCL000822_tumor-classifications.tsv")
-```
-
-```{r}
-# read in all files 
-sce <- readr::read_rds(sce_file)
-
-# read in marker genes table 
-marker_genes_df <- readr::read_tsv(marker_genes) |> 
-  # account for genes being from multiple sources
-  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
-  dplyr::distinct()
-
-marker_gene_results_df <- readr::read_tsv(marker_gene_results_file)
-
-# copykat output
-copykat_results_df <- readr::read_tsv(copykat_predictions_file) |> 
-  # only select no_ref classification and rename to be consistent with naming of other classifications
-  dplyr::select(barcodes, copykat_classification = no_ref) |> 
-  dplyr::mutate(
-    copykat_classification = dplyr::case_when(
-      copykat_classification == "aneuploid" ~ "Tumor",
-      copykat_classification == "diploid" ~ "Normal",
-      .default = copykat_classification
-    )
-  )
-
-# infercnv output 
-infercnv_results_df <- readr::read_tsv(infercnv_predictions_file)
-infercnv_metadata_df <- read.delim(infercnv_metadata_file) |> 
-  tibble::rownames_to_column("barcodes")
-
-# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
-chr_columns_to_keep <- which(!stringr::str_detect(names(infercnv_metadata_df), "chrMT|chrGL"))
-infercnv_metadata_df <- infercnv_metadata_df[, chr_columns_to_keep]
-
-# gene set scores 
-geneset_scores_df <- readr::read_tsv(geneset_scores_file)
-```
-
-```{r}
-# grab all coldata and add in all classifications and results to use for plotting throughout notebook 
-# get list of marker genes to use 
-marker_genes <- marker_genes_df |> 
-  dplyr::filter(cell_type == "tumor") |> 
-  dplyr::pull(ensembl_gene_id)
-
-# get the gene expression counts for all marker genes
-marker_gene_exp <- logcounts(sce[marker_genes, ]) |>
-  as.matrix() |>
-  t() |>
-  as.data.frame() |>
-  tibble::rownames_to_column("barcodes")
-
-# pull out the UMAP coordinates, add in marker genes and classifications 
-# and make a data frame to use for plotting
-classification_df <- sce |> 
-  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
-  # replace UMAP.1 with UMAP1
-  dplyr::rename_with(
-        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
-      ) |> 
-  # add in marker gene expression to dataframe 
-  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
-  # combine all genes into a single column for easy faceting
-  tidyr::pivot_longer(
-    cols = starts_with("ENSG"),
-    names_to = "ensembl_gene_id",
-    values_to = "gene_expression"
-  ) |> 
-  # join with marker gene df to get gene symbols for plotting 
-  dplyr::left_join(marker_genes_df, by = "ensembl_gene_id") |>
-  # get rid of excess columns
-  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression) |> 
-  # add in marker gene classifications
-  dplyr::left_join(marker_gene_results_df, by = "barcodes") |> 
-  # add in copykat results 
-  dplyr::left_join(copykat_results_df, by = "barcodes") |> 
-  # add in infercnv results and metadata
-  dplyr::left_join(infercnv_results_df, by = "barcodes") |> 
-  dplyr::left_join(infercnv_metadata_df, by = "barcodes") |> 
-  # gene set scores 
-  dplyr::left_join(geneset_scores_df, by = "barcodes")
-
-```
-
 ## Identify tumor cell population
@@
       # if normal/tumor mismatch set to unknown 
-      .default = "Unknown"
-    ))
+      .default = "Ambiguous"
+    ), 
+    # set factor order to use for plotting
+    combined_cnv_classification = forcats::fct_relevel(combined_cnv_classification, "Tumor", "Normal", "Ambiguous"))
 ```
@@
   classification = annotation_df$combined_cnv_classification,
-  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))
+  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
 ```
@@
 annotation <- ComplexHeatmap::rowAnnotation(classification = annotation_df$combined_cnv_classification,
-                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))
+                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
 
@@
 
-It looks like we have a pretty clear distinction between the `proportion_scaled_dupli_chr8` for normal and tumor cells and we could call any cell with a value > 0.25 as tumor cells and those that are less than 0.25 as normal cells.
-
-```{r}
-new_classification_df <- classification_df |> 
-  # remove duplicat
-  dplyr::select(-c(gene_symbol, ensembl_gene_id, gene_expression)) |> 
-  dplyr::distinct() |> 
-  dplyr::mutate(
-    combined_cnv_classification = dplyr::case_when(
-      combined_cnv_classification == "Unknown" & proportion_scaled_dupli_chr8 >= 0.25 ~ "Tumor",
-      combined_cnv_classification == "Unknown" & proportion_scaled_dupli_chr8 < 0.25 ~ "Normal",
-      .default = combined_cnv_classification
-    )
-  )
-```
-
-```{r}
-# account for new annotations 
-annotation <- ComplexHeatmap::rowAnnotation(classification = new_classification_df$combined_cnv_classification,
-                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))
-
-# recreate the cnv heatmap 
-plot_cnv_heatmap(new_classification_df,
-                 cnv_col = "proportion_scaled_dupli",
-                 annotation = annotation,
-                 legend_title = "Copy number \ngain")
-```
-
-It looks like we now have a prety good separation between tumor and normal cells based mostly on Chr8 gain. 
-Before we finish, we'll make a UMAP showing where the final annotations are.
-
-```{r}
-# look at which cells are tumor/normal
-new_classification_df |> 
-  ggplot(aes(x = UMAP1, y = UMAP2, color = combined_cnv_classification)) +
-  geom_point(alpha = 0.5, size = 0.5)
-```
+It looks like we have a pretty clear distinction between the `proportion_scaled_dupli_chr8` for normal and tumor cells, but there are a few bumps indicating some normal cells do have higher values for CNVs at Chr8 that are more similar tumor cells.
+Because of this, we will keep the "Ambiguous" cells as is and not try to classify them.
 

### 2024-06-14 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 # classified by both as normal are normal 
-# in between are classified as unknown
+# in between are classified as ambiguous
 classification_df <- classification_df |> 
@@
       cnv_proportion_classification == "Reference" & copykat_classification == "Normal" ~ "Normal",
-      # if normal/tumor mismatch set to unknown 
+      # if normal/tumor mismatch set to ambiguous 
       .default = "Ambiguous"
@@
 The remainder of the notebook will focus on validating that these in fact are tumor cells and all other cells are classified as normal. 
-We will also look at where the unknown cells fall in relation to tumor vs. normal.
+We will also look at where the ambiguous cells fall in relation to tumor vs. normal.
 
@@
 
-Another note is that the "Unknown" class appears to have marker gene expression more similar to the normal cells.
+Another note is that the "Ambiguous" class appears to have marker gene expression more similar to the normal cells.
 
@@
 As expected, the tumor cells appear to have more marker genes that are expressed than normal cells. 
-The "Unknown" cells appear to have a small group of cells that cluster with the tumor cells on the right hand side of the plot.
+The "Ambiguous" cells appear to have a small group of cells that cluster with the tumor cells on the right hand side of the plot.
 
@@
 
-We also see that the "Unknown" cells identified in the combined classification mostly mirror the gene set expression scores for normal cells with maybe a small group of cells having more similar gene set scores to tumor cells.
+We also see that the "Ambiguous" cells identified in the combined classification mostly mirror the gene set expression scores for normal cells with maybe a small group of cells having more similar gene set scores to tumor cells.
 
@@
 
-Again, most of the "Unknown" tumor cells tend to align with normal cells, with a few grouping in the area where cells transition from tumor to normal.
+Again, most of the "Ambiguous" tumor cells tend to align with normal cells, with a few grouping in the area where cells transition from tumor to normal.
 
@@
 
-In both of these plots, the majority of "Unknown" cells cluster between tumor and normal or with the normal cells.
+In both of these plots, the majority of "Ambiguous" cells cluster between tumor and normal or with the normal cells.
 
@@
 
-In looking at the plots, it looks like there might be some of the "Unknown" cells that should actually be classified as tumor cells. 
+In looking at the plots, it looks like there might be some of the "Ambiguous" cells that should actually be classified as tumor cells. 
 In particular, we expect those that have a gain in Chr8 are tumor cells, so let's just look at Chr8 gains and see if we can find a cut off to use to categorize our "Unknown" cells.
@@
 ```{r}
-#density plot of chr8 gains
+# density plot of chr8 gains
 ggplot(classification_df, aes(x = proportion_scaled_dupli_chr8, fill = combined_cnv_classification)) +

### 2024-06-14 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd

@@
 ```{r}
+# source in helper functions: plot_gene_heatmap() and plot_cnv_heatmap() 
+# create_classification_df() and create_marker_gene_df()
+validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
+source(validation_functions)
+```
+
+```{r}
 # Input files
 sce_file <- file.path(sample_dir, "SCPCL000822_processed.rds")
-marker_genes <- file.path(module_base, "references", "tumor-marker-genes.tsv")
+marker_genes_file <- file.path(module_base, "references", "tumor-marker-genes.tsv")
 
@@
 ```{r}
-# read in all files 
+# read in sce file
 sce <- readr::read_rds(sce_file)
-
-# read in marker genes table 
-marker_genes_df <- readr::read_tsv(marker_genes) |> 
-  # account for genes being from multiple sources
-  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
-  dplyr::distinct()
-
-marker_gene_results_df <- readr::read_tsv(marker_gene_results_file)
-
-# copykat output
-copykat_results_df <- readr::read_tsv(copykat_predictions_file) |> 
-  # only select no_ref classification and rename to be consistent with naming of other classifications
-  dplyr::select(barcodes, copykat_classification = no_ref) |> 
-  dplyr::mutate(
-    copykat_classification = dplyr::case_when(
-      copykat_classification == "aneuploid" ~ "Tumor",
-      copykat_classification == "diploid" ~ "Normal",
-      .default = copykat_classification
-    )
-  )
-
-# infercnv output 
-infercnv_results_df <- readr::read_tsv(infercnv_predictions_file)
-infercnv_metadata_df <- read.delim(infercnv_metadata_file) |> 
-  tibble::rownames_to_column("barcodes")
-
-# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
-chr_columns_to_keep <- which(!stringr::str_detect(names(infercnv_metadata_df), "chrMT|chrGL"))
-infercnv_metadata_df <- infercnv_metadata_df[, chr_columns_to_keep]
-
-# gene set scores 
-geneset_scores_df <- readr::read_tsv(geneset_scores_file)
 ```
@@
 ```{r}
-# grab all coldata and add in all classifications and results to use for plotting throughout notebook 
-# get list of marker genes to use 
-marker_genes <- marker_genes_df |> 
-  dplyr::filter(cell_type == "tumor") |> 
-  dplyr::pull(ensembl_gene_id)
-
-# get the gene expression counts for all marker genes
-marker_gene_exp <- logcounts(sce[marker_genes, ]) |>
-  as.matrix() |>
-  t() |>
-  as.data.frame() |>
-  tibble::rownames_to_column("barcodes")
-
-# pull out the UMAP coordinates, add in marker genes and classifications 
-# and make a data frame to use for plotting
-classification_df <- sce |> 
-  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
-  # replace UMAP.1 with UMAP1
-  dplyr::rename_with(
-        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
-      ) |> 
-  # add in marker gene expression to dataframe 
-  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
-  # combine all genes into a single column for easy faceting
-  tidyr::pivot_longer(
-    cols = starts_with("ENSG"),
-    names_to = "ensembl_gene_id",
-    values_to = "gene_expression"
-  ) |> 
-  # join with marker gene df to get gene symbols for plotting 
-  dplyr::left_join(marker_genes_df, by = "ensembl_gene_id") |>
-  # get rid of excess columns
-  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression) |> 
-  # add in marker gene classifications
-  dplyr::left_join(marker_gene_results_df, by = "barcodes") |> 
-  # add in copykat results 
-  dplyr::left_join(copykat_results_df, by = "barcodes") |> 
-  # add in infercnv results and metadata
-  dplyr::left_join(infercnv_results_df, by = "barcodes") |> 
-  dplyr::left_join(infercnv_metadata_df, by = "barcodes") |> 
-  # gene set scores 
-  dplyr::left_join(geneset_scores_df, by = "barcodes")
-
-```
-
-## Define functions
-
-```{r}
-# create a heatmap where rows are marker genes or gene sets and columns are cells 
-# this function adds a column annotation since cells are columns 
-plot_gene_heatmap <- function(df, 
-                              row_title = "",
-                              legend_title = "", 
-                              annotation = NULL){
-
-  # plot heatmap of marker genes 
-  heatmap <- ComplexHeatmap::Heatmap(
-    df,
-    # set the color scale based on min and max values
-    col = circlize::colorRamp2(seq(min(df), max(df), length = 2), colors = c("white", "#00274C")),
-    border = TRUE,
-    ## Row parameters
-    cluster_rows = TRUE,
-    row_title = row_title,
-    row_title_gp = grid::gpar(fontsize = 12),
-    row_title_side = "left",
-    row_names_side = "left",
-    row_dend_side = "right",
-    row_names_gp = grid::gpar(fontsize = 10),
-    ## Column parameters
-    cluster_columns = TRUE,
-    show_column_names = FALSE,
-    bottom_annotation = annotation,
-    heatmap_legend_param = list(
-      title = legend_title
-    )
+# generate classification df 
+classification_df <- create_classification_df(
+  sce,
+  marker_gene_results_file,
+  copykat_predictions_file, 
+  infercnv_predictions_file,
+  infercnv_metadata_file,
+  geneset_scores_file
 )
 
-  return(heatmap)
-}
-
-# create a heatmap where rows are cells and columns are chromosomes 
-# this function adds a row annotation since cells are rows 
-plot_cnv_heatmap <- function(df,
-                             cnv_col, 
-                             annotation = NULL,
-                             legend_title = ""){
-
-  # barcode by chromosome df 
-  cnv_df <- df |> 
-    dplyr::select(barcodes, starts_with(cnv_col)) |> 
-    unique() |> 
-    tibble::column_to_rownames("barcodes")
-
-  # just get the name of the chromosome as column name 
-  colnames(cnv_df) <- stringr::word(colnames(cnv_df), -1, sep = "_")
-
-  # do a little bit of column reordering before converting to a matrix
-  cnv_df <- cnv_df |> 
-    dplyr::relocate(chr10, .before = chr11) |> 
-    dplyr::relocate(chr19, .before = chr20) |> 
-    dplyr::relocate(chr21, .before = chr22) |> 
-    as.matrix()
-
-  heatmap <- ComplexHeatmap::Heatmap(
-    cnv_df,
-    col = circlize::colorRamp2(seq(min(cnv_df), max(cnv_df), length = 2), colors = c("white", "#00274C")),
-    border = TRUE,
-    ## Row parameters
-    cluster_rows = TRUE,
-    row_title = "",
-    row_title_gp = grid::gpar(fontsize = 12),
-    row_title_side = "left",
-    show_row_names = FALSE,
-    row_dend_side = "right",
-    row_names_gp = grid::gpar(fontsize = 10),
-    ## Column parameters
-    cluster_columns = FALSE,
-    show_column_names = TRUE,
-    right_annotation = annotation,
-    heatmap_legend_param = list(
-      title = legend_title
+# generate marker genes df
+plot_markers_df <- create_marker_gene_df(
+  classification_df, 
+  marker_genes_file
 )
-  )
-  return(heatmap)
-}
-
 ```
@@
 # calculate sum of the scaled gene expression values for plotting 
-marker_density_df <- classification_df |> 
-  dplyr::select(barcodes, gene_symbol, gene_expression, copykat_classification, cnv_proportion_classification, marker_gene_classification, combined_cnv_classification) |> 
-  unique() |> 
-  dplyr::group_by(gene_symbol) |>
-  # get z-scores for each gene
-  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |> 
-  dplyr::ungroup() |> 
-  dplyr::group_by(barcodes) |> 
-  # add all the scores together 
-  dplyr::mutate(sum_exp = sum(transformed_gene_expression)) |>
-  # expand classifications column for easy faceting 
-  dplyr::ungroup() |> 
+marker_density_df <- plot_markers_df |> 
   tidyr::pivot_longer(
@@
 # create matrix with marker genes as rows and barcodes as columns
-marker_gene_heatmap <- classification_df |> 
+marker_gene_heatmap <- plot_markers_df |> 
   dplyr::select(gene_expression, gene_symbol, barcodes) |> 
@@
 geneset_plot_df <- classification_df |> 
-  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean")) |> 
+  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean-")) |> 
   tidyr::pivot_longer(

### 2024-06-17 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
Use cxds threshold of 0.5 and use scale_color_manual instead of scale_color_identity

@@
 By contrast, both `scrublet` and `scDblFinder` identify a threshold based on the given dataset they are processing. 
-This notebook used a doublet threshold of `>=0.75` for `cxds`, which may not be universally suitable (though choosing a universally suitable threshold is not easy either!).
+This notebook used a doublet threshold of `>=0.5` for `cxds`, which may not be universally suitable (though choosing a universally suitable threshold is not easy either!).
 
@@
 theme_set(theme_bw())
+
+# define threshold used to call cxds
+cxds_threshold <- 0.5
 ```
@@
 
-plot_pca_metrics <- function(df, color_column, false_colors) {
+plot_pca_metrics <- function(df, color_column, metric_colors) {
   # Plot PCs colored by performance metric, showing false calls on top
-  # false_colors is a vector of colors used for fn and fp points
+  # metric_colors is a named vector of colors used for coloring tp/tn/fp/fn
   # df is expected to contain columns PC1, PC2, and `color_column`. This should _not_ be provided as a string.
@@
   geom_point(
-    data = dplyr::filter(df, {{color_column}} %in% false_colors), 
+    data = dplyr::filter(df, {{color_column}} %in% c("fp", "fn")), 
     size = 0.75
   ) +
-  scale_color_identity() +
-  theme(legend.position = "none")
+  scale_color_manual(name = "Call type", values = metric_colors) +
+  theme(legend.position = "bottom")
 }
@@
         ) |>
-        # add cxds calls at 0.75 threshold
+        # add cxds calls at `cxds_threshold` threshold
         dplyr::mutate(
           cxds_prediction = dplyr::if_else(
-            cxds_score >= 0.75,
+            cxds_score >= cxds_threshold,
             "doublet",
@@
       sce <- readr::read_rds(sce_file)
-      sce_df <- scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
+      dataset_df <- scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
         tibble::rownames_to_column(var = "barcodes") |>
@@
             consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
-          ), 
-          # set associated plotting colors
-          call_type_color = dplyr::case_when(
-            call_type == "tp" ~ unname(confusion_colors["tp"]),
-            call_type == "tn" ~ unname(confusion_colors["tn"]),
-            call_type == "fp" ~ unname(confusion_colors["fp"]),
-            call_type == "fn" ~ unname(confusion_colors["fn"])
           )
@@
 ```{r, fig.width = 12, fig.height = 6}
-# Make a legend for the confusion-colored PCA
-legend_plot <- data.frame(
-  x = factor(names(confusion_colors), levels = names(confusion_colors)), y = 1:4
-) |>
- ggplot(aes(x = x, y = y, color = x)) + 
-  geom_point(size = 3) + 
-  scale_color_manual(name = "Metric", values = confusion_colors) 
-confusion_legend <- ggpubr::get_legend(legend_plot) |> ggpubr::as_ggplot()
-
 doublet_df_list |>
@@
         df,
-        call_type_color,
-        false_colors = unname(c(confusion_colors["fn"], confusion_colors["fp"]))
+        call_type, 
+        metric_colors = confusion_colors
       )
@@
       # combine and plot
-      plot( p1 + p2 + p3 + confusion_legend + plot_annotation(glue::glue("PCA for {dataset}")) + plot_layout(ncol=4, widths = c(1,1,1,0.25)) )
+      plot( p1 + p2 + p3 + plot_annotation(glue::glue("PCA for {dataset}")) + plot_layout(ncol=4, widths = c(1,1,1)) )
     }

### 2024-06-17 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
remove pull_barcodes helper function

@@
 ```{r}
-pull_barcodes <- function(df, pred_var) {
-  # Helper function to pull out barcodes for doublet calls
-  df$barcodes[df[[pred_var]] == "doublet"]
-}
-
 upset_list <- doublet_df_list |>
@@
       doublet_barcodes <- list(
-        "scDblFinder" = pull_barcodes(df, "scdbl_prediction"),
-        "scrublet"    = pull_barcodes(df, "scrublet_prediction"),
-        "cxds"        = pull_barcodes(df, "cxds_prediction")
+        "scDblFinder" = df$barcodes[df$scdbl_prediction == "doublet"],
+        "scrublet"    = df$barcodes[df$scrublet_prediction == "doublet"],
+        "cxds"        = df$barcodes[df$cxds_prediction == "doublet"]
       )

### 2024-06-17 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_tumor-cell-validation.Rmd
use saved functions

--- 2024-06-12 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_tumor-cell-validation.Rmd
@@
 ```{r}
+# source in helper functions: plot_gene_heatmap() and plot_cnv_heatmap() 
+# create_classification_df() and create_marker_gene_df()
+validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
+source(validation_functions)
+
+# source in helper functions: make_jaccard_matrix() and jaccard() 
+jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
+source(jaccard_functions)
+```
+
+```{r}
 # Input files
 sce_file <- file.path(sample_dir, "SCPCL000824_processed.rds")
-marker_genes <- file.path(module_base, "references", "tumor-marker-genes.tsv")
+marker_genes_file <- file.path(module_base, "references", "tumor-marker-genes.tsv")
 
@@
 ```{r}
-# read in all files 
+# read in sce file 
 sce <- readr::read_rds(sce_file)
-
-# read in marker genes table 
-marker_genes_df <- readr::read_tsv(marker_genes) |> 
-  # account for genes being from multiple sources
-  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
-  dplyr::distinct()
-
-marker_gene_results_df <- readr::read_tsv(marker_gene_results_file)
-
-# copykat output
-copykat_results_df <- readr::read_tsv(copykat_predictions_file) |> 
-  # only select no_ref classification and rename to be consistent with naming of other classifications
-  dplyr::select(barcodes, copykat_classification = no_ref) |> 
-  dplyr::mutate(
-    copykat_classification = dplyr::case_when(
-      copykat_classification == "aneuploid" ~ "Tumor",
-      copykat_classification == "diploid" ~ "Normal",
-      TRUE ~ copykat_classification
-    )
-  )
-
-# infercnv output 
-infercnv_results_df <- readr::read_tsv(infercnv_predictions_file)
-infercnv_metadata_df <- read.delim(infercnv_metadata_file) |> 
-  tibble::rownames_to_column("barcodes")
-
-# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
-chr_columns_to_keep <- which(!stringr::str_detect(names(infercnv_metadata_df), "chrMT|chrGL"))
-infercnv_metadata_df <- infercnv_metadata_df[, chr_columns_to_keep]
-
-# gene set scores 
-geneset_scores_df <- readr::read_tsv(geneset_scores_file)
 ```
@@
 ```{r}
-# grab all coldata and add in all classifications and results to use for plotting throughout notebook 
-# get list of marker genes to use 
-marker_genes <- marker_genes_df |> 
-  dplyr::filter(cell_type == "tumor") |> 
-  dplyr::pull(ensembl_gene_id)
-
-# get the gene expression counts for all marker genes
-marker_gene_exp <- logcounts(sce[marker_genes, ]) |>
-  as.matrix() |>
-  t() |>
-  as.data.frame() |>
-  tibble::rownames_to_column("barcodes")
-
-# pull out the UMAP coordinates and make a data frame to use for plotting
-classification_df <- sce |> 
-  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
-  # replace UMAP.1 with UMAP1
-  dplyr::rename_with(
-        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
-      ) |> 
-  # add in marker gene expression to dataframe 
-  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
-  # combine all genes into a single column for easy faceting
-  tidyr::pivot_longer(
-    cols = starts_with("ENSG"),
-    names_to = "ensembl_gene_id",
-    values_to = "gene_expression"
-  ) |> 
-  # join with marker gene df to get gene symbols for plotting 
-  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
-  # get rid of excess columns
-  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression) |> 
-  # add in marker gene classifications
-  dplyr::left_join(marker_gene_results_df) |> 
-  # add in copykat results 
-  dplyr::left_join(copykat_results_df) |> 
-  # add in infercnv results and metadata
-  dplyr::left_join(infercnv_results_df) |> 
-  dplyr::left_join(infercnv_metadata_df) |> 
-  # gene set scores 
-  dplyr::left_join(geneset_scores_df)
+# generate classification df 
+classification_df <- create_classification_df(
+  sce,
+  marker_gene_results_file,
+  copykat_predictions_file, 
+  infercnv_predictions_file,
+  infercnv_metadata_file,
+  geneset_scores_file
+)
 ```
@@
 ```
+
+```{r}
+# plot tumor cells for each classification method 
+geneset_plot_df <- classification_df |> 
+  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("scaled_mean")) |> 
+  tidyr::pivot_longer(
+    cols = starts_with("scaled_mean"),
+    names_to = "geneset",
+    values_to = "mean_score"
+  ) |> 
+  dplyr::mutate(
+    geneset = stringr::word(geneset, -1, sep = "-")
+  )
+
+ggplot(geneset_plot_df, aes(x = UMAP1, UMAP2, color = mean_score)) +
+  geom_point(size = 0.5, alpha = 0.5) +
+  facet_wrap(vars(geneset)) +
+  scale_color_viridis_c()
+```
+
+Add geneset classification
+
+```{r}
+# any cell that has > 0 expression of at least one geneset 
+classification_df <- classification_df |>
+  dplyr::mutate(
+    geneset_classification = dplyr::if_else(
+      `scaled_mean-RIGGI` > 0 & `scaled_mean-ZHANG` > 0 & `scaled_mean-SILIGAN` > 0,
+      "Tumor",
+      "Normal"
+    ))
+```
+
+Compare gene set classification to marker gene classification 
+```{r}
+classification_columns <- c(
+  "copykat_classification",
+  "cnv_proportion_classification",
+  "geneset_classification"
+)
+
+# create jaccard matrices for SingleR and CellAssign compared to tumor/normal 
+jaccard_matrices <- classification_columns |>
+  purrr::map(\(name) {
+    make_jaccard_matrix(
+      classification_df,
+      "marker_gene_classification",
+      name
+    )
+  }) |> 
+  purrr::set_names("CopyKAT", "InferCNV", "Gene set")
+```
+
+```{r}
+# Set heatmap padding option
+heatmap_padding <- 0.2
+ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))
+
+# list of heatmaps looking at SingleR/ CellAssign vs tumor/normal 
+heatmap <- jaccard_matrices |>
+  purrr::imap(
+    \(classification_mat, classification_method) {
+      ComplexHeatmap::Heatmap(
+        t(classification_mat), # transpose because matrix rows are in common & we want a vertical arrangement
+        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
+        border = TRUE,
+        ## Row parameters
+        cluster_rows = TRUE,
+        row_title = classification_method,
+        row_title_gp = grid::gpar(fontsize = 12),
+        row_title_side = "left",
+        row_names_side = "left",
+        row_dend_side = "right",
+        row_names_gp = grid::gpar(fontsize = 10),
+        ## Column parameters
+        cluster_columns = FALSE,
+        column_title = "Marker genes",
+        column_title_gp = grid::gpar(fontsize = 12),
+        column_names_side = "bottom",
+        column_names_gp = grid::gpar(fontsize = 10),
+        column_names_rot = 90,
+        ## Legend parameters
+        heatmap_legend_param = list(
+          title = "Jaccard index",
+          direction = "vertical",
+          legend_width = unit(1.5, "in")
+        ),
+        show_heatmap_legend = classification_method == "CopyKAT",
+      )
+    }) |>
+  # concatenate vertically into HeatmapList object
+  purrr::reduce(ComplexHeatmap::`%v%`) |>
+  ComplexHeatmap::draw(
+    heatmap_legend_side = "right",
+    # add a margin to the heatmap so labels don't get cut off
+    padding = unit(c(2, 20, 2, 2), "mm")
+  )
+```
+
+Add in classifications for geneset vs. CNV
+
+```{r}
+classification_df <- classification_df |>
+  dplyr::mutate(
+    exp_classification = dplyr::case_when(
+      marker_gene_classification == "Normal" & geneset_classification == "Normal" ~ "Normal",
+      marker_gene_classification == "Tumor" & geneset_classification == "Tumor" ~ "Tumor",
+      .default = "Ambiguous"
+    ), 
+    all_cnv_classification = dplyr::case_when(
+      copykat_classification == "Normal" & cnv_proportion_classification == "Normal" ~ "Normal",
+      copykat_classification == "Normal" & cnv_proportion_classification == "Reference"~ "Normal",
+      copykat_classification == "Tumor" & cnv_proportion_classification == "Tumor" ~ "Tumor",
+      .default = "Ambiguous"
+    )
+  )
+```
+
+```{r}
+geneset_plot_df |> 
+  tidyr::pivot_longer(
+    cols = ends_with("classification"),
+    names_to = "method",
+    values_to = "classification"
+  ) |> 
+  ggplot(aes(x = mean_score, color = classification)) +
+  geom_density(bw = 0.05) +
+  facet_grid(rows = vars(method), 
+             cols = vars(geneset))
+```
+
+```{r}
+# generate marker genes df
+plot_markers_df <- create_marker_gene_df(
+  classification_df, 
+  marker_genes_file
+)
+
+# calculate sum of the scaled gene expression values for plotting 
+marker_density_df <- plot_markers_df |>
+  tidyr::pivot_longer(
+    cols = ends_with("classification"),
+    names_to = "method",
+    values_to = "classification"
+  ) |> 
+  dplyr::mutate(
+    method = dplyr::case_when(
+      method == "marker_gene_classification" ~ "Marker genes only",
+      method == "copykat_classification" ~ "CopyKAT - no reference",
+      method == "cnv_proportion_classification" ~ "InferCNV proportion",
+      method == "exp_classification" ~ "Expression combined",
+      method == "all_cnv_classification" ~ "CopyKAT and InferCNV",
+      method == "geneset_classification" ~ "EWS-FLI1 Gene set"
+    )
+  )
+
+ggplot(marker_density_df, aes(x = sum_exp, color = classification)) +
+  geom_density() +
+  facet_wrap(vars(method))
+```
+
+```{r}
+ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = geneset_classification)) +
+  geom_point(size = 0.5, alpha = 0.5)
+```
+
+CNV heatmap annotated by copykat and marker gene prediction, two annotation columns
+
+```{r}
+annotation_df <- classification_df |> 
+  dplyr::select(barcodes, ends_with("classification")) |> 
+  unique()
+
+# need to make a new annotation, since we want to annotate the rows here 
+annotation <- ComplexHeatmap::rowAnnotation(exp_classification = annotation_df$exp_classification,
+                                            all_cnv_classification = annotation_df$all_cnv_classification)
+
+# plot gains in cnv
+plot_cnv_heatmap(classification_df,
+                 cnv_col = "proportion_scaled_dupli",
+                 annotation = annotation,
+                 legend_title = "Copy number \ngain")
+```
+
+```{r}
+# plot gains in cnv
+plot_cnv_heatmap(classification_df,
+                 cnv_col = "proportion_scaled_loss",
+                 annotation = annotation,
+                 legend_title = "Copy number \nloss")
+```
+
+Density plot with mean proportion CNV from InferCNV on x-axis, colored by prediction, faceted by classification method
+
+```{r}
+classification_df |> 
+  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), mean_cnv_proportion) |> 
+  tidyr::pivot_longer(
+    cols = ends_with("classification"),
+    names_to = "method",
+    values_to = "classification"
+  ) |> 
+  ggplot(aes(x = mean_cnv_proportion, color = classification)) +
+  geom_density(bw = 0.05) +
+  facet_grid(rows = vars(method))
+```
+
+Mean InferCNV proportion on x-axis, mean copyKAT proportion on y-axis, color by scaled sum of marker gene expression
+
+```{r}
+cnv_prop_df <- classification_df |> 
+  dplyr::select(barcodes, ends_with("classification"), starts_with("scaled_mean"), mean_cnv_proportion, mean_cnv_detection) |> 
+  tidyr::pivot_longer(
+    cols = starts_with("scaled_mean"),
+    names_to = "geneset",
+    values_to = "mean_score"
+  ) |> 
+  dplyr::mutate(
+    geneset = stringr::word(geneset, -1, sep = "-")
+  )
+
+ggplot(cnv_prop_df, aes(x = mean_cnv_proportion, y = abs(mean_cnv_detection), color = mean_score)) +
+  geom_point(size = 0.5, alpha = 0.5) +
+  facet_wrap(vars(geneset)) +
+  scale_color_viridis_c()
+```
+
+```{r}
+ggplot(cnv_prop_df, aes(x = mean_score, y = abs(mean_cnv_detection))) +
+  geom_point(size = 0.5, alpha = 0.5) +
+  facet_wrap(vars(geneset)) +
+  scale_color_viridis_c()
+```
+
+faceted density plot with marker gene expression on x axis, colored by copykat and infercnv prediction and faceted by marker gene
+
+```{r}
+geneset_scores_df |> 
+  dplyr::pivot
+
+ggplot(geneset_scores_df, aes(x = mean_score, color = ))
+```
+
+use SingleR to classify as tumor/normal using SCPCS000490
+
+correlation of geneset score and marker gene score 
+```{r}
+geneset_plot_df |>
+
+  dplyr::left_join(marker_density_df) |> 
+  ggplot(aes(x = sum_exp, y = mean_score)) +
+    facet_wrap()
+```

### 2024-06-17 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
WIP: Updating PCAs

@@
                            pred_column,
-                           color_lab) {
+                           title) {
   # Plot PCs colored by singlet/doublet, showing doublets on top
@@
   ) +
-  scale_color_manual(name = color_lab, values = c("black", "lightblue")) + 
+  scale_color_manual(name = "", values = c("black", "gray80")) + 
   geom_point(
@@
   ) +
+  ggtitle(title) +
   theme(
@@
 
+MORE MORE MORE MORE MORE FILL IN THIS TEXT!!!
+
 This section plots the PCA for each dataset, with three color schemes from left to right:
@@
 
-      # First, ground truth
+      # Top row: methods themselves ---
       p1 <- plot_pca_calls(
         df, 
+        color_column = scdbl_prediction, 
+        title = "scDblFinder prediction"
+      )
+
+      p2 <- plot_pca_calls(
+        df, 
+        color_column = scrublet_prediction, 
+        title = "scrublet prediction"
+      )
+
+      p3 <- plot_pca_calls(
+        df, 
+        color_column = cxds_prediction, 
+        color_lab = "cxds prediction"
+      )
+
+      # Bottom row: consensus  --
+      p4 <- plot_pca_calls(
+        df, 
         color_column = ground_truth, 
-        color_lab = "Ground truth"
+        title = "Ground truth"
       )
@@
       # Second, consensus call
-      p2 <- plot_pca_calls(
+      p5 <- plot_pca_calls(
         df, 
         color_column = consensus_call, 
-        color_lab = "Consensus call"
+        title = "Consensus call"
       )
@@
       # Third, call type
-      p3 <- plot_pca_metrics(
+      p6 <- plot_pca_metrics(
         df,
@@
       # combine and plot
-      plot( p1 + p2 + p3 + plot_annotation(glue::glue("PCA for {dataset}")) + plot_layout(ncol=4, widths = c(1,1,1)) )
+      plot( 
+        (p1 + p2 + p3) / (p4 + p5 + p6) + 
+          plot_annotation(glue::glue("PCA for {dataset}")) + 
+          plot_layout(guides = "collect"))
     }

### 2024-06-17 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
hide code and fix variable name typo in notebook

@@
     toc_depth: 3
+    code_folding: hide
 ---
@@
 # export final TSV with annotations 
-final_classifications <- new_classification_df |> 
+final_classifications <- classification_df |> 
   dplyr::select(

### 2024-06-17 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
add more PCAs, per review

@@
   ) +
-  ggtitle(title) +
-  theme(
-    legend.title.position = "top",
-    legend.position = "bottom"
-  )
+  ggtitle(title) 
 }
@@
   scale_color_manual(name = "Call type", values = metric_colors) +
-  theme(legend.position = "bottom")
+  ggtitle("Consensus call metrics")
 }
@@
 
-MORE MORE MORE MORE MORE FILL IN THIS TEXT!!!
+This section plots the PCA for each dataset, clockwise from the top left:
 
-This section plots the PCA for each dataset, with three color schemes from left to right:
+1. `scDblFinder` singlet/doublet calls. 
+2. `scrublet` singlet/doublet calls
+3. `cxds` singlet/doublet calls 
+4. Ground truth single/doublets
+5. Consensus singlet/doublet calls
+6. Points are colored based on comparing the consensus call to the ground truth as one of:
+    - true positive (`tp`), true negative (`tn`), false positive (`fp`), false negative (`fn`)
 
-1. Ground truth doublets are shown in black and singlets in blue
-2. Consensus doublets are shown in black and all remaining droplets in blue
-3. Points are colored based on comparing the consensus call to the ground truth as one of:
-  - true positive (`tp`), true negative (`tn`), false positive (`fp`), false negative (`fn`)
+For the first five panels, doublets are shown in black and singlets in light gray.
 
@@
         color_column = cxds_prediction, 
-        color_lab = "cxds prediction"
+        title = "cxds prediction"
       )
@@
         (p1 + p2 + p3) / (p4 + p5 + p6) + 
-          plot_annotation(glue::glue("PCA for {dataset}")) + 
+          plot_annotation(
+            glue::glue("PCA for {dataset}"), 
+            theme = theme(plot.title = element_text(size = 16))) + 
           plot_layout(guides = "collect"))

### 2024-06-17 analyses/cell-type-ewings/template_notebooks/03-copykat.Rmd

@@
 })
-
-knitr::opts_chunk$set(
-  dev = "jpeg"
-)
 ```
@@
 # output classifications file 
-copykat_results_file <- file.path(params$results_dir, params$sample_id, glue::glue("{params$library_id}_copykat-classifications.tsv"))
+copykat_results_file <- file.path(params$results_dir, glue::glue("{params$library_id}_copykat-classifications.tsv"))
 ```

### 2024-06-17 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd

--- 2024-05-29 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd
@@
     cnv_proportion_classification = cnv_mean_classification,
-    cnv_sum_classification = cnv_classification
+    cnv_sum_classification = cnv_classification,
+    mean_cnv_proportion = mean_proportion
   )

### 2024-06-17 analyses/cell-type-ewings/template_notebooks/03-copykat.Rmd
Revert "Merge branch 'main' into sjspielman/517-module-gha"
This reverts commit b5e9160f58f517604d962dfc92b2606416e3c49b, reversing
changes made to 1f4cd82978d8ca81c36cf5fd3a97ea4b60792526.

@@
 
-all_info_df <- cnv_df |> 
-  dplyr::select(barcodes,
-                mean_cnv_detection) |> 
-  dplyr::group_by(barcodes) |>
-  dplyr::mutate(mean_cnv_detection = mean(mean_cnv_detection)) |> 
-  dplyr::distinct() |> 
-  dplyr::left_join(celltype_df)
-
-readr::write_tsv(all_info_df, copykat_results_file)
+readr::write_tsv(celltype_df, copykat_results_file)
 ```

### 2024-06-17 analyses/cell-type-ewings/scripts/utils/tumor-validation-helpers.R

@@ -1,199 +1,0 @@
-# These are functions used to prepare data frames for plotting when doing tumor cell validataion 
-
-# Data frame prep --------------------------------------------------------------
-
-create_classification_df <- function(
-    sce,
-    marker_gene_results_file, # path to marker gene classifications
-    copykat_predictions_file, # path to copykat predictions
-    infercnv_predictions_file, # path to infercnv predictions
-    infercnv_metadata_file, # path to full infercnv metadata
-    geneset_scores_file # path to gene set scores results
-){
- 
-  # read in marker gene results 
-  marker_gene_results_df <- readr::read_tsv(marker_gene_results_file, show_col_types = FALSE)
-  
-  # copykat output
-  copykat_results_df <- readr::read_tsv(copykat_predictions_file, show_col_types = FALSE) |> 
-    # only select no_ref classification and rename to be consistent with naming of other classifications
-    dplyr::select(barcodes, copykat_classification = no_ref, mean_cnv_detection) |> 
-    dplyr::distinct() |> 
-    dplyr::mutate(
-      copykat_classification = dplyr::case_when(
-        copykat_classification == "aneuploid" ~ "Tumor",
-        copykat_classification == "diploid" ~ "Normal",
-        .default = copykat_classification
-      )
-    )
-  
-  # infercnv output 
-  infercnv_results_df <- readr::read_tsv(infercnv_predictions_file, show_col_types = FALSE)
-  # use read.delim to ensure row names do not get read in as a column 
-  infercnv_metadata_df <- read.delim(infercnv_metadata_file) |> 
-    tibble::rownames_to_column("barcodes")
-  
-  # remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
-  chr_columns_to_keep <- which(!stringr::str_detect(names(infercnv_metadata_df), "chrMT|chrGL"))
-  infercnv_metadata_df <- infercnv_metadata_df[, chr_columns_to_keep]
-  
-  # gene set scores 
-  geneset_scores_df <- readr::read_tsv(geneset_scores_file, show_col_types = FALSE)
-  
-  # pull out the UMAP coordinates and make a data frame to use for plotting
-  classification_df <- sce |> 
-    scuttle::makePerCellDF(use.dimred = "UMAP") |> 
-    # replace UMAP.1 with UMAP1
[…]

### 2024-06-18 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_tumor-cell-validation.Rmd
add in SingleR validation and comments throughout

--- 2024-06-12 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_tumor-cell-validation.Rmd
@@
 ---
-title: "SCPCL000824 - Validation of tumor cells"
+title: "`SCPCL000824` - Validation of tumor cells"
 author: Ally Hawkins
@@
 - First we look at the cells that are annotated as tumor cells by each method. 
-Using this information, we call any cell that is annotated as a tumor cell by both `CopyKAT` and `InferCNV` as a tumor cell. 
-- We then validate that these are in fact tumor cells by doing the following: 
-  - Looking at marker gene expression. 
-  - Looking at expression of EWS-FLI1 target gene sets. 
-  - Confirming presence of known CNVs in Ewing sarcoma
+- For each classification method we look at the expression of marker genes and gene set scores. 
+- We then use the marker gene expression and gene set scores to classify cells as tumor, normal, or ambiguous.
 
@@
 final_annotations_file <- file.path(final_annotations_dir, "SCPCL000824_tumor-classifications.tsv")
+
+# reference files to use with SingleR
+ref_sce_file <- file.path(data_dir, "SCPCP000015", "SCPCS000490", "SCPCL000822_processed.rds")
+ref_labels_file <- file.path(module_base, "cell_type_annotations", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv")
 ```
@@
 ```{r}
-# read in all files 
+# read in sce file 
 sce <- readr::read_rds(sce_file)
 
-# read in marker genes table 
-marker_genes_df <- readr::read_tsv(marker_genes) |> 
-  # account for genes being from multiple sources
-  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
-  dplyr::distinct()
-
-marker_gene_results_df <- readr::read_tsv(marker_gene_results_file)
-
-# copykat output
-copykat_results_df <- readr::read_tsv(copykat_predictions_file) |> 
-  # only select no_ref classification and rename to be consistent with naming of other classifications
-  dplyr::select(barcodes, copykat_classification = no_ref) |> 
-  dplyr::mutate(
-    copykat_classification = dplyr::case_when(
-      copykat_classification == "aneuploid" ~ "Tumor",
-      copykat_classification == "diploid" ~ "Normal",
-      TRUE ~ copykat_classification
-    )
-  )
-
-# infercnv output 
-infercnv_results_df <- readr::read_tsv(infercnv_predictions_file)
-infercnv_metadata_df <- read.delim(infercnv_metadata_file) |> 
-  tibble::rownames_to_column("barcodes")
-
-# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
-chr_columns_to_keep <- which(!stringr::str_detect(names(infercnv_metadata_df), "chrMT|chrGL"))
-infercnv_metadata_df <- infercnv_metadata_df[, chr_columns_to_keep]
-
-# gene set scores 
-geneset_scores_df <- readr::read_tsv(geneset_scores_file)
+# read in ref sce and ref annotations for SingleR later 
+ref_sce <- readr::read_rds(ref_sce_file)
+ref_labels_df <- readr::read_tsv(ref_labels_file)
 ```
@@
 ```{r}
-# grab all coldata and add in all classifications and results to use for plotting throughout notebook 
-# get list of marker genes to use 
-marker_genes <- marker_genes_df |> 
-  dplyr::filter(cell_type == "tumor") |> 
-  dplyr::pull(ensembl_gene_id)
-
-# get the gene expression counts for all marker genes
-marker_gene_exp <- logcounts(sce[marker_genes, ]) |>
-  as.matrix() |>
-  t() |>
-  as.data.frame() |>
-  tibble::rownames_to_column("barcodes")
-
-# pull out the UMAP coordinates and make a data frame to use for plotting
-classification_df <- sce |> 
-  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
-  # replace UMAP.1 with UMAP1
-  dplyr::rename_with(
-        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
+# generate classification df 
+classification_df <- create_classification_df(
+  sce,
+  marker_gene_results_file,
+  copykat_predictions_file, 
+  infercnv_predictions_file,
+  infercnv_metadata_file,
+  geneset_scores_file
 ) |> 
-  # add in marker gene expression to dataframe 
-  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
-  # combine all genes into a single column for easy faceting
-  tidyr::pivot_longer(
-    cols = starts_with("ENSG"),
-    names_to = "ensembl_gene_id",
-    values_to = "gene_expression"
-  ) |> 
-  # join with marker gene df to get gene symbols for plotting 
-  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
-  # get rid of excess columns
-  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression) |> 
-  # add in marker gene classifications
-  dplyr::left_join(marker_gene_results_df) |> 
-  # add in copykat results 
-  dplyr::left_join(copykat_results_df) |> 
-  # add in infercnv results and metadata
-  dplyr::left_join(infercnv_results_df) |> 
-  dplyr::left_join(infercnv_metadata_df) |> 
-  # gene set scores 
-  dplyr::left_join(geneset_scores_df)
+  # remove extra infercnv classification column we won't use 
+  dplyr::select(- cnv_sum_classification)
 ```
 
-## Identify tumor cell population
+## Evaluate tumor cell classifications
 
@@
 # plot tumor cells for each classification method 
-tumor_umap_df <- coldata_df |> 
-  dplyr::select(barcodes, UMAP1, UMAP2, marker_gene_classification, copykat_classification, cnv_proportion_classification) |> 
+tumor_umap_df <- classification_df |> 
+  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification")) |> 
   tidyr::pivot_longer(
@@
 ```
+
+It looks like we don't have a lot of agreement between the marker gene based classification and copy number inference classification. 
+Let's compare how many cells are overlapping between each CNV method and marker gene classifications using the Jaccard index.
+
+```{r}
+classification_columns <- c(
+  "copykat_classification",
+  "cnv_proportion_classification"
+)
+
+# create jaccard matrices for copykat and infercnv comparison to marker gene  
+jaccard_matrices <- classification_columns |>
+  purrr::map(\(name) {
+    make_jaccard_matrix(
+      classification_df,
+      "marker_gene_classification",
+      name
+    )
+  }) |> 
+  purrr::set_names("CopyKAT", "InferCNV")
+```
+
+```{r}
+# Set heatmap padding option
+heatmap_padding <- 0.2
+ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))
+
+# list of heatmaps looking at each cnv method vs marker genes 
+heatmap <- jaccard_matrices |>
+  purrr::imap(
+    \(classification_mat, classification_method) {
+      ComplexHeatmap::Heatmap(
+        t(classification_mat), # transpose because matrix rows are in common & we want a vertical arrangement
+        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
+        border = TRUE,
+        ## Row parameters
+        cluster_rows = TRUE,
+        row_title = classification_method,
+        row_title_gp = grid::gpar(fontsize = 12),
+        row_title_side = "left",
+        row_names_side = "left",
+        row_dend_side = "right",
+        row_names_gp = grid::gpar(fontsize = 10),
+        ## Column parameters
+        cluster_columns = FALSE,
+        column_title = "Marker genes",
+        column_title_gp = grid::gpar(fontsize = 12),
+        column_names_side = "bottom",
+        column_names_gp = grid::gpar(fontsize = 10),
+        column_names_rot = 90,
+        ## Legend parameters
+        heatmap_legend_param = list(
+          title = "Jaccard index",
+          direction = "vertical",
+          legend_width = unit(1.5, "in")
+        ),
+        show_heatmap_legend = classification_method == "CopyKAT",
+      )
+    }) |>
+  # concatenate vertically into HeatmapList object
+  purrr::reduce(ComplexHeatmap::`%v%`) |>
+  ComplexHeatmap::draw(
+    heatmap_legend_side = "right",
+    # add a margin to the heatmap so labels don't get cut off
+    padding = unit(c(2, 20, 2, 2), "mm")
+  )
+```
+As expected based on the UMAPs, there is not a lot of consensus between the CNV methods and marker gene based method.
+
+### Marker gene and gene set expression
+
+We can confirm this by looking at marker gene expression in cells classified as tumor or normal by each method.
+
+```{r}
+# generate marker genes df
+plot_markers_df <- create_marker_gene_df(
+  classification_df, 
+  marker_genes_file
+)
+
+# create a density plot showing the distribution of marker gene expression across classification methods 
+marker_density_df <- plot_markers_df |>
+  tidyr::pivot_longer(
+    cols = ends_with("classification"),
+    names_to = "method",
+    values_to = "classification"
+  ) |> 
+  dplyr::mutate(
+    method = dplyr::case_when(
+      method == "marker_gene_classification" ~ "Marker genes only",
+      method == "copykat_classification" ~ "CopyKAT - no reference",
+      method == "cnv_proportion_classification" ~ "InferCNV proportion"
+    )
+  )
+
+ggplot(marker_density_df, aes(x = sum_exp, color = classification)) +
+  geom_density() +
+  facet_wrap(vars(method))
+```
+
+We see a few things from this plot:
+
+- As expected given that marker gene expression was used to classify cells using the "Marker genes only" method, we see an increase in marker gene expression in tumor cells. 
+- For `InferCNV`, both tumor and normal cells have higher expression of marker genes than the reference. 
+- For `CopyKAT`, the normal cells appear to have higher expression of marker genes.
+
+Let's also look at gene set scores and see if we can use those to help us determine which cells are most likely to be tumor cells.
+
+```{r}
+# plot gene set scores for each cell  
+geneset_plot_df <- classification_df |> 
+  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("scaled_mean")) |> 
+  tidyr::pivot_longer(
+    cols = starts_with("scaled_mean"),
+    names_to = "geneset",
+    values_to = "mean_score"
+  ) |> 
+  dplyr::mutate(
+    geneset = stringr::word(geneset, -1, sep = "-")
+  )
+
+ggplot(geneset_plot_df, aes(x = UMAP1, UMAP2, color = mean_score)) +
+  geom_point(size = 0.5, alpha = 0.5) +
+  facet_wrap(vars(geneset)) +
+  scale_color_viridis_c()
+```
+
+We do see that the scores across all three genesets seem to be pretty consistent. 
+Cells that have higher scores for `RIGGI`, also have higher scores for `SILIGAN` and `ZHANG`.
+
+Let's look at how the distribution of gene set scores coordinates with assignments from the CNV and marker gene methods. 
+Below, we color the distribution based on the classification method used.
+
+```{r}
+geneset_plot_df |> 
+  tidyr::pivot_longer(
+    cols = ends_with("classification"),
+    names_to = "method",
+    values_to = "classification"
+  ) |> 
+  dplyr::mutate(
+    method = dplyr::case_when(
+      method == "marker_gene_classification" ~ "Marker genes only",
+      method == "copykat_classification" ~ "CopyKAT - no reference",
+      method == "cnv_proportion_classification" ~ "InferCNV proportion"
+    )) |> 
+  ggplot(aes(x = mean_score, color = classification)) +
+  geom_density(bw = 0.05) +
+  facet_grid(rows = vars(method), 
+             cols = vars(geneset))
+```
+
+We see a few things from this plot:
+
+- The marker genes classification has the most consistent increase in gene set scores in tumor cells compared to normal cells, although the separation is not super clear it is there. 
+- For the `RIGGI` and `ZHANG` gene sets, the scores are higher in both tumor and normal cells compared to reference cells as classified by `InferCNV`. 
+- For all gene sets, the normal cells have a slightly higher gene set score as classified by `CopyKAT`.
+
+### CNV results
+
+I'm inclined to trust the gene set scores and marker genes over the copy number variations here. 
+Before we use those to classify cells over CNV methods, let's look at the CNVs identified per chromosome and see if any expected CNVs appear (gain in Chr1, Chr8, or Chr12 or loss in Chr16).  
+We will look at the CNVs identified by InferCNV by chromosome.
+
+#### CNV Gains
+
+```{r}
+annotation_df <- classification_df |> 
+  dplyr::select(barcodes, ends_with("classification")) |> 
+  unique()
+
+# include both infercnv and copykat annotations in the cnv heatmap 
+annotation <- ComplexHeatmap::rowAnnotation(infercnv = annotation_df$cnv_proportion_classification,
+                                            copykat = annotation_df$copykat_classification,
+                                            col = list(infercnv = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"),
+                                                       copykat = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
+
+# plot gains in cnv
+plot_cnv_heatmap(classification_df,
+                 cnv_col = "proportion_scaled_dupli",
+                 annotation = annotation,
+                 legend_title = "Copy number \ngain")
+```
+
+#### CNV Losses
+
+```{r}
+# plot gains in cnv
+plot_cnv_heatmap(classification_df,
+                 cnv_col = "proportion_scaled_loss",
+                 annotation = annotation,
+                 legend_title = "Copy number \nloss")
+```
+
+Looking at this we don't see quite the same separation of distinct CNV profiles between tumor and normal that we've seen previously with `SCPCL000822`. 
+As far as known CNVs, we don't really see any gains in Chr1 or chr12. 
+There may be a few cells that have a gain in Chr8, but it is not striking.
+
+Also, the cells being classified as "normal" have more CNVs than I would expect and look very similar to the profiles seen in "tumor" cells here.
+
+## Classifying tumor cells
+
+Given the fact that the marker gene classification and gene set scores tend to line up while copy number information is less clear, let's classify cells using gene expression information (marker genes and gene sets).
+
+First we will classify cells as being tumor or normal for each of the gene sets and compare those annotations to marker gene based annotation. 
+For each gene set we will call tumor cells as cells that have higher than the mean gene set score.
+
+```{r}
+# get mean value for each gene set score, excluding any NAs
+mean_riggi <- mean(classification_df$`scaled_mean-RIGGI`, na.rm=TRUE)
+mean_zhang <- mean(classification_df$`scaled_mean-ZHANG`, na.rm=TRUE)
+mean_siligan <- mean(classification_df$`scaled_mean-SILIGAN`, na.rm=TRUE)
+
+# add classification for each gene set 
+classification_df <- classification_df |> 
+  dplyr::mutate(riggi_classification = dplyr::if_else(`scaled_mean-RIGGI` > mean_riggi, "Tumor", "Normal"),
+                zhang_classification = dplyr::if_else(`scaled_mean-ZHANG` > mean_zhang, "Tumor", "Normal"),
+                siligan_classification = dplyr::if_else(`scaled_mean-SILIGAN` > mean_siligan, "Tumor", "Normal"))
+```
+
+Now we will compare the annotations from marker genes only to using an entire gene set.
+
+```{r}
+classification_columns <- c(
+  "riggi_classification",
+  "zhang_classification",
+  "siligan_classification"
+)
+
+# create jaccard matrices for each gene set compared to marker gene 
+jaccard_matrices <- classification_columns |>
+  purrr::map(\(name) {
+    make_jaccard_matrix(
+      classification_df,
+      "marker_gene_classification",
+      name
+    )
+  }) |> 
+  purrr::set_names("riggi", "zhang", "siligan")
+```
+
+```{r}
+# Set heatmap padding option
+heatmap_padding <- 0.2
+ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))
+
+# list of heatmaps looking at each gene set classification vs. marker genes 
+heatmap <- jaccard_matrices |>
+  purrr::imap(
+    \(classification_mat, classification_method) {
+      ComplexHeatmap::Heatmap(
+        t(classification_mat), # transpose because matrix rows are in common & we want a vertical arrangement
+        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
+        border = TRUE,
+        ## Row parameters
+        cluster_rows = TRUE,
+        row_title = classification_method,
+        row_title_gp = grid::gpar(fontsize = 12),
+        row_title_side = "left",
+        row_names_side = "left",
+        row_dend_side = "right",
+        row_names_gp = grid::gpar(fontsize = 10),
+        ## Column parameters
+        cluster_columns = FALSE,
+        column_title = "Marker genes",
+        column_title_gp = grid::gpar(fontsize = 12),
+        column_names_side = "bottom",
+        column_names_gp = grid::gpar(fontsize = 10),
+        column_names_rot = 90,
+        ## Legend parameters
+        heatmap_legend_param = list(
+          title = "Jaccard index",
+          direction = "vertical",
+          legend_width = unit(1.5, "in")
+        ),
+        show_heatmap_legend = classification_method == "riggi",
+      )
+    }) |>
+  # concatenate vertically into HeatmapList object
+  purrr::reduce(ComplexHeatmap::`%v%`) |>
+  ComplexHeatmap::draw(
+    heatmap_legend_side = "right",
+    # add a margin to the heatmap so labels don't get cut off
+    padding = unit(c(2, 20, 2, 2), "mm")
+  )
+```
+
+Although not perfect, there is some consistency between classifications across all gene sets. 
+Let's classify cells based on both marker genes and gene set score and then validate those classifications.
+
+```{r}
+classification_df <- classification_df |>
+  dplyr::mutate(
+    # first create a combined classification using all gene sets
+    geneset_classification = dplyr::case_when(
+      riggi_classification == "Tumor" & zhang_classification == "Tumor" & siligan_classification == "Tumor" ~ "Tumor",
+      riggi_classification == "Normal" & zhang_classification == "Normal" & siligan_classification == "Normal" ~ "Normal",
+      .default = "Ambiguous"
+    ),
+    # now classify based on both marker genes and gene set scores
+    combined_classification = dplyr::case_when(
+      marker_gene_classification == "Normal" & geneset_classification == "Normal" ~ "Normal",
+      marker_gene_classification == "Tumor" & geneset_classification == "Tumor" ~ "Tumor",
+      .default = "Ambiguous"
+    )
+  )
+```
+
+Below we see which cells are classified as tumor, normal, or ambiguous.
+
+```{r}
+ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = combined_classification)) +
+  geom_point(alpha = 0.5, size = 0.5) +
+  theme_bw()
+```
+
+## Classify tumor cells with `SingleR`
+
+Another approach for classifying tumor cells is to use `SCPCL000822` as a reference with `SingleR`. 
+We feel pretty confident in those annotations as there was clear consistency between marker gene expression, gene set scores, and copy number inference.
+
+```{r}
+# add annotations from reference data to ref sce file
+coldata_df <- colData(ref_sce) |> 
+  as.data.frame() |> 
+  dplyr::left_join(ref_labels_df, by = c("barcodes" = "cell_barcode"))
+
+# replace coldata with updated coldata 
+colData(ref_sce) <- DataFrame(coldata_df, row.names = coldata_df$barcodes)
+```
+
+```{r}
+# train model using reference labels
+singler_model <- SingleR::trainSingleR(
+  ref_sce,
+  labels = ref_sce$tumor_cell_classification,
+  genes = "de",
+  # only use genes found in input sce 
+  restrict = rownames(sce)
+)
+
+# get singler results 
+singler_results <- SingleR::classifySingleR(
+  trained = singler_model,
+  test = sce,
+  fine.tune = TRUE
+)
+```
+
+```{r}
+# add singler results to classification df 
+classification_df$singler_classification <- singler_results$pruned.labels
+
+# umap of singler results
+ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = singler_classification)) +
+  geom_point(size = 0.5, alpha = 0.5)
+```
+
+Interestingly, `SingleR` classifies most of the cells in the lower right as tumor, whereas CNV inference and gene expression were mixed on which cells from that group were actually tumor.
+
+## Validating annotations
+
+Let's look at the expression of each of the marker genes in individual cells and see how that matches up with the combined classification using marker genes and gene set scores and the `SingleR` classification.
+This is a little circular since we used marker gene expression and gene set scores to classify the cells, but it should still give us a better idea of what the expression of these genes look like in each cell compared to all other cells in the sample.
+This will also allow us to compare annotations using gene sets or `SingleR`.
+
+```{r}
+# create annotation df, keeping both geneset alone and combined classification 
+annotation_df <- classification_df |> 
+  dplyr::select(barcodes, combined_classification, singler_classification)
+
+# create matrix with marker genes as rows and barcodes as columns
+marker_gene_heatmap <- plot_markers_df |> 
+  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
+  tidyr::pivot_wider(values_from = gene_expression, 
+                     names_from = barcodes) |> 
+  tibble::column_to_rownames("gene_symbol") |> 
+  as.matrix()
+
+annotation <- ComplexHeatmap::columnAnnotation(
+  combined_classification = annotation_df$combined_classification,
+  singler_classification = annotation_df$singler_classification, 
+  col = list(combined_classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"),
+             singler_classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
+```
+
+```{r}
+# plot heatmap of marker genes 
+plot_gene_heatmap(marker_gene_heatmap, 
+                  row_title = "Marker gene symbol",
+                  legend_title = "Marker gene \nexpression",
+                  annotation = annotation)
+```
+
+For the combined classification, it looks like there is some separation between normal cells that have less marker genes expressed on the right and tumor cells with more marker genes expressed on the left but it isn't that clear.
+For `SingleR` we do see that the tumor cells appear to separate from the normal cells for the most part. 
+We also see that the normal cells line up with the group of cells that have very little marker gene expression.
+
+We will also look at gene set scores.
+
+```{r}
+# make a matrix of gene set by barcode 
+geneset_heatmap <- geneset_plot_df |> 
+  dplyr::select(mean_score, geneset, barcodes) |> 
+  unique() |> 
+  tidyr::pivot_wider(values_from = mean_score, 
+                     names_from = barcodes) |> 
+  tibble::column_to_rownames("geneset") |> 
+  as.matrix()
+
+# plot heatmap of gene set score 
+plot_gene_heatmap(geneset_heatmap,
+                  annotation = annotation,
+                  legend_title = "Gene set \nscore")
+```
+
+We see a similar pattern here where the combined classification does not have quite as clear a distinction between gene set scores of normal and tumor cells as `SingleR` does.
+
+The last thing we will do is check the distribution of mean CNV proportion in tumor cells and normal cells. 
+Do we see an increase in CNV proportions in tumor cells over normal cells when classifying with either `SingleR` or gene sets?
+
+```{r}
+# pivot classification to be able to facet by classification method 
+cnv_df <- classification_df |> 
+tidyr::pivot_longer(
+    cols = ends_with("classification"),
+    names_to = "method",
+    values_to = "classification"
+  ) |> 
+  dplyr::mutate(
+    method = dplyr::case_when(
+      method == "marker_gene_classification" ~ "Marker genes only",
+      method == "copykat_classification" ~ "CopyKAT - no reference",
+      method == "cnv_proportion_classification" ~ "InferCNV proportion",
+      method == "combined_classification" ~ "combined expression", 
+      method == "singler_classification" ~ "SingleR"
+    )
+  ) |>
+  tidyr::drop_na()
+
+ggplot(cnv_df, aes(x = mean_cnv_proportion, color = classification)) +
+  geom_density() +
+  facet_wrap(vars(method))
+```
+
+Here we see that the combined gene set and marker gene classification does show a slight shift in mean cnv proportion in both tumor and ambiguous cells as compared to normal. 
+However, this is not as striking as the bump in cnv proportion seen in tumor cells when classifying with `SingleR`.
+
+## Save classifications
+
+It appears that the `SingleR` classifications encompass more tumor cells that have both higher marker gene expression, higher gene set scores, and higher CNV scores. 
+I'm inclined to trust these annotations a little more than the slightly arbitrary gene set cut offs that we implemented for classification. 
+However, we will save both annotations in case we need them for future use.
+
+```{r}
+# export final TSV with annotations 
+final_classifications <- classification_df |> 
+  dplyr::select(
+    cell_barcode = barcodes,
+    tumor_cell_classification = combined_classification,
+    singler_tumor_classification = singler_classification
+  ) |> 
+  unique()
+
+readr::write_tsv(final_classifications, final_annotations_file)
+```
+
+## Session Info
+
+```{r session info}
+# record the versions of the packages used in this analysis and other environment information
+sessionInfo()
+```

### 2024-06-18 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_tumor-cell-validation.Rmd
rendered notebook

@@
                                             copykat = annotation_df$copykat_classification,
-                                            col = list(infercnv = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"),
-                                                       copykat = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
+                                            col = list(infercnv = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Reference" = "grey"),
+                                                       copykat = c("Tumor" = "#00274C", "Normal" = "#FFCB05")))
 

### 2024-06-18 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
add to conclusion
WIP: transitioning to classifying as ambiguous

@@
 By contrast, both `scrublet` and `scDblFinder` identify a threshold based on the given dataset they are processing. 
-This notebook used a doublet threshold of `>=0.5` for `cxds`, which may not be universally suitable (though choosing a universally suitable threshold is not easy either!).
+This notebook uses a doublet threshold of `>=0.5` for `cxds`, which may not be universally suitable (though choosing a universally suitable threshold is not easy either!).
 
@@
 
+- `consensus_class`, which will be "doublet" if all three methods are doublet, "singlet" if all three methods are singlet, and "ambiguous" if there are any disagreements
 - `consensus_call`, which will be "doublet" if _all_ methods predict "doublet," and "singlet" otherwise
-- `call_type`, which will classify the consensus call as one of "tp", "tn", "fp", or "fn" (true/false positive/negative)
+- `confusion_call`, which will classify the consensus call as one of "tp", "tn", "fp", or "fn" (true/false positive/negative) based on `consensus_call`
 
@@
 
-      # Add a consensus call column
+      # Add consensus columns:
+      # - consensus_class: "doublet" if all three methods are doublet, "singlet" if all three methods are singlet, 
+      #    and "ambiguous" if there are any disagreements
+      # - consensus_call: "doublet" if all three methods are doublet, and "singlet" otherwise
+      # - confusion_call: tp/tn/fp/fn based on the `consensus_call` column
       dataset_df <- dataset_df |>
         dplyr::rowwise() |>
-        dplyr::mutate(consensus_call = dplyr::if_else(
-          all(
-            c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"
-          ),
+        # Add column `consensus_class`
+        dplyr::mutate(
+          consensus_class = dplyr::case_when(
+            consensus_call == "doublet" ~ "doublet",
+            all(c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "singlet") ~ "singlet", 
+            .default = "ambiguous"
+          )
+         ) |>
+        # Add column `consensus_call`
+        dplyr::mutate(
+          consensus_call = dplyr::if_else(
+            all(c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"),
             "doublet", 
             "singlet"
-        )) |>
+          )
+        ) |>
+        # Add column `confusion_call`
         dplyr::mutate(
-          call_type = dplyr::case_when(
+          confusion_call = dplyr::case_when(
             consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
@@
 
-1. `scDblFinder` singlet/doublet calls. 
+1. `scDblFinder` singlet/doublet calls
 2. `scrublet` singlet/doublet calls
@@
 4. Ground truth single/doublets
-5. Consensus singlet/doublet calls
-6. Points are colored based on comparing the consensus call to the ground truth as one of:
-    - true positive (`tp`), true negative (`tn`), false positive (`fp`), false negative (`fn`)
+5. Consensus _call_ droplets are either "doublet" or "singlet"
+6. Consensus _class_: droplets are either "doublet", "singlet", or "ambiguous"
 
@@
 For three out of four datasets, `scDblFinder` predicts a much larger number of doublets compared to other methods.
-For `pdx-MULTI`, however, `cxds` predicts a much larger number of droplets.
+
+Based on the PCAs, it additionally looks like there are many more false negatives in the consensus prediction that `cxds` appears to capture but other methods do not.
 
@@
 Note that, in the [benchmarking paper these datasets were originally analyzed in](https://doi.org/10.1016/j.cels.2020.11.008), `hm-6k` was observed to be one of the "easiest" datasets to classify across methods.
-Consistent with that observation, it has the highest `kappa` value here, although it is still fairly low - though not as low as the other methods, which are very close to 0.
+Consistent with that observation, it has the highest `kappa` value here.
 

### 2024-06-25 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
clean up PCA code

@@
   scale_color_manual(name = "Call type", values = metric_colors) +
-  ggtitle("Consensus call metrics")
+  ggtitle("Consensus class metrics")
 }
@@
         dplyr::rowwise() |>
+        # Add column `consensus_call`
+        dplyr::mutate(
+          consensus_call = dplyr::if_else(
+            all(c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"),
+            "doublet", 
+            "singlet"
+          )
+        ) |>
         # Add column `consensus_class`
@@
             .default = "ambiguous"
-          )
-         ) |>
-        # Add column `consensus_call`
-        dplyr::mutate(
-          consensus_call = dplyr::if_else(
-            all(c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"),
-            "doublet", 
-            "singlet"
           )
@@
 
-      # Top row: methods themselves ---
-      p1 <- plot_pca_calls(
+      pc_color_columns <- c("scDblFInder prediction" = "scdbl_prediction",
+                            "scublet prediction" = "scrublet_prediction",
+                            "cxds prediction" = "cxds_prediction",
+                            "Ground truth" = "ground_truth", 
+                            "Consensus call" = "consensus_call")
+
+      pc_plot_list <- pc_color_columns |> 
+        purrr::imap(
+          \(color_column, plot_title) {
+            plot_pca_calls(
               df, 
-        color_column = scdbl_prediction, 
-        title = "scDblFinder prediction"
+              color_column = !!sym(color_column), 
+              title = plot_title
+            )}
         )
 
-      p2 <- plot_pca_calls(
+      # note that the list name isn't used here
+      pc_plot_list$metrics <- plot_pca_metrics(
         df,
-        color_column = scrublet_prediction, 
-        title = "scrublet prediction"
-      )
-
-      p3 <- plot_pca_calls(
-        df, 
-        color_column = cxds_prediction, 
-        title = "cxds prediction"
-      )
-
-      # Bottom row: consensus  --
-      p4 <- plot_pca_calls(
-        df, 
-        color_column = ground_truth, 
-        title = "Ground truth"
-      )
-
-      # Second, consensus call
-      p5 <- plot_pca_calls(
-        df, 
-        color_column = consensus_call, 
-        title = "Consensus call"
-      )
-
-      # Third, call type
-      p6 <- plot_pca_metrics(
-        df,
-        call_type, 
+        confusion_call, 
         metric_colors = confusion_colors
@@
 
-      # combine and plot
-      plot( 
-        (p1 + p2 + p3) / (p4 + p5 + p6) + 
+      patchwork::wrap_plots(pc_plot_list, pca_metrics) +
         plot_annotation(
@@
             theme = theme(plot.title = element_text(size = 16))) + 
-          plot_layout(guides = "collect"))
+          plot_layout(guides = "collect") 
     }
@@
 
-This section calculates a confusion matrix and associated statistics on the consensus calls.
+This section calculates a confusion matrix and associated statistics on the consensus classes.
 
@@
         cat("Table of consensus calls:")
-        print(table(df$consensus_call))
+        print(table(df$consensus_class))
 

### 2024-06-26 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd
update notebook to use UCell and AUCell

---
title: "`SCPCL000824` - Exploration of gene set based methods for tumor cell assignment"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
---

## Introduction

In this notebook we attempt to identify tumor cells in `SCPCL000824`. 
We first compare tumor cell annotations obtained from marker genes, `CopyKAT`, and `InferCNV` to identify a group of tumor cells we are confident in. 
However, there is a lack of consensus between tumor cells identified by marker gene expression and CNV inference.

In an attempt to identify a group of cells that we are confident are tumor cells, here we test using the following orthogonal marker gene based methods:

- Marker gene expression from `SCPCL000822` to inform an appropriate cut off for marker gene expression in `SCPCL000824`. 
- [`AUCell`](https://www.bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html) with marker genes and three EWS-FLI1 gene sets in both samples. 
- [`UCell`](https://carmonalab.github.io/UCell_demo/UCell_matrix_vignette.html#content) with marker genes and three EWS-FLI1 gene sets in both samples.

## Setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_bw()
)

# quiet messages
options(readr.show_col_types = FALSE)
ComplexHeatmap::ht_opt(message = FALSE)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "2024-05-01")
sample_dir <- file.path(data_dir, "SCPCP000015", "SCPCS000492")

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-ewings")
```

```{r}
# source in helper functions: plot_gene_heatmap() and plot_cnv_heatmap() 
# create_classification_df() and create_marker_gene_df()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)
```

```{r}
# Input files
sce_file <- file.path(sample_dir, "SCPCL000824_processed.rds")
marker_genes_file <- file.path(module_base, "references", "tumor-marker-genes.tsv")

# results from annotation workflow 
results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000492")

marker_gene_results_file <- file.path(results_dir, "SCPCL000824_tumor-normal-classifications.tsv")
copykat_predictions_file <- file.path(results_dir, "SCPCL000824_copykat-classifications.tsv")

infercnv_predictions_file <- file.path(results_dir, "SCPCL000824_infercnv-classifications.tsv")
infercnv_metadata_file <- file.path(results_dir, "infercnv", "SCPCL000824_cnv-metadata.tsv")

geneset_scores_file <- file.path(results_dir, "SCPCL000824_gene-set-scores.tsv")

# output files 
final_annotations_dir <- file.path(module_base, "cell_type_annotations", "SCPCS000492")
fs::dir_create(final_annotations_dir)
final_annotations_file <- file.path(final_annotations_dir, "SCPCL000824_tumor-classifications.tsv")

# reference files to use with SingleR
ref_sce_file <- file.path(data_dir, "SCPCP000015", "SCPCS000490", "SCPCL000822_processed.rds")
ref_labels_file <- file.path(module_base, "cell_type_annotations", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv")
ref_geneset_scores_file <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
```

```{r}
# read in sce file 
sce <- readr::read_rds(sce_file)

# read in ref sce and ref annotations for comparing between samples 
# ref is SCPCL000822
ref_sce <- readr::read_rds(ref_sce_file)
ref_labels_df <- readr::read_tsv(ref_labels_file)
ref_geneset_df <- readr::read_tsv(ref_geneset_scores_file)
```

```{r}
# generate classification df 
classification_df <- create_classification_df(
  sce,
  marker_gene_results_file,
  copykat_predictions_file, 
  infercnv_predictions_file,
  infercnv_metadata_file,
  geneset_scores_file
) |> 
  # remove extra infercnv classification column we won't use 
  dplyr::select(- cnv_sum_classification)
```

## Evaluate tumor cell classifications

Below is a UMAP showing which cells are labeled as tumor cells using each of the various classification methods (marker genes, `CopyKAT`, and `InferCNV`).

```{r}
# plot tumor cells for each classification method 
tumor_umap_df <- classification_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification")) |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "copykat_classification" ~ "CopyKAT - no reference",
      method == "cnv_proportion_classification" ~ "InferCNV proportion"
    )
  )

ggplot(tumor_umap_df, aes(x = UMAP1, UMAP2, color = classification)) +
  geom_point(size = 0.5, alpha = 0.5) +
  facet_wrap(vars(method))
```

It looks like we don't have a lot of agreement between the marker gene based classification and copy number inference classification. 
Although we do see some agreement between the two CNV methods.

### Marker gene and gene set expression

We can confirm this by looking at marker gene expression in cells classified as tumor or normal by each method.

```{r}
# generate marker genes df
plot_markers_df <- create_marker_gene_df(
  sce,
  classification_df, 
  marker_genes_file
)

# create a density plot showing the distribution of marker gene expression across classification methods 
marker_density_df <- plot_markers_df |>
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "copykat_classification" ~ "CopyKAT - no reference",
      method == "cnv_proportion_classification" ~ "InferCNV proportion"
    )
  )

ggplot(marker_density_df, aes(x = sum_transformed_exp, color = classification)) +
  geom_density() +
  facet_wrap(vars(method))
```

We see a few things from this plot:

- As expected given that marker gene expression was used to classify cells using the "Marker genes only" method, we see an increase in marker gene expression in tumor cells. 
However we don't see a clear bimodal distribution, which is something we did see in `SCPCL000822`. 
- For `InferCNV`, both tumor and normal cells have higher expression of marker genes than the reference. 
- For `CopyKAT`, the normal cells appear to have higher expression of marker genes.

Let's also look at gene set scores and see if we can use those to help us determine which cells are most likely to be tumor cells.

```{r}
# plot gene set scores for each cell  
geneset_plot_df <- classification_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean-")) |> 
  tidyr::pivot_longer(
    cols = starts_with("mean"),
    names_to = "geneset",
    values_to = "mean_score"
  ) |> 
  dplyr::mutate(
    geneset = stringr::word(geneset, -1, sep = "-")
  )

ggplot(geneset_plot_df, aes(x = UMAP1, UMAP2, color = mean_score)) +
  geom_point(size = 0.5, alpha = 0.5) +
  facet_wrap(vars(geneset)) +
  scale_color_viridis_c()
```

We do see that the scores across all three genesets seem to be pretty consistent. 
Cells that have higher scores for `RIGGI`, also have higher scores for `SILIGAN` and `ZHANG`.

Let's look at how the distribution of gene set scores coordinates with assignments from the CNV and marker gene methods. 
Below, we color the distribution based on the classification method used.

```{r}
geneset_plot_df |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "copykat_classification" ~ "CopyKAT - no reference",
      method == "cnv_proportion_classification" ~ "InferCNV proportion"
    )) |> 
  ggplot(aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(rows = vars(method), 
             cols = vars(geneset))
```

We see a few things from this plot:

- The marker genes classification has the most consistent increase in gene set scores in tumor cells compared to normal cells, although the separation is not super clear.  
- For the `RIGGI` and `ZHANG` gene sets, the scores are higher in both tumor and normal cells compared to reference cells as classified by `InferCNV`. 
- For all gene sets, the normal cells have a slightly higher gene set score as classified by `CopyKAT`.

## Using gene expression to classify tumor cells

Unlike with `SCPCL000822`, there is no clear consensus between CNV inference methods and marker gene based classification. 
Because these genomes are quiet, we expect CNV inference to be more difficult and less likely to accurately predict the tumor cells. 
Therefore, we should use methods that rely on gene expression.

The rest of this notebook will explore using different gene set based methods to classify tumor cells to see if we can more accurately identify the tumor cells in this sample.

### Comparing marker gene expression in `SCPCL000822` and `SCPCL000824`

With `SCPCL000822` we had a clear separation between marker gene expression in tumor cells and normal cells because there was a bimodal distribution. 
It does not appear that is the case here so it's going to be more difficult to pull out tumor cells.

First we will just compare the distribution of the raw marker gene expression in `SCPCL000822` and `SCPCL000824`. 
To do this, we will get the total marker gene expression by summing all marker genes in a cell and then plot the distribution. 
The first plot will look at the distribution across all cells and the second will show the classification of tumor vs. normal cells.

```{r}
# look at the raw sum of marker gene expression in both 822 and 824
# first create marker genes df for ref sce 
ref_classification_df <- ref_labels_df |> 
  dplyr::rename("barcodes" = "cell_barcode")

ref_markers_df <- create_marker_gene_df(sce = ref_sce, 
                                        classification_df = ref_classification_df,
                                        marker_genes_file) |> 
  dplyr::mutate(sample = "SCPCL000822")

# combine all marker gene data for both samples into one df 
combined_markers_df <- plot_markers_df |> 
  dplyr::select(barcodes, tumor_cell_classification = marker_gene_classification, gene_symbol, gene_expression, transformed_gene_expression, sum_raw_exp, sum_transformed_exp) |> 
  dplyr::mutate(sample = "SCPCL000824") |> 
  dplyr::bind_rows(ref_markers_df)
```

```{r}
# total distribution
ggplot(combined_markers_df, aes(x = sum_raw_exp)) +
  geom_density() +
  facet_grid(rows = vars(sample))

# color by tumor cell classification 
ggplot(combined_markers_df, aes(x = sum_raw_exp, color = tumor_cell_classification)) +
  geom_density() +
  facet_grid(rows = vars(sample))
```

Looking at this, we see that `SCPCL000822` has a bimodal distribution and shows a clear cutoff where cells are tumor vs. normal, but this is not the case for `SCPCL000824`. 
Let's find the local minima in the bimodal distribution for `SCPCL000822` and then use that to classify tumor cells in `SCPCL000824`.

```{r}
# create distribution 
density_data <- density(combined_markers_df$sum_raw_exp)
# find the local minima in the distribution
exp_cutoff <- optimize(approxfun(density_data$x, density_data$y), interval = c(1,10))$minimum
```

```{r}
# add new column with updated marker gene classification 
# use local minima from 822 to define 824
new_classification <- combined_markers_df |> 
  dplyr::filter(sample == "SCPCL000824") |> 
  dplyr::mutate(updated_marker_gene_classification = dplyr::if_else(sum_raw_exp >= exp_cutoff, "Tumor", "Normal")) |> 
  dplyr::select(barcodes, updated_marker_gene_classification)

# add new column to existing classification
classification_df <- classification_df |> 
  dplyr::left_join(new_classification)

# label cells based on new classifiation 
ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = updated_marker_gene_classification)) +
  geom_point(alpha = 0.5, size = 0.5)
```

This looks closer to what I would expect with most cells in the large group of cells corresponding to tumor cells. 
This also captures cells that are labeled as tumor cells by both marker genes and CNV inference.

### Comparing gene set expression in `SCPCL000822` and `SCPCL000824`

Now we will look at the gene set scores for each gene set in `SCPCL000822` and `SCPCL000824`.  
We did not use these to classify `SCPCL000822`, so we won't actually do any classification, but this will show us if the scores for tumor cells are similar to each other across samples.

```{r}
# get geneset score from reference sce 
ref_geneset_df <- ref_geneset_df |> 
  dplyr::select(barcodes, starts_with("mean")) |> 
  tidyr::pivot_longer(
    cols = starts_with("mean"),
    names_to = "geneset",
    values_to = "mean_score"
  ) |> 
  dplyr::mutate(
    geneset = stringr::word(geneset, -1, sep = "-"),
    sample = "SCPCL000822 - reference"
  )

# join with gene set scores from 824 and plot distribution 
geneset_plot_df |> 
  dplyr::select(barcodes, geneset, mean_score) |> 
  dplyr::mutate(sample = "SCPCL000824") |> 
  dplyr::bind_rows(ref_geneset_df) |> 
  ggplot(aes(x = mean_score, color = sample)) +
  geom_density() +
  facet_grid(rows = vars(geneset))
```

Again we see that the gene set scores are bimodal for `SCPCL000822`, at least for `RIGGI` and `ZHANG`. 
However, we don't see that for `SCPCL000824`, but we do see a very similar ranger of values. 
This makes me think that the majority of the cells are in fact tumor cells.

## Classifying tumor cells with AUCell

The next thing we will do is see if we can use a more data driven approach to identify cells with higher marker gene or gene set expression. 
We will use [`AUCell`](https://www.bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html) to do this and will run it on both samples. 
However, `AUCell` relies on having a bimodal distribution in gene set expression, so I expect we may need to do a similar thing where we use cutoffs determined in `SCPCL000822` to call cells in `SCPCL000824`.

```{r}
# names of gene sets to grab
ews_gene_sets <- c(
  "ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION",
  "RIGGI_EWING_SARCOMA_PROGENITOR_UP",
  "SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN"
)

# pull gene sets from msigbdr 
# all gene sets are part of C2, CGP
genes_df <- msigdbr::msigdbr(species= "Homo sapiens", 
                             category = "C2",
                             subcategory = "CGP") |> 
  # only keep relevant gene sets 
  dplyr::filter(gs_name %in% ews_gene_sets)
```

```{r}
# first create named list of genes
# we will need this to run UCell later too 
genes_list <- ews_gene_sets |> 
  purrr::map(\(name){
    genes <- genes_df |> 
      dplyr::filter(gs_name == name) |> 
      dplyr::pull(ensembl_gene)
  }) |> 
  purrr::set_names(ews_gene_sets)

# build GeneSetCollection for AUCell 
msig_gene_sets <- genes_list |> 
  purrr::imap(\(genes, name) GSEABase::GeneSet(genes, setName = name))

# get list of marker genes to add to GeneSets 
marker_genes <- readr::read_tsv(marker_genes_file, show_col_types = FALSE) |> 
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
  dplyr::distinct() |> 
  dplyr::filter(cell_type == "tumor") |> 
  dplyr::pull(ensembl_gene_id)

# turn it into a gene set 
marker_gene_set <- marker_genes |> 
  GSEABase::GeneSet(setName = "markers")

# join msig and marker genes into a collection to input with AUCell 
collection <- GSEABase::GeneSetCollection(c(msig_gene_sets, marker_gene_set))
```

```{r}
# run AUCell
ref_auc <- AUCell::AUCell_run(counts(ref_sce), collection)

# assign cells based on threshold and output plots 
ref_assignment <- AUCell::AUCell_exploreThresholds(ref_auc, assign = TRUE)
```

Here we can see a bimodal distribution of all gene sets except `SILIGAN`, which is to be expected based on the original distribution plots. 
We also see that `AUCell` identifies a threshold based on the local minima of the AUC distribution where all cells above the threshold are classified as expressing the gene set.
Let's compare the assignments from our validated classifications to using `AUCell` with `SCPCL000822`.

```{r}
# compare AUCell classification vs cnv classification in ref
# pull out tumor cells found by AUCell
tumor_cells <- ref_assignment$markers$assignment

ref_classification_df <- ref_classification_df |> 
  # add new column with auc classification 
  dplyr::mutate(auc_classification = dplyr::if_else(barcodes %in% tumor_cells, "Tumor", "Normal"))

# filter out any ambiguous cells 
caret_df <- ref_classification_df |> 
  dplyr::filter(tumor_cell_classification != "Ambiguous")

# compare using a confusion matrix
caret::confusionMatrix(
    table(
      caret_df$tumor_cell_classification, 
      caret_df$auc_classification)
  ) 
```

It looks like we get pretty good agreement between our original classification and the updated `AUCell` classification using the marker gene list.

Now we repeat on `SCPCL000824`.

```{r}
# get new auc and plot thresholds 
new_auc <- AUCell::AUCell_run(counts(sce), collection)
new_assignment <- AUCell::AUCell_exploreThresholds(new_auc, assign = TRUE)
```

Here we do not see any bimodal distributions and that most cells are considered to not be expressing the specified gene sets, which doesn't fit with our previous findings. 
I expect that without a bimodal distribution this model does not do a good job of calculating a threshold.

Instead, we will use the threshold identified by `AUCell` for `SCPCL000822` to identify tumor cells in `SCPCL000824`.  
Here, we label any cells that have passed the threshold set for the marker gene sets as tumor cells.

```{r}
# pull out threshold used for assigning cells from ref sample 
ref_threshold <- ref_assignment$markers$aucThr$selected

# create a vector of tumor cells in 824 using the cutoff from 822
new_tumor_cells <- new_auc@assays@data$AUC |> 
  as.data.frame() |> 
  tibble::rownames_to_column("gene_set") |> 
  tidyr::pivot_longer(!gene_set, 
                      names_to = "barcodes", 
                      values_to = "auc") |> 
  dplyr::filter(gene_set == "markers", auc >= ref_threshold) |>
  dplyr::pull(barcodes)

# add to original classification df 
classification_df <- classification_df |> 
  dplyr::mutate(auc_classification = dplyr::if_else(barcodes %in% new_tumor_cells, "Tumor", "Normal"))
```

```{r}
# visualize which cells are classified as tumor 
ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = auc_classification)) +
  geom_point(alpha = 0.5, size = 0.5)
```

This looks similar to using the updated marker gene cutoff to classify cells where most cells in that bottom group are tumor cells.

## Classifying tumor cells with UCell

Here we will look at using [`UCell`](https://carmonalab.github.io/UCell_demo/UCell_matrix_vignette.html#content) to calculate a gene set score for each cell and then attempt to classify cells based on the distribution of those gene set scores. 
Again, we will look at both `SCPCL000822` and `SCPCL000824`.

```{r}
# create list to use for ucell of gene sets and marker genes
ucell_gene_sets <- c(genes_list, markers = list(marker_genes))

# run ucell on both ref (822) and 824
ref_ucell <- UCell::ScoreSignatures_UCell(counts(ref_sce), features = ucell_gene_sets)
new_ucell <- UCell::ScoreSignatures_UCell(counts(sce), features = ucell_gene_sets)
```

`UCell` returns a signature score for each gene set and does not calculate any thresholds for classification on it's own. 
We will look at the distribution of scores across both samples below and see if we can identify a good cut off to use for classification.

```{r}
# plot distribution of ucell scores for both samples 
ref_ucell <- ref_ucell |> 
  as.data.frame() |> 
  tibble::rownames_to_column("barcodes") |> 
  dplyr::mutate(sample = "SCPCL000822")

ucell_df <- new_ucell |> 
  as.data.frame() |> 
  tibble::rownames_to_column("barcodes") |> 
  dplyr::mutate(sample = "SCPCL000824") |> 
  dplyr::bind_rows(ref_ucell) |> 
  tidyr::pivot_longer(ends_with("UCell"), 
                      names_to = "gene_list", 
                      values_to = "signature_score") |> 
  dplyr::mutate(
    gene_list = stringr::word(gene_list, 1, sep = "_")
  )
```

```{r}
ggplot(ucell_df, aes(x = signature_score, colour = sample)) +
  geom_density() +
  facet_grid(rows = vars(gene_list))
```

Here we see that the marker gene set maybe has a small bimodal distribution in `SCPCL000822`, but it is very small and not quite as striking as looking at the scores from `AUCell` or raw gene expression. 
`RIGGI`, `ZHANG`, and `SILIGAN` both show very similar distributions to the gene set scores distributions from our manual calculations of gene set scores (mean of all genes in the gene set). 
Again, there is a bimodal distribution for `RIGGI` and `ZHANG`, but not for `SILIGAN` and it is only present in `SCPCL000822`.

Let's use the local minima from the `RIGGI` gene set scores for `SCPCL000822` to identify tumor cells. 
First we will look at how using `UCell` compares to our previous classifications for `SCPCL000822`.

```{r}
# create distribution 
density_data <- density(ref_ucell$RIGGI_EWING_SARCOMA_PROGENITOR_UP_UCell)
# find the local minima in the distribution
geneset_cutoff <- optimize(approxfun(density_data$x, density_data$y), interval = c(0.01, 0.2))$minimum

# get a vector of tumor cells from ref 
ref_ucell_tumor_cells <- ucell_df |> 
  dplyr::filter(gene_list == "RIGGI",
                signature_score >= geneset_cutoff,
                sample == "SCPCL000822") |> 
  dplyr::pull(barcodes) |> 
  unique()

# add classification column for comparison to original classification
caret_df <- caret_df |> 
  dplyr::mutate(
    ucell_classification = dplyr::if_else(barcodes %in% ref_ucell_tumor_cells, "Tumor", "Normal")
  )

# confusion matrix between original and ucell 
caret::confusionMatrix(
  table(
    caret_df$tumor_cell_classification, 
    caret_df$ucell_classification 
  )
)
```

It looks like we get pretty good agreement between using `UCell` with the `RIGGI` gene set to the original classification. 
Now we will use the cutoff from `SCPCL000822` to classify tumor cells in `SCPCL000824`.

```{r}
# get list of tumor cells using ucell 
new_ucell_tumor_cells <- ucell_df |> 
  dplyr::filter(gene_list == "RIGGI",
                signature_score >= geneset_cutoff,
                sample == "SCPCL000824") |> 
  dplyr::pull(barcodes) |> 
  unique()

# add ucell classification
classification_df <- classification_df |> 
  dplyr::mutate(
    ucell_classification = dplyr::if_else(barcodes %in% new_ucell_tumor_cells, "Tumor", "Normal")
  )

ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = ucell_classification)) +
  geom_point(alpha = 0.5, size = 0.5)
```

Here we see less cells classified as tumor cells than we saw with the `AUCell` classification. 
This also looks very similar to the original UMAP when we classified cells using marker gene expression.

## Validate annotations

Finally, we will look at the marker gene and gene set scores for every cell across annotation methods.

```{r}
# create annotation df, keeping all classification methods 
annotation_df <- classification_df |> 
  dplyr::select(barcodes, marker_gene_classification, updated_marker_gene_classification, auc_classification, ucell_classification) |> 
  unique()

# create matrix with marker genes as rows and barcodes as columns
marker_gene_heatmap <- plot_markers_df |> 
  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
  tidyr::pivot_wider(values_from = gene_expression, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("gene_symbol") |> 
  as.matrix()

annotation <- ComplexHeatmap::columnAnnotation(
  marker_genes = annotation_df$marker_gene_classification,
  updated_marker_genes = annotation_df$updated_marker_gene_classification,
  AUCell = annotation_df$auc_classification, 
  UCell = annotation_df$ucell_classification,
  col = list(marker_genes = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"),
             updated_marker_genes = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"), 
             AUCell = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"),
             UCell = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
```

```{r}
# plot heatmap of marker genes 
plot_gene_heatmap(marker_gene_heatmap, 
                  row_title = "Marker gene symbol",
                  legend_title = "Marker gene \nexpression",
                  annotation = annotation)
```

Here we see that the updated marker gene classification (using raw gene expression cut offs determined from `SCPCL000822`) and `AUCell` tend to have the clearest sepration between tumor cells and normal cells that line up with expression of individual marker genes.

Now we will create the same plot but with gene set scores.

```{r}
# make a matrix of gene set by barcode 
geneset_heatmap <- geneset_plot_df |> 
  dplyr::select(mean_score, geneset, barcodes) |> 
  unique() |> 
  tidyr::pivot_wider(values_from = mean_score, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("geneset") |> 
  as.matrix()

# plot heatmap of gene set score 
plot_gene_heatmap(geneset_heatmap,
                  annotation = annotation,
                  legend_title = "Gene set \nscore")
```

Again, we see that the updated marker genes and `AUCell` have the most similar assignments and these tend to group by tumor/normal cells. 
Although we do see a similar separation of cells in `UCell`, but there are many more cells that are classified as normal.

## Save annotations

We will go ahead and save the annotations from all gene expression based methods for future use.

```{r}
# export final TSV with annotations 
classifications_output <- classification_df |> 
  dplyr::select(
    cell_barcode = barcodes,
    marker_gene_classification,
    updated_marker_gene_classification, 
    auc_classification,
    ucell_classification
  ) |> 
  unique()

readr::write_tsv(classifications_output, final_annotations_file)
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-06-26 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
small pca plot fix, but plots still not rendering?

@@
 
-      pc_color_columns <- c("scDblFInder prediction" = "scdbl_prediction",
+      pc_color_columns <- c("scDblFinder prediction" = "scdbl_prediction",
                             "scublet prediction" = "scrublet_prediction",
@@
 
-      patchwork::wrap_plots(pc_plot_list, pca_metrics) +
+      patchwork::wrap_plots(pc_plot_list) +
         plot_annotation(

### 2024-06-26 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
need to use purrr::map() to get patchwork to print. results in extra names printing out, but it is what it is probably
Now making the correct PCA plots with single function, and added section for comparing 2 methods

@@
                            color_column, 
-                           pred_column,
+                           plot_type = "calls",
                            title) {
-  # Plot PCs colored by singlet/doublet, showing doublets on top
-  # df is expected to contain columns PC1, PC2, `color_column`, and `pred_column`. These should _not_ be provided as strings
-  ggplot(df) + 
-    aes(x = PC1, 
+  # Plot PCs colored by either:
+  # if plot_type == "calls", color by singlet or doublet, showing doublets on top
+  # if plot_type == "class", color by singlet, doublet, or ambiguous, showing doublets & ambiguous on top
+  # df is expected to contain columns PC1, PC2, `color_column`, which should _not_ be provided as a string
+  plot <- ggplot(df) + 
+    aes(
+      x = PC1, 
       y = PC2, 
-        color = {{color_column}}) +
+      color = {{color_column}}
+    ) +
     geom_point(
@@
     ) +
-  scale_color_manual(name = "", values = c("black", "gray80")) + 
+    ggtitle(title)
+
+  if (plot_type == "calls") {
+    plot <- plot + 
+      scale_color_manual(
+        name = "", 
+        values = c("doublet" = "black", "singlet" = "gray90")) + 
       geom_point(
@@
         size = 0.75
+      )
+  } else if (plot_type == "class") {
+    plot <- plot + 
+      scale_color_manual(
+        name = "", 
+        values = c("ambiguous" = "lightblue", 
+                   "doublet" = "black", 
+                   "singlet" = "gray90")
       ) + 
-  ggtitle(title) 
+      geom_point(
+        data = dplyr::filter(df, consensus_class != "singlet"), 
+        size = 0.75,
+        alpha = 0.7,
+        aes(color = consensus_class),
+      )
+  } else {
+    stop("plot_type must be 'calls' or 'class'")
   }
 
-plot_pca_metrics <- function(df, color_column, metric_colors) {
-  # Plot PCs colored by performance metric, showing false calls on top
-  # metric_colors is a named vector of colors used for coloring tp/tn/fp/fn
-  # df is expected to contain columns PC1, PC2, and `color_column`. This should _not_ be provided as a string.
-  ggplot(df) + 
-    aes(x = PC1, 
-        y = PC2, 
-        color = {{color_column}}) +
-  geom_point(
-    size = 0.75, 
-    alpha = 0.6
-  ) + 
-  geom_point(
-    data = dplyr::filter(df, {{color_column}} %in% c("fp", "fn")), 
-    size = 0.75
-  ) +
-  scale_color_manual(name = "Call type", values = metric_colors) +
-  ggtitle("Consensus class metrics")
+  return(plot)
 }
@@
 
-## Read and prepare input data
+## Explore consensus across methods
+
+### Read and prepare input data
 
@@
 ```{r read_data}
-# used in PCA plots
-confusion_colors <- c(
-  "tp" = "lightblue",
-  "tn" = "pink",
-  "fp" = "blue",
-  "fn" = "firebrick2"
-)
-
 # Read in data for analysis
@@
 
-For the first five panels, doublets are shown in black and singlets in light gray.
+- For the first five panels, doublets are shown in black and singlets in light gray.
+- In the sixth panel showing the consensus class, doublets are shown in black, singlets in light gray, and ambiguous are blue.
 
@@
 doublet_df_list |>
-  purrr::iwalk(
+  purrr::imap(
     \(df, dataset) {
@@
               color_column = !!sym(color_column), 
+              plot_type = "calls",
               title = plot_title
@@
         ) 
-
-      # note that the list name isn't used here
-      pc_plot_list$metrics <- plot_pca_metrics(
+      # plot for consensus_class, which needs three colors
+      # note that the list name isn't actually used here
+      pc_plot_list$consensus_class <- plot_pca_calls(
         df, 
-        confusion_call, 
-        metric_colors = confusion_colors
+        color_column = consensus_class,
+        plot_type = "class",
+        title = "Consensus class"
       )
@@
 
+## Explore consensus of `scDblFinder` and `cxds`
+
+The above analysis suggests that `scrublet` may be more conservative than the other two methods, so including it in a consensus may be increasing the number of false positives.
+Here, we'll repeat the same consensus analyses, but considering _only_ `scDblFinder` and `cxds`.
+Since there are only two methods here, we will create two columns for this data:
+
+### Prepare data
+
+```{r}
+# modify existing consensus columns to only consider 2 methods
+doublet_df_list_2 <- doublet_df_list |>
+  purrr::map(
+    \(dataset_df) {
+
+      dataset_df <- dataset_df |>
+        dplyr::rowwise() |>
+        # Add column `consensus_call`
+        dplyr::mutate(
+          consensus_call = dplyr::if_else(
+            all(c(scdbl_prediction, cxds_prediction) == "doublet"),
+            "doublet", 
+            "singlet"
+          )
+        ) |>
+        # Add column `consensus_class`
+        dplyr::mutate(
+          consensus_class = dplyr::case_when(
+            consensus_call == "doublet" ~ "doublet",
+            all(c(scdbl_prediction, cxds_prediction) == "singlet") ~ "singlet", 
+            .default = "ambiguous"
+          )
+         ) |>
+        # Add column `confusion_call`
+        dplyr::mutate(
+          confusion_call = dplyr::case_when(
+            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
+            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
+            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
+            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
+          )
+        )
+
+      return(dataset_df)
+    }
+  ) |> 
+  purrr::set_names(dataset_names)
+```
+
+### PCA
+
+This section plots the PCA for each dataset, clockwise from the top left:
+
+1. `scDblFinder` singlet/doublet calls
+2. `cxds` singlet/doublet calls
+3. Ground truth single/doublets
+4. Consensus _call_ droplets are either "doublet" or "singlet"
+5. Consensus _class_: droplets are either "doublet", "singlet", or "ambiguous"
+
+For the first four panels, doublets are shown in black and singlets in light gray.
+
+```{r, fig.width = 12, fig.height = 6}
+doublet_df_list_2 |>
+  purrr::imap(
+    \(df, dataset) {
+
+      pc_color_columns <- c("scDblFinder prediction" = "scdbl_prediction",
+                            "cxds prediction" = "cxds_prediction",
+                            "Ground truth" = "ground_truth", 
+                            "Consensus call" = "consensus_call")
+
+      pc_plot_list <- pc_color_columns |> 
+        purrr::imap(
+          \(color_column, plot_title) {
+            plot_pca_calls(
+              df, 
+              color_column = !!sym(color_column), 
+              plot_type = "calls",
+              title = plot_title
+            )}
+        ) 
+      # plot for consensus_class, which needs three colors
+      # note that the list name isn't actually used here
+      pc_plot_list$consensus_class <- plot_pca_calls(
+        df, 
+        color_column = consensus_class,
+        plot_type = "class",
+        title = "Consensus class"
+      )
+
+      patchwork::wrap_plots(pc_plot_list) +
+        plot_annotation(
+            glue::glue("PCA for {dataset}"), 
+            theme = theme(plot.title = element_text(size = 16))) + 
+        plot_layout(guides = "collect")
+    }
+  )
+```
+
+### Performance metrics
+
+This section calculates a confusion matrix and associated statistics on the consensus classes.
+
+```{r}
+metric_df <- doublet_df_list_2 |>
+  purrr::imap( 
+    \(df, dataset) {
+        print(glue::glue("======================== {dataset} ========================"))
+
+        cat("Table of consensus calls:")
+        print(table(df$consensus_class))
+
+        cat("\n\n")
+
+        confusion_result <- caret::confusionMatrix(
+          # truth should be first
+          table(
+            "Truth" = df$ground_truth,
+            "Consensus prediction" = df$consensus_call
+          ), 
+          positive = "doublet"
+        )
+
+        print(confusion_result)
+
+        # Extract information we want to present later in a table
+        tibble::tibble(
+          "Dataset name" = dataset,
+          "Kappa" = round(confusion_result$overall["Kappa"], 3), 
+          "Balanced accuracy" = round(confusion_result$byClass["Balanced Accuracy"], 3)
+        )
+    }
+  ) |>
+  dplyr::bind_rows()
+```
+
 ## Session Info

### 2024-06-27 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
header levels
Notebook rearrangement so upset plots are before all consensus code. Added conclusion for part 2.

@@
 
-## Explore consensus across methods
+## Read and prepare input data
 
-### Read and prepare input data
+First, we'll read in and combine doublet results, as well as PCA coordinates, into a list of data frames for each dataset.
 
-First, we'll read in and combine doublet results into a list of data frames for each dataset.
-We'll also create new columns for each dataset:
-
-- `consensus_class`, which will be "doublet" if all three methods are doublet, "singlet" if all three methods are singlet, and "ambiguous" if there are any disagreements
-- `consensus_call`, which will be "doublet" if _all_ methods predict "doublet," and "singlet" otherwise
-- `confusion_call`, which will classify the consensus call as one of "tp", "tn", "fp", or "fn" (true/false positive/negative) based on `consensus_call`
-
-```{r paths}
+```{r dataset names}
 # find all dataset names to process:
@@
       sce <- readr::read_rds(sce_file)
-      dataset_df <- scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
+
+      scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
         tibble::rownames_to_column(var = "barcodes") |>
@@
         ) 
-
-      # Add consensus columns:
-      # - consensus_class: "doublet" if all three methods are doublet, "singlet" if all three methods are singlet, 
-      #    and "ambiguous" if there are any disagreements
-      # - consensus_call: "doublet" if all three methods are doublet, and "singlet" otherwise
-      # - confusion_call: tp/tn/fp/fn based on the `consensus_call` column
-      dataset_df <- dataset_df |>
-        dplyr::rowwise() |>
-        # Add column `consensus_call`
-        dplyr::mutate(
-          consensus_call = dplyr::if_else(
-            all(c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"),
-            "doublet", 
-            "singlet"
-          )
-        ) |>
-        # Add column `consensus_class`
-        dplyr::mutate(
-          consensus_class = dplyr::case_when(
-            consensus_call == "doublet" ~ "doublet",
-            all(c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "singlet") ~ "singlet", 
-            .default = "ambiguous"
-          )
-         ) |>
-        # Add column `confusion_call`
-        dplyr::mutate(
-          confusion_call = dplyr::case_when(
-            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
-            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
-            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
-            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
-          )
-        )
-
-      return(dataset_df)
     } 
@@
 ```{r}
-upset_list <- doublet_df_list |>
+doublet_df_list |>
   purrr::iwalk(
@@
 
-## Evaluating consensus performance
+## Explore consensus across methods
 
-This section visualizes and evaluates the consensus doublet calls across each dataset.
+Next, well explore the consensus of doublet predictions across methods. 
+To this end, we'll create some new columns for each dataset:
+
+- `consensus_class`, which will be "doublet" if all three methods are doublet, "singlet" if all three methods are singlet, and "ambiguous" if there are any disagreements
+- `consensus_call`, which will be "doublet" if _all_ methods predict "doublet," and "singlet" otherwise
+- `confusion_call`, which will classify the consensus call as one of "tp", "tn", "fp", or "fn" (true/false positive/negative) based on `consensus_call`
+
+```{r}
+doublet_df_list <- doublet_df_list |>
+  purrr::map(
+    \(dataset_df) {
+
+      dataset_df |>
+        dplyr::rowwise() |>
+        # Add column `consensus_call`
+        #  "doublet" if all three methods are doublet, and "singlet" otherwise
+        dplyr::mutate(
+          consensus_call = dplyr::if_else(
+            all(c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"),
+            "doublet", 
+            "singlet"
+          )
+        ) |>
+        # Add column `consensus_class`
+        #  "doublet" if all three methods are doublet, "singlet" if all three methods are singlet, 
+        #  and "ambiguous" if there are any disagreements
+        dplyr::mutate(
+          consensus_class = dplyr::case_when(
+            consensus_call == "doublet" ~ "doublet",
+            all(c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "singlet") ~ "singlet", 
+            .default = "ambiguous"
+          )
+         ) |>
+        # Add column `confusion_call`
+        #  tp/tn/fp/fn based on the `consensus_call` column
+        dplyr::mutate(
+          confusion_call = dplyr::case_when(
+            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
+            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
+            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
+            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
+          )
+        )
+      }
+  ) 
+```
 
@@
               title = plot_title
-            )}
+            ) + theme(legend.position = "none")}
         ) 
@@
 
-This section calculates a confusion matrix and associated statistics on the consensus classes.
+This section shows a table of the consensus class counts and calculates a confusion matrix with associated statistics from the consensus calls:
 
@@
 
-        cat("Table of consensus calls:")
+        cat("Table of consensus class counts:")
         print(table(df$consensus_class))
@@
 
-## Conclusions
+### Conclusions
 
@@
 # modify existing consensus columns to only consider 2 methods
-doublet_df_list_2 <- doublet_df_list |>
+doublet_df_list <- doublet_df_list |>
   purrr::map(
@@
         dplyr::rowwise() |>
-        # Add column `consensus_call`
+        # Update column `consensus_call`
         dplyr::mutate(
@@
         ) |>
-        # Add column `consensus_class`
+        # Add Update `consensus_class`
         dplyr::mutate(
@@
          ) |>
-        # Add column `confusion_call`
+        # Add Update `confusion_call`
         dplyr::mutate(
@@
 
-For the first four panels, doublets are shown in black and singlets in light gray.
-
 ```{r, fig.width = 12, fig.height = 6}
-doublet_df_list_2 |>
+doublet_df_list |>
   purrr::imap(
@@
 
-This section calculates a confusion matrix and associated statistics on the consensus classes.
+This section shows a table of the consensus class counts and calculates a confusion matrix with associated statistics from the consensus calls:
 
 ```{r}
-metric_df <- doublet_df_list_2 |>
+metric_df <- doublet_df_list |>
   purrr::imap( 
@@
 
+### Conclusions
+
+Considering only `scDblFinder` and `cxds`, we see several particular differences in statistics:
+
+- Balanced accuracy for `hm-6k` and `HMEC-orig-MULTI` are about the same here as for the consensus among all methods, but accuracy has substantially _decreased_ for `pbmc-1B-dm` and `pdx-MULTI1`
+- By contrast, kappa values have increased for all datasets, with the most marked increase for `pdx-MULTI`.
+
+Even so, only `hm-6k` (the "easy" dataset) has both a high kappa and high balanced accuracy.
+While `HMEC-orig-MULTI`'s balanced accuracy is fairly high, its kappa value is not.
+
+```{r}
+metric_df
+```
+
 ## Session Info

### 2024-06-27 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_singler-classification.Rmd
notebook for classifying with SIngleR

---
title: "`SCPCL000824` - Using `SingleR` for classification of tumor cells"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
---

## Introduction

## Setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_bw()
)

# quiet messages
options(readr.show_col_types = FALSE)
ComplexHeatmap::ht_opt(message = FALSE)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "2024-05-01")
sample_dir <- file.path(data_dir, "SCPCP000015", "SCPCS000492")

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-ewings")
```

```{r}
# source in helper functions: plot_gene_heatmap() and plot_cnv_heatmap() 
# create_classification_df() and create_marker_gene_df()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)

# source in helper functions for make_jaccard_matrix() and jaccard()
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)
```

```{r}
# Input files
sce_file <- file.path(sample_dir, "SCPCL000824_processed.rds")
marker_genes_file <- file.path(module_base, "references", "tumor-marker-genes.tsv")

# annotations from SCPCL000824_tumor-gene-expression-exploration.Rmd
annotations_dir <- file.path(module_base, "cell_type_annotations", "SCPCS000492")
annotations_file <- file.path(annotations_dir, "SCPCL000824_tumor-classifications.tsv.gz")

results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000492")
geneset_scores_file <- file.path(results_dir, "SCPCL000824_gene-set-scores.tsv")

# singler results 
full_singler_results <- file.path(module_base, "scratch", "SingleR", "SCPCL000824_singler-results.rds")
singler_annotations_file <- file.path(annotations_dir, "SCPCL000824_singler-classifications.tsv")
```

```{r}
# read in sce file 
sce <- readr::read_rds(sce_file)

# read in annotations
annotations_df <- readr::read_tsv(annotations_file)
singler_annotations_df <- readr::read_tsv(singler_annotations_file)

# read in full singler results 
singler_results_list <- readr::read_rds(full_singler_results)

# read in gene set scores
geneset_scores_df <- readr::read_tsv(geneset_scores_file)
```

```{r}
# generate classification df
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, singler_celltype_annotation) |> 
  dplyr::left_join(annotations_df, by = c("barcodes" = "cell_barcode")) |> 
  dplyr::left_join(singler_annotations_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes") |>
  # get the top cell types for each classification method for plotting later 
  dplyr::mutate(
    across(
      ends_with("_classification"),
      \(x) {
        x |>
          forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
          forcats::fct_infreq() |>
          forcats::fct_relevel("All remaining cell types", after = Inf)
      },
      .names = "{.col}_lumped"
    )
  )
```

## SingleR

```{r}
annotations_to_plot <- c(
  "blueprint_tumor_classification_lumped",
  "updated_singler_classification_lumped",
  "updated_singler_blueprint_classification_lumped"
)

annotations_to_plot |> 
  purrr::map(\(annotation_column){
    ggplot(
    classification_df,
    aes(x = UMAP1, y = UMAP2, color = .data[[annotation_column]])
  ) +
    # set points for all "other" points
    geom_point(
      data = dplyr::select(
        classification_df, -.data[[annotation_column]]
      ),
      color = "gray80",
      alpha = 0.5,
      size = 0.5
    ) +
    # set points for desired cell type
    geom_point(size = 0.5, alpha = 0.5) +
    facet_wrap(
      vars(.data[[annotation_column]]),
      ncol = 3
    ) +
    scale_color_brewer(palette = "Dark2") +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    guides(
      color = guide_legend(
        title = "Cell types",
        # more visible points in legend
        override.aes = list(
          alpha = 1,
          size = 1.5
        )
      )
    )
  })
```

Compare tumor vs. other in all classification methods vs. updated

```{r}
plot_jaccard <- function(classification_df,
                         singler_results_column,
                         methods_to_compare
                         ){

  # create jaccard matrices for SingleR and CellAssign compared to aneuploid/diploid 
  jaccard_matrices <- methods_to_compare |>
    purrr::map(\(name) {
      make_jaccard_matrix(
        classification_df,
        singler_results_column,
        name
      )
    })

  # Set heatmap padding option
  heatmap_padding <- 0.2
  ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))
  # list of heatmaps looking at SingleR/ CellAssign vs tumor/normal 
  heatmap <- jaccard_matrices |>
    purrr::imap(
      \(celltype_mat, celltype_method) {
        ComplexHeatmap::Heatmap(
          t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
          col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
          border = TRUE,
          ## Row parameters
          cluster_rows = TRUE,
          row_title = celltype_method,
          row_title_gp = grid::gpar(fontsize = 12),
          row_title_side = "left",
          row_names_side = "left",
          row_dend_side = "right",
          row_names_gp = grid::gpar(fontsize = 10),
          ## Column parameters
          cluster_columns = TRUE,
          column_title = singler_results_column,
          column_title_gp = grid::gpar(fontsize = 12),
          column_names_side = "bottom",
          column_names_gp = grid::gpar(fontsize = 10),
          column_names_rot = 90,
          ## Legend parameters
          heatmap_legend_param = list(
            title = "Jaccard index",
            direction = "vertical",
            legend_width = unit(1.5, "in")
          ),
          show_heatmap_legend = celltype_method == "UCell",
        )
      }) |>
    # concatenate vertically into HeatmapList object
    purrr::reduce(ComplexHeatmap::`%v%`) |>
    ComplexHeatmap::draw(
      heatmap_legend_side = "right",
      # add a margin to the heatmap so labels don't get cut off
      padding = unit(c(2, 20, 2, 2), "mm")
    )

  return(heatmap)

}
```

```{r}
 original_methods <- c(
    "updated_marker_gene_classification",
    "auc_classification",
    "ucell_classification"
  ) |> 
  purrr::set_names("Marker genes", "AUCell", "UCell")

heatmap_list <- annotations_to_plot |> 
  purrr::walk(\(annotation_column) {plot_jaccard(classification_df, annotation_column, original_methods)})
```

```{r}
methods <- c(
    "blueprint_tumor_classification_lumped",
    "updated_singler_classification_lumped",
    "updated_singler_blueprint_classification_lumped"
  ) |> 
  purrr::set_names("Blueprint", "Original", "Original and Blueprint")

heatmap_list <- plot_jaccard(classification_df, "singler_celltype_annotation", methods)
```

Validate marker gene and gene set expression across all three methods

```{r}
plot_markers_df <- create_marker_gene_df(
  sce,
  classification_df, 
  marker_genes_file
)
```

```{r}
# create a density plot showing the distribution of marker gene expression across classification methods 
marker_density_df <- plot_markers_df |>
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "updated_marker_gene_classification" ~ "Marker gene cutoff from 822",
      method == "auc_classification" ~ "AUCell",
      method == "ucell_classification" ~ "UCell",
      method == "blueprint_tumor_classification" ~ "Blueprint",
      method == "updated_singler_classification" ~ "Original SingleR with tumor",
      method == "updated_singler_blueprint_classification" ~ "Tumor + SingleR + Blueprint"
    ),
    classification = dplyr::if_else(classification == "Tumor", "Tumor", "Other")
  )

ggplot(marker_density_df, aes(x = sum_transformed_exp, color = classification)) +
  geom_density() +
  facet_wrap(vars(method))
```

```{r}
# plot gene set scores for each cell  
geneset_plot_df <- classification_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean-")) |> 
  tidyr::pivot_longer(
    cols = starts_with("mean"),
    names_to = "geneset",
    values_to = "mean_score"
  ) |> 
  dplyr::mutate(
    geneset = stringr::word(geneset, -1, sep = "-")
  ) |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "updated_marker_gene_classification" ~ "Marker gene cutoff from 822",
      method == "auc_classification" ~ "AUCell",
      method == "ucell_classification" ~ "UCell",
      method == "blueprint_tumor_classification" ~ "Blueprint",
      method == "updated_singler_classification" ~ "Original SingleR with tumor",
      method == "updated_singler_blueprint_classification" ~ "Tumor + SingleR + Blueprint"
    ),
    classification = dplyr::if_else(classification == "Tumor", "Tumor", "Other")
    )

ggplot(geneset_plot_df, aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(rows = vars(method), 
             cols = vars(geneset))

```

Look at consensus across all three references. 
For normal cells, identify some top markers and see if they are expressed in that group of cells compared to all other cells.

### 2024-07-01 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_singler-classification.Rmd
move heatmaps

@@
 ## Introduction
+
+This notebook looks at using `SingleR` to annotate tumor cells in `SCPCL000824`. 
+We used three different references with `SingleR`:
+
+1. Both the full `BlueprintEncodeData` reference from `celldex` and `SCPCL000822` where cells are either annotated as tumor or normal. 
+Ambiguous cells are removed from `SCPCL000822`. 
+2. `SCPCL000822` where all tumor cells are labeled as tumor and all other cells keep the original annotation obtained from running `SingleR` as part of `scpca-nf`.  
+3. `SCPCL000822` with tumor cells labeled as in #2 combined with the full `BlueprintEncodeData` reference.
 
@@
 
+# gene set scores to use for validation 
 results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000492")
@@
 full_singler_results <- file.path(module_base, "scratch", "SingleR", "SCPCL000824_singler-results.rds")
-singler_annotations_file <- file.path(annotations_dir, "SCPCL000824_singler-classifications.tsv")
+singler_annotations_file <- file.path(annotations_dir, "SCPCL000824_singler-classifications.tsv.gz")
 ```
@@
 annotations_df <- readr::read_tsv(annotations_file)
-singler_annotations_df <- readr::read_tsv(singler_annotations_file)
+singler_annotations_df <- readr::read_tsv(singler_annotations_file) |> unique()
 
@@
 
+## Define functions
+
 ```{r}
-# generate classification df
+# function to plot jaccard matrices to compare one set of annotations to other methods 
+# `annotation_column` will be the columns and will be compared to each method listed in `methods_to_compare` 
+plot_jaccard <- function(classification_df,
+                         annotation_column, # single column of annotations to compare to all methods
+                         methods_to_compare, # list of methods to compare to annotations
+                         column_title, # title for columns
+                         legend_match # what legend to keep, should be a name in `methods_to_compare`
+){
+
+  # create jaccard matrices 
+  jaccard_matrices <- methods_to_compare |>
+    purrr::map(\(name) {
+      make_jaccard_matrix(
+        classification_df,
+        annotation_column,
+        name
+      )
+    })
+
+  heatmap <- make_heatmap_list(jaccard_matrices, column_title, legend_match)
+
+  return(heatmap)
+
+}
+```
+
+```{r}
+# generate classification df to use for plots 
 classification_df <- sce |> 
@@
   dplyr::select(barcodes, UMAP1, UMAP2, singler_celltype_annotation) |> 
+  # join with previous annotations, singler results, and gene set scores 
   dplyr::left_join(annotations_df, by = c("barcodes" = "cell_barcode")) |> 
@@
 
-## SingleR
+## SingleR results
+
+### UMAPs of cell assignments
+
+The first thing we will do is look at the annotated cell types on a UMAP. 
+The top 7 cell types are shown for each reference used and all other cells are lumped together.
 
 ```{r}
-annotations_to_plot <- c(
+# list of columns with singler results for UMAPs
+singler_annotation_columns <- c(
   "blueprint_tumor_classification_lumped",
@@
 
-annotations_to_plot |> 
+singler_annotation_columns |> 
   purrr::map(\(annotation_column){
@@
 
-Compare tumor vs. other in all classification methods vs. updated
+It appears that most cells in that bottom group of cells are being called tumor cells regardless of the reference that is used. 
+The major differences here are in how the normal cells are labeled. 
+For right now, let's focus on just identifying the tumor cells and see how similar each of the references are to each other before we compare to the other methods that we have used.
+
+### Compare tumor cell annotations between references used
 
 ```{r}
-plot_jaccard <- function(classification_df,
-                         singler_results_column,
-                         methods_to_compare
-                         ){
+methods <- c(
+    "blueprint_tumor_classification_lumped",
+    "updated_singler_classification_lumped",
+    "updated_singler_blueprint_classification_lumped"
+  ) |> 
+  purrr::set_names("Blueprint", "Original", "Original and Blueprint")
 
-  # create jaccard matrices for SingleR and CellAssign compared to aneuploid/diploid 
-  jaccard_matrices <- methods_to_compare |>
-    purrr::map(\(name) {
-      make_jaccard_matrix(
-        classification_df,
-        singler_results_column,
-        name
-      )
-    })
+heatmap_list <- methods |>
+  purrr::imap(\(method, name){
 
-  # Set heatmap padding option
-  heatmap_padding <- 0.2
-  ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))
-  # list of heatmaps looking at SingleR/ CellAssign vs tumor/normal 
-  heatmap <- jaccard_matrices |>
-    purrr::imap(
-      \(celltype_mat, celltype_method) {
-        ComplexHeatmap::Heatmap(
-          t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
-          col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
-          border = TRUE,
-          ## Row parameters
-          cluster_rows = TRUE,
-          row_title = celltype_method,
-          row_title_gp = grid::gpar(fontsize = 12),
-          row_title_side = "left",
-          row_names_side = "left",
-          row_dend_side = "right",
-          row_names_gp = grid::gpar(fontsize = 10),
-          ## Column parameters
-          cluster_columns = TRUE,
-          column_title = singler_results_column,
-          column_title_gp = grid::gpar(fontsize = 12),
-          column_names_side = "bottom",
-          column_names_gp = grid::gpar(fontsize = 10),
-          column_names_rot = 90,
-          ## Legend parameters
-          heatmap_legend_param = list(
-            title = "Jaccard index",
-            direction = "vertical",
-            legend_width = unit(1.5, "in")
-          ),
-          show_heatmap_legend = celltype_method == "UCell",
-        )
-      }) |>
-    # concatenate vertically into HeatmapList object
-    purrr::reduce(ComplexHeatmap::`%v%`) |>
-    ComplexHeatmap::draw(
-      heatmap_legend_side = "right",
-      # add a margin to the heatmap so labels don't get cut off
-      padding = unit(c(2, 20, 2, 2), "mm")
+    plot_jaccard(classification_df,
+                 annotation_column = method, # single column of annotations to compare to all methods
+                 methods_to_compare = methods, # list of methods to compare to annotations
+                 column_title = name,
+                 legend_match = "Original" # what legend to keep, should be a name in `methods_to_compare`
     )
 
-  return(heatmap)
+  }) 
+heatmap_list
+```
 
-}
+Upset plot looking at overlapping tumor cells called by all three
+
+```{r}
+tumor_cells <- list(
+  "Blueprint" = classification_df$barcodes[classification_df$blueprint_tumor_classification == "Tumor"],
+  "Original" = classification_df$barcodes[classification_df$updated_singler_classification == "Tumor"],
+  "Original and Blueprint" = classification_df$barcodes[classification_df$updated_singler_blueprint_classification == "Tumor"]
+)
+
+UpSetR::upset(UpSetR::fromList(tumor_cells), order.by = "freq") |> print()
 ```
+
+```{r}
+classification_df <- classification_df |> 
+  dplyr::mutate(singler_consensus_call = dplyr::case_when(
+    blueprint_tumor_classification == "Tumor" & updated_singler_classification == "Tumor" & updated_singler_blueprint_classification == "Tumor" ~ "Tumor",
+    blueprint_tumor_classification == "Tumor" | updated_singler_classification == "Tumor" | updated_singler_blueprint_classification == "Tumor" ~ "Ambiguous",
+    .default = "Normal"
+  ))
+
+ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = singler_consensus_call)) +
+  geom_point(alpha = 0.5, size = 0.5)
+```
+
+Compare normal calls to original SingleR normal calls 
+```{r}
+original_celltypes <- classification_df$singler_celltype_annotation |> unique()
+normal_celltypes <- classification_df |> 
+  dplyr::filter(updated_singler_blueprint_classification_lumped != "Tumor") |> 
+  dplyr::pull(updated_singler_blueprint_classification_lumped) |> 
+  unique()
+
+shared_celltypes <- intersect(original_celltypes, normal_celltypes)
+
+caret_df <- classification_df |> 
+  dplyr::filter(singler_celltype_annotation %in% shared_celltypes |
+                updated_singler_blueprint_classification %in% shared_celltypes)
+
+caret::confusionMatrix(
+    table(
+      caret_df$singler_celltype_annotation, 
+      caret_df$updated_singler_blueprint_classification)
+  )s
+```
+
+### Compare tumor cell annotations with `SingleR` to marker gene based methods
 
@@
 For normal cells, identify some top markers and see if they are expressed in that group of cells compared to all other cells. 
+confusion matrix across all shared normal cell types between OG singleR and new singleR

### 2024-07-01 analyses/cell-type-ewings/scripts/run-singler.R

@@ -1,6 +1,11 @@
 #!/usr/bin/env Rscript
 
 # this script is used to run SingleR using a reference SCE + BlueprintEncode from celldex 
+# SingleR is run three times using different references. 
+# 1. Both the full BlueprintEncodeData reference from celldex and the ref SCE where cells are either annotated as tumor or normal 
+# Ambiguous cells are removed from the reference 
+# 2. Ref SCE where all tumor cells are labeled as tumor and all other cells keep the original annotation obtained from running SingleR as part of scpca-nf 
+# 3. Ref SCE with tumor cells labeled as in #2 combined with the full BlueprintEncodeData reference 
 
 project_root <- here::here()
 renv::load(project_root)
@@ -175,7 +180,8 @@ annotations_df <- singler_results_list |>
     return(df)
     
   }) |> 
-  purrr::reduce(dplyr::inner_join, by = "barcodes")
+  purrr::reduce(dplyr::inner_join, by = "barcodes") |> 
+  unique()
 
 # save results 
 readr::write_rds(singler_results_list, full_singler_output)

### 2024-07-01 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd
move notebooks and tables

--- 2024-06-26 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd
@@
 fs::dir_create(final_annotations_dir)
-final_annotations_file <- file.path(final_annotations_dir, "SCPCL000824_tumor-classifications.tsv")
+final_annotations_file <- file.path(final_annotations_dir, "SCPCL000824_tumor-classifications.tsv.gz")
 

### 2024-07-01 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd
add AUCell density plot and more explanatory text

@@
 - [`UCell`](https://carmonalab.github.io/UCell_demo/UCell_matrix_vignette.html#content) with marker genes and three EWS-FLI1 gene sets in both samples.
+
+Throughout this notebook, we will use `SCPCL000822` as a reference. 
+We have classified tumor cells to be those identified as tumor cells by both `InferCNV` and `CopyKAT` and validated those classifications in `SCPCL000822_tumor-cell-validation.Rmd`.
 
@@
 ref_sce_file <- file.path(data_dir, "SCPCP000015", "SCPCS000490", "SCPCL000822_processed.rds")
-ref_labels_file <- file.path(module_base, "cell_type_annotations", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv")
+ref_labels_file <- file.path(module_base, "results", "annotation_tables", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv")
 ref_geneset_scores_file <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
@@
 With `SCPCL000822` we had a clear separation between marker gene expression in tumor cells and normal cells because there was a bimodal distribution. 
-It does not appear that is the case here so it's going to be more difficult to pull out tumor cells.
+That does not appear to be the case here so it's going to be more difficult to pull out tumor cells.
 
@@
 To do this, we will get the total marker gene expression by summing all marker genes in a cell and then plot the distribution.
-The first plot will look at the distribution across all cells and the second will show the classification of tumor vs. normal cells.
 
@@
   facet_grid(rows = vars(sample))
-
-# color by tumor cell classification 
-ggplot(combined_markers_df, aes(x = sum_raw_exp, color = tumor_cell_classification)) +
-  geom_density() +
-  facet_grid(rows = vars(sample))
 ```
 
-Looking at this, we see that `SCPCL000822` has a bimodal distribution and shows a clear cutoff where cells are tumor vs. normal, but this is not the case for `SCPCL000824`. 
+Looking at this, we see that `SCPCL000822` has a bimodal distribution, but this is not the case for `SCPCL000824`. 
+Additionally, most of the distribution for `SCPCL000824` lies within the upper distribution for `SCPCL000822`. 
+This would be consistent with our hypothesis that most of the cells in `SCPCL000824` are tumor cells. 
 Let's find the local minima in the bimodal distribution for `SCPCL000822` and then use that to classify tumor cells in `SCPCL000824`.
@@
 Now we will look at the gene set scores for each gene set in `SCPCL000822` and `SCPCL000824`.  
+The gene set scores correspond to the mean expression of all genes in a given gene set for a single cell with no scaling. 
+The normalized gene expression was used as input to calculating the mean.
+
 We did not use these to classify `SCPCL000822`, so we won't actually do any classification, but this will show us if the scores for tumor cells are similar to each other across samples.
@@
 Again we see that the gene set scores are bimodal for `SCPCL000822`, at least for `RIGGI` and `ZHANG`. 
-However, we don't see that for `SCPCL000824`, but we do see a very similar ranger of values. 
+However, we don't see that for `SCPCL000824`, but we do see a very similar range of values. 
 This makes me think that the majority of the cells are in fact tumor cells.
@@
 We will use [`AUCell`](https://www.bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html) to do this and will run it on both samples.
-However, `AUCell` relies on having a bimodal distribution in gene set expression, so I expect we may need to do a similar thing where we use cutoffs determined in `SCPCL000822` to call cells in `SCPCL000824`.
+
+`AUCell` can be used to identify cells with an active gene set given a list of genes or gene signature. 
+From the [vignette](https://bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html):
+
+> AUCell uses the “Area Under the Curve” (AUC) to calculate whether a critical subset of the input gene set is enriched within the expressed genes for each cell.
+The distribution of AUC scores across all the cells allows exploring the relative expression of the signature.
+Since the scoring method is ranking-based, AUCell is independent of the gene expression units and the normalization procedure.
+In addition, since the cells are evaluated individually, it can easily be applied to bigger datasets, subsetting the expression matrix if needed.
+
+We will use `AUCell` with four different gene sets:
+
+- [`ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION`](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION.html)
+- [`RIGGI_EWING_SARCOMA_PROGENITOR_UP`](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/RIGGI_EWING_SARCOMA_PROGENITOR_UP.html?ex=1)
+- [`SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN`](https://www.gsea-msigdb.org/gsea/msigdb/cards/SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN)
+- Custom tumor marker gene list found in [`references/tumor-marker-genes.tsv`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/references/tumor-marker-genes.tsv)
+
+For each gene set above, `AUCell` will identify cells likely to have expression of that gene set returning a distribution of `auc` scores and a threshold for each gene set.
+
+One caveat is that `AUCell` relies on having a bimodal distribution in gene set expression, so I expect we may need to do a similar thing where we use cutoffs determined in `SCPCL000822` to call cells in `SCPCL000824`.
 
@@
 
+For each gene set, several possible thresholds are calculated. 
+The dotted lines over the histogram indicate the distribution and the corresponding vertical dotted line (same color) indicates the possible threshold. 
+The thicker line indicates the chosen threshold by the algorithm. 
+This corresponds to the highest value that reduces the false positives.
+
 Here we can see a bimodal distribution of all gene sets except `SILIGAN`, which is to be expected based on the original distribution plots.
-We also see that `AUCell` identifies a threshold based on the local minima of the AUC distribution where all cells above the threshold are classified as expressing the gene set.
-Let's compare the assignments from our validated classifications to using `AUCell` with `SCPCL000822`.
+
+For the remaining analysis we will use the results from running `AUCell` with only the marker gene list. 
+Let's compare the assignments from our validated classifications (determined by taking the consensus between `InferCNV`, and `CopyKAT` for `SCPCL000822`) to using `AUCell` with `SCPCL000822`.
 
@@
 
-Instead, we will use the threshold identified by `AUCell` for `SCPCL000822` to identify tumor cells in `SCPCL000824`.  
+```{r}
+ref_auc_df <- ref_auc@assays@data$AUC |> 
+  as.data.frame() |> 
+  tibble::rownames_to_column("gene_set") |> 
+  tidyr::pivot_longer(!gene_set, 
+                      names_to = "barcodes", 
+                      values_to = "auc") |> 
+  #dplyr::filter(gene_set == "markers") |> 
+  dplyr::mutate(sample = "SCPCL000822")
+
+new_auc_df <- new_auc@assays@data$AUC |> 
+  as.data.frame() |> 
+  tibble::rownames_to_column("gene_set") |> 
+  tidyr::pivot_longer(!gene_set, 
+                      names_to = "barcodes", 
+                      values_to = "auc") |> 
+  #dplyr::filter(gene_set == "markers") |> 
+  dplyr::mutate(sample = "SCPCL000824")
+
+all_auc_df <- dplyr::bind_rows(list(ref_auc_df, new_auc_df))
+
+ggplot(all_auc_df, aes(x = auc, color = sample)) +
+  geom_density() +
+  facet_grid(rows = vars(gene_set))
+```
+
+Looking at the AUC across both samples for all gene sets, we see that they are in the same range of values and tend to mirror each other. 
+The main difference here is the lack of bimodal distribution in `SCPCL000824`.
+
+Let's use the threshold identified by `AUCell` for `SCPCL000822` to identify tumor cells in `SCPCL000824`.  
 Here, we label any cells that have passed the threshold set for the marker gene sets as tumor cells.
@@
 # create a vector of tumor cells in 824 using the cutoff from 822
-new_tumor_cells <- new_auc@assays@data$AUC |> 
-  as.data.frame() |> 
-  tibble::rownames_to_column("gene_set") |> 
-  tidyr::pivot_longer(!gene_set, 
-                      names_to = "barcodes", 
-                      values_to = "auc") |> 
+new_tumor_cells <- new_auc_df |> 
   dplyr::filter(gene_set == "markers", auc >= ref_threshold) |>
@@
 This looks similar to using the updated marker gene cutoff to classify cells where most cells in that bottom group are tumor cells.  
+As a reminder, the updated marker gene classification is the classification of tumor cells based on the marker gene expression cutoff determined in `SCPCL000822`. 
+We can confirm this by calculating the confusion matrix.
+
+```{r}
+# compare using a confusion matrix
+caret::confusionMatrix(
+    table(
+      classification_df$updated_marker_gene_classification, 
+      classification_df$auc_classification)
+  ) 
+```
 
@@
 Here we will look at using [`UCell`](https://carmonalab.github.io/UCell_demo/UCell_matrix_vignette.html#content) to calculate a gene set score for each cell and then attempt to classify cells based on the distribution of those gene set scores.
+
+The [`UCell` score is calculated as follows](https://doi.org/10.1016/j.csbj.2021.06.043):
+
+1. Calculate a ranked list of genes for each cell in the dataset. 
+2. The Mann-Whitney U statistic is calculated for each cell by subsetting the ranked list by genes of interest in each cell.
+
 Again, we will look at both `SCPCL000822` and `SCPCL000824`.
@@
 # create list to use for ucell of gene sets and marker genes
-ucell_gene_sets <- c(genes_list, markers = list(marker_genes))
+ucell_gene_sets <- c(genes_list, markers = list(marker_genes)) |> 
+  # only keep genes that are found in reference 
+  purrr::map(\(geneset){
+    intersect(geneset, rownames(sce))
+  })
 
@@
   geom_density() +
-  facet_grid(rows = vars(gene_list))
+  facet_grid(rows = vars(gene_list),
+             scales = "free_y")
 ```
 
-Here we see that the marker gene set maybe has a small bimodal distribution in `SCPCL000822`, but it is very small and not quite as striking as looking at the scores from `AUCell` or raw gene expression.
+Here we see that the marker gene set maybe has a bimodal distribution in `SCPCL000822` and no bimodal distribution in `SCPCL000824`. 
+However the score in `SCPCL000824` seems to lie in the upper part of the distribution for `SCPCL000822`.
+
 `RIGGI`, `ZHANG`, and `SILIGAN` both show very similar distributions to the gene set scores distributions from our manual calculations of gene set scores (mean of all genes in the gene set). 
@@
 
-Let's use the local minima from the `RIGGI` gene set scores for `SCPCL000822` to identify tumor cells. 
-First we will look at how using `UCell` compares to our previous classifications for `SCPCL000822`.
+Let's use the local minima from the marker gene set scores for `SCPCL000822` to identify tumor cells. 
+First we will look at how using `UCell` compares to our previous classifications for `SCPCL000822` (consensus between `CopyKAT` and `InferCNV`).
 
@@
 # create distribution 
-density_data <- density(ref_ucell$RIGGI_EWING_SARCOMA_PROGENITOR_UP_UCell)
+density_data <- density(ref_ucell$markers_UCell)
 # find the local minima in the distribution
@@
 ref_ucell_tumor_cells <- ucell_df |> 
-  dplyr::filter(gene_list == "RIGGI",
+  dplyr::filter(gene_list == "markers",
                 signature_score >= geneset_cutoff,
@@
 new_ucell_tumor_cells <- ucell_df |> 
-  dplyr::filter(gene_list == "RIGGI",
+  dplyr::filter(gene_list == "markers",
                 signature_score >= geneset_cutoff,
@@
 
-Here we see less cells classified as tumor cells than we saw with the `AUCell` classification. 
-This also looks very similar to the original UMAP when we classified cells using marker gene expression.
+This looks similar to what we get with both the updated marker gene and `AUCell` classification. 
+We can confirm that by calculating the confusion matrix.
+
+```{r}
+caret::confusionMatrix(
+  table(
+    classification_df$updated_marker_gene_classification, 
+    classification_df$ucell_classification
+  )
+)
+```
 

### 2024-07-01 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd
rendered notebook

@@
 # output files 
-final_annotations_dir <- file.path(module_base, "cell_type_annotations", "SCPCS000492")
+final_annotations_dir <- file.path(module_base, "results", "annotation_tables", "SCPCS000492")
 fs::dir_create(final_annotations_dir)

### 2024-07-01 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd

--- 2024-06-17 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
@@
 ---
-title: "SCPCL000822 - Validation of tumor cells"
+title: "`SCPCL000822` - Validation of tumor cells"
 author: Ally Hawkins
@@
   - Looking at marker gene expression. 
-  - Looking at expression of EWS-FLI1 target gene sets. 
+  - Looking at expression of `EWS-FLI1` target gene sets. 
   - Confirming presence of known CNVs in Ewing sarcoma
@@
 
-Next, we will look at the gene set scores for three different EWS-FLI1 gene sets. 
+Next, we will look at the gene set scores for three different `EWS-FLI1` gene sets. heirarchical
 The below plots will all use the `mean-GENESET` columns.
@@
 Here we see a gain in Chr8 and Chr1 in tumor cells compared to normal cells. 
-We also see other chromosomes with amplifications in tumor cells compared to normal cells and that the tumor and normal cells are nicely separated in the heirarchical clustering.
+We also see other chromosomes with amplifications in tumor cells compared to normal cells and that the tumor and normal cells are nicely separated in the hierarchical clustering.
 

### 2024-07-02 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_singler-classification.Rmd

--- 2024-07-01 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_singler-classification.Rmd
@@
 
-## Define functions
-
-```{r}
-# function to plot jaccard matrices to compare one set of annotations to other methods 
-# `annotation_column` will be the columns and will be compared to each method listed in `methods_to_compare` 
-plot_jaccard <- function(classification_df,
-                         annotation_column, # single column of annotations to compare to all methods
-                         methods_to_compare, # list of methods to compare to annotations
-                         column_title, # title for columns
-                         legend_match # what legend to keep, should be a name in `methods_to_compare`
-){
-
-  # create jaccard matrices 
-  jaccard_matrices <- methods_to_compare |>
-    purrr::map(\(name) {
-      make_jaccard_matrix(
-        classification_df,
-        annotation_column,
-        name
-      )
-    })
-
-  heatmap <- make_heatmap_list(jaccard_matrices, column_title, legend_match)
-
-  return(heatmap)
-
-}
-```
-
 ```{r}
@@
   "updated_singler_blueprint_classification_lumped"
-)
+) |> 
+  purrr::set_names("Blueprint", "Original", "Original + Blueprint")
 
@@
 
-```{r}
-methods <- c(
-    "blueprint_tumor_classification_lumped",
-    "updated_singler_classification_lumped",
-    "updated_singler_blueprint_classification_lumped"
-  ) |> 
-  purrr::set_names("Blueprint", "Original", "Original and Blueprint")
-
-heatmap_list <- methods |>
-  purrr::imap(\(method, name){
-
-    plot_jaccard(classification_df,
-                 annotation_column = method, # single column of annotations to compare to all methods
-                 methods_to_compare = methods, # list of methods to compare to annotations
-                 column_title = name,
-                 legend_match = "Original" # what legend to keep, should be a name in `methods_to_compare`
-    )
-
-  }) 
-heatmap_list
-```
-
-Upset plot looking at overlapping tumor cells called by all three
+First we will just look at the overlap in the tumor cells between all three methods.
 
@@
 ```
+
+Let's see where the shared cells are and if they are where we would expect them to be based on marker gene expression plots we made previously (bottom right group of cells).
 
@@
 
-Compare normal calls to original SingleR normal calls 
+It looks like the cells that are normal are in the smaller clusters and the larger cluster is mostly tumor cells with some that are ambiguous across `SingleR` references.
+
+### Compare normal cell annotations between references used
+
+Next we will compare all of the top cell types identified in each method to the top cell types identified in the other methods. 
+This will show us the overlap in tumor cells and normal cell types identified.
+
+```{r, fig.height=7}
+methods <- c(
+  "updated_singler_classification_lumped",
+  "updated_singler_blueprint_classification_lumped"
+) |> 
+  purrr::set_names("Original", "Original and Blueprint")
+
+# compare blueprint to original and origintal + blueprint
+plot_jaccard(classification_df,
+             annotation_column = "blueprint_tumor_classification_lumped",
+             methods_to_compare = methods,
+             column_title = "Blueprint",
+             legend_match = "Original") |> 
+  ComplexHeatmap::draw(
+    heatmap_legend_side = "right",
+    # add a margin to the heatmap so labels don't get cut off
+    padding = unit(c(2, 20, 2, 2), "mm")
+  )
+
+# compare original to original + blueprint
+plot_jaccard(classification_df,
+             annotation_column = "updated_singler_classification_lumped",
+             methods_to_compare = c("Original and Blueprint" = "updated_singler_blueprint_classification_lumped"),
+             column_title = "Original",
+             legend_match = "Original")
+
+```
+
+- There seems to be mostly consensus across all three methods on which cells are tumor cells. 
+- When only using Blueprint, a lot of cells are categorized as "Normal". 
+This means they match the non-tumor cells identified in `SCPCL000822` more than they match the `Blueprint` reference. 
+Additionally, with only Blueprint, there is a group of cells that are called as tumor but actually appear to match endothelial cells present in `SCPCL000822`.
+- The cells that aren't called as either Tumor or Normal in the Blueprint only reference mostly match up to only `erythrocyte` in the reference that just uses `SCPCL000822`. 
+These same cells match to their corresponding Blueprint cell type when combining `SCPCL000822` and Blueprint (e.g., fibroblasts, neurons, and epithelial cells are identifed in both). 
+- When comparing using just `SCPCL000822` to `SCPCL000822` and Blueprint, we see a overlap in endothelial cells and chrondrocytes. 
+We also see that for the most part the top cell types are the same except when using blueprint you see the addition of epithelial cells.
+
+Next, we will measure how well the updated classification matches the original `SingleR` class identified in `SCPCL000824`.
+
 ```{r}
-original_celltypes <- classification_df$singler_celltype_annotation |> unique()
-normal_celltypes <- classification_df |> 
-  dplyr::filter(updated_singler_blueprint_classification_lumped != "Tumor") |> 
-  dplyr::pull(updated_singler_blueprint_classification_lumped) |> 
+singler_annotation_columns |> 
+  purrr::map(\(classification_column){
+
+    # first get list of tumor cells for this classificationn
+    # we want to remove these since we are only interested in comparing normal cell types
+    tumor_cells <- classification_df |> 
+      dplyr::filter({{classification_column}} == "Tumor") |> 
+      dplyr::pull(barcodes)
+
+    # get list of shared cell types with SingleR
+    original_celltypes <- caret_df$singler_celltype_annotation |> 
+      unique()
+    new_celltypes <- caret_df[[classification_column]] |> 
       unique()
 
-shared_celltypes <- intersect(original_celltypes, normal_celltypes)
+    shared_celltypes <- intersect(original_celltypes, new_celltypes)
 
-caret_df <- classification_df |> 
-  dplyr::filter(singler_celltype_annotation %in% shared_celltypes |
-                updated_singler_blueprint_classification %in% shared_celltypes)
+    # don't filter, just label as tumor? or other normal?  
+    caret_df <- caret_df |> 
+      # first remove tumor cells 
+      dplyr::filter(!barcodes %in% tumor_cells) |> 
+      # then update to make sure all cell types are shared
+      # anything not in shared cell types are classified as "other" 
+      dplyr::mutate(singler_caret = dplyr::if_else(singler_celltype_annotation %in% shared_celltypes, singler_celltype_annotation, "Other"),
+                    new_ref_caret = dplyr::if_else(!!sym(classification_column) %in% shared_celltypes, !!sym(classification_column), "Other"))
 
@@
       table(
-      caret_df$singler_celltype_annotation, 
-      caret_df$updated_singler_blueprint_classification)
-  )s
+        caret_df$singler_caret, 
+        caret_df$new_ref_caret)
+    )
+
+  })
 ```
+
+Here we see that the highest Kappa, although still pretty low, is with using the original `SingleR` annotations from `SCPCL000822` and use of Blueprint.
 
@@
 
-```{r}
+```{r fig.height=7}
  original_methods <- c(
@@
 
-heatmap_list <- annotations_to_plot |> 
-  purrr::walk(\(annotation_column) {plot_jaccard(classification_df, annotation_column, original_methods)})
+# heatmap comparing each singler annotation type to each of the marker gene based methods 
+heatmap_list <- singler_annotation_columns |> 
+  purrr::iwalk(\(annotation_column, name) {
+
+    # first make singler annotations normal or tumor for easier visualization
+    plot_df <- classification_df |> 
+      dplyr::mutate(new_annotation = dplyr::if_else(!!sym(annotation_column) == "Tumor", "Tumor", "Normal"))
+
+    plot_jaccard(
+      classification_df = plot_df, 
+      annotation_column = "new_annotation", 
+      methods_to_compare = original_methods,
+      column_title = name,
+      legend_match = "UCell"
+    ) |> 
+      ComplexHeatmap::draw(
+        heatmap_legend_side = "right",
+        # add a margin to the heatmap so labels don't get cut off
+        padding = unit(c(2, 20, 2, 2), "mm")
+      )
+  })
 ```
 
-```{r}
-methods <- c(
-    "blueprint_tumor_classification_lumped",
-    "updated_singler_classification_lumped",
-    "updated_singler_blueprint_classification_lumped"
-  ) |> 
-  purrr::set_names("Blueprint", "Original", "Original and Blueprint")
+Generally, these methods agree on which cells are tumor cells.
 
-heatmap_list <- plot_jaccard(classification_df, "singler_celltype_annotation", methods)
-```
+It looks like using Blueprint alone is the most similar to all marker based methods. 
+However, it looks like `UCell` categorizes more cells as normal that are classified as tumor with any of the `SingleR` methods.
 
-Validate marker gene and gene set expression across all three methods
+For both original and original and blueprint, it looks like there is a good amount of cells that `SingleR` classifies as normal, but the marker gene based methods classify as tumor.
+
+### Validate marker gene expression in `SingleR` annotations
+
+The last thing we will do is confirm that the cells that are identified as tumor cells with `SingleR` have increased expression of marker genes and higher gene set scores. 
+We would expect this to be the case, but we also want to see if we are missing calling any tumor cells that have higher gene expression.
+
+For these plots we will also look at all the marker based classifications so we can compare across all options.
 
 ```{r}
+# first get marker gene expression 
 plot_markers_df <- create_marker_gene_df(
@@
   geom_density() +
-  facet_wrap(vars(method))
+  facet_wrap(vars(method)) +
+  labs(
+    title = "Marker gene expression",
+    x = "Sum of marker genes"
+  )
 ```
 
-```{r}
+```{r fig.height=10}
 # plot gene set scores for each cell  
@@
   facet_grid(rows = vars(method), 
-             cols = vars(geneset))
+             cols = vars(geneset)) +
+  labs(
+    title = "Gene set scores",
+    x = "Mean gene set score"
+  )
 
@@
 
-Look at consensus across all three references. 
-For normal cells, identify some top markers and see if they are expressed in that group of cells compared to all other cells. 
-confusion matrix across all shared normal cell types between OG singleR and new singleR
+Looking at these plots, generally Blueprint only has the best separation between tumor cells and all other cells for both gene set scores and marker gene expression. 
+This also seems to mirror the plots shown with `AUCell` as the classifier.
+
+It looks like we may be mis-classifying some cells when using the original `SingleR` assignments for SCPCL000822 with or without Blueprint. 
+There is some overlap in the cells that are identified as tumor and other non-tumor cells.
+
+## Conclusions
+
+## Session Info
+
+```{r session info}
+# record the versions of the packages used in this analysis and other environment information
+sessionInfo()
+```

### 2024-07-02 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 Now we will look at the gene set scores for each gene set in `SCPCL000822` and `SCPCL000824`.  
-The gene set scores correspond to the mean expression of all genes in a given gene set for a single cell with no scaling. 
-The normalized gene expression was used as input to calculating the mean.
+The gene set scores for each cell are the mean normalized expression of all genes in a given gene set with no scaling.
 
@@
 
-One caveat is that `AUCell` relies on having a bimodal distribution in gene set expression, so I expect we may need to do a similar thing where we use cutoffs determined in `SCPCL000822` to call cells in `SCPCL000824`.
+One caveat is that `AUCell` relies on having a bimodal distribution in gene set expression to find a threshold, so I expect we may need to do a similar thing where we use thresholds determined in `SCPCL000822` to call cells in `SCPCL000824`.
 
@@
 
-# assign cells based on threshold and output plots 
-ref_assignment <- AUCell::AUCell_exploreThresholds(ref_auc, assign = TRUE)
+# assign cells for each gene set based on thresholds and output plots 
+ref_assignments <- AUCell::AUCell_exploreThresholds(ref_auc, assign = TRUE)
 ```
@@
 
-Let's use the threshold identified by `AUCell` for `SCPCL000822` to identify tumor cells in `SCPCL000824`.  
-Here, we label any cells that have passed the threshold set for the marker gene sets as tumor cells.
+Let's use the marker gene threshold identified by `AUCell` for `SCPCL000822` to identify tumor cells in `SCPCL000824`.  
+Here, we label any cells that have passed the threshold set for the marker gene set as tumor cells.
 
@@
 It looks like we get pretty good agreement between using `UCell` with the `RIGGI` gene set to the original classification. 
-Now we will use the cutoff from `SCPCL000822` to classify tumor cells in `SCPCL000824`.
+Now we will use the marker gene set cutoff from `SCPCL000822` to classify tumor cells in `SCPCL000824`.
 

### 2024-07-02 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_singler-classification.Rmd
move annotations

@@
 # annotations from SCPCL000824_tumor-gene-expression-exploration.Rmd
-annotations_dir <- file.path(module_base, "cell_type_annotations", "SCPCS000492")
+annotations_dir <- file.path(module_base, "results", "annotation_tables", "SCPCS000492")
 annotations_file <- file.path(annotations_dir, "SCPCL000824_tumor-classifications.tsv.gz")
@@
 annotations_df <- readr::read_tsv(annotations_file)
-singler_annotations_df <- readr::read_tsv(singler_annotations_file) |> unique()
+singler_annotations_df <- readr::read_tsv(singler_annotations_file)
 

### 2024-07-02 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_singler-classification.Rmd
add explanations and conclusions

@@
 
-1. Both the full `BlueprintEncodeData` reference from `celldex` and `SCPCL000822` where cells are either annotated as tumor or normal. 
+1. `Blueprint`: Both the full `BlueprintEncodeData` reference from `celldex` and `SCPCL000822` where cells are either annotated as tumor or normal. 
 Ambiguous cells are removed from `SCPCL000822`.
-2. `SCPCL000822` where all tumor cells are labeled as tumor and all other cells keep the original annotation obtained from running `SingleR` as part of `scpca-nf`.  
-3. `SCPCL000822` with tumor cells labeled as in #2 combined with the full `BlueprintEncodeData` reference.
+2. `Original`: `SCPCL000822` where all tumor cells are labeled as tumor and all other cells keep the original annotation obtained from running `SingleR` as part of `scpca-nf`.  
+3. `Original and Blueprint`: `SCPCL000822` with tumor cells labeled as in #2 combined with the full `BlueprintEncodeData` reference.
 
@@
 
-```{r}
+```{r, warning=FALSE}
 # generate classification df to use for plots 
@@
         x |>
+          # there are warnings because we have some that have less than 7 cell types (ones that only did tumor/normal)
+          # this is why we ignore warnings in this chunk 
           forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
@@
 singler_annotation_columns |> 
-  purrr::map(\(annotation_column){
+  purrr::imap(\(annotation_column, title){
     ggplot(
@@
       color = guide_legend(
-        title = "Cell types",
+        title = title,
         # more visible points in legend
@@
 The major differences here are in how the normal cells are labeled. 
-For right now, let's focus on just identifying the tumor cells and see how similar each of the references are to each other before we compare to the other methods that we have used.
+For right now, let's focus on just identifying the tumor cells and see how similar each of the references are to each other before we look into normal cells and compare to the other methods that we have used.
 
-### Compare tumor cell annotations between references used
+### Compare tumor cell annotations between between `SingleR` references used
 
@@
 ```
+Generally, it looks like there is a large group of cells that are identified as tumor regardless of the reference, with a few cells that are specifically tumor in only one or two references.
 
@@
 It looks like the cells that are normal are in the smaller clusters and the larger cluster is mostly tumor cells with some that are ambiguous across `SingleR` references. 
+This is what I would expect given the marker gene expression in these cells.
 
@@
 
-- There seems to be mostly consensus across all three methods on which cells are tumor cells. 
+A few things that we see in these plots:
+
+- There seems to be mostly consensus across all three methods on which cells are tumor cells, which is consistent with the upset plot. 
 - When only using Blueprint, a lot of cells are categorized as "Normal". 
-This means they match the non-tumor cells identified in `SCPCL000822` more than they match the `Blueprint` reference. 
+This means they match the non-tumor cells identified in `SCPCL000822` more than they match the normal cell types in the `Blueprint` reference. 
 Additionally, with only Blueprint, there is a group of cells that are called as tumor but actually appear to match endothelial cells present in `SCPCL000822`.
-- The cells that aren't called as either Tumor or Normal in the Blueprint only reference mostly match up to only `erythrocyte` in the reference that just uses `SCPCL000822`. 
+- The cells that aren't called as either Tumor or Normal in the Blueprint only reference mostly match up to `erythrocyte` in the reference that just uses `SCPCL000822` (original). 
 These same cells match to their corresponding Blueprint cell type when combining `SCPCL000822` and Blueprint (e.g., fibroblasts, neurons, and epithelial cells are identifed in both). 
-- When comparing using just `SCPCL000822` to `SCPCL000822` and Blueprint, we see a overlap in endothelial cells and chrondrocytes. 
-We also see that for the most part the top cell types are the same except when using blueprint you see the addition of epithelial cells.
+- When comparing using just `SCPCL000822` (original) to `SCPCL000822` and Blueprint (original and blueprint), we see an overlap in endothelial cells and chrondrocytes. 
+We also see that for the most part the top cell types identified are the same except when using blueprint you see the addition of epithelial cells.
 
 Next, we will measure how well the updated classification matches the original `SingleR` class identified in `SCPCL000824`. 
+We will create a confusion matrix between each of these `SingleR` references and the original `SingleR` classifications from `scpca-nf`. 
+Note that we will not include tumor cells here and will only look at shared cell types that are identified in both. 
+Any cells that are not in the shared cell types are classified as "other".
+
+For all matrices, the rows are the original `SingleR` annotations and the columns are from the new annotations.
 
@@
     # get list of shared cell types with SingleR
-    original_celltypes <- caret_df$singler_celltype_annotation |> 
+    original_celltypes <- classification_df$singler_celltype_annotation |> 
       unique()
-    new_celltypes <- caret_df[[classification_column]] |> 
+    new_celltypes <- classification_df[[classification_column]] |> 
       unique()
@@
 
-    # don't filter, just label as tumor? or other normal?  
-    caret_df <- caret_df |> 
+    caret_df <- classification_df |> 
       # first remove tumor cells 
@@
 
-Here we see that the highest Kappa, although still pretty low, is with using the original `SingleR` annotations from `SCPCL000822` and use of Blueprint.
+Here we see that the highest Kappa, although still very low, is with using the original `SingleR` annotations from `SCPCL000822` and use of Blueprint. 
+Generally, I would say there is not a lot of agreement with the original `SingleR` annotations. 
+Identifying the correct classes of the "normal" cells I think is out of scope for this notebook and it is ultimately more important to identify a way to label tumor cells.
+
+I tend to favor using one of these options:
+
+- `Blueprint`: `SCPCL000822` with cells labeled as either tumor or normal in combination with Blueprint. 
+- `Original and Blueprint`: `SCPCL000822` with tumor cells labeled and all other cells labeled with the original `SingleR` annotations in combination with `Blueprint`.
+
+I don't like using just the `Original`, where the only reference comes from a single sample, `SCPCL000822`. 
+It's very likely that the types of normal cells will differ across samples, so I think we want to be able to use `Blueprint` to help classify any of those cells.
+
+Another option would be to try running with just the tumor cells from `SCPCL000822` as a reference in combination with `Blueprint`, removing the "normal" cells from the reference. 
+But either of the above options I think would be sufficient to labeling and pulling out tumor cells. 
+Let's confirm this by comparing to our previously used marker gene based methods for classifying this sample.
 
 ### Compare tumor cell annotations with `SingleR` to marker gene based methods
+
+Below we will compare the annotations obtained using the different references and `SingleR` to the marker gene based annotations identified in `SCPCL000824_tumor-gene-expression-exploration.Rmd`. 
+First we will look at the Jaccard similarity between each of the `SingleR` references and all of the marker gene based methods. 
+All cells will be categorized as either tumor or normal.
 
@@
 For both original and original and blueprint, it looks like there is a good amount of cells that `SingleR` classifies as normal, but the marker gene based methods classify as tumor. 
+This tells me that using `SingleR` may be slightly more conservative than the marker gene based methods.
 
@@
       method == "marker_gene_classification" ~ "Marker genes only",
-      method == "updated_marker_gene_classification" ~ "Marker gene cutoff from 822",
+      method == "updated_marker_gene_classification" ~ "Marker genes from 822",
       method == "auc_classification" ~ "AUCell",
@@
       method == "blueprint_tumor_classification" ~ "Blueprint",
-      method == "updated_singler_classification" ~ "Original SingleR with tumor",
-      method == "updated_singler_blueprint_classification" ~ "Tumor + SingleR + Blueprint"
+      method == "updated_singler_classification" ~ "Original",
+      method == "updated_singler_blueprint_classification" ~ "Original + Blueprint"
     ),
@@
   geom_density() +
-  facet_wrap(vars(method)) +
+  facet_wrap(vars(method), 
+             scales = "free_y") +
   labs(
@@
   facet_grid(rows = vars(method), 
-             cols = vars(geneset)) +
+             cols = vars(geneset),
+             scales = "free_y") +
   labs(
@@
 It looks like we may be mis-classifying some cells when using the original `SingleR` assignments for SCPCL000822 with or without Blueprint. 
-There is some overlap in the cells that are identified as tumor and other non-tumor cells.
+There is some overlap in the marker gene and gene set scores for the cells that are identified as tumor and other non-tumor cells for both original and original and blueprint.
 
 ## Conclusions
+
+- For the most part, all `SingleR` references identify a similar set of tumor cells. 
+However there are some cells that are only identified in one or two methods making the full set of tumor cells unique across references. 
+- All three references identify the same top normal cell types with some overlapping sets. 
+- There does not appear to be a consensus between the original `SingleR` annotations and the new annotations for any of the references. 
+- Using `SCPCL000822` with tumor and normal labels in combination with `Blueprint` seems to be the most consistent with using the marker gene based methods for calling tumor cells. 
+Additionally, this reference showed the greatest distinction between tumor cells and all other cells in marker gene expression and gene set scores.
 

### 2024-07-02 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
Apply suggestions from code review
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

@@
 
-Next, well explore the consensus of doublet predictions across methods. 
+Next, we will explore the consensus of doublet predictions across methods. 
 To this end, we'll create some new columns for each dataset:
@@
       pc_color_columns <- c("scDblFinder prediction" = "scdbl_prediction",
-                            "scublet prediction" = "scrublet_prediction",
+                            "scrublet prediction" = "scrublet_prediction",
                             "cxds prediction" = "cxds_prediction",

### 2024-07-02 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
move vectors out of maps, and use one mutate

@@
         dplyr::rowwise() |>
+
+        dplyr::mutate(
           # Add column `consensus_call`
           # "doublet" if all three methods are doublet, and "singlet" otherwise          
-        dplyr::mutate(
           consensus_call = dplyr::if_else(
@@
             "singlet"
-          )
-        ) |>
+          ),
           # Add column `consensus_class`
@@
           #  and "ambiguous" if there are any disagreements
-        dplyr::mutate(
           consensus_class = dplyr::case_when(
@@
             .default = "ambiguous"
-          )
-         ) |>
+          ), 
           # Add column `confusion_call`
           #  tp/tn/fp/fn based on the `consensus_call` column
-        dplyr::mutate(
           confusion_call = dplyr::case_when(
@@
 ```{r, fig.width = 12, fig.height = 6}
-doublet_df_list |>
-  purrr::imap(
-    \(df, dataset) {
-
 pc_color_columns <- c("scDblFinder prediction" = "scdbl_prediction",
@@
                       "Consensus call" = "consensus_call")
+doublet_df_list |>
+  purrr::imap(
+    \(df, dataset) {
 
@@
                       "Consensus call" = "consensus_call")
+
+doublet_df_list |>
+  purrr::imap(
+    \(df, dataset) {
 

### 2024-07-02 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
restore the metric PCA plots, and make legend points bigger all around

@@
     ) +
-    ggtitle(title)
+    ggtitle(title) +
+    guides(color = guide_legend(override.aes = list(size = 2)))
 
@@
       scale_color_manual(
-        name = "", 
-        values = c("ambiguous" = "lightblue", 
+        name = "Consensus", 
+        values = c("ambiguous" = "yellow", 
                    "doublet" = "black", 
@@
   return(plot)
+}
+
+plot_pca_metrics <- function(df, color_column) {
+  # Plot PCs colored by performance metric, showing false calls on top
+  # metric_colors is a named vector of colors used for coloring tp/tn/fp/fn
+
+  # used in PCA plots
+  metric_colors <- c(
+    "tp" = "lightblue",
+    "tn" = "pink",
+    "fp" = "blue",
+    "fn" = "firebrick2"
+  )
+
+  ggplot(df) + 
+    aes(x = PC1, 
+        y = PC2, 
+        color = {{color_column}}) +
+  geom_point(
+    size = 0.75, 
+    alpha = 0.6
+  ) + 
+  geom_point(
+    data = dplyr::filter(df, {{color_column}} %in% c("fp", "fn")), 
+    size = 0.75
+  ) +
+  scale_color_manual(name = "Call type", values = metric_colors) +
+  ggtitle("Consensus class metrics") +
+  guides(color = guide_legend(override.aes = list(size = 2)))
+
 }
@@
 6. Consensus _class_: droplets are either "doublet", "singlet", or "ambiguous"
+7. Metrics based on the consensus call
 
 - For the first five panels, doublets are shown in black and singlets in light gray.
-- In the sixth panel showing the consensus class, doublets are shown in black, singlets in light gray, and ambiguous are blue.
+- In the sixth panel showing the consensus class, doublets are shown in black, singlets in light gray, and ambiguous are yellow
+- In the seventh panel showing metrics, colors are:
+  - True positive: light blue
+  - True negative: pink
+  - False positive: blue
+  - False negative: red
 
-```{r, fig.width = 12, fig.height = 6}
+```{r, fig.width = 12, fig.height = 9}
 pc_color_columns <- c("scDblFinder prediction" = "scdbl_prediction",
@@
         )
+
       # plot for consensus_class, which needs three colors
@@
         title = "Consensus class"
+      )
+
+      # plot metrics
+      pc_plot_list$metrics <- plot_pca_metrics(
+        df, 
+        color_column = confusion_call
       )
@@
 5. Consensus _class_: droplets are either "doublet", "singlet", or "ambiguous"
+6. Metrics based on the consensus call
 

### 2024-07-02 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
plot theming

@@
 
-theme_set(theme_bw())
+theme_set(
+  theme_bw() + 
+    theme(
+      legend.title = element_text(size = rel(1.1)),
+      legend.text = element_text(size = rel(1.1))
+  )
+)
 
@@
   scale_color_manual(name = "Call type", values = metric_colors) +
-  ggtitle("Consensus class metrics") +
+  ggtitle("Consensus call metrics") +
   guides(color = guide_legend(override.aes = list(size = 2)))
-
 }

### 2024-07-03 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd

@@
   dplyr::mutate(updated_marker_gene_classification = dplyr::if_else(sum_raw_exp >= exp_cutoff, "Tumor", "Normal")) |> 
-  dplyr::select(barcodes, updated_marker_gene_classification)
+  dplyr::select(barcodes, updated_marker_gene_classification) |> 
+  unique()
 
@@
 This corresponds to the highest value that reduces the false positives. 
+See the [`AUCell` vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html#determine-the-cells-with-the-given-gene-signatures-or-active-gene-sets) for more information.
 
@@
 # pull out tumor cells found by AUCell
-tumor_cells <- ref_assignment$markers$assignment
+tumor_cells <- ref_assignments$markers$assignment
 
@@
 caret_df <- ref_classification_df |> 
-  dplyr::filter(tumor_cell_classification != "Ambiguous")
+  dplyr::filter(tumor_cell_classification != "Ambiguous") |> 
+  # make sure positive class is tumor
+  dplyr::mutate(
+    tumor_cell_classification = forcats::fct_relevel(tumor_cell_classification, "Tumor"),
+    auc_classification = forcats::fct_relevel(auc_classification, "Tumor")
+  )
 
@@
 
+Here the original tumor cell classifications are the rows and the `AUCell` classification are the columns. 
 It looks like we get pretty good agreement between our original classification and the updated `AUCell` classification using the marker gene list.
@@
 
-all_auc_df <- dplyr::bind_rows(list(ref_auc_df, new_auc_df))
+all_auc_df <- dplyr::bind_rows(list(ref_auc_df, new_auc_df)) |> 
+  # simplify the name for plotting
+  dplyr::mutate(gene_set = stringr::word(gene_set, 1, sep = "_"))
 
@@
   geom_density() +
-  facet_grid(rows = vars(gene_set))
+  facet_grid(rows = vars(gene_set),
+             scales = "free_y")
 ```
 
-Looking at the AUC across both samples for all gene sets, we see that they are in the same range of values and tend to mirror each other. 
-The main difference here is the lack of bimodal distribution in `SCPCL000824`.
+Looking at the AUC across both samples for all gene sets, we see that they are in the same range of values. 
+The main difference here is the lack of bimodal distribution in `SCPCL000824` and we see that most of the time the peak for `SCPCL000824` seems to be in between the two peaks for `SCPCL000822`.
 
-Let's use the marker gene threshold identified by `AUCell` for `SCPCL000822` to identify tumor cells in `SCPCL000824`.  
+The exception to this is with the marker gene list where the peak in `SCPCL000824` lines up with the second peak in `SCPCL000822`. 
+Because of that we will use the marker gene threshold identified by `AUCell` for `SCPCL000822` to identify tumor cells in `SCPCL000824`.  
 Here, we label any cells that have passed the threshold set for the marker gene set as tumor cells.
@@
 # pull out threshold used for assigning cells from ref sample 
-ref_threshold <- ref_assignment$markers$aucThr$selected
+ref_threshold <- ref_assignments$markers$aucThr$selected
 
@@
 We can confirm this by calculating the confusion matrix. 
+Here the marker gene annotations are the rows and the annotations from `AUCell` are the columns.
 
 ```{r}
+classification_df <- classification_df |> 
+  dplyr::mutate(
+    updated_marker_gene_classification = forcats::fct_relevel(updated_marker_gene_classification, "Tumor"),
+    auc_classification = forcats::fct_relevel(auc_classification, "Tumor")
+  )
+
 # compare using a confusion matrix
@@
   dplyr::mutate(
-    ucell_classification = dplyr::if_else(barcodes %in% ref_ucell_tumor_cells, "Tumor", "Normal")
+    ucell_classification = dplyr::if_else(barcodes %in% ref_ucell_tumor_cells, "Tumor", "Normal"), 
+    tumor_cell_classification = forcats::fct_relevel(tumor_cell_classification, "Tumor"),
+    ucell_classification = forcats::fct_relevel(ucell_classification, "Tumor")
   )
@@
 
+Here the rows correspond to the original tumor annotations and the columns are the `UCell` annotations. 
 It looks like we get pretty good agreement between using `UCell` with the `RIGGI` gene set to the original classification. 
@@
   dplyr::mutate(
-    ucell_classification = dplyr::if_else(barcodes %in% new_ucell_tumor_cells, "Tumor", "Normal")
+    ucell_classification = dplyr::if_else(barcodes %in% new_ucell_tumor_cells, "Tumor", "Normal"), 
+    # relevel for confusion matrix later 
+    updated_marker_gene_classification = forcats::fct_relevel(updated_marker_gene_classification, "Tumor"),
+    ucell_classification = forcats::fct_relevel(ucell_classification, "Tumor")
   )
@@
 We can confirm that by calculating the confusion matrix.
+Here the rows correspond to the updated marker gene annotations and the columns are the `UCell` annotations.
 

### 2024-07-05 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_singler-classification.Rmd
add tumor only and comparison of reference labels

@@
 This notebook looks at using `SingleR` to annotate tumor cells in `SCPCL000824`. 
-We used three different references with `SingleR`:
+We used four different references with `SingleR`:
 
@@
 Ambiguous cells are removed from `SCPCL000822`.
-2. `Original`: `SCPCL000822` where all tumor cells are labeled as tumor and all other cells keep the original annotation obtained from running `SingleR` as part of `scpca-nf`.  
-3. `Original and Blueprint`: `SCPCL000822` with tumor cells labeled as in #2 combined with the full `BlueprintEncodeData` reference.
+2. `Tumor only`: Both the full `BlueprintEncodeData` reference from `celldex` and only the tumor cells from `SCPCL000822`. 
+3. `Original`: `SCPCL000822` where all tumor cells are labeled as tumor and all other cells keep the original annotation obtained from running `SingleR` as part of `scpca-nf`.  
+4. `Original and Blueprint`: `SCPCL000822` with tumor cells labeled as in #2 combined with the full `BlueprintEncodeData` reference.
 
@@
       \(x) {
-        x |>
+          # give NA's a unique label so we can easily track and plot them later
+          # they will get grouped with all remaining cell types unless they are a top cell type
+        dplyr::if_else(is.na(x), "Not labeled", x) |>
           # there are warnings because we have some that have less than 7 cell types (ones that only did tumor/normal)
@@
   "blueprint_tumor_classification_lumped",
+  "tumor_only_classification_lumped",
   "updated_singler_classification_lumped",
@@
 ) |> 
-  purrr::set_names("Blueprint", "Original", "Original + Blueprint")
+  purrr::set_names("Blueprint", "Tumor only", "Original", "Original + Blueprint")
 
@@
         alpha = 0.5,
-      size = 0.5
+        size = 0.1
       ) +
       # set points for desired cell type
-    geom_point(size = 0.5, alpha = 0.5) +
+      geom_point(size = 0.1, alpha = 0.5) +
       facet_wrap(
@@
 
-First we will just look at the overlap in the tumor cells between all three methods.
+First we will just look at the overlap in the tumor cells between all four methods.
 
@@
   "Blueprint" = classification_df$barcodes[classification_df$blueprint_tumor_classification == "Tumor"],
+  "Tumor only" = classification_df$barcodes[classification_df$tumor_only_classification == "Tumor"],
   "Original" = classification_df$barcodes[classification_df$updated_singler_classification == "Tumor"],
@@
   dplyr::mutate(singler_consensus_call = dplyr::case_when(
-    blueprint_tumor_classification == "Tumor" & updated_singler_classification == "Tumor" & updated_singler_blueprint_classification == "Tumor" ~ "Tumor",
-    blueprint_tumor_classification == "Tumor" | updated_singler_classification == "Tumor" | updated_singler_blueprint_classification == "Tumor" ~ "Ambiguous",
+    blueprint_tumor_classification == "Tumor" & tumor_only_classification == "Tumor" & updated_singler_classification == "Tumor" & updated_singler_blueprint_classification == "Tumor" ~ "Tumor",
+    blueprint_tumor_classification == "Tumor" | tumor_only_classification == "Tumor" | updated_singler_classification == "Tumor" | updated_singler_blueprint_classification == "Tumor" ~ "Ambiguous",
     .default = "Normal"
@@
 
-It looks like the cells that are normal are in the smaller clusters and the larger cluster is mostly tumor cells with some that are ambiguous across `SingleR` references. 
+It looks like the cells that are normal or ambiguous are in the smaller clusters and the larger cluster is mostly tumor cells with some that are ambiguous across `SingleR` references. 
 This is what I would expect given the marker gene expression in these cells. 
+This plot looked different when we didn't include the classifications from the tumor only reference, where most of the smaller clusters were considered normal and not ambiguous. 
+It looks like using just the tumor cells from `SCPCL000822` with Blueprint is calling more cells as tumor than the other methods.
 
@@
 methods <- c(
+  "tumor_only_classification_lumped",
   "updated_singler_classification_lumped",
@@
 ) |> 
-  purrr::set_names("Original", "Original and Blueprint")
+  purrr::set_names("Tumor only", "Original", "Original and Blueprint")
 
-# compare blueprint to original and origintal + blueprint
+# compare blueprint to tumor, original and original + blueprint
 plot_jaccard(classification_df,
@@
 
-# compare original to original + blueprint
+# compare original to tumor, original + blueprint
+subset_methods <- c(
+  "tumor_only_classification_lumped",
+  "updated_singler_blueprint_classification_lumped"
+) |> 
+  purrr::set_names("Tumor only", "Original and Blueprint")
+
 plot_jaccard(classification_df,
              annotation_column = "updated_singler_classification_lumped",
-             methods_to_compare = c("Original and Blueprint" = "updated_singler_blueprint_classification_lumped"),
+             methods_to_compare = subset_methods,
              column_title = "Original",
              legend_match = "Original")
+
+# compare tumor to original and blueprint
+plot_jaccard(classification_df,
+             annotation_column = "tumor_only_classification_lumped",
+             methods_to_compare = c("Original and Blueprint" = "updated_singler_blueprint_classification_lumped"),
+             column_title = "Tumor only",
+             legend_match = "Tumor only")
 
@@
 
-- There seems to be mostly consensus across all three methods on which cells are tumor cells, which is consistent with the upset plot. 
+- There seems to be mostly consensus across all four methods on which cells are tumor cells, which is consistent with the upset plot. 
 - When only using Blueprint, a lot of cells are categorized as "Normal". 
@@
 We also see that for the most part the top cell types identified are the same except when using blueprint you see the addition of epithelial cells. 
+- Using tumor only appears to label some cells as tumor that are labeled as normal with other methods. 
+For example, comparing tumor only to both original and original and blueprint shows that tumor cells are classified as endothelial cells using either of the other two methods.
 
@@
 
-I tend to favor using one of these options:
+I tend to favor using the original and Blueprint option that uses tumor cell labels and `SingleR` labels from `SCPCL000822` along with `BlueprintEncodeData`.
 
-- `Blueprint`: `SCPCL000822` with cells labeled as either tumor or normal in combination with Blueprint. 
-- `Original and Blueprint`: `SCPCL000822` with tumor cells labeled and all other cells labeled with the original `SingleR` annotations in combination with `Blueprint`.
-
+It's possible that the Blueprint with tumor and normal cells from `SCPCL000822` is absorbing bad calls into the broader `Normal` category, so we probably want to avoid that. 
+The tumor only appears to be calling more cells as tumor than may be we want to be? 
+These same cells are categorized as endothelial cells with both Original and Original and blueprint. 
 I don't like using just the `Original`, where the only reference comes from a single sample, `SCPCL000822`. 
@@
 
-Another option would be to try running with just the tumor cells from `SCPCL000822` as a reference in combination with `Blueprint`, removing the "normal" cells from the reference. 
-But either of the above options I think would be sufficient to labeling and pulling out tumor cells. 
+Let's look into the original and blueprint more, specifically let's see if the non tumor cells tend to align with the normal cells from `SCPCL000822` or the normal cells from `BlueprintEncodeData` more.
+
+```{r}
+# label cells with original or blueprint label
+classification_df <- classification_df |> 
+  dplyr::mutate(
+    orig_ref = dplyr::if_else(
+      stringr::str_detect(updated_singler_blueprint_ontology, "CL:"),
+      "Blueprint", 
+      "SCPCL000822"
+    ), 
+    combined_annotation = glue::glue("{updated_singler_blueprint_classification}-{orig_ref}")
+  )
+```
+
+Below is a table that shows the number of cells for each cell type that correspond to each reference. 
+The table is sorted by the number of cells assigned to the original reference from `SCPCL000822`.
+
+```{r}
+# make a table showing which cells are assigned to which reference 
+celltype_table <- table(classification_df$updated_singler_blueprint_classification, 
+      classification_df$orig_ref)
+# sort by SCPCL000822
+celltype_table[order(celltype_table[,2], decreasing = TRUE), ]
+```
+
+From this table, we see a number of cell normal cell types that have more cells that match the `SCPCL000822` reference over the Blueprint reference including endothelial cells and chondrocytes. 
+It looks like fibroblasts and neurons better resemble the cells in Blueprint over `SCPCL000822`. 
+And finally, there are a number of cells that are only from Blueprint, like astrocytes and epithelial cells.
+
+Based on the data shown so far, I think using the Original with blueprint would be sufficient to labeling and pulling out tumor cells. 
 Let's confirm this by comparing to our previously used marker gene based methods for classifying this sample.
@@
 However, it looks like `UCell` categorizes more cells as normal that are classified as tumor with any of the `SingleR` methods. 
+Also, using the tumor cells only categorizes more cells as tumor that are classified as normal with other marker gene based methods.
 
 For both original and original and blueprint, it looks like there is a good amount of cells that `SingleR` classifies as normal, but the marker gene based methods classify as tumor. 
-This tells me that using `SingleR` may be slightly more conservative than the marker gene based methods.
+This tells me that using `SingleR` may be slightly more conservative than the marker gene based methods other than when using tumor cells alone.
 
@@
       method == "blueprint_tumor_classification" ~ "Blueprint",
+      method == "tumor_only_classification" ~ "Tumor only",
       method == "updated_singler_classification" ~ "Original",
@@
 However there are some cells that are only identified in one or two methods making the full set of tumor cells unique across references. 
-- All three references identify the same top normal cell types with some overlapping sets. 
+- All four references identify the same top normal cell types with some overlapping sets. 
 - There does not appear to be a consensus between the original `SingleR` annotations and the new annotations for any of the references. 

### 2024-07-08 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_singler-classification.Rmd

@@
 ```
+
 Generally, it looks like there is a large group of cells that are identified as tumor regardless of the reference, with a few cells that are specifically tumor in only one or two references.

### 2024-07-09 analyses/cell-type-ewings/template_notebooks/cnv-workflow/01-marker-genes.Rmd
move cnv output

--- 2024-06-05 analyses/cell-type-ewings/template_notebooks/01-marker-genes.Rmd
@@
   library_id: SCPCL000822
-  results_dir: "annotate_tumor_cells_output"
+  results_dir: "cnv_annotation"
   reference_cell_file: "reference-cells.tsv"

### 2024-07-09 analyses/cell-type-ewings/template_notebooks/cnv-workflow/02-cellassign.Rmd
move cnv output

--- 2024-06-05 analyses/cell-type-ewings/template_notebooks/02-cellassign.Rmd
@@
   panglao_predictions: "panglao-endo-fibro_predictions.tsv"
-  results_dir: "annotate_tumor_cells_output"
+  results_dir: "cnv_annotation"
 ---

### 2024-07-09 analyses/cell-type-ewings/template_notebooks/cnv-workflow/03-copykat.Rmd
move cnv output

--- 2024-06-18 analyses/cell-type-ewings/template_notebooks/03-copykat.Rmd
@@
   reference_cell_file: ".reference-cells.tsv"
-  results_dir: "annotate_tumor_cells_output"
+  results_dir: "cnv_annotation"
 ---

### 2024-07-10 analyses/cell-type-ewings/template_notebooks/cnv-workflow/01-marker-genes.Rmd
don't load Rproj in notebooks
don't use find git root

@@
 ```{r}
-project_root <- here::here()
-renv::load(project_root)
-
 knitr::opts_chunk$set(
@@
 ```{r base paths}
-# The base path for the OpenScPCA repository, found by its (hidden) .git directory
-repository_base <- rprojroot::find_root(rprojroot::is_git_root)
-
 # The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "current")
+data_dir <- file.path("..", "..", "data", "current")
 sample_dir <- file.path(data_dir, "SCPCP000015", params$sample_id)
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "cell-type-ewings")
+module_base <- file.path("analyses", "cell-type-ewings")
 ```

### 2024-07-10 analyses/cell-type-ewings/template_notebooks/cnv-workflow/04-infercnv.Rmd
don't use find git root

--- 2024-06-18 analyses/cell-type-ewings/template_notebooks/04-infercnv.Rmd
@@
   infercnv_dir: "infercnv"
-  results_dir: "annotate_tumor_cells_output"
+  results_dir: "cnv_annotation"
 ---

### 2024-07-10 analyses/cell-type-ewings/template_notebooks/cnv-workflow/01-marker-genes.Rmd
test changing criterion for find root

--- 2024-07-09 analyses/cell-type-ewings/template_notebooks/cnv-workflow/01-marker-genes.Rmd
@@
 # The base path for the OpenScPCA repository, found by its (hidden) .git directory
-repository_base <- rprojroot::find_root(rprojroot::is_git_root)
+repository_base <- rprojroot::find_root(rprojroot::has_file("LICENSE.md"))
 

### 2024-07-10 analyses/cell-type-ewings/template_notebooks/cnv-workflow/01-marker-genes.Rmd
update criterion for all notebooks

@@
 ```{r base paths}
-# The base path for the OpenScPCA repository, found by its (hidden) .git directory
+# The base path for the OpenScPCA repository, found by LICENSE.md file
+# can't use .git because GHA doesn't find .git
 repository_base <- rprojroot::find_root(rprojroot::has_file("LICENSE.md"))

### 2024-07-10 analyses/cell-type-ewings/template_notebooks/cnv-workflow/02-cellassign.Rmd
update criterion for all notebooks

--- 2024-07-09 analyses/cell-type-ewings/template_notebooks/cnv-workflow/02-cellassign.Rmd
@@
 ```{r base paths}
-# The base path for the OpenScPCA repository, found by its (hidden) .git directory
-repository_base <- rprojroot::find_root(rprojroot::is_git_root)
+# The base path for the OpenScPCA repository, found by LICENSE.md file
+# can't use .git because GHA doesn't find .git
+repository_base <- rprojroot::find_root(rprojroot::has_file("LICENSE.md"))
 

### 2024-07-10 analyses/cell-type-ewings/template_notebooks/cnv-workflow/02-cellassign.Rmd

--- 2024-07-10 analyses/cell-type-ewings/template_notebooks/cnv-workflow/02-cellassign.Rmd
@@
 ```{r base paths}
+# The path to this module
+module_base <- rprojroot::find_root(rprojroot::is_renv_project)
+
+# repo base is two levels up from module 
+repository_base <- dirname(dirname(module_base))
+
 # The current data directory, found within the repository base directory
-data_dir <- file.path("..", "..", "data", "current")
+data_dir <- file.path(repository_base, "data", "current")
 sample_dir <- file.path(data_dir, "SCPCP000015", params$sample_id)
-
-# The path to this module
-module_base <- file.path("analyses", "cell-type-ewings")
 ```

### 2024-07-10 analyses/cell-type-ewings/template_notebooks/auc-workflow/01-auc-results.Rmd
Reapply "add notebook for looking at AUC results"
This reverts commit 920083841ca440a447c8865f694a9b135e9df960.

---
title: "Summary of tumor cell classification with `AUCell` for `r params$library_id`"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  library_id: NULL
  sce_file: NULL
  aucell_results_file: NULL
  ref_auc_results: NULL
  geneset_scores_file: NULL
  marker_genes_file: NULL
  auc_threshold: NULL
---

This report summarizes the classification of `r params$library_id` using [`AUCell`](https://www.bioconductor.org/packages/release/bioc/html/AUCell.html). 
Cells were classified as tumor if they had an AUC value greater than or equal to `r params$auc_threshold`.

## Setup

```{r}
# check that sce and results files exist 
stopifnot(
  "sce file does not exist" = file.exists(params$sce_file),
  "AUCell results file does not exist" = file.exists(params$aucell_results_file), 
  "reference results from AUCell does not exist" = file.exists(params$ref_auc_results), 
  "Marker genes file does not exist" = file.exists(params$marker_genes_file), 
  "Gene set scores file does not exist" = file.exists(params$geneset_scores_file)
)
```

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)

# quiet messages
options(readr.show_col_types = FALSE)
ComplexHeatmap::ht_opt(message = FALSE)

# set colors for tumor and normal cells throughout report 
plot_colors <- c("Tumor" = "#00274C", "Normal" = "#FFCB05")
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-ewings")

# source in helper functions: plot_gene_heatmap() and plot_cnv_heatmap() 
# create_classification_df() and create_marker_gene_df()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)
```

```{r}
# Read in files
sce <- readr::read_rds(params$sce_file)
auc_results_df <- readr::read_tsv(params$aucell_results_file)
geneset_scores_df <- readr::read_tsv(params$geneset_scores_file)

# reference auc results file 
ref_auc_df <- readr::read_tsv(params$ref_auc_results)
```

```{r}
# generate classification df to use for plots 
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, singler_celltype_annotation) |> 
  # join with previous annotations, singler results, and gene set scores 
  dplyr::left_join(auc_results_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes")

# get marker gene expression 
markers_df <- create_marker_gene_df(
  sce,
  classification_df, 
  params$marker_genes_file
)
```

## AUCell classification UMAP

Below is a UMAP showing cells that are labeled as either `Tumor` or `Normal` using `AUCell`.

```{r}
ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = auc_classification)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_manual(values = plot_colors) +
  labs(
    color = "AUCell classification"
  )
```

## Comparison of AUC distribution

The below plot compares the distribution of AUC values in the query library as compared to the reference library used to determine the AUC threshold.

```{r}
all_auc_df <- list(
  "reference" = ref_auc_df, 
  "query" = auc_results_df
) |> 
  dplyr::bind_rows(.id = "sample")

ggplot(all_auc_df, aes(x = auc, color = sample)) +
  geom_density() + 
  geom_vline(xintercept = params$auc_threshold, linetype = "dashed") +
  labs(
    x = "AUC",
    color = ""
  )
```

## Expression of marker genes

The below plots summarize marker gene expression in tumor and normal cells as classified by `AUCell`.

### Density plot

The `logcounts` for each gene in the marker gene list is summed for each cell and shown on the x-axis.
We would expect tumor cells to have higher expression of marker genes than normal cells.

```{r}
# create a density plot showing the distribution of marker gene expression across classification methods 
markers_df |> 
  dplyr::select(barcodes, sum_raw_exp, auc_classification) |> 
  dplyr::distinct() |> 
  ggplot(aes(x = sum_raw_exp, color = auc_classification)) +
  geom_density() +
  scale_color_manual(values = plot_colors) +
  labs(
    x = "Raw sum of marker gene logcounts",
    color = "AUCell classification"
  )
```

### Heatmap

Here every row is a marker gene and every column is a cell with colors corresponding to the expression of a specified marker gene.  
The annotation bar indicates which cells are tumor or normal cells as classified by `AUCell`. 
We expect to see higher expression of individual marker genes in tumor cells compared to normal cells.

```{r}
# create matrix with marker genes as rows and barcodes as columns
marker_gene_heatmap <- markers_df |> 
  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
  tidyr::pivot_wider(values_from = gene_expression, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("gene_symbol") |> 
  as.matrix()

annotation <- ComplexHeatmap::columnAnnotation(
  AUCell = classification_df$auc_classification, 
  col = list(AUCell = plot_colors)
)
```

```{r}
# plot heatmap of marker genes 
plot_gene_heatmap(marker_gene_heatmap, 
                  row_title = "Marker gene symbol",
                  legend_title = "Marker gene \nexpression",
                  annotation = annotation)
```

## Gene set scores

The below plots summarize the gene set scores in tumor and normal cells as classified by `AUCell`.

Gene set scores are determined by calculating the mean expression of all genes in a given gene set. 
Scores are calculated for the following gene sets from `MsigDb`:

  - [`ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION`](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION.html)
  - [`RIGGI_EWING_SARCOMA_PROGENITOR_UP`](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/RIGGI_EWING_SARCOMA_PROGENITOR_UP.html?ex=1)
  - [`SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN`](https://www.gsea-msigdb.org/gsea/msigdb/cards/SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN)

### Density plot

```{r}
# prep data frame for plotting gene set scores 
geneset_plot_df <- classification_df |> 
  dplyr::select(barcodes, auc_classification, starts_with("mean-")) |> 
  # make one column containing all gene set scores 
  tidyr::pivot_longer(
    cols = starts_with("mean"),
    names_to = "geneset",
    values_to = "mean_score"
  ) |> 
  # simplify gene set name for labels
  dplyr::mutate(
    geneset = stringr::word(geneset, -1, sep = "-")
  )

# make density plot showing distribution of gene set scores for tumor/normal 
ggplot(geneset_plot_df, aes(x = mean_score, color = auc_classification)) +
  geom_density(bw = 0.05) +
  facet_grid(rows = vars(geneset),
             scales = "free_y") +
  scale_color_manual(values = plot_colors) +
  labs(
    x = "Gene set score",
    color = "AUCell classification"
  )
```

### Heatmap

Here every column is a cell and every row is a gene set with colors corresponding to the gene set score. 
The annotation bar indicates which cells are tumor or normal cells as classified by `AUCell`. 
We expect to see higher gene set scores in tumor cells compared to normal cells.

```{r}
# make a matrix of gene set by barcode 
geneset_heatmap <- geneset_plot_df |> 
  dplyr::select(mean_score, geneset, barcodes) |> 
  unique() |> 
  tidyr::pivot_wider(values_from = mean_score, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("geneset") |> 
  as.matrix()

# plot heatmap of gene set score 
plot_gene_heatmap(geneset_heatmap,
                  annotation = annotation,
                  legend_title = "Gene set \nscore")
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-07-10 analyses/cell-type-ewings/template_notebooks/auc-workflow/01-auc-results.Rmd

@@
   dplyr::left_join(auc_results_df, by = "barcodes") |> 
-  dplyr::left_join(geneset_scores_df, by = "barcodes")
+  dplyr::left_join(geneset_scores_df, by = "barcodes") |> 
+  dplyr::mutate(
+    # make a factor so tumor always appears first 
+    auc_classification = forcats::fct_relevel(auc_classification, "Tumor")
+  )
 
@@
 ```{r}
-all_auc_df <- list(
+all_auc_df <- dplyr::bind_rows(
   "reference" = ref_auc_df, 
-  "query" = auc_results_df
+  "query" = auc_results_df,
+  .id = "sample"
 ) |>
-  dplyr::bind_rows(.id = "sample")
+  dplyr::mutate(
+    sample = ifelse(sample == "query", paste("query:", params$library_id), sample)
+  )
 
@@
 # create matrix with marker genes as rows and barcodes as columns
-marker_gene_heatmap <- markers_df |> 
+marker_gene_matrix <- markers_df |> 
   dplyr::select(gene_expression, gene_symbol, barcodes) |> 
@@
 # plot heatmap of marker genes 
-plot_gene_heatmap(marker_gene_heatmap, 
+plot_gene_heatmap(marker_gene_matrix, 
                   row_title = "Marker gene symbol",
@@
 # make a matrix of gene set by barcode 
-geneset_heatmap <- geneset_plot_df |> 
+geneset_matrix <- geneset_plot_df |> 
   dplyr::select(mean_score, geneset, barcodes) |> 
@@
 # plot heatmap of gene set score 
-plot_gene_heatmap(geneset_heatmap,
+plot_gene_heatmap(geneset_matrix,
                   annotation = annotation,

### 2024-07-11 analyses/cell-type-ewings/template_notebooks/auc-workflow/01-auc-results.Rmd
account for no marker gene expression

@@
 params:
-  library_id: NULL
-  sce_file: NULL
-  aucell_results_file: NULL
-  ref_auc_results: NULL
-  geneset_scores_file: NULL
-  marker_genes_file: NULL
-  auc_threshold: NULL
+  library_id: SCPCS000490
+  sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
+  aucell_results_file: "../../results/aucell_annotation_test/SCPCS000490/SCPCL000822_auc-classifications.tsv"
+  ref_auc_results: "../../results/aucell_annotation_test/SCPCS000490/SCPCL000822_auc-classifications.tsv"
+  geneset_scores_file: "../../results/aucell_annotation_test/SCPCS000490/SCPCL000822_gene-set-scores.tsv"
+  marker_genes_file: "../../references/tumor-marker-genes.tsv"
+  auc_threshold: 0.009131468
 ---
@@
 )
+
+# test if _any_ marker genes have expression
+if(sum(markers_df$gene_expression) == 0){
+  has_marker_gene_exp <- FALSE
+  message("No marker gene expression detected, so all plots of marker genes will be skipped.")
+} else {
+  has_marker_gene_exp <- TRUE
+}
 ```
@@
 
-```{r}
+```{r, eval=has_marker_gene_exp}
 # create a density plot showing the distribution of marker gene expression across classification methods 
@@
 
-```{r}
+```{r, eval=has_marker_gene_exp}
 # plot heatmap of marker genes 

### 2024-07-11 analyses/cell-type-ewings/template_notebooks/auc-workflow/01-auc-results.Rmd
go back to null variables

@@
 params:
-  library_id: SCPCS000490
-  sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
-  aucell_results_file: "../../results/aucell_annotation_test/SCPCS000490/SCPCL000822_auc-classifications.tsv"
-  ref_auc_results: "../../results/aucell_annotation_test/SCPCS000490/SCPCL000822_auc-classifications.tsv"
-  geneset_scores_file: "../../results/aucell_annotation_test/SCPCS000490/SCPCL000822_gene-set-scores.tsv"
-  marker_genes_file: "../../references/tumor-marker-genes.tsv"
-  auc_threshold: 0.009131468
+  library_id: NULL
+  sce_file: NULL
+  aucell_results_file: NULL
+  ref_auc_results: NULL
+  geneset_scores_file: NULL
+  marker_genes_file: NULL
+  auc_threshold: NULL
 ---

### 2024-07-11 analyses/cell-type-ewings/template_notebooks/auc-workflow/01-auc-results.Rmd

@@
 ```{r base paths}
-# The base path for the OpenScPCA repository, found by its (hidden) .git directory
-repository_base <- rprojroot::find_root(rprojroot::is_git_root)
-
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "cell-type-ewings")
+module_base <- rprojroot::find_root(rprojroot::is_renv_project)
 

### 2024-07-12 analyses/cell-type-ewings/template_notebooks/auc-workflow/01-auc-results.Rmd
add plotting of marker gene cutoff

@@
   dplyr::left_join(auc_results_df, by = "barcodes") |> 
-  dplyr::left_join(geneset_scores_df, by = "barcodes") |> 
-  dplyr::mutate(
-    # make a factor so tumor always appears first 
-    auc_classification = forcats::fct_relevel(auc_classification, "Tumor")
-  )
+  dplyr::left_join(geneset_scores_df, by = "barcodes")
 
@@
 
+```{r}
+# add marker gene classification classifying cells based on expression of at least 1 marker gene  
+marker_gene_class_df <- markers_df |> 
+  dplyr::select(barcodes, sum_raw_exp) |> 
+  dplyr::distinct() |> 
+  dplyr::mutate(
+    marker_gene_classification = dplyr::if_else(
+      sum_raw_exp > 0,
+      "Tumor",
+      "Normal"
+
+    )
+  )
+
+classification_df <- classification_df |> 
+  dplyr::left_join(marker_gene_class_df, by = "barcodes") |> 
+  tidyr::pivot_longer(ends_with("classification"),
+                      names_to = "method",
+                      values_to = "classification") |> 
+  dplyr::mutate(
+    # make a factor so tumor always appears first 
+    classification = forcats::fct_relevel(classification, "Tumor")
+  )
+```
+
 ## AUCell classification UMAP
@@
 ```{r}
-ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = auc_classification)) +
+ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = classification)) +
   geom_point(size = 0.5, alpha = 0.5) +
+  facet_wrap(vars(method)) +
   scale_color_manual(values = plot_colors) +
   labs(
-    color = "AUCell classification"
-  )
+    color = ""
+  ) +
+  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
 ```
@@
 # create a density plot showing the distribution of marker gene expression across classification methods 
-markers_df |> 
-  dplyr::select(barcodes, sum_raw_exp, auc_classification) |> 
-  dplyr::distinct() |> 
-  ggplot(aes(x = sum_raw_exp, color = auc_classification)) +
+classification_df |> 
+  ggplot(aes(x = sum_raw_exp, color = classification)) +
   geom_density() +
+  facet_wrap(vars(method),
+             scales = "free_y") +
   scale_color_manual(values = plot_colors) +
@@
     x = "Raw sum of marker gene logcounts",
-    color = "AUCell classification"
+    color = ""
   ) 
@@
 
+annotation_df <- classification_df |> 
+  dplyr::select(barcodes, classification, method) |> 
+  tidyr::pivot_wider(names_from = method, 
+                     values_from = classification)
+
 annotation <- ComplexHeatmap::columnAnnotation(
-  AUCell = classification_df$auc_classification, 
-  col = list(AUCell = plot_colors)
+  AUCell = annotation_df$auc_classification, 
+  marker_genes = annotation_df$marker_gene_classification,
+  col = list(AUCell = plot_colors,
+             marker_genes = plot_colors)
 )
@@
 geneset_plot_df <- classification_df |> 
-  dplyr::select(barcodes, auc_classification, starts_with("mean-")) |> 
+  dplyr::select(barcodes, classification, method, starts_with("mean-")) |> 
   # make one column containing all gene set scores 
@@
 # make density plot showing distribution of gene set scores for tumor/normal 
-ggplot(geneset_plot_df, aes(x = mean_score, color = auc_classification)) +
+ggplot(geneset_plot_df, aes(x = mean_score, color = classification)) +
   geom_density(bw = 0.05) +
   facet_grid(rows = vars(geneset),
+             cols = vars(method),
              scales = "free_y") +
@@
     x = "Gene set score",
-    color = "AUCell classification"
+    color = ""
   )

### 2024-07-12 analyses/cell-type-ewings/template_notebooks/auc-workflow/01-auc-results.Rmd
update comments and text

@@
 ```{r}
-# add marker gene classification classifying cells based on expression of at least 1 marker gene  
+# add marker gene classification based on expression of at least 1 marker gene  
 marker_gene_class_df <- markers_df |> 
@@
 
+# add classification to main classification data frame
 classification_df <- classification_df |> 
@@
 
-Below is a UMAP showing cells that are labeled as either `Tumor` or `Normal` using `AUCell`.
+Below is a UMAP showing cells that are labeled as either `Tumor` or `Normal` using `AUCell` and marker gene expression.
 
@@
 
-The below plots summarize marker gene expression in tumor and normal cells as classified by `AUCell`.
+The below plots summarize marker gene expression in tumor and normal cells as classified by `AUCell` and marker gene expression. 
+Note that we expect to see clear separation of gene expression for the marker gene classification as classification is based purely on expression.
 
@@
 Here every row is a marker gene and every column is a cell with colors corresponding to the expression of a specified marker gene.  
-The annotation bar indicates which cells are tumor or normal cells as classified by `AUCell`. 
+The annotation bar indicates which cells are tumor or normal cells as classified by `AUCell` or marker gene expression. 
 We expect to see higher expression of individual marker genes in tumor cells compared to normal cells.
@@
 
-The below plots summarize the gene set scores in tumor and normal cells as classified by `AUCell`.
+The below plots summarize the gene set scores in tumor and normal cells as classified by `AUCell` and marker gene expression.
 
@@
 Here every column is a cell and every row is a gene set with colors corresponding to the gene set score. 
-The annotation bar indicates which cells are tumor or normal cells as classified by `AUCell`. 
+The annotation bar indicates which cells are tumor or normal cells as classified by `AUCell` and marker gene expression. 
 We expect to see higher gene set scores in tumor cells compared to normal cells.

### 2024-07-12 analyses/cell-type-ewings/template_notebooks/auc-workflow/01-auc-results.Rmd
export marker gene classification

@@
   marker_genes_file: NULL
+  output_file: NULL
   auc_threshold: NULL
@@
 
-This report summarizes the classification of `r params$library_id` using [`AUCell`](https://www.bioconductor.org/packages/release/bioc/html/AUCell.html). 
-Cells were classified as tumor if they had an AUC value greater than or equal to `r params$auc_threshold`.
+This report summarizes the classification of `r params$library_id` using two methods:
+
+- [`AUCell`](https://www.bioconductor.org/packages/release/bioc/html/AUCell.html): Cells are classified as tumor if they had an AUC value greater than or equal to `r params$auc_threshold`. 
+- Marker gene expression: Cells were classified as tumor if they expressed any tumor marker genes.
 
@@
 
+## Save marker gene classification
+
+```{r}
+classification_df |> 
+  dplyr::filter(method == "marker_gene_classification") |> 
+  dplyr::select(barcodes, marker_gene_classification = classification) |> 
+  readr::write_tsv(params$output_file)
+```
+
 ## Session Info

### 2024-07-12 analyses/cell-type-ewings/aucell-annotation.sh

@@ -58,6 +58,7 @@ for sample_id in $sample_ids; do
         # define output files
         auc_results="${sample_results_dir}/${library_id}_auc-classifications.tsv"
         geneset_results="${sample_results_dir}/${library_id}_gene-set-scores.tsv"
+        marker_gene_results=${sample_results_dir}/${library_id}_marker-gene-classifications.tsv"
 
         # only run AUCell if sample is NOT the ref sample
         if [[ ($sample_id != $ref_sample) && ($library_id != $ref_library) && (! -f $auc_results) ]]; then
@@ -89,6 +90,7 @@ for sample_id in $sample_ids; do
                         ref_auc_results = '$ref_auc_results', \
                         geneset_scores_file = '$geneset_results', \
                         marker_genes_file = '$marker_genes_file', \
+                        output_file = '${marker_gene_results}', \
                         auc_threshold = $auc_threshold), \
           envir = new.env()) \
         "

### 2024-07-12 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000752/SCPCL001114_tumor-gene-expression-exploration.Rmd
start notebook for validation

--- 2024-07-02 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd
@@
 ---
-title: "`SCPCL000824` - Exploration of gene set based methods for tumor cell assignment"
+title: "`SCPCL001114` - Exploration of gene set based methods for tumor cell assignment"
 author: Ally Hawkins
@@
 
-In this notebook we attempt to identify tumor cells in `SCPCL000824`. 
-We first compare tumor cell annotations obtained from marker genes, `CopyKAT`, and `InferCNV` to identify a group of tumor cells we are confident in. 
-However, there is a lack of consensus between tumor cells identified by marker gene expression and CNV inference.
-
-In an attempt to identify a group of cells that we are confident are tumor cells, here we test using the following orthogonal marker gene based methods:
+In this notebook we attempt to identify tumor cells in `SCPCL001114` using the following orthogonal marker gene based methods:
 
@@
 # The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "2024-05-01")
-sample_dir <- file.path(data_dir, "SCPCP000015", "SCPCS000492")
+data_dir <- file.path(repository_base, "data", "current")
+sample_dir <- file.path(data_dir, "SCPCP000015", "SCPCS000752")
 
@@
 # Input files
-sce_file <- file.path(sample_dir, "SCPCL000824_processed.rds")
+sce_file <- file.path(sample_dir, "SCPCL001114_processed.rds")
 marker_genes_file <- file.path(module_base, "references", "tumor-marker-genes.tsv")
@@
 # results from annotation workflow 
-results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000492")
-
-marker_gene_results_file <- file.path(results_dir, "SCPCL000824_tumor-normal-classifications.tsv")
-copykat_predictions_file <- file.path(results_dir, "SCPCL000824_copykat-classifications.tsv")
-
-infercnv_predictions_file <- file.path(results_dir, "SCPCL000824_infercnv-classifications.tsv")
-infercnv_metadata_file <- file.path(results_dir, "infercnv", "SCPCL000824_cnv-metadata.tsv")
-
-geneset_scores_file <- file.path(results_dir, "SCPCL000824_gene-set-scores.tsv")
+results_dir <- file.path(module_base, "results", "aucell_annotation", "SCPCS000752")
+auc_results_file <- file.path(results_dir, "SCPCL001114_auc-classifications.tsv")
+geneset_scores_file <- file.path(results_dir, "SCPCL001114_gene-set-scores.tsv")
 
@@
 
-# reference files to use with SingleR
+# reference files to use
 ref_sce_file <- file.path(data_dir, "SCPCP000015", "SCPCS000490", "SCPCL000822_processed.rds")
 ref_labels_file <- file.path(module_base, "results", "annotation_tables", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv")
-ref_geneset_scores_file <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
+ref_geneset_scores_file <- file.path(module_base, "results", "cnv_annotation", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
 ```
@@
 sce <- readr::read_rds(sce_file)
+auc_results_df <- readr::read_tsv(auc_results_file)
+geneset_scores_df <- readr::read_tsv(geneset_scores_file)
 
@@
 ```{r}
-# generate classification df 
-classification_df <- create_classification_df(
-  sce,
-  marker_gene_results_file,
-  copykat_predictions_file, 
-  infercnv_predictions_file,
-  infercnv_metadata_file,
-  geneset_scores_file
+# generate classification df to use for plots 
+classification_df <- sce |> 
+  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
+  # replace UMAP.1 with UMAP1
+  dplyr::rename_with(
+    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
   ) |>
-  # remove extra infercnv classification column we won't use 
-  dplyr::select(- cnv_sum_classification)
-```
+  # get rid of excess columns
+  dplyr::select(barcodes, UMAP1, UMAP2, singler_celltype_annotation) |> 
+  # join with previous annotations, singler results, and gene set scores 
+  dplyr::left_join(auc_results_df, by = "barcodes") |> 
+  dplyr::left_join(geneset_scores_df, by = "barcodes")
 
-## Evaluate tumor cell classifications
-
-Below is a UMAP showing which cells are labeled as tumor cells using each of the various classification methods (marker genes, `CopyKAT`, and `InferCNV`).
-
-```{r}
-# plot tumor cells for each classification method 
-tumor_umap_df <- classification_df |> 
-  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification")) |> 
-  tidyr::pivot_longer(
-    cols = ends_with("classification"),
-    names_to = "method",
-    values_to = "classification"
-  ) |> 
-  dplyr::mutate(
-    method = dplyr::case_when(
-      method == "marker_gene_classification" ~ "Marker genes only",
-      method == "copykat_classification" ~ "CopyKAT - no reference",
-      method == "cnv_proportion_classification" ~ "InferCNV proportion"
-    )
-  )
-
-ggplot(tumor_umap_df, aes(x = UMAP1, UMAP2, color = classification)) +
-  geom_point(size = 0.5, alpha = 0.5) +
-  facet_wrap(vars(method))
-```
-
-It looks like we don't have a lot of agreement between the marker gene based classification and copy number inference classification. 
-Although we do see some agreement between the two CNV methods.
-
-### Marker gene and gene set expression
-
-We can confirm this by looking at marker gene expression in cells classified as tumor or normal by each method.
-
-```{r}
 # generate marker genes df
@@
 )
-
-# create a density plot showing the distribution of marker gene expression across classification methods 
-marker_density_df <- plot_markers_df |>
-  tidyr::pivot_longer(
-    cols = ends_with("classification"),
-    names_to = "method",
-    values_to = "classification"
-  ) |> 
-  dplyr::mutate(
-    method = dplyr::case_when(
-      method == "marker_gene_classification" ~ "Marker genes only",
-      method == "copykat_classification" ~ "CopyKAT - no reference",
-      method == "cnv_proportion_classification" ~ "InferCNV proportion"
-    )
-  )
-
-ggplot(marker_density_df, aes(x = sum_transformed_exp, color = classification)) +
-  geom_density() +
-  facet_wrap(vars(method))
 ```
-
-We see a few things from this plot:
-
-- As expected given that marker gene expression was used to classify cells using the "Marker genes only" method, we see an increase in marker gene expression in tumor cells. 
-However we don't see a clear bimodal distribution, which is something we did see in `SCPCL000822`. 
-- For `InferCNV`, both tumor and normal cells have higher expression of marker genes than the reference. 
-- For `CopyKAT`, the normal cells appear to have higher expression of marker genes.
-
-Let's also look at gene set scores and see if we can use those to help us determine which cells are most likely to be tumor cells.
-
-```{r}
-# plot gene set scores for each cell  
-geneset_plot_df <- classification_df |> 
-  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean-")) |> 
-  tidyr::pivot_longer(
-    cols = starts_with("mean"),
-    names_to = "geneset",
-    values_to = "mean_score"
-  ) |> 
-  dplyr::mutate(
-    geneset = stringr::word(geneset, -1, sep = "-")
-  )
-
-ggplot(geneset_plot_df, aes(x = UMAP1, UMAP2, color = mean_score)) +
-  geom_point(size = 0.5, alpha = 0.5) +
-  facet_wrap(vars(geneset)) +
-  scale_color_viridis_c()
-```
-
-We do see that the scores across all three genesets seem to be pretty consistent. 
-Cells that have higher scores for `RIGGI`, also have higher scores for `SILIGAN` and `ZHANG`.
-
-Let's look at how the distribution of gene set scores coordinates with assignments from the CNV and marker gene methods. 
-Below, we color the distribution based on the classification method used.
-
-```{r}
-geneset_plot_df |> 
-  tidyr::pivot_longer(
-    cols = ends_with("classification"),
-    names_to = "method",
-    values_to = "classification"
-  ) |> 
-  dplyr::mutate(
-    method = dplyr::case_when(
-      method == "marker_gene_classification" ~ "Marker genes only",
-      method == "copykat_classification" ~ "CopyKAT - no reference",
-      method == "cnv_proportion_classification" ~ "InferCNV proportion"
-    )) |> 
-  ggplot(aes(x = mean_score, color = classification)) +
-  geom_density(bw = 0.05) +
-  facet_grid(rows = vars(method), 
-             cols = vars(geneset))
-```
-
-We see a few things from this plot:
-
-- The marker genes classification has the most consistent increase in gene set scores in tumor cells compared to normal cells, although the separation is not super clear.  
-- For the `RIGGI` and `ZHANG` gene sets, the scores are higher in both tumor and normal cells compared to reference cells as classified by `InferCNV`. 
-- For all gene sets, the normal cells have a slightly higher gene set score as classified by `CopyKAT`.
-
-## Using gene expression to classify tumor cells
-
-Unlike with `SCPCL000822`, there is no clear consensus between CNV inference methods and marker gene based classification. 
-Because these genomes are quiet, we expect CNV inference to be more difficult and less likely to accurately predict the tumor cells. 
-Therefore, we should use methods that rely on gene expression.
-
-The rest of this notebook will explore using different gene set based methods to classify tumor cells to see if we can more accurately identify the tumor cells in this sample.
-
-### Comparing marker gene expression in `SCPCL000822` and `SCPCL000824`
-
-With `SCPCL000822` we had a clear separation between marker gene expression in tumor cells and normal cells because there was a bimodal distribution. 
-That does not appear to be the case here so it's going to be more difficult to pull out tumor cells.
-
-First we will just compare the distribution of the raw marker gene expression in `SCPCL000822` and `SCPCL000824`. 
-To do this, we will get the total marker gene expression by summing all marker genes in a cell and then plot the distribution.
 
@@
 combined_markers_df <- plot_markers_df |> 
-  dplyr::select(barcodes, tumor_cell_classification = marker_gene_classification, gene_symbol, gene_expression, transformed_gene_expression, sum_raw_exp, sum_transformed_exp) |> 
-  dplyr::mutate(sample = "SCPCL000824") |> 
+  dplyr::select(barcodes, gene_symbol, gene_expression, transformed_gene_expression, sum_raw_exp, sum_transformed_exp) |> 
+  dplyr::mutate(sample = "SCPCL001114") |> 
   dplyr::bind_rows(ref_markers_df)
@@
 ```
-
-Looking at this, we see that `SCPCL000822` has a bimodal distribution, but this is not the case for `SCPCL000824`. 
-Additionally, most of the distribution for `SCPCL000824` lies within the upper distribution for `SCPCL000822`. 
-This would be consistent with our hypothesis that most of the cells in `SCPCL000824` are tumor cells. 
-Let's find the local minima in the bimodal distribution for `SCPCL000822` and then use that to classify tumor cells in `SCPCL000824`.
 
@@
 new_classification <- combined_markers_df |> 
-  dplyr::filter(sample == "SCPCL000824") |> 
-  dplyr::mutate(updated_marker_gene_classification = dplyr::if_else(sum_raw_exp >= exp_cutoff, "Tumor", "Normal")) |> 
-  dplyr::select(barcodes, updated_marker_gene_classification)
+  dplyr::filter(sample == "SCPCL001114") |> 
+  dplyr::mutate(marker_gene_classification = dplyr::if_else(sum_raw_exp >= exp_cutoff, "Tumor", "Normal")) |> 
+  dplyr::select(barcodes, marker_gene_classification) |> 
+  unique()
 
@@
 # label cells based on new classifiation 
-ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = updated_marker_gene_classification)) +
+ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = marker_gene_classification)) +
   geom_point(alpha = 0.5, size = 0.5)
 ```
-
-This looks closer to what I would expect with most cells in the large group of cells corresponding to tumor cells. 
-This also captures cells that are labeled as tumor cells by both marker genes and CNV inference.
-
-### Comparing gene set expression in `SCPCL000822` and `SCPCL000824`
-
-Now we will look at the gene set scores for each gene set in `SCPCL000822` and `SCPCL000824`.  
-The gene set scores for each cell are the mean normalized expression of all genes in a given gene set with no scaling.
-
-We did not use these to classify `SCPCL000822`, so we won't actually do any classification, but this will show us if the scores for tumor cells are similar to each other across samples.
 
@@
 ref_geneset_df <- ref_geneset_df |> 
-  dplyr::select(barcodes, starts_with("mean")) |> 
+  dplyr::select(barcodes, starts_with("mean-")) |> 
   tidyr::pivot_longer(
-    cols = starts_with("mean"),
+    cols = starts_with("mean-"),
     names_to = "geneset",
@@
 
+# plot gene set scores for each cell  
+geneset_plot_df <- classification_df |> 
+  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean-")) |> 
+  tidyr::pivot_longer(
+    cols = starts_with("mean-"),
+    names_to = "geneset",
+    values_to = "mean_score"
+  ) |> 
+  dplyr::mutate(
+    geneset = stringr::word(geneset, -1, sep = "-")
+  )
+
 # join with gene set scores from 824 and plot distribution 
@@
   dplyr::select(barcodes, geneset, mean_score) |> 
-  dplyr::mutate(sample = "SCPCL000824") |> 
+  dplyr::mutate(sample = "SCPCL001114") |> 
   dplyr::bind_rows(ref_geneset_df) |> 
@@
 
-Again we see that the gene set scores are bimodal for `SCPCL000822`, at least for `RIGGI` and `ZHANG`. 
-However, we don't see that for `SCPCL000824`, but we do see a very similar range of values. 
-This makes me think that the majority of the cells are in fact tumor cells.
-
-## Classifying tumor cells with AUCell
-
-The next thing we will do is see if we can use a more data driven approach to identify cells with higher marker gene or gene set expression. 
-We will use [`AUCell`](https://www.bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html) to do this and will run it on both samples.
-
-`AUCell` can be used to identify cells with an active gene set given a list of genes or gene signature. 
-From the [vignette](https://bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html):
-
-> AUCell uses the “Area Under the Curve” (AUC) to calculate whether a critical subset of the input gene set is enriched within the expressed genes for each cell.
-The distribution of AUC scores across all the cells allows exploring the relative expression of the signature.
-Since the scoring method is ranking-based, AUCell is independent of the gene expression units and the normalization procedure.
-In addition, since the cells are evaluated individually, it can easily be applied to bigger datasets, subsetting the expression matrix if needed.
-
-We will use `AUCell` with four different gene sets:
-
-- [`ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION`](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION.html)
-- [`RIGGI_EWING_SARCOMA_PROGENITOR_UP`](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/RIGGI_EWING_SARCOMA_PROGENITOR_UP.html?ex=1)
-- [`SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN`](https://www.gsea-msigdb.org/gsea/msigdb/cards/SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN)
-- Custom tumor marker gene list found in [`references/tumor-marker-genes.tsv`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/references/tumor-marker-genes.tsv)
-
-For each gene set above, `AUCell` will identify cells likely to have expression of that gene set returning a distribution of `auc` scores and a threshold for each gene set.
-
-One caveat is that `AUCell` relies on having a bimodal distribution in gene set expression to find a threshold, so I expect we may need to do a similar thing where we use thresholds determined in `SCPCL000822` to call cells in `SCPCL000824`.
+## Compare AUCell classification to marker gene classification
 
 ```{r}
-# names of gene sets to grab
-ews_gene_sets <- c(
-  "ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION",
-  "RIGGI_EWING_SARCOMA_PROGENITOR_UP",
-  "SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN"
+classification_df <- classification_df |> 
+  dplyr::mutate(
+    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor"),
+    auc_classification = forcats::fct_relevel(auc_classification, "Tumor")
   )
 
-# pull gene sets from msigbdr 
-# all gene sets are part of C2, CGP
-genes_df <- msigdbr::msigdbr(species= "Homo sapiens", 
-                             category = "C2",
-                             subcategory = "CGP") |> 
-  # only keep relevant gene sets 
-  dplyr::filter(gs_name %in% ews_gene_sets)
+# compare using a confusion matrix
+caret::confusionMatrix(
+    table(
+      classification_df$marker_gene_classification, 
+      classification_df$auc_classification)
+  ) 
 ```
 
+## UCell
+
 ```{r}
-# first create named list of genes
-# we will need this to run UCell later too 
-genes_list <- ews_gene_sets |> 
-  purrr::map(\(name){
-    genes <- genes_df |> 
-      dplyr::filter(gs_name == name) |> 
-      dplyr::pull(ensembl_gene)
-  }) |> 
-  purrr::set_names(ews_gene_sets)
-
-# build GeneSetCollection for AUCell 
-msig_gene_sets <- genes_list |> 
-  purrr::imap(\(genes, name) GSEABase::GeneSet(genes, setName = name))
-
 # get list of marker genes to add to GeneSets 
@@
 
-# turn it into a gene set 
-marker_gene_set <- marker_genes |> 
-  GSEABase::GeneSet(setName = "markers")
-
-# join msig and marker genes into a collection to input with AUCell 
-collection <- GSEABase::GeneSetCollection(c(msig_gene_sets, marker_gene_set))
-```
-
-```{r}
-# run AUCell
-ref_auc <- AUCell::AUCell_run(counts(ref_sce), collection)
-
-# assign cells for each gene set based on thresholds and output plots 
-ref_assignments <- AUCell::AUCell_exploreThresholds(ref_auc, assign = TRUE)
-```
-
-For each gene set, several possible thresholds are calculated. 
-The dotted lines over the histogram indicate the distribution and the corresponding vertical dotted line (same color) indicates the possible threshold. 
-The thicker line indicates the chosen threshold by the algorithm. 
-This corresponds to the highest value that reduces the false positives.
-
-Here we can see a bimodal distribution of all gene sets except `SILIGAN`, which is to be expected based on the original distribution plots.
-
-For the remaining analysis we will use the results from running `AUCell` with only the marker gene list. 
-Let's compare the assignments from our validated classifications (determined by taking the consensus between `InferCNV`, and `CopyKAT` for `SCPCL000822`) to using `AUCell` with `SCPCL000822`.
-
-```{r}
-# compare AUCell classification vs cnv classification in ref
-# pull out tumor cells found by AUCell
-tumor_cells <- ref_assignment$markers$assignment
-
-ref_classification_df <- ref_classification_df |> 
-  # add new column with auc classification 
-  dplyr::mutate(auc_classification = dplyr::if_else(barcodes %in% tumor_cells, "Tumor", "Normal"))
-
-# filter out any ambiguous cells 
-caret_df <- ref_classification_df |> 
-  dplyr::filter(tumor_cell_classification != "Ambiguous")
-
-# compare using a confusion matrix
-caret::confusionMatrix(
-    table(
-      caret_df$tumor_cell_classification, 
-      caret_df$auc_classification)
-  ) 
-```
-
-It looks like we get pretty good agreement between our original classification and the updated `AUCell` classification using the marker gene list.
-
-Now we repeat on `SCPCL000824`.
-
-```{r}
-# get new auc and plot thresholds 
-new_auc <- AUCell::AUCell_run(counts(sce), collection)
-new_assignment <- AUCell::AUCell_exploreThresholds(new_auc, assign = TRUE)
-```
-
-Here we do not see any bimodal distributions and that most cells are considered to not be expressing the specified gene sets, which doesn't fit with our previous findings. 
-I expect that without a bimodal distribution this model does not do a good job of calculating a threshold.
-
-```{r}
-ref_auc_df <- ref_auc@assays@data$AUC |> 
-  as.data.frame() |> 
-  tibble::rownames_to_column("gene_set") |> 
-  tidyr::pivot_longer(!gene_set, 
-                      names_to = "barcodes", 
-                      values_to = "auc") |> 
-  #dplyr::filter(gene_set == "markers") |> 
-  dplyr::mutate(sample = "SCPCL000822")
-
-new_auc_df <- new_auc@assays@data$AUC |> 
-  as.data.frame() |> 
-  tibble::rownames_to_column("gene_set") |> 
-  tidyr::pivot_longer(!gene_set, 
-                      names_to = "barcodes", 
-                      values_to = "auc") |> 
-  #dplyr::filter(gene_set == "markers") |> 
-  dplyr::mutate(sample = "SCPCL000824")
-
-all_auc_df <- dplyr::bind_rows(list(ref_auc_df, new_auc_df))
-
-ggplot(all_auc_df, aes(x = auc, color = sample)) +
-  geom_density() +
-  facet_grid(rows = vars(gene_set))
-```
-
-Looking at the AUC across both samples for all gene sets, we see that they are in the same range of values and tend to mirror each other. 
-The main difference here is the lack of bimodal distribution in `SCPCL000824`.
-
-Let's use the marker gene threshold identified by `AUCell` for `SCPCL000822` to identify tumor cells in `SCPCL000824`.  
-Here, we label any cells that have passed the threshold set for the marker gene set as tumor cells.
-
-```{r}
-# pull out threshold used for assigning cells from ref sample 
-ref_threshold <- ref_assignment$markers$aucThr$selected
-
-# create a vector of tumor cells in 824 using the cutoff from 822
-new_tumor_cells <- new_auc_df |> 
-  dplyr::filter(gene_set == "markers", auc >= ref_threshold) |>
-  dplyr::pull(barcodes)
-
-# add to original classification df 
-classification_df <- classification_df |> 
-  dplyr::mutate(auc_classification = dplyr::if_else(barcodes %in% new_tumor_cells, "Tumor", "Normal"))
-```
-
-```{r}
-# visualize which cells are classified as tumor 
-ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = auc_classification)) +
-  geom_point(alpha = 0.5, size = 0.5)
-```
-
-This looks similar to using the updated marker gene cutoff to classify cells where most cells in that bottom group are tumor cells.  
-As a reminder, the updated marker gene classification is the classification of tumor cells based on the marker gene expression cutoff determined in `SCPCL000822`. 
-We can confirm this by calculating the confusion matrix.
-
-```{r}
-# compare using a confusion matrix
-caret::confusionMatrix(
-    table(
-      classification_df$updated_marker_gene_classification, 
-      classification_df$auc_classification)
-  ) 
-```
-
-## Classifying tumor cells with UCell
-
-Here we will look at using [`UCell`](https://carmonalab.github.io/UCell_demo/UCell_matrix_vignette.html#content) to calculate a gene set score for each cell and then attempt to classify cells based on the distribution of those gene set scores.
-
-The [`UCell` score is calculated as follows](https://doi.org/10.1016/j.csbj.2021.06.043):
-
-1. Calculate a ranked list of genes for each cell in the dataset. 
-2. The Mann-Whitney U statistic is calculated for each cell by subsetting the ranked list by genes of interest in each cell.
-
-Again, we will look at both `SCPCL000822` and `SCPCL000824`.
-
-```{r}
 # create list to use for ucell of gene sets and marker genes
-ucell_gene_sets <- c(genes_list, markers = list(marker_genes)) |> 
+ucell_gene_sets <- c(markers = list(marker_genes)) |> 
   # only keep genes that are found in reference 
@@
   tibble::rownames_to_column("barcodes") |> 
-  dplyr::mutate(sample = "SCPCL000824") |> 
+  dplyr::mutate(sample = "SCPCL001114") |> 
   dplyr::bind_rows(ref_ucell) |> 
@@
 
-Here we see that the marker gene set maybe has a bimodal distribution in `SCPCL000822` and no bimodal distribution in `SCPCL000824`. 
-However the score in `SCPCL000824` seems to lie in the upper part of the distribution for `SCPCL000822`.
+Boo this looks bad
 
-`RIGGI`, `ZHANG`, and `SILIGAN` both show very similar distributions to the gene set scores distributions from our manual calculations of gene set scores (mean of all genes in the gene set). 
-Again, there is a bimodal distribution for `RIGGI` and `ZHANG`, but not for `SILIGAN` and it is only present in `SCPCL000822`.
+## Consensus between AUCell and marker genes
 
-Let's use the local minima from the marker gene set scores for `SCPCL000822` to identify tumor cells. 
-First we will look at how using `UCell` compares to our previous classifications for `SCPCL000822` (consensus between `CopyKAT` and `InferCNV`).
+## Validation plots
 
-```{r}
-# create distribution 
-density_data <- density(ref_ucell$markers_UCell)
-# find the local minima in the distribution
-geneset_cutoff <- optimize(approxfun(density_data$x, density_data$y), interval = c(0.01, 0.2))$minimum
-
-# get a vector of tumor cells from ref 
-ref_ucell_tumor_cells <- ucell_df |> 
-  dplyr::filter(gene_list == "markers",
-                signature_score >= geneset_cutoff,
-                sample == "SCPCL000822") |> 
-  dplyr::pull(barcodes) |> 
-  unique()
-
-# add classification column for comparison to original classification
-caret_df <- caret_df |> 
-  dplyr::mutate(
-    ucell_classification = dplyr::if_else(barcodes %in% ref_ucell_tumor_cells, "Tumor", "Normal")
-  )
-
-# confusion matrix between original and ucell 
-caret::confusionMatrix(
-  table(
-    caret_df$tumor_cell_classification, 
-    caret_df$ucell_classification 
-  )
-)
-```
-
-It looks like we get pretty good agreement between using `UCell` with the `RIGGI` gene set to the original classification. 
-Now we will use the marker gene set cutoff from `SCPCL000822` to classify tumor cells in `SCPCL000824`.
-
-```{r}
-# get list of tumor cells using ucell 
-new_ucell_tumor_cells <- ucell_df |> 
-  dplyr::filter(gene_list == "markers",
-                signature_score >= geneset_cutoff,
-                sample == "SCPCL000824") |> 
-  dplyr::pull(barcodes) |> 
-  unique()
-
-# add ucell classification
-classification_df <- classification_df |> 
-  dplyr::mutate(
-    ucell_classification = dplyr::if_else(barcodes %in% new_ucell_tumor_cells, "Tumor", "Normal")
-  )
-
-ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = ucell_classification)) +
-  geom_point(alpha = 0.5, size = 0.5)
-```
-
-This looks similar to what we get with both the updated marker gene and `AUCell` classification. 
-We can confirm that by calculating the confusion matrix.
-
-```{r}
-caret::confusionMatrix(
-  table(
-    classification_df$updated_marker_gene_classification, 
-    classification_df$ucell_classification
-  )
-)
-```
-
-## Validate annotations
-
-Finally, we will look at the marker gene and gene set scores for every cell across annotation methods.
-
-```{r}
-# create annotation df, keeping all classification methods 
-annotation_df <- classification_df |> 
-  dplyr::select(barcodes, marker_gene_classification, updated_marker_gene_classification, auc_classification, ucell_classification) |> 
-  unique()
-
-# create matrix with marker genes as rows and barcodes as columns
-marker_gene_heatmap <- plot_markers_df |> 
-  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
-  tidyr::pivot_wider(values_from = gene_expression, 
-                     names_from = barcodes) |> 
-  tibble::column_to_rownames("gene_symbol") |> 
-  as.matrix()
-
-annotation <- ComplexHeatmap::columnAnnotation(
-  marker_genes = annotation_df$marker_gene_classification,
-  updated_marker_genes = annotation_df$updated_marker_gene_classification,
-  AUCell = annotation_df$auc_classification, 
-  UCell = annotation_df$ucell_classification,
-  col = list(marker_genes = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"),
-             updated_marker_genes = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"), 
-             AUCell = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"),
-             UCell = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
-```
-
-```{r}
-# plot heatmap of marker genes 
-plot_gene_heatmap(marker_gene_heatmap, 
-                  row_title = "Marker gene symbol",
-                  legend_title = "Marker gene \nexpression",
-                  annotation = annotation)
-```
-
-Here we see that the updated marker gene classification (using raw gene expression cut offs determined from `SCPCL000822`) and `AUCell` tend to have the clearest sepration between tumor cells and normal cells that line up with expression of individual marker genes.
-
-Now we will create the same plot but with gene set scores.
-
-```{r}
-# make a matrix of gene set by barcode 
-geneset_heatmap <- geneset_plot_df |> 
-  dplyr::select(mean_score, geneset, barcodes) |> 
-  unique() |> 
-  tidyr::pivot_wider(values_from = mean_score, 
-                     names_from = barcodes) |> 
-  tibble::column_to_rownames("geneset") |> 
-  as.matrix()
-
-# plot heatmap of gene set score 
-plot_gene_heatmap(geneset_heatmap,
-                  annotation = annotation,
-                  legend_title = "Gene set \nscore")
-```
-
-Again, we see that the updated marker genes and `AUCell` have the most similar assignments and these tend to group by tumor/normal cells. 
-Although we do see a similar separation of cells in `UCell`, but there are many more cells that are classified as normal.
-
-## Save annotations
-
-We will go ahead and save the annotations from all gene expression based methods for future use.
-
-```{r}
-# export final TSV with annotations 
-classifications_output <- classification_df |> 
-  dplyr::select(
-    cell_barcode = barcodes,
-    marker_gene_classification,
-    updated_marker_gene_classification, 
-    auc_classification,
-    ucell_classification
-  ) |> 
-  unique()
-
-readr::write_tsv(classifications_output, final_annotations_file)
-```
-
-## Session Info
-
-```{r session info}
-# record the versions of the packages used in this analysis and other environment information
-sessionInfo()
-```
+## Session info

### 2024-07-12 analyses/cell-type-ewings/template_notebooks/auc-workflow/01-auc-results.Rmd
use number of genes detected and change output file name

@@
   marker_genes_file: NULL
-  output_file: NULL
+  marker_class_file: NULL
   auc_threshold: NULL
+  marker_threshold: 1
 ---
@@
 ```{r}
-# add marker gene classification based on expression of at least 1 marker gene  
+# get a list of tumor cells based on those that have number of marker genes expressed 
+# greater than the threshold 
+tumor_cells <- markers_df |> 
+  dplyr::mutate(gene_detected = dplyr:if_else(gene_expression > 0 , 1, 0)) |> 
+  dplyr::group_by(barcodes) |> 
+  dplyr::summarize(total_genes = sum(gene_detected)) |> 
+  dplyr::filter(total_genes >= params$marker_threshold) |> 
+  dplyr::pull(barcodes)
+
+# create a data frame with raw exp and classification 
+# will add this to the larger data frame
 marker_gene_class_df <- markers_df |> 
@@
     marker_gene_classification = dplyr::if_else(
-      sum_raw_exp > 0,
+      barcodes %in% tumor_cells,
       "Tumor",
@@
   dplyr::select(barcodes, marker_gene_classification = classification) |> 
-  readr::write_tsv(params$output_file)
+  readr::write_tsv(params$marker_class_file)
 ```

### 2024-07-12 analyses/cell-type-ewings/aucell-annotation.sh

@@ -90,7 +90,7 @@ for sample_id in $sample_ids; do
                         ref_auc_results = '$ref_auc_results', \
                         geneset_scores_file = '$geneset_results', \
                         marker_genes_file = '$marker_genes_file', \
-                        output_file = '${marker_gene_results}', \
+                        marker_class_file = '${marker_gene_results}', \
                         auc_threshold = $auc_threshold), \
           envir = new.env()) \
         "

### 2024-07-15 analyses/cell-type-ewings/template_notebooks/auc-workflow/01-auc-results.Rmd

@@
 tumor_cells <- markers_df |> 
-  dplyr::mutate(gene_detected = dplyr:if_else(gene_expression > 0 , 1, 0)) |> 
+  dplyr::mutate(gene_detected = as.integer(gene_expression > 0)) |> 
   dplyr::group_by(barcodes) |> 

### 2024-07-15 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd

@@
       scrub_tsv <- file.path(result_dir, glue::glue("{dataset}_scrublet.tsv"))
-      sce_file <- file.path(data_dir, dataset, glue::glue("{dataset}_sce.rds"))
+      sce_file <- file.path(data_dir, dataset, glue::glue("{dataset}.rds"))
 
@@
 
-For three out of four datasets, `scDblFinder` predicts a much larger number of doublets compared to other methods.
+For two out of four datasets, `scDblFinder` predicts a much larger number of doublets compared to other methods.
 

### 2024-07-15 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd

@@
   data_dir, 
-  glue::glue("{params$dataset}_sce.rds")
+  glue::glue("{params$dataset}.rds")
 )
@@
     table(
-      "Truth" = doublet_df$ground_truth,
-      "Prediction" = doublet_df[[pred_col]]
+      "Truth" = df$ground_truth,
+      "Prediction" = df[[pred_col]]
     ), 

### 2024-07-15 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000752/SCPCL001114_tumor-gene-expression-exploration.Rmd
compare across methods and add validation

@@
 - Marker gene expression from `SCPCL000822` to inform an appropriate cut off for marker gene expression in `SCPCL000824`. 
-- [`AUCell`](https://www.bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html) with marker genes and three EWS-FLI1 gene sets in both samples. 
-- [`UCell`](https://carmonalab.github.io/UCell_demo/UCell_matrix_vignette.html#content) with marker genes and three EWS-FLI1 gene sets in both samples.
+- [`AUCell`](https://www.bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html) with marker genes in both samples. 
+These results were obtained from running through `aucell-annotation.sh`.
+- [`UCell`](https://carmonalab.github.io/UCell_demo/UCell_matrix_vignette.html#content) with marker genes in both samples.
 
@@
 auc_results_file <- file.path(results_dir, "SCPCL001114_auc-classifications.tsv")
+marker_results_file <- file.path(results_dir, "SCPCL001114_marker-gene-classifications.tsv")
 geneset_scores_file <- file.path(results_dir, "SCPCL001114_gene-set-scores.tsv")
@@
 # output files 
-final_annotations_dir <- file.path(module_base, "results", "annotation_tables", "SCPCS000492")
+final_annotations_dir <- file.path(module_base, "results", "annotation_tables", "SCPCS000752")
 fs::dir_create(final_annotations_dir)
-final_annotations_file <- file.path(final_annotations_dir, "SCPCL000824_tumor-classifications.tsv.gz")
+final_annotations_file <- file.path(final_annotations_dir, "SCPCL001114_tumor-classifications.tsv.gz")
 
@@
 geneset_scores_df <- readr::read_tsv(geneset_scores_file)
+marker_results_df <- readr::read_tsv(marker_results_file)
 
@@
   dplyr::left_join(auc_results_df, by = "barcodes") |> 
+  dplyr::left_join(marker_results_df, by = "barcodes") |> 
   dplyr::left_join(geneset_scores_df, by = "barcodes")
@@
 ```
+
+## Using marker gene expression from `SCPCL000822` to define tumor cells in `SCPCL001114`
+
+With `SCPCL000822` we had a clear separation between marker gene expression in tumor cells and normal cells because there was a bimodal distribution. 
+First we will just compare the distribution of the raw marker gene expression in `SCPCL000822` and `SCPCL001114`. 
+To do this, we will get the total marker gene expression by summing all marker genes in a cell and then plot the distribution.
 
@@
 
+Looking at this, we see that `SCPCL000822` has a bimodal distribution, but this is not the case for `SCPCL001114`. 
+Additionally, most of the distribution for `SCPCL001114` lies within the upper distribution for `SCPCL000822`. 
+This would be consistent with our hypothesis that most of the cells in `SCPCL001114` are tumor cells. 
+Let's find the local minima in the bimodal distribution for `SCPCL000822` and then use that to classify tumor cells in `SCPCL001114`.
+
 ```{r}
@@
 # add new column with updated marker gene classification 
-# use local minima from 822 to define 824
+# use local minima from 822 to define 1114
 new_classification <- combined_markers_df |> 
   dplyr::filter(sample == "SCPCL001114") |> 
-  dplyr::mutate(marker_gene_classification = dplyr::if_else(sum_raw_exp >= exp_cutoff, "Tumor", "Normal")) |> 
-  dplyr::select(barcodes, marker_gene_classification) |> 
+  dplyr::mutate(ref_marker_gene_classification = dplyr::if_else(sum_raw_exp >= exp_cutoff, "Tumor", "Normal")) |> 
+  dplyr::select(barcodes, ref_marker_gene_classification) |> 
   unique()
@@
 # label cells based on new classifiation 
-ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = marker_gene_classification)) +
+ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = ref_marker_gene_classification)) +
   geom_point(alpha = 0.5, size = 0.5)
 ```
+
+## Comparing gene set expression in `SCPCL000822` and `SCPCL001114`
+
+Now we will look at the gene set scores for each gene set in `SCPCL000822` and `SCPCL001114`.  
+The gene set scores for each cell are the mean normalized expression of all genes in a given gene set with no scaling.
+
+We did not use these to classify `SCPCL000822`, so we won't actually do any classification, but this will show us if the scores for tumor cells are similar to each other across samples.
 
@@
 
-## Compare AUCell classification to marker gene classification
+Again we see that the gene set scores are bimodal for `SCPCL000822`, at least for `RIGGI` and `ZHANG`. 
+However, we don't see that for `SCPCL001114`, but we do see that the majority of the distribution seems to line up with the upper distribution in `SCPCL000822`.
+Or at least this is the case for `RIGGI` and `ZHANG`.
+This makes me think that the majority of the cells are in fact tumor cells.
+
+## Comparison to other classification methods
+
+This section will compare using the marker gene cutoff determined from `SCPCL000822` to other classification methods. 
+First we will compare the calls between each individual method and the marker gene method. 
+Then, we will look at all methods together and plot the expression of marker genes and gene set scores for normal and tumor cells.
+
+### `AUCell`
+
+[`AUCell`](https://www.bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html) was run previously using the `aucell-annotation.sh` workflow. 
+In that workflow, we ran `AUCell` on both a reference library, `SCPCL000822` and the query library, `SCPCL001114`. 
+The AUC value determined by running `AUCell` with `SCPCL000822` was used as a cutoff for identifying tumor cells in `SCPCL001114`.
+
+Here we will compare the classifications obtained by using `AUCell`. 
+The classification using marker gene expression from `SCPCL000822` is the rows and the classification using `AUCell` is the columns.
 
@@
   dplyr::mutate(
-    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor"),
+    ref_marker_gene_classification = forcats::fct_relevel(ref_marker_gene_classification, "Tumor"),
     auc_classification = forcats::fct_relevel(auc_classification, "Tumor")
@@
     table(
-      classification_df$marker_gene_classification, 
+      classification_df$ref_marker_gene_classification, 
       classification_df$auc_classification)
@@
 
-## UCell
+It looks like using `AUCell` results in more cells being called normal than with the marker gene cutoff determined by `SCPCL000822`.
+
+### Expression of at least one marker gene
+
+In the `aucell-annotation.sh` workflow, we looked at marker gene expression and classified any cell that expresses at least 1 marker gene as a tumor cell. 
+Let's compare the annotations obtained here using a cutoff determined by `SCPCL000822` to setting the cutoff of 1 marker gene.
+
+First we will look at a confusion matrix between both methods. 
+Here the classification using expression of at least 1 marker gene is the rows and the updated classification based on `SCPCL000822` is the columns.
 
 ```{r}
-# get list of marker genes to add to GeneSets 
+classification_df <- classification_df |> 
+  dplyr::mutate(
+    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
+  )
+
+# compare using a confusion matrix
+caret::confusionMatrix(
+    table(
+      classification_df$marker_gene_classification, 
+      classification_df$ref_marker_gene_classification)
+  ) 
+```
+
+It looks like using `SCPCL000822` to define the cutoff results in more cells being called normal than with the strict cutoff of expressing at least 1 marker gene.
+
+### `UCell`
+
+Here we will look at using [`UCell`](https://carmonalab.github.io/UCell_demo/UCell_matrix_vignette.html#content) to calculate a gene set score for each cell and then attempt to classify cells based on the distribution of those gene set scores.
+
+```{r}
+# get list of marker genes
 marker_genes <- readr::read_tsv(marker_genes_file, show_col_types = FALSE) |> 
@@
 
-# create list to use for ucell of gene sets and marker genes
-ucell_gene_sets <- c(markers = list(marker_genes)) |> 
-  # only keep genes that are found in reference 
-  purrr::map(\(geneset){
-    intersect(geneset, rownames(sce))
-  })
+# create list to use for ucell
+ucell_gene_sets <- c(markers = list(marker_genes))
 
-# run ucell on both ref (822) and 824
+# run ucell on both ref (822) and 1114
 ref_ucell <- UCell::ScoreSignatures_UCell(counts(ref_sce), features = ucell_gene_sets)
@@
 
-Boo this looks bad
+It looks like there is a bimodal distribution for `SCPCL000822` and not for `SCPCL001114` which is consistent with all the other marker gene methods we have looked at. 
+However, here we see that the signature scores for `SCPCL001114` tend to fall in the middle of the upper and lower distribution for `SCPCL000822`. 
+Because of this I don't think using `UCell` for this library would be very helpful.
 
-## Consensus between AUCell and marker genes
+### Consensus between `AUCell` and marker genes
 
-## Validation plots
+We already saw that all cells called as tumor in `AUCell` were also called as tumor using the marker gene cutoff from `SCPCL000822`. 
+However, there are a group of normal cells that are called as tumor using marker genes but normal using `AUCell`. 
+Here we will take the consensus between both methods. 
+We will define any cells that are tumor in both as `Tumor`, normal in both as `Normal`, and any cells that are called differently in both as `Ambiguous`.
+
+```{r}
+# add new column with consensus to classification df 
+classification_df <- classification_df |> 
+  dplyr::mutate(consensus_classification = dplyr::case_when(
+    ref_marker_gene_classification == "Tumor" & auc_classification == "Tumor" ~ "Tumor",
+    ref_marker_gene_classification == "Normal" & auc_classification == "Normal" ~ "Normal",
+    .default = "Ambiguous"
+  ))
+```
+
+## Validate classifications from all methods
+
+First let's look at which cells are called tumor or normal on the UMAP across all classification methods.
+
+```{r}
+method_levels <- c(
+  "1 marker gene" = "marker_gene_classification",
+  "Ref marker gene" = "ref_marker_gene_classification",
+  "AUCell" = "auc_classification", 
+  "Consensus" = "consensus_classification"
+)
+
+classification_df |> 
+  tidyr::pivot_longer(
+    cols = ends_with("classification"),
+    names_to = "method",
+    values_to = "classification"
+  ) |> 
+  dplyr::mutate(
+    method = forcats::fct_recode(method, !!!method_levels)) |> 
+  ggplot(aes(x = UMAP1, y = UMAP2, color = classification)) +
+  geom_point(size = 0.1, alpha = 0.5) +
+  facet_wrap(vars(method)) +
+  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
+```
+
+Next, we compare the expression of marker genes and gene set scores across tumor and normal cells for each classification method. 
+We expect to see higher expression of marker genes and gene set scores in tumor cells.
+
+```{r}
+# pull out any new classification methods we added 
+new_classification_df <- classification_df |> 
+  dplyr::select(barcodes, ref_marker_gene_classification, consensus_classification)
+
+# create a density plot showing the distribution of marker gene expression across classification methods 
+marker_density_df <- plot_markers_df |>
+  # add new methods 
+  dplyr::left_join(new_classification_df, by = c("barcodes")) |> 
+  tidyr::pivot_longer(
+    cols = ends_with("classification"),
+    names_to = "method",
+    values_to = "classification"
+  ) |> 
+  dplyr::mutate(
+    method = forcats::fct_recode(method, !!!method_levels)
+  )
+
+ggplot(marker_density_df, aes(x = sum_transformed_exp, color = classification)) +
+  geom_density() +
+  facet_wrap(vars(method))
+```
+
+```{r fig.height=7}
+geneset_plot_df |> 
+  # add new methods 
+  dplyr::left_join(new_classification_df, by = c("barcodes", "ref_marker_gene_classification")) |> 
+  tidyr::pivot_longer(
+    cols = ends_with("classification"),
+    names_to = "method",
+    values_to = "classification"
+  ) |> 
+  dplyr::mutate(
+    method = forcats::fct_recode(method, !!!method_levels)) |> 
+  ggplot(aes(x = mean_score, color = classification)) +
+  geom_density(bw = 0.05) +
+  facet_grid(rows = vars(method), 
+             cols = vars(geneset))
+```
+
+Looking at the above plots, we see the most distinct separation in either of the marker gene methods. 
+`AUCell` appears to have the most overlap between distributions in normal and tumor cells. 
+We also see that the consensus shows the "ambiguous" cells falling right in the middle. 
+This tells us that the group of cells being called normal in `AUCell` and tumor in the marker gene methods that appear to have lower expression than those that are called tumor by both.
+
+We can look at the same information but on an individual cell level. 
+In the below plots we expect to see some separate clustering of tumor and normal cells.
+
+```{r}
+# create annotation df, keeping all classification methods 
+annotation_df <- classification_df |> 
+  dplyr::select(barcodes, ends_with("classification")) |> 
+  unique()
+
+plot_colors <- c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")
+
+annotation <- ComplexHeatmap::columnAnnotation(
+  marker_genes = annotation_df$marker_gene_classification,
+  ref_marker_genes = annotation_df$ref_marker_gene_classification,
+  AUCell = annotation_df$auc_classification, 
+  consensus = annotation_df$consensus_classification,
+  col = list(marker_genes = plot_colors,
+             ref_marker_genes = plot_colors, 
+             AUCell = plot_colors,
+             consensus = plot_colors))
+```
+
+```{r}
+# create matrix with marker genes as rows and barcodes as columns
+marker_gene_heatmap <- plot_markers_df |> 
+  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
+  tidyr::pivot_wider(values_from = gene_expression, 
+                     names_from = barcodes) |> 
+  tibble::column_to_rownames("gene_symbol") |> 
+  as.matrix()
+
+# plot heatmap of marker genes 
+plot_gene_heatmap(marker_gene_heatmap, 
+                  row_title = "Marker gene symbol",
+                  legend_title = "Marker gene \nexpression",
+                  annotation = annotation)
+```
+
+```{r}
+# make a matrix of gene set by barcode 
+geneset_heatmap <- geneset_plot_df |> 
+  dplyr::select(mean_score, geneset, barcodes) |> 
+  unique() |> 
+  tidyr::pivot_wider(values_from = mean_score, 
+                     names_from = barcodes) |> 
+  tibble::column_to_rownames("geneset") |> 
+  as.matrix()
+
+# plot heatmap of gene set score 
+plot_gene_heatmap(geneset_heatmap,
+                  annotation = annotation,
+                  legend_title = "Gene set \nscore")
+```
+
+Again we see that the ambigous cells tend to lie in the middle between tumor and normal cells. 
+We also see that `AUCell` probably has the most "normal" cells that are mixing in with the tumor cells, while using a strict marker gene cutoff of expressing 1 marker gene leads to the biggest separation.
+
+Moving forward I tend to favor using either just the reference based marker gene cutoff or the consensus between that and `AUCell`. 
+Because we see such variable expression in our marker genes, I'm not sold on the idea of setting a firm cutoff at expression of 1 gene, while the other methods look at expression of all genes together.
+
+## Export results
+
+We will save the classification from all methods for future use.
+
+```{r}
+# export final TSV with annotations 
+classifications_output <- classification_df |> 
+  dplyr::select(
+    cell_barcode = barcodes,
+    marker_gene_classification,
+    ref_marker_gene_classification, 
+    auc_classification,
+    consensus_classification
+  )
+
+readr::write_tsv(classifications_output, final_annotations_file)
+```
 
 ## Session info
+
+```{r session info}
+# record the versions of the packages used in this analysis and other environment information
+sessionInfo()
+```

### 2024-07-15 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
spelling fixes

@@
 # results from annotation workflow 
-results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000490")
+results_dir <- file.path(module_base, "results", "cnv_annotation", "SCPCS000490")
 
@@
 
-Next, we will look at the gene set scores for three different `EWS-FLI1` gene sets. heirarchical
+Next, we will look at the gene set scores for three different `EWS-FLI1` gene sets. hierarchical
 The below plots will all use the `mean-GENESET` columns.

### 2024-07-15 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd
spelling fixes

@@
 # results from annotation workflow 
-results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000492")
+results_dir <- file.path(module_base, "results", "cnv_annotation", "SCPCS000492")
 
@@
 ref_labels_file <- file.path(module_base, "results", "annotation_tables", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv")
-ref_geneset_scores_file <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
+ref_geneset_scores_file <- file.path(module_base, "results", "cnv_annotation", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
 ```
@@
 
-We do see that the scores across all three genesets seem to be pretty consistent. 
+We do see that the scores across all three gene sets seem to be pretty consistent. 
 Cells that have higher scores for `RIGGI`, also have higher scores for `SILIGAN` and `ZHANG`.
@@
 
-## Classifying tumor cells with AUCell
+## Classifying tumor cells with `AUCell`
 
@@
 
-> AUCell uses the “Area Under the Curve” (AUC) to calculate whether a critical subset of the input gene set is enriched within the expressed genes for each cell.
+> `AUCell` uses the “Area Under the Curve” (AUC) to calculate whether a critical subset of the input gene set is enriched within the expressed genes for each cell.
 The distribution of AUC scores across all the cells allows exploring the relative expression of the signature.
@@
 
-## Classifying tumor cells with UCell
+## Classifying tumor cells with `UCell`
 
@@
 
-Here we see that the updated marker gene classification (using raw gene expression cut offs determined from `SCPCL000822`) and `AUCell` tend to have the clearest sepration between tumor cells and normal cells that line up with expression of individual marker genes.
+Here we see that the updated marker gene classification (using raw gene expression cut offs determined from `SCPCL000822`) and `AUCell` tend to have the clearest separation between tumor cells and normal cells that line up with expression of individual marker genes.
 

### 2024-07-16 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
update output tables for 822 and 824

@@
 # output files 
-final_annotations_dir <- file.path(module_base, "cell_type_annotations", "SCPCS000490")
+final_annotations_dir <- file.path(module_base, "results", "annotation_tables", "SCPCS000490")
 fs::dir_create(final_annotations_dir)
-final_annotations_file <- file.path(final_annotations_dir, "SCPCL000822_tumor-classifications.tsv")
+final_annotations_file <- file.path(final_annotations_dir, "SCPCL000822_tumor-classifications.tsv.gz")
 ```
@@
 plot_markers_df <- create_marker_gene_df(
+  sce,
   classification_df, 
@@
 
-ggplot(marker_density_df, aes(x = sum_exp, color = classification)) +
+ggplot(marker_density_df, aes(x = sum_raw_exp, color = classification)) +
   geom_density() +
@@
 
-Next, we will look at the gene set scores for three different `EWS-FLI1` gene sets. hierarchical
+Next, we will look at the gene set scores for three different `EWS-FLI1` gene sets. heirarchical
 The below plots will all use the `mean-GENESET` columns.
@@
   dplyr::select(
-    cell_barcode = barcodes,
-    tumor_cell_classification = combined_cnv_classification
+    "cell_barcode" = barcodes,
+    combined_cnv_classification
   ) |> 

### 2024-07-16 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd
update output tables for 822 and 824

--- 2024-07-03 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd
@@
 ref_sce_file <- file.path(data_dir, "SCPCP000015", "SCPCS000490", "SCPCL000822_processed.rds")
-ref_labels_file <- file.path(module_base, "results", "annotation_tables", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv")
-ref_geneset_scores_file <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
+ref_labels_file <- file.path(module_base, "results", "annotation_tables", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv.gz")
+ref_geneset_scores_file <- file.path(module_base, "results", "cnv_annotation", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
 ```
@@
 caret_df <- ref_classification_df |> 
-  dplyr::filter(tumor_cell_classification != "Ambiguous") |> 
+  dplyr::filter(combined_cnv_classification != "Ambiguous") |> 
   # make sure positive class is tumor
   dplyr::mutate(
-    tumor_cell_classification = forcats::fct_relevel(tumor_cell_classification, "Tumor"),
+    cnv_classification = forcats::fct_relevel(combined_cnv_classification, "Tumor"),
     auc_classification = forcats::fct_relevel(auc_classification, "Tumor")
@@
     table(
-      caret_df$tumor_cell_classification, 
+      caret_df$cnv_classification, 
       caret_df$auc_classification)
@@
     ucell_classification = dplyr::if_else(barcodes %in% ref_ucell_tumor_cells, "Tumor", "Normal"), 
-    tumor_cell_classification = forcats::fct_relevel(tumor_cell_classification, "Tumor"),
+    tumor_cell_classification = forcats::fct_relevel(combined_cnv_classification, "Tumor"),
     ucell_classification = forcats::fct_relevel(ucell_classification, "Tumor")
@@
 Finally, we will look at the marker gene and gene set scores for every cell across annotation methods. 
+We will also add in annotations that are the consensus between the updated marker gene and `AUCell`. 
+Cells that are called tumor in both will be "Tumor" and normal in both will be called "Normal". 
+All other cells will be labeled as "Ambiguous".
+
+```{r}
+classification_df <- classification_df |> 
+  dplyr::mutate(consensus_classification = dplyr::case_when(
+    updated_marker_gene_classification == "Tumor" & auc_classification == "Tumor" ~ "Tumor",
+    updated_marker_gene_classification == "Normal" & auc_classification == "Normal" ~ "Normal",
+    .default = "Ambiguous"
+  ))
+```
 
@@
 annotation_df <- classification_df |> 
-  dplyr::select(barcodes, marker_gene_classification, updated_marker_gene_classification, auc_classification, ucell_classification) |> 
+  dplyr::select(barcodes, marker_gene_classification, updated_marker_gene_classification, auc_classification, ucell_classification, consensus_classification) |> 
   unique()
@@
 
+plot_colors <- c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")
+
 annotation <- ComplexHeatmap::columnAnnotation(
@@
   UCell = annotation_df$ucell_classification,
-  col = list(marker_genes = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"),
-             updated_marker_genes = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"), 
-             AUCell = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"),
-             UCell = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
+  consensus = annotation_df$consensus_classification,
+  col = list(marker_genes = plot_colors,
+             updated_marker_genes =plot_colors, 
+             AUCell = plot_colors,
+             UCell = plot_colors,
+             consensus = plot_colors))
 ```
@@
     auc_classification,
-    ucell_classification
+    ucell_classification,
+    consensus_classification
   ) |> 

### 2024-07-16 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000752/SCPCL001114_tumor-gene-expression-exploration.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 
-In this notebook we attempt to identify tumor cells in `SCPCL001114` using the following orthogonal marker gene based methods:
+In this notebook we attempt to identify tumor cells in `SCPCL001114` using the following marker gene based methods:
 
@@
     table(
-      classification_df$marker_gene_classification, 
-      classification_df$ref_marker_gene_classification)
+      classification_df$ref_marker_gene_classification,
+      classification_df$marker_gene_classification)
   ) 

### 2024-07-16 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000752/SCPCL001114_tumor-gene-expression-exploration.Rmd
use two marker genes

@@
 ref_sce_file <- file.path(data_dir, "SCPCP000015", "SCPCS000490", "SCPCL000822_processed.rds")
-ref_labels_file <- file.path(module_base, "results", "annotation_tables", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv")
+ref_labels_file <- file.path(module_base, "results", "annotation_tables", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv.gz")
 ref_geneset_scores_file <- file.path(module_base, "results", "cnv_annotation", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
@@
   dplyr::left_join(marker_results_df, by = "barcodes") |> 
-  dplyr::left_join(geneset_scores_df, by = "barcodes")
+  dplyr::left_join(geneset_scores_df, by = "barcodes") |> 
+  # rename since we are going to be looking at multiple marker gene methods 
+  dplyr::rename(
+    "one_marker_gene_classification" = marker_gene_classification
+  )
 
@@
 First we will look at a confusion matrix between both methods. 
-Here the classification using expression of at least 1 marker gene is the rows and the updated classification based on `SCPCL000822` is the columns.
+Here the updated classification based on `SCPCL000822` is the rows and the classification using expression of at least 1 marker gene is the columns.
 
@@
   dplyr::mutate(
-    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
+    one_marker_gene_classification = forcats::fct_relevel(one_marker_gene_classification, "Tumor")
   )
@@
       classification_df$ref_marker_gene_classification,
-      classification_df$marker_gene_classification)
+      classification_df$one_marker_gene_classification)
   ) 
@@
 It looks like using `SCPCL000822` to define the cutoff results in more cells being called normal than with the strict cutoff of expressing at least 1 marker gene.
+
+### Expression of at least two marker genes
+
+Below we will look at classifying cells based on expression of two marker genes rather than one.
+
+```{r}
+# get list of tumor cells with > 2 marker genes expressed 
+tumor_cells <- plot_markers_df |> 
+  dplyr::mutate(gene_detected = as.integer(gene_expression > 0)) |> 
+  dplyr::group_by(barcodes) |> 
+  dplyr::summarize(total_genes = sum(gene_detected)) |> 
+  dplyr::filter(total_genes >= 2) |> 
+  dplyr::pull(barcodes)
+
+classification_df <- classification_df |> 
+  dplyr::mutate(
+    two_marker_gene_classification = dplyr::if_else(
+      barcodes %in% tumor_cells,
+      "Tumor",
+      "Normal"
+    ))
+```
+
+```{r}
+ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = two_marker_gene_classification)) +
+  geom_point(alpha = 0.5, size = 0.5)
+```
+
+Now let's compare the annotations obtained using a cutoff determined by `SCPCL000822` to setting the cutoff of 2 marker genes. 
+Here the updated classification based on `SCPCL000822` is the rows and the classification using expression of at least 2 marker genes is the columns.
+
+```{r}
+classification_df <- classification_df |> 
+  dplyr::mutate(
+    two_marker_gene_classification = forcats::fct_relevel(two_marker_gene_classification, "Tumor")
+  )
+
+# compare using a confusion matrix
+caret::confusionMatrix(
+    table(
+      classification_df$ref_marker_gene_classification,
+      classification_df$two_marker_gene_classification)
+  ) 
+```
+
+It looks like using `SCPCL000822` to define the cutoff results in more cells being called normal than with the strict cutoff of expressing at least 2 marker genes.
 
@@
 method_levels <- c(
-  "1 marker gene" = "marker_gene_classification",
+  "1 marker gene" = "one_marker_gene_classification",
+  "2 marker genes" = "two_marker_gene_classification",
   "Ref marker gene" = "ref_marker_gene_classification",
@@
 new_classification_df <- classification_df |> 
-  dplyr::select(barcodes, ref_marker_gene_classification, consensus_classification)
+  dplyr::select(barcodes, ref_marker_gene_classification, consensus_classification, two_marker_gene_classification)
 
@@
 annotation <- ComplexHeatmap::columnAnnotation(
-  marker_genes = annotation_df$marker_gene_classification,
+  one_marker = annotation_df$one_marker_gene_classification,
+  two_marker = annotation_df$two_marker_gene_classification,
   ref_marker_genes = annotation_df$ref_marker_gene_classification,
@@
   consensus = annotation_df$consensus_classification,
-  col = list(marker_genes = plot_colors,
+  col = list(one_marker = plot_colors,
+             two_marker = plot_colors,
              ref_marker_genes = plot_colors, 
@@
     cell_barcode = barcodes,
-    marker_gene_classification,
+    one_marker_gene_classification,
+    two_marker_gene_classification,
     ref_marker_gene_classification, 

### 2024-07-17 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_singler-classification.Rmd

@@
 # gene set scores to use for validation 
-results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000492")
+results_dir <- file.path(module_base, "results", "cnv_annotation", "SCPCS000492")
 geneset_scores_file <- file.path(results_dir, "SCPCL000824_gene-set-scores.tsv")

### 2024-07-17 analyses/cell-type-ewings/template_notebooks/cnv-workflow/01-marker-genes.Rmd

--- 2024-07-10 analyses/cell-type-ewings/template_notebooks/cnv-workflow/01-marker-genes.Rmd
@@
 ```{r base paths}
+# The path to this module
+module_base <- rprojroot::find_root(rprojroot::is_renv_project)
+
+# repo base is two levels up from module
+repository_base <- dirname(dirname(module_base))
+
 # The current data directory, found within the repository base directory
-data_dir <- file.path("..", "..", "data", "current")
+data_dir <- file.path(repository_base, "data", "current")
 sample_dir <- file.path(data_dir, "SCPCP000015", params$sample_id)
-
-# The path to this module
-module_base <- file.path("analyses", "cell-type-ewings")
 ```

### 2024-07-17 analyses/cell-type-ewings/template_notebooks/cnv-workflow/04-infercnv.Rmd

--- 2024-07-10 analyses/cell-type-ewings/template_notebooks/cnv-workflow/04-infercnv.Rmd
@@
 
-Those cells in gray (bottom left) are cells used as a reference. 
+Those cells in grey (bottom left) are cells used as a reference. 
 If we were to use total CNVs we would need some sort of cut off for the number. 
@@
 
-We will call any tumor cell as those with a total number of CNVs greater than the mean(number of CNVs).
+We will call any tumor cell as those with a total number of cnvs greater than the mean(number of cnvs).
 
@@
     cnv_proportion_classification = cnv_mean_classification,
-    cnv_sum_classification = cnv_classification,
-    mean_cnv_proportion = mean_proportion
+    cnv_sum_classification = cnv_classification
   )

### 2024-07-17 analyses/cell-type-ewings/annotate-tumor-cells.sh
Revert "merge main into reproducibility-docs and fix conflicts"
This reverts commit ac0ef609b60c5db0b8c0c88c999f7debb081fe7d, reversing
changes made to 7087ee3a069da705b7e5ae3dab92a31150460381.

@@ -1,0 +1,253 @@
+#!/bin/bash
+
+####
+: '
+This workflow will identify tumor cells in a single library.
+Tumor cell annotations are obtained by:
+
+- manually identifying cells that express marker genes
+- Calculating gene set scores for three EWS-FLI1 target gene sets
+- Running CellAssign with a list of tumor marker genes
+- Running CopyKAT to identify aneuploid cells
+- Running InverCNV to identify potential tumor cells (those with a high proportion of CNVs)
+
+Before running this workflow, you must download both the processed `SingleCellExperiment` objects and `AnnData` objects with `download-data.py`.
+Do this for any samples you would like to run through this workflow.
+
+The following arguments are optional and can be used to run this workflow on additional samples (Default sample is "SCPCS000490"):
+
+sample_id: Unique sample ID (name of folder containing libray data)
+normal_celltypes: Comma separated list of cell types annotated with either `SingleR` or `CellAssign` to use as a reference list of normal cells.
+tumor_celltypes: Comma separated list of cell typs annotated with either `SingleR` or `CellAssign` that are expected to align with tumor cells.
+Any cell types used here will be used for comparing to tumor cells annotated in this workflow.
+data_dir: Directory containing SCE and AnnData objects to use for a given project.
+This is the full path to the project folder.
+Default is `../../data/current/SCPCP000015`.
+workflow_results_dir: Full path to folder to store outputs from the workflow.
+By default, all results will be stored in `results/annotate_tumor_cells_output`.
+
+A set of reports summarizing the tumor cell annotations and a TSV file containing annotations from all used methods will be saved to `results/annotate_tumor_cells_output`.
+
+Example of running the workflow with a different sample:
+
+sample_id="SCPCS000491" ./annotate-tumor-cells.sh
+
+'
+####
+
+set -euo pipefail
+
+# this script lives in the root of the module directory
+# use this to define other default paths
+cd $(dirname "$0")
+module_directory=$(pwd)
+
+# input variables
+sample_id=${sample_id:-"SCPCS000490"}
+normal_celltypes=${normal_celltypes:-"Endothelial cells,endothelial cell"}
+tumor_celltypes=${tumor_celltypes:-"Pulmonary vascular smooth muscle cells,smooth muscle cell"}
+data_dir=${data_dir:-"../../data/current/SCPCP000015"}
[…]

### 2024-07-17 analyses/cell-type-ewings/cnv-annotation.sh

@@ -1,246 +1,0 @@
-#!/bin/bash
-
-####
-: '
-This workflow will identify tumor cells in a single library.
-Tumor cell annotations are obtained by:
-
-- manually identifying cells that express marker genes
-- Calculating gene set scores for three EWS-FLI1 target gene sets
-- Running CellAssign with a list of tumor marker genes
-- Running CopyKAT to identify aneuploid cells
-- Running InverCNV to identify potential tumor cells (those with a high proportion of CNVs)
-
-Before running this workflow, you must download both the processed `SingleCellExperiment` objects and `AnnData` objects with `download-data.py`.
-Do this for any samples you would like to run through this workflow.
-
-The following arguments are optional and can be used to run this workflow on additional samples (Default sample is "SCPCS000490"):
-
-sample_id: Unique sample ID (name of folder containing libray data)
-normal_celltypes: Comma separated list of cell types annotated with either `SingleR` or `CellAssign` to use as a reference list of normal cells.
-tumor_celltypes: Comma separated list of cell typs annotated with either `SingleR` or `CellAssign` that are expected to align with tumor cells.
-Any cell types used here will be used for comparing to tumor cells annotated in this workflow.
-data_dir: Directory containing SCE and AnnData objects to use for a given project.
-This is the full path to the project folder.
-Default is `../../data/current/SCPCP000015`.
-workflow_results_dir: Full path to folder to store outputs from the workflow.
-By default, all results will be stored in `results/cnv_annotation`.
-
-A set of reports summarizing the tumor cell annotations and a TSV file containing annotations from all used methods will be saved to `results/cnv_annotation`.
-
-Example of running the workflow with a different sample:
-
-sample_id="SCPCS000491" ./cnv-annotation.sh
-
-'
-####
-
-set -euo pipefail
-
-# this script lives in the root of the module directory
-# use this to define other default paths
-cd $(dirname "$0")
-module_directory=$(pwd)
-
-# input variables
-sample_id=${sample_id:-"SCPCS000490"}
-normal_celltypes=${normal_celltypes:-"Endothelial cells,endothelial cell"}
-tumor_celltypes=${tumor_celltypes:-"Pulmonary vascular smooth muscle cells,smooth muscle cell"}
-data_dir=${data_dir:-"../../data/current/SCPCP000015"}
[…]

### 2024-07-17 analyses/cell-type-ewings/scripts/auc-workflow/02-calculate-gene-set-scores.R

@@ -1,121 +1,0 @@
-#!/usr/bin/env Rscript
-
-# This script is used to calculate gene set scores for a set of EWS-FLI1 gene sets from MsigDB
-# ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION - https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION.html
-# RIGGI_EWING_SARCOMA_PROGENITOR_UP - https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/RIGGI_EWING_SARCOMA_PROGENITOR_UP.html?ex=1
-# SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN https://www.gsea-msigdb.org/gsea/msigdb/cards/SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN
-
-# output is a TSV where each cell is a row and each gene set score is a column 
-# calculates the sum, mean, z-scaled(sum), and z-scaled(mean) for each geneset 
-
-project_root <- here::here()
-
-suppressPackageStartupMessages({
-  library(optparse)
-  library(SingleCellExperiment)
-})
-
-option_list <- list(
-  make_option(
-    opt_str = c("--sce_file"),
-    type = "character",
-    help = "Path to RDS file containing a processed SingleCellExperiment object from scpca-nf"
-  ),
-  make_option(
-    opt_str = c("--output_file"),
-    type = "character",
-    default = NULL,
-    help = "Full path to TSV file to save gene set scores"
-  )
-)
-
-# Parse options
-opt <- parse_args(OptionParser(option_list = option_list))
-
-# Define function for calculating gene set score -------------------------------
-
-# function to calculate sum and mean of all genes in a given geneset 
-# genes_df is a dataframe with one row per gene and a column `gs_name`
-# All genes that have `gs_name == geneset_name` are included in the calculation
-calculate_score <- function(sce, genes_df, geneset_name){
-  
-  # pull out genes that belong to specified geneset 
-  genes <- genes_df |> 
-    dplyr::filter(gs_name == geneset_name,
-                  ensembl_gene %in% rownames(sce)) |> 
-    dplyr::pull(ensembl_gene)
-  
-  # get expression data for all genes in geneset
-  gene_exp <- logcounts(sce[genes, ]) |>
-    as.matrix() |>
-    t() |>
-    as.data.frame()
-  
[…]

### 2024-07-17 analyses/cell-type-ewings/scripts/calculate-gene-set-scores.R

@@ -1,0 +1,124 @@
+#!/usr/bin/env Rscript
+
+# This script is used to calculate gene set scores for a set of EWS-FLI1 gene sets from MsigDB
+# ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION - https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION.html
+# RIGGI_EWING_SARCOMA_PROGENITOR_UP - https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/RIGGI_EWING_SARCOMA_PROGENITOR_UP.html?ex=1
+# SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN https://www.gsea-msigdb.org/gsea/msigdb/cards/SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN
+
+# output is a TSV where each cell is a row and each gene set score is a column 
+# calculates the sum, mean, z-scaled(sum), and z-scaled(mean) for each geneset 
+
+project_root <- here::here()
+renv::load(project_root)
+
+library(optparse)
+library(SingleCellExperiment)
+
+option_list <- list(
+  make_option(
+    opt_str = c("--sce_file"),
+    type = "character",
+    help = "Path to RDS file containing a processed SingleCellExperiment object from scpca-nf"
+  ),
+  make_option(
+    opt_str = c("--results_dir"),
+    type = "character",
+    default = NULL,
+    help = "Full path to folder to save gene set scores"
+  )
+)
+
+# Parse options
+opt <- parse_args(OptionParser(option_list = option_list))
+
+# Define function for calculating gene set score -------------------------------
+
+# function to calculate sum and mean of all genes in a given geneset 
+# genes_df is a dataframe with one row per gene and a column `gs_name`
+# All genes that have `gs_name == geneset_name` are included in the calculation
+calculate_score <- function(sce, genes_df, geneset_name){
+  
+  # pull out genes that belong to specified geneset 
+  genes <- genes_df |> 
+    dplyr::filter(gs_name == geneset_name,
+                  ensembl_gene %in% rownames(sce)) |> 
+    dplyr::pull(ensembl_gene)
+  
+  # get expression data for all genes in geneset
+  gene_exp <- logcounts(sce[genes, ]) |>
+    as.matrix() |>
+    t() |>
+    as.data.frame()
+  
+  # create data frame with sum, mean, and scaled values 
[…]

### 2024-07-17 analyses/cell-type-ewings/scripts/cnv-workflow/00-generate-cellassign-refs.R

@@ -1,145 +1,0 @@
-#!/usr/bin/env Rscript
-
-# this script is used to create binary gene x cell marker genes matrices
-# for use as references with CellAssign
-
-project_root <- here::here()
-
-library(optparse)
-
-option_list <- list(
-  make_option(
-    opt_str = "--all_markers",
-    type = "character",
-    default = file.path(project_root, "references", "visser-all-marker-genes.tsv"),
-    help = "Path to marker gene file with markers for tumor cells and all normal cells from Visser et al"
-  ),
-  make_option(
-    opt_str = c("--panglao_file"),
-    type = "character",
-    default = file.path(project_root, "references", "PanglaoDB_markers_2020-03-27.tsv")
-  ),
-  make_option(
-    opt_str = c("--gtf_file"),
-    type = "character",
-    default = "s3://scpca-references/homo_sapiens/ensembl-104/annotation/Homo_sapiens.GRCh38.104.gtf.gz",
-    help = "URI to gtf file."
-  ),
-  make_option(
-    opt_str = c("--output_dir"),
-    type = "character",
-    default = file.path(project_root, "references", "cellassign_refs"),
-    help = "Directory to save all reference matrix files"
-  ),
-  make_option(
-    opt_str = c("--scratch_dir"),
-    type = "character",
-    default = file.path(project_root, "scratch"),
-    help = "Path to store copied GTF file"
-  )
-)
-
-opt <- parse_args(OptionParser(option_list = option_list))
-
-# Functions --------------------------------------------------------------------
-
-# function for creating a binary matrix with cell types as columns and marker genes as rows 
-build_binary_df <- function(marker_genes_df) {
-  
-  # check column names exist 
-  stopifnot("cell_type and ensembl_gene_id must be columns in marker genes file" = 
-              c("cell_type", "ensembl_gene_id") %in% colnames(marker_genes_df))
-  
-  # build binary df 
-  binary_df <- marker_genes_df |>
-    dplyr::select(cell_type, ensembl_gene_id) |> 
-    unique() |> 
-    tidyr::pivot_wider(
-      id_cols = ensembl_gene_id,
-      names_from = cell_type,
-      values_from = cell_type,
-      values_fn = length,
-      values_fill = 0
-    ) |>
-    # add a column with no marker genes
[…]

### 2024-07-17 analyses/cell-type-ewings/scripts/cnv-workflow/00-make-gene-order-file.R

@@ -1,58 +1,0 @@
-#!/usr/bin/env Rscript
-
-project_root <- here::here()
-
-library(optparse)
-
-option_list <- list(
-  make_option(
-    opt_str = c("--gtf_file"),
-    type = "character",
-    default = "s3://scpca-references/homo_sapiens/ensembl-104/annotation/Homo_sapiens.GRCh38.104.gtf.gz",
-    help = "URI to gtf file to create gene order file."
-  ),
-  make_option(
-    opt_str = c("--local_ref_dir"),
-    type = "character",
-    default = file.path(project_root, "references", "infercnv_refs"), 
-    help = "Directory to use for saving GTF file and gene order file."
-  ), 
-  make_option(
-    opt_str = c("--scratch_dir"),
-    type = "character",
-    default = file.path(project_root, "scratch"),
-    help = "Path to store copied GTF file"
-  )
-)
-
-# Parse options
-opt <- parse_args(OptionParser(option_list = option_list))
-
-# create ref directory if doesn't already exist 
-fs::dir_create(opt$local_ref_dir)
-
-# sync gtf file to local directory 
-gtf_filename <- basename(opt$gtf_file)
-
-local_gtf_file <- file.path(opt$scratch_dir, gtf_filename)
-if(!file.exists(local_gtf_file)){
-  sync_call <- glue::glue('aws s3 cp {opt$gtf_file} {local_gtf_file} --no-sign-request')
-  system(sync_call)
-}
-
-# define gene order file name using gtf file 
-gtf_basename <- stringr::str_remove(gtf_filename, ".gtf.gz")
-gene_order_filename <- glue::glue("{gtf_basename}.gene_order.txt")
-gene_order_file <- file.path(opt$local_ref_dir, gene_order_filename)
-
-# read in gtf file
-gtf <- rtracklayer::import(local_gtf_file, feature.type = "gene")
-
-# format gene order file 
-gtf_df <- gtf |> 
-  as.data.frame() |> 
-  dplyr::select(gene_id, seqnames, start, end) |> 
-  dplyr::mutate(seqnames = glue::glue("chr{seqnames}"))
-
-# export gene order file 
-readr::write_tsv(gtf_df, gene_order_file, col_names = FALSE)

### 2024-07-17 analyses/cell-type-ewings/scripts/cnv-workflow/01-select-cell-types.R

@@ -1,95 +1,0 @@
-#!/usr/bin/env Rscript
-
-# This script is used to generate a table of reference cells for a given ScPCA libray that are annotated by 
-# SingleR and/or CellAssign to have a specific label
-# The output is a table that contains columns for barcodes, reference cell class, 
-# and then columns for each of the original cell type annotations for that cell 
-# The `reference_cell_class` column will contain either tumor or normal 
-
-project_root <- here::here()
-
-suppressPackageStartupMessages({
-  library(optparse)
-  library(SingleCellExperiment)
-})
-
-option_list <- list(
-  make_option(
-    opt_str = c("--sce_file"),
-    type = "character",
-    help = "Path to RDS file containing a processed SingleCellExperiment object from scpca-nf"
-  ),
-  make_option(
-    opt_str = c("--normal_cells"),
-    type = "character",
-    default = "",
-    help = "Comma separated list of cell types to pull from annotations present and save as normal cells."
-  ),
-  make_option(
-    opt_str = c("--tumor_cells"),
-    type = "character",
-    default = "",
-    help = "Comma separated list of cell types to pull from annotations present and save as tumor cells."
-  ),
-  make_option(
-    opt_str = c("--output_filename"),
-    type = "character",
-    help = "Full path to file to store saved table of cell barcodes."
-  )
-)
-
-# Parse options
-opt <- parse_args(OptionParser(option_list = option_list))
-
-# Set up -----------------------------------------------------------------------
-
-stopifnot(
-  # make sure path to sce file exists
-  "sce_file does not exist" = file.exists(opt$sce_file),
-
-  # check that either normal or tumor cell types are provided 
-  "Either normal_cells or tumor_cells must be provided" = 
-    opt$normal_cells != "" || opt$tumor_cells != ""
-)       
-
-# read in sce file
-sce <- readr::read_rds(opt$sce_file)
-
-# Create table of reference cells ----------------------------------------------
-
-# possible annotation columns 
-annotation_columns <- c(
-  "singler_celltype_annotation",
-  "cellassign_celltype_annotation"
-)
-
-# get dataframe with barcodes and associated annotation 
-coldata_df <- colData(sce) |>
-  as.data.frame() |> 
-  dplyr::select(barcodes, all_of(annotation_columns)) |> 
-  tidyr::pivot_longer(all_of(annotation_columns),
-                      names_to = "celltype_method",
[…]

### 2024-07-17 analyses/cell-type-ewings/scripts/cnv-workflow/03-run-copykat.R

@@ -1,118 +1,0 @@
-#!/usr/bin/env Rscript
-
-# This script is used to run CopyKAT on a given ScPCA library 
-# The output will be found in `outdir`/`results_prefix` and all files will be labeled with the `library_id`
-
-project_root <- here::here()
-
-library(optparse)
-library(copykat)
-library(SingleCellExperiment)
-
-option_list <- list(
-  make_option(
-    opt_str = c("--sce_file"),
-    type = "character",
-    help = "Path to RDS file containing a processed SingleCellExperiment object from scpca-nf"
-  ),
-  make_option(
-    opt_str = c("--reference_cell_file"),
-    type = "character",
-    default = NULL,
-    help = "Optional file that contains a table of cell barcodes. 
-      Any cells with Normal in the reference_cell_class column will be used as a reference for CopyKAT"
-  ),
-  make_option(
-    opt_str = c("--results_dir"),
-    type = "character",
-    default = NULL,
-    help = "Full path to folder to save final CopyKAT results"
-  ),
-  make_option(
-    opt_str = c("--scratch_dir"),
-    type = "character",
-    default = file.path(project_root, "scratch", "copykat"),
-    help = "path to folder where all intermediate CopyKAT results live"
-  ),
-  make_option(
-    opt_str = c("-t", "--threads"),
-    type = "integer",
-    default = 1,
-    help = "Number of multiprocessing threads to use."
-  )
-)
-
-# Parse options
-opt <- parse_args(OptionParser(option_list = option_list))
-
-# Set up -----------------------------------------------------------------------
-
-# make sure path to sce file exists
-stopifnot("sce_file does not exist" = file.exists(opt$sce_file))
-
-# check that results dir was provided 
-stopifnot("Must provide a --results_dir to save results" = !is.null(opt$results_dir))
-
-# read in sce file
-sce <- readr::read_rds(opt$sce_file)
-
-# get library id to use as prefix for all output file names 
-library_id <- metadata(sce)$library_id
-
-# contstruct and create output folder if not already present 
-fs::dir_create(opt$results_dir)
-
-# define scratch directory 
-scratch_dir <- file.path(opt$scratch_dir, library_id)
-fs::dir_create(scratch_dir)
-
-# path to output copykat object 
-copykat_output_obj <- file.path(opt$results_dir, glue::glue("{library_id}_final-copykat.rds"))
-
[…]

### 2024-07-17 analyses/cell-type-ewings/scripts/cnv-workflow/04-run-infercnv.R

@@ -1,158 +1,0 @@
-#!/usr/bin/env Rscript
-
-# this script is used to run inferCNV on a processed SCE file 
-
-project_root <- here::here()
-
-library(optparse)
-library(SingleCellExperiment)
-
-option_list <- list(
-  make_option(
-    opt_str = c("--sce_file"),
-    type = "character",
-    help = "Path to RDS file containing a processed SingleCellExperiment object from scpca-nf"
-  ),
-  make_option(
-    opt_str = c("--annotations_file"),
-    type = "character",
-    help = "Path to save annotations file as tab delimited .txt file with no column headers. 
-      First column is cell name and second column is annotations (normal or unknown)."
-  ),
-  make_option(
-    opt_str = c("--gene_order_file"),
-    type = "character",
-    default = file.path(project_root, "references", "infercnv_refs", "Homo_sapiens.GRCh38.104.gene_order.txt"),
-    help = "Path to gene order file as tab delimited .txt file with no column headers.
-      Columns are: Ensembl gene id, chr, start, stop."
-  ),
-  make_option(
-    opt_str = c("--reference_cell_file"),
-    type = "character",
-    default = NULL,
-    help = "Optional file that contains a table of cell barcodes. 
-      Any cells with Normal in the reference_cell_class column will be used as a reference for InferCNV"
-  ),
-  make_option(
-    opt_str = c("--output_dir"),
-    type = "character",
-    default = NULL,
-    help = "Folder to save final infercnv results"
-  ),
-  make_option(
-    opt_str = c("--scratch_dir"),
-    type = "character",
-    default = file.path(project_root, "scratch", "infercnv"),
-    help = "path to scratch directory to store intermediate files from infercnv"
-  ), 
-  make_option(
-    opt_str = c("-t", "--threads"),
-    type = "integer",
-    default = 1,
-    help = "Number of multiprocessing threads to use."
-  )
-)
-
-# Parse options
-opt <- parse_args(OptionParser(option_list = option_list))
-
-
-# Set up -----------------------------------------------------------------------
-
-# make sure path to sce file exists
-stopifnot("--sce_file does not exist." = file.exists(opt$sce_file))
-
-# read in sce file
-sce <- readr::read_rds(opt$sce_file)
-
-# get library id and construct scratch directory
-library_id <- metadata(sce)$library_id
-
-# create output and scratch directory if not already present for library id
-fs::dir_create(opt$output_dir)
-
[…]

### 2024-07-17 analyses/cell-type-ewings/scripts/generate-cellassign-refs.R

@@ -1,0 +1,146 @@
+#!/usr/bin/env Rscript
+
+# this script is used to create binary gene x cell marker genes matrices
+# for use as references with CellAssign
+
+project_root <- here::here()
+renv::load(project_root)
+
+library(optparse)
+
+option_list <- list(
+  make_option(
+    opt_str = "--all_markers",
+    type = "character",
+    default = file.path(project_root, "references", "visser-all-marker-genes.tsv"),
+    help = "Path to marker gene file with markers for tumor cells and all normal cells from Visser et al"
+  ),
+  make_option(
+    opt_str = c("--panglao_file"),
+    type = "character",
+    default = file.path(project_root, "references", "PanglaoDB_markers_2020-03-27.tsv")
+  ),
+  make_option(
+    opt_str = c("--gtf_file"),
+    type = "character",
+    default = "s3://scpca-references/homo_sapiens/ensembl-104/annotation/Homo_sapiens.GRCh38.104.gtf.gz",
+    help = "URI to gtf file."
+  ),
+  make_option(
+    opt_str = c("--output_dir"),
+    type = "character",
+    default = file.path(project_root, "references", "cellassign_refs"),
+    help = "Directory to save all reference matrix files"
+  ),
+  make_option(
+    opt_str = c("--scratch_dir"),
+    type = "character",
+    default = file.path(project_root, "scratch"),
+    help = "Path to store copied GTF file"
+  )
+)
+
+opt <- parse_args(OptionParser(option_list = option_list))
+
+# Functions --------------------------------------------------------------------
+
+# function for creating a binary matrix with cell types as columns and marker genes as rows 
+build_binary_df <- function(marker_genes_df) {
+  
+  # check column names exist 
+  stopifnot("cell_type and ensembl_gene_id must be columns in marker genes file" = 
+              c("cell_type", "ensembl_gene_id") %in% colnames(marker_genes_df))
+  
+  # build binary df 
+  binary_df <- marker_genes_df |>
+    dplyr::select(cell_type, ensembl_gene_id) |> 
+    unique() |> 
+    tidyr::pivot_wider(
+      id_cols = ensembl_gene_id,
+      names_from = cell_type,
+      values_from = cell_type,
+      values_fn = length,
+      values_fill = 0
+    ) |>
+    # add a column with no marker genes
[…]

### 2024-07-17 analyses/cell-type-ewings/scripts/make-gene-order-file.R

@@ -1,0 +1,59 @@
+#!/usr/bin/env Rscript
+
+project_root <- here::here()
+renv::load(project_root)
+
+library(optparse)
+
+option_list <- list(
+  make_option(
+    opt_str = c("--gtf_file"),
+    type = "character",
+    default = "s3://scpca-references/homo_sapiens/ensembl-104/annotation/Homo_sapiens.GRCh38.104.gtf.gz",
+    help = "URI to gtf file to create gene order file."
+  ),
+  make_option(
+    opt_str = c("--local_ref_dir"),
+    type = "character",
+    default = file.path(project_root, "references", "infercnv_refs"), 
+    help = "Directory to use for saving GTF file and gene order file."
+  ), 
+  make_option(
+    opt_str = c("--scratch_dir"),
+    type = "character",
+    default = file.path(project_root, "scratch"),
+    help = "Path to store copied GTF file"
+  )
+)
+
+# Parse options
+opt <- parse_args(OptionParser(option_list = option_list))
+
+# create ref directory if doesn't already exist 
+fs::dir_create(opt$local_ref_dir)
+
+# sync gtf file to local directory 
+gtf_filename <- basename(opt$gtf_file)
+
+local_gtf_file <- file.path(opt$scratch_dir, gtf_filename)
+if(!file.exists(local_gtf_file)){
+  sync_call <- glue::glue('aws s3 cp {opt$gtf_file} {local_gtf_file} --no-sign-request')
+  system(sync_call)
+}
+
+# define gene order file name using gtf file 
+gtf_basename <- stringr::str_remove(gtf_filename, ".gtf.gz")
+gene_order_filename <- glue::glue("{gtf_basename}.gene_order.txt")
+gene_order_file <- file.path(opt$local_ref_dir, gene_order_filename)
+
+# read in gtf file
+gtf <- rtracklayer::import(local_gtf_file, feature.type = "gene")
+
+# format gene order file 
+gtf_df <- gtf |> 
+  as.data.frame() |> 
+  dplyr::select(gene_id, seqnames, start, end) |> 
+  dplyr::mutate(seqnames = glue::glue("chr{seqnames}"))
+
+# export gene order file 
+readr::write_tsv(gtf_df, gene_order_file, col_names = FALSE)

### 2024-07-17 analyses/cell-type-ewings/scripts/run-copykat.R

@@ -1,0 +1,119 @@
+#!/usr/bin/env Rscript
+
+# This script is used to run CopyKAT on a given ScPCA library 
+# The output will be found in `outdir`/`results_prefix` and all files will be labeled with the `library_id`
+
+project_root <- here::here()
+renv::load(project_root)
+
+library(optparse)
+library(copykat)
+library(SingleCellExperiment)
+
+option_list <- list(
+  make_option(
+    opt_str = c("--sce_file"),
+    type = "character",
+    help = "Path to RDS file containing a processed SingleCellExperiment object from scpca-nf"
+  ),
+  make_option(
+    opt_str = c("--reference_cell_file"),
+    type = "character",
+    default = NULL,
+    help = "Optional file that contains a table of cell barcodes. 
+      Any cells with Normal in the reference_cell_class column will be used as a reference for CopyKAT"
+  ),
+  make_option(
+    opt_str = c("--results_dir"),
+    type = "character",
+    default = NULL,
+    help = "Full path to folder to save final CopyKAT results"
+  ),
+  make_option(
+    opt_str = c("--scratch_dir"),
+    type = "character",
+    default = file.path(project_root, "scratch", "copykat"),
+    help = "path to folder where all intermediate CopyKAT results live"
+  ),
+  make_option(
+    opt_str = c("-t", "--threads"),
+    type = "integer",
+    default = 1,
+    help = "Number of multiprocessing threads to use."
+  )
+)
+
+# Parse options
+opt <- parse_args(OptionParser(option_list = option_list))
+
+# Set up -----------------------------------------------------------------------
+
+# make sure path to sce file exists
+stopifnot("sce_file does not exist" = file.exists(opt$sce_file))
+
+# check that results dir was provided 
+stopifnot("Must provide a --results_dir to save results" = !is.null(opt$results_dir))
+
+# read in sce file
+sce <- readr::read_rds(opt$sce_file)
+
+# get library id to use as prefix for all output file names 
+library_id <- metadata(sce)$library_id
+
+# contstruct and create output folder if not already present 
+fs::dir_create(opt$results_dir)
+
+# define scratch directory 
+scratch_dir <- file.path(opt$scratch_dir, library_id)
+fs::dir_create(scratch_dir)
+
+# path to output copykat object 
[…]

### 2024-07-17 analyses/cell-type-ewings/scripts/select-cell-types.R

@@ -1,0 +1,96 @@
+#!/usr/bin/env Rscript
+
+# This script is used to generate a table of reference cells for a given ScPCA libray that are annotated by 
+# SingleR and/or CellAssign to have a specific label
+# The output is a table that contains columns for barcodes, reference cell class, 
+# and then columns for each of the original cell type annotations for that cell 
+# The `reference_cell_class` column will contain either tumor or normal 
+
+project_root <- here::here()
+renv::load(project_root)
+
+suppressPackageStartupMessages({
+  library(optparse)
+  library(SingleCellExperiment)
+})
+
+option_list <- list(
+  make_option(
+    opt_str = c("--sce_file"),
+    type = "character",
+    help = "Path to RDS file containing a processed SingleCellExperiment object from scpca-nf"
+  ),
+  make_option(
+    opt_str = c("--normal_cells"),
+    type = "character",
+    default = "",
+    help = "Comma separated list of cell types to pull from annotations present and save as normal cells."
+  ),
+  make_option(
+    opt_str = c("--tumor_cells"),
+    type = "character",
+    default = "",
+    help = "Comma separated list of cell types to pull from annotations present and save as tumor cells."
+  ),
+  make_option(
+    opt_str = c("--output_filename"),
+    type = "character",
+    help = "Full path to file to store saved table of cell barcodes."
+  )
+)
+
+# Parse options
+opt <- parse_args(OptionParser(option_list = option_list))
+
+# Set up -----------------------------------------------------------------------
+
+stopifnot(
+  # make sure path to sce file exists
+  "sce_file does not exist" = file.exists(opt$sce_file),
+
+  # check that either normal or tumor cell types are provided 
+  "Either normal_cells or tumor_cells must be provided" = 
+    opt$normal_cells != "" || opt$tumor_cells != ""
+)       
+
+# read in sce file
+sce <- readr::read_rds(opt$sce_file)
+
+# Create table of reference cells ----------------------------------------------
+
+# possible annotation columns 
+annotation_columns <- c(
+  "singler_celltype_annotation",
+  "cellassign_celltype_annotation"
+)
+
+# get dataframe with barcodes and associated annotation 
+coldata_df <- colData(sce) |>
+  as.data.frame() |> 
+  dplyr::select(barcodes, all_of(annotation_columns)) |> 
+  tidyr::pivot_longer(all_of(annotation_columns),
[…]

### 2024-07-17 analyses/cell-type-ewings/scripts/utils/jaccard-functions.R

@@ -48,72 +48,3 @@ make_jaccard_matrix <- function(celltype_df, colname1, colname2) {
   
   return(jaccard_matrix)
 }
-
-# function that turns jaccard matrices into a list of heatmaps 
-make_heatmap_list <- function(jaccard_matrices, column_title, legend_match){
- 
-  # Set heatmap padding option
-  heatmap_padding <- 0.2
-  ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))
-  # list of heatmaps looking at SingleR/ CellAssign vs tumor/normal 
-  heatmap <- jaccard_matrices |>
-    purrr::imap(
-      \(celltype_mat, celltype_method) {
-        ComplexHeatmap::Heatmap(
-          t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
-          col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
-          border = TRUE,
-          ## Row parameters
-          cluster_rows = TRUE,
-          row_title = celltype_method,
-          row_title_gp = grid::gpar(fontsize = 12),
-          row_title_side = "left",
-          row_names_side = "left",
-          row_dend_side = "right",
-          row_names_gp = grid::gpar(fontsize = 10),
-          ## Column parameters
-          cluster_columns = TRUE,
-          column_title = column_title,
-          column_title_gp = grid::gpar(fontsize = 12),
-          column_names_side = "bottom",
-          column_names_gp = grid::gpar(fontsize = 10),
-          column_names_rot = 90,
-          ## Legend parameters
-          heatmap_legend_param = list(
-            title = "Jaccard index",
-            direction = "vertical",
-            legend_width = unit(1.5, "in")
-          ),
-          show_heatmap_legend = celltype_method == legend_match,
-        )
-      }) |>
-    # concatenate vertically into HeatmapList object
-    purrr::reduce(ComplexHeatmap::`%v%`)
-  
-  return(heatmap)
-}
-
-# function to plot jaccard matrices to compare one set of annotations to other methods 
-# `annotation_column` will be the columns and will be compared to each method listed in `methods_to_compare` 
-plot_jaccard <- function(classification_df,
-                         annotation_column, # single column of annotations to compare to all methods
-                         methods_to_compare, # list of methods to compare to annotations
-                         column_title, # title for columns
[…]

### 2024-07-17 analyses/cell-type-ewings/scripts/utils/tumor-validation-helpers.R

@@ -1,201 +1,0 @@
-# These are functions used to prepare data frames for plotting when doing tumor cell validataion 
-
-# Data frame prep --------------------------------------------------------------
-
-create_classification_df <- function(
-    sce,
-    marker_gene_results_file, # path to marker gene classifications
-    copykat_predictions_file, # path to copykat predictions
-    infercnv_predictions_file, # path to infercnv predictions
-    infercnv_metadata_file, # path to full infercnv metadata
-    geneset_scores_file # path to gene set scores results
-){
- 
-  # read in marker gene results 
-  marker_gene_results_df <- readr::read_tsv(marker_gene_results_file, show_col_types = FALSE)
-  
-  # copykat output
-  copykat_results_df <- readr::read_tsv(copykat_predictions_file, show_col_types = FALSE) |> 
-    # only select no_ref classification and rename to be consistent with naming of other classifications
-    dplyr::select(barcodes, copykat_classification = no_ref, mean_cnv_detection) |> 
-    dplyr::distinct() |> 
-    dplyr::mutate(
-      copykat_classification = dplyr::case_when(
-        copykat_classification == "aneuploid" ~ "Tumor",
-        copykat_classification == "diploid" ~ "Normal",
-        .default = copykat_classification
-      )
-    )
-  
-  # infercnv output 
-  infercnv_results_df <- readr::read_tsv(infercnv_predictions_file, show_col_types = FALSE)
-  # use read.delim to ensure row names do not get read in as a column 
-  infercnv_metadata_df <- read.delim(infercnv_metadata_file) |> 
-    tibble::rownames_to_column("barcodes")
-  
-  # remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
-  chr_columns_to_keep <- which(!stringr::str_detect(names(infercnv_metadata_df), "chrMT|chrGL"))
-  infercnv_metadata_df <- infercnv_metadata_df[, chr_columns_to_keep]
-  
-  # gene set scores 
-  geneset_scores_df <- readr::read_tsv(geneset_scores_file, show_col_types = FALSE)
-  
-  # pull out the UMAP coordinates and make a data frame to use for plotting
-  classification_df <- sce |> 
-    scuttle::makePerCellDF(use.dimred = "UMAP") |> 
-    # replace UMAP.1 with UMAP1
[…]

### 2024-07-17 analyses/doublet-detection/run_doublet-detection-benchmark.sh

@@ -1,62 +1,0 @@
-#!/bin/bash
-
-# This script runs the benchmarking portion of the `doublet-detection` module
-
-set -euo pipefail
-
-
-# Set up --------------
-
-# Ensure script is being run from its directory
-MODULE_DIR=$(dirname "${BASH_SOURCE[0]}")
-cd ${MODULE_DIR}
-
-TEMPLATE_NB_DIR="template-notebooks" # directory with template notebooks
-EXPLORE_NB_DIR="exploratory-notebooks"  # directory with exploratory notebooks
-
-# Create benchmark directories
-DATA_DIR="scratch/benchmark-datasets"
-RESULTS_DIR="results/benchmark-results"
-RESULTS_NB_DIR="${RESULTS_DIR}/rendered-notebooks"
-mkdir -p ${DATA_DIR}
-mkdir -p ${RESULTS_DIR}
-mkdir -p ${RESULTS_NB_DIR}
-
-
-# define benchmarking datasets to use
-bench_datasets=("hm-6k" "pbmc-1B-dm" "pdx-MULTI" "HMEC-orig-MULTI")
-
-# Download and unzip `real_datasets.zip` archive from https://doi.org/10.5281/zenodo.4562782
-# Files are saved in $DATA_DIR/raw
-wget https://zenodo.org/records/4562782/files/real_datasets.zip
-unzip real_datasets.zip -d ${DATA_DIR}/raw
-rm real_datasets.zip
-
-for dataset in "${bench_datasets[@]}"; do
-
-    # formatted SCE and AnnData files will be saved here
-    DATASET_DIR=${DATA_DIR}/$dataset
-    mkdir -p $DATASET_DIR
-
-    # Read raw downloaded data and export SCE, AnnData files
-    ./scripts/00_format-benchmark-data.R --dataset ${dataset} --input_dir ${DATA_DIR}/raw --output_dir ${DATASET_DIR}
-
-    # Infer doublets with scDblFinder
-    ./scripts/01a_run-scdblfinder.R --input_sce_file ${DATASET_DIR}/${dataset}.rds --results_dir ${RESULTS_DIR} --benchmark
-
-    # Infer doublets with scrublet
-    ./scripts/01b_run-scrublet.py --input_anndata_file ${DATASET_DIR}/${dataset}.h5ad --results_dir ${RESULTS_DIR}
-
-    # Explore each individual set of doublet results
-    Rscript -e "rmarkdown::render('${TEMPLATE_NB_DIR}/02_explore-benchmark-results.Rmd',
[…]

### 2024-07-17 analyses/doublet-detection/run_doublet-detection-scpca.sh

@@ -1,33 +1,0 @@
-#!/bin/bash
-
-# This script runs doublet detection over ScPCA data for a given project
-# Usage: ./run_doublet-detection-scpca.sh {scpca project id}
-
-set -euo pipefail
-
-
-# Ensure script is being run from its directory
-MODULE_DIR=$(dirname "${BASH_SOURCE[0]}")
-cd ${MODULE_DIR}
-
-PROJECT_ID=$1
-
-# Define directories
-DATA_DIR="../../data/current"
-RESULTS_DIR="results/scpca-results/${PROJECT_ID}"
-mkdir -p ${RESULTS_DIR}
-
-# Detect doublets on each processed SCE file in each sample directory
-for SAMPLE_DIR in ${DATA_DIR}/${PROJECT_ID}/SCPCS*; do
-    SAMPLE_ID=$(basename $SAMPLE_DIR)
-    echo "Processing ${SAMPLE_ID}..."
-
-    SAMPLE_RESULTS_DIR=${RESULTS_DIR}/${SAMPLE_ID}
-    mkdir -p ${SAMPLE_RESULTS_DIR}
-
-    for SCE_FILE in ${SAMPLE_DIR}/*_processed.rds; do
-        Rscript scripts/01a_run-scdblfinder.R \
-            --input_sce_file ${SCE_FILE} \
-            --results_dir ${SAMPLE_RESULTS_DIR}
-    done
-done

### 2024-07-17 analyses/doublet-detection/run_doublet-detection.sh

@@ -1,0 +1,55 @@
+#!/bin/bash
+
+set -euo pipefail
+
+# Set up --------------
+
+# Ensure script is being run from its directory
+MODULE_DIR=$(dirname "${BASH_SOURCE[0]}")
+cd ${MODULE_DIR}
+
+NB_DIR=template-notebooks # directory with template notebooks
+###########################################################################################
+########## Step 1: Benchmark doublet detection methods on ground truth datasets ###########
+###########################################################################################
+
+# Create benchmark directories
+BENCH_DATA_DIR=scratch/benchmark-datasets
+BENCH_RESULTS_DIR=results/benchmark-results
+BENCH_NB_DIR=${BENCH_RESULTS_DIR}/rendered-notebooks
+mkdir -p ${BENCH_DATA_DIR}
+mkdir -p ${BENCH_RESULTS_DIR}
+mkdir -p ${BENCH_NB_DIR}
+
+
+# define benchmarking datasets to use
+bench_datasets=("hm-6k" "pbmc-1B-dm" "pdx-MULTI" "HMEC-orig-MULTI")
+
+# Download and unzip `real_datasets.zip` archive from https://doi.org/10.5281/zenodo.4562782
+# Files are saved in $BENCH_DATA_DIR/raw
+wget https://zenodo.org/records/4562782/files/real_datasets.zip
+unzip real_datasets.zip -d ${BENCH_DATA_DIR}/raw
+rm real_datasets.zip
+
+for dataset in "${bench_datasets[@]}"; do
+
+    # formatted SCE and AnnData files will be saved here
+    DATASET_DIR=${BENCH_DATA_DIR}/$dataset
+    mkdir -p $DATASET_DIR
+
+    # Read raw downloaded data and export SCE, AnnData files
+    ./scripts/00_format-benchmark-data.R --dataset_name ${dataset} --input_dir ${BENCH_DATA_DIR}/raw --output_dir ${DATASET_DIR}
+
+    # Infer doublets with scDblFinder
+    ./scripts/01a_run-scdblfinder.R --dataset_name ${dataset} --data_dir ${DATASET_DIR} --results_dir ${BENCH_RESULTS_DIR}
+
+    # Infer doublets with scrublet
+    ./scripts/01b_run-scrublet.py --dataset_name ${dataset} --data_dir ${DATASET_DIR} --results_dir ${BENCH_RESULTS_DIR}
+
+    # Explore each individual set of doublet results
[…]

### 2024-07-17 analyses/doublet-detection/scripts/00_format-benchmark-data.R

@@ -12,7 +12,10 @@
 #   - In the SCE file, this is in the colData slot
 #   - In the AnnData file, this is in the obs slot
 
-# Load libraries
+# Load renv environment and libraries
+project_root <- rprojroot::find_root(rprojroot::is_renv_project)
+renv::load(project_root)
+
 library(SingleCellExperiment)
 library(optparse)
 
@@ -44,8 +47,8 @@ if (!file.exists(input_file)) {
   )
 }
 
-output_sce_file <- file.path(opts$output_dir, glue::glue("{opts$dataset_name}.rds"))
-output_anndata_file <- file.path(opts$output_dir, glue::glue("{opts$dataset_name}.h5ad"))
+output_sce_file <- file.path(opts$output_dir, glue::glue("{opts$dataset_name}_sce.rds"))
+output_anndata_file <- file.path(opts$output_dir, glue::glue("{opts$dataset_name}_anndata.h5ad"))
 
 dat <- readRDS(input_file)
 mat <- dat[[1]] # raw counts matrix

### 2024-07-17 analyses/doublet-detection/scripts/01a_run-scdblfinder.R

@@ -1,24 +1,24 @@
 #!/usr/bin/env Rscript
 
 # This script runs scDblFinder on an SCE file and exports a TSV file of results.
-# If the SCE has fewer than 10 droplets, results will not be calculated and the exported TSV will contain NA values.
 
-# Load libraries ------
-suppressPackageStartupMessages({
-  library(SingleCellExperiment)
-  library(optparse)
-})
+
+# Load libraries and renv environment ------
+project_root <- rprojroot::find_root(rprojroot::is_renv_project)
+renv::load(project_root)
+
+library(SingleCellExperiment)
+library(optparse)
+
 
 # Functions -----------
 
 #' Run scDblFinder on an SCE object
 #'
 #' @param sce SCE object to run scDblFinder on
-#' @param cores Number of cores to specify to BiocParallel::MulticoreParam()
-#' @param sample_var Required for SCE object with multiple samples, the SCE colData variable that indicates sample of origin.
-#'   This option should _not_ be used for a multiplexed library which has not been demultiplexed, but is suitable for merged objects.
-#' @param random_seed Required for SCE object with multiple samples, a random seed to ensure reproducibility when running on multiple samples.
-#' @param columns_to_keep Vector of columns in the final scDblFinder table to keep in the returned table. No checking is done on these column names.
+#' @param cores Number of cores to specify to BiocParallel::MulticoreParam(
+#' @param sample_var Required for SCE object with multiple samples, the SCE colData variable that indicates sample of origin
+#' @param random_seed Required for SCE object with multiple samples, a random seed to ensure reproducibility when running on multiple samples
 #' @param ... Additional arguments to pass to scDblFinder::scDblFinder
 #'
 #' @return Data frame object with scDblFinder inferences
@@ -26,7 +26,6 @@ run_scdblfinder <- function(sce,
                             cores = 4,
                             sample_var = NULL,
                             random_seed = NULL,
-                            columns_to_keep = c("barcodes", "score", "class"),
                             ...) {
 
   # first check multiple samples, which requires a random seed
@@ -38,7 +37,7 @@ run_scdblfinder <- function(sce,
     }
     if (is.null(random_seed)) {
[…]

### 2024-07-17 analyses/doublet-detection/scripts/01b_run-scrublet.py

@@ -38,10 +38,16 @@ def main() -> None:
         description="Detect doublets on a set of AnnData objects using scrublet.",
     )
     parser.add_argument(
-        "--input_anndata_file",
+        "--dataset_name",
         type=str,
         required=True,
-        help="Path to the input AnnData file to process."
+        help="Name of dataset to process, where the associated file is expected to be named `{name}_anndata.h5ad`."
+    )
+    parser.add_argument(
+        "--data_dir",
+        type=Path,
+        required=True,
+        help="The directory containing H5AD input file."
     )
     parser.add_argument(
         "--results_dir",
@@ -57,22 +63,21 @@ def main() -> None:
     )
     args = parser.parse_args()
 
-    # Define and check files, directories
     args.results_dir.mkdir(parents = True, exist_ok = True)
 
-    input_anndata_file = Path(args.input_anndata_file)
-    if not input_anndata_file.exists():
+    # Run scrublet and export the results
+    input_anndata = args.data_dir / (args.dataset_name + "_anndata.h5ad")
+    result_tsv = args.results_dir / (args.dataset_name + "_scrublet.tsv")
+
+    if not input_anndata.exists():
         print(
             "The input AnnData file could not be found at:",
-            args.input_anndata_file,
+            input_anndata,
             file=sys.stderr
         )
         sys.exit(1)
-    result_tsv = args.results_dir / (input_anndata_file.name.replace(".h5ad", "_scrublet.tsv"))
 
-
-    # Run scrublet and export the results
-    adata = anndata.read_h5ad(input_anndata_file)
+    adata = anndata.read_h5ad(input_anndata)
     scrub_results = run_scrublet(adata, args.random_seed)
     scrub_results.to_csv(result_tsv, sep="\t", index=False )
 

### 2024-07-17 analyses/metacells/script-template.py

@@ -1,81 +1,0 @@
-#!/usr/bin/env python3
-
-# metacells Analysis Script
-# Author Name
-# Date
-#
-
-# This file is a template for an analysis script written in Python.
-#
-# It is a good idea to start with a brief introduction to the contents of the
-# script, including the purpose of the analysis, the data used, and the methods
-# applied.
-#
-# Please also include examples of of how the script should be invoked, with any
-# required arguments or options.
-#
-# Replace this set of comment with your own introduction, and be sure to update
-# the Title, Author Name, and Date at the top of the document.
-# Don't forget to rename this file as well!
-#
-# Include comments in your code to explain your steps and to document any
-# assumptions you are making.
-
-# Load modules
-#
-# Load required Python modules at the top of your script
-# We have included the standard `pathlib` module and the `session_info` module
-# that we will be using at the bottom of this notebook to record the versions of
-# the modules used in this analysis.
-#
-# Do not install modules here; only load them with `import` statements.
-# Avoid renaming modules with `as` statements, unless you are performing a
-# standard renaming (e.g., `import pandas as pd`).
-
-import contextlib
-import pathlib
-
-import session_info
-
-# Define functions
-
-
-def find_git_root():
-    """Find the root directory of the git repository from the calling location of this script"""
-    repo_root = pathlib.Path.cwd()
-    while not (repo_root / ".git").is_dir():  # search for the .git directory
-        repo_root = repo_root.parent
-        if repo_root == "/":
-            raise FileNotFoundError("Could not find the repository root directory")
-    return repo_root
-
-
-# Below we set some standard paths that may be useful for your script
-# find the repository root directory
-repo_root = find_git_root()
-
-# set module path (using pathlib)
-module_root = repo_root / "analyses" / "metacells"
-
-# set current data directory
-data_dir = repo_root / "data" / "current"
-
-# set results and plots directories (using the analysis project file to find root)
-results_dir = module_root / "results"
-plots_dir = module_root / "plots"
-
-# Set other paths as needed:
-
-# Input files
-
-# Output files
-session_info_path = results_dir / "session_info.txt"
-
-# Main body of the script goes here
-
-
-# As the last step, record the versions of the modules and dependencies
-# that were used in this analysis
-with open(session_info_path, "w") as f:
-    with contextlib.redirect_stdout(f):  # direct the session_info output to a file
-        session_info.show(dependencies=True)

### 2024-07-17 create-analysis-module.py

@@ -215,15 +215,6 @@ def main() -> None:
             ["Rscript", "-e", renv_script],
             cwd=module_dir,
         )
-
-        # Set .Rprofile to not activate renv in an OpenScPCA Docker image
-        (module_dir / ".Rprofile").write_text(
-            "# Don't activate renv in an OpenScPCA docker image\n"
-            "if (Sys.getenv('OPENSCPCA_DOCKER') != 'TRUE') {\n"
-            "  source('renv/activate.R')\n"
-            "}\n"
-        )
-
         # make the components directory and add a dependencies.R file
         component_dir = module_dir / "components"
         component_dir.mkdir(exist_ok=True)

### 2024-07-17 download-data.py

@@ -151,31 +151,6 @@ def add_parent_dirs(patterns: List[str], dirs: List[str]) -> List[str]:
     return parent_patterns
 
 
-def update_symlink(data_dir: pathlib.Path, target: str, dryrun: bool = False) -> None:
-    """
-    Update the {data_dir}/current symlink to direct to target.
-    If `dryrun` is True, this function will print the expected outcome only.
-    """
-    target_path = data_dir / target
-
-    # Ensure the target path exists before updating symlink
-    if not target_path.exists():
-        print(
-            f"\nThe requested target directory '{target}' does not exist, so the symlink will not be changed.\n",
-            "Please instead download the desired release with the `--release` flag.",
-            file=sys.stderr,
-        )
-        sys.exit(1)
-    else:
-        if dryrun:
-            print(f"\nThe 'current' symlink would be updated to point to '{target}'.")
-        else:
-            current_symlink = data_dir / "current"
-            current_symlink.unlink(missing_ok=True)
-            current_symlink.symlink_to(target_path)
-            print(f"Updated 'current' symlink to point to '{target}'.")
-
-
 def download_release_data(
     bucket: str,
     release: str,
@@ -269,11 +244,12 @@ def download_release_data(
 
     ### Update current link to point to new or test data if required ###
     # only do this if we are using test data or the specified release is "current" or "latest", not for specific dates
-    if update_current:
-        if not dryrun:
-            update_symlink(data_dir, release)
-        else:
-            print(f"\nThe 'current' symlink would be updated to point to '{release}'.")
+    if update_current and not dryrun:
+        # update the current symlink
+        current_symlink = data_dir / "current"
+        current_symlink.unlink(missing_ok=True)
+        current_symlink.symlink_to(release)
+        print(f"Updated 'current' symlink to point to '{release}'.")
 
 
 def main() -> None:
@@ -335,7 +311,7 @@ def main() -> None:
         "--test-data",
         action="store_true",
         help="Download test data from the test bucket and direct the `current` symlink to the test data directory."
-        " To switch back, rerun this script with either the `--release current` option or the `--update-symlink` flag.",
+        " To switch back, rerun this script with the `--release current` option.",
     )
     parser.add_argument(
         "--metadata-only",
@@ -370,15 +346,7 @@ def main() -> None:
         default="",
         help="The AWS profile to use for the download. Uses the current default profile if undefined.",
     )
-    parser.add_argument(
-        "--update-symlink",
[…]

### 2024-07-17 analyses/cell-type-ewings/template_notebooks/cnv-workflow/04-infercnv.Rmd
fix up some of the ewings files with diffs

@@
 
-Those cells in grey (bottom left) are cells used as a reference. 
+Those cells in gray (bottom left) are cells used as a reference.
 If we were to use total CNVs we would need some sort of cut off for the number.
@@
     cnv_proportion_classification = cnv_mean_classification,
-    cnv_sum_classification = cnv_classification
+    mean_cnv_proportion = mean_proportion
   )

### 2024-07-17 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000752/SCPCL001114_tumor-gene-expression-exploration.Rmd
move consensus to be last

@@
   dplyr::mutate(
-    method = forcats::fct_recode(method, !!!method_levels)) |> 
+    method = forcats::fct_recode(method, !!!method_levels),
+    method = forcats::fct_relevel(method, "Consensus", after = Inf)) |> 
   ggplot(aes(x = UMAP1, y = UMAP2, color = classification)) +
@@
   dplyr::mutate(
-    method = forcats::fct_recode(method, !!!method_levels)
+    method = forcats::fct_recode(method, !!!method_levels),
+    method = forcats::fct_relevel(method, "Consensus", after = Inf)
   )
@@
   dplyr::mutate(
-    method = forcats::fct_recode(method, !!!method_levels)) |> 
+    method = forcats::fct_recode(method, !!!method_levels), 
+    method = forcats::fct_relevel(method, "Consensus", after = Inf)) |> 
   ggplot(aes(x = mean_score, color = classification)) +

### 2024-07-18 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd

@@
 - `consensus_call`, which will be "doublet" if _all_ methods predict "doublet," and "singlet" otherwise
-- `confusion_call`, which will classify the consensus call as one of "tp", "tn", "fp", or "fn" (true/false positive/negative) based on `consensus_call`
+- `confusion_call`, which will classify the consensus call as one of `tp`, `tn`, `fp`, or `fn` (true/false positive/negative) based on `consensus_call`
 

### 2024-07-19 analyses/cell-type-ewings/template_notebooks/singler-workflow/01-singler-results.Rmd
template for evaluating singler

---
title: "Summary of tumor cell classification with `SingleR` for `r params$library_id`"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  library_id: NULL
  sce_file: NULL
  singler_results_file: NULL
  geneset_scores_file: NULL
  marker_genes_file: NULL
---

This report summarizes the classification of `r params$library_id` using the following approach:

- Tumor cells were identified by using `AUCell`. 
These cells are labeled as `tumor_original`. 
- All other cells were labeled using `SingleR` with a combination of three references: 
  - All tumor cells annotated by `AUCell` for all other libraries in `SCPCP000015`. 
  Any cells that are labeled as tumor using this reference are labeled as `tumor_new`. 
  - `BlueprintEncodeData` from `celldex`
  - `HumanPrimaryCellAtlasData` from `celldex`

## Setup

```{r}
# check that sce and results files exist 
stopifnot(
  "sce file does not exist" = file.exists(params$sce_file),
  "SingleR results file does not exist" = file.exists(params$singler_results_file), 
  "Gene set scores file does not exist" = file.exists(params$geneset_scores_file),
  "Marker genes file does not exist" = file.exists(params$marker_genes_file)
)
```

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_bw()
)

# quiet messages
options(readr.show_col_types = FALSE)
ComplexHeatmap::ht_opt(message = FALSE)
```

```{r base paths}
# The path to this module
module_base <- rprojroot::find_root(rprojroot::is_renv_project)

# source in helper functions: plot_gene_heatmap() and plot_cnv_heatmap() 
# create_classification_df() and create_marker_gene_df()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)
```

```{r}
# Read in files
sce <- readr::read_rds(params$sce_file)
singler_results_df <- readr::read_tsv(params$singler_results_file)
geneset_scores_df <- readr::read_tsv(params$geneset_scores_file)
```

## Prep for plots

```{r}
# original library annotation 
orig_tumor <- glue::glue("tumor-{params$library_id}")

# generate classification df to use for plots 
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2) |> 
  # join with previous annotations, singler results, and gene set scores 
  dplyr::left_join(singler_results_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes") |> 
  dplyr::mutate(
    # first remove other library IDs from tumor annotations
    # we only care about whether or not it was from the original library or a new library
    singler_tumor_annotation = dplyr::case_when(
      singler_tumor_annotation == orig_tumor ~ "tumor_original",
      stringr::str_detect(singler_tumor_annotation, "tumor") ~ "tumor_new",
      .default = singler_tumor_annotation
    ),
    # get the top cell types for plotting later 
    singler_tumor_annotation_lumped = dplyr::if_else(is.na(singler_tumor_annotation), "Not labeled", singler_tumor_annotation) |>
          # there are warnings because we have some that have less than 7 cell types (ones that only did tumor/normal)
          # this is why we ignore warnings in this chunk 
          forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
          forcats::fct_infreq() |>
          forcats::fct_relevel("All remaining cell types", after = Inf), 
    # make a label that combines tumor_new with tumor_original
    singler_combined_tumor = dplyr::if_else(
      singler_tumor_annotation_lumped %in% c("tumor_original", "tumor_new"), 
      "tumor", 
      singler_tumor_annotation_lumped
      ) |> 
      forcats::fct_relevel("tumor", after = 0) |> 
      forcats::fct_relevel("All remaining cell types", after = Inf),
    # make a label that is just tumor or normal
    tumor_normal_classification = dplyr::if_else(singler_combined_tumor == "tumor", "tumor", "normal") |> 
      forcats::fct_relevel("tumor", after = 0)
  )
```

```{r}
# read in marker genes table 
marker_genes_df <- readr::read_tsv(params$marker_genes_file, show_col_types = FALSE) |> 
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
  dplyr::distinct()

# get list of all cell types found
cell_types <- unique(marker_genes_df$cell_type)

# get the sum of expression of all genes for each cell type
gene_exp_df <- cell_types |> 
  purrr::map(\(type){
    calculate_sum_markers(marker_genes_df, type)
  }) |> 
  purrr::reduce(dplyr::inner_join, by = "barcodes")

# join sum expression columns with classification df 
classification_df <- classification_df |> 
  dplyr::left_join(gene_exp_df)
```

```{r}
# get columns that correspond to marker genes and gene sets to help with plotting later 
marker_gene_columns <- colnames(classification_df)[which(endsWith(colnames(classification_df), "_sum"))]
geneset_columns <- colnames(classification_df)[which(startsWith(colnames(classification_df), "mean-"))]
```

## UMAP of all cell type annotations

Here we show the annotated cell types on a UMAP. 
The top 7 cell types are shown for each reference used and all other cells are lumped together.

```{r, fig.height=8, fig.width=8}
ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = singler_tumor_annotation_lumped)) +
  # set points for all "other" points
  geom_point(
    data = dplyr::select(
      classification_df, -singler_tumor_annotation_lumped
    ),
    color = "gray80",
    alpha = 0.5,
    size = 0.1
  ) +
  # set points for desired cell type
  geom_point(size = 0.1, alpha = 0.5) +
  facet_wrap(
    vars(singler_tumor_annotation_lumped),
    ncol = 3
  ) +
  scale_color_brewer(palette = "Dark2") +
  # remove axis numbers and background grid
  scale_x_continuous(labels = NULL, breaks = NULL) +
  scale_y_continuous(labels = NULL, breaks = NULL) +
  guides(
    color = guide_legend(
      title = "Cell type",
      # more visible points in legend
      override.aes = list(
        alpha = 1,
        size = 1.5
      )
    )) +
  theme(
        legend.position = c(0.67, 0.33),
        legend.justification = c("left", "top"),
        legend.title.align = 0.5,
        # use slightly smaller legend text, which helps legend fit and prevents
        #  long wrapped labels from bunching up
        legend.text = element_text(size = rel(0.85)),
        legend.key.height = unit(0.75, "cm"),
        aspect.ratio = 1
      )
```

## Evaluate output from `SingleR`

The next three sections will evaluate the annotations by looking at expression of marker genes and gene set scores.

The gene expression for all marker genes for tumor cells, endothelial cells, mesenchymal-like cells, and immune cells was summed together for each cell. 
Marker genes were obtained from [Visser et al.](https://doi.org/10.1158/2767-9764.CRC-23-0027).

Gene set scores are determined by calculating the mean expression of all genes in a given gene set. 
Scores are calculated for the following gene sets from `MsigDb`:

  - [`ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION`](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION.html)
  - [`RIGGI_EWING_SARCOMA_PROGENITOR_UP`](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/RIGGI_EWING_SARCOMA_PROGENITOR_UP.html?ex=1)
  - [`SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN`](https://www.gsea-msigdb.org/gsea/msigdb/cards/SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN)

For both marker gene expression and gene set scores, the following plots will be shown:

- Heatmap showing gene expression or gene set scores across all cells. 
Here the rows correspond to the gene list and the columns are the cells. 
The annotations are shown below the heatmap. 
- Density plot showing gene expression or gene set scores across all cells. 
Each row is a cell type and the expression or score is plotted on the x-axis.

### Tumor vs. Normal

In this section we show just the cells that are considered tumor (including both `tumor_original` and `tumor_new`) and normal, lumping all non-tumor cell types together.

**Marker gene expression**

```{r}
full_celltype_heatmap(classification_df, marker_gene_columns, "tumor_normal_classification")
```

```{r}
plot_density(classification_df,
             "tumor_sum",
             "tumor_normal_classification")
```

**Gene set scores**

```{r}
full_celltype_heatmap(classification_df, geneset_columns, "tumor_normal_classification")
```

```{r}
geneset_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "tumor_normal_classification")
  })
```

### Tumor vs. all other cell types

In this section we show all tumor cells together (including both `tumor_original` and `tumor_new`) and the top 5 most represented normal cell types.

**Marker gene expression**

```{r}
full_celltype_heatmap(classification_df, marker_gene_columns, "singler_combined_tumor")
```

```{r}
marker_gene_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_combined_tumor")
  })
```

**Gene set scores**

```{r}
full_celltype_heatmap(classification_df, geneset_columns, "singler_combined_tumor")
```

```{r}
geneset_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_combined_tumor")
  })
```

### Compare newly annotated tumor cells to existing tumor cells

In this section we show both tumor cell types, `tumor_original` and `tumor_new`, separately and the top 5 most represented normal cell types.

**Marker gene expression**

```{r}
full_celltype_heatmap(classification_df, marker_gene_columns, "singler_tumor_annotation_lumped")
```

```{r}
marker_gene_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_tumor_annotation_lumped")
  })
```

**Gene set scores**

```{r}
full_celltype_heatmap(classification_df, geneset_columns, "singler_tumor_annotation_lumped")
```

```{r}
geneset_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_tumor_annotation_lumped")
  })
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-07-22 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
use params to handle dataset names, and only run confusion matrices if there are two categories in the table

@@
 ---
+params: 
+  datasets: "hm-6k pbmc-1B-dm pdx-MULTI HMEC-orig-MULTI"
 title: "Comparison among doublet predictions on benchmarking datasets"
@@
 # find all dataset names to process:
-dataset_names <- list.files(result_dir, pattern = "*_scrublet.tsv") |>
-  stringr::str_remove("_scrublet.tsv")
+dataset_names <- stringr::str_split_1(params$datasets, pattern = " ")
 ```
@@
 
+        # run confusion matrix only if there are both singlets and doublets
+        if (length(unique(df$consensus_call)) == 2) {
             confusion_result <- caret::confusionMatrix(
@@
           # Extract information we want to present later in a table
+          return(
             tibble::tibble(
@@
             ) 
+          )
+        }
     }
@@
 
+        # run confusion matrix only if there are both singlets and doublets
+        if (length(unique(df$consensus_call)) == 2) {
           confusion_result <- caret::confusionMatrix(
@@
             )
+          )
+        }
     }

### 2024-07-22 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd
only calculate metrics if there are sufficient counts, and warn otherwise

@@
   # df is expected to contain `pred_col`, should should be provided as a string
-  dub_tab <- table(df[[pred_col]])
+
+  # If there are no predicted doublets, make sure the tab has a 0 count
+  dub_tab <- table( factor(df[[pred_col]], levels = c("singlet", "doublet") ))
   dub_percent <- round(dub_tab[["doublet"]]/sum(dub_tab), 4) * 100
@@
   # The `pred_col` argument _should_ be provided as a strings
+  # This function will return NULL if there are insufficient counts to calculate metrics
 
-  caret::confusionMatrix(
     # truth should be first
-    table(
+    confusion_table <- table(
       "Truth" = df$ground_truth,
       "Prediction" = df[[pred_col]]
-    ), 
+    )
+
+    if (!all(dim(confusion_table) == c(2,2))) {
+      print("Metrics could not be calculated; insufficient counts.")
+      return(NULL)
+    } else {
+      return(
+        caret::confusionMatrix(
+          confusion_table, 
           positive = "doublet"
         )
+      )
+    }
 

### 2024-07-22 analyses/cell-type-ewings/template_notebooks/singler-workflow/01-singler-results.Rmd
add in singler vs. aucell comparison

@@
 ```{r}
-# original library annotation 
-orig_tumor <- glue::glue("tumor-{params$library_id}")
-
 # generate classification df to use for plots 
@@
   scuttle::makePerCellDF(use.dimred = "UMAP") |> 
-  # replace UMAP.1 with UMAP1
-  dplyr::rename_with(
-    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
-  ) |>
-  # get rid of excess columns
-  dplyr::select(barcodes, UMAP1, UMAP2) |> 
+  # replace UMAP.1 with UMAP1 and get rid of excess columns
+  dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2) |>
   # join with previous annotations, singler results, and gene set scores 
@@
   dplyr::mutate(
+    # set factor order for aucell 
+    aucell_annotation = forcats::fct_relevel(aucell_annotation, "tumor", after= 0), 
     # first remove other library IDs from tumor annotations
     # we only care about whether or not it was from the original library or a new library
-    singler_tumor_annotation = dplyr::case_when(
-      singler_tumor_annotation == orig_tumor ~ "tumor_original",
-      stringr::str_detect(singler_tumor_annotation, "tumor") ~ "tumor_new",
-      .default = singler_tumor_annotation
-    ),
+    singler_annotation = dplyr::case_when(
+      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
+      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
+      .default = singler_annotation
+    ) |> 
+      forcats::fct_relevel("tumor", after = 0),
     # get the top cell types for plotting later 
-    singler_tumor_annotation_lumped = dplyr::if_else(is.na(singler_tumor_annotation), "Not labeled", singler_tumor_annotation) |>
+    singler_lumped = singler_annotation |>
           # there are warnings because we have some that have less than 7 cell types (ones that only did tumor/normal)
@@
           forcats::fct_relevel("All remaining cell types", after = Inf), 
-    # make a label that combines tumor_new with tumor_original
-    singler_combined_tumor = dplyr::if_else(
-      singler_tumor_annotation_lumped %in% c("tumor_original", "tumor_new"), 
-      "tumor", 
-      singler_tumor_annotation_lumped
+    # make a label that is just tumor or normal
+    singler_tumor_normal = dplyr::if_else(!(singler_annotation %in% c("tumor", "unknown")), "normal", singler_annotation) |> 
+      forcats::fct_relevel("tumor", after = 0), 
+    # get the consensus between aucell and singler 
+    consensus = dplyr::case_when(
+      singler_annotation == "tumor" & aucell_annotation == "tumor" ~ "tumor_both",
+      singler_annotation == "tumor" & aucell_annotation != "tumor" ~ "tumor_singler",
+      singler_annotation != "tumor" & aucell_annotation == "tumor" ~ "tumor_aucell",
+      .default = "normal"
     ) |> 
-      forcats::fct_relevel("tumor", after = 0) |> 
-      forcats::fct_relevel("All remaining cell types", after = Inf),
-    # make a label that is just tumor or normal
-    tumor_normal_classification = dplyr::if_else(singler_combined_tumor == "tumor", "tumor", "normal") |> 
-      forcats::fct_relevel("tumor", after = 0)
+      forcats::fct_relevel("normal", after = Inf)
   )
@@
 classification_df <- classification_df |> 
-  dplyr::left_join(gene_exp_df)
+  dplyr::left_join(gene_exp_df, by = "barcodes")
 ```
@@
 
-## UMAP of all cell type annotations
+## UMAPs of all cell type annotations
 
@@
 ```{r, fig.height=8, fig.width=8}
-ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = singler_tumor_annotation_lumped)) +
-  # set points for all "other" points
-  geom_point(
-    data = dplyr::select(
-      classification_df, -singler_tumor_annotation_lumped
-    ),
-    color = "gray80",
-    alpha = 0.5,
-    size = 0.1
-  ) +
-  # set points for desired cell type
-  geom_point(size = 0.1, alpha = 0.5) +
-  facet_wrap(
-    vars(singler_tumor_annotation_lumped),
-    ncol = 3
-  ) +
-  scale_color_brewer(palette = "Dark2") +
-  # remove axis numbers and background grid
-  scale_x_continuous(labels = NULL, breaks = NULL) +
-  scale_y_continuous(labels = NULL, breaks = NULL) +
-  guides(
-    color = guide_legend(
-      title = "Cell type",
-      # more visible points in legend
-      override.aes = list(
-        alpha = 1,
-        size = 1.5
+plot_faceted_umap(classification_df, singler_lumped)
+```
+
+## Comparison between `SingleR` and `AUCell`
+
+Below we show the cells that are annotated as tumor by both methods, just `AUCell` and just `SingleR` along with all normal cells.
+
+```{r}
+plot_faceted_umap(classification_df, consensus)
+```
+
+We then calculate the confusion matrix between tumor and normal cells called in `SingleR` and `AUCell`. 
+The rows here correspond to the `AUCell` annotation and columns are from `SingleR`.
+
+```{r}
+caret_df <- classification_df |> 
+  dplyr::filter(singler_tumor_normal != "unknown") |> 
+  dplyr::mutate(singler_tumor_normal = droplevels(singler_tumor_normal, "unknown"))
+
+caret::confusionMatrix(
+  table(
+    caret_df$aucell_annotation,
+    caret_df$singler_tumor_normal
   )
-    )) +
-  theme(
-        legend.position = c(0.67, 0.33),
-        legend.justification = c("left", "top"),
-        legend.title.align = 0.5,
-        # use slightly smaller legend text, which helps legend fit and prevents
-        #  long wrapped labels from bunching up
-        legend.text = element_text(size = rel(0.85)),
-        legend.key.height = unit(0.75, "cm"),
-        aspect.ratio = 1
 )
@@
 
-## Evaluate output from `SingleR`
+## Validate output from `SingleR`
 
-The next three sections will evaluate the annotations by looking at expression of marker genes and gene set scores.
+The next three sections will evaluate the annotations output obtained from using `SingleR` by looking at expression of marker genes and gene set scores.
 
@@
 
-In this section we show just the cells that are considered tumor (including both `tumor_original` and `tumor_new`) and normal, lumping all non-tumor cell types together.
+In this section we show just the cells that are considered tumor and normal, lumping all non-tumor cell types together.
 
@@
 ```{r}
-full_celltype_heatmap(classification_df, marker_gene_columns, "tumor_normal_classification")
+full_celltype_heatmap(classification_df, marker_gene_columns, "singler_tumor_normal")
 ```
@@
              "tumor_sum",
-             "tumor_normal_classification")
+             "singler_tumor_normal")
 ```
@@
 ```{r}
-full_celltype_heatmap(classification_df, geneset_columns, "tumor_normal_classification")
+full_celltype_heatmap(classification_df, geneset_columns, "singler_tumor_normal")
 ```
@@
                  column,
-                 "tumor_normal_classification")
+                 "singler_tumor_normal")
   })
@@
 
-In this section we show all tumor cells together (including both `tumor_original` and `tumor_new`) and the top 5 most represented normal cell types.
+In this section we show all tumor cells and the top 5 most represented normal cell types.
 
@@
 ```{r}
-full_celltype_heatmap(classification_df, marker_gene_columns, "singler_combined_tumor")
+full_celltype_heatmap(classification_df, marker_gene_columns, "singler_lumped")
 ```
@@
                  column,
-                 "singler_combined_tumor")
+                 "singler_lumped")
   })
@@
 ```{r}
-full_celltype_heatmap(classification_df, geneset_columns, "singler_combined_tumor")
+full_celltype_heatmap(classification_df, geneset_columns, "singler_lumped")
 ```
@@
 
-### Compare newly annotated tumor cells to existing tumor cells
+### Tumor cells annotatd by `SingleR` and `AUCell`
 
-In this section we show both tumor cell types, `tumor_original` and `tumor_new`, separately and the top 5 most represented normal cell types.
+Here we compare the marker gene expression and gene set scores for cells annotated as tumor by both methods, just `SingleR`, or just `AUCell` and compare to all normal cells.
 
@@
 ```{r}
-full_celltype_heatmap(classification_df, marker_gene_columns, "singler_tumor_annotation_lumped")
+full_celltype_heatmap(classification_df, marker_gene_columns, "consensus")
 ```
@@
 ```{r}
-marker_gene_columns |> 
-  purrr::map(\(column){
 plot_density(classification_df,
-                 column,
-                 "singler_tumor_annotation_lumped")
-  })
+             "tumor_sum",
+             "consensus")
 ```
@@
 ```{r}
-full_celltype_heatmap(classification_df, geneset_columns, "singler_tumor_annotation_lumped")
+full_celltype_heatmap(classification_df, geneset_columns, "consensus")
 ```
@@
                  column,
-                 "singler_tumor_annotation_lumped")
+                 "consensus")
   })

### 2024-07-22 analyses/cell-type-ewings/template_notebooks/singler-workflow/01-singler-results.Rmd
some comment updates

@@
 - Tumor cells were identified by using `AUCell`. 
-These cells are labeled as `tumor_original`. 
-- All other cells were labeled using `SingleR` with a combination of three references: 
+- All cells were labeled using `SingleR` with a combination of three references: 
   - All tumor cells annotated by `AUCell` for all other libraries in `SCPCP000015`. 
-  Any cells that are labeled as tumor using this reference are labeled as `tumor_new`. 
   - `BlueprintEncodeData` from `celldex`

### 2024-07-23 analyses/cell-type-ewings/exploratory_analysis/02-cellassign-tumor-cell-classification.Rmd

--- 2024-05-13 analyses/cell-type-ewings/02-cellassign-tumor-cell-classification.Rmd
@@
 # all markers 
-all_markers_file <- file.path(module_base, "references", "all-marker-genes.tsv")
+all_markers_file <- file.path(module_base, "references", "visser-all-marker-genes.tsv")
 # panglao markers
@@
 
-The last thing I'm trying here is to run CellAssign using some of the same reference that we used when running CellAssign as part of scpca-nf. 
+The last thing I'm trying here is to run CellAssign using some of the same reference that we used when running CellAssign as part of `scpca-nf`. 
 In particular, the majority of non-muscle cells (which we believe to be tumor cells) were identified as either Endothelial cells or Fibroblasts.

### 2024-07-23 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_singler-classification.Rmd

@@
 - The cells that aren't called as either Tumor or Normal in the Blueprint only reference mostly match up to `erythrocyte` in the reference that just uses `SCPCL000822` (original). 
-These same cells match to their corresponding Blueprint cell type when combining `SCPCL000822` and Blueprint (e.g., fibroblasts, neurons, and epithelial cells are identifed in both). 
-- When comparing using just `SCPCL000822` (original) to `SCPCL000822` and Blueprint (original and blueprint), we see an overlap in endothelial cells and chrondrocytes. 
+These same cells match to their corresponding Blueprint cell type when combining `SCPCL000822` and Blueprint (e.g., fibroblasts, neurons, and epithelial cells are identiied in both). 
+- When comparing using just `SCPCL000822` (original) to `SCPCL000822` and Blueprint (original and blueprint), we see an overlap in endothelial cells and chondrocytes. 
 We also see that for the most part the top cell types identified are the same except when using blueprint you see the addition of epithelial cells. 
@@
 
-It looks like we may be mis-classifying some cells when using the original `SingleR` assignments for SCPCL000822 with or without Blueprint. 
+It looks like we may be incorrectly classifying some cells when using the original `SingleR` assignments for `SCPCL000822` with or without Blueprint. 
 There is some overlap in the marker gene and gene set scores for the cells that are identified as tumor and other non-tumor cells for both original and original and blueprint.

### 2024-07-23 analyses/cell-type-ewings/template_notebooks/cnv-workflow/04-infercnv.Rmd

@@
     cnv_proportion_classification = cnv_mean_classification,
+    cnv_sum_classification = cnv_classification,
     mean_cnv_proportion = mean_proportion

### 2024-07-23 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd

@@
 
+```{r}
+# Check if scores are present, to determine if certain chunks should be run
+has_scdbl <- !all(is.na(doublet_df$scdbl_score))
+has_cxds <- !all(is.na(doublet_df$cxds_score))
+has_scrublet <- !all(is.na(doublet_df$scrublet_score))
+```
+
 ## Ground truth
@@
 
-```{r fig.width = 8}
+```{r, eval = has_scdbl, fig.width = 8}
 
@@
 
-```{r fig.width = 8}
+```{r, eval = has_scrublet, fig.width = 8}
 d_plot <- plot_pca(
@@
 
-```{r}
+```{r, eval = has_cxds}
 # just need a single score pca for this case

### 2024-07-23 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
spacing to trigger styling

@@
       y = PC2,
-      color = {{color_column}}
+      color = {{ color_column }}
     ) +
@@
         name = "",
-        values = c("doublet" = "black", "singlet" = "gray90")) +
+        values = c("doublet" = "black", "singlet" = "gray90")
+      ) +
       geom_point(
-        data = dplyr::filter(df, {{color_column}} == "doublet"),
+        data = dplyr::filter(df, {{ color_column }} == "doublet"),
         color = "black",
@@
         name = "Consensus",
-        values = c("ambiguous" = "yellow",
+        values = c(
+          "ambiguous" = "yellow",
           "doublet" = "black",
-                   "singlet" = "gray90")
+          "singlet" = "gray90"
+        )
       ) +
@@
   ggplot(df) +
-    aes(x = PC1,
+    aes(
+      x = PC1,
       y = PC2,
-        color = {{color_column}}) +
+      color = {{ color_column }}
+    ) +
     geom_point(
@@
     geom_point(
-    data = dplyr::filter(df, {{color_column}} %in% c("fp", "fn")),
+      data = dplyr::filter(df, {{ color_column }} %in% c("fp", "fn")),
       size = 0.75
@@
     \(dataset) {
-
       scdbl_tsv <- file.path(result_dir, glue::glue("{dataset}_scdblfinder.tsv"))
@@
           PC1 = PCA.1,
-                      PC2 = PCA.2) |>
+          PC2 = PCA.2
+        ) |>
         dplyr::left_join(
@@
     \(df, dataset) {
-
       doublet_barcodes <- list(
@@
         y = 0.95,
-        gp = grid::gpar(fontsize=16)
+        gp = grid::gpar(fontsize = 16)
       )
-
     }
   )
-
 ```
@@
     \(dataset_df) {
-
       dataset_df |>
         dplyr::rowwise() |>
-
         dplyr::mutate(
@@
 ```{r, fig.width = 12, fig.height = 9}
-pc_color_columns <- c("scDblFinder prediction" = "scdbl_prediction",
+pc_color_columns <- c(
+  "scDblFinder prediction" = "scdbl_prediction",
   "scrublet prediction" = "scrublet_prediction",
@@
   "Ground truth" = "ground_truth",
-                      "Consensus call" = "consensus_call")
+  "Consensus call" = "consensus_call"
+)
 doublet_df_list |>
@@
     \(df, dataset) {
-
       pc_plot_list <- pc_color_columns |>
@@
               title = plot_title
-            ) + theme(legend.position = "none")}
+            ) + theme(legend.position = "none")
+          }
         )
@@
           glue::glue("PCA for {dataset}"),
-            theme = theme(plot.title = element_text(size = 16))) +
+          theme = theme(plot.title = element_text(size = 16))
+        ) +
         plot_layout(guides = "collect")
@@
       # run confusion matrix only if there are both singlets and doublets
-        if (length(unique(df$consensus_call)) == 2) {
+      if ("singlet" %in% df$consensus_call & "doublet" %in% df$consensus_call) {
         confusion_result <- caret::confusionMatrix(
@@
     \(dataset_df) {
-
       dataset_df <- dataset_df |>
@@
 ```{r, fig.width = 12, fig.height = 6}
-pc_color_columns <- c("scDblFinder prediction" = "scdbl_prediction",
+pc_color_columns <- c(
+  "scDblFinder prediction" = "scdbl_prediction",
   "cxds prediction" = "cxds_prediction",
   "Ground truth" = "ground_truth",
-                      "Consensus call" = "consensus_call")
+  "Consensus call" = "consensus_call"
+)
 

### 2024-07-23 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd
spacing to trigger styling

@@
 )
-
 ```
@@
   # If there are no predicted doublets, make sure the tab has a 0 count
-  dub_tab <- table( factor(df[[pred_col]], levels = c("singlet", "doublet") ))
-  dub_percent <- round(dub_tab[["doublet"]]/sum(dub_tab), 4) * 100
+  dub_tab <- table(factor(df[[pred_col]], levels = c("singlet", "doublet")))
+  dub_percent <- round(dub_tab[["doublet"]] / sum(dub_tab), 4) * 100
 
@@
   p <- ggplot(df) +
-    aes(x = PC1,
+    aes(
+      x = PC1,
       y = PC2,
-        color = {{color_column}}) +
+      color = {{ color_column }}
+    ) +
     geom_point(alpha = 0.5) +
@@
       geom_point(
-      data = dplyr::filter(df, {{color_column}} == "doublet"),
+        data = dplyr::filter(df, {{ color_column }} == "doublet"),
         color = "black"
@@
       geom_point(
-      data = dplyr::filter(df, {{pred_column}} == "doublet"),
-      aes(color = {{color_column}})
+        data = dplyr::filter(df, {{ pred_column }} == "doublet"),
+        aes(color = {{ color_column }})
       )
@@
 
-plot_jitter_scores <- function(df, score_col, pred_col, method_name){
+plot_jitter_scores <- function(df, score_col, pred_col, method_name) {
   # Plot jitter plot of score distributions and predictions, colored by ground truth
@@
   ggplot(df) +
-    aes(x = ground_truth,
-        y = {{score_col}},
-        color = {{pred_col}}) +
+    aes(
+      x = ground_truth,
+      y = {{ score_col }},
+      color = {{ pred_col }}
+    ) +
     geom_jitter(
@@
 
-plot_density_scores <- function(df, score_col, method_name){
+plot_density_scores <- function(df, score_col, method_name) {
   # Plot density plot of score distributions, colored by ground truth
@@
   ggplot(df) +
-    aes(x = {{score_col}},
-        fill = ground_truth) +
+    aes(
+      x = {{ score_col }},
+      fill = ground_truth
+    ) +
     geom_density(alpha = 0.7) +
@@
 
-    if (!all(dim(confusion_table) == c(2,2))) {
+  if (!all(dim(confusion_table) == c(2, 2))) {
     print("Metrics could not be calculated; insufficient counts.")
@@
   }
-
 }
-
 ```
@@
     PC1 = PCA.1,
-                PC2 = PCA.2)
+    PC2 = PCA.2
+  )
 
@@
 ```{r, eval = has_scdbl, fig.width = 8}
-
 d_plot <- plot_pca(
@@
 ```{r fig.width = 10}
-
 cxds_cols <- names(doublet_df)[grep("^cxds_prediction_", names(doublet_df))]

### 2024-07-23 analyses/cell-type-ewings/template_notebooks/singler-workflow/01-singler-results.Rmd
use patchwork

@@
 
-```{r}
+```{r, fig.height=10}
 geneset_columns |> 
@@
                  "singler_tumor_normal")
-  })
+  }) |> 
+  patchwork::wrap_plots(ncol = 1)
 ```
@@
 
-```{r}
+```{r, fig.height=10}
 marker_gene_columns |> 
@@
                  "singler_lumped")
-  })
+  }) |> 
+  patchwork::wrap_plots(ncol = 1)
 ```
@@
                  "consensus")
-  })
+  }) |> 
+  patchwork::wrap_plots(ncol = 1)
 ```

### 2024-07-23 analyses/cell-type-ewings/scripts/utils/tumor-validation-helpers.R

@@ -291,7 +291,7 @@ plot_density <- function(classification_df,
   # pull out gene set name to create the plot title 
   geneset_name <- stringr::str_remove(gene_exp_column, "_sum|mean-")
   
-  ggplot(classification_df) +
+  plot <- ggplot(classification_df) +
     aes(x = !!sym(gene_exp_column)) + # marker gene set column 
     geom_density(
       fill = "grey65",
@@ -320,6 +320,8 @@ plot_density <- function(classification_df,
       panel.grid.major.y = element_blank(),
       panel.spacing = unit(0.02, "in")
     )
+  
+  return(plot)
 }
 
 

### 2024-07-23 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd

--- 2024-07-22 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
@@
         # run confusion matrix only if there are both singlets and doublets
-        if (length(unique(df$consensus_call)) == 2) {
+        if ("singlet" %in% df$consensus_call & "doublet" %in% df$consensus_call) {
             confusion_result <- caret::confusionMatrix(
@@
         # run confusion matrix only if there are both singlets and doublets
-        if (length(unique(df$consensus_call)) == 2) {
+        if ("singlet" %in% df$consensus_call & "doublet" %in% df$consensus_call) {
           confusion_result <- caret::confusionMatrix(

### 2024-07-23 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_singler-classification.Rmd
Styled code

@@
   column_title = "Blueprint",
-             legend_match = "Original") |> 
+  legend_match = "Original"
+) |>
   ComplexHeatmap::draw(
@@
   column_title = "Original",
-             legend_match = "Original")
+  legend_match = "Original"
+)
 
@@
   column_title = "Tumor only",
-             legend_match = "Tumor only")
-
+  legend_match = "Tumor only"
+)
 ```
@@
   purrr::map(\(classification_column){
-
     # first get list of tumor cells for this classificationn
@@
     tumor_cells <- classification_df |>
-      dplyr::filter({{classification_column}} == "Tumor") |> 
+      dplyr::filter({{ classification_column }} == "Tumor") |>
       dplyr::pull(barcodes)
@@
       # anything not in shared cell types are classified as "other"
-      dplyr::mutate(singler_caret = dplyr::if_else(singler_celltype_annotation %in% shared_celltypes, singler_celltype_annotation, "Other"),
-                    new_ref_caret = dplyr::if_else(!!sym(classification_column) %in% shared_celltypes, !!sym(classification_column), "Other"))
+      dplyr::mutate(
+        singler_caret = dplyr::if_else(singler_celltype_annotation %in% shared_celltypes, singler_celltype_annotation, "Other"),
+        new_ref_caret = dplyr::if_else(!!sym(classification_column) %in% shared_celltypes, !!sym(classification_column), "Other")
+      )
 
@@
         caret_df$singler_caret,
-        caret_df$new_ref_caret)
+        caret_df$new_ref_caret
       )
-
+    )
   })
@@
 # make a table showing which cells are assigned to which reference
-celltype_table <- table(classification_df$updated_singler_blueprint_classification, 
-      classification_df$orig_ref)
+celltype_table <- table(
+  classification_df$updated_singler_blueprint_classification,
+  classification_df$orig_ref
+)
 # sort by SCPCL000822
-celltype_table[order(celltype_table[,2], decreasing = TRUE), ]
+celltype_table[order(celltype_table[, 2], decreasing = TRUE), ]
 ```
@@
   purrr::iwalk(\(annotation_column, name) {
-
     # first make singler annotations normal or tumor for easier visualization
@@
   facet_wrap(vars(method),
-             scales = "free_y") +
+    scales = "free_y"
+  ) +
   labs(
@@
   geom_density(bw = 0.05) +
-  facet_grid(rows = vars(method), 
+  facet_grid(
+    rows = vars(method),
     cols = vars(geneset),
-             scales = "free_y") +
+    scales = "free_y"
+  ) +
   labs(
@@
   )
-
 ```

### 2024-07-23 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000752/SCPCL001114_tumor-gene-expression-exploration.Rmd
Styled code

@@
 
-ref_markers_df <- create_marker_gene_df(sce = ref_sce, 
+ref_markers_df <- create_marker_gene_df(
+  sce = ref_sce,
   classification_df = ref_classification_df,
-                                        marker_genes_file) |> 
+  marker_genes_file
+) |>
   dplyr::mutate(sample = "SCPCL000822")
@@
 # find the local minima in the distribution
-exp_cutoff <- optimize(approxfun(density_data$x, density_data$y), interval = c(1,10))$minimum
+exp_cutoff <- optimize(approxfun(density_data$x, density_data$y), interval = c(1, 10))$minimum
 ```
@@
     classification_df$ref_marker_gene_classification,
-      classification_df$auc_classification)
+    classification_df$auc_classification
+  )
 )
@@
     classification_df$ref_marker_gene_classification,
-      classification_df$one_marker_gene_classification)
+    classification_df$one_marker_gene_classification
+  )
 )
@@
       "Normal"
-    ))
+    )
+  )
 ```
@@
     classification_df$ref_marker_gene_classification,
-      classification_df$two_marker_gene_classification)
+    classification_df$two_marker_gene_classification
+  )
 )
@@
     names_to = "gene_list",
-                      values_to = "signature_score") |> 
+    values_to = "signature_score"
+  ) |>
   dplyr::mutate(
@@
   geom_density() +
-  facet_grid(rows = vars(gene_list),
-             scales = "free_y")
+  facet_grid(
+    rows = vars(gene_list),
+    scales = "free_y"
+  )
 ```
@@
     method = forcats::fct_recode(method, !!!method_levels),
-    method = forcats::fct_relevel(method, "Consensus", after = Inf)) |> 
+    method = forcats::fct_relevel(method, "Consensus", after = Inf)
+  ) |>
   ggplot(aes(x = UMAP1, y = UMAP2, color = classification)) +
@@
     method = forcats::fct_recode(method, !!!method_levels),
-    method = forcats::fct_relevel(method, "Consensus", after = Inf)) |> 
+    method = forcats::fct_relevel(method, "Consensus", after = Inf)
+  ) |>
   ggplot(aes(x = mean_score, color = classification)) +
   geom_density(bw = 0.05) +
-  facet_grid(rows = vars(method), 
-             cols = vars(geneset))
+  facet_grid(
+    rows = vars(method),
+    cols = vars(geneset)
+  )
 ```
@@
   consensus = annotation_df$consensus_classification,
-  col = list(one_marker = plot_colors,
+  col = list(
+    one_marker = plot_colors,
     two_marker = plot_colors,
@@
     AUCell = plot_colors,
-             consensus = plot_colors))
+    consensus = plot_colors
+  )
+)
 ```
@@
   dplyr::select(gene_expression, gene_symbol, barcodes) |>
-  tidyr::pivot_wider(values_from = gene_expression, 
-                     names_from = barcodes) |> 
+  tidyr::pivot_wider(
+    values_from = gene_expression,
+    names_from = barcodes
+  ) |>
   tibble::column_to_rownames("gene_symbol") |>
@@
   legend_title = "Marker gene \nexpression",
-                  annotation = annotation)
+  annotation = annotation
+)
 ```
@@
   unique() |>
-  tidyr::pivot_wider(values_from = mean_score, 
-                     names_from = barcodes) |> 
+  tidyr::pivot_wider(
+    values_from = mean_score,
+    names_from = barcodes
+  ) |>
   tibble::column_to_rownames("geneset") |>
@@
   annotation = annotation,
-                  legend_title = "Gene set \nscore")
+  legend_title = "Gene set \nscore"
+)
 ```

### 2024-07-23 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_singler-classification.Rmd
Fix spelling errors

--- 2024-07-23 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_singler-classification.Rmd
@@
 - The cells that aren't called as either Tumor or Normal in the Blueprint only reference mostly match up to `erythrocyte` in the reference that just uses `SCPCL000822` (original). 
-These same cells match to their corresponding Blueprint cell type when combining `SCPCL000822` and Blueprint (e.g., fibroblasts, neurons, and epithelial cells are identiied in both). 
+These same cells match to their corresponding Blueprint cell type when combining `SCPCL000822` and Blueprint (e.g., fibroblasts, neurons, and epithelial cells are identified in both). 
 - When comparing using just `SCPCL000822` (original) to `SCPCL000822` and Blueprint (original and blueprint), we see an overlap in endothelial cells and chondrocytes. 

### 2024-07-23 analyses/cell-type-ewings/exploratory_analysis/01-marker-gene-tumor-classification.Rmd
fix up some of the ewings files with diffs
Clean up notebooks in a few more places, directly grabbing from main to avoid whitespace diffs

@@
 # do all marker genes so differences between tumor/normal or only a subset?
-ggplot(celltype_df, aes(x=transformed_gene_expression, color = sum_classification)) +
+ggplot(celltype_df, aes(x = transformed_gene_expression, color = sum_classification)) +
   geom_density() +
@@
   groups = sce$sum_classification,
-                              pval.type = "all") 
+  pval.type = "all"
+)
 ```
@@
   theme_classic()
-
 ```
@@
       )
-    }) |>
+    }
+  ) |>
   # concatenate vertically into HeatmapList object
@@
 .
-- The top marker genes for the "tumor" cells contain PRKCB and MAPT, two of the marker genes in our gene list. 
+- The top marker genes for the "tumor" cells contain `PRKCB` and `MAPT`, two of the marker genes in our gene list. 
 - It's likely that `ADGRG2`, `HES1`, `NROB1`, and `STEAP1` will not be informative in assigning tumor cells. 

### 2024-07-23 analyses/cell-type-ewings/template_notebooks/cnv-workflow/02-cellassign.Rmd
fix up some of the ewings files with diffs

@@
 # function to get assigned cell type based on max prediction returned by CellAssign
-get_celltype_assignment <- function(predictions){
-
+get_celltype_assignment <- function(predictions) {
   # if no predictions, assign every cell to NA
-  if(length(colnames(predictions)) == 1){
+  if (length(colnames(predictions)) == 1) {
     celltype_assignments <- data.frame(
@@
 # check if marker gene annotations are present
-if(all(is.na(classifications_df$marker_gene_classification))){
+if (all(is.na(classifications_df$marker_gene_classification))) {
   has_marker_gene <- FALSE
@@
 
-if(all(is.na(tumor_celltype_assignments$celltype))){
+if (all(is.na(tumor_celltype_assignments$celltype))) {
   has_tumor_assignments <- FALSE
@@
 # define annotations to use in all heatmaps
-if(has_marker_gene){
+if (has_marker_gene) {
   # get annotations for heatmaps
@@
   as.matrix() |>
-  pheatmap::pheatmap(show_rownames = FALSE,
-                     annotation_row = annotation_df)
+  pheatmap::pheatmap(
+    show_rownames = FALSE,
+    annotation_row = annotation_df
+  )
 ```
@@
 # make sure that we only run if both classes are found in both cell types
-if(all(c("other", "tumor") %in% celltype_df$cellassign_annotations) &&
-   all(c("Tumor", "Normal") %in% celltype_df$marker_gene_classification)){
+if (all(c("other", "tumor") %in% celltype_df$cellassign_annotations) &&
+  all(c("Tumor", "Normal") %in% celltype_df$marker_gene_classification)) {
   # prep annotations for confusion matrix
@@
       celltype_df$marker_gene_classification,
-      celltype_df$cellassign_annotations)
+      celltype_df$cellassign_annotations
+    )
   )
@@
 
-if(all(is.na(visser_celltype_assignments$celltype))){
+if (all(is.na(visser_celltype_assignments$celltype))) {
   has_visser_assignments <- FALSE
@@
 # if present, compare across both references
-if(has_endothelial){
+if (has_endothelial) {
   # prep annotations for confusion matrix
@@
 
-  if(all("Endothelial cells", "Other") %in% celltype_df$visser_cellassign_annotations &&
-     all("Endothelial cells", "Other") %in% celltype_df$cellassign_celltype_annotation){
+  if (all("Endothelial cells", "Other") %in% celltype_df$visser_cellassign_annotations &&
+    all("Endothelial cells", "Other") %in% celltype_df$cellassign_celltype_annotation) {
     caret::confusionMatrix(
@@
         celltype_df$caret_cellassign_annotations,
-        celltype_df$caret_visser_annotations)
+        celltype_df$caret_visser_annotations
+      )
     )
@@
       theme_bw()
-
   })
@@
 
-if(all(is.na(panglao_celltype_assignments$celltype))){
+if (all(is.na(panglao_celltype_assignments$celltype))) {
   has_panglao_assignments <- FALSE
@@
   theme_bw()
-
 ```
@@
     visser_cellassign_annotations,
-                panglao_cellassign_annotations)
+    panglao_cellassign_annotations
+  )
 

### 2024-07-23 analyses/cell-type-ewings/exploratory_analysis/03-copykat.Rmd

@@
 full_cnv_df <- full_ck_results_df |>
-  tidyr::pivot_longer(cols = -c(
+  tidyr::pivot_longer(
+    cols = -c(
       reference_used,
@@
     names_to = "barcodes",
-    values_to = "cnv_detection") |> 
+    values_to = "cnv_detection"
+  ) |>
   dplyr::group_by(chrom, barcodes, reference_used) |>
@@
   theme_bw() +
-      facet_grid(rows = vars(chrom), 
-                 cols = vars(reference_used))
+  facet_grid(
+    rows = vars(chrom),
+    cols = vars(reference_used)
+  )
 ```
@@
         df$marker_gene_classification,
-        df$copykat)
+        df$copykat
+      )
     )
@@
   purrr::map(\(df) {
-
     make_jaccard_matrix(

### 2024-07-23 analyses/cell-type-ewings/template_notebooks/cnv-workflow/01-marker-genes.Rmd

@@
 # read in normal and tumor cells for later comparisons
-if(file.exists(params$reference_cell_file)){
+if (file.exists(params$reference_cell_file)) {
   ref_cells_df <- readr::read_tsv(params$reference_cell_file)
@@
 # test if _any_ marker genes have expression
-if(sum(umap_df$gene_expression) == 0){
+if (sum(umap_df$gene_expression) == 0) {
   has_marker_gene_exp <- FALSE
@@
   groups = sce$sum_classification,
-                              pval.type = "all")
+  pval.type = "all"
+)
 ```
@@
 # use already identified normal/tumor cell types
-if(!is.null(ref_cells_df)){
-
+if (!is.null(ref_cells_df)) {
   # get list of normal cells
@@
   # only create this plot if normal and tumor cells are provided
-  if(length(normal_cells) > 0 && length(tumor_cells) > 0){
-
+  if (length(normal_cells) > 0 && length(tumor_cells) > 0) {
     # indicate cells that are part of normal and tumor lists
@@
 ```{r}
-if(has_marker_gene_exp){
+if (has_marker_gene_exp) {
   # get a dataframe with barcodes, sum_classification and cluster_classification
@@
   readr::write_tsv(celltypes_df, output_classifications_file)
-} else{
+} else {
   # save marker gene classification as NA if no expression

### 2024-07-23 analyses/cell-type-ewings/template_notebooks/singler-workflow/01-singler-results.Rmd
Fix spelling errors

--- 2024-07-22 analyses/cell-type-ewings/template_notebooks/singler-workflow/01-singler-results.Rmd
@@
     # set factor order for aucell
-    aucell_annotation = forcats::fct_relevel(aucell_annotation, "tumor", after= 0), 
+    aucell_annotation = forcats::fct_relevel(aucell_annotation, "tumor", after = 0),
     # first remove other library IDs from tumor annotations
@@
 ```{r}
-plot_density(classification_df,
+plot_density(
+  classification_df,
   "tumor_sum",
-             "singler_tumor_normal")
+  "singler_tumor_normal"
+)
 ```
@@
 
-```{r}
+```{r, fig.height=10}
 geneset_columns |>
   purrr::map(\(column){
-    plot_density(classification_df,
+    plot_density(
+      classification_df,
       column,
-                 "singler_tumor_normal")
-  })
+      "singler_tumor_normal"
+    )
+  }) |>
+  patchwork::wrap_plots(ncol = 1)
 ```
@@
 
-```{r}
+```{r, fig.height=10}
 marker_gene_columns |>
   purrr::map(\(column){
-    plot_density(classification_df,
+    plot_density(
+      classification_df,
       column,
-                 "singler_lumped")
-  })
+      "singler_lumped"
+    )
+  }) |>
+  patchwork::wrap_plots(ncol = 1)
 ```
@@
 
-```{r}
+```{r, fig.height=10}
 geneset_columns |>
   purrr::map(\(column){
-    plot_density(classification_df,
+    plot_density(
+      classification_df,
       column,
-                 "singler_lumped")
-  })
+      "singler_lumped"
+    )
+  }) |>
+  patchwork::wrap_plots(ncol = 1)
 ```
 
-### Tumor cells annotatd by `SingleR` and `AUCell`
+### Tumor cells annotated by `SingleR` and `AUCell`
 
@@
 ```{r}
-plot_density(classification_df,
+plot_density(
+  classification_df,
   "tumor_sum",
-             "consensus")
+  "consensus"
+)
 ```
@@
 
-```{r}
+```{r, fig.height=10}
 geneset_columns |>
   purrr::map(\(column){
-    plot_density(classification_df,
+    plot_density(
+      classification_df,
       column,
-                 "consensus")
-  })
+      "consensus"
+    )
+  }) |>
+  patchwork::wrap_plots(ncol = 1)
 ```

### 2024-07-23 analyses/cell-type-ewings/scripts/auc-singler-workflow/04-run-singler.R
move scripts and notebooks

@@ -27,7 +27,7 @@ option_list <- list(
       The `ref_tumor_label` column in the `colData` will be used for the reference labels."
   ),
   make_option(
-    opt_str = c("--output_annotations_file"),
+    opt_str = c("--output_file"),
     type = "character",
     help = "Path to TSV file to save annotations as obtained from `SingleR` for input SCE file."
   ),
@@ -36,6 +36,12 @@ option_list <- list(
     type = "integer",
     default = 4,
     help = "Number of multiprocessing threads to use."
+  ),
+  make_option(
+    opt_str = c("--seed"),
+    type = "integer",
+    default = 2024,
+    help = "A random seed for reproducibility."
   )
 )
 
@@ -44,6 +50,9 @@ opt <- parse_args(OptionParser(option_list = option_list))
 
 # Set up -----------------------------------------------------------------------
 
+# set seed
+set.seed(opt$seed)
+
 # make sure input files exist
 stopifnot(
   "SCE file does not exist" = file.exists(opt$sce_file),
@@ -51,7 +60,7 @@ stopifnot(
 )
 
 # create output directory if it doesn't exist
-output_dir <- dirname(opt$output_annotations_file)
+output_dir <- dirname(opt$output_file)
 fs::dir_create(output_dir)
 
 # set up multiprocessing params
@@ -128,4 +137,4 @@ results_df <- data.frame(
   )
 
 # export tsv file with barcodes, aucell_annotation, singler_annotation and singler_ontology columns
-readr::write_tsv(results_df, opt$output_annotations_file)
+readr::write_tsv(results_df, opt$output_file)

### 2024-07-24 analyses/cell-type-ewings/template_notebooks/auc-singler-workflow/02-singler-results.Rmd
move scripts and notebooks
add missing variable

--- 2024-07-23 analyses/cell-type-ewings/template_notebooks/singler-workflow/01-singler-results.Rmd
@@
   purrr::map(\(type){
-    calculate_sum_markers(marker_genes_df, type)
+    calculate_sum_markers(marker_genes_df, sce, type)
   }) |>
@@
 
-### Tumor cells annotated by `SingleR` and `AUCell`
+### Tumor cells annotatd by `SingleR` and `AUCell`
 

### 2024-07-24 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd
fix up some of the ewings files with diffs

@@
   purrr::walk(\(file)
-    stopifnot("Missing PNG file" = file.exists(file))
-  )
+  stopifnot("Missing PNG file" = file.exists(file)))
 
@@
       rowSums(across(has_cnv_cols))
-    ))
+    )
+  )
 ```
@@
 # plot the total number of CNVs
-ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = has_cnv)) +
+ggplot(cnv_df, aes(x = UMAP1, y = UMAP2, color = has_cnv)) +
   geom_point(alpha = 0.5, size = 0.5) +
@@
 # get mean value for has_cnv, excluding any NAs
-mean_cnv <- mean(cnv_df$has_cnv, na.rm=TRUE)
+mean_cnv <- mean(cnv_df$has_cnv, na.rm = TRUE)
 
@@
 cnv_df <- cnv_df |>
-  dplyr::mutate(cnv_classification = dplyr::case_when(is.na(has_cnv) ~ "Reference",
+  dplyr::mutate(cnv_classification = dplyr::case_when(
+    is.na(has_cnv) ~ "Reference",
     has_cnv > mean_cnv ~ "Tumor",
-                                                      has_cnv <= mean_cnv ~ "Normal"))
+    has_cnv <= mean_cnv ~ "Normal"
+  ))
 
-ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = cnv_classification)) +
+ggplot(cnv_df, aes(x = UMAP1, y = UMAP2, color = cnv_classification)) +
   geom_point(alpha = 0.5, size = 0.5) +
@@
       weighted.mean(across(proportion_cols), chr_weights),
-      NA)) |> 
+      NA
+    )
+  ) |>
   dplyr::ungroup()
@@
 ```{r}
-ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = mean_proportion)) +
+ggplot(cnv_df, aes(x = UMAP1, y = UMAP2, color = mean_proportion)) +
   geom_point(alpha = 0.5, size = 0.5) +
@@
 cnv_df <- cnv_df |>
-  dplyr::mutate(cnv_mean_classification = dplyr::case_when(is.na(mean_proportion) ~ "Reference",
+  dplyr::mutate(cnv_mean_classification = dplyr::case_when(
+    is.na(mean_proportion) ~ "Reference",
     mean_proportion > all_mean_proportion ~ "Tumor",
-                                                           mean_proportion <= all_mean_proportion ~ "Normal"))
+    mean_proportion <= all_mean_proportion ~ "Normal"
+  ))
 
-ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = cnv_mean_classification)) +
+ggplot(cnv_df, aes(x = UMAP1, y = UMAP2, color = cnv_mean_classification)) +
   geom_point(alpha = 0.5, size = 0.5) +
@@
   dplyr::select(barcodes, UMAP1, UMAP2, cnv_mean_classification, all_of(proportion_cols)) |>
-  tidyr::pivot_longer(cols = all_of(proportion_cols),
+  tidyr::pivot_longer(
+    cols = all_of(proportion_cols),
     names_to = "chromosome",
-    values_to = "proportion") |> 
+    values_to = "proportion"
+  ) |>
   # remove extra string in front of chr
@@
 ggplot(proportion_df, aes(x = proportion, color = cnv_mean_classification)) +
-  geom_density(bounds = c(0,1), bw = 0.05) +
+  geom_density(bounds = c(0, 1), bw = 0.05) +
   theme_bw() +
@@
 ewings_chr_df <- proportion_df |>
-  dplyr::filter(chromosome %in% c(1,8,12,16))
+  dplyr::filter(chromosome %in% c(1, 8, 12, 16))
 
-ggplot(ewings_chr_df, aes(x = UMAP1, y=UMAP2 , color = proportion)) +
+ggplot(ewings_chr_df, aes(x = UMAP1, y = UMAP2, color = proportion)) +
   geom_point(alpha = 0.5, size = 0.5) +
@@
   scale_color_viridis_c()
-
 ```

### 2024-07-24 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd

@@
     # set factor order to use for plotting
-    combined_cnv_classification = forcats::fct_relevel(combined_cnv_classification, "Tumor", "Normal", "Ambiguous"))
+    combined_cnv_classification = forcats::fct_relevel(combined_cnv_classification, "Tumor", "Normal", "Ambiguous")
+  )
 ```
@@
   classification = annotation_df$combined_cnv_classification,
-  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
+  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"))
+)
 ```
@@
   geom_density(bw = 0.05) +
-  facet_grid(rows = vars(method), 
-             cols = vars(geneset))
+  facet_grid(
+    rows = vars(method),
+    cols = vars(geneset)
+  )
 ```
@@
 # need to make a new annotation, since we want to annotate the rows here
-annotation <- ComplexHeatmap::rowAnnotation(classification = annotation_df$combined_cnv_classification,
-                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
+annotation <- ComplexHeatmap::rowAnnotation(
+  classification = annotation_df$combined_cnv_classification,
+  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"))
+)
 
@@
   annotation = annotation,
-                 legend_title = "Copy number \ngain")
+  legend_title = "Copy number \ngain"
+)
 ```
@@
   annotation = annotation,
-                 legend_title = "Copy number \nloss")
+  legend_title = "Copy number \nloss"
+)
 ```

### 2024-07-24 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd

@@
   # remove extra infercnv classification column we won't use
-  dplyr::select(- cnv_sum_classification)
+  dplyr::select(-cnv_sum_classification)
 ```
@@
       method == "cnv_proportion_classification" ~ "InferCNV proportion"
-    )) |> 
+    )
+  ) |>
   ggplot(aes(x = mean_score, color = classification)) +
   geom_density(bw = 0.05) +
-  facet_grid(rows = vars(method), 
-             cols = vars(geneset))
+  facet_grid(
+    rows = vars(method),
+    cols = vars(geneset)
+  )
 ```
@@
 # all gene sets are part of C2, CGP
-genes_df <- msigdbr::msigdbr(species= "Homo sapiens", 
+genes_df <- msigdbr::msigdbr(
+  species = "Homo sapiens",
   category = "C2",
-                             subcategory = "CGP") |> 
+  subcategory = "CGP"
+) |>
   # only keep relevant gene sets
@@
     caret_df$cnv_classification,
-      caret_df$auc_classification)
+    caret_df$auc_classification
+  )
 )
@@
     names_to = "barcodes",
-                      values_to = "auc") |> 
-  #dplyr::filter(gene_set == "markers") |> 
+    values_to = "auc"
+  ) |>
+  # dplyr::filter(gene_set == "markers") |>
   dplyr::mutate(sample = "SCPCL000822")
@@
     names_to = "barcodes",
-                      values_to = "auc") |> 
-  #dplyr::filter(gene_set == "markers") |> 
+    values_to = "auc"
+  ) |>
+  # dplyr::filter(gene_set == "markers") |>
   dplyr::mutate(sample = "SCPCL000824")
@@
   geom_density() +
-  facet_grid(rows = vars(gene_set),
-             scales = "free_y")
+  facet_grid(
+    rows = vars(gene_set),
+    scales = "free_y"
+  )
 ```
@@
     classification_df$updated_marker_gene_classification,
-      classification_df$auc_classification)
+    classification_df$auc_classification
+  )
 )
@@
 ref_ucell_tumor_cells <- ucell_df |>
-  dplyr::filter(gene_list == "markers",
+  dplyr::filter(
+    gene_list == "markers",
     signature_score >= geneset_cutoff,
-                sample == "SCPCL000822") |> 
+    sample == "SCPCL000822"
+  ) |>
   dplyr::pull(barcodes) |>
@@
 new_ucell_tumor_cells <- ucell_df |>
-  dplyr::filter(gene_list == "markers",
+  dplyr::filter(
+    gene_list == "markers",
     signature_score >= geneset_cutoff,
-                sample == "SCPCL000824") |> 
+    sample == "SCPCL000824"
+  ) |>
   dplyr::pull(barcodes) |>
@@
   consensus = annotation_df$consensus_classification,
-  col = list(marker_genes = plot_colors,
-             updated_marker_genes =plot_colors, 
+  col = list(
+    marker_genes = plot_colors,
+    updated_marker_genes = plot_colors,
     AUCell = plot_colors,
     UCell = plot_colors,
-             consensus = plot_colors))
+    consensus = plot_colors
+  )
+)
 ```

### 2024-07-24 analyses/cell-type-ewings/template_notebooks/cnv-workflow/03-copykat.Rmd
fix up some of the ewings files with diffs

--- 2024-07-10 analyses/cell-type-ewings/template_notebooks/cnv-workflow/03-copykat.Rmd
@@
 # check if marker gene annotations are present
-if(all(is.na(manual_classifications_df$marker_gene_classification))){
+if (all(is.na(manual_classifications_df$marker_gene_classification))) {
   has_marker_gene <- FALSE
@@
       "reference",
-      copykat.pred)
+      copykat.pred
+    )
   )
@@
 filtered_cnv_df <- cnv_df |>
-  dplyr::filter(!(copykat.pred %in% c("not.defined", "reference")),
+  dplyr::filter(
+    !(copykat.pred %in% c("not.defined", "reference")),
     # filter any low confidence calls
-                !stringr::str_detect(copykat.pred, "low.conf"))
+    !stringr::str_detect(copykat.pred, "low.conf")
+  )
 
@@
 celltype_df <- cnv_df |>
-  dplyr::select(barcodes, 
+  dplyr::select(
+    barcodes,
     reference_used,
-                copykat.pred) |> 
+    copykat.pred
+  ) |>
   dplyr::distinct() |>
@@
 all_info_df <- cnv_df |>
-  dplyr::select(barcodes,
-                mean_cnv_detection) |> 
+  dplyr::select(
+    barcodes,
+    mean_cnv_detection
+  ) |>
   dplyr::group_by(barcodes) |>

### 2024-07-24 analyses/cell-type-ewings/template_notebooks/auc-singler-workflow/02-singler-results.Rmd
account for only normal or only tumor

@@
 
+# get the cell types in both
+# these should be tumor normal but sometimes that's not the case
+singler_cell_types <- unique(caret_df$singler_tumor_normal) |> as.character()
+aucell_cell_types <- unique(caret_df$aucell_annotation) |> as.character()
+
+# only calculate confusion matrix if all cell types are the same
+if (all(singler_cell_types == aucell_cell_types)) {
   caret::confusionMatrix(
@@
   )
+}
 ```

### 2024-07-26 analyses/cell-type-ewings/template_notebooks/auc-singler-workflow/01-auc-results.Rmd

--- 2024-07-15 analyses/cell-type-ewings/template_notebooks/auc-workflow/01-auc-results.Rmd
@@
 # test if _any_ marker genes have expression
-if(sum(markers_df$gene_expression) == 0){
+if (sum(markers_df$gene_expression) == 0) {
   has_marker_gene_exp <- FALSE
@@
       "Normal"
-
     )
@@
     names_to = "method",
-                      values_to = "classification") |> 
+    values_to = "classification"
+  ) |>
   dplyr::mutate(
@@
 
-## AUCell classification UMAP
+## `AUCell` classification UMAP
 
@@
   facet_wrap(vars(method),
-             scales = "free_y") +
+    scales = "free_y"
+  ) +
   scale_color_manual(values = plot_colors) +
@@
   dplyr::select(barcodes, classification, method) |>
-  tidyr::pivot_wider(names_from = method, 
-                     values_from = classification)
+  tidyr::pivot_wider(
+    names_from = method,
+    values_from = classification
+  )
 
@@
   marker_genes = annotation_df$marker_gene_classification,
-  col = list(AUCell = plot_colors,
-             marker_genes = plot_colors)
+  col = list(
+    AUCell = plot_colors,
+    marker_genes = plot_colors
+  )
 )
@@
   geom_density(bw = 0.05) +
-  facet_grid(rows = vars(geneset),
+  facet_grid(
+    rows = vars(geneset),
     cols = vars(method),
-             scales = "free_y") +
+    scales = "free_y"
+  ) +
   scale_color_manual(values = plot_colors) +

### 2024-08-01 analyses/cell-type-wilms-tumor-06/notebook/01-clustering_SCPCS000169.Rmd
changes to issue #679

---
title: "Evaluation of clustering parameters - SCPCS000169"
author: maud-p
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: kable
params:
  sample_name: ~
  sample_path: ~
  out_rds_path: ~
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE)
```

## Introduction

This a template for an analysis notebook using RMarkdown.

In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000169) of the Wilms Tumor dataset (SCPCP000006).

This correspond to the step 2 of the proposed analysis: clustering of cells across a set of parameters for few samples

- https://github.com/AlexsLemonade/OpenScPCA-analysis/discussions/635
- https://github.com/AlexsLemonade/OpenScPCA-analysis/issues/671

### Packages

Load required packages in the following chunk, if needed.
Do not install packages here; only load them with the `library()` function.

```{r packages, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(Seurat)      # to use Seurat workflow

library(tidyr)
library(dplyr)
library(DT)          # for table visualization

library(ggplot2)     # for visualization
library(patchwork)   # for visualization
library(ggplotify)
library(SCpubr)      # for visualization
library(viridis)     # for visualization - colors

library(edgeR)       # For pseudobulk based differential expression (DE) analysis
library(DElegate)    # for pseudobulk based DE analysis and identification of marker genes

library(Azimuth)     # For label transfer
```

#### Some config set-up / threshold

We store in a config list names "cfg" parameters used all along the analysis to filter for p-value, log fold change, percentage of expression, etc.

```{r cfg}
cfg = list()
cfg$padj_thershold = 0.05
cfg$lfc_threshold = 1
cfg$rate1_threshold = 0.5
```

#### Base directories

In the chunk below, we include some common directories that you might need, finding their locations using the very useful [`rprojroot` package](https://rprojroot.r-lib.org/).

Note for the DataLab, not sure what to do with this, I decided to use paths below.

```{r base paths, eval=FALSE, include=FALSE}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current")

# The path to this module
module_base <- file.path(repository_base, "analyses", "{{openscpca_module}}")
```

#### Input and output files

In our case, the SCPCP000006 dataset downloaded the 25-06-2024 is locally saved in /mnt_data/Wilms ALSF/SCPCP000006_2024-06-25.
Please mount your local folder of data accordingly or change the path below. Ideally, the mnt_data is set up in the config.yaml file

```{r paths}
# input
# please mount your local folder of data accordingly or change the path below. ideally, the mnt_data is set up in the config.yaml file
path_to_data <- "~/mnt_data/Wilms ALSF/SCPCP000006_2024-06-25"

# get all files in the data directory
filelist <- list.files(path_to_data, full.names = TRUE)

# select the 40 Wilms tumor single nucleus folders only
filelist <- filelist[grepl("/mnt_data/Wilms ALSF/SCPCP000006_2024-06-25/SCPC", filelist)]

# output
## html report for each of the Wilms tumor sample will be saved in 
path_to_report <- "~/notebook/01-clustering"
## some plots from the report will be saved in 
path_to_plot <- "~/plots"
## after final decision on the clustering strategy, processed RDS files will be saved in 
path_to_result <- "~/results"
```

```{r paths2}
## to be modified when rendered
i = 2
filedir <- filelist[i] 
sample <- gsub("/home/rstudio/mnt_data/Wilms ALSF/SCPCP000006_2024-06-25/", "", filedir)

metadata <- read.table("SCPCP000006_metadata.tsv", sep = "\t", header=TRUE)

metadata[metadata$scpca_sample_id == sample,]
```

### Description of the analysis workflow

We perform the following analysis to assess for the quality of clustering:

[1] We perform some quality check to assess any QC-induced clustering (nFeature, nCount, percent.mito).

[2] We add cell cycle information, as we know that in a specific cell cycle state, the transcriptional program is mostly/exclusively related to cell cycle genes and the identity of cells is difficult to determine. We expect these cells to cluster together in a cluster of proliferating cells.

[3] We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method.

[4] We look at specific marker genes that we reported in the table marker.sets/CellType_metadata.csv to check the relevance of the clustering.

[5] We plot pca/umap reduction grouping with available annotations (singler_, cellassign_). We expect at least immune cells to be correctly label and fall into a few set of clusters.

[6] We run label transfer (Azimuth) to transfer annotation from the fetal kidney atlas human reference. We plot pca/umap reduction grouping with latest labels. We expect it to be the most representative of the cell types in the sample.

### Selected parameters

Please note: to keep the notebook as straight as possible, we decided to show the analysis for the selected set of parameters:

- SCTransform (default parameters)
- RunPCA (default) 
- RunUMAP (dims = 1:50) 
- FindNeighbors (dims = 1:50) 
- FindClusters (default)

### Tested parameters

These other parameters have been tested but not selected:

- SCTransform (default parameters) --> RunPCA (default) --> RunUMAP (dims = 1:30) --> FindNeighbors (dims = 1:30) --> FindClusters (default)

## Input and output files

For this analysis, we worked with the _processed.rds data. 
We builded a Seurat object based on the counts data and re-perform the analysis [normalization --> reduction --> clustering] following the Seurat workflow.

We transferred meta.data to keep:

[x] QC data computed by the DataLab

[x] annotation data computed by the DataLab

[x] raw annotation and gene_symbol conversion

## Analysis content

### Load the processed SingleCellExperiment rds object

```{r load}
file <- list.files(filedir)
file <- file[grepl("_processed.rds", file)]

# open the processed rds object
sce <- readRDS(paste0(filedir, "/", file))
```

### Run Seurat pipeline

```{r seurat_wf, fig.width=8, fig.height=4, out.width='100%'}
# convert to seurat
srat <- CreateSeuratObject(counts = counts(sce),
                                    assay = "RNA",
                                    project = file)

# convert colData and rowData to data.frame for use in the Seurat object
cell_metadata <- as.data.frame(colData(sce))
row_metadata <- as.data.frame(rowData(sce))

# add cell metadata (colData) from SingleCellExperiment to Seurat
srat@meta.data <- cell_metadata

# add row metadata (rowData) from SingleCellExperiment to Seurat
srat[["RNA"]]@meta.data <- row_metadata

# add metadata from SingleCellExperiment to Seurat
srat@misc <- metadata(sce)

# Normalization
options(future.globals.maxSize= 891289600)
srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)

# dimensionality reduction
srat <- RunPCA(srat, verbose = F)
srat <- RunUMAP(srat, dims = 1:50, verbose = F)

# clustering
srat    <- FindNeighbors(srat, dims = 1:50, verbose = F)
srat    <- FindClusters(srat, verbose = T)

```

### Visualize seurat clusters

```{r fig.width=8, fig.height=4, out.width='100%'}

d2 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")
d1 <- SCpubr::do_DimPlot(srat, reduction="pca", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - pca")

d1|d2

```

The umap clustering is expected for a triphasic Wilms tumor sample. The 3 main clusters might be:

- blastema cancer cells
- epithelium cancer and/or normal cells
- stroma cancer and/or normal cells

The 2 small clusters C11 and C13 might be immune and endothelial cells.

To be checked in the following analysis!

### Quick check of the QC values
```{r qc, fig.width=10, fig.height=2.5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = c("subsets_mito_percent"), ncol = 3, group.by = "seurat_clusters", legend.position = "none") 
```

### Cell cycle information
```{r cell_cycle, fig.height=20, fig.width=30}
s.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$s.genes]
g2m.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$g2m.genes]

srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE)
```

```{r fig.width=8, fig.height=4, out.width='100%'}

d2 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "Phase", label = TRUE) + ggtitle("Phase - umap")
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")

d1|d2

```

```{r fig.width=10, fig.height=2.5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = c("S.Score", "G2M.Score"), ncol = 2, group.by = "seurat_clusters", legend.position = "none") 
```

Cluster 3 might be composed of cycling cells, in higher proportion than the other clusters.

### Look at specific genes

Here, we open the table of marker genes marker-sets/CellType_metadata.csv.

```{r marker_genes, fig.width=12, fig.height=6, out.width='100%'}

  # open the set of reference marker genes

CellType_metadata <- read.table("marker-sets/CellType_metadata.csv", header = TRUE, sep = ",")

DT::datatable(CellType_metadata, caption = ("CellType_metadata"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))
```

#### Malignant markers

```{r fig.width=20, fig.height=5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("malignant")])] , ncol = 6, group.by = "seurat_clusters", legend.position = "none") 
```
The set of clusters 0-1-3-4-6-10-12-14 is NCAM1+ (ENSG00000149294) which is a marker of blastema area.
C2 is positive for COL6A3 which is a marker of stroma Wilms tumor cells (also normal stroma)

```{r fig.width=20, fig.height=15, out.width='100%'}
 SCpubr::do_FeaturePlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("malignant")])])
```

#### Immune markers

```{r fig.width=20, fig.height=5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("immune")])]
                       , ncol = 6, group.by = "seurat_clusters", legend.position = "none") 
```

```{r fig.width=20, fig.height=15, out.width='100%'}
 SCpubr::do_FeaturePlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("immune")])])
```
C11 is PTPRC+ (ENSG00000081237) and must be a cluster of immune cells. Wilms tumor are generally cold tumor, especially untreated Wilms tumor. According to the clinical data, the present sample is untreated (treatment = upfront resection). We do not expect high percentage of immune cells. The size of the cluster would therefore make sense.

#### Other markers

```{r fig.width=20, fig.height=5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[! CellType_metadata$cell_class %in% c("immune", "malignant")])]
                       , ncol = 6, group.by = "seurat_clusters", legend.position = "none") 
```

```{r fig.width=15, fig.height=5, out.width='100%'}
 SCpubr::do_FeaturePlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[! CellType_metadata$cell_class %in% c("immune", "malignant")])], ncol = 3)
```

C13 is VWF+ (ENSG00000110799) and must be a cluster of endothelial cells.

### Find marker genes for each of the seurat clusters

In addition to the list of known marker genes, we used an unbiased approach to find transcripts that characterized the different clusters.
We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method. 
Please check the preprint from Chistoph Hafemeister: https://www.biorxiv.org/content/10.1101/2023.03.28.534443v1
and tool described here: https://github.com/cancerbits/DElegate

```{r fig.width=10, fig.height=10, out.width='100%'}
feature_conversion <- srat@assays$RNA@meta.data
de_results   <- DElegate::FindAllMarkers2(srat, group_column = "seurat_clusters")

#filter the most relevant markers
s.markers <- de_results[de_results$padj < cfg$padj_thershold & de_results$log_fc > cfg$lfc_threshold & de_results$rate1 > cfg$rate1_threshold,]

# add gene symbol for easiest interpretation of the result
s.markers$gene_ids <- s.markers$feature
s.markers <- left_join(s.markers,feature_conversion, by = c( "gene_ids") )
identical(s.markers$feature, s.markers$gene_ids)     # check the quality of the merge, must be true

DT::datatable(s.markers, caption = ("marker genes"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))

s.markers <- na.omit(s.markers)
s.markers %>%
    group_by(group1) %>%
    top_n(n =  5, wt = log_fc) -> top5

# subset for plotting
cells <- WhichCells(srat, downsample = 100)
ss <- subset(srat, cells = cells)
ss <- ScaleData(ss, c( top5$feature))

p1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")
p2 <- DoHeatmap(ss, features = top5$feature,  cells = cells, group.by = "seurat_clusters") + NoLegend() + 
  scale_fill_gradientn(colors =  c("#01665e","#35978f",'darkslategray3', "#f7f7f7", "#fee391","#fec44f","#F9AD03")) 
p3 <- ggplot(srat@meta.data, aes(seurat_clusters, fill = seurat_clusters)) + geom_bar() + NoLegend()
p4 <- VlnPlot(srat, features = c("nFeature_RNA", "nCount_RNA", "subsets_mito_percent"), ncol = 3)

common_title <- sprintf("Unsupervised clustering %s, %d cells", srat@meta.data$orig.ident[1], ncol(srat))
show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) / p4) + plot_layout(widths = c(1, 2)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))
DT::datatable(top5[,c(1,9,11,12)], caption = ("top 5 marker genes"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))

```

The cluster that we found seems to make sense as they can be defined by a few number of specific genes each. I don't know most of the marker genes found, but few point below that could support the analysis and annotation.

UNC5D (C1, dependency receptor), has been described in cancer and shown to regulate p53-dependent apoptosis (https://pubmed.ncbi.nlm.nih.gov/24691657/). We know that especially in blastema cell, p53 signaling play an important role.

PDCH7 (C7, protocadherin 7) is expressed in human kidney renal tubule (normal kidney epithelium, see https://www.proteinatlas.org/ENSG00000169851-PCDH7/tissue/kidney) and has been associated with oncogenic function in lung cancer (https://pubmed.ncbi.nlm.nih.gov/30409919/, https://www.frontiersin.org/journals/pharmacology/articles/10.3389/fphar.2023.1217213/full).

EBF1 (C8, Early B Cell factor 1) has been shown to play a role in kidney development, esp. glomerular, while expressed in stroma progenitor (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6727263/).

ROR1 (C8, Receptor tyrosine kinase-like orphan receptor 1) ROR1 is only detectable in embryonic tissue and generally absent in mature tissue. It has been associated with cancer (https://academic.oup.com/proteincell/article/5/7/496/6831810) and cancer stroma (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5367847/).

ERBB4 (C9, EGF receptor family) is expressed during kidney development of epithelial tubule and uteric bud (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3269924/)

FLT1 (C13, vascular endothelial growth factor receptor 1), PCAM and ELMO1 in the top5 of C13 marker genes confirm the identification of endothelial cells (https://www.ahajournals.org/doi/pdf/10.1161/CIRCRESAHA.109.213983, https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4986701/)

### Annotation from SingleR (no cancer or kidney specific dataset)

Here, we quickly checked annotations that are present in the _processed rds object. However, the automated annotation have not been performed using a cancer specific reference or a kidney reference. We do not expect a nice labelling of the cells as the overlap of cell types between the reference and the query dataset is poor. This support the need to do a proper label transfer from the fetal kidney atlas, which is imho the best reference that can be applied to a Wilms tumor query.

```{r fig.width=15, fig.height=10, out.width='100%'}

d2 <-DimPlot(srat, group.by = "singler_celltype_annotation", reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()

DT::datatable(table(srat$seurat_clusters, srat$singler_celltype_annotation), caption = ("table of SingleR annotation per seurat clusters"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))

d3 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "singler_celltype_annotation",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

d2|d3
```

### Annotation from cellassign (no cancer or kidney specific dataset)

```{r fig.width=15, fig.height=10, out.width='100%'}

d2 <-DimPlot(srat, group.by = "cellassign_celltype_annotation", reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()

DT::datatable(table(srat$seurat_clusters, srat$cellassign_celltype_annotation)
, caption = ("table of cellassign annotation per seurat clusters"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))

d3 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "cellassign_celltype_annotation",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

d2|d3
```

Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, it confirm once again that C13 is a cluster of endothelial cells and C11 a cluster of immune cells.

### Azimuth annotation from fetal kidney

Note to the DataLab : where should I save the fetal kidney reference? In S3 like results as it is a quite big object or in the folder marker-sets?

For more information related to the reference, please go to https://www.kidneycellatlas.org/
You will find:

- interactive viewer

- h5ad files to download.

Please note that as Wilms tumor have been described to be closer to fetal kidney as mature kidney, we only used the fetal kidney atlas as the reference. 
Also check : https://www.science.org/doi/10.1126/science.aat5031

```{r fig.height=6, fig.width=6, message=FALSE, warnings=FALSE}

# access to the azimuth kidney reference 
##### to be updated depending on the location of the reference file 
ref_dir = "~/mnt_resources/datasets/Kidney_cell_atlas/fetal_full/Azimuth_Compatible_Fetal_full"

reference <- LoadReference(ref_dir)
# Load the query object for mapping

# Preprocess with SCTransform
srat <-  RunAzimuth(srat, ref_dir, assay = 'RNA')
rm(reference)
gc()

DT::datatable(table(srat$seurat_clusters, srat$predicted.compartment), caption = ("table of azimuth compartment annotation per seurat clusters"),               extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))

DT::datatable(table(srat$seurat_clusters, srat$predicted.celltype), caption = ("table of azimuth annotation per seurat clusters"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))

```

```{r fig.height=5, fig.width=20, warnings=FALSE}

d1 <- DimPlot(srat, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + NoLegend()
d2 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
d3 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "predicted.celltype", label = TRUE, repel = TRUE) + NoLegend()

d1 | d2 | d3
```

Looking at the predicted compartment, we are satisfied that the immune, endothelium and stroma compartment are specific to a set of clusters. It is expected that the fetal__nephron compartment match to both blastema and epithelial cancer and normal kidney. It is thus OK to have it labeling two distinct set of clusters.

Looking at the predicted cell types, we observed that the set of cluster [0,1,3,4,6,10,12,14] is labeled as cap mesenchyme, an early developmental stage of nephron, while the set of clusters [5,7,9] is labeled with more differentiated epithelial units of the kidney. 
Based on this observation and the expression of NCAM1, UNC5D (C1) and PDCH7 (C7), ERBB4 (C9), we can conclude that the set of cluster [0,1,3,4,6,10,12,14] is mostly blastema cancer cells and the set of clusters [5,7,9] is eipthelial cancer or normal cells.

```{r fig.height=5, fig.width=20, warnings=FALSE}

f1 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "predicted.compartment",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

f2 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "predicted.celltype",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

f1 | f2

```

## Summary and conclusion

Taking together, we can conclude:

- C13 = endothelium (with high confidency)
- C11 = immune cells (with high confidency)

- C2 and C8, that form a unit in the umap reduction, are stromal cells, either cancer or normal (with high confidency)
- C[0-1-3-4-6-10-12-14] that form a unit in the umap reduction, are label as cap mesenchyme (<=> stem) and NCAM+ must be blastema cancer cells.
- C[5-7-9] that form a unit in the umap reduction, are label as epithelial cells (SingleR) and differentiated kidney epithelium subunit (Fetal kidney atlas) must be epithelial (cancer or normal)

The primary aim of this script was to assess the quality of clustering and play with clustering parameters. On this sample, we are happy with the clustering, as we could identify:
- one cluster of immune cells (cell marker PTPRC, SingleR, Azimuth label transfer fro the fetal kidney) 
- one cluster of endothelial cells (cell marker VWF, SingleR, Azimuth label transfer fro the fetal kidney)
- one set of clusters of stromal cells
- one set of cluster of epithelial cells
- one set of cluster of blastema cancer cells.

Additionally, we showed the feasibility to run label transfer using runAzimuth (part of step 3 of the proposed analysis:  https://github.com/maud-p/OpenScPCA-analysis/issues/1).

We also showed the easy and robust identification of normal cells (endothelial and immune cells) that would serve as reference for the CNV inference (inferCNV, step 4 of the analysis).

## Next step

[ ] We will adapt this template to be render on all the 40 Wilms tumor samples of the dataset.

[ ] We will save for each sample the rds file

[ ] We will run inferCNV for each sample to decide the malignant/normal status of some stroma and epithelial cluster, and confirm the blastema annotation.

The next step will provide us a better understanding of the entire cohort. We will then have to set up a strategy to annotate each sample. Open questions are:

[ ] should we annotate single cell                      
or   
[] consider applying similar annotations to all cells in a cluster?

[ ] manual annotation of each cluster / each patient    
or   
[] automated annotation using some threshold?

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-08-01 analyses/cell-type-ewings/exploratory_analysis/05-cluster-exploration.Rmd
notebook to explore clusters in 822 and 824

---
title: "Clustering SCPCL000822 and SCPCL000824"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
---

This notebook does some initial exploration of clustering in two libraries from `SCPCP000015`: `SCPCL000822` and `SCPCL000824`.

- Louvain clusters are calculated using a range of parameters specifying nearest neighbors (5, 10, 15, 20, 25, 30, 35, and 40). 
- Metrics are then calculated to evaluate clustering results across all parameters: 
  - Silhouette width 
  - Cluster purity 
  - Cluster stability

The clusters are then compared to the results from running `SingleR` in the `aucell-singler-annotation.sh` workflow.
We then look at marker gene expression across the clusters and assign each cluster to a cell type.

## Setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_bw()
)

# set seed
set.seed(2024)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", "SCPCP000015")

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-ewings")

# path to marker genes file
marker_genes_file <- file.path(module_base, "references", "visser-all-marker-genes.tsv")
```

```{r}
# source in helper functions for make_jaccard_matrix() and jaccard()
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)

# source in helper functions: plot_gene_heatmap() and plot_cnv_heatmap()
# create_classification_df() and create_marker_gene_df()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)
```

## Define functions

```{r, assign_clusters_functions}
# get clusters for a specified value of k
get_clusters <- function(pcs, k) {
  clusters <- bluster::clusterRows(
    pcs,
    bluster::NNGraphParam(
      k = k,
      type = "jaccard",
      cluster.fun = "louvain"
    )
  )

  return(clusters)
}

# define a function to perform clustersweep and get clusters across multiple values of k
cluster_sweep <- function(sce) {
  # first perform clustering across parameters
  cluster_results <- bluster::clusterSweep(reducedDim(sce, "PCA"),
    bluster::NNGraphParam(),
    k = as.integer(seq(5, 40, 5)),
    cluster.fun = "louvain",
    type = "jaccard"
  )

  # turn results into a data frame
  cluster_df <- cluster_results$clusters |>
    as.data.frame() |>
    # add barcode column
    dplyr::mutate(barcodes = colnames(sce)) |>
    # combine all cluster results into one column
    tidyr::pivot_longer(
      cols = ends_with("jaccard"),
      names_to = "params",
      values_to = "cluster"
    ) |>
    # separate out parameters, nn, function, and type into their own columns
    dplyr::mutate(
      nn_param = stringr::word(params, 1, sep = "_") |>
        stringr::str_replace("k.", "k_"),
      cluster_fun = stringr::word(params, 2, sep = "_") |>
        stringr::str_remove("cluster.fun."),
      cluster_type = stringr::word(params, -1, sep = "_") |>
        stringr::str_remove("type.")
    ) |>
    # remove combined params column
    dplyr::select(-params)

  return(cluster_df)
}
```

```{r, cluster_stats_functions}
# cluster statistics functions
# get silhouette width and cluster purity for each cluster
# calculates values across all nn_param options used to determine clustering
# all_cluster_results must have nn_param column
get_cluster_stats <- function(sce,
                              id,
                              all_cluster_results) {
  pcs <- reducedDim(sce, "PCA")

  # split clustering results by library id and param used
  split_clusters <- all_cluster_results |>
    dplyr::filter(library_id == id) |>
    split(all_cluster_results$nn_param)

  # for each nn_param get cluster width and purity
  all_stats_df <- split_clusters |>
    purrr::map(\(df){
      sil_df <- bluster::approxSilhouette(pcs, df$cluster) |>
        as.data.frame() |>
        tibble::rownames_to_column("barcodes")

      purity_df <- bluster::neighborPurity(pcs, df$cluster) |>
        as.data.frame() |>
        tibble::rownames_to_column("barcodes")

      # join into one data frame to return
      stats_df <- sil_df |>
        dplyr::left_join(purity_df, by = "barcodes")

      return(stats_df)
    }) |>
    dplyr::bind_rows(.id = "nn_param")

  return(all_stats_df)
}

# calculate cluster stability for a single set of clusters using ari
# bootstrap and get ari for clusters compared to sampled clusters
# re-clusters and gets ari across 20 iterations
get_ari <- function(pcs,
                    clusters,
                    k) {
  ari <- c()
  for (iter in 1:20) {
    # sample cells with replacement
    sample_cells <- sample(nrow(pcs), nrow(pcs), replace = TRUE)
    resampled_pca <- pcs[sample_cells, , drop = FALSE]

    # perform clustering on sampled cells
    resampled_clusters <- get_clusters(resampled_pca, k)

    # calculate ARI between new clustering and original clustering
    ari[iter] <- pdfCluster::adj.rand.index(resampled_clusters, clusters[sample_cells])
  }

  ari_df <- data.frame(
    ari = ari,
    k_value = k
  )
}

# get cluster stability for each nn_param cluster results are available for
# calculated the ari between original clusters and shuffled clusters 20 times
get_cluster_stability <- function(sce,
                                  id,
                                  all_cluster_results) {
  pcs <- reducedDim(sce, "PCA")

  # split clustering results by library id and param used
  cluster_df_list <- all_cluster_results |>
    dplyr::filter(library_id == id) |>
    split(all_cluster_results$nn_param)

  # for each parameter, get ari values
  cluster_stability_df <- cluster_df_list |>
    purrr::imap(\(df, k_value){
      # make sure k is numeric and remove extra k_
      k <- stringr::str_remove(k_value, "k_") |>
        as.numeric()

      get_ari(pcs, df$cluster, k)
    }) |>
    dplyr::bind_rows()

  return(cluster_stability_df)
}
```

```{r, plotting_functions}
# heatmap comparing cluster assignments to SingleR labels
cluster_celltype_heatmap <- function(cluster_classification_df,
                                     id) {
  # get a jaccard mtx for each cluster param
  jaccard_df_list <- cluster_classification_df |>
    dplyr::filter(library_id == id) |>
    split(cluster_classification_df$nn_param) |>
    purrr::map(\(df) {
      make_jaccard_matrix(
        df,
        "cluster",
        "singler_lumped"
      )
    })

  # turn into heatmap list
  make_heatmap_list(jaccard_df_list, column_title = glue::glue("{id}-clusters"), legend_match = "k_5")
}

# Density plot looking at marker gene expression across all clusters
# each panel is from a different marker gene list
# each row shows marker gene expression for that cluster
plot_marker_genes <- function(cluster_classification_df,
                              k_value,
                              id) {
  # pick clustering to use and select those columns
  final_clusters_df <- cluster_classification_df |>
    dplyr::filter(nn_param == k_value) |>
    dplyr::filter(library_id == id)

  # grab columns that contain marker gene sums
  marker_gene_columns <- colnames(final_clusters_df)[which(endsWith(colnames(final_clusters_df), "_sum"))]

  # create individual density plots and combine into one
  marker_gene_columns |>
    purrr::map(\(column){
      plot_density(
        final_clusters_df,
        column,
        "cluster"
      )
    }) |>
    patchwork::wrap_plots(ncol = 1) + patchwork::plot_annotation(glue::glue("{id}-{k_value}-clusters"))
}
```

## Prepare data

```{r}
# sample and library ids
sample_ids <- c("SCPCS000490", "SCPCS000492")
library_ids <- c("SCPCL000822", "SCPCL000824")

# define input sce files
sce_file_names <- glue::glue("{library_ids}_processed.rds")
sce_files <- file.path(data_dir, sample_ids, sce_file_names) |>
  purrr::set_names(library_ids)

# define classification files
singler_results_dir <- file.path(module_base, "results", "aucell_singler_annotation")
classification_file_names <- glue::glue("{library_ids}_singler-classifications.tsv")
classification_files <- file.path(singler_results_dir, sample_ids, classification_file_names) |>
  purrr::set_names(library_ids)

# define output files
cluster_results_dir <- file.path(module_base, "results", "clustering", sample_ids)
cluster_results_dir |>
  purrr::map(fs::dir_create)

cluster_file_names <- glue::glue("{library_ids}_cluster-results.tsv")
cluster_output_files <- file.path(cluster_results_dir, cluster_file_names) |>
  purrr::set_names(library_ids)
```

```{r}
# read in both SCEs
sce_list <- sce_files |>
  purrr::map(readr::read_rds)

# read in classification data frame with SingleR results and combine into one
classification_df <- classification_files |>
  purrr::map(readr::read_tsv) |>
  dplyr::bind_rows(.id = "library_id")
```

```{r}
# read in marker genes table
marker_genes_df <- readr::read_tsv(marker_genes_file, show_col_types = FALSE) |>
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |>
  dplyr::distinct()

# get list of all cell types found
cell_types <- unique(marker_genes_df$cell_type)

# get the sum of expression of all genes for each cell type
gene_exp_df <- sce_list |>
  purrr::map(\(sce) {
    cell_types |>
      purrr::map(\(type){
        calculate_sum_markers(marker_genes_df, sce, type)
      }) |>
      purrr::reduce(dplyr::inner_join, by = "barcodes")
  }) |>
  dplyr::bind_rows(.id = "library_id")

# join sum expression columns with classification df
classification_df <- classification_df |>
  dplyr::left_join(gene_exp_df, by = c("barcodes", "library_id"))
```

## Clustering

Below we perform Louvain, Jaccard clustering, varying `k`. 
The minimum `k` is 5 and the maximum `k` is 40 with a step size of 5.

```{r}
# perform clustering across k = 5-40 with increments of 5
all_cluster_results <- sce_list |>
  purrr::map(get_clusters) |>
  dplyr::bind_rows(.id = "library_id")
```

```{r}
# get umap embeddings and combine into a data frame with cluster assignments
umap_df <- sce_list |>
  purrr::map(\(sce){
    df <- sce |>
      scuttle::makePerCellDF(use.dimred = "UMAP") |>
      # replace UMAP.1 with UMAP1 and get rid of excess columns
      dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2)
  }) |>
  dplyr::bind_rows(.id = "library_id") |>
  dplyr::left_join(all_cluster_results, by = c("library_id", "barcodes"))
```

Below we visualize the cluster assignments using each parameter on a UMAP. 
This isn't particularly useful because of clusters, but it can be helpful to see any values of K that may have obvious over or under clustering.

```{r, fig.height=15, fig.width=10}
# look at clustering results across all params
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cluster)) +
  geom_point(alpha = 0.5, size = 0.1) +
  facet_grid(
    cols = vars(library_id),
    rows = vars(nn_param)
  ) +
  theme(aspect.ratio = 1)
```

### Clustering statistics

Below we calculate a series of statistics:

- Average silhouette width: This metric evaluates cluster separation. 
Cells with large positive silhouette widths are closer to other cells in the same cluster than to cells in different clusters. 
- Average cluster purity: This metric also evaluates cluster separation and tells us the proportion of neighboring cells that are assigned to the same cluster. 
Higher purity values indicate that clusters are well separated. 
- Cluster stability: This evaluates how stable the clustering is to input data. 
Higher values of cluster stability indicate more reproducible clusters.

```{r, warning=FALSE}
# get a combined stats dataframe with purity and width for all clusters
all_stats_df <- sce_list |>
  purrr::imap(\(sce, library_id){
    get_cluster_stats(sce, library_id, all_cluster_results)
  }) |>
  dplyr::bind_rows(.id = "library_id")
```

```{r}
# summarize the silhouette width and purity for each nn_param used
summary_stats_df <- all_stats_df |>
  dplyr::group_by(library_id, nn_param) |>
  dplyr::summarise(
    mean_width = mean(width),
    mean_purity = mean(purity)
  ) |>
  dplyr::mutate(
    # make sure the order is correct
    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
  )
```

```{r}
# silhouette width for different params
ggplot(summary_stats_df, aes(x = nn_param, y = mean_width)) +
  geom_point() +
  facet_wrap(vars(library_id),
    scales = "free_y"
  ) +
  labs(y = "Average silhouette width")
```

```{r}
# cluster purity for different params
ggplot(summary_stats_df, aes(x = nn_param, y = mean_purity)) +
  geom_point() +
  facet_wrap(vars(library_id),
    scales = "free_y"
  ) +
  labs(y = "Average cluster purity")
```

Here we see that both silhouette width and cluster purity peak around k=20 or k=25. 
For `SCPCL000822`, increasing k after 25 does not change the width or purity, 
For `SCPCL000824`, increasing k after 20-25 decreases the width and keeps purity around the same.

```{r, warning=FALSE, message=FALSE}
# calculate cluster stability
stability_df <- sce_list |>
  purrr::imap(\(sce, id){
    get_cluster_stability(sce, id, all_cluster_results)
  }) |>
  dplyr::bind_rows(.id = "library_id") |>
  dplyr::mutate(
    k_value = as.factor(k_value),
    # make sure that k = 5 comes first
    k_value = forcats::fct_relevel(k_value, "5", after = 0)
  )

# plot stability across all values of k
ggplot(stability_df, aes(x = k_value, y = ari)) +
  geom_point() +
  facet_grid(rows = vars(library_id)) +
  title("Cluster stability")
```

Here we see that cluster stability increases as k increases for both libraries. 
We see a plateau start to appear around k = 20 and k = 25.

## Compare clusters to cell types from `SingleR`

Now we will compare the clustering results to the cell type assignments obtained from `SingleR` in the `aucell-singler-annotation.sh` workflow. 
To compare results we will calculate the Jaccard similarity index between clusters and cell types. 
We expect that good clustering will line up with the cell type assignments so that there is ~ 1 cell type per cluster.

For the plots, we will only display the top 7 cell types and all other cells will be lumped together into `All remaining cell types`.

```{r}
cluster_classification_df <- all_cluster_results |>
  # add in classifications from singler
  dplyr::left_join(classification_df, by = c("library_id", "barcodes")) |>
  dplyr::mutate(
    # first grab anything that is tumor and label it tumor
    # NA should be unknown
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later
    singler_lumped = singler_annotation |>
      forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf)
  )
```

```{r, fig.height=10, fig.width=7}
# get heatmap for each library showing cluster assignments vs. singler assignments
names(sce_list) |>
  purrr::map(\(id){
    cluster_celltype_heatmap(
      cluster_classification_df,
      id
    )
  })
```

From these plots, it's obvious that K = 5 is over clustering and cell types are spread across a lot of clusters. 
However, for the other values of K, there doesn't seem to be any striking difference that are obvious by eye.

## Marker gene expression across clusters

We will also look at the expression of marker genes across each cluster. 
In these plots, each row shows the distribution of the specified marker genes in that cluster. 
Each panel is labeled with the marker gene expression being plotted.

Based on the above results, we will only show these plots for a value of K = 25.

```{r, fig.height=10, fig.width=5}
# plot to look at marker gene expression for all cell types across all clusters
# choose k=25 based on previous stats
names(sce_list) |>
  purrr::map(\(id) {
    plot_marker_genes(cluster_classification_df,
      k_value = "k_25",
      id
    )
  })
```

For both samples we see very clear separation between clusters that express tumor marker genes and each of the normal cell marker genes.

## Assign cell types to clusters

Finally, we will assign cell types to all cells in a cluster based on a cluster having high expression of a set of marker genes.

- Any clusters with high tumor marker gene expression will be labeled "tumor". 
- Any clusters with high immune marker gene expression also correspond to clusters that are predominantly "macrophage" in the heatmaps, so they will be labeled as "macrophage". 
- Any clusters with high endothelial marker gene expression also correspond to clusters that are predominantly "endothelial cell" in the heatmaps, so they will be labeled as "macrophage". 
- There is one cluster in `SCPCL000822` that has high mesenchymal gene expression and is predominantly chondrocytes so we will label this cell as "chondrocytes". 
- All other clusters with high mesenchymal gene expression appear to have a mix of chondrocytes, fibroblasts, and smooth muscle cells so we will just label these as "mesenchymal-like cells".

```{r}
# get umap embeddings for plotting later
umap_df <- umap_df |>
  dplyr::select(library_id, barcodes, UMAP1, UMAP2) |>
  dplyr::distinct()

# assign clusters based on marker gene expression and cell types that are prominent in that cluster
# use the jaccard matrices to help guide assignments
cluster_umap_df <- cluster_classification_df |>
  dplyr::left_join(umap_df, by = c("library_id", "barcodes")) |>
  dplyr::filter(nn_param == "k_25") |>
  dplyr::mutate(
    cluster_celltype = dplyr::case_when(
      library_id == "SCPCL000822" & cluster %in% c(1, 2, 4, 5, 9) ~ "tumor",
      library_id == "SCPCL000822" & cluster == 7 ~ "endothelial cell",
      # cluster 3 has predominantly chondrocytes and has MSC marker gene expression
      library_id == "SCPCL000822" & cluster == 3 ~ "chondrocyte",
      # clusters with high "immune" marker gene expression are macrophage
      library_id == "SCPCL000822" & cluster == 6 ~ "macrophage",
      # cluster 8 has high MSC marker gene expression and a mix of MSC like cell types
      library_id == "SCPCL000822" & cluster == 8 ~ "mesenchymal-like cell",
      library_id == "SCPCL000824" & cluster %in% c(1, 2, 3, 4, 5, 6, 11) ~ "tumor",
      library_id == "SCPCL000824" & cluster == 8 ~ "endothelial cell",
      # cluster 10 has high immune markers and is mostly macrohpages
      library_id == "SCPCL000824" & cluster == 10 ~ "macrophage",
      # cluster 7 has high MSC marker gene expression and a mix of MSC like cell types
      library_id == "SCPCL000824" & cluster == 7 ~ "mesenchymal-like cell",
      # if they didn't have high expression of any markers, we label them as unknown
      .default = "unknown"
    ),
    # make tumor first
    cluster_celltype = forcats::fct_relevel(cluster_celltype, "tumor", after = 0)
  )
```

```{r}
# look at cell type assignments on the UMAP
ggplot(cluster_umap_df, aes(x = UMAP1, y = UMAP2, color = cluster_celltype)) +
  geom_point(alpha = 0.5, size = 0.5) +
  facet_grid(cols = vars(library_id)) +
  theme(aspect.ratio = 1) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5))) +
  scale_color_brewer(palette = "Dark2")
```

We can also validate that these cells do in fact have higher expression of the marker genes as compared to other cells by plotting the marker gene expression on UMAPs.

```{r}
# grab columns that contain marker gene sums
marker_gene_columns <- colnames(cluster_umap_df)[which(endsWith(colnames(cluster_umap_df), "_sum"))]

# create individual density plots and combine into one
marker_gene_columns |>
  purrr::map(\(column){
    genelist_name <- stringr::str_remove(column, "_sum")

    ggplot(cluster_umap_df, aes(x = UMAP1, y = UMAP2, color = !!sym(column))) +
      geom_point(alpha = 0.5, size = 0.5) +
      facet_grid(cols = vars(library_id)) +
      theme(aspect.ratio = 1) +
      scale_color_viridis_c() +
      labs(
        title = genelist_name
      )
  })
```

## Save clustering and annotations

Here we will save the assigned clusters and cell type annotations for each library. 
Each file will have the following columns:

- `barcodes`: Unique cell barcode
- `cluster`: Assigned cluster
- `cluster_celltype`: Cell type assigned based on clustering and marker gene expression 
- `nn_param`: Number of nearest neighbors used as a parameter for clustering
- `singler_ontology`: Cell ontology identifier for assigned cell type by `SingleR`
- `singler_annotation`: Cell type assigned by `SingleR`
- `consensus_celltype`: Consensus between cluster assignments and `SingleR` assignments for tumor cells. 
Here any cell in a cluster with tumor cells will be labeled with `tumor` and all other cells will be labeled with the `SingleR` annotation.

```{r}
cluster_output_files |>
  purrr::iwalk(\(file, id){
    cluster_umap_df |>
      dplyr::filter(library_id == id) |>
      dplyr::select(barcodes, cluster, cluster_celltype, nn_param, singler_ontology, singler_annotation) |>
      dplyr::mutate(consensus_celltype = dplyr::if_else(cluster_celltype == "tumor", "tumor", singler_annotation)) |>
      readr::write_tsv(file)
  })
```

## Session info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-08-01 analyses/cell-type-ewings/exploratory_analysis/05-cluster-exploration.Rmd
render notebook

@@
 cluster_results_dir |>
-  purrr::map(fs::dir_create)
+  purrr::walk(fs::dir_create)
 
@@
 all_cluster_results <- sce_list |>
-  purrr::map(get_clusters) |>
+  purrr::map(cluster_sweep) |>
   dplyr::bind_rows(.id = "library_id")
@@
   facet_grid(rows = vars(library_id)) +
-  title("Cluster stability")
+  labs(title = "Cluster stability")
 ```

### 2024-08-02 analyses/cell-type-wilms-tumor-06/notebook/01-clustering_SCPCS000169.Rmd
Update 01-clustering_SCPCS000169.Rmd
Because of data download with the download-data.py method. Update paths

@@
 
-In the chunk below, we include some common directories that you might need, finding their locations using the very useful [`rprojroot` package](https://rprojroot.r-lib.org/).
-
-Note for the DataLab, not sure what to do with this, I decided to use paths below.
-
-```{r base paths, eval=FALSE, include=FALSE}
+```{r base paths, eval=TRUE, include=TRUE}
 # The base path for the OpenScPCA repository, found by its (hidden) .git directory
@@
 # The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "current")
+data_dir <- file.path(repository_base, "data", "2024-07-08", "SCPCP000006")
 
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "{{openscpca_module}}")
+module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
 ```
@@
 
-In our case, the SCPCP000006 dataset downloaded the 25-06-2024 is locally saved in /mnt_data/Wilms ALSF/SCPCP000006_2024-06-25.
-Please mount your local folder of data accordingly or change the path below. Ideally, the mnt_data is set up in the config.yaml file
-
 ```{r paths}
 # input
-# please mount your local folder of data accordingly or change the path below. ideally, the mnt_data is set up in the config.yaml file
-path_to_data <- "~/mnt_data/Wilms ALSF/SCPCP000006_2024-06-25"
+# use download-data.py to download the data
 
 # get all files in the data directory
-filelist <- list.files(path_to_data, full.names = TRUE)
+filelist <- list.files(data_dir, full.names = TRUE)
 
 # select the 40 Wilms tumor single nucleus folders only
-filelist <- filelist[grepl("/mnt_data/Wilms ALSF/SCPCP000006_2024-06-25/SCPC", filelist)]
+filelist <- filelist[! grepl(".tsv", filelist)]
 
@@
 ## html report for each of the Wilms tumor sample will be saved in 
-path_to_report <- "~/notebook/01-clustering"
+path_to_report <- file.path(module_base, "01-clustering")
 ## some plots from the report will be saved in 
-path_to_plot <- "~/plots"
+path_to_plot <- file.path(module_base, "/plots")
 ## after final decision on the clustering strategy, processed RDS files will be saved in 
-path_to_result <- "~/results"
+path_to_result <- file.path(module_base, "/results")
 ```
@@
 filedir <- filelist[i] 
-sample <- gsub("/home/rstudio/mnt_data/Wilms ALSF/SCPCP000006_2024-06-25/", "", filedir)
+sample <- gsub(paste0(data_dir, "/"), "", filedir)
 
-metadata <- read.table("SCPCP000006_metadata.tsv", sep = "\t", header=TRUE)
+metadata <- read.table(file.path(data_dir, "/single_cell_metadata.tsv"), sep = "\t", header=TRUE)
 

### 2024-08-02 analyses/cell-type-wilms-tumor-06/notebook/01-clustering_SCPCS000169.Rmd
change fetal kidney reference path

@@
 
-In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000169) of the Wilms Tumor dataset (SCPCP000006).
+In this notebook, we will set up parameters in the seurat workflow 
+[normalization --> reduction --> clustering] 
+for one Wilms tumor sample (SCPCS000169) of the Wilms Tumor dataset (SCPCP000006).
 
@@
 
-[5] We plot pca/umap reduction grouping with available annotations (singler_, cellassign_). We expect at least immune cells to be correctly label and fall into a few set of clusters.
+[5] We plot pca/umap reduction grouping with available annotations (singler_, cellassign_). 
+We expect at least immune cells to be correctly label and fall into a few set of clusters.
 
 [6] We run label transfer (Azimuth) to transfer annotation from the fetal kidney atlas human reference. 
-We plot pca/umap reduction grouping with latest labels.
-We expect it to be the most representative of the cell types in the sample.
+We plot pca/umap reduction grouping with latest labels. We expect it to be the most representative of the cell types in the sample.
 
@@
 For this analysis, we worked with the _processed.rds data. 
-We builded a Seurat object based on the counts data and re-perform the analysis [normalization --> reduction --> clustering] following the Seurat workflow.
+We builded a Seurat object based on the counts data and re-perform the analysis 
+[normalization --> reduction --> clustering] following the Seurat workflow.
 
@@
 
-Note to the DataLab : where should I save the fetal kidney reference? In S3 like results as it is a quite big object or in the folder marker-sets?
+Note to the DataLab : where should I save the fetal kidney reference? 
+In S3 like results as it is a quite big object or in the folder marker-sets?
+The reference in now in the module folder, marker-sets
+For more information related to the reference, please go to https://www.kidneycellatlas.org/
 
-For more information related to the reference, please go to https://www.kidneycellatlas.org/
 You will find:
@@
 ##### to be updated depending on the location of the reference file 
-ref_dir = "~/mnt_resources/datasets/Kidney_cell_atlas/fetal_full/Azimuth_Compatible_Fetal_full"
+ref_dir = file.path(module_base, "marker-sets/Azimuth_Compatible_Fetal_full/")
 
@@
 
-The primary aim of this script was to assess the quality of clustering and play with clustering parameters. On this sample, we are happy with the clustering, as we could identify:
+The primary aim of this script was to assess the quality of clustering and play with clustering parameters. 
+On this sample, we are happy with the clustering, as we could identify:
 - one cluster of immune cells (cell marker PTPRC, SingleR, Azimuth label transfer fro the fetal kidney) 

### 2024-08-02 analyses/cell-type-ewings/exploratory_analysis/05-cluster-exploration.Rmd
Apply suggestions from code review
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 # cluster statistics functions
+
 # get silhouette width and cluster purity for each cluster
@@
 # get cluster stability for each nn_param cluster results are available for
-# calculated the ari between original clusters and shuffled clusters 20 times
 get_cluster_stability <- function(sce,
@@
 ggplot(stability_df, aes(x = k_value, y = ari)) +
-  geom_point() +
+  geom_jitter(width = 0.1) +
   facet_grid(rows = vars(library_id)) +
@@
 From these plots, it's obvious that K = 5 is over clustering and cell types are spread across a lot of clusters. 
-However, for the other values of K, there doesn't seem to be any striking difference that are obvious by eye.
+However, for the other values of K, there don't seem to be any striking differences that are obvious by eye.
 
@@
 
-# create individual density plots and combine into one
+# create individual umaps and combine into one
 marker_gene_columns |>

### 2024-08-02 analyses/cell-type-ewings/exploratory_analysis/05-cluster-exploration.Rmd
update plots based on review

@@
 
-# source in helper functions: plot_gene_heatmap() and plot_cnv_heatmap()
-# create_classification_df() and create_marker_gene_df()
+# source in helper functions: plot_density() and calculate_sum_markers()
 validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
@@
 ```{r, assign_clusters_functions}
-# get clusters for a specified value of k
+# get louvain, jaccard clusters for a specified value of k (nearest neighbors)
 get_clusters <- function(pcs, k) {
@@
 ```{r, plotting_functions}
+# plot individual stats for clusters, either purity or width
+plot_cluster_stats <- function(all_stats_df,
+                               stat_column,
+                               id) {
+  plot_df <- all_stats_df |>
+    dplyr::filter(library_id == id)
+
+  ggplot(plot_df, aes(x = nn_param, y = {{ stat_column }})) +
+    # ggforce::geom_sina(size = .2) +
+    ggbeeswarm::geom_quasirandom(method = "smiley", size = 0.1) +
+    stat_summary(
+      aes(group = nn_param),
+      color = "red",
+      # median and quartiles for point range
+      fun = "median",
+      fun.min = function(x) {
+        quantile(x, 0.25)
+      },
+      fun.max = function(x) {
+        quantile(x, 0.75)
+      }
+    )
+}
+
 # heatmap comparing cluster assignments to SingleR labels
@@
   # turn into heatmap list
-  make_heatmap_list(jaccard_df_list, column_title = glue::glue("{id}-clusters"), legend_match = "k_5")
+  make_heatmap_list(jaccard_df_list, column_title = glue::glue("{id}-clusters"), legend_match = "k_5", cluster_rows = FALSE)
 }
@@
   purrr::map(cluster_sweep) |>
-  dplyr::bind_rows(.id = "library_id")
+  dplyr::bind_rows(.id = "library_id") |>
+  dplyr::mutate(
+    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
+  )
 ```
@@
 ```{r, fig.height=15, fig.width=10}
-# look at clustering results across all params
-ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cluster)) +
+# look at clustering results for each library across all params
+library_ids |>
+  purrr::map(\(id){
+    umap_df |>
+      dplyr::filter(library_id == id) |>
+      ggplot(aes(x = UMAP1, y = UMAP2, color = cluster)) +
       geom_point(alpha = 0.5, size = 0.1) +
-  facet_grid(
-    cols = vars(library_id),
-    rows = vars(nn_param)
+      facet_wrap(vars(nn_param)) +
+      theme(
+        aspect.ratio = 1
       ) +
-  theme(aspect.ratio = 1)
+      labs(title = id) +
+      guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))
+  })
 ```
@@
   }) |>
-  dplyr::bind_rows(.id = "library_id")
-```
-
-```{r}
-# summarize the silhouette width and purity for each nn_param used
-summary_stats_df <- all_stats_df |>
-  dplyr::group_by(library_id, nn_param) |>
-  dplyr::summarise(
-    mean_width = mean(width),
-    mean_purity = mean(purity)
-  ) |>
+  dplyr::bind_rows(.id = "library_id") |>
   dplyr::mutate(
@@
 # silhouette width for different params
-ggplot(summary_stats_df, aes(x = nn_param, y = mean_width)) +
-  geom_point() +
-  facet_wrap(vars(library_id),
-    scales = "free_y"
-  ) +
-  labs(y = "Average silhouette width")
+library_ids |>
+  purrr::map(\(id){
+    plot_cluster_stats(all_stats_df, width, id)
+  })
 ```
@@
 # cluster purity for different params
-ggplot(summary_stats_df, aes(x = nn_param, y = mean_purity)) +
-  geom_point() +
-  facet_wrap(vars(library_id),
-    scales = "free_y"
-  ) +
-  labs(y = "Average cluster purity")
+library_ids |>
+  purrr::map(\(id){
+    plot_cluster_stats(all_stats_df, purity, id)
+  })
 ```
@@
   facet_grid(rows = vars(library_id)) +
-  labs(title = "Cluster stability")
+  labs(title = "Cluster stability") +
+  stat_summary(
+    aes(group = k_value),
+    color = "red",
+    # median and quartiles for point range
+    fun = "median",
+    fun.min = function(x) {
+      quantile(x, 0.25)
+    },
+    fun.max = function(x) {
+      quantile(x, 0.75)
+    }
+  )
 ```
@@
       forcats::fct_infreq() |>
-      forcats::fct_relevel("All remaining cell types", after = Inf)
+      forcats::fct_relevel("All remaining cell types", after = Inf),
+    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
   )
@@
 # get heatmap for each library showing cluster assignments vs. singler assignments
-names(sce_list) |>
+library_ids |>
   purrr::map(\(id){
@@
 # choose k=25 based on previous stats
-names(sce_list) |>
+library_ids |>
   purrr::map(\(id) {
@@
 - All other clusters with high mesenchymal gene expression appear to have a mix of chondrocytes, fibroblasts, and smooth muscle cells so we will just label these as "mesenchymal-like cells". 
+- Any clusters that do not have high expression of any marker gene lists are labeled as "unknown".
+This corresponds to cluster 9 and 12 in `SCPCL000824`.
 
@@
 
-```{r}
+```{r, fig.height=7, fig.width=7}
 # look at cell type assignments on the UMAP
@@
   facet_grid(cols = vars(library_id)) +
-  theme(aspect.ratio = 1) +
+  theme(
+    aspect.ratio = 1,
+    legend.position = "bottom"
+  ) +
   guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5))) +
@@
 
-```{r}
+```{r, fig.height=7, fig.width=7}
 # grab columns that contain marker gene sums
@@
 
-# create individual umaps and combine into one
-marker_gene_columns |>
-  purrr::map(\(column){
-    genelist_name <- stringr::str_remove(column, "_sum")
+# for each sample show a faceted umap by cell type
+# color is gene expression of all marker genes for that cell type
+library_ids |>
+  purrr::map(\(id){
+    plot_df <- cluster_umap_df |>
+      dplyr::filter(library_id == id) |>
+      tidyr::pivot_longer(
+        cols = all_of(marker_gene_columns),
+        names_to = "cell_type",
+        values_to = "gene_expression"
+      ) |>
+      dplyr::mutate(cell_type = stringr::str_remove(cell_type, "_sum"))
 
-    ggplot(cluster_umap_df, aes(x = UMAP1, y = UMAP2, color = !!sym(column))) +
-      geom_point(alpha = 0.5, size = 0.5) +
-      facet_grid(cols = vars(library_id)) +
+    ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
+      geom_point(alpha = 0.5, size = 0.2) +
+      facet_wrap(vars(cell_type)) +
       theme(aspect.ratio = 1) +
@@
       labs(
-        title = genelist_name
+        title = id
       )

### 2024-08-02 analyses/cell-type-ewings/exploratory_analysis/05-cluster-exploration.Rmd
add id to title and render

@@
       }
-    )
+    ) +
+    labs(title = id)
 }

### 2024-08-05 analyses/cell-type-ewings/exploratory_analysis/02-cellassign-tumor-cell-classification.Rmd

@@
                             cell_type_col) {
-
   binary_df <- marker_genes_df |>
-    dplyr::select({{cell_type_col}}, {{gene_col}}) |> 
+    dplyr::select({{ cell_type_col }}, {{ gene_col }}) |>
     unique() |>
     tidyr::pivot_wider(
-      id_cols = {{gene_col}},
-      names_from = {{cell_type_col}},
-      values_from = {{cell_type_col}},
+      id_cols = {{ gene_col }},
+      names_from = {{ cell_type_col }},
+      values_from = {{ cell_type_col }},
       values_fn = length,
@@
 # function to get assigned cell type based on max prediction returned by CellAssign
-get_celltype_assignment <- function(predictions){
+get_celltype_assignment <- function(predictions) {
   # get individual cell type assignments
@@
   gene_col = "ensembl_gene_id",
-                                  cell_type_col = "cell_type")
+  cell_type_col = "cell_type"
+)
 
@@
     celltype_df$marker_gene_classification,
-    celltype_df$cellassign_annotations)
+    celltype_df$cellassign_annotations
+  )
 )
@@
   gene_col = "ensembl_gene_id",
-                                   cell_type_col = "cell_type")
+  cell_type_col = "cell_type"
+)
 
@@
     celltype_df$caret_cellassign_annotations,
-    celltype_df$caret_visser_annotations)
+    celltype_df$caret_visser_annotations
+  )
 )
@@
   dplyr::bind_rows(tumor_markers_df)
-
 ```
@@
   gene_col = "ensembl_gene_id",
-                                          cell_type_col = "cell_type")
+  cell_type_col = "cell_type"
+)
 
@@
 
-panglao_tumor_ref_mtx <- panglao_tumor_ref_mtx[-which(rowsums > 1),]
+panglao_tumor_ref_mtx <- panglao_tumor_ref_mtx[-which(rowsums > 1), ]
 

### 2024-08-05 analyses/cell-type-ewings/template_notebooks/cnv-workflow/04-infercnv.Rmd

@@
 # png ref files
-png_file <-file.path(params$infercnv_dir, glue::glue("{params$library_id}_infercnv.png"))
+png_file <- file.path(params$infercnv_dir, glue::glue("{params$library_id}_infercnv.png"))
 

### 2024-08-05 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd

--- 2024-07-23 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
@@
         tibble::rownames_to_column(var = "barcodes") |>
-        dplyr::select(barcodes,
+        dplyr::select(
+          barcodes,
           ground_truth = ground_truth_doublets,

### 2024-08-05 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd

@@
   tibble::rownames_to_column(var = "barcodes") |>
-  dplyr::select(barcodes,
+  dplyr::select(
+    barcodes,
     ground_truth = ground_truth_doublets,

### 2024-08-05 analyses/cell-type-ewings/template_notebooks/auc-singler-workflow/03-clustering.Rmd
clustering template

---
title: "Clustering exploration for `r params$library_id`"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  library_id: "SCPCL000822"
  sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
  singler_results_file: "../../results/aucell_singler_annotation/SCPCS000490/SCPCL000822_singler-classifications.tsv"
  cluster_results_file: "../../results/aucell_singler_annotation/SCPCS000490/SCPCL000822_clusters.tsv" 
  marker_genes_file: "../../references/visser-all-marker-genes.tsv"
---

This notebook summarizes clustering for `r params$library_id`.

- Louvain clusters are calculated using a range of parameters specifying nearest neighbors (5, 10, 15, 20, 25, 30, 35, and 40). 
- Metrics are then calculated to evaluate clustering results across all parameters: 
  - Silhouette width 
  - Cluster purity 
  - Cluster stability 
- The clusters are then compared to the results from running `SingleR` in the `aucell-singler-annotation.sh` workflow.
- We then look at marker gene expression across the clusters and assign each cluster to a cell type.

## Setup

```{r}
# check that sce and results files exist
stopifnot(
  "sce file does not exist" = file.exists(params$sce_file),
  "SingleR results file does not exist" = file.exists(params$singler_results_file),
  "Marker genes file does not exist" = file.exists(params$marker_genes_file)
)
```

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)

# quiet messages
options(readr.show_col_types = FALSE)
ComplexHeatmap::ht_opt(message = FALSE)
```

```{r base paths}
# The path to this module
module_base <- rprojroot::find_root(rprojroot::is_renv_project)

# source in helper functions: calculate_sum_markers()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)

# source in clustering functions
clustering_functions <- file.path(module_base, "scripts", "utils", "clustering-functions.R")
source(clustering_functions)
```

```{r}
# read in input sce files
sce <- readr::read_rds(params$sce_file)

# read in singler results
singler_classification_df <- readr::read_tsv(params$singler_results_file)
```

## Clustering

Below we perform Louvain, Jaccard clustering, varying `k`. 
The minimum `k` is 5 and the maximum `k` is 40 with a step size of 5.

```{r}
# perform clustering across k = 5-40 with increments of 5
all_cluster_results <- cluster_sweep(sce) |>
  dplyr::mutate(
    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
  )
```

```{r}
# get umap embeddings and combine into a data frame with cluster assignments
umap_df <- sce |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  # replace UMAP.1 with UMAP1 and get rid of excess columns
  dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2) |>
  dplyr::left_join(all_cluster_results, by = "barcodes")
```

Below we visualize the cluster assignments using each parameter on a UMAP. 
This can be helpful to see any values of K that may have obvious over or under clustering.

```{r, fig.height=15, fig.width=10}
# look at clustering results for each library across all params
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cluster)) +
  geom_point(alpha = 0.5, size = 0.1) +
  facet_wrap(vars(nn_param)) +
  theme(
    aspect.ratio = 1
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))
```

## Clustering statistics

Below we calculate a series of statistics:

- Average silhouette width: This metric evaluates cluster separation. 
Cells with large positive silhouette widths are closer to other cells in the same cluster than to cells in different clusters. 
- Average cluster purity: This metric also evaluates cluster separation and tells us the proportion of neighboring cells that are assigned to the same cluster. 
Higher purity values indicate that clusters are well separated. 
- Cluster stability: This evaluates how stable the clustering is to input data. 
Higher values of cluster stability indicate more reproducible clusters.

```{r, warning=FALSE}
# get a combined stats dataframe with purity and width for all clusters
all_stats_df <- get_cluster_stats(sce, all_cluster_results) |>
  dplyr::mutate(
    # make sure the order is correct
    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
  )
```

### Silhouette width

```{r}
# silhouette width for different params
plot_cluster_stats(all_stats_df, width)
```

### Cluster purity

```{r}
# cluster purity for different params
plot_cluster_stats(all_stats_df, purity)
```

### Cluster stability

```{r}
# calculate cluster stability
stability_df <- get_cluster_stability(sce, all_cluster_results) |>
  dplyr::mutate(
    k_value = as.factor(k_value),
    # make sure that k = 5 comes first
    k_value = forcats::fct_relevel(k_value, "5", after = 0)
  )

# plot stability across all values of k
ggplot(stability_df, aes(x = k_value, y = ari)) +
  geom_jitter(width = 0.1) +
  labs(title = "Cluster stability") +
  stat_summary(
    aes(group = k_value),
    color = "red",
    # median and quartiles for point range
    fun = "median",
    fun.min = function(x) {
      quantile(x, 0.25)
    },
    fun.max = function(x) {
      quantile(x, 0.75)
    }
  )
```

## Compare clusters to cell types from `SingleR`

Now we will compare the clustering results to the cell type assignments obtained from `SingleR` in the `aucell-singler-annotation.sh` workflow. 
To compare results we will calculate the Jaccard similarity index between clusters and cell types. 
We expect that good clustering will line up with the cell type assignments so that there is ~ 1 cell type per cluster.

For the plots, we will only display the top 7 cell types and all other cells will be lumped together into `All remaining cell types`.

```{r}
cluster_classification_df <- all_cluster_results |>
  # add in classifications from singler
  dplyr::left_join(singler_classification_df, by = "barcodes") |>
  dplyr::mutate(
    # first grab anything that is tumor and label it tumor
    # NA should be unknown
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later
    singler_lumped = singler_annotation |>
      forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf),
    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
  )
```

```{r, fig.height=10, fig.width=7}
# get heatmap for showing cluster assignments vs. singler assignments
cluster_celltype_heatmap(cluster_classification_df)
```

## Marker gene expression across clusters

We will also look at the expression of marker genes across each cluster. 
In these plots, each row shows the distribution of the specified marker genes in that cluster. 
Each panel is labeled with the marker gene expression being plotted.

```{r}
# read in marker genes table
marker_genes_df <- readr::read_tsv(params$marker_genes_file, show_col_types = FALSE) |>
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |>
  dplyr::distinct()

# get list of all cell types found
cell_types <- unique(marker_genes_df$cell_type)

# get the sum of expression of all genes for each cell type
gene_exp_df <- cell_types |>
  purrr::map(\(type){
    calculate_sum_markers(marker_genes_df, sce, type)
  }) |>
  purrr::reduce(dplyr::inner_join, by = "barcodes")

# join sum expression columns with clustering results
cluster_exp_df <- all_cluster_results |>
  dplyr::left_join(gene_exp_df, by = "barcodes")
```

```{r, fig.height=10, fig.width=5}
# plot to look at marker gene expression for all cell types across all clusters and all params
k_params <- unique(cluster_exp_df$nn_param) |> as.character()

k_params |>
  purrr::map(\(k) {
    plot_marker_genes(cluster_exp_df,
      k_value = k
    )
  })
```

## Session info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-08-05 analyses/cell-type-ewings/exploratory_analysis/05-cluster-exploration.Rmd
spelling updates

@@
 ---
-title: "Clustering SCPCL000822 and SCPCL000824"
+title: "Clustering `SCPCL000822` and `SCPCL000824`"
 author: Ally Hawkins

### 2024-08-05 analyses/cell-type-ewings/template_notebooks/auc-singler-workflow/03-clustering.Rmd
add comments and document

@@
 params:
-  library_id: "SCPCL000822"
-  sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
-  singler_results_file: "../../results/aucell_singler_annotation/SCPCS000490/SCPCL000822_singler-classifications.tsv"
-  cluster_results_file: "../../results/aucell_singler_annotation/SCPCS000490/SCPCL000822_clusters.tsv" 
-  marker_genes_file: "../../references/visser-all-marker-genes.tsv"
+  library_id: NULL
+  sce_file: NULL
+  singler_results_file: NULL
+  cluster_results_file: NULL
+  marker_genes_file: NULL
 ---
@@
 
-- Louvain clusters are calculated using a range of parameters specifying nearest neighbors (5, 10, 15, 20, 25, 30, 35, and 40). 
+- Louvain, Jaccard clusters are calculated using a range of parameters specifying nearest neighbors (5, 10, 15, 20, 25, 30, 35, and 40). 
 - Metrics are then calculated to evaluate clustering results across all parameters: 
@@
 
-```{r}
+```{r, warning=FALSE, message=FALSE}
 # calculate cluster stability
@@
 
-For the plots, we will only display the top 7 cell types and all other cells will be lumped together into `All remaining cell types`.
+For the plots, we will only display the top 7 cell types identified by `SingleR` and all other cells will be lumped together into `All remaining cell types`.
 
@@
 
-We will also look at the expression of marker genes across each cluster. 
+Finally, we will look at the expression of marker genes across each cluster for each value of K. 
 In these plots, each row shows the distribution of the specified marker genes in that cluster. 
@@
 # plot to look at marker gene expression for all cell types across all clusters and all params
-k_params <- unique(cluster_exp_df$nn_param) |> as.character()
+k_params <- unique(cluster_exp_df$nn_param) |>
+  as.character()
 

### 2024-08-06 analyses/cell-type-ewings/template_notebooks/auc-singler-workflow/03-clustering.Rmd
Apply suggestions from code review
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 Below we visualize the cluster assignments using each parameter on a UMAP. 
-This can be helpful to see any values of K that may have obvious over or under clustering.
+This can be helpful to see any values of `k` that may have obvious over or under clustering.
 
@@
 Cells with large positive silhouette widths are closer to other cells in the same cluster than to cells in different clusters. 
+Higher values indicate tighter clusters.
 - Average cluster purity: This metric also evaluates cluster separation and tells us the proportion of neighboring cells that are assigned to the same cluster. 
@@
 
-Finally, we will look at the expression of marker genes across each cluster for each value of K. 
+Finally, we will look at the expression of marker genes across each cluster for each value of `k`. 
 In these plots, each row shows the distribution of the specified marker genes in that cluster. 
@@
 # read in marker genes table
-marker_genes_df <- readr::read_tsv(params$marker_genes_file, show_col_types = FALSE) |>
+marker_genes_df <- readr::read_tsv(params$marker_genes_file) |>
   # account for genes being from multiple sources

### 2024-08-06 analyses/cell-type-ewings/template_notebooks/clustering-workflow/01-clustering.Rmd
some comment updates and figure size change

--- 2024-08-06 analyses/cell-type-ewings/template_notebooks/auc-singler-workflow/03-clustering.Rmd
@@
   theme(
-    aspect.ratio = 1
-  ) +
-  guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))
+    aspect.ratio = 1,
+    legend.position = "none"
+  )
 ```
@@
 - Average cluster purity: This metric also evaluates cluster separation and tells us the proportion of neighboring cells that are assigned to the same cluster. 
-Higher purity values indicate that clusters are well separated. 
+Purity values range from 0-1 with higher purity values indicating clusters that are well separated. 
 - Cluster stability: This evaluates how stable the clustering is to input data. 
-Higher values of cluster stability indicate more reproducible clusters.
+Stability values range from 01- with higher values of cluster stability indicating more reproducible clusters.
 
@@
 
-```{r, fig.height=10, fig.width=5}
+```{r, fig.height=7, fig.width=7}
 # plot to look at marker gene expression for all cell types across all clusters and all params

### 2024-08-07 analyses/cell-type-wilms-tumor-06/01-clustering_SCPCS000168.Rmd
add 01-clustering abalysis

--- 2024-08-02 analyses/cell-type-wilms-tumor-06/notebook/01-clustering_SCPCS000169.Rmd
@@
 ---
-title: "Evaluation of clustering parameters - SCPCS000169"
+title: "Evaluation of clustering parameters - SCPCS000168"
 author: maud-p
@@
 
-In this notebook, we will set up parameters in the seurat workflow 
-[normalization --> reduction --> clustering] 
-for one Wilms tumor sample (SCPCS000169) of the Wilms Tumor dataset (SCPCP000006).
+In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000168) of the Wilms Tumor dataset (SCPCP000006).
 
@@
 library(Azimuth)     # For label transfer
+
+library(msigdbr)     # for gene set enrichment analysis or pathway enrichment
+library(clusterProfiler)
+
+library(data.table)
 ```
@@
 cfg$rate1_threshold = 0.5
+set.seed(12345)
 ```
@@
 #### Input and output files
+
+For this analysis, we worked with the _processed.rds data. 
+We builded a Seurat object based on the counts data and re-perform the analysis 
+[normalization --> reduction --> clustering] following the Seurat workflow.
+
+We transferred meta.data to keep:
+
+[x] QC data computed by the DataLab
+
+[x] annotation data computed by the DataLab
+
+[x] raw annotation and gene_symbol conversion
 
@@
 ## to be modified when rendered
-i = 2
+i = 1
 filedir <- filelist[i] 
@@
 ```
+
+Here we have a sample of anaplastic histology, untreated. Few thoughts:
+
+- We thus expect anaplasia-associated patterns (TP53, MYCN as an example) to be major driver here.
+
+- We do not expect high immune infiltration (induced by chemotherapy).
+
+- We do not expect chemotherapy induced DNA damages, but maybe some DNA-repair pathways up in anaplstic cells.
 
@@
 
-[3] We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. 
+[3] We look at specific marker genes that we reported in the table marker.sets/CellType_metadata.csv to check the relevance of the clustering.
+
+[4] We look at specific pathways that we reported in the table marker.sets/Pathways_metadata.csv to check the relevance of the clustering.
+
+[5] We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. 
 DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method.
 
-[4] We look at specific marker genes that we reported in the table marker.sets/CellType_metadata.csv to check the relevance of the clustering.
+[6] We perform enrichment analysis of marker genes for each seurat clusters. 
+We defined all the genes from the seurat object as the universe and used the MSigDB gene sets.
 
-[5] We plot pca/umap reduction grouping with available annotations (singler_, cellassign_). 
+[7] We plot pca/umap reduction grouping with available annotations (singler_, cellassign_). 
 We expect at least immune cells to be correctly label and fall into a few set of clusters.
 
-[6] We run label transfer (Azimuth) to transfer annotation from the fetal kidney atlas human reference. 
-We plot pca/umap reduction grouping with latest labels. We expect it to be the most representative of the cell types in the sample.
+[8] We run label transfer (Azimuth) to transfer annotation from the fetal kidney atlas human reference. 
+We plot pca/umap reduction grouping with latest labels. 
+We expect it to be the most representative of the cell types in the sample.
 
@@
 
-### Tested parameters
-
-These other parameters have been tested but not selected:
-
-- SCTransform (default parameters) --> RunPCA (default) --> RunUMAP (dims = 1:30) --> FindNeighbors (dims = 1:30) --> FindClusters (default)
-
-## Input and output files
-
-For this analysis, we worked with the _processed.rds data. 
-We builded a Seurat object based on the counts data and re-perform the analysis 
-[normalization --> reduction --> clustering] following the Seurat workflow.
-
-We transferred meta.data to keep:
-
-[x] QC data computed by the DataLab
-
-[x] annotation data computed by the DataLab
-
-[x] raw annotation and gene_symbol conversion
+Note: Other parameters have been previously tested, but we would like to show in the following report that the one selected is performing good.
 
@@
                                     assay = "RNA",
-                                    project = file)
+                                    project = sample
+                           )
 
@@
 
-```{r fig.width=8, fig.height=4, out.width='100%'}
+```{r fig.width=20, fig.height=5, out.width='100%'}
 
@@
 d1 <- SCpubr::do_DimPlot(srat, reduction="pca", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - pca")
+v <- SCpubr::do_ViolinPlot(srat, features = c( "subsets_mito_percent"), ncol = 1, group.by = "seurat_clusters", legend.position = "none")
 
-d1|d2
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(2,2,4))
 
-```
-
-The umap clustering is expected for a triphasic Wilms tumor sample. The 3 main clusters might be:
-
-- blastema cancer cells
-- epithelium cancer and/or normal cells
-- stroma cancer and/or normal cells
-
-The 2 small clusters C11 and C13 might be immune and endothelial cells.
-
-To be checked in the following analysis!
-
-### Quick check of the QC values
-```{r qc, fig.width=10, fig.height=2.5, out.width='100%'}
- SCpubr::do_ViolinPlot(srat, features = c("subsets_mito_percent"), ncol = 3, group.by = "seurat_clusters", legend.position = "none") 
 ```
@@
 
-d1|d2
+b <- SCpubr::do_BarPlot(sample = srat,
+                         group.by = "Phase",
+                         split.by = "seurat_clusters",
+                         position = "fill",
+                         font.size = 10,
+                         legend.ncol = 3) +
+                         ggtitle("% cells")+
+                         xlab(file)
+d1 + d2 + b + plot_layout(ncol = 3, widths = c(2,2,4))
 
@@
 ```
-
-Cluster 3 might be composed of cycling cells, in higher proportion than the other clusters.
 
@@
 
+```{r fig.width=25, fig.height=5, out.width='100%'}
+ SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("malignant")])] , ncol = 9, group.by = "seurat_clusters", legend.position = "none")
+```
+All clustersexcept C13 are IGF2+ (ENSG00000167244).
+
 ```{r fig.width=20, fig.height=5, out.width='100%'}
- SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("malignant")])] , ncol = 6, group.by = "seurat_clusters", legend.position = "none")
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000167244", pt.size = 0.2) + ggtitle("IGF2")
+v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000167244", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
 ```
-The set of clusters 0-1-3-4-6-10-12-14 is NCAM1+ (ENSG00000149294) which is a marker of blastema area.
-C2 is positive for COL6A3 which is a marker of stroma Wilms tumor cells (also normal stroma)
+The set of clusters 0-1-4-8-9-11-12 is NCAM1+ (ENSG00000149294), which is a blastemal marker.
 
-```{r fig.width=20, fig.height=15, out.width='100%'}
- SCpubr::do_FeaturePlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("malignant")])])
+```{r fig.width=20, fig.height=5, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000149294",  pt.size = 0.2) + ggtitle("NCAM1")
+v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000149294", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```
+
+All clusters except of C13 are WT1+ (ENSG00000184937).
+
+```{r fig.width=20, fig.height=5, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000184937", pt.size = 0.2) + ggtitle("WT1")
+v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000184937", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```
+
+C10 is positive for COL6A3 which is a marker of stroma Wilms tumor cells (also normal stroma)
+
+```{r fig.width=20, fig.height=5, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000163359", pt.size = 0.2) + ggtitle("COL6A3")
+v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000163359", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+
 ```
@@
 
-```{r fig.width=20, fig.height=15, out.width='100%'}
- SCpubr::do_FeaturePlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("immune")])])
-```
-C11 is PTPRC+ (ENSG00000081237) and must be a cluster of immune cells.
+C13 is PTPRC+ (ENSG00000081237) and must be a cluster of immune cells.
 Wilms tumor are generally cold tumor, especially untreated Wilms tumor.
@@
 The size of the cluster would therefore make sense.
+
+```{r fig.width=20, fig.height=5, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000081237", pt.size = 0.2) + ggtitle("PTPRC = CD45")
+v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000081237", ncol = 1, group.by = "seurat_clusters", legend.position = "none", pt.size = 2)
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```
 
@@
 
-```{r fig.width=15, fig.height=5, out.width='100%'}
- SCpubr::do_FeaturePlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[! CellType_metadata$cell_class %in% c("immune", "malignant")])], ncol = 3)
+The set of clusters 8-11 is EPCAM+ (ENSG00000119888), which is a marker of epithelial (malignant and normal) kidney cells.
+
+```{r fig.width=20, fig.height=5, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000119888", pt.size = 0.2) + ggtitle("EPCAM")
+v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000119888", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
 ```
 
-C13 is VWF+ (ENSG00000110799) and must be a cluster of endothelial cells.
+### Look at specific pathways
+
+Here, we open the table of marker genes marker-sets/Pathway_metadata.csv.
+
+```{r marker_pathways, fig.width=12, fig.height=6, out.width='100%'}
+# open the set of reference marker genes
+
+Pathway_metadata <- read.csv("marker-sets/Pathway_metadata.csv", header = TRUE, sep = ",")
+
+DT::datatable(Pathway_metadata, caption = ("CellType_metadata"), 
+              extensions = 'Buttons', 
+              options = list(  dom = 'Bfrtip',
+
+                               buttons = c( 'csv', 'excel')))
+```
+#### TP53 pathway
+
+here we will calculate a TP53 score using AddMduleScore and the genes of the HALLMARK_P53_PATHWAY gene set.
+
+```{r fig.width=10, fig.height=10, out.width='100%'}
+
+## define genesets
+hallmarks <- msigdbr(species = "human", category = "H")
+
+TP53_list = hallmarks %>% 
+  filter(gs_name == "HALLMARK_P53_PATHWAY") %>%
+  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()
+
+TP53_list <- list(TP53_list$human_ensembl_gene)
+
+srat <- AddModuleScore(srat, features = TP53_list, name = "TP53_score")
+```
+
+```{r fig.width=20, fig.height=5, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "TP53_score1", pt.size = 0.2) + ggtitle("TP53 score")
+v <- SCpubr::do_ViolinPlot(srat, features = "TP53_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```
+"Normal" cells (immune, C13) have a slightly higher TP53 score.
+
+#### DNA repair pathway
+
+here we will calculate a DNA_repair score using AddMduleScore and the genes of the HALLMARK_DNA_REPAIR gene set.
+
+```{r fig.width=10, fig.height=10, out.width='100%'}
+
+DNA_repair = hallmarks %>% 
+  filter(gs_name == "HALLMARK_DNA_REPAIR") %>%
+  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()
+
+DNA_repair_list <- list(DNA_repair$human_ensembl_gene)
+
+srat <- AddModuleScore(srat, features = DNA_repair_list, name = "DNA_repair_score")
+```
+
+```{r fig.width=20, fig.height=5, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "DNA_repair_score1", pt.size = 0.2) + ggtitle("DNA_repair score")
+v <- SCpubr::do_ViolinPlot(srat, features = "DNA_repair_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```
+I am not sure how relevant the DNA repair score is. To be tested on more samples.
+
+#### DROSHA target genes
+
+```{r fig.width=10, fig.height=10, out.width='100%'}
+
+## define genesets
+c3 <- msigdbr(species = "human", category = "C3", subcategory = "TFT:GTRD")
+
+DROSHA_list = c3 %>% 
+  filter(gs_name == "DROSHA_TARGET_GENES") %>%
+  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()
+
+DROSHA_list <- list(DROSHA_list$human_ensembl_gene)
+
+srat <- AddModuleScore(srat, features = DROSHA_list, name = "DROSHA_score")
+```
+
+```{r fig.width=20, fig.height=5, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "DROSHA_score1", pt.size = 0.2) + ggtitle("DROSHA score")
+v <- SCpubr::do_ViolinPlot(srat, features = "DROSHA_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```
+
+#### DICER1 target genes
+
+```{r fig.width=10, fig.height=10, out.width='100%'}
+DICER1_list = c3 %>% 
+  filter(gs_name == "DICER1_TARGET_GENES") %>%
+  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()
+
+DICER1_list <- list(DICER1_list$human_ensembl_gene)
+
+srat <- AddModuleScore(srat, features = DICER1_list, name = "DICER1_score")
+```
+
+```{r fig.width=20, fig.height=5, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "DICER1_score1", pt.size = 0.2) + ggtitle("DICER1 score")
+v <- SCpubr::do_ViolinPlot(srat, features = "DICER1_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```
+Not sure what to do with these scores.
 
@@
 
+# Select top 5 genes for heatmap plotting
 s.markers <- na.omit(s.markers)
@@
 ss <- subset(srat, cells = cells)
-ss <- ScaleData(ss, c( top5$feature))
+ss <- ScaleData(ss, features = top5$feature)
 
-p1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")
+p1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
 p2 <- DoHeatmap(ss, features = top5$feature,  cells = cells, group.by = "seurat_clusters") + NoLegend() + 
@@
 p3 <- ggplot(srat@meta.data, aes(seurat_clusters, fill = seurat_clusters)) + geom_bar() + NoLegend()
-p4 <- VlnPlot(srat, features = c("nFeature_RNA", "nCount_RNA", "subsets_mito_percent"), ncol = 3)
 
 common_title <- sprintf("Unsupervised clustering %s, %d cells", srat@meta.data$orig.ident[1], ncol(srat))
-show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) / p4) + plot_layout(widths = c(1, 2)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))
-DT::datatable(top5[,c(1,9,11,12)], caption = ("top 5 marker genes"), 
+show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) ) + plot_layout(widths = c(1, 2)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))
+DT::datatable(top5[, c(1, 9, 11, 12)], caption = ("top 5 marker genes"), 
               extensions = 'Buttons', 
@@
 
-UNC5D (C1, dependency receptor), has been described in cancer and shown to regulate p53-dependent apoptosis (https://pubmed.ncbi.nlm.nih.gov/24691657/). 
-We know that especially in blastema cell, p53 signaling play an important role.
+PTPRC (C13, CD45), define the immune cluster.
 
-PDCH7 (C7, protocadherin 7) is expressed in human kidney renal tubule (normal kidney epithelium, see https://www.proteinatlas.org/ENSG00000169851-PCDH7/tissue/kidney) and has been associated with oncogenic function in lung cancer (https://pubmed.ncbi.nlm.nih.gov/30409919/, https://www.frontiersin.org/journals/pharmacology/articles/10.3389/fphar.2023.1217213/full).
+### Enrichment analysis of marker genes
 
-EBF1 (C8, Early B Cell factor 1) has been shown to play a role in kidney development, esp. glomerular, while expressed in stroma progenitor (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6727263/).
+Here we perform enrichment analysis of the marker genes found in the previous section for each Seurat cluster.
 
-ROR1 (C8, Receptor tyrosine kinase-like orphan receptor 1) ROR1 is only detectable in embryonic tissue and generally absent in mature tissue.
-It has been associated with cancer (https://academic.oup.com/proteincell/article/5/7/496/6831810) and cancer stroma (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5367847/).
+We defined as universe/background all the genes expressed in the dataset, meaning the rownames of the Seurat object.
 
-ERBB4 (C9, EGF receptor family) is expressed during kidney development of epithelial tubule and uteric bud (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3269924/)
+We used three gene sets from MSiGDB:
 
-FLT1 (C13, vascular endothelial growth factor receptor 1), PCAM and ELMO1 in the top5 of C13 marker genes confirm the identification of endothelial cells (https://www.ahajournals.org/doi/pdf/10.1161/CIRCRESAHA.109.213983, https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4986701/)
+- the hallmark gene sets  are coherently expressed signatures derived by aggregating many MSigDB gene sets to represent well-defined biological states or processes.
+- the C3 : regulatory target gene sets  based on gene target predictions for microRNA seed sequences and predicted transcription factor binding sites.
+- the C8 : cell type signature gene sets  curated from cluster markers identified in single-cell sequencing studies of human tissue.
+
+We used enricher function from clusterProfiler to perform enrichment analysis.
+
+```{r fig.width=10, fig.height=10, out.width='100%'}
+# define background genes = universe for enrichment
+background <-  row_metadata$gene_ids
+
+## define genesets
+c8 <- msigdbr(species = "human", category = "C8")
+
+msigdbr_hallmarks = hallmarks %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
+msigdbr_c3 = c3 %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
+msigdbr_c8 = c8 %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
+```
+
+#### Hallmarks MSiGDB gene sets
+
+```{r fig.width=25, fig.height=6, out.width='100%'}
+tmp_H <- NULL
+for(i in unique(srat$seurat_clusters)){
+signature = s.markers$feature[s.markers$group1 == i]
+ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_hallmarks)
+tmp <- NULL
+
+  if(!is.null(ego_module)){
+    if(dim(ego_module)[1]>0){
+    b_H <- barplot(ego_module, showCategory = 15, font.size = 20, label_format = 40)  
+    tmp <- b_H$data
+    tmp$set <- "Hallmarks"
+    tmp$cluster <- i
+    }
+  }
+tmp_H <- rbind(tmp_H, tmp)
+}
+
+ggplot(tmp_H, aes(Count, ID)) +
+    geom_bar(stat = "identity", aes(fill=cluster))+
+
+  theme_classic()+ 
+    geom_text(
+        aes(label = (paste( "Padj = ", round(p.adjust,2)))),
+        color = "black",
+        size = 3,
+        hjust=1,
+        position = position_dodge(0.5)
+    )+
+   theme(text = element_text(size=14))+
+  facet_wrap(facets=c("cluster"), ncol = length(unique(tmp_H$cluster)))
+```
+
+MYC enrichment is expected in anaplastic samples.
+
+#### C3 MSiGDB regulatory target gene sets
+
+```{r fig.width=25, fig.height=6, out.width='100%'}
+
+tmp_H <- NULL
+for(i in unique(srat$seurat_clusters)){
+signature = s.markers$feature[s.markers$group1 == i]
+ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_c3)
+tmp <- NULL
+
+  if(!is.null(ego_module)){
+    if(dim(ego_module)[1]>0){
+    b_H <- barplot(ego_module, showCategory = 15, font.size = 20, label_format = 40)  
+    tmp <- b_H$data
+    tmp$ID <- factor(tmp$ID, levels = c(tmp$ID[order(tmp$Count)]))
+    tmp$set <- "c8"
+    tmp$cluster <- i
+    }
+  }
+tmp_H <- rbind(tmp_H, tmp[1:5,])
+}
+
+ggplot(tmp_H, aes(Count, ID)) +
+    geom_bar(stat = "identity", aes(fill=cluster))+
+
+  theme_classic()+ 
+    geom_text(
+        aes(label = (paste( "Padj = ", round(p.adjust,2)))),
+        color = "black",
+        size = 3,
+        hjust=1,
+        position = position_dodge(0.5)
+    )+
+   theme(text = element_text(size=14))+
+  facet_wrap(facets=c("cluster"), ncol = length(unique(tmp_H$cluster)))
+```
+
+#### C8 MSiGDB cell type signature gene sets
+
+```{r fig.width=30, fig.height=15, out.width='100%'}
+
+tmp_H <- NULL
+for(i in unique(srat$seurat_clusters)){
+signature = s.markers$feature[s.markers$group1 == i]
+ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_c8)
+tmp <- NULL
+
+  if(!is.null(ego_module)){
+    if(dim(ego_module)[1]>0){
+    b_H <- barplot(ego_module, showCategory = 15, font.size = 20, label_format = 40)  
+    tmp <- b_H$data
+    tmp$ID <- factor(tmp$ID, levels = c(tmp$ID[order(tmp$Count)]))
+    tmp$set <- "c8"
+    tmp$cluster <- i
+    }
+  }
+tmp_H <- rbind(tmp_H, tmp[1:5,])
+}
+
+ggplot(tmp_H, aes(Count, ID)) +
+    geom_bar(stat = "identity", aes(fill=cluster))+
+
+  theme_classic()+ 
+    geom_text(
+        aes(label = (paste( "Padj = ", round(p.adjust,2)))),
+        color = "black",
+        size = 3,
+        hjust=1,
+        position = position_dodge(0.5)
+    )+
+   theme(text = element_text(size=14))+
+  facet_wrap(facets=c("cluster"), ncol = length(unique(tmp_H$cluster)))
+```
+
+- C10 is enriched in adulte differentiated cell types/pathway of the kidney.
+This could suggest that C10 are from normal kidney. 
+To be checked with inferCNV profile.
+
+- C2, C5are enriched in muscle pathways, suggesting a stromal origin.
 
@@
 
-Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, it confirm once again that C13 is a cluster of endothelial cells and C11 a cluster of immune cells.
+Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, it give the first impression that C13 a cluster of immune cells.
 
@@
 In S3 like results as it is a quite big object or in the folder marker-sets?
-The reference in now in the module folder, marker-sets
+
 For more information related to the reference, please go to https://www.kidneycellatlas.org/
-
 You will find:
@@
 Also check : https://www.science.org/doi/10.1126/science.aat5031
+
+This part is imho one of the most important step that allow us to have a quick and reliable idea of the composition of the different clusters. 
+The predicted compartment are defined into 4 categories:
+
+- endotheliúm
+- stroma
+- fetal nephron 
+- immune
+
+As for SingleR and CellAssign, the annotation of immune cells and endothelial cells is straightforward. 
+The stroma compartment should then contain normal and cancer stromal cells.
+The fetal nephron compartment contain blastema cancer cells a well as normal and cancer epithelial cells.
+
+Further segregation of cancer versus normal cells will be achieved using a combination of markers/pathways (see above) and inferred CNV (to be done).
 
@@
 # Preprocess with SCTransform
-srat <-  RunAzimuth(srat, ref_dir, assay = 'RNA')
+# Note, the RunAzimuth change a bit the Seurat object, changing for example the rownames to Symbol. 
+# We thus choose to rename the object to keep the 
+data <-  RunAzimuth(srat, ref_dir, assay = 'RNA')
 rm(reference)
@@
 
-DT::datatable(table(srat$seurat_clusters, srat$predicted.compartment), caption = ("table of azimuth compartment annotation per seurat clusters"),               extensions = 'Buttons', 
+DT::datatable(table(data$seurat_clusters, data$predicted.compartment), caption = ("table of azimuth compartment annotation per seurat clusters"),               extensions = 'Buttons', 
               options = list(  dom = 'Bfrtip',
@@
 
-DT::datatable(table(srat$seurat_clusters, srat$predicted.celltype), caption = ("table of azimuth annotation per seurat clusters"), 
+DT::datatable(table(data$seurat_clusters, data$predicted.celltype), caption = ("table of azimuth annotation per seurat clusters"), 
               extensions = 'Buttons', 
@@
 
-d1 <- DimPlot(srat, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + NoLegend()
-d2 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
-d3 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "predicted.celltype", label = TRUE, repel = TRUE) + NoLegend()
+d1 <- DimPlot(data, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + NoLegend()
+d2 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
+d3 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.celltype", label = TRUE, repel = TRUE) + NoLegend()
 
@@
 
-Looking at the predicted cell types, we observed that the set of cluster [0,1,3,4,6,10,12,14] is labeled as cap mesenchyme, an early developmental stage of nephron, while the set of clusters [5,7,9] is labeled with more differentiated epithelial units of the kidney.
-Based on this observation and the expression of NCAM1, UNC5D (C1) and PDCH7 (C7), ERBB4 (C9), we can conclude that the set of cluster [0,1,3,4,6,10,12,14] is mostly blastema cancer cells and the set of clusters [5,7,9] is eipthelial cancer or normal cells.
+Looking at the predicted cell types, we observed that most of the clusters are labeled as cap mesenchyme, an early developmental stage of nephron.
+This suggest that these clusters are either blastema or primitive epithelium.
 
@@
 
-f1 <- SCpubr::do_BarPlot(sample = srat,
+f1 <- SCpubr::do_BarPlot(sample = data,
                          group.by = "predicted.compartment",
@@
 
-f2 <- SCpubr::do_BarPlot(sample = srat,
+f2 <- SCpubr::do_BarPlot(sample = data,
                          group.by = "predicted.celltype",
@@
 
-- C13 = endothelium (with high confidency)
-- C11 = immune cells (with high confidency)
+- C13 = immune cells (with high confidency)
 
-- C2 and C8, that form a unit in the umap reduction, are stromal cells, either cancer or normal (with high confidency)
-- C[0-1-3-4-6-10-12-14] that form a unit in the umap reduction, are label as cap mesenchyme (<=> stem) and NCAM+ must be blastema cancer cells.
-- C[5-7-9] that form a unit in the umap reduction, are label as epithelial cells (SingleR) and differentiated kidney epithelium subunit (Fetal kidney atlas) must be epithelial (cancer or normal)
+- C10, is composed of stroma (normal or cancer) cells and endothelium cells. Might be a cluster of normal cells, a bit mixed. To be checked with inferCNV.
 
-The primary aim of this script was to assess the quality of clustering and play with clustering parameters. 
-On this sample, we are happy with the clustering, as we could identify:
-- one cluster of immune cells (cell marker PTPRC, SingleR, Azimuth label transfer fro the fetal kidney) 
-- one cluster of endothelial cells (cell marker VWF, SingleR, Azimuth label transfer fro the fetal kidney)
-- one set of clusters of stromal cells
-- one set of cluster of epithelial cells
-- one set of cluster of blastema cancer cells.
+- the two other set of clusters are cancer cells, with positivity for some Wilms tumor markers such as NCAM, WT1, EPCAM.
 

### 2024-08-07 analyses/cell-type-wilms-tumor-06/01-clustering_SCPCS000169.Rmd
add 01-clustering abalysis

--- 2024-08-07 analyses/cell-type-wilms-tumor-06/01-clustering_SCPCS000168.Rmd
@@
 ---
-title: "Evaluation of clustering parameters - SCPCS000168"
+title: "Evaluation of clustering parameters - SCPCS000169"
 author: maud-p
@@
 
-In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000168) of the Wilms Tumor dataset (SCPCP000006).
+In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000169) of the Wilms Tumor dataset (SCPCP000006).
 
@@
 ## to be modified when rendered
-i = 1
+i = 2
 filedir <- filelist[i] 
@@
 
-Here we have a sample of anaplastic histology, untreated. Few thoughts:
+Here we have a sample of favorable histology, untreated. Few thoughs:
 
-- We thus expect anaplasia-associated patterns (TP53, MYCN as an example) to be major driver here.
+- We thus do not expect anaplasia-associated patterns (TP53 as an example) to be major driver here.
 
@@
 
-- We do not expect chemotherapy induced DNA damages, but maybe some DNA-repair pathways up in anaplstic cells.
+- We do not expect chemotherapy induced DNA damages.
 
@@
 
+The umap clustering is expected for a triphasic Wilms tumor sample. The 3 main clusters might be:
+
+- blastema cancer cells
+- epithelium cancer and/or normal cells
+- stroma cancer and/or normal cells
+
+The two small clusters C11 and C12 might be immune and endothelial cells.
+
+To be checked in the following analysis!
+
 ### Cell cycle information
@@
 
+C1 might be composed of cycling cells, in higher proportion than the other clusters.
+
 ### Look at specific genes
@@
 ```
-All clustersexcept C13 are IGF2+ (ENSG00000167244).
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000167244", pt.size = 0.2) + ggtitle("IGF2")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000167244", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
-```
-The set of clusters 0-1-4-8-9-11-12 is NCAM1+ (ENSG00000149294), which is a blastemal marker.
+The set of clusters 0-1-3-4-7-9-10-13 is NCAM1+ (ENSG00000149294) which is a marker of blastema area.
 
@@
 
-All clusters except of C13 are WT1+ (ENSG00000184937).
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000184937", pt.size = 0.2) + ggtitle("WT1")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000184937", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
-```
-
-C10 is positive for COL6A3 which is a marker of stroma Wilms tumor cells (also normal stroma)
+C2-7 is positive for COL6A3 which is a marker of stroma Wilms tumor cells (also normal stroma)
 
@@
 
-C13 is PTPRC+ (ENSG00000081237) and must be a cluster of immune cells.
+C11 is PTPRC+ (ENSG00000081237) and must be a cluster of immune cells.
 Wilms tumor are generally cold tumor, especially untreated Wilms tumor.
@@
 
-The set of clusters 8-11 is EPCAM+ (ENSG00000119888), which is a marker of epithelial (malignant and normal) kidney cells.
+C12 is VWF+ (ENSG00000110799) and must be a cluster of endothelial cells.
 
@@
 d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000119888", pt.size = 0.2) + ggtitle("EPCAM")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000119888", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000110799", pt.size = 0.2) + ggtitle("VWF")
+v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000110799", ncol = 1, group.by = "seurat_clusters", legend.position = "none", pt.size = 2)
 
@@
 ```
-"Normal" cells (immune, C13) have a slightly higher TP53 score.
+"Normal" cells (immune, endothelial) have a slightly higher TP53 score.
+I am however not sure how relevant the TP53 score is for Favorable sample as the one we are analysing here.
+To be tested on an anaplastic sample.
 
@@
 ```
-I am not sure how relevant the DNA repair score is. To be tested on more samples.
+I am not sure how relevant the DNA repair score is for Favorable sample as the one we are analysing here. To be tested on an anaplastic sample.
 
@@
 ```
-Not sure what to do with these scores.
+
+C10 might have a slightly altered DROSHA/DICER1 ratio compared to the other cluster, but this might be due to the fact that C10 is mostly in G2M phase (?).
 
@@
 
-PTPRC (C13, CD45), define the immune cluster.
+UNC5D (C0, dependency receptor), has been described in cancer and shown to regulate p53-dependent apoptosis (https://pubmed.ncbi.nlm.nih.gov/24691657/).
+We know that especially in blastema cell, p53 signaling play an important role.
+
+PDCH7 (C6, protocadherin 7) is expressed in human kidney renal tubule (normal kidney epithelium, see https://www.proteinatlas.org/ENSG00000169851-PCDH7/tissue/kidney) and has been associated with oncogenic function in lung cancer (https://pubmed.ncbi.nlm.nih.gov/30409919/, https://www.frontiersin.org/journals/pharmacology/articles/10.3389/fphar.2023.1217213/full).
+
+EBF1 (C7, Early B Cell factor 1) has been shown to play a role in kidney development, esp. glomerular, while expressed in stroma progenitor (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6727263/).
+
+ERBB4 (C8, EGF receptor family) is expressed during kidney development of epithelial tubule and uteric bud (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3269924/)
+
+FLT1 (C12, vascular endothelial growth factor receptor 1), PCAM and ELMO1 in the top5 of C13 marker genes confirm the identification of endothelial cells (https://www.ahajournals.org/doi/pdf/10.1161/CIRCRESAHA.109.213983, https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4986701/)
 
@@
 
-```{r fig.width=25, fig.height=6, out.width='100%'}
+```{r fig.width=25, fig.height=3, out.width='100%'}
 tmp_H <- NULL
@@
 
-MYC enrichment is expected in anaplastic samples.
+- C2 and C13 are enriched for EMT signature (Warning count = 3) which makes sense with the hypothesis that they are stroma cluster.
 
@@
 
+- C5, which is though to be more epithelial, has a specific enrichment pattern of PRMD4, F10 and BCL6B targets.
+Might be usefull to look more into detail in the lit. if this pattern comes more often accross other samples.
+
+- C13 has a specific enrichement of HMGB2 targets, which is though to be associated with "stemness". 
+Would makes sense with the hypothesis tht C13 is blastema.
+
 #### C8 MSiGDB cell type signature gene sets
@@
 
-- C10 is enriched in adulte differentiated cell types/pathway of the kidney.
-This could suggest that C10 are from normal kidney. 
+- C0 is enriched in fetal metanephric cells, which makes sense with the hypothesis that C0 is blastema.
+
+- C2, C7 is enriched in adult kidney mesenchymal cells (interstitial, muscle and mesenglial).
+As the associated pathways are from "adult" reference, this might suggests that the cells are from normal kidney. 
 To be checked with inferCNV profile.
 
-- C2, C5are enriched in muscle pathways, suggesting a stromal origin.
+- C5, C6, C8 is enriched in adulte differentiated cell types/pathway of the kidney.
+This could suggest that C5 are from normal kidney. 
+To be checked with inferCNV profile.
+
+- C12 is enriched in endothelial gene sets, which fits whith the hypothesis that C12 are endothelial cells.
 
@@
 
-Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, it give the first impression that C13 a cluster of immune cells.
+Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, it give the first impression that C12 is a cluster of endothelial cells and C11 a cluster of immune cells.
 
@@
 # Preprocess with SCTransform
-# Note, the RunAzimuth change a bit the Seurat object, changing for example the rownames to Symbol. 
-# We thus choose to rename the object to keep the 
-data <-  RunAzimuth(srat, ref_dir, assay = 'RNA')
+srat <-  RunAzimuth(srat, ref_dir, assay = 'RNA')
 rm(reference)
@@
 
-DT::datatable(table(data$seurat_clusters, data$predicted.compartment), caption = ("table of azimuth compartment annotation per seurat clusters"),               extensions = 'Buttons', 
+DT::datatable(table(srat$seurat_clusters, srat$predicted.compartment), caption = ("table of azimuth compartment annotation per seurat clusters"),               extensions = 'Buttons', 
               options = list(  dom = 'Bfrtip',
@@
 
-DT::datatable(table(data$seurat_clusters, data$predicted.celltype), caption = ("table of azimuth annotation per seurat clusters"), 
+DT::datatable(table(srat$seurat_clusters, srat$predicted.celltype), caption = ("table of azimuth annotation per seurat clusters"), 
               extensions = 'Buttons', 
@@
 
-d1 <- DimPlot(data, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + NoLegend()
-d2 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
-d3 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.celltype", label = TRUE, repel = TRUE) + NoLegend()
+d1 <- DimPlot(srat, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + NoLegend()
+d2 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
+d3 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "predicted.celltype", label = TRUE, repel = TRUE) + NoLegend()
 
@@
 
-Looking at the predicted cell types, we observed that most of the clusters are labeled as cap mesenchyme, an early developmental stage of nephron.
-This suggest that these clusters are either blastema or primitive epithelium.
+Looking at the predicted cell types, we observed that the set of cluster [0,1,3,4,9,10,14] is labeled as cap mesenchyme, an early developmental stage of nephron, while the set of clusters [2,7,13] is labeled with more differentiated epithelial units of the kidney.
+This suggest that:
+
+- clusters [0,1,3,4,9,10,14] are more stem, either primitive epithelium or blastem cancer cells. 
+- clusters [2,7,13] are more differentiated, either epithelial cancer cells or normal cells.
 
@@
 
-f1 <- SCpubr::do_BarPlot(sample = data,
+f1 <- SCpubr::do_BarPlot(sample = srat,
                          group.by = "predicted.compartment",
@@
 
-f2 <- SCpubr::do_BarPlot(sample = data,
+f2 <- SCpubr::do_BarPlot(sample = srat,
                          group.by = "predicted.celltype",
@@
 
-- C13 = immune cells (with high confidency)
+- C12 = endothelium (with high confidency)
+- C11 = immune cells (with high confidency)
 
-- C10, is composed of stroma (normal or cancer) cells and endothelium cells. Might be a cluster of normal cells, a bit mixed. To be checked with inferCNV.
+- C2. C7, C13, that form a unit in the umap reduction, are stromal cells, either cancer or normal (with high confidency)
+- C[0-1-3-4-9-10-14] that form a unit in the umap reduction, are label as cap mesenchyme (<=> stem) and NCAM+ must be blastema cancer cells.
+- C[5-6-8] that form a unit in the umap reduction, are label as epithelial cells (SingleR) and differentiated kidney epithelium subunit (Fetal kidney atlas) must be epithelial (cancer or normal)
 
-- the two other set of clusters are cancer cells, with positivity for some Wilms tumor markers such as NCAM, WT1, EPCAM.
+The primary aim of this script was to assess the quality of clustering and play with clustering parameters. On this sample, we are happy with the clustering, as we could identify:
+
+- one cluster of immune cells (cell marker PTPRC, SingleR, Azimuth label transfer fro the fetal kidney) 
+- one cluster of endothelial cells (cell marker VWF, SingleR, Azimuth label transfer fro the fetal kidney)
+- one set of clusters of stromal cells
+- one set of cluster of epithelial cells
+- one set of cluster of blastema cancer cells.
 

### 2024-08-07 analyses/cell-type-wilms-tumor-06/01-clustering_SCPCS000170.Rmd
add 01-clustering abalysis

--- 2024-08-07 analyses/cell-type-wilms-tumor-06/01-clustering_SCPCS000168.Rmd
@@
 ---
-title: "Evaluation of clustering parameters - SCPCS000168"
+title: "Evaluation of clustering parameters - SCPCS000170"
 author: maud-p
@@
 
-In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000168) of the Wilms Tumor dataset (SCPCP000006).
+In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000170) of the Wilms Tumor dataset (SCPCP000006).
 
@@
 ## to be modified when rendered
-i = 1
+i = 3
 filedir <- filelist[i] 
@@
 
-Here we have a sample of anaplastic histology, untreated. Few thoughts:
+Here we have a sample of favorable histology, untreated. Few thoughts:
 
-- We thus expect anaplasia-associated patterns (TP53, MYCN as an example) to be major driver here.
+- We thus do not expect anaplasia-associated patterns (TP53 as an example) to be major driver here.
 
@@
 # Normalization
-options(future.globals.maxSize= 891289600)
+options(future.globals.maxSize= 891289600000)
 srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)
@@
 ```
+
+C9 and C13 might be a specific subtype, immune, endothelium, healthy cells (?)
 
@@
 ```
-All clustersexcept C13 are IGF2+ (ENSG00000167244).
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000167244", pt.size = 0.2) + ggtitle("IGF2")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000167244", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
-```
-The set of clusters 0-1-4-8-9-11-12 is NCAM1+ (ENSG00000149294), which is a blastemal marker.
+Almost all clusters are NCAM1+ (ENSG00000149294) which is a marker of blastema area.
 
@@
 
-All clusters except of C13 are WT1+ (ENSG00000184937).
+C9 is positive for COL6A3 which is a marker of stroma Wilms tumor cells (also normal stroma)
+
+```{r fig.width=20, fig.height=5, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000163359", pt.size = 0.2) + ggtitle("COL6A3")
+v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000163359", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+
+```
+
+Almost all clusters are WT1+ (ENSG00000184937).
 
@@
 
-C10 is positive for COL6A3 which is a marker of stroma Wilms tumor cells (also normal stroma)
+Almost all clusters are IGF2+ (ENSG00000167244).
 
@@
 d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000163359", pt.size = 0.2) + ggtitle("COL6A3")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000163359", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000167244", pt.size = 0.2) + ggtitle("IGF2")
+v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000167244", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
 
 d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
-
 ```
@@
 
-The set of clusters 8-11 is EPCAM+ (ENSG00000119888), which is a marker of epithelial (malignant and normal) kidney cells.
+The set of clusters [0-2-3-4-5-6-11-12] is EPCAM+ (ENSG00000119888), which is a marker of epithelial (malignant and normal) kidney cells.
 
@@
 v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000119888", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```
+
+C9 is slightly VWF+ (ENSG00000110799) and might contain few endothelial cells.
+
+```{r fig.width=20, fig.height=5, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000110799", pt.size = 0.2) + ggtitle("VWF")
+v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000110799", ncol = 1, group.by = "seurat_clusters", legend.position = "none", pt.size = 2)
 
@@
 ```
-Not sure what to do with these scores.
 
@@
 
-The cluster that we found seems to make sense as they can be defined by a few number of specific genes each.
-I don't know most of the marker genes found, but few point below that could support the analysis and annotation.
-
-PTPRC (C13, CD45), define the immune cluster.
+- PAX8 (C2), associated woth blastema and epithelial tumor cells (https://pubmed.ncbi.nlm.nih.gov/34306127/)
+- CD74 and HLA-DRA in C3 suggest the presence of immune cells in C13.
 
@@
 
-```{r fig.width=25, fig.height=6, out.width='100%'}
+```{r fig.width=25, fig.height=7, out.width='100%'}
 tmp_H <- NULL
@@
 
-MYC enrichment is expected in anaplastic samples.
+- C9 is enriched for EMT signature (Warning count = 5) which makes sense with the hypothesis that they are stroma cluster.
 
@@
 
-- C10 is enriched in adulte differentiated cell types/pathway of the kidney.
-This could suggest that C10 are from normal kidney. 
-To be checked with inferCNV profile.
-
-- C2, C5are enriched in muscle pathways, suggesting a stromal origin.
-
 ### Annotation from SingleR (no cancer or kidney specific dataset)
@@
 
-Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, it give the first impression that C13 a cluster of immune cells.
+Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, we can say that C13 a cluster of immune cells.
 
@@
 
-Looking at the predicted compartment, we are satisfied that the immune, endothelium and stroma compartment are specific to a set of clusters. 
+Looking at the predicted compartment, we are satisfied that the immune compartment is specific to C13. 
+Endothelium and stroma compartment are mixed in C9.
+
 It is expected that the fetal__nephron compartment match to both blastema and epithelial cancer and normal kidney.
-It is thus OK to have it labeling two distinct set of clusters.
 
-Looking at the predicted cell types, we observed that most of the clusters are labeled as cap mesenchyme, an early developmental stage of nephron.
-This suggest that these clusters are either blastema or primitive epithelium.
+Looking at the predicted cell types, we observed that the set of cluster [0,1,2,3,4,5,6,7,8,10,11,12] is labeled as cap mesenchyme, an early developmental stage of nephron
+This suggest that clusters [0,1,2,3,4,5,6,7,8,10,11,12] are more stem, either primitive epithelium or blastem cancer cells.
 
@@
 
+- C9 = stroma + endothelium (with high confidency)
 - C13 = immune cells (with high confidency)
 
-- C10, is composed of stroma (normal or cancer) cells and endothelium cells. Might be a cluster of normal cells, a bit mixed. To be checked with inferCNV.
-
-- the two other set of clusters are cancer cells, with positivity for some Wilms tumor markers such as NCAM, WT1, EPCAM.
-
-Additionally, we showed the feasibility to run label transfer using runAzimuth (part of step 3 of the proposed analysis:  https://github.com/maud-p/OpenScPCA-analysis/issues/1).
-
-We also showed the easy and robust identification of normal cells (endothelial and immune cells) that would serve as reference for the CNV inference (inferCNV, step 4 of the analysis).
+- C2. C7, C13, that form a unit in the umap reduction, are stromal cells, either cancer or normal (with high confidency)
+- [0,1,2,3,4,5,6,7,8,10,11,12] are cancer cells, either blastema or primitive epithelium.
 

### 2024-08-07 analyses/cell-type-wilms-tumor-06/01_fetal_reference_kidney.Rmd
add 01-clustering analysis
few lines to get an idea how I build the azimuth compatible fetal kidney reference
Must be improved + add a way to build proper docker container for this.
Or I can maybe directly load somewhere the azimuth compatible reference?

---
title: "Build azimuth compatible fetal kidney reference from the kidney cell atlas"
author: "Maud PLASCHKA"
date: '2024-08-07'
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

The aim is to build an azimuth compatible reference for fetal kidney from the kidney cell atlas.
Download the Fetal_full_v3.h5ad from : https://www.kidneycellatlas.org/fetal-kidney-global
And save it in path_to_resource folder

```{r }
# run with docker image maud_R_Python
    library("reticulate")
    library("scater")
    library("Seurat")
    sc <- import("scanpy")
```
```{r }

file = paste0(path_to_resource, "/Fetal_full_v3.h5ad")

adata <- sc$read_h5ad(file)

counts <- t(adata$layers["counts"])
colnames(counts) <- adata$obs_names$to_list()
rownames(counts) <- adata$var_names$to_list()
counts <- Matrix::Matrix(as.matrix(counts), sparse = T)
seurat <- CreateSeuratObject(counts)
seurat <- AddMetaData(seurat,  adata$obs)
embedding <- adata$obsm["X_umap"]
rownames(embedding) <- adata$obs_names$to_list()
colnames(embedding) <- c("umap_1", "umap_2")
seurat[["umap"]] <- CreateDimReducObject(embedding, key = "umap_")
saveRDS(seurat, file = paste0(path_to_resources,"/Fetal_full.rds"))
```

# Create a azimuth compatible reference

```{r echo=TRUE, fig.height=10, fig.width=14, message=FALSE, warning=FALSE, out.width='100%'}
# run with my standard docker image maud
library(Seurat)
library(sctransform)
library(Azimuth)

seurat <- readRDS( file = paste0(path_to_resources,"/Fetal_full.rds"))
seurat <- subset(seurat, subset = (nCount_RNA > 0) )

DimPlot(seurat, group.by = "compartment")
DimPlot(seurat, group.by = "celltype")

s <- SCTransform(seurat, verbose = FALSE, method = "glmGamPoi", conserve.memory = TRUE)
s <- RunPCA(s, npcs = 50, verbose = FALSE)

s <- RunUMAP(s, dims = 1:50, verbose = FALSE, return.model = TRUE)
DimPlot(s, reduction = "umap", group.by = "compartment") 
DimPlot(s, group.by = "celltype")

Fetal_kidney <- AzimuthReference(
  s,
  refUMAP = "umap",
  refDR = "pca",
  refAssay = "SCT",
  dims = 1:50,
  k.param = 31,
  plotref = "umap",
  plot.metadata = NULL,
  ori.index = NULL,
  colormap = NULL,
  assays = NULL,
  metadata = c("compartment", "celltype"),
  reference.version = "0.0.0",
  verbose = FALSE
)

ref_dir = paste0(path_to_resources)

# save reference in rds file
SaveAnnoyIndex(object = Fetal_kidney[["refdr.annoy.neighbors"]], file = file.path(ref_dir, "idx.annoy"))
saveRDS(object = Fetal_kidney, file = file.path(ref_dir, "ref.Rds"))

```

### 2024-08-08 analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
Add 02-clustering template analysis

--- 2024-08-07 analyses/cell-type-wilms-tumor-06/01-clustering_SCPCS000170.Rmd
@@
 ---
-title: "Evaluation of clustering parameters - SCPCS000170"
+title: "Evaluation of clustering parameters"
 author: maud-p
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE,
-                      message=FALSE,
-                      warnings=FALSE)
+knitr::opts_chunk$set(
+	echo = TRUE,
+	message = FALSE,
+	warnings = FALSE,
+	error = TRUE
+)
 ```
@@
 
-In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000170) of the Wilms Tumor dataset (SCPCP000006).
+In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000169) of the Wilms Tumor dataset (SCPCP000006).
 
@@
 
-### Packages
+In order to assess the clustering quality, we look into some marker genes, pathways enrichment and label transfer.
 
-Load required packages in the following chunk, if needed.
-Do not install packages here; only load them with the `library()` function.
-
-```{r packages, message=FALSE, warning=FALSE}
-library(SingleCellExperiment)
-library(Seurat)      # to use Seurat workflow
-
-library(tidyr)
-library(dplyr)
-library(DT)          # for table visualization
-
-library(ggplot2)     # for visualization
-library(patchwork)   # for visualization
-library(ggplotify)
-library(SCpubr)      # for visualization
-library(viridis)     # for visualization - colors
-
-library(edgeR)       # For pseudobulk based differential expression (DE) analysis
-library(DElegate)    # for pseudobulk based DE analysis and identification of marker genes
-
-library(Azimuth)     # For label transfer
-
-library(msigdbr)     # for gene set enrichment analysis or pathway enrichment
-library(clusterProfiler)
-
-library(data.table)
-```
-
-#### Some config set-up / threshold
-
-We store in a config list names "cfg" parameters used all along the analysis to filter for p-value, log fold change, percentage of expression, etc.
-
-```{r cfg}
-cfg = list()
-cfg$padj_thershold = 0.05
-cfg$lfc_threshold = 1
-cfg$rate1_threshold = 0.5
-set.seed(12345)
-```
-
-#### Base directories
-
-```{r base paths, eval=TRUE, include=TRUE}
-# The base path for the OpenScPCA repository, found by its (hidden) .git directory
-repository_base <- rprojroot::find_root(rprojroot::is_git_root)
-
-# The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "2024-07-08", "SCPCP000006")
-
-# The path to this module
-module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
-```
-
-#### Input and output files
-
-For this analysis, we worked with the _processed.rds data. 
-We builded a Seurat object based on the counts data and re-perform the analysis 
-[normalization --> reduction --> clustering] following the Seurat workflow.
-
-We transferred meta.data to keep:
-
-[x] QC data computed by the DataLab
-
-[x] annotation data computed by the DataLab
-
-[x] raw annotation and gene_symbol conversion
-
-```{r paths}
-# input
-# use download-data.py to download the data
-
-# get all files in the data directory
-filelist <- list.files(data_dir, full.names = TRUE)
-
-# select the 40 Wilms tumor single nucleus folders only
-filelist <- filelist[! grepl(".tsv", filelist)]
-
-# output
-## html report for each of the Wilms tumor sample will be saved in 
-path_to_report <- file.path(module_base, "01-clustering")
-## some plots from the report will be saved in 
-path_to_plot <- file.path(module_base, "/plots")
-## after final decision on the clustering strategy, processed RDS files will be saved in 
-path_to_result <- file.path(module_base, "/results")
-```
+## Sample metadata
 
@@
 ## to be modified when rendered
-i = 3
 filedir <- filelist[i] 
@@
 ```
-
-Here we have a sample of favorable histology, untreated. Few thoughts:
-
-- We thus do not expect anaplasia-associated patterns (TP53 as an example) to be major driver here.
-
-- We do not expect high immune infiltration (induced by chemotherapy).
-
-- We do not expect chemotherapy induced DNA damages.
 
@@
 
-[8] We run label transfer (Azimuth) to transfer annotation from the fetal kidney atlas human reference. 
+[ ] Next step, aim for a next PR: We will label transfer (Azimuth) to transfer annotation from the fetal kidney atlas human reference. 
 We plot pca/umap reduction grouping with latest labels. 
@@
 # Normalization
-options(future.globals.maxSize= 891289600000)
+options(future.globals.maxSize= 8912896000000)
 srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)
@@
 
-C9 and C13 might be a specific subtype, immune, endothelium, healthy cells (?)
+We expect up to 5 set of clusters:
+
+- blastema cancer cells
+- epithelium cancer and/or normal cells
+- stroma cancer and/or normal cells
+- immune cells
+- endothelial cells
 
@@
 
-C1 might be composed of cycling cells, in higher proportion than the other clusters.
-
 ### Look at specific genes
@@
 Here, we open the table of marker genes marker-sets/CellType_metadata.csv. 
+Note: we do not expect to have a clear and nice pattern of expression for all of the following markers in every tumor. 
+This is just ti get a few idea.
 
 ```{r marker_genes, fig.width=12, fig.height=6, out.width='100%'}
-
-  # open the set of reference marker genes
-
-CellType_metadata <- read.table("marker-sets/CellType_metadata.csv", header = TRUE, sep = ",")
 
@@
 ```
-Almost all clusters are NCAM1+ (ENSG00000149294) which is a marker of blastema area.
+
+##### NCAM1
 
@@
 
-C9 is positive for COL6A3 which is a marker of stroma Wilms tumor cells (also normal stroma)
+##### COL6A3
 
@@
 
-Almost all clusters are WT1+ (ENSG00000184937).
+##### IGF2
+
+```{r fig.width=20, fig.height=5, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000167244", pt.size = 0.2) + ggtitle("IGF2")
+v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000167244", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```
+
+##### WT1
 
@@
 
-Almost all clusters are IGF2+ (ENSG00000167244).
+##### EPCAM
 
@@
 d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000167244", pt.size = 0.2) + ggtitle("IGF2")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000167244", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000119888", pt.size = 0.2) + ggtitle("EPCAM")
+v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000119888", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
 
@@
 
-C13 is PTPRC+ (ENSG00000081237) and must be a cluster of immune cells.
-Wilms tumor are generally cold tumor, especially untreated Wilms tumor.
-According to the clinical data, the present sample is untreated (treatment = upfront resection).
-We do not expect high percentage of immune cells.
-The size of the cluster would therefore make sense.
+##### PTPRC = CD45
 
@@
 
-The set of clusters [0-2-3-4-5-6-11-12] is EPCAM+ (ENSG00000119888), which is a marker of epithelial (malignant and normal) kidney cells.
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000119888", pt.size = 0.2) + ggtitle("EPCAM")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000119888", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
-```
-
-C9 is slightly VWF+ (ENSG00000110799) and might contain few endothelial cells.
+##### VWF, endothelial cells
 
@@
 ```{r marker_pathways, fig.width=12, fig.height=6, out.width='100%'}
-# open the set of reference marker genes
 
-Pathway_metadata <- read.csv("marker-sets/Pathway_metadata.csv", header = TRUE, sep = ",")
-
-DT::datatable(Pathway_metadata, caption = ("CellType_metadata"), 
+DT::datatable(Pathway_metadata, caption = ("Pathway_metadata"), 
               extensions = 'Buttons', 
@@
 ```
+
 #### TP53 pathway
@@
 ```
+
 "Normal" cells (immune, endothelial) have a slightly higher TP53 score.
-I am however not sure how relevant the TP53 score is for Favorable sample as the one we are analysing here.
-To be tested on an anaplastic sample.
 
@@
 ```
-I am not sure how relevant the DNA repair score is for Favorable sample as the one we are analysing here. To be tested on an anaplastic sample.
+
+Note: Chemo-treated samples should have higher DNA-damage scores.
 
@@
 
-- PAX8 (C2), associated woth blastema and epithelial tumor cells (https://pubmed.ncbi.nlm.nih.gov/34306127/)
-- CD74 and HLA-DRA in C3 suggest the presence of immune cells in C13.
-
 ### Enrichment analysis of marker genes
@@
 
-```{r fig.width=25, fig.height=7, out.width='100%'}
+```{r fig.width=25, fig.height=10, out.width='100%'}
 tmp_H <- NULL
@@
 
-- C9 is enriched for EMT signature (Warning count = 5) which makes sense with the hypothesis that they are stroma cluster.
+EMT signature should be enriched in stroma cluster.
+E2F/proliferation should be enriched in blastema cluster.
+MYC(N), TP53 must be enriched in blastema cluster.
 
@@
 
-```{r fig.width=25, fig.height=6, out.width='100%'}
+```{r fig.width=25, fig.height=10, out.width='100%'}
 
@@
 ```
+
+Here to check of we catch and MIR pattern (DROSHA; DICER1; other?)
 
@@
 ```
+
+The MSigDB C8 gene set is quite relevant for kidney and nephroblastoma annotations. 
+Epithelial (cancer and normal) cells should be enriched in mature/adulte kidney pathways while blastema cancer cells will show enrichment of fetal kidney development pathway / cap mesenchyme.
 
@@
 
-Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, we can say that C13 a cluster of immune cells.
+Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, it give the first impression that C12 is a cluster of endothelial cells and C11 a cluster of immune cells.
 
@@
 
-Note to the DataLab : where should I save the fetal kidney reference? 
-In S3 like results as it is a quite big object or in the folder marker-sets?
+To be added in a next PR.
 
@@
 
-```{r fig.height=6, fig.width=6, message=FALSE, warnings=FALSE}
-
-# access to the azimuth kidney reference 
-##### to be updated depending on the location of the reference file 
-ref_dir = file.path(module_base, "marker-sets/Azimuth_Compatible_Fetal_full/")
-
-reference <- LoadReference(ref_dir)
-# Load the query object for mapping
-
-# Preprocess with SCTransform
-# Note, the RunAzimuth change a bit the Seurat object, changing for example the rownames to Symbol. 
-# We thus choose to rename the object to keep the 
-data <-  RunAzimuth(srat, ref_dir, assay = 'RNA')
-rm(reference)
-gc()
-
-DT::datatable(table(data$seurat_clusters, data$predicted.compartment), caption = ("table of azimuth compartment annotation per seurat clusters"),               extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-              buttons = c( 'csv', 'excel')))
-
-DT::datatable(table(data$seurat_clusters, data$predicted.celltype), caption = ("table of azimuth annotation per seurat clusters"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-              buttons = c( 'csv', 'excel')))
-
-```
-
-```{r fig.height=5, fig.width=20, warnings=FALSE}
-
-d1 <- DimPlot(data, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + NoLegend()
-d2 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
-d3 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.celltype", label = TRUE, repel = TRUE) + NoLegend()
-
-d1 | d2 | d3
-```
-
-Looking at the predicted compartment, we are satisfied that the immune compartment is specific to C13. 
-Endothelium and stroma compartment are mixed in C9.
-
-It is expected that the fetal__nephron compartment match to both blastema and epithelial cancer and normal kidney.
-
-Looking at the predicted cell types, we observed that the set of cluster [0,1,2,3,4,5,6,7,8,10,11,12] is labeled as cap mesenchyme, an early developmental stage of nephron
-This suggest that clusters [0,1,2,3,4,5,6,7,8,10,11,12] are more stem, either primitive epithelium or blastem cancer cells.
-
-```{r fig.height=5, fig.width=20, warnings=FALSE}
-
-f1 <- SCpubr::do_BarPlot(sample = data,
-                         group.by = "predicted.compartment",
-                         split.by = "seurat_clusters",
-                         position = "fill",
-                         font.size = 10,
-                         legend.ncol = 3) +
-                         ggtitle("% cells")+
-                         xlab(file)
-
-f2 <- SCpubr::do_BarPlot(sample = data,
-                         group.by = "predicted.celltype",
-                         split.by = "seurat_clusters",
-                         position = "fill",
-                         font.size = 10,
-                         legend.ncol = 3) +
-                         ggtitle("% cells")+
-                         xlab(file)
-
-f1 | f2
-
-```
-
-## Summary and conclusion
-
-Taking together, we can conclude:
-
-- C9 = stroma + endothelium (with high confidency)
-- C13 = immune cells (with high confidency)
-
-- C2. C7, C13, that form a unit in the umap reduction, are stromal cells, either cancer or normal (with high confidency)
-- [0,1,2,3,4,5,6,7,8,10,11,12] are cancer cells, either blastema or primitive epithelium.
-
 ## Next step
 
-[ ] We will adapt this template to be render on all the 40 Wilms tumor samples of the dataset.
+-[x] We will adapt this template to be render on all the 40 Wilms tumor samples of the dataset.
 
-[ ] We will save for each sample the rds file
+-[ ] We will save for each sample the rds file
 
-[ ] We will run inferCNV for each sample to decide the malignant/normal status of some stroma and epithelial cluster, and confirm the blastema annotation.
+-[ ] We will run inferCNV for each sample to decide the malignant/normal status of some stroma and epithelial cluster, and confirm the blastema annotation.
 

### 2024-08-08 analyses/cell-type-wilms-tumor-06/render_01-clustering.R

@@ -1,0 +1,92 @@
+# This script is to run the 01-clustering.Rmd template notebook over all samples
+
+## Load libraries
+
+library(SingleCellExperiment)
+library(Seurat)      # to use Seurat workflow
+
+library(tidyr)
+library(dplyr)
+library(DT)          # for table visualization
+
+library(ggplot2)     # for visualization
+library(patchwork)   # for visualization
+library(ggplotify)
+library(SCpubr)      # for visualization
+library(assertthat)  # for visualization, to be able to use do_DimPlot
+library(viridis)     # for visualization - colors
+
+library(edgeR)       # For pseudobulk based differential expression (DE) analysis
+library(DElegate)    # for pseudobulk based DE analysis and identification of marker genes
+
+library(Azimuth)     # For label transfer
+
+library(msigdbr)     # for gene set enrichment analysis or pathway enrichment
+library(clusterProfiler)
+
+library(data.table)
+
+## define some config set-up / threshold
+
+# We store in a config list names "cfg" parameters used all along the analysis to filter for p-value, log fold change, percentage of expression, etc.
+
+cfg = list()
+cfg$padj_thershold = 0.05
+cfg$lfc_threshold = 1
+cfg$rate1_threshold = 0.5
+set.seed(12345)
+
+
+## Define the base directories
+
+# The base path for the OpenScPCA repository, found by its (hidden) .git directory
+repository_base <- rprojroot::find_root(rprojroot::is_git_root)
+
+# The current data directory, found within the repository base directory
+data_dir <- file.path(repository_base, "data", "2024-07-08", "SCPCP000006")
+
+# The path to this module
+module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
+
+
+## Input and output files
+
+#For this analysis, we worked with the _processed.rds data. 
+#We builded a Seurat object based on the counts data and re-perform the analysis 
+#[normalization --> reduction --> clustering] following the Seurat workflow.
+
+# input
+# use download-data.py to download the data
+
+# get all files in the data directory
+filelist <- list.files(data_dir, full.names = TRUE)
+
+# select the 40 Wilms tumor single nucleus folders only
+filelist <- filelist[! grepl(".tsv", filelist)]
+
+# output
+## html report for each of the Wilms tumor sample will be saved in 
+path_to_report <- file.path(module_base, "notebook")
+## some plots from the report will be saved in 
[…]

### 2024-08-09 analyses/cell-type-wilms-tumor-06/00_fetal_reference_kidney.Rmd

---
title: "Build azimuth compatible fetal kidney reference from the kidney cell atlas"
author: "Maud PLASCHKA"
date: '2024-08-07'
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE)
```

Introduction

The aim is to build an azimuth compatible reference for fetal kidney from the kidney cell atlas.
The rds data can be download using the download link https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds

## Packages

Load required packages in the following chunk, if needed.
Do not install packages here; only load them with the `library()` function.

```{r packages, message=FALSE, warning=FALSE}
library("Seurat")
library(sctransform)
library(Azimuth)
library(SCpubr)
library(tidyverse)
library(patchwork)
library(SeuratWrappers)
```

## Base directories

```{r base paths, eval=TRUE, include=TRUE}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "2024-07-08", "SCPCP000006")

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
```

## Input files

rds data can be downloaded using the URLhttps://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds

Please note that this download link permanently references the current version of the dataset (08/2024). 
If this dataset is updated, a new download link will be created that permanently references the next version of this dataset.

We save it in the marker-sets folder of the module.
Note to the DataLab: should we save it somewhere else? 
I suggest this rds file could be placed here transiently and removed once the reference is build?

```{r path_to_data}
path_to_data <- file.path(module_base, "marker-sets/fetal_full.rds")

url = "https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds"

download.file(url, path_to_data)
```
## Output file

The azimuth compatible fetal kidney reference will be saved in the marker-sets folder from the module.

```{r path_to_output}
path_to_output <- file.path(module_base, "marker-sets")
```

## Some config set-up / threshold

We store in a config list names "cfg" parameters used all along the analysis to filter for p-value, log fold change, percentage of expression, etc.

```{r cfg}
cfg = list()
cfg$padj_thershold = 0.05
cfg$lfc_threshold = 1
cfg$rate1_threshold = 0.5

set.seed(12345)
```

# Create a azimuth compatible reference

```{r pre_process, echo=TRUE, fig.height=7, fig.width=12, message=FALSE, warning=FALSE, out.width='100%'}
seurat <- readRDS(path_to_data)

d2 <- do_DimPlot(seurat, reduction = "umap", dims = c(1,2), group.by = "compartment", label = TRUE, repel = TRUE) + NoLegend() + ggtitle("umap reduction before SCTransform")
d3 <- do_DimPlot(seurat, reduction = "umap", dims = c(1,2), group.by = "cell_type", label = TRUE, repel = TRUE) + NoLegend()

d2 | d3

options(future.globals.maxSize= 891289600000)
s <- SCTransform(seurat, verbose = FALSE, method = "glmGamPoi", conserve.memory = TRUE)
s <- RunPCA(s, npcs = 50, verbose = FALSE)
s <- RunUMAP(s, dims = 1:50, verbose = FALSE, return.model = TRUE)
```

```{r create_ref, echo=TRUE, fig.height=7, fig.width=12, message=FALSE, warning=FALSE, out.width='100%'}
options(future.globals.maxSize= 891289600000)
Fetal_kidney <- AzimuthReference(
  s,
  refUMAP = "umap",
  refDR = "pca",
  refAssay = "SCT",
  dims = 1:50,
  k.param = 31,
  plotref = "umap",
  plot.metadata = NULL,
  ori.index = NULL,
  colormap = NULL,
  assays = NULL,
  metadata = c("compartment", "cell_type"),
  reference.version = "0.0.0",
  verbose = FALSE
)

d2 <- do_DimPlot(Fetal_kidney, reduction = "refUMAP", dims = c(1,2), group.by = "compartment", label = TRUE, repel = TRUE) + NoLegend() + ggtitle("umap reduction after SCTransform and Seurat workflow")
d3 <- do_DimPlot(Fetal_kidney, reduction = "refUMAP", dims = c(1,2), group.by = "cell_type", label = TRUE, repel = TRUE) + NoLegend()

d2 | d3

# save reference in rds file
SaveAnnoyIndex(object = Fetal_kidney[["refdr.annoy.neighbors"]], file = file.path(path_to_output, "idx.annoy"))
saveRDS(object = Fetal_kidney, file = file.path(path_to_output, "ref.Rds"))

# finally, we can remove the fetal_full.rds file from the marker-sets directory
file.remove(path_to_data)
```
## Characterization of compartment and cell types in the reference

Here, we use an unbiased approach to find transcripts that characterized the different compartments and cell types.

This is just to get markers genes of the different population, in case some could be of interest for the Wilms tumor annotations.

We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. 
DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method. 
Please check the preprint from Chistoph Hafemeister: https://www.biorxiv.org/content/10.1101/2023.03.28.534443v1
and tool described here: https://github.com/cancerbits/DElegate

### Find marker genes for each of the compartment

```{r markers_compatment, fig.width=8, fig.height=7, out.width='100%'}
de_results   <- DElegate::FindAllMarkers2(s, group_column = "compartment")

#filter the most relevant markers
s.markers <- de_results[de_results$padj < cfg$padj_thershold & de_results$log_fc > cfg$lfc_threshold & de_results$rate1 > cfg$rate1_threshold,]

DT::datatable(s.markers, caption = ("marker genes"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))

# Select top 5 genes for heatmap plotting
s.markers <- na.omit(s.markers)
s.markers %>%
    group_by(group1) %>%
    top_n(n =  5, wt = log_fc) -> top5

# subset for plotting
Idents(s) <- s$compartment
cells <- WhichCells(s, downsample = 100)
ss <- subset(s, cells = cells)
ss <- ScaleData(ss, features = top5$feature)

p1 <- SCpubr::do_DimPlot(s, reduction="umap", group.by = "compartment", label = TRUE, repel = TRUE) + ggtitle("compartment")
p2 <- DoHeatmap(ss, features = top5$feature,  cells = cells, group.by = "compartment") + NoLegend() + 
  scale_fill_gradientn(colors =  c("#01665e","#35978f",'darkslategray3', "#f7f7f7", "#fee391","#fec44f","#F9AD03")) 
p3 <- ggplot(s@meta.data, aes(compartment, fill = compartment)) + geom_bar() + NoLegend()

common_title <- sprintf("Unsupervised clustering %s, %d cells", s@meta.data$orig.ident[1], ncol(s))
show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) ) + plot_layout(widths = c(1, 2)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))

```

### Find marker genes for each of the cell types

```{r markers_cell, fig.width=15, fig.height=17, out.width='100%'}
de_results   <- DElegate::FindAllMarkers2(s, group_column = "cell_type")

#filter the most relevant markers
s.markers <- de_results[de_results$padj < cfg$padj_thershold & de_results$log_fc > cfg$lfc_threshold & de_results$rate1 > cfg$rate1_threshold,]

DT::datatable(s.markers, caption = ("marker genes"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))

# Select top 5 genes for heatmap plotting
s.markers <- na.omit(s.markers)
s.markers %>%
    group_by(group1) %>%
    top_n(n =  5, wt = log_fc) -> top5

# subset for plotting
Idents(s) <- s$cell_type
cells <- WhichCells(s, downsample = 100)
ss <- subset(s, cells = cells)
ss <- ScaleData(ss, features = top5$feature)

p1 <- SCpubr::do_DimPlot(s, reduction="umap", group.by = "cell_type", label = TRUE, repel = TRUE) + ggtitle("cell_type") + NoLegend()
p2 <- DoHeatmap(ss, features = top5$feature,  cells = cells, group.by = "cell_type") + NoLegend() + 
  scale_fill_gradientn(colors =  c("#01665e","#35978f",'darkslategray3', "#f7f7f7", "#fee391","#fec44f","#F9AD03")) 
p3 <- ggplot(s@meta.data, aes(cell_type, fill = cell_type)) + geom_bar() + NoLegend() + scale_x_discrete(guide = guide_axis(angle = 90))

common_title <- sprintf("Unsupervised clustering %s, %d cells", s@meta.data$orig.ident[1], ncol(s))
show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) ) + plot_layout(widths = c(1, 1)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))

```

### 2024-08-09 analyses/cell-type-wilms-tumor-06/00_fetal_reference_kidney.Rmd
Update analyses/cell-type-wilms-tumor-06/00_fetal_reference_kidney.Rmd
Co-authored-by: Jaclyn Taroni <19534205+jaclyn-taroni@users.noreply.github.com>

@@
 # The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "2024-07-08", "SCPCP000006")
+data_dir <- file.path(repository_base, "data", "current", "SCPCP000006")
 
@@
 ```{r path_to_output}
-path_to_output <- file.path(module_base, "marker-sets")
+path_to_output <- file.path(module_base, "results", "references")
 ```

### 2024-08-09 analyses/cell-type-wilms-tumor-06/00_fetal_reference_kidney.Rmd
Update analyses/cell-type-wilms-tumor-06/00_fetal_reference_kidney.Rmd
Co-authored-by: Jaclyn Taroni <19534205+jaclyn-taroni@users.noreply.github.com>

@@
 
-We save it in the marker-sets folder of the module.
-Note to the DataLab: should we save it somewhere else? 
-I suggest this rds file could be placed here transiently and removed once the reference is build?
+We place it in the `scratch` folder of the module because it is a temporary file.
 
 ```{r path_to_data}
-path_to_data <- file.path(module_base, "marker-sets/fetal_full.rds")
+path_to_data <- file.path(module_base, "scratch", fetal_full.rds")
 

### 2024-08-09 analyses/cell-type-wilms-tumor-06/00_fetal_reference_kidney.Rmd
Update analyses/cell-type-wilms-tumor-06/00_fetal_reference_kidney.Rmd
Co-authored-by: Jaclyn Taroni <19534205+jaclyn-taroni@users.noreply.github.com>

@@
 date: '2024-08-07'
+params:
+  ref_url: "https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds"
+  padj_threshold: 0.05
+  lfc_threshold: 1
+  rate1_threshold: 0.5
+  seed: 12345
 output: 
@@
 ```{r cfg}
-cfg = list()
-cfg$padj_thershold = 0.05
-cfg$lfc_threshold = 1
-cfg$rate1_threshold = 0.5
+cfg <- list()
+cfg$padj_threshold <- params$padj_threshold
+cfg$lfc_threshold <- params$lfc_threshold
+cfg$rate1_threshold <- params$rate1_threshold
 
-set.seed(12345)
+set.seed(params$seed)
 ```

### 2024-08-09 analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 ---
-title: "Evaluation of clustering parameters"
+title: "Exploration of clustering results"
 author: maud-p

### 2024-08-09 analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
-- SCTransform (default parameters)
-- RunPCA (default) 
-- RunUMAP (dims = 1:50) 
-- FindNeighbors (dims = 1:50) 
-- FindClusters (default)
+- `SCTransform` (default parameters)
+- `RunPCA` (default) 
+- `RunUMAP` (dims = 1:50) 
+- `FindNeighbors` (dims = 1:50) 
+- `FindClusters` (default)
 
@@
 srat@misc <- metadata(sce)
+rm(sce)
 

### 2024-08-13 analyses/cell-type-wilms-tumor-06/notebook_template/01_fetal_reference_Cao_Azimuth.Rmd
add to PR #706

---
title: "Reference label transfer exploration for `r params$sample_id`"
author: "Maud PLASCHKA"
date: '2024-08-07'
params:
  current: "2024-07-08"
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000176"
  seed: 12345
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE)
```

## Introduction

The aim is to find an appropriate reference to perform label transfer as a step of the annotation of the Wilms tumor dataset for the sample `param$sample_id` =

```{r}
params$sample_id
```

The use of the right reference is crucial. 
It is recommended that the cell types in the reference is representative to the cell types to be annotated in the query.

Wilms tumors can contain up to three histologies that resemble fetal kidney: blastema, stroma, and epithelia [1-2].
Because of their histological similarity to fetal kidneys, Wilms tumors are thought to arise from developmental derangements in embryonic renal progenitors.

We thus decided to test and compare fetal (kidney) references that could be use in the analysis module.
You can find below the three workflow/reference that we plan to test.

### Human fetal kidney atlas Stewart et al.

We first wanted to try the human fetal kidney atlas to transfer label into the Wilms tumor samples using azimuth. 
You can find more about the human kidney atlas here: https://www.kidneycellatlas.org/ [3]

Using this reference, we test two workflows for label transfer:

- `Azimuth v5`,
- `Seurat v4` as described here  https://satijalab.org/seurat/articles/integration_mapping.html#cell-type-classification-using-an-integrated-reference

### Azimuth Human fetal atlas Cao et al.

Azimuth also provide a human fetal atlas as a reference [4].

The data can be found on Zenodo: 
https://zenodo.org/records/4738021#.YJIW4C2ZNQI

The reference contain cells from 15 organs including kidney from fetal samples. 
Here we will use `Azimuth` to transfer labels from the reference.

##########################################################################################################
In this notebook, we will test the human fetal Azimuth reference from Cao et al. using Azimuth . 
##########################################################################################################

### Packages

Load required packages in the following chunk, if needed.
Do not install packages here; only load them with the `library()` function.

```{r packages, message=FALSE, warning=FALSE}
library(SingleCellExperiment)

library("Seurat")
library(sctransform)
library(Azimuth)
library(SeuratData)

library(assertthat)
library(viridis)
library(SCpubr)
library(tidyverse)
library(patchwork)

set.seed(params$seed)
```

### Base directories

```{r base paths, eval=TRUE, include=TRUE}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", params$current, params$scpca_project_id)

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
```

### Input files

#### Reference

We install and load the reference using `Azimuth`.
```{r path_to_reference}
#Check the names of the Azimuth available data and reference
AvailableData()

# Install the fetal reference 
InstallData("fetusref")

ref <- SeuratData::LoadData("fetusref", type = "azimuth")
```

#### Query

In this notebook, we test the label transfer from the azimuth fetal atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`

```{r path_to_query}
filelist <- list.files(data_dir, full.names = TRUE)

# select the 40 Wilms tumor single nucleus folders only
filedir <- filelist[grepl(params$sample_id, filelist)]
```

### Output file

Reports will be saved in the `notebook/00-reference` directory.

### Load and pre-process the query data

```{r load, message=FALSE, warning=FALSE}
file <- list.files(filedir)
file <- file[grepl("_processed.rds", file)]

# open the processed rds object
sce <- readRDS(paste0(filedir, "/", file))
```

```{r seurat_wf, fig.height=4, fig.width=8, warning=FALSE, out.width='100%'}
# convert to seurat
srat <- CreateSeuratObject(counts = counts(sce),
                                    assay = "RNA",
                                    project = params$sample_id
                           )

# convert colData and rowData to data.frame for use in the Seurat object
cell_metadata <- as.data.frame(colData(sce))
row_metadata <- as.data.frame(rowData(sce))

# add cell metadata (colData) from SingleCellExperiment to Seurat
srat@meta.data <- cell_metadata

# add row metadata (rowData) from SingleCellExperiment to Seurat
srat[["RNA"]]@meta.data <- row_metadata

# add metadata from SingleCellExperiment to Seurat
srat@misc <- metadata(sce)

# Normalization
options(future.globals.maxSize= 8912896000000)
srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)

# dimensionality reduction
srat <- RunPCA(srat, verbose = F)
srat <- RunUMAP(srat, dims = 1:50, verbose = F)

# clustering
srat    <- FindNeighbors(srat, dims = 1:50, verbose = F)
srat    <- FindClusters(srat, verbose = T)

```

### Label transfer from fetal kidney reference using Azimuth

```{r fig.height=6, fig.width=6, message=FALSE, warnings=FALSE}
DefaultAssay(srat) <- "RNA"
s <- RunAzimuth(srat, reference ="fetusref" )
```

```{r fig.height=15, fig.width=8, warnings=FALSE}

d1 <- DimPlot(s, reduction = "umap", dims = c(1,2), group.by = "predicted.organ", label = TRUE, repel = TRUE) + NoLegend()
d2 <- DimPlot(s, reduction = "umap", dims = c(1,2), group.by = "predicted.annotation.l1", label = TRUE, repel = TRUE) + NoLegend()
d3 <- DimPlot(s, reduction = "umap", dims = c(1,2), group.by = "predicted.annotation.l2", label = TRUE, repel = TRUE) + NoLegend()

f1 <- SCpubr::do_BarPlot(sample = s,
                         group.by = "predicted.organ",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 4) +
                         ggtitle("% cells")+
                         xlab(file)

f2 <- SCpubr::do_BarPlot(sample = s,
                         group.by = "predicted.annotation.l1",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 2) +
                         ggtitle("% cells")+
                         xlab(file)

f3 <- SCpubr::do_BarPlot(sample = s,
                         group.by = "predicted.annotation.l2",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 2) +
                         ggtitle("% cells")+
                         xlab(file)

((d1/f1) | (d2/f2) ) 
```

## Session info

```{r }
sessionInfo()
```

## References

- [1] https://www.ncbi.nlm.nih.gov/books/NBK373356/

- [2] https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9915828/

- [3] https://www.science.org/doi/10.1126/science.aat5031

- [4] https://www.science.org/doi/10.1126/science.aba7721

### 2024-08-13 analyses/cell-type-wilms-tumor-06/notebook_template/01_fetal_reference_Stewart_Azimuthv5.Rmd
add to PR #706

---
title: "Reference label transfer exploration for `r params$sample_id`"
author: "Maud PLASCHKA"
date: '2024-08-13'
params:
  current: '2024-07-08'
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000176"
  ref_url: "https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds"
  seed: 12345
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE
                      )
```

## Introduction

The aim is to find an appropriate reference to perform label transfer as a step of the annotation of the Wilms tumor dataset for the sample `param$sample_id` =

```{r}
params$sample_id
```

The use of the right reference is crucial. 
It is recommended that the cell types in the reference is representative to the cell types to be annotated in the query.

Wilms tumors can contain up to three histologies that resemble fetal kidney: blastema, stroma, and epithelia [1-2].
Because of their histological similarity to fetal kidneys, Wilms tumors are thought to arise from developmental derangements in embryonic renal progenitors.

We thus decided to test and compare fetal (kidney) references that could be use in the analysis module.
You can find below the three workflow/reference that we plan to test.

### Human fetal kidney atlas Stewart et al.

We first wanted to try the human fetal kidney atlas to transfer label into the Wilms tumor samples using azimuth. 
You can find more about the human kidney atlas here: https://www.kidneycellatlas.org/ [3]

Using this reference, we test two workflows for label transfer:

- `Azimuth v5`,
- `Seurat v4` as described here  https://satijalab.org/seurat/articles/integration_mapping.html#cell-type-classification-using-an-integrated-reference

### Azimuth Human fetal atlas Cao et al.

Azimuth also provide a human fetal atlas as a reference [4].

The data can be found on Zenodo: 
https://zenodo.org/records/4738021#.YJIW4C2ZNQI

The reference contain cells from 15 organs including kidney from fetal samples. 
Here we will use `Azimuth` to transfer labels from the reference.

##########################################################################################################
In this notebook, we will test the human fetal kidney atlas from Stewart et al. using the Azimuth workflow. 
##########################################################################################################

### Packages

Load required packages in the following chunk, if needed.
Do not install packages here; only load them with the `library()` function.

```{r packages, message=FALSE, warning=FALSE}
library("Seurat")
library(sctransform)
library(Azimuth)
library(SCpubr)
library(tidyverse)
library(patchwork)
library(SingleCellExperiment)

set.seed(params$seed)
```

### Base directories

```{r base paths, eval=TRUE}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", params$current, params$scpca_project_id)

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
```

### Input files

#### Reference

rds data can be downloaded using the URL defined in `param$ref_url`.

Please note that this download link permanently references the current version of the dataset (08/2024). 
If this dataset is updated, a new download link will be created that permanently references the next version of this dataset.

We place it in the `scratch` folder of the module because it is a temporary file.

```{r path_to_reference}
path_to_reference <- file.path(module_base, "scratch", "fetal_full.rds")
```

#### Query

In this notebook, we test the label transfer from the fetal kidney atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`

```{r path_to_query}
filelist <- list.files(data_dir, full.names = TRUE)

# select the 40 Wilms tumor single nucleus folders only
filedir <- filelist[grepl(params$sample_id, filelist)]
```

### Output file

This notebook has two main outputs

#### Azimuth compatible reference

The reference data are downloaded from the `params$ref_url` and transiently saved in the `scratch`folder.
The azimuth compatible fetal kidney reference will be saved in the `results` folder from the module, `reference` subfolder.

Please note that the creation of the Azimuth reference cannot be done within RMarkdown. 
We thus set `include = FALSE` to the corresponding chunk.
Please, before running this code either :

- download the azimuth compatible reference from S3 bucket or
- run manualy chunk by chunk the report to create and save the reference once.

```{r path_to_output}
path_to_output <- file.path(module_base, "results", "references")
```

#### The html notebook saved afetr rendering this notebook (knit).
Will be saved in the `notebook/00-reference` directory.

## Using the Human Kidney Atlas - Fetal as a reference

### Create a azimuth compatible reference

Note: not shown here because the `AzimuthReference` function do not allow knitr.
The reference should be build "manually" prior to knitr the report.

```{r create_ref, eval=FALSE, fig.height=7, fig.width=12, message=FALSE, warning=FALSE, include=FALSE, out.width='100%'}
download.file(params$ref_url, path_to_reference)
seurat <- readRDS(path_to_reference)

options(future.globals.maxSize= 891289600000)
s <- SCTransform(seurat, verbose = FALSE, method = "glmGamPoi", conserve.memory = TRUE)
s <- RunPCA(s, npcs = 50, verbose = FALSE)
s <- RunUMAP(s, dims = 1:50, verbose = FALSE, return.model = TRUE)

Fetal_kidney <- AzimuthReference(
  s,
  refUMAP = "umap",
  refDR = "pca",
  refAssay = "SCT",
  dims = 1:50,
  k.param = 31,
  plotref = "umap",
  plot.metadata = NULL,
  ori.index = NULL,
  colormap = NULL,
  assays = NULL,
  metadata = c("compartment", "cell_type"),
  reference.version = "0.0.0",
  verbose = FALSE
)

d2 <- do_DimPlot(Fetal_kidney, reduction = "refUMAP", dims = c(1,2), group.by = "compartment", label = TRUE, repel = TRUE) + NoLegend() + ggtitle("umap reduction after SCTransform and Seurat workflow")
d3 <- do_DimPlot(Fetal_kidney, reduction = "refUMAP", dims = c(1,2), group.by = "cell_type", label = TRUE, repel = TRUE) + NoLegend()

d2 | d3

if (file.exists(path_to_output)){
   SaveAnnoyIndex(object = Fetal_kidney[["refdr.annoy.neighbors"]], file = file.path(path_to_output, "idx.annoy"))
    saveRDS(object = Fetal_kidney, file = file.path(path_to_output, "ref.Rds"))
} else {
    SaveAnnoyIndex(object = Fetal_kidney[["refdr.annoy.neighbors"]], file = file.path(path_to_output, "idx.annoy"))
    saveRDS(object = Fetal_kidney, file = file.path(path_to_output, "ref.Rds"))

}

# finally, we can remove the fetal_full.rds file from the marker-sets directory
file.remove(path_to_reference)
```

### Load and pre-process the query data

```{r load, message=FALSE, warning=FALSE}
file <- list.files(filedir)
file <- file[grepl("_processed.rds", file)]

# open the processed rds object
sce <- readRDS(paste0(filedir, "/", file))
```

```{r seurat_wf, fig.height=4, fig.width=8, warning=FALSE, out.width='100%'}
# convert to seurat
srat <- CreateSeuratObject(counts = counts(sce),
                                    assay = "RNA",
                                    project = params$sample_id
                           )

# convert colData and rowData to data.frame for use in the Seurat object
cell_metadata <- as.data.frame(colData(sce))
row_metadata <- as.data.frame(rowData(sce))

# add cell metadata (colData) from SingleCellExperiment to Seurat
srat@meta.data <- cell_metadata

# add row metadata (rowData) from SingleCellExperiment to Seurat
srat[["RNA"]]@meta.data <- row_metadata

# add metadata from SingleCellExperiment to Seurat
srat@misc <- metadata(sce)

# Normalization
options(future.globals.maxSize= 8912896000000)
srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)

# dimensionality reduction
srat <- RunPCA(srat, verbose = F)
srat <- RunUMAP(srat, dims = 1:50, verbose = F)

# clustering
srat    <- FindNeighbors(srat, dims = 1:50, verbose = F)
srat    <- FindClusters(srat, verbose = T)

```

### Azimuth annotation from fetal kidney

```{r fig.height=6, fig.width=6, message=FALSE, warnings=FALSE}

data <-  RunAzimuth(srat, path_to_output, assay = 'RNA')

```

```{r fig.height=8, fig.width=8, warnings=FALSE}

d1 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
d2 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.cell_type", label = TRUE, repel = TRUE) + NoLegend()

f1 <- SCpubr::do_BarPlot(sample = data,
                         group.by = "predicted.compartment",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

f2 <- SCpubr::do_BarPlot(sample = data,
                         group.by = "predicted.cell_type",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

(d1/f1) | (d2/f2) 
```

Note:

For some reason, the "cap-mesenchyme cells" has been renamed in cellxgene in "mesenchymal cells".

The cap mesenchyme is a cap of condensed metanephric mesenchyme, comprised of cells which epithelialize and sequentially form the pretubular aggregate (PA), renal vesicle (RV), C-, and S-shaped bodies, and finally the mature nephron. 
The CM contains nephron progenitor cells.

This can be confusing and we just need to pay attention that :

- the fetal nephron / mesenchymal cells are cap-mesenchyme cells. 
In our case, cap-mesenchyme contains blastema and primitive epitheliul cancer cells.

- the stroma / mesenchymal stem are likelly mesenchymal cancer or normal cells.

## References

- [1] https://www.ncbi.nlm.nih.gov/books/NBK373356/

- [2] https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9915828/

- [3] https://www.science.org/doi/10.1126/science.aat5031

- [4] https://www.science.org/doi/10.1126/science.aba7721

## Session info

```{r }
sessionInfo()
```

### 2024-08-13 analyses/cell-type-wilms-tumor-06/notebook_template/01_fetal_reference_Stewart_Seuratv4.Rmd
add to PR #706

--- 2024-08-13 analyses/cell-type-wilms-tumor-06/notebook_template/01_fetal_reference_Stewart_Azimuthv5.Rmd
@@
 params:
-  current: '2024-07-08'
+  current: "2024-07-08"
   scpca_project_id: "SCPCP000006"
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE,
-                      message=FALSE,
-                      warnings=FALSE
+knitr::opts_chunk$set(
+	echo = TRUE,
+	message = FALSE,
+	warnings = FALSE
 )
@@
 ##########################################################################################################
-In this notebook, we will test the human fetal kidney atlas from Stewart et al. using the Azimuth workflow. 
+In this notebook, we will test the human fetal kidney atlas from Stewart et al. using the Seurat v4 workflow. 
 ##########################################################################################################
@@
 
-```{r packages, message=FALSE, warning=FALSE}
+```{r packages, echo=TRUE, message=FALSE, warning=FALSE}
 library("Seurat")
 library(sctransform)
-library(Azimuth)
 library(SCpubr)
@@
 
-```{r base paths, eval=TRUE}
+```{r base paths, eval=TRUE, include=TRUE}
 # The base path for the OpenScPCA repository, found by its (hidden) .git directory
@@
 path_to_reference <- file.path(module_base, "scratch", "fetal_full.rds")
+
+download.file(params$ref_url, path_to_reference)
 ```
@@
 
-This notebook has two main outputs
-
-#### Azimuth compatible reference
-
-The reference data are downloaded from the `params$ref_url` and transiently saved in the `scratch`folder.
-The azimuth compatible fetal kidney reference will be saved in the `results` folder from the module, `reference` subfolder.
-
-Please note that the creation of the Azimuth reference cannot be done within RMarkdown. 
-We thus set `include = FALSE` to the corresponding chunk.
-Please, before running this code either :
-
-- download the azimuth compatible reference from S3 bucket or
-- run manualy chunk by chunk the report to create and save the reference once.
-
-```{r path_to_output}
-path_to_output <- file.path(module_base, "results", "references")
-```
-
-#### The html notebook saved afetr rendering this notebook (knit).
-Will be saved in the `notebook/00-reference` directory.
+Reports will be saved in the `notebook\00-reference` directory.
 
 ## Using the Human Kidney Atlas - Fetal as a reference
-
-### Create a azimuth compatible reference
-
-Note: not shown here because the `AzimuthReference` function do not allow knitr.
-The reference should be build "manually" prior to knitr the report.
-
-```{r create_ref, eval=FALSE, fig.height=7, fig.width=12, message=FALSE, warning=FALSE, include=FALSE, out.width='100%'}
-download.file(params$ref_url, path_to_reference)
-seurat <- readRDS(path_to_reference)
-
-options(future.globals.maxSize= 891289600000)
-s <- SCTransform(seurat, verbose = FALSE, method = "glmGamPoi", conserve.memory = TRUE)
-s <- RunPCA(s, npcs = 50, verbose = FALSE)
-s <- RunUMAP(s, dims = 1:50, verbose = FALSE, return.model = TRUE)
-
-Fetal_kidney <- AzimuthReference(
-  s,
-  refUMAP = "umap",
-  refDR = "pca",
-  refAssay = "SCT",
-  dims = 1:50,
-  k.param = 31,
-  plotref = "umap",
-  plot.metadata = NULL,
-  ori.index = NULL,
-  colormap = NULL,
-  assays = NULL,
-  metadata = c("compartment", "cell_type"),
-  reference.version = "0.0.0",
-  verbose = FALSE
-)
-
-d2 <- do_DimPlot(Fetal_kidney, reduction = "refUMAP", dims = c(1,2), group.by = "compartment", label = TRUE, repel = TRUE) + NoLegend() + ggtitle("umap reduction after SCTransform and Seurat workflow")
-d3 <- do_DimPlot(Fetal_kidney, reduction = "refUMAP", dims = c(1,2), group.by = "cell_type", label = TRUE, repel = TRUE) + NoLegend()
-
-d2 | d3
-
-if (file.exists(path_to_output)){
-   SaveAnnoyIndex(object = Fetal_kidney[["refdr.annoy.neighbors"]], file = file.path(path_to_output, "idx.annoy"))
-    saveRDS(object = Fetal_kidney, file = file.path(path_to_output, "ref.Rds"))
-} else {
-    SaveAnnoyIndex(object = Fetal_kidney[["refdr.annoy.neighbors"]], file = file.path(path_to_output, "idx.annoy"))
-    saveRDS(object = Fetal_kidney, file = file.path(path_to_output, "ref.Rds"))
-
-}
-
-# finally, we can remove the fetal_full.rds file from the marker-sets directory
-file.remove(path_to_reference)
-```
 
@@
 
-### Azimuth annotation from fetal kidney
+```{r fig.height=4, fig.width=4, warnings=FALSE}
+
+DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "seurat_clusters", label = TRUE, repel = TRUE) + NoLegend()
+
+```
+
+### Load the reference
+
+```{r pre_process,  fig.height=7, fig.width=12, message=FALSE, warning=FALSE, out.width='100%'}
+seurat <- readRDS(path_to_reference)
+
+options(future.globals.maxSize= 891289600000)
+s <- SCTransform(seurat, verbose = FALSE, method = "glmGamPoi", conserve.memory = TRUE)
+s <- RunPCA(s, npcs = 50, verbose = FALSE)
+s <- RunUMAP(s, dims = 1:50, verbose = FALSE, return.model = TRUE)
+```
+
+```{r create_ref, fig.height=7, fig.width=12, message=FALSE, warning=FALSE, out.width='100%'}
+d2 <- do_DimPlot(s, reduction = "umap", dims = c(1,2), group.by = "compartment", label = TRUE, repel = TRUE) + NoLegend() + ggtitle("umap reduction after SCTransform and Seurat workflow")
+d3 <- do_DimPlot(s, reduction = "umap", dims = c(1,2), group.by = "cell_type", label = TRUE, repel = TRUE) + NoLegend()
+
+d2 | d3
+```
+
+### Label transfer from fetal kidney reference using Seurat v4
 
@@
 
-data <-  RunAzimuth(srat, path_to_output, assay = 'RNA')
+anchors <- FindTransferAnchors(reference = s, query = srat, dims = 1:50,
+    reference.reduction = "pca")
+predictions <- TransferData(anchorset = anchors, refdata = s$cell_type, dims = 1:50)
+srat <- AddMetaData(srat, metadata = predictions$predicted.id, col.name =  "predicted.cell_type")
 
+predictions <- TransferData(anchorset = anchors, refdata = s$compartment, dims = 1:50)
+srat <- AddMetaData(srat, metadata = predictions$predicted.id, col.name =  "predicted.compartment")
 ```
@@
 
-d1 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
-d2 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.cell_type", label = TRUE, repel = TRUE) + NoLegend()
+d1 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
+d2 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "predicted.cell_type", label = TRUE, repel = TRUE) + NoLegend()
 
-f1 <- SCpubr::do_BarPlot(sample = data,
+f1 <- SCpubr::do_BarPlot(sample = srat,
                          group.by = "predicted.compartment",
@@
 
-f2 <- SCpubr::do_BarPlot(sample = data,
+f2 <- SCpubr::do_BarPlot(sample = srat,
                          group.by = "predicted.cell_type",
@@
 
+## Session info
+
+```{r }
+sessionInfo()
+```
+
 ## References
@@
 - [4] https://www.science.org/doi/10.1126/science.aba7721
-
-## Session info
-
-```{r }
-sessionInfo()
-```

### 2024-08-13 analyses/cell-type-wilms-tumor-06/00_reference.R

@@ -1,0 +1,34 @@
+# get list of samples in the library
+metadata <- read.table("sample_metadata.tsv", sep = "\t", header = TRUE)
+
+
+# Render the reports for (all) samples in the project
+for (i in metadata$scpca_sample_id[9:11]) {
+  # Label transfer from the Cao reference using Azimuth
+  rmarkdown::render(input = "notebook_template/01_fetal_reference_Cao_Azimuth.Rmd", 
+                    params = list(scpca_project_id = metadata$scpca_project_id[metadata$scpca_sample_id ==i], sample_id = i),
+                    output_format = "html_document",
+                    output_file = paste0("01_fetal_reference_Cao_Azimuth_",i, ".html"),
+                    output_dir = "notebook/00-reference")
+  
+  # Label transfer from the Stewart reference using Seurat
+  rmarkdown::render(input = "notebook_template/01_fetal_reference_Stewart_Seuratv4.Rmd", 
+                    params = list(scpca_project_id = metadata$scpca_project_id[metadata$scpca_sample_id ==i],sample_id = i),
+                    output_format = "html_document",
+                    output_file = paste0("01_fetal_reference_Stewart_Seuratv4_",i, ".html"),
+                    output_dir = "notebook/00-reference")
+  
+  # Label transfer from the Stewart reference using Azimuth
+  # Note this report need the download of the azimuth compatible reference in results s3 bucket.
+  # I would be in favor to remove this part in the module analysis as the result with Seurat v4 are similar.
+  # I just left it here now for information
+  rmarkdown::render(input = "notebook_template/01_fetal_reference_Stewart_Azimuthv5.Rmd", 
+                    params = list(scpca_project_id = metadata$scpca_project_id[metadata$scpca_sample_id ==i], sample_id= i),
+                    output_format = "html_document",
+                    output_file = paste0("01_fetal_reference_Stewart_Azimuthv5_",i, ".html"),
+                    output_dir = "notebook/00-reference")
+  
+  }
+
+
+

### 2024-08-14 analyses/cell-type-wilms-tumor-06/00_fetal_reference_kidney.Rmd
add to PR #706
switch to selfcontained mode to render the plots

@@
     df_print: paged
+    self_contained: yes
+    mode: selfcontained
 ---
@@
 ```{r path_to_data}
-path_to_data <- file.path(module_base, "scratch", fetal_full.rds")
+path_to_data <- file.path(module_base, "scratch", "fetal_full.rds")
 

### 2024-08-14 analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>
Add to PR #704

@@
 ---
-title: "Exploration of clustering results"
+title: "Exploration of clustering results for `r params$sample_id`"
 author: maud-p
@@
     df_print: kable
+    self_contained: yes
+    mode: selfcontained 
 params:
-  sample_name: ~
-  sample_path: ~
-  out_rds_path: ~
+  current: "2024-07-08"
+  scpca_project_id: "SCPCP000006"
+  sample_id: "SCPCS000176"
+  seed: 12345
+  padj_thershold: 0.05  # Threshold for adjusted p-value
+  lfc_threshold: 1      # Threshold for log fold change
+  rate1_threshold: 0.5  # Threshold for rate (percentage) of cells expressing the marker genes in group1
 ---
@@
 	message = FALSE,
-	warnings = FALSE,
-	error = TRUE
+	warnings = FALSE
 )
@@
 
-This a template for an analysis notebook using RMarkdown.
+This a template for an analysis notebook using `RMarkdown`.
 
-In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000169) of the Wilms Tumor dataset (SCPCP000006).
+In this notebook, we will perform the `Seurat` workflow [normalization --> reduction --> clustering] and explore the clustering results.
 
@@
 In order to assess the clustering quality, we look into some marker genes, pathways enrichment and label transfer.
+This approach would provide us a rapid idea of the quality of the clustering:
 
-## Sample metadata
-
-```{r paths2}
-## to be modified when rendered
-filedir <- filelist[i] 
-sample <- gsub(paste0(data_dir, "/"), "", filedir)
-
-metadata <- read.table(file.path(data_dir, "/single_cell_metadata.tsv"), sep = "\t", header=TRUE)
-
-metadata[metadata$scpca_sample_id == sample,]
-```
+- in the case each cluster do have a set of specific marker genes, we could expect each of the cluster to have at least a different phenotype,
+- if marker genes are mostly shared between clusters, we might have overclustered,
+- if no marker genes are found, the quality of cells in the clusters might be impaired (high mitochondrial content? ribosomal content?)
 
@@
 
-[4] We look at specific pathways that we reported in the table marker.sets/Pathways_metadata.csv to check the relevance of the clustering.
+[4] We look at specific pathways to check the relevance of the clustering.
 
-[5] We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. 
-DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method.
+[5] We run `DElegate::FindAllMarkers2` to find markers of the different clusters and manually check if they do make sense. 
+`DElegate::FindAllMarkers2` is an improved version of `Seurat::FindAllMarkers` based on pseudobulk differential expression method.
 
-[6] We perform enrichment analysis of marker genes for each seurat clusters. 
-We defined all the genes from the seurat object as the universe and used the MSigDB gene sets.
+[6] We perform enrichment analysis of marker genes for each Seurat clusters. 
+We defined all the genes from the Seurat object as the universe and used the MSigDB gene sets.
 
-[7] We plot pca/umap reduction grouping with available annotations (singler_, cellassign_). 
+[7] We plot `pca`/`umap` reduction grouping with available annotations (`singler_`, `cellassign_`). 
 We expect at least immune cells to be correctly label and fall into a few set of clusters.
 
-[ ] Next step, aim for a next PR: We will label transfer (Azimuth) to transfer annotation from the fetal kidney atlas human reference. 
+[ ] Next step, aim for a next PR: We will transfer annotations from two human fetal (kidney) references (Stewart et al., Cao et al.). 
 We plot pca/umap reduction grouping with latest labels. 
 We expect it to be the most representative of the cell types in the sample.
+
+### Packages
+
+Load required packages in the following chunk, if needed.
+Do not install packages here; only load them with the `library()` function.
+
+```{r packages, message=FALSE, warning=FALSE}
+library(SingleCellExperiment)
+
+library("Seurat")
+
+#library(Azimuth)            # Will be required once we add the labl transfer from Azimuth fetal reference
+#library(SeuratData)         # Will be required once we add the labl transfer from Azimuth fetal reference
+
+library(viridis)            # for plotting with viridis colors
+library(SCpubr)             # for plotting 
+library(tidyverse)
+library(patchwork)
+
+library(msigdbr)
+library(enrichplot)
+library(clusterProfiler)
+
+library(org.Hs.eg.db)
+
+set.seed(params$seed)
+```
+
+### Base directories
+
+```{r base paths}
+# The base path for the OpenScPCA repository, found by its (hidden) .git directory
+repository_base <- rprojroot::find_root(rprojroot::is_git_root)
+
+# The current data directory, found within the repository base directory
+data_dir <- file.path(repository_base, "data", params$current, params$scpca_project_id)
+
+# The path to this module
+module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
+```
 
@@
 
+### Functions
+
+Here we defined function that will be used multiple time all along the notebook.
+
+#### Visualize seurat clusters and markers genes
+
+For a Seurat object `object`and a features `features`, the function `visualize_feature` will plot `FeaturePlot` and `ViolinPlot`
+
+- `object` is the Seurat object
+
+- `features` the gene or quantitative value to be plotted
+
+- `group.by` is the metadata used for grouping the violin plots
+
+```{r }
+
+visualize_feature <- function(object, features, group.by = "seurat_clusters"){
+
+                  feature_symbol = AnnotationDbi::select(org.Hs.eg.db, 
+                                keys=features, 
+                                columns="SYMBOL", 
+                                keytype="ENSEMBL")
+
+                  d <- SCpubr::do_FeaturePlot(object, 
+                                              features = feature_symbol$ENSEMBL, 
+                                              pt.size = 0.2, 
+                                              legend.width = 0.5, 
+                                              legend.length = 5, 
+                                              legend.position = "right") + ggtitle(feature_symbol$SYMBOL)
+
+                  b <- SCpubr::do_ViolinPlot(srat, 
+                                             features = feature_symbol$ENSEMBL, 
+                                             ncol = 1, 
+                                             group.by = group.by, 
+                                             legend.position = "none") + ylab(feature_symbol$SYMBOL)
+
+                  return(d + b + plot_layout(ncol = 2, widths = c(2,4)))
+
+}
+
+```
+
+#### Visualize seurat clusters and metadata
+
+For a Seurat object `object`and a metadata `metadata`, the function `visualize_metadata` will plot `FeaturePlot` and `BarPlot`
+
+- `object` is the Seurat object
+
+- `metadata` the gene or quantitative value to be plotted
+
+- `group.by` is the metadata used for grouping the violin plots
+
+```{r }
+
+visualize_metadata <- function(object, meta, group.by){
+
+  if(class(object@meta.data[,meta]) == "numeric"){
+     d <- SCpubr::do_FeaturePlot(object, 
+                                              features = meta, 
+                                              pt.size = 0.2, 
+                                              legend.width = 0.5, 
+                                              legend.length = 5, 
+                                              legend.position = "right") + ggtitle(meta)
+
+    b <- SCpubr::do_ViolinPlot(srat, 
+                                             features = meta, 
+                                             ncol = 1, 
+                                             group.by = group.by, 
+                                             legend.position = "none")
+
+   return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
+  }
+
+  else{
+
+  d <- SCpubr::do_DimPlot(object, reduction="umap", group.by = group.by, label = TRUE, repel = TRUE) + ggtitle(paste0(meta," - umap")) + theme(text=element_text(size=18))
+
+  b <- SCpubr::do_BarPlot(sample = object,
+                         group.by = meta,
+                         split.by = group.by,
+                         position = "fill",
+                         font.size = 10,
+                         legend.ncol = 3) +
+                         ggtitle("% cells")+
+                         xlab(file) +
+                         theme(text=element_text(size=18))
+  return(d + b + plot_layout(ncol = 2, widths = c(2,4)))
+  }
+
+}
+
+```
+
+#### Calculate `ModuleScore` from a `MSigDB` dataset
+
+For a Seurat object `object`, the function `MSigDB_score` will calculate a score using `AddModuleScore()` function for a MSigDB gene set `gs_name` in the category `category`
+
+- `object` is the Seurat object to calculate the score. 
+Score will be added in the metadata of this object.
+
+- `category` is a MSigDB collection (https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp).
+values in `c("H", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")`
+
+- `gs_name` is the name of a MSigDB gene set, e.g. `"HALLMARK_P53_PATHWAY"`
+
+- `name` is the name of the module
+
+```{r }
+
+MSigDB_score <- function(object, category, gs_name, name){
+
+        set <- msigdbr(species = "human", category = category)
+
+        set_list <- set %>% 
+                    dplyr::filter(gs_name == gs_name) %>%
+                    dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()
+
+        set_list <- list(set_list$human_ensembl_gene)
+
+        object <- AddModuleScore(object, features = set_list, name = name)
+
+return(object)
+}
+
+```
+
+#### Enrichment analysis
+
+`Enrichment_plot`aim to perform enrichment of the marker genes for each of the Seurat clusters and summarize the results in a `dotplot`.
+
+- `category` is the MSigDB category or collection to be used
+values in `c("H", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")`
+
+- `signatures` is a list of marker genes per cluster
+
+- `background`is the universe used for enrichment
+
+```{r }
+
+Enrichment_plot <- function(category, signatures, background){
+
+    ## define genesets
+    gene_set <- msigdbr(species = "human", category = category)
+    msigdbr_set <-  gene_set %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
+
+    cclust<-compareCluster(geneCluster = signatures, 
+               fun = enricher,
+               TERM2GENE = msigdbr_set,
+               universe=background)
+    d <- dotplot(cclust,showCategory=15) + scale_y_discrete(labels=function(x) str_wrap(x, width=40))
+    return(d)  
+}
+
+```
+
 ## Analysis content
@@
 ```{r load}
+filelist <- list.files(data_dir, full.names = TRUE)
+
+# select the 40 Wilms tumor single nucleus folders only
+filedir <- filelist[grepl(params$sample_id, filelist)]
+
 file <- list.files(filedir)
@@
 
-```{r seurat_wf, fig.width=8, fig.height=4, out.width='100%'}
+```{r seurat_wf}
 # convert to seurat
@@
                                     assay = "RNA",
-                                    project = sample
+                                    project = params$sample_id
                            )
@@
 ### Cell cycle information
-```{r cell_cycle, fig.height=20, fig.width=30}
+```{r cell_cycle}
 s.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$s.genes]
@@
 
-```{r fig.width=20, fig.height=5, out.width='100%'}
-
-d2 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "Phase", label = TRUE) + ggtitle("Phase - umap")
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")
-
-b <- SCpubr::do_BarPlot(sample = srat,
-                         group.by = "Phase",
-                         split.by = "seurat_clusters",
-                         position = "fill",
-                         font.size = 10,
-                         legend.ncol = 3) +
-                         ggtitle("% cells")+
-                         xlab(file)
-d1 + d2 + b + plot_layout(ncol = 3, widths = c(2,2,4))
-
+```{r fig.width=10, fig.height=4, out.width='100%'}
+visualize_metadata(srat, meta = "Phase", group.by = "seurat_clusters")
 ```
 
-```{r fig.width=10, fig.height=2.5, out.width='100%'}
- SCpubr::do_ViolinPlot(srat, features = c("S.Score", "G2M.Score"), ncol = 2, group.by = "seurat_clusters", legend.position = "none") 
+```{r fig.width=8, fig.height=2.5, out.width='100%'}
+visualize_metadata(srat, meta  = "S.Score", group.by = "seurat_clusters")
+visualize_metadata(srat, meta = "G2M.Score", group.by = "seurat_clusters")
 ```
@@
 ```{r marker_genes, fig.width=12, fig.height=6, out.width='100%'}
+CellType_metadata <- read_csv(file.path(module_base, "marker-sets", "CellType_metadata.csv"))
 
@@
 
-#### Malignant markers
-
-```{r fig.width=25, fig.height=5, out.width='100%'}
- SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("malignant")])] , ncol = 9, group.by = "seurat_clusters", legend.position = "none")
-```
-
-##### NCAM1
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000149294", pt.size = 0.2) + ggtitle("NCAM1")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000149294", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
-```
-
-##### COL6A3
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000163359", pt.size = 0.2) + ggtitle("COL6A3")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000163359", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
-
-```
-
-##### IGF2
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000167244", pt.size = 0.2) + ggtitle("IGF2")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000167244", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
-```
-
-##### WT1
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000184937", pt.size = 0.2) + ggtitle("WT1")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000184937", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
-```
-
-##### EPCAM
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000119888", pt.size = 0.2) + ggtitle("EPCAM")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000119888", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
-```
-
-#### Immune markers
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
- SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("immune")])]
-                       , ncol = 6, group.by = "seurat_clusters", legend.position = "none") 
-```
-
-##### PTPRC = CD45
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000081237", pt.size = 0.2) + ggtitle("PTPRC = CD45")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000081237", ncol = 1, group.by = "seurat_clusters", legend.position = "none", pt.size = 2)
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
-```
-
-#### Other markers
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
- SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[! CellType_metadata$cell_class %in% c("immune", "malignant")])]
-                       , ncol = 6, group.by = "seurat_clusters", legend.position = "none") 
-```
-
-##### VWF, endothelial cells
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000110799", pt.size = 0.2) + ggtitle("VWF")
-v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000110799", ncol = 1, group.by = "seurat_clusters", legend.position = "none", pt.size = 2)
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```{r fig.width=8, fig.height=2.5, out.width='100%'}
+for(feature in CellType_metadata$ENSEMBL_ID){
+  if(feature %in% rownames(srat)){
+    tmp <-  visualize_feature(srat, features = feature, group.by = "seurat_clusters")
+    print(tmp)
+  }
+}
 ```
@@
 ### Look at specific pathways
-
-Here, we open the table of marker genes marker-sets/Pathway_metadata.csv.
-
-```{r marker_pathways, fig.width=12, fig.height=6, out.width='100%'}
-
-DT::datatable(Pathway_metadata, caption = ("Pathway_metadata"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-
-                               buttons = c( 'csv', 'excel')))
-```
 
@@
 
-```{r fig.width=10, fig.height=10, out.width='100%'}
-
-## define genesets
-hallmarks <- msigdbr(species = "human", category = "H")
-
-TP53_list = hallmarks %>% 
-  filter(gs_name == "HALLMARK_P53_PATHWAY") %>%
-  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()
-
-TP53_list <- list(TP53_list$human_ensembl_gene)
-
-srat <- AddModuleScore(srat, features = TP53_list, name = "TP53_score")
-```
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "TP53_score1", pt.size = 0.2) + ggtitle("TP53 score")
-v <- SCpubr::do_ViolinPlot(srat, features = "TP53_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```{r fig.height=2.5, fig.width=8, warning=FALSE, out.width='100%'}
+srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_P53_PATHWAY", name = "TP53_score")
+visualize_metadata(srat, meta = "TP53_score1", group.by = "seurat_clusters" )
 ```
@@
 
-```{r fig.width=10, fig.height=10, out.width='100%'}
-
-DNA_repair = hallmarks %>% 
-  filter(gs_name == "HALLMARK_DNA_REPAIR") %>%
-  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()
-
-DNA_repair_list <- list(DNA_repair$human_ensembl_gene)
-
-srat <- AddModuleScore(srat, features = DNA_repair_list, name = "DNA_repair_score")
-```
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "DNA_repair_score1", pt.size = 0.2) + ggtitle("DNA_repair score")
-v <- SCpubr::do_ViolinPlot(srat, features = "DNA_repair_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```{r fig.height=2.5, fig.width=8, warning=FALSE, out.width='100%'}
+srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_DNA_REPAIR", name = "DNA_repair_score")
+visualize_metadata(srat, meta = "DNA_repair_score1", group.by = "seurat_clusters" )
 ```
@@
 
-```{r fig.width=10, fig.height=10, out.width='100%'}
-
-## define genesets
-c3 <- msigdbr(species = "human", category = "C3", subcategory = "TFT:GTRD")
-
-DROSHA_list = c3 %>% 
-  filter(gs_name == "DROSHA_TARGET_GENES") %>%
-  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()
-
-DROSHA_list <- list(DROSHA_list$human_ensembl_gene)
-
-srat <- AddModuleScore(srat, features = DROSHA_list, name = "DROSHA_score")
-```
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "DROSHA_score1", pt.size = 0.2) + ggtitle("DROSHA score")
-v <- SCpubr::do_ViolinPlot(srat, features = "DROSHA_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```{r fig.height=2.5, fig.width=8, warning=FALSE, out.width='100%'}
+srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DROSHA_TARGET_GENES", name = "DROSHA_score")
+visualize_metadata(srat, meta = "DROSHA_score1", group.by = "seurat_clusters" )
 ```
@@
 
-```{r fig.width=10, fig.height=10, out.width='100%'}
-DICER1_list = c3 %>% 
-  filter(gs_name == "DICER1_TARGET_GENES") %>%
-  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()
-
-DICER1_list <- list(DICER1_list$human_ensembl_gene)
-
-srat <- AddModuleScore(srat, features = DICER1_list, name = "DICER1_score")
-```
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-d2 <- SCpubr::do_FeaturePlot(srat, features = "DICER1_score1", pt.size = 0.2) + ggtitle("DICER1 score")
-v <- SCpubr::do_ViolinPlot(srat, features = "DICER1_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
+```{r fig.height=2.5, fig.width=8, warning=FALSE, out.width='100%'}
+srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DICER1_TARGET_GENES", name = "DICER1_score")
+visualize_metadata(srat, meta = "DICER1_score1", group.by = "seurat_clusters" )
 ```
@@
 
+Of note, we won't use it for annotation, this is just here to get an idea!
+
 ```{r fig.width=10, fig.height=10, out.width='100%'}
@@
 #filter the most relevant markers
-s.markers <- de_results[de_results$padj < cfg$padj_thershold & de_results$log_fc > cfg$lfc_threshold & de_results$rate1 > cfg$rate1_threshold,]
+s.markers <- de_results[de_results$padj < params$padj_thershold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold,]
 
@@
 
-```{r fig.width=10, fig.height=10, out.width='100%'}
+```{r enrichment}
 # define background genes = universe for enrichment
@@
 
-## define genesets
-c8 <- msigdbr(species = "human", category = "C8")
-
-msigdbr_hallmarks = hallmarks %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
-msigdbr_c3 = c3 %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
-msigdbr_c8 = c8 %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
+# Define gene signature per cluster
+signatures <- list()
+for( i in unique(s.markers$group1)){
+  signatures[[paste0("cluster ",i)]] <- s.markers$feature[s.markers$group1 == i]
+}
 ```
@@
 
-```{r fig.width=25, fig.height=10, out.width='100%'}
-tmp_H <- NULL
-for(i in unique(srat$seurat_clusters)){
-signature = s.markers$feature[s.markers$group1 == i]
-ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_hallmarks)
-tmp <- NULL
-
-  if(!is.null(ego_module)){
-    if(dim(ego_module)[1]>0){
-    b_H <- barplot(ego_module, showCategory = 15, font.size = 20, label_format = 40)  
-    tmp <- b_H$data
-    tmp$set <- "Hallmarks"
-    tmp$cluster <- i
+```{r fig.width=15, fig.height=8, out.width='100%'}
+tryCatch(
+  expr =  {
+    Enrichment_plot(category = "H", signatures = signatures, background = background)
+  },
+  error = function(e) {
+    print("No enrichment")
   }
-  }
-tmp_H <- rbind(tmp_H, tmp)
-}
-
-ggplot(tmp_H, aes(Count, ID)) +
-    geom_bar(stat = "identity", aes(fill=cluster))+
-
-  theme_classic()+ 
-    geom_text(
-        aes(label = (paste( "Padj = ", round(p.adjust,2)))),
-        color = "black",
-        size = 3,
-        hjust=1,
-        position = position_dodge(0.5)
-    )+
-   theme(text = element_text(size=14))+
-  facet_wrap(facets=c("cluster"), ncol = length(unique(tmp_H$cluster)))
+)
 ```
@@
 
-```{r fig.width=25, fig.height=10, out.width='100%'}
-
-tmp_H <- NULL
-for(i in unique(srat$seurat_clusters)){
-signature = s.markers$feature[s.markers$group1 == i]
-ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_c3)
-tmp <- NULL
-
-  if(!is.null(ego_module)){
-    if(dim(ego_module)[1]>0){
-    b_H <- barplot(ego_module, showCategory = 15, font.size = 20, label_format = 40)  
-    tmp <- b_H$data
-    tmp$ID <- factor(tmp$ID, levels = c(tmp$ID[order(tmp$Count)]))
-    tmp$set <- "c8"
-    tmp$cluster <- i
+```{r fig.width=20, fig.height=25, out.width='100%'}
+tryCatch(
+  expr =  {
+Enrichment_plot(category = "C3", signatures = signatures, background = background)
+  },
+  error = function(e) {
+    print("No enrichment")
   }
-  }
-tmp_H <- rbind(tmp_H, tmp[1:5,])
-}
-
-ggplot(tmp_H, aes(Count, ID)) +
-    geom_bar(stat = "identity", aes(fill=cluster))+
-
-  theme_classic()+ 
-    geom_text(
-        aes(label = (paste( "Padj = ", round(p.adjust,2)))),
-        color = "black",
-        size = 3,
-        hjust=1,
-        position = position_dodge(0.5)
-    )+
-   theme(text = element_text(size=14))+
-  facet_wrap(facets=c("cluster"), ncol = length(unique(tmp_H$cluster)))
+)
 ```
-
 Here to check of we catch and MIR pattern (DROSHA; DICER1; other?)
@@
 
-```{r fig.width=30, fig.height=15, out.width='100%'}
-
-tmp_H <- NULL
-for(i in unique(srat$seurat_clusters)){
-signature = s.markers$feature[s.markers$group1 == i]
-ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_c8)
-tmp <- NULL
-
-  if(!is.null(ego_module)){
-    if(dim(ego_module)[1]>0){
-    b_H <- barplot(ego_module, showCategory = 15, font.size = 20, label_format = 40)  
-    tmp <- b_H$data
-    tmp$ID <- factor(tmp$ID, levels = c(tmp$ID[order(tmp$Count)]))
-    tmp$set <- "c8"
-    tmp$cluster <- i
+```{r fig.width=20, fig.height=25, out.width='100%'}
+tryCatch(
+  expr =  {
+Enrichment_plot(category = "C8", signatures = signatures, background = background)
+  },
+  error = function(e) {
+    print("No enrichment")
   }
-  }
-tmp_H <- rbind(tmp_H, tmp[1:5,])
-}
-
-ggplot(tmp_H, aes(Count, ID)) +
-    geom_bar(stat = "identity", aes(fill=cluster))+
-
-  theme_classic()+ 
-    geom_text(
-        aes(label = (paste( "Padj = ", round(p.adjust,2)))),
-        color = "black",
-        size = 3,
-        hjust=1,
-        position = position_dodge(0.5)
-    )+
-   theme(text = element_text(size=14))+
-  facet_wrap(facets=c("cluster"), ncol = length(unique(tmp_H$cluster)))
+)
 ```
@@
 ```{r fig.width=15, fig.height=10, out.width='100%'}
-
-d2 <-DimPlot(srat, group.by = "singler_celltype_annotation", reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()
-
-DT::datatable(table(srat$seurat_clusters, srat$singler_celltype_annotation), caption = ("table of SingleR annotation per seurat clusters"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-
-                               buttons = c( 'csv', 'excel')))
-
-d3 <- SCpubr::do_BarPlot(sample = srat,
-                         group.by = "singler_celltype_annotation",
-                         split.by = "seurat_clusters",
-                         position = "fill",
-                         font.size = 10,
-                         legend.ncol = 3) +
-                         ggtitle("% cells")+
-                         xlab(file)
-
-d2|d3
+visualize_metadata(srat, meta = "singler_celltype_annotation", group.by = "seurat_clusters")
 ```
@@
 ```{r fig.width=15, fig.height=10, out.width='100%'}
-
-d2 <-DimPlot(srat, group.by = "cellassign_celltype_annotation", reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()
-
-DT::datatable(table(srat$seurat_clusters, srat$cellassign_celltype_annotation)
-, caption = ("table of cellassign annotation per seurat clusters"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-
-                               buttons = c( 'csv', 'excel')))
-
-d3 <- SCpubr::do_BarPlot(sample = srat,
-                         group.by = "cellassign_celltype_annotation",
-                         split.by = "seurat_clusters",
-                         position = "fill",
-                         font.size = 10,
-                         legend.ncol = 3) +
-                         ggtitle("% cells")+
-                         xlab(file)
-
-d2|d3
+visualize_metadata(srat, meta = "cellassign_celltype_annotation", group.by = "seurat_clusters")
 ```
@@
 
-To be added in a next PR.
+To be added in a next PR. Look at opened PR #706
 
@@
 
-- endotheliúm
+- endothelium
 - stroma
@@
 
-## Next step
-
--[x] We will adapt this template to be render on all the 40 Wilms tumor samples of the dataset.
-
--[ ] We will save for each sample the rds file
-
--[ ] We will run inferCNV for each sample to decide the malignant/normal status of some stroma and epithelial cluster, and confirm the blastema annotation.
-
-The next step will provide us a better understanding of the entire cohort. We will then have to set up a strategy to annotate each sample. Open questions are:
-
-[ ] should we annotate single cell                      
-or   
-[] consider applying similar annotations to all cells in a cluster?
-
-[ ] manual annotation of each cluster / each patient    
-or   
-[] automated annotation using some threshold?
-
 ## Session Info

### 2024-08-14 analyses/cell-type-wilms-tumor-06/render_01-clustering.R

@@ -1,92 +1,0 @@
-# This script is to run the 01-clustering.Rmd template notebook over all samples
-
-## Load libraries
-
-library(SingleCellExperiment)
-library(Seurat)      # to use Seurat workflow
-
-library(tidyr)
-library(dplyr)
-library(DT)          # for table visualization
-
-library(ggplot2)     # for visualization
-library(patchwork)   # for visualization
-library(ggplotify)
-library(SCpubr)      # for visualization
-library(assertthat)  # for visualization, to be able to use do_DimPlot
-library(viridis)     # for visualization - colors
-
-library(edgeR)       # For pseudobulk based differential expression (DE) analysis
-library(DElegate)    # for pseudobulk based DE analysis and identification of marker genes
-
-library(Azimuth)     # For label transfer
-
-library(msigdbr)     # for gene set enrichment analysis or pathway enrichment
-library(clusterProfiler)
-
-library(data.table)
-
-## define some config set-up / threshold
-
-# We store in a config list names "cfg" parameters used all along the analysis to filter for p-value, log fold change, percentage of expression, etc.
-
-cfg = list()
-cfg$padj_thershold = 0.05
-cfg$lfc_threshold = 1
-cfg$rate1_threshold = 0.5
-set.seed(12345)
-
-
-## Define the base directories
-
-# The base path for the OpenScPCA repository, found by its (hidden) .git directory
-repository_base <- rprojroot::find_root(rprojroot::is_git_root)
-
-# The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "current", "SCPCP000006")
-
-# The path to this module
-module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
-
-
-## Input and output files
-
-#For this analysis, we worked with the _processed.rds data. 
-#We builded a Seurat object based on the counts data and re-perform the analysis 
-#[normalization --> reduction --> clustering] following the Seurat workflow.
-
-# input
-# use download-data.py to download the data
-
-# get all files in the data directory
-filelist <- list.files(data_dir, full.names = TRUE)
-
-# select the 40 Wilms tumor single nucleus folders only
-filelist <- filelist[! grepl(".tsv", filelist)]
-
-# output
-## html report for each of the Wilms tumor sample will be saved in 
-path_to_report <- file.path(module_base, "notebook")
-## some plots from the report will be saved in 
[…]

### 2024-08-14 analyses/cell-type-wilms-tumor-06/scripts/render_01-clustering.R

@@ -1,0 +1,16 @@
+# get list of samples in the library
+root_dir <- rprojroot::find_root(rprojroot::is_git_root)
+sample_metadata_file <- file.path(root_dir, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
+metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
+
+for (i in metadata$scpca_sample_id[9:11]) {
+  
+  rmarkdown::render(input = "notebook-template/01-clustering.Rmd", 
+                    params = list(scpca_project_id = metadata$scpca_project_id[metadata$scpca_sample_id ==i], sample_id = i),
+                    output_format = "html_document",
+                    output_file = paste0("01-clustering_",i, ".html"),
+                    output_dir = "notebook/01-clustering")
+  
+  
+}
+

### 2024-08-14 analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 ```{r load}
-filelist <- list.files(data_dir, full.names = TRUE)
+sample_dir <- file.path(data_dir, params$sample_id)
+sample_files <- list.files(sample_dir)
 
-# select the 40 Wilms tumor single nucleus folders only
-filedir <- filelist[grepl(params$sample_id, filelist)]
-
-file <- list.files(filedir)
-file <- file[grepl("_processed.rds", file)]
+file <- sample_files[grepl("_processed.rds", sample_files)]
 
 # open the processed rds object
-sce <- readRDS(paste0(filedir, "/", file))
+sce <- readRDS(file.path(sample_dir, file))
 ```
@@
 # Normalization
-options(future.globals.maxSize= 8912896000000)
 srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)
@@
 Note: we do not expect to have a clear and nice pattern of expression for all of the following markers in every tumor. 
-This is just ti get a few idea.
+This is just to get a few ideas.
 

### 2024-08-16 analyses/cell-type-wilms-tumor-06/notebook_template/00b_characterize_fetal_kidney_reference_Stewart.Rmd
changes to #706

---
title: "Characterize the fetal kidney reference from the kidney cell atlas"
author: "Maud PLASCHKA"
date: '2024-08-07'
params:
  url: https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds
  padj_threshold: 0.05
  lfc_threshold: 1
  rate1_threshold: 0.5
  seed: 12345
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
    self_contained: yes
    mode: selfcontained
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE)
```

# Introduction

The aim is to characterize the human fetal kidney from the kidney cell atlas.
You can find more about the human kidney atlas here: https://www.kidneycellatlas.org/ [1]
The rds data can be download using the download link https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds
The azimuth compatible reference has been downloaded and created in the `R script` `download-and-create-fetal-kidney-ref.R`

## Packages

Load required packages in the following chunk, if needed.
Do not install packages here; only load them with the `library()` function.

```{r packages, message=FALSE, warning=FALSE}
library("Seurat")
library(Azimuth)
library(SCpubr)
library(tidyverse)
library(patchwork)

set.seed(params$seed)
options(future.globals.maxSize= 891289600000)
```

## Base directories

```{r base paths, eval=TRUE, include=TRUE}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")

```

## Input files

The input file is the output of the `R script` `download-and-create-fetal-kidney-ref.R`

```{r path_to_data}
path_to_data <- file.path(
  module_base,
  "scratch",
  "fetal_kidney.rds"
)
download.file(url = params$url, destfile = path_to_data)
```

## Output file

We will save the result of the differential expression analysis in results/references/00b_marker_genes_fetal_kidney_Stewart.csv
Notebook is saved in the `notebook/00-reference` directory

```{r path_to_output}
path_to_output <- file.path(module_base, "results", "references")
```

# Analysis

## Load the reference

```{r pre_process, echo=TRUE, fig.height=7, fig.width=12, message=FALSE, warning=FALSE, out.width='100%'}
fetal_kidney <- readRDS(path_to_data)

d1 <- do_DimPlot(fetal_kidney, reduction = "umap", dims = c(1,2), group.by = "compartment", label = TRUE, repel = TRUE) + NoLegend() 
d2 <- do_DimPlot(fetal_kidney, reduction = "umap", dims = c(1,2), group.by = "cell_type", label = TRUE, repel = TRUE) + NoLegend()

d1 | d2
```

## Characterization of compartment and cell types in the reference

Here, we use an unbiased approach to find transcripts that characterized the different compartments and cell types.

This is just to get markers genes of the different population, in case some could be of interest for the Wilms tumor annotations.

We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. 
DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method. 
Please check the preprint from Chistoph Hafemeister: https://www.biorxiv.org/content/10.1101/2023.03.28.534443v1
and tool described here: https://github.com/cancerbits/DElegate

### Find marker genes for each of the compartment

```{r markers_compatment, fig.width=8, fig.height=7, out.width='100%'}
de_results   <- DElegate::FindAllMarkers2(fetal_kidney, group_column = "compartment",)

#filter the most relevant markers
s.markers <- de_results[de_results$padj < params$padj_threshold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold,]

DT::datatable(s.markers, caption = ("marker genes"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
                               buttons = c( 'csv', 'excel')))

# Select top 5 genes for heatmap plotting
s.markers <- na.omit(s.markers)
s.markers %>%
    group_by(group1) %>%
    top_n(n =  5, wt = log_fc) -> top5

# subset for plotting
Idents(fetal_kidney) <- fetal_kidney$compartment
cells <- WhichCells(fetal_kidney, downsample = 100)
ss <- subset(fetal_kidney, cells = cells)
ss <- ScaleData(ss, features = top5$feature)

p1 <- SCpubr::do_DimPlot(fetal_kidney, reduction="umap", group.by = "compartment", label = TRUE, repel = TRUE) + ggtitle("compartment")
p2 <- DoHeatmap(ss, features = top5$feature,  cells = cells, group.by = "compartment") + NoLegend() + 
  scale_fill_gradientn(colors =  c("#01665e","#35978f",'darkslategray3', "#f7f7f7", "#fee391","#fec44f","#F9AD03")) 
p3 <- ggplot(fetal_kidney@meta.data, aes(compartment, fill = compartment)) + geom_bar() + NoLegend()

common_title <- sprintf("Unsupervised clustering %s, %d cells", fetal_kidney@meta.data$orig.ident[1], ncol(fetal_kidney))
show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) ) + plot_layout(widths = c(1, 2)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))

write_csv(de_results, file = file.path(path_to_output, "00a_marker_compartment_fetal_kidney_Stewart.csv"))

```

### Find marker genes for each of the cell types

```{r markers_cell, fig.width=17, fig.height=20, out.width='100%'}
de_results   <- DElegate::FindAllMarkers2(fetal_kidney, group_column = "cell_type")

#filter the most relevant markers
s.markers <- de_results[de_results$padj < params$padj_threshold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold,]

DT::datatable(s.markers, caption = ("marker genes"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
                               buttons = c( 'csv', 'excel')))

# Select top 5 genes for heatmap plotting
s.markers <- na.omit(s.markers)
s.markers %>%
    group_by(group1) %>%
    top_n(n =  5, wt = log_fc) -> top5

# subset for plotting
Idents(fetal_kidney) <- fetal_kidney$cell_type
cells <- WhichCells(fetal_kidney, downsample = 100)
ss <- subset(fetal_kidney, cells = cells)
ss <- ScaleData(ss, features = top5$feature)

p1 <- SCpubr::do_DimPlot(fetal_kidney, reduction="umap", group.by = "cell_type", label = TRUE, repel = TRUE) + ggtitle("cell_type") + NoLegend()
p2 <- DoHeatmap(ss, features = top5$feature,  cells = cells, group.by = "cell_type") + NoLegend() + 
  scale_fill_gradientn(colors =  c("#01665e","#35978f",'darkslategray3', "#f7f7f7", "#fee391","#fec44f","#F9AD03")) 
p3 <- ggplot(fetal_kidney@meta.data, aes(cell_type, fill = cell_type)) + geom_bar() + NoLegend() + scale_x_discrete(guide = guide_axis(angle = 90))

common_title <- sprintf("Unsupervised clustering %s, %d cells", fetal_kidney@meta.data$orig.ident[1], ncol(fetal_kidney))
show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) ) + plot_layout(widths = c(1, 1)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))

write_csv( de_results, file = file.path(path_to_output, "00a_marker_cell-type_fetal_kidney_Stewart.csv"))

```

## References

- [1] https://www.science.org/doi/10.1126/science.aat5031

## Session info

```{r }
sessionInfo()
```

### 2024-08-16 analyses/cell-type-wilms-tumor-06/notebook_template/01_seurat-processing.Rmd
changes to #706

---
title: "Seurat processing for `r params$sample_id`"
author: "Maud PLASCHKA"
date: '2024-08-07'
params:
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000176"
  seed: 12345
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE)
```

## Introduction

The aim is to pre-process the sample `param$sample_id` =

```{r}
params$sample_id
```

Here we perform the `Seurat` workflow [normalization --> reduction --> clustering] and explore the clustering results.

This correspond to the step 2 of the proposed analysis: clustering of cells across a set of parameters for few samples

- https://github.com/AlexsLemonade/OpenScPCA-analysis/discussions/635
- https://github.com/AlexsLemonade/OpenScPCA-analysis/issues/671

This notebook will be updated with PR #704 to explore the clustering.

### Packages

Load required packages in the following chunk, if needed.
Do not install packages here; only load them with the `library()` function.

```{r packages, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library("Seurat")

set.seed(params$seed)
```

### Base directories

```{r base paths, eval=TRUE, include=TRUE}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", params$scpca_project_id)

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
```

### Input files

In this notebook, we pre-processed the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`

```{r path_to_query}
filelist <- list.files(data_dir, full.names = TRUE)

# select the 40 Wilms tumor single nucleus folders only
filedir <- filelist[grepl(params$sample_id, filelist)]

```

### Output file

Reports will be saved in the `notebook/01-Seurat` directory. 
The pre-processed `Seurat` object per samples are saved in the `result` folder.
They will be the `input` for the next `notebook-template`.
```{r path_to_output}
output_dir <- file.path(module_base, "results", params$sample_id)
```

### Functions

#### Convert `SingleCellExperiment` object to `Seurat`

```{r sce_to_seurat}
convert_sce_to_seurat <- function(sce, 
                                  sample_id) {
  # Convert a SingleCellExperiment from the ScPCA Portal to a Seurat object
  # 
  # Arguments:
  #	  sce: A SingleCellExperiment from the ScPCA Portal
  #   sample_id: The sample identifier supplied as the project to 
  #              CreateSeuratObject() (character)
  #
  # Returns: A Seurat Object
  # Check class of input
  stopifnot(class(sce) == "SingleCellExperiment")

  srat <- CreateSeuratObject(counts = counts(sce),
                             assay = "RNA",
                             project = sample_id
  )

  # convert colData and rowData to data.frame for use in the Seurat object
  cell_metadata <- as.data.frame(colData(sce))
  row_metadata <- as.data.frame(rowData(sce))

  # add cell metadata (colData) from SingleCellExperiment to Seurat
  srat@meta.data <- cell_metadata

  # add row metadata (rowData) from SingleCellExperiment to Seurat
  srat[["RNA"]]@meta.data <- row_metadata

  # add metadata from SingleCellExperiment to Seurat
  srat@misc <- metadata(sce)

  # return Seurat object
  return(srat)

}

```

#### Perform `Seurat` workflow

`seurat_wf` run the `Seurat` workflow from a `Seurat`object `srat`:

```{r seurat_wf}
perform_norm_dimred_cluster <- function(srat,
                                        transform_method = "glmGamPoi",
                                        conserve_memory = TRUE,
                                        umap_dims = 1:50, 
                                        neighbors_dims = 1:50, 
                                        verbose = FALSE) {

  # Perform normalization, dimensionality reduction, and clustering on a 
  # Seurat object
  #
  # Arguments:
  #   srat: A Seurat object
  #   transform_method: The method passed to SCTransform(); glmGamPoi by default
  #   conserve_memory: Logical specifying whether or not SCTransform() should
  #                    conserve memory
  #   umap_dims: A vector of integers passed to RunUMAP() to specify the number 
  #              of dimensions; 1:50 by default
  #   neighbors_dims: A vector of integers passed to FindNeighbors() to specify
  #                   the dimensions of reduction to use as input
  #   verbose: Logical specifying whether or not functions should be verbose
  #
  # Returns: 
  #   A Seurat object containing all the contents of the normalization, PCA, 
  #   UMAP, neighbors, and clustering steps
  #

  # Check class
  stopifnot(class(srat) == "Seurat")

  # Normalization
  options(future.globals.maxSize= 8912896000000)
  srat <- SCTransform(srat, 
                      method = transform_method,
                      verbose = verbose, 
                      conserve.memory = conserve_memory)

  # dimensionality reduction
  srat <- RunPCA(srat, verbose = verbose)
  srat <- RunUMAP(srat, dims = umap_dims, verbose = verbose)

  # clustering
  srat    <- FindNeighbors(srat, dims = neighbors_dims, verbose = verbose)
  srat    <- FindClusters(srat, verbose = verbose)

  # Return
  return(srat)

}

```

## Analysis

### Load and pre-process the query data

```{r load, message=FALSE, warning=FALSE}
file <- list.files(filedir)
file <- file[grepl("_processed.rds", file)]
# open the processed rds object
sce <- readRDS(paste0(filedir, "/", file))
```

```{r run_seurat_wf, fig.height=4, fig.width=8, warning=FALSE, out.width='100%'}

srat <- convert_sce_to_seurat(sce, 
                                  sample_id = params$sample_id)

srat <- perform_norm_dimred_cluster(srat,
                                        transform_method = "glmGamPoi",
                                        conserve_memory = TRUE,
                                        umap_dims = 1:50, 
                                        neighbors_dims = 1:50, 
                                        verbose = FALSE)
```

## Save the `Seurat`object

```{r save}
saveRDS(object = srat, file = file.path(output_dir, paste0("01-Seurat_", params$sample_id,".Rds")))
```

## Session info

```{r }
sessionInfo()
```

### 2024-08-16 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
changes to #706

--- 2024-08-13 analyses/cell-type-wilms-tumor-06/notebook_template/01_fetal_reference_Cao_Azimuth.Rmd
@@
 ---
-title: "Reference label transfer exploration for `r params$sample_id`"
+title: "Reference label transfer from the fetal full reference (Cao et al.) for `r params$sample_id`"
 author: "Maud PLASCHKA"
@@
 params:
-  current: "2024-07-08"
   scpca_project_id: "SCPCP000006"
@@
 
-The aim is to find an appropriate reference to perform label transfer as a step of the annotation of the Wilms tumor dataset for the sample `param$sample_id` =
+The aim is to perform label transfer as a step of the annotation of the Wilms tumor dataset for the sample `param$sample_id` =
 
@@
 We thus decided to test and compare fetal (kidney) references that could be use in the analysis module.
-You can find below the three workflow/reference that we plan to test.
+You can find below the two references that we plan to use
 
@@
 
-We first wanted to try the human fetal kidney atlas to transfer label into the Wilms tumor samples using azimuth. 
+Stewart et al. created and a human fetal kidney atlas.
+We will use `Azimuth` to transfer label into the Wilms tumor samples. 
 You can find more about the human kidney atlas here: https://www.kidneycellatlas.org/ [3]
-
-Using this reference, we test two workflows for label transfer:
-
-- `Azimuth v5`,
-- `Seurat v4` as described here  https://satijalab.org/seurat/articles/integration_mapping.html#cell-type-classification-using-an-integrated-reference
 
@@
 
-##########################################################################################################
-In this notebook, we will test the human fetal Azimuth reference from Cao et al. using Azimuth . 
-##########################################################################################################
+In this notebook, we perform label transfer from the human fetal Azimuth reference from Cao et al.
 
@@
 
-# The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", params$current, params$scpca_project_id)
-
 # The path to this module
@@
 
-In this notebook, we test the label transfer from the azimuth fetal atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`
+In this notebook, we test the label transfer from the azimuth fetal atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`.
+
+We work with the pre-processed `Seurat`object saved in the `results` directory.
 
 ```{r path_to_query}
-filelist <- list.files(data_dir, full.names = TRUE)
-
-# select the 40 Wilms tumor single nucleus folders only
-filedir <- filelist[grepl(params$sample_id, filelist)]
+data_dir <- file.path(module_base, "results", params$sample_id)
 ```
@@
 
-Reports will be saved in the `notebook/00-reference` directory.
+Reports will be saved in the `notebook/02-reference` directory. 
+The pre-processed and annotated `Seurat` object per samples are saved in the `result` folder.
+```{r path_to_output}
+output_dir <- file.path(module_base, "results", params$sample_id)
+```
 
-### Load and pre-process the query data
+## Analysis
+
+### Load the pre-process data (query)
 
 ```{r load, message=FALSE, warning=FALSE}
-file <- list.files(filedir)
-file <- file[grepl("_processed.rds", file)]
-
 # open the processed rds object
-sce <- readRDS(paste0(filedir, "/", file))
-```
-
-```{r seurat_wf, fig.height=4, fig.width=8, warning=FALSE, out.width='100%'}
-# convert to seurat
-srat <- CreateSeuratObject(counts = counts(sce),
-                                    assay = "RNA",
-                                    project = params$sample_id
-                           )
-
-# convert colData and rowData to data.frame for use in the Seurat object
-cell_metadata <- as.data.frame(colData(sce))
-row_metadata <- as.data.frame(rowData(sce))
-
-# add cell metadata (colData) from SingleCellExperiment to Seurat
-srat@meta.data <- cell_metadata
-
-# add row metadata (rowData) from SingleCellExperiment to Seurat
-srat[["RNA"]]@meta.data <- row_metadata
-
-# add metadata from SingleCellExperiment to Seurat
-srat@misc <- metadata(sce)
-
-# Normalization
-options(future.globals.maxSize= 8912896000000)
-srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)
-
-# dimensionality reduction
-srat <- RunPCA(srat, verbose = F)
-srat <- RunUMAP(srat, dims = 1:50, verbose = F)
-
-# clustering
-srat    <- FindNeighbors(srat, dims = 1:50, verbose = F)
-srat    <- FindClusters(srat, verbose = T)
-
+srat <- readRDS(file.path(data_dir, paste0("01-Seurat_", params$sample_id,".Rds")))
 ```
@@
 
-```{r fig.height=6, fig.width=6, message=FALSE, warnings=FALSE}
+```{r run_azimuth, message=FALSE, warnings=FALSE}
 DefaultAssay(srat) <- "RNA"
 s <- RunAzimuth(srat, reference ="fetusref" )
+
+# We transfer the annotation to the pre-processed `Seurat`iobject as we don't want to keep changes done on the query by `RunAzimuth`
+metadata_to_trasfer <- s@meta.data[, c("predicted.annotation.l1.score", "predicted.annotation.l1", "predicted.annotation.l2.score", "predicted.annotation.l2", "predicted.organ.score", "predicted.organ")]
+srat <- AddMetaData(srat, metadata_to_trasfer, col.name = c("fetal_full_predicted.annotation.l1.score", "fetal_full_predicted.annotation.l1", "fetal_full_predicted.annotation.l2.score", "fetal_full_predicted.annotation.l2", "fetal_full_predicted.organ.score", "fetal_full_predicted.organ"))
 ```
 
-```{r fig.height=15, fig.width=8, warnings=FALSE}
+```{r plot_azimuth, fig.height=15, fig.width=8, warnings=FALSE}
 
-d1 <- DimPlot(s, reduction = "umap", dims = c(1,2), group.by = "predicted.organ", label = TRUE, repel = TRUE) + NoLegend()
-d2 <- DimPlot(s, reduction = "umap", dims = c(1,2), group.by = "predicted.annotation.l1", label = TRUE, repel = TRUE) + NoLegend()
-d3 <- DimPlot(s, reduction = "umap", dims = c(1,2), group.by = "predicted.annotation.l2", label = TRUE, repel = TRUE) + NoLegend()
+d1 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_full_predicted.organ", label = TRUE, repel = TRUE) + NoLegend()
+d2 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_full_predicted.annotation.l1", label = TRUE, repel = TRUE) + NoLegend()
+d3 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_full_predicted.annotation.l2", label = TRUE, repel = TRUE) + NoLegend()
 
-f1 <- SCpubr::do_BarPlot(sample = s,
-                         group.by = "predicted.organ",
+f1 <- SCpubr::do_BarPlot(sample = srat,
+                         group.by = "fetal_full_predicted.organ",
                          split.by = "seurat_clusters",
@@
 
-f2 <- SCpubr::do_BarPlot(sample = s,
-                         group.by = "predicted.annotation.l1",
+f2 <- SCpubr::do_BarPlot(sample = srat,
+                         group.by = "fetal_full_predicted.annotation.l1",
                          split.by = "seurat_clusters",
@@
                          ggtitle("% cells")+
-                         xlab(file)
+                         xlab(params$sample_id)
 
-f3 <- SCpubr::do_BarPlot(sample = s,
-                         group.by = "predicted.annotation.l2",
+f3 <- SCpubr::do_BarPlot(sample = srat,
+                         group.by = "fetal_full_predicted.annotation.l2",
                          split.by = "seurat_clusters",
@@
                          ggtitle("% cells")+
-                         xlab(file)
+                         xlab(params$sample_id)
 
 ((d1/f1) | (d2/f2) ) 
+```
+
+## Save the `Seurat`object
+
+```{r save}
+saveRDS(object = srat, file = file.path(output_dir, paste0("00a-fetal_full_label-transfer_",params$sample_id,".Rds")))
 ```

### 2024-08-16 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
changes to #706

--- 2024-08-13 analyses/cell-type-wilms-tumor-06/notebook_template/01_fetal_reference_Stewart_Azimuthv5.Rmd
@@
 ---
-title: "Reference label transfer exploration for `r params$sample_id`"
+title: "Reference label transfer of the fetal kidney reference (Stewart et al.) for `r params$sample_id`"
 author: "Maud PLASCHKA"
@@
 params:
-  current: '2024-07-08'
   scpca_project_id: "SCPCP000006"
   sample_id: "SCPCS000176"
-  ref_url: "https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds"
   seed: 12345
@@
 
-##########################################################################################################
-In this notebook, we will test the human fetal kidney atlas from Stewart et al. using the Azimuth workflow. 
-##########################################################################################################
+In this notebook, we perform label transfer from the human fetal kidney atlas from Stewart et al.
 
@@
 # The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", params$current, params$scpca_project_id)
+data_dir <- file.path(repository_base, "data", "current", params$scpca_project_id)
 
@@
 
-rds data can be downloaded using the URL defined in `param$ref_url`.
-
-Please note that this download link permanently references the current version of the dataset (08/2024). 
-If this dataset is updated, a new download link will be created that permanently references the next version of this dataset.
-
-We place it in the `scratch` folder of the module because it is a temporary file.
+The reference has been download and pre-process in the `R Script` `download-and-create-fetal-kidney-ref.R`
 
 ```{r path_to_reference}
-path_to_reference <- file.path(module_base, "scratch", "fetal_full.rds")
+path_to_reference <- file.path(module_base, "results", "references")
 ```
@@
 In this notebook, we test the label transfer from the fetal kidney atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id` 
+We work with the pre-processed `Seurat`object and labeled with the fetal kidney reference saved in the `results` directory.
 
 ```{r path_to_query}
-filelist <- list.files(data_dir, full.names = TRUE)
-
-# select the 40 Wilms tumor single nucleus folders only
-filedir <- filelist[grepl(params$sample_id, filelist)]
+data_dir <- file.path(module_base, "results", params$sample_id)
 ```
@@
 
-This notebook has two main outputs
-
-#### Azimuth compatible reference
-
-The reference data are downloaded from the `params$ref_url` and transiently saved in the `scratch`folder.
-The azimuth compatible fetal kidney reference will be saved in the `results` folder from the module, `reference` subfolder.
-
-Please note that the creation of the Azimuth reference cannot be done within RMarkdown. 
-We thus set `include = FALSE` to the corresponding chunk.
-Please, before running this code either :
-
-- download the azimuth compatible reference from S3 bucket or
-- run manualy chunk by chunk the report to create and save the reference once.
+Reports will be saved in the `notebook/02-reference` directory. 
+The pre-processed and annotated `Seurat` object per samples are saved in the `result` folder.
 
 ```{r path_to_output}
-path_to_output <- file.path(module_base, "results", "references")
+output_dir <- file.path(module_base, "results", params$sample_id)
 ```
 
-#### The html notebook saved afetr rendering this notebook (knit).
-Will be saved in the `notebook/00-reference` directory.
+## Analysis
 
-## Using the Human Kidney Atlas - Fetal as a reference
-
-### Create a azimuth compatible reference
-
-Note: not shown here because the `AzimuthReference` function do not allow knitr.
-The reference should be build "manually" prior to knitr the report.
-
-```{r create_ref, eval=FALSE, fig.height=7, fig.width=12, message=FALSE, warning=FALSE, include=FALSE, out.width='100%'}
-download.file(params$ref_url, path_to_reference)
-seurat <- readRDS(path_to_reference)
-
-options(future.globals.maxSize= 891289600000)
-s <- SCTransform(seurat, verbose = FALSE, method = "glmGamPoi", conserve.memory = TRUE)
-s <- RunPCA(s, npcs = 50, verbose = FALSE)
-s <- RunUMAP(s, dims = 1:50, verbose = FALSE, return.model = TRUE)
-
-Fetal_kidney <- AzimuthReference(
-  s,
-  refUMAP = "umap",
-  refDR = "pca",
-  refAssay = "SCT",
-  dims = 1:50,
-  k.param = 31,
-  plotref = "umap",
-  plot.metadata = NULL,
-  ori.index = NULL,
-  colormap = NULL,
-  assays = NULL,
-  metadata = c("compartment", "cell_type"),
-  reference.version = "0.0.0",
-  verbose = FALSE
-)
-
-d2 <- do_DimPlot(Fetal_kidney, reduction = "refUMAP", dims = c(1,2), group.by = "compartment", label = TRUE, repel = TRUE) + NoLegend() + ggtitle("umap reduction after SCTransform and Seurat workflow")
-d3 <- do_DimPlot(Fetal_kidney, reduction = "refUMAP", dims = c(1,2), group.by = "cell_type", label = TRUE, repel = TRUE) + NoLegend()
-
-d2 | d3
-
-if (file.exists(path_to_output)){
-   SaveAnnoyIndex(object = Fetal_kidney[["refdr.annoy.neighbors"]], file = file.path(path_to_output, "idx.annoy"))
-    saveRDS(object = Fetal_kidney, file = file.path(path_to_output, "ref.Rds"))
-} else {
-    SaveAnnoyIndex(object = Fetal_kidney[["refdr.annoy.neighbors"]], file = file.path(path_to_output, "idx.annoy"))
-    saveRDS(object = Fetal_kidney, file = file.path(path_to_output, "ref.Rds"))
-
-}
-
-# finally, we can remove the fetal_full.rds file from the marker-sets directory
-file.remove(path_to_reference)
-```
-
-### Load and pre-process the query data
-
+### Load the pre-processed `Seurat` object
 ```{r load, message=FALSE, warning=FALSE}
-file <- list.files(filedir)
-file <- file[grepl("_processed.rds", file)]
-
 # open the processed rds object
-sce <- readRDS(paste0(filedir, "/", file))
-```
-
-```{r seurat_wf, fig.height=4, fig.width=8, warning=FALSE, out.width='100%'}
-# convert to seurat
-srat <- CreateSeuratObject(counts = counts(sce),
-                                    assay = "RNA",
-                                    project = params$sample_id
-                           )
-
-# convert colData and rowData to data.frame for use in the Seurat object
-cell_metadata <- as.data.frame(colData(sce))
-row_metadata <- as.data.frame(rowData(sce))
-
-# add cell metadata (colData) from SingleCellExperiment to Seurat
-srat@meta.data <- cell_metadata
-
-# add row metadata (rowData) from SingleCellExperiment to Seurat
-srat[["RNA"]]@meta.data <- row_metadata
-
-# add metadata from SingleCellExperiment to Seurat
-srat@misc <- metadata(sce)
-
-# Normalization
-options(future.globals.maxSize= 8912896000000)
-srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)
-
-# dimensionality reduction
-srat <- RunPCA(srat, verbose = F)
-srat <- RunUMAP(srat, dims = 1:50, verbose = F)
-
-# clustering
-srat    <- FindNeighbors(srat, dims = 1:50, verbose = F)
-srat    <- FindClusters(srat, verbose = T)
-
+srat <- readRDS(file.path(data_dir, paste0("00a-fetal_full_label-transfer_",params$sample_id,".Rds")))
 ```
@@
 
-```{r fig.height=6, fig.width=6, message=FALSE, warnings=FALSE}
+```{r run_azimuth, message=FALSE, warnings=FALSE}
 
-data <-  RunAzimuth(srat, path_to_output, assay = 'RNA')
-
+data <-  RunAzimuth(srat, path_to_reference, assay = 'RNA')
+# We transfer the annotation to the pre-processed `Seurat`object as we don't want to keep changes done on the query by `RunAzimuth`
+metadata_to_trasfer <- data@meta.data[, c("predicted.compartment.score", "predicted.compartment", "predicted.cell_type.score", "predicted.cell_type")]
+srat <- AddMetaData(srat, metadata_to_trasfer, col.name = c("fetal_kidney_predicted.compartment.score", "fetal_kidney_predicted.compartment", "fetal_kidney_predicted.cell_type.score", "fetal_kidney_predicted.cell_type"))
 ```
 
-```{r fig.height=8, fig.width=8, warnings=FALSE}
+```{r plot_azimuth, fig.height=8, fig.width=8, warnings=FALSE}
 
-d1 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
-d2 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.cell_type", label = TRUE, repel = TRUE) + NoLegend()
+d1 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_kidney_predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
+d2 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_kidney_predicted.cell_type", label = TRUE, repel = TRUE) + NoLegend()
 
-f1 <- SCpubr::do_BarPlot(sample = data,
-                         group.by = "predicted.compartment",
+f1 <- SCpubr::do_BarPlot(sample = srat,
+                         group.by = "fetal_kidney_predicted.compartment",
                          split.by = "seurat_clusters",
@@
                          ggtitle("% cells")+
-                         xlab(file)
+                         xlab(params$sample_id)
 
-f2 <- SCpubr::do_BarPlot(sample = data,
-                         group.by = "predicted.cell_type",
+f2 <- SCpubr::do_BarPlot(sample = srat,
+                         group.by = "fetal_kidney_predicted.cell_type",
                          split.by = "seurat_clusters",
@@
                          ggtitle("% cells")+
-                         xlab(file)
+                         xlab(params$sample_id)
 
@@
 
+## Save the `Seurat`object
+
+```{r save}
+saveRDS(object = srat, file = file.path(output_dir, paste0("00b-fetal_kidey_label-transfer_",params$sample_id,".Rds")))
+```
+
+## Session info
+
+```{r }
+sessionInfo()
+```
+
 ## References

### 2024-08-16 analyses/cell-type-wilms-tumor-06/00_reference.R

@@ -1,36 +1,0 @@
-# get list of samples in the library
-root_dir <- rprojroot::find_root(rprojroot::is_git_root)
-sample_metadata_file <- file.path(root_dir, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
-metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
-
-
-# Render the reports for (all) samples in the project
-for (i in metadata$scpca_sample_id[9:11]) {
-  # Label transfer from the Cao reference using Azimuth
-  rmarkdown::render(input = "notebook_template/01_fetal_reference_Cao_Azimuth.Rmd", 
-                    params = list(scpca_project_id = metadata$scpca_project_id[metadata$scpca_sample_id ==i], sample_id = i),
-                    output_format = "html_document",
-                    output_file = paste0("01_fetal_reference_Cao_Azimuth_",i, ".html"),
-                    output_dir = "notebook/00-reference")
-  
-  # Label transfer from the Stewart reference using Seurat
-  rmarkdown::render(input = "notebook_template/01_fetal_reference_Stewart_Seuratv4.Rmd", 
-                    params = list(scpca_project_id = metadata$scpca_project_id[metadata$scpca_sample_id ==i],sample_id = i),
-                    output_format = "html_document",
-                    output_file = paste0("01_fetal_reference_Stewart_Seuratv4_",i, ".html"),
-                    output_dir = "notebook/00-reference")
-  
-  # Label transfer from the Stewart reference using Azimuth
-  # Note this report need the download of the azimuth compatible reference in results s3 bucket.
-  # I would be in favor to remove this part in the module analysis as the result with Seurat v4 are similar.
-  # I just left it here now for information
-  rmarkdown::render(input = "notebook_template/01_fetal_reference_Stewart_Azimuthv5.Rmd", 
-                    params = list(scpca_project_id = metadata$scpca_project_id[metadata$scpca_sample_id ==i], sample_id= i),
-                    output_format = "html_document",
-                    output_file = paste0("01_fetal_reference_Stewart_Azimuthv5_",i, ".html"),
-                    output_dir = "notebook/00-reference")
-  
-  }
-
-
-

### 2024-08-16 analyses/cell-type-wilms-tumor-06/00_run_workflow.R

@@ -1,0 +1,62 @@
+#!/usr/bin/env Rscript
+
+# Run the Label transfer from two fetal references ------------------------------
+
+# get list of samples in the library --------------------------------------------
+root_dir <- rprojroot::find_root(rprojroot::is_git_root)
+sample_metadata_file <- file.path(root_dir, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
+metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
+
+# set path to this module--------------------------------------------------------
+module_base <- file.path(root_dir, "analyses", "cell-type-wilms-tumor-06")
+
+# Download and create the fetal kidney reference (Stewart et al) ----------
+source(file.path(module_base,"scripts", "download-and-create-fetal-kidney-ref.R"))
+
+# Characterize the two fetal references -----------------------------------------
+
+# Characterize the fetal full reference (Cao et al.)
+# To be done, next PR
+
+# Characterize the fetal kidney reference (Stewart et al.)
+rmarkdown::render(input = file.path(module_base, "notebook_template", "00b_characterize_fetal_kidney_reference_Stewart.Rmd"),
+                  output_format = "html_document",
+                  output_file = "00b_characterization_fetal_kidney_reference_Stewart.html",
+                  output_dir = file.path(module_base, "notebook","00-reference"))
+
+
+# Run the workflow for (all) samples in the project -----------------------------
+for (i in metadata$scpca_sample_id[1:11]) {
+ 
+  # create a directory to save the pre-processed and labeled `Seurat` objects
+  dir.create(file.path(module_base, "results", i))
+  # create a directory to save the notebooks
+  dir.create(file.path(module_base, "notebook", i))
+  
+  
+  # Pre-process the data - `Seurat` workflow
+  rmarkdown::render(input = file.path(module_base, "notebook_template", "01_seurat-processing.Rmd"),
+                    params = list(scpca_project_id = metadata$scpca_project_id[metadata$scpca_sample_id ==i], sample_id = i),
+                    output_format = "html_document",
+                    output_file = paste0("01_Clustering_",i, ".html"),
+                    output_dir = file.path(module_base, "notebook", i))
+  
+  # Label transfer from the Cao reference using Azimuth
+  rmarkdown::render(input = file.path(module_base, "notebook_template", "02a_label-transfer_fetal_full_reference_Cao.Rmd"),
[…]

### 2024-08-19 analyses/cell-type-wilms-tumor-06/notebook_template/00b_characterize_fetal_kidney_reference_Stewart.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/00b_characterize_fetal_kidney_reference_Stewart.Rmd
Co-authored-by: Jaclyn Taroni <19534205+jaclyn-taroni@users.noreply.github.com>

@@
 params:
-  url: https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds
   padj_threshold: 0.05

### 2024-08-19 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
Co-authored-by: Jaclyn Taroni <19534205+jaclyn-taroni@users.noreply.github.com>

@@
 # We transfer the annotation to the pre-processed `Seurat`object as we don't want to keep changes done on the query by `RunAzimuth`
-metadata_to_trasfer <- data@meta.data[, c("predicted.compartment.score", "predicted.compartment", "predicted.cell_type.score", "predicted.cell_type")]
-srat <- AddMetaData(srat, metadata_to_trasfer, col.name = c("fetal_kidney_predicted.compartment.score", "fetal_kidney_predicted.compartment", "fetal_kidney_predicted.cell_type.score", "fetal_kidney_predicted.cell_type"))
+metadata_vec <- c("predicted.compartment.score", "predicted.compartment", "predicted.cell_type.score", "predicted.cell_type")
+metadata_to_trasfer <- data@meta.data[, metadata_vec]
+srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_kidney_", metadata_vec))
 ```
@@
 ```{r save}
-saveRDS(object = srat, file = file.path(output_dir, paste0("00b-fetal_kidey_label-transfer_",params$sample_id,".Rds")))
+saveRDS(object = srat, file = file.path(output_dir, paste0("00b-fetal_kidney_label-transfer_",params$sample_id,".Rds")))
 ```

### 2024-08-19 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
Changes to PR#706

@@
 DefaultAssay(srat) <- "RNA"
+options(future.globals.maxSize= 8912896000000)
 s <- RunAzimuth(srat, reference ="fetusref" )
@@
 # We transfer the annotation to the pre-processed `Seurat`iobject as we don't want to keep changes done on the query by `RunAzimuth`
-metadata_to_trasfer <- s@meta.data[, c("predicted.annotation.l1.score", "predicted.annotation.l1", "predicted.annotation.l2.score", "predicted.annotation.l2", "predicted.organ.score", "predicted.organ")]
-srat <- AddMetaData(srat, metadata_to_trasfer, col.name = c("fetal_full_predicted.annotation.l1.score", "fetal_full_predicted.annotation.l1", "fetal_full_predicted.annotation.l2.score", "fetal_full_predicted.annotation.l2", "fetal_full_predicted.organ.score", "fetal_full_predicted.organ"))
+metadata_vec <- c("predicted.annotation.l1.score", "predicted.annotation.l1", "predicted.annotation.l2.score", "predicted.annotation.l2", "predicted.organ.score", "predicted.organ")
+
+metadata_to_trasfer <- s@meta.data[, metadata_vec]
+
+srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_full_", metadata_vec))
 ```

### 2024-08-19 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
Changes to PR#706

--- 2024-08-16 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
@@
 ```{r run_azimuth, message=FALSE, warnings=FALSE}
+DefaultAssay(srat) <- "RNA"
+options(future.globals.maxSize= 8912896000000)
 
@@
 # We transfer the annotation to the pre-processed `Seurat`object as we don't want to keep changes done on the query by `RunAzimuth`
-metadata_to_trasfer <- data@meta.data[, c("predicted.compartment.score", "predicted.compartment", "predicted.cell_type.score", "predicted.cell_type")]
-srat <- AddMetaData(srat, metadata_to_trasfer, col.name = c("fetal_kidney_predicted.compartment.score", "fetal_kidney_predicted.compartment", "fetal_kidney_predicted.cell_type.score", "fetal_kidney_predicted.cell_type"))
+# We transfer the annotation to the pre-processed `Seurat`iobject as we don't want to keep changes done on the query by `RunAzimuth`
+metadata_vec <- c("predicted.compartment.score", "predicted.compartment", "predicted.cell_type.score", "predicted.cell_type")
+metadata_to_trasfer <- s@meta.data[, metadata_vec]
+srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_full_", metadata_vec))
+
 ```

### 2024-08-19 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd

--- 2024-08-19 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
@@
 metadata_vec <- c("predicted.compartment.score", "predicted.compartment", "predicted.cell_type.score", "predicted.cell_type")
-metadata_to_trasfer <- data@meta.data[, metadata_vec]
-srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_kidney_", metadata_vec))
+metadata_to_trasfer <- s@meta.data[, metadata_vec]
+srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_full_", metadata_vec))
 ```

### 2024-08-21 analyses/cell-type-wilms-tumor-06/00_run_workflow.R
few changes/typos PR #706

@@ -41,21 +41,21 @@ for (sample_id in metadata$scpca_sample_id) {
                     params = list(scpca_project_id = project_id, sample_id = sample_id),
                     output_format = "html_document",
                     output_file = paste0("01_seurat_processing_", sample_id, ".html"),
-                    output_dir = file.path(module_base, "notebook",  sample_id))
+                    output_dir = file.path(notebook_output_dir,  sample_id))
   
   # Label transfer from the Cao reference using Azimuth
   rmarkdown::render(input = file.path(module_base, "notebook_template", "02a_label-transfer_fetal_full_reference_Cao.Rmd"),
                     params = list(scpca_project_id = project_id, sample_id = sample_id),
                     output_format = "html_document",
                     output_file = paste0("02a_fetal_all_reference_Cao_", sample_id, ".html"),
-                    output_dir = file.path(module_base, "notebook", i))
+                    output_dir = file.path(notebook_output_dir, sample_id))
   
   # Label transfer from the Stewart reference using Seurat
   rmarkdown::render(input = file.path(module_base, "notebook_template", "02b_label-transfer_fetal_kidney_reference_Stewart.Rmd"),
                     params = list(scpca_project_id = project_id, sample_id = sample_id),
                     output_format = "html_document",
                     output_file = paste0("02b_fetal_kidney_reference_Stewart_", sample_id, ".html"),
-                    output_dir = file.path(module_base, "notebook", sample_id))
+                    output_dir = file.path(notebook_output_dir, sample_id))
   
 
   }

### 2024-08-28 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
few changes/typos PR #706
Update 02a_label-transfer_fetal_full_reference_Cao.Rmd

@@
 DefaultAssay(srat) <- "RNA"
-options(future.globals.maxSize= 8912896000000)
-s <- RunAzimuth(srat, reference ="fetusref" )
+options(future.globals.maxSize= 891289600000000)
+s <- RunAzimuth(srat, reference ="fetusref", k.weight = 30 )
 
@@
 ```{r save}
-saveRDS(object = srat, file = file.path(output_dir, paste0("00a-fetal_full_label-transfer_",params$sample_id,".Rds")))
+saveRDS(object = srat, file = file.path(output_dir, paste0("02a-fetal_full_label-transfer_",params$sample_id,".Rds")))
 ```

### 2024-08-28 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
Update 02a_label-transfer_fetal_full_reference_Cao.Rmd
test similar to https://github.com/satijalab/seurat/issues/4803

@@
 options(future.globals.maxSize= 891289600000000)
-s <- RunAzimuth(srat, reference ="fetusref", k.weight = 30 )
+s <- RunAzimuth(srat, reference ="fetusref", k.weight =  NA)
 

### 2024-08-28 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
Update 02a_label-transfer_fetal_full_reference_Cao.Rmd

@@
 options(future.globals.maxSize= 891289600000000)
-s <- RunAzimuth(srat, reference ="fetusref", k.weight =  NA)
+s <- RunAzimuth.Seurat(srat, reference ="fetusref")
 

### 2024-08-28 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
Update 02a_label-transfer_fetal_full_reference_Cao.Rmd

@@
 options(future.globals.maxSize= 891289600000000)
-s <- RunAzimuth.Seurat(srat, reference ="fetusref")
+s <- Azimuth::RunAzimuth(srat, reference ="fetusref")
 

### 2024-08-28 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
Update 02a_label-transfer_fetal_full_reference_Cao.Rmd

@@
 ```{r packages, message=FALSE, warning=FALSE}
-library(SingleCellExperiment)
-
 library("Seurat")
+library(SeuratData)
 library(sctransform)
 library(Azimuth)
-library(SeuratData)
-
-library(assertthat)
-library(viridis)
 library(SCpubr)
@@
 library(patchwork)
+library(SingleCellExperiment)
 

### 2024-08-28 analyses/cluster-utils/notebook-template.Rmd
add module with renv and some readme filled in

--- 2024-04-24 analyses/ewings-cell-type/notebook-template.Rmd
@@
 ---
-title: "ewings-cell-type Analysis"
+title: "cluster-utils Analysis"
 author:
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
+module_base <- file.path(repository_base, "analyses", "cluster-utils")
 ```

### 2024-08-29 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
Try to debug k.weight

@@
   seed: 12345
+  is_ci: FALSE
 output:
@@
                       warnings=FALSE)
+```
+
+Determine whether or not this notebook is being run in GitHub Actions.
+
+```{r setup_ci}
+# Adapted from: https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/e854738b57b16c3725673ad63b4ffd20ec48ad19/analyses/gene-set-enrichment-analysis/02-model-gsea.Rmd#L44-L52
+# Assigning params$is_ci to running_in_ci avoids a locked binding error
+running_in_ci <- params$is_ci
+
+# Are we testing? In case of a non 0/1 number, we recast as logical, and then ensure logical.
+if (running_in_ci %in% c(0,1)) running_in_ci <- as.logical(running_in_ci)
+if (!(is.logical(running_in_ci))) {
+  stop("\n\nERROR: The parameter `is_ci` should be FALSE/TRUE (or 0/1).")
+}
 ```
@@
 options(future.globals.maxSize= 891289600000000)
-s <- Azimuth::RunAzimuth(srat, reference ="fetusref")
+
+# If running in CI, there will be fewer cells and we must set k.weight in
+# RunAzimuth() to a lower value
+if (running_in_ci) {
+  k_weight <- 2
+} else {
+  k_weight <- 50  # The default value
+}
+
+print(k_weight)
+s <- Azimuth::RunAzimuth(srat, reference ="fetusref", k.weight = k_weight)
 

### 2024-08-29 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
Try to debug k.weight

@@
                       )
+```
+
+Determine whether or not this notebook is being run in GitHub Actions.
+
+```{r setup_ci}
+# Adapted from: https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/e854738b57b16c3725673ad63b4ffd20ec48ad19/analyses/gene-set-enrichment-analysis/02-model-gsea.Rmd#L44-L52
+# Assigning params$is_ci to running_in_ci avoids a locked binding error
+running_in_ci <- params$is_ci
+
+# Are we testing? In case of a non 0/1 number, we recast as logical, and then ensure logical.
+if (running_in_ci %in% c(0,1)) running_in_ci <- as.logical(running_in_ci)
+if (!(is.logical(running_in_ci))) {
+  stop("\n\nERROR: The parameter `is_ci` should be FALSE/TRUE (or 0/1).")
+}
 ```
@@
 # open the processed rds object
-srat <- readRDS(file.path(data_dir, paste0("00a-fetal_full_label-transfer_",params$sample_id,".Rds")))
+srat <- readRDS(file.path(data_dir, paste0("02a-fetal_full_label-transfer_",params$sample_id,".Rds")))
 ```
@@
 
-data <-  RunAzimuth(srat, path_to_reference, assay = 'RNA')
+# If running in CI, there will be fewer cells and we must set k.weight in
+# RunAzimuth() to a lower value
+if (running_in_ci) {
+  k_weight <- 10
+} else {
+  k_weight <- 50  # The default value
+}
+
+data <-  RunAzimuth(srat, path_to_reference, assay = 'RNA', k.weight = k_weight)
 # We transfer the annotation to the pre-processed `Seurat`object as we don't want to keep changes done on the query by `RunAzimuth`
 metadata_vec <- c("predicted.compartment.score", "predicted.compartment", "predicted.cell_type.score", "predicted.cell_type")
-metadata_to_trasfer <- s@meta.data[, metadata_vec]
-srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_full_", metadata_vec))
+metadata_to_trasfer <- data@meta.data[, metadata_vec]
+srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_kidney_", metadata_vec))
 ```
@@
 ```{r save}
-saveRDS(object = srat, file = file.path(output_dir, paste0("00b-fetal_kidney_label-transfer_",params$sample_id,".Rds")))
+saveRDS(object = srat, file = file.path(output_dir, paste0("02b-fetal_kidney_label-transfer_",params$sample_id,".Rds")))
 ```

### 2024-08-29 analyses/cell-type-wilms-tumor-06/00_run_workflow.R

@@ -1,5 +1,27 @@
 #!/usr/bin/env Rscript
 
+# USAGE:
+# Rscript 00_run_workflow.R
+#
+# USAGE in CI:
+# Rscript 00_run_workflow.R --testing
+
+# Set up options ----------------------------------------------------------------
+
+library(optparse)
+
+option_list <- list(
+  make_option(
+    opt_str = c("--testing"),
+    type = "logical",
+    action = "store_true",
+    help = "Use this flag when running on test data"
+  )
+)
+
+opts <- parse_args(OptionParser(option_list = option_list))
+running_ci <- opts$testing
+
 # Run the Label transfer from two fetal references ------------------------------
 
 # get list of samples in the library --------------------------------------------
@@ -12,7 +34,7 @@ metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
 module_base <- file.path(root_dir, "analyses", "cell-type-wilms-tumor-06")
 
 # Download and create the fetal kidney reference (Stewart et al) ----------
-# source(file.path(module_base,"scripts", "download-and-create-fetal-kidney-ref.R"))
+system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "download-and-create-fetal-kidney-ref.R")))
 
 # Characterize the two fetal references -----------------------------------------
 
@@ -29,7 +51,7 @@ rmarkdown::render(input = file.path(notebook_template_dir, "00b_characterize_fet
 
 # Run the workflow for (all) samples in the project -----------------------------
 for (sample_id in metadata$scpca_sample_id) {
- 
+  
   # create a directory to save the pre-processed and labeled `Seurat` objects
   dir.create(file.path(module_base, "results", sample_id))
   # create a directory to save the notebooks
@@ -45,20 +67,19 @@ for (sample_id in metadata$scpca_sample_id) {
   
   # Label transfer from the Cao reference using Azimuth
   rmarkdown::render(input = file.path(module_base, "notebook_template", "02a_label-transfer_fetal_full_reference_Cao.Rmd"),
-                    params = list(scpca_project_id = project_id, sample_id = sample_id),
+                    params = list(scpca_project_id = project_id, sample_id = sample_id, is_ci = running_ci),
                     output_format = "html_document",
                     output_file = paste0("02a_fetal_all_reference_Cao_", sample_id, ".html"),
                     output_dir = file.path(notebook_output_dir, sample_id))
   
   # Label transfer from the Stewart reference using Seurat
[…]

### 2024-08-29 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
Debug RunAzimuth

--- 2024-08-28 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
@@
 
+### Define function
+
+Here we define a function `RunAzimuth_small` to run the `Azimuth` label transfer on small samples (< 100 cells).
+
+We take the `source` code of the function, change some default and add the `ksmooth`parameter.
+
+We add the parameter `ksmooth`.
+`ksmooth` is not a parameter of `RunAzimuth` but need to be define as in `FindTransferAnchors`, `mapping.score.k` is reinitialized to `ksmooth`. 
+If not defined here, `MappingScore`will fail
+
+We set up the `k.weight` default to 10 (instead of 50).
+
+We set up the `mapping.score.k` default to 80 (instead of 100).
+
+```{r RunAzimuth_small, message=FALSE, warning=FALSE}
+
+RunAzimuth_small <- function(query = query,
+reference =reference,
+query.modality = "RNA",
+  annotation.levels = NULL,
+  umap.name = "ref.umap",
+  do.adt = FALSE,
+  verbose = TRUE,
+  assay = NULL,
+  k.weight = 10, # or any value lower than the number of anchors/cells
+  n.trees = 20,
+  mapping.score.k = 80,
+  ksmooth = 80 
+){
+    if (dir.exists(reference)) {
+      reference <- LoadReference(reference)$map
+    } else {
+      reference <- tolower(reference)
+      if (reference %in% InstalledData()$Dataset) {
+        # only get the `map` object since no plotting is performed
+        reference <- LoadData(reference, type = "azimuth")$map
+      } else if (reference %in% AvailableData()$Dataset) {
+        InstallData(reference)
+        # only get the `map` object since no plotting is performed
+        reference <- LoadData(reference, type = "azimuth")$map
+      } else {
+        possible.references <- AvailableData()$Dataset[grepl("*ref", AvailableData()$Dataset)]
+        print("Choose one of:")
+        print(possible.references)
+        stop(paste("Could not find a reference for", reference))
+      }
+      # handle expected new parameters in uwot models beginning in v0.1.13
+      if (!"num_precomputed_nns" %in% names(Misc(reference[["refUMAP"]])$model)) {
+        Misc(reference[["refUMAP"]], slot="model")$num_precomputed_nns <- 1
+      }
+      key.pattern = "^[^_]*_"
+      new.colnames <- gsub(pattern = key.pattern, 
+                           replacement = Key(reference[["refDR"]]), 
+                           x = colnames(Loadings(
+                             object = reference[["refDR"]],
+                             projected = FALSE)))
+      colnames(Loadings(object = reference[["refDR"]], 
+                        projected = FALSE)) <- new.colnames
+    }
+    dims <- as.double(slot(reference, "neighbors")$refdr.annoy.neighbors@alg.info$ndim)
+    if (isTRUE(do.adt) && !("ADT" %in% Assays(reference))) {
+      warning("Cannot impute an ADT assay because the reference does not have antibody data")
+      do.adt = FALSE
+    }
+    reference.version <- ReferenceVersion(reference)
+    azimuth.version <- as.character(packageVersion(pkg = "Azimuth"))
+    seurat.version <- as.character(packageVersion(pkg = "Seurat"))
+    meta.data <- names(slot(reference, "meta.data"))
+
+    # is annotation levels are not specify, gather all levels of annotation
+    if (is.null(annotation.levels)) {
+      annotation.levels <- names(slot(object = reference, name = "meta.data"))
+      annotation.levels <- annotation.levels[!grepl(pattern = "^nCount", x = annotation.levels)]
+      annotation.levels <- annotation.levels[!grepl(pattern = "^nFeature", x = annotation.levels)]
+      annotation.levels <- annotation.levels[!grepl(pattern = "^ori", x = annotation.levels)]
+    }
+
+    # Change the file path based on where the query file is located on your system.
+    query <- ConvertGeneNames(
+      object = query,
+      reference.names = rownames(x = reference),
+      homolog.table = 'https://seurat.nygenome.org/azimuth/references/homologs.rds'
+    )
+
+    # Calculate nCount_RNA and nFeature_RNA if the query does not
+    # contain them already
+    if (!all(c("nCount_RNA", "nFeature_RNA") %in% c(colnames(x = query[[]])))) {
+      calcn <- as.data.frame(x = Seurat:::CalcN(object = query[["RNA"]]))
+      colnames(x = calcn) <- paste(
+        colnames(x = calcn),
+        assay,
+        sep = '_'
+      )
+      query <- AddMetaData(
+        object = query,
+        metadata = calcn
+      )
+      rm(calcn)
+    }
+
+    # Calculate percent mitochondrial genes if the query contains genes
+    # matching the regular expression "^MT-"
+    if (any(grepl(pattern = '^MT-', x = rownames(x = query)))) {
+      query <- PercentageFeatureSet(
+        object = query,
+        pattern = '^MT-',
+        col.name = 'percent.mt',
+        assay = assay
+      )
+    }
+    # Find anchors between query and reference
+    anchors <- FindTransferAnchors(
+      reference = reference,
+      query = query,
+      k.filter =  NA,
+      reference.neighbors = "refdr.annoy.neighbors",
+      reference.assay = "refAssay",
+      query.assay = assay,
+      reference.reduction = "refDR",
+      normalization.method = "SCT",
+      features = rownames(Loadings(reference[["refDR"]])),
+      dims = 1:dims,
+      n.trees = n.trees,
+      mapping.score.k = mapping.score.k,
+      verbose = verbose
+    )
+    # Transferred labels are in metadata columns named "predicted.*"
+    # The maximum prediction score is in a metadata column named "predicted.*.score"
+    # The prediction scores for each class are in an assay named "prediction.score.*"
+    # The imputed assay is named "impADT" if computed
+    refdata <- lapply(X = annotation.levels, function(x) {
+      reference[[x, drop = TRUE]]
+    })
+    names(x = refdata) <- annotation.levels
+
+    if (isTRUE(do.adt)) {
+      refdata[["impADT"]] <- GetAssayData(
+        object = reference[["ADT"]],
+        slot = "data"
+      )
+    }
+
+    query <- TransferData(
+      reference = reference,
+      query = query,
+      query.assay = assay,
+      dims = 1:dims,
+      anchorset = anchors,
+      refdata = refdata,
+      n.trees = 20,
+      store.weights = TRUE,
+      k.weight = k.weight,
+      verbose = verbose
+    )
+    # Calculate the embeddings of the query data on the reference SPCA
+    query <- IntegrateEmbeddings(
+      anchorset = anchors,
+      reference = reference,
+      query = query,
+      query.assay = assay,
+      reductions = "pcaproject",
+      reuse.weights.matrix = TRUE,
+      verbose = verbose
+    )
+    # Calculate the query neighbors in the reference
+    # with respect to the integrated embeddings
+    query[["query_ref.nn"]] <- FindNeighbors(
+      object = Embeddings(reference[["refDR"]]),
+      query = Embeddings(query[["integrated_dr"]]),
+      return.neighbor = TRUE,
+      l2.norm = TRUE,
+      verbose = verbose
+    )
+    # The reference used in the app is downsampled compared to the reference on which
+    # the UMAP model was computed. This step, using the helper function NNTransform,
+    # corrects the Neighbors to account for the downsampling.
+    query <- Azimuth:::NNTransform(
+      object = query,
+      meta.data = reference[[]]
+    )
+    # Project the query to the reference UMAP.
+    query[[umap.name]] <- RunUMAP(
+      object = query[["query_ref.nn"]],
+      reduction.model = reference[["refUMAP"]],
+      reduction.key = 'UMAP_',
+      verbose = verbose
+    )
+
+    # Calculate mapping score and add to metadata
+    query <- AddMetaData(
+      object = query,
+      metadata = MappingScore(anchors = anchors, ndim = dims,  ksmooth = 80),
+      col.name = "mapping.score"
+    )
+
+}
+
+```
+
 ### Base directories
@@
 options(future.globals.maxSize= 891289600000000)
-s <- Azimuth::RunAzimuth(srat, reference ="fetusref")
+
+if(dim(srat)[2] < 100){
+ s <- RunAzimuth_small(srat, reference ="fetusref", k.weight = 10, mapping.score.k = 80, ksmooth = 80)
+} else {
+ s <- RunAzimuth(srat, reference ="fetusref")
+
+}
 

### 2024-08-29 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
Debug RunAzimuth

--- 2024-08-19 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
@@
 DefaultAssay(srat) <- "RNA"
-options(future.globals.maxSize= 8912896000000)
+options(future.globals.maxSize= 891289600000000)
 
-data <-  RunAzimuth(srat, path_to_reference, assay = 'RNA')
-# We transfer the annotation to the pre-processed `Seurat`object as we don't want to keep changes done on the query by `RunAzimuth`
+if(dim(srat)[2] < 100){
+ s <- RunAzimuth_small(srat, reference =path_to_reference, assay = 'RNA',k.weight = 10, mapping.score.k = 80, ksmooth = 80)
+} else {
+ s <- RunAzimuth(srat, reference =path_to_reference,assay = 'RNA' )
+
+}
+
+# We transfer the annotation to the pre-processed `Seurat`iobject as we don't want to keep changes done on the query by `RunAzimuth`
 metadata_vec <- c("predicted.compartment.score", "predicted.compartment", "predicted.cell_type.score", "predicted.cell_type")
-metadata_to_trasfer <- s@meta.data[, metadata_vec]
-srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_full_", metadata_vec))
+metadata_to_trasfer <- data@meta.data[, metadata_vec]
+srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_kidney_", metadata_vec))
 ```

### 2024-08-29 analyses/cell-type-wilms-tumor-06/00_run_workflow.R

@@ -67,14 +67,14 @@ for (sample_id in metadata$scpca_sample_id) {
   
   # Label transfer from the Cao reference using Azimuth
   rmarkdown::render(input = file.path(module_base, "notebook_template", "02a_label-transfer_fetal_full_reference_Cao.Rmd"),
-                    params = list(scpca_project_id = project_id, sample_id = sample_id, is_ci = running_ci),
+                    params = list(scpca_project_id = project_id, sample_id = sample_id),
                     output_format = "html_document",
                     output_file = paste0("02a_fetal_all_reference_Cao_", sample_id, ".html"),
                     output_dir = file.path(notebook_output_dir, sample_id))
   
   # Label transfer from the Stewart reference using Seurat
   rmarkdown::render(input = file.path(module_base, "notebook_template", "02b_label-transfer_fetal_kidney_reference_Stewart.Rmd"),
-                    params = list(scpca_project_id = project_id, sample_id = sample_id, is_ci = running_ci),
+                    params = list(scpca_project_id = project_id, sample_id = sample_id),
                     output_format = "html_document",
                     output_file = paste0("02b_fetal_kidney_reference_Stewart_", sample_id, ".html"),
                     output_dir = file.path(notebook_output_dir, sample_id))

### 2024-08-29 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
Try debug RunAzimuth

@@
 
-if(dim(srat)[2] < 100){
+if(dim(srat)[2] < 500){
  s <- RunAzimuth_small(srat, reference ="fetusref", k.weight = 10, mapping.score.k = 80, ksmooth = 80)
@@
 }
-
+print(dim(srat)[2])
 # We transfer the annotation to the pre-processed `Seurat`iobject as we don't want to keep changes done on the query by `RunAzimuth`

### 2024-08-29 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
correct ERROR in notebook 02b_

@@
   scpca_project_id: "SCPCP000006"
-  sample_id: "SCPCS000176"
+  sample_id: "SCPCS000168"
   seed: 12345
@@
 
-if(dim(srat)[2] < 100){
+if(dim(srat)[2] < 500){
  s <- RunAzimuth_small(srat, reference =path_to_reference, assay = 'RNA',k.weight = 10, mapping.score.k = 80, ksmooth = 80)
@@
 }
+print(dim(srat)[2])
 
@@
 metadata_vec <- c("predicted.compartment.score", "predicted.compartment", "predicted.cell_type.score", "predicted.cell_type")
-metadata_to_trasfer <- data@meta.data[, metadata_vec]
+metadata_to_trasfer <- s@meta.data[, metadata_vec]
 srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_kidney_", metadata_vec))

### 2024-08-29 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
Try debug RunAzimuth

--- 2024-08-29 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
@@
 
-if(dim(srat)[2] < 100){
+print(dim(srat)[2])
+# this would need to be change for CI or whole sample
  s <- RunAzimuth_small(srat, reference ="fetusref", k.weight = 10, mapping.score.k = 80, ksmooth = 80)
-} else {
- s <- RunAzimuth(srat, reference ="fetusref")
-
-}
 

### 2024-08-29 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
Try debug RunAzimuth

@@
 
-if(dim(srat)[2] < 500){
+# this would need to be change for CI or whole sample
  s <- RunAzimuth_small(srat, reference =path_to_reference, assay = 'RNA',k.weight = 10, mapping.score.k = 80, ksmooth = 80)
-} else {
- s <- RunAzimuth(srat, reference =path_to_reference,assay = 'RNA' )
-
-}
-print(dim(srat)[2])
 

### 2024-08-29 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
Expand timeout for reference download
Try debug RunAzimuth

@@
     )
+
+    print("FindTransferAnchors done")
     # Transferred labels are in metadata columns named "predicted.*"
@@
     )
+        print("TransferData done")
+
     # Calculate the embeddings of the query data on the reference SPCA
@@
     )
+            print("IntegrateEmbeddings done")
+
     # Calculate the query neighbors in the reference
@@
     )
+                print("FindNeighbors done")
+
     # The reference used in the app is downsampled compared to the reference on which
@@
     )
+
+                    print("NNTransform done")
+
     # Project the query to the reference UMAP.
@@
     )
+                        print("RunUMAP done")
 
     # Calculate mapping score and add to metadata
-    query <- AddMetaData(
-      object = query,
-      metadata = MappingScore(anchors = anchors, ndim = dims,  ksmooth = 80),
-      col.name = "mapping.score"
-    )
+  #  query <- AddMetaData(
+   #   object = query,
+    #  metadata = MappingScore(anchors = anchors, ndim = dims,  ksmooth = mapping.score.k, kanchors = (k.weight+1)),
+     # col.name = "mapping.score"
+    #)
+
+    return(query)
 
@@
 ```{r path_to_reference}
+options(timeout=600)
+
 #Check the names of the Azimuth available data and reference

### 2024-09-03 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd

--- 2024-08-29 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
@@
 
-### Define function
-
-Here we define a function `RunAzimuth_small` to run the `Azimuth` label transfer on small samples (< 100 cells).
-
-We take the `source` code of the function, change some default and add the `ksmooth`parameter.
-
-We add the parameter `ksmooth`.
-`ksmooth` is not a parameter of `RunAzimuth` but need to be define as in `FindTransferAnchors`, `mapping.score.k` is reinitialized to `ksmooth`. 
-If not defined here, `MappingScore`will fail
-
-We set up the `k.weight` default to 10 (instead of 50).
-
-We set up the `mapping.score.k` default to 80 (instead of 100).
-
-```{r RunAzimuth_small, message=FALSE, warning=FALSE}
-
-RunAzimuth_small <- function(query = query,
-reference =reference,
-query.modality = "RNA",
-  annotation.levels = NULL,
-  umap.name = "ref.umap",
-  do.adt = FALSE,
-  verbose = TRUE,
-  assay = NULL,
-  k.weight = 10, # or any value lower than the number of anchors/cells
-  n.trees = 20,
-  mapping.score.k = 80,
-  ksmooth = 80 
-){
-    if (dir.exists(reference)) {
-      reference <- LoadReference(reference)$map
-    } else {
-      reference <- tolower(reference)
-      if (reference %in% InstalledData()$Dataset) {
-        # only get the `map` object since no plotting is performed
-        reference <- LoadData(reference, type = "azimuth")$map
-      } else if (reference %in% AvailableData()$Dataset) {
-        InstallData(reference)
-        # only get the `map` object since no plotting is performed
-        reference <- LoadData(reference, type = "azimuth")$map
-      } else {
-        possible.references <- AvailableData()$Dataset[grepl("*ref", AvailableData()$Dataset)]
-        print("Choose one of:")
-        print(possible.references)
-        stop(paste("Could not find a reference for", reference))
-      }
-      # handle expected new parameters in uwot models beginning in v0.1.13
-      if (!"num_precomputed_nns" %in% names(Misc(reference[["refUMAP"]])$model)) {
-        Misc(reference[["refUMAP"]], slot="model")$num_precomputed_nns <- 1
-      }
-      key.pattern = "^[^_]*_"
-      new.colnames <- gsub(pattern = key.pattern, 
-                           replacement = Key(reference[["refDR"]]), 
-                           x = colnames(Loadings(
-                             object = reference[["refDR"]],
-                             projected = FALSE)))
-      colnames(Loadings(object = reference[["refDR"]], 
-                        projected = FALSE)) <- new.colnames
-    }
-    dims <- as.double(slot(reference, "neighbors")$refdr.annoy.neighbors@alg.info$ndim)
-    if (isTRUE(do.adt) && !("ADT" %in% Assays(reference))) {
-      warning("Cannot impute an ADT assay because the reference does not have antibody data")
-      do.adt = FALSE
-    }
-    reference.version <- ReferenceVersion(reference)
-    azimuth.version <- as.character(packageVersion(pkg = "Azimuth"))
-    seurat.version <- as.character(packageVersion(pkg = "Seurat"))
-    meta.data <- names(slot(reference, "meta.data"))
-
-    # is annotation levels are not specify, gather all levels of annotation
-    if (is.null(annotation.levels)) {
-      annotation.levels <- names(slot(object = reference, name = "meta.data"))
-      annotation.levels <- annotation.levels[!grepl(pattern = "^nCount", x = annotation.levels)]
-      annotation.levels <- annotation.levels[!grepl(pattern = "^nFeature", x = annotation.levels)]
-      annotation.levels <- annotation.levels[!grepl(pattern = "^ori", x = annotation.levels)]
-    }
-
-    # Change the file path based on where the query file is located on your system.
-    query <- ConvertGeneNames(
-      object = query,
-      reference.names = rownames(x = reference),
-      homolog.table = 'https://seurat.nygenome.org/azimuth/references/homologs.rds'
-    )
-
-    # Calculate nCount_RNA and nFeature_RNA if the query does not
-    # contain them already
-    if (!all(c("nCount_RNA", "nFeature_RNA") %in% c(colnames(x = query[[]])))) {
-      calcn <- as.data.frame(x = Seurat:::CalcN(object = query[["RNA"]]))
-      colnames(x = calcn) <- paste(
-        colnames(x = calcn),
-        assay,
-        sep = '_'
-      )
-      query <- AddMetaData(
-        object = query,
-        metadata = calcn
-      )
-      rm(calcn)
-    }
-
-    # Calculate percent mitochondrial genes if the query contains genes
-    # matching the regular expression "^MT-"
-    if (any(grepl(pattern = '^MT-', x = rownames(x = query)))) {
-      query <- PercentageFeatureSet(
-        object = query,
-        pattern = '^MT-',
-        col.name = 'percent.mt',
-        assay = assay
-      )
-    }
-    # Find anchors between query and reference
-    anchors <- FindTransferAnchors(
-      reference = reference,
-      query = query,
-      k.filter =  NA,
-      reference.neighbors = "refdr.annoy.neighbors",
-      reference.assay = "refAssay",
-      query.assay = assay,
-      reference.reduction = "refDR",
-      normalization.method = "SCT",
-      features = rownames(Loadings(reference[["refDR"]])),
-      dims = 1:dims,
-      n.trees = n.trees,
-      mapping.score.k = mapping.score.k,
-      verbose = verbose
-    )
-    # Transferred labels are in metadata columns named "predicted.*"
-    # The maximum prediction score is in a metadata column named "predicted.*.score"
-    # The prediction scores for each class are in an assay named "prediction.score.*"
-    # The imputed assay is named "impADT" if computed
-    refdata <- lapply(X = annotation.levels, function(x) {
-      reference[[x, drop = TRUE]]
-    })
-    names(x = refdata) <- annotation.levels
-
-    if (isTRUE(do.adt)) {
-      refdata[["impADT"]] <- GetAssayData(
-        object = reference[["ADT"]],
-        slot = "data"
-      )
-    }
-
-    query <- TransferData(
-      reference = reference,
-      query = query,
-      query.assay = assay,
-      dims = 1:dims,
-      anchorset = anchors,
-      refdata = refdata,
-      n.trees = 20,
-      store.weights = TRUE,
-      k.weight = k.weight,
-      verbose = verbose
-    )
-    # Calculate the embeddings of the query data on the reference SPCA
-    query <- IntegrateEmbeddings(
-      anchorset = anchors,
-      reference = reference,
-      query = query,
-      query.assay = assay,
-      reductions = "pcaproject",
-      reuse.weights.matrix = TRUE,
-      verbose = verbose
-    )
-    # Calculate the query neighbors in the reference
-    # with respect to the integrated embeddings
-    query[["query_ref.nn"]] <- FindNeighbors(
-      object = Embeddings(reference[["refDR"]]),
-      query = Embeddings(query[["integrated_dr"]]),
-      return.neighbor = TRUE,
-      l2.norm = TRUE,
-      verbose = verbose
-    )
-    # The reference used in the app is downsampled compared to the reference on which
-    # the UMAP model was computed. This step, using the helper function NNTransform,
-    # corrects the Neighbors to account for the downsampling.
-    query <- Azimuth:::NNTransform(
-      object = query,
-      meta.data = reference[[]]
-    )
-    # Project the query to the reference UMAP.
-    query[[umap.name]] <- RunUMAP(
-      object = query[["query_ref.nn"]],
-      reduction.model = reference[["refUMAP"]],
-      reduction.key = 'UMAP_',
-      verbose = verbose
-    )
-
-    # Calculate mapping score and add to metadata
-    query <- AddMetaData(
-      object = query,
-      metadata = MappingScore(anchors = anchors, ndim = dims,  ksmooth = 80),
-      col.name = "mapping.score"
-    )
-
-}
-
-```
-
 ### Base directories
@@
 DefaultAssay(srat) <- "RNA"
-options(future.globals.maxSize= 891289600000000)
+options(future.globals.maxSize= 8912896000000)
 
-if(dim(srat)[2] < 100){
- s <- RunAzimuth_small(srat, reference =path_to_reference, assay = 'RNA',k.weight = 10, mapping.score.k = 80, ksmooth = 80)
-} else {
- s <- RunAzimuth(srat, reference =path_to_reference,assay = 'RNA' )
-
-}
-
-# We transfer the annotation to the pre-processed `Seurat`iobject as we don't want to keep changes done on the query by `RunAzimuth`
+data <-  RunAzimuth(srat, path_to_reference, assay = 'RNA')
+# We transfer the annotation to the pre-processed `Seurat`object as we don't want to keep changes done on the query by `RunAzimuth`
 metadata_vec <- c("predicted.compartment.score", "predicted.compartment", "predicted.cell_type.score", "predicted.cell_type")

### 2024-09-04 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Add clustering exploration

--- 2024-08-14 analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
@@
 ---
-title: "Exploration of clustering results for `r params$sample_id`"
-author: maud-p
-date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
+title: "Clustering exploration for `r params$sample_id`"
+author: "Maud PLASCHKA"
+date: '2024-08-13'
+params:
+  scpca_project_id: "SCPCP000006"
+  sample_id: "SCPCS000176"
+  seed: 12345
+  padj_thershold: 0.05  # Threshold for adjusted p-value
+  lfc_threshold: 1      # Threshold for log fold change
+  rate1_threshold: 0.5  # Threshold for rate (percentage) of cells expressing the marker genes in group1
 output: 
@@
     highlight: pygments
-    df_print: kable
-    self_contained: yes
-    mode: selfcontained 
-params:
-  current: "2024-07-08"
-  scpca_project_id: "SCPCP000006"
-  sample_id: "SCPCS000176"
-  seed: 12345
-  padj_thershold: 0.05  # Threshold for adjusted p-value
-  lfc_threshold: 1      # Threshold for log fold change
-  rate1_threshold: 0.5  # Threshold for rate (percentage) of cells expressing the marker genes in group1
+    df_print: paged
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(
-	echo = TRUE,
-	message = FALSE,
-	warnings = FALSE
+knitr::opts_chunk$set(echo = TRUE,
+                      message=FALSE,
+                      warnings=FALSE
                       )
@@
 
-This a template for an analysis notebook using `RMarkdown`.
+The aim is to explore the clustering and label transfers for the sample `param$sample_id` =
 
-In this notebook, we will perform the `Seurat` workflow [normalization --> reduction --> clustering] and explore the clustering results.
+```{r}
+params$sample_id
+```
 
-This correspond to the step 2 of the proposed analysis: clustering of cells across a set of parameters for few samples
+In order to explore the clustering results, we look into some marker genes, pathways enrichment and label transfer.
 
-- https://github.com/AlexsLemonade/OpenScPCA-analysis/discussions/635
-- https://github.com/AlexsLemonade/OpenScPCA-analysis/issues/671
-
-In order to assess the clustering quality, we look into some marker genes, pathways enrichment and label transfer.
 This approach would provide us a rapid idea of the quality of the clustering:
@@
 - if no marker genes are found, the quality of cells in the clusters might be impaired (high mitochondrial content? ribosomal content?)
-
-### Description of the analysis workflow
-
-We perform the following analysis to assess for the quality of clustering:
-
-[1] We perform some quality check to assess any QC-induced clustering (nFeature, nCount, percent.mito).
-
-[2] We add cell cycle information, as we know that in a specific cell cycle state, the transcriptional program is mostly/exclusively related to cell cycle genes and the identity of cells is difficult to determine.
-We expect these cells to cluster together in a cluster of proliferating cells.
-
-[3] We look at specific marker genes that we reported in the table marker.sets/CellType_metadata.csv to check the relevance of the clustering.
-
-[4] We look at specific pathways to check the relevance of the clustering.
-
-[5] We run `DElegate::FindAllMarkers2` to find markers of the different clusters and manually check if they do make sense. 
-`DElegate::FindAllMarkers2` is an improved version of `Seurat::FindAllMarkers` based on pseudobulk differential expression method.
-
-[6] We perform enrichment analysis of marker genes for each Seurat clusters. 
-We defined all the genes from the Seurat object as the universe and used the MSigDB gene sets.
-
-[7] We plot `pca`/`umap` reduction grouping with available annotations (`singler_`, `cellassign_`). 
-We expect at least immune cells to be correctly label and fall into a few set of clusters.
-
-[ ] Next step, aim for a next PR: We will transfer annotations from two human fetal (kidney) references (Stewart et al., Cao et al.). 
-We plot pca/umap reduction grouping with latest labels. 
-We expect it to be the most representative of the cell types in the sample.
 
@@
 ```{r packages, message=FALSE, warning=FALSE}
-library(SingleCellExperiment)
+library("Seurat")
+library(sctransform)
 
-library("Seurat")
-
-#library(Azimuth)            # Will be required once we add the labl transfer from Azimuth fetal reference
-#library(SeuratData)         # Will be required once we add the labl transfer from Azimuth fetal reference
-
-library(viridis)            # for plotting with viridis colors
 library(SCpubr)             # for plotting 
@@
 library(patchwork)
-
 library(msigdbr)
@@
 library(clusterProfiler)
-
 library(org.Hs.eg.db)
+library(ggalluvial)
 
@@
 
-```{r base paths}
+```{r base paths, eval=TRUE}
 # The base path for the OpenScPCA repository, found by its (hidden) .git directory
@@
 
-### Selected parameters
+### Input files
 
-Please note: to keep the notebook as straight as possible, we decided to show the analysis for the selected set of parameters:
+In this notebook, we are working with the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`.
+We work with the pre-processed and labeled `Seurat` object that is the output of `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd` saved in the `results` directory.
 
-- `SCTransform` (default parameters)
-- `RunPCA` (default) 
-- `RunUMAP` (dims = 1:50) 
-- `FindNeighbors` (dims = 1:50) 
-- `FindClusters` (default)
+```{r path_to_query}
+result_dir <- file.path(module_base, "results", params$sample_id)
+```
 
-Note: Other parameters have been previously tested, but we would like to show in the following report that the one selected is performing good.
+To explore the clustering results, we look into some marker genes reported here:
+
+```{r marker_genes, fig.width=12, fig.height=6, out.width='100%'}
+CellType_metadata <- read_csv(file.path(module_base, "marker-sets", "CellType_metadata.csv"))
+
+DT::datatable(CellType_metadata, caption = ("CellType_metadata"), 
+              extensions = 'Buttons', 
+              options = list(  dom = 'Bfrtip',
+                               buttons = c( 'csv', 'excel')))
+```
+
+### Output file
+
+Reports will be saved in the `notebook` directory. 
+The pre-processed and annotated `Seurat` object per samples are saved in the `result` folder.
+
+## Analysis
+
+### Load the pre-processed `Seurat` object
+```{r load, message=FALSE, warning=FALSE}
+# open the processed rds object
+srat <- readRDS(file.path(result_dir, paste0("02b-fetal_kidney_label-transfer_",params$sample_id,".Rds")))
+DefaultAssay(srat) <- "SCT"
+
+```
 
@@
 ```{r }
-
 visualize_feature <- function(object, features, group.by = "seurat_clusters"){
@@
                                               legend.position = "right") + ggtitle(feature_symbol$SYMBOL)
-
                   b <- SCpubr::do_ViolinPlot(srat, 
@@
                                              group.by = group.by, 
-                                             legend.position = "none") + ylab(feature_symbol$SYMBOL)
+                                             legend.position = "none",
+                                             assay = "SCT") + ylab(feature_symbol$SYMBOL)
 
                   return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
-
 }
@@
 ```{r }
-
 visualize_metadata <- function(object, meta, group.by){
@@
                                               legend.position = "right") + ggtitle(meta)
-
     b <- SCpubr::do_ViolinPlot(srat, 
@@
   d <- SCpubr::do_DimPlot(object, reduction="umap", group.by = group.by, label = TRUE, repel = TRUE) + ggtitle(paste0(meta," - umap")) + theme(text=element_text(size=18))
-
   b <- SCpubr::do_BarPlot(sample = object,
@@
                          ggtitle("% cells")+
-                         xlab(file) +
+                         xlab(print(group.by)) +
                          theme(text=element_text(size=18))
@@
 ```{r }
-
 MSigDB_score <- function(object, category, gs_name, name){
@@
         set <- msigdbr(species = "human", category = category)
-
         set_list <- set %>% 
@@
                     dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()
-
         set_list <- list(set_list$human_ensembl_gene)
-
         object <- AddModuleScore(object, features = set_list, name = name)
-
 return(object)
 }
-
 ```
@@
 ```{r }
-
 Enrichment_plot <- function(category, signatures, background){
-
     ## define genesets
+   tryCatch( expr =  {
                gene_set <- msigdbr(species = "human", category = category)
                msigdbr_set <-  gene_set %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
-
                cclust<-compareCluster(geneCluster = signatures, 
@@
                return(d)  
+   },
+   error = function(e) { print("No enrichment") } )
+
 }
@@
 
-### Load the processed SingleCellExperiment rds object
-
-```{r load}
-sample_dir <- file.path(data_dir, params$sample_id)
-sample_files <- list.files(sample_dir)
-
-file <- sample_files[grepl("_processed.rds", sample_files)]
-
-# open the processed rds object
-sce <- readRDS(file.path(sample_dir, file))
-```
-
-### Run Seurat pipeline
-
-```{r seurat_wf}
-# convert to seurat
-srat <- CreateSeuratObject(counts = counts(sce),
-                                    assay = "RNA",
-                                    project = params$sample_id
-                           )
-
-# convert colData and rowData to data.frame for use in the Seurat object
-cell_metadata <- as.data.frame(colData(sce))
-row_metadata <- as.data.frame(rowData(sce))
-
-# add cell metadata (colData) from SingleCellExperiment to Seurat
-srat@meta.data <- cell_metadata
-
-# add row metadata (rowData) from SingleCellExperiment to Seurat
-srat[["RNA"]]@meta.data <- row_metadata
-
-# add metadata from SingleCellExperiment to Seurat
-srat@misc <- metadata(sce)
-rm(sce)
-
-# Normalization
-srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)
-
-# dimensionality reduction
-srat <- RunPCA(srat, verbose = F)
-srat <- RunUMAP(srat, dims = 1:50, verbose = F)
-
-# clustering
-srat    <- FindNeighbors(srat, dims = 1:50, verbose = F)
-srat    <- FindClusters(srat, verbose = T)
-
-```
-
 ### Visualize seurat clusters
-
-```{r fig.width=20, fig.height=5, out.width='100%'}
-
-d2 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")
-d1 <- SCpubr::do_DimPlot(srat, reduction="pca", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - pca")
-v <- SCpubr::do_ViolinPlot(srat, features = c( "subsets_mito_percent"), ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + v + plot_layout(ncol = 3, widths = c(2,2,4))
-
-```
 
@@
 
+```{r fig.width=20, fig.height=10, out.width='100%'}
+d2 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")
+d1 <- SCpubr::do_DimPlot(srat, reduction="pca", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - pca")
+v1 <- SCpubr::do_ViolinPlot(srat, features = c( "subsets_mito_percent"), ncol = 1, group.by = "seurat_clusters", legend.position = "none") 
+v2 <- SCpubr::do_ViolinPlot(srat, features = c( "detected"), ncol = 1, group.by = "seurat_clusters", legend.position = "none") 
+v3 <- SCpubr::do_ViolinPlot(srat, features = c( "sum"), ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+
+d1 + d2 + (v1 + v2 +v3 + plot_layout(ncol=1) ) + plot_layout(ncol = 3, widths = c(2,2,4))
+```
+
 ### Cell cycle information
@@
 g2m.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$g2m.genes]
-
 srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE)
@@
 Note: we do not expect to have a clear and nice pattern of expression for all of the following markers in every tumor. 
-This is just to get a few ideas.
-
-```{r marker_genes, fig.width=12, fig.height=6, out.width='100%'}
-CellType_metadata <- read_csv(file.path(module_base, "marker-sets", "CellType_metadata.csv"))
-
-DT::datatable(CellType_metadata, caption = ("CellType_metadata"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-
-                               buttons = c( 'csv', 'excel')))
-```
+This is just ti get a few idea.
 
 ```{r fig.width=8, fig.height=2.5, out.width='100%'}
-for(feature in CellType_metadata$ENSEMBL_ID){
-  if(feature %in% rownames(srat)){
+for(feature in CellType_metadata$ENSEMBL_ID[CellType_metadata$ENSEMBL_ID %in% rownames(srat@assays$SCT)]){
+ tryCatch(
+  expr =  {
     tmp <-  visualize_feature(srat, features = feature, group.by = "seurat_clusters")
     print(tmp)
+  },
+  error = {
+    expr= print("feature not found")
     }
+ )
 }
@@
 ```
-
-"Normal" cells (immune, endothelial) have a slightly higher TP53 score.
 
@@
 de_results   <- DElegate::FindAllMarkers2(srat, group_column = "seurat_clusters")
-
 #filter the most relevant markers
 s.markers <- de_results[de_results$padj < params$padj_thershold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold,]
-
 # add gene symbol for easiest interpretation of the result
@@
 identical(s.markers$feature, s.markers$gene_ids)     # check the quality of the merge, must be true
-
 DT::datatable(s.markers, caption = ("marker genes"), 
@@
               options = list(  dom = 'Bfrtip',
-
                                buttons = c( 'csv', 'excel')))
-
 # Select top 5 genes for heatmap plotting
@@
     top_n(n =  5, wt = log_fc) -> top5
-
 # subset for plotting
@@
 ss <- ScaleData(ss, features = top5$feature)
-
 p1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
@@
 p3 <- ggplot(srat@meta.data, aes(seurat_clusters, fill = seurat_clusters)) + geom_bar() + NoLegend()
-
 common_title <- sprintf("Unsupervised clustering %s, %d cells", srat@meta.data$orig.ident[1], ncol(srat))
@@
               options = list(  dom = 'Bfrtip',
-
                                buttons = c( 'csv', 'excel')))
-
 ```
@@
 # define background genes = universe for enrichment
-background <-  row_metadata$gene_ids
-
+background <-  rownames(srat)
 # Define gene signature per cluster
@@
 ```{r fig.width=15, fig.height=8, out.width='100%'}
-tryCatch(
-  expr =  {
+
     Enrichment_plot(category = "H", signatures = signatures, background = background)
-  },
-  error = function(e) {
-    print("No enrichment")
-  }
-)
+
 ```
@@
 
-#### C3 MSiGDB regulatory target gene sets
-
-```{r fig.width=20, fig.height=25, out.width='100%'}
-tryCatch(
-  expr =  {
-Enrichment_plot(category = "C3", signatures = signatures, background = background)
-  },
-  error = function(e) {
-    print("No enrichment")
-  }
-)
-```
-Here to check of we catch and MIR pattern (DROSHA; DICER1; other?)
-
 #### C8 MSiGDB cell type signature gene sets
@@
 ```{r fig.width=20, fig.height=25, out.width='100%'}
-tryCatch(
-  expr =  {
 Enrichment_plot(category = "C8", signatures = signatures, background = background)
-  },
-  error = function(e) {
-    print("No enrichment")
-  }
-)
 ```
@@
 
-### Annotation from SingleR (no cancer or kidney specific dataset)
+### Cell annotation
+
+#### Annotation from SingleR (no cancer or kidney specific dataset)
 
@@
 
-### Annotation from cellassign (no cancer or kidney specific dataset)
+#### Annotation from cellassign (no cancer or kidney specific dataset)
 
@@
 
-Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, it give the first impression that C12 is a cluster of endothelial cells and C11 a cluster of immune cells.
+#### Annotation from the fetal reference (Cao et al)
 
-### Azimuth annotation from fetal kidney
+```{r fig.width=15, fig.height=10, out.width='100%'}
+visualize_metadata(srat, meta = "fetal_full_predicted.organ", group.by = "seurat_clusters")
+visualize_metadata(srat, meta = "fetal_full_predicted.annotation.l1", group.by = "seurat_clusters")
+visualize_metadata(srat, meta = "fetal_full_predicted.annotation.l2", group.by = "seurat_clusters")
+```
 
-To be added in a next PR. Look at opened PR #706
+#### Annotation from the kidney reference (Stewart et al)
 
-For more information related to the reference, please go to https://www.kidneycellatlas.org/
-You will find:
+```{r fig.width=15, fig.height=10, out.width='100%'}
+visualize_metadata(srat, meta = "fetal_kidney_predicted.compartment", group.by = "seurat_clusters")
+visualize_metadata(srat, meta = "fetal_kidney_predicted.cell_type", group.by = "seurat_clusters")
+```
 
-- interactive viewer
+### Compare annotations with marker genes
 
-- h5ad files to download.
+```{r fig.width=15, fig.height=15, out.width='100%'}
 
-Please note that as Wilms tumor have been described to be closer to fetal kidney as mature kidney, we only used the fetal kidney atlas as the reference. 
-Also check : https://www.science.org/doi/10.1126/science.aat5031
+markers = list("malignant" = CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "malignant"],
+            "immune"= CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "immune"],
+            "endothelium"= CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "endothelium"],
+            "non-malignant"= CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "non-malignant"])
 
-This part is imho one of the most important step that allow us to have a quick and reliable idea of the composition of the different clusters. 
-The predicted compartment are defined into 4 categories:
+p1 <- SCpubr::do_DotPlot(sample = srat,
+                         group.by =  "fetal_kidney_predicted.compartment",
+                         features = markers, 
+                         axis.text.x.angle = 90, 
+                         flip = TRUE, 
+                         plot.title = "fetal_kidney_predicted.compartment"
+                         )
 
-- endothelium
-- stroma
-- fetal nephron 
-- immune
+p2 <- SCpubr::do_DotPlot(sample = srat,
+                         group.by =  "fetal_full_predicted.annotation.l1",
+                         features = markers, 
+                         axis.text.x.angle = 90, 
+                         flip = TRUE,
+                         plot.title = "fetal_full_predicted.annotation.l1"
+                          )
 
-As for SingleR and CellAssign, the annotation of immune cells and endothelial cells is straightforward. 
-The stroma compartment should then contain normal and cancer stromal cells.
-The fetal nephron compartment contain blastema cancer cells a well as normal and cancer epithelial cells.
+p3 <- SCpubr::do_DotPlot(sample = srat,
+                         group.by =   "seurat_clusters",
+                         features = markers, 
+                         axis.text.x.angle = 90, 
+                         flip = TRUE,
+                         plot.title =  "seurat_clusters"
+                          )
 
-Further segregation of cancer versus normal cells will be achieved using a combination of markers/pathways (see above) and inferred CNV (to be done).
+p4 <- SCpubr::do_DotPlot(sample = srat,
+                         group.by =   "cellassign_celltype_annotation",
+                         features = markers, 
+                         axis.text.x.angle = 90, 
+                         flip = TRUE,
+                         plot.title = "cellassign_celltype_annotation"
+                          )
+(p1|p2)/(p3|p4)
+
+```
+
+### Compare annotations between them
+
+```{r fig.width=15, fig.height=15, out.width='100%'}
+p1 <- SCpubr::do_AlluvialPlot(sample = srat,
+                            first_group = "fetal_kidney_predicted.compartment",
+                            last_group = "seurat_clusters",
+                            repel = TRUE,
+
+                            )
+
+p2 <- SCpubr::do_AlluvialPlot(sample = srat,
+                            first_group = "fetal_full_predicted.annotation.l1",
+                            last_group = "seurat_clusters",
+                            repel = TRUE
+                            )
+
+p3 <- SCpubr::do_AlluvialPlot(sample = srat,
+                            first_group = "singler_celltype_annotation",
+                            last_group = "seurat_clusters",
+                            repel = TRUE
+                            )
+
+p4 <- SCpubr::do_AlluvialPlot(sample = srat,
+                            first_group = "cellassign_celltype_annotation",
+                            last_group = "seurat_clusters",
+                            repel = TRUE
+                            )
+
+(p1|p2)/(p3|p4)
+```
+```{r fig.width=10, fig.height=15, out.width='100%'}
+SCpubr::do_AlluvialPlot(sample = srat,
+                            first_group = "fetal_kidney_predicted.compartment",
+                            last_group = "fetal_full_predicted.annotation.l1",
+                            repel = TRUE
+
+                                                        )
+
+SCpubr::do_AlluvialPlot(sample = srat,
+                            first_group = "fetal_kidney_predicted.cell_type",
+                            last_group = "fetal_full_predicted.annotation.l2",
+                            repel = TRUE
+                                                        )
+
+```
 

### 2024-09-04 analyses/cell-type-wilms-tumor-06/00_run_workflow.R

@@ -85,5 +85,16 @@ for (sample_id in metadata$scpca_sample_id) {
   }
 }
 
+for (sample_id in metadata$scpca_sample_id) {
+ if (!running_ci) {
+    # Cluster exploration
+    rmarkdown::render(input = file.path(notebook_template_dir, "03_clustering_exploration.Rmd"),
+                      params = list(scpca_project_id = project_id, sample_id = sample_id),
+                      output_format = "html_document",
+                      output_file = paste0("03_clustering_exploration", sample_id, ".html"),
+                      output_dir = file.path(notebook_output_dir, sample_id))
+  }
+}
+
 
 

### 2024-09-05 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
-The aim is to explore the clustering and label transfers for the sample `param$sample_id` =
-
-```{r}
-params$sample_id
-```
+The aim is to explore the clustering and label transfers for the sample `r param$sample_id`.
 
@@
 ### Packages
-
-Load required packages in the following chunk, if needed.
-Do not install packages here; only load them with the `library()` function.
 

### 2024-09-05 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
-                  feature_symbol = AnnotationDbi::select(org.Hs.eg.db, 
+                  feature_symbol <- AnnotationDbi::select(org.Hs.eg.db, 
                                 keys=features, 
@@
 
-  if(class(object@meta.data[,meta]) == "numeric"){
+  if(is.numeric(object@meta.data[,meta])){
      d <- SCpubr::do_FeaturePlot(object, 

### 2024-09-05 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 for(feature in CellType_metadata$ENSEMBL_ID[CellType_metadata$ENSEMBL_ID %in% rownames(srat@assays$SCT)]){
- tryCatch(
-  expr =  {
-    tmp <-  visualize_feature(srat, features = feature, group.by = "seurat_clusters")
-    print(tmp)
-  },
-  error = {
-    expr= print("feature not found")
-    }
- )
+   print(visualize_feature(srat, features = feature, group.by = "seurat_clusters"))
 }

### 2024-09-05 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 s.markers <- left_join(s.markers,feature_conversion, by = c( "gene_ids") )
-identical(s.markers$feature, s.markers$gene_ids)     # check the quality of the merge, must be true
+stopifnot(
+  "Error joining gene ids and feature names" = identical(s.markers$feature, s.markers$gene_ids)   
+)
 DT::datatable(s.markers, caption = ("marker genes"), 

### 2024-09-05 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Add to PR #750

--- 2024-09-05 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
@@
 
-The aim is to explore the clustering and label transfers for the sample `r param$sample_id`.
+The aim is to explore the clustering and label transfers for the sample `r params$sample_id`.
 
@@
 library(ggalluvial)
-
+library(ggsankey)
 set.seed(params$seed)
@@
 
-## Analysis
-
-### Load the pre-processed `Seurat` object
-```{r load, message=FALSE, warning=FALSE}
-# open the processed rds object
-srat <- readRDS(file.path(result_dir, paste0("02b-fetal_kidney_label-transfer_",params$sample_id,".Rds")))
-DefaultAssay(srat) <- "SCT"
-
-```
-
-### Functions
+## Functions
 
 Here we defined function that will be used multiple time all along the notebook.
-
-#### Visualize seurat clusters and markers genes
-
-For a Seurat object `object`and a features `features`, the function `visualize_feature` will plot `FeaturePlot` and `ViolinPlot`
-
-- `object` is the Seurat object
-
-- `features` the gene or quantitative value to be plotted
-
-- `group.by` is the metadata used for grouping the violin plots
-
-```{r }
-visualize_feature <- function(object, features, group.by = "seurat_clusters"){
-
-                  feature_symbol <- AnnotationDbi::select(org.Hs.eg.db, 
-                                keys=features, 
-                                columns="SYMBOL", 
-                                keytype="ENSEMBL")
-
-                  d <- SCpubr::do_FeaturePlot(object, 
-                                              features = feature_symbol$ENSEMBL, 
-                                              pt.size = 0.2, 
-                                              legend.width = 0.5, 
-                                              legend.length = 5, 
-                                              legend.position = "right") + ggtitle(feature_symbol$SYMBOL)
-                  b <- SCpubr::do_ViolinPlot(srat, 
-                                             features = feature_symbol$ENSEMBL, 
-                                             ncol = 1, 
-                                             group.by = group.by, 
-                                             legend.position = "none",
-                                             assay = "SCT") + ylab(feature_symbol$SYMBOL)
-
-                  return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
-}
-
-```
 
@@
 }
+```
+
+#### Visualize seurat clusters and markers genes
+
+For a Seurat object `object`and a features `features`, the function `visualize_feature` will plot `FeaturePlot` and `ViolinPlot`
+
+- `object` is the Seurat object
+
+- `features` the gene or quantitative value to be plotted
+
+- `group.by` is the metadata used for grouping the violin plots
+
+```{r }
+visualize_feature <- function(object, features, group.by = "seurat_clusters"){
+
+                  feature_symbol <- AnnotationDbi::select(org.Hs.eg.db, 
+                                keys=features, 
+                                columns="SYMBOL", 
+                                keytype="ENSEMBL")
+
+                  d <- SCpubr::do_FeaturePlot(object, 
+                                              features = feature_symbol$ENSEMBL, 
+                                              pt.size = 0.2, 
+                                              legend.width = 0.5, 
+                                              legend.length = 5, 
+                                              legend.position = "right") + ggtitle(feature_symbol$SYMBOL)
+                  b <- SCpubr::do_ViolinPlot(srat, 
+                                             features = feature_symbol$ENSEMBL, 
+                                             ncol = 1, 
+                                             group.by = group.by, 
+                                             legend.position = "none",
+                                             assay = "SCT") + ylab(feature_symbol$SYMBOL)
+
+                  return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
+}
+
 ```
@@
 
-## Analysis content
+#### do_SankeyPlot
+
+`do_SankeyPlot` is largely inspired by the source code from `SCpubr` as the function has been removed from the package.
+https://github.com/enblacar/SCpubr/blob/v1.1.2-dev-stable/R/do_SankeyPlot.R
+
+- `first_group` is a metadata variable. First group of nodes of the sankey plot.
+- `last_group` is a metadata variable. Last group of nodes of the sankey plot.
+- `middle_groups` is a metadata variable. Vector of groups of nodes of the sankey plot.
+- `type` is a type of plot to make. One of:
+
+  - `sankey`: Generates a sankey plot.
+  - `alluvial`: Generated an Alluvial plot, a kind of sankey plot where all groups have the same height.
+
+- `width` Width of the nodes.
+- `space` Vertical space between the nodes. It appears to be equal to a single cell. Use big numbers to see a difference (like, 1000 or 10000).
+- `position` GGplot2 position.
+- `node.fill` Color to fill the nodes.
+- `node.color` Color for the contour of the nodes.
+- `flow.alpha` Alpha of the connections.
+- `flow.color` Color for the contour of the connections.
+- `text_size` Size of the labels.
+- `text_color` Color of the labels.
+- `smooth` How smooth the connections are.
+- `use_labels` Whether to use labels or text for the node names.
+- `hjust` General hjust for the labels.
+
+```{r }
+
+do_SankeyPlot <- function(sample,
+                          first_group,
+                          last_group,
+                          type = "sankey",
+                          middle_groups = NULL,
+                          width = 0.1,
+                          space = ifelse(type == "sankey", 0.05 * ncol(sample), 0),
+                          position = "identity",
+                          node.fill = "white",
+                          node.color = "white",
+                          flow.alpha = 0.75,
+                          flow.color = "black",
+                          text_size = 3,
+                          text_color = "black",
+                          font.size = 18,
+                          font.type = "sans",
+                          smooth = 8,
+                          use_labels = FALSE,
+                          hjust = NULL,
+                          colors.first = NULL,
+                          colors.middle = NULL,
+                          colors.last = NULL,
+                          plot.title = NULL,
+                          plot.subtitle = NULL,
+                          plot.caption = NULL){
+
+  # Check logical parameters.
+  logical_list <- list("use_labels" = use_labels)
+  # Check numeric parameters.
+  numeric_list <- list("width" = width,
+                       "space" = space,
+                       "flow.alpha" = flow.alpha,
+                       "text_size" = text_size,
+                       "font.size" = font.size,
+                       "smooth" = smooth,
+                       "hjust" = hjust)
+  # Check character parameters.
+  character_list <- list("first_group" = first_group,
+                         "last_group" = last_group,
+                         "middle_groups" = middle_groups,
+                         "type" = type,
+                         "position" = position,
+                         "node.color" = node.color,
+                         "flow.color" = flow.color,
+                         "text_color" = text_color,
+                         "font.type" = font.type,
+                         "colors.first" = colors.first,
+                         "colors.middle" = colors.middle,
+                         "colors.last" = colors.last,
+                         "node.fill" = node.fill,
+                         "plot.title" = plot.title,
+                         "plot.subtitle" = plot.subtitle,
+                         "plot.caption" = plot.caption)
+
+  `%>%` <- magrittr::`%>%`
+
+  data <- suppressWarnings({sample@meta.data %>%
+                            dplyr::select(dplyr::all_of(c(first_group, middle_groups, last_group))) %>%
+                            tibble::rownames_to_column(var = "cell") %>%
+                            dplyr::select(-.data$cell) %>%
+                            ggsankey::make_long(dplyr::all_of(c(first_group, middle_groups, last_group))) %>%
+                            dplyr::rowwise() %>%
+                            dplyr::mutate(hjust = if(.data$x %in% middle_groups){0.5}
+                                          else if (.data$x == last_group){0}
+                                          else if (.data$x == first_group){1})})
+  if (!is.null(hjust)){data$hjust <- hjust}
+
+  colors.first <- SCpubr::do_ColorPalette(colors.use = "steelblue",
+                        n = length(unique(sample@meta.data[, first_group])))
+    if (is.factor(sample@meta.data[, first_group])){
+      names(colors.first) <- levels(sample@meta.data[, first_group])
+    } else {
+      names(colors.first) <- sort(unique(sample@meta.data[, first_group]))
+    }
+
+  colors.last <- SCpubr::do_ColorPalette(colors.use = "steelblue",
+                        n = length(unique(sample@meta.data[, last_group])))
+    if (is.factor(sample@meta.data[, last_group])){
+      names(colors.last) <- levels(sample@meta.data[, last_group])
+    } else {
+      names(colors.last) <- sort(unique(sample@meta.data[, last_group]))
+    }
+
+  colors.use <- c(colors.first, colors.last)
+  func_use <- ifelse(isTRUE(use_labels), ggsankey::geom_sankey_label, ggsankey::geom_sankey_text)
+
+  p <- data %>%
+
+       ggplot2::ggplot(mapping = ggplot2::aes(x = .data$x,
+                                              next_x = .data$next_x,
+                                              node = .data$node,
+                                              next_node = .data$next_node,
+                                              fill = factor(.data$node),
+                                              label = .data$node,
+                                              hjust = .data$hjust)) +
+       ggsankey::geom_sankey(flow.alpha = flow.alpha,
+                             node.color = node.color,
+                             node.fill = node.fill,
+                             color = flow.color,
+                             width = width,
+                             position = position,
+                             type = type,
+                             space = space) +
+       func_use(size = text_size,
+                color = text_color,
+                fontface = "bold",
+                position = position,
+                type = type,
+                space = space) +
+       ggplot2::scale_fill_manual(values = colors.use) +
+       ggplot2::xlab("") +
+       ggplot2::ylab("") +
+       ggplot2::labs(title = plot.title,
+                     subtitle = plot.subtitle,
+                     caption = plot.caption) +
+       ggplot2::theme_minimal(base_size = font.size) +
+       ggplot2::theme(axis.title = ggplot2::element_text(color = "black",
+                                                         face = "bold"),
+                      axis.line.x = ggplot2::element_blank(),
+                      axis.text.x = ggplot2::element_text(color = "black",
+                                                          face = "bold",
+                                                          angle = 0,
+                                                          hjust = 0.5,
+                                                          vjust = 1),
+                      axis.text.x.top = ggplot2::element_text(color = "black",
+                                                              face = "bold",
+                                                              angle = 0,
+                                                              hjust = 0.5,
+                                                              vjust = 1),
+                      axis.text.y = ggplot2::element_blank(),
+                      axis.ticks = ggplot2::element_blank(),
+                      panel.grid.major = ggplot2::element_blank(),
+                      plot.title.position = "plot",
+                      plot.title = ggplot2::element_text(face = "bold", hjust = 0),
+                      plot.subtitle = ggplot2::element_text(hjust = 0),
+                      plot.caption = ggplot2::element_text(hjust = 1),
+                      panel.grid = ggplot2::element_blank(),
+                      text = ggplot2::element_text(family = font.type),
+                      plot.caption.position = "plot",
+                      legend.text = ggplot2::element_text(face = "bold"),
+                      legend.position = "none",
+                      legend.title = ggplot2::element_text(face = "bold"),
+                      legend.justification = "center",
+                      plot.margin = ggplot2::margin(t = 10, r = 10, b = 10, l = 10),
+                      plot.background = ggplot2::element_rect(fill = "white", color = "white"),
+                      panel.background = ggplot2::element_rect(fill = "white", color = "white"),
+                      legend.background = ggplot2::element_rect(fill = "white", color = "white"),
+                      strip.text =ggplot2::element_text(color = "black", face = "bold"))
+
+  return(p)
+}
+```
+
+#### do_Table_Heatmap
+
+`do_Table_Heatmap` shows heatmap of counts of cells in groups of metadata
+
+- `data` seurat object 
+- `first_group` is the name of the first metadata to group the cells
+- `last_group` is the name of the second metadata to group the cells
+
+```{r fig.width=10, fig.height=4, out.width='100%'}
+do_Table_Heatmap <- function(data, first_group, last_group ){
+
+  df <- data@meta.data %>%
+  mutate_if(sapply(data@meta.data, is.character), as.factor)  %>% 
+  group_by( !!sym(first_group), !!sym(last_group))%>% 
+  summarise(Nb = n())
+
+p <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = Nb)) +
+   geom_tile() +
+   scale_fill_viridis(discrete=FALSE)  +
+   theme_bw() +
+   theme(text = element_text(size = 20))
+
+return(p)
+}
+```
+
+## Analysis
+
+### Load the pre-processed `Seurat` object
+```{r load, message=FALSE, warning=FALSE}
+# open the processed rds object
+srat <- readRDS(file.path(result_dir, paste0("02b-fetal_kidney_label-transfer_",params$sample_id,".Rds")))
+DefaultAssay(srat) <- "SCT"
+```
 
@@
 
-```{r fig.width=10, fig.height=4, out.width='100%'}
+```{r fig.width=20, fig.height=4, out.width='100%'}
 visualize_metadata(srat, meta = "Phase", group.by = "seurat_clusters")
@@
 
-```{r fig.width=8, fig.height=2.5, out.width='100%'}
+```{r fig.width=20, fig.height=2.5, out.width='100%'}
 visualize_metadata(srat, meta  = "S.Score", group.by = "seurat_clusters")
@@
 
-```{r fig.width=8, fig.height=2.5, out.width='100%'}
+```{r fig.width=20, fig.height=4, out.width='100%'}
 for(feature in CellType_metadata$ENSEMBL_ID[CellType_metadata$ENSEMBL_ID %in% rownames(srat@assays$SCT)]){
@@
 
-```{r fig.height=2.5, fig.width=8, warning=FALSE, out.width='100%'}
+```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
 srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_P53_PATHWAY", name = "TP53_score")
@@
 
-```{r fig.height=2.5, fig.width=8, warning=FALSE, out.width='100%'}
+```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
 srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_DNA_REPAIR", name = "DNA_repair_score")
@@
 
-```{r fig.height=2.5, fig.width=8, warning=FALSE, out.width='100%'}
+```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
 srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DROSHA_TARGET_GENES", name = "DROSHA_score")
@@
 
-```{r fig.height=2.5, fig.width=8, warning=FALSE, out.width='100%'}
+```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
 srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DICER1_TARGET_GENES", name = "DICER1_score")
@@
 In addition to the list of known marker genes, we used an unbiased approach to find transcripts that characterized the different clusters.
+
 We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. 
@@
 s.markers <- left_join(s.markers,feature_conversion, by = c( "gene_ids") )
-identical(s.markers$feature, s.markers$gene_ids)     # check the quality of the merge, must be true
+stopifnot(
+  "Error joining gene ids and feature names" = identical(s.markers$feature, s.markers$gene_ids)   
+)
+```
+
+```{r fig.width=10, fig.height=10, out.width='100%'}
+
 DT::datatable(s.markers, caption = ("marker genes"), 
@@
                                buttons = c( 'csv', 'excel')))
+
+```
+
+```{r fig.width=10, fig.height=10, out.width='100%'}
+
 # Select top 5 genes for heatmap plotting
@@
 show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) ) + plot_layout(widths = c(1, 2)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))
+
+```
+
+```{r fig.width=10, fig.height=10, out.width='100%'}
 DT::datatable(top5[, c(1, 9, 11, 12)], caption = ("top 5 marker genes"), 
@@
                                buttons = c( 'csv', 'excel')))
+
 ```
@@
 
-```{r fig.width=15, fig.height=8, out.width='100%'}
+```{r fig.width=20, fig.height=8, out.width='100%'}
 
@@
 
-```{r fig.width=15, fig.height=10, out.width='100%'}
+```{r fig.width=20, fig.height=10, out.width='100%'}
 visualize_metadata(srat, meta = "singler_celltype_annotation", group.by = "seurat_clusters")
@@
 
-```{r fig.width=15, fig.height=10, out.width='100%'}
+```{r fig.width=20, fig.height=10, out.width='100%'}
 visualize_metadata(srat, meta = "cellassign_celltype_annotation", group.by = "seurat_clusters")
@@
 
-```{r fig.width=15, fig.height=10, out.width='100%'}
+```{r fig.width=20, fig.height=10, out.width='100%'}
 visualize_metadata(srat, meta = "fetal_full_predicted.organ", group.by = "seurat_clusters")
@@
 
-```{r fig.width=15, fig.height=10, out.width='100%'}
+```{r fig.width=20, fig.height=10, out.width='100%'}
 visualize_metadata(srat, meta = "fetal_kidney_predicted.compartment", group.by = "seurat_clusters")
@@
 
-```{r fig.width=15, fig.height=15, out.width='100%'}
+```{r fig.width=20, fig.height=15, out.width='100%'}
 
@@
 
-```{r fig.width=15, fig.height=15, out.width='100%'}
-p1 <- SCpubr::do_AlluvialPlot(sample = srat,
+#### Sankey plots
+
+```{r fig.width=20, fig.height=20, out.width='100%'}
+
+p1 <- do_SankeyPlot(sample = srat,
                             first_group = "fetal_kidney_predicted.compartment",
                             last_group = "seurat_clusters",
-                            repel = TRUE,
-
+                             type = "sankey",text_size = 5, use_labels = TRUE
                             )
 
-p2 <- SCpubr::do_AlluvialPlot(sample = srat,
+p2 <- do_SankeyPlot(sample = srat,
                             first_group = "fetal_full_predicted.annotation.l1",
                             last_group = "seurat_clusters",
-                            repel = TRUE
+                             type = "sankey",text_size = 5, use_labels = TRUE
                             )
 
-p3 <- SCpubr::do_AlluvialPlot(sample = srat,
+p3 <- do_SankeyPlot(sample = srat,
                             first_group = "singler_celltype_annotation",
                             last_group = "seurat_clusters",
-                            repel = TRUE
+                             type = "sankey",text_size = 4, use_labels = TRUE
                             )
 
-p4 <- SCpubr::do_AlluvialPlot(sample = srat,
+p4 <- do_SankeyPlot(sample = srat,
                             first_group = "cellassign_celltype_annotation",
                             last_group = "seurat_clusters",
-                            repel = TRUE
+                             type = "sankey",text_size = 5, use_labels = TRUE
                             )
@@
 ```
-```{r fig.width=10, fig.height=15, out.width='100%'}
-SCpubr::do_AlluvialPlot(sample = srat,
+
+```{r fig.width=30, fig.height=10, out.width='100%'}
+p1 <- do_SankeyPlot(sample = srat,
                             first_group = "fetal_kidney_predicted.compartment",
                             last_group = "fetal_full_predicted.annotation.l1",
-                            repel = TRUE
+                            type = "sankey",text_size = 6, use_labels = TRUE
+ )
+
+p2 <- do_SankeyPlot(sample = srat,
+                            first_group = "fetal_kidney_predicted.compartment",
+                            last_group = "fetal_kidney_predicted.cell_type",
+                            type = "sankey",text_size = 6, use_labels = TRUE                                                        )
+
+p1 | p2
+```
+#### Heatmaps
+
+```{r fig.width=20, fig.height=15, out.width='100%'}
+p1 <- do_Table_Heatmap( srat,
+                            last_group = "fetal_kidney_predicted.compartment",
+                            first_group = "seurat_clusters"
 
@@
 
-SCpubr::do_AlluvialPlot(sample = srat,
-                            first_group = "fetal_kidney_predicted.cell_type",
-                            last_group = "fetal_full_predicted.annotation.l2",
-                            repel = TRUE
+p2 <- do_Table_Heatmap( srat,
+                            last_group = "fetal_full_predicted.annotation.l1",
+                            first_group = "seurat_clusters"
                             )
 
+p3 <- do_Table_Heatmap( srat,
+                            last_group = "singler_celltype_annotation",
+                            first_group = "seurat_clusters"
+                            )
+
+p4 <- do_Table_Heatmap( srat,
+                            last_group = "cellassign_celltype_annotation",
+                            first_group = "seurat_clusters"
+                            )
+
+(p1|p2)/(p3|p4)
 ```

### 2024-09-05 analyses/cell-type-wilms-tumor-06/00_run_workflow.R

@@ -59,7 +59,6 @@ for (sample_id in metadata$scpca_sample_id) {
   # create a directory to save the notebooks
   dir.create(file.path(module_base, "notebook", sample_id), showWarnings = FALSE)
 
-
   # Pre-process the data - `Seurat` workflow
   rmarkdown::render(input = file.path(notebook_template_dir, "01_seurat-processing.Rmd"),
                     params = list(scpca_project_id = project_id, sample_id = sample_id),
@@ -81,20 +80,18 @@ for (sample_id in metadata$scpca_sample_id) {
                       output_format = "html_document",
                       output_file = paste0("02b_fetal_kidney_reference_Stewart_", sample_id, ".html"),
                       output_dir = file.path(notebook_output_dir, sample_id))
-
-  }
-}
-
-for (sample_id in metadata$scpca_sample_id) {
- if (!running_ci) {
+    
     # Cluster exploration
     rmarkdown::render(input = file.path(notebook_template_dir, "03_clustering_exploration.Rmd"),
                       params = list(scpca_project_id = project_id, sample_id = sample_id),
                       output_format = "html_document",
                       output_file = paste0("03_clustering_exploration_", sample_id, ".html"),
                       output_dir = file.path(notebook_output_dir, sample_id))
+
   }
 }
+
+
 
 
 

### 2024-09-09 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd

--- 2024-08-29 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
@@
 options(future.globals.maxSize= 891289600000000)
+s <- Azimuth::RunAzimuth(srat, reference ="fetusref")
 
-print(dim(srat)[2])
-# this would need to be change for CI or whole sample
- s <- RunAzimuth_small(srat, reference ="fetusref", k.weight = 10, mapping.score.k = 80, ksmooth = 80)
-
-# We transfer the annotation to the pre-processed `Seurat`iobject as we don't want to keep changes done on the query by `RunAzimuth`
+# We transfer the annotation to the pre-processed `Seurat` object as we don't want to keep changes done on the query by `RunAzimuth`
 metadata_vec <- c("predicted.annotation.l1.score", "predicted.annotation.l1", "predicted.annotation.l2.score", "predicted.annotation.l2", "predicted.organ.score", "predicted.organ")

### 2024-09-10 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
    geom_tile() +
-   scale_fill_viridis(discrete=FALSE)  +
+   scale_fill_viridis_c()  +
    theme_bw() +

### 2024-09-10 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
-markers = list("malignant" = CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "malignant"],
+markers = list("malignant" = CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "malignant")],
             "immune"= CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "immune"],

### 2024-09-10 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

--- 2024-09-10 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
@@
 
-markers = list("malignant" = CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "malignant"],
-            "immune"= CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "immune"],
+markers = list("malignant" = CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "malignant")],
+            "immune"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "immune")],
             "endothelium"= CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "endothelium"],

### 2024-09-10 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

--- 2024-09-10 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
@@
 
-markers = list("malignant" = CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "malignant"],
-            "immune"= CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "immune"],
-            "endothelium"= CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "endothelium"],
+markers = list("malignant" = CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "malignant")],
+            "immune"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "immune")],
+            "endothelium"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "endothelium")],
             "non-malignant"= CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "non-malignant"])

### 2024-09-10 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

--- 2024-09-10 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
@@
 markers = list("malignant" = CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "malignant")],
-            "immune"= CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "immune"],
-            "endothelium"= CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "endothelium"],
-            "non-malignant"= CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class == "non-malignant"])
+            "immune"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "immune")],
+            "endothelium"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "endothelium")],
+            "non-malignant"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "non-malignant")])
 

### 2024-09-10 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
-here we will calculate a TP53 score using AddMduleScore and the genes of the HALLMARK_P53_PATHWAY gene set.
+here we will calculate a TP53 score using `AddModuleScore` and the genes of the HALLMARK_P53_PATHWAY gene set.
 
@@
 
-here we will calculate a DNA_repair score using AddMduleScore and the genes of the HALLMARK_DNA_REPAIR gene set.
+here we will calculate a DNA_repair score using `AddModuleScore` and the genes of the HALLMARK_DNA_REPAIR gene set.
 

### 2024-09-10 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
         set_list <- list(set_list$human_ensembl_gene)
+       suppressWarnings({
             object <- AddModuleScore(object, features = set_list, name = name)
+       })
 return(object)
@@
 ```
+
 #### Heatmaps

### 2024-09-12 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>
add to PR #750

@@
 ```{r setup, include=FALSE}
+
 knitr::opts_chunk$set(echo = TRUE,
@@
                       )
+
 ```
@@
 
+```{r}
+
+subdiagnosis <- readr::read_tsv(
+  file.path("..", "..", "..", "data", "current", params$scpca_project_id, "single_cell_metadata.tsv"),
+  show_col_types = FALSE
+  ) |>
+  dplyr::filter(scpca_sample_id == params$sample_id) |>
+  dplyr::pull(subdiagnosis)
+
+```
+
 The aim is to explore the clustering and label transfers for the sample `r params$sample_id`.
+This sample has a(n) `r subdiagnosis` subdiagnosis.
 
@@
 ```{r packages, message=FALSE, warning=FALSE}
-library("Seurat")
+library(Seurat)
 library(sctransform)
-
 library(SCpubr)             # for plotting 
@@
 library(ggalluvial)
-library(ggsankey)
+
 set.seed(params$seed)
@@
 
-In this notebook, we are working with the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`.
+In this notebook, we are working with the Wilms tumor sample defined in `r params$sample_id` from the Wilms tumor dataset `r params$scpca_project_id`.
 We work with the pre-processed and labeled `Seurat` object that is the output of `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd` saved in the `results` directory.
@@
 
-#### do_SankeyPlot
-
-`do_SankeyPlot` is largely inspired by the source code from `SCpubr` as the function has been removed from the package.
-https://github.com/enblacar/SCpubr/blob/v1.1.2-dev-stable/R/do_SankeyPlot.R
-
-- `first_group` is a metadata variable. First group of nodes of the sankey plot.
-- `last_group` is a metadata variable. Last group of nodes of the sankey plot.
-- `middle_groups` is a metadata variable. Vector of groups of nodes of the sankey plot.
-- `type` is a type of plot to make. One of:
-
-  - `sankey`: Generates a sankey plot.
-  - `alluvial`: Generated an Alluvial plot, a kind of sankey plot where all groups have the same height.
-
-- `width` Width of the nodes.
-- `space` Vertical space between the nodes. It appears to be equal to a single cell. Use big numbers to see a difference (like, 1000 or 10000).
-- `position` GGplot2 position.
-- `node.fill` Color to fill the nodes.
-- `node.color` Color for the contour of the nodes.
-- `flow.alpha` Alpha of the connections.
-- `flow.color` Color for the contour of the connections.
-- `text_size` Size of the labels.
-- `text_color` Color of the labels.
-- `smooth` How smooth the connections are.
-- `use_labels` Whether to use labels or text for the node names.
-- `hjust` General hjust for the labels.
-
-```{r }
-
-do_SankeyPlot <- function(sample,
-                          first_group,
-                          last_group,
-                          type = "sankey",
-                          middle_groups = NULL,
-                          width = 0.1,
-                          space = ifelse(type == "sankey", 0.05 * ncol(sample), 0),
-                          position = "identity",
-                          node.fill = "white",
-                          node.color = "white",
-                          flow.alpha = 0.75,
-                          flow.color = "black",
-                          text_size = 3,
-                          text_color = "black",
-                          font.size = 18,
-                          font.type = "sans",
-                          smooth = 8,
-                          use_labels = FALSE,
-                          hjust = NULL,
-                          colors.first = NULL,
-                          colors.middle = NULL,
-                          colors.last = NULL,
-                          plot.title = NULL,
-                          plot.subtitle = NULL,
-                          plot.caption = NULL){
-
-  # Check logical parameters.
-  logical_list <- list("use_labels" = use_labels)
-  # Check numeric parameters.
-  numeric_list <- list("width" = width,
-                       "space" = space,
-                       "flow.alpha" = flow.alpha,
-                       "text_size" = text_size,
-                       "font.size" = font.size,
-                       "smooth" = smooth,
-                       "hjust" = hjust)
-  # Check character parameters.
-  character_list <- list("first_group" = first_group,
-                         "last_group" = last_group,
-                         "middle_groups" = middle_groups,
-                         "type" = type,
-                         "position" = position,
-                         "node.color" = node.color,
-                         "flow.color" = flow.color,
-                         "text_color" = text_color,
-                         "font.type" = font.type,
-                         "colors.first" = colors.first,
-                         "colors.middle" = colors.middle,
-                         "colors.last" = colors.last,
-                         "node.fill" = node.fill,
-                         "plot.title" = plot.title,
-                         "plot.subtitle" = plot.subtitle,
-                         "plot.caption" = plot.caption)
-
-  `%>%` <- magrittr::`%>%`
-
-  data <- suppressWarnings({sample@meta.data %>%
-                            dplyr::select(dplyr::all_of(c(first_group, middle_groups, last_group))) %>%
-                            tibble::rownames_to_column(var = "cell") %>%
-                            dplyr::select(-.data$cell) %>%
-                            ggsankey::make_long(dplyr::all_of(c(first_group, middle_groups, last_group))) %>%
-                            dplyr::rowwise() %>%
-                            dplyr::mutate(hjust = if(.data$x %in% middle_groups){0.5}
-                                          else if (.data$x == last_group){0}
-                                          else if (.data$x == first_group){1})})
-  if (!is.null(hjust)){data$hjust <- hjust}
-
-  colors.first <- SCpubr::do_ColorPalette(colors.use = "steelblue",
-                        n = length(unique(sample@meta.data[, first_group])))
-    if (is.factor(sample@meta.data[, first_group])){
-      names(colors.first) <- levels(sample@meta.data[, first_group])
-    } else {
-      names(colors.first) <- sort(unique(sample@meta.data[, first_group]))
-    }
-
-  colors.last <- SCpubr::do_ColorPalette(colors.use = "steelblue",
-                        n = length(unique(sample@meta.data[, last_group])))
-    if (is.factor(sample@meta.data[, last_group])){
-      names(colors.last) <- levels(sample@meta.data[, last_group])
-    } else {
-      names(colors.last) <- sort(unique(sample@meta.data[, last_group]))
-    }
-
-  colors.use <- c(colors.first, colors.last)
-  func_use <- ifelse(isTRUE(use_labels), ggsankey::geom_sankey_label, ggsankey::geom_sankey_text)
-
-  p <- data %>%
-
-       ggplot2::ggplot(mapping = ggplot2::aes(x = .data$x,
-                                              next_x = .data$next_x,
-                                              node = .data$node,
-                                              next_node = .data$next_node,
-                                              fill = factor(.data$node),
-                                              label = .data$node,
-                                              hjust = .data$hjust)) +
-       ggsankey::geom_sankey(flow.alpha = flow.alpha,
-                             node.color = node.color,
-                             node.fill = node.fill,
-                             color = flow.color,
-                             width = width,
-                             position = position,
-                             type = type,
-                             space = space) +
-       func_use(size = text_size,
-                color = text_color,
-                fontface = "bold",
-                position = position,
-                type = type,
-                space = space) +
-       ggplot2::scale_fill_manual(values = colors.use) +
-       ggplot2::xlab("") +
-       ggplot2::ylab("") +
-       ggplot2::labs(title = plot.title,
-                     subtitle = plot.subtitle,
-                     caption = plot.caption) +
-       ggplot2::theme_minimal(base_size = font.size) +
-       ggplot2::theme(axis.title = ggplot2::element_text(color = "black",
-                                                         face = "bold"),
-                      axis.line.x = ggplot2::element_blank(),
-                      axis.text.x = ggplot2::element_text(color = "black",
-                                                          face = "bold",
-                                                          angle = 0,
-                                                          hjust = 0.5,
-                                                          vjust = 1),
-                      axis.text.x.top = ggplot2::element_text(color = "black",
-                                                              face = "bold",
-                                                              angle = 0,
-                                                              hjust = 0.5,
-                                                              vjust = 1),
-                      axis.text.y = ggplot2::element_blank(),
-                      axis.ticks = ggplot2::element_blank(),
-                      panel.grid.major = ggplot2::element_blank(),
-                      plot.title.position = "plot",
-                      plot.title = ggplot2::element_text(face = "bold", hjust = 0),
-                      plot.subtitle = ggplot2::element_text(hjust = 0),
-                      plot.caption = ggplot2::element_text(hjust = 1),
-                      panel.grid = ggplot2::element_blank(),
-                      text = ggplot2::element_text(family = font.type),
-                      plot.caption.position = "plot",
-                      legend.text = ggplot2::element_text(face = "bold"),
-                      legend.position = "none",
-                      legend.title = ggplot2::element_text(face = "bold"),
-                      legend.justification = "center",
-                      plot.margin = ggplot2::margin(t = 10, r = 10, b = 10, l = 10),
-                      plot.background = ggplot2::element_rect(fill = "white", color = "white"),
-                      panel.background = ggplot2::element_rect(fill = "white", color = "white"),
-                      legend.background = ggplot2::element_rect(fill = "white", color = "white"),
-                      strip.text =ggplot2::element_text(color = "black", face = "bold"))
-
-  return(p)
-}
-```
-
 #### do_Table_Heatmap
 
-`do_Table_Heatmap` shows heatmap of counts of cells in groups of metadata
+`do_Table_Heatmap` shows heatmap of counts of cells for combinations of two metadata variables
 
@@
   group_by( !!sym(first_group), !!sym(last_group))%>% 
-  summarise(Nb = n())
+  summarise(Nb = log(n()))
 
@@
    theme_bw() +
-   theme(text = element_text(size = 20))
+   theme(text = element_text(size = 20)) +
+   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+   guides(fill=guide_legend(title=" Number of \n cells (log)"))
 
@@
 
-```{r fig.width=20, fig.height=15, out.width='100%'}
+Here we evaluate with marker genes the identification of endothelial, immune cells and other cancer and non-cancerous sub-populations.
+
+```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
 
@@
 p2 <- SCpubr::do_DotPlot(sample = srat,
+                         group.by =  "fetal_kidney_predicted.cell_type",
+                         features = markers, 
+                         axis.text.x.angle = 90, 
+                         flip = TRUE,
+                         plot.title = "fetal_kidney_predicted.cell_type"
+                          )
+
+p3 <- SCpubr::do_DotPlot(sample = srat,
                          group.by =   "fetal_full_predicted.annotation.l1",
@@
 
-p3 <- SCpubr::do_DotPlot(sample = srat,
+p4 <- SCpubr::do_DotPlot(sample = srat,
+                         group.by =  "fetal_full_predicted.annotation.l2",
+                         features = markers, 
+                         axis.text.x.angle = 90, 
+                         flip = TRUE,
+                         plot.title = "fetal_full_predicted.annotation.l2"
+                          )
+
+p5 <- SCpubr::do_DotPlot(sample = srat,
                          group.by =   "seurat_clusters",
@@
 
-p4 <- SCpubr::do_DotPlot(sample = srat,
+p6 <- SCpubr::do_DotPlot(sample = srat,
                          group.by =  "cellassign_celltype_annotation",
@@
                           )
-(p1|p2)/(p3|p4)
+
+(p1|p2)/(p3|p4) /(p5|p6)
 
@@
 
-#### Sankey plots
+#### Annotation versus `Seurat` clusters
 
-```{r fig.width=20, fig.height=20, out.width='100%'}
-
-p1 <- do_SankeyPlot(sample = srat,
-                            first_group = "fetal_kidney_predicted.compartment",
-                            last_group = "seurat_clusters",
-                             type = "sankey",text_size = 5, use_labels = TRUE
-                            )
-
-p2 <- do_SankeyPlot(sample = srat,
-                            first_group = "fetal_full_predicted.annotation.l1",
-                            last_group = "seurat_clusters",
-                             type = "sankey",text_size = 5, use_labels = TRUE
-                            )
-
-p3 <- do_SankeyPlot(sample = srat,
-                            first_group = "singler_celltype_annotation",
-                            last_group = "seurat_clusters",
-                             type = "sankey",text_size = 4, use_labels = TRUE
-                            )
-
-p4 <- do_SankeyPlot(sample = srat,
-                            first_group = "cellassign_celltype_annotation",
-                            last_group = "seurat_clusters",
-                             type = "sankey",text_size = 5, use_labels = TRUE
-                            )
-
-(p1|p2)/(p3|p4)
-```
-
-```{r fig.width=30, fig.height=10, out.width='100%'}
-p1 <- do_SankeyPlot(sample = srat,
-                            first_group = "fetal_kidney_predicted.compartment",
-                            last_group = "fetal_full_predicted.annotation.l1",
-                            type = "sankey",text_size = 6, use_labels = TRUE
- )
-
-p2 <- do_SankeyPlot(sample = srat,
-                            first_group = "fetal_kidney_predicted.compartment",
-                            last_group = "fetal_kidney_predicted.cell_type",
-                            type = "sankey",text_size = 6, use_labels = TRUE                                                        )
-
-p1 | p2
-```
-
-#### Heatmaps
+Here we check wether one (or more) `Seurat` cluster do correspond to one label. 
+This would allow us to be more confident in the annotation of immune or endothelial cells for example.
 
@@
 
+#### Comparison of the labels from the two fetal references
+
+The use of the right reference is crucial for label transfer and annotation. 
+It is recommended that the cell types in the reference is representative to the cell types to be annotated in the query.
+
+Wilms tumors can contain up to three histologies that resemble fetal kidney: blastema, stroma, and epithelia [[1](https://www.ncbi.nlm.nih.gov/books/NBK373356/), [2](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9915828/)].
+Because of their histological similarity to fetal kidneys, Wilms tumors are thought to arise from developmental derangements in embryonic renal progenitors.
+
+In the previous step of the analysis, we selected two fetal references containing kidney samples and transfer the labels.
+
+Here we aim to understand the overlap and differences between the two references.
+
+##### Human fetal kidney atlas Stewart et al.
+
+Stewart et al. created a [human fetal kidney atlas](https://www.kidneycellatlas.org/) [[3](https://www.science.org/doi/10.1126/science.aat5031)].
+
+This reference contains only fetal kidney cells and has been precisely annotated by kidney experts.
+
+##### Azimuth Human fetal atlas Cao et al.
+
+`Azimuth` also provides a human fetal atlas as a reference [[4](https://www.science.org/doi/10.1126/science.aba7721)].
+
+The data can be found on [Zenodo](https://zenodo.org/records/4738021#.YJIW4C2ZNQI).
+
+The reference contain cells from 15 organs including kidney from fetal samples.
+
+We were first nicely surprised that most of the cells in the Wilms tumor dataset mapped to "kidney" organ (see previous report `02a_fetal_all_reference_Cao_{sample_id}.html`).
+
+##### first level annotation
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+p1 <- do_Table_Heatmap( srat,
+                            last_group = "fetal_kidney_predicted.compartment",
+                            first_group = "fetal_full_predicted.annotation.l1"
+
+                            )
+p2 <- do_Table_Heatmap( srat,
+                            last_group = "fetal_kidney_predicted.cell_type",
+                            first_group = "fetal_full_predicted.annotation.l1"
+
+                            )
+p1|p2
+```
+
+##### second level annotation
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+p2 <- do_Table_Heatmap( srat,
+                            last_group = "fetal_kidney_predicted.cell_type",
+                            first_group = "fetal_full_predicted.annotation.l2"
+
+                            )
+p2
+```
+
 ## Session Info

### 2024-09-12 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 ```{r fig.width=20, fig.height=10, out.width='100%'}
+d1 <- SCpubr::do_DimPlot(srat, reduction="pca", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - pca")
 d2 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")
-d1 <- SCpubr::do_DimPlot(srat, reduction="pca", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - pca")
 v1 <- SCpubr::do_ViolinPlot(srat, features = c( "subsets_mito_percent"), ncol = 1, group.by = "seurat_clusters", legend.position = "none") 

### 2024-09-12 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 We do not expect a nice labelling of the cells as the overlap of cell types between the reference and the query dataset is poor. 
-This support the need to do a proper label transfer from the fetal kidney atlas, which is imho the best reference that can be applied to a Wilms tumor query.
+This support the need to do a proper label transfer from the fetal kidney atlas, which appears to be the best reference that can be applied to a Wilms tumor query.
 

### 2024-09-12 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
-```{r fig.width=20, fig.height=8, out.width='100%'}
+```{r fig.width=20, fig.height=15, out.width='100%'}
 p1 <- do_Table_Heatmap( srat,

### 2024-09-12 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
-```{r fig.width=20, fig.height=8, out.width='100%'}
+```{r fig.width=20, fig.height=15, out.width='100%'}
 p2 <- do_Table_Heatmap( srat,

### 2024-09-12 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
change to PR#750

@@
 
+#### Introduction to the four annotation strategies
+
+The use of the right reference is crucial for label transfer and annotation. 
+It is recommended that the cell types in the reference is representative to the cell types to be annotated in the query.
+
+Wilms tumors can contain up to three histologies that resemble fetal kidney: blastema, stroma, and epithelia [[1](https://www.ncbi.nlm.nih.gov/books/NBK373356/), [2](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9915828/)].
+Because of their histological similarity to fetal kidneys, Wilms tumors are thought to arise from developmental derangements in embryonic renal progenitors.
+
+For these reasons, we expect Wilms tumor cells to map to fetal cells, especially fetal kidney cells.
+
+Here, we investigate annotations resulting from 4 different strategies:
+
+* the `_processed.rds` objects already contain automated annotations computed by the [Data Lab](https://scpca.readthedocs.io/en/latest/getting_started.html):
+
+    + Annotations determined by SingleR, an automated reference-based method [Looney et al. 2019](https://www.nature.com/articles/s41590-018-0276-y).
+
+    + Annotations determined by CellAssign, an automated marker-gene based method [Zhang et al. 2019](https://www.nature.com/articles/s41592-019-0529-1).
+
+* the label transfers from two fetal references using `Azimuth`, that have been performed in the previous analysis
+
+    + Using the human fetal atlas as a reference: 
+    This is the reference developed by [Cao et al.](https://www.science.org/doi/10.1126/science.aba7721) provided by `Azimuth` for label transfer.
+    The reference contain cells from 15 organs including kidney from fetal samples.
+    The label transfer have been performed in the notebook `02a_fetal_full_reference_Cao_{sample_id}.Rmd`
+
+    + Using the human fetal kidney atlas: 
+    [Stewart et al.](https://www.science.org/doi/10.1126/science.aat5031) created a [human fetal kidney atlas](https://www.kidneycellatlas.org/). 
+    This reference contains only fetal kidney cells and has been precisely annotated by kidney experts. 
+    The label transfer have been performed in the notebook `02b_fetal_kidney_reference_Stewart_{sample_id}.Rmd`
+
 #### Annotation from SingleR (no cancer or kidney specific dataset)
-
-Here, we quickly checked annotations that are present in the _processed rds object. 
-However, the automated annotation have not been performed using a cancer specific reference or a kidney reference. 
-We do not expect a nice labelling of the cells as the overlap of cell types between the reference and the query dataset is poor. 
-This support the need to do a proper label transfer from the fetal kidney atlas, which appears to be the best reference that can be applied to a Wilms tumor query.
 
@@
 
+The fetal full reference (Cao et al.) provides three levels of annotations:
+
+- `fetal_full_predicted.organ` is one of the 15 organs used for building the reference. 
+We were nicely surprised that most of the cells from the Wilms tumor cohort maped to fetal kidney cells.
+
+- `fetal_full_predicted.annotation.l1` holds predicted cell type annotation
+
+- `fetal_full_predicted.annotation.l2` holds the combination of the predicted organ and predicted cell type annotation
+
 ```{r fig.width=20, fig.height=10, out.width='100%'}
@@
 #### Annotation from the kidney reference (Stewart et al)
+
+The fetal kidney reference (Stewart et al.) provides two levels of annotations:
+
+- `fetal_kidney_predicted.compartment` is one of the 4 main compartments composing a fetal kidney.
+We expect immune and endothelial cells to be healthy (non-canerous) cells identified easily with high confidency. 
+We expect stroma and fetal nephron compartment to contain both normal and malignant cells.
+
+    - `immune` cells
+
+    - `stroma` cells
+
+    - `fetal_nephron` cells
+
+    - `endothelial` cells
+
+- `fetal_full_predicted.annotation.l1` holds predicted cell type annotations that details further the four compartments, as for example:
+
+    - `immune` subtypes: myeloid cells, lymphocytes, ...
+
+    - `stroma` subtypes: mesenchymal stem cells, fibroblasts, ...
+
+    - `fetal_nephron` subtypes: podocytes, uteric bud, ...
+
+    - `endothelial` cells
 
@@
 
-The use of the right reference is crucial for label transfer and annotation. 
-It is recommended that the cell types in the reference is representative to the cell types to be annotated in the query.
-
-Wilms tumors can contain up to three histologies that resemble fetal kidney: blastema, stroma, and epithelia [[1](https://www.ncbi.nlm.nih.gov/books/NBK373356/), [2](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9915828/)].
-Because of their histological similarity to fetal kidneys, Wilms tumors are thought to arise from developmental derangements in embryonic renal progenitors.
-
-In the previous step of the analysis, we selected two fetal references containing kidney samples and transfer the labels.
-
-Here we aim to understand the overlap and differences between the two references.
-
-##### Human fetal kidney atlas Stewart et al.
-
-Stewart et al. created a [human fetal kidney atlas](https://www.kidneycellatlas.org/) [[3](https://www.science.org/doi/10.1126/science.aat5031)].
-
-This reference contains only fetal kidney cells and has been precisely annotated by kidney experts.
-
-##### Azimuth Human fetal atlas Cao et al.
-
-`Azimuth` also provides a human fetal atlas as a reference [[4](https://www.science.org/doi/10.1126/science.aba7721)].
-
-The data can be found on [Zenodo](https://zenodo.org/records/4738021#.YJIW4C2ZNQI).
-
-The reference contain cells from 15 organs including kidney from fetal samples.
-
-We were first nicely surprised that most of the cells in the Wilms tumor dataset mapped to "kidney" organ (see previous report `02a_fetal_all_reference_Cao_{sample_id}.html`).
-
 ##### first level annotation
@@
 
+#### Comparison of the  fetal reference with `singleR` and `cellassign`
+
+##### first level annotation
+
+```{r fig.width=20, fig.height=15, out.width='100%'}
+p1 <- do_Table_Heatmap( srat,
+                            last_group = "fetal_kidney_predicted.compartment",
+                            first_group = "cellassign_celltype_annotation"
+
+                            )
+p2 <- do_Table_Heatmap( srat,
+                            last_group = "fetal_kidney_predicted.compartment",
+                            first_group = "singler_celltype_annotation"
+
+                            )
+p1|p2
+```
+
+##### second level annotation
+
+```{r fig.width=20, fig.height=30, out.width='100%'}
+p1 <- do_Table_Heatmap( srat,
+                            last_group = "fetal_kidney_predicted.cell_type",
+                            first_group = "cellassign_celltype_annotation"
+
+                            )
+
+p1 <- do_Table_Heatmap( srat,
+                            last_group = "fetal_kidney_predicted.cell_type",
+                            first_group = "singler_celltype_annotation"
+
+                            )
+
+p1/p2
+```
+
 ## Session Info

### 2024-09-13 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Add notebook 03- to PR#750

@@
 
-    - `fetal_nephron` subtypes: podocytes, uteric bud, ...
+    - `fetal_nephron` subtypes: podocytes, uteric bud, mesenchymal cells (that appear to be cap mesenchyme cells) ...
 
@@
 
-```{r fig.width=20, fig.height=15, out.width='100%'}
+```{r fig.width=20, fig.height=10, out.width='100%'}
 p1 <- do_Table_Heatmap( srat,
@@
 
-p1 <- do_Table_Heatmap( srat,
+p2 <- do_Table_Heatmap( srat,
                             last_group = "fetal_kidney_predicted.cell_type",

### 2024-09-13 analyses/cell-type-wilms-tumor-14/notebook-template.Rmd
init module skeleton

--- 2024-08-28 analyses/cluster-utils/notebook-template.Rmd
@@
 ---
-title: "cluster-utils Analysis"
+title: "cell-type-wilms-tumor-14 Analysis"
 author:
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "cluster-utils")
+module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-14")
 ```

### 2024-09-13 analyses/cell-type-dsrct/notebook-template.Rmd
initial commit include loading files, analyzing gene expression, and saving a processed RDS

--- 2024-08-28 analyses/cluster-utils/notebook-template.Rmd
@@
 ---
-title: "cluster-utils Analysis"
+title: "cell-type-dsrct Analysis"
 author:
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "cluster-utils")
+module_base <- file.path(repository_base, "analyses", "cell-type-dsrct")
 ```

### 2024-09-13 analyses/cell-type-dsrct/scripts/utils/jaccard-functions.R

@@ -1,0 +1,119 @@
+# These functions are taken directly from the supplemental cell type report produced in scpca-nf
+# https://github.com/AlexsLemonade/scpca-nf/blob/main/templates/qc_report/celltypes_supplemental_report.rmd
+
+#' Function to calculate Jaccard similarity on two vectors
+#'
+#' @param vec1 First vector
+#' @param vec2 Second vector
+#'
+#' @return Jaccard similarity between the vectors
+jaccard <- function(vec1, vec2) {
+  length(intersect(vec1, vec2)) / length(union(vec1, vec2))
+}
+
+
+# Wrapper function to calculate jaccard similarity matrix for two categorical variables
+#'
+#' @param celltype_df The celltype_df data frame which must contain these columns:
+#'   `colname1`, `colname2`, and `barcodes`
+#' @param colname1 Column name, as a string, of first categorical variable of interest
+#' @param colname2 Column name, as a string, of second categorical variable of interest
+#'
+#' @return Jaccard similarity matrix for the two columns. `colname1` values will
+#'   be row names and `colname2` values will be column names in the final matrix
+make_jaccard_matrix <- function(celltype_df, colname1, colname2) {
+  # make lists of barcodes for each category, named by the category
+  id1_list <- split(celltype_df$barcodes, celltype_df[[colname1]])
+  id2_list <- split(celltype_df$barcodes, celltype_df[[colname2]])
+  
+  # create the grid of comparisons
+  cross_df <- tidyr::expand_grid(id1 = names(id1_list), id2 = names(id2_list))
+  
+  # calculate a single Jaccard index for each combination using split lists & ids
+  jaccard_scores <- cross_df |>
+    purrr::pmap_dbl(\(id1, id2){
+      jaccard(id1_list[[id1]], id2_list[[id2]])
+    })
+  
+  # add scores to the comparison grid and convert to matrix
+  jaccard_matrix <- cross_df |>
+    dplyr::mutate(jaccard = jaccard_scores) |>
+    # convert to matrix
+    tidyr::pivot_wider(
+      names_from = "id2",
+      values_from = "jaccard"
+    ) |>
+    tibble::column_to_rownames(var = "id1") |>
+    as.matrix()
+  
+  return(jaccard_matrix)
+}
+
+# function that turns jaccard matrices into a list of heatmaps 
+make_heatmap_list <- function(jaccard_matrices, column_title, legend_match){
[…]

### 2024-09-13 analyses/cell-type-nonETP-ALL-03/notebook-template.Rmd
init module

--- 2024-08-28 analyses/cluster-utils/notebook-template.Rmd
@@
 ---
-title: "cluster-utils Analysis"
+title: "cell-type-nonETP-ALL-03 Analysis"
 author:
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "cluster-utils")
+module_base <- file.path(repository_base, "analyses", "cell-type-nonETP-ALL-03")
 ```

### 2024-09-16 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
+- `nbins` is the number of bins for `AddModuleScore` to use. Its default value is 24, which matches Seurat's default value.
 ```{r }
-MSigDB_score <- function(object, category, gs_name, name){
+MSigDB_score <- function(object, category, gs_name, name, nbin = 24){
 
@@
        suppressWarnings({
-            object <- AddModuleScore(object, features = set_list, name = name)
+            object <- AddModuleScore(object, features = set_list, name = name, nbin = nbin)
        })

### 2024-09-16 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 DefaultAssay(srat) <- "SCT"
+
+# Add number of bins to use with Seurat::AddModuleScore().
+# One sample in this project fails with the default number of bins, so we need to override it
+if (params$sample_id == "SCPCS000197") {
+  seurat_nbins <- 23 # default is 24, so this is very close
+} else {
+  seurat_nbins <- 24
+}
 ```

### 2024-09-16 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 ```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
-srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_P53_PATHWAY", name = "TP53_score")
+srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_P53_PATHWAY", name = "TP53_score", nbin = seurat_nbins)
 visualize_metadata(srat, meta = "TP53_score1", group.by = "seurat_clusters" )

### 2024-09-16 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 ```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
-srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_DNA_REPAIR", name = "DNA_repair_score")
+srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_DNA_REPAIR", name = "DNA_repair_score", nbin = seurat_nbins)
 visualize_metadata(srat, meta = "DNA_repair_score1", group.by = "seurat_clusters" )

### 2024-09-16 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 ```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
-srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DROSHA_TARGET_GENES", name = "DROSHA_score")
+srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DROSHA_TARGET_GENES", name = "DROSHA_score", nbin = seurat_nbins)
 visualize_metadata(srat, meta = "DROSHA_score1", group.by = "seurat_clusters" )

### 2024-09-16 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 ```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
-srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DICER1_TARGET_GENES", name = "DICER1_score")
+srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DICER1_TARGET_GENES", name = "DICER1_score", nbin = seurat_nbins)
 visualize_metadata(srat, meta = "DICER1_score1", group.by = "seurat_clusters" )

### 2024-09-16 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 g2m.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$g2m.genes]
-srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE)
+srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE, nbin = seurat_nbins)
 ```

### 2024-09-19 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 - `fetal_full_predicted.organ` is one of the 15 organs used for building the reference. 
-We were nicely surprised that most of the cells from the Wilms tumor cohort maped to fetal kidney cells.
+We were nicely surprised that, for many of the samples, most of the cells from the Wilms tumor cohort mapped to fetal kidney cells.
 
@@
 
-- `fetal_full_predicted.annotation.l1` holds predicted cell type annotations that details further the four compartments, as for example:
+- `fetal_kidney_predicted.cell_type` holds predicted cell type annotations that details further the four compartments, as for example:
 

### 2024-09-23 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
04_annotation_Across_Samples_exploration

--- 2024-09-19 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
@@
 ---
-title: "Clustering exploration for `r params$sample_id`"
+title: "Annotation exploration for `r params$scpca_project_id`"
 author: "Maud PLASCHKA"
@@
   seed: 12345
-  padj_thershold: 0.05  # Threshold for adjusted p-value
-  lfc_threshold: 1      # Threshold for log fold change
-  rate1_threshold: 0.5  # Threshold for rate (percentage) of cells expressing the marker genes in group1
 output: 
@@
 
-```{r}
+The aim is to explore the annotations and label transfers for all of the samples in `r params$scpca_project_id`.
 
-subdiagnosis <- readr::read_tsv(
-  file.path("..", "..", "..", "data", "current", params$scpca_project_id, "single_cell_metadata.tsv"),
-  show_col_types = FALSE
-  ) |>
-  dplyr::filter(scpca_sample_id == params$sample_id) |>
-  dplyr::pull(subdiagnosis)
-
-```
-
-The aim is to explore the clustering and label transfers for the sample `r params$sample_id`.
-This sample has a(n) `r subdiagnosis` subdiagnosis.
-
-In order to explore the clustering results, we look into some marker genes, pathways enrichment and label transfer.
-
-This approach would provide us a rapid idea of the quality of the clustering:
-
-- in the case each cluster do have a set of specific marker genes, we could expect each of the cluster to have at least a different phenotype,
-- if marker genes are mostly shared between clusters, we might have overclustered,
-- if no marker genes are found, the quality of cells in the clusters might be impaired (high mitochondrial content? ribosomal content?)
+In order to explore the label transfer results, we look into some marker genes, table and percentages of cells in each annotation groups (from label transfers).
 
@@
 library(patchwork)
-library(msigdbr)
-library(enrichplot)
-library(clusterProfiler)
+
+library(reshape2)
 library(org.Hs.eg.db)
-library(ggalluvial)
 
@@
 module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
+
+result_dir <- file.path(module_base, "results")
 ```
@@
 
-In this notebook, we are working with the Wilms tumor sample defined in `r params$sample_id` from the Wilms tumor dataset `r params$scpca_project_id`.
+In this notebook, we are working with all of the samples in `r params$scpca_project_id`.
+
 We work with the pre-processed and labeled `Seurat` object that is the output of `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd` saved in the `results` directory.
 
+The sample metadata can be found in `sample_metadata_file` in the `data` folder.
+
 ```{r path_to_query}
-result_dir <- file.path(module_base, "results", params$sample_id)
+sample_metadata_file <- file.path(repository_base, "data", "current", params$scpca_project_id, "single_cell_metadata.tsv")
+
+metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
 ```
 
-To explore the clustering results, we look into some marker genes reported here:
+To explore the annotation results, we look into some marker genes reported here:
 
@@
 
-Reports will be saved in the `notebook` directory. 
-The pre-processed and annotated `Seurat` object per samples are saved in the `result` folder.
+The report will be saved in the `notebook` directory.
 
@@
                                               features = meta, 
+                                              reduction = "umap.rpca",
                                               pt.size = 0.2, 
@@
 
-  d <- SCpubr::do_DimPlot(object, reduction="umap", group.by = group.by, label = TRUE, repel = TRUE) + ggtitle(paste0(meta," - umap")) + theme(text=element_text(size=18))
+  d <- SCpubr::do_DimPlot(object, reduction="umap.rpca", group.by = meta, label = TRUE, repel = TRUE) + ggtitle(paste0(meta," - umap")) + theme(text=element_text(size=18)) + NoLegend()
   b <- SCpubr::do_BarPlot(sample = object,
@@
                          theme(text=element_text(size=18))
-  return(d + b + plot_layout(ncol = 2, widths = c(2,4)))
+  return(d + b + plot_layout(ncol = 1, heights = c(4,2)))
   }
@@
                                               features = feature_symbol$ENSEMBL,
+                                              reduction = "umap.rpca",
                                               pt.size = 0.2, 
@@
                   return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
-}
-
-```
-
-#### Calculate `ModuleScore` from a `MSigDB` dataset
-
-For a Seurat object `object`, the function `MSigDB_score` will calculate a score using `AddModuleScore()` function for a MSigDB gene set `gs_name` in the category `category`
-
-- `object` is the Seurat object to calculate the score. 
-Score will be added in the metadata of this object.
-
-- `category` is a MSigDB collection (https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp).
-values in `c("H", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")`
-
-- `gs_name` is the name of a MSigDB gene set, e.g. `"HALLMARK_P53_PATHWAY"`
-
-- `name` is the name of the module
-
-- `nbins` is the number of bins for `AddModuleScore` to use. Its default value is 24, which matches Seurat's default value.
-```{r }
-MSigDB_score <- function(object, category, gs_name, name, nbin = 24){
-
-        set <- msigdbr(species = "human", category = category)
-        set_list <- set %>% 
-                    dplyr::filter(gs_name == gs_name) %>%
-                    dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()
-        set_list <- list(set_list$human_ensembl_gene)
-       suppressWarnings({
-            object <- AddModuleScore(object, features = set_list, name = name, nbin = nbin)
-       })
-return(object)
-}
-```
-
-#### Enrichment analysis
-
-`Enrichment_plot`aim to perform enrichment of the marker genes for each of the Seurat clusters and summarize the results in a `dotplot`.
-
-- `category` is the MSigDB category or collection to be used
-values in `c("H", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")`
-
-- `signatures` is a list of marker genes per cluster
-
-- `background`is the universe used for enrichment
-
-```{r }
-Enrichment_plot <- function(category, signatures, background){
-    ## define genesets
-   tryCatch( expr =  {
-               gene_set <- msigdbr(species = "human", category = category)
-               msigdbr_set <-  gene_set %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
-               cclust<-compareCluster(geneCluster = signatures, 
-               fun = enricher,
-               TERM2GENE = msigdbr_set,
-               universe=background)
-               d <- dotplot(cclust,showCategory=15) + scale_y_discrete(labels=function(x) str_wrap(x, width=40))
-               return(d)  
-   },
-   error = function(e) { print("No enrichment") } )
-
 }
@@
 # open the processed rds object
-srat <- readRDS(file.path(result_dir, paste0("02b-fetal_kidney_label-transfer_",params$sample_id,".Rds")))
-DefaultAssay(srat) <- "SCT"
+filelist <- list.files(result_dir, 
+                       recursive = TRUE, 
+                       pattern = "02b-fetal_kidney_label-transfer_", 
+                       full.names = TRUE)
 
-# Add number of bins to use with Seurat::AddModuleScore().
-# One sample in this project fails with the default number of bins, so we need to override it
-if (params$sample_id == "SCPCS000197") {
-  seurat_nbins <- 23 # default is 24, so this is very close
-} else {
-  seurat_nbins <- 24
-}
-```
+sample_id <- list.files(result_dir, 
+                       recursive = TRUE, 
+                       pattern = "02b-fetal_kidney_label-transfer_", 
+                       full.names = FALSE)
+sample_id <- colsplit(sample_id, pattern = "/", names = c("sample_id", "suffixe"))
+sample_id <- sample_id[,1]
 
-### Visualize seurat clusters
-
-We expect up to 5 set of clusters:
-
-- blastema cancer cells
-- epithelium cancer and/or normal cells
-- stroma cancer and/or normal cells
-- immune cells
-- endothelial cells
-
-```{r fig.width=20, fig.height=10, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="pca", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - pca")
-d2 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")
-v1 <- SCpubr::do_ViolinPlot(srat, features = c( "subsets_mito_percent"), ncol = 1, group.by = "seurat_clusters", legend.position = "none") 
-v2 <- SCpubr::do_ViolinPlot(srat, features = c( "detected"), ncol = 1, group.by = "seurat_clusters", legend.position = "none") 
-v3 <- SCpubr::do_ViolinPlot(srat, features = c( "sum"), ncol = 1, group.by = "seurat_clusters", legend.position = "none")
-
-d1 + d2 + (v1 + v2 +v3 + plot_layout(ncol=1) ) + plot_layout(ncol = 3, widths = c(2,2,4))
-```
-
-### Cell cycle information
-```{r cell_cycle}
-s.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$s.genes]
-g2m.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$g2m.genes]
-srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE, nbin = seurat_nbins)
-```
-
-```{r fig.width=20, fig.height=4, out.width='100%'}
-visualize_metadata(srat, meta = "Phase", group.by = "seurat_clusters")
-```
-
-```{r fig.width=20, fig.height=2.5, out.width='100%'}
-visualize_metadata(srat, meta  = "S.Score", group.by = "seurat_clusters")
-visualize_metadata(srat, meta = "G2M.Score", group.by = "seurat_clusters")
-```
-
-### Look at specific genes
-
-Here, we open the table of marker genes marker-sets/CellType_metadata.csv. 
-Note: we do not expect to have a clear and nice pattern of expression for all of the following markers in every tumor. 
-This is just ti get a few idea.
-
-```{r fig.width=20, fig.height=4, out.width='100%'}
-for(feature in CellType_metadata$ENSEMBL_ID[CellType_metadata$ENSEMBL_ID %in% rownames(srat@assays$SCT)]){
-   print(visualize_feature(srat, features = feature, group.by = "seurat_clusters"))
-}
-```
-
-### Look at specific pathways
-
-#### TP53 pathway
-
-here we will calculate a TP53 score using `AddModuleScore` and the genes of the HALLMARK_P53_PATHWAY gene set.
-
-```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
-srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_P53_PATHWAY", name = "TP53_score", nbin = seurat_nbins)
-visualize_metadata(srat, meta = "TP53_score1", group.by = "seurat_clusters" )
-```
-
-#### DNA repair pathway
-
-here we will calculate a DNA_repair score using `AddModuleScore` and the genes of the HALLMARK_DNA_REPAIR gene set.
-
-```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
-srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_DNA_REPAIR", name = "DNA_repair_score", nbin = seurat_nbins)
-visualize_metadata(srat, meta = "DNA_repair_score1", group.by = "seurat_clusters" )
-```
-
-Note: Chemo-treated samples should have higher DNA-damage scores.
-
-#### DROSHA target genes
-
-```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
-srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DROSHA_TARGET_GENES", name = "DROSHA_score", nbin = seurat_nbins)
-visualize_metadata(srat, meta = "DROSHA_score1", group.by = "seurat_clusters" )
-```
-
-#### DICER1 target genes
-
-```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
-srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DICER1_TARGET_GENES", name = "DICER1_score", nbin = seurat_nbins)
-visualize_metadata(srat, meta = "DICER1_score1", group.by = "seurat_clusters" )
-```
-
-### Find marker genes for each of the seurat clusters
-
-In addition to the list of known marker genes, we used an unbiased approach to find transcripts that characterized the different clusters.
-
-We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. 
-DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method. 
-Please check the preprint from Chistoph Hafemeister: https://www.biorxiv.org/content/10.1101/2023.03.28.534443v1
-and tool described here: https://github.com/cancerbits/DElegate
-
-Of note, we won't use it for annotation, this is just here to get an idea!
-
-```{r fig.width=10, fig.height=10, out.width='100%'}
-feature_conversion <- srat@assays$RNA@meta.data
-de_results   <- DElegate::FindAllMarkers2(srat, group_column = "seurat_clusters")
-#filter the most relevant markers
-s.markers <- de_results[de_results$padj < params$padj_thershold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold,]
-# add gene symbol for easiest interpretation of the result
-s.markers$gene_ids <- s.markers$feature
-s.markers <- left_join(s.markers,feature_conversion, by = c( "gene_ids") )
-stopifnot(
-  "Error joining gene ids and feature names" = identical(s.markers$feature, s.markers$gene_ids)   
-)
-```
-
-```{r fig.width=10, fig.height=10, out.width='100%'}
-
-DT::datatable(s.markers, caption = ("marker genes"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-                               buttons = c( 'csv', 'excel')))
+sce <- lapply(filelist, function(x) readRDS(x))
 
@@
 
-```{r fig.width=10, fig.height=10, out.width='100%'}
+### Merge `Seurat` objects (with integration)
 
-# Select top 5 genes for heatmap plotting
-s.markers <- na.omit(s.markers)
-s.markers %>%
-    group_by(group1) %>%
-    top_n(n =  5, wt = log_fc) -> top5
-# subset for plotting
-cells <- WhichCells(srat, downsample = 100)
-ss <- subset(srat, cells = cells)
-ss <- ScaleData(ss, features = top5$feature)
-p1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
-p2 <- DoHeatmap(ss, features = top5$feature,  cells = cells, group.by = "seurat_clusters") + NoLegend() + 
-  scale_fill_gradientn(colors =  c("#01665e","#35978f",'darkslategray3', "#f7f7f7", "#fee391","#fec44f","#F9AD03")) 
-p3 <- ggplot(srat@meta.data, aes(seurat_clusters, fill = seurat_clusters)) + geom_bar() + NoLegend()
-common_title <- sprintf("Unsupervised clustering %s, %d cells", srat@meta.data$orig.ident[1], ncol(srat))
-show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) ) + plot_layout(widths = c(1, 2)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))
+We followed the standard `Seurat` workflow described here to merge and integrate the 40 samples from the Wilms tumor dataset `r params$scpca_project_id`.
+For more information, see the vignette [Merge objects (with integration)](https://satijalab.org/seurat/articles/essential_commands.html#merge-objects-with-integration).
+We have slight changes as we already performed normalization using [`SCTransform`](https://github.com/satijalab/seurat/issues/2814).
 
+```{r process, message=FALSE, warning=FALSE}
+options(future.globals.maxSize= 8912896000000)
+
+srat <- merge(sce[[1]], y = sce[2:length(sce)], 
+              add.cell.ids = sample_id, 
+              project = params$scpca_project_id)
+
+DefaultAssay(object = srat) <- "SCT"
+
+VariableFeatures(srat[["SCT"]]) <- rownames(srat[["SCT"]]@scale.data)
+
+srat <- RunPCA(srat)
+
+srat <- IntegrateLayers(object = srat, 
+                        method = RPCAIntegration,
+                        orig.reduction = "pca", 
+                        new.reduction = 'rpca',
+                        normalization.method="SCT", verbose = FALSE)
+
+srat <- FindNeighbors(srat, dims = 1:30, reduction = "rpca")
+srat <- FindClusters(srat)
+srat <- RunUMAP(srat, reduction = "rpca", dims = 1:30, reduction.name = "umap.rpca")
 ```
 
-```{r fig.width=10, fig.height=10, out.width='100%'}
-DT::datatable(top5[, c(1, 9, 11, 12)], caption = ("top 5 marker genes"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-                               buttons = c( 'csv', 'excel')))
+### Quick look at the integration
 
+This is not the main point of this notebook, we expect to work on integration later on, one of the final step of the analysis. 
+We simply here would like to have a rough idea of cell/sample distribution.
+
+```{r fig.width=15, fig.height=10, out.width='100%'}
+tmp <- colsplit(colnames(srat), pattern = "_", names = c("sample_id", "barcode"))
+srat <- AddMetaData(srat, tmp$sample_id, col.name = "sample_id")
+
+d1 <- SCpubr::do_DimPlot(srat, reduction="umap.rpca", group.by = "sample_id", label = FALSE) + ggtitle("sample_id - umap.rpca reduction")
+
+d2 <- SCpubr::do_DimPlot(srat, reduction="umap.rpca", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("seurat_clusters - umap.rpca reduction")
+
+d1 | d2
 ```
-
-### Enrichment analysis of marker genes
-
-Here we perform enrichment analysis of the marker genes found in the previous section for each Seurat cluster.
-
-We defined as universe/background all the genes expressed in the dataset, meaning the rownames of the Seurat object.
-
-We used three gene sets from MSiGDB:
-
-- the hallmark gene sets  are coherently expressed signatures derived by aggregating many MSigDB gene sets to represent well-defined biological states or processes.
-- the C3 : regulatory target gene sets  based on gene target predictions for microRNA seed sequences and predicted transcription factor binding sites.
-- the C8 : cell type signature gene sets  curated from cluster markers identified in single-cell sequencing studies of human tissue.
-
-We used enricher function from clusterProfiler to perform enrichment analysis.
-
-```{r enrichment}
-# define background genes = universe for enrichment
-background <-  rownames(srat)
-# Define gene signature per cluster
-signatures <- list()
-for( i in unique(s.markers$group1)){
-  signatures[[paste0("cluster ",i)]] <- s.markers$feature[s.markers$group1 == i]
-}
-```
-
-#### Hallmarks MSiGDB gene sets
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-Enrichment_plot(category = "H", signatures = signatures, background = background)
-
-```
-
-EMT signature should be enriched in stroma cluster.
-E2F/proliferation should be enriched in blastema cluster.
-MYC(N), TP53 must be enriched in blastema cluster.
-
-#### C8 MSiGDB cell type signature gene sets
-
-```{r fig.width=20, fig.height=25, out.width='100%'}
-Enrichment_plot(category = "C8", signatures = signatures, background = background)
-```
-
-The MSigDB C8 gene set is quite relevant for kidney and nephroblastoma annotations. 
-Epithelial (cancer and normal) cells should be enriched in mature/adulte kidney pathways while blastema cancer cells will show enrichment of fetal kidney development pathway / cap mesenchyme.
 
@@
 
-```{r fig.width=20, fig.height=10, out.width='100%'}
+```{r fig.width=20, fig.height=20, out.width='100%'}
 visualize_metadata(srat, meta = "singler_celltype_annotation", group.by = "seurat_clusters")
@@
 
-```{r fig.width=20, fig.height=10, out.width='100%'}
+```{r fig.width=10, fig.height=20, out.width='100%'}
 visualize_metadata(srat, meta = "cellassign_celltype_annotation", group.by = "seurat_clusters")
@@
 
-```{r fig.width=20, fig.height=10, out.width='100%'}
+```{r fig.width=15, fig.height=20, out.width='100%'}
 visualize_metadata(srat, meta = "fetal_full_predicted.organ", group.by = "seurat_clusters")
+```
+
+```{r fig.width=30, fig.height=20, out.width='100%'}
+tmp <- table(srat$fetal_full_predicted.organ)
+percent.of.kidney <- tmp["Kidney"] / sum(tmp) * 100
+```
+
+Of note, `r percent.of.kidney` % of the cells are labeled as kidney cells.
+
+```{r fig.width=15, fig.height=20, out.width='100%'}
 visualize_metadata(srat, meta = "fetal_full_predicted.annotation.l1", group.by = "seurat_clusters")
@@
 
-```{r fig.width=20, fig.height=10, out.width='100%'}
+```{r fig.width=10, fig.height=10, out.width='100%'}
 visualize_metadata(srat, meta = "fetal_kidney_predicted.compartment", group.by = "seurat_clusters")
+```
+
+```{r fig.width=30, fig.height=20, out.width='100%'}
+tmp <- table(srat$fetal_kidney_predicted.compartment)
+percent.of.immune <- 100*tmp["immune"]/sum(tmp)
+percent.of.endothelial <- 100*tmp["endothelium"]/sum(tmp)
+```
+
+Of note, `r percent.of.immune` % of the cells are labeled as immune cells and `r percent.of.endothelial` % of the cells are labeled as endothelial cells.
+
+Let's have a look at the repartition of cells per patient:
+
+```{r fig.width=10, fig.height=50, out.width='100%'}
+table(srat$sample_id, srat$fetal_kidney_predicted.compartment)
+```
+
+```{r fig.width=10, fig.height=20, out.width='100%'}
 visualize_metadata(srat, meta = "fetal_kidney_predicted.cell_type", group.by = "seurat_clusters")
@@
 
+#### "fetal_kidney_predicted.compartment"
+
 Here we evaluate with marker genes the identification of endothelial, immune cells and other cancer and non-cancerous sub-populations.
 
-```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
+```{r fig.height=8, fig.width=8, message=FALSE, warning=FALSE, out.width='100%'}
 
@@
 
-p1 <- SCpubr::do_DotPlot(sample = srat,
+SCpubr::do_DotPlot(sample = srat,
                          group.by =  "fetal_kidney_predicted.compartment",
@@
                          ) 
+```
 
-p2 <- SCpubr::do_DotPlot(sample = srat,
+##### Endothelial cells
+
+We look at the endothelial marker "ENSG00000110799" = "VWF"
+
+```{r fig.height=4, fig.width=8, message=FALSE, warning=FALSE, out.width='100%'}
+visualize_feature(object = srat,
+                  features = "ENSG00000110799",
+                  group.by = "fetal_kidney_predicted.compartment")
+```
+
+##### Immune cells
+
+We look at the immune marker "ENSG00000081237" = "PTPRC" alias "CD45"
+
+```{r fig.height=4, fig.width=8, message=FALSE, warning=FALSE, out.width='100%'}
+visualize_feature(object = srat,
+                  features = "ENSG00000081237",
+                  group.by = "fetal_kidney_predicted.compartment")
+```
+#### "fetal_kidney_predicted.cell_type"
+
+```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
+
+SCpubr::do_DotPlot(sample = srat,
                          group.by =  "fetal_kidney_predicted.cell_type",
@@
 
-p3 <- SCpubr::do_DotPlot(sample = srat,
+```
+#### "fetal_full_predicted.annotation.l1"
+
+```{r fig.height=10, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
+
+SCpubr::do_DotPlot(sample = srat,
                          group.by =   "fetal_full_predicted.annotation.l1",
@@
                           )
+```
+#### "cellassign_celltype_annotation"
 
-p4 <- SCpubr::do_DotPlot(sample = srat,
-                         group.by =  "fetal_full_predicted.annotation.l2",
-                         features = markers, 
-                         axis.text.x.angle = 90, 
-                         flip = TRUE,
-                         plot.title = "fetal_full_predicted.annotation.l2"
-                          )
-
-p5 <- SCpubr::do_DotPlot(sample = srat,
-                         group.by =   "seurat_clusters",
-                         features = markers, 
-                         axis.text.x.angle = 90, 
-                         flip = TRUE,
-                         plot.title =  "seurat_clusters"
-                          )
-
-p6 <- SCpubr::do_DotPlot(sample = srat,
+```{r fig.height=15, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
+SCpubr::do_DotPlot(sample = srat,
                          group.by =  "cellassign_celltype_annotation",
@@
                           )                         
-
-(p1|p2)/(p3|p4) /(p5|p6)
-
 ```
@@
 
-```{r fig.width=20, fig.height=15, out.width='100%'}
+```{r fig.width=25, fig.height=25, out.width='100%'}
 p1 <- do_Table_Heatmap( srat,
@@
 
-```{r fig.width=20, fig.height=15, out.width='100%'}
+```{r fig.width=30, fig.height=12, out.width='100%'}
 p1 <- do_Table_Heatmap( srat,
@@
 
-```{r fig.width=20, fig.height=15, out.width='100%'}
+```{r fig.width=30, fig.height=15, out.width='100%'}
 p2 <- do_Table_Heatmap( srat,
@@
 
-```{r fig.width=20, fig.height=10, out.width='100%'}
+```{r fig.width=30, fig.height=10, out.width='100%'}
 p1 <- do_Table_Heatmap( srat,
@@
 
-```{r fig.width=20, fig.height=30, out.width='100%'}
+```{r fig.width=30, fig.height=30, out.width='100%'}
 p1 <- do_Table_Heatmap( srat,

### 2024-09-23 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
changes to PR#776

@@
 library(Seurat)
-library(sctransform)
-library(SCpubr)             # for plotting 
 library(tidyverse)
 library(patchwork)
-
-library(reshape2)
-library(org.Hs.eg.db)
 
@@
 
-We work with the pre-processed and labeled `Seurat` object that is the output of `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd` saved in the `results` directory.
+The sample metadata can be found in `sample_metadata_file` in the `data` folder.
 
-The sample metadata can be found in `sample_metadata_file` in the `data` folder.
+We extracted from the pre-processed and labeled `Seurat` object (that is the output of `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd` saved in the `results` directory)
+the following information per cell:
+
+- the predicted compartment in `fetal_kidney_predicted.compartment`
+
+- the predicted organ in `fetal_full_predicted.organ`
+
+- the sample identifier in `sample_id`
+
+- the `counts`of the endothelial marker VWF = "ENSG00000110799"
+
+- the `counts`of the immune marker PTPRC = "ENSG00000081237"
 
@@
 metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
-```
 
-To explore the annotation results, we look into some marker genes reported here:
+# Create a data frames of all annotations
+cell_type_df <- metadata$scpca_sample_id |>
+  purrr::map(
+    # For each sample_id, do the following:
+    \(sample_id) {
 
-```{r marker_genes, fig.width=12, fig.height=6, out.width='100%'}
-CellType_metadata <- read_csv(file.path(module_base, "marker-sets", "CellType_metadata.csv"))
+      # Read in the Seurat object
+      srat <- readRDS(
+        file.path(result_dir, sample_id, paste0("02b-fetal_kidney_label-transfer_", sample_id, ".Rds"))
+      )
 
-DT::datatable(CellType_metadata, caption = ("CellType_metadata"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-                               buttons = c( 'csv', 'excel')))
+      # Create and return a data frame from the Seurat object with relevant annotations
+      # this data frame will have four columns: barcode, sample_id, compartment, organ
+      data.frame(
+        compartment = srat$fetal_kidney_predicted.compartment,
+        organ = srat$fetal_full_predicted.organ, 
+        PTPRC = FetchData(object = srat, vars = "ENSG00000081237", layer = "counts"),
+        VWF = FetchData(object = srat, vars = "ENSG00000110799", layer = "counts")
+        ) |>
+        tibble::rownames_to_column("barcode") |>
+        dplyr::mutate(sample_id = sample_id)
+    }
+  ) |>
+  # now combine all dataframes to make one big one
+  dplyr::bind_rows()
+
 ```
@@
 
-Here we defined function that will be used multiple time all along the notebook.
-
-#### Visualize seurat clusters and metadata
-
-For a Seurat object `object`and a metadata `metadata`, the function `visualize_metadata` will plot `FeaturePlot` and `BarPlot`
-
-- `object` is the Seurat object
-
-- `metadata` the gene or quantitative value to be plotted
-
-- `group.by` is the metadata used for grouping the violin plots
-
-```{r }
-visualize_metadata <- function(object, meta, group.by){
-
-  if(is.numeric(object@meta.data[,meta])){
-     d <- SCpubr::do_FeaturePlot(object, 
-                                              features = meta, 
-                                              reduction = "umap.rpca",
-                                              pt.size = 0.2, 
-                                              legend.width = 0.5, 
-                                              legend.length = 5, 
-                                              legend.position = "right") + ggtitle(meta)
-    b <- SCpubr::do_ViolinPlot(srat, 
-                                             features = meta, 
-                                             ncol = 1, 
-                                             group.by = group.by, 
-                                             legend.position = "none")
-
-   return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
-  }
-
-  else{
-
-  d <- SCpubr::do_DimPlot(object, reduction="umap.rpca", group.by = meta, label = TRUE, repel = TRUE) + ggtitle(paste0(meta," - umap")) + theme(text=element_text(size=18)) + NoLegend()
-  b <- SCpubr::do_BarPlot(sample = object,
-                         group.by = meta,
-                         split.by = group.by,
-                         position = "fill",
-                         font.size = 10,
-                         legend.ncol = 3) +
-                         ggtitle("% cells")+
-                         xlab(print(group.by)) +
-                         theme(text=element_text(size=18))
-  return(d + b + plot_layout(ncol = 1, heights = c(4,2)))
-  }
-
-}
-```
-
-#### Visualize seurat clusters and markers genes
-
-For a Seurat object `object`and a features `features`, the function `visualize_feature` will plot `FeaturePlot` and `ViolinPlot`
-
-- `object` is the Seurat object
-
-- `features` the gene or quantitative value to be plotted
-
-- `group.by` is the metadata used for grouping the violin plots
-
-```{r }
-visualize_feature <- function(object, features, group.by = "seurat_clusters"){
-
-                  feature_symbol <- AnnotationDbi::select(org.Hs.eg.db, 
-                                keys=features, 
-                                columns="SYMBOL", 
-                                keytype="ENSEMBL")
-
-                  d <- SCpubr::do_FeaturePlot(object, 
-                                              features = feature_symbol$ENSEMBL,
-                                              reduction = "umap.rpca",
-                                              pt.size = 0.2, 
-                                              legend.width = 0.5, 
-                                              legend.length = 5, 
-                                              legend.position = "right") + ggtitle(feature_symbol$SYMBOL)
-                  b <- SCpubr::do_ViolinPlot(srat, 
-                                             features = feature_symbol$ENSEMBL, 
-                                             ncol = 1, 
-                                             group.by = group.by, 
-                                             legend.position = "none",
-                                             assay = "SCT") + ylab(feature_symbol$SYMBOL)
-
-                  return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
-}
-
-```
-
 #### do_Table_Heatmap
@@
 
-- `data` seurat object 
 - `first_group` is the name of the first metadata to group the cells
@@
 ```{r fig.width=10, fig.height=4, out.width='100%'}
-do_Table_Heatmap <- function(data, first_group, last_group ){
+do_Table_Heatmap <- function(first_group, last_group ){
 
-  df <- data@meta.data %>%
-  mutate_if(sapply(data@meta.data, is.character), as.factor)  %>% 
+  df <- cell_type_df %>%
   group_by( !!sym(first_group), !!sym(last_group))%>% 
-  summarise(Nb = log(n()))
+  summarise(Nb = n())  %>%
+  mutate(C = sum(Nb)) %>%
+  mutate(percent = Nb/C*100)
 
-p <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = Nb)) +
+p1 <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = Nb)) +
    geom_tile() +
@@
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
-   guides(fill=guide_legend(title=" Number of \n cells (log)"))
+   guides(fill=guide_legend(title=" Number of \n cells"))
+
+p2 <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = percent)) +
+   geom_tile() +
+   scale_fill_viridis_c()  +
+   theme_bw() +
+   theme(text = element_text(size = 20)) +
+   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+   guides(fill=guide_legend(title=" Percentage of \n cells"))
+
+p <- p1 + p2 + plot_layout(ncol = 2)
+return(p)
+}
+```
+
+#### do_Feature_mean
+
+`do_Feature_mean` shows heatmap of mean expression of a feature grouped by a metadata.
+
+- `group.by` is the name of the metadata to group the cells
+- `feature` is the name of the gene to average the expression
+
+```{r fig.width=10, fig.height=4, out.width='100%'}
+do_Feature_mean <- function(group.by, feature ){
+
+  df <- cell_type_df %>%
+  group_by(sample_id, !!sym(group.by))%>% 
+  summarise(m = mean(!!sym(feature)))
+
+p <- ggplot(df, aes(x= sample_id, y = !!sym(group.by), fill = m)) +
+   geom_tile() +
+   scale_fill_viridis_c()  +
+   theme_bw() +
+   theme(text = element_text(size = 20)) +
+   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+   guides(fill=guide_legend(title=paste0(feature)))
 
@@
 
-### Load the pre-processed `Seurat` object
-```{r load, message=FALSE, warning=FALSE}
-# open the processed rds object
-filelist <- list.files(result_dir, 
-                       recursive = TRUE, 
-                       pattern = "02b-fetal_kidney_label-transfer_", 
-                       full.names = TRUE)
+### Total number of cells per samples
 
-sample_id <- list.files(result_dir, 
-                       recursive = TRUE, 
-                       pattern = "02b-fetal_kidney_label-transfer_", 
-                       full.names = FALSE)
-sample_id <- colsplit(sample_id, pattern = "/", names = c("sample_id", "suffixe"))
-sample_id <- sample_id[,1]
+```{r fig.width=5, fig.height=10, out.width='100%', results='asis'}
+tmp <- table(cell_type_df$sample_id)
+knitr::kable(tmp)
+```
+### Predicted organ
 
-sce <- lapply(filelist, function(x) readRDS(x))
-
-```
-
-### Merge `Seurat` objects (with integration)
-
-We followed the standard `Seurat` workflow described here to merge and integrate the 40 samples from the Wilms tumor dataset `r params$scpca_project_id`.
-For more information, see the vignette [Merge objects (with integration)](https://satijalab.org/seurat/articles/essential_commands.html#merge-objects-with-integration).
-We have slight changes as we already performed normalization using [`SCTransform`](https://github.com/satijalab/seurat/issues/2814).
-
-```{r process, message=FALSE, warning=FALSE}
-options(future.globals.maxSize= 8912896000000)
-
-srat <- merge(sce[[1]], y = sce[2:length(sce)], 
-              add.cell.ids = sample_id, 
-              project = params$scpca_project_id)
-
-DefaultAssay(object = srat) <- "SCT"
-
-VariableFeatures(srat[["SCT"]]) <- rownames(srat[["SCT"]]@scale.data)
-
-srat <- RunPCA(srat)
-
-srat <- IntegrateLayers(object = srat, 
-                        method = RPCAIntegration,
-                        orig.reduction = "pca", 
-                        new.reduction = 'rpca',
-                        normalization.method="SCT", verbose = FALSE)
-
-srat <- FindNeighbors(srat, dims = 1:30, reduction = "rpca")
-srat <- FindClusters(srat)
-srat <- RunUMAP(srat, reduction = "rpca", dims = 1:30, reduction.name = "umap.rpca")
-```
-
-### Quick look at the integration
-
-This is not the main point of this notebook, we expect to work on integration later on, one of the final step of the analysis. 
-We simply here would like to have a rough idea of cell/sample distribution.
-
-```{r fig.width=15, fig.height=10, out.width='100%'}
-tmp <- colsplit(colnames(srat), pattern = "_", names = c("sample_id", "barcode"))
-srat <- AddMetaData(srat, tmp$sample_id, col.name = "sample_id")
-
-d1 <- SCpubr::do_DimPlot(srat, reduction="umap.rpca", group.by = "sample_id", label = FALSE) + ggtitle("sample_id - umap.rpca reduction")
-
-d2 <- SCpubr::do_DimPlot(srat, reduction="umap.rpca", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("seurat_clusters - umap.rpca reduction")
-
-d1 | d2
-```
-
-### Cell annotation
-
-#### Introduction to the four annotation strategies
-
-The use of the right reference is crucial for label transfer and annotation. 
-It is recommended that the cell types in the reference is representative to the cell types to be annotated in the query.
-
-Wilms tumors can contain up to three histologies that resemble fetal kidney: blastema, stroma, and epithelia [[1](https://www.ncbi.nlm.nih.gov/books/NBK373356/), [2](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9915828/)].
-Because of their histological similarity to fetal kidneys, Wilms tumors are thought to arise from developmental derangements in embryonic renal progenitors.
-
-For these reasons, we expect Wilms tumor cells to map to fetal cells, especially fetal kidney cells.
-
-Here, we investigate annotations resulting from 4 different strategies:
-
-* the `_processed.rds` objects already contain automated annotations computed by the [Data Lab](https://scpca.readthedocs.io/en/latest/getting_started.html):
-
-    + Annotations determined by SingleR, an automated reference-based method [Looney et al. 2019](https://www.nature.com/articles/s41590-018-0276-y).
-
-    + Annotations determined by CellAssign, an automated marker-gene based method [Zhang et al. 2019](https://www.nature.com/articles/s41592-019-0529-1).
-
-* the label transfers from two fetal references using `Azimuth`, that have been performed in the previous analysis
-
-    + Using the human fetal atlas as a reference: 
+The predicted organ comes from label transfer from the human fetal atlas. 
 This is the reference developed by [Cao et al.](https://www.science.org/doi/10.1126/science.aba7721) provided by `Azimuth` for label transfer.
@@
 The label transfer have been performed in the notebook `02a_fetal_full_reference_Cao_{sample_id}.Rmd`
-
-    + Using the human fetal kidney atlas: 
-    [Stewart et al.](https://www.science.org/doi/10.1126/science.aat5031) created a [human fetal kidney atlas](https://www.kidneycellatlas.org/). 
-    This reference contains only fetal kidney cells and has been precisely annotated by kidney experts. 
-    The label transfer have been performed in the notebook `02b_fetal_kidney_reference_Stewart_{sample_id}.Rmd`
-
-#### Annotation from SingleR (no cancer or kidney specific dataset)
-
-```{r fig.width=20, fig.height=20, out.width='100%'}
-visualize_metadata(srat, meta = "singler_celltype_annotation", group.by = "seurat_clusters")
-```
-
-#### Annotation from cellassign (no cancer or kidney specific dataset)
-
-```{r fig.width=10, fig.height=20, out.width='100%'}
-visualize_metadata(srat, meta = "cellassign_celltype_annotation", group.by = "seurat_clusters")
-```
-
-#### Annotation from the fetal reference (Cao et al)
 
@@
 
-```{r fig.width=15, fig.height=20, out.width='100%'}
-visualize_metadata(srat, meta = "fetal_full_predicted.organ", group.by = "seurat_clusters")
+```{r fig.width=25, fig.height=6,  results='asis'}
+tmp <- table(cell_type_df$sample_id, cell_type_df$organ)
+
+knitr::kable(tmp)
 ```
 
-```{r fig.width=30, fig.height=20, out.width='100%'}
-tmp <- table(srat$fetal_full_predicted.organ)
-percent.of.kidney <- tmp["Kidney"] / sum(tmp) * 100
+```{r fig.width=25, fig.height=6}
+do_Table_Heatmap(first_group = "sample_id", last_group = "organ")
 ```
 
-Of note, `r percent.of.kidney` % of the cells are labeled as kidney cells.
+### Predicted compartment
 
-```{r fig.width=15, fig.height=20, out.width='100%'}
-visualize_metadata(srat, meta = "fetal_full_predicted.annotation.l1", group.by = "seurat_clusters")
-visualize_metadata(srat, meta = "fetal_full_predicted.annotation.l2", group.by = "seurat_clusters")
-```
-
-#### Annotation from the kidney reference (Stewart et al)
+The predicted compartment is the result of the label transfer fron the human fetal kidney atlas: 
+[Stewart et al.](https://www.science.org/doi/10.1126/science.aat5031) created a [human fetal kidney atlas](https://www.kidneycellatlas.org/). 
+This reference contains only fetal kidney cells and has been precisely annotated by kidney experts. 
+The label transfer have been performed in the notebook `02b_fetal_kidney_reference_Stewart_{sample_id}.Rmd`
 
@@
 
-```{r fig.width=10, fig.height=10, out.width='100%'}
-visualize_metadata(srat, meta = "fetal_kidney_predicted.compartment", group.by = "seurat_clusters")
+We think that the first level of annotation `fetal_kidney_predicted.compartment` can be useful in identifying non-cancer cells, at least the endothelial and immune cells.
+
+```{r fig.width=25, fig.height=3, results='asis'}
+tmp <- table(cell_type_df$sample_id, cell_type_df$compartment)
+
+knitr::kable(tmp)
 ```
 
-```{r fig.width=30, fig.height=20, out.width='100%'}
-tmp <- table(srat$fetal_kidney_predicted.compartment)
-percent.of.immune <- 100*tmp["immune"]/sum(tmp)
-percent.of.endothelial <- 100*tmp["endothelium"]/sum(tmp)
+```{r fig.width=25, fig.height=3}
+
+do_Table_Heatmap(first_group = "sample_id", last_group = "compartment")
 ```
 
-Of note, `r percent.of.immune` % of the cells are labeled as immune cells and `r percent.of.endothelial` % of the cells are labeled as endothelial cells.
+### Evaluate annotations with marker genes
 
-Let's have a look at the repartition of cells per patient:
-
-```{r fig.width=10, fig.height=50, out.width='100%'}
-table(srat$sample_id, srat$fetal_kidney_predicted.compartment)
-```
-
-```{r fig.width=10, fig.height=20, out.width='100%'}
-visualize_metadata(srat, meta = "fetal_kidney_predicted.cell_type", group.by = "seurat_clusters")
-```
-
-### Compare annotations with marker genes
-
-#### "fetal_kidney_predicted.compartment"
-
-Here we evaluate with marker genes the identification of endothelial, immune cells and other cancer and non-cancerous sub-populations.
-
-```{r fig.height=8, fig.width=8, message=FALSE, warning=FALSE, out.width='100%'}
-
-markers = list("malignant" = CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "malignant")],
-            "immune"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "immune")],
-            "endothelium"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "endothelium")],
-            "non-malignant"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "non-malignant")])
-
-SCpubr::do_DotPlot(sample = srat,
-                         group.by =  "fetal_kidney_predicted.compartment",
-                         features = markers, 
-                         axis.text.x.angle = 90, 
-                         flip = TRUE, 
-                         plot.title = "fetal_kidney_predicted.compartment"
-                         ) 
-```
+Here we evaluate with marker genes the identification of endothelial, immune cells sub-populations.
 
@@
 
-```{r fig.height=4, fig.width=8, message=FALSE, warning=FALSE, out.width='100%'}
-visualize_feature(object = srat,
-                  features = "ENSG00000110799",
-                  group.by = "fetal_kidney_predicted.compartment")
+```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
+do_Feature_mean(  feature = "ENSG00000110799",
+                  group.by = "compartment")
 ```
@@
 
-```{r fig.height=4, fig.width=8, message=FALSE, warning=FALSE, out.width='100%'}
-visualize_feature(object = srat,
-                  features = "ENSG00000081237",
-                  group.by = "fetal_kidney_predicted.compartment")
-```
-#### "fetal_kidney_predicted.cell_type"
-
-```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
-
-SCpubr::do_DotPlot(sample = srat,
-                         group.by =  "fetal_kidney_predicted.cell_type",
-                         features = markers, 
-                         axis.text.x.angle = 90, 
-                         flip = TRUE,
-                         plot.title = "fetal_kidney_predicted.cell_type"
-                          )
-
-```
-#### "fetal_full_predicted.annotation.l1"
-
-```{r fig.height=10, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
-
-SCpubr::do_DotPlot(sample = srat,
-                         group.by =   "fetal_full_predicted.annotation.l1",
-                         features = markers, 
-                         axis.text.x.angle = 90, 
-                         flip = TRUE,
-                         plot.title =  "fetal_full_predicted.annotation.l1"
-                          )
-```
-#### "cellassign_celltype_annotation"
-
-```{r fig.height=15, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
-SCpubr::do_DotPlot(sample = srat,
-                         group.by =  "cellassign_celltype_annotation",
-                         features = markers, 
-                         axis.text.x.angle = 90, 
-                         flip = TRUE,
-                         plot.title = "cellassign_celltype_annotation"
-                          )                         
-```
-
-### Compare annotations between them
-
-#### Annotation versus `Seurat` clusters
-
-Here we check wether one (or more) `Seurat` cluster do correspond to one label. 
-This would allow us to be more confident in the annotation of immune or endothelial cells for example.
-
-```{r fig.width=25, fig.height=25, out.width='100%'}
-p1 <- do_Table_Heatmap( srat,
-                            last_group = "fetal_kidney_predicted.compartment",
-                            first_group = "seurat_clusters"
-
-                            )
-
-p2 <- do_Table_Heatmap( srat,
-                            last_group = "fetal_full_predicted.annotation.l1",
-                            first_group = "seurat_clusters"
-                            )
-
-p3 <- do_Table_Heatmap( srat,
-                            last_group = "singler_celltype_annotation",
-                            first_group = "seurat_clusters"
-                            )
-
-p4 <- do_Table_Heatmap( srat,
-                            last_group = "cellassign_celltype_annotation",
-                            first_group = "seurat_clusters"
-                            )
-
-(p1|p2)/(p3|p4)
-```
-
-#### Comparison of the labels from the two fetal references
-
-##### first level annotation
-
-```{r fig.width=30, fig.height=12, out.width='100%'}
-p1 <- do_Table_Heatmap( srat,
-                            last_group = "fetal_kidney_predicted.compartment",
-                            first_group = "fetal_full_predicted.annotation.l1"
-
-                            )
-p2 <- do_Table_Heatmap( srat,
-                            last_group = "fetal_kidney_predicted.cell_type",
-                            first_group = "fetal_full_predicted.annotation.l1"
-
-                            )
-p1|p2
-```
-
-##### second level annotation
-
-```{r fig.width=30, fig.height=15, out.width='100%'}
-p2 <- do_Table_Heatmap( srat,
-                            last_group = "fetal_kidney_predicted.cell_type",
-                            first_group = "fetal_full_predicted.annotation.l2"
-
-                            )
-p2
-```
-
-#### Comparison of the  fetal reference with `singleR` and `cellassign`
-
-##### first level annotation
-
-```{r fig.width=30, fig.height=10, out.width='100%'}
-p1 <- do_Table_Heatmap( srat,
-                            last_group = "fetal_kidney_predicted.compartment",
-                            first_group = "cellassign_celltype_annotation"
-
-                            )
-p2 <- do_Table_Heatmap( srat,
-                            last_group = "fetal_kidney_predicted.compartment",
-                            first_group = "singler_celltype_annotation"
-
-                            )
-p1|p2
-```
-
-##### second level annotation
-
-```{r fig.width=30, fig.height=30, out.width='100%'}
-p1 <- do_Table_Heatmap( srat,
-                            last_group = "fetal_kidney_predicted.cell_type",
-                            first_group = "cellassign_celltype_annotation"
-
-                            )
-
-p2 <- do_Table_Heatmap( srat,
-                            last_group = "fetal_kidney_predicted.cell_type",
-                            first_group = "singler_celltype_annotation"
-
-                            )
-
-p1/p2
+```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
+do_Feature_mean(  feature = "ENSG00000081237",
+                  group.by = "compartment")
 ```

### 2024-09-24 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 date: '2024-08-13'
-params:
-  scpca_project_id: "SCPCP000006"
-  sample_id: "SCPCS000176"
-  seed: 12345
 output: 

### 2024-09-24 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
-The aim is to explore the annotations and label transfers for all of the samples in `r params$scpca_project_id`.
+The aim is to explore the annotations and label transfers for all of the samples in SCPCP000006.
 
@@
 
-set.seed(params$seed)
 ```
@@
 
-The fetal full reference (Cao et al.) provides three levels of annotations:
-
-- `fetal_full_predicted.organ` is one of the 15 organs used for building the reference. 
-We were nicely surprised that, for many of the samples, most of the cells from the Wilms tumor cohort mapped to fetal kidney cells.
-
-- `fetal_full_predicted.annotation.l1` holds predicted cell type annotation
-
-- `fetal_full_predicted.annotation.l2` holds the combination of the predicted organ and predicted cell type annotation
+This notebook looks at the `fetal_full_predicted.organ` annotation from this reference.
 

### 2024-09-24 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 author: "Maud PLASCHKA"
-date: '2024-08-13'
+date: "`r Sys.Date()`"
 output: 

### 2024-09-24 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
-- `fetal_kidney_predicted.cell_type` holds predicted cell type annotations that details further the four compartments, as for example:
-
-    - `immune` subtypes: myeloid cells, lymphocytes, ...
-
-    - `stroma` subtypes: mesenchymal stem cells, fibroblasts, ...
-
-    - `fetal_nephron` subtypes: podocytes, uteric bud, mesenchymal cells (that appear to be cap mesenchyme cells) ...
-
-    - `endothelial` cells
-
-We think that the first level of annotation `fetal_kidney_predicted.compartment` can be useful in identifying non-cancer cells, at least the endothelial and immune cells.
-
 ```{r fig.width=25, fig.height=3, results='asis'}

### 2024-09-25 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
changes to PR#776

@@
 ---
-title: "Annotation exploration for `r params$scpca_project_id`"
+title: "Annotation exploration for SCPCP000006"
 author: "Maud PLASCHKA"
@@
 library(tidyverse)
+library(tidyr)
 library(patchwork)
-
+library(DT)
 ```
@@
 # The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "current", params$scpca_project_id)
+data_dir <- file.path(repository_base, "data", "current", "SCPCP000006")
 
@@
 
-In this notebook, we are working with all of the samples in `r params$scpca_project_id`.
+In this notebook, we are working with all of the samples in SCPCP000006.
 
@@
 ```{r path_to_query}
-sample_metadata_file <- file.path(repository_base, "data", "current", params$scpca_project_id, "single_cell_metadata.tsv")
+sample_metadata_file <- file.path(repository_base, "data", "current","SCPCP000006", "single_cell_metadata.tsv")
 
@@
 
-#### do_Table_Heatmap
-
-`do_Table_Heatmap` shows heatmap of counts of cells for combinations of two metadata variables
-
-- `first_group` is the name of the first metadata to group the cells
-- `last_group` is the name of the second metadata to group the cells
-
-```{r fig.width=10, fig.height=4, out.width='100%'}
-do_Table_Heatmap <- function(first_group, last_group ){
-
-  df <- cell_type_df %>%
-  group_by( !!sym(first_group), !!sym(last_group))%>% 
-  summarise(Nb = n())  %>%
-  mutate(C = sum(Nb)) %>%
-  mutate(percent = Nb/C*100)
-
-p1 <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = Nb)) +
-   geom_tile() +
-   scale_fill_viridis_c()  +
-   theme_bw() +
-   theme(text = element_text(size = 20)) +
-   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
-   guides(fill=guide_legend(title=" Number of \n cells"))
-
-p2 <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = percent)) +
-   geom_tile() +
-   scale_fill_viridis_c()  +
-   theme_bw() +
-   theme(text = element_text(size = 20)) +
-   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
-   guides(fill=guide_legend(title=" Percentage of \n cells"))
-
-p <- p1 + p2 + plot_layout(ncol = 2)
-return(p)
-}
-```
-
 #### do_Feature_mean
@@
 
+- `df` is the name of the table containing metadata and feature expression (counts) per cells
 - `group.by` is the name of the metadata to group the cells
@@
 ```{r fig.width=10, fig.height=4, out.width='100%'}
-do_Feature_mean <- function(group.by, feature ){
+do_Feature_mean <- function(df, group.by, feature ){
 
-  df <- cell_type_df %>%
+  df <- df %>%
   group_by(sample_id, !!sym(group.by))%>% 
@@
    theme(text = element_text(size = 20)) +
-   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+   theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5)) +
    guides(fill=guide_legend(title=paste0(feature)))
+
+return(p)
+}
+```
+
+#### do_Feature_boxplot
+
+`do_Feature_boxplot` shows boxplot of expression of a feature grouped by a metadata.
+
+- `table` is the name of the table containing metadata and feature expression (counts) per cells
+- `group.by` is the name of the metadata to group the cells
+- `feature` is the name of the gene to average the expression
+
+```{r fig.width=10, fig.height=4, out.width='100%'}
+do_Feature_boxplot <- function(df, group.by, feature, split.by ){
+
+  df <- df %>%
+    mutate(!!sym(group.by), group = factor(!!sym(group.by), levels = unique(!!sym(group.by))))
+
+  xlabs <- paste(levels(df$group),"\n(N=",table(df$group),")",sep="")
+
+p <- ggplot(df, 
+    aes(x = !!sym(group.by), y = !!sym(feature), fill = !!sym(group.by))
+  ) + 
+  geom_boxplot(size = 0.5, size.outlier = 0.25) +
+  facet_wrap(vars(!!sym(split.by)), scales = "free", ncol = 4) + 
+  theme_bw() +
+  theme(
+    axis.text.x = element_text(size = 12, angle = 0, hjust = 0.5, vjust = 0), 
+    legend.position = "none"
+  ) +scale_x_discrete(labels=xlabs)
 
@@
 
-### Total number of cells per samples
-
-```{r fig.width=5, fig.height=10, out.width='100%', results='asis'}
-tmp <- table(cell_type_df$sample_id)
-knitr::kable(tmp)
-```
-### Predicted organ
+### Total number of cells per samples and predicted organ
 
@@
 
-```{r fig.width=25, fig.height=6,  results='asis'}
-tmp <- table(cell_type_df$sample_id, cell_type_df$organ)
+Here, we are summarizing per sample:
 
-knitr::kable(tmp)
-```
+- the percentage of cells labeled as kidney, after label transfer from the fetal full reference [Cao et al.](https://www.science.org/doi/10.1126/science.aba7721),
 
-```{r fig.width=25, fig.height=6}
-do_Table_Heatmap(first_group = "sample_id", last_group = "organ")
+- the percentage of cells not labeled as kidney, after label transfer from the fetal full reference [Cao et al.](https://www.science.org/doi/10.1126/science.aba7721),
+
+- the total number of cells per sample.
+
+```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
+kidney_count_df <- cell_type_df |>
+  # make a new variable that tells us if it's kidney or not
+  dplyr::mutate(organ_type = ifelse(organ == "Kidney", "kidney", "not_kidney"))  |>
+  # count how many kidney or not
+  dplyr::count(sample_id, organ_type) |>
+  # make the data frame wide and use a value of 0 if a count is missing
+  tidyr::pivot_wider(names_from = organ_type, values_from = n, values_fill = 0) |>
+  # add a column for total, and convert other counts to percentages
+  dplyr::mutate(
+    total = kidney + not_kidney, 
+    kidney = round(kidney/total, digits = 4)*100, 
+    not_kidney = round(not_kidney/total, digits = 4)*100) |>
+  # arrange in descending order of percentage of kidney cells
+  dplyr::arrange(desc(kidney))
+
+DT::datatable(kidney_count_df, 
+              caption = "percentage of cells mapping to the predicted organ kidney",
+              extensions = 'Buttons', 
+              options = list(  dom = 'Bfrtip',
+                               buttons = c( 'csv', 'excel')))
 ```
@@
 ```{r fig.width=25, fig.height=3, results='asis'}
-tmp <- table(cell_type_df$sample_id, cell_type_df$compartment)
+compartment_df <- cell_type_df |>
+  # count how many per compartment
+  dplyr::count(sample_id, compartment) |>
+  # make the table wide and use a value of 0 if a count is missing
+  tidyr::pivot_wider(names_from = compartment, values_from = n, values_fill = 0) |>
+  # order the columns how we want to show them
+  dplyr::select(sample_id, fetal_nephron, stroma, endothelium, immune)
 
-knitr::kable(tmp)
-```
-
-```{r fig.width=25, fig.height=3}
-
-do_Table_Heatmap(first_group = "sample_id", last_group = "compartment")
+DT::datatable(compartment_df, 
+              caption = "counts of cells in each compartment",
+              extensions = 'Buttons', 
+              options = list(  dom = 'Bfrtip',
+                               buttons = c( 'csv', 'excel')))
 ```
@@
 
-```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
-do_Feature_mean(  feature = "ENSG00000110799",
+```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
+do_Feature_mean( df = cell_type_df,
+                  feature = "ENSG00000110799",
                   group.by = "compartment")
+```
+
+```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
+do_Feature_boxplot( df = cell_type_df,
+                    feature = "ENSG00000110799",
+                    group.by = "compartment",
+                    split.by = "sample_id")
+
 ```
@@
 
-```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
-do_Feature_mean(  feature = "ENSG00000081237",
+```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
+do_Feature_mean(  df = cell_type_df,
+                  feature = "ENSG00000081237",
                   group.by = "compartment")
+```
+
+```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
+do_Feature_boxplot( df = cell_type_df,
+                    feature = "ENSG00000081237",
+                    group.by = "compartment",
+                    split.by = "sample_id")
+
 ```

### 2024-09-25 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
PR#776

@@
                       message=FALSE,
-                      warnings=FALSE
-                      )
+                      warnings=FALSE,
+                      knitr.digits.signif = FALSE)
 
@@
 
-- `table` is the name of the table containing metadata and feature expression (counts) per cells
+- `df` is the name of the table containing metadata and feature expression (counts) per cells
 - `group.by` is the name of the metadata to group the cells
@@
   df <- df %>%
-    mutate(!!sym(group.by), group = factor(!!sym(group.by), levels = unique(!!sym(group.by))))
-
-  xlabs <- paste(levels(df$group),"\n(N=",table(df$group),")",sep="")
+    mutate(!!sym(group.by), group  = factor(!!sym(group.by), levels = c("fetal_nephron", "stroma", "immune", "endothelium")))
 
   p <- ggplot(df, 
-    aes(x = !!sym(group.by), y = !!sym(feature), fill = !!sym(group.by))
+    aes(x = group, y = !!sym(feature), fill = group)
   ) + 
   geom_boxplot(size = 0.5, size.outlier = 0.25) +
-  facet_wrap(vars(!!sym(split.by)), scales = "free", ncol = 4) + 
+  facet_wrap(vars(!!sym(split.by)), scale = "free_y", ncol = 4) + 
   theme_bw() +
@@
     legend.position = "none"
-  ) +scale_x_discrete(labels=xlabs)
+  )
 
@@
 ```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
+
 kidney_count_df <- cell_type_df |>
@@
     total = kidney + not_kidney, 
-    kidney = round(kidney/total, digits = 4)*100, 
-    not_kidney = round(not_kidney/total, digits = 4)*100) |>
+    kidney = round(kidney/total,  4)*100, 
+    not_kidney = round(not_kidney/total, 4)*100) |>
   # arrange in descending order of percentage of kidney cells
@@
     - `endothelial` cells
+
+For each compartment, we sumarized the percentage of cells that do match kidney annotation or not:
+
+```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
+
+compartment_count_df <- cell_type_df |>
+  # make a new variable that tells us if it's kidney or not
+  dplyr::mutate(organ_type = ifelse(organ == "Kidney", "kidney", "not_kidney"))  |>
+  # count how many kidney or not
+  dplyr::count(compartment, organ_type) |>
+  # make the data frame wide and use a value of 0 if a count is missing
+  tidyr::pivot_wider(names_from = organ_type, values_from = n, values_fill = 0) |>
+  # add a column for total, and convert other counts to percentages
+  dplyr::mutate(
+    total = kidney + not_kidney, 
+    kidney = round(kidney/total,  4)*100, 
+    not_kidney = round(not_kidney/total, 4)*100) |>
+  # arrange in descending order of percentage of kidney cells
+  dplyr::arrange(desc(kidney))
+
+DT::datatable(compartment_count_df, 
+              caption = "percentage of cells mapping to the predicted organ kidney",
+              extensions = 'Buttons', 
+              options = list(  dom = 'Bfrtip',
+                               buttons = c( 'csv', 'excel')))
+```   
+What is the predicted organ of cells that are not labeled as kidney cell?
+
+```{r fig.width=25, fig.height=3, results='asis'}
+compartment_df <- cell_type_df |>
+  # count how many per compartment
+  dplyr::count(organ, compartment) |>
+  # make the table wide and use a value of 0 if a count is missing
+  tidyr::pivot_wider(names_from = compartment, values_from = n, values_fill = 0) |>
+  # order the columns how we want to show them
+  dplyr::select(organ, fetal_nephron, stroma, endothelium, immune)
+
+DT::datatable(compartment_df, 
+              caption = "counts of cells in each compartment",
+              extensions = 'Buttons', 
+              options = list(  dom = 'Bfrtip',
+                               buttons = c( 'csv', 'excel')))
+```
+
+We also checked the number of cell in each compartment per sample, to assess the presence/absence of non-cancer cells (endothelia and immune) that could help the inference of copy number alterations.
 

### 2024-09-27 analyses/metacells/notebooks/testing-seacells.Rmd
Initial notebook tests

---
title: "Testing SEACells results"
output: html_notebook
---

## Setup

```{r setup}
library(ggplot2)
library(reticulate)
reticulate::use_condaenv("openscpca-metacells")
```

```{python imports}
import pathlib

import matplotlib
import matplotlib.pyplot as plt
import scanpy as sc
import seaborn as sns
import SEACells
```

## Path definitions

```{r paths}
root_dir <- rprojroot::find_root(rprojroot::is_git_root)
data_dir <- file.path(root_dir, "data", "current")

infile <- file.path(data_dir, "SCPCP000001/SCPCS000001/SCPCL000001_processed_rna.h5ad")
```

## Load data
```{python load_data}
ad = sc.read(r.infile)

# put the highly variable genes in a column of adata.var
ad.var["highly_variable"] = ad.var.gene_ids.isin(ad.uns["highly_variable_genes"])
```

## Run SEACells

```{python run_seacells}
n_SEACells = round(ad.n_obs / 75)
n_waypoint_eigs = 10

model = SEACells.core.SEACells(
    ad,
    build_kernel_on="X_pca",
    n_SEACells=n_SEACells,
    n_waypoint_eigs=n_waypoint_eigs,
    convergence_epsilon=1e-5,
)
```
```{python}
model.construct_kernel_matrix()
M = model.kernel_matrix
```

```{python}
sns.clustermap(M.toarray()[:500, :500])
plt.show()
```

```{python}
model.initialize_archetypes()
SEACells.plot.plot_initialization(ad, model)
```

```{python}
model.fit(min_iter=10, max_iter=50)
model.plot_convergence()
```

```{python}
SEACells.plot.plot_2D(ad, key="X_umap", colour_metacells=True)
```

```{python}
SEACells.plot.plot_SEACell_sizes(ad, bins=10)
```

```{python}
purity = SEACells.evaluate.compute_celltype_purity(
    ad, "singler_celltype_ontology"
)
```

```{r purity_plot}
ggplot(py$purity, aes(y = singler_celltype_ontology_purity)) +
  geom_boxplot() +
  labs(title = "Celltype purity",
       y = "Celltype purity (singler)")
```

### 2024-09-27 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>
PR#776

@@
 
-For each compartment, we sumarized the percentage of cells that do match kidney annotation or not:
+For each compartment, we sumarized the percentage of cells that do match kidney annotation or not. 
+Please note that this table is <u>not</u> sample-specific but contains all samples pooled into one.
 
@@
 ```
+
 What is the predicted organ of cells that are not labeled as kidney cell?
+Please note that this table is <u>not</u> sample-specific but contains all samples pooled into one.
 
@@
 Here we evaluate with marker genes the identification of endothelial, immune cells sub-populations.
+We expect that cells labeled as endothelial and immune cells will have higher expression of these markers compared to other cell types.
 
@@
 
+## Conclusion
+
+### Assessment of the quality of label transfer
+
+Looking at the `predicted.organ` label (from the human fetal atlas, [Cao et al.](https://www.science.org/doi/10.1126/science.aba7721)), we can be confident about the label transfer.
+
+The majority of fetal nephron cell (92%) has been predicted as kidney.
+The fact that the other `compartments` (i.e.
+stroma, endothelial and immune) do not really match to kidney cells (less than one third predicted as kidney) shouldn't be a concern and and shouldn't be interpreted as a poor label transfer or annotation.
+
+It is indeed known that Wilms tumor stroma (sometimes) shows (unexpected) differentiation into cell types such as skeletal muscle cells, fat tissue, cartilage, bone and even glial cells  [[1](https://academic.oup.com/jnci/article/89/15/1089/2526288?login=false)-[2](https://www.ncbi.nlm.nih.gov/books/NBK373364/)].
+For that reason, we are not surprised that most stroma cells are not predicted as kidney cells.
+
+Regarding the immune compartment, it is not of a surprise that cancer cells, and/or treatment, modulate the immune microenvironment, via the attraction of immune cells that are not usually in the kidney and/or induction of a cancer-associated phenotype.
+For that reason, we are not surprised that immune cells can be resembling immune cells from other organs.
+
+### Next step
+
+The next step of the analysis is to test whether `copyKAT` can help us distinguishing cancer cells from non-malignant cell types.
+[`copyKAT`](https://github.com/navinlabcode/copykat) (Copynumber Karyotyping of Tumors) is a computational tool using integrative Bayesian approaches to identify genome-wide aneuploidy in single cells to separate tumor cells from normal cells, using high-throughput sc-RNAseq data.
+The underlying logic for calculating DNA copy number events from RNAseq data is that gene expression levels of many adjacent genes can provide depth information to infer genomic copy number in that region.
+
+Of note, it is known that `CopyKAT` has difficulty in predicting tumor and normal cells in the cases of pediatric and liquid tumors that have a few CNAs, such as Wilms tumors. 
+`CopyKAT` provides two ways to bypass this:
+
+- by providing a vector of cell names of known normal cells from the same dataset
+
+- or by try to search for T cells.
+
+For that reason, we will first try to run `CopyKAT` on 2-5 samples that have a decent amount (>100 cells when possible) of normal cells (endothelial and immune cells) and for which we are quite confident with the label transfer:
+
+- sample `SCPCS000194` has > `85` % of cells predicted as kidney and `234` + `83` endothelium and immune cells.
+
+- sample `SCPCS000179` has > `94` % of cells predicted as kidney and `25` + `111` endothelium and immune cells.
+
+- sample `SCPCS000184` has > `96` % of cells predicted as kidney and `39` + `70` endothelium and immune cells.
+
+- sample `SCPCS000205` has > `89` % of cells predicted as kidney and `92` + `76` endothelium and immune cells.
+
+- sample `SCPCS0000208` has > `95` % of cells predicted as kidney and `18` + `35` endothelium and immune cells.
+
+For the selected samples, we will run `copyKAT` with and without providing a list of normal cells. 
+That will be useful to us to see how consistent results are.
+Also it will inform us about how we might analyze samples that aren't predicted to have that many (or sometimes any at all!) endothelial or immune cells.
+
 ## Session Info

### 2024-10-08 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd
explore CNV results

---
title: "Copykat and infercnv CNV results exploration for `r params$sample_id`"
author: "Maud PLASCHKA"
date: '2024-08-13'
params:
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000194"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE
                      )

```

## Introduction

```{r}

subdiagnosis <- readr::read_tsv(
  file.path("..", "..", "..", "data", "current", params$scpca_project_id, "single_cell_metadata.tsv"),
  show_col_types = FALSE
  ) |>
  dplyr::filter(scpca_sample_id == params$sample_id) |>
  dplyr::pull(subdiagnosis)

```

This notebook explores using [`CopyKAT`](https://github.com/navinlabcode/copykat) or [`infercnv`](https://github.com/broadinstitute/inferCNV/wiki) to estimate tumor and normal cells in `r params$sample_id` from SCPCP000006. 
This sample has a(n) `r subdiagnosis` subdiagnosis.

`CopyKAT` was run using the `05_copyKAT.R` script with and without a normal reference, using 2 different methods to calculate the distance, namely `euclidean` or `spearman`.

`infercnv` was run using the `06_inferCNV.R` script with and without a normal reference. We also tested the impact of the subselection of normal cells using either immune, and/or endothelial cells as healthy reference.

In this notebook, we just want to compare the heatmaps of CNV profiles, and evaluate how comparable are the methods and how snesible they are to key parameters such as choice of distance or choice of healthy reference.

### Base directories

```{r base paths, eval=TRUE}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
```

### Input files

The input for this notebook are the results of `05_copyKAT.R` and `06_inferCNV.R`

```{r}

result_dir <- file.path(module_base, "results", params$sample_id)

```

### CopyKAT results

Below we look at the heatmaps produced by `CopyKAT`.

#### Heatmaps without reference

##### `distance` = `euclidean`

![](`r file.path(result_dir, "05_copyKAT", "noref", "euclidean", glue::glue("{params$sample_id}_copykat_heatmap.jpeg"))`)

##### `distance` = `spearman`

![](`r file.path(result_dir, "05_copyKAT", "noref", "spearman", glue::glue("{params$sample_id}_copykat_heatmap.jpeg"))`)

#### Heatmasp with endothelial cells as reference

##### `distance` = `euclidean`

![](`r file.path(result_dir, "05_copyKAT", "ref", "euclidean", glue::glue("{params$sample_id}_copykat_heatmap.jpeg"))`)

##### `distance` = `spearman`

![](`r file.path(result_dir, "05_copyKAT", "ref", "spearman", glue::glue("{params$sample_id}_copykat_heatmap.jpeg"))`)

### infercnv results

#### Heatmaps without reference
![](`r file.path(result_dir, "06_infercnv", "none", "infercnv.png")`)

#### Heatmaps with immune cells as reference
![](`r file.path(result_dir, "06_infercnv", "immune", "infercnv.png")`)

#### Heatmaps with endothelium cells as reference
![](`r file.path(result_dir, "06_infercnv", "endothelium", "infercnv.png")`)

#### Heatmaps with immune and endothelium cells as reference
![](`r file.path(result_dir, "06_infercnv", "both", "infercnv.png")`)

## Conclusions

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-10-08 analyses/cell-type-dsrct/exploratory_analysis/01-marker-gene-tumor-classification.Rmd

---
title: "EDA of DSRCT Samples"
author: Danh Truong
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
---

## Introduction

This notebook looks explores the data from the DSRCT sample set, `SCPCP000013`. 
We then see if we can use expression of DSRCT-specific genes  to manually classify tumor and normal cells.
The main goal of this notebook is only to identify tumor cells, identification and labeling of the other cells is a separate question that we do not answer here.

- First we look at expression of each of the DSRCT-specific genes  across all cells. 
- Then we use a z-transform prior to summing expression of all DSRCT-specific genes  
Cells with a z-score for any DSRCT-specific genes  > 0 are classified as tumor cells. 
- We anticipate that normal cells will not express DSRCT-specific genes.

## Setup
```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_bw()
)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "2024-07-08")
#sample_dir <- file.path(data_dir, "SCPCP000013", params$sample_id)

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-DSRCT")
```

```{r}
# source in helper functions: make_jaccard_matrix() and jaccard() 
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)
```

```{r}
metadata_file <- file.path(data_dir, "SCPCP000013", "single_cell_metadata.tsv")
metadata <- read.csv(metadata_file,  sep = '\t')
metadata_DSRCT <- dplyr::filter(metadata, diagnosis == 'Desmoplastic small round cell tumor')

```

```{r paths}
sce_file_list <- list()
for (i in 1:dim(metadata_DSRCT)[1])
{
  # Input files
  sce_file <- paste0(metadata_DSRCT$scpca_library_id[i], '_processed.rds')
  sce_file_list[[i]] <- file.path(data_dir, "SCPCP000013", metadata_DSRCT$scpca_sample_id[i], sce_file)
}

marker_genes <- file.path(module_base, "references", "tumor-marker-genes.tsv")

# output tumor/normal classifications
results_dir <- file.path(module_base, "results", "marker_gene_analysis")
fs::dir_create(results_dir)

#classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
#output_classifications_file <- file.path(results_dir, classifications_filename)
```

Read in each data as a separate object `SingleCellExperiment` in the list. 
```{r}
sce_list <- lapply(sce_file_list, function(x){
  # read in processed sce
  sce <- readr::read_rds(x)
})

#adding the sample id to the name of each SingleCellExperiment
names(sce_list) <- metadata_DSRCT$scpca_sample_id

# read in marker genes table 
marker_genes_df <- readr::read_tsv(marker_genes) |> 
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
  dplyr::distinct()

marker_genes_df

marker_genes <- marker_genes_df |> 
  dplyr::filter(cell_type == "tumor") |> 
  dplyr::pull(ensembl_gene_id)
```

## Analysis content

### Explore marker gene expression

The first thing we do here is just create a faceted UMAP showing the expression of each marker gene for tumor cells.

```{r}
umap_df_list <- lapply(sce_list, function(x) {
  # get the gene expression counts for all marker genes
  marker_gene_exp <- logcounts(x[marker_genes, ]) |>
    as.matrix() |>
    t() |>
    as.data.frame() |>
    tibble::rownames_to_column("barcodes")

  # pull out the UMAP coordinates and make a data frame to use for plotting
  umap_df <- x |>
    scuttle::makePerCellDF(use.dimred = "UMAP") |>
    # replace UMAP.1 with UMAP1
    dplyr::rename_with(\(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")) |>
    # add in marker gene expression to dataframe
    dplyr::left_join(marker_gene_exp, by = "barcodes") |>
    # combine all genes into a single column for easy faceting
    tidyr::pivot_longer(cols = starts_with("ENSG"),
                        names_to = "ensembl_gene_id",
                        values_to = "gene_expression") |>
    # join with marker gene df to get gene symbols for plotting
    dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
    dplyr::select(barcodes,
                  UMAP1,
                  UMAP2,
                  gene_symbol,
                  ensembl_gene_id,
                  gene_expression,
                  cluster)
})

names(umap_df_list) <- metadata_DSRCT$scpca_sample_id
```

```{r, fig.width=8}
for(i in 1:length(umap_df_list)){
  # faceted umap showing a umap panel for each marker gene
  p <- ggplot(umap_df_list[[i]], aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
    geom_point(alpha = 0.8, size = 0.1) +
    facet_wrap(vars(gene_symbol)) +
    scale_color_viridis_c() +
    labs(color = "Log-normalized gene expression") +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    theme(
      aspect.ratio = 1,
      legend.position = "bottom",
      axis.title = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
    ggtitle(names(umap_df_list)[i])
  print(p)
}

```

In my experience, ST6GALNAC5 is a strong marker for DSRCT in single-cell data. As can be seen, several of the samples have expression of ST6GALNAC5, but some do not, such as SCPCS000731 and SCPCS000729. In fact, these SCPCS000729 contains  low number of cells. These  samples in particular are the PDX samples and I will be excluding SCPCS000729 from further analyses.

```{r}
umap_df_list_excluded <- umap_df_list[!(names(umap_df_list) %in% c('SCPCS000731'))]
```

We can also look at the distributions for each marker gene. 
I would expect to see some sort of bimodal distribution separating cells that do and do not have expression of the marker gene. What is clear from these plots that ST6GALNAC5, CACNA2D2, PTPRQ, IQCJ-SCHIP1,  show a bi modal distribution among the cells. To some extent, we do see expression of the other markers.

```{r}
for(i in 1:length(umap_df_list_excluded)) {
  p <- ggplot(umap_df_list_excluded[[i]],
              aes(x = gene_expression, fill = gene_symbol)) +
    geom_density() +
    facet_wrap(vars(gene_symbol), scales = 'free_y') +
    theme(legend.position = "none") + 
    ggtitle(names(umap_df_list_excluded)[i])
  print(p)
}
```

Although we see some slight semblance of a bimodal distribution for most marker genes, it is hard to see and we cannot directly compare gene expression values across genes. 
This would make it hard to identify a cut off to categorize cells as expression or not expressing the marker gene.

Now we will transform each of the gene expression vectors by generating z-scores. 
Then we might be able to find a cut off we can use across samples for if marker genes are present in a cell or not.

```{r}
umap_df_list_excluded_scaled <- lapply(umap_df_list_excluded, function(x){
  x   |>
  dplyr::group_by(gene_symbol) |>
  # get z-scores for each gene
  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |>
  dplyr::ungroup()
})

```

Now we can create the same density plot but looking at z-scores. Interestingly, in some samples, like SPCS000489, we see that ST6GALNAC5 has negative values. Presumably, these are tumor cells but there may be a skew due to the number of tumor cells compared to normal.

```{r}

for(i in 1:length(umap_df_list_excluded_scaled)) {
  p <- ggplot(umap_df_list_excluded_scaled[[i]],
              aes(x = transformed_gene_expression, fill = gene_symbol)) +
    geom_density() +
    facet_wrap(vars(gene_symbol), scales = 'free_y') +
    theme(legend.position = "none") + 
    ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)
}
```

It looks like some marker genes have distinct groups of cells with z-score > 0, while other marker genes may not be as informative.

### Classify tumor cells using marker genes only

Let's try and use the marker gene expression to classify tumor cells. 
It looks like we could use a cutoff of z-score > 0 to count that cell as a tumor cell.

We could either count any cell that expresses at least one marker gene > 0 as a tumor cell, or look at the combined expression. 
Let's start with classifying tumor cells as tumor if any marker gene is present (z-score > 0). 
This also means the sum of all z-scores > 0.

Below, we can get the sum of the transformed gene expression of all marker genes and plot in a single UMAP.

```{r}
# calculate sum gene expression across all marker genes in list 
marker_sum_exp <- lapply(umap_df_list_excluded_scaled, function(x){
  x |>
  dplyr::group_by(barcodes) |> 
  dplyr::mutate(sum_exp = sum(transformed_gene_expression, na.rm = T)) |> 
  dplyr::select(barcodes, UMAP1, UMAP2, sum_exp, cluster) |> 
  dplyr::distinct()
})

# plot mean gene expression

for(i in 1:length(marker_sum_exp)) {
  p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_exp)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_viridis_c() +
    ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)
}

```
Similar to the individual plots, it looks like there is one group of cells on the bottom right that has the highest marker gene expression. 
We would anticipate that these are most likely to be the tumor cells.

Now let's classify any cell that has a sum of marker genes > 0 (after z-transformation) as tumor cells.

```{r}
# classify tumor cells based on presence of any marker genes 
marker_sum_exp <- lapply(marker_sum_exp, function(x){
  x |> 
  dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))})

for(i in 1:length(marker_sum_exp)) {
  p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_classification)) +
  geom_point(size = 0.5, alpha = 1) +
  ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)
}

```

This gives us a rough idea of cells that may be classified as tumor cells. However, I believe re-doing this analysis using the cluster level data may give us better results. This is highly dependent on whether the clusters were generated with enough granularity.

```{r}
# calculate sum gene expression across all marker genes in list 
marker_sum_exp <- lapply(umap_df_list_excluded_scaled, function(x){
  x |> 
  dplyr::group_by(cluster) |> #change to cluster
  dplyr::mutate(sum_exp = sum(transformed_gene_expression, na.rm = T)) |> 
  dplyr::select(barcodes, UMAP1, UMAP2, sum_exp, cluster) |> 
  dplyr::distinct()
})

# plot mean gene expression

for(i in 1:length(marker_sum_exp)) {
  p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_exp)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_viridis_c() +
    ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)
}

```
It looks like there are group of cells with high marker gene expression. We  anticipate that these are most likely to be the tumor cells. The groups with low or negative values are likely normal cells.

Now let's classify clusters that has a sum of marker genes > 0 (after z-transformation) as tumor cell clusters.

```{r}
# classify tumor cells based on presence of any marker genes 
marker_sum_exp <- lapply(marker_sum_exp, function(x){
  d <- density(x$sum_exp)

  x |> 
  dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))})

for(i in 1:length(marker_sum_exp)) {
  p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_classification)) +
  geom_point(size = 0.5, alpha = 1) +
  ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)
}

```

Based on my experiences, the above plots show that we are unable to adequately determine the tumor cells from the normal cells. This can be seen in SCPCS000729, which is most likely all tumor cells, since it is collected from a patient-derived xenograft.

## Conclusions

- We do see variation in DSRCT-specific gene expression across cells, suggesting that there may be gene drop-out reducing the quality of the data. 
- `ST6GALNAC5`, `PTRPQ`, and `IQCJ-SCHIP1` are good markers for the DSRCT cells within this data. The other markers are not as highly expressed either due to heterogeneity or data quality. 
- For the next steps, we may try to use `SingleR` or `CellAssign` to identify the normal cells and then work to identify remaining cells as tumor cells. Normal cells may have more definitive markers than tumor cells.

## Save outputs

```{r}
# get an RDS of the processed data
saveRDS(sce_file_list, '../results/SCPCP000013_sce_file_list.rds')
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-10-08 analyses/cell-type-wilms-tumor-06/notebook_template/05_cnv_copykat_exploration.Rmd

---
title: "Copykat CNV results exploration for `r params$sample_id`"
author: "Maud PLASCHKA"
date: '2024-08-13'
params:
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000194"
  distance: "spearman"
  seed: 12345
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE
                      )

```

## Introduction

```{r}

subdiagnosis <- readr::read_tsv(
  file.path("..", "..", "..", "data", "current", params$scpca_project_id, "single_cell_metadata.tsv"),
  show_col_types = FALSE
  ) |>
  dplyr::filter(scpca_sample_id == params$sample_id) |>
  dplyr::pull(subdiagnosis)

```

This notebook explores using [`CopyKAT`](https://github.com/navinlabcode/copykat) to estimate tumor and normal cells in `r params$sample_id` from SCPCP000006. 
This sample has a(n) `r subdiagnosis` subdiagnosis.

`CopyKAT` was run using the `copyKAT.R` script with and without a normal reference, using `r params$distance` to calculate distance in `copyKAT`.

These results are read into this notebook and used to:

- Visualize diploid and aneuploid cells on the UMAP. 
- Evaluate common copy number gains and losses in Wilms tumor. 
- Calculate the confusion matrix comparing manual annotations of tumor cells to using `CopyKAT` to annotate tumor cells. 
- Compare the annotations from `CopyKAT` to cell type annotations using label transfer and the fetal (kidney) references.

### Packages

```{r packages, message=FALSE, warning=FALSE}
library(Seurat)
library(SCpubr)             # for plotting 
library(tidyverse)
library(patchwork)

set.seed(params$seed)
```

### Base directories

```{r base paths, eval=TRUE}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", params$scpca_project_id)

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
```

### Input files

In this notebook, we are working with the Wilms tumor sample defined in `r params$sample_id` from the Wilms tumor dataset `r params$scpca_project_id`.
We work with the pre-processed and labeled `Seurat` object that is the output of `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd` saved in the `results` directory.

```{r}

result_dir <- file.path(module_base, "results", params$sample_id)

# predictions files 
predictions_file <- glue::glue("{params$sample_id}_copykat_prediction.txt")

predictions_paths <- c(
  no_ref = file.path(result_dir, "05_copyKAT", "noref", params$distance, predictions_file),
  with_ref = file.path(result_dir, "05_copyKAT", "ref", params$distance, predictions_file)
)

# full results gene by cell 
full_ck_result_file <- glue::glue("{params$sample_id}_copykat_CNA_results.txt")

full_ck_result_paths <- c(
  no_ref = file.path(result_dir, "05_copyKAT", "noref", params$distance, full_ck_result_file),
  with_ref = file.path(result_dir, "05_copyKAT", "ref", params$distance, full_ck_result_file)
)

```

### Output file

Reports will be saved in the `notebook` directory. 
The pre-processed and annotated `Seurat` object per samples are saved in the `result` folder.

## Functions

Here we defined function that will be used multiple time all along the notebook.

## Analysis

### Load the pre-processed `Seurat` object
```{r load, message=FALSE, warning=FALSE}
# open the processed rds object
srat <- readRDS(file.path(result_dir, paste0("02b-fetal_kidney_label-transfer_", params$sample_id,".Rds")))
DefaultAssay(srat) <- "SCT"
```

### CopyKAT results

Below we look at the heatmaps produced by `CopyKAT`.

#### Heatmap without reference

![](`r file.path(result_dir, "copyKAT", "noref", params$distance, glue::glue("{params$sample_id}_copykat_heatmap.jpeg"))`)

#### Heatmap with endothelial cells as reference

![](`r file.path(result_dir, "copyKAT", "ref", params$distance, glue::glue("{params$sample_id}_copykat_heatmap.jpeg"))`)

#### UMAP

Below we prepare and plot a UMAP that shows which cells are classified as diploid, aneuploid, and not defined by `CopyKAT`. 
We show a side by side UMAP with results from running `CopyKAT` both with and without a reference of normal cells.

```{r}
# read in ck predictions from both reference types (no_normal and with_normal)
ck_results_df <- predictions_paths |> 
  purrr::map(readr::read_tsv) |> 
  dplyr::bind_rows(.id = "reference_used")

# get umap coordinate
umap_df <- srat[["umap"]]@cell.embeddings |>
  as.data.frame() |>
  rownames_to_column('barcodes')

cnv_df <- umap_df |>
   dplyr::left_join(ck_results_df, by = c("barcodes" = "cell.names"))

```

```{r}
ggplot(cnv_df, aes(x = umap_1, y = umap_2, color = copykat.pred)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  facet_wrap(vars(reference_used))
```

### Validate common CNAs found in Wilms tumor

To validate some of these annotations, we can also look at some [commonly found copy number variations](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-wilms-tumor-06#the-table-geneticalterations_metadatacsv-contains-the-following-column-and-information) in Wilms tumor patients:

- Loss of Chr1p
- Gain of Chr1q
- Loss of Chr11p13
- Loss of Chr11p15
- Loss of Chr16q

Although these are the most frequent, there are patients who do not have any of these alterations and patients that only have some of these alterations. 
See [Tirode et al.,](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4264969/) and [Crompton et al.](https://doi.org/10.1158/2159-8290.CD-13-1037).

`CopyKAT` outputs a matrix that contains the estimated copy numbers for each gene in each cell. 
We can read that in and look at the mean estimated copy numbers for each chromosome across each cell. 
We might expect that tumor cells would show an increased estimated copy number in Chr1q, and/or a loss of Chr1p, Chr11p and Chr16q.

```{r}
# read in full gene by cell copy number detection results
full_ck_results_df <- full_ck_result_paths |>
  purrr::map(readr::read_tsv) |>
  dplyr::bind_rows(.id = "reference_used")

# for every cell, calculate the mean detection level across all genes in a given chromosome
full_cnv_df <- full_ck_results_df |>
  tidyr::pivot_longer(
    cols = -c(
      reference_used,
      chrom
    ),
    names_to = "barcodes",
    values_to = "cnv_detection"
  ) |>
  dplyr::group_by(chrom, barcodes, reference_used) |>
  dplyr::summarise(mean_cnv_detection = mean(cnv_detection))

# join with cnv info
cnv_df <- cnv_df |>
  dplyr::left_join(full_cnv_df, by = c("barcodes", "reference_used")) |>
  dplyr::filter(!is.na(chrom))
```

Let's look at the distribution of CNV estimation in cells that are called aneuploid and diploid by `CopyKAT`.

```{r, fig.height=15, fig.width=10}
# create faceted density plots showing estimation of CNV detection across each chr of interest
# colored by aneuploid/diploid estimation
ggplot(cnv_df, aes(x = mean_cnv_detection, color = copykat.pred)) +
  geom_density() +
  theme_bw() +
  facet_grid(
    rows = vars(chrom),
    cols = vars(reference_used)
  )
```

## Conclusions

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-10-09 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
Add initial code files, without much code yet but to set the stage for later

--- 2024-09-13 analyses/cell-type-dsrct/notebook-template.Rmd
@@
 ---
-title: "cell-type-dsrct Analysis"
-author:
-  - Your name here
+title: "Performing and evaluating graph-based clustering with rOpenScPCA"
 date: "`r Sys.Date()`"
-output: html_notebook
+output:
+  html_document:
+    df_print: paged
 ---
@@
 ## Introduction
-
-This a template for an analysis notebook using R Markdown.
-
-It is a good idea to start with a brief introduction to the analysis, including the purpose of the analysis, the data used, and the methods applied.
-Replace this text with your own introduction, and be sure to update the `title:` and `author:` fields in the YAML header at the top of the document.
-
-Don't forget to rename this file as well!
 
@@
 
-Load required packages in the following chunk, if needed.
-Do not install packages here; only load them with the `library()` function.
-
 ```{r packages}
 # load required packages
-# e.g. library(tidyverse)
+library(rOpenScPCA)
+suppressPackageStartupMessages({
+  library(SingleCellExperiment)
+  library(Seurat)
+  library(dplyr)
+  library(ggplot2)
+})
 ```
-
-You may also want to use the `package::function()` syntax within your code for better tracking of which functions come from which packages, and to avoid conflicts between functions with the same name in different packages.
 
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "cell-type-dsrct")
+module_base <- file.path(repository_base, "analyses", "hell-c")
 ```

### 2024-10-09 analyses/hello-clusters/run_hello-clusters.sh

@@ -1,0 +1,11 @@
+#!/bin/bash
+
+# This script renders this module's example notebooks that.
+# This script is primarily used for testing.
+# Please refer to the individual notebooks in this module for examples of performing and evaluating clustering with rOpenScPCA.
+
+set -euo pipefail
+
+# Ensure script is being run from its directory
+MODULE_DIR=$(dirname "${BASH_SOURCE[0]}")
+cd ${MODULE_DIR}

### 2024-10-15 analyses/cell-type-dsrct/exploratory_analysis/01-marker-gene-tumor-classification.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 ```{r paths}
-sce_file_list <- list()
-for (i in 1:dim(metadata_DSRCT)[1])
-{
-  # Input files
-  sce_file <- paste0(metadata_DSRCT$scpca_library_id[i], '_processed.rds')
-  sce_file_list[[i]] <- file.path(data_dir, "SCPCP000013", metadata_DSRCT$scpca_sample_id[i], sce_file)
-}
+sce_file_list <- file.path(
+  data_dir, "SCPCP000013",  
+  metadata_DSRCT$scpca_sample_id, 
+  paste0(metadata_DSRCT$scpca_library_id, "_processed.rds")
+)
 
@@
 ```{r}
-sce_list <- lapply(sce_file_list, function(x){
-  # read in processed sce
-  sce <- readr::read_rds(x)
-})
+sce_list <- lapply(sce_file_list, readr::read_rds)
 
 #adding the sample id to the name of each SingleCellExperiment
-names(sce_list) <- metadata_DSRCT$scpca_sample_id
+names(sce_list) <- metadata_DSRCT$scpca_library_id
 

### 2024-10-15 analyses/cell-type-wilms-tumor-06/notebook_template/05_copykat_exploration.Rmd
Start exploring copykat results

--- 2024-10-08 analyses/cell-type-wilms-tumor-06/notebook_template/05_cnv_copykat_exploration.Rmd
@@
 params:
-  scpca_project_id: "SCPCP000006"
   sample_id: "SCPCS000194"
-  distance: "spearman"
   seed: 12345
@@
 subdiagnosis <- readr::read_tsv(
-  file.path("..", "..", "..", "data", "current", params$scpca_project_id, "single_cell_metadata.tsv"),
+  file.path("..", "..", "..", "data", "current", "SCPCP000006", "single_cell_metadata.tsv"),
   show_col_types = FALSE
@@
 
-`CopyKAT` was run using the `copyKAT.R` script with and without a normal reference, using `r params$distance` to calculate distance in `copyKAT`.
+`CopyKAT` was run using the `copyKAT.R` script with and without a normal reference, using either an euclidean or statistical (spearman) method to calculate distance in `copyKAT`.
 
@@
 
-In this notebook, we are working with the Wilms tumor sample defined in `r params$sample_id` from the Wilms tumor dataset `r params$scpca_project_id`.
+In this notebook, we are working with the Wilms tumor sample defined in `r params$sample_id` from the Wilms tumor dataset "SCPCP000006".
 We work with the pre-processed and labeled `Seurat` object that is the output of `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd` saved in the `results` directory.
@@
 result_dir <- file.path(module_base, "results", params$sample_id)
+predictions_paths <- list()
+full_ck_result_paths <- list()
+for(ref_value in c("ref", "noref")){
+  for(distance_value in c("euclidean", "spearman")){
+    predictions_file <- glue::glue("05_copykat_", {params$sample_id}, "_",ref_value,"_distance-", distance_value, "_copykat_prediction.txt")
+    full_ck_result_file <- glue::glue("05_copykat_", {params$sample_id}, "_",ref_value,"_distance-", distance_value, "_copykat_CNA_results.txt")
 
-# predictions files 
-predictions_file <- glue::glue("{params$sample_id}_copykat_prediction.txt")
-
-predictions_paths <- c(
-  no_ref = file.path(result_dir, "05_copyKAT", "noref", params$distance, predictions_file),
-  with_ref = file.path(result_dir, "05_copyKAT", "ref", params$distance, predictions_file)
-)
-
-# full results gene by cell 
-full_ck_result_file <- glue::glue("{params$sample_id}_copykat_CNA_results.txt")
-
-full_ck_result_paths <- c(
-  no_ref = file.path(result_dir, "05_copyKAT", "noref", params$distance, full_ck_result_file),
-  with_ref = file.path(result_dir, "05_copyKAT", "ref", params$distance, full_ck_result_file)
-)
-
+    predictions_paths[[glue::glue(ref_value, "_", distance_value)]] <- file.path(result_dir, "05_copyKAT", ref_value, distance_value, predictions_file)
+    full_ck_result_paths[[glue::glue(ref_value, "_", distance_value)]] <- file.path(result_dir, "05_copyKAT", ref_value, distance_value, full_ck_result_file)
+  }
+}
 ```
@@
 
-![](`r file.path(result_dir, "copyKAT", "noref", params$distance, glue::glue("{params$sample_id}_copykat_heatmap.jpeg"))`)
+##### Euclidean distance
+
+![](`r file.path(result_dir, "05_copyKAT", "noref", "euclidean", glue::glue("05_copykat_", params$sample_id, "_noref_distance-euclidean_copykat_heatmap.png"))`)
+
+##### Spearman distance
+
+![](`r file.path(result_dir, "05_copyKAT", "noref", "spearman", glue::glue("05_copykat_", params$sample_id, "_noref_distance-spearman_copykat_heatmap.png"))`)
 
@@
 
-![](`r file.path(result_dir, "copyKAT", "ref", params$distance, glue::glue("{params$sample_id}_copykat_heatmap.jpeg"))`)
+##### Euclidean distance
+
+![](`r file.path(result_dir, "05_copyKAT", "ref", "euclidean", glue::glue("05_copykat_", params$sample_id, "_ref_distance-euclidean_copykat_heatmap.png"))`)
+
+##### Spearman distance
+
+![](`r file.path(result_dir, "05_copyKAT", "ref", "spearman", glue::glue("05_copykat_", params$sample_id, "_ref_distance-spearman_copykat_heatmap.png"))`)
 
@@
    dplyr::filter(!is.na(chrom))
+
 ```
@@
 
+From the heatmap of CNV and the mean CNV detection plots, it is difficult to see any CNV pattern that drive the identification of aneuploid cells. 
+The assignment of the aneuploidy/diploidy value might relies on very few CNV and/or an arbitrary threshold. 
+This might be why the assignement of aneuploidy/diploidy values differs between condition (and between runs!!).
+
 ## Session Info

### 2024-10-15 analyses/cell-type-wilms-tumor-06/scripts/explore-cnv-methods.R

@@ -4,6 +4,10 @@ repository_base <- rprojroot::find_root(rprojroot::is_git_root)
 # The path to this module
 module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
 
+# paths to the notebook templates and outputs directory
+notebook_template_dir <- file.path(module_base, "notebook_template")
+notebook_output_dir <- file.path(module_base, "notebook")
+
 # We run copyKAT and infercnv for a subselection of samples selected in "04_annotation_Across_Samples_exploration.Rmd" with and without healthy cells as reference
 
 for (sample_id in c("SCPCS000179",
@@ -38,3 +42,21 @@ for (sample_id in c("SCPCS000179",
   system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "06_infercnv.R"), " --sample_id ", sample_id, " --reference ", "none"))
   
 }
+
+
+# We explore `copykat` result rendering one notebook per sample tested
+
+for (sample_id in c("SCPCS000179",
+                    "SCPCS000184",
+                    "SCPCS000194", 
+                    "SCPCS000205",
+                    "SCPCS000208")){
+  
+  rmarkdown::render(input = file.path(notebook_template_dir, "05_copykat_exploration.Rmd"),
+                    params = list(sample_id = sample_id, seed = 12345),
+                    output_format = "html_document",
+                    output_file = paste0("05_copykat_exploration_", sample_id,".html"),
+                    output_dir = file.path(notebook_output_dir, sample_id))
+
+}
+

### 2024-10-15 analyses/cell-type-wilms-tumor-14/exploratory_analysis/03_copykat_SCPCL000850.Rmd
add copycat scripts

--- 2024-09-13 analyses/cell-type-dsrct/notebook-template.Rmd
@@
 ---
-title: "cell-type-dsrct Analysis"
-author:
-  - Your name here
+title: "Exploratory analysis using Copykat"
+author: "Jingxuan Chen"
 date: "`r Sys.Date()`"
-output: html_notebook
+output: html_document
 ---
@@
 
-Load required packages in the following chunk, if needed.
-Do not install packages here; only load them with the `library()` function.
-
 ```{r packages}
-# load required packages
-# e.g. library(tidyverse)
+library(dplyr)
+library(Seurat)
+library(copykat)
+library(ggplot2)
 ```
-
-You may also want to use the `package::function()` syntax within your code for better tracking of which functions come from which packages, and to avoid conflicts between functions with the same name in different packages.
 
@@
 
-In the chunk below, we include some common directories that you might need, finding their locations using the very useful [`rprojroot` package](https://rprojroot.r-lib.org/).
-
 ```{r base paths}
 # The base path for the OpenScPCA repository, found by its (hidden) .git directory
-repository_base <- rprojroot::find_root(rprojroot::is_git_root)
-
-# The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "current")
+path_repo <- rprojroot::find_root(rprojroot::is_git_root)
 
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "cell-type-dsrct")
+path_anal <- file.path(path_repo,"analyses","cell-type-wilms-tumor-14") 
 ```
@@
 
-Set paths to input and output directories and files in the chunk below.
+Only run for one sample
 
 ```{r paths}
+library <- "SCPCL000850"
 # Input files

### 2024-10-15 analyses/cell-type-wilms-tumor-06/notebook_template/05_copykat_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/05_copykat_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 author: "Maud PLASCHKA"
-date: '2024-08-13'
+date: "`r Sys.Date()`"
 params:

### 2024-10-15 analyses/cell-type-wilms-tumor-06/notebook_template/05_copykat_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/05_copykat_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
-`CopyKAT` was run using the `copyKAT.R` script with and without a normal reference, using either an euclidean or statistical (spearman) method to calculate distance in `copyKAT`.
+`CopyKAT` was run using the `05_copyKAT.R` script using either an euclidean or statistical (spearman) method to calculate distance in `copyKAT`. 
+`CopyKAT` was run with and without a normal reference.
+Immune and endothelial cells as identified by label transfer were used as the references cells where applicable.
 

### 2024-10-15 analyses/cell-type-wilms-tumor-06/notebook_template/00b_characterize_fetal_kidney_reference_Stewart.Rmd
use 'expected' kidney file name, but also make it a param for the notebook for consistency. notebook got styled too

@@
   seed: 12345
+  fetal_kidney_path: "scratch/fetal_kidney.rds"
 output: 
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE,
-                      message=FALSE,
-                      warnings=FALSE)
+knitr::opts_chunk$set(
+  echo = TRUE,
+  message = FALSE,
+  warnings = FALSE
+)
 ```
@@
 set.seed(params$seed)
-options(future.globals.maxSize= 891289600000)
+options(future.globals.maxSize = 891289600000)
 ```
@@
 module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
-
 ```
@@
 
-The input file is the output of the `R script` `download-and-create-fetal-kidney-ref.R`
-
-```{r path_to_data}
-path_to_data <- file.path(
-  module_base,
-  "scratch",
-  "fetal_kidney.rds"
-)
-download.file(url = params$url, destfile = path_to_data)
-```
+The input file `r params$fetal_kidney_path` is the output of the `R script` `prepare-fetal-references.R`.
 
@@
 ```{r pre_process, echo=TRUE, fig.height=7, fig.width=12, message=FALSE, warning=FALSE, out.width='100%'}
-fetal_kidney <- readRDS(path_to_data)
+fetal_kidney <- readRDS(params$fetal_kidney_path)
 
-d1 <- do_DimPlot(fetal_kidney, reduction = "umap", dims = c(1,2), group.by = "compartment", label = TRUE, repel = TRUE) + NoLegend() 
-d2 <- do_DimPlot(fetal_kidney, reduction = "umap", dims = c(1,2), group.by = "cell_type", label = TRUE, repel = TRUE) + NoLegend()
+d1 <- do_DimPlot(fetal_kidney, reduction = "umap", dims = c(1, 2), group.by = "compartment", label = TRUE, repel = TRUE) + NoLegend()
+d2 <- do_DimPlot(fetal_kidney, reduction = "umap", dims = c(1, 2), group.by = "cell_type", label = TRUE, repel = TRUE) + NoLegend()
 
@@
 ```{r markers_compatment, fig.width=8, fig.height=7, out.width='100%'}
-de_results   <- DElegate::FindAllMarkers2(fetal_kidney, group_column = "compartment",)
+de_results <- DElegate::FindAllMarkers2(fetal_kidney, group_column = "compartment", )
 
-#filter the most relevant markers
-s.markers <- de_results[de_results$padj < params$padj_threshold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold,]
+# filter the most relevant markers
+s.markers <- de_results[de_results$padj < params$padj_threshold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold, ]
 
-DT::datatable(s.markers, caption = ("marker genes"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-                               buttons = c( 'csv', 'excel')))
+DT::datatable(s.markers,
+  caption = ("marker genes"),
+  extensions = "Buttons",
+  options = list(
+    dom = "Bfrtip",
+    buttons = c("csv", "excel")
+  )
+)
 
@@
 
-p1 <- SCpubr::do_DimPlot(fetal_kidney, reduction="umap", group.by = "compartment", label = TRUE, repel = TRUE) + ggtitle("compartment")
+p1 <- SCpubr::do_DimPlot(fetal_kidney, reduction = "umap", group.by = "compartment", label = TRUE, repel = TRUE) + ggtitle("compartment")
 p2 <- DoHeatmap(ss, features = top5$feature, cells = cells, group.by = "compartment") + NoLegend() +
-  scale_fill_gradientn(colors =  c("#01665e","#35978f",'darkslategray3', "#f7f7f7", "#fee391","#fec44f","#F9AD03")) 
-p3 <- ggplot(fetal_kidney@meta.data, aes(compartment, fill = compartment)) + geom_bar() + NoLegend()
+  scale_fill_gradientn(colors = c("#01665e", "#35978f", "darkslategray3", "#f7f7f7", "#fee391", "#fec44f", "#F9AD03"))
+p3 <- ggplot(fetal_kidney@meta.data, aes(compartment, fill = compartment)) +
+  geom_bar() +
+  NoLegend()
 
 common_title <- sprintf("Unsupervised clustering %s, %d cells", fetal_kidney@meta.data$orig.ident[1], ncol(fetal_kidney))
-show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) ) + plot_layout(widths = c(1, 2)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))
+show((((p1 / p3) + plot_layout(heights = c(3, 2)) | p2)) + plot_layout(widths = c(1, 2)) + plot_layout(heights = c(3, 1)) + plot_annotation(title = common_title))
 
 write_csv(de_results, file = file.path(path_to_output, "00a_marker_compartment_fetal_kidney_Stewart.csv"))
-
 ```
@@
 
-#filter the most relevant markers
-s.markers <- de_results[de_results$padj < params$padj_threshold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold,]
+# filter the most relevant markers
+s.markers <- de_results[de_results$padj < params$padj_threshold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold, ]
 
-DT::datatable(s.markers, caption = ("marker genes"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-                               buttons = c( 'csv', 'excel')))
+DT::datatable(s.markers,
+  caption = ("marker genes"),
+  extensions = "Buttons",
+  options = list(
+    dom = "Bfrtip",
+    buttons = c("csv", "excel")
+  )
+)
 
@@
 
-p1 <- SCpubr::do_DimPlot(fetal_kidney, reduction="umap", group.by = "cell_type", label = TRUE, repel = TRUE) + ggtitle("cell_type") + NoLegend()
+p1 <- SCpubr::do_DimPlot(fetal_kidney, reduction = "umap", group.by = "cell_type", label = TRUE, repel = TRUE) + ggtitle("cell_type") + NoLegend()
 p2 <- DoHeatmap(ss, features = top5$feature, cells = cells, group.by = "cell_type") + NoLegend() +
-  scale_fill_gradientn(colors =  c("#01665e","#35978f",'darkslategray3', "#f7f7f7", "#fee391","#fec44f","#F9AD03")) 
-p3 <- ggplot(fetal_kidney@meta.data, aes(cell_type, fill = cell_type)) + geom_bar() + NoLegend() + scale_x_discrete(guide = guide_axis(angle = 90))
+  scale_fill_gradientn(colors = c("#01665e", "#35978f", "darkslategray3", "#f7f7f7", "#fee391", "#fec44f", "#F9AD03"))
+p3 <- ggplot(fetal_kidney@meta.data, aes(cell_type, fill = cell_type)) +
+  geom_bar() +
+  NoLegend() +
+  scale_x_discrete(guide = guide_axis(angle = 90))
 
 common_title <- sprintf("Unsupervised clustering %s, %d cells", fetal_kidney@meta.data$orig.ident[1], ncol(fetal_kidney))
-show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) ) + plot_layout(widths = c(1, 1)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))
+show((((p1 / p3) + plot_layout(heights = c(3, 2)) | p2)) + plot_layout(widths = c(1, 1)) + plot_layout(heights = c(3, 1)) + plot_annotation(title = common_title))
 
-write_csv( de_results, file = file.path(path_to_output, "00a_marker_cell-type_fetal_kidney_Stewart.csv"))
-
+write_csv(de_results, file = file.path(path_to_output, "00a_marker_cell-type_fetal_kidney_Stewart.csv"))
 ```

### 2024-10-15 analyses/cell-type-wilms-tumor-06/00_run_workflow.R

@@ -36,7 +36,7 @@ module_base <- file.path(root_dir, "analyses", "cell-type-wilms-tumor-06")
 
 # Download and create the references for label transfer, and download the homolog file for ID mapping ----------
 kidney_ref_url <- "https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds"
-kidney_ref_file <- file.path(module_base, "scratch", "kidney_ref.rds")
+kidney_ref_file <- file.path(module_base, "scratch", "fetal_kidney.rds")
 if (!file.exists(kidney_ref_file)) {
   download.file(kidney_ref_url, kidney_ref_file)
 }
@@ -63,7 +63,8 @@ rmarkdown::render(
   input = file.path(notebook_template_dir, "00b_characterize_fetal_kidney_reference_Stewart.Rmd"),
   output_format = "html_document",
   output_file = "00b_characterization_fetal_kidney_reference_Stewart.html",
-  output_dir = file.path(notebook_output_dir, "00-reference")
+  output_dir = file.path(notebook_output_dir, "00-reference"),
+  params = list(fetal_kidney_path = kidney_ref_file)
 )
 
 

### 2024-10-16 analyses/cell-type-wilms-tumor-06/notebook_template/00b_characterize_fetal_kidney_reference_Stewart.Rmd
fix relative path

@@
   seed: 12345
-  fetal_kidney_path: "scratch/fetal_kidney.rds"
+  fetal_kidney_path: "../scratch/fetal_kidney.rds"
 output: 

### 2024-10-16 analyses/cell-type-wilms-tumor-06/00_run_workflow.sh

@@ -45,7 +45,7 @@ Rscript -e "rmarkdown::render('${notebook_template_dir}/00b_characterize_fetal_k
     output_format = 'html_document',
     output_file = '00b_characterization_fetal_kidney_reference_Stewart.html',
     output_dir = '${notebook_output_dir}/00-reference',
-    params = list(fetal_kidney_path = '${kidney_ref_file}'))"
+    params = list(fetal_kidney_path = '../${kidney_ref_file}'))" # path relative to notebook
 
 
 # Run the label transfer and cluster exploration for all samples in the project

### 2024-10-16 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd
Start exploring infercnv results

@@
 ---
-title: "Copykat and infercnv CNV results exploration for `r params$sample_id`"
+title: "`infercnv` results exploration for `r params$sample_id`"
 author: "Maud PLASCHKA"
-date: '2024-08-13'
+date: "`r Sys.Date()`"
 params:
@@
   sample_id: "SCPCS000194"
+  seed: 12345
 output: 
@@
 
-This notebook explores using [`CopyKAT`](https://github.com/navinlabcode/copykat) or [`infercnv`](https://github.com/broadinstitute/inferCNV/wiki) to estimate tumor and normal cells in `r params$sample_id` from SCPCP000006.
+This notebook explores using [`infercnv`](https://github.com/broadinstitute/inferCNV/wiki) to estimate tumor and normal cells in `r params$sample_id` from SCPCP000006. 
 This sample has a(n) `r subdiagnosis` subdiagnosis.
-
-`CopyKAT` was run using the `05_copyKAT.R` script with and without a normal reference, using 2 different methods to calculate the distance, namely `euclidean` or `spearman`.
 
@@
 
-In this notebook, we just want to compare the heatmaps of CNV profiles, and evaluate how comparable are the methods and how snesible they are to key parameters such as choice of distance or choice of healthy reference.
+In this notebook, we just want to compare the heatmaps of CNV profiles, and evaluate how comparable are the methods and how sensible they are to key parameters such as selection of healthy reference.
+
+### Libraries
+
+```{r}
+
+library(infercnv)
+library(SCpubr)
+library(ggplot2)
+library(patchwork)
+
+```
+
+### Functions
+
+Here we defined function that will be used multiple time all along the notebook.
+
+#### Visualize CNV grouped by clusters or other metadata
+
+For a Seurat object `object`, the function `Do_CNV_heatmap` load the `infercnv` object created with the script `06_infercnv.R` using `reference_value` as a reference and call the function `SCpubr::do_CopyNumberVariantPlot` to plot the mean CNV score in each group defined in `group.by`.
+
+- `object` is the Seurat object
+
+- `reference_value` is the selection of normal cells used for `infercnv`
+
+- `group.by` is the metadata used for grouping the violin plots
+
+```{r}
+
+Do_CNV_heatmap <- function(object, reference_value, group.by){
+
+  infercnv_obj <- readRDS(file.path(result_dir, reference_value , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, ".rds")))
+
+  chromosome_locations = SCpubr::human_chr_locations
+
+  out <- SCpubr::do_CopyNumberVariantPlot(sample = srat,
+                                        infercnv_object = infercnv_obj,
+                                        using_metacells = FALSE,
+                                        chromosome_locations = chromosome_locations,
+                                        return_object = FALSE,
+                                        group.by = group.by
+                                        )
+  return(out)
+}
+
+```
+
+#### Visualize seurat clusters and metadata
+
+For a Seurat object `object`and a metadata `metadata`, the function `visualize_metadata` will plot `FeaturePlot` and `BarPlot`
+
+- `object` is the Seurat object
+
+- `metadata` the gene or quantitative value to be plotted
+
+- `group.by` is the metadata used for grouping the violin plots
+
+```{r }
+visualize_metadata <- function(object, meta, group.by){
+
+  if(is.numeric(object@meta.data[,meta])){
+     d <- SCpubr::do_FeaturePlot(object, 
+                                              features = meta, 
+                                              pt.size = 0.2, 
+                                              legend.width = 0.5, 
+                                              legend.length = 5, 
+                                              legend.position = "right") + ggtitle(meta)
+    b <- SCpubr::do_ViolinPlot(srat, 
+                                             features = meta, 
+                                             ncol = 1, 
+                                             group.by = group.by, 
+                                             legend.position = "none")
+
+   return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
+  }
+
+  else{
+
+  d <- SCpubr::do_DimPlot(object, reduction="umap", group.by = group.by, label = TRUE, repel = TRUE) + ggtitle(paste0(meta," - umap")) + theme(text=element_text(size=18))
+  b <- SCpubr::do_BarPlot(sample = object,
+                         group.by = meta,
+                         split.by = group.by,
+                         position = "fill",
+                         font.size = 10,
+                         legend.ncol = 3) +
+                         ggtitle("% cells")+
+                         xlab(print(group.by)) +
+                         theme(text=element_text(size=18))
+  return(d + b + plot_layout(ncol = 2, widths = c(2,4)))
+  }
+
+}
+```
+
+#### Visualize seurat clusters and markers genes
+
+For a Seurat object `object`and a features `features`, the function `visualize_feature` will plot `FeaturePlot` and `ViolinPlot`
+
+- `object` is the Seurat object
+
+- `features` the gene or quantitative value to be plotted
+
+- `group.by` is the metadata used for grouping the violin plots
+
+```{r }
+visualize_feature <- function(object, features, group.by ){
+
+                  d <- SCpubr::do_FeaturePlot(object, 
+                                              features = features, 
+                                              pt.size = 0.2, 
+                                              legend.width = 0.5, 
+                                              legend.length = 5, 
+                                              legend.position = "right") + ggtitle(features)
+                  b <- SCpubr::do_ViolinPlot(srat, 
+                                             features = features, 
+                                             ncol = 1, 
+                                             group.by = group.by, 
+                                             legend.position = "none",
+                                             assay = "SCT") + ylab(features)
+
+                  return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
+}
+```
+
+#### Visualize CNV density
+
+For a Seurat object `object`and a features `features`, the function `visualize_feature` will plot `FeaturePlot` and `ViolinPlot`
+
+- `object` is the Seurat object
+
+- `features` the gene or quantitative value to be plotted
+
+- `group.by` is the metadata used for grouping the violin plots
+
+```{r }
+visualize_density <- function(object, features, group.by ){
+
+                  d <- SCpubr::do_RidgePlot(object, 
+                                            feature = features, 
+                                            group.by = group.by, 
+                                             legend.position = "none",
+                                             assay = "SCT") + ylab(features)
+
+                  return(d) 
+}
+
+```
 
@@
 
-The input for this notebook are the results of `05_copyKAT.R` and `06_inferCNV.R`
+The input for this notebook are the results of `06_inferCNV.R`
 
@@
 
-result_dir <- file.path(module_base, "results", params$sample_id)
+result_dir <- file.path(module_base, "results", params$sample_id, "06_infercnv")
 
@@
 
-### CopyKAT results
+#### `Seurat` object
 
-Below we look at the heatmaps produced by `CopyKAT`.
+We load the `Seurat` object generated in `06_infercnv.R`
 
-#### Heatmaps without reference
+```{r}
 
-##### `distance` = `euclidean`
+srat <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")))
 
-![](`r file.path(result_dir, "05_copyKAT", "noref", "euclidean", glue::glue("{params$sample_id}_copykat_heatmap.jpeg"))`)
+```
 
-##### `distance` = `spearman`
+## Analysis
 
-![](`r file.path(result_dir, "05_copyKAT", "noref", "spearman", glue::glue("{params$sample_id}_copykat_heatmap.jpeg"))`)
+### Heatmap of infercnv results
 
-#### Heatmasp with endothelial cells as reference
+Here we plot the output of `infercnv` as heatmaps of CNV.
+We first look at the png file generated by the `infercnv` function.
+We then used the `infercnv object` to look at mean CNV value across compartments (immune, endothelial, stroma and fetal nephron).
 
-##### `distance` = `euclidean`
+#### Without reference
 
-![](`r file.path(result_dir, "05_copyKAT", "ref", "euclidean", glue::glue("{params$sample_id}_copykat_heatmap.jpeg"))`)
+```{r fig.width=20, fig.height=8, out.width='100%'}
 
-##### `distance` = `spearman`
+reference_value = "reference-none"
 
-![](`r file.path(result_dir, "05_copyKAT", "ref", "spearman", glue::glue("{params$sample_id}_copykat_heatmap.jpeg"))`)
+```
 
-### infercnv results
+![](`r file.path(result_dir, reference_value , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, "_heatmap.png"))`)
 
-#### Heatmaps without reference
-![](`r file.path(result_dir, "06_infercnv", "none", "infercnv.png")`)
+```{r fig.width=20, fig.height=8, out.width='100%'}
 
-#### Heatmaps with immune cells as reference
-![](`r file.path(result_dir, "06_infercnv", "immune", "infercnv.png")`)
+Do_CNV_heatmap(object = srat, reference_value, group.by = "fetal_kidney_predicted.compartment")
 
-#### Heatmaps with endothelium cells as reference
-![](`r file.path(result_dir, "06_infercnv", "endothelium", "infercnv.png")`)
+```
 
-#### Heatmaps with immune and endothelium cells as reference
-![](`r file.path(result_dir, "06_infercnv", "both", "infercnv.png")`)
+#### With immune cells as reference
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+reference_value = "reference-immune"
+
+```
+
+![](`r file.path(result_dir, reference_value , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, "_heatmap.png"))`)
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+Do_CNV_heatmap(object = srat, reference_value, group.by = "fetal_kidney_predicted.compartment")
+
+```
+
+#### With endothelium cells as reference
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+reference_value = "reference-endothelium"
+
+```
+
+![](`r file.path(result_dir, reference_value , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, "_heatmap.png"))`)
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+Do_CNV_heatmap(object = srat, reference_value, group.by = "fetal_kidney_predicted.compartment")
+
+```
+
+#### With immune and endothelium cells as reference
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+reference_value = "reference-both"
+
+```
+
+![](`r file.path(result_dir, reference_value , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, "_heatmap.png"))`)
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+Do_CNV_heatmap(object = srat, reference_value, group.by = "fetal_kidney_predicted.compartment")
+
+```
+
+### Explore infercnv results generated with immune and endothelial cells as reference, using a HMM i6 model
+
+#### Load the `Seurat` object
+
+We load the `Seurat` object generated in `06_infercnv.R`
+
+```{r}
+
+srat <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_", params$sample_id, "_reference-both.rds")))
+
+```
+
+#### Plot UMAP of known Wilms tumor related alteration
+
+As a reminder, we plot the label transfer from the fetal kidney atlas.
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+visualize_metadata(srat, "seurat_clusters", "fetal_kidney_predicted.compartment")
+
+```
+
+#### Feature plot and repartition of the CNV per chromosome
+
+For each chromosome, we look at the repartition of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
+`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+
+This should allow a quick visual check that immune and endothelial cells do not expressed any CNV. If so, we should check if genes driving the CNV score are not key immune or endothelium genes.
+
+Also, we should visualize if one area of the umap do not have CNV and could be identify as additional normal cell subset.
+
+```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
+for(i in 1:22){
+  print(visualize_feature(srat, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
+}
+```
+
+###  Distribution of CNV estimation in the Wilms tumor copartments
+
+For each chromosome, we look at the distribution of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
+`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+
+We are quite confident that immune and endothelial cells are well identified by label transfer done in `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd`. 
+The distribution of CNV for endothelial and immune cells should thus be a single peak center on 0.
+
+We do not know if fetal nephron and stroma cells are a mix of normal and cancer cells. 
+Would they be a group of normal cells, we should expect a single peak center on 0 for every chromosome. 
+As we expect to have a large number of cancer with heterogeneous CNV, we should see multiple peaks.
+
+```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%'}
+for(i in 1:22){
+  print(visualize_density(srat, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
+}
+```
 
 ## Conclusions
+
+- Providing no reference is not a good option, as we think that most of the cells are cancer cells with few CNV. 
+Without healthy reference, `infercnv` take the mean of expression across the cells as a reference. 
+In case of events that are well shared by _a majority_ of the cells, they will be smoothed.
+
+- For some CNV and specific chromosome, the choice a the reference do not affect the result: ch18, ch17.
+- Some CNV values differs depending on the choice of the reference, for chr5 for example.
+
+This is an issue if we want to have the exact CNV pattern.
+
+There are three samples with a very low (<5) number of immune + endothelial cells: SCPCS000197 (4 endothelial), SCPCS000190 (1 immune), SCPCS000177	(2 immune) for which it might be problematic to run `infercnv`.
+
+We should maybe try to run `infercnv` on one sample, gradualy downsampling the reference to 50, 10, 5, 1 cell to see how it impact the results. 
+For this try, I would suggest running with the HMM i3 model, to have an additional idea of the feasibility to extract CNV features from `infercnv`.
 

### 2024-10-16 analyses/cell-type-wilms-tumor-06/scripts/explore-cnv-methods.R

@@ -50,5 +50,13 @@ for (sample_id in c("SCPCS000179",
                     output_file = paste0("05_copykat_exploration_", sample_id,".html"),
                     output_dir = file.path(notebook_output_dir, sample_id))
 
+  # We explore `infercnv` result rendering one notebook per sample tested
+
+  rmarkdown::render(input = file.path(notebook_template_dir, "06_cnv_infercnv_exploration.Rmd"),
+                  params = list(sample_id = sample_id, seed = 12345),
+                  output_format = "html_document",
+                  output_file = paste0("06_infercnv_exploration_", sample_id,".html"),
+                  output_dir = file.path(notebook_output_dir, sample_id))
+
 }
 

### 2024-10-17 analyses/cell-type-wilms-tumor-14/exploratory_analysis/03_cnv/03_copykat_SCPCL000850.Rmd
re arrange explore folder

---
title: "Exploratory analysis using Copykat"
author: "Jingxuan Chen"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

Here we test `CopyKat` to analyze the CNV profile for Wilms tumor sample (SCPCL000850), aiming to identify potential tumor cells from normal cells.

Before running this notebook, one should first generate `CopyKat` outputs as in `03_runCopyKat.R`.

## Setup

### Packages

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(Seurat)
  library(copykat)
  library(ggplot2)
})
```

### Paths

#### Base directories

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
path_repo <- rprojroot::find_root(rprojroot::is_git_root)

# The path to this module
path_anal <- file.path(path_repo,"analyses","cell-type-wilms-tumor-14") 
```

#### Input and output files

Only run for one sample

```{r paths}
library_id <- "SCPCL000850"
scratch_out_dir <- file.path(path_anal, "scratch", "03_cnv","copykat")
dir.create(scratch_out_dir, showWarnings = FALSE, recursive = TRUE)
library_out_dir <- file.path(scratch_out_dir, library_id)
dir.create(library_out_dir, showWarnings = FALSE, recursive = TRUE)

## Input files
# load pre-processed sample objs & anchor transfer results
obj <- SeuratObject::LoadSeuratRds( file.path(path_anal,"scratch","00_preprocessing_rds",paste0(library_id,".rdsSeurat")) )
level <- "compartment"
predictions <- read.csv( file.path(path_anal, "results", "01_anchor_transfer_seurat", paste0(library_id, "_", level,".csv")) ) 
obj <- AddMetaData(object = obj, metadata = predictions)
copykat_result <- readr::read_rds( file = file.path(library_out_dir, paste0(library_id, "_copykat_resultobj.rds")) )
copykat_result_noref <- readr::read_rds( file = file.path(library_out_dir, paste0(library_id, "_noref_copykat_resultobj.rds")) )

```

```{r images}
heatmap_wref <- file.path(library_out_dir, paste0(library_id, "_copykat_heatmap.jpeg"))
heatmap_noref <- file.path(library_out_dir, paste0(library_id, "_noref_copykat_heatmap.jpeg"))
```

## Analysis content

#### Run CopyKat with normal cells

Here, "immune" cells annotated with anchor transfer were used as reference normal cells. Here is the heatmap generated by the software:

```{r, echo=FALSE, out.width = '100%'}
knitr::include_graphics(heatmap_wref)
```

Frequency of predicted `aneuploid`, `diploid`, and `not.defined` categories.

```{r}
copykat_df <- copykat_result$prediction %>%
  select(copykat.pred) %>% 
  rename(c("ref.copykat" = "copykat.pred"))
table(copykat_df)
```

Plot CopyKat identifications onto UMAP.

```{r}
obj <- AddMetaData(object = obj, metadata = copykat_df)
DimPlot(obj, group.by = "ref.copykat", alpha = 0.3)
```

#### Run CopyKat without normal cells

Here, no normal cells were provided to CopyKat.

```{r, echo=FALSE, out.width = '100%'}
knitr::include_graphics(heatmap_noref)
```

Frequency of predicted `aneuploid`, `diploid`, and `not.defined` categories.

```{r}
copykat_noref_df <- copykat_result_noref$prediction %>%
  select(copykat.pred) %>% 
  rename(c("noref.copykat" = "copykat.pred"))
table(copykat_noref_df)
```

Plot CopyKat identifications onto UMAP.

```{r}
obj <- AddMetaData(object = obj, metadata = copykat_noref_df)
DimPlot(obj, group.by = "noref.copykat", alpha = 0.3)
```

In this Wilms tumor sample (SCPCL000850), it's likely that CopyKat is not working well: (i) The result with or without specifying normal cells are not consistent; (ii) From the heatmap, no much differences in CNV profile could be observed.

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-10-17 analyses/cell-type-wilms-tumor-14/exploratory_analysis/03_cnv/03_infercnv_SCPCL000850.Rmd
re arrange explore folder

---
title: "Exploratory CNV analysis using inferCNV"
author: "Jingxuan Chen"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

Here we test `inferCNV` to analyze the CNV profile for Wilms tumor sample (SCPCL000850), aiming to identify potential tumor cells from normal cells.

Before running this notebook, one should first generate `inferCNV` outputs as in `03_runInferCNV.R`.

## Setup

### Packages

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(Seurat)
  library(infercnv)
  library(ggplot2)
})
```

### Paths

#### Base directories

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
path_repo <- rprojroot::find_root(rprojroot::is_git_root)

# The path to this module
path_anal <- file.path(path_repo,"analyses","cell-type-wilms-tumor-14") 
```

#### Input and output files

Only run for one sample

```{r paths}
library_id <- "SCPCL000850"
scratch_out_dir <- file.path(path_anal, "scratch", "03_cnv","infercnv")
dir.create(scratch_out_dir, showWarnings = FALSE, recursive = TRUE)
library_out_dir <- file.path(scratch_out_dir, library_id)
dir.create(library_out_dir, showWarnings = FALSE, recursive = TRUE)

## Input files
# load pre-processed sample objs & anchor transfer results
obj <- SeuratObject::LoadSeuratRds( file.path(path_anal,"scratch","00_preprocessing_rds",paste0(library_id,".rdsSeurat")) )
level <- "compartment"
predictions <- read.csv( file.path(path_anal, "results", "01_anchor_transfer_seurat", paste0(library_id, "_", level,".csv")) ) 
obj <- AddMetaData(object = obj, metadata = predictions)
infercnv_result <- readr::read_rds( file = file.path(library_out_dir, "run.final.infercnv_obj") )
# add infercnv output to seurat object
obj_cnv <- infercnv::add_to_seurat(
  seurat_obj = obj,
  infercnv_output_path = library_out_dir
)
```

```{r images}
heatmap_wref <- file.path(library_out_dir, "infercnv.png")
normal_level <- "immune"
```

## Analysis content

Here, "immune" cells annotated with anchor transfer were used as reference normal cells. Here is the heatmap generated by the software:

```{r, echo=FALSE, out.width = '100%'}
knitr::include_graphics(heatmap_wref)
```

We could observe some different CNV profiles between "fetal_nephron" and "stroma", especially chr1 and chr11. However, it's hard to observe any different CNV profiles within the "fetal_nephron" subgroups.

`inferCNV` outputs results for each Chromosome. Thus, I applied several ways to summarize the overall CNV profile.

#### CNV score - proportion of genes with CNV

Based on [this](https://www.biostars.org/p/9573777/) Biostars discussion, I calculated a `cnv_score`, which basically indicates the proportion of CNVs across all genes.

```{r}
## summary strategy 1 https://www.biostars.org/p/9573777/
scores <- apply(infercnv_result@expr.data, 2 ,function(x){ sum(x < 0.95 | x > 1.05)/length(x) }) %>% 
  as.data.frame() %>%
  mutate(pred = obj$predicted.id) %>%
  mutate(cnv_score = ifelse(
    pred == normal_level,
    NA,
    .
  ))
```

Ideally, we should observe bimodal distribution for this score, indicating the CNV difference between normal and tumor cells. However, here we could get a distribution close to normal.
```{r}
hist(scores$cnv_score, breaks = 100)
```

By plotting the CNV score onto UMAP, no clear pattern is shown.
```{r}
obj <- AddMetaData(object = obj, metadata = scores)
FeaturePlot(obj, features = "cnv_score", alpha = 0.3) +
  scale_color_viridis_c()
```

#### Summary based upon number of chromosomes that have CNV
This strategy to summary CNV results is based on [`cell-type-ewings`](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-ewings/template_notebooks/cnv-workflow) module.

```{r}
## summary strategy 2 ewings analysis, based on number of chr that have cnv
cnv_df <- obj_cnv@meta.data %>%
  select(matches("predicted.id") | starts_with("has_cnv_chr") & !matches("has_cnv_chrMT")) %>%
  mutate(count_cnv_chr = ifelse(predicted.id == normal_level,
                            NA,
                            rowSums(across(starts_with("has_cnv_chr")))
                            ) )
```

```{r}
hist(cnv_df$count_cnv_chr)
```

```{r}
obj <- AddMetaData(object = obj, metadata = cnv_df)
FeaturePlot(obj, features = "count_cnv_chr", alpha = 0.3) +
  scale_color_viridis_c()
```

#### Summary based upon weighted means of CNV proportion
This strategy to summary CNV results is based on [`cell-type-ewings`](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-ewings/template_notebooks/cnv-workflow) module.

```{r}
## summary strategy 3 ewings analysis, based on number of chr that have cn
chr_weights <- infercnv_result@gene_order |> 
  as.data.frame() |> 
  dplyr::count(chr) |> 
  # only keep chr 1-22, drops MT and GL chr
  dplyr::filter(chr %in% glue::glue("chr{seq(1,22)}")) |> 
  dplyr::pull(n)

cnv_df <- obj_cnv@meta.data %>%
  select(matches("predicted.id") | starts_with("proportion_scaled_cnv_chr") & !ends_with("chrMT")) %>%
  rowwise() %>%
  mutate(weight_mean = ifelse(
    predicted.id == normal_level,
    NA,
    weighted.mean(across(starts_with("proportion_scaled_cnv_chr")), chr_weights/sum(chr_weights))
  ) )
```

```{r}
hist(cnv_df$weight_mean)
```

```{r}
obj <- AddMetaData(object = obj, metadata = cnv_df)
FeaturePlot(obj, features = "weight_mean", alpha = 0.3) +
  scale_color_viridis_c()
```

In conclusion, the heatmap generated by `inferCNV` indicates different CNV profiles between `fetal_nephron` and `stroma`, especially in Chr1. However, none of the three summary strategies seem working.

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-10-17 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
Update 02a notebook to use new functions, and style

@@
   seed: 12345
+  homologs_file: "../scratch/homologs.rds"
+  testing: FALSE
 output: 
@@
 ```{r packages, message=FALSE, warning=FALSE}
-library("Seurat")
+library(Seurat)
 library(SeuratData)
 library(sctransform)
-library(Azimuth)
 library(SCpubr)
@@
 
+```{r}
+# load functions for label transfer
+source(
+  file.path(
+    module_base,
+    "notebook_template",
+    "utils",
+    "label-transfer-functions.R"
+  )
+)
+```
+
 ### Input files
@@
 
-We install and load the reference using `Azimuth`.
+Load the Azimuth reference which has been prepared for label transfer.
+
 ```{r path_to_reference}
-#Check the names of the Azimuth available data and reference
-AvailableData()
+path_to_ref <- file.path(
+  module_base,
+  "results",
+  "references",
+  "cao_formatted_ref.rds"
+)
+if (!file.exists(path_to_ref)) {
+  stop("Reference file could not be found. Make sure `scripts/prepare-fetal-references.R` has been run first.")
+}
+ref <- readRDS(path_to_ref)
 
-# Install the fetal reference 
-InstallData("fetusref")
-
-ref <- SeuratData::LoadData("fetusref", type = "azimuth")
+# Pull out information from the reference object we need for label transfer
+reference <- ref$reference
+refdata <- ref$refdata
+dims <- ref$dims
+annotation_levels <- ref$annotation_levels
 ```
@@
 # open the processed rds object
-srat <- readRDS(file.path(data_dir, paste0("01-Seurat_", params$sample_id,".Rds")))
+srat <- readRDS(file.path(data_dir, paste0("01-Seurat_", params$sample_id, ".Rds")))
+
+# prepare the query for label transfer
+DefaultAssay(srat) <- "RNA"
+srat <- prepare_query(srat, rownames(reference), params$homologs_file)
 ```
@@
 ```{r run_azimuth, message=FALSE, warnings=FALSE}
-DefaultAssay(srat) <- "RNA"
-options(future.globals.maxSize= 891289600000000)
-s <- Azimuth::RunAzimuth(srat, reference ="fetusref")
+options(future.globals.maxSize = 891289600000000)
 
-# We transfer the annotation to the pre-processed `Seurat` object as we don't want to keep changes done on the query by `RunAzimuth`
-metadata_vec <- c("predicted.annotation.l1.score", "predicted.annotation.l1", "predicted.annotation.l2.score", "predicted.annotation.l2", "predicted.organ.score", "predicted.organ")
+# determine k.weight based CI
+if (params$testing) {
+  k.weight <- 10 # only for test datasets
+} else {
+  k.weight <- 50 # Azimuth default
+}
+s <- transfer_labels(
+  srat,
+  reference,
+  dims,
+  refdata,
+  k.weight = k.weight
+)
 
-metadata_to_trasfer <- s@meta.data[, metadata_vec]
+# We transfer the annotation to the pre-processed `Seurat` object as we don't want to keep changes done on the query by Azimuth
+annotation_columns <- c(
+  glue::glue("predicted.{annotation_levels}"),
+  glue::glue("predicted.{annotation_levels}.score")
+)
+metadata_to_trasfer <- s@meta.data[, annotation_columns]
 
-srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_full_", metadata_vec))
+srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_full_", annotation_columns))
 ```
@@
 ```{r plot_azimuth, fig.height=15, fig.width=8, warnings=FALSE}
+d1 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_full_predicted.organ", label = TRUE, repel = TRUE) + NoLegend()
+d2 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_full_predicted.annotation.l1", label = TRUE, repel = TRUE) + NoLegend()
+d3 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_full_predicted.annotation.l2", label = TRUE, repel = TRUE) + NoLegend()
 
-d1 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_full_predicted.organ", label = TRUE, repel = TRUE) + NoLegend()
-d2 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_full_predicted.annotation.l1", label = TRUE, repel = TRUE) + NoLegend()
-d3 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_full_predicted.annotation.l2", label = TRUE, repel = TRUE) + NoLegend()
-
-f1 <- SCpubr::do_BarPlot(sample = srat,
+f1 <- SCpubr::do_BarPlot(
+  sample = srat,
   group.by = "fetal_full_predicted.organ",
@@
   font.size = 10,
-                         legend.ncol = 4) +
-                         ggtitle("% cells")+
+  legend.ncol = 4
+) +
+  ggtitle("% cells") +
   xlab(params$sample_id)
 
-f2 <- SCpubr::do_BarPlot(sample = srat,
+f2 <- SCpubr::do_BarPlot(
+  sample = srat,
   group.by = "fetal_full_predicted.annotation.l1",
@@
   font.size = 10,
-                         legend.ncol = 2) +
-                         ggtitle("% cells")+
+  legend.ncol = 2
+) +
+  ggtitle("% cells") +
   xlab(params$sample_id)
 
-f3 <- SCpubr::do_BarPlot(sample = srat,
+f3 <- SCpubr::do_BarPlot(
+  sample = srat,
   group.by = "fetal_full_predicted.annotation.l2",
@@
   font.size = 10,
-                         legend.ncol = 2) +
-                         ggtitle("% cells")+
+  legend.ncol = 2
+) +
+  ggtitle("% cells") +
   xlab(params$sample_id)
 
-((d1/f1) | (d2/f2) ) 
+((d1 / f1) | (d2 / f2))
 ```
@@
 ```{r save}
-saveRDS(object = srat, file = file.path(output_dir, paste0("02a-fetal_full_label-transfer_",params$sample_id,".Rds")))
+saveRDS(object = srat, file = file.path(output_dir, paste0("02a-fetal_full_label-transfer_", params$sample_id, ".Rds")))
 ```

### 2024-10-17 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
use 0/1 for testing variable

@@
   homologs_file: "../scratch/homologs.rds"
-  testing: FALSE
+  testing: 0
 output: 

### 2024-10-17 analyses/cell-type-wilms-tumor-06/00_run_workflow.sh

@@ -74,7 +74,7 @@ for sample_dir in ${data_dir}/${project_id}/SCPCS*; do
 
         # Label transfer from the Cao reference using Azimuth
         Rscript -e "rmarkdown::render('${notebook_template_dir}/02a_label-transfer_fetal_full_reference_Cao.Rmd',
-                        params = list(scpca_project_id = '${project_id}', sample_id = '${sample_id}'),
+                        params = list(scpca_project_id = '${project_id}', sample_id = '${sample_id}', testing = 1),
                         output_format = 'html_document',
                         output_file = '02a_fetal_all_reference_Cao_${sample_id}.html',
                         output_dir = '${sample_notebook_dir}')"

### 2024-10-17 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
Update 02b notebook to use new functions, and style

@@
   seed: 12345
+  homologs_file: "../scratch/homologs.rds"
+  testing: 0
 output: 
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE,
-                      message=FALSE,
-                      warnings=FALSE
+knitr::opts_chunk$set(
+  echo = TRUE,
+  message = FALSE,
+  warnings = FALSE
 )
@@
 ```{r packages, message=FALSE, warning=FALSE}
-library("Seurat")
+library(Seurat)
 library(sctransform)
-library(Azimuth)
 library(SCpubr)
@@
 ```{r path_to_reference}
-path_to_reference <- file.path(module_base, "results", "references")
+path_to_ref <- file.path(
+  module_base,
+  "results",
+  "references",
+  "stewart_formatted_ref.rds"
+)
+if (!file.exists(path_to_ref)) {
+  stop("Reference file could not be found. Make sure `scripts/prepare-fetal-references.R` has been run first.")
+}
+ref <- readRDS(path_to_ref)
+
+# Pull out information from the reference object we need for label transfer
+reference <- ref$reference
+refdata <- ref$refdata
+dims <- ref$dims
+annotation_levels <- ref$annotation_levels
 ```
@@
 # open the processed rds object
-srat <- readRDS(file.path(data_dir, paste0("02a-fetal_full_label-transfer_",params$sample_id,".Rds")))
+srat <- readRDS(file.path(data_dir, paste0("02a-fetal_full_label-transfer_", params$sample_id, ".Rds")))
+
+# prepare the query for label transfer
+DefaultAssay(srat) <- "RNA"
+srat <- prepare_query(srat, rownames(reference), params$homologs_file)
 ```
@@
 ```{r run_azimuth, message=FALSE, warnings=FALSE}
-DefaultAssay(srat) <- "RNA"
-options(future.globals.maxSize= 8912896000000)
+options(future.globals.maxSize = 8912896000000)
 
-data <-  RunAzimuth(srat, path_to_reference, assay = 'RNA')
-# We transfer the annotation to the pre-processed `Seurat`object as we don't want to keep changes done on the query by `RunAzimuth`
-metadata_vec <- c("predicted.compartment.score", "predicted.compartment", "predicted.cell_type.score", "predicted.cell_type")
-metadata_to_trasfer <- data@meta.data[, metadata_vec]
-srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_kidney_", metadata_vec))
+# determine k.weight based CI
+if (params$testing) {
+  k.weight <- 10 # only for test datasets
+} else {
+  k.weight <- 50 # Azimuth default
+}
+s <- transfer_labels(
+  srat,
+  reference,
+  dims,
+  refdata,
+  k.weight = k.weight
+)
+
+# We transfer the annotation to the pre-processed `Seurat` object as we don't want to keep changes done on the query by Azimuth
+annotation_columns <- c(
+  glue::glue("predicted.{annotation_levels}"),
+  glue::glue("predicted.{annotation_levels}.score")
+)
+metadata_to_trasfer <- s@meta.data[, annotation_columns]
+
+srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_kidney_", annotation_columns))
 ```
@@
 ```{r plot_azimuth, fig.height=8, fig.width=8, warnings=FALSE}
+d1 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_kidney_predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
+d2 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_kidney_predicted.cell_type", label = TRUE, repel = TRUE) + NoLegend()
 
-d1 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_kidney_predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
-d2 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_kidney_predicted.cell_type", label = TRUE, repel = TRUE) + NoLegend()
-
-f1 <- SCpubr::do_BarPlot(sample = srat,
+f1 <- SCpubr::do_BarPlot(
+  sample = srat,
   group.by = "fetal_kidney_predicted.compartment",
@@
   font.size = 10,
-                         legend.ncol = 3) +
-                         ggtitle("% cells")+
+  legend.ncol = 3
+) +
+  ggtitle("% cells") +
   xlab(params$sample_id)
 
-f2 <- SCpubr::do_BarPlot(sample = srat,
+f2 <- SCpubr::do_BarPlot(
+  sample = srat,
   group.by = "fetal_kidney_predicted.cell_type",
@@
   font.size = 10,
-                         legend.ncol = 3) +
-                         ggtitle("% cells")+
+  legend.ncol = 3
+) +
+  ggtitle("% cells") +
   xlab(params$sample_id)
 
-(d1/f1) | (d2/f2) 
+(d1 / f1) | (d2 / f2)
 ```
@@
 ```{r save}
-saveRDS(object = srat, file = file.path(output_dir, paste0("02b-fetal_kidney_label-transfer_",params$sample_id,".Rds")))
+saveRDS(object = srat, file = file.path(output_dir, paste0("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")))
 ```

### 2024-10-18 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd
changes to PR#828

@@
   scpca_project_id: "SCPCP000006"
-  sample_id: "SCPCS000194"
+  sample_id: "SCPCS000179"
   seed: 12345
@@
 library(patchwork)
+library(Seurat)
 
@@
 
-- `reference_value` is the selection of normal cells used for `infercnv`
+- `infercnv_obj` is the `infercnv` object
 
@@
 
-Do_CNV_heatmap <- function(object, reference_value, group.by){
-
-  infercnv_obj <- readRDS(file.path(result_dir, reference_value , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, ".rds")))
+Do_CNV_heatmap <- function(object, infercnv_obj, group.by){
 
@@
 
-  out <- SCpubr::do_CopyNumberVariantPlot(sample = srat,
+  out <- SCpubr::do_CopyNumberVariantPlot(sample = object,
                                         infercnv_object = infercnv_obj,
@@
                                         )
+  out <- out + 
+    ggtitle(glue::glue("Copy Number Variant Plot, ", reference_value)) +
+    ylab(label = "")
+
   return(out)
@@
 
+```
+
+#### Calculate a global CNV score per cell to check the general distribution
+
+For a `Seurat` object an `infercnv` object created with the script `06_infercnv.R` using `reference_value` as a reference, the function `Do_CNV_score` load the `infercnv` object and calculate a CNV score per cell.
+The score is calculated based on the [biostar discussion](https://www.biostars.org/p/9573777/).
+It
+
+- `reference_value` is the selection of normal cells used for `infercnv`
+
+```{r}
+
+Do_CNV_score <- function(seurat_oject,infercnv_obj,reference_value){
+
+  score <- apply(infercnv_obj@expr.data,2,function(x){ sum(x < 0.90 | x > 1.10)/length(x) })
+
+  seurat_obj <- AddMetaData(seurat_oject, score, col.name = glue::glue("CNV-score_", reference_value))
+  return(seurat_obj)
+}
 ```
@@
 
-- `metadata` the gene or quantitative value to be plotted
+- `meta` the gene or quantitative value to be plotted
 
@@
                                               legend.position = "right") + ggtitle(meta)
-    b <- SCpubr::do_ViolinPlot(srat, 
+    b <- SCpubr::do_ViolinPlot(object, 
                                              features = meta, 
@@
                                               legend.position = "right") + ggtitle(features)
-                  b <- SCpubr::do_ViolinPlot(srat, 
+                  b <- SCpubr::do_ViolinPlot(object, 
                                              features = features, 
@@
 
-```{r fig.width=20, fig.height=8, out.width='100%'}
+![](`r file.path(result_dir, "reference-none_HMM-no" , glue::glue("06_infercnv_", params$sample_id, "_", "reference-none_HMM-no", "_heatmap.png"))`)
 
-reference_value = "reference-none"
+#### With immune cells as reference
+
+![](`r file.path(result_dir, "reference-immune_HMM-no" , glue::glue("06_infercnv_", params$sample_id, "_", "reference-immune_HMM-no", "_heatmap.png"))`)
+
+#### With endothelium cells as reference
+
+![](`r file.path(result_dir, "reference-endothelium_HMM-no" , glue::glue("06_infercnv_", params$sample_id, "_", "reference-endothelium_HMM-no", "_heatmap.png"))`)
+
+#### With immune and endothelium cells as reference
+
+![](`r file.path(result_dir, "reference-both_HMM-no" , glue::glue("06_infercnv_", params$sample_id, "_", "reference-both_HMM-no", "_heatmap.png"))`)
+
+### Summarize CNV per chromosome and compartment
+
+```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%'}
+infercnv_obj <- list()
+for(reference_value in c("reference-none", "reference-immune", "reference-endothelium", "reference-both")){
+  infercnv_obj[[reference_value]] <- readRDS(file.path(result_dir, glue::glue(reference_value, "_HMM-no") , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, "_HMM-no.rds")))
+  print(Do_CNV_heatmap(object = srat, infercnv_obj = infercnv_obj[[reference_value]], group.by = "fetal_kidney_predicted.compartment"))
+}
 
@@
 
-![](`r file.path(result_dir, reference_value , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, "_heatmap.png"))`)
+### Try to calculate  single CNV score and see if it can be use to define cells with CNV
 
-```{r fig.width=20, fig.height=8, out.width='100%'}
+```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
 
-Do_CNV_heatmap(object = srat, reference_value, group.by = "fetal_kidney_predicted.compartment")
+for(reference_value in c("reference-none", "reference-immune", "reference-endothelium", "reference-both")){
+  srat <- Do_CNV_score(srat,infercnv_obj = infercnv_obj[[reference_value]], reference_value)
+  p1 <- visualize_feature(srat, glue::glue("CNV-score_", reference_value) , group.by = "fetal_kidney_predicted.compartment" )
+  p2 <- visualize_density(srat, features=glue::glue("CNV-score_", reference_value), group.by = "fetal_kidney_predicted.compartment")
+  print(p1 + p2 + plot_layout(ncol = 3, widths = c(1,2,2)))
+}
 
@@
 
-#### With immune cells as reference
+This unique CNV score do not look so promising. 
+We might have to select chromosomes we would like to look at, i.e. the one relevant for Wilms tumor.
 
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-reference_value = "reference-immune"
-
-```
-
-![](`r file.path(result_dir, reference_value , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, "_heatmap.png"))`)
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-Do_CNV_heatmap(object = srat, reference_value, group.by = "fetal_kidney_predicted.compartment")
-
-```
-
-#### With endothelium cells as reference
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-reference_value = "reference-endothelium"
-
-```
-
-![](`r file.path(result_dir, reference_value , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, "_heatmap.png"))`)
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-Do_CNV_heatmap(object = srat, reference_value, group.by = "fetal_kidney_predicted.compartment")
-
-```
-
-#### With immune and endothelium cells as reference
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-reference_value = "reference-both"
-
-```
-
-![](`r file.path(result_dir, reference_value , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, "_heatmap.png"))`)
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-Do_CNV_heatmap(object = srat, reference_value, group.by = "fetal_kidney_predicted.compartment")
-
-```
-
-### Explore infercnv results generated with immune and endothelial cells as reference, using a HMM i6 model
+### Explore infercnv results generated with immune and endothelial cells as reference, using a HMM i3 model
 
@@
 
-srat <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_", params$sample_id, "_reference-both.rds")))
+srat <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-both.rds")))
 
@@
 
-###  Distribution of CNV estimation in the Wilms tumor copartments
+####  Distribution of CNV estimation in the Wilms tumor copartments
 
@@
 
-```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%'}
+```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
+for(i in 1:22){
+  print(visualize_density(srat, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
+}
+```
+
+### Explore infercnv results generated with immune and endothelial cells as reference, using a HMM i6 model
+
+#### Load the `Seurat` object
+
+We load the `Seurat` object generated in `06_infercnv.R`
+
+```{r}
+
+srat <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i6_", params$sample_id, "_reference-both.rds")))
+
+```
+
+#### Feature plot and repartition of the CNV per chromosome
+
+For each chromosome, we look at the repartition of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
+`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+
+This should allow a quick visual check that immune and endothelial cells do not expressed any CNV. If so, we should check if genes driving the CNV score are not key immune or endothelium genes.
+
+Also, we should visualize if one area of the umap do not have CNV and could be identify as additional normal cell subset.
+
+```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
+for(i in 1:22){
+  print(visualize_feature(srat, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
+}
+```
+
+####  Distribution of CNV estimation in the Wilms tumor copartments
+
+For each chromosome, we look at the distribution of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
+`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+
+We are quite confident that immune and endothelial cells are well identified by label transfer done in `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd`. 
+The distribution of CNV for endothelial and immune cells should thus be a single peak center on 0.
+
+We do not know if fetal nephron and stroma cells are a mix of normal and cancer cells. 
+Would they be a group of normal cells, we should expect a single peak center on 0 for every chromosome. 
+As we expect to have a large number of cancer with heterogeneous CNV, we should see multiple peaks.
+
+```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
 for(i in 1:22){
@@
 
-We should maybe try to run `infercnv` on one sample, gradualy downsampling the reference to 50, 10, 5, 1 cell to see how it impact the results. 
-For this try, I would suggest running with the HMM i3 model, to have an additional idea of the feasibility to extract CNV features from `infercnv`.
+I am not sure if some endothelial cells might be missclassified? 
+To be careful when we look at % of alterations within a compartment, as there are sometimes very few immune and/or endothelial cells, some % can be easily high without being meaningful.
+
+HMM-i3 model looks "cleaner" than the HMM-i6 model, with less CNV found in immune and endothelial cells.
 

### 2024-10-18 analyses/cell-type-wilms-tumor-06/scripts/explore-cnv-methods.R

@@ -16,6 +16,10 @@ for (sample_id in c("SCPCS000179",
                     "SCPCS000205",
                     "SCPCS000208")){
   
+  ##############################################################################
+  #################################`copykat`  ##################################
+  ##############################################################################
+  
   # We run and explore copykat using euclidian distance parameter and normal cell as reference
   system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "05_copyKAT.R"), " --sample_id ", sample_id, " --n_core ", 32, " --distance ", "euclidean", " --use_reference ", "ref"))
   
@@ -23,7 +27,6 @@ for (sample_id in c("SCPCS000179",
   
   system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "05_copyKAT.R"), " --sample_id ", sample_id, " --n_core ", 32, " --distance ", "spearman", " --use_reference ", "ref"))
 
-  
   # We run and explore copykat using euclidian distance parameter and without normal cell as reference
   system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "05_copyKAT.R"), " --sample_id ", sample_id, " --n_core ", 32, " --distance ", "euclidean", " --use_reference ", "noref"))
   
@@ -31,27 +34,39 @@ for (sample_id in c("SCPCS000179",
   
   system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "05_copyKAT.R"), " --sample_id ", sample_id, " --n_core ", 32, " --distance ", "spearman", " --use_reference ", "noref"))
   
+  # We explore `copykat` result rendering one notebook per sample tested
   
-  # We run and explore infercnv using immune cells as reference
-  system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "06_infercnv.R"), " --sample_id ", sample_id, " --reference ", "immune"))
-
-  # We run and explore infercnv using endothelial cells as reference
-  system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "06_infercnv.R"), " --sample_id ", sample_id, " --reference ", "endothelium"))
-
-  # We run and explore infercnv using no normal reference
-  system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "06_infercnv.R"), " --sample_id ", sample_id, " --reference ", "none"))
-  
-# We explore `copykat` result rendering one notebook per sample tested
-
-
   rmarkdown::render(input = file.path(notebook_template_dir, "05_copykat_exploration.Rmd"),
[…]

### 2024-10-18 analyses/cell-type-wilms-tumor-06/notebook_template/05_copykat_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/05_copykat_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
   - Evaluate common copy number gains and losses in Wilms tumor. 
-  - Calculate the confusion matrix comparing manual annotations of tumor cells to using `CopyKAT` to annotate tumor cells. 
   - Compare the annotations from `CopyKAT` to cell type annotations using label transfer and the fetal (kidney) references.
@@
 
-From the heatmap of CNV and the mean CNV detection plots, it is difficult to see any CNV pattern that drive the identification of aneuploid cells. 
+From the heatmap of CNV and the mean CNV detection plots, there does not appear to be any pattern that drives the identification of aneuploid cells. 
 The assignment of the aneuploidy/diploidy value might relies on very few CNV and/or an arbitrary threshold. 

### 2024-10-19 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd
corrections PR#828

@@
 `infercnv` was run using the `06_inferCNV.R` script with and without a normal reference.
-We also tested the impact of the subselection of normal cells using either immune, and/or endothelial cells as healthy reference.
+We tested the impact of the sub-selection of normal cells using either immune, and/or endothelial cells as healthy reference.
+In addition, we are exploring the use of the [HMM based CNV Prediction Methods](https://github.com/broadinstitute/infercnv/wiki/inferCNV-HMM-based-CNV-Prediction-Methods).
 
-In this notebook, we just want to compare the heatmaps of CNV profiles, and evaluate how comparable are the methods and how sensible they are to key parameters such as selection of healthy reference.
+`infercnv` currently support two models for HMM-based CNV prediction, what we refer to as the i3 and i6 models.
+These are set in the 'infercnv::run()' as HMM_type='i3' or HMM_type='i6' (i6 is default).
+Each method operates on the 'preliminary infercnv object' which has been processed through the standard inferCNV processing routines, involving subtraction of signal corresponding to 'normal (reference)' cells and smoothing operations.
+
+- i3 HMM is a three-state CNV model representing deletion, neutral, and amplification states. 
+- i6 HMM: a six-state CNV model that predicts the following CNV levels:
+
+   - state 1 : 0x = complete loss
+   - state 2 : 0.5x = loss of one copy
+   - state 3 : 1x = neutral
+   - state 4 : 1.5x = addition of one copy
+   - state 5 : 2x = addition of two copies
+   - state 6 : 3x = essentially a placeholder for >2x copies but modeled as 3x.
+
+Of note, running `infercnv` with a i6 HMM model is very long, ~2hours of running time per sample, while i3 HMM model only requires ~10min per sample.
+
+In this notebook, we want to compare the heatmaps of CNV profiles, and evaluate how comparable are the methods and how sensible they are to key parameters such as selection of healthy reference.
 
@@
 
-  chromosome_locations = SCpubr::human_chr_locations
-
   out <- SCpubr::do_CopyNumberVariantPlot(sample = object,
@@
                                         using_metacells = FALSE,
-                                        chromosome_locations = chromosome_locations,
+                                        chromosome_locations = SCpubr::human_chr_locations,
                                         return_object = FALSE,
@@
 
-For a `Seurat` object an `infercnv` object created with the script `06_infercnv.R` using `reference_value` as a reference, the function `Do_CNV_score` load the `infercnv` object and calculate a CNV score per cell.
+For a `Seurat` object an `infercnv` object created with the script `06_infercnv.R` using `reference_value` as a reference, the function `Do_CNV_score` calculate a CNV score per cell.
 The score is calculated based on the [biostar discussion](https://www.biostars.org/p/9573777/).
-It
+The function `Do_CNV_score` returns the `Seurat` object with an additional metadata named `CNV-score_{reference_value}.`
 
 - `reference_value` is the selection of normal cells used for `infercnv`
+- `seurat_object` and `infercnv_obj` are the `Seurat`and `infercnv` object created with the script `06_infercnv.R`
 
@@
 
-  score <- apply(infercnv_obj@expr.data,2,function(x){ sum(x < 0.90 | x > 1.10)/length(x) })
+  score <- apply(infercnv_obj@expr.data,2,function(x){ sum(x < 0.95 | x > 1.05)/length(x) })
 
@@
 
-visualize_metadata(srat, "seurat_clusters", "fetal_kidney_predicted.compartment")
+visualize_metadata(srat, "fetal_kidney_predicted.compartment", "seurat_clusters")
 

### 2024-10-19 analyses/cell-type-wilms-tumor-06/scripts/06_infercnv.R

@@ -96,12 +96,6 @@ if(opts$reference == "both"){
   normal_cells <- opts$reference
 }
 
-# create output directory if it does not exist
-dir.create(output_dir, recursive = TRUE)
-
-# retrieve the gene order file created in `06a_build-geneposition.R`
-gene_order_file <- file.path(module_base, "results", "references", "gencode_v19_gene_pos.txt")
-
 # We only run the CNV HMM prediction model if the reference is "both"
 HMM_logical = TRUE
 HMM_type <- opts$HMM
@@ -110,6 +104,12 @@ if(opts$HMM == "no"){
   HMM_logical <- FALSE
   HMM_type <- NULL
 }
+
+# create output directory if it does not exist
+dir.create(output_dir, recursive = TRUE)
+
+# retrieve the gene order file created in `06a_build-geneposition.R`
+gene_order_file <- file.path(module_base, "results", "references", "gencode_v19_gene_pos.txt")
 
 # Run infercnv ------------------------------------------------------------------
 # create inferCNV object and run method
@@ -134,7 +134,7 @@ infercnv_obj <- infercnv::run(
   save_final_rds = TRUE
 )
 
-if(HMM_logical == TRUE){
+if(HMM_logical){
 # Add `infercnv` data to the `Seurat` object  
 srat = infercnv::add_to_seurat(infercnv_output_path=output_dir,
                                      seurat_obj=srat,

### 2024-10-19 analyses/cell-type-wilms-tumor-06/scripts/explore-cnv-methods.R

@@ -65,7 +65,7 @@ for (sample_id in c("SCPCS000179",
 
   # We run and explore infercnv using both endothelial and immune cells as reference and i6 HMM model
   
- #  system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "06_infercnv.R"), " --sample_id ", sample_id, " --reference ", "both", " --HMM ", "i6")) 
+   system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "06_infercnv.R"), " --sample_id ", sample_id, " --reference ", "both", " --HMM ", "i6")) 
   
   rmarkdown::render(input = file.path(notebook_template_dir, "06_cnv_infercnv_exploration.Rmd"),
                   params = list(sample_id = sample_id, seed = 12345),

### 2024-10-21 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Add files via upload

@@
 date: "`r Sys.Date()`"
+params: 
+  mapping_score_thr: 0.85
 output: 
@@
         compartment = srat$fetal_kidney_predicted.compartment,
+        compartment_score = srat$fetal_kidney_predicted.compartment.score,
         organ = srat$fetal_full_predicted.organ, 
@@
 
+cell_type_df <- merge(cell_type_df, metadata, by.x = "sample_id", by.y = "scpca_sample_id")
+
+cell_type_df$pass_mapping_QC <- FALSE 
+cell_type_df$pass_mapping_QC[cell_type_df$compartment_score > params$mapping_score_thr] <- TRUE
 ```
@@
 - `df` is the name of the table containing metadata and feature expression (counts) per cells
+- `feature` is the name of the gene to average the expression
 - `group.by` is the name of the metadata to group the cells
-- `feature` is the name of the gene to average the expression
+- `split.by` is the name of the metadata to split the plots
 
@@
   )
+
+return(p)
+}
+```
+
+#### do_Feature_densityplot
+
+`do_Feature_densityplot` shows boxplot of expression of a feature grouped by a metadata.
+
+- `df` is the name of the table containing metadata and feature expression (counts) per cells
+- `group.by` is the name of the metadata to group the cells
+- `feature` is the name of the gene to average the expression
+
+```{r fig.width=10, fig.height=4, out.width='100%'}
+do_Feature_densityplot <- function(df, group.by, feature, split.by = NULL  ){
+
+  df <- df %>%
+    mutate(!!sym(group.by), group  = factor(!!sym(group.by), levels = c("fetal_nephron", "stroma", "immune", "endothelium")))
+
+  p <- ggplot(df, 
+    aes(x = !!sym(feature),  fill = group)
+  ) + 
+  geom_density() +
+  facet_wrap(vars(!!sym(group.by)), scale = "free_y", ncol = 4) + 
+  theme_bw()
 
@@
 ```
+### Label transfer (mapping) scores for the four compartments
+
+```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
+p <- do_Feature_densityplot( df = cell_type_df,
+                    feature = "compartment_score",
+                    group.by = "compartment"
+                    )
+p +  geom_vline(xintercept=params$mapping_score_thr)
+```
+
+Some cells have an excellent mapping score (>0,85) while other performed worth. Let's have a look at the repartiction per patient.
+
+```{r fig.height=15, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
+p <- do_Feature_boxplot( df = cell_type_df,
+                    feature = "compartment_score",
+                  group.by = "compartment",
+                    split.by = "sample_id")
+
+p + geom_hline(yintercept = params$mapping_score_thr)
+
+```
+
+Regarding the endothelial cells, some patient have very high mapping scores while others perform poorly. 
+Let's have a look if this can be linked to some metadata/clinical data.
+
+##### Treatment
+
+```{r fig.height=3, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
+p <- do_Feature_boxplot( df = cell_type_df,
+                    feature = "compartment_score",
+                  group.by = "compartment",
+                    split.by = "treatment")
+
+p + geom_hline(yintercept = params$mapping_score_thr)
+
+```
+
+##### Subdiagnosis
+
+```{r fig.height=3, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
+p <- do_Feature_boxplot( df = cell_type_df,
+                    feature = "compartment_score",
+                  group.by = "compartment",
+                    split.by = "subdiagnosis")
+
+p + geom_hline(yintercept = params$mapping_score_thr)
+
+```
+
+##### Disease timing
+
+```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
+p <- do_Feature_boxplot( df = cell_type_df,
+                    feature = "compartment_score",
+                  group.by = "compartment",
+                    split.by = "disease_timing")
+
+p + geom_hline(yintercept = params$mapping_score_thr)
+
+```
+
+We do not really see a pattern linking the label transfer score and some meta/clinical data. 
+For the later CNV analysis, we can apply a threshold of `r params$mapping_score_thr` for the `fetal_kidney_predicted.compartment.score` while selecting for normal cells, i.e.
+`immune` and/or `endothelial` cells.
+
+We finally look for each patient and compartment, the number of cells that have a mapping score > `r params$mapping_score_thr` (TRUE).
+
+```{r fig.width=25, fig.height=3, results='asis'}
+scores_df <- cell_type_df |>
+  # count how many per compartment
+  dplyr::count(sample_id, compartment, pass_mapping_QC) |>
+  # make the table wide and use a value of 0 if a count is missing
+  tidyr::pivot_wider(names_from = pass_mapping_QC, values_from = n, values_fill = 0)
+
+DT::datatable(scores_df, 
+              caption = "counts of cells that have a good/poor mapping in each compartment",
+              extensions = 'Buttons', 
+              options = list(  dom = 'Bfrtip',
+                               buttons = c( 'csv', 'excel')))
+```
 

### 2024-10-21 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd
improve PR#828 and try inter-patient normal reference

@@
   scpca_project_id: "SCPCP000006"
-  sample_id: "SCPCS000179"
+  sample_id: "SCPCS000184"
   seed: 12345
@@
 
-`infercnv` was run using the `06_inferCNV.R` script with and without a normal reference.
+`infercnv` was run using the `06_inferCNV.R` script with and without a normal reference, from the same patient or from an inter-patient pull of normal cells.
 We tested the impact of the sub-selection of normal cells using either immune, and/or endothelial cells as healthy reference.
@@
 library(Seurat)
-
 ```
@@
 
+#### With immune and endothelium cells from all upfront resection samples spiked-in as reference
+
+![](`r file.path(result_dir, "reference-pull_HMM-i3" , glue::glue("06_infercnv_", params$sample_id, "_", "reference-pull_HMM-i3", "_heatmap.png"))`)
+
 ### Summarize CNV per chromosome and compartment
@@
 ```
+These heatmaps emphasize the importance of the selection of normal cells prior the inference of CNV.
+The normal reference should contain as much cell types as possible, in order to minimize false positive CNV.
+In our case, we should take immune and entodethial cells when possible.
 
-### Try to calculate  single CNV score and see if it can be use to define cells with CNV
+Of note: By default if no reference is provided, `infercnv` take the mean of expression as normal reference. 
+The risk is that the main cell population (in our case the fetal nephron compartment) might be mistaken as the normal baseline.
+
+### Summary CNV score
+
+We want to calculate a single CNV score and asess if/how it can be use to define cells with CNV versus stable/normal cells. 
+We defined the score as discribed in the [biostar discussion](https://www.biostars.org/p/9573777/).
+
+We would expect:
+
+- immune and endothelial cells to have a low CNV score
+
+- fetal nephron and stroma cells to have low to high CNV score.
+
+Ideally, the CNV score distribution should show two peaks, one for the normal cells with no CNV, one for cancer cells with more genomic rearrangements.
 
@@
 
-This unique CNV score do not look so promising. 
-We might have to select chromosomes we would like to look at, i.e. the one relevant for Wilms tumor.
+This unique CNV score does not look promising. 
+We will not use this score in future notebooks.
 
-### Explore infercnv results generated with immune and endothelial cells as reference, using a HMM i3 model
+We might have to select chromosomes we would like to look at, i.e. the one relevant for Wilms tumor (1p, 1q, chr17).
+
+### HMM-i3 inference prediction with both immune and endothelium cells as reference
+
+We then explore infercnv results generated with immune and endothelial cells as reference, using a [HMM-i3 prediction models](https://github.com/broadinstitute/infercnv/wiki/infercnv-i3-HMM-type).
 
@@
 
-srat <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-both.rds")))
+srat_i3 <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-both.rds")))
 
+```
+
+#### Plot UMAP of known Wilms tumor related alteration
+
+As a reminder, we plot the label transfer from the fetal kidney atlas.
+
+Immune and endothelial cells shouldn't show any CNV.
+
+The stroma and fetal nephron compartments may contain normal cells with no CNV or cancer cells with low to high CNV profiles.
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+visualize_metadata(srat_i3, "fetal_kidney_predicted.compartment", "seurat_clusters")
+
+```
+
+#### Feature plot and repartition of the CNV per chromosome
+
+For each chromosome, we look at the repartition of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
+`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+
+This should allow a quick visual check that immune and endothelial cells do not expressed any CNV. 
+If not, we should check if genes driving the CNV score are not key immune or endothelium genes.
+
+The `umap` reduction visualization might allow the identification of cluster(s) of normal cells with no CNV.
+
+```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
+for(i in 1:22){
+  print(visualize_feature(srat_i3, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
+}
+```
+
+####  CNV distribution in the Wilms tumor compartments
+
+For each chromosome, we look at the distribution of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
+`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+
+We are quite confident that immune and endothelial cells are well identified by label transfer done in `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd`. 
+The distribution of CNV for endothelial and immune cells should thus be a single peak center on 0.
+
+We do not know if fetal nephron and stroma cells are a mix of normal and cancer cells. 
+Would they be a group of normal cells, we should expect a single peak center on 0 for every chromosome. 
+As we expect to have a large number of cancer with heterogeneous CNV, we should see multiple peaks.
+
+```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
+for(i in 1:22){
+  print(visualize_density(srat_i3, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
+}
+```
+
+#### DotPlot
+
+The `Dotplot` representation summarizes the percentage of cells in each compartment with cnv in each of the 22 chromosomes.
+
+The HMM-i3 prediction model with intra-patient reference (composed of immune and endothelium cells) perform really well in identifying Wilms tumor relevant CNV.
+No false positive CNV are observed in the immune and endothelium compartment.
+
+```{r fig.width=15, fig.height=5, out.width='100%'}
+
+DotPlot(srat_i3, features =  colnames(srat_i3@meta.data)[grepl("has_cnv_chr", colnames(srat_i3@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+```
+
+#### CNV score
+
+Here, we try to calculate a single CNV score and assess its potential in iditenfying cells with CNV versus normal cells without CNV.
+
+We simply checked for each chromosome if the cell `has_cnv_chr`. 
+Would the cell have one or more chromosome with CNV, the global `has_cnv_score` will be TRUE.
+Else, the cell will have a `has_cnv_score` set to FALSE.
+
+```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
+meta <- srat_i3@meta.data
+
+meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
+meta$has_cnv_score[meta$has_cnv_score > 1] <- "CNV"
+
+# we know that some immune cells might have false positive CNV 
+meta$has_cnv_score[meta$has_cnv_score %in% c(0,1)] <- "no CNV"
+
+srat_i3 <- AddMetaData(srat_i3, meta$has_cnv_score, col.name =  "has_cnv_score")
+
+```
+
+We identified clusters with no/low CNV while the majority of the `Seurat` cluster present CNV on more than one chromosome.
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+visualize_metadata(srat_i3, "has_cnv_score", "seurat_clusters")
+
+```
+
+Immune and endothelial cells have a majority of cells with no/low CNV profile, which confirm their annotation as normal cells. 
+The fetal nephron and stroma copartments are mainly composed of cancer cells with more than one chromosome affected by a CNV.
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+visualize_metadata(srat_i3, "has_cnv_score", "fetal_kidney_predicted.compartment")
+
+```
+
+### HMM-i6 prediction of CNV with intra-patient normal reference (immune and endothelial)
+
+We then explore `infercnv` results generated with immune and endothelial cells as reference, using a [HMM-i6 model](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type).
+
+#### Load the `Seurat` object
+
+We load the `Seurat` object generated in `06_infercnv.R`
+
+```{r}
+
+srat_i6 <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i6_", params$sample_id, "_reference-both.rds")))
+
+```
+
+#### Feature plot and repartition of the CNV per chromosome
+
+For each chromosome, we look at the repartition of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
+`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+
+```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
+for(i in 1:22){
+  print(visualize_feature(srat_i6, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
+}
+```
+
+####  Distribution of CNV estimation in the Wilms tumor copartments
+
+For each chromosome, we look at the distribution of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
+`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+
+We are quite confident that immune and endothelial cells are well identified by label transfer done in `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd`. 
+The distribution of CNV for endothelial and immune cells should thus be a single peak center on 0.
+
+We do not know if fetal nephron and stroma cells are a mix of normal and cancer cells. 
+Would they be a group of normal cells, we should expect a single peak center on 0 for every chromosome. 
+As we expect to have a large number of cancer with heterogeneous CNV, we should see multiple peaks.
+
+```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
+for(i in 1:22){
+  print(visualize_density(srat_i6, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
+}
+```
+
+#### DotPlot
+
+The `Dotplot` representation summarizes the percentage of cells in each compartment with cnv in each of the 22 chromosomes.
+
+The HMM-i3 prediction model with intra-patient reference (composed of immune and endothelium cells) perform really well in identifying Wilms tumor relevant CNV.
+No false positive CNV are observed in the immune and endothelium compartment.
+
+```{r fig.width=15, fig.height=5, out.width='100%'}
+
+DotPlot(srat_i6, features =  colnames(srat_i6@meta.data)[grepl("has_cnv_chr", colnames(srat_i6@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+```
+
+#### CNV score
+
+Finally, we try to calculate  single CNV score and assess its potential in iditenfying cells with CNV versus normal cells without CNV.
+
+We simply checked for each chromosome if the cell `has_cnv_chr`. 
+Would the cell have one or more chromosome with CNV, the global `has_cnv_score` will be TRUE.
+Else, the cell will have a `has_cnv_score` set to FALSE.
+
+```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
+meta <- srat_i6@meta.data
+
+meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
+meta$has_cnv_score[meta$has_cnv_score > 1] <- "CNV"
+
+# we know that some immune cells might have false positive CNV 
+meta$has_cnv_score[meta$has_cnv_score %in% c(0,1)] <- "no CNV"
+
+srat_i6 <- AddMetaData(srat_i6, meta$has_cnv_score, col.name =  "has_cnv_score")
+
+```
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+visualize_metadata(srat_i6, "has_cnv_score", "seurat_clusters")
+
+```
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+visualize_metadata(srat_i6, "has_cnv_score", "fetal_kidney_predicted.compartment")
+
+```
+
+### HMM-i3 CNV prediction model with an inter-patient reference of immune cells (endothelial and immune)
+
+We lastly explore `infercnv` results generated with a pull of immune and endothelial cells from different Wilms tumor samples as reference, using a HMM-i3 prediction model.
+
+The selection of normal cells to be spiked-in as a normal reference prior running `infercnv` has been performed in `06a_build-normal-cell_reference.R.
+We selected samples that haven't been pre-treated with chemotherapies.
+Indeed, even if normal cells shouldn't be affected by chemotherapy in terms of CNV, we are not sure how chemotherapies can affect the CNV profile of normal cells.
+We thus decided to only take endothelial and immune cells from non-treated samples to build the reference of normal cells.
+
+#### Load the `Seurat` object
+
+We load the `Seurat` object generated in `06_infercnv.R` and transfer the `infercnv` metadata to the main `Seurat` object of sample `r params$sample_id` saved in `r glue::glue("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")`.
+
+Such a transfer is required as the merged `Seurat` object + spike-in cells does not contained the `umap` reduction.
+
+```{r}
+seurat_full <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-pull.rds")))
+# subset non spike-in cells
+keep_cells <- colnames(seurat_full)[!grepl("spike", colnames(seurat_full))]
+seurat <- subset(seurat_full, cells = keep_cells)
+
+meta <- seurat@meta.data
+
+# merge the new info into the srat metadata
+srat <- AddMetaData(srat, meta[,grepl("_cnv_chr", colnames(meta))], col.name = colnames(meta[grepl("_cnv_chr", colnames(meta))]))
 ```
@@
 
-### Explore infercnv results generated with immune and endothelial cells as reference, using a HMM i6 model
+#### DotPlot
 
-#### Load the `Seurat` object
+The `Dotplot` representation summarizes the percentage of cells in each compartment with cnv in each of the 22 chromosomes.
 
-We load the `Seurat` object generated in `06_infercnv.R`
+The HMM-i3 prediction model with inter-patient reference (composed of immune and endothelium cells) perform quite well in identifying Wilms tumor relevant CNV.
+However, sample variation induced false positive CNV such as CNV on chromosome 5 and 13 observed in the endothelium compartment.
 
-```{r}
+```{r fig.width=15, fig.height=5, out.width='100%'}
 
-srat <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i6_", params$sample_id, "_reference-both.rds")))
+DotPlot(srat, features =  colnames(srat@meta.data)[grepl("proportion_cnv_chr", colnames(srat@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 
@@
 
-#### Feature plot and repartition of the CNV per chromosome
+#### CNV score
 
-For each chromosome, we look at the repartition of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
-`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+Finally, we try to calculate  single CNV score and assess its potential in iditenfying cells with CNV versus normal cells without CNV.
 
-This should allow a quick visual check that immune and endothelial cells do not expressed any CNV. If so, we should check if genes driving the CNV score are not key immune or endothelium genes.
+We simply checked for each chromosome if the cell `has_cnv_chr`. 
+Would the cell have three or more chromosome with CNV, the global `has_cnv_score` will be TRUE.
+Else, the cell will have a `has_cnv_score` set to FALSE.
 
-Also, we should visualize if one area of the umap do not have CNV and could be identify as additional normal cell subset.
+Of note, as we know that the use of a ~universal inter-patient reference of normal cells can induce false positive CNV in some chromosomes, we decided to allow up to 2 CNV in normal cells.
 
-```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
-for(i in 1:22){
-  print(visualize_feature(srat, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
-}
+```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
+meta <- srat@meta.data
+meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
+meta$has_cnv_score[meta$has_cnv_score > 2] <- "CNV"
+
+# we know that some immune cells might have false positive CNV 
+meta$has_cnv_score[meta$has_cnv_score %in% c(0,1, 2)] <- "no CNV"
+
+srat <- AddMetaData(srat, meta$has_cnv_score, col.name =  "has_cnv_score")
+
 ```
 
-####  Distribution of CNV estimation in the Wilms tumor copartments
+```{r fig.width=20, fig.height=8, out.width='100%'}
 
-For each chromosome, we look at the distribution of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
-`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+visualize_metadata(srat, "has_cnv_score", "seurat_clusters")
 
-We are quite confident that immune and endothelial cells are well identified by label transfer done in `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd`. 
-The distribution of CNV for endothelial and immune cells should thus be a single peak center on 0.
+```
 
-We do not know if fetal nephron and stroma cells are a mix of normal and cancer cells. 
-Would they be a group of normal cells, we should expect a single peak center on 0 for every chromosome. 
-As we expect to have a large number of cancer with heterogeneous CNV, we should see multiple peaks.
+```{r fig.width=20, fig.height=8, out.width='100%'}
 
-```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
-for(i in 1:22){
-  print(visualize_density(srat, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
-}
+visualize_metadata(srat, "has_cnv_score", "fetal_kidney_predicted.compartment")
+
+```
+
+## Comparisons of inter- and intra-patient global cnv score with HMM prediction model
+
+We compare here the binary CNV scores calculated with the three HMM prediction models:
+
+- HMM-i3 with inter-patient endothelial and immunce cells as reference
+
+- HMM-i3 with intra-patient endothelial and immunce cells as reference
+
+- HMM-i6 with intra-patient endothelial and immunce cells as reference
+
+```{r fig.width=20, fig.height=8, out.width='100%'}
+
+p1 <- SCpubr::do_DimPlot(srat, "has_cnv_score", plot.title = "inter-patient reference, HMM-i3", reduction = "umap")
+
+p2 <- SCpubr::do_DimPlot(srat_i3, "has_cnv_score", plot.title = "intra-patient reference, HMM-i3", reduction = "umap")
+
+p3 <- SCpubr::do_DimPlot(srat_i6, "has_cnv_score", plot.title = "intra-patient reference, HMM-i6", reduction = "umap")
+
+p1 + p2 + p3 + plot_layout(ncol = 3)
+
 ```
@@
 
-- Providing no reference is not a good option, as we think that most of the cells are cancer cells with few CNV. 
-Without healthy reference, `infercnv` take the mean of expression across the cells as a reference. 
-In case of events that are well shared by _a majority_ of the cells, they will be smoothed.
+- We should run `infercnv` with the most diverse normal reference as possible, including different cell types.
+Providing no reference is not a good option, as we think that most of the cells are cancer cells with few CNV. 
+In our case, we advise taking at least immune and endothelial cells as normal reference.
 
-- For some CNV and specific chromosome, the choice a the reference do not affect the result: ch18, ch17.
-- Some CNV values differs depending on the choice of the reference, for chr5 for example.
+- The HMM prediction models help exploring the `infercnv` results. 
+In this notebook, we have compared three HMM prediction models:
 
-This is an issue if we want to have the exact CNV pattern.
+   + HMM-i3 with inter-patient endothelial and immunce cells as reference
 
-There are three samples with a very low (<5) number of immune + endothelial cells: SCPCS000197 (4 endothelial), SCPCS000190 (1 immune), SCPCS000177	(2 immune) for which it might be problematic to run `infercnv`.
+   + HMM-i3 with intra-patient endothelial and immunce cells as reference
 
-I am not sure if some endothelial cells might be missclassified? 
-To be careful when we look at % of alterations within a compartment, as there are sometimes very few immune and/or endothelial cells, some % can be easily high without being meaningful.
+   + HMM-i6 with intra-patient endothelial and immunce cells as reference
 
-HMM-i3 model looks "cleaner" than the HMM-i6 model, with less CNV found in immune and endothelial cells.
+- Globally, the three scores seems to drive similar conclusions, with the majority of fetal nephron and stroma cells being cancer cells, at least in the sample selected.
+
+   + The HMM-i3 model with inter-patient endothelial and immunce cells as reference has the advantage to be usable for all Wilms tumor samples, including the ones with a very low number of immune and/or endothelial cells.
+
+   + The HMM-i3 model with intra-patient endothelial and immunce cells as reference seems to be the cleaner, ~fast to run (10 minutes per samples) and is more precise than the HMM-i3 with the inter-patient reference.
+
+   + The HMM-i6 model with intra-patient endothelial and immunce cells as reference is very slow (~2 hours per sample) and couldn't be used for the entire cohorte.
+It is more noisy than the i3 version. However, it could have the potential to detect cancer cells with very low CNV profile.
+
+- Surprisingly, running `infercnv` with emdothelial and immune cells from (i) the same patient or (ii) a set of Wilms tumor patients do not seem to affect drastically the results. 
+Some false positive CNV might occur in every patient due to the inter-patient variability. 
+By comparing the results in conditions (i) and (ii), we should be able to understand which false positive are recurrent and do not take them into account.
 

### 2024-10-21 analyses/cell-type-wilms-tumor-06/scripts/explore-cnv-methods.R

@@ -66,7 +66,10 @@ for (sample_id in c("SCPCS000179",
   # We run and explore infercnv using both endothelial and immune cells as reference and i6 HMM model
   
    system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "06_infercnv.R"), " --sample_id ", sample_id, " --reference ", "both", " --HMM ", "i6")) 
-  
+   
+   # We run and explore infercnv using both endothelial and immune cells from all non-treated Wilms tumor patients as reference and i3 HMM model
+   system(command = glue::glue("Rscript ", file.path(module_base,"scripts", "06_infercnv.R"), " --sample_id ", sample_id, " --reference ", "pull", " --HMM ", "i3"))   
+   
   rmarkdown::render(input = file.path(notebook_template_dir, "06_cnv_infercnv_exploration.Rmd"),
                   params = list(sample_id = sample_id, seed = 12345),
                   output_format = "html_document",

### 2024-10-22 analyses/cell-type-dsrct/exploratory_analysis/01-marker-gene-tumor-classification.Rmd
updated the notebooks
updated the notebook to include plot outputs
updated the reference to include papers and internal reference
added plot outputs

@@
 module_base <- file.path(repository_base, "analyses", "cell-type-DSRCT")
-```
-
-```{r}
-# source in helper functions: make_jaccard_matrix() and jaccard() 
-jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
-source(jaccard_functions)
 ```
@@
   print(p)
+
+  #save plots
+  ggsave(
+    filename = paste0(
+      '../plots/marker_expression_',
+      names(umap_df_list)[i],
+      '.png'
+    ),
+    plot = p,
+    width = 6,
+    height = 6,
+    units = 'in',
+    dpi = 150
+  )
 }
@@
   print(p)
+
+  #save plots
+  ggsave(
+    filename = paste0(
+      '../plots/marker_distribution_',
+      names(umap_df_list)[i],
+      '.png'
+    ),
+    plot = p,
+    width = 6,
+    height = 6,
+    units = 'in',
+    dpi = 150
+  )
 }
@@
     ggtitle(names(umap_df_list_excluded_scaled)[i])
+
   print(p)
+
+  #save plots
+  ggsave(
+    filename = paste0(
+      '../plots/marker_distribution_zscaled_',
+      names(umap_df_list)[i],
+      '.png'
+    ),
+    plot = p,
+    width = 6,
+    height = 6,
+    units = 'in',
+    dpi = 150
+  )
+
 }
@@
   print(p)
+
+  #save plots
+  ggsave(
+    filename = paste0(
+      '../plots/UMAP_marker_expression_',
+      names(umap_df_list)[i],
+      '.png'
+    ),
+    plot = p,
+    width = 6,
+    height = 6,
+    units = 'in',
+    dpi = 150
+  )
 }
@@
   print(p)
+
+  #save plots
+  ggsave(
+    filename = paste0(
+      '../plots/UMAP_classification_',
+      names(umap_df_list)[i],
+      '.png'
+    ),
+    plot = p,
+    width = 6,
+    height = 6,
+    units = 'in',
+    dpi = 150
+  )
 }
@@
   print(p)
+
+  #save plots
+  ggsave(
+    filename = paste0('../plots/UMAP_marker_expression_cluster_', names(umap_df_list)[i], '.png'),
+    plot = p,
+    width = 6,
+    height = 6,
+    units = 'in',
+    dpi = 150
+  )
 }
@@
   print(p)
+
+  #save plots
+  ggsave(
+    filename = paste0(
+      '../plots/UMAP_classification_cluster_',
+      names(umap_df_list)[i],
+      '.png'
+    ),
+    plot = p,
+    width = 6,
+    height = 6,
+    units = 'in',
+    dpi = 150
+  )
+
 }
@@
 ## Save outputs 
-
 ```{r}

### 2024-10-22 analyses/cell-type-dsrct/exploratory_analysis/01-marker-gene-tumor-classification.Rmd
updated the notebooks
updated the notebook to include plot outputs
updated the reference to include papers and internal reference

@@
     ggtitle(names(umap_df_list)[i])
+
+  tiff(filename = paste0('../plots/marker_expression_', names(umap_df_list)[i], '.tiff'), width = 6, height = 6, units = 'in', res = 150)
   print(p)
-
-  #save plots
-  ggsave(
-    filename = paste0(
-      '../plots/marker_expression_',
-      names(umap_df_list)[i],
-      '.png'
-    ),
-    plot = p,
-    width = 6,
-    height = 6,
-    units = 'in',
-    dpi = 150
-  )
+  dev.off()
 }
@@
   #save plots
-  ggsave(
-    filename = paste0(
-      '../plots/marker_distribution_',
-      names(umap_df_list)[i],
-      '.png'
-    ),
-    plot = p,
-    width = 6,
-    height = 6,
-    units = 'in',
-    dpi = 150
-  )
+  tiff(filename = paste0('../plots/marker_distribution_', names(umap_df_list)[i], '.tiff'), width = 6, height = 6, units = 'in', res = 150)
+  print(p)
+  dev.off()
 }
@@
   #save plots
-  ggsave(
-    filename = paste0(
-      '../plots/marker_distribution_zscaled_',
-      names(umap_df_list)[i],
-      '.png'
-    ),
-    plot = p,
-    width = 6,
-    height = 6,
-    units = 'in',
-    dpi = 150
-  )
+  tiff(filename = paste0('../plots/marker_distribution_zscaled_', names(umap_df_list)[i], '.tiff'), width = 6, height = 6, units = 'in', res = 150)
+  print(p)
+  dev.off()
 
@@
     #save plots
-  ggsave(
-    filename = paste0(
-      '../plots/UMAP_marker_expression_',
-      names(umap_df_list)[i],
-      '.png'
-    ),
-    plot = p,
-    width = 6,
-    height = 6,
-    units = 'in',
-    dpi = 150
-  )
+  tiff(filename = paste0('../plots/UMAP_marker_expression_', names(umap_df_list)[i], '.tiff'), width = 6, height = 6, units = 'in', res = 150)
+  print(p)
+  dev.off()
 }
@@
 #save plots
-  ggsave(
-    filename = paste0(
-      '../plots/UMAP_classification_',
-      names(umap_df_list)[i],
-      '.png'
-    ),
-    plot = p,
-    width = 6,
-    height = 6,
-    units = 'in',
-    dpi = 150
-  )
+  tiff(filename = paste0('../plots/UMAP_classification_', names(umap_df_list)[i], '.tiff'), width = 6, height = 6, units = 'in', res = 150)
+  print(p)
+  dev.off()
 }
@@
   #save plots
-  ggsave(
-    filename = paste0('../plots/UMAP_marker_expression_cluster_', names(umap_df_list)[i], '.png'),
-    plot = p,
-    width = 6,
-    height = 6,
-    units = 'in',
-    dpi = 150
-  )
+  tiff(filename = paste0('../plots/UMAP_marker_expression_cluster_', names(umap_df_list)[i], '.tiff'), width = 6, height = 6, units = 'in', res = 150)
+  print(p)
+  dev.off()
 }
@@
   print(p)
-
-  #save plots
-  ggsave(
-    filename = paste0(
-      '../plots/UMAP_classification_cluster_',
-      names(umap_df_list)[i],
-      '.png'
-    ),
-    plot = p,
-    width = 6,
-    height = 6,
-    units = 'in',
-    dpi = 150
-  )
+  tiff(filename = paste0('../plots/UMAP_classification_cluster_', names(umap_df_list)[i], '.tiff'), width = 6, height = 6, units = 'in', res = 150)
+  print(p)
+  dev.off()
 

### 2024-10-22 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
   # now combine all dataframes to make one big one
-  dplyr::bind_rows()
-
-cell_type_df <- merge(cell_type_df, metadata, by.x = "sample_id", by.y = "scpca_sample_id")
-
-cell_type_df$pass_mapping_QC <- FALSE 
-cell_type_df$pass_mapping_QC[cell_type_df$compartment_score > params$mapping_score_thr] <- TRUE
+  dplyr::bind_rows() |>
+  # Combine with metadata
+  dplyr::left_join(metadata, by = c("sample_id" = "scpca_sample_id")) |>
+  # Create column for whether the compartment score passes threshold
+  dplyr::mutate(pass_mapping_QC = compartment_score > params$mapping_score_thr)
 ```

### 2024-10-22 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 ```{r fig.width=10, fig.height=4, out.width='100%'}
-do_Feature_densityplot <- function(df, group.by, feature, split.by = NULL  ){
+do_Feature_densityplot <- function(df, group.by, feature){
 
@@
 
-Some cells have an excellent mapping score (>0,85) while other performed worth. Let's have a look at the repartiction per patient.
+Some cells have an excellent mapping score (>0,85) while other performed worse. Let's have a look at the compartments per patient.
 

### 2024-10-22 analyses/cell-type-dsrct/exploratory_analysis/01-marker-gene-tumor-classification.Rmd

--- 2024-10-22 analyses/cell-type-dsrct/exploratory_analysis/01-marker-gene-tumor-classification.Rmd
@@
     ggtitle(names(umap_df_list)[i])
-  print(p)
 

### 2024-10-22 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 ---
-title: "Annotation exploration for SCPCP000006"
+title: "Annotation exploration for SCPCP000006, mapping score threshold `r params$mapping_score_thr`"
 author: "Maud PLASCHKA"
@@
 In order to explore the label transfer results, we look into some marker genes, table and percentages of cells in each annotation groups (from label transfers).
+
+We also looked into the mapping scores for each of the compartments and explored the potential effects of score thresholds.
+In this notebook, we fixed the mapping score threshold to `r params$mapping_score_thr`.
 
@@
       # Create and return a data frame from the Seurat object with relevant annotations
-      # this data frame will have four columns: barcode, sample_id, compartment, organ
+      # this data frame will have six columns: barcode, compartment, organ, compartment_score, VWF and PTPRC
       data.frame(
@@
         PTPRC = FetchData(object = srat, vars = "ENSG00000081237", layer = "counts"),
-        VWF = FetchData(object = srat, vars = "ENSG00000110799", layer = "counts")
+        VWF = FetchData(object = srat, vars = "ENSG00000110799", layer = "counts"),
+        umap = srat@reductions$umap@cell.embeddings
         ) |>
@@
   dplyr::mutate(pass_mapping_QC = compartment_score > params$mapping_score_thr)
+
+# we create a data frames of annotation of cells that pass the mapping score threshold
+cell_type_df_pass <- cell_type_df  |>
+  dplyr::filter(pass_mapping_QC)
+
+# we create a data frames of annotation of cells that don't pass the mapping score threshold
+cell_type_df_nopass <- cell_type_df  |>
+  dplyr::filter(!pass_mapping_QC)
 ```
@@
 
+The vertical line drawn correspods to the threshold explored in the notebook.
+
 ```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
@@
 
-Some cells have an excellent mapping score (>0,85) while other performed worse. Let's have a look at the compartments per patient.
+Some cells have an excellent mapping score (> `r params$mapping_score_thr`) while other performed worse. Let's have a look at the compartments per patient.
 
@@
 
-p + geom_hline(yintercept = params$mapping_score_thr)
+p + geom_hline(yintercept = params$mapping_score_thr)  + coord_cartesian(ylim = c(0.25,1))
 
@@
 
-p + geom_hline(yintercept = params$mapping_score_thr)
+p + geom_hline(yintercept = params$mapping_score_thr) + coord_cartesian(ylim = c(0.25,1))
 
@@
 We do not really see a pattern linking the label transfer score and some meta/clinical data.
-For the later CNV analysis, we can apply a threshold of `r params$mapping_score_thr` for the `fetal_kidney_predicted.compartment.score` while selecting for normal cells, i.e.
-`immune` and/or `endothelial` cells.
 
@@
                     feature = "ENSG00000110799",
-                    group.by = "compartment")
+                    group.by = "compartment") +
+  ggtitle("VWF expression averaged by compartment for all cells")
+```
+
+```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
+do_Feature_mean( df = cell_type_df_pass,
+                    feature = "ENSG00000110799",
+                    group.by = "compartment") +
+  ggtitle("VWF expression averaged by compartment for cells passing the mapping score threshold")
+```
+
+```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
+do_Feature_mean( df = cell_type_df_nopass,
+                    feature = "ENSG00000110799",
+                    group.by = "compartment") +
+  ggtitle("VWF expression averaged by compartment for cells that don't pass the mapping score threshold")
 ```
@@
                     group.by = "compartment",
-                    split.by = "sample_id")
+                    split.by = "sample_id")  +
+  ggtitle("boxplot of VWF expression for all cells")
 
@@
                     feature = "ENSG00000081237",
-                  group.by = "compartment")
+                    group.by = "compartment") +
+  ggtitle("PTPRC expression averaged by compartment for all cells")
+```
+
+```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
+do_Feature_mean( df = cell_type_df_pass,
+                    feature = "ENSG00000081237",
+                    group.by = "compartment") +
+  ggtitle("PTPRC expression averaged by compartment for cells passing the mapping score threshold")
+```
+
+```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
+do_Feature_mean( df = cell_type_df_nopass,
+                    feature = "ENSG00000081237",
+                    group.by = "compartment") +
+  ggtitle("PTPRC expression averaged by compartment for cellsthat are not passing the mapping score threshold")
 ```
@@
                     group.by = "compartment",
-                    split.by = "sample_id")
+                    split.by = "sample_id")  +
+  ggtitle("boxplot of PTPRC expression for all cells")
+
+```
+
+### Umap reduction
+
+We look at the umap reduction per sample.
+
+#### All cells
+
+Point are colored per compartment:
+
+- red = endothelial
+
+- green = fetal nephron
+
+- cyan = immune
+
+- purple = stroma
+
+Black cross show cell that do not pass the mapping threshold.
+
+```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
+ggplot(cell_type_df, aes( x = umap.umap_1, y = umap.umap_2, color = compartment), shape = 19, size = 2)+
+  geom_point() +
+  geom_point(data = cell_type_df_nopass, mapping = aes( x = umap.umap_1, y = umap.umap_2),  shape = 4, color = "black", alpha = 0.3) +
+  facet_wrap(facets = ~sample_id)
+```
+
+#### Cells passing the mapping score threshold
+
+Point are colored per compartment. 
+Here we only plotted cells that do pass the mapping score threshold.
+
+```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
+ggplot(cell_type_df_pass, aes( x = umap.umap_1, y = umap.umap_2, color = compartment), shape = 19, size = 2)+
+  geom_point() +
+  facet_wrap(facets = ~sample_id)
 
@@
 
-For that reason, we will first try to run `CopyKAT` on 2-5 samples that have a decent amount (>100 cells when possible) of normal cells (endothelial and immune cells) and for which we are quite confident with the label transfer:
+For that reason, we will first try to run `CopyKAT` on 2-5 samples that have a majority of cells mapping to kidney and a decent amount (>100 cells when possible) of normal cells (endothelial and immune cells) and for which we are quite confident with the label transfer:
 
-- sample `SCPCS000194` has > `85` % of cells predicted as kidney and `234` + `83` endothelium and immune cells.
+- sample `SCPCS000194`
 
-- sample `SCPCS000179` has > `94` % of cells predicted as kidney and `25` + `111` endothelium and immune cells.
+- sample `SCPCS000179`
 
-- sample `SCPCS000184` has > `96` % of cells predicted as kidney and `39` + `70` endothelium and immune cells.
+- sample `SCPCS000184`
 
-- sample `SCPCS000205` has > `89` % of cells predicted as kidney and `92` + `76` endothelium and immune cells.
+- sample `SCPCS000205`
 
-- sample `SCPCS0000208` has > `95` % of cells predicted as kidney and `18` + `35` endothelium and immune cells.
+- sample `SCPCS0000208`
 

### 2024-10-23 analyses/cell-type-dsrct/exploratory_analysis/01-marker-gene-tumor-classification.Rmd
Apply suggestions from code review
modified the saves

Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 umap_df_list <- lapply(sce_list, function(x) {
-  # get the gene expression counts for all marker genes
-  marker_gene_exp <- logcounts(x[marker_genes, ]) |>
-    as.matrix() |>
-    t() |>
-    as.data.frame() |>
-    tibble::rownames_to_column("barcodes")
-
-  # pull out the UMAP coordinates and make a data frame to use for plotting
+# pull out the UMAP coordinates and genes and make a data frame to use for plotting
   umap_df <- x |>
-    scuttle::makePerCellDF(use.dimred = "UMAP") |>
+    scuttle::makePerCellDF(features = marker_genes, use.dimred = "UMAP") |>
     # replace UMAP.1 with UMAP1
     dplyr::rename_with(\(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")) |>
-    # add in marker gene expression to dataframe
-    dplyr::left_join(marker_gene_exp, by = "barcodes") |>
     # combine all genes into a single column for easy faceting
@@
 
-names(umap_df_list) <- metadata_DSRCT$scpca_sample_id
 ```
@@
     filename = paste0(
-      '../plots/UMAP_marker_expression_',
+      module_base,
+      '/plots/UMAP_marker_expression_',
       names(umap_df_list)[i],
@@
     filename = paste0(
-      '../plots/UMAP_classification_',
+      module_base,
+      '/plots/UMAP_classification_',
       names(umap_df_list)[i],
@@
   ggsave(
-    filename = paste0('../plots/UMAP_marker_expression_cluster_', names(umap_df_list)[i], '.png'),
+    filename = paste0(module_base, '/plots/UMAP_marker_expression_cluster_', names(umap_df_list)[i], '.png'),
     plot = p,
@@
     filename = paste0(
+      module_base,
       '../plots/UMAP_classification_cluster_',

### 2024-10-28 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
use separate query variable to prevent feature loss, and rm when done for space

@@
 # prepare the query for label transfer
+# we don't want to overwrite the srat object since `prepare_query`
+#  removes features that are not present in the reference
 DefaultAssay(srat) <- "RNA"
-srat <- prepare_query(srat, rownames(reference), params$homologs_file)
+query <- prepare_query(srat, rownames(reference), params$homologs_file)
 ```
@@
 }
-s <- transfer_labels(
-  srat,
+query_labeled <- transfer_labels(
+  query,
   reference,
@@
 )
-metadata_to_trasfer <- s@meta.data[, annotation_columns]
+metadata_to_trasfer <- query_labeled@meta.data[, annotation_columns]
 
 srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_full_", annotation_columns))
+
+rm(query)
+rm(query_labeled)
 ```

### 2024-10-28 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Update README.md file

@@
    theme(text = element_text(size = 20)) +
-   theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5)) +
-   guides(fill=guide_legend(title=paste0(feature)))
+   theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5), title = element_text(size = rel(0.75))) +
+   guides(fill=guide_colourbar(title=paste0(feature)))
 

### 2024-10-28 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
update notebook and reports

@@
                     group.by = "compartment") +
-  ggtitle("PTPRC expression averaged by compartment for cellsthat are not passing the mapping score threshold")
+  ggtitle("PTPRC expression averaged by compartment for cells that are not passing the mapping score threshold")
 ```

### 2024-10-28 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
update mapping_score --> predicted.score

--- 2024-10-28 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
@@
 ---
-title: "Annotation exploration for SCPCP000006, mapping score threshold `r params$mapping_score_thr`"
+title: "Annotation exploration for SCPCP000006, `predicted.score` threshold `r params$predicted.score_thr`"
 author: "Maud PLASCHKA"
@@
 params: 
-  mapping_score_thr: 0.85
+  predicted.score_thr: 0.85
 output: 
@@
 
-We also looked into the mapping scores for each of the compartments and explored the potential effects of score thresholds.
-In this notebook, we fixed the mapping score threshold to `r params$mapping_score_thr`.
+We also looked into the `predicted.score` for each of the compartments and explored the potential effects of score thresholds.
+In this notebook, we fixed the `predicted.score` threshold to `r params$predicted.score_thr`.
 
@@
   # Create column for whether the compartment score passes threshold
-  dplyr::mutate(pass_mapping_QC = compartment_score > params$mapping_score_thr)
+  dplyr::mutate(pass_mapping_QC = compartment_score > params$predicted.score_thr)
 
-# we create a data frames of annotation of cells that pass the mapping score threshold
+# we create a data frames of annotation of cells that pass the `predicted.score` threshold
 cell_type_df_pass <- cell_type_df  |>
@@
 
-# we create a data frames of annotation of cells that don't pass the mapping score threshold
+# we create a data frames of annotation of cells that don't pass the `predicted.score` threshold
 cell_type_df_nopass <- cell_type_df  |>
@@
 ```
-### Label transfer (mapping) scores for the four compartments
+
+### Label transfer predicted.score for the four compartments
 
@@
                     )
-p +  geom_vline(xintercept=params$mapping_score_thr)
+p +  geom_vline(xintercept=params$predicted.score_thr)
 ```
 
-Some cells have an excellent mapping score (> `r params$mapping_score_thr`) while other performed worse. Let's have a look at the compartments per patient.
+Some cells have an excellent `predicted.score` (> `r params$predicted.score_thr`) while other performed worse. Let's have a look at the compartments per patient.
 
@@
 
-p + geom_hline(yintercept = params$mapping_score_thr)  + coord_cartesian(ylim = c(0.25,1))
+p + geom_hline(yintercept = params$predicted.score_thr)  + coord_cartesian(ylim = c(0.25,1))
 
@@
 
-Regarding the endothelial cells, some patient have very high mapping scores while others perform poorly. 
+Regarding the endothelial cells, some patient have very high `predicted.score` while others perform poorly. 
 Let's have a look if this can be linked to some metadata/clinical data.
@@
 
-p + geom_hline(yintercept = params$mapping_score_thr) + coord_cartesian(ylim = c(0.25,1))
+p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25,1))
 
@@
 
-We finally look for each patient and compartment, the number of cells that have a mapping score > `r params$mapping_score_thr` (TRUE).
+We finally look for each patient and compartment, the number of cells that have a `predicted.score` > `r params$predicted.score_thr` (TRUE).
 
@@
                     group.by = "compartment") +
-  ggtitle("VWF expression averaged by compartment for cells passing the mapping score threshold")
+  ggtitle("VWF expression averaged by compartment for cells passing the `predicted.score` threshold")
 ```
@@
                     group.by = "compartment") +
-  ggtitle("VWF expression averaged by compartment for cells that don't pass the mapping score threshold")
+  ggtitle("VWF expression averaged by compartment for cells that don't pass the `predicted.score` threshold")
 ```
@@
                     group.by = "compartment") +
-  ggtitle("PTPRC expression averaged by compartment for cells passing the mapping score threshold")
+  ggtitle("PTPRC expression averaged by compartment for cells passing the `predicted.score` threshold")
 ```
@@
                     group.by = "compartment") +
-  ggtitle("PTPRC expression averaged by compartment for cellsthat are not passing the mapping score threshold")
+  ggtitle("PTPRC expression averaged by compartment for cells that are not passing the predicted score threshold")
 ```
@@
 
-Black cross show cell that do not pass the mapping threshold.
+Black cross show cell that do not pass the `predicted.score` threshold.
 
@@
 
-#### Cells passing the mapping score threshold
+#### Cells passing the `predicted.score` threshold
 
 Point are colored per compartment. 
-Here we only plotted cells that do pass the mapping score threshold.
+Here we only plotted cells that do pass the `predicted.score` threshold.
 

### 2024-10-28 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd
few changes to PR828

@@
 
+- `reference_value` indicates the reference used for `infercnv`, either endothelial, immune, both or pool.
+
 ```{r}
 
-Do_CNV_heatmap <- function(object, infercnv_obj, group.by){
+Do_CNV_heatmap <- function(object, infercnv_obj, group.by, reference_value){
 
@@
 
-- `features` the gene or quantitative value to be plotted
+- `feature` the gene or quantitative value to be plotted
 
@@
 ```{r }
-visualize_feature <- function(object, features, group.by ){
+visualize_feature <- function(object, feature, group.by ){
 
                   d <- SCpubr::do_FeaturePlot(object, 
-                                              features = features, 
+                                              features = feature, 
                                               pt.size = 0.2, 
@@
                                               legend.length = 5, 
-                                              legend.position = "right") + ggtitle(features)
+                                              legend.position = "right") + ggtitle(as.character(feature))
                   b <- SCpubr::do_ViolinPlot(object, 
-                                             features = features, 
+                                             features = feature, 
                                              ncol = 1, 
@@
                                              legend.position = "none",
-                                             assay = "SCT") + ylab(features)
+                                             assay = "SCT") + ylab(as.character(feature))
 
@@
   infercnv_obj[[reference_value]] <- readRDS(file.path(result_dir, glue::glue(reference_value, "_HMM-no") , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, "_HMM-no.rds")))
-  print(Do_CNV_heatmap(object = srat, infercnv_obj = infercnv_obj[[reference_value]], group.by = "fetal_kidney_predicted.compartment"))
+  print(Do_CNV_heatmap(object = srat, infercnv_obj = infercnv_obj[[reference_value]], group.by = "fetal_kidney_predicted.compartment", reference_value = reference_value))
 }
@@
   srat <- Do_CNV_score(srat,infercnv_obj = infercnv_obj[[reference_value]], reference_value)
-  p1 <- visualize_feature(srat, glue::glue("CNV-score_", reference_value) , group.by = "fetal_kidney_predicted.compartment" )
+  p1 <- visualize_feature(srat, feature =  glue::glue("CNV-score_", reference_value) , group.by = "fetal_kidney_predicted.compartment" )
   p2 <- visualize_density(srat, features=glue::glue("CNV-score_", reference_value), group.by = "fetal_kidney_predicted.compartment")
@@
 for(i in 1:22){
-  print(visualize_feature(srat_i3, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
+  print(visualize_feature(srat_i3, feature=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
 }
@@
 for(i in 1:22){
-  print(visualize_feature(srat_i6, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
+  print(visualize_feature(srat_i6, features=c(glue::glue("proportion_cnv_chr", i)), group.by = "fetal_kidney_predicted.compartment"))
 }
@@
 
-The selection of normal cells to be spiked-in as a normal reference prior running `infercnv` has been performed in `06a_build-normal-cell_reference.R.
+The selection of normal cells to be spiked-in as a normal reference prior running `infercnv` has been performed in `06b_build-normal-cell_reference.R.
 We selected samples that haven't been pre-treated with chemotherapies.
 Indeed, even if normal cells shouldn't be affected by chemotherapy in terms of CNV, we are not sure how chemotherapies can affect the CNV profile of normal cells.
-We thus decided to only take endothelial and immune cells from non-treated samples to build the reference of normal cells.
+We thus decided to only take endothelial and immune cells from non-treated samples whose label transfer scores are greater than 0.85 to build the reference of normal cells.
 
@@
 for(i in 1:22){
-  print(visualize_feature(srat, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
+  print(visualize_feature(srat, feature=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
 }

### 2024-10-28 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd
few changes to PR828

@@
 for(i in 1:22){
-  print(visualize_feature(srat_i6, features=c(glue::glue("proportion_cnv_chr", i)), group.by = "fetal_kidney_predicted.compartment"))
+  print(visualize_feature(srat_i6, feature=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
 }

### 2024-10-28 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd
update notebook

@@
 
+#### Wrapper function to explore `infercnv` HMM cnv prediction
+
+The `wrapper_explore_hmm` take as input the `infercnv_obj` generated with `infercnv` HMM cnv predictions.
+The wrapper allows the following steps and plots:
+
+##### Feature plot and repartition of the CNV per chromosome
+
+For each chromosome, we look at the repartition of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
+`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+
+#####  Distribution of CNV estimation in the Wilms tumor copartments
+
+For each chromosome, we look at the distribution of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
+`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+
+We are quite confident that immune and endothelial cells are well identified by label transfer done in `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd`. 
+The distribution of CNV for endothelial and immune cells should thus be a single peak center on 0.
+
+We do not know if fetal nephron and stroma cells are a mix of normal and cancer cells. 
+Would they be a group of normal cells, we should expect a single peak center on 0 for every chromosome. 
+As we expect to have a large number of cancer with heterogeneous CNV, we should see multiple peaks.
+
+##### DotPlot
+
+The `Dotplot` representation summarizes the percentage of cells in each compartment with cnv in each of the 22 chromosomes.
+
+##### CNV score
+
+Finally, we try to calculate  single CNV score and assess its potential in identifying cells with CNV versus normal cells without CNV.
+
+We simply checked for each chromosome if the cell `has_cnv_chr`. 
+Would the cell have more than `cnv_threshold` chromosome with CNV, the global `has_cnv_score` will be TRUE.
+Else, the cell will have a `has_cnv_score` set to FALSE.
+
+```{r}
+
+wrapper_explore_hmm <- function(infercnv_obj, cnv_threshold = 1){
+
+  p <- list()
+  #### Plot UMAP of known Wilms tumor related alteration
+  p[["compartment"]] <- visualize_metadata(infercnv_obj, "fetal_kidney_predicted.compartment", "seurat_clusters")
+  #### Feature plot and repartition of the CNV per chromosome
+  for(i in 1:22){
+  tmp1 <- visualize_feature(infercnv_obj, feature=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment")
+  tmp2 <- visualize_density(infercnv_obj, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment")
+  p[[glue::glue("proportion_cnv_chr", i)]] <- tmp1 + tmp2 + plot_layout(ncol = 3, widths = c(1,2,2))
+  }
+  #### DotPlot
+  p[["Dotplot"]] <- DotPlot(infercnv_obj, features =  colnames(infercnv_obj@meta.data)[grepl("has_cnv_chr", colnames(infercnv_obj@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+  #### CNV score
+  meta <- infercnv_obj@meta.data
+  meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
+  meta$has_cnv_score[meta$has_cnv_score > cnv_threshold] <- "CNV"
+
+  # we know that some immune cells might have false positive CNV 
+  meta$has_cnv_score[meta$has_cnv_score %in% c(0:cnv_threshold)] <- "no CNV"
+  infercnv_obj <- AddMetaData(infercnv_obj, meta$has_cnv_score, col.name =  "has_cnv_score") 
+  p[["infercnv_obj"]] <- infercnv_obj
+
+  p[["cnv_seurat"]] <- visualize_metadata(infercnv_obj, "has_cnv_score", "seurat_clusters")
+  p[["cnv_compartment"]] <- visualize_metadata(infercnv_obj, "has_cnv_score", "fetal_kidney_predicted.compartment")
+
+return(p)
+
+}
+
+```
+
 ### Base directories
@@
 
-#### Load the `Seurat` object
-
 We load the `Seurat` object generated in `06_infercnv.R`
@@
 
+#### Load the `Seurat` object
 srat_i3 <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-both.rds")))
@@
 
-#### Plot UMAP of known Wilms tumor related alteration
+and explore the CNV results using the `wrapper_explore_hmm` fucntion.
 
-As a reminder, we plot the label transfer from the fetal kidney atlas.
-
-Immune and endothelial cells shouldn't show any CNV.
-
-The stroma and fetal nephron compartments may contain normal cells with no CNV or cancer cells with low to high CNV profiles.
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-visualize_metadata(srat_i3, "fetal_kidney_predicted.compartment", "seurat_clusters")
-
-```
-
-#### Feature plot and repartition of the CNV per chromosome
-
-For each chromosome, we look at the repartition of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
-`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
-
-This should allow a quick visual check that immune and endothelial cells do not expressed any CNV. 
-If not, we should check if genes driving the CNV score are not key immune or endothelium genes.
-
-The `umap` reduction visualization might allow the identification of cluster(s) of normal cells with no CNV.
-
-```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
-for(i in 1:22){
-  print(visualize_feature(srat_i3, feature=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
-}
-```
-
-####  CNV distribution in the Wilms tumor compartments
-
-For each chromosome, we look at the distribution of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
-`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
-
-We are quite confident that immune and endothelial cells are well identified by label transfer done in `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd`. 
-The distribution of CNV for endothelial and immune cells should thus be a single peak center on 0.
-
-We do not know if fetal nephron and stroma cells are a mix of normal and cancer cells. 
-Would they be a group of normal cells, we should expect a single peak center on 0 for every chromosome. 
-As we expect to have a large number of cancer with heterogeneous CNV, we should see multiple peaks.
-
-```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
-for(i in 1:22){
-  print(visualize_density(srat_i3, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
-}
-```
-
-#### DotPlot
-
-The `Dotplot` representation summarizes the percentage of cells in each compartment with cnv in each of the 22 chromosomes.
-
-The HMM-i3 prediction model with intra-patient reference (composed of immune and endothelium cells) perform really well in identifying Wilms tumor relevant CNV.
-No false positive CNV are observed in the immune and endothelium compartment.
-
-```{r fig.width=15, fig.height=5, out.width='100%'}
-
-DotPlot(srat_i3, features =  colnames(srat_i3@meta.data)[grepl("has_cnv_chr", colnames(srat_i3@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
-
-```
-
-#### CNV score
-
-Here, we try to calculate a single CNV score and assess its potential in iditenfying cells with CNV versus normal cells without CNV.
-
-We simply checked for each chromosome if the cell `has_cnv_chr`. 
-Would the cell have one or more chromosome with CNV, the global `has_cnv_score` will be TRUE.
-Else, the cell will have a `has_cnv_score` set to FALSE.
-
-```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
-meta <- srat_i3@meta.data
-
-meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
-meta$has_cnv_score[meta$has_cnv_score > 1] <- "CNV"
-
-# we know that some immune cells might have false positive CNV 
-meta$has_cnv_score[meta$has_cnv_score %in% c(0,1)] <- "no CNV"
-
-srat_i3 <- AddMetaData(srat_i3, meta$has_cnv_score, col.name =  "has_cnv_score")
-
-```
-
-We identified clusters with no/low CNV while the majority of the `Seurat` cluster present CNV on more than one chromosome.
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-visualize_metadata(srat_i3, "has_cnv_score", "seurat_clusters")
-
-```
-
-Immune and endothelial cells have a majority of cells with no/low CNV profile, which confirm their annotation as normal cells. 
-The fetal nephron and stroma copartments are mainly composed of cancer cells with more than one chromosome affected by a CNV.
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-visualize_metadata(srat_i3, "has_cnv_score", "fetal_kidney_predicted.compartment")
-
+```{r fig.width=16, fig.height=8, out.width='100%', out.height='100%', warning=FALSE}
+p <- list()
+p <- wrapper_explore_hmm(srat_i3, cnv_threshold = 1)
+p
 ```
@@
 We then explore `infercnv` results generated with immune and endothelial cells as reference, using a [HMM-i6 model](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type).
-
-#### Load the `Seurat` object
 
@@
 
-#### Feature plot and repartition of the CNV per chromosome
-
-For each chromosome, we look at the repartition of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
-`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
-
-```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
-for(i in 1:22){
-  print(visualize_feature(srat_i6, feature=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
-}
-```
-
-####  Distribution of CNV estimation in the Wilms tumor copartments
-
-For each chromosome, we look at the distribution of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
-`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
-
-We are quite confident that immune and endothelial cells are well identified by label transfer done in `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd`. 
-The distribution of CNV for endothelial and immune cells should thus be a single peak center on 0.
-
-We do not know if fetal nephron and stroma cells are a mix of normal and cancer cells. 
-Would they be a group of normal cells, we should expect a single peak center on 0 for every chromosome. 
-As we expect to have a large number of cancer with heterogeneous CNV, we should see multiple peaks.
-
-```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
-for(i in 1:22){
-  print(visualize_density(srat_i6, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
-}
-```
-
-#### DotPlot
-
-The `Dotplot` representation summarizes the percentage of cells in each compartment with cnv in each of the 22 chromosomes.
-
-The HMM-i3 prediction model with intra-patient reference (composed of immune and endothelium cells) perform really well in identifying Wilms tumor relevant CNV.
-No false positive CNV are observed in the immune and endothelium compartment.
-
-```{r fig.width=15, fig.height=5, out.width='100%'}
-
-DotPlot(srat_i6, features =  colnames(srat_i6@meta.data)[grepl("has_cnv_chr", colnames(srat_i6@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
-
-```
-
-#### CNV score
-
-Finally, we try to calculate  single CNV score and assess its potential in iditenfying cells with CNV versus normal cells without CNV.
-
-We simply checked for each chromosome if the cell `has_cnv_chr`. 
-Would the cell have one or more chromosome with CNV, the global `has_cnv_score` will be TRUE.
-Else, the cell will have a `has_cnv_score` set to FALSE.
-
-```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
-meta <- srat_i6@meta.data
-
-meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
-meta$has_cnv_score[meta$has_cnv_score > 1] <- "CNV"
-
-# we know that some immune cells might have false positive CNV 
-meta$has_cnv_score[meta$has_cnv_score %in% c(0,1)] <- "no CNV"
-
-srat_i6 <- AddMetaData(srat_i6, meta$has_cnv_score, col.name =  "has_cnv_score")
-
-```
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-visualize_metadata(srat_i6, "has_cnv_score", "seurat_clusters")
-
-```
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-visualize_metadata(srat_i6, "has_cnv_score", "fetal_kidney_predicted.compartment")
-
+```{r fig.width=16, fig.height=8, out.width='100%', out.height='100%', warning=FALSE}
+q <- list()
+q <- wrapper_explore_hmm(srat_i6, cnv_threshold = 1)
+q
 ```
@@
 # subset non spike-in cells
-keep_cells <- colnames(seurat_full)[!grepl("spike", colnames(seurat_full))]
+keep_cells <- colnames(srat)
 seurat <- subset(seurat_full, cells = keep_cells)
 
-meta <- seurat@meta.data
+meta <- seurat@meta.data[colnames(srat),]
 
 # merge the new info into the srat metadata
-srat <- AddMetaData(srat, meta[,grepl("_cnv_chr", colnames(meta))], col.name = colnames(meta[grepl("_cnv_chr", colnames(meta))]))
+srat_pull <- AddMetaData(srat, meta[,grepl("_cnv_chr", colnames(meta))], col.name = colnames(meta[grepl("_cnv_chr", colnames(meta))]))
 ```
 
-#### Plot UMAP of known Wilms tumor related alteration
-
-As a reminder, we plot the label transfer from the fetal kidney atlas.
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-visualize_metadata(srat, "fetal_kidney_predicted.compartment", "seurat_clusters")
-
-```
-
-#### Feature plot and repartition of the CNV per chromosome
-
-For each chromosome, we look at the repartition of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
-`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
-
-This should allow a quick visual check that immune and endothelial cells do not expressed any CNV. If so, we should check if genes driving the CNV score are not key immune or endothelium genes.
-
-Also, we should visualize if one area of the umap do not have CNV and could be identify as additional normal cell subset.
-
-```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
-for(i in 1:22){
-  print(visualize_feature(srat, feature=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
-}
-```
-
-####  Distribution of CNV estimation in the Wilms tumor copartments
-
-For each chromosome, we look at the distribution of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
-`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
-
-We are quite confident that immune and endothelial cells are well identified by label transfer done in `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd`. 
-The distribution of CNV for endothelial and immune cells should thus be a single peak center on 0.
-
-We do not know if fetal nephron and stroma cells are a mix of normal and cancer cells. 
-Would they be a group of normal cells, we should expect a single peak center on 0 for every chromosome. 
-As we expect to have a large number of cancer with heterogeneous CNV, we should see multiple peaks.
-
-```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
-for(i in 1:22){
-  print(visualize_density(srat, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
-}
-```
-
-#### DotPlot
-
-The `Dotplot` representation summarizes the percentage of cells in each compartment with cnv in each of the 22 chromosomes.
-
-The HMM-i3 prediction model with inter-patient reference (composed of immune and endothelium cells) perform quite well in identifying Wilms tumor relevant CNV.
-However, sample variation induced false positive CNV such as CNV on chromosome 5 and 13 observed in the endothelium compartment.
-
-```{r fig.width=15, fig.height=5, out.width='100%'}
-
-DotPlot(srat, features =  colnames(srat@meta.data)[grepl("proportion_cnv_chr", colnames(srat@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
-
-```
-
-#### CNV score
-
-Finally, we try to calculate  single CNV score and assess its potential in iditenfying cells with CNV versus normal cells without CNV.
-
-We simply checked for each chromosome if the cell `has_cnv_chr`. 
-Would the cell have three or more chromosome with CNV, the global `has_cnv_score` will be TRUE.
-Else, the cell will have a `has_cnv_score` set to FALSE.
-
-Of note, as we know that the use of a ~universal inter-patient reference of normal cells can induce false positive CNV in some chromosomes, we decided to allow up to 2 CNV in normal cells.
-
-```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
-meta <- srat@meta.data
-meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
-meta$has_cnv_score[meta$has_cnv_score > 2] <- "CNV"
-
-# we know that some immune cells might have false positive CNV 
-meta$has_cnv_score[meta$has_cnv_score %in% c(0,1, 2)] <- "no CNV"
-
-srat <- AddMetaData(srat, meta$has_cnv_score, col.name =  "has_cnv_score")
-
-```
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-visualize_metadata(srat, "has_cnv_score", "seurat_clusters")
-
-```
-
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
-visualize_metadata(srat, "has_cnv_score", "fetal_kidney_predicted.compartment")
-
+```{r fig.width=16, fig.height=8, out.width='100%', out.height='100%', warning=FALSE}
+pull <- list()
+pull <- wrapper_explore_hmm(srat_pull, cnv_threshold = 3)
+pull
 ```
@@
 
-p1 <- SCpubr::do_DimPlot(srat, "has_cnv_score", plot.title = "inter-patient reference, HMM-i3", reduction = "umap")
+p1 <- SCpubr::do_DimPlot(pull$infercnv_obj, "has_cnv_score", plot.title = "inter-patient reference, HMM-i3", reduction = "umap")
 
-p2 <- SCpubr::do_DimPlot(srat_i3, "has_cnv_score", plot.title = "intra-patient reference, HMM-i3", reduction = "umap")
+p2 <- SCpubr::do_DimPlot(p$infercnv_obj, "has_cnv_score", plot.title = "intra-patient reference, HMM-i3", reduction = "umap")
 
-p3 <- SCpubr::do_DimPlot(srat_i6, "has_cnv_score", plot.title = "intra-patient reference, HMM-i6", reduction = "umap")
+p3 <- SCpubr::do_DimPlot(q$infercnv_obj, "has_cnv_score", plot.title = "intra-patient reference, HMM-i6", reduction = "umap")
 

### 2024-10-29 analyses/cell-type-dsrct/exploratory_analysis/01-marker-gene-tumor-classification.Rmd

@@
 # The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "2024-07-08")
+data_dir <- file.path(repository_base, "data", "current")
 #sample_dir <- file.path(data_dir, "SCPCP000013", params$sample_id)
@@
     filename = paste0(
-      '../plots/marker_expression_',
+      module_base,
+      '/plots/marker_expression_',
       names(umap_df_list)[i],
@@
     filename = paste0(
-      '../plots/marker_distribution_',
+      module_base,
+      '/plots/marker_distribution_',
       names(umap_df_list)[i],
@@
     filename = paste0(
-      '../plots/marker_distribution_zscaled_',
+      module_base,
+      '/plots/marker_distribution_zscaled_',
       names(umap_df_list)[i],
@@
 Let's start with classifying tumor cells as tumor if any marker gene is present (z-score > 0).
-This also means the sum of all z-scores > 0.
 
@@
       module_base,
-      '../plots/UMAP_classification_cluster_',
+      '/plots/UMAP_classification_cluster_',
       names(umap_df_list)[i],
@@
 saveRDS(sce_file_list, '../results/SCPCP000013_sce_file_list.rds')
+# get an RDS of the UMAP data with marker genes
+saveRDS(umap_df_list, file.path(module_base, 'results/SCPCP000013_umap_df_list.rds'))
 ```

### 2024-10-29 analyses/cell-type-wilms-tumor-06/supplemental-notebooks/compare-label-transfer-approaches.Rmd
Add supplemental notebook with results comparing azimuth to adapted azimuth

---
title: "Compare label transfer results between Azimuth and Azimuth-adapted strategy"
author: Stephanie Spielman, Data Lab
output: 
  html_notebook: 
    toc: yes
    toc_float: yes
params:
  seed: 12345
---

The goal of this notebook is to compare label transfer results between:

- Label transfer code with Azimuth currently in `main` at commit `6af112d`. These results are referred to as `"azimuth"`.
- Label transfer code adapted from Azimuth. These results are referred to as `"adapted_azimuth"`.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
options(future.globals.maxSize = 891289600000000)

suppressPackageStartupMessages({
  library(tidyverse)
  library(patchwork)
  library(Seurat)
})

repository_base <- rprojroot::find_root(rprojroot::is_git_root)
module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
result_dir <- file.path(module_base, "results")

# functions to perform label transfer with azimuth-adapted approach
source(
  file.path(module_base, "notebook_template", "utils", "label-transfer-functions.R")
)

# Output files
full_results_file <- file.path(module_base, "scratch", "compare-label-transfer_fetal-full.rds")
kidney_results_file <- file.path(module_base, "scratch", "compare-label-transfer_fetal-kidney.rds")
```

## Functions

```{r functions}
# Make a heatmap of counts for label transfer strategies
plot_count_heatmap <- function(df, title, sample_id) {
  all_preds <- union(df$azimuth, df$adapted_azimuth)

  plotme <- data.frame(
    azimuth = all_preds,
    adapted_azimuth = all_preds
  ) |>
    expand(azimuth, adapted_azimuth) |>
    mutate(n = NA_integer_) |>
    anti_join(distinct(df)) |>
    bind_rows(
      df |> count(azimuth, adapted_azimuth)
    ) |>
    arrange(azimuth) |>
    mutate(
      color = case_when(
        is.na(n) ~ "white",
        n <= 20 ~ "grey90",
        n <= 50 ~ "lightblue",
        n <= 100 ~ "cornflowerblue",
        n <= 500 ~ "red",
        n <= 1000 ~ "yellow2",
        .default = "yellow"
      )
    )

  ggplot(plotme) +
    aes(x = azimuth, y = adapted_azimuth, fill = color, label = n) +
    geom_tile(alpha = 0.5) +
    geom_abline(color = "firebrick", alpha = 0.5) +
    geom_text(size = 3.5) +
    # scale_fill_viridis_c(name = "count", na.value = "grey90") +
    scale_fill_identity() +
    theme_bw() +
    theme(
      axis.text.y = element_text(size = 7),
      axis.text.x = element_text(angle = 30, size = 7, hjust = 1),
      legend.position = "bottom",
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    labs(
      title = glue::glue("{sample_id}: {str_to_title(title)}")
    )
}

# Wrapper function to compare results between approaches
# Makes two plots:
# - heatmap comparing counts for cell labels between approaches
# - density plot of annotation scores for labels that agree and disagree between approaches
compare <- function(df, compare_column, score_column, title) {
  spread_df <- df |>
    select({{ compare_column }}, barcode, version) |>
    pivot_wider(names_from = version, values_from = {{ compare_column }})

  heatmap <- plot_count_heatmap(spread_df, title, unique(df$sample_id))

  disagree_barcodes <- spread_df |>
    filter(azimuth != adapted_azimuth) |>
    pull(barcode)

  df2 <- df |>
    mutate(
      agree = ifelse(barcode %in% disagree_barcodes, "labels disagree", "labels agree"),
      agree = fct_relevel(agree, "labels disagree", "labels agree")
    )

  density_plot <- ggplot(df2) +
    aes(x = {{ score_column }}, fill = agree) +
    geom_density(alpha = 0.6) +
    theme_bw() +
    ggtitle(
      glue::glue("Disagree count: {length(disagree_barcodes)} out of {nrow(spread_df)}")
    ) +
    theme(legend.position = "bottom")

  print(heatmap + density_plot + plot_layout(widths = c(2, 1)))
}
```

## Label transfer

This section both:

- Reads in existing Azimuth label transfer results
- Performs label transfer with Azimuth-adapted approach

If results are already available, we read in the files rather than regenerating results.

```{r}
# sample ids to process
sample_ids <- c("SCPCS000179", "SCPCS000184", "SCPCS000194", "SCPCS000205", "SCPCS000208")

# read in seurat input objects, as needed
if ((!file.exists(full_results_file)) || (!file.exists(kidney_results_file))) {
  srat_objects <- sample_ids |>
    purrr::map(
      \(id) {
        srat <- readRDS(
          file.path(result_dir, id, glue::glue("01-Seurat_{id}.Rds"))
        )
        DefaultAssay(srat) <- "RNA"

        return(srat)
      }
    )
  names(srat_objects) <- sample_ids
}
```

### Label transfer for fetal full

```{r}
if (!file.exists(full_results_file)) {
  # read reference
  ref <- readRDS(file.path(
    module_base,
    "results",
    "references",
    "cao_formatted_ref.rds"
  ))
  full_reference <- ref$reference
  full_refdata <- ref$refdata
  full_dims <- ref$dims
  full_annotation_columns <- c(
    glue::glue("predicted.{ref$annotation_levels}"),
    glue::glue("predicted.{ref$annotation_levels}.score")
  )

  fetal_full <- srat_objects |>
    purrr::imap(
      \(srat, id) {
        # Perform label transfer with new code
        set.seed(params$seed)
        query <- prepare_query(srat, rownames(full_reference), file.path(module_base, "scratch", "homologs.rds"))
        query <- transfer_labels(
          query,
          full_reference,
          full_dims,
          full_refdata
        )

        # Read in results from existing Azimuth label transfer code
        srat_02a <- readRDS(
          file.path(result_dir, id, glue::glue("02a-fetal_full_label-transfer_{id}.Rds"))
        )

        # create final data frame with all annotations
        query@meta.data[, full_annotation_columns] |>
          tibble::rownames_to_column(var = "barcode") |>
          mutate(
            sample_id = id,
            version = "adapted_azimuth"
          ) |>
          # existing results
          bind_rows(
            data.frame(
              sample_id = id,
              barcode = colnames(srat_02a),
              version = "azimuth",
              predicted.annotation.l1 = srat_02a$fetal_full_predicted.annotation.l1,
              predicted.annotation.l1.score = srat_02a$fetal_full_predicted.annotation.l1.score,
              predicted.annotation.l2 = srat_02a$fetal_full_predicted.annotation.l2,
              predicted.annotation.l2.score = srat_02a$fetal_full_predicted.annotation.l2.score,
              predicted.organ = srat_02a$fetal_full_predicted.organ,
              predicted.organ.score = srat_02a$fetal_full_predicted.organ.score
            )
          )
      }
    )
  write_rds(fetal_full, full_results_file)
} else {
  fetal_full <- read_rds(full_results_file)
}
```

### Label transfer for fetal kidney

```{r}
if (!file.exists(kidney_results_file)) {
  # read reference
  ref <- readRDS(file.path(
    module_base,
    "results",
    "references",
    "stewart_formatted_ref.rds"
  ))

  # Pull out information from the reference object we need for label transfer
  kidney_reference <- ref$reference
  kidney_refdata <- ref$refdata
  kidney_dims <- ref$dims
  kidney_annotation_columns <- c(
    glue::glue("predicted.{ref$annotation_levels}"),
    glue::glue("predicted.{ref$annotation_levels}.score")
  )

  fetal_kidney <- srat_objects |>
    purrr::imap(
      \(srat, id) {
        # Perform label transfer with new code
        set.seed(params$seed)
        query <- prepare_query(srat, rownames(kidney_reference), file.path(module_base, "scratch", "homologs.rds"))
        query <- transfer_labels(
          query,
          kidney_reference,
          kidney_dims,
          kidney_refdata
        )

        # Read in results from existing Azimuth label transfer code
        srat_02b <- readRDS(
          file.path(result_dir, id, glue::glue("02b-fetal_kidney_label-transfer_{id}.Rds"))
        )

        # create final data frame with all annotations
        query@meta.data[, kidney_annotation_columns] |>
          tibble::rownames_to_column(var = "barcode") |>
          mutate(
            sample_id = id,
            version = "adapted_azimuth"
          ) |>
          # existing results
          bind_rows(
            data.frame(
              sample_id = id,
              barcode = colnames(srat_02b),
              version = "azimuth",
              predicted.compartment = srat_02b$fetal_kidney_predicted.compartment,
              predicted.compartment.score = srat_02b$fetal_kidney_predicted.compartment.score,
              predicted.cell_type = srat_02b$fetal_kidney_predicted.cell_type,
              predicted.cell_type.score = srat_02b$fetal_kidney_predicted.cell_type.score
            )
          )
      }
    )

  write_rds(fetal_kidney, kidney_results_file)
} else {
  fetal_kidney <- read_rds(kidney_results_file)
}
```

## Compare results

We expect:
- The majority of annotations match between approaches, with heatmap counts primarily falling along the diagonal
- Any annotations that disagree should have low scores

### Fetal full reference

Note that results from the L2 reference are not plotted because they are not used in cell type annotation.

```{r fig.height=8, fig.width=14}
fetal_full |>
  purrr::walk(
    \(dat) {
      compare(dat, predicted.annotation.l1, predicted.annotation.l1.score, "l1")
      compare(dat, predicted.organ, predicted.organ.score, "organ")
    }
  )
```

### Fetal kidney reference

```{r fig.height=8, fig.width=14}
fetal_kidney |>
  purrr::walk(
    \(dat) {
      compare(dat, predicted.compartment, predicted.compartment.score, "compartment")
      compare(dat, predicted.cell_type, predicted.cell_type.score, "cell_type")
    }
  )
```

## Conclusions

The vast majority of the time, labels agree. 
Generally speaking, when labels do not agree, their annotation scores are much lower, which is as expected.

Additional notable differences are shown in tables below:

### Fetal full reference:

- The Azimuth-adapted approach occasionally calls kidney or kidney-related cells as intestine or intestine epithelial
- Some other kidney-related differences are noted:

| Sample | Reference | Count | Azimuth | Azimuth-adapted |
|--------|-----------|-------|---------|-----------------|
| SCPSC000179 | L1 | 70 | Metanephritic cells | Intestinal epithelial cells | 
| SCPSC000179 | Organ | 64 | Kidney | Intestine | 
| SCPSC000179 | Organ | 20 | Lung | Kidney | 
| SCPSC000194 | L1 | 60 | Stromal cells | Mesangial cells | 
| SCPSC000194 | Organ | 35 | Kidney | Intestine | 
| SCPSC000194 | Organ | 36 | Lung | Kidney | 
| SCPSC000205 | Organ | 56 | Kidney | Intestine |
| SCPSC000208 | L1 | 101 | Mesangial cells | Metanephritic cells |  
| SCPSC000208 | L1 | 101 | Intestinal epithelial cells | Metanephritic cells |  
| SCPSC000208 | Organ | 149 | Kidney | Intestine |

### Fetal kidney reference:

- There are a small but notable number of cells flipped between mesenchyme and kidney cells, in particular for sample SCPSC000184.
This is the main discrepancy.
- Most of the cell type differences are not necessarily biologically meaningful for our purposes, as listed below. These are not noted in the table.
   - `kidney cell` vs `podocyte`
   - `kidney epithelial cell` vs `kidney cell` 
   - `mesenchymal cell` vs `mesenchymal stem cell` 
- There are a decent number of times when stroma and fetal nephron are flipped, but this makes sense given that we expect many of the stroma may be tumor.
- Disagreeing annotation scores when using the cell type reference were often higher, but many of the disagreements for this reference were not meaningful (e.g. `kidney epithelial cell` vs `kidney cell`).

| Sample | Reference | Count | Azimuth | Azimuth-adapted |
|--------|-----------|-------|---------|-----------------|
| SCPSC000179 | cell type | 202 | mesenchymal cell | kidney epithelial cell | 
| SCPSC000184 | compartment | 111 | fetal nephron |  stroma | 
| SCPSC000184 | cell type | 536 | kidney epithelial cell | mesenchymal cell | 
| SCPSC000194 | compartment | 565 | fetal nephron |  stroma | 
| SCPSC000194 | cell type  | 89 | kidney epithelial cell | mesenchymal cell | 
| SCPSC000205 | compartment  | 684 | fetal nephron |  stroma | 
| SCPSC000208 | compartment  | 2111 | fetal nephron |  stroma |

## Session Info

```{r}
sessionInfo()
```

### 2024-10-29 analyses/cell-type-wilms-tumor-06/notebook/07_annotation_Across_Samples_exploration.Rmd
Draft annotation

---
title: "Annotation exploration for SCPCP000006"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params:
  predicted.celltype.threshold: 0.85
  cnv_threshold: 0
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE,
                      knitr.digits.signif = FALSE)

```

## Introduction

The aim is to combine label transfer and CNV inference to annotate Wilms tumor samples in SCPCP000006.
The proposed annotation will be based on the combination of:

- the label transfer from the fetal kidney reference (Stewart et al.), in particular the `fetal_kidney_predicted.compartment` and `fetal_kidney:predicted.cell_type`, as well as the `mapping.score` for each compartment,

- the predicted CNV calculated using intra-sample endothelial and immune cells (`--reference both`) as normal reference

In a second time, we will explore and validate the chosen annotation.

We will use some of the [markers genes](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-wilms-tumor-06#the-table-celltype_metadatacsv-contains-the-following-column-and-information) to validate visually the annotations.

The analysis can be summarized as the following:

_Where `cnv.thr` and `map.thr` need to be discussed_

first   level annotation | second level   annotation | selection of   the cells | marker genes   for validation | cnv validation
-- | -- | -- | -- | --
normal | endothelial | compartment == "endothelium"     & mapping_score > map.thr     & cnv_score < cnv.thr | WVF | no   cnv
normal | immune | compartment == "immune"     & mapping_score > map.thr     & cnv_score < cnv.thr | PTPRC, CD163, CD68 | no cnv
normal | kidney | cell_type %in% c("kidney cell",   "kidney epithelial", "podocyte")     & mapping_score > map.thr     & cnv_score < cnv.thr | CDH1, PODXL, LTL | no cnv
normal | stroma | compartment == "stroma"     & mapping_score > map.thr     & cnv_score < cnv.thr | VIM | no cnv
cancer | stroma | compartment == "stroma"     & cnv_score > cnv.thr | VIM | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
cancer | blastema | compartment ==   "fetal_nephron"     & cell_type == "mesenchymal cell"     & cnv_score > cnv.thr | CITED1 | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
cancer | epithelial | compartment ==   "fetal_nephron"     & cell_type != "mesenchymal cell"     & cnv_score > cnv.thr | CDH1 | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
unknown | - | the rest of the cells | - | proportion_cnv_chr     -1     -4     -11     -16     -17     -18

### Packages

```{r packages, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(tidyr)
library(patchwork)
library(DT)
```

### Base directories

```{r base paths, eval=TRUE}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", "SCPCP000006")

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")

result_dir <- file.path(module_base, "results")
```

### Input files

In this notebook, we are working with all of the samples in SCPCP000006.

The sample metadata can be found in `sample_metadata_file` in the `data` folder.

We extracted from the pre-processed and labeled `Seurat` object (that is the output of `06_infercnv_HMM-i3_SCPCS000169_reference-both.rds` saved in the `results` directory)
the following information per cell:

- the predicted compartment in `fetal_kidney_predicted.compartment` and `fetal_kidney_predicted.compartment.score`

- the predicted compartment in `fetal_kidney_predicted.cell_type` and `fetal_kidney_predicted.cell_type.score`

- the sample identifier in `sample_id`

- the proportion of estimated CNV per chromosome, {i} in 1 to 22 `proportion_cnv_chr{i}`

- the `counts`of [markers genes](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-wilms-tumor-06#the-table-celltype_metadatacsv-contains-the-following-column-and-information)

```{r path_to_query}
sample_metadata_file <- file.path(repository_base, "data", "current","SCPCP000006", "single_cell_metadata.tsv")

metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)

# Create a data frames of all annotations
cell_type_df <- metadata$scpca_sample_id |>
  purrr::map(
    # For each sample_id, do the following:
    \(sample_id) {

      # Read in the Seurat object
     if(file.exists(file.path(result_dir, sample_id, paste0("06_infercnv_HMM-i3_", sample_id, "_reference-both.rds")))){

       srat <- readRDS(
        file.path(result_dir, sample_id, paste0("06_infercnv_HMM-i3_", sample_id, "_reference-both.rds"))
      )

      # Create and return a data frame from the Seurat object with relevant annotations
      # this data frame will have four columns: barcode, sample_id, compartment, organ
      data.frame(
        # label transfer from the fetal kidney reference
        cell_type = srat$fetal_kidney_predicted.cell_type,
        compartment = srat$fetal_kidney_predicted.compartment,

        # predicted.scores from the label transfer from the fetal kidney reference
        cell_type.score = srat$fetal_kidney_predicted.cell_type.score,
        compartment.score = srat$fetal_kidney_predicted.compartment.score,

        # cell embedding
        umap = srat@reductions$umap@cell.embeddings,

        # marker genes
        PTPRC = FetchData(object = srat, vars = "ENSG00000081237", layer = "counts"),
        VWF = FetchData(object = srat, vars = "ENSG00000110799", layer = "counts"),
        VIM = FetchData(object = srat, vars = "ENSG00000026025", layer = "counts"),
        CITED1 = FetchData(object = srat, vars = "ENSG00000125931", layer = "counts"),
        CDH1 = FetchData(object = srat, vars = "ENSG00000039068", layer = "counts"),
        PODXL = FetchData(object = srat, vars = "ENSG00000128567", layer = "counts"),
        COL6A3 = FetchData(object = srat, vars = "ENSG00000163359", layer = "counts"),
        SIX2 = FetchData(object = srat, vars = "ENSG00000170577", layer = "counts"),
        NCAM1 = FetchData(object = srat, vars = "ENSG00000149294", layer = "counts"),
        THY1 = FetchData(object = srat, vars = "ENSG00000154096", layer = "counts"),

        # proportion of cnv per chromosome
        proportion_cnv_chr1 = srat$proportion_cnv_chr1,
        proportion_cnv_chr2 = srat$proportion_cnv_chr2,
        proportion_cnv_chr3 = srat$proportion_cnv_chr3,
        proportion_cnv_chr4 = srat$proportion_cnv_chr4,
        proportion_cnv_chr5 = srat$proportion_cnv_chr5,
        proportion_cnv_chr6 = srat$proportion_cnv_chr6,
        proportion_cnv_chr7 = srat$proportion_cnv_chr7,
        proportion_cnv_chr8 = srat$proportion_cnv_chr8,
        proportion_cnv_chr9 = srat$proportion_cnv_chr9,
        proportion_cnv_chr10 = srat$proportion_cnv_chr10,
        proportion_cnv_chr11 = srat$proportion_cnv_chr11,
        proportion_cnv_chr12 = srat$proportion_cnv_chr12,
        proportion_cnv_chr13 = srat$proportion_cnv_chr13,
        proportion_cnv_chr14 = srat$proportion_cnv_chr14,
        proportion_cnv_chr15 = srat$proportion_cnv_chr15,
        proportion_cnv_chr16 = srat$proportion_cnv_chr16,
        proportion_cnv_chr17 = srat$proportion_cnv_chr17,
        proportion_cnv_chr18 = srat$proportion_cnv_chr18,
        proportion_cnv_chr19 = srat$proportion_cnv_chr19,
        proportion_cnv_chr20 = srat$proportion_cnv_chr20,
        proportion_cnv_chr21 = srat$proportion_cnv_chr21,
        proportion_cnv_chr22 = srat$proportion_cnv_chr22,

        # cnv global estimation per chromosome
        has_cnv_chr1 = srat$has_cnv_chr1,
        has_cnv_chr2 = srat$has_cnv_chr2,
        has_cnv_chr3 = srat$has_cnv_chr3,
        has_cnv_chr4 = srat$has_cnv_chr4,
        has_cnv_chr5 = srat$has_cnv_chr5,
        has_cnv_chr6 = srat$has_cnv_chr6,
        has_cnv_chr7 = srat$has_cnv_chr7,
        has_cnv_chr8 = srat$has_cnv_chr8,
        has_cnv_chr9 = srat$has_cnv_chr9,
        has_cnv_chr10 = srat$has_cnv_chr10,
        has_cnv_chr11 = srat$has_cnv_chr11,
        has_cnv_chr12 = srat$has_cnv_chr12,
        has_cnv_chr13 = srat$has_cnv_chr13,
        has_cnv_chr14 = srat$has_cnv_chr14,
        has_cnv_chr15 = srat$has_cnv_chr15,
        has_cnv_chr16 = srat$has_cnv_chr16,
        has_cnv_chr17 = srat$has_cnv_chr17,
        has_cnv_chr18 = srat$has_cnv_chr18,
        has_cnv_chr19 = srat$has_cnv_chr19,
        has_cnv_chr20 = srat$has_cnv_chr20,
        has_cnv_chr21 = srat$has_cnv_chr21,
        has_cnv_chr22 = srat$has_cnv_chr22
        ) |>
        tibble::rownames_to_column("barcode") |>
        dplyr::mutate(sample_id = sample_id)
     }
    }
  ) |>
  # now combine all dataframes to make one big one
  dplyr::bind_rows()

```

### Output file

The report will be saved in the `notebook` directory.

## Functions

#### do_Feature_mean

`do_Feature_mean` shows heatmap of mean expression of a feature grouped by a metadata.

- `df` is the name of the table containing metadata and feature expression (counts) per cells
- `group.by` is the name of the metadata to group the cells
- `feature` is the name of the gene to average the expression

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_mean <- function(df, group.by, feature ){

  df <- df %>%
  group_by(sample_id, !!sym(group.by))%>% 
  summarise(m = mean(!!sym(feature)))

p <- ggplot(df, aes(x= sample_id, y = !!sym(group.by), fill = m)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20)) +
   theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5)) +
   guides(fill=guide_colourbar(title=paste0(feature)))

return(p)
}
```

#### do_Feature_boxplot

`do_Feature_boxplot` shows boxplot of expression of a feature grouped by a metadata.

- `df` is the name of the table containing metadata and feature expression (counts) per cells
- `group.by` is the name of the metadata to group the cells
- `feature` is the name of the gene to average the expression

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_boxplot <- function(df, group.by, feature, split.by ){

  df <- df %>%
    mutate(!!sym(group.by), group  = factor(!!sym(group.by), levels = c("fetal_nephron", "stroma", "immune", "endothelium")))

  p <- ggplot(df, 
    aes(x = group, y = !!sym(feature), fill = group)
  ) + 
  geom_boxplot(size = 0.5, size.outlier = 0.25) +
  facet_wrap(vars(!!sym(split.by)), scale = "free_y", ncol = 4) + 
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 12, angle = 0, hjust = 0.5, vjust = 0), 
    legend.position = "none"
  )

return(p)
}
```

## Analysis

### Global cnv score

As done in `06_cnv_infercnv_exploration.Rmd`, we calculate  single CNV score and assess its potential in identifying cells with CNV versus normal cells without CNV.

We simply checked for each chromosome if the cell `has_cnv_chr`. 
Would the cell have more than `cnv_threshold` chromosome with CNV, the global `has_cnv_score` will be TRUE.
Else, the cell will have a `has_cnv_score` set to FALSE.

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}

#### CNV score
  cell_type_df$has_cnv_score <- rowSums(cell_type_df[,grepl("has_cnv_chr", colnames(cell_type_df))])
  cell_type_df$has_cnv_score[cell_type_df$has_cnv_score > params$cnv_threshold] <- "CNV"

  # we know that some immune cells might have false positive CNV 
  cell_type_df$has_cnv_score[cell_type_df$has_cnv_score %in% c(0:params$cnv_threshold)] <- "no CNV"

table(cell_type_df$has_cnv_score)
```

### First level annotation

At first, we like to indicate in the `first.level_annotation` if a cell is normal, cancer or unknown.

- _normal_ cells can be observe in all four compartments (`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have cnv. 
We only allow a bit of flexibility in terms of cnv profile for immune and endothelium cells that have a high predicted score.
Indeed, we know that false positive cnv can be observed in a cell type specific manner.

The threshold used for the `predicted.score` is defined as a parameter of this notebook as `r params$predicted.celltype.threshold`.
The threshold used for the identification of cnv is also defined in the params of the notebook as `r params$cnv_threshold`.

- _cancer_ cells are either from the `stroma` or `fetal nephron` compartments and must have at least few cnv.

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}

cell_type_df$first.level_annotation <- "unknown"

# Define normal cells
# We first pick up the immune and endothelial cells annotated via the label transfer compartments under the condition that the predicted score is above the threshold
cell_type_df$first.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron", "stroma") &
                                      cell_type_df$has_cnv_score == "no CNV"]                          <- "normal"

cell_type_df$first.level_annotation[  cell_type_df$compartment %in% c("endothelium", "immune") &
                                      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
                                      cell_type_df$has_cnv_score == "no CNV") ]                        <- "normal"

# Define cancer cells
cell_type_df$first.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron", "stroma") &
                                      cell_type_df$has_cnv_score == "CNV" ]                            <- "cancer"

```

Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` _cancer_ cells, `r table(cell_type_df$first.level_annotation)["normal"]` _normal_ cells.
Only `r table(cell_type_df$first.level_annotation)["unknown"]`cells remain _unknown_

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes( x = umap.umap_1, y = umap.umap_2, color = first.level_annotation), shape = 19, size = 1)+
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw()  +
  theme(text = element_text(size=22))

```

### Second level annotation

#### Normal cells

- Normal cells from the `fetal nephron` compartment must be normal kidney cells.

- Normal cells from the `stroma` compartment must be normal stroma cells.

- Immune and endothelial cells have been already identified by label transfer.

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron") &
                                      cell_type_df$has_cnv_score == "no CNV"]                          <- "kidney"

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("stroma") &
                                      cell_type_df$has_cnv_score == "no CNV"]                          <- "normal stroma"

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("endothelium") &
                                      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
                                      cell_type_df$has_cnv_score == "no CNV") ]                        <- "endothelium"

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("immune") &
                                      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
                                      cell_type_df$has_cnv_score == "no CNV") ]                        <- "immune"

```

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df[cell_type_df$first.level_annotation == "normal",], aes( x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1)+
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw()  +
  theme(text = element_text(size=22))
```

#### Cancer cells

Wilms tumor cancer cells can be:

- _cancer stroma_: We define as _cancer stroma_ all cancer cells from the stroma compartment.

- _blastema_,: we defined as _bastema_ every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`. 
We know that these _mesenchymal_ cells are cells from the cap mesenchyme that are not expected to be in a mature kidney. 
These blastema cells should express higher _CITED1_.

- _cancer epithelium_: we defined as _cancer epithelium_ all cancer cells that are neither stroma nor blastemal cells. 
We expect these cells to express epithelial markers. 
Their predicted cell type should correspond to more mature kidney epithelial subunits.

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("stroma") &
                                      cell_type_df$has_cnv_score == "CNV"]                          <- "cancer stroma"

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron") &
                                      cell_type_df$has_cnv_score == "CNV" &
                                        cell_type_df$cell_type == "mesenchymal cell"]                          <- "blastema"

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron") &
                                      cell_type_df$has_cnv_score == "CNV" &
                                        cell_type_df$cell_type != "mesenchymal cell"]                          <- "cancer epithelium"

```

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer",], aes( x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 0.1)+
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size=22))
```

#### Cancer and normal cells

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes( x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1)+
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw()  +
  theme(text = element_text(size=22))
```

### Validation cancer versus normal based on the cnv profile

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
  for(i in 1:22){
    print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("proportion_cnv_chr", i)))
  }

```

### Validation of second level annotation using marker genes

#### Immune, _PTPRC_ expression

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000081237")

```

#### Endothelium, _VWF_ expression

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000110799")

```

#### Stroma, _Vimentin_ expression

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000026025")

```

#### Stroma, _COL6A3_ expression

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000163359")

```

#### Stroma, _THY1_ expression

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000154096")

```

#### Blastema, _CITED1_ expression

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000125931")

```

#### Blastema, _NCAM1_ expression

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000149294")

```

#### stemness marker (blastema and primitive epithelium), _SIX2_ expression

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000170577")

```

#### Epithelium, _CDH1_ expression

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000039068")

```

#### Epithelium, _PODXL_ expression

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000128567")

```

## Conclusion

- Combining label transfer and CNV inference we annotated 37/40 Wilms tumor samples in SCPCP000006.

- For the three remaining samples, we need to check why we do not have the output of `06_infercnv.R --HMM i3 --reference both`. 
It might be that we don't have enough normal cell for the reference.

- The heatmaps of cnv proportion and marker genes support our annotations, but signals with some marker genes are very low.
Also, there is no universal marker for each entity of Wilms tumor that cover all tumor cells from all patient. 
This makes the validation of the annotations quite difficult.

- However, we could try to take the problem from the other side, and used the current annotation to perform differential expression analysis and try to find marker genes that are consistent across patient and Wilms tumor histologies.

- In each histology (i.e. epithelial and stroma), the distinction between cancer and non cancer cell is difficult (as expected). 
In this analysis, we suggested to rely on the cnv score to assess the normality of the cell.
Here again, we could try to run differential expression analysis and compare epithelial (resp. stroma) cancer versus non-cancer cells across patient, aiming to find a share transcriptional program allowing the classification cancer versus normal.

- In our annotation, we haven't taken into account the favorable/anaplastic status of the sample. 
However, as anaplasia can occur in every (but do not has to) wilms tumor histology, I am not sure how to integrate the information into the annotation.

- This notebook could be finally rendered using different parameters, i.e. threshold for the cnv score and predicted score to use.

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-10-30 analyses/cell-type-wilms-tumor-06/notebook/07_annotation_Across_Samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/07_annotation_Across_Samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
-- the label transfer from the fetal kidney reference (Stewart et al.), in particular the `fetal_kidney_predicted.compartment` and `fetal_kidney:predicted.cell_type`, as well as the `mapping.score` for each compartment,
+- the label transfer from the fetal kidney reference (Stewart et al.), in particular the `fetal_kidney_predicted.compartment` and `fetal_kidney:predicted.cell_type`, as well as the `prediction.score` for each compartment,
 

### 2024-10-30 analyses/cell-type-wilms-tumor-06/notebook/07_annotation_Across_Samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/07_annotation_Across_Samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 library(tidyverse)
-library(tidyr)
 library(patchwork)
@@
 
-We extracted from the pre-processed and labeled `Seurat` object (that is the output of `06_infercnv_HMM-i3_SCPCS000169_reference-both.rds` saved in the `results` directory)
+We extracted from the pre-processed and labeled `Seurat` object from `results/{sample_id}/06_infercnv_HMM-i3_SCPCS000169_reference-both.rds`.
 the following information per cell:

### 2024-10-30 analyses/cell-type-wilms-tumor-06/notebook/07_annotation_Across_Samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/07_annotation_Across_Samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 # We first pick up the immune and endothelial cells annotated via the label transfer compartments under the condition that the predicted score is above the threshold
-cell_type_df$first.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron", "stroma") &
-                                      cell_type_df$has_cnv_score == "no CNV"]                          <- "normal"
-
-cell_type_df$first.level_annotation[  cell_type_df$compartment %in% c("endothelium", "immune") &
-                                      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
-                                      cell_type_df$has_cnv_score == "no CNV") ]                        <- "normal"
-
-# Define cancer cells
-cell_type_df$first.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron", "stroma") &
-                                      cell_type_df$has_cnv_score == "CNV" ]                            <- "cancer"
+cell_type_df <- cell_type_df |>
+  mutate(first.level_annotation = case_when(
+    # assign normal/cancer based on condition
+    compartment %in% c("fetal_nephron", "stroma") && has_cnv_score == "no CNV" ~ "normal",
+    compartment %in% c("endothelium", "immune") && has_cnv_score == "no CNV" ~ "normal",
+    compartment %in% c("fetal_nephron", "stroma") && has_cnv_score == "CNV" ~ "cancer",
+    # all other cells are unknown
+    .default ~ "unknown"
+  ))
 

### 2024-10-30 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
changes to PR844

--- 2024-10-30 analyses/cell-type-wilms-tumor-06/notebook/07_annotation_Across_Samples_exploration.Rmd
@@
 ---
-title: "Annotation exploration for SCPCP000006"
+title: "Combined annotation exploration for SCPCP000006"
 author: "Maud PLASCHKA"
@@
 
-_Where `cnv.thr` and `map.thr` need to be discussed_
+_Where `cnv.thr` and `pred.thr` need to be discussed_
 
@@
 -- | -- | -- | -- | --
-normal | endothelial | compartment == "endothelium"     & mapping_score > map.thr     & cnv_score < cnv.thr | WVF | no   cnv
-normal | immune | compartment == "immune"     & mapping_score > map.thr     & cnv_score < cnv.thr | PTPRC, CD163, CD68 | no cnv
-normal | kidney | cell_type %in% c("kidney cell",   "kidney epithelial", "podocyte")     & mapping_score > map.thr     & cnv_score < cnv.thr | CDH1, PODXL, LTL | no cnv
-normal | stroma | compartment == "stroma"     & mapping_score > map.thr     & cnv_score < cnv.thr | VIM | no cnv
+normal | endothelial | compartment == "endothelium"     & predicted.score > pred.thr     & cnv_score < cnv.thr | WVF | no   cnv
+normal | immune | compartment == "immune"     & predicted.score > pred.thr     & cnv_score < cnv.thr | PTPRC, CD163, CD68 | no cnv
+normal | kidney | cell_type %in% c("kidney cell",   "kidney epithelial", "podocyte")     & predicted.score > pred.thr     & cnv_score < cnv.thr | CDH1, PODXL, LTL | no cnv
+normal | stroma | compartment == "stroma"     & predicted.score > pred.thr     & cnv_score < cnv.thr | VIM | no cnv
 cancer | stroma | compartment == "stroma"     & cnv_score > cnv.thr | VIM | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
@@
 
-#### do_Feature_boxplot
-
-`do_Feature_boxplot` shows boxplot of expression of a feature grouped by a metadata.
-
-- `df` is the name of the table containing metadata and feature expression (counts) per cells
-- `group.by` is the name of the metadata to group the cells
-- `feature` is the name of the gene to average the expression
-
-```{r fig.width=10, fig.height=4, out.width='100%'}
-do_Feature_boxplot <- function(df, group.by, feature, split.by ){
-
-  df <- df %>%
-    mutate(!!sym(group.by), group  = factor(!!sym(group.by), levels = c("fetal_nephron", "stroma", "immune", "endothelium")))
-
-  p <- ggplot(df, 
-    aes(x = group, y = !!sym(feature), fill = group)
-  ) + 
-  geom_boxplot(size = 0.5, size.outlier = 0.25) +
-  facet_wrap(vars(!!sym(split.by)), scale = "free_y", ncol = 4) + 
-  theme_bw() +
-  theme(
-    axis.text.x = element_text(size = 12, angle = 0, hjust = 0.5, vjust = 0), 
-    legend.position = "none"
-  )
-
-return(p)
-}
-```
-
 ## Analysis
@@
 
-#### CNV score
-  cell_type_df$has_cnv_score <- rowSums(cell_type_df[,grepl("has_cnv_chr", colnames(cell_type_df))])
-  cell_type_df$has_cnv_score[cell_type_df$has_cnv_score > params$cnv_threshold] <- "CNV"
-
-  # we know that some immune cells might have false positive CNV 
-  cell_type_df$has_cnv_score[cell_type_df$has_cnv_score %in% c(0:params$cnv_threshold)] <- "no CNV"
+cell_type_df <- cell_type_df |>
+  mutate(has_cnv_score = rowSums(cell_type_df[,grepl("has_cnv_chr", colnames(cell_type_df))])) |>
+  mutate(has_cnv_score = case_when(
+    has_cnv_score > params$cnv_threshold ~ "CNV",
+    has_cnv_score <= params$cnv_threshold ~ "no CNV"
+  ))
 
@@
 
-cell_type_df$first.level_annotation <- "unknown"
-
 # Define normal cells
@@
     # assign normal/cancer based on condition
-    compartment %in% c("fetal_nephron", "stroma") && has_cnv_score == "no CNV" ~ "normal",
-    compartment %in% c("endothelium", "immune") && has_cnv_score == "no CNV" ~ "normal",
-    compartment %in% c("fetal_nephron", "stroma") && has_cnv_score == "CNV" ~ "cancer",
-    # all other cells are unknown
-    .default ~ "unknown"
+    compartment %in% c("fetal_nephron", "stroma") & has_cnv_score == "no CNV" ~ "normal",
+    compartment %in% c("endothelium", "immune") & cell_type.score > params$predicted.celltype.threshold ~ "normal",
+    compartment %in% c("fetal_nephron", "stroma") & has_cnv_score == "CNV" ~ "cancer",
+    .default = "unknown"
+
   ))
@@
 
-```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
-
-cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron") &
-                                      cell_type_df$has_cnv_score == "no CNV"]                          <- "kidney"
-
-cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("stroma") &
-                                      cell_type_df$has_cnv_score == "no CNV"]                          <- "normal stroma"
-
-cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("endothelium") &
-                                      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
-                                      cell_type_df$has_cnv_score == "no CNV") ]                        <- "endothelium"
-
-cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("immune") &
-                                      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
-                                      cell_type_df$has_cnv_score == "no CNV") ]                        <- "immune"
-
-```
-
-```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
-ggplot(cell_type_df[cell_type_df$first.level_annotation == "normal",], aes( x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1)+
-  geom_point() +
-  facet_wrap(facets = ~sample_id, ncol = 5) +
-  theme_bw()  +
-  theme(text = element_text(size=22))
-```
-
 #### Cancer cells
@@
 ```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
+cell_type_df <- cell_type_df |>
+  mutate(second.level_annotation = case_when(
+    # assign normal cells based on condition
+   cell_type_df$compartment %in% c("fetal_nephron") &
+                                      cell_type_df$has_cnv_score == "no CNV"                          ~ "kidney", 
+   cell_type_df$compartment %in% c("stroma") &
+                                      cell_type_df$has_cnv_score == "no CNV"                          ~ "normal stroma",
+   cell_type_df$compartment %in% c("endothelium") &
+                                      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
+                                      cell_type_df$has_cnv_score == "no CNV")                         ~ "endothelium",
+   cell_type_df$compartment %in% c("immune") &
+                                      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
+                                      cell_type_df$has_cnv_score == "no CNV")                         ~ "immune",
 
-cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("stroma") &
-                                      cell_type_df$has_cnv_score == "CNV"]                          <- "cancer stroma"
+   # assign cancer cells based on condition
+   cell_type_df$compartment %in% c("stroma") &
+          cell_type_df$has_cnv_score == "CNV"                         ~ "cancer stroma",
 
-cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron") &
+          cell_type_df$compartment %in% c("fetal_nephron") &
           cell_type_df$has_cnv_score == "CNV" &
-                                        cell_type_df$cell_type == "mesenchymal cell"]                          <- "blastema"
+          cell_type_df$cell_type == "mesenchymal cell"                ~ "blastema",
 
-cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron") &
+          cell_type_df$compartment %in% c("fetal_nephron") &
           cell_type_df$has_cnv_score == "CNV" &
-                                        cell_type_df$cell_type != "mesenchymal cell"]                          <- "cancer epithelium"
+          cell_type_df$cell_type != "mesenchymal cell"                ~ "cancer epithelium",
+    .default = "unknown"
 
+  ))
+
+```
+
+```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
+ggplot(cell_type_df[cell_type_df$first.level_annotation == "normal",], aes( x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1)+
+  geom_point() +
+  facet_wrap(facets = ~sample_id, ncol = 5) +
+  theme_bw()  +
+  theme(text = element_text(size=22))
 ```

### 2024-10-31 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd

@@
   p <- list()
+
   #### Plot UMAP of known Wilms tumor related alteration
   p[["compartment"]] <- visualize_metadata(infercnv_obj, "fetal_kidney_predicted.compartment", "seurat_clusters")
+
   #### Feature plot and repartition of the CNV per chromosome
@@
   }
+
   #### DotPlot
@@
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
   #### CNV score

### 2024-10-31 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
update code to create a final table of annotations

@@
 ```{r setup, include=FALSE}
-
-knitr::opts_chunk$set(echo = TRUE,
-                      message=FALSE,
-                      warnings=FALSE,
-                      knitr.digits.signif = FALSE)
-
+knitr::opts_chunk$set(
+  echo = TRUE,
+  message = FALSE,
+  warnings = FALSE,
+  knitr.digits.signif = FALSE
+)
 ```
@@
 
+### Output files
+
+```{r}
+# cell type annotation table
+annotations_tsv <- file.path(result_dir, "SCPCP000006-annotations.tsv")
+```
+
 ### Input files
@@
 ```{r path_to_query}
-sample_metadata_file <- file.path(repository_base, "data", "current","SCPCP000006", "single_cell_metadata.tsv")
+sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
 
@@
 # Create a data frames of all annotations
-cell_type_df <- metadata$scpca_sample_id |>
+cell_type_df <- unique(metadata$scpca_sample_id) |>
   purrr::map(
@@
     \(sample_id) {
+      input_file <- file.path(
+        result_dir,
+        sample_id,
+        glue::glue("06_infercnv_HMM-i3_{sample_id}_reference-both.rds")
+      )
 
       # Read in the Seurat object
-     if(file.exists(file.path(result_dir, sample_id, paste0("06_infercnv_HMM-i3_", sample_id, "_reference-both.rds")))){
-
-       srat <- readRDS(
-        file.path(result_dir, sample_id, paste0("06_infercnv_HMM-i3_", sample_id, "_reference-both.rds"))
-      )
+      if (file.exists(input_file)) {
+        srat <- readRDS(input_file)
 
@@
   dplyr::bind_rows()
-
 ```
@@
 ```{r fig.width=10, fig.height=4, out.width='100%'}
-do_Feature_mean <- function(df, group.by, feature ){
-
+do_Feature_mean <- function(df, group.by, feature) {
   df <- df %>%
-  group_by(sample_id, !!sym(group.by))%>% 
+    group_by(sample_id, !!sym(group.by)) %>%
     summarise(m = mean(!!sym(feature)))
 
-p <- ggplot(df, aes(x= sample_id, y = !!sym(group.by), fill = m)) +
+  p <- ggplot(df, aes(x = sample_id, y = !!sym(group.by), fill = m)) +
     geom_tile() +
@@
     theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5)) +
-   guides(fill=guide_colourbar(title=paste0(feature)))
+    guides(fill = guide_colourbar(title = paste0(feature)))
 
@@
 ```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
-
 cell_type_df <- cell_type_df |>
-  mutate(has_cnv_score = rowSums(cell_type_df[,grepl("has_cnv_chr", colnames(cell_type_df))])) |>
+  mutate(has_cnv_score = rowSums(cell_type_df[, grepl("has_cnv_chr", colnames(cell_type_df))])) |>
   mutate(has_cnv_score = case_when(
@@
 ```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
-
 # Define normal cells
@@
     .default = "unknown"
-
   ))
-
 ```
@@
 ```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
-ggplot(cell_type_df, aes( x = umap.umap_1, y = umap.umap_2, color = first.level_annotation), shape = 19, size = 1)+
+ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = first.level_annotation), shape = 19, size = 1) +
   geom_point() +
@@
   theme_bw() +
-  theme(text = element_text(size=22))
-
+  theme(text = element_text(size = 22))
 ```
@@
       cell_type_df$has_cnv_score == "CNV" ~ "cancer stroma",
-
     cell_type_df$compartment %in% c("fetal_nephron") &
@@
       cell_type_df$cell_type == "mesenchymal cell" ~ "blastema",
-
     cell_type_df$compartment %in% c("fetal_nephron") &
@@
 ```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
-ggplot(cell_type_df[cell_type_df$first.level_annotation == "normal",], aes( x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1)+
+ggplot(cell_type_df[cell_type_df$first.level_annotation == "normal", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1) +
   geom_point() +
@@
   theme_bw() +
-  theme(text = element_text(size=22))
+  theme(text = element_text(size = 22))
 ```
@@
 ```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
-ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer",], aes( x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 0.1)+
+ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 0.1) +
   geom_point() +
@@
 ```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
-ggplot(cell_type_df, aes( x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1)+
+ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1) +
   geom_point() +
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-  for(i in 1:22){
+for (i in 1:22) {
   print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("proportion_cnv_chr", i)))
 }
-
 ```
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-
 do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000081237")
-
 ```
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-
 do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000110799")
-
 ```
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-
 do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000026025")
-
 ```
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-
 do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000163359")
-
 ```
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-
 do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000154096")
-
 ```
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-
 do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000125931")
-
 ```
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-
 do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000149294")
-
 ```
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-
 do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000170577")
-
 ```
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-
 do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000039068")
-
 ```
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
+do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000128567")
+```
 
-    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000128567")
+## Create annotation table for export
 
+This section creates the cell type annotation table for export.
+
+```{r}
+annotations_table <- cell_type_df |>
+  select(
+    cell_barcode = barcode,
+    scpca_sample_id = sample_id,
+    tumor_cell_classification = first.level_annotation,
+    cell_type_assignment = second.level_annotation
+  ) |>
+  mutate(
+    # change cancer --> tumor, but keep the other labels
+    tumor_cell_classification = ifelse(
+      tumor_cell_classification == "cancer", "tumor", tumor_cell_classification
+    ),
+    cell_type_assignment = str_replace_all(
+      cell_type_assignment,
+      "cancer ",
+      "tumor "
+    )
+  )
+
+write_tsv(annotations_table, annotations_tsv)
+```
+
+Confirm how many samples we have annotations for:
+```{r}
+length(unique(annotations_table$scpca_sample_id))
 ```
@@
 
-- Combining label transfer and CNV inference we annotated 37/40 Wilms tumor samples in SCPCP000006.
-
-- For the three remaining samples, we need to check why we do not have the output of `06_infercnv.R --HMM i3 --reference both`. 
-It might be that we don't have enough normal cell for the reference.
+- Combining label transfer and CNV inference we annotated 38/40 Wilms tumor samples in SCPCP000006. For the two remaining samples, we need to investigate inferCNV further:
+    - SCPCS000190 only has 1 normal cell, so we cannot specify a reference for inferCNV
+    - SCPCS000203 does not run properly through inferCNV; cells are removed from the inferCNV inference, and why requires further investigation
 

### 2024-11-01 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
render final notebook

@@
 - Combining label transfer and CNV inference we annotated 38/40 Wilms tumor samples in SCPCP000006. For the two remaining samples, we need to investigate inferCNV further:
-    - SCPCS000190 only has 1 normal cell, so we cannot specify a reference for inferCNV
+    - SCPCS000190 only has 1 normal cell, which is insuffient as a reference for inferCNV
     - SCPCS000203 does not run properly through inferCNV; cells are removed from the inferCNV inference, and why requires further investigation

### 2024-11-01 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Include infercnv results for SCPCS000190 by using none as a reference, so we can get some draft annotations for it

--- 2024-10-31 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
@@
     \(sample_id) {
+      # This sample was run with "none" for reference
+      if (sample_id == "SCPCS000190") {
+        reference <- "none"
+      } else {
+        reference <- "both"
+      }
       input_file <- file.path(
@@
         sample_id,
-        glue::glue("06_infercnv_HMM-i3_{sample_id}_reference-both.rds")
+        glue::glue("06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds")
       )
@@
 
-- Combining label transfer and CNV inference we annotated 38/40 Wilms tumor samples in SCPCP000006. For the two remaining samples, we need to investigate inferCNV further:
-    - SCPCS000190 only has 1 normal cell, so we cannot specify a reference for inferCNV
-    - SCPCS000203 does not run properly through inferCNV; cells are removed from the inferCNV inference, and why requires further investigation
+- Combining label transfer and CNV inference we annotated 39/40 Wilms tumor samples in SCPCP000006.
+For the remaining samples (SCPCS000203), we need to investigate inferCNV further, since for some reason inferCNV removes cells from the inference and errors.
 

### 2024-11-01 analyses/cell-type-wilms-tumor-06/00_run_workflow.R

@@ -120,15 +120,22 @@ if (!running_ci) {
 
   # Run infercnv for all samples with HMM i3 and using "both" as the reference
   for (sample_id in sample_ids) {
-    # We currently skip samples SCPCS000190 and SCPSC000203:
-    # - SCPCS000190: insufficient cells to use as a reference
-    # - SCPSC000203: inferCNV is returning an error that requires further debugging:
+    # We run SCPCS000190 with "none" instead of "both" since it does not have
+    # enough "normal" cells to use as a reference
+    if (sample_id == "SCPCS000190") {
+      reference <- "none"
+    } else {
+      reference <- "both"
+    }
+
+    # We currently skip SCPSC000203 since infer returns an error that requires further debugging:
     #   Cell names in Seurat object and infercnv results do not match
-    if (sample_id %in% c("SCPCS000190", "SCPCS000203")) {
+    if (sample_id == "SCPCS000203") {
       next
     }
+
     # don't repeat inference on selection of samples it's already been run on
-    output_file <- file.path(module_base, "results", sample_id, glue::glue("06_infercnv_HMM-i3_{sample_id}_reference-both.rds"))
+    output_file <- file.path(module_base, "results", sample_id, glue::glue("06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds"))
     if (!file.exists(output_file)) {
       system(command = glue::glue("Rscript {file.path(module_base, 'scripts', '06_infercnv.R')} --sample_id {sample_id} --reference {reference} --HMM i3"))
     }

### 2024-11-01 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Found infercnv argument to get infercnv running for the sample with the missing annotations, and updated code accordingly to run this sample

@@
       # Read in the Seurat object
-      if (file.exists(input_file)) {
       srat <- readRDS(input_file)
@@
         dplyr::mutate(sample_id = sample_id)
-      }
     }
@@
 
-- Combining label transfer and CNV inference we annotated 39/40 Wilms tumor samples in SCPCP000006.
-For the remaining samples (SCPCS000203), we need to investigate inferCNV further, since for some reason inferCNV removes cells from the inference and errors.
+- Combining label transfer and CNV inference we have produced draft annotations for all 40 Wilms tumor samples in SCPCP000006
 

### 2024-11-01 analyses/cell-type-wilms-tumor-06/00_run_workflow.R

@@ -128,12 +128,6 @@ if (!running_ci) {
       reference <- "both"
     }
 
-    # We currently skip SCPSC000203 since infer returns an error that requires further debugging:
-    #   Cell names in Seurat object and infercnv results do not match
-    if (sample_id == "SCPCS000203") {
-      next
-    }
-
     # don't repeat inference on selection of samples it's already been run on
     output_file <- file.path(module_base, "results", sample_id, glue::glue("06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds"))
     if (!file.exists(output_file)) {

### 2024-11-01 analyses/cell-type-wilms-tumor-06/scripts/06_infercnv.R

@@ -164,7 +164,9 @@ infercnv_obj <- infercnv::CreateInfercnvObject(
   raw_counts_matrix = as.matrix(counts),
   annotations_file = annot_df,
   ref_group_names = normal_cells,
-  gene_order_file = gene_order_file
+  gene_order_file = gene_order_file,
+  # ensure all cells are included
+  min_max_counts_per_cell = c(-Inf, +Inf)
 )
 
 infercnv_obj <- infercnv::run(

### 2024-11-01 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
Specify RNA assay in the actual label transfer notebooks

@@
 srat <- readRDS(file.path(data_dir, paste0("01-Seurat_", params$sample_id, ".Rds")))
+srat_assay <- "RNA"
 
@@
 #  removes features that are not present in the reference
-DefaultAssay(srat) <- "RNA"
-query <- prepare_query(srat, rownames(reference), params$homologs_file)
+DefaultAssay(srat) <- srat_assay
+query <- prepare_query(
+  srat,
+  rownames(reference),
+  srat_assay,
+  params$homologs_file
+)
 ```
@@
   refdata,
+  query.assay = srat_assay,
   k.weight = k.weight

### 2024-11-01 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
Specify RNA assay in the actual label transfer notebooks

@@
 srat <- readRDS(file.path(data_dir, paste0("02a-fetal_full_label-transfer_", params$sample_id, ".Rds")))
+srat_assay <- "RNA"
 
 # prepare the query for label transfer
-DefaultAssay(srat) <- "RNA"
-srat <- prepare_query(srat, rownames(reference), params$homologs_file)
+# we don't want to overwrite the srat object since `prepare_query`
+#  removes features that are not present in the reference
+DefaultAssay(srat) <- srat_assay
+query <- prepare_query(
+  srat,
+  rownames(reference),
+  srat_assay,
+  params$homologs_file
+)
 ```

### 2024-11-01 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
why was s still there? and fix a typo

@@
 )
-metadata_to_trasfer <- query_labeled@meta.data[, annotation_columns]
+metadata_to_transfer <- query_labeled@meta.data[, annotation_columns]
 
-srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_full_", annotation_columns))
+srat <- AddMetaData(srat, metadata_to_transfer, col.name = paste0("fetal_full_", annotation_columns))
 

### 2024-11-01 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
why was s still there? and fix a typo

@@
 )
-metadata_to_trasfer <- s@meta.data[, annotation_columns]
+metadata_to_transfer <- query_labeled@meta.data[, annotation_columns]
 
-srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_kidney_", annotation_columns))
+srat <- AddMetaData(srat, metadata_to_transfer, col.name = paste0("fetal_kidney_", annotation_columns))
 ```

### 2024-11-04 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd

@@
 ```{r setup, include=FALSE}
-
-knitr::opts_chunk$set(echo = TRUE,
-                      message=FALSE,
-                      warnings=FALSE
+knitr::opts_chunk$set(
+  echo = TRUE,
+  message = FALSE,
+  warnings = FALSE
 )
-
 ```
@@
 ```{r}
-
 subdiagnosis <- readr::read_tsv(
@@
   dplyr::pull(subdiagnosis)
-
 ```
@@
 ```{r}
-
 library(infercnv)
@@
 ```{r}
-
-Do_CNV_heatmap <- function(object, infercnv_obj, group.by, reference_value){
-
-  out <- SCpubr::do_CopyNumberVariantPlot(sample = object,
+Do_CNV_heatmap <- function(object, infercnv_obj, group.by, reference_value) {
+  out <- SCpubr::do_CopyNumberVariantPlot(
+    sample = object,
     infercnv_object = infercnv_obj,
@@
 ```{r}
-
-Do_CNV_score <- function(seurat_oject,infercnv_obj,reference_value){
-
-  score <- apply(infercnv_obj@expr.data,2,function(x){ sum(x < 0.95 | x > 1.05)/length(x) })
+Do_CNV_score <- function(seurat_oject, infercnv_obj, reference_value) {
+  score <- apply(infercnv_obj@expr.data, 2, function(x) {
+    sum(x < 0.95 | x > 1.05) / length(x)
+  })
 
@@
 ```{r }
-visualize_metadata <- function(object, meta, group.by){
-
-  if(is.numeric(object@meta.data[,meta])){
+visualize_metadata <- function(object, meta, group.by) {
+  if (is.numeric(object@meta.data[, meta])) {
     d <- SCpubr::do_FeaturePlot(object,
@@
       legend.length = 5,
-                                              legend.position = "right") + ggtitle(meta)
+      legend.position = "right"
+    ) + ggtitle(meta)
     b <- SCpubr::do_ViolinPlot(object,
@@
       group.by = group.by,
-                                             legend.position = "none")
+      legend.position = "none"
+    )
 
-   return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
-  }
-
-  else{
-
-  d <- SCpubr::do_DimPlot(object, reduction="umap", group.by = group.by, label = TRUE, repel = TRUE) + ggtitle(paste0(meta," - umap")) + theme(text=element_text(size=18))
-  b <- SCpubr::do_BarPlot(sample = object,
+    return(d + b + plot_layout(ncol = 2, widths = c(2, 4)))
+  } else {
+    d <- SCpubr::do_DimPlot(object, reduction = "umap", group.by = group.by, label = TRUE, repel = TRUE) + ggtitle(paste0(meta, " - umap")) + theme(text = element_text(size = 18))
+    b <- SCpubr::do_BarPlot(
+      sample = object,
       group.by = meta,
@@
       font.size = 10,
-                         legend.ncol = 3) +
-                         ggtitle("% cells")+
+      legend.ncol = 3
+    ) +
+      ggtitle("% cells") +
       xlab(print(group.by)) +
-                         theme(text=element_text(size=18))
-  return(d + b + plot_layout(ncol = 2, widths = c(2,4)))
+      theme(text = element_text(size = 18))
+    return(d + b + plot_layout(ncol = 2, widths = c(2, 4)))
   }
-
 }
@@
 ```{r }
-visualize_feature <- function(object, feature, group.by ){
-
+visualize_feature <- function(object, feature, group.by) {
   d <- SCpubr::do_FeaturePlot(object,
@@
     legend.length = 5,
-                                              legend.position = "right") + ggtitle(as.character(feature))
+    legend.position = "right"
+  ) + ggtitle(as.character(feature))
   b <- SCpubr::do_ViolinPlot(object,
@@
     legend.position = "none",
-                                             assay = "SCT") + ylab(as.character(feature))
+    assay = "SCT"
+  ) + ylab(as.character(feature))
 
-                  return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
+  return(d + b + plot_layout(ncol = 2, widths = c(2, 4)))
 }
@@
 ```{r }
-visualize_density <- function(object, features, group.by ){
-
+visualize_density <- function(object, features, group.by) {
   d <- SCpubr::do_RidgePlot(object,
@@
     legend.position = "none",
-                                             assay = "SCT") + ylab(features)
+    assay = "SCT"
+  ) + ylab(features)
 
@@
 ```{r}
-
-wrapper_explore_hmm <- function(infercnv_obj, cnv_threshold = 1){
-
+wrapper_explore_hmm <- function(infercnv_obj, cnv_threshold = 1) {
   p <- list()
@@
   #### Feature plot and repartition of the CNV per chromosome
-  for(i in 1:22){
-    tmp1 <- visualize_feature(infercnv_obj, feature=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment")
-    tmp2 <- visualize_density(infercnv_obj, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment")
-    p[[glue::glue("proportion_cnv_chr", i)]] <- tmp1 + tmp2 + plot_layout(ncol = 3, widths = c(1,2,2))
+  for (i in 1:22) {
+    tmp1 <- visualize_feature(infercnv_obj, feature = glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment")
+    tmp2 <- visualize_density(infercnv_obj, features = glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment")
+    p[[glue::glue("proportion_cnv_chr", i)]] <- tmp1 + tmp2 + plot_layout(ncol = 3, widths = c(1, 2, 2))
   }
@@
   p[["Dotplot"]] <- DotPlot(infercnv_obj, features = colnames(infercnv_obj@meta.data)[grepl("has_cnv_chr", colnames(infercnv_obj@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
 
   #### CNV score
-  meta <- infercnv_obj@meta.data
-  meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
-  meta$has_cnv_score[meta$has_cnv_score > cnv_threshold] <- "CNV"
+  meta <- infercnv_obj@meta.data |>
+    mutate(has_cnv_score = rowSums(meta[, grepl("has_cnv_chr", colnames(meta))])) |>
+    mutate(has_cnv_score = case_when(
+      has_cnv_score > cnv_threshold ~ "CNV",
+      has_cnv_score <= cnv_threshold ~ "no CNV"
+    ))
 
-  # we know that some immune cells might have false positive CNV 
-  meta$has_cnv_score[meta$has_cnv_score %in% c(0:cnv_threshold)] <- "no CNV"
   infercnv_obj <- AddMetaData(infercnv_obj, meta$has_cnv_score, col.name = "has_cnv_score")
@@
   return(p)
-
 }
-
 ```
@@
 ```{r}
-
 result_dir <- file.path(module_base, "results", params$sample_id, "06_infercnv")
-
 ```
@@
 ```{r}
-
 srat <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")))
-
 ```
@@
 infercnv_obj <- list()
-for(reference_value in c("reference-none", "reference-immune", "reference-endothelium", "reference-both")){
-  infercnv_obj[[reference_value]] <- readRDS(file.path(result_dir, glue::glue(reference_value, "_HMM-no") , glue::glue("06_infercnv_", params$sample_id, "_", reference_value, "_HMM-no.rds")))
+for (reference_value in c("reference-none", "reference-immune", "reference-endothelium", "reference-both")) {
+  infercnv_obj[[reference_value]] <- readRDS(file.path(result_dir, glue::glue(reference_value, "_HMM-no"), glue::glue("06_infercnv_", params$sample_id, "_", reference_value, "_HMM-no.rds")))
   print(Do_CNV_heatmap(object = srat, infercnv_obj = infercnv_obj[[reference_value]], group.by = "fetal_kidney_predicted.compartment", reference_value = reference_value))
 }
-
 ```
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
-
-for(reference_value in c("reference-none", "reference-immune", "reference-endothelium", "reference-both")){
-  srat <- Do_CNV_score(srat,infercnv_obj = infercnv_obj[[reference_value]], reference_value)
-  p1 <- visualize_feature(srat, feature =  glue::glue("CNV-score_", reference_value) , group.by = "fetal_kidney_predicted.compartment" )
-  p2 <- visualize_density(srat, features=glue::glue("CNV-score_", reference_value), group.by = "fetal_kidney_predicted.compartment")
-  print(p1 + p2 + plot_layout(ncol = 3, widths = c(1,2,2)))
+for (reference_value in c("reference-none", "reference-immune", "reference-endothelium", "reference-both")) {
+  srat <- Do_CNV_score(srat, infercnv_obj = infercnv_obj[[reference_value]], reference_value)
+  p1 <- visualize_feature(srat, feature = glue::glue("CNV-score_", reference_value), group.by = "fetal_kidney_predicted.compartment")
+  p2 <- visualize_density(srat, features = glue::glue("CNV-score_", reference_value), group.by = "fetal_kidney_predicted.compartment")
+  print(p1 + p2 + plot_layout(ncol = 3, widths = c(1, 2, 2)))
 }
-
 ```
@@
 ```{r}
-
 #### Load the `Seurat` object
 srat_i3 <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-both.rds")))
-
 ```
@@
 ```{r}
-
 srat_i6 <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i6_", params$sample_id, "_reference-both.rds")))
-
 ```
@@
 
-meta <- seurat@meta.data[colnames(srat),]
+meta <- seurat@meta.data[colnames(srat), ]
 
 # merge the new info into the srat metadata
-srat_pull <- AddMetaData(srat, meta[,grepl("_cnv_chr", colnames(meta))], col.name = colnames(meta[grepl("_cnv_chr", colnames(meta))]))
+srat_pull <- AddMetaData(srat, meta[, grepl("_cnv_chr", colnames(meta))], col.name = colnames(meta[grepl("_cnv_chr", colnames(meta))]))
 ```
@@
 ```{r fig.width=20, fig.height=8, out.width='100%'}
-
 p1 <- SCpubr::do_DimPlot(pull$infercnv_obj, "has_cnv_score", plot.title = "inter-patient reference, HMM-i3", reduction = "umap")
@@
 p1 + p2 + p3 + plot_layout(ncol = 3)
-
 ```

### 2024-11-04 analyses/cell-type-wilms-tumor-06/notebook_template/00b_characterize_fetal_kidney_reference_Stewart.Rmd
textual changes

@@
 The rds data can be download using the download link https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds
-The azimuth compatible reference has been downloaded and created in the `R script` `download-and-create-fetal-kidney-ref.R`
+The reference was been downloaded and created in the script `scripts/prepare-fetal-references.R`.
 
@@
 
-Load required packages in the following chunk, if needed.
-Do not install packages here; only load them with the `library()` function.
-
 ```{r packages, message=FALSE, warning=FALSE}
-library("Seurat")
+library(Seurat)
 library(Azimuth)

### 2024-11-05 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update 03 notebook for testing and regenerate 03 notebooks with current data release. notebook also styled

@@
   rate1_threshold: 0.5  # Threshold for rate (percentage) of cells expressing the marker genes in group1
+  testing: 0            # When testing, certain parameter values have to be changed, and certain chunks are skipped
 output:
@@
 
+options(future.globals.maxSize = 8912896000000)
 set.seed(params$seed)
@@
 
-DT::datatable(CellType_metadata, caption = ("CellType_metadata"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-                               buttons = c( 'csv', 'excel')))
+DT::datatable(CellType_metadata,
+  caption = ("CellType_metadata"),
+  extensions = "Buttons",
+  options = list(
+    dom = "Bfrtip",
+    buttons = c("csv", "excel")
+  )
+)
 ```
@@
       legend.length = 5,
-                                              legend.position = "right") + ggtitle(meta)
+      legend.position = "right"
+    ) + ggtitle(meta)
     b <- SCpubr::do_ViolinPlot(srat,
@@
 ```{r }
-visualize_feature <- function(object, features, group.by = "seurat_clusters"){
-
+visualize_feature <- function(object, features, group.by = "seurat_clusters") {
   feature_symbol <- AnnotationDbi::select(org.Hs.eg.db,
-                                keys=features, 
-                                columns="SYMBOL", 
-                                keytype="ENSEMBL")
+    keys = features,
+    columns = "SYMBOL",
+    keytype = "ENSEMBL"
+  )
 
@@
     legend.length = 5,
-                                              legend.position = "right") + ggtitle(feature_symbol$SYMBOL)
+    legend.position = "right"
+  ) + ggtitle(feature_symbol$SYMBOL)
   b <- SCpubr::do_ViolinPlot(srat,
@@
     legend.position = "none",
-                                             assay = "SCT") + ylab(feature_symbol$SYMBOL)
+    assay = "SCT"
+  ) + ylab(feature_symbol$SYMBOL)
 
-                  return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
+  return(d + b + plot_layout(ncol = 2, widths = c(2, 4)))
 }
-
 ```
@@
 ```{r }
-MSigDB_score <- function(object, category, gs_name, name, nbin = 24){
-
+MSigDB_score <- function(object, category, gs_name, name, nbin = 24) {
   set <- msigdbr(species = "human", category = category)
@@
     dplyr::filter(gs_name == gs_name) %>%
-                    dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()
+    dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>%
+    as.data.frame()
   set_list <- list(set_list$human_ensembl_gene)
@@
 ```{r }
-Enrichment_plot <- function(category, signatures, background){
+Enrichment_plot <- function(category, signatures, background) {
   ## define genesets
-   tryCatch( expr =  {
+  tryCatch(
+    expr = {
       gene_set <- msigdbr(species = "human", category = category)
-               msigdbr_set <-  gene_set %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
-               cclust<-compareCluster(geneCluster = signatures, 
+      msigdbr_set <- gene_set %>%
+        dplyr::distinct(gs_name, ensembl_gene) %>%
+        as.data.frame()
+      cclust <- compareCluster(
+        geneCluster = signatures,
         fun = enricher,
         TERM2GENE = msigdbr_set,
-               universe=background)
-               d <- dotplot(cclust,showCategory=15) + scale_y_discrete(labels=function(x) str_wrap(x, width=40))
+        universe = background
+      )
+      d <- dotplot(cclust, showCategory = 15) + scale_y_discrete(labels = function(x) str_wrap(x, width = 40))
       return(d)
     },
-   error = function(e) { print("No enrichment") } )
-
+    error = function(e) {
+      print("No enrichment")
     }
-
+  )
+}
 ```
@@
 ```{r fig.width=10, fig.height=4, out.width='100%'}
-do_Table_Heatmap <- function(data, first_group, last_group ){
-
+do_Table_Heatmap <- function(data, first_group, last_group) {
   df <- data@meta.data %>%
     mutate_if(sapply(data@meta.data, is.character), as.factor) %>%
-  group_by( !!sym(first_group), !!sym(last_group))%>% 
+    group_by(!!sym(first_group), !!sym(last_group)) %>%
     summarise(Nb = log(n()))
 
-p <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = Nb)) +
+  p <- ggplot(df, aes(x = !!sym(first_group), y = !!sym(last_group), fill = Nb)) +
     geom_tile() +
@@
     theme(text = element_text(size = 20)) +
-   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
-   guides(fill=guide_legend(title=" Number of \n cells (log)"))
+    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
+    guides(fill = guide_legend(title = " Number of \n cells (log)"))
 
@@
 # open the processed rds object
-srat <- readRDS(file.path(result_dir, paste0("02b-fetal_kidney_label-transfer_",params$sample_id,".Rds")))
+srat <- readRDS(file.path(result_dir, paste0("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")))
 DefaultAssay(srat) <- "SCT"
@@
 # One sample in this project fails with the default number of bins, so we need to override it
-if (params$sample_id == "SCPCS000197") {
+if (params$testing) {
+  seurat_nbins <- 5
+} else if (params$sample_id == "SCPCS000197") {
   seurat_nbins <- 23 # default is 24, so this is very close
@@
 ```{r fig.width=20, fig.height=10, out.width='100%'}
-d1 <- SCpubr::do_DimPlot(srat, reduction="pca", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - pca")
-d2 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")
-v1 <- SCpubr::do_ViolinPlot(srat, features = c( "subsets_mito_percent"), ncol = 1, group.by = "seurat_clusters", legend.position = "none") 
-v2 <- SCpubr::do_ViolinPlot(srat, features = c( "detected"), ncol = 1, group.by = "seurat_clusters", legend.position = "none") 
-v3 <- SCpubr::do_ViolinPlot(srat, features = c( "sum"), ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+d1 <- SCpubr::do_DimPlot(srat, reduction = "pca", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - pca")
+d2 <- SCpubr::do_DimPlot(srat, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")
+v1 <- SCpubr::do_ViolinPlot(srat, features = c("subsets_mito_percent"), ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+v2 <- SCpubr::do_ViolinPlot(srat, features = c("detected"), ncol = 1, group.by = "seurat_clusters", legend.position = "none")
+v3 <- SCpubr::do_ViolinPlot(srat, features = c("sum"), ncol = 1, group.by = "seurat_clusters", legend.position = "none")
 
-d1 + d2 + (v1 + v2 +v3 + plot_layout(ncol=1) ) + plot_layout(ncol = 3, widths = c(2,2,4))
+d1 + d2 + (v1 + v2 + v3 + plot_layout(ncol = 1)) + plot_layout(ncol = 3, widths = c(2, 2, 4))
 ```
@@
 ```{r fig.width=20, fig.height=4, out.width='100%'}
-for(feature in CellType_metadata$ENSEMBL_ID[CellType_metadata$ENSEMBL_ID %in% rownames(srat@assays$SCT)]){
+for (feature in CellType_metadata$ENSEMBL_ID[CellType_metadata$ENSEMBL_ID %in% rownames(srat@assays$SCT)]) {
   print(visualize_feature(srat, features = feature, group.by = "seurat_clusters"))
@@
 ```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
-srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_P53_PATHWAY", name = "TP53_score", nbin = seurat_nbins)
-visualize_metadata(srat, meta = "TP53_score1", group.by = "seurat_clusters" )
+srat <- MSigDB_score(object = srat, category = "H", gs_name = "HALLMARK_P53_PATHWAY", name = "TP53_score", nbin = seurat_nbins)
+visualize_metadata(srat, meta = "TP53_score1", group.by = "seurat_clusters")
 ```
@@
 ```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
-srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_DNA_REPAIR", name = "DNA_repair_score", nbin = seurat_nbins)
-visualize_metadata(srat, meta = "DNA_repair_score1", group.by = "seurat_clusters" )
+srat <- MSigDB_score(object = srat, category = "H", gs_name = "HALLMARK_DNA_REPAIR", name = "DNA_repair_score", nbin = seurat_nbins)
+visualize_metadata(srat, meta = "DNA_repair_score1", group.by = "seurat_clusters")
 ```
@@
 ```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
-srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DROSHA_TARGET_GENES", name = "DROSHA_score", nbin = seurat_nbins)
-visualize_metadata(srat, meta = "DROSHA_score1", group.by = "seurat_clusters" )
+srat <- MSigDB_score(object = srat, category = "C3", gs_name = "DROSHA_TARGET_GENES", name = "DROSHA_score", nbin = seurat_nbins)
+visualize_metadata(srat, meta = "DROSHA_score1", group.by = "seurat_clusters")
 ```
@@
 ```{r fig.height=4, fig.width=20, warning=FALSE, out.width='100%'}
-srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DICER1_TARGET_GENES", name = "DICER1_score", nbin = seurat_nbins)
-visualize_metadata(srat, meta = "DICER1_score1", group.by = "seurat_clusters" )
+srat <- MSigDB_score(object = srat, category = "C3", gs_name = "DICER1_TARGET_GENES", name = "DICER1_score", nbin = seurat_nbins)
+visualize_metadata(srat, meta = "DICER1_score1", group.by = "seurat_clusters")
 ```
@@
 
-```{r fig.width=10, fig.height=10, out.width='100%'}
+```{r fig.width=10, fig.height=10, out.width='100%', eval=!params$testing}
+# this chunk is not run in testing (CI) since there are generally too few clusters for the method to work
 feature_conversion <- srat@assays$RNA@meta.data
 de_results <- DElegate::FindAllMarkers2(srat, group_column = "seurat_clusters")
-#filter the most relevant markers
-s.markers <- de_results[de_results$padj < params$padj_thershold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold,]
+# filter the most relevant markers
+s.markers <- de_results[de_results$padj < params$padj_thershold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold, ]
 # add gene symbol for easiest interpretation of the result
 s.markers$gene_ids <- s.markers$feature
-s.markers <- left_join(s.markers,feature_conversion, by = c( "gene_ids") )
+s.markers <- left_join(s.markers, feature_conversion, by = c("gene_ids"))
 stopifnot(
@@
 
-```{r fig.width=10, fig.height=10, out.width='100%'}
-
-DT::datatable(s.markers, caption = ("marker genes"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-                               buttons = c( 'csv', 'excel')))
-
+```{r fig.width=10, fig.height=10, out.width='100%', eval=!params$testing}
+DT::datatable(s.markers,
+  caption = ("marker genes"),
+  extensions = "Buttons",
+  options = list(
+    dom = "Bfrtip",
+    buttons = c("csv", "excel")
+  )
+)
 ```
 
-```{r fig.width=10, fig.height=10, out.width='100%'}
-
+```{r fig.width=10, fig.height=10, out.width='100%', eval=!params$testing}
 # Select top 5 genes for heatmap plotting
@@
 ss <- ScaleData(ss, features = top5$feature)
-p1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
+p1 <- SCpubr::do_DimPlot(srat, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
 p2 <- DoHeatmap(ss, features = top5$feature, cells = cells, group.by = "seurat_clusters") + NoLegend() +
-  scale_fill_gradientn(colors =  c("#01665e","#35978f",'darkslategray3', "#f7f7f7", "#fee391","#fec44f","#F9AD03")) 
-p3 <- ggplot(srat@meta.data, aes(seurat_clusters, fill = seurat_clusters)) + geom_bar() + NoLegend()
+  scale_fill_gradientn(colors = c("#01665e", "#35978f", "darkslategray3", "#f7f7f7", "#fee391", "#fec44f", "#F9AD03"))
+p3 <- ggplot(srat@meta.data, aes(seurat_clusters, fill = seurat_clusters)) +
+  geom_bar() +
+  NoLegend()
 common_title <- sprintf("Unsupervised clustering %s, %d cells", srat@meta.data$orig.ident[1], ncol(srat))
-show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) ) + plot_layout(widths = c(1, 2)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))
-
+show((((p1 / p3) + plot_layout(heights = c(3, 2)) | p2)) + plot_layout(widths = c(1, 2)) + plot_layout(heights = c(3, 1)) + plot_annotation(title = common_title))
 ```
 
-```{r fig.width=10, fig.height=10, out.width='100%'}
-DT::datatable(top5[, c(1, 9, 11, 12)], caption = ("top 5 marker genes"), 
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-                               buttons = c( 'csv', 'excel')))
-
+```{r fig.width=10, fig.height=10, out.width='100%', eval=!params$testing}
+DT::datatable(top5[, c(1, 9, 11, 12)],
+  caption = ("top 5 marker genes"),
+  extensions = "Buttons",
+  options = list(
+    dom = "Bfrtip",
+    buttons = c("csv", "excel")
+  )
+)
 ```
@@
 
-```{r enrichment}
+```{r enrichment, eval=!params$testing}
 # define background genes = universe for enrichment
@@
 signatures <- list()
-for( i in unique(s.markers$group1)){
-  signatures[[paste0("cluster ",i)]] <- s.markers$feature[s.markers$group1 == i]
+for (i in unique(s.markers$group1)) {
+  signatures[[paste0("cluster ", i)]] <- s.markers$feature[s.markers$group1 == i]
 }
@@
 
-```{r fig.width=20, fig.height=8, out.width='100%'}
-
+```{r fig.width=20, fig.height=8, out.width='100%', eval=!params$testing}
 Enrichment_plot(category = "H", signatures = signatures, background = background)
-
 ```
@@
 
-```{r fig.width=20, fig.height=25, out.width='100%'}
+```{r fig.width=20, fig.height=25, out.width='100%', eval=!params$testing}
 Enrichment_plot(category = "C8", signatures = signatures, background = background)
@@
 ```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
+markers <- list(
+  "malignant" = CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "malignant")],
+  "immune" = CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "immune")],
+  "endothelium" = CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "endothelium")],
+  "non-malignant" = CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "non-malignant")]
+)
 
-markers = list("malignant" = CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "malignant")],
-            "immune"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "immune")],
-            "endothelium"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "endothelium")],
-            "non-malignant"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "non-malignant")])
-
-p1 <- SCpubr::do_DotPlot(sample = srat,
+p1 <- SCpubr::do_DotPlot(
+  sample = srat,
   group.by = "fetal_kidney_predicted.compartment",
@@
 
-p2 <- SCpubr::do_DotPlot(sample = srat,
+p2 <- SCpubr::do_DotPlot(
+  sample = srat,
   group.by = "fetal_kidney_predicted.cell_type",
@@
 
-p3 <- SCpubr::do_DotPlot(sample = srat,
+p3 <- SCpubr::do_DotPlot(
+  sample = srat,
   group.by = "fetal_full_predicted.annotation.l1",
@@
 
-p4 <- SCpubr::do_DotPlot(sample = srat,
+p4 <- SCpubr::do_DotPlot(
+  sample = srat,
   group.by = "fetal_full_predicted.annotation.l2",
@@
 
-p5 <- SCpubr::do_DotPlot(sample = srat,
+p5 <- SCpubr::do_DotPlot(
+  sample = srat,
   group.by = "seurat_clusters",
@@
 
-p6 <- SCpubr::do_DotPlot(sample = srat,
+p6 <- SCpubr::do_DotPlot(
+  sample = srat,
   group.by = "cellassign_celltype_annotation",
@@
 
-(p1|p2)/(p3|p4) /(p5|p6)
-
+(p1 | p2) / (p3 | p4) / (p5 | p6)
 ```
@@
 ```{r fig.width=20, fig.height=15, out.width='100%'}
-p1 <- do_Table_Heatmap( srat,
+p1 <- do_Table_Heatmap(srat,
   last_group = "fetal_kidney_predicted.compartment",
   first_group = "seurat_clusters"
-
 )
 
-p2 <- do_Table_Heatmap( srat,
+p2 <- do_Table_Heatmap(srat,
   last_group = "fetal_full_predicted.annotation.l1",
@@
 
-p3 <- do_Table_Heatmap( srat,
+p3 <- do_Table_Heatmap(srat,
   last_group = "singler_celltype_annotation",
@@
 
-p4 <- do_Table_Heatmap( srat,
+p4 <- do_Table_Heatmap(srat,
   last_group = "cellassign_celltype_annotation",
@@
 
-(p1|p2)/(p3|p4)
+(p1 | p2) / (p3 | p4)
 ```
@@
 ```{r fig.width=20, fig.height=15, out.width='100%'}
-p1 <- do_Table_Heatmap( srat,
+p1 <- do_Table_Heatmap(srat,
   last_group = "fetal_kidney_predicted.compartment",
   first_group = "fetal_full_predicted.annotation.l1"
-
 )
-p2 <- do_Table_Heatmap( srat,
+p2 <- do_Table_Heatmap(srat,
   last_group = "fetal_kidney_predicted.cell_type",
   first_group = "fetal_full_predicted.annotation.l1"
-
 )
-p1|p2
+p1 | p2
 ```
@@
 ```{r fig.width=20, fig.height=15, out.width='100%'}
-p2 <- do_Table_Heatmap( srat,
+p2 <- do_Table_Heatmap(srat,
   last_group = "fetal_kidney_predicted.cell_type",
   first_group = "fetal_full_predicted.annotation.l2"
-
 )
@@
 ```{r fig.width=20, fig.height=10, out.width='100%'}
-p1 <- do_Table_Heatmap( srat,
+p1 <- do_Table_Heatmap(srat,
   last_group = "fetal_kidney_predicted.compartment",
   first_group = "cellassign_celltype_annotation"
-
 )
-p2 <- do_Table_Heatmap( srat,
+p2 <- do_Table_Heatmap(srat,
   last_group = "fetal_kidney_predicted.compartment",
   first_group = "singler_celltype_annotation"
-
 )
-p1|p2
+p1 | p2
 ```
@@
 ```{r fig.width=20, fig.height=30, out.width='100%'}
-p1 <- do_Table_Heatmap( srat,
+p1 <- do_Table_Heatmap(srat,
   last_group = "fetal_kidney_predicted.cell_type",
   first_group = "cellassign_celltype_annotation"
-
 )
 
-p2 <- do_Table_Heatmap( srat,
+p2 <- do_Table_Heatmap(srat,
   last_group = "fetal_kidney_predicted.cell_type",
   first_group = "singler_celltype_annotation"
-
 )
 
-p1/p2
+p1 / p2
 ```

### 2024-11-05 analyses/cell-type-wilms-tumor-06/supplemental-notebooks/compare-label-transfer-approaches.Rmd

@@
 
+  # Perform label transfer with new code
+  assay <- "RNA"
   fetal_full <- srat_objects |>
@@
       \(srat, id) {
-        # Perform label transfer with new code
         set.seed(params$seed)
-        query <- prepare_query(srat, rownames(full_reference), file.path(module_base, "scratch", "homologs.rds"))
+
+        query <- prepare_query(
+          srat,
+          rownames(full_reference),
+          assay,
+          file.path(module_base, "scratch", "homologs.rds")
+        )
         query <- transfer_labels(
@@
           full_dims,
-          full_refdata
+          full_refdata,
+          query.assay = assay
         )
@@
 
+  # Perform label transfer with new code
+  assay <- "RNA"
   fetal_kidney <- srat_objects |>
@@
       \(srat, id) {
-        # Perform label transfer with new code
         set.seed(params$seed)
-        query <- prepare_query(srat, rownames(kidney_reference), file.path(module_base, "scratch", "homologs.rds"))
+
+        query <- prepare_query(
+          srat,
+          rownames(kidney_reference),
+          assay,
+          file.path(module_base, "scratch", "homologs.rds")
+        )
         query <- transfer_labels(
@@
           kidney_dims,
-          kidney_refdata
+          kidney_refdata,
+          query.assay = assay
         )
@@
 | SCPSC000208 | L1 | 101 | Mesangial cells | Metanephritic cells |  
-| SCPSC000208 | L1 | 101 | Intestinal epithelial cells | Metanephritic cells |  
+| SCPSC000208 | L1 | 75 | Intestinal epithelial cells | Metanephritic cells |  
 | SCPSC000208 | Organ | 149 | Kidney | Intestine |
@@
 
-- There are a small but notable number of cells flipped between mesenchyme and kidney cells, in particular for sample SCPSC000184.
-This is the main discrepancy.
-- Most of the cell type differences are not necessarily biologically meaningful for our purposes, as listed below. These are not noted in the table.
+- Most of the cell type differences are not in the table below because they are not necessarily biologically meaningful for our purposes:
    - `kidney cell` vs `podocyte`
@@
    - `mesenchymal cell` vs `mesenchymal stem cell`
-- There are a decent number of times when stroma and fetal nephron are flipped, but this makes sense given that we expect many of the stroma may be tumor.
-- Disagreeing annotation scores when using the cell type reference were often higher, but many of the disagreements for this reference were not meaningful (e.g.
-`kidney epithelial cell` vs `kidney cell`).
 
@@
 |--------|-----------|-------|---------|-----------------|
-| SCPSC000179 | cell type | 202 | mesenchymal cell | kidney epithelial cell | 
-| SCPSC000184 | compartment | 111 | fetal nephron |  stroma | 
-| SCPSC000184 | cell type | 536 | kidney epithelial cell | mesenchymal cell | 
-| SCPSC000194 | compartment | 565 | fetal nephron |  stroma | 
-| SCPSC000194 | cell type  | 89 | kidney epithelial cell | mesenchymal cell | 
-| SCPSC000205 | compartment  | 684 | fetal nephron |  stroma | 
-| SCPSC000208 | compartment  | 2111 | fetal nephron |  stroma |
+| SCPSC000179 | cell type | 94 | mesenchymal cell | kidney epithelial cell | 
+| SCPSC000205 | compartment  | 52 | fetal nephron |  stroma |
 

### 2024-11-05 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd

--- 2024-10-28 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
@@
   seed: 12345
-  homologs_file: "../scratch/homologs.rds"
-  testing: 0
 output:
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(
-  echo = TRUE,
-  message = FALSE,
-  warnings = FALSE
-)
+knitr::opts_chunk$set(echo = TRUE,
+                      message=FALSE,
+                      warnings=FALSE)
 ```
@@
 ```{r packages, message=FALSE, warning=FALSE}
-library(Seurat)
+library("Seurat")
 library(SeuratData)
 library(sctransform)
+library(Azimuth)
 library(SCpubr)
@@
 
-```{r}
-# load functions for label transfer
-source(
-  file.path(
-    module_base,
-    "notebook_template",
-    "utils",
-    "label-transfer-functions.R"
-  )
-)
-```
-
 ### Input files
@@
 
-Load the Azimuth reference which has been prepared for label transfer.
+We install and load the reference using `Azimuth`.
+```{r path_to_reference}
+#Check the names of the Azimuth available data and reference
+AvailableData()
 
-```{r path_to_reference}
-path_to_ref <- file.path(
-  module_base,
-  "results",
-  "references",
-  "cao_formatted_ref.rds"
-)
-if (!file.exists(path_to_ref)) {
-  stop("Reference file could not be found. Make sure `scripts/prepare-fetal-references.R` has been run first.")
-}
-ref <- readRDS(path_to_ref)
+# Install the fetal reference
+# InstallData("fetusref")
 
-# Pull out information from the reference object we need for label transfer
-reference <- ref$reference
-refdata <- ref$refdata
-dims <- ref$dims
-annotation_levels <- ref$annotation_levels
+ref <- SeuratData::LoadData("fetusref", type = "azimuth")
 ```
@@
 # open the processed rds object
-srat <- readRDS(file.path(data_dir, paste0("01-Seurat_", params$sample_id, ".Rds")))
-
-# prepare the query for label transfer
-# we don't want to overwrite the srat object since `prepare_query`
-#  removes features that are not present in the reference
-DefaultAssay(srat) <- "RNA"
-query <- prepare_query(srat, rownames(reference), params$homologs_file)
+srat <- readRDS(file.path(data_dir, paste0("01-Seurat_", params$sample_id,".Rds")))
 ```
@@
 ```{r run_azimuth, message=FALSE, warnings=FALSE}
-options(future.globals.maxSize = 891289600000000)
+DefaultAssay(srat) <- "RNA"
+options(future.globals.maxSize= 891289600000000)
+s <- Azimuth::RunAzimuth(srat, reference ="fetusref")
 
-# determine k.weight based CI
-if (params$testing) {
-  k.weight <- 10 # only for test datasets
-} else {
-  k.weight <- 50 # Azimuth default
-}
-query_labeled <- transfer_labels(
-  query,
-  reference,
-  dims,
-  refdata,
-  k.weight = k.weight
-)
+# We transfer the annotation to the pre-processed `Seurat` object as we don't want to keep changes done on the query by `RunAzimuth`
+metadata_vec <- c("predicted.annotation.l1.score", "predicted.annotation.l1", "predicted.annotation.l2.score", "predicted.annotation.l2", "predicted.organ.score", "predicted.organ")
 
-# We transfer the annotation to the pre-processed `Seurat` object as we don't want to keep changes done on the query by Azimuth
-annotation_columns <- c(
-  glue::glue("predicted.{annotation_levels}"),
-  glue::glue("predicted.{annotation_levels}.score")
-)
-metadata_to_trasfer <- query_labeled@meta.data[, annotation_columns]
+metadata_to_transfer <- s@meta.data[, metadata_vec]
 
-srat <- AddMetaData(srat, metadata_to_trasfer, col.name = paste0("fetal_full_", annotation_columns))
-
-rm(query)
-rm(query_labeled)
+srat <- AddMetaData(srat, metadata_to_transfer, col.name = paste0("fetal_full_", metadata_vec))
 ```
@@
 ```{r plot_azimuth, fig.height=15, fig.width=8, warnings=FALSE}
-d1 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_full_predicted.organ", label = TRUE, repel = TRUE) + NoLegend()
-d2 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_full_predicted.annotation.l1", label = TRUE, repel = TRUE) + NoLegend()
-d3 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_full_predicted.annotation.l2", label = TRUE, repel = TRUE) + NoLegend()
 
-f1 <- SCpubr::do_BarPlot(
-  sample = srat,
+d1 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_full_predicted.organ", label = TRUE, repel = TRUE) + NoLegend()
+d2 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_full_predicted.annotation.l1", label = TRUE, repel = TRUE) + NoLegend()
+d3 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_full_predicted.annotation.l2", label = TRUE, repel = TRUE) + NoLegend()
+
+f1 <- SCpubr::do_BarPlot(sample = srat,
                          group.by = "fetal_full_predicted.organ",
@@
                          font.size = 10,
-  legend.ncol = 4
-) +
-  ggtitle("% cells") +
+                         legend.ncol = 4) +
+                         ggtitle("% cells")+
                          xlab(params$sample_id)
 
-f2 <- SCpubr::do_BarPlot(
-  sample = srat,
+f2 <- SCpubr::do_BarPlot(sample = srat,
                          group.by = "fetal_full_predicted.annotation.l1",
@@
                          font.size = 10,
-  legend.ncol = 2
-) +
-  ggtitle("% cells") +
+                         legend.ncol = 2) +
+                         ggtitle("% cells")+
                          xlab(params$sample_id)
 
-f3 <- SCpubr::do_BarPlot(
-  sample = srat,
+f3 <- SCpubr::do_BarPlot(sample = srat,
                          group.by = "fetal_full_predicted.annotation.l2",
@@
                          font.size = 10,
-  legend.ncol = 2
-) +
-  ggtitle("% cells") +
+                         legend.ncol = 2) +
+                         ggtitle("% cells")+
                          xlab(params$sample_id)
 
-((d1 / f1) | (d2 / f2))
+((d1/f1) | (d2/f2) )
 ```
@@
 ```{r save}
-saveRDS(object = srat, file = file.path(output_dir, paste0("02a-fetal_full_label-transfer_", params$sample_id, ".Rds")))
+saveRDS(object = srat, file = file.path(output_dir, paste0("02a-fetal_full_label-transfer_",params$sample_id,".Rds")))
 ```

### 2024-11-05 analyses/cell-type-wilms-tumor-14/exploratory_analysis/03_cnv/03_infercnv_SCPCL000850.Rmd

@@
 level <- "compartment"
-predictions <- read.csv( file.path(path_anal, "results", "01_anchor_transfer_seurat", paste0(library_id, "_", level,".csv")) ) 
+predictions <- read.csv( file.path(path_anal, "results", "01_anchor_transfer_seurat", "RNA", paste0(library_id, "_", level,".csv")) ) 
 obj <- AddMetaData(object = obj, metadata = predictions)

### 2024-11-06 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
re-render the 07 notebook with updated results

@@
 
-- the predicted CNV calculated using intra-sample endothelial and immune cells (`--reference both`) as normal reference
+- the predicted CNV calculated using intra-sample endothelial and immune cells (`--reference both`) as normal reference; no reference was used for samples with fewer than 3 predicted normal cells.
 
@@
 metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
+sample_ids <- metadata |>
+  dplyr::filter(seq_unit != "spot") |>
+  dplyr::pull(scpca_sample_id) |>
+  unique()
 
 # Create a data frames of all annotations
-cell_type_df <- unique(metadata$scpca_sample_id) |>
+cell_type_df <- sample_ids |>
   purrr::map(
@@
     \(sample_id) {
-      # This sample was run with "none" for reference
-      if (sample_id == "SCPCS000190") {
+      # These samples were run with "none" as the reference
+      none_reference_samples <- c("SCPCS000177", "SCPCS000180", "SCPCS000181", "SCPCS000190", "SCPCS000197")
+      if (sample_id %in% none_reference_samples) {
         reference <- "none"

### 2024-11-06 analyses/cell-type-wilms-tumor-06/notebook_template/00b_characterize_fetal_kidney_reference_Stewart.Rmd

@@
 The rds data can be download using the download link https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds
-The reference was been downloaded and created in the script `scripts/prepare-fetal-references.R`.
+The reference was downloaded and created in the script `scripts/prepare-fetal-references.R`.
 
@@
 library(Seurat)
-library(Azimuth)
 library(SCpubr)

### 2024-11-06 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Fix subdiagnosis bug by adding unique, and fix several typos while we're here

@@
   dplyr::filter(scpca_sample_id == params$sample_id) |>
-  dplyr::pull(subdiagnosis)
+  dplyr::pull(subdiagnosis) |>
+  unique()
 ```
@@
 - in the case each cluster do have a set of specific marker genes, we could expect each of the cluster to have at least a different phenotype,
-- if marker genes are mostly shared between clusters, we might have overclustered,
+- if marker genes are mostly shared between clusters, we might have over-clustered,
 - if no marker genes are found, the quality of cells in the clusters might be impaired (high mitochondrial content? ribosomal content?)
@@
 
-For a Seurat object `object`, the function `MSigDB_score` will calculate a score using `AddModuleScore()` function for a MSigDB gene set `gs_name` in the category `category`
+For a Seurat object `object`, the function `MSigDB_score` will calculate a score using `AddModuleScore()` function for a `MSigDB` gene set `gs_name` in the category `category`
 
@@
 
-- `category` is a MSigDB collection (https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp).
+- `category` is a `MSigDB` collection (https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp).
 values in `c("H", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")`
 
-- `gs_name` is the name of a MSigDB gene set, e.g. `"HALLMARK_P53_PATHWAY"`
+- `gs_name` is the name of a `MSigDB` gene set, e.g. `"HALLMARK_P53_PATHWAY"`
 
@@
 
-- `category` is the MSigDB category or collection to be used
+- `category` is the `MSigDB` category or collection to be used
 values in `c("H", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")`
@@
 
-Here, we open the table of marker genes marker-sets/CellType_metadata.csv.
+Here, we open the table of marker genes `marker-sets/CellType_metadata.csv`.
 Note: we do not expect to have a clear and nice pattern of expression for all of the following markers in every tumor.
@@
 
-#### DROSHA target genes
+#### `DROSHA` target genes
 
@@
 
-#### DICER1 target genes
+#### `DICER1` target genes
 
@@
 
-We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense.
-DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method.
-Please check the preprint from Chistoph Hafemeister: https://www.biorxiv.org/content/10.1101/2023.03.28.534443v1
+We run `DElegate::FindAllMarkers2` to find markers of the different clusters and manually check if they do make sense.
+`DElegate::FindAllMarkers2` is an improved version of `Seurat::FindAllMarkers` based on pseudobulk differential expression method.
+Please check the preprint from Hafemeister: https://www.biorxiv.org/content/10.1101/2023.03.28.534443v1
 and tool described here: https://github.com/cancerbits/DElegate
@@
 
-We used three gene sets from MSiGDB:
+We used three gene sets from `MSigDB`:
 
-- the hallmark gene sets  are coherently expressed signatures derived by aggregating many MSigDB gene sets to represent well-defined biological states or processes.
+- the hallmark gene sets  are coherently expressed signatures derived by aggregating many `MSigDB` gene sets to represent well-defined biological states or processes.
 - the C3 : regulatory target gene sets  based on gene target predictions for microRNA seed sequences and predicted transcription factor binding sites.
@@
 
-We used enricher function from clusterProfiler to perform enrichment analysis.
+We used enricher function from `clusterProfiler` to perform enrichment analysis.
 
@@
 
-#### Hallmarks MSiGDB gene sets
+#### Hallmarks `MSigDB` gene sets
 
@@
 
-EMT signature should be enriched in stroma cluster.
-E2F/proliferation should be enriched in blastema cluster.
-MYC(N), TP53 must be enriched in blastema cluster.
+`EMT` signature should be enriched in stroma cluster.
+`E2F`/proliferation should be enriched in blastema cluster.
+`MYC(N)`, `TP53` must be enriched in blastema cluster.
 
-#### C8 MSiGDB cell type signature gene sets
+#### C8 `MSigDB` cell type signature gene sets
 
@@
 
-The MSigDB C8 gene set is quite relevant for kidney and nephroblastoma annotations.
-Epithelial (cancer and normal) cells should be enriched in mature/adulte kidney pathways while blastema cancer cells will show enrichment of fetal kidney development pathway / cap mesenchyme.
+The `MSigDB` C8 gene set is quite relevant for kidney and nephroblastoma annotations.
+Epithelial (cancer and normal) cells should be enriched in mature/adult kidney pathways while blastema cancer cells will show enrichment of fetal kidney development pathway / cap mesenchyme.
 
@@
 
-Wilms tumors can contain up to three histologies that resemble fetal kidney: blastema, stroma, and epithelia [[1](https://www.ncbi.nlm.nih.gov/books/NBK373356/), [2](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9915828/)].
+Wilms tumors can contain up to three histologies that resemble fetal kidney: blastema, stroma, and epithelia.
+[[1](https://www.ncbi.nlm.nih.gov/books/NBK373356/), [2](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9915828/)].
 Because of their histological similarity to fetal kidneys, Wilms tumors are thought to arise from developmental derangements in embryonic renal progenitors.
@@
 
-    + Annotations determined by SingleR, an automated reference-based method [Looney et al. 2019](https://www.nature.com/articles/s41590-018-0276-y).
+    + Annotations determined by `SingleR`, an automated reference-based method [Looney et al. 2019](https://www.nature.com/articles/s41590-018-0276-y).
 
-    + Annotations determined by CellAssign, an automated marker-gene based method [Zhang et al. 2019](https://www.nature.com/articles/s41592-019-0529-1).
+    + Annotations determined by `CellAssign`, an automated marker-gene based method [Zhang et al. 2019](https://www.nature.com/articles/s41592-019-0529-1).
 
@@
 
-#### Annotation from SingleR (no cancer or kidney specific dataset)
+#### Annotation from `SingleR` (no cancer or kidney specific dataset)
 
@@
 
-#### Annotation from cellassign (no cancer or kidney specific dataset)
+#### Annotation from `CellAssign` (no cancer or kidney specific dataset)
 
@@
 - `fetal_kidney_predicted.compartment` is one of the 4 main compartments composing a fetal kidney.
-We expect immune and endothelial cells to be healthy (non-canerous) cells identified easily with high confidency.
+We expect immune and endothelial cells to be healthy (non-cancerous) cells identified easily with high confidence.
 We expect stroma and fetal nephron compartment to contain both normal and malignant cells.
@@
 
-Here we check wether one (or more) `Seurat` cluster do correspond to one label.
+Here we check whether one (or more) `Seurat` cluster do correspond to one label.
 This would allow us to be more confident in the annotation of immune or endothelial cells for example.
@@
 
-#### Comparison of the  fetal reference with `singleR` and `cellassign`
+#### Comparison of the  fetal reference with `SingleR` and `CellAssign`
 

### 2024-11-07 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Fix typos and sample_id loop bug, and the file was also styled

@@
 
-- the `counts`of the endothelial marker VWF = "ENSG00000110799"
+- the `counts` of the endothelial marker `VWF = "ENSG00000110799"`
 
-- the `counts`of the immune marker PTPRC = "ENSG00000081237"
+- the `counts` of the immune marker `PTPRC = "ENSG00000081237"`
 
 ```{r path_to_query}
-sample_metadata_file <- file.path(repository_base, "data", "current","SCPCP000006", "single_cell_metadata.tsv")
+sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
 
-metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
+metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE) |>
+  dplyr::filter(seq_unit != "spot")
 
@@
     \(sample_id) {
-
       # Read in the Seurat object
@@
 
-`do_Feature_mean` shows heatmap of mean expression of a feature grouped by a metadata.
+`do_Feature_mean` shows a heatmap of mean expression of a feature grouped by a metadata.
 
@@
     theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5), title = element_text(size = rel(0.75))) +
-   guides(fill=guide_colourbar(title=paste0(feature)))
+    guides(fill = guide_colourbar(title = paste0(feature)))
 
@@
 ```{r fig.width=10, fig.height=4, out.width='100%'}
-do_Feature_boxplot <- function(df, group.by, feature, split.by ){
-
+do_Feature_boxplot <- function(df, group.by, feature, split.by) {
   df <- df %>%
@@
 
-  p <- ggplot(df, 
+  p <- ggplot(
+    df,
     aes(x = group, y = !!sym(feature), fill = group)
@@
 ```{r fig.width=10, fig.height=4, out.width='100%'}
-do_Feature_densityplot <- function(df, group.by, feature){
-
+do_Feature_densityplot <- function(df, group.by, feature) {
   df <- df %>%
@@
 
-  p <- ggplot(df, 
+  p <- ggplot(
+    df,
     aes(x = !!sym(feature), fill = group)
@@
 ```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
-
 kidney_count_df <- cell_type_df |>
@@
     total = kidney + not_kidney,
-    kidney = round(kidney/total,  4)*100, 
-    not_kidney = round(not_kidney/total, 4)*100) |>
+    kidney = round(kidney / total, 4) * 100,
+    not_kidney = round(not_kidney / total, 4) * 100
+  ) |>
   # arrange in descending order of percentage of kidney cells
@@
   caption = "percentage of cells mapping to the predicted organ kidney",
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-                               buttons = c( 'csv', 'excel')))
+  extensions = "Buttons",
+  options = list(
+    dom = "Bfrtip",
+    buttons = c("csv", "excel")
+  )
+)
 ```
@@
 
-The predicted compartment is the result of the label transfer fron the human fetal kidney atlas: 
+The predicted compartment is the result of the label transfer from the human fetal kidney atlas: 
 [Stewart et al.](https://www.science.org/doi/10.1126/science.aat5031) created a [human fetal kidney atlas](https://www.kidneycellatlas.org/). 
@@
 - `fetal_kidney_predicted.compartment` is one of the 4 main compartments composing a fetal kidney.
-We expect immune and endothelial cells to be healthy (non-canerous) cells identified easily with high confidency. 
+We expect immune and endothelial cells to be healthy (non-cancerous) cells identified easily with high confidence. 
 We expect stroma and fetal nephron compartment to contain both normal and malignant cells.
@@
 
-For each compartment, we sumarized the percentage of cells that do match kidney annotation or not. 
-Please note that this table is <u>not</u> sample-specific but contains all samples pooled into one.
+For each compartment, we summarized the percentage of cells that do match kidney annotation or not. 
+Please note that this table is _not_ sample-specific but contains all samples pooled into one.
 
 ```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
-
 compartment_count_df <- cell_type_df |>
@@
   caption = "counts of cells in each compartment",
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-                               buttons = c( 'csv', 'excel')))
+  extensions = "Buttons",
+  options = list(
+    dom = "Bfrtip",
+    buttons = c("csv", "excel")
+  )
+)
 ```
@@
 ```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
-p <- do_Feature_densityplot( df = cell_type_df,
+p <- do_Feature_densityplot(
+  df = cell_type_df,
   feature = "compartment_score",
@@
 )
-p +  geom_vline(xintercept=params$predicted.score_thr)
+p + geom_vline(xintercept = params$predicted.score_thr)
 ```
@@
 ```{r fig.height=15, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
-p <- do_Feature_boxplot( df = cell_type_df,
+p <- do_Feature_boxplot(
+  df = cell_type_df,
   feature = "compartment_score",
   group.by = "compartment",
-                    split.by = "sample_id")
+  split.by = "sample_id"
+)
 
-p + geom_hline(yintercept = params$predicted.score_thr)  + coord_cartesian(ylim = c(0.25,1))
-
+p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25, 1))
 ```
@@
 ```{r fig.height=3, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
-p <- do_Feature_boxplot( df = cell_type_df,
+p <- do_Feature_boxplot(
+  df = cell_type_df,
   feature = "compartment_score",
   group.by = "compartment",
-                    split.by = "treatment")
+  split.by = "treatment"
+)
 
-p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25,1))
-
+p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25, 1))
 ```
@@
 ```{r fig.height=3, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
-p <- do_Feature_boxplot( df = cell_type_df,
+p <- do_Feature_boxplot(
+  df = cell_type_df,
   feature = "compartment_score",
   group.by = "compartment",
-                    split.by = "subdiagnosis")
+  split.by = "subdiagnosis"
+)
 
-p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25,1))
-
+p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25, 1))
 ```
@@
 ```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
-p <- do_Feature_boxplot( df = cell_type_df,
+p <- do_Feature_boxplot(
+  df = cell_type_df,
   feature = "compartment_score",
   group.by = "compartment",
-                    split.by = "disease_timing")
+  split.by = "disease_timing"
+)
 
-p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25,1))
-
+p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25, 1))
 ```
@@
   caption = "counts of cells that have a good/poor mapping in each compartment",
-              extensions = 'Buttons', 
-              options = list(  dom = 'Bfrtip',
-                               buttons = c( 'csv', 'excel')))
+  extensions = "Buttons",
+  options = list(
+    dom = "Bfrtip",
+    buttons = c("csv", "excel")
+  )
+)
 ```
@@
 
-We look at the endothelial marker "ENSG00000110799" = "VWF"
+We look at the endothelial marker `"ENSG00000110799" = "VWF"`
 
 ```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
-do_Feature_mean( df = cell_type_df,
+do_Feature_mean(
+  df = cell_type_df,
   feature = "ENSG00000110799",
-                    group.by = "compartment") +
+  group.by = "compartment"
+) +
   ggtitle("VWF expression averaged by compartment for all cells")
@@
 ```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
-do_Feature_mean( df = cell_type_df_pass,
+do_Feature_mean(
+  df = cell_type_df_pass,
   feature = "ENSG00000110799",
-                    group.by = "compartment") +
+  group.by = "compartment"
+) +
   ggtitle("VWF expression averaged by compartment for cells passing the `predicted.score` threshold")
@@
 ```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
-do_Feature_mean( df = cell_type_df_nopass,
+do_Feature_mean(
+  df = cell_type_df_nopass,
   feature = "ENSG00000110799",
-                    group.by = "compartment") +
+  group.by = "compartment"
+) +
   ggtitle("VWF expression averaged by compartment for cells that don't pass the `predicted.score` threshold")
@@
 ```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
-do_Feature_boxplot( df = cell_type_df,
+do_Feature_boxplot(
+  df = cell_type_df,
   feature = "ENSG00000110799",
   group.by = "compartment",
-                    split.by = "sample_id")  +
+  split.by = "sample_id"
+) +
   ggtitle("boxplot of VWF expression for all cells")
-
 ```
@@
 
-We look at the immune marker "ENSG00000081237" = "PTPRC" alias "CD45"
+We look at the immune marker `"ENSG00000081237" = "PTPRC"` alias `"CD45"`
 
 ```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
-do_Feature_mean( df = cell_type_df,
+do_Feature_mean(
+  df = cell_type_df,
   feature = "ENSG00000081237",
-                    group.by = "compartment") +
+  group.by = "compartment"
+) +
   ggtitle("PTPRC expression averaged by compartment for all cells")
@@
 ```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
-do_Feature_mean( df = cell_type_df_pass,
+do_Feature_mean(
+  df = cell_type_df_pass,
   feature = "ENSG00000081237",
-                    group.by = "compartment") +
+  group.by = "compartment"
+) +
   ggtitle("PTPRC expression averaged by compartment for cells passing the `predicted.score` threshold")
@@
 ```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
-do_Feature_mean( df = cell_type_df_nopass,
+do_Feature_mean(
+  df = cell_type_df_nopass,
   feature = "ENSG00000081237",
-                    group.by = "compartment") +
+  group.by = "compartment"
+) +
   ggtitle("PTPRC expression averaged by compartment for cells that are not passing the predicted score threshold")
@@
 ```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
-do_Feature_boxplot( df = cell_type_df,
+do_Feature_boxplot(
+  df = cell_type_df,
   feature = "ENSG00000081237",
   group.by = "compartment",
-                    split.by = "sample_id")  +
+  split.by = "sample_id"
+) +
   ggtitle("boxplot of PTPRC expression for all cells")
-
 ```
@@
 ```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
-ggplot(cell_type_df, aes( x = umap.umap_1, y = umap.umap_2, color = compartment), shape = 19, size = 2)+
+ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = compartment), shape = 19, size = 2) +
   geom_point() +
-  geom_point(data = cell_type_df_nopass, mapping = aes( x = umap.umap_1, y = umap.umap_2),  shape = 4, color = "black", alpha = 0.3) +
+  geom_point(data = cell_type_df_nopass, mapping = aes(x = umap.umap_1, y = umap.umap_2), shape = 4, color = "black", alpha = 0.3) +
   facet_wrap(facets = ~sample_id)
@@
 ```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
-ggplot(cell_type_df_pass, aes( x = umap.umap_1, y = umap.umap_2, color = compartment), shape = 19, size = 2)+
+ggplot(cell_type_df_pass, aes(x = umap.umap_1, y = umap.umap_2, color = compartment), shape = 19, size = 2) +
   geom_point() +
   facet_wrap(facets = ~sample_id)
-
 ```
@@
 The next step of the analysis is to test whether `copyKAT` can help us distinguishing cancer cells from non-malignant cell types.
-[`copyKAT`](https://github.com/navinlabcode/copykat) (Copynumber Karyotyping of Tumors) is a computational tool using integrative Bayesian approaches to identify genome-wide aneuploidy in single cells to separate tumor cells from normal cells, using high-throughput sc-RNAseq data.
+[`copyKAT`](https://github.com/navinlabcode/copykat) is a computational tool using integrative Bayesian approaches to identify genome-wide aneuploidy in single cells to separate tumor cells from normal cells, using high-throughput scRNAseq data.
 The underlying logic for calculating DNA copy number events from RNAseq data is that gene expression levels of many adjacent genes can provide depth information to infer genomic copy number in that region.

### 2024-11-07 analyses/metacells/notebooks/testing-seacells.Rmd
update testing notebook and add rendering to script

@@
 output: html_notebook
+params:
+  project_id: "SCPCP000001"
+  sample_id: "SCPCS000001"
+  library_id: "SCPCL000001"
 ---
@@
 import pathlib
+import pickle
 
@@
 data_dir <- file.path(root_dir, "data", "current")
+base_dir <- file.path(root_dir, "analyses", "metacells")
+results_dir <- file.path(base_dir, "results", params$project_id, params$sample_id)
 
-infile <- file.path(data_dir, "SCPCP000001/SCPCS000001/SCPCL000001_processed_rna.h5ad")
+ad_file <- file.path(data_dir, params$project_id, params$sample_id, paste0(params$library_id, "_processed_rna.h5ad"))
+sc_anndata_file <- file.path(results_dir, paste0(params$library_id, "_seacells.h5ad"))
+sc_model_file <- file.path(results_dir, paste0(params$library_id, "_seacells_model.pkl"))
+
 ```
 
-## Load data
+## Load SEACells results
+
 ```{python load_data}
-ad = sc.read(r.infile)
-
-# put the highly variable genes in a column of adata.var
-ad.var["highly_variable"] = ad.var.gene_ids.isin(ad.uns["highly_variable_genes"])
-```
-
-## Run SEACells
-
-```{python run_seacells}
-n_SEACells = round(ad.n_obs / 75)
-n_waypoint_eigs = 10
-
-model = SEACells.core.SEACells(
-    ad,
-    build_kernel_on="X_pca",
-    n_SEACells=n_SEACells,
-    n_waypoint_eigs=n_waypoint_eigs,
-    convergence_epsilon=1e-5,
-)
-```
-```{python}
-model.construct_kernel_matrix()
-M = model.kernel_matrix
+ad = sc.read(r.ad_file)
+sc_ad = sc.read(r.sc_anndata_file)
+with open(r.sc_model_file, "rb") as f:
+  sc_model = pickle.load(f)
 ```
@@
 ```{python}
-sns.clustermap(M.toarray()[:500, :500])
+_ = sns.clustermap(sc_model.kernel_matrix.toarray()[:500, :500])
 plt.show()
@@
 ```{python}
-model.initialize_archetypes()
-SEACells.plot.plot_initialization(ad, model)
+SEACells.plot.plot_initialization(sc_ad, sc_model)
 ```
@@
 ```{python}
-model.fit(min_iter=10, max_iter=50)
-model.plot_convergence()
+sc_model.plot_convergence()
 ```
@@
 ```{python}
-SEACells.plot.plot_2D(ad, key="X_umap", colour_metacells=True)
+SEACells.plot.plot_2D(sc_ad, key="X_umap", colour_metacells=True)
 ```
@@
 ```{python}
-SEACells.plot.plot_SEACell_sizes(ad, bins=10)
+SEACells.plot.plot_SEACell_sizes(sc_ad, bins=10)
 ```
@@
 purity = SEACells.evaluate.compute_celltype_purity(
-    ad, "singler_celltype_ontology"
+    sc_ad, "singler_celltype_ontology"
 )

### 2024-11-07 analyses/metacells/run-metacells.sh

@@ -17,9 +17,16 @@ results_dir="results/${project}/${sample}"
 mkdir -p ${results_dir}
 
 # run the SEACells script
+echo "Calculating SEACells..."
 python scripts/run-seacells.py \
     ${test_file} \
     --adata_out ${results_dir}/${library}_seacells.h5ad \
     --model_out ${results_dir}/${library}_seacells_model.pkl \
     --logfile ${results_dir}/${library}_seacells.log \
     --seed 2024
+
+echo "Rendering test notebook"
+Rscript -e "rmarkdown::render(
+    'notebooks/testing-seacells.Rmd',
+    params=list(project_id=''${project}', sample_id='${sample}', library_id='${library}')
+  )"

### 2024-11-07 analyses/metacells/notebooks/testing-seacells.Rmd
Add text to notebook and render

@@
 ---
+
+## Introduction
+
+After running SEACells on the data, we want to do some evaluation of how well it might have worked on various samples.
+This notebook is to pilot what some of the more and less useful visualizations might be, as well as to test running in a mixed R/Python environment.
+Will this result in the best of both worlds?
 
@@
 
-## Path definitions
+### Path definitions
 
@@
 
-ad_file <- file.path(data_dir, params$project_id, params$sample_id, paste0(params$library_id, "_processed_rna.h5ad"))
+# these files were created by the `scripts/run-seacells.py` script
 sc_anndata_file <- file.path(results_dir, paste0(params$library_id, "_seacells.h5ad"))
@@
 
-## Load SEACells results
+### Load SEACells results
 
 ```{python load_data}
-ad = sc.read(r.ad_file)
+# the updated AnnData file
 sc_ad = sc.read(r.sc_anndata_file)
+# the SEACells model object
 with open(r.sc_model_file, "rb") as f:
@@
 ```
+
+## Visualizations
+
+### Kernel matrix
+
+The first visualization we will look at is the kernel matrix.
+This should show a strong block-diagonal structure.
 
@@
 
+### Model fitting
+
+We can look at where SEACells started with the metacells as plotted on the UMAP grid, and how the model converged.
+
+First, where it started:
+
 ```{python}
@@
 ```
+
+Now, how the convergence went:
 
@@
 
+Finally, we can see where the metacells ended up on the UMAP grid:
+
 ```{python}
@@
 
+We can also plot how many cells were assigned to each metacell:
+
 ```{python}
@@
 ```
+
+### Cell type purity
+
+The first real measure of how well SEACells worked may be to look at the cell type purity of the metacells.
+We will use the cell types as assigned by singleR to do this.
+
+If SEACells is working well, we should expect lots of high values here, indicating that most of the cells that contribute to a metacell were from the same initial cell type.
 
@@
 ```
+
+That looks pretty good, but if we wanted to evaluate whether it was better than chance, we would need to do some randomization, assigning cells to metacells randomly and seeing whether this method is outperfroming that.
+The main concern would be that if there are not many different cell types, we will expect to get very high purity scores.
+
+But that evaluation will be in a later revision of this notebook.

### 2024-11-08 analyses/metacells/notebooks/testing-seacells.Rmd
try not specifying conda environment for reticulate

@@
 library(reticulate)
-reticulate::use_condaenv("openscpca-metacells")
+# reticulate::use_condaenv("openscpca-metacells")
 ```
@@
 sc_model_file <- file.path(results_dir, paste0(params$library_id, "_seacells_model.pkl"))
-
 ```
@@
   geom_boxplot() +
-  labs(title = "Celltype purity",
-       y = "Celltype purity (singler)")
+  labs(
+    title = "Celltype purity",
+    y = "Celltype purity (singler)"
+  )
 ```

### 2024-11-08 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>
WIP: began sketching out notebook to eval clustering

@@
 date: "`r Sys.Date()`"
+author: Data Lab
+params:
+  seed: 2024
 output: 
-  html_document:
+  html_notebook: 
+    toc: yes
+    toc_float: yes
     df_print: paged
@@
 ## Introduction
+
+This notebook provides examples of how to use the `rOpenScPCA` package to:
+
+- Perform clustering on an SCE object or Seurat object
+- Add clustering results to the object
+- Evaluate clustering results using a variety of metrics, including:
+  - Average silhouette width
+  - Neighborhood purity
+  - Cluster stability, as measured with the adjusted Rand index
+
+This notebook will use the `SCPCS000001` sample for demonstration and assumes this sample is present in the `OpenScPCA-analysis/data/current` directory.
+Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about downloading OpenScPCA data.
 
@@
 ```{r packages}
-# load required packages
 library(rOpenScPCA)
+
 suppressPackageStartupMessages({
@@
 })
+
+# Set ggplot theme
+theme_set(theme_bw())
 ```
 
-### Paths
-
-Setting paths to directories and files for input and output at the start of a notebook can be helpful for organization and reproducibility.
-You may not know all of the paths you need when you start a notebook, but once you do, keeping them all in one place is helpful, both for yourself in the future and for other users of the notebook.
-
-#### Base directories
-
-In the chunk below, we include some common directories that you might need, finding their locations using the very useful [`rprojroot` package](https://rprojroot.r-lib.org/).
+#### Paths
 
 ```{r base paths}
-# The base path for the OpenScPCA repository, found by its (hidden) .git directory
+# The base path for the OpenScPCA repository
 repository_base <- rprojroot::find_root(rprojroot::is_git_root)
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "hell-c")
+module_base <- file.path(repository_base, "analyses", "hello-clusters")
 ```
@@
 ```{r paths}
-# Input files
+# Path to processed SCE file for sample SCPCS000001
+input_sce_file <- file.path(data_dir, "SCPCP000001", "SCPCS000001", "SCPCL000001_processed.rds")
 
-# Output files
+# Path to output file where clustering results will be saved
+cluster_tsv <- file.path(module_base, "results", "SCPCS000001-cluster-results.tsv")
 ```
 
-## Analysis content
+### Read in data
 
-Organize the remainder of your content into sections and subsections as appropriate for your analysis.
+```{r}
+sce <- readRDS(input_sce_file)
 
-Add a new code chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I* on Windows or Linux, *Cmd+Option+I* on a Mac.
-Be sure to comment your code and explain your steps as you go.
+# Convert to Seurat object for demonstration
+seurat_obj <- as.Seurat(
+  sce,
+  counts = "counts",
+  data = "logcounts"
+)
+```
 
-When you save the notebook, an HTML file containing the code and output will be saved alongside it.
-To preview the HTML file, click the *Preview* button or press *Ctrl+Shift+K* on Windows or Linux, *Cmd+Shift+K* on a Mac.
+## Perform clustering
 
-The preview shows you a rendered HTML copy of the contents of the editor.
-Consequently, *Preview* does not run any R code chunks.
-Instead, the output of the chunk when it was last run in the editor is displayed.
+This section will show how to perform clustering with the function `rOpenScPCA::calculate_clusters`.
 
-It is usually a good idea to rerun the entire notebook before sharing any output, to ensure that the notebook and output files are fully up to date.
-You can do this by clicking on the *Run* menu in the toolbar and clicking on *Restart R and Run All Chunks*.
+By default, this function will use the following parameters:
+
+- Louvain algorithm
+- Jaccard weighting
+- 10 nearest neighbors
+- A resolution parameter of 1
+
+These parameters can be customized, as follows:
+
+- `algorithm` can be one of
+  - `"louvain"`, `"walktrap"`, or `"leiden"`
+- etc.
+
+This function will return a table with both cluster identities and specified parameter values.
+
+## Cluster on an SCE
+
+Make sure to provide the seed!
+
+The function assumes the PC name is `"PCA"`. If it's not, you can provide this ifnormation with the `pc_name` argument.
+
+```{r}
+cluster_results <- calculate_clusters(
+  sce,
+  seed = params$seed
+)
+cluster_results
+```
+
+If you want to include these in your SCE object colData, you can add columns as shown in the following chunk.
+
+_Caution_! There already exists a column `cluster` which represents an automated clustering which was _not evaluated_ or parameterized according to the dataset. 
+You may therefore want to create a new column, for example `openscpca_cluster`
+
+```{r}
+# First, extract the colData into a separate data frame for manipulation
+sce_coldata <- colData(sce) |>
+  as.data.frame()
+
+# Second, let's rename the cluster column in `cluster_results` to avoid conflict
+# But, we don't want to overwrite the results themselves
+cluster_results_renamed <- cluster_results |>
+  dplyr::rename(openscpca_cluster = cluster)
+
+# Third, add columns as desired using the barcodes and cell_id columns to join
+sce_coldata <- sce_coldata |>
+  dplyr::left_join(
+    cluster_results_renamed,
+    by = c("barcodes" = "cell_id")
+  )
+
+# Finally, add the modified colData back into the SCE, ensuring the rownames are retained
+colData(sce) <- DataFrame(
+  sce_coldata,
+  row.names = sce_coldata$barcodes
+)
+
+colData(sce)
+```
+
+## Cluster on a Seurat object
+
+```{r}
+cluster_results <- calculate_clusters(
+  seurat_obj,
+  seed = params$seed,
+  # Since we converted this from an SCE object, its PC name is `"PCA"`
+  pc_name = "PCA"
+)
+
+cluster_results
+```
+
+If you want to include these in your SCE object colData, you can add columns as shown in the following chunk.
+
+_Caution_! There already exists a column `cluster` which represents an automated clustering which was _not evaluated_ or parameterized according to the dataset. 
+You may therefore want to create a new column, for example `openscpca_cluster`
+
+```{r}
+# Let's rename the cluster column in `cluster_results` to avoid conflict and/or ambiguity
+cluster_results <- cluster_results |>
+  dplyr::rename(openscpca_cluster = cluster)
+
+seurat_obj <- AddMetaData(seurat_obj, cluster_results)
+```
+
+## Evaluate clustering results
+
+For this section, we'll just use the SCE object.
+
+### Silhouette width
+
+Silhouette width values range from -1 to 1, where a value close to 1 means a given cell is very closely matched to its own cluster, and lower values mean a cell is not well matched to its own cluster.
+
+We'll use the function `rOpenScPCA::calculate_silhouette` to calculate the silhouette width, and then visualize the results.
+This function assumes columns called `cell_id` and `cluster`.
+
+```{r}
+silhouette_results <- calculate_silhouette(
+  sce,
+  cluster_results
+)
+
+silhouette_results
+```
+
+```{r}
+ave_silhouette_width <- mean(silhouette_results$silhouette_width)
+ggplot(silhouette_results) +
+  aes(x = silhouette_width) +
+  geom_density(fill = "lightblue") +
+  geom_vline(xintercept = ave_silhouette_width, color = "red", size = 1)
+```
+
+For most clusters, it is below 0.25, and for cluster 5 it is actually negative.
+We also see it is higher for clusters 11 and 12, reaching up to about 0.5.
+
+Overall, these values suggest that clustering is not fantastic and many cells may not well matched. 
+We can explore this further by visualizing distributions for each cluster:
+
+```{r}
+ggplot(silhouette_results) +
+  aes(x = cluster, y = silhouette_width) +
+  geom_violin() +
+  geom_jitter(size = 0.75, alpha = 0.25) +
+  stat_summary(color = "red")
+```
+Cluster 5 looks particularly rough, but cluster 11 seems the most reliable.
+
+### Neighborhood purity
+
+We'll use the function `rOpenScPCA::calculate_purity` to calculate the neighborhood purity, and then visualize the results.
+This function assumes columns called `cell_id` and `cluster`.
+
+```{r}
+purity_results <- calculate_purity(
+  sce,
+  cluster_results
+)
+
+purity_results
+```
+
+```{r}
+ave_purity <- mean(purity_results$purity)
+
+ggplot(purity_results) +
+  aes(x = purity) +
+  geom_density(fill = "lightblue") +
+  geom_vline(xintercept = ave_purity, color = "red", size = 1)
+```
+Not bad! Let's look at individual clusters:
+
+```{r}
+ggplot(purity_results) +
+  aes(x = cluster, y = purity) +
+  geom_violin() +
+  geom_jitter(size = 0.75, alpha = 0.25) +
+  stat_summary(color = "red")
+```
+With this measure, clusters 1, 3, and 6 all have average values below 0.75, but the rest are high - in particular clusters 11, 12, and 13.
+
+### Cluster stability
+
+We'll use the function `rOpenScPCA::calculate_stability` to calculate the neighborhood purity, and then visualize the results.
+This function assumes columns called `cell_id` and `cluster`.
 

### 2024-11-11 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
Continued WIP: begin to flesh out sections, some reorg as it comes together

@@
 - Add clustering results to the object
-- Evaluate clustering results using a variety of metrics, including:
-  - Average silhouette width
+- Evaluate clustering results using several metrics including:
+  - Silhouette width
   - Neighborhood purity
-  - Cluster stability, as measured with the adjusted Rand index
+  - Cluster stability, as measured with the Adjusted Rand index
 
@@
 
-```{r}
+To begin, we'll read in the `SingleCellExperiment` (SCE) object and convert it to a Seurat object.
+This will allow us to demonstrate `rOpenScPCA` usage with both SCE and Seurat objects.
+
+```{r readrds}
 sce <- readRDS(input_sce_file)
@@
 
-By default, this function will use the following parameters:
+By default, this function will calculate clusters using the following parameters:
 
@@
 
-- `algorithm` can be one of
+- The `algorithm` can be one of:
   - `"louvain"`, `"walktrap"`, or `"leiden"`
-- etc.
+- The `weighting` can be one of:
+  - `"jaccard"`, `"rank"`, or `"number"`
+- The nearest neighbors parameter can be customized with the `nn` argument
+- The resolution parameter can be customized with the `resolution` argument
+  - This parameter is only used by Louvain and Leiden algorithms
+- If the Leiden algorithm is used, its default objective function parameter will be `"CPM"`, but you can also set  `objective_function = "modularity"` instead.
+- You can provide additional parameters as a list to the `cluster_args` argument.
+  - Please refer to the [`igraph` documentation](https://igraph.org/r/html/latest) to learn more about what additional parameters can be provided to each clustering algorithm.
+  - Note that `cluster_args` only accepts single-length arguments (no vectors or lists).
 
-This function will return a table with both cluster identities and specified parameter values.
+This function will return a table with the following columns:
 
-## Cluster on an SCE
+- `cell_id`: Unique cell identifiers, e.g. cell barcodes
+- `cluster`: A factor column with the cluster identities
+- There will be one column for each clustering parameter used
 
-Make sure to provide the seed!
+### Clustering from an SCE object
 
-The function assumes the PC name is `"PCA"`. If it's not, you can provide this ifnormation with the `pc_name` argument.
+When performing clustering on an SCE object, `rOpenScPCA::calculate_clusters` assumes that the principal components are stored in a reduced dimension slot named `"PCA"`.
+If your 
+The function assumes the PC name is `"PCA"`. 
+If your SCE has a different name for the principal components, you can provide the name with the argument `pc_name`.
 
-```{r}
+We also strongly recommend providing a random sed with the `seed` argument to ensure reproducibility.
+
+```{r cluster sce}
+# Calculate clusters with default parameters
 cluster_results <- calculate_clusters(
@@
 )
+
+# Print resulting table
 cluster_results
@@
 
-If you want to include these in your SCE object colData, you can add columns as shown in the following chunk.
+## Clustering a Seurat object
 
-_Caution_! There already exists a column `cluster` which represents an automated clustering which was _not evaluated_ or parameterized according to the dataset. 
-You may therefore want to create a new column, for example `openscpca_cluster`
-
-```{r}
-# First, extract the colData into a separate data frame for manipulation
-sce_coldata <- colData(sce) |>
-  as.data.frame()
-
-# Second, let's rename the cluster column in `cluster_results` to avoid conflict
-# But, we don't want to overwrite the results themselves
-cluster_results_renamed <- cluster_results |>
-  dplyr::rename(openscpca_cluster = cluster)
-
-# Third, add columns as desired using the barcodes and cell_id columns to join
-sce_coldata <- sce_coldata |>
-  dplyr::left_join(
-    cluster_results_renamed,
-    by = c("barcodes" = "cell_id")
-  )
-
-# Finally, add the modified colData back into the SCE, ensuring the rownames are retained
-colData(sce) <- DataFrame(
-  sce_coldata,
-  row.names = sce_coldata$barcodes
-)
-
-colData(sce)
-```
-
-## Cluster on a Seurat object
-
-```{r}
+```{r cluster seurat}
 cluster_results <- calculate_clusters(
@@
 
-If you want to include these in your SCE object colData, you can add columns as shown in the following chunk.
-
-_Caution_! There already exists a column `cluster` which represents an automated clustering which was _not evaluated_ or parameterized according to the dataset. 
-You may therefore want to create a new column, for example `openscpca_cluster`
-
-```{r}
-# Let's rename the cluster column in `cluster_results` to avoid conflict and/or ambiguity
-cluster_results <- cluster_results |>
-  dplyr::rename(openscpca_cluster = cluster)
-
-seurat_obj <- AddMetaData(seurat_obj, cluster_results)
-```
-
 ## Evaluate clustering results
@@
 
-Silhouette width values range from -1 to 1, where a value close to 1 means a given cell is very closely matched to its own cluster, and lower values mean a cell is not well matched to its own cluster.
+Silhouette width is a common metric that measures how well separated clusters are by, for each cell, comparing the average distance to all cells in the same cluster, and all cells in other clusters.
+This value ranges from -1 to 1.
+Cells in well-separated clusters should have high silhouette values closer to 1.
+You can read more about silhouette width purity from the [_Orchestrating Single Cell Analysis with Bioconductor_ book](https://bioconductor.org/books/3.19/OSCA.advanced/clustering-redux.html#silhouette-width).
 
@@
 
-```{r}
+```{r silhouette}
 silhouette_results <- calculate_silhouette(
   sce,
-  cluster_results
+  cluster_results,
+  cluster_col = "openscpca_cluster"
 )
@@
 
-```{r}
+```{r plot silhouette densit}
 ave_silhouette_width <- mean(silhouette_results$silhouette_width)
@@
 
-```{r}
+```{r plot silhouette violin}
 ggplot(silhouette_results) +
@@
 ```
+
 Cluster 5 looks particularly rough, but cluster 11 seems the most reliable.
@@
 
+Neighborhood purity is defined, for each cell, as the proportion of neighboring cells that are assigned to the same cluster.
+This value ranges from 0 to 1.
+Cells in well-separated clusters should have high purity values closer to 1, since there should be minimal overlap between member and neighboring cells.
+You can read more about neighborhood purity from the [_Orchestrating Single Cell Analysis with Bioconductor_ book](https://bioconductor.org/books/3.19/OSCA.advanced/clustering-redux.html#cluster-purity).
+
 We'll use the function `rOpenScPCA::calculate_purity` to calculate the neighborhood purity, and then visualize the results.
+This function requires 
 This function assumes columns called `cell_id` and `cluster`.
 
-```{r}
+```{r purity}
 purity_results <- calculate_purity(
@@
 
-```{r}
+```{r plot purity density}
 ave_purity <- mean(purity_results$purity)
@@
 
-```{r}
+```{r plot purity violin}
 ggplot(purity_results) +
@@
 
-We'll use the function `rOpenScPCA::calculate_stability` to calculate the neighborhood purity, and then visualize the results.
-This function assumes columns called `cell_id` and `cluster`.
+Another approach to exploring cluster quality is how stable the clusters themselves are using bootstrapping. 
+Given a set of original clusters, we can compare the bootstrapped cluster identities to original ones using the Adjusted Rand Index (ARI), which measures the similarity of two data clusterings.
+ARI ranges from -1 to 1, where:
+
+- A value of 1 indicates they are completely overlapping
+- A value of -1 indicates they are completely distinct
+- A value of 0 indicates a random relationship
+
+We expect that highly stable clusterings have ARI values closer to 1 across a set of bootstrap replicates.
+
+You can read more about the Adjusted Rand Index from the [_Orchestrating Single Cell Analysis with Bioconductor_ book](https://bioconductor.org/books/release/OSCA.advanced/clustering-redux.html#adjusted-rand-index).
+
+We'll use the function `rOpenScPCA::calculate_stability` to calculate the stability of clusters using a bootstrapping approach, and then visualize the results.
+
+- By default, this function performs 20 bootstrap replicates, but this can be customized using the argument `replicates`.
+- This function will return a data frame with columns `replicate`, `ari`, and additional columns for the clustering parameters used when calculating bootstrap clusters.
+
+```{r stability, warning=FALSE}
+stability_results <- calculate_stability(
+  sce,
+  cluster_results,
+  seed = params$seed
+)
+
+stability_results
+```
+
+We can visualize these results, for example, with a density plot with a rug to emphasize individual values:
+
+```{r plot stability}
+ggplot(stability_results) +
+  aes(x = ari) +
+  geom_density(fill = "lightblue") +
+  geom_rug(color = "red", size = 1)
+```
+
+All bootstrap replicates produced values around 0.7, which is relatively close to 1. 
+This result suggests that these clusters are fairly stable.
+
+#### Using non-default clustering parameters
+
+When calculating bootstrap clusters, `rOpenScPCA::calculate_stability` uses `rOpenScPCA::calculate_clusters` with default parameters. 
+If your original clusters were not calculated with these defaults, you should pass those customized values into this function as well to ensure a fair comparison between your original clusters and the bootstrap clusters.
+
+```{r stability custom parameters, warning=FALSE}
+# Calculate clusters with non-default parameters
+cluster_df_leiden <- calculate_clusters(
+  sce,
+  algorithm = "leiden",
+  resolution = 0.5,
+  nn = 15,
+  seed = params$seed
+)
+
+# Now, pass in the same arguments customizing parameters here
+stability_results_leiden <- calculate_stability(
+  sce,
+  cluster_df_leiden,
+  algorithm = "leiden",
+  resolution = 0.5,
+  nn = 15,
+  seed = params$seed
+)
+
+stability_results_leiden
+```
+
+## Saving clustering results
+
+Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv`), or you can add the results into your scRNA-seq object.
+
+_A word of caution!_
+Objects from the ScPCA Portal already contain a column called `cluster` with results from an automated clustering which was _not evaluated_ or parameterized according to the dataset. 
+To avoid ambiguity with these new clustering results, you may want to rename columns.
+
+The subsequent examples will show you how to save the `cluster_results` data frame to a given object, as an example.
+We'll begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results` data frame before adding them to the given object.
+
+```{r rename columns}
+# First, rename columns in `cluster_results` to avoid ambiguity
+cluster_results_renamed <- cluster_results |>
+  # add the prefix openscpca_ to all columns
+  dplyr::rename_with(~ paste0("ropenscpca_", .x))
+
+cluster_results_renamed
+```
+
+### Saving results to an SCE object
+
+Follow these steps save the clustering results to an SCE object's `colData`:
+
+```{r add to sce}
+# First, extract the colData into a separate data frame for manipulation
+sce_coldata <- colData(sce) |>
+  as.data.frame()
+
+# Third, add columns as desired using the barcodes and cell_id columns to join
+sce_coldata <- sce_coldata |>
+  dplyr::left_join(
+    cluster_results_renamed,
+    by = c("barcodes" = "ropenscpca_cell_id")
+  )
+
+# Finally, add the modified colData back into the SCE
+# Use the row.names argument to ensure rownames (barcodes) are retained; don't forget this step!
+colData(sce) <- DataFrame(
+  sce_coldata,
+  row.names = sce_coldata$barcodes
+)
+
+# Print updated colData slot
+colData(sce)
+```
+
+### Saving results to a Seurat object
+
+Follow these steps save the clustering results to a Seurat object's cell metadata:
+
+```{r add to seurat}
+# Add the cluster result data frame to the Seurat object metadata
+seurat_obj <- AddMetaData(seurat_obj, cluster_results_openscpca)
+
+# Print the cell metadata
+seurat_obj@meta.data
+```
 

### 2024-11-11 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
WIP: much progress. Code basically complete and most text complete

@@
 
-- Perform clustering on an SCE object or Seurat object
-- Add clustering results to the object
+- Perform clustering on an SCE or Seurat object
 - Evaluate clustering results using several metrics including:
@@
   - Cluster stability, as measured with the Adjusted Rand index
+- Save clustering results to an SCE or Seurat object
 
-This notebook will use the `SCPCS000001` sample for demonstration and assumes this sample is present in the `OpenScPCA-analysis/data/current` directory.
-Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about downloading OpenScPCA data.
+This notebook will demonstrate use of `rOpenScPCA` with the `SCPCS000001` sample from project `SCPCP000001`, which is assumed present in the `OpenScPCA-analysis/data/current/SCPCP000001` directory.
+Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about obtaining ScPCA data.
 
@@
 
-#### Paths
+### Paths
 
@@
 
-#### Input and output files
-
-Set paths to input and output directories and files in the chunk below.
+### Input and output files
 
@@
 input_sce_file <- file.path(data_dir, "SCPCP000001", "SCPCS000001", "SCPCL000001_processed.rds")
-
-# Path to output file where clustering results will be saved
-cluster_tsv <- file.path(module_base, "results", "SCPCS000001-cluster-results.tsv")
 ```
@@
 
-```{r readrds}
+```{r read data}
+# Read the SCE file
 sce <- readRDS(input_sce_file)
 
-# Convert to Seurat object for demonstration
+# Convert to a Seurat object for demonstration
 seurat_obj <- as.Seurat(
@@
 
-### Clustering from an SCE object
+Default use of this function is shown below.
+
+### Clustering on an SCE object
 
@@
 
-We also strongly recommend providing a random sed with the `seed` argument to ensure reproducibility.
+We also strongly recommend providing a random seed with the `seed` argument to ensure reproducibility.
 
@@
 
-## Clustering a Seurat object
+The following chunk shows an example of performing clustering with several non-default parameters:
+
+```{r cluster sce nondefault}
+# Calculate clusters with non-default parameters
+cluster_results <- calculate_clusters(
+  sce,
+  algorithm = "leiden",
+  nn = 15,
+  objective_function = "modularity",
+  seed = params$seed
+)
+
+# Print resulting table
+cluster_results
+```
+
+### Clustering on a Seurat object
+
+When performing clustering on a Seurat object, `rOpenScPCA::calculate_clusters` assumes that the principal components are stored in a reduced dimension slot named `"pca"`.
+If your Seurat has a different name for the principal components, you can provide the name with the argument `pc_name`.
+In this example, because we created the Seurat object from the SCE object, its principal components are named `"PCA"`, so we will provide that information.
+
+We also strongly recommend providing a random seed with the `seed` argument to ensure reproducibility.
 
@@
   seurat_obj,
-  seed = params$seed,
   # Since we converted this from an SCE object, its PC name is `"PCA"`
-  pc_name = "PCA"
+  pc_name = "PCA",
+  seed = params$seed
 )
@@
 
-For this section, we'll just use the SCE object.
+This section demonstrates how to use several functions for evaluating cluster quality and reliability. 
+All examples will use the SCE object for demonstration, but a Seurat object can be provided instead and will provide the same results.
+Again, as needed, you may need to specify the `pc_name` argument if the principal components do not use the object type's default name (`"PCA"` for SCE and `"pca"` for Seurat).
 
@@
 
-We'll use the function `rOpenScPCA::calculate_silhouette` to calculate the silhouette width, and then visualize the results.
-This function assumes columns called `cell_id` and `cluster`.
+We'll use the function `rOpenScPCA::calculate_silhouette` to calculate the silhouette width, and then we'll visualize the results.
+
+This function takes the following arguments:
+- An SCE or Seurat object that contains principal components
+- A data frame with, at least, columns representing unique cell ids and cluster assignments.
+  - By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
+  - You can override these defaults using the arguments `cell_id_col` and `cluster_col`
+- Optionally, `pc_name` if your PCA matrix does not have the default name for the object type (`"PCA"` for SCE and `"pca"` for Seurat)
+
+This function will return the inputted data frame with two additional columns:
+- `silhouette_width`: The calculated silhouette width for the cell
+- `other`: The closet cluster to the cell besides the cluster to which it belongs, as used in the silhouette width calculation
 
@@
   sce,
-  cluster_results,
-  cluster_col = "openscpca_cluster"
+  cluster_results
 )
@@
 
-```{r plot silhouette densit}
-ave_silhouette_width <- mean(silhouette_results$silhouette_width)
+```{r plot silhouette density}
+# Calculate mean silhouette width
+mean_silhouette_width <- mean(silhouette_results$silhouette_width)
+
+# Visualize values as a density plot, showing the mean value as a vertical line
 ggplot(silhouette_results) +
@@
   geom_density(fill = "lightblue") +
-  geom_vline(xintercept = ave_silhouette_width, color = "red", size = 1)
+  geom_vline(xintercept = mean_silhouette_width, color = "red", size = 1)
 ```
 
-For most clusters, it is below 0.25, and for cluster 5 it is actually negative.
-We also see it is higher for clusters 11 and 12, reaching up to about 0.5.
+Looking at the full distribution of silhouette width values, we see most are positive which is encouraging, but the values are mostly below 0.5, and the mean is around 0.2 which is somewhat low.
+Silhouette width therefore suggests that, while these clusters do have some structure, they are not very tight.
 
-Overall, these values suggest that clustering is not fantastic and many cells may not well matched. 
-We can explore this further by visualizing distributions for each cluster:
+We can also explore the silhouette width results on a per-cluster basis:
 
@@
   geom_jitter(size = 0.75, alpha = 0.25) +
+  # show the mean in value
   stat_summary(color = "red")
@@
 
-Cluster 5 looks particularly rough, but cluster 11 seems the most reliable.
+This visualization shows us that cluster 5 in particular has the lowest silhouette width values, and clusters 11 and 12 have the highest.
 
@@
 We'll use the function `rOpenScPCA::calculate_purity` to calculate the neighborhood purity, and then visualize the results.
-This function requires 
-This function assumes columns called `cell_id` and `cluster`.
+
+This function takes the following arguments:
+- An SCE or Seurat object that contains principal components
+- A data frame with, at least, columns representing unique cell ids and cluster assignments.
+  - By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
+  - You can override these defaults using the arguments `cell_id_col` and `cluster_col`
+- Optionally, `pc_name` if your PCA matrix does not have the default name for the object type (`"PCA"` for SCE and `"pca"` for Seurat)
+
+This function will return the inputted data frame with two additional columns:
+- `purity`: The neighborhood purity for the cell
+- `maximum`: The cluster with the highest proportion of observations neighboring the cell
 
@@
 ```{r plot purity density}
-ave_purity <- mean(purity_results$purity)
+mean_purity <- mean(purity_results$purity)
 
+# Visualize values as a density plot, showing the mean value as a vertical line
 ggplot(purity_results) +
@@
   geom_density(fill = "lightblue") +
-  geom_vline(xintercept = ave_purity, color = "red", size = 1)
+  geom_vline(xintercept = mean_purity, color = "red", size = 1)
 ```
-Not bad! Let's look at individual clusters:
+This metric shows a relatively high average neighborhood purity, suggesting that neighboring cells tend to belong to the same cluster.
+
+We can again visualize this on a per-cluster basis:
 
@@
   geom_jitter(size = 0.75, alpha = 0.25) +
+  # show the mean in value
   stat_summary(color = "red")
 ```
-With this measure, clusters 1, 3, and 6 all have average values below 0.75, but the rest are high - in particular clusters 11, 12, and 13.
+
+With this view, we see that clusters 1, 3, and 6 all have average values below 0.75, but the rest are higher, in particular clusters 11, 12, and 13.
+It's worth noting that both silhouette width and neighborhood purity found that clusters 11 and 12 were among the tightest clusters.
 
@@
 
+This function takes the following arguments:
+- An SCE or Seurat object that contains principal components
+- A data frame with, at least, columns representing unique cell ids and cluster assignments.
+  - By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
+  - You can override these defaults using the arguments `cell_id_col` and `cluster_col`
+- Optionally, `pc_name` if your PCA matrix does not have the default name for the object type (`"PCA"` for SCE and `"pca"` for Seurat)
+TODO UPDATE HERE!!!!!!!!!!!!!!!!!!!!
+
 - By default, this function performs 20 bootstrap replicates, but this can be customized using the argument `replicates`.
 - This function will return a data frame with columns `replicate`, `ari`, and additional columns for the clustering parameters used when calculating bootstrap clusters.
+
+Since this function uses random sampling when performing bootstrapping, it's important to pass in a random seed for reproducibility.
 

### 2024-11-12 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
final touchups to complete the first draft of this notebook, and left some TODO comments to come back to

@@
 
-This notebook will demonstrate use of `rOpenScPCA` with the `SCPCS000001` sample from project `SCPCP000001`, which is assumed present in the `OpenScPCA-analysis/data/current/SCPCP000001` directory.
+To demonstrate these tasks, this notebook will use the sample `SCPCS000001` from project `SCPCP000001`, which is assumed present in the `OpenScPCA-analysis/data/current/SCPCP000001` directory.
 Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about obtaining ScPCA data.
@@
 
-- Louvain algorithm
-- Jaccard weighting
-- 10 nearest neighbors
-- A resolution parameter of 1
+* Louvain algorithm
+* Jaccard weighting
+* 10 nearest neighbors
+* A resolution parameter of 1
 
@@
 
-- The `algorithm` can be one of:
-  - `"louvain"`, `"walktrap"`, or `"leiden"`
-- The `weighting` can be one of:
-  - `"jaccard"`, `"rank"`, or `"number"`
-- The nearest neighbors parameter can be customized with the `nn` argument
-- The resolution parameter can be customized with the `resolution` argument
-  - This parameter is only used by Louvain and Leiden algorithms
-- If the Leiden algorithm is used, its default objective function parameter will be `"CPM"`, but you can also set  `objective_function = "modularity"` instead.
-- You can provide additional parameters as a list to the `cluster_args` argument.
-  - Please refer to the [`igraph` documentation](https://igraph.org/r/html/latest) to learn more about what additional parameters can be provided to each clustering algorithm.
-  - Note that `cluster_args` only accepts single-length arguments (no vectors or lists).
+* The `algorithm` can be one of:
+  * `"louvain"`, `"walktrap"`, or `"leiden"`
+* The `weighting` can be one of:
+  * `"jaccard"`, `"rank"`, or `"number"`
+* The nearest neighbors parameter can be customized with the `nn` argument
+* The resolution parameter can be customized with the `resolution` argument
+  * This parameter is only used by Louvain and Leiden algorithms
+* If the Leiden algorithm is used, its default objective function parameter will be `"CPM"`, but you can also set  `objective_function = "modularity"` instead.
+* You can provide additional parameters as a list to the `cluster_args` argument.
+  * Please refer to the [`igraph` documentation](https://igraph.org/r/html/latest) to learn more about what additional parameters can be provided to each clustering algorithm.
+  * Note that `cluster_args` only accepts single-length arguments (no vectors or lists).
 
@@
 
-- `cell_id`: Unique cell identifiers, e.g. cell barcodes
-- `cluster`: A factor column with the cluster identities
-- There will be one column for each clustering parameter used
-
-Default use of this function is shown below.
+* `cell_id`: Unique cell identifiers, e.g. cell barcodes
+* `cluster`: A factor column with the cluster identities
+* There will be one column for each clustering parameter used
 
@@
 This section demonstrates how to use several functions for evaluating cluster quality and reliability. 
-All examples will use the SCE object for demonstration, but a Seurat object can be provided instead and will provide the same results.
-Again, as needed, you may need to specify the `pc_name` argument if the principal components do not use the object type's default name (`"PCA"` for SCE and `"pca"` for Seurat).
+All examples will use the SCE object for demonstration, but a Seurat object can be provided instead.
+
+All functions presented in this section take the following required arguments:
+
+* An SCE or Seurat object that contains principal components
+  * If the PCA matrix name is not the default for the object type (`"PCA"` for SCE and `"pca"` for Seurat), you can use `pc_name` to specify its name.
+* A data frame with, at least, columns representing unique cell ids and cluster assignments.
+  * By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
+  * You can override these defaults using the arguments `cell_id_col` and `cluster_col`
 
@@
 
-This function takes the following arguments:
-- An SCE or Seurat object that contains principal components
-- A data frame with, at least, columns representing unique cell ids and cluster assignments.
-  - By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
-  - You can override these defaults using the arguments `cell_id_col` and `cluster_col`
-- Optionally, `pc_name` if your PCA matrix does not have the default name for the object type (`"PCA"` for SCE and `"pca"` for Seurat)
+This function will return the inputted data frame with two additional columns:
 
-This function will return the inputted data frame with two additional columns:
-- `silhouette_width`: The calculated silhouette width for the cell
-- `other`: The closet cluster to the cell besides the cluster to which it belongs, as used in the silhouette width calculation
+* `silhouette_width`: The calculated silhouette width for the cell
+* `silhouette_other`: The closet cluster to the cell besides the cluster to which it belongs, as used in the silhouette width calculation
 
@@
 
-Looking at the full distribution of silhouette width values, we see most are positive which is encouraging, but the values are mostly below 0.5, and the mean is around 0.2 which is somewhat low.
+Looking at the full distribution of silhouette width values, we see most are positive which is encouraging, but the values are mostly below 0.5, and the mean is less than 0.25.
 Silhouette width therefore suggests that, while these clusters do have some structure, they are not very tight.
@@
   geom_jitter(size = 0.75, alpha = 0.25) +
-  # show the mean in value
+  # show the mean +/- SE in red
   stat_summary(color = "red")
@@
 
-This function takes the following arguments:
-- An SCE or Seurat object that contains principal components
-- A data frame with, at least, columns representing unique cell ids and cluster assignments.
-  - By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
-  - You can override these defaults using the arguments `cell_id_col` and `cluster_col`
-- Optionally, `pc_name` if your PCA matrix does not have the default name for the object type (`"PCA"` for SCE and `"pca"` for Seurat)
+This function will return the inputted data frame with two additional columns:
 
-This function will return the inputted data frame with two additional columns:
-- `purity`: The neighborhood purity for the cell
-- `maximum`: The cluster with the highest proportion of observations neighboring the cell
+* `purity`: The neighborhood purity for the cell
+* `maximum_neighbor`: The cluster with the highest proportion of observations neighboring the cell
 
@@
 
-With this view, we see that clusters 1, 3, and 6 all have average values below 0.75, but the rest are higher, in particular clusters 11, 12, and 13.
+In this view, we see that clusters 1, 3, and 6 all have average values below 0.75, but the rest are higher, in particular clusters 11, 12, and 13.
 It's worth noting that both silhouette width and neighborhood purity found that clusters 11 and 12 were among the tightest clusters.
@@
 
-- A value of 1 indicates they are completely overlapping
-- A value of -1 indicates they are completely distinct
-- A value of 0 indicates a random relationship
+* A value of 1 indicates they are completely overlapping
+* A value of -1 indicates they are completely distinct
+* A value of 0 indicates a random relationship
 
@@
 We'll use the function `rOpenScPCA::calculate_stability` to calculate the stability of clusters using a bootstrapping approach, and then visualize the results.
+By default, this function performs 20 bootstrap replicates, but this can be customized using the argument `replicates`.
 
-This function takes the following arguments:
-- An SCE or Seurat object that contains principal components
-- A data frame with, at least, columns representing unique cell ids and cluster assignments.
-  - By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
-  - You can override these defaults using the arguments `cell_id_col` and `cluster_col`
-- Optionally, `pc_name` if your PCA matrix does not have the default name for the object type (`"PCA"` for SCE and `"pca"` for Seurat)
-TODO UPDATE HERE!!!!!!!!!!!!!!!!!!!!
+This function will return a data frame with columns `replicate`, `ari`, and additional columns for the clustering parameters used when calculating bootstrap clusters.
 
-- By default, this function performs 20 bootstrap replicates, but this can be customized using the argument `replicates`.
-- This function will return a data frame with columns `replicate`, `ari`, and additional columns for the clustering parameters used when calculating bootstrap clusters.
-
-Since this function uses random sampling when performing bootstrapping, it's important to pass in a random seed for reproducibility.
+Since this function uses random sampling when performing bootstrapping, it's also important to pass in a random seed for reproducibility with the argument `seed`.
 
@@
   sce,
-  cluster_results,
+  cluster_results$cluster, # TODO!!!!! Update once the rOpenScPCA hash is updated in this branch
   seed = params$seed
@@
 
-All bootstrap replicates produced values around 0.7, which is relatively close to 1. 
-This result suggests that these clusters are fairly stable.
+All bootstrap replicates produced values around 0.7, which is relatively high and close to 1, which suggests that the these clusters are stable and robust to starting conditions.
 
@@
   sce,
-  cluster_df_leiden,
+  cluster_df_leiden$cluster, # TODO!!!!! Update once the rOpenScPCA hash is updated in this branch
   algorithm = "leiden",
@@
 Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv`), or you can add the results into your scRNA-seq object.
+The subsequent examples will show you, as an example, how to save the `cluster_results` data frame to a given object.
 
 _A word of caution!_
-Objects from the ScPCA Portal already contain a column called `cluster` with results from an automated clustering which was _not evaluated_ or parameterized according to the dataset. 
-To avoid ambiguity with these new clustering results, you may want to rename columns.
+Objects from the ScPCA Portal already contain a column called `cluster` with results from an automated clustering.
+These automatic clusters were not evaluated, and their parameters were not optimized for any given library.
 
-The subsequent examples will show you how to save the `cluster_results` data frame to a given object, as an example.
-We'll begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results` data frame before adding them to the given object.
+To avoid ambiguity between the existing and new clustering results, you may want to rename incoming columns.
+We'll therefore begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results` data frame before adding them to a given object.
 
@@
 
-Follow these steps save the clustering results to an SCE object's `colData`:
+Follow these steps to save the clustering results to an SCE object's `colData`:
 
@@
 # Add the cluster result data frame to the Seurat object metadata
-seurat_obj <- AddMetaData(seurat_obj, cluster_results_openscpca)
+seurat_obj <- AddMetaData(seurat_obj, cluster_results_renamed)
 

### 2024-11-12 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
bump ropenscpca for calculate_stability usage

@@
   geom_density(fill = "lightblue") +
-  geom_vline(xintercept = mean_silhouette_width, color = "red", size = 1)
+  geom_vline(xintercept = mean_silhouette_width, color = "red", linewidth = 1)
 ```
@@
   geom_density(fill = "lightblue") +
-  geom_vline(xintercept = mean_purity, color = "red", size = 1)
+  geom_vline(xintercept = mean_purity, color = "red", linewidth = 1)
 ```
@@
   sce,
-  cluster_results$cluster, # TODO!!!!! Update once the rOpenScPCA hash is updated in this branch
+  cluster_results,
   seed = params$seed
@@
   geom_density(fill = "lightblue") +
-  geom_rug(color = "red", size = 1)
+  geom_rug(color = "red", linewidth = 1)
 ```
@@
   sce,
-  cluster_df_leiden$cluster, # TODO!!!!! Update once the rOpenScPCA hash is updated in this branch
+  cluster_df_leiden,
   algorithm = "leiden",

### 2024-11-12 analyses/cell-type-wilms-tumor-06/notebook_template/05_copykat_exploration.Rmd

@@
 params:
-  sample_id: "SCPCS000194"
+  sample_id: "SCPCS000179"
   seed: 12345
@@
 ```{r}
-
 result_dir <- file.path(module_base, "results", params$sample_id)
@@
 full_ck_result_paths <- list()
-for(ref_value in c("ref", "noref")){
-  for(distance_value in c("euclidean", "spearman")){
-    predictions_file <- glue::glue("05_copykat_", {params$sample_id}, "_",ref_value,"_distance-", distance_value, "_copykat_prediction.txt")
-    full_ck_result_file <- glue::glue("05_copykat_", {params$sample_id}, "_",ref_value,"_distance-", distance_value, "_copykat_CNA_results.txt")
+for (ref_value in c("ref", "noref")) {
+  for (distance_value in c("euclidean", "spearman")) {
+    predictions_file <- glue::glue(
+      "05_copykat_",
+      {
+        params$sample_id
+      },
+      "_",
+      ref_value,
+      "_distance-",
+      distance_value,
+      "_copykat_prediction.txt"
+    )
+    full_ck_result_file <- glue::glue(
+      "05_copykat_",
+      {
+        params$sample_id
+      },
+      "_",
+      ref_value,
+      "_distance-",
+      distance_value,
+      "_copykat_CNA_results.txt"
+    )
 
@@
 # open the processed rds object
-srat <- readRDS(file.path(result_dir, paste0("02b-fetal_kidney_label-transfer_", params$sample_id,".Rds")))
+srat <- readRDS(file.path(result_dir, paste0("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")))
 DefaultAssay(srat) <- "SCT"
@@
 
-![](`r file.path(result_dir, "05_copyKAT", "noref", "euclidean", glue::glue("05_copykat_", params$sample_id, "_noref_distance-euclidean_copykat_heatmap.png"))`)
+![](`r file.path(result_dir, "05_copyKAT", "noref", "euclidean", glue::glue("05_copykat_", params$sample_id, "_noref_distance-euclidean_copykat_heatmap.jpg"))`)
 
@@
 
-![](`r file.path(result_dir, "05_copyKAT", "noref", "spearman", glue::glue("05_copykat_", params$sample_id, "_noref_distance-spearman_copykat_heatmap.png"))`)
+![](`r file.path(result_dir, "05_copyKAT", "noref", "spearman", glue::glue("05_copykat_", params$sample_id, "_noref_distance-spearman_copykat_heatmap.jpg"))`)
 
@@
 
-![](`r file.path(result_dir, "05_copyKAT", "ref", "euclidean", glue::glue("05_copykat_", params$sample_id, "_ref_distance-euclidean_copykat_heatmap.png"))`)
+![](`r file.path(result_dir, "05_copyKAT", "ref", "euclidean", glue::glue("05_copykat_", params$sample_id, "_ref_distance-euclidean_copykat_heatmap.jpg"))`)
 
@@
 
-![](`r file.path(result_dir, "05_copyKAT", "ref", "spearman", glue::glue("05_copykat_", params$sample_id, "_ref_distance-spearman_copykat_heatmap.png"))`)
+![](`r file.path(result_dir, "05_copyKAT", "ref", "spearman", glue::glue("05_copykat_", params$sample_id, "_ref_distance-spearman_copykat_heatmap.jpg"))`)
 
@@
   as.data.frame() |>
-  rownames_to_column('barcodes')
+  rownames_to_column("barcodes")
 
@@
   dplyr::left_join(ck_results_df, by = c("barcodes" = "cell.names"))
-
 ```
@@
   dplyr::filter(!is.na(chrom))
-
 ```

### 2024-11-12 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd

@@
   scpca_project_id: "SCPCP000006"
-  sample_id: "SCPCS000184"
+  sample_id: "SCPCS000179"
   seed: 12345
@@
   #### CNV score
-  meta <- infercnv_obj@meta.data |>
-    mutate(has_cnv_score = rowSums(meta[, grepl("has_cnv_chr", colnames(meta))])) |>
-    mutate(has_cnv_score = case_when(
+  meta <- infercnv_obj@meta.data
+  meta <- meta |>
+    dplyr::mutate(has_cnv_score = rowSums(meta[, grepl("has_cnv_chr", colnames(meta))])) |>
+    dplyr::mutate(has_cnv_score = dplyr::case_when(
       has_cnv_score > cnv_threshold ~ "CNV",

### 2024-11-13 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
debugging to be able to run in CI if needed: We need to only loop over available samples, and not assume that all compartments are present when printing the count table

@@
   predicted.score_thr: 0.85
+  sample_ids: ""
 output: 
@@
 sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
+metadata <- read_tsv(sample_metadata_file)
 
-metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE) |>
-  dplyr::filter(seq_unit != "spot")
+# Determine which sample ids to loop over, which differs if we are testing
+if (params$sample_ids == "") {
+  sample_ids <- metadata |>
+    dplyr::filter(seq_unit != "spot") |>
+    dplyr::pull(scpca_sample_id) |>
+    unique()
+} else {
+  sample_ids <- stringr::str_split_1(params$sample_ids, ",")
+  stopifnot(
+    "Could not determine the sample_ids from the provided param value." =
+      all(sample_ids %in% metadata$scpca_sample_id)
+  )
+}
 
 # Create a data frames of all annotations
-cell_type_df <- metadata$scpca_sample_id |>
+cell_type_df <- sample_ids |>
   purrr::map(
@@
   # make the table wide and use a value of 0 if a count is missing
-  tidyr::pivot_wider(names_from = compartment, values_from = n, values_fill = 0) |>
-  # order the columns how we want to show them
-  dplyr::select(organ, fetal_nephron, stroma, endothelium, immune)
+  tidyr::pivot_wider(names_from = compartment, values_from = n, values_fill = 0)
 
@@
   # make the table wide and use a value of 0 if a count is missing
-  tidyr::pivot_wider(names_from = compartment, values_from = n, values_fill = 0) |>
-  # order the columns how we want to show them
-  dplyr::select(sample_id, fetal_nephron, stroma, endothelium, immune)
+  tidyr::pivot_wider(names_from = compartment, values_from = n, values_fill = 0)
 

### 2024-11-13 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
debugging to be able to run in CI: We need to only loop over available samples

--- 2024-11-06 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
@@
   cnv_threshold: 0
+  sample_ids: ""
 output: 
@@
 
-We extracted from the pre-processed and labeled `Seurat` object from `results/{sample_id}/06_infercnv_HMM-i3_SCPCS000169_reference-both.rds`.
+We extracted from the pre-processed and labeled `Seurat` object from `results/{sample_id}/06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds`.
 the following information per cell:
@@
 
-- the proportion of estimated CNV per chromosome, {i} in 1 to 22 `proportion_cnv_chr{i}`
+- the proportion of estimated CNV per chromosome, `{i}` in 1 to 22 `proportion_cnv_chr{i}`
 
-- the `counts`of [markers genes](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-wilms-tumor-06#the-table-celltype_metadatacsv-contains-the-following-column-and-information)
+- the `counts` of [markers genes](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-wilms-tumor-06#the-table-celltype_metadatacsv-contains-the-following-column-and-information)
 
@@
 sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
+metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
 
-metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
+# Determine which sample ids to loop over, which differs if we are testing
+if (params$sample_ids == "") {
   sample_ids <- metadata |>
@@
     unique()
+} else {
+  sample_ids <- stringr::str_split_1(params$sample_ids, ",")
+  stopifnot(
+    "Could not determine the sample_ids from the provided param value." =
+      all(sample_ids %in% metadata$scpca_sample_id)
+  )
+}
 

### 2024-11-13 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
response to reviews: add parentheses for functions, just use backticks for pca names, and add missing wording

@@
 title: "Performing and evaluating graph-based clustering with rOpenScPCA"
-date: "`r Sys.Date()`"
+date: `r Sys.Date()`
 author: Data Lab
@@
 
-This section will show how to perform clustering with the function `rOpenScPCA::calculate_clusters`.
+This section will show how to perform clustering with the function `rOpenScPCA::calculate_clusters()`.
 
@@
 * The `algorithm` can be one of:
-  * `"louvain"`, `"walktrap"`, or `"leiden"`
+  * `louvain`, `walktrap`, or `leiden`
 * The `weighting` can be one of:
-  * `"jaccard"`, `"rank"`, or `"number"`
+  * `jaccard`, `rank`, or `number`
 * The nearest neighbors parameter can be customized with the `nn` argument
@@
   * This parameter is only used by Louvain and Leiden algorithms
-* If the Leiden algorithm is used, its default objective function parameter will be `"CPM"`, but you can also set  `objective_function = "modularity"` instead.
+* If the Leiden algorithm is used, its default objective function parameter will be `CPM`, but you can also set  `objective_function = "modularity` instead.
 * You can provide additional parameters as a list to the `cluster_args` argument.
@@
 
-When performing clustering on an SCE object, `rOpenScPCA::calculate_clusters` assumes that the principal components are stored in a reduced dimension slot named `"PCA"`.
-If your 
-The function assumes the PC name is `"PCA"`. 
-If your SCE has a different name for the principal components, you can provide the name with the argument `pc_name`.
+When performing clustering on an SCE object, `rOpenScPCA::calculate_clusters()` assumes that the principal components are stored in a reduced dimension slot named `PCA`.
+The function assumes the PC name is `PCA`. 
+If your SCE object has a different name for the principal components, you can provide the name with the argument `pc_name`.
 
@@
 
-When performing clustering on a Seurat object, `rOpenScPCA::calculate_clusters` assumes that the principal components are stored in a reduced dimension slot named `"pca"`.
-If your Seurat has a different name for the principal components, you can provide the name with the argument `pc_name`.
-In this example, because we created the Seurat object from the SCE object, its principal components are named `"PCA"`, so we will provide that information.
+When performing clustering on a Seurat object, `rOpenScPCA::calculate_clusters()` assumes that the principal components are stored in a reduced dimension slot named `pca`.
+If your Seurat object has a different name for the principal components, you can provide the name with the argument `pc_name`.
+In this example, because we created the Seurat object from the SCE object, its principal components are named `PCA`, so we will provide that information.
 
@@
   seurat_obj,
-  # Since we converted this from an SCE object, its PC name is `"PCA"`
+  # Since we converted this from an SCE object, its PC name is `PCA`
   pc_name = "PCA",
@@
 * An SCE or Seurat object that contains principal components
-  * If the PCA matrix name is not the default for the object type (`"PCA"` for SCE and `"pca"` for Seurat), you can use `pc_name` to specify its name.
+  * If the PCA matrix name is not the default for the object type (`PCA` for SCE and `pca` for Seurat), you can use `pc_name` to specify its name.
 * A data frame with, at least, columns representing unique cell ids and cluster assignments.
-  * By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
+  * By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters()`
   * You can override these defaults using the arguments `cell_id_col` and `cluster_col`
@@
 
-We'll use the function `rOpenScPCA::calculate_silhouette` to calculate the silhouette width, and then we'll visualize the results.
+We'll use the function `rOpenScPCA::calculate_silhouette()` to calculate the silhouette width, and then we'll visualize the results.
 
@@
 
-We'll use the function `rOpenScPCA::calculate_purity` to calculate the neighborhood purity, and then visualize the results.
+We'll use the function `rOpenScPCA::calculate_purity()` to calculate the neighborhood purity, and then visualize the results.
 
@@
 
-We'll use the function `rOpenScPCA::calculate_stability` to calculate the stability of clusters using a bootstrapping approach, and then visualize the results.
+We'll use the function `rOpenScPCA::calculate_stability()` to calculate the stability of clusters using a bootstrapping approach, and then visualize the results.
 By default, this function performs 20 bootstrap replicates, but this can be customized using the argument `replicates`.
@@
 
-When calculating bootstrap clusters, `rOpenScPCA::calculate_stability` uses `rOpenScPCA::calculate_clusters` with default parameters. 
+When calculating bootstrap clusters, `rOpenScPCA::calculate_stability()` uses `rOpenScPCA::calculate_clusters()` with default parameters. 
 If your original clusters were not calculated with these defaults, you should pass those customized values into this function as well to ensure a fair comparison between your original clusters and the bootstrap clusters.
@@
 
-Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv`), or you can add the results into your scRNA-seq object.
+Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv()`), or you can add the results into your scRNA-seq object.
 The subsequent examples will show you, as an example, how to save the `cluster_results` data frame to a given object.

### 2024-11-13 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
one seed to rule them all, and fix yaml

@@
 title: "Performing and evaluating graph-based clustering with rOpenScPCA"
-date: `r Sys.Date()`
+date: "`r Sys.Date()`"
 author: Data Lab
@@
 theme_set(theme_bw())
+```
+
+### Set the random seed
+
+Because clustering involves random samples, it is important to set the random seed at the top of your analysis script or notebook to ensure reproducibilty.
+
+```{r set seed}
+set.seed(params$seed)
 ```
@@
 
-We also strongly recommend providing a random seed with the `seed` argument to ensure reproducibility.
-
 ```{r cluster sce}
 # Calculate clusters with default parameters
-cluster_results <- calculate_clusters(
-  sce,
-  seed = params$seed
-)
+cluster_results <- calculate_clusters(sce)
 
@@
 
-We also strongly recommend providing a random seed with the `seed` argument to ensure reproducibility.
-
 ```{r cluster seurat}
@@
   # Since we converted this from an SCE object, its PC name is `PCA`
-  pc_name = "PCA",
-  seed = params$seed
+  pc_name = "PCA"
 )

### 2024-11-14 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
found a few more chunks not to run with test data

@@
 - `nbins` is the number of bins for `AddModuleScore` to use. Its default value is 24, which matches Seurat's default value.
-```{r }
+```{r}
 MSigDB_score <- function(object, category, gs_name, name, nbin = 24) {
@@
 ### Cell cycle information
-```{r cell_cycle}
+```{r cell_cycle, eval=!params$testing}
 s.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$s.genes]
@@
 
-```{r fig.width=20, fig.height=4, out.width='100%'}
+```{r fig.width=20, fig.height=4, out.width='100%', eval=!params$testing}
 visualize_metadata(srat, meta = "Phase", group.by = "seurat_clusters")
@@
 
-```{r fig.width=20, fig.height=2.5, out.width='100%'}
+```{r fig.width=20, fig.height=2.5, out.width='100%', eval=!params$testing}
 visualize_metadata(srat, meta = "S.Score", group.by = "seurat_clusters")
@@
 
-```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
+```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%', eval=!params$testing}
 markers <- list(

### 2024-11-14 analyses/cell-type-wilms-tumor-06/00_run_workflow.sh

@@ -133,7 +133,7 @@ if [[ $RUN_EXPLORATORY -eq 1 ]]; then
   # Run notebook template to explore label transfer and clustering for all samples at once
   for score_threshold in 0.5 0.75 0.85 0.95; do
     Rscript -e "rmarkdown::render('${notebook_output_dir}/04_annotation_Across_Samples_exploration.Rmd',
-                    params = list(predicted.score_thr = ${score_threshold}),
+                    params = list(predicted.score_thr = ${score_threshold}, testing = ${IS_CI}),
                     output_format = 'html_document',
                     output_file = '04_annotation_Across_Samples_exploration_predicted.score_threshold_${score_threshold}.html',
                     output_dir = '${notebook_output_dir}')"
@@ -149,6 +149,7 @@ fi
 # Run infercnv for all samples with HMM i3 and using "both" as the reference, where possible
 for sample_dir in ${data_dir}/${project_id}/SCPCS*; do
     sample_id=$(basename $sample_dir)
+    results_dir=results/${sample_id}
 
     # These samples do not have sufficient normal cells to run with a reference in infercnv
     samples_no_reference=("SCPCS000177" "SCPCS000180" "SCPCS000181" "SCPCS000190" "SCPCS000197")
@@ -170,7 +171,7 @@ done
 
 # Render notebook to make draft annotations
 Rscript -e "rmarkdown::render('${notebook_output_dir}/07_combined_annotation_across_samples_exploration.Rmd',
-                        params = list(predicted.celltype.threshold = 0.85, cnv_threshold = 0),
+                        params = list(predicted.celltype.threshold = 0.85, cnv_threshold = 0, testing = ${IS_CI}),
                         output_format = 'html_document',
                         output_file = '07_combined_annotation_across_samples_exploration.html',
                         output_dir = '${notebook_output_dir}')"

### 2024-11-14 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
bug fix for CI; it's ok to skip missing samples if we are testing

--- 2024-11-07 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
@@
   predicted.score_thr: 0.85
+  testing: 0
 output: 
@@
     \(sample_id) {
+      input_file <- file.path(
+        result_dir,
+        sample_id,
+        glue::glue("02b-fetal_kidney_label-transfer_{sample_id}.Rds")
+      )
+
+      # The file may not be present if this is being run in CI, so check first
+      # We should error out if the file does not exist and we are NOT testing
+      if (!file.exists(input_file) & !params$testing) {
+        stop("input RDS file not found.")
+      }
+
       # Read in the Seurat object
-      srat <- readRDS(
-        file.path(result_dir, sample_id, paste0("02b-fetal_kidney_label-transfer_", sample_id, ".Rds"))
-      )
+      srat <- readRDS(input_file)
 
@@
   # order the columns how we want to show them
-  dplyr::select(organ, fetal_nephron, stroma, endothelium, immune)
+  # use any_of in case columns are missing, which will be the case for test data
+  dplyr::select(dplyr::any_of(c("organ", "fetal_nephron", "stroma", "endothelium", "immune")))
 
@@
   # order the columns how we want to show them
-  dplyr::select(sample_id, fetal_nephron, stroma, endothelium, immune)
+  # use any_of in case columns are missing, which will be the case for test data
+  dplyr::select(dplyr::any_of(c("organ", "fetal_nephron", "stroma", "endothelium", "immune")))
 

### 2024-11-14 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
bug fix for CI; it's ok to skip missing samples if we are testing

--- 2024-11-06 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
@@
   cnv_threshold: 0
+  testing: 0
 output: 
@@
       }
+
       input_file <- file.path(
@@
       )
+
+      # The file may not be present if this is being run in CI, so check first
+      # We should error out if the file does not exist and we are NOT testing
+      if (!file.exists(input_file) & !params$testing) {
+        stop("input RDS file not found.")
+      }
 

### 2024-11-14 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd

@@
 sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
+metadata <- read_tsv(sample_metadata_file)
 
-metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE) |>
-  dplyr::filter(seq_unit != "spot")
+sample_ids <- metadata |>
+  dplyr::filter(seq_unit != "spot") |>
+  dplyr::pull(scpca_sample_id) |>
+  unique()
 
 # Create a data frames of all annotations
-cell_type_df <- metadata$scpca_sample_id |>
+cell_type_df <- sample_ids |>
   purrr::map(

### 2024-11-14 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd

@@
 sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
+metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
 
-metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
 sample_ids <- metadata |>

### 2024-11-14 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
revert some of the sample_id code

--- 2024-11-14 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
@@
 sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
-
-metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE) |>
+metadata <- read_tsv(sample_metadata_file) |>
   dplyr::filter(seq_unit != "spot")
@@
 # Create a data frames of all annotations
-cell_type_df <- metadata$scpca_sample_id |>
+cell_type_df <- metadata$scpca_sample_ids |>
   purrr::map(

### 2024-11-14 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
fix logic bug, re-render

@@
 
+# These samples were run with "none" as the reference
+none_reference_samples <- c("SCPCS000177", "SCPCS000180", "SCPCS000181", "SCPCS000190", "SCPCS000197")
+
 # Create a data frames of all annotations
@@
     \(sample_id) {
-      # These samples were run with "none" as the reference
-      none_reference_samples <- c("SCPCS000177", "SCPCS000180", "SCPCS000181", "SCPCS000190", "SCPCS000197")
       if (sample_id %in% none_reference_samples) {
@@
 
-      # The file may not be present if this is being run in CI, so check first
+      # The file may not be present if this is being run in CI, which is ok.
+      # If we are not running in CI and the file doesn't exist, we should error out
       # We should error out if the file does not exist and we are NOT testing
-      if (!file.exists(input_file) & !params$testing) {
-        stop("input RDS file not found.")
+      if (!file.exists(input_file)) {
+        if (params$testing) {
+          return(NULL)
+        } else {
+          stop("Input RDS file does not exist.")
+        }
       }

### 2024-11-14 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
WIP: rearranging notebook

@@
 date: "`r Sys.Date()`"
-author: Data Lab
+author: "Data Lab"
 params:
@@
 
-- Perform clustering on an SCE or Seurat object
-- Evaluate clustering results using several metrics including:
+- Perform clustering
+- Evaluate clustering results using several metrics, including:
   - Silhouette width
   - Neighborhood purity
-  - Cluster stability, as measured with the Adjusted Rand index
-- Save clustering results to an SCE or Seurat object
+  - Cluster stability, as measured with the Adjusted Rand Index
+- Save clustering results
+
+While this notebook does demonstrate how to use individual functions that evaluate a single clustering result, a full clustering evaluation would entail comparison these metrics across different clusterings from different parameterizations.
 
@@
 
-Because clustering involves random samples, it is important to set the random seed at the top of your analysis script or notebook to ensure reproducibilty.
+Because clustering involves random sampling, it is important to set the random seed at the top of your analysis script or notebook to ensure reproducibilty.
 
@@
 
-### Input and output files
+### Input file
 
@@
 
-To begin, we'll read in the `SingleCellExperiment` (SCE) object and convert it to a Seurat object.
-This will allow us to demonstrate `rOpenScPCA` usage with both SCE and Seurat objects.
+To begin, we'll read in the `SingleCellExperiment` (SCE) object.
+We'll also extract its PCA matrix, which we'll use to demonstrate how to calculate and evaluate clusters.
 
@@
 
-# Convert to a Seurat object for demonstration
-seurat_obj <- as.Seurat(
-  sce,
-  counts = "counts",
-  data = "logcounts"
-)
+# Extract the PCA matrix
+pca_matrix <- reducedDim(sce, "PCA")
 ```
@@
 
-By default, this function will calculate clusters using the following parameters:
+This function takes a PCA matrix with rownames representing unique cell ids (e.g., barcodes) as its primary argument.
+By default it will calculate clusters using the following parameters:
 
@@
 
-These parameters can be customized, as follows:
+This function will return a table with the following columns:
+
+* `cell_id`: Unique cell identifiers, obtained from the PCA matrix's row names
+* `cluster`: A factor column with the cluster identities
+* There will be one column for each clustering parameter used
+
+### Clustering with default parameters
+
+```{r cluster sce}
+# Calculate clusters with default parameters
+cluster_results_df <- calculate_clusters(pca_matrix)
+
+# Print resulting table
+cluster_results_df
+```
+
+### Clustering with non-default parameters
+
+Parameters used for clustering can be customized with these arguments:
 
@@
 
-This function will return a table with the following columns:
-
-* `cell_id`: Unique cell identifiers, e.g. cell barcodes
-* `cluster`: A factor column with the cluster identities
-* There will be one column for each clustering parameter used
-
-### Clustering on an SCE object
-
-When performing clustering on an SCE object, `rOpenScPCA::calculate_clusters()` assumes that the principal components are stored in a reduced dimension slot named `PCA`.
-The function assumes the PC name is `PCA`. 
-If your SCE object has a different name for the principal components, you can provide the name with the argument `pc_name`.
-
-```{r cluster sce}
-# Calculate clusters with default parameters
-cluster_results <- calculate_clusters(sce)
-
-# Print resulting table
-cluster_results
-```
-
-The following chunk shows an example of performing clustering with several non-default parameters:
+For example:
 
@@
 # Calculate clusters with non-default parameters
-cluster_results <- calculate_clusters(
-  sce,
+cluster_results_df <- calculate_clusters(
+  pca_matrix,
   algorithm = "leiden",
   nn = 15,
-  objective_function = "modularity",
-  seed = params$seed
+  objective_function = "modularity"
 )
@@
 # Print resulting table
-cluster_results
-```
-
-### Clustering on a Seurat object
-
-When performing clustering on a Seurat object, `rOpenScPCA::calculate_clusters()` assumes that the principal components are stored in a reduced dimension slot named `pca`.
-If your Seurat object has a different name for the principal components, you can provide the name with the argument `pc_name`.
-In this example, because we created the Seurat object from the SCE object, its principal components are named `PCA`, so we will provide that information.
-
-```{r cluster seurat}
-cluster_results <- calculate_clusters(
-  seurat_obj,
-  # Since we converted this from an SCE object, its PC name is `PCA`
-  pc_name = "PCA"
-)
-
-cluster_results
+cluster_results_df
 ```
@@
 This section demonstrates how to use several functions for evaluating cluster quality and reliability. 
-All examples will use the SCE object for demonstration, but a Seurat object can be provided instead.
+It's important to note that a full evaluation of clustering results would compare these metrics across a set of clustering results, with the aim of identifying an optimal parameterization.
 
@@
 
-* An SCE or Seurat object that contains principal components
-  * If the PCA matrix name is not the default for the object type (`PCA` for SCE and `pca` for Seurat), you can use `pc_name` to specify its name.
-* A data frame with, at least, columns representing unique cell ids and cluster assignments.
+* A PCA matrix with rownames representing unique cell ids (e.g., barcodes)
+* A data frame with, at least, columns representing unique cell ids and cluster assignments
   * By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters()`
@@
 
-We'll use the function `rOpenScPCA::calculate_silhouette()` to calculate the silhouette width, and then we'll visualize the results.
+We'll use the function `rOpenScPCA::calculate_silhouette()` to calculate the silhouette width.
 
@@
 ```{r silhouette}
+# calculate the silhouette width for each cell
 silhouette_results <- calculate_silhouette(
-  sce,
-  cluster_results
+  pca_matrix,
+  cluster_results_df
 )
 
+# print result
 silhouette_results
 ```
-
-```{r plot silhouette density}
-# Calculate mean silhouette width
-mean_silhouette_width <- mean(silhouette_results$silhouette_width)
-
-# Visualize values as a density plot, showing the mean value as a vertical line
-ggplot(silhouette_results) +
-  aes(x = silhouette_width) +
-  geom_density(fill = "lightblue") +
-  geom_vline(xintercept = mean_silhouette_width, color = "red", linewidth = 1)
-```
-
-Looking at the full distribution of silhouette width values, we see most are positive which is encouraging, but the values are mostly below 0.5, and the mean is less than 0.25.
-Silhouette width therefore suggests that, while these clusters do have some structure, they are not very tight.
-
-We can also explore the silhouette width results on a per-cluster basis:
-
-```{r plot silhouette violin}
-ggplot(silhouette_results) +
-  aes(x = cluster, y = silhouette_width) +
-  geom_violin() +
-  geom_jitter(size = 0.75, alpha = 0.25) +
-  # show the mean +/- SE in red
-  stat_summary(color = "red")
-```
-
-This visualization shows us that cluster 5 in particular has the lowest silhouette width values, and clusters 11 and 12 have the highest.
 
@@
 
-We'll use the function `rOpenScPCA::calculate_purity()` to calculate the neighborhood purity, and then visualize the results.
+We'll use the function `rOpenScPCA::calculate_purity()` to calculate the neighborhood purity.
 
@@
 ```{r purity}
+# calculate the neighborhood purity for each cell
 purity_results <- calculate_purity(
-  sce,
-  cluster_results
+  pca_matrix,
+  cluster_results_df
 )
 
+# print result
 purity_results
 ```
-
-```{r plot purity density}
-mean_purity <- mean(purity_results$purity)
-
-# Visualize values as a density plot, showing the mean value as a vertical line
-ggplot(purity_results) +
-  aes(x = purity) +
-  geom_density(fill = "lightblue") +
-  geom_vline(xintercept = mean_purity, color = "red", linewidth = 1)
-```
-This metric shows a relatively high average neighborhood purity, suggesting that neighboring cells tend to belong to the same cluster.
-
-We can again visualize this on a per-cluster basis:
-
-```{r plot purity violin}
-ggplot(purity_results) +
-  aes(x = cluster, y = purity) +
-  geom_violin() +
-  geom_jitter(size = 0.75, alpha = 0.25) +
-  # show the mean in value
-  stat_summary(color = "red")
-```
-
-In this view, we see that clusters 1, 3, and 6 all have average values below 0.75, but the rest are higher, in particular clusters 11, 12, and 13.
-It's worth noting that both silhouette width and neighborhood purity found that clusters 11 and 12 were among the tightest clusters.
 
@@
 
-We'll use the function `rOpenScPCA::calculate_stability()` to calculate the stability of clusters using a bootstrapping approach, and then visualize the results.
+We'll use the function `rOpenScPCA::calculate_stability()` to calculate the cluster stability.
 By default, this function performs 20 bootstrap replicates, but this can be customized using the argument `replicates`.
@@
 
-Since this function uses random sampling when performing bootstrapping, it's also important to pass in a random seed for reproducibility with the argument `seed`.
-
 ```{r stability, warning=FALSE}
+# calculate the stability of clusters
 stability_results <- calculate_stability(
   sce,
-  cluster_results,
-  seed = params$seed
+  cluster_results_df
 )
 
+# print the result
 stability_results
 ```
-
-We can visualize these results, for example, with a density plot with a rug to emphasize individual values:
-
-```{r plot stability}
-ggplot(stability_results) +
-  aes(x = ari) +
-  geom_density(fill = "lightblue") +
-  geom_rug(color = "red", linewidth = 1)
-```
-
-All bootstrap replicates produced values around 0.7, which is relatively high and close to 1, which suggests that the these clusters are stable and robust to starting conditions.
 
@@
 
-```{r stability custom parameters, warning=FALSE}
+```{r stability custom parameters}
 # Calculate clusters with non-default parameters
@@
   resolution = 0.5,
-  nn = 15,
-  seed = params$seed
+  nn = 15
 )
@@
 
+## Working with objects directly
+
+The above examples present how to work with a pca matrix, but you can also pass in an SCE or Seurat object with PCs to use instead.
+When doing so, this package will assume a pc_name, and you will need to override this as needed.
+
+When performing clustering on an SCE object, `rOpenScPCA::calculate_clusters()` assumes that the principal components are stored in a reduced dimension slot named `PCA`.
+The function assumes the PC name is `PCA`. 
+If your SCE object has a different name for the principal components, you can provide the name with the argument `pc_name`.
+
+When performing clustering on a Seurat object, `rOpenScPCA::calculate_clusters()` assumes that the principal components are stored in a reduced dimension slot named `pca`.
+If your Seurat object has a different name for the principal components, you can provide the name with the argument `pc_name`.
+In this example, because we created the Seurat object from the SCE object, its principal components are named `PCA`, so we will provide that information.
+
+```{r}
+# Convert to a Seurat object for demonstration
+seurat_obj <- as.Seurat(
+  sce,
+  counts = "counts",
+  data = "logcounts"
+)
+```
+
+### Clustering on a Seurat object
+
+```{r cluster seurat}
+cluster_results_df <- calculate_clusters(
+  seurat_obj,
+  # Since we converted this from an SCE object, its PC name is `PCA`
+  pc_name = "PCA"
+)
+
+cluster_results_df
+```
+
 ## Saving clustering results
@@
 Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv()`), or you can add the results into your scRNA-seq object.
-The subsequent examples will show you, as an example, how to save the `cluster_results` data frame to a given object.
+The subsequent examples will show you, as an example, how to save the `cluster_results_df` data frame to a given object.
 
@@
 To avoid ambiguity between the existing and new clustering results, you may want to rename incoming columns.
-We'll therefore begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results` data frame before adding them to a given object.
+We'll therefore begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results_df` data frame before adding them to a given object.
 
 ```{r rename columns}
-# First, rename columns in `cluster_results` to avoid ambiguity
-cluster_results_renamed <- cluster_results |>
+# First, rename columns in `cluster_results_df` to avoid ambiguity
+cluster_results_renamed_df <- cluster_results_df |>
   # add the prefix openscpca_ to all columns
@@
 
-cluster_results_renamed
+cluster_results_renamed_df
 ```
@@
   dplyr::left_join(
-    cluster_results_renamed,
+    cluster_results_renamed_df,
     by = c("barcodes" = "ropenscpca_cell_id")
@@
 # Add the cluster result data frame to the Seurat object metadata
-seurat_obj <- AddMetaData(seurat_obj, cluster_results_renamed)
+seurat_obj <- AddMetaData(seurat_obj, cluster_results_renamed_df)
 

### 2024-11-14 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
Continuing notebook reorg, added code for a seurat section for which glmGamPoi is now needed in renv

@@
 To begin, we'll read in the `SingleCellExperiment` (SCE) object.
-We'll also extract its PCA matrix, which we'll use to demonstrate how to calculate and evaluate clusters.
 
@@
 sce <- readRDS(input_sce_file)
+```
 
+We'll also extract its PCA matrix, which we'll use to demonstrate how to calculate and evaluate clusters.
+
+```{r extract pca data}
 # Extract the PCA matrix
 pca_matrix <- reducedDim(sce, "PCA")
+```
+
+If you are working with a Seurat object, you can extract the PCA matrix like this:
+
+```{r convert to seurat}
+# First, convert the SCE to a Seurat object for demonstration
+
+# Here, the PCA matrix has the same name as a
 ```
@@
 
-* A PCA matrix with rownames representing unique cell ids (e.g., barcodes)
+* A PCA matrix with row names representing unique cell ids (e.g., barcodes)
 * A data frame with, at least, columns representing unique cell ids and cluster assignments
@@
 
-## Working with objects directly
+## Evaluating pre-calculated clusters
 
-The above examples present how to work with a pca matrix, but you can also pass in an SCE or Seurat object with PCs to use instead.
-When doing so, this package will assume a pc_name, and you will need to override this as needed.
+Even if you perform clustering with other tools, you can still use the `rOpenScPCA` functions to evaluate your clusters.
+For example, if you are analyzing your data with a Seurat pipeline, you might also calculate clusters with Seurat, and you can evaluate those with `rOpenScPCA`.
+In addition, ScPCA data already contains a column called `cluster` with results from an automated clustering.
 
-When performing clustering on an SCE object, `rOpenScPCA::calculate_clusters()` assumes that the principal components are stored in a reduced dimension slot named `PCA`.
-The function assumes the PC name is `PCA`. 
-If your SCE object has a different name for the principal components, you can provide the name with the argument `pc_name`.
+In this section, we'll present how you can calculate the silhouette width and neighborhood purity from existing cluster assignments.
 
-When performing clustering on a Seurat object, `rOpenScPCA::calculate_clusters()` assumes that the principal components are stored in a reduced dimension slot named `pca`.
-If your Seurat object has a different name for the principal components, you can provide the name with the argument `pc_name`.
-In this example, because we created the Seurat object from the SCE object, its principal components are named `PCA`, so we will provide that information.
+### Evaluating Seurat clusters
 
-```{r}
-# Convert to a Seurat object for demonstration
-seurat_obj <- as.Seurat(
-  sce,
-  counts = "counts",
-  data = "logcounts"
+First, we'll prepare a Seurat object and perform clustering for demonstration:
+
+```{r process seurat object}
+# Convert to a Seurat object
+seurat_obj <- CreateSeuratObject(counts = counts(sce), assay = "RNA")
+
+# Process the object with a Seurat pipeline to obtain clusters
+seurat_obj <- seurat_obj |>
+  SCTransform() |>
+  RunPCA() |>
+  FindNeighbors(dims = 1:30) |>
+  FindClusters()
+
+# Print resulting clusters
+head(seurat_obj$seurat_clusters)
+```
+
+Next, we'll prepare inputs for `rOpenScPCA` functions:
+
+- The PCA matrix
+- A data frame with columns `cell_id` and `cluster`
+
+```{r prepare seurat input}
+# Extract the PCA matrix
+seurat_pca_matrix <- Embeddings(seurat_obj, reduction = "pca")
+
+# Create a data frame
+seurat_cluster_df <- data.frame(
+  cell_id = names(seurat_obj$seurat_clusters),
+  cluster = seurat_obj$seurat_clusters
 )
@@
 
-### Clustering on a Seurat object
+Now, we can run `calculate_silhouette()` and `calculate_purity()`:
 
-```{r cluster seurat}
-cluster_results_df <- calculate_clusters(
-  seurat_obj,
-  # Since we converted this from an SCE object, its PC name is `PCA`
-  pc_name = "PCA"
+```{r seurat silhouette}
+seurat_silhouette_df <- calculate_silhouette(
+  seurat_pca_matrix,
+  seurat_cluster_df
+)
+seurat_silhouette_df
+```
+
+```{r seurat purity}
+seurat_purity_df <- calculate_purity(
+  seurat_pca_matrix,
+  seurat_cluster_df
+)
+seurat_purity_df
+```
+
+### Evaluating SCE clusters
+
+```{r}
+```
+
+* Louvain algorithm
+* Jaccard weighting
+* 20 nearest neighbors
+* Resolution of 1
+
+Seurat parameters:
+* louvain
+* resolution of 0.8
+*
+
+There are a few common circumstances where you might encounter this:
+
+```{r}
+# Print the cluster column from the SCE object
+head(colData(sce)$cluster)
+```
+
+In this section, we'll demonstrate how you can run the `rOpenScPCA` cluster evaluation functions on these clusters.
+
+First, you should prepare a data frame two columns: `cell_id` containing cell barcodes, and `cluster` containing the cluster identities.
+
+```{r}
+scpca_cluster_df <- data.frame(
+  cell_id = colnames(sce),
+  cluster = colData(sce)$cluster
 )
 
-cluster_results_df
+scpca_cluster_df
+```
+
+Then, you can proceed to run evaluation functions using this data frame, and the PCA matrix from the object.
+
+```{r}
+# Calculate silhouette width
+scpca_silhouette_df <- calculate_silhouette(pca_matrix, scpca_cluster_df)
+scpca_silhouette_df
+
+# Calculate neighborhood purity
+scpca_purity_df <- calculate_purity(pca_matrix, scpca_cluster_df)
+scpca_purity_df
+```
+
+When running `calculate_stability()`, it is important to specify the same parameters originally used to build the clusters:
+
+```{r}
+scpca_stability_df <- calculate_stability(
+  pca_matrix,
+  scpca_cluster_df,
+  # provide ScPCA clustering parameters
+  algorithm = "louvain",
+  weighting = "jaccard",
+  nn = 20,
+  resolution = 1
+)
+scpca_stability_df
+```
+
+### Working with objects directly
+
+As presented above, `rOpenScPCA` clustering functions take a PCA matrix with row names representing unique cell ids as their first argument.
+
+Instead of a matrix, you can alternatively pass in an SCE or Seurat object that contains PCs.
+For example:
+
+```{r}
+# Calculate clusters from an SCE object using default parameters
+cluster_results_df <- calculate_clusters(sce)
+```
+
+`rOpenScPCA` assumes that the PCA matrix is named `PCA` in SCE objects, and `pca` in Seurat objects.
+If the PCA matrix you want to use in the object has a different name, you can provide the argument `pc_name`.
+For example:
+
+```{r}
+# First, we'll rename the PCA matrix for demonstration
+reducedDimNames(sce) <- c("PCA_matrix", "UMAP")
+
+# Calculate clusters from an SCE object using default parameters
+cluster_results_df <- calculate_clusters(
+  sce,
+  pc_name = "PCA_matrix"
+)
 ```
@@
 
-Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv()`), or you can add the results into your scRNA-seq object.
+Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv()`), or you can add the results into your SCE or Seurat object.
 The subsequent examples will show you, as an example, how to save the `cluster_results_df` data frame to a given object.

### 2024-11-14 analyses/hello-clusters/components/dependencies.R

@@ -1,2 +1,4 @@
 # R dependencies not captured by `renv`
 # library("missing_package")
+
+library(glmGamPoi)

### 2024-11-14 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
WIP: delete extra text, and dont do stability with seurat

@@
 
-### Evaluating SCE clusters
+We do not recommend using `rOpenScPCA::calculate_stability()` on Seurat clusters due to differences in the underlying clustering approach between Seurat and the `bluster` package which `rOpenScPCA` uses.
 
-```{r}
-```
-
-* Louvain algorithm
-* Jaccard weighting
-* 20 nearest neighbors
-* Resolution of 1
-
-Seurat parameters:
-* louvain
-* resolution of 0.8
-*
-
-There are a few common circumstances where you might encounter this:
-
-```{r}
-# Print the cluster column from the SCE object
-head(colData(sce)$cluster)
-```
-
-In this section, we'll demonstrate how you can run the `rOpenScPCA` cluster evaluation functions on these clusters.
+### Evaluating ScPCA clusters
 
@@
 cluster_results_df <- calculate_clusters(sce)
+
+# Calculate clusters from a Seurat object using default parameters
+cluster_results_df <- calculate_clusters(seurat_obj)
 ```

### 2024-11-15 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
typo

@@
 # Create a data frames of all annotations
-cell_type_df <- metadata$scpca_sample_ids |>
+cell_type_df <- metadata$scpca_sample_id |>
   purrr::map(

### 2024-11-15 analyses/cell-type-wilms-tumor-06/notebook_template/00b_characterize_fetal_kidney_reference_Stewart.Rmd
fix typos in 00b and 02b, and re-render 00b

@@
 You can find more about the human kidney atlas here: https://www.kidneycellatlas.org/ [1]
-The rds data can be download using the download link https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds
+The rds data can be download using the download link: <https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds>.
 The reference was downloaded and created in the script `scripts/prepare-fetal-references.R`.
@@
 
-The input file `r params$fetal_kidney_path` is the output of the `R script` `prepare-fetal-references.R`.
+The input file `r params$fetal_kidney_path` is the output of the script `prepare-fetal-references.R`.
 
@@
 
-We will save the result of the differential expression analysis in results/references/00b_marker_genes_fetal_kidney_Stewart.csv
-Notebook is saved in the `notebook/00-reference` directory
+We will save the result of the differential expression analysis in `results/references/00b_marker_genes_fetal_kidney_Stewart.csv`.
+Notebook is saved in the `notebook/00-reference` directory.
 
@@
 
-We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. 
-DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method. 
-Please check the preprint from Chistoph Hafemeister: https://www.biorxiv.org/content/10.1101/2023.03.28.534443v1
+We run `DElegate::FindAllMarkers2()` to find markers of the different clusters and manually check if they do make sense. 
+`DElegate::FindAllMarkers2()` is an improved version of `Seurat::FindAllMarkers()` based on pseudobulk differential expression method. 
+Please check the preprint from Hafemeister and Halbritter: https://www.biorxiv.org/content/10.1101/2023.03.28.534443v2
 and tool described here: https://github.com/cancerbits/DElegate

### 2024-11-15 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd
fix typos on 06 notebook, but do not regenerate

@@
 `infercnv` currently support two models for HMM-based CNV prediction, what we refer to as the i3 and i6 models.
-These are set in the 'infercnv::run()' as HMM_type='i3' or HMM_type='i6' (i6 is default).
-Each method operates on the 'preliminary infercnv object' which has been processed through the standard inferCNV processing routines, involving subtraction of signal corresponding to 'normal (reference)' cells and smoothing operations.
+These are set in the `infercnv::run()` as HMM_type='i3' or HMM_type='i6' (i6 is default).
+Each method operates on the preliminary `infercnv` object which has been processed through the standard inferCNV processing routines, involving subtraction of signal corresponding to "normal (reference)" cells and smoothing operations.
 
@@
 For a `Seurat` object an `infercnv` object created with the script `06_infercnv.R` using `reference_value` as a reference, the function `Do_CNV_score` calculate a CNV score per cell.
-The score is calculated based on the [biostar discussion](https://www.biostars.org/p/9573777/).
+The score is calculated based on the [this discussion](https://www.biostars.org/p/9573777/).
 The function `Do_CNV_score` returns the `Seurat` object with an additional metadata named `CNV-score_{reference_value}.`
@@
 
-#### Visualize seurat clusters and metadata
+#### Visualize Seurat clusters and metadata
 
-For a Seurat object `object`and a metadata `metadata`, the function `visualize_metadata` will plot `FeaturePlot` and `BarPlot`
+For a Seurat object `object` and a metadata `metadata`, the function `visualize_metadata` will plot `FeaturePlot` and `BarPlot`
 
@@
 
-#### Wrapper function to explore `infercnv` HMM cnv prediction
+#### Wrapper function to explore `infercnv` HMM CNV prediction
 
-The `wrapper_explore_hmm` take as input the `infercnv_obj` generated with `infercnv` HMM cnv predictions.
+The `wrapper_explore_hmm` take as input the `infercnv_obj` generated with `infercnv` HMM CNV predictions.
 The wrapper allows the following steps and plots:
@@
 
-#####  Distribution of CNV estimation in the Wilms tumor copartments
+#####  Distribution of CNV estimation in the Wilms tumor compartments
 
@@
 
-##### DotPlot
+##### Dot Plot
 
-The `Dotplot` representation summarizes the percentage of cells in each compartment with cnv in each of the 22 chromosomes.
+The `Dotplot` representation summarizes the percentage of cells in each compartment with CNV in each of the 22 chromosomes.
 
@@
 
-### Heatmap of infercnv results
+### Heatmap of `infercnv` results
 
 Here we plot the output of `infercnv` as heatmaps of CNV.
-We first look at the png file generated by the `infercnv` function.
+We first look at the PNG file generated by the `infercnv` function.
 We then used the `infercnv object` to look at mean CNV value across compartments (immune, endothelial, stroma and fetal nephron).
@@
 The normal reference should contain as much cell types as possible, in order to minimize false positive CNV.
-In our case, we should take immune and entodethial cells when possible.
+In our case, we should take immune and endothelial cells when possible.
 
@@
 
-We want to calculate a single CNV score and asess if/how it can be use to define cells with CNV versus stable/normal cells. 
-We defined the score as discribed in the [biostar discussion](https://www.biostars.org/p/9573777/).
+We want to calculate a single CNV score and asesss if/how it can be use to define cells with CNV versus stable/normal cells. 
+We defined the score as described in the [biostar discussion](https://www.biostars.org/p/9573777/).
 
@@
 
-We then explore infercnv results generated with immune and endothelial cells as reference, using a [HMM-i3 prediction models](https://github.com/broadinstitute/infercnv/wiki/infercnv-i3-HMM-type).
+We then explore `infercnv` results generated with immune and endothelial cells as reference, using a [HMM-i3 prediction models](https://github.com/broadinstitute/infercnv/wiki/infercnv-i3-HMM-type).
 
@@
 
-and explore the CNV results using the `wrapper_explore_hmm` fucntion.
+and explore the CNV results using the `wrapper_explore_hmm` function.
 
@@
 
-## Comparisons of inter- and intra-patient global cnv score with HMM prediction model
+## Comparisons of inter- and intra-patient global CNV score with HMM prediction model
 
@@
 
-- HMM-i3 with inter-patient endothelial and immunce cells as reference
+- HMM-i3 with inter-patient endothelial and immune cells as reference
 
-- HMM-i3 with intra-patient endothelial and immunce cells as reference
+- HMM-i3 with intra-patient endothelial and immune cells as reference
 
-- HMM-i6 with intra-patient endothelial and immunce cells as reference
+- HMM-i6 with intra-patient endothelial and immune cells as reference
 
@@
 
-   + HMM-i3 with inter-patient endothelial and immunce cells as reference
+   + HMM-i3 with inter-patient endothelial and immune cells as reference
 
-   + HMM-i3 with intra-patient endothelial and immunce cells as reference
+   + HMM-i3 with intra-patient endothelial and immune cells as reference
 
-   + HMM-i6 with intra-patient endothelial and immunce cells as reference
+   + HMM-i6 with intra-patient endothelial and immune cells as reference
 
@@
 
-   + The HMM-i3 model with inter-patient endothelial and immunce cells as reference has the advantage to be usable for all Wilms tumor samples, including the ones with a very low number of immune and/or endothelial cells.
+   + The HMM-i3 model with inter-patient endothelial and immune cells as reference has the advantage to be usable for all Wilms tumor samples, including the ones with a very low number of immune and/or endothelial cells.
 
-   + The HMM-i3 model with intra-patient endothelial and immunce cells as reference seems to be the cleaner, ~fast to run (10 minutes per samples) and is more precise than the HMM-i3 with the inter-patient reference.
+   + The HMM-i3 model with intra-patient endothelial and immune cells as reference seems to be the cleaner, ~fast to run (10 minutes per samples) and is more precise than the HMM-i3 with the inter-patient reference.
 
-   + The HMM-i6 model with intra-patient endothelial and immunce cells as reference is very slow (~2 hours per sample) and couldn't be used for the entire cohorte.
+   + The HMM-i6 model with intra-patient endothelial and immune cells as reference is very slow (~2 hours per sample) and couldn't be used for the entire cohort.
 It is more noisy than the i3 version. However, it could have the potential to detect cancer cells with very low CNV profile.
 
-- Surprisingly, running `infercnv` with emdothelial and immune cells from (i) the same patient or (ii) a set of Wilms tumor patients do not seem to affect drastically the results. 
+- Surprisingly, running `infercnv` with endothelial and immune cells from (i) the same patient or (ii) a set of Wilms tumor patients do not seem to affect drastically the results. 
 Some false positive CNV might occur in every patient due to the inter-patient variability. 

### 2024-11-15 analyses/cell-type-wilms-tumor-06/notebook_template/05_copykat_exploration.Rmd
fix typos on 05 notebook, but do not regenerate

@@
 ---
-title: "Copykat CNV results exploration for `r params$sample_id`"
+title: "CopyKAT CNV results exploration for `r params$sample_id`"
 author: "Maud PLASCHKA"
@@
 The assignment of the aneuploidy/diploidy value might relies on very few CNV and/or an arbitrary threshold. 
-This might be why the assignement of aneuploidy/diploidy values differs between condition (and between runs!!).
+This might be why the assignment of aneuploidy/diploidy values differs between condition (and between runs!!).
 

### 2024-11-15 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
fix 07 typos and table formatting, and regenerate notebook

@@
 
-first   level annotation | second level   annotation | selection of   the cells | marker genes   for validation | cnv validation
--- | -- | -- | -- | --
-normal | endothelial | compartment == "endothelium"     & predicted.score > pred.thr     & cnv_score < cnv.thr | WVF | no   cnv
-normal | immune | compartment == "immune"     & predicted.score > pred.thr     & cnv_score < cnv.thr | PTPRC, CD163, CD68 | no cnv
-normal | kidney | cell_type %in% c("kidney cell",   "kidney epithelial", "podocyte")     & predicted.score > pred.thr     & cnv_score < cnv.thr | CDH1, PODXL, LTL | no cnv
-normal | stroma | compartment == "stroma"     & predicted.score > pred.thr     & cnv_score < cnv.thr | VIM | no cnv
-cancer | stroma | compartment == "stroma"     & cnv_score > cnv.thr | VIM | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
-cancer | blastema | compartment ==   "fetal_nephron"     & cell_type == "mesenchymal cell"     & cnv_score > cnv.thr | CITED1 | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
-cancer | epithelial | compartment ==   "fetal_nephron"     & cell_type != "mesenchymal cell"     & cnv_score > cnv.thr | CDH1 | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
-unknown | - | the rest of the cells | - | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
+| first level annotation | second level annotation | selection of the cells | marker genes for validation | CNV validation |
+| ---------------------- | ----------------------- | ---------------------- | --------------------------- | --------------- |
+| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr  & cnv_score < cnv.thr` | `VWF`| no CNV |
+| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr & cnv_score < cnv.thr` | `PTPRC`, `CD163`, `CD68`| no CNV |
+| normal                 | kidney                  | `cell_type %in% c("kidney cell","kidney epithelial", "podocyte") & predicted.score > pred.thr & cnv_score < cnv.thr` | `CDH1`, `PODXL`, `LTL`| no CNV |
+| normal                 | stroma                  | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr`| `VIM`| no CNV |
+| cancer                 | stroma                  | `compartment == "stroma" & cnv_score > cnv.thr` | `VIM`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| cancer                 | blastema                | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr`  | `CITED1`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| cancer                 | epithelial              | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr`  | `CDH1`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| unknown                | -                       | the rest of the cells  | -  | -|
 
@@
 
-### Global cnv score
+### Global CNV score
 
@@
 
-- _normal_ cells can be observe in all four compartments (`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have cnv.
-We only allow a bit of flexibility in terms of cnv profile for immune and endothelium cells that have a high predicted score.
-Indeed, we know that false positive cnv can be observed in a cell type specific manner.
+- _normal_ cells can be observe in all four compartments (`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV
+We only allow a bit of flexibility in terms of CNV profile for immune and endothelium cells that have a high predicted score.
+Indeed, we know that false positive CNV can be observed in a cell type specific manner.
 
 The threshold used for the `predicted.score` is defined as a parameter of this notebook as `r params$predicted.celltype.threshold`.
-The threshold used for the identification of cnv is also defined in the params of the notebook as `r params$cnv_threshold`.
+The threshold used for the identification of CNV is also defined as the notebook parameter `r params$cnv_threshold`.
 
-- _cancer_ cells are either from the `stroma` or `fetal nephron` compartments and must have at least few cnv.
+- _cancer_ cells are either from the `stroma` or `fetal nephron` compartments and must have at least few CNV
 
@@
 
-- _blastema_,: we defined as _bastema_ every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`.
+- _blastema_,: we defined as _blastema_ every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`.
 We know that these _mesenchymal_ cells are cells from the cap mesenchyme that are not expected to be in a mature kidney.
@@
 
-### Validation cancer versus normal based on the cnv profile
+### Validation cancer versus normal based on the CNV profile
 
@@
 
-#### Immune, _PTPRC_ expression
+#### Immune, `PTPRC` expression
 
@@
 
-#### Endothelium, _VWF_ expression
+#### Endothelium, `VWF` expression
 
@@
 
-#### Stroma, _Vimentin_ expression
+#### Stroma, `Vimentin` expression
 
@@
 
-#### Stroma, _COL6A3_ expression
+#### Stroma, `COL6A3` expression
 
@@
 
-#### Stroma, _THY1_ expression
+#### Stroma, `THY1` expression
 
@@
 
-#### Blastema, _CITED1_ expression
+#### Blastema, `CITED1` expression
 
@@
 
-#### Blastema, _NCAM1_ expression
+#### Blastema, `NCAM1` expression
 
@@
 
-#### stemness marker (blastema and primitive epithelium), _SIX2_ expression
+#### stemness marker (blastema and primitive epithelium), `SIX2` expression
 
@@
 
-#### Epithelium, _CDH1_ expression
+#### Epithelium, `CDH1` expression
 
@@
 
-#### Epithelium, _PODXL_ expression
+#### Epithelium, `PODXL` expression
 
@@
 
-- The heatmaps of cnv proportion and marker genes support our annotations, but signals with some marker genes are very low.
+- The heatmaps of CNV proportion and marker genes support our annotations, but signals with some marker genes are very low.
 Also, there is no universal marker for each entity of Wilms tumor that cover all tumor cells from all patient.
@@
 - In each histology (i.e. epithelial and stroma), the distinction between cancer and non cancer cell is difficult (as expected).
-In this analysis, we suggested to rely on the cnv score to assess the normality of the cell.
+In this analysis, we suggested to rely on the CNV score to assess the normality of the cell.
 Here again, we could try to run differential expression analysis and compare epithelial (resp.
@@
 - In our annotation, we haven't taken into account the favorable/anaplastic status of the sample.
-However, as anaplasia can occur in every (but do not has to) wilms tumor histology, I am not sure how to integrate the information into the annotation.
+However, as anaplasia can occur in every (but do not has to) Wilms tumor histology, I am not sure how to integrate the information into the annotation.
 
-- This notebook could be finally rendered using different parameters, i.e. threshold for the cnv score and predicted score to use.
+- This notebook could be finally rendered using different parameters, i.e. threshold for the CNV score and predicted score to use.
 

### 2024-11-15 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
Finish notebook rearrangement

@@
 ---
-title: "Performing and evaluating graph-based clustering with rOpenScPCA"
+title: "Performing graph-based clustering with rOpenScPCA"
 date: "`r Sys.Date()`"
@@
 
-This notebook provides examples of how to use the `rOpenScPCA` package to:
+This notebook provides examples of how to use functions in `rOpenScPCA` that:
 
-- Perform clustering
-- Evaluate clustering results using several metrics, including:
-  - Silhouette width
-  - Neighborhood purity
-  - Cluster stability, as measured with the Adjusted Rand Index
-- Save clustering results
+* Perform clustering
+* Calculate QC metrics on clustering including:
+  * Silhouette width
+  * Neighborhood purity
+  * Cluster stability, as measured with the Adjusted Rand Index
+* Calculate QC metrics obtained with other tools, such as `Seurat`
+* Save clustering results to an SCE or `Seurat`
 
-While this notebook does demonstrate how to use individual functions that evaluate a single clustering result, a full clustering evaluation would entail comparison these metrics across different clusterings from different parameterizations.
+While this notebook demonstrates how to use individual functions that calculate helpful metrics for evaluating clustering results, a full evaluation would compare these metrics across different clusterings from different parameterizations.
 
-To demonstrate these tasks, this notebook will use the sample `SCPCS000001` from project `SCPCP000001`, which is assumed present in the `OpenScPCA-analysis/data/current/SCPCP000001` directory.
+This notebook will use the sample `SCPCS000001` from project `SCPCP000001`, which is assumed present in the `OpenScPCA-analysis/data/current/SCPCP000001` directory, for all examples.
 Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about obtaining ScPCA data.
@@
   library(dplyr)
-  library(ggplot2)
 })
-
-# Set ggplot theme
-theme_set(theme_bw())
 ```
@@
 
-### Read in data
+## Read in and prepare data
 
 To begin, we'll read in the `SingleCellExperiment` (SCE) object.
+We'll also establish a corresponding processed Seurat object from its raw counts that we'll use for some examples.
 
@@
 
+```{r create process seurat, message = FALSE}
+# Convert to a Seurat object
+seurat_obj <- CreateSeuratObject(counts = counts(sce), assay = "RNA")
+
+# Process the object with a Seurat pipeline to obtain clusters
+seurat_obj <- seurat_obj |>
+  SCTransform() |>
+  RunPCA() |>
+  FindNeighbors() |>
+  FindClusters()
+```
+
 We'll also extract its PCA matrix, which we'll use to demonstrate how to calculate and evaluate clusters.
+We'll show how to do this for both SCE and Seurat objects:
 
 ```{r extract pca data}
-# Extract the PCA matrix
+# Extract the PCA matrix from an SCE object
+# Moving forward, we'll use this version for most examples
 pca_matrix <- reducedDim(sce, "PCA")
-```
 
-If you are working with a Seurat object, you can extract the PCA matrix like this:
-
-```{r convert to seurat}
-# First, convert the SCE to a Seurat object for demonstration
-
-# Here, the PCA matrix has the same name as a
+# Extract the PCA matrix from a Seurat object
+seurat_pca_matrix <- Embeddings(seurat_obj, reduction = "pca")
 ```
@@
   * This parameter is only used by Louvain and Leiden algorithms
-* If the Leiden algorithm is used, its default objective function parameter will be `CPM`, but you can also set  `objective_function = "modularity` instead.
+* If the Leiden algorithm is used, its default objective function parameter will be `CPM`, but you can also set  `objective_function = "modularity"` instead.
 * You can provide additional parameters as a list to the `cluster_args` argument.
@@
 
-## Evaluate clustering results
+## Calculate QC metrics on clusters
 
@@
 stability_results <- calculate_stability(
-  sce,
+  pca_matrix,
   cluster_results_df
@@
 cluster_df_leiden <- calculate_clusters(
-  sce,
+  pca_matrix,
   algorithm = "leiden",
@@
 stability_results_leiden <- calculate_stability(
-  sce,
+  pca_matrix,
   cluster_df_leiden,
@@
 
-## Evaluating pre-calculated clusters
+##  Calculate QC metrics on existing clusters
 
 Even if you perform clustering with other tools, you can still use the `rOpenScPCA` functions to evaluate your clusters.
-For example, if you are analyzing your data with a Seurat pipeline, you might also calculate clusters with Seurat, and you can evaluate those with `rOpenScPCA`.
-In addition, ScPCA data already contains a column called `cluster` with results from an automated clustering.
 
-In this section, we'll present how you can calculate the silhouette width and neighborhood purity from existing cluster assignments.
+In this section, we'll present examples of how you can calculate the silhouette width, neighborhood purity, and cluster stability from existing cluster assignments.
 
@@
 
-First, we'll prepare a Seurat object and perform clustering for demonstration:
+If you are analyzing your data with a Seurat pipeline that includes calculating clusters, you can use `rOpenScPCA` to evaluate them.
+We previously created a Seurat object called `seurat_obj` with clusters:
 
-```{r process seurat object}
-# Convert to a Seurat object
-seurat_obj <- CreateSeuratObject(counts = counts(sce), assay = "RNA")
-
-# Process the object with a Seurat pipeline to obtain clusters
-seurat_obj <- seurat_obj |>
-  SCTransform() |>
-  RunPCA() |>
-  FindNeighbors(dims = 1:30) |>
-  FindClusters()
-
-# Print resulting clusters
+```{r show seurat clusters}
 head(seurat_obj$seurat_clusters)
@@
 
-Next, we'll prepare inputs for `rOpenScPCA` functions:
+To calculate QC metrics on these clusters, we'll need two pieces of information:
 
@@
 
+We'll create those here:
+
 ```{r prepare seurat input}
-# Extract the PCA matrix
+# Extract the PCA matrix from the Seurat object
 seurat_pca_matrix <- Embeddings(seurat_obj, reduction = "pca")
 
-# Create a data frame
+# Create a data frame for input
 seurat_cluster_df <- data.frame(
@@
 
-First, you should prepare a data frame two columns: `cell_id` containing cell barcodes, and `cluster` containing the cluster identities.
+ScPCA cell metadata already contains a column called `cluster` with results from an automated clustering. 
+These clusters were calculated using `bluster`, the same tool that `rOpenScPCA` uses, with the following parameters:
 
-```{r}
+* The Louvain algorithm
+* Jaccard weighting
+* 20 nearest neighbors
+
+In this example, we'll show how to use the cluster evaluation functions on these clusters.
+
+To begin, we'll prepare a data frame with two columns: `cell_id` containing cell barcodes, and `cluster` containing the cluster identities.
+
+```{r prepare scpca data frame}
 scpca_cluster_df <- data.frame(
@@
 
-Then, you can proceed to run evaluation functions using this data frame, and the PCA matrix from the object.
+We can run evaluation functions using this data frame and the PCA matrix from the object.
 
-```{r}
+```{r scpca silhouette}
 # Calculate silhouette width
@@
 scpca_silhouette_df
+```
 
+```{r scpca purity}
 # Calculate neighborhood purity
@@
 
-When running `calculate_stability()`, it is important to specify the same parameters originally used to build the clusters:
+When running `calculate_stability()`, we'll specify the same parameters originally used to build the clusters.
+We only recommend evaluating clusters with `calculate_stability()` if you know the original parameters used.
 
-```{r}
+```{r scpca stability}
 scpca_stability_df <- calculate_stability(
@@
   weighting = "jaccard",
-  nn = 20,
-  resolution = 1
+  nn = 20
 )
@@
 
-Instead of a matrix, you can alternatively pass in an SCE or Seurat object that contains PCs.
-For example:
+Instead of a matrix, you can alternatively pass in an SCE or Seurat object that contains a matrix.
+We show examples of this below with `rOpenScPCA::calculate_clusters()`, but this will also work for any of the evaluation functions as well:
 
-```{r}
+```{r run on sce}
 # Calculate clusters from an SCE object using default parameters
 cluster_results_df <- calculate_clusters(sce)
+cluster_results_df
+```
 
+```{r run on seurat}
 # Calculate clusters from a Seurat object using default parameters
 cluster_results_df <- calculate_clusters(seurat_obj)
+cluster_results_df
 ```
 
-`rOpenScPCA` assumes that the PCA matrix is named `PCA` in SCE objects, and `pca` in Seurat objects.
+`rOpenScPCA` assumes that the PCA matrix is named `PCA` in SCE objects, and `pca` in Seurat objects (as they were in the above examples).
 If the PCA matrix you want to use in the object has a different name, you can provide the argument `pc_name`.
-For example:
+We can see this below with an SCE object, for example:
 
@@
 reducedDimNames(sce) <- c("PCA_matrix", "UMAP")
+reducedDimNames(sce)
 
@@
 )
+cluster_results_df
 ```
@@
 Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv()`), or you can add the results into your SCE or Seurat object.
-The subsequent examples will show you, as an example, how to save the `cluster_results_df` data frame to a given object.
+The subsequent examples will show you how to save the `cluster_results_df` data frame to an SCE and a Seurat object.
 
@@
 To avoid ambiguity between the existing and new clustering results, you may want to rename incoming columns.
+
 We'll therefore begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results_df` data frame before adding them to a given object.
@@
 
-# Third, add columns as desired using the barcodes and cell_id columns to join
+# Second, add columns as desired using the barcodes and cell_id columns to join
 sce_coldata <- sce_coldata |>
@@
 
-# Finally, add the modified colData back into the SCE
+# Third, add the modified colData back into the SCE
 # Use the row.names argument to ensure rownames (barcodes) are retained; don't forget this step!
@@
 
-# Print updated colData slot
 colData(sce)

### 2024-11-15 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
fix header depth
fix wording

@@
 * Perform clustering
-* Calculate QC metrics on clustering including:
+* Calculate QC metrics on clusters, including:
   * Silhouette width
@@
   * Cluster stability, as measured with the Adjusted Rand Index
-* Calculate QC metrics obtained with other tools, such as `Seurat`
+* Calculate QC metrics on clusters obtained with other tools, such as `Seurat`
 * Save clustering results to an SCE or `Seurat`
@@
 
-### Working with objects directly
+## Working with objects directly
 

### 2024-11-18 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
fix typos in 00b and 02b, and re-render 00b

@@
 
-For some reason, the "cap-mesenchyme cells" has been renamed in cellxgene in "mesenchymal cells".
+For some reason, the "cap-mesenchyme cells" has been renamed in CELLxGENE as "mesenchymal cells".
 
-The cap mesenchyme is a cap of condensed metanephric mesenchyme, comprised of cells which epithelialize and sequentially form the pretubular aggregate (PA), renal vesicle (RV), C-, and S-shaped bodies, and finally the mature nephron.
+The cap mesenchyme is a cap of condensed metanephric mesenchyme, comprised of cells which epithelialize and sequentially form the peritubular aggregate (PA), renal vesicle (RV), C-, and S-shaped bodies, and finally the mature nephron.
 The CM contains nephron progenitor cells.
@@
 - the fetal nephron / mesenchymal cells are cap-mesenchyme cells. 
-In our case, cap-mesenchyme contains blastema and primitive epitheliul cancer cells.
+In our case, cap-mesenchyme contains blastema and primitive epithelial cancer cells.
 
-- the stroma / mesenchymal stem are likelly mesenchymal cancer or normal cells.
+- the stroma / mesenchymal stem are likely mesenchymal cancer or normal cells.
 

### 2024-11-18 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd

@@
 
-#### Visualize seurat clusters and metadata
+#### Visualize Seurat clusters and metadata
 
@@
 
-#### Visualize seurat clusters and markers genes
+#### Visualize Seurat clusters and markers genes
 
@@
 
-- `data` seurat object
+- `data` Seurat object
 - `first_group` is the name of the first metadata to group the cells
@@
 
-### Visualize seurat clusters
+### Visualize Seurat clusters
 
@@
 
-#### TP53 pathway
+#### `TP53` pathway
 
-here we will calculate a TP53 score using `AddModuleScore` and the genes of the HALLMARK_P53_PATHWAY gene set.
+here we will calculate a `TP53` score using `AddModuleScore` and the genes of the HALLMARK_P53_PATHWAY gene set.
 
@@
 
-### Find marker genes for each of the seurat clusters
+### Find marker genes for each of the Seurat clusters
 

### 2024-11-19 analyses/cell-type-ewings/template_notebooks/clustering-workflow/01-clustering-metrics.Rmd
add leiden clustering to report

---
title: "Clustering exploration for `r params$library_id`"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  library_id: "SCPCL000822"
  sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
  cluster_results_file: "SCPCL000822_clusters.tsv"
---

This notebook summarizes clustering metrics for `r params$library_id`.

Louvain clusters are calculated using the following range of parameters:

- Nearest neighbors: 5, 10, 15, 20, 25, 30, 35, and 40
- Resolution: 0.5, 1, 1.5

Leiden clusters are calcualted using the following range of parameters:

- Objective function: CPM and modularity 
- Nearest neighbors: 5, 10, 15, 20, 25, 30, 35, and 40
- Resolution: 0.5, 1, 1.5

- Metrics are then calculated to evaluate clustering results across all parameters: 
  - Silhouette width 
  - Cluster purity 
  - Cluster stability

## Setup

```{r}
# check that sce file exists
stopifnot(
  "sce file does not exist" = file.exists(params$sce_file)
)
```

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)

# set seed
set.seed(2024)
```

```{r base paths}
# The path to this module
module_base <- rprojroot::find_root(rprojroot::is_renv_project)

# source in clustering functions
clustering_functions <- file.path(module_base, "scripts", "utils", "clustering-functions.R")
source(clustering_functions)
```

```{r}
# read in input sce files
sce <- readr::read_rds(params$sce_file)
```

## Calculate clusters

```{r}
# louvain clustering across all parameters 
louvain_results <- rOpenScPCA::sweep_clusters(
  sce,
  algorithm = "louvain",
  nn = seq(5, 40, 5),
  resolution = c(0.5, 1, 1.5)
) |> 
  dplyr::bind_rows() |> 
  # add in objective function so we can combine with leiden results
  dplyr::mutate(objective_function = NA_character_)

# leiden clustering
leiden_results <- rOpenScPCA::sweep_clusters(
  sce,
  algorithm = "leiden",
  nn = seq(5, 40, 5),
  resolution = c(0.5, 1, 1.5),
  objective_function = c("CPM", "modularity")
) |> 
  dplyr::bind_rows()
```

```{r}
# join all clustering results 
all_cluster_results <- dplyr::bind_rows(list(louvain_results, leiden_results)) |> 
  # add column for method + objective function 
  # use this for splitting plots throughout report
  dplyr::mutate(cluster_method = glue::glue(
    "{algorithm}-{objective_function}"
  ) |> 
    stringr::str_remove("-NA")
  )
```

## Visualize clusters with UMAP

```{r}
# get umap embeddings and combine into a data frame with cluster assignments
umap_df <- sce |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  # replace UMAP.1 with UMAP1 and get rid of excess columns
  dplyr::select("cell_id" = barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2) |>
  dplyr::left_join(all_cluster_results, by = "cell_id")

split_umap_df <- umap_df |> 
  # add a column that shows the combination of nn/res so we get stats for each unique combo of params 
  dplyr::mutate(cluster_params = glue::glue("{nn}nn-{resolution}res")) |> 
  split(umap_df$cluster_method)
```

Below we visualize the cluster assignments using each parameter on a UMAP. 
This can be helpful to see if any parameters lead to any obvious over or under clustering.

```{r, fig.height=15, fig.width=10}
# look at clustering results for each library across all params
split_umap_df |> 
  purrr::imap(\(df, method){

    ggplot(df, aes(x = UMAP1, y = UMAP2, color = cluster)) +
      geom_point(alpha = 0.5, size = 0.1) +
      facet_grid(rows = vars(nn),
                 cols = vars(resolution),
                 labeller = labeller(nn = ~ glue::glue("{.}-nn"),
                                     resolution = ~ glue::glue("{.}-res"))) +
      theme(
        aspect.ratio = 1,
        legend.position = "none"
      ) +
      labs(title = method)

  })
```

## Clustering statistics

Below we calculate a series of statistics:

- Average silhouette width: This metric evaluates cluster separation. 
Cells with large positive silhouette widths are closer to other cells in the same cluster than to cells in different clusters. 
Higher values indicate tighter clusters.
- Average cluster purity: This metric also evaluates cluster separation and tells us the proportion of neighboring cells that are assigned to the same cluster. 
Purity values range from 0-1 with higher purity values indicating clusters that are well separated. 
- Cluster stability: This evaluates how stable the clustering is to input data. 
Stability values range from 01- with higher values of cluster stability indicating more reproducible clusters.

```{r, warning=FALSE}
# get a combined stats dataframe with purity and width for all clusters
all_stats <- split_umap_df |> 
  purrr::map(\(df) {get_cluster_stats(sce,df)})
```

### Silhouette width

```{r}
# silhouette width for different params
all_stats |> 
  purrr::imap(\(df, cluster_method){plot_cluster_stats(df, width, cluster_method)})
```

### Cluster purity

```{r}
# cluster purity for different params
all_stats |> 
  purrr::imap(\(df, cluster_method){plot_cluster_stats(df, purity, cluster_method)})
```

### Cluster stability

```{r, warning=FALSE, message=FALSE}
# calculate cluster stability
stability_stats <- split_umap_df |> 
  purrr::map(\(df){ get_cluster_stability(sce, df)})
```

```{r, warning=FALSE, message=FALSE}
# plot stability across all parameters
stability_stats |> 
  purrr::imap(\(df, cluster_method){

    # plot stability across all values of k
    ggplot(df, aes(x = nn, y = ari)) +
      geom_jitter(width = 0.1) +
      facet_wrap(vars(resolution),
                 labeller = labeller(resolution = ~ glue::glue("{.}-res"))) +
      labs(title = "Cluster stability") +
      stat_summary(
        aes(group = nn),
        color = "red",
        # median and quartiles for point range
        fun = "median",
        fun.min = function(x) {
          quantile(x, 0.25)
        },
        fun.max = function(x) {
          quantile(x, 0.75)
        }
      )

  })
```

## Export clusters

```{r}
# export all clustering results
readr::write_tsv(all_cluster_results, params$cluster_results_file)
```

## Session info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2024-11-19 analyses/cell-type-ewings/template_notebooks/cnv-workflow/04-infercnv.Rmd
account for weird chromosomes showing up

@@
 # remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate
-chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL"))
+# only keep coldata columns and chr1-22 columns
+chr_columns <- paste0("chr", seq(1,22), collapse = "|")
+all_colnames_keep <- paste0(c(names(colData(sce)), "UMAP1", "UMAP2|"), collapse = "|") |> 
+  paste0(chr_columns, collapse = "|")
+chr_columns_to_keep <- which(stringr::str_detect(names(coldata_df), all_colnames_keep))
 all_cnv_df <- coldata_df[, chr_columns_to_keep]

### 2024-11-19 analyses/cell-type-consensus/notebook-template.Rmd
initiate consensus module

--- 2024-09-13 analyses/cell-type-dsrct/notebook-template.Rmd
@@
 ---
-title: "cell-type-dsrct Analysis"
+title: "cell-type-consensus Analysis"
 author:
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "cell-type-dsrct")
+module_base <- file.path(repository_base, "analyses", "cell-type-consensus")
 ```

### 2024-11-20 analyses/cell-type-ewings/template_notebooks/clustering-workflow/01-clustering-metrics.Rmd
adjust range and add boxes around umaps

@@
         aspect.ratio = 1,
-        legend.position = "none"
+        legend.position = "none", 
+        panel.border = element_rect(color = "black", fill = NA)
       ) +
@@
 - Average silhouette width: This metric evaluates cluster separation. 
+Values range from -1 to 1.
 Cells with large positive silhouette widths are closer to other cells in the same cluster than to cells in different clusters. 
@@
 - Cluster stability: This evaluates how stable the clustering is to input data. 
-Stability values range from 01- with higher values of cluster stability indicating more reproducible clusters.
+Stability values range from 0-1 with higher values of cluster stability indicating more reproducible clusters.
 

### 2024-11-20 analyses/cell-type-ewings/template_notebooks/clustering-workflow/01-clustering-metrics.Rmd
use different resolution for cpm

@@
 
-# leiden clustering
-leiden_results <- rOpenScPCA::sweep_clusters(
+# leiden cpm clustering
+leiden_cpm_results <- rOpenScPCA::sweep_clusters(
+  sce,
+  algorithm = "leiden",
+  nn = seq(5, 40, 5),
+  resolution = c(0.001, .005, .01),
+  objective_function = "CPM"
+) |> 
+  dplyr::bind_rows()
+
+# leiden modularity clustering
+leiden_mod_results <- rOpenScPCA::sweep_clusters(
   sce,
@@
   resolution = c(0.5, 1, 1.5),
-  objective_function = c("CPM", "modularity")
+  objective_function = "modularity"
 ) |> 
@@
 # join all clustering results 
-all_cluster_results <- dplyr::bind_rows(list(louvain_results, leiden_results)) |> 
+all_cluster_results <- dplyr::bind_rows(list(louvain_results, leiden_cpm_results, leiden_mod_results)) |> 
   # add column for method + objective function 

### 2024-11-20 analyses/cell-type-wilms-tumor-06/notebook_template/00b_characterize_fetal_kidney_reference_Stewart.Rmd

@@
   seed: 12345
-  fetal_kidney_path: "../scratch/fetal_kidney.rds"
+  fetal_kidney_path: "../scratch/fetal_kidney_seurat.rds"
 output:
@@
 
-The aim is to characterize the human fetal kidney from the kidney cell atlas.
-You can find more about the human kidney atlas here: https://www.kidneycellatlas.org/ [1]
-The rds data can be download using the download link: <https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds>.
+The aim is to characterize the human fetal kidney from the kidney cell atlas obtained from CELLxGENE.
+You can find more about the human kidney atlas here: https://www.kidneycellatlas.org/ [1].
 The reference was downloaded and created in the script `scripts/prepare-fetal-references.R`.

### 2024-11-20 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd

@@
   seed: 12345
-  homologs_file: "../scratch/homologs.rds"
   testing: 0
@@
   srat_assay,
-  params$homologs_file
+  convert_gene_names = FALSE
 )

### 2024-11-20 analyses/cell-type-ewings/template_notebooks/clustering-workflow/01-clustering-metrics.Rmd
add parameters as actual parameters

@@
   library_id: "SCPCL000822"
-  sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
+  sce_file: "../../../../data/2024-07-08/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
   cluster_results_file: "SCPCL000822_clusters.tsv"
+  nn_range: !r seq(5, 40, 5)
+  louvain_res_range: !r c(0.5, 1, 1.5)
+  mod_res_range: !r c(0.5, 1, 1.5)
+  cpm_res_range: !r c(0.001, 0.005, 0.01)
 ---
@@
 
-- Nearest neighbors: 5, 10, 15, 20, 25, 30, 35, and 40
-- Resolution: 0.5, 1, 1.5
+- Nearest neighbors: `r paste0(params$nn_range, collapse = ", ")`
+- Resolution:`r paste0(params$louvain_res_range, collapse = ", ")`
 
-Leiden clusters are calcualted using the following range of parameters:
+Leiden clusters are calculated using modularity as the objective function with the following range of parameters:
 
-- Objective function: CPM and modularity 
-- Nearest neighbors: 5, 10, 15, 20, 25, 30, 35, and 40
-- Resolution: 0.5, 1, 1.5
+- Objective function: modularity 
+- Nearest neighbors: `r paste0(params$nn_range, collapse = ", ")`
+- Resolution: `r paste0(params$mod_res_range, collapse = ", ")`
+
+Leiden clusters are calculated using CPM as the objective function with the following range of parameters:
+
+- Objective function: CPM 
+- Nearest neighbors: `r paste0(params$nn_range, collapse = ", ")`
+- Resolution: `r paste0(params$cpm_res_range, collapse = ", ")`
 
@@
   algorithm = "louvain",
-  nn = seq(5, 40, 5),
-  resolution = c(0.5, 1, 1.5)
+  nn = params$nn_range,
+  resolution = params$louvain_res_range
 ) |> 
@@
   algorithm = "leiden",
-  nn = seq(5, 40, 5),
-  resolution = c(0.001, .005, .01),
+  nn = params$nn_range,
+  resolution = params$cpm_res_range,
   objective_function = "CPM"
@@
   algorithm = "leiden",
-  nn = seq(5, 40, 5),
-  resolution = c(0.5, 1, 1.5),
+  nn = params$nn_range,
+  resolution = params$mod_res_range,
   objective_function = "modularity"

### 2024-11-20 analyses/cell-type-ewings/template_notebooks/clustering-workflow/01-clustering-metrics.Rmd
change back to current

@@
   library_id: "SCPCL000822"
-  sce_file: "../../../../data/2024-07-08/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
+  sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
   cluster_results_file: "SCPCL000822_clusters.tsv"

### 2024-11-21 analyses/cell-type-ewings/template_notebooks/clustering-workflow/02-clustering-celltypes.Rmd

--- 2024-08-06 analyses/cell-type-ewings/template_notebooks/clustering-workflow/01-clustering.Rmd
@@
 ---
-title: "Clustering exploration for `r params$library_id`"
+title: "Comparison of clusters to cell types for `r params$library_id`"
 author: Ally Hawkins
@@
 params:
-  library_id: NULL
-  sce_file: NULL
-  singler_results_file: NULL
-  cluster_results_file: NULL
-  marker_genes_file: NULL
+  library_id: "SCPCL000822"
+  sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
+  singler_results_file: "../../results/aucell_singler_annotation/SCPCS000490/SCPCL000822_singler-classifications.tsv"
+  cluster_results_file: "../../results/cluster_exploration/SCPCS000490/SCPCL000822_clusters.tsv"
+  marker_genes_file: "../../references/visser-all-marker-genes.tsv"
 ---
 
-This notebook summarizes clustering for `r params$library_id`.
+TODO: Update this notebook to take as input a clustering algorithm, nn (range of nn?), resolution, and optional objective function to display in SingleR and marker gene plots
 
-- Louvain, Jaccard clusters are calculated using a range of parameters specifying nearest neighbors (5, 10, 15, 20, 25, 30, 35, and 40). 
-- Metrics are then calculated to evaluate clustering results across all parameters: 
-  - Silhouette width 
-  - Cluster purity 
-  - Cluster stability 
 - The clusters are then compared to the results from running `SingleR` in the `aucell-singler-annotation.sh` workflow.
@@
 ComplexHeatmap::ht_opt(message = FALSE)
+
+# set seed
+set.seed(2024)
 ```
@@
 # read in singler results
-singler_classification_df <- readr::read_tsv(params$singler_results_file)
+singler_classification_df <- readr::read_tsv(params$singler_results_file) |>
+  dplyr::rename("cell_id" = "barcodes")
 ```
 
-## Clustering
-
-Below we perform Louvain, Jaccard clustering, varying `k`. 
-The minimum `k` is 5 and the maximum `k` is 40 with a step size of 5.
-
-```{r}
-# perform clustering across k = 5-40 with increments of 5
-all_cluster_results <- cluster_sweep(sce) |>
-  dplyr::mutate(
-    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
-  )
-```
-
-```{r}
-# get umap embeddings and combine into a data frame with cluster assignments
-umap_df <- sce |>
-  scuttle::makePerCellDF(use.dimred = "UMAP") |>
-  # replace UMAP.1 with UMAP1 and get rid of excess columns
-  dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2) |>
-  dplyr::left_join(all_cluster_results, by = "barcodes")
-```
-
-Below we visualize the cluster assignments using each parameter on a UMAP. 
-This can be helpful to see any values of `k` that may have obvious over or under clustering.
-
-```{r, fig.height=15, fig.width=10}
-# look at clustering results for each library across all params
-ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cluster)) +
-  geom_point(alpha = 0.5, size = 0.1) +
-  facet_wrap(vars(nn_param)) +
-  theme(
-    aspect.ratio = 1,
-    legend.position = "none"
-  )
-```
-
-## Clustering statistics
-
-Below we calculate a series of statistics:
-
-- Average silhouette width: This metric evaluates cluster separation. 
-Cells with large positive silhouette widths are closer to other cells in the same cluster than to cells in different clusters. 
-Higher values indicate tighter clusters.
-- Average cluster purity: This metric also evaluates cluster separation and tells us the proportion of neighboring cells that are assigned to the same cluster. 
-Purity values range from 0-1 with higher purity values indicating clusters that are well separated. 
-- Cluster stability: This evaluates how stable the clustering is to input data. 
-Stability values range from 01- with higher values of cluster stability indicating more reproducible clusters.
-
-```{r, warning=FALSE}
-# get a combined stats dataframe with purity and width for all clusters
-all_stats_df <- get_cluster_stats(sce, all_cluster_results) |>
-  dplyr::mutate(
-    # make sure the order is correct
-    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
-  )
-```
-
-### Silhouette width
-
-```{r}
-# silhouette width for different params
-plot_cluster_stats(all_stats_df, width)
-```
-
-### Cluster purity
-
-```{r}
-# cluster purity for different params
-plot_cluster_stats(all_stats_df, purity)
-```
-
-### Cluster stability
-
-```{r, warning=FALSE, message=FALSE}
-# calculate cluster stability
-stability_df <- get_cluster_stability(sce, all_cluster_results) |>
-  dplyr::mutate(
-    k_value = as.factor(k_value),
-    # make sure that k = 5 comes first
-    k_value = forcats::fct_relevel(k_value, "5", after = 0)
-  )
-
-# plot stability across all values of k
-ggplot(stability_df, aes(x = k_value, y = ari)) +
-  geom_jitter(width = 0.1) +
-  labs(title = "Cluster stability") +
-  stat_summary(
-    aes(group = k_value),
-    color = "red",
-    # median and quartiles for point range
-    fun = "median",
-    fun.min = function(x) {
-      quantile(x, 0.25)
-    },
-    fun.max = function(x) {
-      quantile(x, 0.75)
-    }
-  )
-```
-
-## Compare clusters to cell types from `SingleR`
+## Clusters vs. `SingleR` {.tabset}
 
@@
   # add in classifications from singler
-  dplyr::left_join(singler_classification_df, by = "barcodes") |>
+  dplyr::left_join(singler_classification_df, by = "cell_id") |>
   dplyr::mutate(
@@
       forcats::fct_infreq() |>
-      forcats::fct_relevel("All remaining cell types", after = Inf),
-    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
+      forcats::fct_relevel("All remaining cell types", after = Inf)
   )
+
+cluster_classification_df |> 
+  split(cluster_classification_df$resolution) |> 
+  purrr::map(\(df){cluster_celltype_heatmap(df)})
+
 ```
@@
 
-## Marker gene expression across clusters
+## Marker gene expression {.tabset}
 

### 2024-11-21 analyses/cell-type-ewings/template_notebooks/clustering-workflow/01-clustering-metrics.Rmd
remove extra cluster code

--- 2024-11-20 analyses/cell-type-ewings/template_notebooks/clustering-workflow/01-clustering-metrics.Rmd
@@
   sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
-  cluster_results_file: "SCPCL000822_clusters.tsv"
+  cluster_results_file: "../../results/clustering/SCPCS000490/SCPCL000822_cluster-results.tsv"
 ---
@@
 
-Louvain clusters are calculated using the following range of parameters:
-
-- Nearest neighbors: 5, 10, 15, 20, 25, 30, 35, and 40
-- Resolution: 0.5, 1, 1.5
-
-Leiden clusters are calcualted using the following range of parameters:
-
-- Objective function: CPM and modularity 
-- Nearest neighbors: 5, 10, 15, 20, 25, 30, 35, and 40
-- Resolution: 0.5, 1, 1.5
-
-- Metrics are then calculated to evaluate clustering results across all parameters: 
+The following metrics are calculated and shown across all parameters: 
   - Silhouette width 
@@
 sce <- readr::read_rds(params$sce_file)
+
+# read in results file 
+cluster_results_df <- readr::read_tsv(params$cluster_results_file)
 ```
 
-## Calculate clusters
+## Clustering summary
 
-```{r}
-# louvain clustering across all parameters 
-louvain_results <- rOpenScPCA::sweep_clusters(
-  sce,
-  algorithm = "louvain",
-  nn = seq(5, 40, 5),
-  resolution = c(0.5, 1, 1.5)
-) |> 
-  dplyr::bind_rows() |> 
-  # add in objective function so we can combine with leiden results
-  dplyr::mutate(objective_function = NA_character_)
+```{r, results='asis'}
+cluster_results_df |> 
+  split(cluster_results_df$cluster_method) |> 
+  purrr::imap(\(df, method){
+    res_print <- unique(df$resolution) |> 
+      paste0(collapse = ", ")
+    nn_print <- unique(df$nn) |> 
+      paste0(collapse = ", ")
+    glue::glue("
+               {method} clustering was performed with the following parameters:
 
-# leiden cpm clustering
-leiden_cpm_results <- rOpenScPCA::sweep_clusters(
-  sce,
-  algorithm = "leiden",
-  nn = seq(5, 40, 5),
-  resolution = c(0.001, .005, .01),
-  objective_function = "CPM"
-) |> 
-  dplyr::bind_rows()
-
-# leiden modularity clustering
-leiden_mod_results <- rOpenScPCA::sweep_clusters(
-  sce,
-  algorithm = "leiden",
-  nn = seq(5, 40, 5),
-  resolution = c(0.5, 1, 1.5),
-  objective_function = "modularity"
-) |> 
-  dplyr::bind_rows()
-```
-
-```{r}
-# join all clustering results 
-all_cluster_results <- dplyr::bind_rows(list(louvain_results, leiden_cpm_results, leiden_mod_results)) |> 
-  # add column for method + objective function 
-  # use this for splitting plots throughout report
-  dplyr::mutate(cluster_method = glue::glue(
-    "{algorithm}-{objective_function}"
-  ) |> 
-    stringr::str_remove("-NA")
+                - Nearest neighbors: {nn_print}
+                - Resolution: {res_print}"
                )
+  })
 ```
@@
   dplyr::select("cell_id" = barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2) |>
-  dplyr::left_join(all_cluster_results, by = "cell_id")
+  dplyr::left_join(cluster_results_df, by = "cell_id")
 
@@
 
-## Export clusters
-
-```{r}
-# export all clustering results
-readr::write_tsv(all_cluster_results, params$cluster_results_file)
-```
-
 ## Session info

### 2024-11-25 analyses/cell-type-dsrct/exploratory_analysis/01-marker-gene-tumor-classification.Rmd
fix typos and re-render notebook

--- 2024-10-23 analyses/cell-type-dsrct/exploratory_analysis/01-marker-gene-tumor-classification.Rmd
@@
 ---
-title: "EDA of DSRCT Samples"
+title: "Exploratory data analysis of DSRCT Samples"
 author: Danh Truong
@@
 # The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "2024-07-08")
-#sample_dir <- file.path(data_dir, "SCPCP000013", params$sample_id)
+data_dir <- file.path(repository_base, "data", "current")
+# sample_dir <- file.path(data_dir, "SCPCP000013", params$sample_id)
 
@@
 metadata_file <- file.path(data_dir, "SCPCP000013", "single_cell_metadata.tsv")
-metadata <- read.csv(metadata_file,  sep = '\t')
-metadata_DSRCT <- dplyr::filter(metadata, diagnosis == 'Desmoplastic small round cell tumor')
-
+metadata <- read.csv(metadata_file, sep = "\t")
+metadata_DSRCT <- dplyr::filter(metadata, diagnosis == "Desmoplastic small round cell tumor")
 ```
@@
 
-#classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
-#output_classifications_file <- file.path(results_dir, classifications_filename)
+# classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
+# output_classifications_file <- file.path(results_dir, classifications_filename)
 ```
@@
 
-#adding the sample id to the name of each SingleCellExperiment
+# adding the sample id to the name of each SingleCellExperiment
 names(sce_list) <- metadata_DSRCT$scpca_library_id
@@
     # combine all genes into a single column for easy faceting
-    tidyr::pivot_longer(cols = starts_with("ENSG"),
+    tidyr::pivot_longer(
+      cols = starts_with("ENSG"),
       names_to = "ensembl_gene_id",
-                        values_to = "gene_expression") |>
+      values_to = "gene_expression"
+    ) |>
     # join with marker gene df to get gene symbols for plotting
     dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
-    dplyr::select(barcodes,
+    dplyr::select(
+      barcodes,
       UMAP1,
@@
       gene_expression,
-                  cluster)
+      cluster
+    )
 })
-
 ```
@@
 ```{r, fig.width=8}
-for(i in 1:length(umap_df_list)){
+for (i in 1:length(umap_df_list)) {
   # faceted umap showing a umap panel for each marker gene
@@
 
-  #save plots
+  # save plots
   ggsave(
     filename = paste0(
-      '../plots/marker_expression_',
+      module_base,
+      "/plots/marker_expression_",
       names(umap_df_list)[i],
-      '.png'
+      ".png"
     ),
@@
     height = 6,
-    units = 'in',
+    units = "in",
     dpi = 150
@@
 }
-
 ```
 
-In my experience, ST6GALNAC5 is a strong marker for DSRCT in single-cell data.
-As can be seen, several of the samples have expression of ST6GALNAC5, but some do not, such as SCPCS000731 and SCPCS000729.
+In my experience, `ST6GALNAC5` is a strong marker for DSRCT in single-cell data.
+As can be seen, several of the samples have expression of `ST6GALNAC5`, but some do not, such as SCPCS000731 and SCPCS000729.
 In fact, these SCPCS000729 contains  low number of cells.
@@
 ```{r}
-umap_df_list_excluded <- umap_df_list[!(names(umap_df_list) %in% c('SCPCS000731'))]
+umap_df_list_excluded <- umap_df_list[!(names(umap_df_list) %in% c("SCPCS000731"))]
 ```
@@
 I would expect to see some sort of bimodal distribution separating cells that do and do not have expression of the marker gene.
-What is clear from these plots that ST6GALNAC5, CACNA2D2, PTPRQ, IQCJ-SCHIP1,  show a bi modal distribution among the cells.
+What is clear from these plots that `ST6GALNAC5`, `CACNA2D2`, `PTPRQ`, `IQCJ-SCHIP1`,  show a bi modal distribution among the cells.
 To some extent, we do see expression of the other markers.
@@
 ```{r}
-for(i in 1:length(umap_df_list_excluded)) {
-  p <- ggplot(umap_df_list_excluded[[i]],
-              aes(x = gene_expression, fill = gene_symbol)) +
+for (i in 1:length(umap_df_list_excluded)) {
+  p <- ggplot(
+    umap_df_list_excluded[[i]],
+    aes(x = gene_expression, fill = gene_symbol)
+  ) +
     geom_density() +
-    facet_wrap(vars(gene_symbol), scales = 'free_y') +
+    facet_wrap(vars(gene_symbol), scales = "free_y") +
     theme(legend.position = "none") +
@@
 
-  #save plots
+  # save plots
   ggsave(
     filename = paste0(
-      '../plots/marker_distribution_',
+      module_base,
+      "/plots/marker_distribution_",
       names(umap_df_list)[i],
-      '.png'
+      ".png"
     ),
@@
 ```{r}
-umap_df_list_excluded_scaled <- lapply(umap_df_list_excluded, function(x){
+umap_df_list_excluded_scaled <- lapply(umap_df_list_excluded, function(x) {
   x |>
@@
 })
-
 ```
@@
 Now we can create the same density plot but looking at z-scores.
-Interestingly, in some samples, like SPCS000489, we see that ST6GALNAC5 has negative values.
+Interestingly, in some samples, like SCPCS000489, we see that `ST6GALNAC5` has negative values.
 Presumably, these are tumor cells but there may be a skew due to the number of tumor cells compared to normal.
@@
 ```{r}
-
-for(i in 1:length(umap_df_list_excluded_scaled)) {
-  p <- ggplot(umap_df_list_excluded_scaled[[i]],
-              aes(x = transformed_gene_expression, fill = gene_symbol)) +
+for (i in 1:length(umap_df_list_excluded_scaled)) {
+  p <- ggplot(
+    umap_df_list_excluded_scaled[[i]],
+    aes(x = transformed_gene_expression, fill = gene_symbol)
+  ) +
     geom_density() +
-    facet_wrap(vars(gene_symbol), scales = 'free_y') +
+    facet_wrap(vars(gene_symbol), scales = "free_y") +
     theme(legend.position = "none") +
@@
 
-  #save plots
+  # save plots
   ggsave(
     filename = paste0(
-      '../plots/marker_distribution_zscaled_',
+      module_base,
+      "/plots/marker_distribution_zscaled_",
       names(umap_df_list)[i],
-      '.png'
+      ".png"
     ),
@@
     height = 6,
-    units = 'in',
+    units = "in",
     dpi = 150
   )
-
 }
@@
 # calculate sum gene expression across all marker genes in list
-marker_sum_exp <- lapply(umap_df_list_excluded_scaled, function(x){
+marker_sum_exp <- lapply(umap_df_list_excluded_scaled, function(x) {
   x |>
@@
 
-for(i in 1:length(marker_sum_exp)) {
+for (i in 1:length(marker_sum_exp)) {
   p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_exp)) +
@@
 
-  #save plots
+  # save plots
   ggsave(
@@
       module_base,
-      '/plots/UMAP_marker_expression_',
+      "/plots/UMAP_marker_expression_",
       names(umap_df_list)[i],
-      '.png'
+      ".png"
     ),
@@
 # classify tumor cells based on presence of any marker genes
-marker_sum_exp <- lapply(marker_sum_exp, function(x){
+marker_sum_exp <- lapply(marker_sum_exp, function(x) {
   x |>
-  dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))})
+    dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))
+})
 
-for(i in 1:length(marker_sum_exp)) {
+for (i in 1:length(marker_sum_exp)) {
   p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_classification)) +
@@
       module_base,
-      '/plots/UMAP_classification_',
+      "/plots/UMAP_classification_",
       names(umap_df_list)[i],
-      '.png'
+      ".png"
     ),
@@
 # calculate sum gene expression across all marker genes in list
-marker_sum_exp <- lapply(umap_df_list_excluded_scaled, function(x){
+marker_sum_exp <- lapply(umap_df_list_excluded_scaled, function(x) {
   x |>
-  dplyr::group_by(cluster) |> #change to cluster
+    dplyr::group_by(cluster) |> # change to cluster
     dplyr::mutate(sum_exp = sum(transformed_gene_expression, na.rm = T)) |>
@@
 
-  #save plots
+  # save plots
   ggsave(
-    filename = paste0(module_base, '/plots/UMAP_marker_expression_cluster_', names(umap_df_list)[i], '.png'),
+    filename = paste0(module_base, "/plots/UMAP_marker_expression_cluster_", names(umap_df_list)[i], ".png"),
     plot = p,
@@
 # classify tumor cells based on presence of any marker genes
-marker_sum_exp <- lapply(marker_sum_exp, function(x){
+marker_sum_exp <- lapply(marker_sum_exp, function(x) {
   d <- density(x$sum_exp)
@@
   x |>
-  dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))})
+    dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))
+})
 
-for(i in 1:length(marker_sum_exp)) {
+for (i in 1:length(marker_sum_exp)) {
   p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_classification)) +
@@
       module_base,
-      '../plots/UMAP_classification_cluster_',
+      "/plots/UMAP_classification_cluster_",
       names(umap_df_list)[i],
-      '.png'
+      ".png"
     ),
@@
     height = 6,
-    units = 'in',
+    units = "in",
     dpi = 150
   )
-
 }
-
 ```
@@
 # get an RDS of the processed data
-saveRDS(sce_file_list, '../results/SCPCP000013_sce_file_list.rds')
+saveRDS(sce_file_list, "../results/SCPCP000013_sce_file_list.rds")
+# get an RDS of the UMAP data with marker genes
+saveRDS(umap_df_list, file.path(module_base, "results/SCPCP000013_umap_df_list.rds"))
 ```

### 2024-11-25 analyses/cell-type-wilms-tumor-06/supplemental-notebooks/compare-label-transfer-approaches.Rmd

@@
 - Label transfer code with Azimuth currently in `main` at commit `6af112d`. These results are referred to as `"azimuth"`.
-- Label transfer code adapted from Azimuth. These results are referred to as `"adapted_azimuth"`.
+- Label transfer code adapted from Azimuth at commit `3e2f90`. These results are referred to as `"adapted_azimuth"`.
+- All code was run performed with the `2024-08-22` OpenScPCA data release.
 
@@
 |--------|-----------|-------|---------|-----------------|
-| SCPSC000179 | L1 | 70 | Metanephritic cells | Intestinal epithelial cells | 
-| SCPSC000179 | Organ | 64 | Kidney | Intestine | 
-| SCPSC000179 | Organ | 20 | Lung | Kidney | 
-| SCPSC000194 | L1 | 60 | Stromal cells | Mesangial cells | 
-| SCPSC000194 | Organ | 35 | Kidney | Intestine | 
-| SCPSC000194 | Organ | 36 | Lung | Kidney | 
-| SCPSC000205 | Organ | 56 | Kidney | Intestine |
-| SCPSC000208 | L1 | 101 | Mesangial cells | Metanephritic cells |  
-| SCPSC000208 | L1 | 75 | Intestinal epithelial cells | Metanephritic cells |  
-| SCPSC000208 | Organ | 149 | Kidney | Intestine |
+| `SCPCS000179` | L1 | 70 | Metanephric cells | Intestinal epithelial cells |
+| `SCPCS000179` | Organ | 64 | Kidney | Intestine |
+| `SCPCS000179` | Organ | 20 | Lung | Kidney |
+| `SCPCS000194` | L1 | 60 | Stromal cells | Mesangial cells |
+| `SCPCS000194` | Organ | 35 | Kidney | Intestine |
+| `SCPCS000194` | Organ | 36 | Lung | Kidney |
+| `SCPCS000205` | Organ | 56 | Kidney | Intestine |
+| `SCPCS000208` | L1 | 101 | Mesangial cells | Metanephric cells |
+| `SCPCS000208` | L1 | 75 | Intestinal epithelial cells | Metanephric cells |
+| `SCPCS000208` | Organ | 149 | Kidney | Intestine |
 
@@
 |--------|-----------|-------|---------|-----------------|
-| SCPSC000179 | cell type | 94 | mesenchymal cell | kidney epithelial cell | 
-| SCPSC000205 | compartment  | 52 | fetal nephron |  stroma |
+| `SCPCS000179` | cell type | 94 | mesenchymal cell | kidney epithelial cell |
+| `SCPCS000205` | compartment  | 52 | fetal nephron |  stroma |
 

### 2024-11-25 analyses/cell-type-wilms-tumor-14/exploratory_analysis/03_cnv/03_copykat_SCPCL000850.Rmd

--- 2024-10-17 analyses/cell-type-wilms-tumor-14/exploratory_analysis/03_cnv/03_copykat_SCPCL000850.Rmd
@@
 ---
-title: "Exploratory analysis using Copykat"
+title: "Exploratory analysis using CopyKAT"
 author: "Jingxuan Chen"
@@
 
-Here we test `CopyKat` to analyze the CNV profile for Wilms tumor sample (SCPCL000850), aiming to identify potential tumor cells from normal cells.
+Here we test `CopyKAT` to analyze the CNV profile for Wilms tumor sample (SCPCL000850), aiming to identify potential tumor cells from normal cells.
 
-Before running this notebook, one should first generate `CopyKat` outputs as in `03_runCopyKat.R`.
+Before running this notebook, one should first generate `CopyKAT` outputs as in `03_run`CopyKAT`.R`.
 
@@
 # The path to this module
-path_anal <- file.path(path_repo,"analyses","cell-type-wilms-tumor-14") 
+path_anal <- file.path(path_repo, "analyses", "cell-type-wilms-tumor-14")
 ```
@@
 library_id <- "SCPCL000850"
-scratch_out_dir <- file.path(path_anal, "scratch", "03_cnv","copykat")
+scratch_out_dir <- file.path(path_anal, "scratch", "03_cnv", "copykat")
 dir.create(scratch_out_dir, showWarnings = FALSE, recursive = TRUE)
@@
 # load pre-processed sample objs & anchor transfer results
-obj <- SeuratObject::LoadSeuratRds( file.path(path_anal,"scratch","00_preprocessing_rds",paste0(library_id,".rdsSeurat")) )
+obj <- SeuratObject::LoadSeuratRds(file.path(path_anal, "scratch", "00_preprocessing_rds", paste0(library_id, ".rdsSeurat")))
 level <- "compartment"
-predictions <- read.csv( file.path(path_anal, "results", "01_anchor_transfer_seurat", paste0(library_id, "_", level,".csv")) ) 
+predictions <- read.csv(file.path(path_anal, "results", "01_anchor_transfer_seurat", "RNA", paste0(library_id, "_", level, ".csv")))
 obj <- AddMetaData(object = obj, metadata = predictions)
-copykat_result <- readr::read_rds( file = file.path(library_out_dir, paste0(library_id, "_copykat_resultobj.rds")) )
-copykat_result_noref <- readr::read_rds( file = file.path(library_out_dir, paste0(library_id, "_noref_copykat_resultobj.rds")) )
-
+copykat_result <- readr::read_rds(file = file.path(library_out_dir, paste0(library_id, "_copykat_resultobj.rds")))
+copykat_result_noref <- readr::read_rds(file = file.path(library_out_dir, paste0(library_id, "_noref_copykat_resultobj.rds")))
 ```
@@
 
-#### Run CopyKat with normal cells
+#### Run CopyKAT with normal cells
 
@@
 
-Plot CopyKat identifications onto UMAP.
+Plot `CopyKAT` identifications onto UMAP.
 
@@
 
-#### Run CopyKat without normal cells
+#### Run CopyKAT without normal cells
 
-Here, no normal cells were provided to CopyKat.
+Here, no normal cells were provided to `CopyKAT`.
 
@@
 
-In this Wilms tumor sample (SCPCL000850), it's likely that CopyKat is not working well: (i) The result with or without specifying normal cells are not consistent; (ii) From the heatmap, no much differences in CNV profile could be observed.
+In this Wilms tumor sample (SCPCL000850), it's likely that `CopyKAT` is not working well: (i) The result with or without specifying normal cells are not consistent; (ii) From the heatmap, no much differences in CNV profile could be observed.
 

### 2024-11-25 analyses/cell-type-wilms-tumor-06/notebooks/01_seurat-processing.Rmd
move everything over to notebook, out of notebook_template

--- 2024-08-16 analyses/cell-type-wilms-tumor-06/notebook_template/01_seurat-processing.Rmd
@@
 - https://github.com/AlexsLemonade/OpenScPCA-analysis/issues/671
-
-This notebook will be updated with PR #704 to explore the clustering.
 
@@
 filedir <- filelist[grepl(params$sample_id, filelist)]
-
 ```
@@
 
-  srat <- CreateSeuratObject(counts = counts(sce),
+  srat <- CreateSeuratObject(
+    counts = counts(sce),
     assay = "RNA",
@@
   return(srat)
-
 }
-
 ```
@@
                                         verbose = FALSE) {
-
   # Perform normalization, dimensionality reduction, and clustering on a
@@
   # Normalization
-  options(future.globals.maxSize= 8912896000000)
+  options(future.globals.maxSize = 8912896000000)
   srat <- SCTransform(srat,
@@
     verbose = verbose,
-                      conserve.memory = conserve_memory)
+    conserve.memory = conserve_memory
+  )
 
@@
 ```{r run_seurat_wf, fig.height=4, fig.width=8, warning=FALSE, out.width='100%'}
-
 srat <- convert_sce_to_seurat(sce,
-                                  sample_id = params$sample_id)
+  sample_id = params$sample_id
+)
 
@@
   neighbors_dims = 1:50,
-                                        verbose = FALSE)
+  verbose = FALSE
+)
 ```
@@
 ```{r save}
-saveRDS(object = srat, file = file.path(output_dir, paste0("01-Seurat_", params$sample_id,".Rds")))
+saveRDS(object = srat, file = file.path(output_dir, paste0("01-Seurat_", params$sample_id, ".Rds")))
 ```

### 2024-11-25 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd

@@
 ---
-title: "`infercnv` results exploration for `r params$sample_id`"
+title: "`inferCNV` results exploration for `r params$sample_id`"
 author: "Maud PLASCHKA"
@@
 
-This notebook explores using [`infercnv`](https://github.com/broadinstitute/inferCNV/wiki) to estimate tumor and normal cells in `r params$sample_id` from SCPCP000006. 
+This notebook explores using [`inferCNV`](https://github.com/broadinstitute/inferCNV/wiki) to estimate tumor and normal cells in `r params$sample_id` from SCPCP000006.
 This sample has a(n) `r subdiagnosis` subdiagnosis.
 
-`infercnv` was run using the `06_inferCNV.R` script with and without a normal reference, from the same patient or from an inter-patient pull of normal cells.
+`inferCNV` was run using the `06_inferCNV.R` script with and without a normal reference, from the same patient or from an inter-patient pull of normal cells.
 We tested the impact of the sub-selection of normal cells using either immune, and/or endothelial cells as healthy reference.
@@
 
-`infercnv` currently support two models for HMM-based CNV prediction, what we refer to as the i3 and i6 models.
+`inferCNV` currently support two models for HMM-based CNV prediction, what we refer to as the i3 and i6 models.
 These are set in the `infercnv::run()` as HMM_type='i3' or HMM_type='i6' (i6 is default).
-Each method operates on the preliminary `infercnv` object which has been processed through the standard inferCNV processing routines, involving subtraction of signal corresponding to "normal (reference)" cells and smoothing operations.
+Each method operates on the preliminary `inferCNV` object which has been processed through the standard `inferCNV` processing routines, involving subtraction of signal corresponding to "normal (reference)" cells and smoothing operations.
 
@@
 
-Of note, running `infercnv` with a i6 HMM model is very long, ~2hours of running time per sample, while i3 HMM model only requires ~10min per sample.
+Of note, running `inferCNV` with a i6 HMM model is very long, ~2hours of running time per sample, while i3 HMM model only requires ~10min per sample.
 
@@
 
-For a Seurat object `object`, the function `Do_CNV_heatmap` load the `infercnv` object created with the script `06_infercnv.R` using `reference_value` as a reference and call the function `SCpubr::do_CopyNumberVariantPlot` to plot the mean CNV score in each group defined in `group.by`.
+For a Seurat object `object`, the function `Do_CNV_heatmap` load the `inferCNV` object created with the script `06_infercnv.R` using `reference_value` as a reference and call the function `SCpubr::do_CopyNumberVariantPlot` to plot the mean CNV score in each group defined in `group.by`.
 
@@
 
-- `infercnv_obj` is the `infercnv` object
+- `infercnv_obj` is the `inferCNV` object
 
@@
 
-- `reference_value` indicates the reference used for `infercnv`, either endothelial, immune, both or pool.
+- `reference_value` indicates the reference used for `inferCNV`, either endothelial, immune, both or pool.
 
@@
 
-For a `Seurat` object an `infercnv` object created with the script `06_infercnv.R` using `reference_value` as a reference, the function `Do_CNV_score` calculate a CNV score per cell.
+For a `Seurat` object an `inferCNV` object created with the script `06_infercnv.R` using `reference_value` as a reference, the function `Do_CNV_score` calculate a CNV score per cell.
 The score is calculated based on the [this discussion](https://www.biostars.org/p/9573777/).
@@
 
-- `reference_value` is the selection of normal cells used for `infercnv`
-- `seurat_object` and `infercnv_obj` are the `Seurat`and `infercnv` object created with the script `06_infercnv.R`
+- `reference_value` is the selection of normal cells used for `inferCNV`
+- `seurat_object` and `infercnv_obj` are the `Seurat`and `inferCNV` object created with the script `06_infercnv.R`
 
@@
 
-#### Visualize seurat clusters and markers genes
+#### Visualize Seurat clusters and markers genes
 
-For a Seurat object `object`and a features `features`, the function `visualize_feature` will plot `FeaturePlot` and `ViolinPlot`
+For a Seurat object `object` and a features `features`, the function `visualize_feature` will plot `FeaturePlot` and `ViolinPlot`
 
@@
 
-#### Wrapper function to explore `infercnv` HMM CNV prediction
+#### Wrapper function to explore `inferCNV` HMM CNV prediction
 
-The `wrapper_explore_hmm` take as input the `infercnv_obj` generated with `infercnv` HMM CNV predictions.
+The `wrapper_explore_hmm` take as input the `infercnv_obj` generated with `inferCNV` HMM CNV predictions.
 The wrapper allows the following steps and plots:
@@
 For each chromosome, we look at the repartition of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
-`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+`proportion_cnv_` is the proportion in number of genes that are part of any CNV/loss/duplication in the given chr.
 
@@
 For each chromosome, we look at the distribution of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
-`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.
+`proportion_cnv_` is the proportion in number of genes that are part of any CNV/loss/duplication in the given chr.
 
@@
 
-### Heatmap of `infercnv` results
+### Heatmap of `inferCNV` results
 
-Here we plot the output of `infercnv` as heatmaps of CNV.
-We first look at the PNG file generated by the `infercnv` function.
+Here we plot the output of `inferCNV` as heatmaps of CNV.
+We first look at the PNG file generated by the `inferCNV` function.
 We then used the `infercnv object` to look at mean CNV value across compartments (immune, endothelial, stroma and fetal nephron).
@@
 
-Of note: By default if no reference is provided, `infercnv` take the mean of expression as normal reference. 
+Of note: By default if no reference is provided, `inferCNV` take the mean of expression as normal reference.
 The risk is that the main cell population (in our case the fetal nephron compartment) might be mistaken as the normal baseline.
@@
 
-We want to calculate a single CNV score and asesss if/how it can be use to define cells with CNV versus stable/normal cells. 
-We defined the score as described in the [biostar discussion](https://www.biostars.org/p/9573777/).
+We want to calculate a single CNV score and assess if/how it can be used to define cells with CNV versus stable/normal cells.
+We defined the score as described in the [discussion](https://www.biostars.org/p/9573777/).
 
@@
 
-We then explore `infercnv` results generated with immune and endothelial cells as reference, using a [HMM-i3 prediction models](https://github.com/broadinstitute/infercnv/wiki/infercnv-i3-HMM-type).
+We then explore `inferCNV` results generated with immune and endothelial cells as reference, using a [HMM-i3 prediction models](https://github.com/broadinstitute/infercnv/wiki/infercnv-i3-HMM-type).
 
@@
 
-We then explore `infercnv` results generated with immune and endothelial cells as reference, using a [HMM-i6 model](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type).
+We then explore `inferCNV` results generated with immune and endothelial cells as reference, using a [HMM-i6 model](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type).
 
@@
 
-We lastly explore `infercnv` results generated with a pull of immune and endothelial cells from different Wilms tumor samples as reference, using a HMM-i3 prediction model.
+We lastly explore `inferCNV` results generated with a pull of immune and endothelial cells from different Wilms tumor samples as reference, using a HMM-i3 prediction model.
 
-The selection of normal cells to be spiked-in as a normal reference prior running `infercnv` has been performed in `06b_build-normal-cell_reference.R.
+The selection of normal cells to be spiked-in as a normal reference prior running `inferCNV` has been performed in `06b_build-normal-cell_reference.R.
 We selected samples that haven't been pre-treated with chemotherapies.
@@
 
-We load the `Seurat` object generated in `06_infercnv.R` and transfer the `infercnv` metadata to the main `Seurat` object of sample `r params$sample_id` saved in `r glue::glue("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")`.
+We load the `Seurat` object generated in `06_infercnv.R` and transfer the `inferCNV` metadata to the main `Seurat` object of sample `r params$sample_id` saved in `r glue::glue("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")`.
 
@@
 
-- We should run `infercnv` with the most diverse normal reference as possible, including different cell types.
+- We should run `inferCNV` with the most diverse normal reference as possible, including different cell types.
 Providing no reference is not a good option, as we think that most of the cells are cancer cells with few CNV.
@@
 
-- The HMM prediction models help exploring the `infercnv` results. 
+- The HMM prediction models help exploring the `inferCNV` results.
 In this notebook, we have compared three HMM prediction models:
@@
 
-- Surprisingly, running `infercnv` with endothelial and immune cells from (i) the same patient or (ii) a set of Wilms tumor patients do not seem to affect drastically the results. 
+- Surprisingly, running `inferCNV` with endothelial and immune cells from (i) the same patient or (ii) a set of Wilms tumor patients do not seem to affect drastically the results.
 Some false positive CNV might occur in every patient due to the inter-patient variability.

### 2024-11-26 analyses/cell-type-ewings/template_notebooks/clustering-workflow/01-clustering-metrics.Rmd

@@
 stopifnot(
-  "sce file does not exist" = file.exists(params$sce_file)
+  "sce file does not exist" = file.exists(params$sce_file), 
+  "clustering results file does not exist" = file.exists(params$cluster_results_file)
 )

### 2024-11-26 analyses/cell-type-wilms-tumor-06/notebooks/02a_label-transfer_fetal_full_reference_Cao.Rmd
Debugs related to reorg into notebooks, and new data release which can have fewer than 500 cells

--- 2024-11-20 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
@@
 ---
-title: "Reference label transfer of the fetal kidney reference (Stewart et al.) for `r params$sample_id`"
+title: "Reference label transfer from the fetal full reference (Cao et al.) for `r params$sample_id`"
 author: "Maud PLASCHKA"
-date: '2024-08-13'
+date: '2024-08-07'
 params:
@@
   seed: 12345
+  homologs_file: "../scratch/homologs.rds"
   testing: 0
@@
 
-In this notebook, we perform label transfer from the human fetal kidney atlas from Stewart et al.
+In this notebook, we perform label transfer from the human fetal Azimuth reference from Cao et al.
 
@@
 library(Seurat)
+library(SeuratData)
 library(sctransform)
@@
 
-```{r base paths, eval=TRUE}
+```{r base paths, eval=TRUE, include=TRUE}
 # The base path for the OpenScPCA repository, found by its (hidden) .git directory
 repository_base <- rprojroot::find_root(rprojroot::is_git_root)
-
-# The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "current", params$scpca_project_id)
 
@@
     module_base,
-    "notebook_template",
+    "notebooks",
     "utils",
@@
 
-The reference has been download and pre-process in the `R Script` `download-and-create-fetal-kidney-ref.R`
+Load the Azimuth reference which has been prepared for label transfer.
 
@@
   "references",
-  "stewart_formatted_ref.rds"
+  "cao_formatted_ref.rds"
 )
@@
 
-In this notebook, we test the label transfer from the fetal kidney atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id` 
-We work with the pre-processed `Seurat`object and labeled with the fetal kidney reference saved in the `results` directory.
+In this notebook, we test the label transfer from the azimuth fetal atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`.
+
+We work with the pre-processed `Seurat`object saved in the `results` directory.
 
@@
 The pre-processed and annotated `Seurat` object per samples are saved in the `result` folder.
-
 ```{r path_to_output}
@@
 
-### Load the pre-processed `Seurat` object
+### Load the pre-process data (query)
+
 ```{r load, message=FALSE, warning=FALSE}
 # open the processed rds object
-srat <- readRDS(file.path(data_dir, paste0("02a-fetal_full_label-transfer_", params$sample_id, ".Rds")))
+srat <- readRDS(file.path(data_dir, paste0("01-Seurat_", params$sample_id, ".Rds")))
 srat_assay <- "RNA"
@@
 query <- prepare_query(
-  srat,
-  rownames(reference),
-  srat_assay,
-  convert_gene_names = FALSE
+  query = srat,
+  reference_rownames = rownames(reference),
+  assay = srat_assay,
+  homolog_file = params$homologs_file
 )
@@
 
-### Azimuth annotation from fetal kidney
+### Label transfer from fetal kidney reference using Azimuth
 
 ```{r run_azimuth, message=FALSE, warnings=FALSE}
-options(future.globals.maxSize = 8912896000000)
+options(future.globals.maxSize = 891289600000000)
 
@@
 
-srat <- AddMetaData(srat, metadata_to_transfer, col.name = paste0("fetal_kidney_", annotation_columns))
+srat <- AddMetaData(srat, metadata_to_transfer, col.name = paste0("fetal_full_", annotation_columns))
+
+rm(query)
+rm(query_labeled)
 ```
 
-```{r plot_azimuth, fig.height=8, fig.width=8, warnings=FALSE}
-d1 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_kidney_predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
-d2 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_kidney_predicted.cell_type", label = TRUE, repel = TRUE) + NoLegend()
+```{r plot_azimuth, fig.height=15, fig.width=8, warnings=FALSE}
+d1 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_full_predicted.organ", label = TRUE, repel = TRUE) + NoLegend()
+d2 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_full_predicted.annotation.l1", label = TRUE, repel = TRUE) + NoLegend()
+d3 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_full_predicted.annotation.l2", label = TRUE, repel = TRUE) + NoLegend()
 
@@
   sample = srat,
-  group.by = "fetal_kidney_predicted.compartment",
+  group.by = "fetal_full_predicted.organ",
   split.by = "seurat_clusters",
@@
   font.size = 10,
-  legend.ncol = 3
+  legend.ncol = 4
 ) +
@@
   sample = srat,
-  group.by = "fetal_kidney_predicted.cell_type",
+  group.by = "fetal_full_predicted.annotation.l1",
   split.by = "seurat_clusters",
@@
   font.size = 10,
-  legend.ncol = 3
+  legend.ncol = 2
 ) +
@@
 
-(d1 / f1) | (d2 / f2)
+f3 <- SCpubr::do_BarPlot(
+  sample = srat,
+  group.by = "fetal_full_predicted.annotation.l2",
+  split.by = "seurat_clusters",
+  position = "fill",
+  font.size = 10,
+  legend.ncol = 2
+) +
+  ggtitle("% cells") +
+  xlab(params$sample_id)
+
+((d1 / f1) | (d2 / f2))
 ```
-
-Note:
-
-For some reason, the "cap-mesenchyme cells" has been renamed in CELLxGENE as "mesenchymal cells".
-
-The cap mesenchyme is a cap of condensed metanephric mesenchyme, comprised of cells which epithelialize and sequentially form the peritubular aggregate (PA), renal vesicle (RV), C-, and S-shaped bodies, and finally the mature nephron.
-The CM contains nephron progenitor cells.
-
-This can be confusing and we just need to pay attention that :
-
-- the fetal nephron / mesenchymal cells are cap-mesenchyme cells. 
-In our case, cap-mesenchyme contains blastema and primitive epithelial cancer cells.
-
-- the stroma / mesenchymal stem are likely mesenchymal cancer or normal cells.
 
@@
 ```{r save}
-saveRDS(object = srat, file = file.path(output_dir, paste0("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")))
+saveRDS(object = srat, file = file.path(output_dir, paste0("02a-fetal_full_label-transfer_", params$sample_id, ".Rds")))
 ```

### 2024-11-26 analyses/cell-type-wilms-tumor-06/notebooks/03_clustering_exploration.Rmd
Debugs related to reorg into notebooks, and new data release which can have fewer than 500 cells

--- 2024-11-18 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
@@
   suppressWarnings({
+    # AddModuleScore is known to throw an error when there are too few cells, e.g.:
+    # https://github.com/satijalab/seurat/issues/5814 (the source of this error is the call to AddModuleScore)
+    # When we have fewer than ~500 cells, this error will occur. To avoid this, we override the `ctrl` (default 100) parameter to a lower value (15)
+    if (ncol(object) <= 500) {
+      object <- AddModuleScore(object, features = set_list, name = name, nbin = nbin, ctrl = 15)
+    } else {
       object <- AddModuleScore(object, features = set_list, name = name, nbin = nbin)
+    }
   })
@@
 g2m.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$g2m.genes]
+
+# CellCycleSorting is known to throw an error when there are too few cells, e.g.:
+# https://github.com/satijalab/seurat/issues/5814
+# When we have fewer than ~500 cells, this error will occur. To avoid this, we override the `ctrl` (default 100) parameter to a lower value (15)
+if (ncol(srat) <= 500) {
+  srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE, nbin = seurat_nbins, ctrl = 15)
+} else {
   srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE, nbin = seurat_nbins)
+}
 ```
@@
 
-We used enricher function from `clusterProfiler` to perform enrichment analysis.
+We used `clusterProfiler::enricher()` to perform enrichment analysis.
 

### 2024-11-26 analyses/cell-type-wilms-tumor-06/supplemental-notebooks/compare-label-transfer-approaches.Rmd
Debugs related to reorg into notebooks, and new data release which can have fewer than 500 cells

@@
 source(
-  file.path(module_base, "notebook_template", "utils", "label-transfer-functions.R")
+  file.path(module_base, "notebooks", "utils", "label-transfer-functions.R")
 )

### 2024-12-05 analyses/cell-type-ewings/template_notebooks/clustering-workflow/02-clustering-celltypes.Rmd
update report to use a single resolution and algorithm

@@
   singler_results_file: "../../results/aucell_singler_annotation/SCPCS000490/SCPCL000822_singler-classifications.tsv"
-  cluster_results_file: "../../results/cluster_exploration/SCPCS000490/SCPCL000822_clusters.tsv"
+  cluster_results_file: "../../results/clustering/SCPCS000490/SCPCL000822_cluster-results.tsv"
   marker_genes_file: "../../references/visser-all-marker-genes.tsv"
+  algorithm: "leiden"
+  objective_function: "modularity"
+  resolution: 1
 ---
 
-TODO: Update this notebook to take as input a clustering algorithm, nn (range of nn?), resolution, and optional objective function to display in SingleR and marker gene plots
+This notebook will compare cluster assignments to cell type assignments and look at expression of tumor and normal marker genes. 
+We expect that "good" clusters will be compromised of cells belonging to the same cell type and express marker genes for a single cell type (e.g., only tumor or only immune markers).
 
-- The clusters are then compared to the results from running `SingleR` in the `aucell-singler-annotation.sh` workflow.
-- We then look at marker gene expression across the clusters and assign each cluster to a cell type.
+Clustering was calculated using the following parameters:
+
+- Algorithm: `r params$algorithm` 
+- Objective function: `r params$objective_function`
+- Resolution: `r params$resolution`
+
+Results will be shown for all values of nearest neighbors used to perform clustering found in the cluster results file.
 
@@
   "SingleR results file does not exist" = file.exists(params$singler_results_file),
+  "Clustering results file does not exist" = file.exists(params$cluster_results_file),
   "Marker genes file does not exist" = file.exists(params$marker_genes_file)
@@
 
-# source in helper functions: calculate_sum_markers()
+# source in helper functions: calculate_sum_markers() and plot_faceted_umap()
 validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
@@
 # read in singler results
-singler_classification_df <- readr::read_tsv(params$singler_results_file) |>
-  dplyr::rename("cell_id" = "barcodes")
+singler_classification_df <- readr::read_tsv(params$singler_results_file)
+
+# read in clustering results 
+clusters_df <- readr::read_tsv(params$cluster_results_file) |> 
+  # filter to specified parameters 
+  dplyr::filter(algorithm == params$algorithm,
+                objective_function == params$objective_function,
+                resolution == params$resolution) |> 
+  # modify the nn column to be a character for easy plot labels
+  dplyr::mutate(cluster = as.factor(cluster),
+                nn_char = glue::glue("nn-{nn}") |> 
+                  # make sure 5 comes first 
+                  forcats::fct_relevel("nn-5", after = 0)) |> 
+  # rename barcode column for joining
+  dplyr::rename(barcodes = "cell_id")
 ```
 
-## Clusters vs. `SingleR` {.tabset}
+## UMAP of cluster assignments
+
+```{r}
+# get umap embeddings and combine into a data frame with cluster assignments
+umap_df <- sce |>
+  scuttle::makePerCellDF(use.dimred = "UMAP") |>
+  # replace UMAP.1 with UMAP1 and get rid of excess columns
+  dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2) |>
+  dplyr::left_join(clusters_df, by = "barcodes")
+```
+
+```{r fig.height=10, fig.width=10}
+ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cluster)) +
+  geom_point(alpha = 0.5, size = 0.1) +
+  facet_wrap(vars(nn),
+             labeller = labeller(nn = ~ glue::glue("{.}-nn"))) +
+  theme(
+    aspect.ratio = 1,
+    legend.position = "none", 
+    panel.border = element_rect(color = "black", fill = NA)
+  )
+```
+
+## Clusters vs. `SingleR`
 
 Now we will compare the clustering results to the cell type assignments obtained from `SingleR` in the `aucell-singler-annotation.sh` workflow. 
-To compare results we will calculate the Jaccard similarity index between clusters and cell types. 
-We expect that good clustering will line up with the cell type assignments so that there is ~ 1 cell type per cluster.
-
 For the plots, we will only display the top 7 cell types identified by `SingleR` and all other cells will be lumped together into `All remaining cell types`.
@@
 ```{r}
-cluster_classification_df <- all_cluster_results |>
+# prep classifications and clusters for plots 
+cluster_classification_df <- umap_df |>
   # add in classifications from singler
-  dplyr::left_join(singler_classification_df, by = "cell_id") |>
+  dplyr::left_join(singler_classification_df, by = "barcodes") |>
   dplyr::mutate(
@@
   )
-
-cluster_classification_df |> 
-  split(cluster_classification_df$resolution) |> 
-  purrr::map(\(df){cluster_celltype_heatmap(df)})
-
 ```
 
-```{r, fig.height=10, fig.width=7}
+First, we just look at the cell type assignments on the UMAP.
+
+```{r fig.height=10, fig.width=10}
+celltype_plot_df <- cluster_classification_df |> 
+  dplyr::select(barcodes, UMAP1, UMAP2, singler_lumped)
+
+plot_faceted_umap(celltype_plot_df, singler_lumped)
+```
+
+To compare results we will calculate the Jaccard similarity index between clusters and cell types. 
+We expect that good clustering will line up with the cell type assignments so that there is ~ 1 cell type per cluster.
+
+```{r, fig.height=15, fig.width=7}
 # get heatmap for showing cluster assignments vs. singler assignments
@@
 
-## Marker gene expression {.tabset}
+## Marker gene expression
 
@@
 # join sum expression columns with clustering results
-cluster_exp_df <- all_cluster_results |>
+cluster_exp_df <- clusters_df |>
   dplyr::left_join(gene_exp_df, by = "barcodes")
@@
 # plot to look at marker gene expression for all cell types across all clusters and all params
-k_params <- unique(cluster_exp_df$nn_param) |>
-  as.character()
+k_params <- unique(cluster_exp_df$nn_char)
 
@@
   purrr::map(\(k) {
+    k_num <- stringr::str_remove(k, "nn-")
+
     plot_marker_genes(cluster_exp_df,
-      k_value = k
+      k_value = k_num
     )

### 2024-12-05 analyses/cell-type-ewings/template_notebooks/clustering-workflow/02-clustering-celltypes.Rmd
Apply suggestions from code review
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 - Algorithm: `r params$algorithm` 
-- Objective function: `r params$objective_function`
+- Objective function (if Leiden): `r params$objective_function`
 - Resolution: `r params$resolution`
@@
   dplyr::mutate(cluster = as.factor(cluster),
-                nn_char = glue::glue("nn-{nn}") |> 
                 # make sure 5 comes first 
-                  forcats::fct_relevel("nn-5", after = 0)) |> 
+                nn_char = forcats::fct_relevel(glue::glue("nn-{nn}"), "nn-5", after = 0)
+         ) |>
   # rename barcode column for joining

### 2024-12-06 analyses/cell-type-ewings/template_notebooks/clustering-workflow/02-clustering-celltypes.Rmd
use nn not k

@@
 
+## Number of clusters
+
+Below is a table showing the number of clusters found for each value of `nn`.
+
+```{r}
+clusters_df |> 
+  dplyr::select(nn, cluster) |> 
+  dplyr::group_by(nn) |> 
+  dplyr::summarise(num_of_clusters = length(unique(cluster)))
+```
+
 ## UMAP of cluster assignments
@@
 
-Finally, we will look at the expression of marker genes across each cluster for each value of `k`. 
+Finally, we will look at the expression of marker genes across each cluster for each value of `nn`. 
 In these plots, each row shows the distribution of the specified marker genes in that cluster. 

### 2024-12-06 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
Add missing chunk name, and regenerate with script

@@
 author: "Data Lab"
-params:
-  seed: 2024
 output: 
@@
 ```{r set seed}
-set.seed(params$seed)
+set.seed(2024)
 ```
@@
 
-```{r}
+```{r use pc_name}
 # First, we'll rename the PCA matrix for demonstration

### 2024-12-12 analyses/cell-type-consensus/exploratory-notebooks/01-reference-exploration.Rmd
consensus labels from references notebook

---
title: "Summary of cell type ontologies in reference files"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
    df_print: "kable"
---

This notebook aims to identify a set of consensus labels between cell types in the PanglaoDB and Blueprint Encode references.

- First I look at the entire cell type ontology and summarize the total ancestors and descendants. 
- Then we find all possible combinations of cell type labels between PanglaoDB and Blueprint Encode and find the latest common ancestor (LCA) for all combinations.
- The total descendants for each LCA is used to define a cutoff for consensus terms we may want to use.
- I also explored the pairs for some terms in depth to look at borderline LCA terms. 
- Finally, I calculated the similarity index for all pairs.

## Setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_bw()
)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The path to this module
ref_dir <- file.path(repository_base, "analyses", "cell-type-consensus", "references")

# path to ref file for panglao 
panglao_file <- file.path(ref_dir, "panglao-cell-type-ontologies.tsv")
```

```{r}
# grab obo file 
cl_ont <- ontologyIndex::get_ontology("http://purl.obolibrary.org/obo/cl-basic.obo")

# read in panglao file 
panglao_df <- readr::read_tsv(panglao_file) |>
  # rename columns to have panglao in them for easy joining later 
  dplyr::select(
    panglao_ontology = "ontology_id",
    panglao_annotation = "human_readable_value"
  )

# grab singler ref from celldex
blueprint_ref <- celldex::BlueprintEncodeData() 
# get ontologies and human readable name into data frame
blueprint_df <- data.frame(
  blueprint_ontology = blueprint_ref$label.ont,
  blueprint_annotation_main = blueprint_ref$label.main,
  blueprint_annotation_fine = blueprint_ref$label.fine
) |> 
  unique()
```

## Full cell ontology

Below I will calculate the total number of ancestors and the total number of descendants for each term in the full cell type ontology and then show the distributions for those statistics.
This will give us an idea of the range of values we expect to see when looking at the PanglaoDB and Blueprint Encode references.

```{r}
# turn cl_ont into data frame with one row per term 
cl_df <- data.frame(
  cl_ontology = cl_ont$id,
  cl_annotation = cl_ont$name
) |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # list all ancestors and descendants calculate total
    ancestors = list(ontologyIndex::get_ancestors(cl_ont, cl_ontology)),
    total_ancestors = length(ancestors),
    descendants = list(ontologyIndex::get_descendants(cl_ont, cl_ontology)),
    total_descendants = length(descendants)
  )
```

The vertical lines in the below plot indicate the value for cell types of varying granularity.

```{r}
celltypes_of_interest <- c("eukaryotic cell", "lymphocyte", "leukocyte", "hematopoietic cell", "T cell", "endothelial cell", "smooth muscle cell", "memory T cell")
line_df <- cl_df |> 
  dplyr::filter(cl_annotation %in% celltypes_of_interest) |> 
  dplyr::select(cl_annotation, total_descendants, total_ancestors) |> 
  unique()

# group any labels that have the same number of ancestors 
ancestor_labels_df <- line_df |> 
  dplyr::group_by(total_ancestors) |> 
  dplyr::summarise(cl_annotation = paste(cl_annotation, collapse = ","))
```

```{r}
# make density plots showing distribution of ancestors and descendants 
ggplot(cl_df, aes(x = total_ancestors)) +
  geom_density(fill = "#00274C", alpha = 0.5) +
  geom_vline(data = ancestor_labels_df,
             mapping = aes(xintercept = total_ancestors), 
             lty = 2) +
  geom_text(
    data = ancestor_labels_df,
    mapping = aes(x = total_ancestors, y = 0.04, label = cl_annotation),
    angle = 90, 
    vjust = -0.5
  ) +
  labs(
    x = "Number of ancestors",
    y = "Density"
  )
```

Generally it looks like as the cell types get more specific we see a greater number of ancestors.
However, the range of values is small and we see some cell types have the same value and probably not the same level of granularity.

Below we will look at total number of descendants.

```{r}
ggplot(cl_df, aes(x = total_descendants)) +
  geom_density(fill = "#FFCB05", alpha = 0.5) +
  geom_vline(data = line_df,
             mapping = aes(xintercept = total_descendants), 
             lty = 2) +
  geom_text(
    data = line_df,
    mapping = aes(x = total_descendants, y = 0.6, label = cl_annotation),
    angle = 90, 
    vjust = -0.5
  ) +
  labs(
    x = "Number of descendants",
    y = "Density"
  )
```

It looks like most cell types have very few descendants, so let's zoom into the area below 500 to get a better look.

```{r}
ggplot(cl_df, aes(x = total_descendants)) +
  geom_density(fill = "#FFCB05", alpha = 0.5) +
  geom_vline(data = line_df,
             mapping = aes(xintercept = total_descendants), 
             lty = 2) +
  geom_text(
    data = line_df,
    mapping = aes(x = total_descendants, y = 0.6, label = cl_annotation),
    angle = 90, 
    vjust = -0.5
  ) +
  labs(
    x = "Number of descendants",
    y = "Density"
  ) +
  xlim(c(0,500))
```

Here we see a much larger range of values and that cell types become more general as the number of descendants goes up. 
However, this distribution alone is probably not helpful in determining a cutoff. 
The next section we will look at this distribution specifically for cell types present in our references, PanglaoDB and Blueprint encode.

## Latest common ancestor (LCA) between PanglaoDB and Blueprint encode

This section will look at identifying the latest common ancestor (LCA) between all possible combinations of terms from PanglaoDB (used for assigning cell types with `CellAssign`) and the `BlueprintEncodeData` reference from `celldex` (used for assigning cell types with `SingleR`). 
The LCA refers to the latest term in the cell ontology heirarchy that is common between two terms. 
I will use the [`ontoProc::findCommonAncestors()` function](https://rdrr.io/bioc/ontoProc/man/findCommonAncestors.html) to get the LCA for each combination.

Note that it is possible to have more than one LCA for a set of terms. 
To start, I will keep all LCA terms found.

For each LCA, I will again look at the total number of ancestors and descendants and see if I can identify an appropriate cutoff. 
Ultimately, I would like to see if we can use that cutoff to decide if we should keep the LCA term as the consensus label or use "Unknown".

```{r}
# first set up the graph from cl ont
parent_terms <- cl$parents
cl_graph <- igraph::make_graph(rbind(unlist(parent_terms), rep(names(parent_terms), lengths(parent_terms))))
```

```{r}
# get a data frame with all combinations of panglao and blueprint terms
# one row for each combination 
all_ref_df <- expand.grid(panglao_df$panglao_ontology,
                          blueprint_df$blueprint_ontology) |> 
  dplyr::rename(
    panglao_ontology = "Var1",
    blueprint_ontology = "Var2"
  ) |> 
  # add in the human readable values for each ontology term
  dplyr::left_join(blueprint_df, by = "blueprint_ontology") |> 
  dplyr::left_join(panglao_df, by = "panglao_ontology") |> 
  tidyr::drop_na() |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # least common shared ancestor
    lca = list(rownames(ontoProc::findCommonAncestors(blueprint_ontology, panglao_ontology, g = g)))
  )

lca_df <- all_ref_df |> 
  dplyr::mutate(
    total_lca = length(lca), # max is three terms 
    lca = paste0(lca, collapse = ",") # make it easier to split the df 
  ) |>
  # split each lca term into its own column 
  tidyr::separate(lca, into = c("lca_1", "lca_2", "lca_3"), sep = ",") |> 
  tidyr::pivot_longer(
    cols = dplyr::starts_with("lca"),
    names_to = "lca_number",
    values_to = "lca"
  ) |> 
  tidyr::drop_na() |> 
  dplyr::select(-lca_number) |> 
  # account for any cases where the ontology IDs are exact matches 
  # r complains about doing this earlier since the lca column holds lists until now
  dplyr::mutate(lca = dplyr::if_else(blueprint_ontology == panglao_ontology, blueprint_ontology, lca)) |> 
  # join in information for each of the lca terms including name, number of ancestors and descendants 
  dplyr::left_join(cl_df, by = c("lca" = "cl_ontology")) 
```

### Distribution of ancestors and descendants

```{r}
ggplot(lca_df, aes(x = total_ancestors)) +
  geom_density() +
  geom_vline(data = ancestor_labels_df,
             mapping = aes(xintercept = total_ancestors), 
             lty = 2) +
  geom_text(
    data = ancestor_labels_df,
    mapping = aes(x = total_ancestors, y = 0.6, label = cl_annotation),
    angle = 90, 
    vjust = -0.5
  ) +
  labs(
    x = "Total number of ancestors",
    y = "Density"
  )
```

```{r}
ggplot(lca_df, aes(x = total_descendants)) +
  geom_density() +
  geom_vline(data = line_df,
             mapping = aes(xintercept = total_descendants), 
             lty = 2) +
  geom_text(
    data = line_df,
    mapping = aes(x = total_descendants, y = 0.002, label = cl_annotation),
    angle = 90, 
    vjust = -0.5
  ) +
  labs(
    x = "Total number of descendants",
    y = "Density"
  )
```

Let's zoom into the area below 1000, since we already know we would want to exlude anything above that based on this plot.

```{r}
ggplot(lca_df, aes(x = total_descendants)) +
  geom_density() +
  geom_vline(data = line_df,
             mapping = aes(xintercept = total_descendants), 
             lty = 2) +
  geom_text(
    data = line_df,
    mapping = aes(x = total_descendants, y = 0.002, label = cl_annotation),
    angle = 90, 
    vjust = -0.5
  ) +
  xlim(c(0, 1000)) +
  labs(
    x = "Total number of descendants",
    y = "Density"
  )
```

We can use the vertical lines for cells of interest to help us define a potential cutoff based on the granularity we would like to see in our consensus label. 
We want to be able to label things like T cell, but we don't want to label anything as lymphocyte as that's probably not helpful. 
I don't see any obvious cutoffs that may be present in the total number of ancestors, but the number of descendants is likely to be informative. 
I think it might be a good idea to start by drawing a line at the local maxima between the T cell and lymphocyte lines on the number of descendants graph.

### Defining a cutoff for number of descendants

First we will find the value for the first peak shown in the distribution. 
This is likely to be a good cutoff for deciding which LCA labels to keep.

```{r}
peak_idx <- splus2R::peaks(lca_df$total_descendants)
cutoff <- lca_df$total_descendants[peak_idx] |> 
  min() # find the smallest peak and use that as the cutoff for number of descendants 
```

Below is the list of all consensus cell type labels that we will be keeping if we were to just use this cutoff.

```{r}
celltypes_to_keep <- lca_df |> 
  dplyr::filter(total_descendants <= cutoff) |> 
  dplyr::pull(cl_annotation) |> 
  unique()

celltypes_to_keep
```

We can also look at all the cell types we are keeping and the total number of descendants to see if there are any that may be we don't want to include because the term is too braod.

```{r}
# pull out the cell types and total descendants for cell types to keep
plot_celltype_df <- lca_df |>
  dplyr::filter(cl_annotation %in% celltypes_to_keep) |>
  dplyr::select(cl_annotation, total_descendants) |> 
  unique()

# bar chart showing total number of descendants for each cell type
ggplot(plot_celltype_df, aes(x = reorder(cl_annotation, total_descendants), y = total_descendants)) +
  geom_bar(stat = "identity") +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  labs(
    x = "cell type",
    y = "Total descendants"
  )
```

There are a few terms that I think might be more broad than we want like `blood cell`, `bone cell`, `supporting cell`, and `lining cell`. 
I'm on the fence about keeping `myeloid leukocyte` and `progenitor cell`. 
I think if we wanted to remove those terms we could move our cutoff to be the same number of descendants as `T cell`, since we do want to keep that.

One could also argue to remove `stromal cell` or `extracellular matrix secreting cell`.

Below are tables that look specifically at the combinations of cell type annotations that resulted in some of the terms that I might consider removing.

```{r}
print_df <- lca_df |> 
  dplyr::select(blueprint_annotation_main, blueprint_annotation_fine, panglao_annotation, total_lca, cl_annotation)

# blood cell
print_df |> 
  dplyr::filter(cl_annotation == "blood cell") 
```

I think I'm in favor of not having a "blood cell" label, since I'm not sure that it's helpful. 
Also, if two different methods label something a platelet and a neutrophil, then perhaps that label is inaccurate and it's really a tumor cell.

```{r}
# bone cell
print_df |> 
  dplyr::filter(cl_annotation == "bone cell")
```

I think I would also remove bone cell, since hematopoietic stem cells and osteoclasts seem pretty different to me.

```{r}
# myeloid leukocyte cell
print_df |> 
  dplyr::filter(cl_annotation == "myeloid leukocyte")
```

I'm torn on this one, because I do think it's helpful to know if something is of the myeloid lineage, but if we aren't keeping lymphocyte then I would argue we shouldn't keep myeloid leukocyte.

```{r}
# progenitor cell
print_df |> 
  dplyr::filter(cl_annotation == "progenitor cell") |> 
  head(n=15) # there's a lot of these so let's only print out some 
```

Same with `progenitor cell`, I do think it could be helpful to know that something may be a progenitor cell, but when you have a cell with the label for HSC and the label for cells like monocytes or osteoblasts, then maybe we are talking about a tumor cell instead.

Along those same lines, I think the below terms, `lining cell` and `supporting cell`, are too broad even though they have few descendants.

```{r}
# lining cell 
print_df |> 
  dplyr::filter(cl_annotation == "lining cell")
```

```{r}
# supporting cell 
print_df |> 
  dplyr::filter(cl_annotation == "supporting cell")
```

We can also look at what cell type labels we are excluding when using this cut off to see if there are any terms we might actually want to keep instead.

```{r}
lca_df |> 
  dplyr::filter(total_descendants > cutoff) |> 
  dplyr::pull(cl_annotation) |> 
  unique()
```

The only term in this list that I would be concerned about losing is "neuron". 
Let's look at those combinations.

```{r}
# blood cell
print_df |> 
  dplyr::filter(cl_annotation == "neuron")
```

It looks like there are a lot of types of neurons in the PanglaoDB reference and only "neuron" as a term in Blueprint. 
Even though neuron has ~ 500 descendants, I think we should keep these labels.

### Removing anything with more than 1 LCA

One thing I noticed when looking at the labels that have less than the cutoff is that most of them are from scenarios where we have multiple LCAs. 
Maybe in the case where we have multiple LCAs we are already too broad and we should just eliminate those matches from the beginning. 
Here I'm looking at the total number of descendants for all terms that show up because a term has multiple LCAs.

```{r}
lca_df |> 
  dplyr::filter(total_lca > 1) |> 
  dplyr::select(cl_annotation, total_descendants) |> 
  unique()
```

It looks like most of these terms are pretty broad and are either much higher than the cutoff or right around the cutoff with a few exceptions.
Things like "bone cell" and "supporting cell" have few descendants, but I would still argue these are very broad terms and not useful.

I'm going to filter out any matches that show two LCA terms first and then use the cutoff to define labels we would keep. 
I'll also look to see what cell types we lose when we add this extra filtering step to be sure they are ones that we want to lose.

```{r}
# remove any combinations with more than one lca
filtered_lca_df <- lca_df |> 
  dplyr::filter(total_lca < 2)

# get a list of cell types to keep based on cutoff 
updated_celltypes <- filtered_lca_df |> 
  dplyr::filter(total_descendants <= cutoff) |> 
  dplyr::pull(cl_annotation) |> 
  unique()

# which cell types are now missing from the list to keep
setdiff(celltypes_to_keep, updated_celltypes)
```

It looks like I am losing a few terms I already said were not specific and then a few other terms, like "hematopoietic precursor cell" and "perivascular cell". I'll look at both of those to confirm we would not want them.

```{r}
print_df |> 
  dplyr::filter(cl_annotation == "hematopoietic precursor cell")
```

It looks like here we should be keeping these matches because both references have these labels as hematopoietic stem and progenitor cells. 
I think in the context of pediatric cancer having this label would be helpful, so maybe we shouldn't remove all terms that have 2 LCAs.

Let's look at what the other LCA is for an example set.

```{r}
lca_df |> 
  dplyr::filter(panglao_ontology == "CL:0000037" & blueprint_ontology == "CL:0000050") |> 
  dplyr::select(blueprint_annotation_main, blueprint_annotation_fine, panglao_annotation, cl_annotation)
```

It looks like these terms have both `hematopoietic precursor cell` and `progenitor cell` as LCAs. 
Personally, I would keep the term for `hematopoietic precursor cell` because I think it's more informative and specific to the type of progenitor cell.

```{r}
print_df |> 
  dplyr::filter(cl_annotation == "perivascular cell")
```

I would remove `perivascular cell`, since the cell type labels from PanglaoDB and Blueprint are pretty different from each other.

## Similarity index

An alternative approach would be to calculate the [similarity index](https://cran.r-project.org/web/packages/ontologySimilarity/vignettes/ontologySimilarity-introduction.html) between each set of terms and define a cutoff for which set of terms are similar. 
This is a value on a 0-1 scale where 0 indicates no similarity and 1 indicates the terms are equal.

Although this could provide a metric that we could use to define similar cell types, we would still have to identify the label to use which would most likely be the LCA. 
Even if the similarity index is close to 1, if the LCA term is not informative then I don't know that we would want to use that.

However, we could use this to finalize the actual pairs of terms that we trust. 
For example, if the LCA for a pair is `T cell` we can look at the similarity index to confirm that specific pair of terms has high similarity.

Below I'll calculate the similarity index for each set of terms and plot the distribution. 
Then we will look at the values for pairs that have an LCA that pass the total descendants threshold we set to see if those pairs have a higher similarity index.

```{r}
information_content <- ontologySimilarity::descendants_IC(cl_ont)

# get similarity index for each set of terms 
si_df <- lca_df |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    similarity_index = ontologySimilarity::get_sim_grid(ontology = cl_ont,
                                                        term_sets = list(panglao_ontology, blueprint_ontology)) |> 
      ontologySimilarity::get_sim()
  )
```

```{r}
si_df <- si_df |> 
  dplyr::mutate(
    lca_threshold = dplyr::if_else(total_descendants < cutoff, "PASS", "FAIL")
  )

ggplot(si_df, aes(x = similarity_index, fill = lca_threshold)) +
  geom_density(bw = 0.05, alpha = 0.5) +
  labs(
    x = "Similarity index",
    y = "Density"
  )
```

This looks as I expected with most of the pairs that pass the total descendants cutoff having a higher similarity index than those that do not pass. 
There is still some overlap though so perhaps even if a set of terms shares an LCA that passes the threshold, the actual terms being compared may be further apart than we would like.

Now let's look at the similarity index for various LCA terms. 
Here each LCA term is its own plot and the vertical lines are the similarity index for each pair of terms that results in that LCA.

```{r}
celltypes_to_plot <- c("myeloid leukocyte", "T cell", "cell", "supporting cell", "B cell")

celltypes_to_plot |> 
  purrr::map(\(celltype){
    line_df <- si_df |> 
      dplyr::filter(cl_annotation == celltype) |> 
      dplyr::select(cl_annotation, similarity_index) |> 
      unique()

    ggplot(si_df, aes(x = similarity_index)) +
      geom_density() +
      geom_vline(data = line_df,
                 mapping = aes(xintercept = similarity_index), 
                 lty = 2) +
      labs(
        x = "Similarity index",
        y = "Density",
        title = celltype
      )

  })
```

It looks like terms that are more granular like T and B cell have higher similarity index values than terms that are less granular which is what we would expect. 
However, within terms like myeloid leukocyte and even T cell we do see a range of values. 
We could dig deeper into which pairs are resulting in which similarity index values if we wanted to, but I think that might be a future direction if we feel like the similarity index is something that could be useful.

## Conclusions

Based on these findings, I think it might be best to create a reference that has all possible pairs of labels between PanglaoDB and Blueprint Encode and the resulting consensus label for those pairs. 
To do this we could come up with a whitelist of LCA terms that we would be comfortable including and all other cell types would be unknowns. 
I would use the following criteria to come up with my whitelist:

- Pairs should not have more than 1 LCA, with the exception of the matches that have the label hematopoietic precursor cell. 
- The LCA should have equal to or less than 170 total descendants. 
- We whould include the term for `neuron` even though it has 500 descendants. 
- Terms that are too broad (like `supporting cell`, `blood cell`, `bone cell`, `lining cell`) should be removed.

Alternatively, rather than eliminate terms that are too broad we could look at the similarity index for individual matches and decide on a case by case basis if those should be allowed. 
Although I still think having a term that is too braod, even if it's a good match, is not super informative.

## Session info

```{r}
sessionInfo()
```

### 2024-12-12 analyses/cell-type-consensus/exploratory-notebooks/01-reference-exploration.Rmd
add some headers for tables

@@
 
+#### Blood cell
+
 ```{r}
 print_df <- lca_df |> 
-  dplyr::select(blueprint_annotation_main, blueprint_annotation_fine, panglao_annotation, total_lca, cl_annotation)
+  dplyr::select(blueprint_ontology, blueprint_annotation_main, blueprint_annotation_fine, panglao_ontology, panglao_annotation, total_lca, lca, cl_annotation)
 
@@
 
+#### Bone cell
+
 ```{r}
@@
 
+#### Myeloid leukocyte
+
 ```{r}
@@
 I'm torn on this one, because I do think it's helpful to know if something is of the myeloid lineage, but if we aren't keeping lymphocyte then I would argue we shouldn't keep myeloid leukocyte.
+
+#### Progenitor cell
 
@@
 
+#### Lining cell
+
 ```{r}
@@
 
+#### Supporting cell
+
 ```{r}
@@
 ```
+
+### Discarded cell types
 
@@
 Let's look at those combinations.
+
+#### Neuron
 
@@
   dplyr::select(cl_annotation, total_descendants) |> 
-  unique()
+  unique() |>
+  dplyr::arrange(total_descendants)
 ```
@@
 
+#### Hematopoietic precursor cell
+
 ```{r}
@@
 Personally, I would keep the term for `hematopoietic precursor cell` because I think it's more informative and specific to the type of progenitor cell.
+
+#### Perivascular cell
 

### 2024-12-13 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 
-We'll also extract its PCA matrix, which we'll use to demonstrate how to calculate and evaluate clusters.
+For the initial cluster calculations and evaluations, we will use the PCA matrix extracted from the SCE object.
+It's also possible to use an SCE object or a Seurat object directly, which we will demonstrate later.
 We'll show how to do this for both SCE and Seurat objects:
@@
 
-# Print resulting table
-cluster_results_df
+# Print the first rows of the resulting table
+head(cluster_results_df)
 ```
@@
 
-Even if you perform clustering with other tools, you can still use the `rOpenScPCA` functions to evaluate your clusters.
+If you already have clustering results calculated with other tools, you can still use the `rOpenScPCA` functions to evaluate your clusters.
 
@@
   cell_id = colnames(sce),
-  cluster = colData(sce)$cluster
+  cluster = sce$cluster
 )
@@
 cluster_results_renamed_df <- cluster_results_df |>
-  # add the prefix openscpca_ to all columns
-  dplyr::rename_with(~ paste0("ropenscpca_", .x))
+  # add the prefix "ropenscpca_" to all columns
+  dplyr::rename_with(\(x) paste0("ropenscpca_", x))
 

### 2024-12-13 analyses/cell-type-ewings/template_notebooks/auc-singler-workflow/02-singler-results.Rmd

@@
 
+```{r}
+# check that marker genes are expressed, otherwise turn off those plots
+total_exp <- sum(classification_df[marker_gene_columns])
+if(total_exp > 0){
+  show_marker_gene_plots <- TRUE
+} else {
+  show_marker_gene_plots <- FALSE
+  message("No expression of marker genes in this library. No plots will be displayed in sections labeled 'Marker gene expression'.")
+}
+
+```
+
+```{r}
+# check that gene set scores aren't all 0, otherwise turn off those plots
+total_score <- sum(classification_df[geneset_columns])
+if(total_score > 0){
+  show_gene_set_plots <- TRUE
+} else {
+  show_gene_set_plots <- FALSE
+  message("Genes present in provided gene sets are not expressed in this library. No plots will be displayed in sections labeled 'Gene set scores'.")
+}
+
+```
+
 ### Tumor vs. Normal
@@
 
-```{r}
+```{r, eval=show_marker_gene_plots}
 full_celltype_heatmap(classification_df, marker_gene_columns, "singler_tumor_normal") 
@@
 
-```{r}
+```{r, eval=show_marker_gene_plots}
 plot_density(
@@
 
-```{r}
+```{r, eval=show_gene_set_plots}
 full_celltype_heatmap(classification_df, geneset_columns, "singler_tumor_normal")
@@
 
-```{r, fig.height=10}
+```{r, fig.height=10, eval=show_gene_set_plots}
 geneset_columns |>
@@
 
-```{r}
+```{r, eval=show_marker_gene_plots}
 full_celltype_heatmap(classification_df, marker_gene_columns, "singler_lumped")
@@
 
-```{r, fig.height=10}
+```{r, fig.height=10, eval=show_marker_gene_plots}
 marker_gene_columns |>
@@
 
-```{r}
+```{r, eval=show_gene_set_plots}
 full_celltype_heatmap(classification_df, geneset_columns, "singler_lumped")
@@
 
-### Tumor cells annotatd by `SingleR` and `AUCell`
+### Tumor cells annotated by `SingleR` and `AUCell`
 
@@
 
-```{r}
+```{r, eval=show_marker_gene_plots}
 full_celltype_heatmap(classification_df, marker_gene_columns, "consensus")
@@
 
-```{r}
+```{r, eval=show_gene_set_plots}
 full_celltype_heatmap(classification_df, geneset_columns, "consensus")

### 2024-12-13 analyses/cell-type-ewings/template_notebooks/clustering-workflow/01-clustering-metrics.Rmd

@@
   cluster_results_file: "../../results/clustering/SCPCS000490/SCPCL000822_cluster-results.tsv"
+  threads: 1
 ---
@@
 # read in results file 
-cluster_results_df <- readr::read_tsv(params$cluster_results_file)
+cluster_results_df <- readr::read_tsv(params$cluster_results_file) |> 
+  dplyr::mutate(cluster = as.factor(cluster))
 ```
@@
 stability_stats <- split_umap_df |> 
-  purrr::map(\(df){ get_cluster_stability(sce, df)})
+  purrr::map(\(df){ get_cluster_stability(sce, df, params$threads)})
 ```

### 2024-12-13 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd

--- 2024-11-01 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
@@
 query <- prepare_query(
-  srat,
-  rownames(reference),
-  srat_assay,
-  params$homologs_file
+  query = srat,
+  reference_rownames = rownames(reference),
+  assay = srat_assay,
+  homolog_file = params$homologs_file
 )

### 2024-12-13 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd

@@
 g2m.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$g2m.genes]
+# CellCycleSorting is known to throw an error when there are too few cells, e.g.:
+# https://github.com/satijalab/seurat/issues/5814
+# When we have fewer than ~500 cells, this error will occur. To avoid this, we override the `ctrl` (default 100) parameter to a lower value (15)
+if (ncol(srat) <= 500) {
+  srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE, nbin = seurat_nbins, ctrl = 15)
+} else {
   srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE, nbin = seurat_nbins)
+}
 ```

### 2024-12-13 analyses/cell-type-wilms-tumor-06/supplemental_notebooks/characterize_fetal_kidney_reference_Stewart.Rmd

--- 2024-11-20 analyses/cell-type-wilms-tumor-06/notebook_template/00b_characterize_fetal_kidney_reference_Stewart.Rmd
@@
 
-The input file `r params$fetal_kidney_path` is the output of the script `prepare-fetal-references.R`.
+The input file `params$fetal_kidney_path` is the output of the script `scripts/prepare-fetal-references.R`.
 
-## Output file
+## Output files
 
-We will save the result of the differential expression analysis in `results/references/00b_marker_genes_fetal_kidney_Stewart.csv`.
-Notebook is saved in the `notebook/00-reference` directory.
+We will save the result of the differential expression analysis in `results/references/marker_<cell-type,compartment>_fetal_kidney_Stewart.csv`.
 
@@
 
-write_csv(de_results, file = file.path(path_to_output, "00a_marker_compartment_fetal_kidney_Stewart.csv"))
+write_csv(de_results, file = file.path(path_to_output, "marker_compartment_fetal_kidney_Stewart.csv"))
 ```
@@
 
-write_csv(de_results, file = file.path(path_to_output, "00a_marker_cell-type_fetal_kidney_Stewart.csv"))
+write_csv(de_results, file = file.path(path_to_output, "marker_cell-type_fetal_kidney_Stewart.csv"))
 ```

### 2024-12-13 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
respond to reviews, and use rOpenScPCA:: throughout

@@
 
-### Set the random seed
-
-Because clustering involves random sampling, it is important to set the random seed at the top of your analysis script or notebook to ensure reproducibilty.
-
-```{r set seed}
-set.seed(2024)
-```
-
 ### Paths
@@
 
-### Input file
-
-```{r paths}
+```{r input file path}
 # Path to processed SCE file for sample SCPCS000001
 input_sce_file <- file.path(data_dir, "SCPCP000001", "SCPCS000001", "SCPCL000001_processed.rds")
+```
+
+### Set the random seed
+
+Because clustering involves random sampling, it is important to set the random seed at the top of your analysis script or notebook to ensure reproducibilty.
+
+```{r set seed}
+set.seed(2024)
 ```
@@
 
-```{r create process seurat, message = FALSE}
-# Convert to a Seurat object
-seurat_obj <- CreateSeuratObject(counts = counts(sce), assay = "RNA")
-
-# Process the object with a Seurat pipeline to obtain clusters
-seurat_obj <- seurat_obj |>
-  SCTransform() |>
-  RunPCA() |>
-  FindNeighbors() |>
-  FindClusters()
-```
-
 For the initial cluster calculations and evaluations, we will use the PCA matrix extracted from the SCE object.
 It's also possible to use an SCE object or a Seurat object directly, which we will demonstrate later.
-We'll show how to do this for both SCE and Seurat objects:
 
@@
 # Extract the PCA matrix from an SCE object
-# Moving forward, we'll use this version for most examples
 pca_matrix <- reducedDim(sce, "PCA")
-
-# Extract the PCA matrix from a Seurat object
-seurat_pca_matrix <- Embeddings(seurat_obj, reduction = "pca")
 ```
@@
 # Calculate clusters with default parameters
-cluster_results_df <- calculate_clusters(pca_matrix)
+cluster_results_df <- rOpenScPCA::calculate_clusters(pca_matrix)
 
@@
 # Calculate clusters with non-default parameters
-cluster_results_df <- calculate_clusters(
+cluster_results_df <- rOpenScPCA::calculate_clusters(
   pca_matrix,
@@
 # calculate the silhouette width for each cell
-silhouette_results <- calculate_silhouette(
+silhouette_results <- rOpenScPCA::calculate_silhouette(
   pca_matrix,
@@
 
-# print result
-silhouette_results
+# Print the first rows of the resulting table
+head(silhouette_results)
 ```
@@
 # calculate the neighborhood purity for each cell
-purity_results <- calculate_purity(
+purity_results <- rOpenScPCA::calculate_purity(
   pca_matrix,
@@
 
-# print result
-purity_results
+# Print the first rows of the resulting table
+head(purity_results)
 ```
@@
 # calculate the stability of clusters
-stability_results <- calculate_stability(
+stability_results <- rOpenScPCA::calculate_stability(
   pca_matrix,
@@
 # Calculate clusters with non-default parameters
-cluster_df_leiden <- calculate_clusters(
+cluster_df_leiden <- rOpenScPCA::calculate_clusters(
   pca_matrix,
@@
 # Now, pass in the same arguments customizing parameters here
-stability_results_leiden <- calculate_stability(
+stability_results_leiden <- rOpenScPCA::calculate_stability(
   pca_matrix,
@@
 If you are analyzing your data with a Seurat pipeline that includes calculating clusters, you can use `rOpenScPCA` to evaluate them.
-We previously created a Seurat object called `seurat_obj` with clusters:
 
-```{r show seurat clusters}
-head(seurat_obj$seurat_clusters)
+To demonstrate this, we'll first create a Seurat object from our SCE object. 
+We'll use the function `rOpenScPCA::sce_to_seurat()` for this, setting `use_symbols = FALSE` to preserve Ensembl ids as the feature names rather than converting to gene symbols.
+
+```{r sce to seurat}
+# Convert the SCE to a Seurat object using rOpenScPCA
+seurat_obj <- rOpenScPCA::sce_to_seurat(sce, use_symbols = FALSE)
+
+seurat_obj
+```
+
+The `rOpenScPCA::sce_to_seurat()` function will carry over the existing SCE column `cluster` into the Seurat object, which we'll use here for demonstration. 
+If you calculate clusters within the Seurat ecosystem, you may need to use a different column name here, e.g. `seurat_clusters`.
+
+```{r seurat cluster}
+head(seurat_obj$cluster)
 ```
@@
 seurat_cluster_df <- data.frame(
-  cell_id = names(seurat_obj$seurat_clusters),
-  cluster = seurat_obj$seurat_clusters
+  cell_id = colnames(seurat_obj),
+  cluster = seurat_obj$cluster
 )
@@
 ```{r seurat silhouette}
-seurat_silhouette_df <- calculate_silhouette(
+seurat_silhouette_df <- rOpenScPCA::calculate_silhouette(
   seurat_pca_matrix,
@@
 )
-seurat_silhouette_df
+head(seurat_silhouette_df)
 ```
@@
 ```{r seurat purity}
-seurat_purity_df <- calculate_purity(
+seurat_purity_df <- rOpenScPCA::calculate_purity(
   seurat_pca_matrix,
@@
 )
-seurat_purity_df
+head(seurat_purity_df)
 ```
@@
 ScPCA cell metadata already contains a column called `cluster` with results from an automated clustering. 
-These clusters were calculated using `bluster`, the same tool that `rOpenScPCA` uses, with the following parameters:
+These clusters were calculated using `bluster`, the same tool that `rOpenScPCA` uses. 
+We can see the parameters it used from the SCE object's metadata; all other parameters were left at their default values.
 
-* The Louvain algorithm
-* Jaccard weighting
-* 20 nearest neighbors
+```{r extract cluster params}
+# Print the clustering algorithm used
+metadata(sce)$cluster_algorithm
+
+# Print the weighting scheme
+metadata(sce)$cluster_weighting
+
+# Print the number of nearest neighbors
+metadata(sce)$cluster_nn
+```
 
@@
 # Calculate silhouette width
-scpca_silhouette_df <- calculate_silhouette(pca_matrix, scpca_cluster_df)
-scpca_silhouette_df
+scpca_silhouette_df <- rOpenScPCA::calculate_silhouette(
+  pca_matrix,
+  scpca_cluster_df
+)
+head(scpca_silhouette_df)
 ```
@@
 # Calculate neighborhood purity
-scpca_purity_df <- calculate_purity(pca_matrix, scpca_cluster_df)
-scpca_purity_df
+scpca_purity_df <- rOpenScPCA::calculate_purity(
+  pca_matrix,
+  scpca_cluster_df
+)
+head(scpca_purity_df)
 ```
 
-When running `calculate_stability()`, we'll specify the same parameters originally used to build the clusters.
-We only recommend evaluating clusters with `calculate_stability()` if you know the original parameters used.
+When running `rOpenScPCA::calculate_stability()`, we'll specify the same parameters originally used to build the clusters.
+We only recommend evaluating clusters with `rOpenScPCA::calculate_stability()` if you know the original parameters used.
 
 ```{r scpca stability}
-scpca_stability_df <- calculate_stability(
+scpca_stability_df <- rOpenScPCA::calculate_stability(
   pca_matrix,
@@
 # Calculate clusters from an SCE object using default parameters
-cluster_results_df <- calculate_clusters(sce)
+cluster_results_df <- rOpenScPCA::calculate_clusters(sce)
 cluster_results_df
@@
 # Calculate clusters from a Seurat object using default parameters
-cluster_results_df <- calculate_clusters(seurat_obj)
+cluster_results_df <- rOpenScPCA::calculate_clusters(seurat_obj)
 cluster_results_df
@@
 If the PCA matrix you want to use in the object has a different name, you can provide the argument `pc_name`.
-We can see this below with an SCE object, for example:
-
-```{r use pc_name}
-# First, we'll rename the PCA matrix for demonstration
-reducedDimNames(sce) <- c("PCA_matrix", "UMAP")
-reducedDimNames(sce)
-
-# Calculate clusters from an SCE object using default parameters
-cluster_results_df <- calculate_clusters(
-  sce,
-  pc_name = "PCA_matrix"
-)
-cluster_results_df
-```
 
@@
 
-Follow these steps save the clustering results to a Seurat object's cell metadata:
+Follow these steps save the clustering results to a Seurat object's cell metadata.
+Note that the Seurat function `AddMetaData()`, used below, assumes that the row order in the provided data frame matches the order of cells in the Seurat object you're adding metadata to.
 

### 2024-12-13 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
rearrange calculating qc metrics on existing clusters

@@
 )
-
-# Print the first rows of the resulting table
-head(cluster_results_df)
 ```
@@
 )
-
-stability_results_leiden
 ```
 
-## Calculate QC metrics on existing clusters
+## Working with objects directly
+
+As presented above, `rOpenScPCA` clustering functions take a PCA matrix with row names representing unique cell ids as their first argument.
+
+Instead of a matrix, you can alternatively pass in an SCE or Seurat object that contains a matrix.
+
+We show an example of this below with and SCE object and `rOpenScPCA::calculate_clusters()`, but this will also work for any of the evaluation functions as well and has the same syntax for Seurat objects.
+
+```{r run on sce}
+# Calculate clusters from an SCE object using default parameters
+cluster_results_df <- rOpenScPCA::calculate_clusters(sce)
+cluster_results_df
+```
+
+`rOpenScPCA` assumes that the PCA matrix is named `PCA` in SCE objects, and `pca` in Seurat objects.
+If the PCA matrix you want to use in the object has a different name, you can provide the argument `pc_name`.
+
+## Calculating QC metrics on existing clusters
 
@@
 
-In this section, we'll present examples of how you can calculate the silhouette width, neighborhood purity, and cluster stability from existing cluster assignments.
+In this section, we'll present examples of how you can calculate the silhouette width, neighborhood purity, and cluster stability from existing cluster assignments within objects.
 
@@
 
-To demonstrate this, we'll first create a Seurat object from our SCE object. 
-We'll use the function `rOpenScPCA::sce_to_seurat()` for this, setting `use_symbols = FALSE` to preserve Ensembl ids as the feature names rather than converting to gene symbols.
+To demonstrate this, we'll convert our SCE object to a Seurat using the function `rOpenScPCA::sce_to_seurat()`.
+We'll set `use_symbols = FALSE` to preserve Ensembl ids as the feature names rather than converting to gene symbols.
+Then, we'll use a simple Seurat pipeline to obtain clusters.
 
-```{r sce to seurat}
-# Convert the SCE to a Seurat object using rOpenScPCA
+```{r sce to seurat, message = FALSE}
+# Convert the SCE to a Seurat object using rOpenScPCA, preserving Ensembl ids as feature names
 seurat_obj <- rOpenScPCA::sce_to_seurat(sce, use_symbols = FALSE)
+
+# Calculate clusters with Seurat
+seurat_obj <- seurat_obj |>
+  SCTransform() |>
+  RunPCA() |>
+  FindNeighbors() |>
+  FindClusters()
 
@@
 
-The `rOpenScPCA::sce_to_seurat()` function will carry over the existing SCE column `cluster` into the Seurat object, which we'll use here for demonstration. 
-If you calculate clusters within the Seurat ecosystem, you may need to use a different column name here, e.g. `seurat_clusters`.
-
-```{r seurat cluster}
-head(seurat_obj$cluster)
-```
-
-To calculate QC metrics on these clusters, we'll need two pieces of information:
-
-- The PCA matrix
-- A data frame with columns `cell_id` and `cluster`
-
-We'll create those here:
+To calculate QC metrics on these clusters, we'll need to create a data frame with columns `cell_id` and `cluster`:
 
 ```{r prepare seurat input}
-# Extract the PCA matrix from the Seurat object
-seurat_pca_matrix <- Embeddings(seurat_obj, reduction = "pca")
-
 # Create a data frame for input
@@
   cell_id = colnames(seurat_obj),
-  cluster = seurat_obj$cluster
+  cluster = seurat_obj$seurat_clusters
 )
+
+head(seurat_cluster_df)
 ```
 
-Now, we can run `calculate_silhouette()` and `calculate_purity()`:
+Now, we can run `rOpenScPCA::calculate_silhouette()` and `rOpenScPCA::calculate_purity()` using this data frame and the Seurat object:
 
@@
 seurat_silhouette_df <- rOpenScPCA::calculate_silhouette(
-  seurat_pca_matrix,
+  seurat_obj,
   seurat_cluster_df
 )
-head(seurat_silhouette_df)
 ```
@@
 seurat_purity_df <- rOpenScPCA::calculate_purity(
-  seurat_pca_matrix,
+  seurat_obj,
   seurat_cluster_df
 )
-head(seurat_purity_df)
 ```
@@
 
-scpca_cluster_df
+head(scpca_cluster_df)
 ```
 
-We can run evaluation functions using this data frame and the PCA matrix from the object.
+We can run evaluation functions using this data frame and the SCE object.
 
@@
 scpca_silhouette_df <- rOpenScPCA::calculate_silhouette(
-  pca_matrix,
+  sce,
   scpca_cluster_df
 )
-head(scpca_silhouette_df)
 ```
@@
 scpca_purity_df <- rOpenScPCA::calculate_purity(
-  pca_matrix,
+  sce,
   scpca_cluster_df
 )
-head(scpca_purity_df)
 ```
@@
 When running `rOpenScPCA::calculate_stability()`, we'll specify the same parameters originally used to build the clusters.
-We only recommend evaluating clusters with `rOpenScPCA::calculate_stability()` if you know the original parameters used.
+Generally speaking, we only recommend evaluating clusters with `rOpenScPCA::calculate_stability()` if you know the original parameters used.
 
@@
 scpca_stability_df <- rOpenScPCA::calculate_stability(
-  pca_matrix,
+  sce,
   scpca_cluster_df,
@@
 )
-scpca_stability_df
 ```
-
-## Working with objects directly
-
-As presented above, `rOpenScPCA` clustering functions take a PCA matrix with row names representing unique cell ids as their first argument.
-
-Instead of a matrix, you can alternatively pass in an SCE or Seurat object that contains a matrix.
-We show examples of this below with `rOpenScPCA::calculate_clusters()`, but this will also work for any of the evaluation functions as well:
-
-```{r run on sce}
-# Calculate clusters from an SCE object using default parameters
-cluster_results_df <- rOpenScPCA::calculate_clusters(sce)
-cluster_results_df
-```
-
-```{r run on seurat}
-# Calculate clusters from a Seurat object using default parameters
-cluster_results_df <- rOpenScPCA::calculate_clusters(seurat_obj)
-cluster_results_df
-```
-
-`rOpenScPCA` assumes that the PCA matrix is named `PCA` in SCE objects, and `pca` in Seurat objects (as they were in the above examples).
-If the PCA matrix you want to use in the object has a different name, you can provide the argument `pc_name`.
 

### 2024-12-13 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
speeling and less printing

@@
 
-Because clustering involves random sampling, it is important to set the random seed at the top of your analysis script or notebook to ensure reproducibilty.
+Because clustering involves random sampling, it is important to set the random seed at the top of your analysis script or notebook to ensure reproducibility.
 
@@
 
-colData(sce)
+head(colData(sce))
 ```
@@
 # Print the cell metadata
-seurat_obj@meta.data
+head(seurat_obj@meta.data)
 ```

### 2024-12-16 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
add some simple plots with a color i like that may not survive review

@@
   library(dplyr)
+  library(ggplot2)
 })
+
+# Set ggplot theme for plots
+theme_set(theme_bw())
 ```
@@
 
+We can visualize these results by plotting silhouette width across clusters as violin plots, for example:
+
+```{r violin silhouette}
+ggplot(silhouette_results) +
+  aes(x = cluster, y = silhouette_width) +
+  geom_violin(fill = "darkolivegreen3") +
+  labs(x = "Cluster", y = "Silhouette width")
+```
+
 ### Neighborhood purity
@@
 head(purity_results)
+```
+
+We can visualize these results by plotting purity clusters as violin plots, for example:
+
+```{r violin purity}
+ggplot(purity_results) +
+  aes(x = cluster, y = purity) +
+  geom_violin(fill = "darkolivegreen3") +
+  labs(x = "Cluster", y = "Neighborhood purity")
 ```
@@
 stability_results
+```
+
+We can visualize these results by plotting stability as a density plot, for example:
+
+```{r ari density}
+ggplot(stability_results) +
+  aes(x = ari) +
+  geom_density(color = "grey30", fill = "darkolivegreen3") +
+  labs(x = "Adjusted rand index across bootstrap replicates")
 ```
@@
 To demonstrate this, we'll convert our SCE object to a Seurat using the function `rOpenScPCA::sce_to_seurat()`.
-We'll set `use_symbols = FALSE` to preserve Ensembl ids as the feature names rather than converting to gene symbols.
 Then, we'll use a simple Seurat pipeline to obtain clusters.
+<!-- TODO: We will want to reference this module for further documentation on this function: https://github.com/AlexsLemonade/OpenScPCA-analysis/issues/945 -->
 
 ```{r sce to seurat, message = FALSE}
-# Convert the SCE to a Seurat object using rOpenScPCA, preserving Ensembl ids as feature names
-seurat_obj <- rOpenScPCA::sce_to_seurat(sce, use_symbols = FALSE)
+# Convert the SCE to a Seurat object using rOpenScPCA
+seurat_obj <- rOpenScPCA::sce_to_seurat(sce)
 

### 2024-12-16 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
simpler saving

@@
 
-# Calculate clusters with Seurat
+# Calculate clusters with Seurat using a standard Seurat pipeline, for example
 seurat_obj <- seurat_obj |>
@@
 Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv()`), or you can add the results into your SCE or Seurat object.
-The subsequent examples will show you how to save the `cluster_results_df` data frame to an SCE and a Seurat object.
+The subsequent examples will demonstrate saving the cluster assignments stored in `cluster_results_df$cluster` to an SCE and a Seurat object.
 
@@
 These automatic clusters were not evaluated, and their parameters were not optimized for any given library.
-
-To avoid ambiguity between the existing and new clustering results, you may want to rename incoming columns.
-
-We'll therefore begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results_df` data frame before adding them to a given object.
-
-```{r rename columns}
-# First, rename columns in `cluster_results_df` to avoid ambiguity
-cluster_results_renamed_df <- cluster_results_df |>
-  # add the prefix "ropenscpca_" to all columns
-  dplyr::rename_with(\(x) paste0("ropenscpca_", x))
-
-cluster_results_renamed_df
-```
+To avoid ambiguity between the existing and new clustering results, we'll name the new column `ropenscpca_cluster`.
 
@@
 
-Follow these steps to save the clustering results to an SCE object's `colData`:
+We can add columns to an SCE object's `colData` table by directly creating a column in the object with `$`.
+Before we do so, we'll confirm that the clusters are in the same order as the SCE object by comparing cell ids:
+
+```{r check sce order}
+all.equal(
+  colnames(sce),
+  cluster_results_df$cell_id
+)
+```
 
 ```{r add to sce}
-# First, extract the colData into a separate data frame for manipulation
-sce_coldata <- colData(sce) |>
-  as.data.frame()
-
-# Second, add columns as desired using the barcodes and cell_id columns to join
-sce_coldata <- sce_coldata |>
-  dplyr::left_join(
-    cluster_results_renamed_df,
-    by = c("barcodes" = "ropenscpca_cell_id")
-  )
-
-# Third, add the modified colData back into the SCE
-# Use the row.names argument to ensure rownames (barcodes) are retained; don't forget this step!
-colData(sce) <- DataFrame(
-  sce_coldata,
-  row.names = sce_coldata$barcodes
-)
-
-head(colData(sce))
+# Add cluster results to the colData
+sce$ropenscpca_cluster <- cluster_results_df$cluster
 ```
@@
 
-Follow these steps save the clustering results to a Seurat object's cell metadata.
-Note that the Seurat function `AddMetaData()`, used below, assumes that the row order in the provided data frame matches the order of cells in the Seurat object you're adding metadata to.
+We can add columns to an Seurat object's cell metadata table by directly creating a column in the object with `$` (note that you can also use the Seurat function `AddMetaData()`).
+Before we do so, we'll confirm that the clusters are in the same order as the Seurat object by comparing cell ids:
+
+```{r check seurat order}
+all.equal(
+  colnames(seurat_obj),
+  cluster_results_df$cell_id
+)
+```
 
 ```{r add to seurat}
-# Add the cluster result data frame to the Seurat object metadata
-seurat_obj <- AddMetaData(seurat_obj, cluster_results_renamed_df)
-
-# Print the cell metadata
-head(seurat_obj@meta.data)
+# Add cluster results to the cell metadata
+seurat_obj$ropenscpca_cluster <- cluster_results_df$cluster
 ```

### 2024-12-17 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
moar colors

@@
   aes(x = cluster, y = silhouette_width) +
-  geom_violin(fill = "darkolivegreen3") +
+  geom_violin(fill = "darkmagenta") +
   labs(x = "Cluster", y = "Silhouette width")
@@
   aes(x = ari) +
-  geom_density(color = "grey30", fill = "darkolivegreen3") +
+  geom_density(color = "grey30", fill = "lightslateblue") +
   labs(x = "Adjusted rand index across bootstrap replicates")

### 2024-12-17 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
update how metadata clustering params are presented, printed, and used

@@
 These clusters were calculated using `bluster`, the same tool that `rOpenScPCA` uses. 
-We can see the parameters it used from the SCE object's metadata; all other parameters were left at their default values.
+The specifications used for this clustering are stored in the SCE object's metadata, as follows; note that all other clustering parameters were left at their default values.
+
+* `metadata(sce)$cluster_algorithm`: The clustering algorithm used
+* `metadata(sce)$cluster_weighting`: The weighting scheme used
+* `metadata(sce)$cluster_nn`: The number of nearest neighbors used
+
+You can see all their values here:
 
 ```{r extract cluster params}
-# Print the clustering algorithm used
-metadata(sce)$cluster_algorithm
-
-# Print the weighting scheme
-metadata(sce)$cluster_weighting
-
-# Print the number of nearest neighbors
-metadata(sce)$cluster_nn
+# Print clustering specifications
+metadata(sce)[c("cluster_algorithm", "cluster_weighting", "cluster_nn")]
 ```
@@
 
-When running `rOpenScPCA::calculate_stability()`, we'll specify the same parameters originally used to build the clusters.
+When running `rOpenScPCA::calculate_stability()`, we'll specify the same parameters originally used to build the clusters by extracting them from the metadata.
+We'll need to ensure the provided arguments are lowercase, as well.
+
 Generally speaking, we only recommend evaluating clusters with `rOpenScPCA::calculate_stability()` if you know the original parameters used.
@@
   scpca_cluster_df,
-  # provide ScPCA clustering parameters
-  algorithm = "louvain",
-  weighting = "jaccard",
-  nn = 20
+  # provide ScPCA clustering parameters by extracting from the SCE metadata
+  algorithm = tolower(metadata(sce)$cluster_algorithm),
+  weighting = tolower(metadata(sce)$cluster_weighting),
+  nn = metadata(sce)$cluster_nn
 )

### 2024-12-17 analyses/cell-type-consensus/exploratory-notebooks/01-reference-exploration.Rmd
Apply suggestions from code review
Co-authored-by: Jaclyn Taroni <19534205+jaclyn-taroni@users.noreply.github.com>

@@
 This section will look at identifying the latest common ancestor (LCA) between all possible combinations of terms from PanglaoDB (used for assigning cell types with `CellAssign`) and the `BlueprintEncodeData` reference from `celldex` (used for assigning cell types with `SingleR`).
-The LCA refers to the latest term in the cell ontology heirarchy that is common between two terms. 
+The LCA refers to the latest term in the cell ontology hierarchy that is common between two terms. 
 I will use the [`ontoProc::findCommonAncestors()` function](https://rdrr.io/bioc/ontoProc/man/findCommonAncestors.html) to get the LCA for each combination.
@@
     # least common shared ancestor
-    lca = list(rownames(ontoProc::findCommonAncestors(blueprint_ontology, panglao_ontology, g = g)))
+    lca = list(rownames(ontoProc::findCommonAncestors(blueprint_ontology, panglao_ontology, g = cl_graph)))
   )
@@
 Alternatively, rather than eliminate terms that are too broad we could look at the similarity index for individual matches and decide on a case by case basis if those should be allowed. 
-Although I still think having a term that is too braod, even if it's a good match, is not super informative.
+Although I still think having a term that is too broad, even if it's a good match, is not super informative.
 

### 2024-12-17 analyses/cell-type-consensus/exploratory-notebooks/01-reference-exploration.Rmd
keep myeloid and epithelial

@@
     total_ancestors = length(ancestors),
-    descendants = list(ontologyIndex::get_descendants(cl_ont, cl_ontology)),
+    descendants = list(ontologyIndex::get_descendants(cl_ont, cl_ontology, exclude_roots = TRUE)),
     total_descendants = length(descendants)
@@
 # first set up the graph from cl ont
-parent_terms <- cl$parents
+parent_terms <- cl_ont$parents
 cl_graph <- igraph::make_graph(rbind(unlist(parent_terms), rep(names(parent_terms), lengths(parent_terms))))
@@
 I'm torn on this one, because I do think it's helpful to know if something is of the myeloid lineage, but if we aren't keeping lymphocyte then I would argue we shouldn't keep myeloid leukocyte. 
+Noting that after discussion we have decided to keep this one since T and B cells are much easier to differentiate based on gene expression alone than cells that are party of the myeloid lineage.
 
@@
 Same with `progenitor cell`, I do think it could be helpful to know that something may be a progenitor cell, but when you have a cell with the label for HSC and the label for cells like monocytes or osteoblasts, then maybe we are talking about a tumor cell instead.
+After discussion, we are going to remove progenitor cells.
 
@@
 
-The only term in this list that I would be concerned about losing is "neuron". 
+The only terms in this list that I would be concerned about losing are "neuron" and epithelial cells. 
 Let's look at those combinations.
@@
 ```{r}
-# blood cell
+# neuron
 print_df |> 
@@
 Even though neuron has ~ 500 descendants, I think we should keep these labels.
+
+#### Epithelial cell
+
+```{r}
+# epithelial cell
+print_df |> 
+  dplyr::filter(cl_annotation == "epithelial cell")
+```
+
+The PanglaoDB cell types seem to be more specific than the ones present in Blueprint Encode, similar to the observation with neurons.
+We should keep epithelial cell.
 
@@
 - The LCA should have equal to or less than 170 total descendants. 
-- We whould include the term for `neuron` even though it has 500 descendants. 
-- Terms that are too broad (like `supporting cell`, `blood cell`, `bone cell`, `lining cell`) should be removed.
+- We whould include the term for `neuron` and `epithelial cell` even though they do not pass the threshold for number of descendants. 
+- Terms that are too broad should be removed. 
+This includes: `lining cell`, `blood cell`, `progenitor cell`, `bone cell`, and `supporting cell`
 

### 2024-12-17 analyses/cell-type-ewings/template_notebooks/clustering-workflow/01-clustering-metrics.Rmd

@@
 
+Before we show any statistics, we will first show the number of clusters for each set of parameters. 
+For any cases where we only see 1 cluster, no distributions will be shown in the plots.
+
+```{r}
+all_stats |> 
+  purrr::imap(\(df, method){
+    df |> 
+      dplyr::select(resolution, nn, cluster) |> 
+      dplyr::group_by(resolution, nn) |> 
+      dplyr::summarise(num_of_clusters = length(unique(cluster))) |>
+      knitr::kable()
+  })
+```
+
 ### Silhouette width

### 2024-12-18 analyses/cell-type-consensus/exploratory-notebooks/01-reference-exploration.Rmd
spell part
Co-authored-by: Jaclyn Taroni <19534205+jaclyn-taroni@users.noreply.github.com>

@@
 I'm torn on this one, because I do think it's helpful to know if something is of the myeloid lineage, but if we aren't keeping lymphocyte then I would argue we shouldn't keep myeloid leukocyte. 
-Noting that after discussion we have decided to keep this one since T and B cells are much easier to differentiate based on gene expression alone than cells that are party of the myeloid lineage.
+Noting that after discussion we have decided to keep this one since T and B cells are much easier to differentiate based on gene expression alone than cells that are part of the myeloid lineage.
 

### 2024-12-18 analyses/cell-type-consensus/exploratory-notebooks/01-reference-exploration.Rmd
add note about keratinocytes

@@
 The PanglaoDB cell types seem to be more specific than the ones present in Blueprint Encode, similar to the observation with neurons.
-We should keep epithelial cell.
+We should keep epithelial cell in the cases where the Blueprint Encode annotation is `Epithelial cells` but not when it is `Keratinocytes`.
 
@@
 - The LCA should have equal to or less than 170 total descendants. 
-- We whould include the term for `neuron` and `epithelial cell` even though they do not pass the threshold for number of descendants. 
+- We should include the term for `neuron` and `epithelial cell` even though they do not pass the threshold for number of descendants. 
+However, `epithelial cell` should only be included if the Blueprint Encode name is `Epithelial cells` and _not_ `Keratinocytes`.
 - Terms that are too broad should be removed. 

### 2024-12-20 analyses/cell-type-consensus/exploratory-notebooks/01-reference-exploration.Rmd
fix module typos and re-render notebook. notebook was also styled by precommit

@@
   geom_density(fill = "#00274C", alpha = 0.5) +
-  geom_vline(data = ancestor_labels_df,
+  geom_vline(
+    data = ancestor_labels_df,
     mapping = aes(xintercept = total_ancestors),
-             lty = 2) +
+    lty = 2
+  ) +
   geom_text(
@@
   geom_density(fill = "#FFCB05", alpha = 0.5) +
-  geom_vline(data = line_df,
+  geom_vline(
+    data = line_df,
     mapping = aes(xintercept = total_descendants),
-             lty = 2) +
+    lty = 2
+  ) +
   geom_text(
@@
   ) +
-  xlim(c(0,500))
+  xlim(c(0, 500))
 ```
@@
 # one row for each combination
-all_ref_df <- expand.grid(panglao_df$panglao_ontology,
-                          blueprint_df$blueprint_ontology) |> 
+all_ref_df <- expand.grid(
+  panglao_df$panglao_ontology,
+  blueprint_df$blueprint_ontology
+) |>
   dplyr::rename(
@@
   geom_density() +
-  geom_vline(data = ancestor_labels_df,
+  geom_vline(
+    data = ancestor_labels_df,
     mapping = aes(xintercept = total_ancestors),
-             lty = 2) +
+    lty = 2
+  ) +
   geom_text(
@@
   geom_density() +
-  geom_vline(data = line_df,
+  geom_vline(
+    data = line_df,
     mapping = aes(xintercept = total_descendants),
-             lty = 2) +
+    lty = 2
+  ) +
   geom_text(
@@
 
-Let's zoom into the area below 1000, since we already know we would want to exlude anything above that based on this plot.
+Let's zoom into the area below 1000, since we already know we would want to exclude anything above that based on this plot.
 
@@
 
-We can also look at all the cell types we are keeping and the total number of descendants to see if there are any that may be we don't want to include because the term is too braod.
+We can also look at all the cell types we are keeping and the total number of descendants to see if there are any that may be we don't want to include because the term is too broad.
 
@@
   dplyr::filter(cl_annotation == "progenitor cell") |>
-  head(n=15) # there's a lot of these so let's only print out some 
+  head(n = 15) # there's a lot of these so let's only print out some
 ```
@@
   dplyr::mutate(
-    similarity_index = ontologySimilarity::get_sim_grid(ontology = cl_ont,
-                                                        term_sets = list(panglao_ontology, blueprint_ontology)) |> 
+    similarity_index = ontologySimilarity::get_sim_grid(
+      ontology = cl_ont,
+      term_sets = list(panglao_ontology, blueprint_ontology)
+    ) |>
       ontologySimilarity::get_sim()
@@
       geom_density() +
-      geom_vline(data = line_df,
+      geom_vline(
+        data = line_df,
         mapping = aes(xintercept = similarity_index),
-                 lty = 2) +
+        lty = 2
+      ) +
       labs(
@@
       )
-
   })

### 2024-12-20 analyses/hello-clusters/02_compare-clustering-parameters.Rmd
WIP: started sweep notebook

---
title: "Comparing clustering parameters with rOpenScPCA"
date: "`r Sys.Date()`"
author: "Data Lab"
output:
  html_notebook:
    toc: yes
    toc_float: yes
    df_print: paged
---

## Introduction

Clustering algorithms have several parameters which can be varied, leading to different clustering results.
A key question when clustering, therefore, is how to identify a set of parameters that lead to robust and reliable clusters that can be used in downstream analysis.

This notebook provides examples of how to use the `rOpenScPCA` package to:

* Calculate clusters across several different parameterizations
* Calculate QC metrics on across parameterizations

Please refer to the [`01_perform-evaluate-clustering.Rmd`](01_perform-evaluate-clustering.Rmd) notebook for a tutorial on using `rOpenScPCA` functions to:

* Calculate clusters from a single parameterization
* Calculate QC metrics on a single set of clusters

This notebook will use the sample `SCPCS000001` from project `SCPCP000001`, which is assumed present in the `OpenScPCA-analysis/data/current/SCPCP000001` directory, for all examples.
Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about obtaining ScPCA data.

## Setup

### Packages

```{r packages}
library(rOpenScPCA)

suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(Seurat)
  library(dplyr)
  library(ggplot2)
})

# Set ggplot theme for plots
theme_set(theme_bw())
```

### Paths

```{r base paths}
# The base path for the OpenScPCA repository
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current")

# The path to this module
module_base <- file.path(repository_base, "analyses", "hello-clusters")
```

```{r input file path}
# Path to processed SCE file for sample SCPCS000001
input_sce_file <- file.path(data_dir, "SCPCP000001", "SCPCS000001", "SCPCL000001_processed.rds")
```

### Set the random seed

Because clustering involves random sampling, it is important to set the random seed at the top of your analysis script or notebook to ensure reproducibility.

```{r set seed}
set.seed(2024)
```

## Read in and prepare data

To begin, we'll read in the `SingleCellExperiment` (SCE) object.

```{r read data}
# Read the SCE file
sce <- readRDS(input_sce_file)
```

For the initial cluster calculations and evaluations, we will use the PCA matrix extracted from the SCE object.
As shown in [`01_perform-evaluate-clustering.Rmd`](01_perform-evaluate-clustering.Rmd), it is also possible to use an SCE object or a Seurat object directly.

```{r extract pca data}
# Extract the PCA matrix from an SCE object
pca_matrix <- reducedDim(sce, "PCA")
```

## Perform clustering across a set of parameters

This section will show how to perform clustering across a set of parameters (aka, "sweep" a set of parameters) with `rOpenScPCA::sweep_clusters()`.

This function takes a PCA matrix with rownames representing unique cell ids (e.g., barcodes) as its primary argument, with additional arguments for cluster parameters. 
This function wraps the `rOpenScPCA::calculate_clusters()` function but allows you to provide a vector of parameter values to perform clustering across, as listed below.
Clusters will be calculated for all combinations of parameters values (where applicable); default values that the function will use for any unspecified parameter values are shown in parentheses

* `algorithm`: Which clustering algorithm to use (Louvain)
* `weighting`: Which weighting scheme to use (Jaccard)
* `nn`: The number of nearest neighbors (10)
* `resolution`: The resolution parameter (1; used only with Louvain and Leiden clustering)
* `objective_function`: The objective function to optimize clusters (CPM; used only with Leiden clustering)

`rOpenScPCA::sweep_clusters()` does allow you to specify values for any other parameters.

This function will return a list of data frames of clustering results. 
Each data frame will have the following columns:

* `cell_id`: Unique cell identifiers, obtained from the PCA matrix's row names
* `cluster`: A factor column with the cluster identities
* There will be one column for each clustering parameter used

To demonstrate this function, we'll calculate clusters using the Louvain algorithm while varying the `nn` parameter:

```{r sweep clusters}
# Define nn parameter values of interest
nn <- seq(10, 30, 10)

# Calculate clusters varying nn, but leaving other parameters at their default values
cluster_results_list <- rOpenScPCA::sweep_clusters(
  pca_matrix,
  nn = nn
)
```

The resulting list has a length of three, one data frame for each `nn` parameter tested:

```{r length cluster_results_list}
length(cluster_results_list)
```

We can look at the first few rows of each data frame using [`purrr::map()`](https://purrr.tidyverse.org/reference/map.html) to iterate over the list:

```{r map cluster_results_list}
cluster_results_list |>
  purrr::map(head)
```

Generally speaking, `purrr::map()` can be used to iterate over this list to analyze or visualize each clustering result on its own.
Complementary to this, we can also combine this list into a single data frame to jointly analyze or visualize all clustering results together:

```{r bind cluster_results_list rows}
cluster_results_df <- dplyr::bind_rows(cluster_results_list)
head(cluster_results_df)
```

## Evaluating clustering results

This section will use `purrr::map()` to iterate over each df to calculate silhouette, purity, and stability. 
We'll also plot results.

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-01-09 analyses/cell-type-ewings/exploratory_analysis/06-tumor-cell-state-assignment.Rmd
initiate notebook for looking at tumor cell states

---
title: "Tumor cell state annotation: `SCPCL000822` and `SCPCL000824`"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
---

## Setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_bw()
)

# set seed
set.seed(2024)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", "SCPCP000015")

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-ewings")

# path to marker genes file
marker_genes_file <- file.path(module_base, "references", "tumor-cell-state-markers.tsv")
```

```{r}
# source in helper functions for make_jaccard_matrix() and jaccard()
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)

# source in helper functions: plot_density() and calculate_sum_markers()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)
```

## Prepare data

```{r}
# sample and library ids
sample_ids <- c("SCPCS000490", "SCPCS000492")
library_ids <- c("SCPCL000822", "SCPCL000824")

# define input sce files
sce_file_names <- glue::glue("{library_ids}_processed.rds")
sce_files <- file.path(data_dir, sample_ids, sce_file_names) |>
  purrr::set_names(library_ids)

# path to cell type classification files
singler_results_dir <- file.path(module_base, "results", "aucell_singler_annotation")
classification_file_names <- glue::glue("{library_ids}_singler-classifications.tsv")
classification_files <- file.path(singler_results_dir, sample_ids, classification_file_names) |>
  purrr::set_names(library_ids)

# path to clustering results files
cluster_results_dir <- file.path(module_base, "results", "clustering", sample_ids)
cluster_file_names <- glue::glue("{library_ids}_cluster-results.tsv")
cluster_files <- file.path(cluster_results_dir, cluster_file_names) |>
  purrr::set_names(library_ids)
```

```{r}
# read in both SCEs
sce_list <- sce_files |>
  purrr::map(readr::read_rds)

# read in classification data frame with SingleR results and combine into one
classification_df <- classification_files |>
  purrr::map(readr::read_tsv) |>
  dplyr::bind_rows(.id = "library_id") |> 
  dplyr::mutate(
    # first grab anything that is tumor and label it tumor
    # NA should be unknown
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later
    singler_lumped = singler_annotation |>
      forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf)
  )

# read in clustering results and select cluster assignments of interest 
# here we keep leiden-mod, 0.5 res, 20 nn for both libraries 
cluster_df <- cluster_files |> 
  purrr::map(readr::read_tsv) |> 
  dplyr::bind_rows(.id = "library_id") |> 
  dplyr::filter(
    cluster_method == "leiden_mod",
    nn == 20,
    resolution == 0.5
  ) |> 
  dplyr::select(
    barcodes = cell_id,
    library_id, 
    cluster
  )
```

```{r}
# read in marker genes table
marker_genes_df <- readr::read_tsv(marker_genes_file, show_col_types = FALSE) |>
  dplyr::select(cell_state, ensembl_gene_id, gene_symbol)

# get list of all cell types found
cell_states <- unique(marker_genes_df$cell_state)

# get the sum of expression of all genes for each cell type
gene_exp_df <- sce_list |>
  purrr::map(\(sce) {
    cell_states |>
      purrr::map(\(state){
        calculate_sum_markers(marker_genes_df, sce, state, cell_state)
      }) |>
      purrr::reduce(dplyr::inner_join, by = "barcodes")
  }) |>
  dplyr::bind_rows(.id = "library_id")

# get umap embeddings and combine into a data frame with gene exp, cluster, and cell type assignments 
umap_df <- sce_list |>
  purrr::map(\(sce){
    df <- sce |>
      scuttle::makePerCellDF(use.dimred = "UMAP") |>
      # replace UMAP.1 with UMAP1 and get rid of excess columns
      dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2)
  }) |>
  dplyr::bind_rows(.id = "library_id") |>
  # add in classifications 
  dplyr::left_join(classification_df, by = c("library_id", "barcodes")) |> 
  dplyr::left_join(gene_exp_df, by = c("barcodes", "library_id")) |> 
  # add in cluster assignments 
  dplyr::left_join(cluster_df, by = c("barcodes", "library_id"))
```

assign cells as tumor cells if they belong to a cluster with high tumor marker gene expression

![SCPCL000822](../plots/SCPCL000822_cluster-marker-gene-exp.png)
![SCPCL000824](../plots/SCPCL000824_cluster-marker-gene-exp.png)
```{r, fig.height=10}
library_ids |>
  purrr::map(\(id){

    p1 <- cluster_classification_df |>
      dplyr::filter(library_id == id) |>
      ggplot(aes(x = UMAP1, y = UMAP2, color = singler_lumped)) +
      geom_point(alpha = 0.5, size = 0.1) +
      theme(
        aspect.ratio = 1
      ) +
      scale_color_brewer(palette = "Dark2") +
      labs(title = id) +
      guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))

    p2 <- cluster_classification_df |>
      dplyr::filter(library_id == id) |>
      ggplot(aes(x = UMAP1, y = UMAP2, color = as.factor(cluster))) +
      geom_point(alpha = 0.5, size = 0.1) +
      theme(
        aspect.ratio = 1
      ) +
      #scale_color_brewer(palette = "Dark2") +
      labs(title = id, color = "cluster") +
      guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))

    patchwork::wrap_plots(p1, p2, nrow = 2)
  })
```

For SCPCL000822, clusters 1, 3, and 4 are tumor 
For SCPCL000824, clustrs 2, 3, 4, 5, 6, 7 are tumor

```{r}
cluster_classification_df <- umap_df |> 
  dplyr::mutate(
    cluster_classification = dplyr::case_when(
      library_id == "SCPCL000822" & cluster %in% c(1, 3, 4) ~ "tumor",
      library_id == "SCPCL000822" & cluster == 6 ~ "endothelial cell",
      # cluster 2 has predominantly chondrocytes and has MSC marker gene expression
      library_id == "SCPCL000822" & cluster == 2 ~ "chondrocyte",
      # clusters with high "immune" marker gene expression are macrophage
      library_id == "SCPCL000822" & cluster == 5 ~ "macrophage",
      # cluster 8 has high MSC marker gene expression and a mix of MSC like cell types
      library_id == "SCPCL000822" & cluster == 7 ~ "mesenchymal-like cell",
      library_id == "SCPCL000824" & cluster %in% c(2, 3, 4, 5, 6, 7) ~ "tumor", 
      library_id == "SCPCL000824" & cluster == 8 ~ "endothelial cell",
      # cluster 9 has high immune markers and is mostly macrohpages
      library_id == "SCPCL000824" & cluster == 9 ~ "macrophage",
      # cluster 1 has high MSC marker gene expression and a mix of MSC like cell types
      library_id == "SCPCL000824" & cluster == 1 ~ "mesenchymal-like cell",
      # if they didn't have high expression of any markers, we label them as unknown
      .default = "unknown"
    )
  )
```

```{r}
library_ids |>
  purrr::map(\(id){
    cluster_classification_df |>
      dplyr::filter(library_id == id) |>
      ggplot(aes(x = UMAP1, y = UMAP2, color = cluster_classification)) +
      geom_point(alpha = 0.5, size = 0.1) +
      theme(
        aspect.ratio = 1
      ) +
      labs(title = id) +
      scale_color_brewer(palette = "Dark2") +
      guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))
  })
```

```{r}
library_ids |>
  purrr::map(\(id){
    cluster_classification_df |>
      dplyr::filter(library_id == id) |>
      ggplot(aes(x = UMAP1, y = UMAP2, color = proliferative_sum)) +
      geom_point(alpha = 0.5, size = 0.1) +
      theme(
        aspect.ratio = 1
      ) +
      labs(title = id) +
      scale_color_viridis_c()
  })
```

pull out just tumor cells

```{r}
tumor_cells_df <- cluster_classification_df |> 
  dplyr::filter(cluster_classification == "tumor")

library_ids |>
  purrr::map(\(id){
    tumor_cells_df |>
      dplyr::filter(library_id == id) |>
      ggplot(aes(x = UMAP1, y = UMAP2, color = `EWS-low_sum`)) +
      geom_point(alpha = 0.5, size = 0.1) +
      theme(
        aspect.ratio = 1
      ) +
      labs(title = id) +
      scale_color_viridis_c()
  })
```

### 2025-01-09 analyses/cell-type-ewings/scripts/utils/tumor-validation-helpers.R

@@ -123,10 +123,11 @@ create_marker_gene_df <- function(
 # output is a data frame with barcodes and `{cell_type}_sum`
 calculate_sum_markers <- function(marker_genes_df,
                                   sce,
-                                  type) {
+                                  type, 
+                                  cell_type_column = cell_type) {
   # get list of marker genes to use
   marker_genes <- marker_genes_df |>
-    dplyr::filter(cell_type == type) |>
+    dplyr::filter({{cell_type_column}} == type) |>
     dplyr::pull(ensembl_gene_id)
 
   # get the gene expression counts for all marker genes

### 2025-01-10 analyses/cell-type-ewings/exploratory_analysis/06-tumor-cell-state-assignment.Rmd
subclustering and more plots

@@
 ---
-title: "Tumor cell state annotation: `SCPCL000822` and `SCPCL000824`"
+title: "Explore expression of marker genes for tumor cell states"
 author: Ally Hawkins
@@
 ---
+
+In this notebook we look at expression of potential markers for tumor cell states in two libraries from `SCPCP000015`: `SCPCL000822` and `SCPCL000824`. 
+The expression of specific marker genes are then used to classify tumor cells into three different tumor cell states: `EWS-low`, `EWS-mid`, and `EWS-high`.
+
+We did this by:
+
+- Evaluating marker gene expression for different tumor cell states across all cells. 
+- Identifying tumor cells and getting new cluster assignments for just tumor cells.
+- Looking at expression of tumor cell state marker genes across clusters and individual cells.
 
@@
 theme_set(
-  theme_bw()
+  theme_classic()
 )
@@
 marker_genes_file <- file.path(module_base, "references", "tumor-cell-state-markers.tsv")
+
+# gene signatures
+aynaud_file <- file.path(module_base, "references", "gene_signatures", "aynaud-ews-targets.tsv")
+wrenn_file <- file.path(module_base, "references","gene_signatures", "wrenn-nt5e-genes.tsv")
+
+# path to save annotations tsv in results 
+results_dir <- file.path(module_base, "results", "tumor-cell-state-classifications")
+fs::dir_create(results_dir)
 ```
@@
 ```{r}
-# source in helper functions for make_jaccard_matrix() and jaccard()
-jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
-source(jaccard_functions)
-
 # source in helper functions: plot_density() and calculate_sum_markers()
@@
 
-```{r}
+Below we set up the data for plotting throughout this notebook.
+
+1. The output from `aucell-singler-annotation.sh` is read in and annotations are lumped so that we can easily plot the top cell types. 
+2.
+Clustering results from `evaluate-clusters.sh` is read in and subset to the desired parameters we would like to use for assigning clusters (leiden, modularity, resolution 0.5, nearest neighbors 20).
+3. The marker genes for tumor cell states and custom gene signatures are used to calculate the sum of all marker genes in each gene set for each library. 
+4. UMAP embeddings for each library is pulled out into a data frame and combined with `SingleR` annotations, cluster assignments, and marker gene expression for all marker gene sets.
+
+```{r, message=FALSE}
 # read in both SCEs
@@
 
-```{r}
+```{r, message=FALSE}
 # read in marker genes table
@@
 
-# get list of all cell types found
-cell_states <- unique(marker_genes_df$cell_state)
+# get individual gene signatures
+aynaud_genes <- readr::read_tsv(aynaud_file) |> 
+  dplyr::mutate(cell_state = "aynaud-ews-high") |>
+  tidyr::drop_na()
 
-# get the sum of expression of all genes for each cell type
+wrenn_genes <- readr::read_tsv(wrenn_file) |> 
+  dplyr::mutate(cell_state = "wrenn-ews-low") |>
+  tidyr::drop_na()
+
+# combine all genes into a single df 
+all_markers_df <- dplyr::bind_rows(list(marker_genes_df, aynaud_genes, wrenn_genes))
+
+# get list of all cell states/ gene lists 
+cell_states <- unique(all_markers_df$cell_state)
+
+# get the sum of expression of all genes for each cell state
 gene_exp_df <- sce_list |>
@@
       purrr::map(\(state){
-        calculate_sum_markers(marker_genes_df, sce, state, cell_state)
+        calculate_sum_markers(all_markers_df, sce, state, cell_state)
       }) |>
@@
 
-assign cells as tumor cells if they belong to a cluster with high tumor marker gene expression
+## Identify tumor cells
+
+To label tumor cell states, we first want to pull out the cells that are tumor cells and then look at the expression of marker genes for each tumor cell state in those cells. 
+We have used a variety of methods to identify tumor cells so we will look at the output from two orthogonal approaches to identify the group of tumor cells we want to further classify into cell states.
+Let's start by looking at the intersection of tumor cells assigned by the `aucell-singler-annotation.sh` workflow and the clusters obtained from the `evaluate-clusters.sh` workflow that have high expression of tumor marker genes.
+
+Below are two plots that we generated as part of the `evaluate-clusters.sh` workflow that tells us which clusters express tumor marker genes for each of these libraries. 
+We can use this information to help identify tumor cells to classify.
 
 ![SCPCL000822](../plots/SCPCL000822_cluster-marker-gene-exp.png)
+
 ![SCPCL000824](../plots/SCPCL000824_cluster-marker-gene-exp.png)
+
 ```{r, fig.height=10}
@@
 
-For SCPCL000822, clusters 1, 3, and 4 are tumor 
-For SCPCL000824, clustrs 2, 3, 4, 5, 6, 7 are tumor
+Looking at the above plots, we can assign some general cell types using clusters. 
+This is the same approach we took in `05-cluster-exploration.Rmd` but with slightly different parameters. 
+Based on the UMAPs and the density plots we see that tumor cells are in cluster 1, 3, and 4 for `SCPCL000822` and cluster 2, 3, 4, 5, 6, and 7 for `SCPCL000824`.
 
@@
 
+Looking at this I think the cells that are classified as tumor cells using the cluster assignments mostly line up with the `SingleR` annotations. 
+So we can label any cells from clusters that have high tumor marker gene expression as tumor cells.
+
+### Look at marker gene expression for tumor cell states in all cells
+
+Below we look at the expression of genes for the tumor cell states across all cells. 
+This will show us how the expression of these gene sets varies across cells in each library and help ensure we have the correct population of tumor cells.
+
+The value for each of the marker gene sets corresponds to the sum of the expression of all genes in the given gene list. 
+(Note that I also looked at mean and it had similar output)
+
+The gene sets that we use are:
+
+- `EWS-low`: Marker genes from `tumor-cell-state-markers.tsv` that correspond to the `EWS-low` cell state. 
+- `EWS-high`: Marker genes from `tumor-cell-state-markers.tsv` that correspond to the `EWS-high` cell state. 
+- `proliferative` : Marker genes from `tumor-cell-state-markers.tsv` that correspond to the `proliferative` cell state. 
+Note that a lot of the Ewing sarcoma literature argues that `EWS-FLI1` high cells are proliferative. 
+- `aynaud-ews-high`: Marker genes from `gene_signatures/aynaud-ews-targets.tsv` that correspond to genes activated by `EWS-FLI1`. 
+- `wrenn-ews-low`: Marker genes from `gene_signatures/wrenn-nt5e-genes.tsv` that correspond to genes repressed by `EWS-FLI1`.
+
+```{r, fig.height=7, fig.width=10}
+# pull out columns that have sum of marker gene expression 
+marker_gene_cols <- colnames(cluster_classification_df)[stringr::str_detect(colnames(cluster_classification_df), "_sum$")]
+
+# look at the expression of all gene sets across both libraries 
+marker_gene_cols |>
+  purrr::map(\(geneset){
+    cluster_classification_df |>
+      ggplot(aes(x = UMAP1, y = UMAP2, color = !!sym(geneset))) +
+      geom_point(alpha = 0.5, size = 0.1) +
+      facet_wrap(vars(library_id)) +
+      scale_color_viridis_c() +
+      theme(
+        aspect.ratio = 1
+      )
+  }) |> 
+  patchwork::wrap_plots(ncol = 2)
+```
+
+It looks like there's a cluster of cells in both libraries that have expression of the `EWS-low` and `wrenn-ews-low` signatures. 
+Those cells correspond to the cells that we are labeling as "chondrocyte" in `SCPCL000822` and "mesenchymal-like" in `SCPCL000824`. 
+The chondrocytes also express high markers of mesenchymal-like genes and the literature talks about tumor cells with low `EWS-FLI1` expression being "MSC-like". 
+I think there's a possibility that these are tumor cells, so let's relabel them.
+
+```{r}
+# create a new tumor classification that takes all tumor clusters plus the clusters with EWS-FLI1 low expression 
+cluster_classification_df <- cluster_classification_df |> 
+  dplyr::mutate(
+    tumor_classification = dplyr::case_when(
+      cluster_classification == "tumor" ~ "tumor",
+      library_id == "SCPCL000822" & cluster_classification == "chondrocyte" ~ "tumor",
+      library_id == "SCPCL000824" & cluster_classification == "mesenchymal-like cell" ~ "tumor",
+      .default = cluster_classification
+    )
+  )
+```
+
+## Classify tumor cell states
+
+Now that we have identified a subset of cells that are tumor cells, we can attempt to group them into various cell states. 
+The main thing that we are interested in is grouping cells into levels of `EWS-FLI1` expression. 
+Let's start by looking at the expression of the gene sets across all tumor cells.
+
+To do this, we will make a density plot that looks at gene set expression by cluster using the cluster assignments calculated using all cells.
+
+```{r}
+tumor_cells_df <- cluster_classification_df |> 
+  dplyr::filter(tumor_classification == "tumor")
+```
+
+```{r, fig.height=10}
+# density plot showing expression of each gene set by cluster
+library_ids |> 
+  purrr::map(\(id){
+
+    density_df <- tumor_cells_df |> 
+      dplyr::filter(library_id == id)
+
+    marker_gene_cols |>
+    purrr::map(\(column){
+      plot_density(
+        density_df,
+        column,
+        "cluster"
+      ) 
+    }) |>
+    patchwork::wrap_plots(ncol = 1) +
+      patchwork::plot_annotation(id)
+
+  })
+```
+
+A few notes from this:
+
+- It doesn't look like very many cells express proliferative markers, so perhaps that gene set will not be helpful for these libraries? 
+- There are some very obvious `EWS-FLI1` low groups of cells. 
+- The `EWS-high` and `aynaud-EWS-high` show clusters that have high expression of `EWS-FLI1` targets, but the `aynaud-EWS-high` appears to have some more separation across clusters. 
+That tells me that gene signature might be more helpful in classifying cells across the `EWS-FLI1` continuum.
+
+### Sub cluster tumor cells
+
+We would expect that tumor cells that are in similar cell states would cluster together. 
+So here I am going to take just the tumor cells and re-assign clusters. 
+I'm using the same parameters we use for the whole sample, leiden, modularity, resolution of 0.5, and nearest neighbors of 20. 
+I think if we really wanted to be stringent we would evaluate clusters here and find the best parameters to use, but I think starting with these is good.
+
+```{r}
+# get a data frame of all tumor cells and re assigned clusters 
+subcluster_df <- sce_list |>
+  purrr::imap(\(sce, id){
+
+    # pull out the barcodes that are from tumor cells
+    tumor_cells <- tumor_cells_df |> 
+      dplyr::filter(library_id == id) |> 
+      dplyr::pull(barcodes)
+
+    # filter the sce to only those barcodes
+    filtered_sce <- sce[ , tumor_cells]
+
+    # grab pcs for clustering
+    pcs <- reducedDim(filtered_sce)
+
+    # get the clusters 
+    clusters <- bluster::clusterRows(
+      pcs,
+      bluster::NNGraphParam(
+        k = 20,
+        type = "jaccard",
+        cluster.fun = "leiden",
+        cluster.args = list(objective_function = "modularity",
+                            resolution = 0.5)
+      )
+    )
+
+    # combine into a data frame with barcodes
+    cluster_df <- data.frame(
+      barcodes = rownames(pcs),
+      sub_cluster = clusters
+    )
+  }) |> 
+  # get clusters for both samples in one data frame
+  dplyr::bind_rows(.id = "library_id")
+
+# add clusters to tumor cells df for plotting
+tumor_cells_df <- tumor_cells_df |> 
+  dplyr::left_join(subcluster_df, by = c("library_id", "barcodes"))
+```
+
+The first thing we will do is plot the cluster assignments on the UMAP.
+
 ```{r}
@@
   purrr::map(\(id){
-    cluster_classification_df |>
+    tumor_cells_df |>
       dplyr::filter(library_id == id) |>
-      ggplot(aes(x = UMAP1, y = UMAP2, color = proliferative_sum)) +
+      ggplot(aes(x = UMAP1, y = UMAP2, color = sub_cluster)) +
       geom_point(alpha = 0.5, size = 0.1) +
@@
       labs(title = id) +
-      scale_color_viridis_c()
+      scale_color_brewer(palette = "Dark2") +
+      guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))
   })
@@
 
-pull out just tumor cells
+Now we will look at the marker gene expression across all clusters for each sample. 
+First using a density plot and then a heatmap.
+
+```{r, fig.height=10}
+library_ids |> 
+  purrr::map(\(id){
+
+    density_df <- tumor_cells_df |> 
+      dplyr::filter(library_id == id)
+
+    marker_gene_cols |>
+    purrr::map(\(column){
+      plot_density(
+        density_df,
+        column,
+        "sub_cluster"
+      ) 
+    }) |>
+    patchwork::wrap_plots(ncol = 1) +
+      patchwork::plot_annotation(id)
+
+  })
+```
 
 ```{r}
-tumor_cells_df <- cluster_classification_df |> 
-  dplyr::filter(cluster_classification == "tumor")
-
 library_ids |> 
   purrr::map(\(id){
-    tumor_cells_df |>
+
+    filtered_df <- tumor_cells_df |> 
+      dplyr::filter(library_id == id)
+
+    full_celltype_heatmap(filtered_df, marker_gene_cols, "sub_cluster", plot_name = id)
+
+  })
+
+```
+
+It looks like for both samples we have some clear division between EWS high and EWS low cells. 
+Additionally, there are some cells with really strong expression of the EWS high markers and others that look like the genes are expressed but to a lesser degree. 
+This fits with the findings from the literature that `EWS-FLI1` expression lies on a continuum.
+
+Between these two plots I think we can try and group cells into EWS low, EWS middle, and EWS high cells. 
+I don't think the proliferative markers are very informative, so I don't think we can use that to categorize any cells.
+
+Based on this I would make the following assignments:
+
+`SCPCL000822`: 
+`EWS-FLI1` low  - 2
+`EWS-FLI1` middle - 1 & 4
+`EWS-FLI1` high - 3
+
+`SCPCL000824` : 
+`EWS-FLI1` low - 1
+`EWS-FLI1` middle - 6 
+`EWS-FLI1` high - 2, 3, 4, 5, 7
+
+```{r}
+# categorize tumor cells based on EWS expression 
+tumor_cells_df <- tumor_cells_df |> 
+  dplyr::mutate(
+    cell_state = dplyr::case_when(
+      library_id == "SCPCL000822" & sub_cluster == 2 ~ "EWS-low",
+      library_id == "SCPCL000822" & sub_cluster %in% c(1, 4) ~ "EWS-mid",
+      library_id == "SCPCL000822" & sub_cluster == 3 ~ "EWS-high",
+      library_id == "SCPCL000824" & sub_cluster == 1 ~ "EWS-low",
+      library_id == "SCPCL000824" & sub_cluster %in% c(6, 8) ~ "EWS-mid",
+      library_id == "SCPCL000824" & sub_cluster %in% c(2, 3, 4, 5, 7) ~ "EWS-high",
+    )
+  )
+```
+
+Let's re-plot the heatmap but label by tumor cell state rather than cluster.
+
+```{r}
+library_ids |> 
+  purrr::map(\(id){
+
+    filtered_df <- tumor_cells_df |> 
+      dplyr::filter(library_id == id)
+
+    full_celltype_heatmap(filtered_df, marker_gene_cols, "cell_state", plot_name = id)
+
+  })
+```
+
+I think that looks pretty reasonable. 
+I'm not sure we can really make more definitive calls given the gene lists that we have, unless we use fancier approaches. 
+We could try something like `AUCell`, but without a nice bimodal distribution, I don't think we would get anything but low/high cells.
+
+## Combined classifications
+
+As a final step, we'll add the tumor cell states back into the full object and look at them on the UMAP. 
+Here I am plotting the classification of cells based on the cluster because otherwise the normal cells get a little wild with colors and I want to be able to see visualize the tumor cells.
+
+```{r}
+# select the columns we want to join back into the main classification df 
+cell_state_classifications <- tumor_cells_df |> 
+  dplyr::select(library_id, barcodes, cell_state, sub_cluster)
+
+# add cell states to classification 
+final_classification_df <- cluster_classification_df |> 
+  dplyr::left_join(cell_state_classifications, by = c("library_id", "barcodes")) |> 
+  dplyr::mutate(full_classification = dplyr::if_else(is.na(cell_state), singler_lumped, cell_state),
+                cluster_classification = dplyr::if_else(is.na(cell_state), cluster_classification, cell_state))
+
+# make some umaps 
+library_ids |>
+  purrr::map(\(id){
+    final_classification_df |>
       dplyr::filter(library_id == id) |>
-      ggplot(aes(x = UMAP1, y = UMAP2, color = `EWS-low_sum`)) +
+      ggplot(aes(x = UMAP1, y = UMAP2, color = cluster_classification)) +
       geom_point(alpha = 0.5, size = 0.1) +
@@
       labs(title = id) +
-      scale_color_viridis_c()
+      scale_color_brewer(palette = "Dark2") +
+      guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))
   })
 ```
+
+```{r}
+# export the final classifications in case we want to use them in the future 
+# save all classification columns 
+final_classification_df |> 
+  split(final_classification_df$library_id) |> 
+  purrr::iwalk(\(df, id){
+
+    filename <- glue::glue("{id}_tumor-cell-state-annotations.tsv")
+
+    df |> 
+      dplyr::select(
+        library_id,
+        barcodes,
+        singler_annotation,
+        aucell_annotation, 
+        cluster, 
+        cluster_classification, 
+        tumor_sub_cluster = sub_cluster, 
+        tumor_cell_state = cell_state,
+        full_classification 
+      ) |> 
+      readr::write_tsv(file.path(results_dir, filename))
+
+  })
+```
+
+## Conclusions
+
+- There is a clear division between cells that are `EWS-FLI` high and `EWS-FLI1` low. 
+- Most tumor cells in both samples are `EWS-FLI` high, which is consistent with the literature. 
+- There may be a class of cells that live in between high and low, but if we want to define those we probably need to be more stringent with our approach and identify a mechanism to bin cells based on expression of both `EWS-FLI1` activated and repressed genes, which is outside the scope of this notebook.
+
+## Session info
+
+```{r session info}
+# record the versions of the packages used in this analysis and other environment information
+sessionInfo()
+```

### 2025-01-10 analyses/cell-type-ewings/scripts/utils/tumor-validation-helpers.R

@@ -250,9 +250,9 @@ full_celltype_heatmap <- function(classification_df,
 
   # create annotation for heatmap
   annotation <- ComplexHeatmap::columnAnnotation(
-    singler = classification_df[[annotation_column]],
+    annotation = classification_df[[annotation_column]],
     col = list(
-      singler = colors
+      annotation = colors
     )
   )
 
@@ -294,7 +294,7 @@ plot_density <- function(classification_df,
     ) +
     labs(
       x = "Gene set expression",
-      y = "Cell type annotation",
+      y = annotation_column,
       title = geneset_name
     ) +
     scale_alpha_identity() +

### 2025-01-10 analyses/cell-type-ewings/exploratory_analysis/06-tumor-cell-state-assignment.Rmd
Apply suggestions from code review
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 3. The marker genes for tumor cell states and custom gene signatures are used to calculate the sum of all marker genes in each gene set for each library. 
-4. UMAP embeddings for each library is pulled out into a data frame and combined with `SingleR` annotations, cluster assignments, and marker gene expression for all marker gene sets.
+4. UMAP embeddings for each library are pulled out into a data frame and combined with `SingleR` annotations, cluster assignments, and marker gene expression for all marker gene sets.
 
@@
     cluster_classification = dplyr::case_when(
+      ############## Library SCPCL000822 ###############
       library_id == "SCPCL000822" & cluster %in% c(1, 3, 4) ~ "tumor",
@@
       library_id == "SCPCL000822" & cluster == 7 ~ "mesenchymal-like cell",
+      ############## Library SCPCL000824 ###############
       library_id == "SCPCL000824" & cluster %in% c(2, 3, 4, 5, 6, 7) ~ "tumor", 
@@
     # filter the sce to only those barcodes
-    filtered_sce <- sce[ , tumor_cells]
+    tumor_sce <- sce[ , tumor_cells]
 
     # grab pcs for clustering
-    pcs <- reducedDim(filtered_sce)
+    pcs <- reducedDim(tumor_sce)
 
@@
 
-Now we will look at the marker gene expression across all clusters for each sample. 
-First using a density plot and then a heatmap.
+Now we will look at the marker gene expression across all clusters for each sample, first using a density plot and then a heatmap.
 

### 2025-01-10 analyses/cell-type-ewings/exploratory_analysis/06-tumor-cell-state-assignment.Rmd
update umaps and heatmaps

@@
 
-    p1 <- cluster_classification_df |>
+    p1 <- umap_df |>
       dplyr::filter(library_id == id) |>
@@
 
-    p2 <- cluster_classification_df |>
+    p2 <- umap_df |>
       dplyr::filter(library_id == id) |>
@@
 
-```{r, fig.height=7, fig.width=10}
+```{r}
 # pull out columns that have sum of marker gene expression 
@@
       )
-  }) |> 
-  patchwork::wrap_plots(ncol = 2)
+  })
 ```
@@
 
-- It doesn't look like very many cells express proliferative markers, so perhaps that gene set will not be helpful for these libraries? 
+- It doesn't look like very many cells express proliferative markers in `SCPCL000822`, but cluster 2 has high expression in `SCPCL000824`. 
+It looks like that might coincide with high `EWS-FLI` seen in that same cluster, but we have to dive deeper to know for sure. 
 - There are some very obvious `EWS-FLI1` low groups of cells. 
@@
 
-Now we will look at the marker gene expression across all clusters for each sample, first using a density plot and then a heatmap.
+Now we will look at the marker gene expression across all clusters for each sample, first using a density plot and then heatmaps.
 
@@
 
+In all of the heatmaps below, the `annotation` column is the `sub_cluster`.
+
 ```{r}
@@
 
-    full_celltype_heatmap(filtered_df, marker_gene_cols, "sub_cluster", plot_name = id)
+    full_celltype_heatmap(filtered_df, marker_gene_cols, "sub_cluster")
 
@@
 
+```
+
+Let's make the same heatmap, but remove the column clustering.
+
+```{r}
+library_ids |> 
+  purrr::map(\(id){
+
+    filtered_df <- tumor_cells_df |> 
+      dplyr::filter(library_id == id) |> 
+      dplyr::arrange(sub_cluster)
+
+    full_celltype_heatmap(filtered_df, marker_gene_cols, "sub_cluster", cluster_columns = FALSE)
+
+  })
 ```
@@
 Between these two plots I think we can try and group cells into EWS low, EWS middle, and EWS high cells.
-I don't think the proliferative markers are very informative, so I don't think we can use that to categorize any cells.
+
+The last thing we'll do is look at any cells that could be marked as "proliferative". 
+`SCPCL000822` doesn't appear to have very much expression of the proliferative markers, but `SCPCL000824` does seem to have one cluster, cluster 2, that has expression of the proliferative cells. 
+Perhaps we should label those cells separately as `EWS-high-proliferative`.
 
@@
 `EWS-FLI1` middle - 6 
-`EWS-FLI1` high - 2, 3, 4, 5, 7
+`EWS-FLI1` high - 3, 4, 5, 7
+`EWS-FLI1` high proliferative - 2
 
@@
       library_id == "SCPCL000824" & sub_cluster == 1 ~ "EWS-low",
-      library_id == "SCPCL000824" & sub_cluster %in% c(6, 8) ~ "EWS-mid",
-      library_id == "SCPCL000824" & sub_cluster %in% c(2, 3, 4, 5, 7) ~ "EWS-high",
+      library_id == "SCPCL000824" & sub_cluster == 6 ~ "EWS-mid",
+      library_id == "SCPCL000824" & sub_cluster == 2 ~ "EWS-high-proliferative",
+      library_id == "SCPCL000824" & sub_cluster %in% c(3, 4, 5, 7) ~ "EWS-high",
     )
@@
 Let's re-plot the heatmap but label by tumor cell state rather than cluster. 
+In this heatmap the annotation column is `cell_state`.
 
@@
 
-    full_celltype_heatmap(filtered_df, marker_gene_cols, "cell_state", plot_name = id)
+    full_celltype_heatmap(filtered_df, marker_gene_cols, "cell_state")
 

### 2025-01-10 analyses/cell-type-ewings/exploratory_analysis/06-tumor-cell-state-assignment.Rmd

@@
 Clustering results from `evaluate-clusters.sh` is read in and subset to the desired parameters we would like to use for assigning clusters (leiden, modularity, resolution 0.5, nearest neighbors 20).
-3. The marker genes for tumor cell states and custom gene signatures are used to calculate the sum of all marker genes in each gene set for each library. 
+3. The marker genes for tumor cell states and custom gene signatures are used to calculate the mean of all marker genes in each gene set for each library. 
 4. UMAP embeddings for each library are pulled out into a data frame and combined with `SingleR` annotations, cluster assignments, and marker gene expression for all marker gene sets.
@@
 
-# get the sum of expression of all genes for each cell state
+# get the mean expression of all genes for each cell state
 gene_exp_df <- sce_list |>
@@
       purrr::map(\(state){
-        calculate_sum_markers(all_markers_df, sce, state, cell_state)
+        calculate_mean_markers(all_markers_df, sce, state, cell_state)
       }) |>
@@
 
-The value for each of the marker gene sets corresponds to the sum of the expression of all genes in the given gene list. 
-(Note that I also looked at mean and it had similar output)
+The value for each of the marker gene sets corresponds to the mean of the expression of all genes in the given gene list.
 
@@
 # pull out columns that have sum of marker gene expression 
-marker_gene_cols <- colnames(cluster_classification_df)[stringr::str_detect(colnames(cluster_classification_df), "_sum$")]
+marker_gene_cols <- colnames(cluster_classification_df)[stringr::str_detect(colnames(cluster_classification_df), "_mean$")]
 
@@
     filtered_df <- tumor_cells_df |> 
-      dplyr::filter(library_id == id)
+      dplyr::filter(library_id == id)|> 
+      dplyr::arrange(cell_state)
 
-    full_celltype_heatmap(filtered_df, marker_gene_cols, "cell_state")
+    full_celltype_heatmap(filtered_df, marker_gene_cols, "cell_state", cluster_columns = FALSE)
 

### 2025-01-13 analyses/hello-clusters/02_compare-clustering-parameters.Rmd
flesh out section to vary one parameter

@@
 
-* Calculate clusters across several different parameterizations
-* Calculate QC metrics on across parameterizations
+* Calculate several versions of clustering results across several different parameterizations
+* Calculate QC metrics on across clustering results
 
@@
 * Calculate clusters from a single parameterization
-* Calculate QC metrics on a single set of clusters
+* Calculate QC metrics on a single set of clusters, as well as explanations of the metrics themselves
 
@@
   library(SingleCellExperiment)
-  library(Seurat)
-  library(dplyr)
   library(ggplot2)
+  library(patchwork)
 })
@@
 
-This function takes a PCA matrix with rownames representing unique cell ids (e.g., barcodes) as its primary argument, with additional arguments for cluster parameters. 
+This function takes a PCA matrix with row names representing unique cell ids (e.g., barcodes) as its primary argument, with additional arguments for cluster parameters. 
 This function wraps the `rOpenScPCA::calculate_clusters()` function but allows you to provide a vector of parameter values to perform clustering across, as listed below.
@@
 # Define nn parameter values of interest
-nn <- seq(10, 30, 10)
+nn_values <- seq(10, 30, 10)
 
@@
   pca_matrix,
-  nn = nn
+  nn = nn_values
 )
@@
 
+It can be helpful (although it is not strictly necessary to keep track) to name this list by the varied `nn` parameter:
+
+```{r set list names}
+names(cluster_results_list) <- glue::glue("nn_{nn_values}")
+```
+
 We can look at the first few rows of each data frame using [`purrr::map()`](https://purrr.tidyverse.org/reference/map.html) to iterate over the list:
@@
 
-Generally speaking, `purrr::map()` can be used to iterate over this list to analyze or visualize each clustering result on its own.
-Complementary to this, we can also combine this list into a single data frame to jointly analyze or visualize all clustering results together:
+Generally speaking, `purrr::map()` can be used to iterate over this list to visualize or analyze each clustering result on its own; we'll use this approach in the following sections.
 
-```{r bind cluster_results_list rows}
-cluster_results_df <- dplyr::bind_rows(cluster_results_list)
-head(cluster_results_df)
+## Visualizing clustering results
+
+When comparing clustering results, it's important to first visualize the different clusterings to build context for interpreting QC metrics.
+
+As one example of why this is important, we generally expect that more robust clusters will have higher values for metrics like silhouette width and neighborhood purity.
+However, we also expect that having fewer clusters in the first place will also lead to higher metrics, regardless of cluster quality: When there are fewer clusters, it is more likely that clusters overlap less with one another just because there aren't many clusters in the first place.
+This means that, when interpreting cluster quality metrics, you should be careful to take more context about the data into consideration and not only rely on the metric values.
+
+We'll therefore visualize these results as UMAPs by iterating over `cluster_results_list` and combining plots with [`patchwork::wrap_plots()`](https://patchwork.data-imaginist.com/reference/wrap_plots.html).
+We'll specifically use [`purrr::imap()`](https://purrr.tidyverse.org/reference/imap.html) to iterate so that we can assign this list's names as plot titles.
+
+For this, we'll begin by extracting a table of UMAP coordinates from our SCE object.
+
+```{r create umap_df}
+umap_df <- reducedDim(sce, "UMAP") |>
+  as.data.frame()
 ```
+
+Next, we'll iterate over `cluster_results_list` to plot the UMAPs.
+
+```{r plot all umaps, fig.width = 12}
+umap_plots <- cluster_results_list |>
+  purrr::imap(
+    \(cluster_df, clustering_name) {
+      # Add a column with cluster assignments to umap_df
+      umap_df_plot <- umap_df |>
+        dplyr::mutate(cluster = cluster_df$cluster)
+
+      # Plot the UMAP, colored by the new cluster variable
+      ggplot(umap_df_plot, aes(x = UMAP1, y = UMAP2, color = cluster)) +
+        geom_point(alpha = 0.6) +
+        labs(title = glue::glue("nearest neighbors: {clustering_name}")) +
+        # We'll add a couple UMAP plot settings here, including equal axes and
+        # turning off the axis ticks and text since UMAP coordinates are not meaningful
+        coord_equal() +
+        theme(
+          axis.ticks = element_blank(),
+          axis.text = element_blank()
+        )
+    }
+  )
+
+# Print the plots with patchwork::wrap_plots()
+patchwork::wrap_plots(umap_plots)
+```
+
+These plots show that the number of clusters decreases as the nearest neighbors parameter increases, with between 9-13 clusters.
 
@@
 
-This section will use `purrr::map()` to iterate over each df to calculate silhouette, purity, and stability. 
-We'll also plot results.
+This section will use `purrr::map()` to iterate over each clustering result data frame to calculate silhouette width, neighborhood purity, and stability, and then visualize results.
+The goal of this code is to identify whether one clustering parameterization produces more reliable clusters.
+
+### Silhouette width and neighborhood purity
+
+Both silhouette width and neighborhood purity are cell-level quantities, so we can calculate them together in the same call to `purrr::map()`.
+Below, we'll iterate over each data frame in `cluster_results_list` to calculate these quantities.
+
+```{r calculate cell level metrics}
+cell_metric_list <- cluster_results_list |>
+  purrr::map(
+    \(cluster_df) {
+      # calculate silhouette width
+      silhouette_df <- rOpenScPCA::calculate_silhouette(pca_matrix, cluster_df)
+
+      # calculate neighbhorhood purity
+      purity_df <- rOpenScPCA::calculate_purity(pca_matrix, cluster_df)
+
+      # Combine into a single data frame
+      dplyr::left_join(silhouette_df, purity_df)
+    }
+  )
+
+# View the first six rows of each clustering result's cell-level QC metrics
+purrr::map(cell_metric_list, head)
+```
+
+To visualize these results, we can combine all data frames in this list into a single overall data frame, where the existing `nn` column will distinguish among conditions.
+
+```{r combine cell metrics list}
+cell_metrics_df <- dplyr::bind_rows(cell_metric_list)
+```
+
+We can visualize silhouette width and neighborhood purity each with boxplots, for example, and use the [`patchwork`](https://patchwork.data-imaginist.com/) package to print them together:
+
+```{r}
+# Plot silhouette width
+silhouette_plot <- ggplot(cell_metrics_df) +
+  aes(x = as.factor(nn), y = silhouette_width, fill = as.factor(nn)) +
+  geom_boxplot() +
+  scale_fill_brewer(palette = "Pastel2") +
+  labs(
+    x = "Number of nearest neighbors",
+    y = "Silhouette width"
+  )
+
+# Plot neighborhood purity width
+purity_plot <- ggplot(cell_metrics_df) +
+  aes(x = as.factor(nn), y = purity, fill = as.factor(nn)) +
+  geom_boxplot() +
+  scale_fill_brewer(palette = "Pastel2") +
+  labs(
+    x = "Number of nearest neighbors",
+    y = "Neighborhood purity"
+  )
+
+# Add together using the patchwork library, without a legend
+silhouette_plot + purity_plot & theme(legend.position = "none")
+```
+While there does not appear to be a salient difference among silhouette width distributions, it does appear that purity is higher with a higher nearest neighbors parameter.
+
+### Stability
+
+Next, we'll calculate stability on the clusters using `rOpenScPCA::calculate_stability()`, specifying the same parameter used for the original cluster calculation at each iteration.
+
+```{r calculate stability}
+stability_list <- cluster_results_list |>
+  purrr::map(
+    \(cluster_df) {
+      nn <- unique(cluster_df$nn)
+
+      # calculate stability, passing in the parameter value used for this iteration
+      rOpenScPCA::calculate_stability(pca_matrix, cluster_df, nn = nn)
+    }
+  )
+```
+
+We'll again combine the output of `stability_list` into a single data frame and plot `ari` values across `nn` parameterizations.
+
+```{r combine plot stability}
+stability_df <- dplyr::bind_rows(stability_list)
+
+ggplot(stability_df) +
+  aes(x = as.factor(nn), y = ari, fill = as.factor(nn)) +
+  geom_boxplot() +
+  scale_fill_brewer(palette = "Pastel2") +
+  labs(
+    x = "Number of nearest neighbors",
+    y = "Adjusted Rand Index"
+  ) +
+  theme(legend.position = "none")
+```
+
+Here, we see that a nearest neighbors value of 20 or 30 leads to more stable clustering results compared to 10.
 

### 2025-01-13 analyses/hello-clusters/02_compare-clustering-parameters.Rmd
add section for varying multiple parameters, and other tweaks

@@
 
-## Perform clustering across a set of parameters
+## Varying a single clustering parameter
 
@@
 
-## Visualizing clustering results
+### Visualizing clustering results
 
@@
 
-```{r plot all umaps, fig.width = 12}
+```{r plot nn umaps, fig.width = 12}
 umap_plots <- cluster_results_list |>
@@
 
-## Evaluating clustering results
+### Evaluating clustering results
 
@@
 
-### Silhouette width and neighborhood purity
+#### Silhouette width and neighborhood purity
 
@@
 ```
+
 While there does not appear to be a salient difference among silhouette width distributions, it does appear that purity is higher with a higher nearest neighbors parameter.
 
-### Stability
+#### Stability
 
@@
 
+## Varying multiple clustering parameters
+
+The previous section demonstrated how to calculate clusters and QC metrics when varying one parameter, but it is possible to vary multiple parameters at once with `rOpenScPCA::sweep_clusters()`.
+In this section, we'll show an overview of how you might write code to vary two parameters at once (here, nearest neighbors and resolution as examples) and visualize results.
+
+```{r sweep two parameters}
+# Define vectors of parameters to vary
+nn_values <- seq(10, 30, 10)
+res_values <- seq(5, 15, 5) / 10
+
+cluster_results_list <- rOpenScPCA::sweep_clusters(
+  pca_matrix,
+  nn = nn_values,
+  resolution = res_values
+)
+```
+
+This resulting list now has 9 different clustering results, for each combination of `nn_values` and `res_values`:
+
+```{r length cluster_results_list two parameters}
+length(cluster_results_list)
+```
+
+### Visualize clusters
+
+Next, we'll iterate over `cluster_results_list` to plot the UMAPs.
+This time, we'll use `purrr::map()` and pull out parameters from each iteration's `cluster_df` to form the UMAP panel title.
+
+```{r plot nn res umaps, fig.height = 14}
+umap_plots <- cluster_results_list |>
+  purrr::map(
+    \(cluster_df) {
+      # Add a column with cluster assignments to umap_df
+      umap_df_plot <- umap_df |>
+        dplyr::mutate(cluster = cluster_df$cluster)
+
+      # Create a title for the UMAP with both parameters
+      umap_title <- glue::glue(
+        "nn: {unique(cluster_df$nn)}; res: {unique(cluster_df$resolution)}"
+      )
+
+      # Plot the UMAP, colored by the new cluster variable
+      ggplot(umap_df_plot, aes(x = UMAP1, y = UMAP2, color = cluster)) +
+        geom_point(alpha = 0.6) +
+        labs(title = umap_title) +
+        # We'll add a couple UMAP-specific plot settings
+        coord_equal() +
+        theme(
+          axis.ticks = element_blank(),
+          axis.text = element_blank(),
+          legend.position = "bottom"
+        )
+    }
+  )
+
+# Print the plots with patchwork::wrap_plots()
+patchwork::wrap_plots(umap_plots)
+```
+
+### Calculate and visualize QC metrics
+
+This section presents one coding strategy to calculate and visualize results when varying two clustering parameters.
+In particular, we use faceting to help display all information in one plot, by placing nearest neighbor values on the X-axis and faceting by resolution values.
+Since silhouette width and neighhorbood purity calculations using generally similar code, we'll just show neighborhood purity here.
+
+#### Neighborhood purity
+
+First, we'll calculate neighborhood purity and combine results into a single data frame.
+
+```{r calculate purity two parameters}
+purity_df <- cluster_results_list |>
+  purrr::map(
+    \(cluster_df) {
+      rOpenScPCA::calculate_purity(pca_matrix, cluster_df)
+    }
+  ) |>
+  dplyr::bind_rows()
+```
+
+We'll add a column `resolution_label` which we'll use to have informative panel titles in the faceted ggplot we make next.
+
+```{r add resolution_label column}
+purity_df <- purity_df |>
+  dplyr::mutate(resolution_label = glue::glue("Resolution: {resolution}"))
+```
+
+```{r visualize purity two parameters}
+ggplot(purity_df) +
+  aes(x = as.factor(nn), y = purity, fill = as.factor(nn)) +
+  geom_boxplot() +
+  scale_fill_brewer(palette = "Pastel2") +
+  # facet by resolution
+  facet_wrap(vars(resolution_label)) +
+  labs(
+    x = "Number of nearest neighbors",
+    y = "Neighborhood purity"
+  ) +
+  theme(legend.position = "none")
+```
+
+### Stability
+
+Similar to above, we'll calculate stability, combine results into a single data frame, add a `resolution_label` column to support plot interpretation, and finally make our plot.
+
+```{r calculate stability two parameters}
+stability_df <- cluster_results_list |>
+  purrr::map(
+    \(cluster_df) {
+      # Extract parameters for this clustering result
+      nn <- unique(cluster_df$nn)
+      resolution <- unique(cluster_df$resolution)
+
+      rOpenScPCA::calculate_stability(
+        pca_matrix,
+        cluster_df,
+        nn = nn,
+        resolution = resolution
+      )
+    }
+  ) |>
+  dplyr::bind_rows()
+
+stability_df <- stability_df |>
+  dplyr::mutate(resolution_label = glue::glue("Resolution: {resolution}"))
+```
+
+```{r visualize stability two parameters}
+ggplot(stability_df) +
+  aes(x = as.factor(nn), y = ari, fill = as.factor(nn)) +
+  geom_boxplot() +
+  scale_fill_brewer(palette = "Pastel2") +
+  # facet by resolution
+  facet_wrap(vars(resolution_label)) +
+  labs(
+    x = "Number of nearest neighbors",
+    y = "Adjusted Rand Index"
+  ) +
+  theme(legend.position = "none")
+```
+
 ## Session Info

### 2025-01-13 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
swap in new rprojroot line so this runs in CI, and re-render notebooks

@@
 # The base path for the OpenScPCA repository
-repository_base <- rprojroot::find_root(rprojroot::is_git_root)
+repository_base <- rprojroot::find_root(rprojroot::has_dir(".github"))
 
@@
 To begin, we'll read in the `SingleCellExperiment` (SCE) object.
-We'll also establish a corresponding processed Seurat object from its raw counts that we'll use for some examples.
 
@@
 Then, we'll use a simple Seurat pipeline to obtain clusters.
-<!-- TODO: We will want to reference this module for further documentation on this function: https://github.com/AlexsLemonade/OpenScPCA-analysis/issues/945 -->
 

### 2025-01-13 analyses/cell-type-ewings/exploratory_analysis/06-tumor-cell-state-assignment.Rmd
respond to reviews

@@
 In this notebook we look at expression of potential markers for tumor cell states in two libraries from `SCPCP000015`: `SCPCL000822` and `SCPCL000824`. 
-The expression of specific marker genes are then used to classify tumor cells into three different tumor cell states: `EWS-low`, `EWS-mid`, and `EWS-high`.
+The expression of specific marker genes are then used to classify tumor cells into three different tumor cell states: `EWS-low`, `EWS-mid`, `EWS-high`, and `EWS-high-proliferative`.
 
@@
 `SCPCL000822`:
+
 `EWS-FLI1` low  - 2
@@
 `SCPCL000824` :
+
 `EWS-FLI1` low - 1
@@
 - There is a clear division between cells that are `EWS-FLI` high and `EWS-FLI1` low. 
+  - We also see a set of cells that are both "proliferative" and `EWS-FLI` high. 
 - Most tumor cells in both samples are `EWS-FLI` high, which is consistent with the literature. 

### 2025-01-13 analyses/cell-type-ewings/scripts/utils/tumor-validation-helpers.R

@@ -175,7 +175,7 @@ calculate_mean_markers <- function(marker_genes_df,
   
   df <- data.frame(
     barcodes = names(mean_exp),
-    sum_exp = mean_exp
+    mean_exp = mean_exp
   )
   
   # get rid of extra " cells" at end of some of the names

### 2025-01-14 analyses/hello-clusters/02_compare-clustering-parameters.Rmd
swap in new rprojroot line so this runs in CI, and re-render notebooks
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 This function wraps the `rOpenScPCA::calculate_clusters()` function but allows you to provide a vector of parameter values to perform clustering across, as listed below.
-Clusters will be calculated for all combinations of parameters values (where applicable); default values that the function will use for any unspecified parameter values are shown in parentheses
+Clusters will be calculated for all combinations of parameters values (where applicable); default values that the function will use for any unspecified parameter values are shown in parentheses.
 
@@
 # Define nn parameter values of interest
-nn_values <- seq(10, 30, 10)
+nn_values <- c(10, 20, 30)
 
@@
       # calculate neighbhorhood purity
-      purity_df <- rOpenScPCA::calculate_purity(pca_matrix, cluster_df)
-
-      # Combine into a single data frame
-      dplyr::left_join(silhouette_df, purity_df)
+      purity_df <- rOpenScPCA::calculate_purity(pca_matrix, silhouette_df)
     }
@@
     \(cluster_df) {
-      nn <- unique(cluster_df$nn)
+      nn <- cluster_df$nn[1] # all rows have the same `nn` parameter, so we'll take the first
 
@@
 # Define vectors of parameters to vary
-nn_values <- seq(10, 30, 10)
-res_values <- seq(5, 15, 5) / 10
+nn_values <- c(10, 20, 30)
+res_values <-c(0.5, 1.0, 1.5)
 
@@
 # Print the plots with patchwork::wrap_plots()
-patchwork::wrap_plots(umap_plots)
+patchwork::wrap_plots(umap_plots, ncol = 3)
 ```
@@
   # facet by resolution
-  facet_wrap(vars(resolution_label)) +
+  facet_wrap(vars(resolution_label), labeller = label_both) +
   labs(

### 2025-01-14 analyses/hello-clusters/02_compare-clustering-parameters.Rmd
Update analyses/hello-clusters/02_compare-clustering-parameters.Rmd
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>
updates in response to review

@@
 
-`rOpenScPCA::sweep_clusters()` does allow you to specify values for any other parameters.
+`rOpenScPCA::sweep_clusters()` does not allow you to specify values for any other parameters.
 
@@
 
-It can be helpful (although it is not strictly necessary to keep track) to name this list by the varied `nn` parameter:
+It can be helpful (although it is not strictly necessary to keep track) to name this list by the varied `nn` parameter.
+In this case, we'll use these names to label plots.
 
@@
 ```{r combine cell metrics list}
-cell_metrics_df <- dplyr::bind_rows(cell_metric_list)
+cell_metrics_df <- purrr::list_rbind(cell_metric_list)
 ```
@@
 While there does not appear to be a salient difference among silhouette width distributions, it does appear that purity is higher with a higher nearest neighbors parameter.
+It's worth noting that this trend in purity values is expected: Higher nearest neighbor parameter values lead to fewer clusters, and neighborhood purity tends to be higher when there are fewer clusters.
 
@@
 ```{r combine plot stability}
-stability_df <- dplyr::bind_rows(stability_list)
+stability_df <- purrr::list_rbind(stability_list)
 
@@
 nn_values <- c(10, 20, 30)
-res_values <-c(0.5, 1.0, 1.5)
+res_values <- c(0.5, 1.0, 1.5)
 
@@
       umap_title <- glue::glue(
-        "nn: {unique(cluster_df$nn)}; res: {unique(cluster_df$resolution)}"
+        "nn: {cluster_df$nn[1]}; res: {cluster_df$resolution[1]}"
       )
@@
   ) |>
-  dplyr::bind_rows()
-```
-
-We'll add a column `resolution_label` which we'll use to have informative panel titles in the faceted ggplot we make next.
-
-```{r add resolution_label column}
-purity_df <- purity_df |>
-  dplyr::mutate(resolution_label = glue::glue("Resolution: {resolution}"))
+  purrr::list_rbind()
 ```
@@
   scale_fill_brewer(palette = "Pastel2") +
-  # facet by resolution
-  facet_wrap(vars(resolution_label), labeller = label_both) +
+  # facet by resolution, labeling panels with both the resolution column name and value
+  facet_wrap(vars(resolution), labeller = label_both) +
   labs(
@@
   ) |>
-  dplyr::bind_rows()
-
-stability_df <- stability_df |>
-  dplyr::mutate(resolution_label = glue::glue("Resolution: {resolution}"))
+  purrr::list_rbind()
 ```
@@
   scale_fill_brewer(palette = "Pastel2") +
-  # facet by resolution
-  facet_wrap(vars(resolution_label)) +
+  facet_wrap(vars(resolution), labeller = label_both) +
   labs(

### 2025-01-14 analyses/cell-type-ewings/exploratory_analysis/06-tumor-cell-state-assignment.Rmd

@@
 
-`EWS-FLI1` low  - 2
-`EWS-FLI1` middle - 1 & 4
-`EWS-FLI1` high - 3
+- `EWS-FLI1` low  - 2
+- `EWS-FLI1` middle - 1 & 4
+- `EWS-FLI1` high - 3
 
@@
 
-`EWS-FLI1` low - 1
-`EWS-FLI1` middle - 6 
-`EWS-FLI1` high - 3, 4, 5, 7
-`EWS-FLI1` high proliferative - 2
+- `EWS-FLI1` low - 1
+- `EWS-FLI1` middle - 6 
+- `EWS-FLI1` high - 3, 4, 5, 7
+- `EWS-FLI1` high proliferative - 2
 

### 2025-01-14 analyses/hello-clusters/02_compare-clustering-parameters.Rmd
update a comment
addidentally deleted an r
ensure legend fits, and render the html

@@
 
-      # calculate neighbhorhood purity
-      purity_df <- rOpenScPCA::calculate_purity(pca_matrix, silhouette_df)
+      # calculate neighbhorhood purity, starting from silhouette_df
+      rOpenScPCA::calculate_purity(pca_matrix, silhouette_df)
     }
@@
           axis.text = element_blank(),
-          legend.position = "bottom"
+          # Ensure legends fit in the figure
+          legend.position = "bottom",
+          legend.key.size = unit(0.2, "cm")
         )

### 2025-01-15 analyses/cell-type-ewings/exploratory_analysis/07-identify-max-rank-aucell.Rmd
notebook for identifying appropriate max rank

---
title: "Identify max rank for AUCell"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
---

This notebook is used to find the appropriate `aucMaxRank` to use with `AUCell` for all samples in `SCPCP00015`.

From the  [`AUCell` vignette]():

> It is important to check that most cells have at least the number of expressed/detected genes that are going to be used to calculate the AUC (aucMaxRank in calcAUC()). The histogram provided by AUCell_buildRankings() allows to quickly check this distribution.

Here we will build the rankings and look at the histogram showing the number of genes detected per cell for each sample and use that information to pick a threshold (percentage of genes) to use across all samples.

## Setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)

# set seed
set.seed(2025)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", "SCPCP000015")

# paths to all processed sce files
sce_files <- list.files(data_dir, pattern = "*processed.rds", full.names = TRUE, recursive = TRUE)
```

```{r}
# read in SCE files 
sce_list <- sce_files |> 
  purrr::map(readr::read_rds)
```

## Filter `SingleCellExperiment` objects

The default `aucMaxRank` is calculated as `ceiling(0.05*nrow(exprMat))` where the `exprMat` is the counts matrix. 
For the processed `SingleCellExperiment` objects in ScPCA we include all genes in the index, regardless of if they are detected in our sample. 
This means the rankings might include a bunch of genes that aren't actually present and the `aucMaxRank` would be calculated based on a set of genes that aren't even there, setting the threshold higher than it should be. 
Because of this, we probably want to filter the `SingleCellExperiment` objects to remove any genes that are not detected in any cells prior to calculating ranks or running `AUCell`.

```{r}
# get a list of filtered sce
# remove any genes that are not detected in any cells 
sce_list <- sce_list |> 
  purrr::map(\(sce){

    genes_to_remove <- rowData(sce)$detected > 0 
    filtered_sce <- sce[rowData(sce)$detected > 0 , ]

  })
```

## `AUCell` build rankings

Below we will look at the distribution of the number of genes detected per cell for each sample using the `AUCell_buildRankings()` function.

```{r}
cell_rankings_list <- sce_list |> 
  purrr::map(\(sce){AUCell::AUCell_buildRankings(counts(sce))})
```

Now we will look at the potential thresholds for `aucMaxRank` using 5% and 1% of genes. 
The numbers printed here represent the total number of genes that would be used as the `aucMaxRank` for a given dataset using the threshold. 
We can then use the histograms to see how many cells have at least that number of genes detected. 
Remember, we want to pick a number of genes that are detected in most cells.
If we picked a max rank higher than the number of genes detected in most cells, the non-detected genes that are randomly ordered would play an outsized role in our AUC values.

### 5%

```{r}
sce_list |> 
  purrr::map_dbl(\(sce){ceiling(nrow(counts(sce))*0.05)})
```

This looks too high based on our histograms. 
We would have a lot of cells where the number of genes detected is lower than the max rank, which is not ideal.

### 1%

```{r}
sce_list |> 
  purrr::map_dbl(\(sce){ceiling(nrow(counts(sce))*0.01)})
```

This looks better to me. 
I also think there's enough variation in some of the samples that we don't want to just pick one max rank for all samples, but instead use a percentage of the total genes detected.

## Conclusions

- Prior to running `AUCell` we should remove any genes that are not detected in any cells.
- Using a cutoff of 1% of the total genes detected seems like a reasonable choice for the `aucMaxRank`.

## Session info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-01-16 analyses/cell-type-ewings/exploratory_analysis/07-identify-max-rank-aucell.Rmd
add missing link
Update analyses/cell-type-ewings/exploratory_analysis/07-identify-max-rank-aucell.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>
make the report pretty

@@
 
-From the  [`AUCell` vignette]():
+From the  [`AUCell` vignette](https://bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html):
 
@@
 
-Here we will build the rankings and look at the histogram showing the number of genes detected per cell for each sample and use that information to pick a threshold (percentage of genes) to use across all samples.
+Here we will look at a histogram showing the number of genes detected per cell for each sample and use that information to pick a threshold (percentage of genes) to use across all samples.
 
@@
     genes_to_remove <- rowData(sce)$detected > 0 
-    filtered_sce <- sce[rowData(sce)$detected > 0 , ]
+    filtered_sce <- sce[genes_to_remove > 0 , ]
 
@@
 
-## `AUCell` build rankings
+## Number of genes detected
 
-Below we will look at the distribution of the number of genes detected per cell for each sample using the `AUCell_buildRankings()` function.
+Below we will look at the distribution of the number of genes detected per cell for each sample.
 
-```{r}
-cell_rankings_list <- sce_list |> 
-  purrr::map(\(sce){AUCell::AUCell_buildRankings(counts(sce))})
+```{r, fig.height = 10, fig.width=10, warning=FALSE}
+# make a list of the `countByCell` as AUCell does
+names(sce_list) <- stringr::str_extract(sce_files, "SCPCS\\d+")
+count_df_list <- sce_list |>
+  purrr::map(counts) |>
+  purrr::map(
+    \(counts) {
+      countByCell <- Matrix::colSums(counts > 0, na.rm = TRUE)
+      data.frame(count_by_cell = countByCell)
+    })
+
+# first, make the plots
+count_df_list |>
+  purrr::imap(
+    \(count_df, sample_id) {
+        count_hist <- ggplot(count_df) + 
+          aes(x = count_by_cell) + 
+          geom_histogram(color = "grey20", fill = "skyblue", bins = 30) + 
+          xlim(0, max(count_df$count_by_cell)) +
+          xlab("Number of genes detected per cell")
+
+        count_box <- ggplot(count_df) + 
+          aes(x = count_by_cell, y = "") + 
+          geom_boxplot(color = "grey20", fill = "skyblue", outlier.size = 0.5) + 
+          theme_void() + 
+          xlim(0, max(count_df$count_by_cell)) + 
+          ggtitle(sample_id)
+
+      count_box / count_hist + patchwork::plot_layout(heights = c(0.2, 1))
+    }
+  ) |>
+  patchwork::wrap_plots(nrow = 5)
+
+# second, print a table of the stats
+count_df_list |>
+  purrr::map(
+     \(count_df) {
+         c(
+          min=min(count_df$count_by_cell), 
+          quantile(count_df$count_by_cell, c(.01,.05, .10, .50, 1))
+        )
+     }) |>
+  dplyr::bind_rows(.id = "sample_id")
 ```
@@
 ```{r}
-sce_list |> 
+# get max rank using 1% threshold
+threshold_list <- sce_list |> 
   purrr::map_dbl(\(sce){ceiling(nrow(counts(sce))*0.05)})
+
+# make it legible
+data.frame(
+  max_rank = threshold_list
+)
 ```
@@
 ```{r}
-sce_list |> 
+# get max rank using 1% threshold
+threshold_list <- sce_list |> 
   purrr::map_dbl(\(sce){ceiling(nrow(counts(sce))*0.01)})
+
+# make it legible
+data.frame(
+  max_rank = threshold_list
+)
 ```

### 2025-01-16 analyses/cell-type-ewings/exploratory_analysis/07-identify-max-rank-aucell.Rmd
remove seed and add comment about using percentage

@@
 )
-
-# set seed
-set.seed(2025)
 ```
@@
 - Using a cutoff of 1% of the total genes detected seems like a reasonable choice for the `aucMaxRank`. 
+This allows us to account for the variation in genes detected per sample, but we should be sure to be careful when interpreting results as using different cutoffs for each sample could impact the AUC values.
 

### 2025-01-22 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd
notebook for exploring consensus results

---
title: "Explore consensus cell types"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: show
---

This notebook summarizes the findings from assigning consensus cell type labels to all ScPCA samples. 
All results from the `cell-type-consensus` module in `OpenScPCA-nf` must be saved to `results` prior to rendering this notebook.

## Functions

```{r}
# function to read in project data frames with all cells in a project
# output is a summarized table with total cells per sample, total cells per annotation, and number of cell types 
summarize_celltypes <- function(file, id){

  # read in data
  df <- readr::read_tsv(file)

  # get total cell count and number of assigned cell types per library
  total_cells_df <- df |> 
    dplyr::group_by(library_id) |> 
    dplyr::summarize(
      total_cells_per_library = length(library_id),
      num_celltypes = length(unique(consensus_annotation))
    )

  summary_df <- df |> 
    dplyr::group_by(library_id, consensus_annotation, consensus_ontology) |> 
    dplyr::summarize(total_cells_per_annotation = length(consensus_annotation)) |>
    dplyr::left_join(total_cells_df, by = "library_id") |> 
    dplyr::mutate(
      # add percentage 
      percent_cells_annotation = round((total_cells_per_annotation / total_cells_per_library) * 100, 2)
    ) |> 
    dplyr::ungroup()

  return(summary_df)

}

# stacked bar chart showing the distribution of the top 9 cell types for each project, including Unknown
celltype_distribution_plot <- function(label,
                                       results_df = all_results_df){

  # filter to get only results for the specified project 
  plot_df <- results_df |> 
    dplyr::filter(project_label == {{label}}) |>
    dplyr::mutate(
      # get most frequently observed cell types across libraries in that project 
      top_celltypes = forcats::fct_lump_n(consensus_annotation, 9, other_level = "All remaining cell types", ties.method = "first") |> 
        # sort by frequency 
        forcats::fct_infreq() |> 
        # make sure all remaining and unknown are last 
        forcats::fct_relevel("All remaining cell types", "Unknown", after = Inf)
    )

  # get color assignments 
  # put grey for unknoqn and all remaining cells at the end of list since they will show up last 
  # first figure out how many cell types are assigned to make sure colors are assigned properly 
  num_assigned_celltypes <- plot_df |> 
    dplyr::filter(!top_celltypes %in% c("Unknown")) |> 
    dplyr::pull(top_celltypes) |> 
    unique() |> 
    length()

  colors <- c(
    palette.colors(palette = "Set1")[1:num_assigned_celltypes], # first get colors for cell types that are not unknown or all remaining
    "grey95" # level for unknown
  ) 
  celltypes <- unique((plot_df$top_celltypes)) |> 
    sort() #make sure they are in the right order 
  names(colors) <- celltypes

  # make a stacked bar chart with top cell types 
  ggplot(plot_df) + 
    aes(
      x = library_id, 
      y = percent_cells_annotation, 
      fill = top_celltypes
    ) +
    geom_col() + 
    scale_y_continuous(expand = c(0,0)) +
    scale_fill_manual(values = colors, name = "cell type") +
    ggtitle(label) +
    theme(axis.text.x = element_blank())

}
```

## Data setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)
module_base <- file.path(repository_base, "analyses", "cell-type-consensus")

# results directory with cell-type-consensus 
results_dir <- file.path(module_base, "results", "cell-type-consensus")

# diagnoses table used for labeling plots 
diagnoses_file <- file.path(module_base, "sample-info", "project-diagnoses.tsv")
```

```{r}
# list all results files 
results_files <- list.files(opt$results_dir, pattern = "*_consensus-cell-types.tsv.gz", full.names = TRUE)

# get project ids from file list  
project_ids <- stringr::str_remove(basename(results_files), "_consensus-cell-types.tsv.gz")
names(results_files) <- project_ids

# remove cell line projects from file list
cell_line_projects <- c("SCPCP000020", "SCPCP000024")
results_files <- results_files[!names(results_files) %in% cell_line_projects]
project_ids <- names(results_files) #update project ids to remove cell line projects
```

```{r, message=FALSE}
# read in diagnoses
diagnoses_df <- readr::read_tsv(diagnoses_file)

# read in results and prep data frame for plotting 
all_results_df <- results_files |> 
  purrr::imap(summarize_celltypes) |> 
  dplyr::bind_rows(.id = "project_id") |> 
  # add in diagnoses 
  dplyr::left_join(diagnoses_df, by = "project_id") |> 
  dplyr::mutate(
    # remove leading 0s
    project_string = stringr::str_remove(project_id, "SCPCP0000"),
    # create a label for plotting
    project_label = glue::glue("{project_string}:{diagnosis}")
  )

```

## Is it all just Unknown?

The first thing we will look at is how many of the cells in each sample are categorized as "Unknown", which means no consensus between `SingleR` and `CellAssign` was identified.

```{r, fig.height=5}
unknown_only <- all_results_df |> 
  dplyr::filter(consensus_annotation == "Unknown")

ggplot(unknown_only, aes(x = project_label, y = percent_cells_annotation)) +
  ggforce::geom_sina(size = 0.1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.margin = margin(10,10,10,10)) +
  labs(
    x = "", 
    y = "Percent of cells annotated as Unknown"
  )

```

It looks like we do have some samples that aren't jus all "Unknown"!
It definitely varies by project, but for most projects we at least see some proportion of samples with assigned cell types.

Let's look at how many samples actually have some cells outside of unknown identified. 
To do this, we will identify all libraries that only have cells called as "Unknown".

```{r}
high_tumor_df <- unknown_only |> 
  dplyr::mutate(no_cells_identified = dplyr::if_else(percent_cells_annotation == 100, TRUE, FALSE)) |> 
  dplyr::group_by(project_label) |> 
  dplyr::summarize(all_unknown = sum(no_cells_identified),
                   classified_cells = sum(!no_cells_identified),
                   percentage_unknown = round(all_unknown/(all_unknown + classified_cells)*100, 2),
                   # add number of libraries for plotting 
                   total_libraries = length(library_id)) |>
  # set order for plots 
  dplyr::mutate(project_label = forcats::fct_reorder(project_label, total_libraries, .desc = TRUE))
```

Which projects have the highest proportion of samples with all "Unknown"?

```{r}
# table with percentage of samples 
high_tumor_df |> 
  dplyr::select(project_label, percentage_unknown) |> 
  dplyr::arrange(desc(percentage_unknown))

```

```{r, fig.height=5}
# pivot for plotting so above and below count is in the same column 
high_tumor_df <- high_tumor_df |> 
  tidyr::pivot_longer(
    cols = c("all_unknown", "classified_cells"),
    names_to = "category",
    values_to = "number_of_samples"
  )

ggplot(high_tumor_df, aes(x = project_label, y = number_of_samples, fill = category)) +
  geom_bar(position = "stack", stat= "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = rel(0.9))) +
  labs(
    x = "", 
    y = "Number of samples",
    fill = ""
  ) +
  scale_fill_manual(values = c( "#FFCB05", "#00274C"))
```

It looks like all projects do have cell types identified that are not "Unknown". 
However, `SCPCP000011` (retinoblastoma), has a fairly high percentage of samples without any consensus labels.

## Number of cell types observed

Below we look at the number of cell types observed in each project for all samples. 
This does not include cells labeled as "Unknown".

```{r}
num_celltypes_df <- all_results_df |> 
  # remove unknown as a cell type 
  dplyr::filter(consensus_annotation != "Unknown") |> 
  dplyr::select(project_label, library_id, num_celltypes) |> 
  unique()

ggplot(num_celltypes_df, aes(x = project_label, y = num_celltypes)) +
  ggforce::geom_sina(size = 0.1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = rel(0.9))) +
  labs(
    x = "", 
    y = "Number of cell types"
  )
```

## Distribution of consensus cell types

Now we look at the distribution of the cell types in each sample. 
For these plots, we will pull out the top 9 cell types for each project. 
All other cells will be labeld with "All remaining cell types".

The top cell types are determined by counting how many libraries each cell type is found in within a project and taking the most frequent types. 
Note that because the top cell types are different for each project, the colors will be different for assigned cell types. 
The only colors that will stay the same are those for "Unknown" and "All remaining cell types".

```{r}
project_labels <- unique(all_results_df$project_label)

project_labels |> 
  purrr::map(celltype_distribution_plot)
```

This looks really promising!
A few observations:

- Cell types identified tend to line up with expectations for the type of tumor. 
For example, leukemia libraries have T and B cells, brain tumors have macrophages, and solid tumors have fibroblasts and muscle cells. 
- Projects that I would expect to be more difficult to classify (sarcomas, wilms, RB) have fewer cells classified then things like brain and leukemia. 
Notably many of the solid tumor projects (4, 5, 12-16, and 23) have a handful of PDX samples where I would expect to see fewer normal cells.

## Most frequently observed cell types

The last thing we will do is look at the most frequently observed cell types across all samples. 
The below table is ordered by the number of libraries the cell type is observed.

```{r}
all_results_df |> 
  dplyr::filter(consensus_annotation != "Unknown") |> 
  dplyr::group_by(consensus_annotation) |> 
  dplyr::summarize(
    total_libraries = length(library_id),
    min_percentage = min(percent_cells_annotation),
    mean_percentage = round(mean(percent_cells_annotation), 2),
    max_percentage = max(percent_cells_annotation)
  ) |> 
  dplyr::arrange(desc(total_libraries))

```

## Session info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-01-23 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd
use alphabet palette

@@
 celltype_distribution_plot <- function(label,
-                                       results_df = all_results_df){
+                                       results_df,
+                                       colors){
 
   # filter to get only results for the specified project 
-  plot_df <- results_df |> 
+    plot_df <- plot_df |> 
       dplyr::filter(project_label == {{label}}) |> 
       dplyr::mutate(
-      # get most frequently observed cell types across libraries in that project 
-      top_celltypes = forcats::fct_lump_n(consensus_annotation, 9, other_level = "All remaining cell types", ties.method = "first") |> 
-        # sort by frequency 
-        forcats::fct_infreq() |> 
-        # make sure all remaining and unknown are last 
+        # relevel factors for specific project 
+        top_celltypes = forcats::fct_infreq(top_celltypes) |> 
           forcats::fct_relevel("All remaining cell types", "Unknown", after = Inf)
       )
-
-  # get color assignments 
-  # put grey for unknoqn and all remaining cells at the end of list since they will show up last 
-  # first figure out how many cell types are assigned to make sure colors are assigned properly 
-  num_assigned_celltypes <- plot_df |> 
-    dplyr::filter(!top_celltypes %in% c("Unknown")) |> 
-    dplyr::pull(top_celltypes) |> 
-    unique() |> 
-    length()
-
-  colors <- c(
-    palette.colors(palette = "Set1")[1:num_assigned_celltypes], # first get colors for cell types that are not unknown or all remaining
-    "grey95" # level for unknown
-  ) 
-  celltypes <- unique((plot_df$top_celltypes)) |> 
-    sort() #make sure they are in the right order 
-  names(colors) <- celltypes
 
@@
 # list all results files 
-results_files <- list.files(opt$results_dir, pattern = "*_consensus-cell-types.tsv.gz", full.names = TRUE)
+results_files <- list.files(results_dir, pattern = "*_consensus-cell-types.tsv.gz", full.names = TRUE)
 
@@
 ```{r}
+plot_df <- all_results_df |> 
+    dplyr::group_by(project_id) |> 
+    dplyr::mutate(
+      # get most frequently observed cell types across libraries in that project 
+      top_celltypes = forcats::fct_lump_n(consensus_annotation, 9, other_level = "All remaining cell types", ties.method = "first") |> 
+        # sort by frequency 
+        forcats::fct_infreq() |> 
+        # make sure all remaining and unknown are last, use this to assign colors in specific order
+        forcats::fct_relevel("All remaining cell types", "Unknown", after = Inf)
+    )
+
+# get all unique cell types ordered by frequency 
+unique_celltypes <- plot_df |> 
+  dplyr::filter(!top_celltypes %in% c("All remaining cell types", "Unknown")) |> 
+  dplyr::pull(top_celltypes) |> 
+  unique() |>
+  sort() |> 
+  as.character()
+
+# get color palette
+colors <- c(
+  palette.colors(palette = "alphabet"),
+  "black", # 1 extra since alphabet is 26 and we have 27, this will be plasma cell which shows up once 
+  "grey60", 
+  "grey95"
+)
+names(colors) <- c(unique_celltypes, "All remaining cell types", "Unknown")
+```
+
+```{r}
 project_labels <- unique(all_results_df$project_label)
@@
 project_labels |> 
-  purrr::map(celltype_distribution_plot)
+  purrr::map(\(label){
+    celltype_distribution_plot(label, plot_df, colors)
+    })
 ```

### 2025-01-23 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 cell_line_projects <- c("SCPCP000020", "SCPCP000024")
-results_files <- results_files[!names(results_files) %in% cell_line_projects]
-project_ids <- names(results_files) #update project ids to remove cell line projects
+project_ids <- setdiff(project_ids, cell_line_projects) # remove cell line projects
+results_files <- results_files[project_ids]
 ```
@@
   ggforce::geom_sina(size = 0.1) +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1),
+  theme(axis.text.x = element_text(angle = 90, hjust = 0.5),
         plot.margin = margin(10,10,10,10)) +
@@
 
-ggplot(num_celltypes_df, aes(x = project_label, y = num_celltypes)) +
-  ggforce::geom_sina(size = 0.1) +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = rel(0.9))) +
-  labs(
-    x = "", 
-    y = "Number of cell types"
-  )
+ggplot(num_celltypes_df, aes(x = num_celltypes)) +
+  geom_histogram(binwidth = 1, center = 0) +
+  facet_wrap(project_label)
 ```
@@
   dplyr::summarize(
-    total_libraries = length(library_id),
+    total_libraries = dplyr::n(),
     min_percentage = min(percent_cells_annotation),
     mean_percentage = round(mean(percent_cells_annotation), 2),
+    median_percentage = median(percent_cells_annotation),
     max_percentage = max(percent_cells_annotation)

### 2025-01-23 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd
adjust plots based on review

@@
 All results from the `cell-type-consensus` module in `OpenScPCA-nf` must be saved to `results` prior to rendering this notebook.
+
+```{r packages}
+suppressPackageStartupMessages({
+  # load required packages
+  library(ggplot2)
+})
+
+# Set default ggplot theme
+theme_set(
+  theme_classic()
+)
+```
 
@@
 }
-
-# stacked bar chart showing the distribution of the top 9 cell types for each project, including Unknown
-celltype_distribution_plot <- function(label,
-                                       results_df,
-                                       colors){
-
-  # filter to get only results for the specified project 
-    plot_df <- plot_df |> 
-      dplyr::filter(project_label == {{label}}) |> 
-      dplyr::mutate(
-        # relevel factors for specific project 
-        top_celltypes = forcats::fct_infreq(top_celltypes) |> 
-          forcats::fct_relevel("All remaining cell types", "Unknown", after = Inf)
-      )
-
-    # make a stacked bar chart with top cell types 
-    ggplot(plot_df) + 
-      aes(
-        x = library_id, 
-        y = percent_cells_annotation, 
-        fill = top_celltypes
-      ) +
-      geom_col() + 
-      scale_y_continuous(expand = c(0,0)) +
-      scale_fill_manual(values = colors, name = "cell type") +
-      ggtitle(label) +
-      theme(axis.text.x = element_blank())
-
-}
 ```
@@
 ## Data setup
-
-```{r packages}
-suppressPackageStartupMessages({
-  # load required packages
-  library(ggplot2)
-})
-
-# Set default ggplot theme
-theme_set(
-  theme_classic()
-)
-```
 
@@
 # list all results files 
-results_files <- list.files(results_dir, pattern = "*_consensus-cell-types.tsv.gz", full.names = TRUE)
+results_files <- list.files(results_dir, pattern = "_consensus-cell-types\\.tsv.\\gz$", full.names = TRUE)
 
@@
   dplyr::mutate(
-    # remove leading 0s
-    project_string = stringr::str_remove(project_id, "SCPCP0000"),
     # create a label for plotting
-    project_label = glue::glue("{project_string}:{diagnosis}")
+    project_label = glue::glue("{project_id}:{diagnosis}")
   )
@@
 
-```{r, fig.height=5}
+```{r, fig.height=7}
 unknown_only <- all_results_df |> 
@@
   ggforce::geom_sina(size = 0.1) +
-  theme(axis.text.x = element_text(angle = 90, hjust = 0.5),
+  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
         plot.margin = margin(10,10,10,10)) +
@@
 
-It looks like we do have some samples that aren't jus all "Unknown"!
+It looks like we do have some samples that aren't just all "Unknown"!
 It definitely varies by project, but for most projects we at least see some proportion of samples with assigned cell types.
@@
 high_tumor_df <- unknown_only |> 
-  dplyr::mutate(no_cells_identified = dplyr::if_else(percent_cells_annotation == 100, TRUE, FALSE)) |> 
+  dplyr::mutate(no_cells_identified = percent_cells_annotation == 100) |> 
   dplyr::group_by(project_label) |> 
@@
 
-```{r, fig.height=5}
-# pivot for plotting so above and below count is in the same column 
-high_tumor_df <- high_tumor_df |> 
-  tidyr::pivot_longer(
-    cols = c("all_unknown", "classified_cells"),
-    names_to = "category",
-    values_to = "number_of_samples"
-  )
-
-ggplot(high_tumor_df, aes(x = project_label, y = number_of_samples, fill = category)) +
-  geom_bar(position = "stack", stat= "identity") +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = rel(0.9))) +
-  labs(
-    x = "", 
-    y = "Number of samples",
-    fill = ""
-  ) +
-  scale_fill_manual(values = c( "#FFCB05", "#00274C"))
-```
-
 It looks like all projects do have cell types identified that are not "Unknown". 
@@
 
-```{r}
+```{r, fig.height=10}
 num_celltypes_df <- all_results_df |> 
+  # add a new line for facet labels 
+  dplyr::mutate(facet_label = glue::glue("{project_id}\n{diagnosis}")) |>
   # remove unknown as a cell type 
   dplyr::filter(consensus_annotation != "Unknown") |> 
-  dplyr::select(project_label, library_id, num_celltypes) |> 
+  dplyr::select(facet_label, library_id, num_celltypes) |> 
   unique()
@@
   geom_histogram(binwidth = 1, center = 0) +
-  facet_wrap(project_label)
+  facet_wrap(vars(facet_label), 
+             ncol = 3) +
+  labs(
+    x = "Number of cell types"
+  ) +
+  theme_bw()
 ```
@@
 For these plots, we will pull out the top 9 cell types for each project. 
-All other cells will be labeld with "All remaining cell types".
+All other cells will be labeled with "All remaining cell types".
 
 The top cell types are determined by counting how many libraries each cell type is found in within a project and taking the most frequent types.
-Note that because the top cell types are different for each project, the colors will be different for assigned cell types. 
-The only colors that will stay the same are those for "Unknown" and "All remaining cell types".
 
@@
 
-```{r}
+```{r, fig.height=60, fig.width=10}
 project_labels <- unique(all_results_df$project_label)
 
+# stacked bar chart showing the distribution of the top 9 cell types for each project, including Unknown
 project_labels |> 
   purrr::map(\(label){
-    celltype_distribution_plot(label, plot_df, colors)
-    })
+
+    project_df <- plot_df |> 
+      dplyr::filter(project_label == {{label}}) |> 
+      dplyr::mutate(
+        # relevel factors for specific project 
+        top_celltypes = forcats::fct_infreq(top_celltypes) |> 
+          forcats::fct_relevel("All remaining cell types", "Unknown", after = Inf)
+      )
+
+    # make a stacked bar chart with top cell types 
+    ggplot(project_df) + 
+      aes(
+        x = library_id, 
+        y = percent_cells_annotation, 
+        fill = top_celltypes
+      ) +
+      geom_col() + 
+      scale_y_continuous(expand = c(0,0)) +
+      scale_fill_manual(values = colors, name = "cell type") +
+      ggtitle(label) +
+      theme(axis.text.x = element_blank())
+
+    }) |>
+  patchwork::wrap_plots(ncol = 1)
 ```

### 2025-01-23 analyses/cell-type-ewings/template_notebooks/celltype-exploration.Rmd
initiate "guide" notebook

---
title: "Template notebook for validating cell type assignments for an individual library in SCPCP000015"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  sample_id: "SCPCS000490"
  library_id: "SCPCL000822" 
  cluster_nn: 20
  cluster_res: 0.5
---

This notebook is meant to be a guide for compiling "final" cell type annotations for an individual library in `SCPCP000015`. 
Results from `aucell-singler-annotation.sh`, `evaulate-clusters.sh`, and `run-aucell-ews-signatures.sh` are all combined and used to assign cell type annotations on a case by case basis.

Instructions for using this guide:

1. Ensure that you have a local copy of the results from `aucell-singler-annotation.sh`, `evaluate-clusters.sh` and `run-aucell-ews-signatures.sh` saved to `results`. 
2. Copy the contents of this notebook to a new notebook titled `<library_id>_celltype-exploration.Rmd` and save in `exploratory_analysis/final_annotation_notebooks`. 
3. Replace the `sample_id` and `library_id` with the correct IDs in the `params` list. 
4. Optionally, you may choose to update the choices for clustering based on the results from `evaluate-clusters.sh`. 
All clusters used will be calculated with the Leiden algorithm and the modularity objective function. 
To modify the nearest neighbors (default: 20) and resolution (default: 0.5) chosen use the `cluster_nn` and `cluster_res` params. 
5. Run through the notebook and update any sections of the notebook marked with `**Manual exploration**`. 
6. Render the completed notebook which will produce the rendered `html` file and a TSV with cell type annotations for that library.

## Setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)

# set seed
set.seed(2024)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", "SCPCP000015")

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-ewings") 
```

```{r}
# path to sce 
sce_file <- file.path(data_dir, params$sample_id, glue::glue("{params$library_id}_processed.rds"))

# path to workflow results
workflow_results_dir <- file.path(module_base, "results")

singler_results_dir <- file.path(workflow_results_dir, "aucell_singler_annotation", params$sample_id)
singler_results_file <- file.path(singler_results_dir,
                                  glue::glue("{params$library_id}_singler-classifications.tsv"))

cluster_results_dir <- file.path(workflow_results_dir, "clustering", params$sample_id)
cluster_results_file <- file.path(cluster_results_dir,
                                  glue::glue("{params$library_id}_cluster-results.tsv"))

aucell_results_dir <- file.path(workflow_results_dir, "aucell-ews-signatures", params$sample_id)
aucell_results_file <- file.path(aucell_results_dir,
                                 glue::glue("{params$library_id}_auc-ews-gene-signatures.tsv"))

# small gene sets
visser_marker_genes_file <- file.path(module_base, "references", "visser-all-marker-genes.tsv")
cell_state_genes_file <- file.path(module_base, "references", "tumor-cell-state-markers.tsv")
```

```{r}
# output file to save final annotations 
results_dir <- file.path(module_base, "results", "final-annotations")
output_file <- file.path(results_dir, glue::glue("{params$library_id}_celltype-annotations.tsv"))
```

```{r}
# source in setup functions prep_results()
setup_functions <- file.path(module_base, "template_notebooks", "utils", "setup-functions.R")
source(setup_functions)

# source in validation functions calculate_mean_markers()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)
```

```{r}
stopifnot(
  "sce file does not exist" = file.exists(sce_file),
  "singler results file does not exist" = file.exists(singler_results_file),
  "cluster results file does not exist" = file.exists(cluster_results_file),
  "aucell results file does not exist" = file.exists(aucell_results_file)
)
```

```{r, message=FALSE}
# read in sce
sce <- readr::read_rds(sce_file)

# read in workflow results
singler_df <- readr::read_tsv(singler_results_file)
cluster_df <- readr::read_tsv(cluster_results_file)
aucell_df <- readr::read_tsv(aucell_results_file)

# read in marker genes and combine into one list 
visser_markers_df <- readr::read_tsv(visser_marker_genes_file) |> 
  dplyr::select(cell_type, ensembl_gene_id) |> 
  unique()

cell_state_markers_df <- readr::read_tsv(cell_state_genes_file) |> 
  dplyr::select(cell_type = cell_state, ensembl_gene_id)

all_markers_df <- dplyr::bind_rows(list(visser_markers_df, cell_state_markers_df))
```

## Prepare data for plotting

```{r}
all_results_df <- prep_results(
  sce, 
  singler_df, 
  cluster_df, 
  aucell_df,
  cluster_nn = params$cluster_nn,
  cluster_res = params$cluster_res
  )

# calculate gene set means for each cell type/cell state
cell_types <- unique(all_markers_df$cell_type)

# get the mean expression of all genes for each cell state
gene_exp_df <- cell_types |>
  purrr::map(\(type){
    calculate_mean_markers(all_markers_df, sce, type, cell_type)
  }) |>
  purrr::reduce(dplyr::inner_join, by = "barcodes")

all_info_df <- all_results_df |> 
  dplyr::left_join(gene_exp_df, by = "barcodes")
```

## Summary of workflow results

TODO: Insert plots that will summarize findings from each of the workflows
- UMAPs of SingleR, clusters, AUC values and custom gene set means 
- Density plots by cluster of AUC values and custom gene set means
- Maybe heatmaps with cluster annotation of AUC scores and custom gene set means

## Re-cluster tumor cells **Manual exploration**

<!-- Use this section to label tumor cells based on the above findings. 
Any cells that are labeled as tumor will then be re-clustered and plots showing only tumor cells
can be created to identify tumor cell states -->

TODO: Functions for re-clustering tumor cells
Show the same plots across the tumor only clusters and assign tumor cell states to each cluster

## Additional exploration **Manual exploration**

<!--This section is for any additional exploration that may be needed to finalize annotations for this library. 
If not using, please delete this section.
For example, here you may want to dive into the normal cell types and make adjustments as needed
-->

## Validate final tumor and normal annotation **Manual exploration**

<!-- This section should be used to update the assignments -->

TODO: Insert plots that will be useful for validation (UMAPs, heatmaps, density plots)

## Prepare annotations **Manual exploration**

<!-- This section should be used to create the final.final table with cell type annotations for export -->

TODO: Code and instructions for exporting such as what columns should be named.

## Session info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-01-24 analyses/cell-type-ewings/template_notebooks/utils/setup-functions.R
Apply suggestions from code review
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@ -1,7 +1,6 @@
 # These functions are used in `celltype-exploration.Rmd` 
 # They are used for reading in and setting up the cell type results
 
-# read in and prepare combined data frame 
 #' Combine workflow results into a single data frame
 #'
 #' @param sce Processed SingleCellExperiment object with UMAP embeddings 

### 2025-01-24 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd

@@
     project_df <- plot_df |> 
-      dplyr::filter(project_label == {{label}}) |> 
+      dplyr::filter(project_label == label) |> 
       dplyr::mutate(

### 2025-01-24 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd

@@
 params:
-  predicted.celltype.threshold: 0.85
+  predicted.celltype.threshold: 0.75
   cnv_threshold: 0
@@
 | ---------------------- | ----------------------- | ---------------------- | --------------------------- | --------------- |
-| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr  & cnv_score < cnv.thr` | `VWF`| no CNV |
-| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr & cnv_score < cnv.thr` | `PTPRC`, `CD163`, `CD68`| no CNV |
-| normal                 | kidney                  | `cell_type %in% c("kidney cell","kidney epithelial", "podocyte") & predicted.score > pred.thr & cnv_score < cnv.thr` | `CDH1`, `PODXL`, `LTL`| no CNV |
+| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr  | `VWF`| no CNV |
+| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr  | `PTPRC`, `CD163`, `CD68`| no CNV |
+| normal                 | kidney                  | `compartment == "fetal_nephron" & predicted.score > pred.thr & cnv_score < cnv.thr` | `CDH1`, `PODXL`, `LTL`| no CNV |
 | normal                 | stroma                  | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr`| `VIM`| no CNV |
@@
 ### Output files
+
+The report will be saved in the `notebook` directory.
+The final annotation file `SCPCP000006-annotations.tsv` will be saved in the `results` directory.
 
@@
 
-### Output file
-
-The report will be saved in the `notebook` directory.
-
 ## Functions
@@
 We simply checked for each chromosome if the cell `has_cnv_chr`.
-Would the cell have more than `cnv_threshold` chromosome with CNV, the global `has_cnv_score` will be TRUE.
-Else, the cell will have a `has_cnv_score` set to FALSE.
+Would the cell have less than `cnv_threshold` chromosome with CNV, the global `has_cnv_score` will be FALSE.
+Would the cell have more than `cnv_threshold` +2 chromosome with CNV, the global `has_cnv_score` will be TRUE.
+The default is set to `unknown`.
+
+The `infercnv` method can lead to false positive CNV.
+We introduced a flexibility of `+2` over the `cnv_threshold` to reduce the risk of misclassification of normal cells as cancer cells.
+
+Please note that some cancer cell might not have any CNV. 
+There is thus a risk of misclassifying cancer cells as normal cells. 
+However, we cannot avoid this risk with the data and tools currently available.
+This is a limitation of our annotations.
 
@@
   mutate(has_cnv_score = case_when(
-    has_cnv_score > params$cnv_threshold ~ "CNV",
-    has_cnv_score <= params$cnv_threshold ~ "no CNV"
+    has_cnv_score > params$cnv_threshold +2 ~ "CNV",
+    has_cnv_score <= params$cnv_threshold ~ "no CNV",
+    .default = "unknown"
   ))
 
-table(cell_type_df$has_cnv_score)
 ```
@@
     # assign normal/cancer based on condition
-    compartment %in% c("fetal_nephron", "stroma") & has_cnv_score == "no CNV" ~ "normal",
-    compartment %in% c("endothelium", "immune") & cell_type.score > params$predicted.celltype.threshold ~ "normal",
-    compartment %in% c("fetal_nephron", "stroma") & has_cnv_score == "CNV" ~ "cancer",
+    compartment %in% c("fetal_nephron", "stroma") & 
+      compartment.score > params$predicted.celltype.threshold & 
+      has_cnv_score == "no CNV" ~ "normal",
+
+    compartment %in% c("endothelium", "immune") &
+      cell_type.score > params$predicted.celltype.threshold ~ "normal",
+
+    compartment %in% c("fetal_nephron", "stroma") & 
+      has_cnv_score == "CNV" ~ "cancer",
+
     .default = "unknown"
   ))
+
 ```
@@
   theme(text = element_text(size = 22))
+
+DT::datatable(table(cell_type_df$first.level_annotation, cell_type_df$sample_id))
+
+```
+
+Strikingly, 5 samples show more normal cells than cancer cells:
+
+- `SCPCS000177`
+- `SCPCS000180`
+- `SCPCS000181`
+- `SCPCS000190`
+- `SCPCS000197`
+
+These five samples do not have enough immune and endothelium cell to be used as a normal reference in `scripts/06_infercnv.R` and we previously decided to run `scripts/06_infercnv.R` without any reference for these samples.
+In that case, the mean expression profile over the sample is taken as the normal reference, biaising the inference of CNV.
+
+For those samples, the annotations based on `infercnv` cannot be trust.
+To avoid any confusion in the following steps of the analysis, we decided to force the annotation of the entire samples to `unknown`.
+
+```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
+
+cell_type_df$first.level_annotation[cell_type_df$sample_id %in% c("SCPCS000177", "SCPCS000180", "SCPCS000181", "SCPCS000190", "SCPCS000197")] <- "unknown"
+
+DT::datatable(table(cell_type_df$first.level_annotation, cell_type_df$sample_id))
+
 ```
@@
     cell_type_df$compartment %in% c("fetal_nephron") &
+      cell_type_df$compartment.score > params$predicted.celltype.threshold &
       cell_type_df$has_cnv_score == "no CNV" ~ "kidney",
+
     cell_type_df$compartment %in% c("stroma") &
+      cell_type_df$compartment.score > params$predicted.celltype.threshold &
       cell_type_df$has_cnv_score == "no CNV" ~ "normal stroma",
+
     cell_type_df$compartment %in% c("endothelium") &
-      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
-        cell_type_df$has_cnv_score == "no CNV") ~ "endothelium",
+      (cell_type_df$compartment.score > params$predicted.celltype.threshold) ~ "endothelium",
     cell_type_df$compartment %in% c("immune") &
-      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
-        cell_type_df$has_cnv_score == "no CNV") ~ "immune",
+      (cell_type_df$compartment.score > params$predicted.celltype.threshold) ~ "immune",
 
@@
   ))
+
+cell_type_df$second.level_annotation[cell_type_df$sample_id %in% c("SCPCS000177", "SCPCS000180", "SCPCS000181", "SCPCS000190", "SCPCS000197")] <- "unknown"
 ```
 
+#### `Second.level_annotation` of cancer and normal cells - `umap` reduction
+
 ```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
-ggplot(cell_type_df[cell_type_df$first.level_annotation == "normal", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1) +
-  geom_point() +
+ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
+  geom_point(size = 0.2, shape = 19) +
   facet_wrap(facets = ~sample_id, ncol = 5) +
@@
 
+#### `Second.level_annotation` of cancer and normal cells without `unknown` - `umap` reduction
+
 ```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
-ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 0.1) +
-  geom_point() +
+tmp <- cell_type_df[cell_type_df$second.level_annotation != "unknown",]
+ggplot(tmp, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
+  geom_point(size = 0.2, shape = 19) +
   facet_wrap(facets = ~sample_id, ncol = 5) +
@@
 
-#### Cancer and normal cells
+#### `Second.level_annotation` of cancer cells only - `umap` reduction
 
 ```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
-ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1) +
-  geom_point() +
+ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
+  geom_point(size = 0.2, shape = 19) +
   facet_wrap(facets = ~sample_id, ncol = 5) +
@@
 
+Per sample, the annotations in the `umap` reduction seem to make sense as we can the three main Wilms tumor components are easily distinguished:
+
+- cancer blastema
+- epithelium
+- stroma
+
 ### Validation cancer versus normal based on the CNV profile
+
+We look for each `second.level_annotation` and `sample_id`, the proportion of cells that has CNV in each of the chromosome.
 
@@
 for (i in 1:22) {
-  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("proportion_cnv_chr", i)))
+  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_cnv_chr", i)))
 }
@@
 annotations_table <- cell_type_df |>
-  select(
+  dplyr::select(
     cell_barcode = barcode,
@@
 
-- Combining label transfer and CNV inference we have produced draft annotations for all 40 Wilms tumor samples in SCPCP000006
+- Combining label transfer and CNV inference we have produced draft annotations for 35/40 Wilms tumor samples in SCPCP000006.
+We decided not to annotate samples for which the `infercnv` couldn't be run with a normal reference, as we don't trust the results in that case and want to avoid misclassification (`SCPCS000177`, `SCPCS000180`, `SCPCS000181`, `SCPCS000190`, `SCPCS000197`).
 

### 2025-01-24 analyses/cell-type-ewings/template_notebooks/celltype-exploration.Rmd
Apply suggestions from code review
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>
some minor review responses

@@
 2. Copy the contents of this notebook to a new notebook titled `<library_id>_celltype-exploration.Rmd` and save in `exploratory_analysis/final_annotation_notebooks`. 
-3. Replace the `sample_id` and `library_id` with the correct IDs in the `params` list. 
+3. Update the `title` in the `yaml` section and replace the `sample_id` and `library_id` with the correct IDs in the `params` list. 
 4. Optionally, you may choose to update the choices for clustering based on the results from `evaluate-clusters.sh`. 
@@
 
-# calculate gene set means for each cell type/cell state
 cell_types <- unique(all_markers_df$cell_type)
@@
 
-## Re-cluster tumor cells **Manual exploration**
+## Re-cluster tumor cells {.manual-exploration}
 
@@
 
-## Additional exploration **Manual exploration**
+## Additional exploration {.manual-exploration}
 
@@
 
-## Validate final tumor and normal annotation **Manual exploration**
+## Validate final tumor and normal annotation {.manual-exploration}
 
@@
 
-## Prepare annotations **Manual exploration**
+## Prepare annotations {.manual-exploration}
 

### 2025-01-27 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 | ---------------------- | ----------------------- | ---------------------- | --------------------------- | --------------- |
-| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr  | `VWF`| no CNV |
+| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr`  | `VWF`| no CNV |
 | normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr  | `PTPRC`, `CD163`, `CD68`| no CNV |

### 2025-01-27 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

--- 2025-01-24 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
@@
 | ---------------------- | ----------------------- | ---------------------- | --------------------------- | --------------- |
-| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr  | `VWF`| no CNV |
-| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr  | `PTPRC`, `CD163`, `CD68`| no CNV |
+| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr`  | `VWF`| no CNV |
+| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr`  | `PTPRC`, `CD163`, `CD68`| no CNV |
 | normal                 | kidney                  | `compartment == "fetal_nephron" & predicted.score > pred.thr & cnv_score < cnv.thr` | `CDH1`, `PODXL`, `LTL`| no CNV |

### 2025-01-27 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
   mutate(has_cnv_score = case_when(
-    has_cnv_score > params$cnv_threshold +2 ~ "CNV",
+    has_cnv_score > (params$cnv_threshold +2) ~ "CNV",
     has_cnv_score <= params$cnv_threshold ~ "no CNV",

### 2025-01-27 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
add parameters for cnv_threshold and change predicted.celltype.threshold default to 0.85

--- 2025-01-27 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
@@
 params:
-  predicted.celltype.threshold: 0.75
-  cnv_threshold: 0
+  predicted.celltype.threshold: 0.85
+  cnv_threshold: !r c(0, 2)
   testing: 0
@@
 We simply checked for each chromosome if the cell `has_cnv_chr`.
-Would the cell have less than `cnv_threshold` chromosome with CNV, the global `has_cnv_score` will be FALSE.
-Would the cell have more than `cnv_threshold` +2 chromosome with CNV, the global `has_cnv_score` will be TRUE.
+Would the cell have less than `r params$cnv_threshold[1]` chromosome with CNV, the global `has_cnv_score` will be FALSE.
+Would the cell have more than `r params$cnv_threshold[2]`  chromosome with CNV, the global `has_cnv_score` will be TRUE.
 The default is set to `unknown`.
@@
 The `infercnv` method can lead to false positive CNV.
-We introduced a flexibility of `+2` over the `cnv_threshold` to reduce the risk of misclassification of normal cells as cancer cells.
+We introduced a flexibility of +`r params$cnv_threshold[2]` over the `cnv_threshold` to reduce the risk of misclassification of normal cells as cancer cells.
 
@@
   mutate(has_cnv_score = case_when(
-    has_cnv_score > params$cnv_threshold +2 ~ "CNV",
-    has_cnv_score <= params$cnv_threshold ~ "no CNV",
+    has_cnv_score > params$cnv_threshold[2] ~ "CNV",
+    has_cnv_score <= params$cnv_threshold[1] ~ "no CNV",
     .default = "unknown"
@@
 The threshold used for the `predicted.score` is defined as a parameter of this notebook as `r params$predicted.celltype.threshold`.
-The threshold used for the identification of CNV is also defined as the notebook parameter `r params$cnv_threshold`.
+The thresholds used for the identification of CNV are also defined as notebook parameters with a minimum of `r params$cnv_threshold[1]` and a maximum of `r params$cnv_threshold[2]`.
 

### 2025-01-27 analyses/cell-type-ewings/template_notebooks/celltype-exploration.Rmd
add section for summary plots

@@
 To modify the nearest neighbors (default: 20) and resolution (default: 0.5) chosen use the `cluster_nn` and `cluster_res` params. 
-5. Run through the notebook and update any sections of the notebook marked with `**Manual exploration**`. 
+5. Run through the notebook and update any sections of the notebook marked with the `{.manual-exploration}` tag. 
 6. Render the completed notebook which will produce the rendered `html` file and a TSV with cell type annotations for that library.
@@
 set.seed(2024)
+
+# quiet messages
+options(readr.show_col_types = FALSE)
+ComplexHeatmap::ht_opt(message = FALSE)
 ```
@@
 
-TODO: Insert plots that will summarize findings from each of the workflows
-- UMAPs of SingleR, clusters, AUC values and custom gene set means 
-- Density plots by cluster of AUC values and custom gene set means
-- Maybe heatmaps with cluster annotation of AUC scores and custom gene set means
+### `aucell-singler-annotation` assignments
+
+The below UMAP shows the top cell types assigned by `SingleR` in the `aucell-singler-annotation.sh` workflow. 
+The top 7 cell types are shown and all other cell types are grouped together as "All remaining cell types".
+
+```{r, fig.height = 5, fig.width=5}
+plot_faceted_umap(all_info_df, singler_lumped, legend_title = "SingleR cell types") +
+  theme(strip.text = element_text(size = 8))
+```
+
+### Cluster assignments
+
+The below UMAP shows the cluster assignments output by `evaluate-clusters.sh` using the following parameters:
+
+- Leiden with modularity 
+- Nearest neighbors: `r params$cluster_nn`
+- Resolution: `r params$cluster_res`
+
+```{r, fig.height = 5, fig.width=5}
+plot_faceted_umap(all_info_df, cluster, legend_title = "Cluster")
+```
+
+### `AUCell` results
+
+The below plots show the AUC values determined by `AUCell` and output from `run-aucell-ews-signatures.sh`. 
+The first plot shows the individual AUC values on the UMAP.
+
+```{r}
+# get the individual thresholds determined by AUCell for each msigdb geneset
+# we want this so we can show the threshold on the density plots 
+auc_threshold_df <- all_info_df |> 
+  tidyr::pivot_longer(starts_with("threshold_auc_"), names_to = "geneset", values_to = "threshold") |> 
+  dplyr::mutate(
+    geneset = stringr::str_remove(geneset, "threshold_auc_")
+  ) |> 
+  dplyr::select(barcodes, geneset, threshold)
+
+# reformat auc data for density plots and UMAPs showing AUC values 
+auc_df <- all_info_df |> 
+  tidyr::pivot_longer(starts_with("auc_"), names_to = "geneset", values_to = "auc_value") |> 
+  dplyr::mutate(
+    geneset = stringr::str_remove(geneset, "auc_")
+  ) |> 
+  dplyr::select(barcodes, UMAP1, UMAP2, geneset, auc_value, cluster) |> 
+  dplyr::left_join(auc_threshold_df, by = c("barcodes", "geneset")) |> 
+  dplyr::mutate(
+    auc_assignment =  auc_value > threshold
+  )
+```
+
+```{r, fig.height=10, fig.width=10}
+expression_umap(auc_df, auc_value)
+```
+
+The below plot shows the distribution of AUC values for each gene set colored based on if the AUC value is above the threshold determined by `AUCell`, indicated with a dotted line.
+
+```{r, fig.height=5}
+
+ggplot(auc_df, aes(x = auc_value, color = auc_assignment, fill = auc_assignment)) +
+  geom_density(alpha = 0.5, bw = 0.01) +
+  facet_wrap(vars(geneset)) +
+  ggplot2::geom_vline(data = auc_df,
+                      mapping = aes(xintercept = threshold),
+                      lty = 2) +
+  theme(
+      aspect.ratio = 1,
+      strip.background = element_rect(fill = "transparent", linewidth = 0.5),
+      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
+    )
+
+```
+
+Here we look at the AUC values for each gene set across all clusters.
+
+```{r, fig.height=10, fig.width=5}
+auc_columns <- colnames(all_info_df)[which(startsWith(colnames(all_info_df), "auc_"))]
+cluster_density_plot(all_info_df, auc_columns, "AUC")
+```
+
+The heatmap below shows the AUC values for all cells and all gene sets. 
+The annotations shown are the `SingleR` cell types and clusters.
+
+```{r}
+annotated_exp_heatmap(
+  all_info_df, 
+  exp_columns = auc_columns, 
+  annotation_column = "singler_lumped", 
+  cluster_column = "cluster", 
+  legend_title = "AUC"
+)
+```
+
+### Mean expression of custom gene sets
+
+Below we look at the mean expression of all genes in each of the custom gene sets we have defined both on a UMAP and in a density plot.
+
+```{r, fig.height=10, fig.width=10}
+# format expression data for UMAP and density plot 
+exp_df <- all_info_df |> 
+  tidyr::pivot_longer(ends_with("_mean"), names_to = "geneset", values_to = "mean") |>
+  dplyr::mutate(tumor_cell_classification = dplyr::if_else(singler_lumped == "tumor", "tumor", "other"))
+
+expression_umap(exp_df, mean)
+```
+
+```{r}
+ggplot(exp_df, aes(x = mean, color = tumor_cell_classification, fill = tumor_cell_classification)) +
+  geom_density(alpha = 0.5, bw = 0.2) +
+  facet_wrap(vars(geneset)) +
+  theme(
+    aspect.ratio = 1,
+    strip.background = element_rect(fill = "transparent", linewidth = 0.5),
+    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
+  )
+```
+
+Here we look at the mean expression for each gene set across all clusters.
+
+```{r, fig.height=7}
+mean_exp_columns <- colnames(all_info_df)[which(endsWith(colnames(all_info_df), "_mean"))]
+cluster_density_plot(all_info_df, mean_exp_columns, "mean gene expression")
+```
+
+The heatmap below shows the mean expression values for all cells and all gene sets. 
+The annotations shown are the `SingleR` cell types and clusters.
+
+```{r}
+annotated_exp_heatmap(
+  all_info_df, 
+  exp_columns = mean_exp_columns, 
+  annotation_column = "singler_lumped", 
+  cluster_column = "cluster", 
+  legend_title = "Mean\ngene expression"
+)
+```
 

### 2025-01-28 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
update params

--- 2025-01-27 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
@@
 params:
-  predicted.celltype.threshold: 0.75
-  cnv_threshold: 0
+  predicted.celltype.threshold: 0.85
+  cnv_threshold_low: 0
+  cnv_threshold_high: 2
   testing: 0
@@
 We simply checked for each chromosome if the cell `has_cnv_chr`.
-Would the cell have less than `cnv_threshold` chromosome with CNV, the global `has_cnv_score` will be FALSE.
-Would the cell have more than `cnv_threshold` +2 chromosome with CNV, the global `has_cnv_score` will be TRUE.
+Would the cell have less than `r params$cnv_threshold_low` chromosome with CNV, the global `has_cnv_score` will be FALSE.
+Would the cell have more than `r params$cnv_threshold_high`  chromosome with CNV, the global `has_cnv_score` will be TRUE.
 The default is set to `unknown`.
@@
 The `infercnv` method can lead to false positive CNV.
-We introduced a flexibility of `+2` over the `cnv_threshold` to reduce the risk of misclassification of normal cells as cancer cells.
+We introduced a flexibility of +`r params$cnv_threshold_high` over the `cnv_threshold` to reduce the risk of misclassification of normal cells as cancer cells.
 
@@
   mutate(has_cnv_score = case_when(
-    has_cnv_score > params$cnv_threshold +2 ~ "CNV",
-    has_cnv_score <= params$cnv_threshold ~ "no CNV",
+    has_cnv_score > params$cnv_threshold_high ~ "CNV",
+    has_cnv_score <= params$cnv_threshold_low ~ "no CNV",
     .default = "unknown"
@@
 The threshold used for the `predicted.score` is defined as a parameter of this notebook as `r params$predicted.celltype.threshold`.
-The threshold used for the identification of CNV is also defined as the notebook parameter `r params$cnv_threshold`.
+The thresholds used for the identification of CNV are also defined as notebook parameters with a minimum of `r params$cnv_threshold_low` and a maximum of `r params$cnv_threshold_high`.
 

### 2025-01-28 analyses/cell-type-ewings/template_notebooks/celltype-exploration.Rmd
add some documentation for new functions

@@
 
-# source in validation functions calculate_mean_markers()
+# source in validation functions 
+# calculate_mean_markers(), plot_faceted_umap()
 validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
 source(validation_functions)
+
+# source in plotting functions 
+# expression_umap(), cluster_density_plot(), and annotated_exp_heatmap()
+plotting_functions <- file.path(module_base, "template_notebooks", "utils", "plotting-functions.R")
+source(plotting_functions)
 ```
@@
 
-```{r, fig.height = 5, fig.width=5}
+```{r, fig.height = 5}
 plot_faceted_umap(all_info_df, singler_lumped, legend_title = "SingleR cell types") +
@@
 
-```{r, fig.height = 5, fig.width=5}
+```{r, fig.height = 5}
 plot_faceted_umap(all_info_df, cluster, legend_title = "Cluster")
@@
 ```{r, fig.height=10, fig.width=10}
-expression_umap(auc_df, auc_value)
+expression_umap(auc_df, auc_value, geneset)
 ```
@@
 
-```{r, fig.height=10, fig.width=5}
+```{r, fig.height=10}
 auc_columns <- colnames(all_info_df)[which(startsWith(colnames(all_info_df), "auc_"))]
@@
   exp_columns = auc_columns, 
-  annotation_column = "singler_lumped", 
+  cell_type_column = "singler_lumped", 
   cluster_column = "cluster", 
@@
 
-expression_umap(exp_df, mean)
+expression_umap(exp_df, mean, geneset)
 ```
@@
   exp_columns = mean_exp_columns, 
-  annotation_column = "singler_lumped", 
+  cell_type_column = "singler_lumped", 
   cluster_column = "cluster", 

### 2025-01-28 analyses/cell-type-ewings/template_notebooks/utils/plotting-functions.R

@@ -2,16 +2,35 @@
 # They are used for creating the summary plots in the report 
 
 
-# create faceted UMAP coloring each cell by an expression value 
+# source in helper plotting functions 
+# plot_density() and plot_gene_heatmap()
+repository_base <- rprojroot::find_root(rprojroot::is_git_root)
+module_base <- file.path(repository_base, "analyses", "cell-type-ewings") 
+validation_helper_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
+source(validation_helper_functions)
+
+
+#' Plot expression for a set of genes or gene sets on the UMAP
+#' UMAP is faceted by gene or gene set and color corresponds to expression value (mean, sum, AUC, etc.)
+#'
+#' @param df Data frame with UMAP embeddings and gene set expression value
+#' @param color_column Column to use for coloring the points in the UMAP. 
+#'   This should correspond to an expression value, like counts, mean, sum, or AUC
+#' @param facet_column Column to use for faceting the UMAP
+#'   This should correspond to the column with a gene or gene set name
+#'
+#' @return UMAP faceted by facet_column and colored by color_column
+#'
 expression_umap <- function(
     df, 
-    color_column) {
-  
-  
+    color_column,
+    facet_column) {
+
   ggplot(df, aes(x = UMAP1, y = UMAP2, color = {{color_column}})) +
     geom_point(size = 0.1, alpha = 0.5) +
     scale_color_viridis_c() +
-    facet_wrap(vars(geneset)) +
+    facet_wrap(vars({{facet_column}})) +
+    # make sure there's a box around every facet 
     theme(
       aspect.ratio = 1,
       strip.background = element_rect(fill = "transparent", linewidth = 0.5),
@@ -20,12 +39,24 @@ expression_umap <- function(
   
 }
 
-# faceted density plots 
+
+#' Density plot showing expression across clusters
+#'
+#' @param df Data frame with clusters for each cell and expresion values to be plotted 
+#'   Must contain the `cluster` column 
+#' @param expression_columns Vector of columns present in the data frame that contain some
+#'   expression value to show on the x-axis of the density plot, such as AUC values or mean expression
+#' @param x_label label to use for labeling all x-axes 
+#'
+#' @return Faceted density plot with clusters as rows and each column from expression_columns
+#'   as a single panel 
+
 cluster_density_plot <- function(
[…]

### 2025-01-29 analyses/cell-type-ewings/template_notebooks/celltype-exploration.Rmd
adjust plots based on review comments

@@
 all_info_df <- all_results_df |> 
-  dplyr::left_join(gene_exp_df, by = "barcodes")
+  dplyr::left_join(gene_exp_df, by = "barcodes") |> 
+  dplyr::arrange(cluster)
 ```
@@
   dplyr::mutate(
-    auc_assignment =  auc_value > threshold
+    in_geneset =  auc_value > threshold
   )
@@
 
-ggplot(auc_df, aes(x = auc_value, color = auc_assignment, fill = auc_assignment)) +
+ggplot(auc_df, aes(x = auc_value, color = in_geneset, fill = in_geneset)) +
   geom_density(alpha = 0.5, bw = 0.01) +
@@
     ) 
-
 ```
@@
 
-```{r}
+```{r, fig.height=8}
 ggplot(exp_df, aes(x = mean, color = tumor_cell_classification, fill = tumor_cell_classification)) +

### 2025-01-29 analyses/cell-type-ewings/template_notebooks/utils/plotting-functions.R

@@ -35,7 +35,10 @@ expression_umap <- function(
       aspect.ratio = 1,
       strip.background = element_rect(fill = "transparent", linewidth = 0.5),
       panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
-    )
+    ) +
+    # remove axis numbers and background grid
+    scale_x_continuous(labels = NULL, breaks = NULL) +
+    scale_y_continuous(labels = NULL, breaks = NULL)
   
 }
 
@@ -131,6 +134,6 @@ annotated_exp_heatmap <- function(
                     row_title = "",
                     legend_title = legend_title,
                     annotation = annotation,
-                    cluster_columns = TRUE
+                    cluster_columns = FALSE
   )
 }

### 2025-01-31 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd
explore pdx vs patient tissue in consensus results

@@
   summary_df <- df |> 
-    dplyr::group_by(library_id, consensus_annotation, consensus_ontology) |> 
+    dplyr::group_by(library_id, sample_type, consensus_annotation, consensus_ontology) |> 
     dplyr::summarize(total_cells_per_annotation = length(consensus_annotation)) |>
@@
 
+Let's also look at the projects that have PDX samples. 
+I expect that fewer cells are able to be classified in PDX samples compared to their patient tissue counterparts.
+
+```{r, fig.height=7}
+# list of projects with pdx 
+pdx_projects <- c(
+  "SCPCP000003",
+  "SCPCP000004",
+  "SCPCP000005",
+  "SCPCP000011",
+  "SCPCP000013",
+  "SCPCP000014",
+  "SCPCP000015",
+  "SCPCP000016",
+  "SCPCP000023"
+)
+
+# filter to pdx projects and adjust the facet label 
+pdx_only <- unknown_only |>
+  dplyr::filter(project_id %in% pdx_projects) |>
+  dplyr::mutate(facet_label = glue::glue("{project_id}\n{diagnosis}"))
+
+# faceted sina showing patient tissue vs. pdx 
+ggplot(pdx_only, aes(x = sample_type, y = percent_cells_annotation)) +
+  ggforce::geom_sina() +
+  stat_summary(fun.y=median, geom="crossbar" , color = "red", linewidth = 0.2) +
+  facet_wrap(vars(facet_label)) +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
+        plot.margin = margin(10,10,10,10)) +
+  labs(
+    x = "", 
+    y = "Percent of cells annotated as Unknown"
+  )
+```
+
+It looks like in `SCPCP000003` and `SCPCP000004` the PDX samples have less cells classified than the patient tissue samples.
+
+For `SCPCP000015` there may be a higher percentage of cells that are classified in a few of the PDX samples. 
+For `SCPCP000023` and `SCPCP000013` there appears to be a handful of PDX and patient tissue samples that have cells that are classified with no real difference between the two sample types. 
+My hunch is that the cells getting classified in `SCPCP000013`, `SCPCP000015`, and `SCPCP000023` PDX samples are muscle cells which are very similar to the tumor cells.
+
+For the remaining projects, there are very few samples that have any cells classified regardless of sample type.
+
 ## Number of cell types observed
@@
       geom_col() + 
+      # split samples based on sample type, patient tissue or pdx 
+      facet_wrap(vars(sample_type), scales ="free") +
       scale_y_continuous(expand = c(0,0)) +
@@
 - Projects that I would expect to be more difficult to classify (sarcomas, wilms, RB) have fewer cells classified then things like brain and leukemia. 
-Notably many of the solid tumor projects (4, 5, 12-16, and 23) have a handful of PDX samples where I would expect to see fewer normal cells.
+- Generally there are very few cells classified in the PDX samples, and if they are classified they tend to be muscle cell types, fibroblasts, and in a handful of cases, endothelial cells. 
+Immune cell types appear to be absent. 
+Notably the projects that have PDX samples are solid tumors and sarcoma projects, where muscle cells or fibroblasts can be similar to tumor cells. 
+- `SCPCP000023` seems to be an exception where a good amount of PDX samples have a high percentage of classified cells, although they are either muscle cells or chondrocytes. 
+My guess is that these cells look very similar to tumor cells, which is why they are being classified.
+
+It's important to note that when we quantify gene expression using `alevin-fry` in `scpca-nf` we only use a human reference, so I would not expect non-malignant cells to appear in the PDX samples at any high frequency.
+Because of this, I think it may be safest to remove these samples when conducting any further analysis.
 

### 2025-02-03 analyses/cell-type-ewings/exploratory_analysis/07-identify-max-rank-aucell.Rmd
address spelling errors

@@
 
-> It is important to check that most cells have at least the number of expressed/detected genes that are going to be used to calculate the AUC (aucMaxRank in calcAUC()).
-The histogram provided by AUCell_buildRankings() allows to quickly check this distribution.
+> It is important to check that most cells have at least the number of expressed/detected genes that are going to be used to calculate the AUC (`aucMaxRank` in `calcAUC`()).
+The histogram provided by `AUCell_buildRankings`() allows to quickly check this distribution.
 
@@
     genes_to_remove <- rowData(sce)$detected > 0 
-    filtered_sce <- sce[genes_to_remove > 0 , ]
+    filtered_sce <- sce[genes_to_remove , ]
 
@@
 Remember, we want to pick a number of genes that are detected in most cells.
-If we picked a max rank higher than the number of genes detected in most cells, the non-detected genes that are randomly ordered would play an outsized role in our AUC values.
+If we picked a max rank higher than the number of genes detected in most cells, the non-detected genes that are randomly ordered would play an oversized role in our AUC values.
 

### 2025-02-03 analyses/cell-type-ewings/exploratory_analysis/07-identify-max-rank-aucell.Rmd
keep outsized

@@
 Remember, we want to pick a number of genes that are detected in most cells.
-If we picked a max rank higher than the number of genes detected in most cells, the non-detected genes that are randomly ordered would play an oversized role in our AUC values.
+If we picked a max rank higher than the number of genes detected in most cells, the non-detected genes that are randomly ordered would play an outsized role in our AUC values.
 

### 2025-02-03 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd

--- 2025-01-24 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd
@@
 For example, leukemia libraries have T and B cells, brain tumors have macrophages, and solid tumors have fibroblasts and muscle cells. 
-- Projects that I would expect to be more difficult to classify (sarcomas, wilms, RB) have fewer cells classified then things like brain and leukemia. 
+- Projects that I would expect to be more difficult to classify (sarcomas, Wilms, RB) have fewer cells classified then things like brain and leukemia. 
 Notably many of the solid tumor projects (4, 5, 12-16, and 23) have a handful of PDX samples where I would expect to see fewer normal cells.

### 2025-02-03 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
notebook for looking at merged results

--- 2025-01-29 analyses/cell-type-ewings/template_notebooks/celltype-exploration.Rmd
@@
 ---
-title: "Template notebook for validating cell type assignments for an individual library in SCPCP000015"
+title: "Cell type assignments for SCPCP000015"
 author: Ally Hawkins
@@
     code_folding: "hide"
-params:
-  sample_id: "SCPCS000490"
-  library_id: "SCPCL000822" 
-  cluster_nn: 20
-  cluster_res: 0.5
 ---
-
-This notebook is meant to be a guide for compiling "final" cell type annotations for an individual library in `SCPCP000015`. 
-Results from `aucell-singler-annotation.sh`, `evaulate-clusters.sh`, and `run-aucell-ews-signatures.sh` are all combined and used to assign cell type annotations on a case by case basis.
-
-Instructions for using this guide:
-
-1. Ensure that you have a local copy of the results from `aucell-singler-annotation.sh`, `evaluate-clusters.sh` and `run-aucell-ews-signatures.sh` saved to `results`. 
-2. Copy the contents of this notebook to a new notebook titled `<library_id>_celltype-exploration.Rmd` and save in `exploratory_analysis/final_annotation_notebooks`. 
-3. Update the `title` in the `yaml` section and replace the `sample_id` and `library_id` with the correct IDs in the `params` list. 
-4. Optionally, you may choose to update the choices for clustering based on the results from `evaluate-clusters.sh`. 
-All clusters used will be calculated with the Leiden algorithm and the modularity objective function. 
-To modify the nearest neighbors (default: 20) and resolution (default: 0.5) chosen use the `cluster_nn` and `cluster_res` params. 
-5. Run through the notebook and update any sections of the notebook marked with the `{.manual-exploration}` tag. 
-6. Render the completed notebook which will produce the rendered `html` file and a TSV with cell type annotations for that library.
 
@@
 # The current data directory, found within the repository base directory
-data_dir <- file.path(repository_base, "data", "current", "SCPCP000015")
+data_dir <- file.path(repository_base, "data", "current", "results", "merge-sce", "SCPCP000015")
 
@@
 # path to sce 
-sce_file <- file.path(data_dir, params$sample_id, glue::glue("{params$library_id}_processed.rds"))
+sce_file <- file.path(data_dir, "SCPCP000015_merged.rds")
 
@@
 
-singler_results_dir <- file.path(workflow_results_dir, "aucell_singler_annotation", params$sample_id)
-singler_results_file <- file.path(singler_results_dir,
-                                  glue::glue("{params$library_id}_singler-classifications.tsv"))
+singler_results_dir <- file.path(workflow_results_dir, "aucell_singler_annotation")
+singler_results_files <- list.files(singler_results_dir, pattern = "*singler-classifications.tsv", full.names = TRUE, recursive = TRUE)
+library_ids <- stringr::str_remove(basename(singler_results_files), "_singler-classifications.tsv")
 
-cluster_results_dir <- file.path(workflow_results_dir, "clustering", params$sample_id)
-cluster_results_file <- file.path(cluster_results_dir,
-                                  glue::glue("{params$library_id}_cluster-results.tsv"))
-
-aucell_results_dir <- file.path(workflow_results_dir, "aucell-ews-signatures", params$sample_id)
-aucell_results_file <- file.path(aucell_results_dir,
-                                 glue::glue("{params$library_id}_auc-ews-gene-signatures.tsv"))
+aucell_results_dir <- file.path(workflow_results_dir, "aucell-ews-signatures")
+aucell_results_file <- file.path(aucell_results_dir, "SCPCP000015_auc-ews-gene-signatures.tsv")
 
@@
 results_dir <- file.path(module_base, "results", "final-annotations")
-output_file <- file.path(results_dir, glue::glue("{params$library_id}_celltype-annotations.tsv"))
+output_file <- file.path(results_dir, glue::glue("SCPCP000015_celltype-annotations.tsv"))
 ```
@@
   "sce file does not exist" = file.exists(sce_file),
-  "singler results file does not exist" = file.exists(singler_results_file),
-  "cluster results file does not exist" = file.exists(cluster_results_file),
   "aucell results file does not exist" = file.exists(aucell_results_file)
@@
 # read in workflow results
-singler_df <- readr::read_tsv(singler_results_file)
-cluster_df <- readr::read_tsv(cluster_results_file)
-aucell_df <- readr::read_tsv(aucell_results_file)
+singler_df <- singler_results_files |> 
+  purrr::set_names(library_ids) |>
+  purrr::map(readr::read_tsv) |> 
+  dplyr::bind_rows(.id = "library_id")
+aucell_df <- readr::read_tsv(aucell_results_file) |> 
+  tidyr::separate(barcodes, "-", into = c("library_id", "barcodes"))
 
@@
   sce, 
-  singler_df, 
-  cluster_df, 
-  aucell_df,
+  singler_df = singler_df, 
+  cluster_df = NULL, 
+  aucell_df = aucell_df,
   cluster_nn = params$cluster_nn,
-  cluster_res = params$cluster_res
-  )
+  cluster_res = params$cluster_res,
+  join_columns = c("barcodes", "library_id")
+  ) |>
+  dplyr::mutate(barcodes = glue::glue("{library_id}-{barcodes}"))
 
@@
 all_info_df <- all_results_df |> 
-  dplyr::left_join(gene_exp_df, by = "barcodes") |> 
-  dplyr::arrange(cluster)
+  dplyr::left_join(gene_exp_df, by = "barcodes")
 ```
@@
   ) |> 
-  dplyr::select(barcodes, UMAP1, UMAP2, geneset, auc_value, cluster) |> 
+  dplyr::select(barcodes, UMAP1, UMAP2, geneset, auc_value) |> 
   dplyr::left_join(auc_threshold_df, by = c("barcodes", "geneset")) |> 
@@
 ```{r}
+all_info_df <- all_info_df |> 
+  dplyr::arrange(singler_lumped)
+head(all_info_df)
+
 annotated_exp_heatmap(
@@
   cell_type_column = "singler_lumped", 
-  cluster_column = "cluster", 
+  cluster_column = "singler_lumped", 
   legend_title = "AUC"
@@
 ggplot(exp_df, aes(x = mean, color = tumor_cell_classification, fill = tumor_cell_classification)) +
-  geom_density(alpha = 0.5, bw = 0.2) +
+  geom_density(alpha = 0.5, bw = 0.1) +
   facet_wrap(vars(geneset)) +
@@
   cell_type_column = "singler_lumped", 
-  cluster_column = "cluster", 
+  cluster_column = "singler_lumped", 
   legend_title = "Mean\ngene expression"

### 2025-02-03 analyses/cell-type-ewings/template_notebooks/utils/plotting-functions.R

@@ -28,7 +28,7 @@ expression_umap <- function(
 
   ggplot(df, aes(x = UMAP1, y = UMAP2, color = {{color_column}})) +
     geom_point(size = 0.1, alpha = 0.5) +
-    scale_color_viridis_c() +
+    scale_color_viridis_c(option = "turbo") +
     facet_wrap(vars({{facet_column}})) +
     # make sure there's a box around every facet 
     theme(

### 2025-02-03 analyses/cell-type-ewings/template_notebooks/utils/setup-functions.R

@@ -14,73 +14,86 @@
 #'
 prep_results <- function(
     sce,
-    singler_df,
-    cluster_df,
-    aucell_df,
+    singler_df = NULL,
+    cluster_df = NULL,
+    aucell_df = NULL,
     cluster_nn = 20,
-    cluster_res = 0.5
+    cluster_res = 0.5,
+    join_columns = c("barcodes")
 ) {
   
   ## grab UMAP 
   umap_df <- sce |>
     scuttle::makePerCellDF(use.dimred = "UMAP") |>
     # replace UMAP.1 with UMAP1 and get rid of excess columns
-    dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2)
+    dplyr::select(all_of(join_columns), UMAP1 = UMAP.1, UMAP2 = UMAP.2)
   
-  ## prep singler data
-  singler_df <- singler_df |> 
-    dplyr::mutate(
-      # first grab anything that is tumor and label it tumor
-      # NA should be unknown
-      singler_annotation = dplyr::case_when(
-        stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
-        is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
-        .default = singler_annotation
+  
+  if(!is.null(cluster_df)){
+    ## prep cluster data 
+    cluster_df <- cluster_df |> 
+      # filter to the clustering results we want to use 
+      dplyr::filter(
+        cluster_method == "leiden_mod",
+        nn == cluster_nn,
+        resolution == cluster_res
+      ) |> 
+      dplyr::select(
+        barcodes = cell_id,
+        cluster
       ) |>
-        forcats::fct_relevel("tumor", after = 0),
-      # get the top cell types for plotting later
-      singler_lumped = singler_annotation |>
-        forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
-        forcats::fct_infreq() |>
-        forcats::fct_relevel("All remaining cell types", after = Inf)
-    )
+      dplyr::mutate(cluster = as.factor(cluster))
+    
+    umap_df <- umap_df |> 
+      dplyr::left_join(cluster_df, by = join_columns)
+  }
   
-  ## prep cluster data 
-  cluster_df <- cluster_df |> 
-    # filter to the clustering results we want to use 
-    dplyr::filter(
-      cluster_method == "leiden_mod",
-      nn == cluster_nn,
-      resolution == cluster_res
-    ) |> 
[…]

### 2025-02-03 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd

--- 2025-01-31 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd
@@
 For example, leukemia libraries have T and B cells, brain tumors have macrophages, and solid tumors have fibroblasts and muscle cells. 
-- Projects that I would expect to be more difficult to classify (sarcomas, wilms, RB) have fewer cells classified then things like brain and leukemia. 
+- Projects that I would expect to be more difficult to classify (sarcomas, Wilms, RB) have fewer cells classified then things like brain and leukemia. 
 - Generally there are very few cells classified in the PDX samples, and if they are classified they tend to be muscle cell types, fibroblasts, and in a handful of cases, endothelial cells. 

### 2025-02-03 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
initial addition of notebook annotating merged object

@@
   aucell_df = aucell_df,
+  consensus_df = consensus_df,
   cluster_nn = params$cluster_nn,
@@
 
-### Cluster assignments
+### Consensus cell type assignments
 
-The below UMAP shows the cluster assignments output by `evaluate-clusters.sh` using the following parameters:
-
-- Leiden with modularity 
-- Nearest neighbors: `r params$cluster_nn`
-- Resolution: `r params$cluster_res`
-
-```{r, fig.height = 5}
-plot_faceted_umap(all_info_df, cluster, legend_title = "Cluster")
+```{r, fig.height=5}
+plot_faceted_umap(all_info_df, consensus_lumped, legend_title = "Consensus cell types") +
+  theme(strip.text = element_text(size = 8))
 ```
@@
 auc_columns <- colnames(all_info_df)[which(startsWith(colnames(all_info_df), "auc_"))]
-cluster_density_plot(all_info_df, auc_columns, "AUC")
+cluster_density_plot(all_info_df, auc_columns, "consensus_lumped", "AUC")
+```
+
+Here we should identify any tumor cells as those that have high expression of `EWS-FLI` targets and low expression of `EWS-FLI` repressed targets. 
+I think we want to use the gene sets where we see some separation in AUC values between cell types to do our assignment.
+
+- `aynaud-ews-targets`: > 0.05
+- `staege`: > 0.25
+- `riggi_down`: < 0.025
+- `migayawa_down`: < 0.03
+
+Intriguingly we see high expression of `hallmark_EMT` and the down gene sets (genes that are repressed by `EWS-FLI`) in fibroblasts, which also happen to have high expression of the `EWS-FLI` low targets identifed by Wrenn et al.
+That paper identifies `EWS-FLI` low cells as cancer associated fibroblasts.
+Let's label these cells accordingly. 
+If cells have `in_geneset=TRUE` for the `hallmark_EMT`, `wrenn-nt5e-genes`, and `migayawa_down` they will be classified as `ews_low_cafs`.
+
+- `wrenn-nt5e-genes`: > 0.1
+- `hallmark_EMT`: > 0.05
+- `migayawa_down`: > 0.03
+- `riggi_down`: > 0.025
+
+```{r}
+all_info_df <- all_info_df |> 
+  dplyr::mutate(
+    final_annotation = dplyr::case_when(
+      `auc_wrenn-nt5e-genes` > 0.1 & auc_miyagawa_down > 0.03 ~ "ews_low_caf",
+      `auc_aynaud-ews-targets` > 0.025 | 
+        consensus_annotation %in% c("Unknown", "muscle cell", "smooth muscle cell") ~ "ews_high",
+      .default = consensus_annotation
+    )
+  )
+
+head(all_info_df)
+table(all_info_df$final_annotation)
 ```
@@
 all_info_df <- all_info_df |> 
-  dplyr::arrange(singler_lumped)
-head(all_info_df)
+  dplyr::arrange(final_annotation)
 
@@
   exp_columns = auc_columns, 
-  cell_type_column = "singler_lumped", 
-  cluster_column = "singler_lumped", 
+  cell_type_column = "consensus_lumped", 
+  cluster_column = "final_annotation", 
   legend_title = "AUC"
+)
+
+# get annotation colors for cell type and cluster 
+cell_types <- unique(all_info_df$final_annotation)
+num_cell_types <- length(cell_types)
+cell_type_colors <- palette.colors(palette = "alphabet") |>
+  head(n = num_cell_types) |>
+  purrr::set_names(cell_types)
+
+# create annotation for heatmap
+annotation <- ComplexHeatmap::columnAnnotation(
+  cell_type = all_info_df$final_annotation,
+  col = list(
+    cell_type = cell_type_colors
+  )
+)
+
+# build matrix for heatmap cells x gene set/ expression columns 
+heatmap_mtx <- all_info_df |>
+  dplyr::select(barcodes, all_of(auc_columns)) |>
+  tibble::column_to_rownames("barcodes") |>
+  as.matrix() |>
+  t()
+
+# remove any extra strings from the expression column names 
+rownames(heatmap_mtx) <- stringr::str_remove(rownames(heatmap_mtx), "auc_|_mean")
+
+# plot heatmap of marker genes
+plot_gene_heatmap(heatmap_mtx,
+                  row_title = "",
+                  legend_title = "final_annotation",
+                  annotation = annotation,
+                  cluster_columns = TRUE
 )
@@
 ggplot(exp_df, aes(x = mean, color = tumor_cell_classification, fill = tumor_cell_classification)) +
-  geom_density(alpha = 0.5, bw = 0.1) +
+  geom_density(alpha = 0.5, bw = 0.5) +
   facet_wrap(vars(geneset)) +
@@
 
-```{r, fig.height=7}
+```{r, fig.height=30, warning=FALSE}
 mean_exp_columns <- colnames(all_info_df)[which(endsWith(colnames(all_info_df), "_mean"))]
-cluster_density_plot(all_info_df, mean_exp_columns, "mean gene expression")
+cluster_density_plot(all_info_df, mean_exp_columns, annotation_column = "consensus_annotation", "mean gene expression")
 ```

### 2025-02-03 analyses/cell-type-ewings/template_notebooks/utils/plotting-functions.R

@@ -28,7 +28,7 @@ expression_umap <- function(
 
   ggplot(df, aes(x = UMAP1, y = UMAP2, color = {{color_column}})) +
     geom_point(size = 0.1, alpha = 0.5) +
-    scale_color_viridis_c() +
+    scale_color_viridis_c(option = "turbo") +
     facet_wrap(vars({{facet_column}})) +
     # make sure there's a box around every facet 
     theme(
@@ -57,6 +57,7 @@ expression_umap <- function(
 cluster_density_plot <- function(
     df,
     expression_columns,
+    annotation_column = "cluster",
     x_label){
   
   # create density plot for each column and combine into one figure 
@@ -65,10 +66,9 @@ cluster_density_plot <- function(
       plot_density(
         df,
         column,
-        "cluster"
+        annotation_column
       ) +
-        labs(y = "Cluster", 
-             x = x_label) +
+        labs(x = x_label) +
         theme(text = element_text(size = 8))
     }) |>
     patchwork::wrap_plots(ncol = 2)

### 2025-02-03 analyses/cell-type-ewings/template_notebooks/utils/setup-functions.R

@@ -14,73 +14,102 @@
 #'
 prep_results <- function(
     sce,
-    singler_df,
-    cluster_df,
-    aucell_df,
+    singler_df = NULL,
+    cluster_df = NULL,
+    aucell_df = NULL,
+    consensus_df = NULL,
     cluster_nn = 20,
-    cluster_res = 0.5
+    cluster_res = 0.5,
+    join_columns = c("barcodes")
 ) {
   
   ## grab UMAP 
   umap_df <- sce |>
     scuttle::makePerCellDF(use.dimred = "UMAP") |>
     # replace UMAP.1 with UMAP1 and get rid of excess columns
-    dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2)
+    dplyr::select(all_of(join_columns), UMAP1 = UMAP.1, UMAP2 = UMAP.2)
   
-  ## prep singler data
-  singler_df <- singler_df |> 
-    dplyr::mutate(
-      # first grab anything that is tumor and label it tumor
-      # NA should be unknown
-      singler_annotation = dplyr::case_when(
-        stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
-        is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
-        .default = singler_annotation
+  
+  if(!is.null(cluster_df)){
+    ## prep cluster data 
+    cluster_df <- cluster_df |> 
+      # filter to the clustering results we want to use 
+      dplyr::filter(
+        cluster_method == "leiden_mod",
+        nn == cluster_nn,
+        resolution == cluster_res
+      ) |> 
+      dplyr::select(
+        barcodes = cell_id,
+        cluster
       ) |>
-        forcats::fct_relevel("tumor", after = 0),
-      # get the top cell types for plotting later
-      singler_lumped = singler_annotation |>
-        forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
-        forcats::fct_infreq() |>
-        forcats::fct_relevel("All remaining cell types", after = Inf)
-    )
+      dplyr::mutate(cluster = as.factor(cluster))
+    
+    umap_df <- umap_df |> 
+      dplyr::left_join(cluster_df, by = join_columns)
+  }
   
-  ## prep cluster data 
-  cluster_df <- cluster_df |> 
-    # filter to the clustering results we want to use 
-    dplyr::filter(
-      cluster_method == "leiden_mod",
[…]

### 2025-02-04 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
incorporate consensus cell types

@@
       .default = consensus_annotation
-    )
+    ),
+    final_lumped = final_annotation |>
+          forcats::fct_lump_n(8, other_level = "All remaining cell types", ties.method = "first") |>
+          forcats::fct_infreq() |>
+          forcats::fct_relevel("All remaining cell types", after = Inf)
   )
@@
   cell_type_column = "consensus_lumped", 
-  cluster_column = "final_annotation", 
+  cluster_column = "final_lumped", 
   legend_title = "AUC"
-)
-
-# get annotation colors for cell type and cluster 
-cell_types <- unique(all_info_df$final_annotation)
-num_cell_types <- length(cell_types)
-cell_type_colors <- palette.colors(palette = "alphabet") |>
-  head(n = num_cell_types) |>
-  purrr::set_names(cell_types)
-
-# create annotation for heatmap
-annotation <- ComplexHeatmap::columnAnnotation(
-  cell_type = all_info_df$final_annotation,
-  col = list(
-    cell_type = cell_type_colors
-  )
-)
-
-# build matrix for heatmap cells x gene set/ expression columns 
-heatmap_mtx <- all_info_df |>
-  dplyr::select(barcodes, all_of(auc_columns)) |>
-  tibble::column_to_rownames("barcodes") |>
-  as.matrix() |>
-  t()
-
-# remove any extra strings from the expression column names 
-rownames(heatmap_mtx) <- stringr::str_remove(rownames(heatmap_mtx), "auc_|_mean")
-
-# plot heatmap of marker genes
-plot_gene_heatmap(heatmap_mtx,
-                  row_title = "",
-                  legend_title = "final_annotation",
-                  annotation = annotation,
-                  cluster_columns = TRUE
 )
@@
 
-```{r, fig.height=8}
-ggplot(exp_df, aes(x = mean, color = tumor_cell_classification, fill = tumor_cell_classification)) +
-  geom_density(alpha = 0.5, bw = 0.5) +
-  facet_wrap(vars(geneset)) +
-  theme(
-    aspect.ratio = 1,
-    strip.background = element_rect(fill = "transparent", linewidth = 0.5),
-    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
-  )
+Here we look at the mean expression for each gene set across all clusters.
+
+```{r, fig.height=7, warning=FALSE}
+mean_exp_columns <- colnames(all_info_df)[which(endsWith(colnames(all_info_df), "_mean"))]
+cluster_density_plot(all_info_df, mean_exp_columns, annotation_column = "final_lumped", "mean gene expression")
 ```
 
-Here we look at the mean expression for each gene set across all clusters.
+What are the other cell types not present here? 
+```{r}
+lumped_celltypes <- unique(all_info_df$final_lumped)
+all_celltypes <- unique(all_info_df$final_annotation)
+missing_celltypes <- setdiff(all_celltypes, lumped_celltypes)
 
-```{r, fig.height=30, warning=FALSE}
-mean_exp_columns <- colnames(all_info_df)[which(endsWith(colnames(all_info_df), "_mean"))]
-cluster_density_plot(all_info_df, mean_exp_columns, annotation_column = "consensus_annotation", "mean gene expression")
+missing_only <- all_info_df |> 
+  dplyr::filter(final_annotation %in% missing_celltypes)
+table(missing_only$final_annotation)
 ```
 
-The heatmap below shows the mean expression values for all cells and all gene sets. 
-The annotations shown are the `SingleR` cell types and clusters.
+I think we can leave these few cells alone and just lump them as "All remaining cell types" for plots as we have been doing.
+
+Now let's make a dot plot! Or a heatmap only looking at key genes! 
+In these plots, lets group all T cells together.
 
 ```{r}
-annotated_exp_heatmap(
-  all_info_df, 
-  exp_columns = mean_exp_columns, 
-  cell_type_column = "singler_lumped", 
-  cluster_column = "singler_lumped", 
-  legend_title = "Mean\ngene expression"
+markers_df <- readr::read_tsv(file.path(module_base, "references", "combined-markers.tsv"))
+genes <- markers_df |>
+  dplyr::pull(ensembl_gene_id)
+
+gene_cts <- logcounts(sce[genes, ]) |>
+  as.matrix() |>
+  t() |> 
+  as.data.frame()
+gene_cts$barcodes <- rownames(gene_cts)
+
+rowData(sce)[genes, ] |> 
+  as.data.frame() |> 
+  dplyr::select(ends_with(".mean")) |> 
+  tibble::rownames_to_column("ensembl_gene_id") |> 
+  tidyr::pivot_longer(!ensembl_gene_id, names_to = "library_id", values_to = "mean")
+
+celltypes_df <- endsWith()celltypes_df <- all_info_df |> 
+  dplyr::select(barcodes, final_lumped)
+
+genes_df <- gene_cts |> 
+  tidyr::pivot_longer(!barcodes, names_to = "ensembl_gene_id", values_to = "gene_exp") |> 
+  dplyr::left_join(celltypes_df, by = c("barcodes")) |> 
+  dplyr::mutate(
+    percent_expression = 
   )
+
+head(genes_df)
 ```
-
-## Re-cluster tumor cells {.manual-exploration}
-
-<!-- Use this section to label tumor cells based on the above findings. 
-Any cells that are labeled as tumor will then be re-clustered and plots showing only tumor cells
-can be created to identify tumor cell states -->
-
-TODO: Functions for re-clustering tumor cells
-Show the same plots across the tumor only clusters and assign tumor cell states to each cluster
-
-## Additional exploration {.manual-exploration}
-
-<!--This section is for any additional exploration that may be needed to finalize annotations for this library. 
-If not using, please delete this section.
-For example, here you may want to dive into the normal cell types and make adjustments as needed
--->
-
-## Validate final tumor and normal annotation {.manual-exploration}
-
-<!-- This section should be used to update the assignments -->
-
-TODO: Insert plots that will be useful for validation (UMAPs, heatmaps, density plots)
-
-## Prepare annotations {.manual-exploration}
-
-<!-- This section should be used to create the final.final table with cell type annotations for export -->
-
-TODO: Code and instructions for exporting such as what columns should be named.
 

### 2025-02-05 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
dotplot and heatmap

@@
 output:
-  html_document:
+  html_notebook:
     toc: true
@@
 ---
+
+This notebook looks at cell type annotations and gene set expression across all samples in `SCPCP00015` and assigns "final" annotations. 
+To do this we are using the merged, but not batch-corrected, object containing the gene expression data for all samples.
+
+The following input is used:
+
+- Annotations obtained by running `SingleR` with tumor cells as a reference output by `aucell-singler-annotation.sh`. 
+- Consensus cell type annotations output from the `cell-type-consensus` module. 
+- AUC values as calculated by `AUCell` for a set of Ewing sarcoma specific gene sets in MSigDB output by `run-aucell-ews-signatures.sh`.
 
@@
 
+consensus_results_dir <- file.path(repository_base, "analyses", "cell-type-consensus", "results", "cell-type-consensus", "SCPCP000015")
+consensus_results_files <- list.files(consensus_results_dir, pattern = "*_consensus-cell-type-assignments.tsv.gz", full.names = TRUE, recursive = TRUE)
+
 # small gene sets
@@
 cell_state_genes_file <- file.path(module_base, "references", "tumor-cell-state-markers.tsv")
+
+# marker genes to be used for validating assignments 
+validation_markers_file <- file.path(module_base, "references", "combined-markers.tsv")
 ```
@@
 results_dir <- file.path(module_base, "results", "final-annotations")
-output_file <- file.path(results_dir, glue::glue("SCPCP000015_celltype-annotations.tsv"))
+fs::dir_create(results_dir)
+output_file <- file.path(results_dir, glue::glue("SCPCP000015_celltype-annotations.tsv.gz"))
 ```
@@
   dplyr::bind_rows(.id = "library_id")
+
 aucell_df <- readr::read_tsv(aucell_results_file) |> 
@@
 
+# consensus cell types 
+consensus_df <- consensus_results_files |> 
+  purrr::map(readr::read_tsv) |> 
+  dplyr::bind_rows()
+
 # read in marker genes and combine into one list 
 visser_markers_df <- readr::read_tsv(visser_marker_genes_file) |> 
-  dplyr::select(cell_type, ensembl_gene_id) |> 
+  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
   unique()
@@
 cell_state_markers_df <- readr::read_tsv(cell_state_genes_file) |> 
-  dplyr::select(cell_type = cell_state, ensembl_gene_id)
+  dplyr::select(cell_type = cell_state, ensembl_gene_id, gene_symbol)
 
@@
 
-## Summary of workflow results
+## Results summary
+
+First, we will just summarize the results from the various inputs and workflows. 
+Then we will look at those results and assign any tumor cells and refine annotations as needed.
 
@@
 
+The below UMAP shows the top cell types for which consensus cell types were assigned.
+Consensus cell types are observed when cell types from `SingleR` and `CellAssign` share a common ancestor. 
+If no consensus is found, the cells are labeled with "Unknown". 
+We anticipate that in many cases, the "Unknown" cells will be tumor cells.
+The top 7 cell types are shown and all other cell types are grouped together as "All remaining cell types".
+
 ```{r, fig.height=5}
@@
 The first plot shows the individual AUC values on the UMAP.
+
+For context:
+
+- Any gene sets labeled with `up` represent `EWS-FLI1` target genes that we expect to be upregulated in `EWS-FLI1` high tumor cells. 
+- Any gene sets labeled with `down` represent `EWS-FLI1` repressed genes that we expect to be downregulated in `EWS-FLI1` high tumor cells and upregulated in `EWS-FLI1` low tumor cells. 
+- The `aynaud-ews-targets` is a list of genes found to be upregulated in `EWS-FLI1` high tumor cells. 
+- The `wrenn-nt5e-genes` is a list of genes found to be upregulated in `EWS-FLI1` low tumor cells/ cancer associated fibroblasts.
+- The `gobp_ECM` and `hallmark_EMT` are gene sets that are hypothesized to be upregulated in `EWS-FLI1` low tumor cells/ cancer associated fibroblasts.
 
@@
 
-Here we look at the AUC values for each gene set across all clusters.
+Here we look at the AUC values for each gene set across all consensus cell types.
 
@@
 
-Here we should identify any tumor cells as those that have high expression of `EWS-FLI` targets and low expression of `EWS-FLI` repressed targets. 
-I think we want to use the gene sets where we see some separation in AUC values between cell types to do our assignment.
+Here we look at the AUC values for each gene set across all `SingleR` cell types.
 
-- `aynaud-ews-targets`: > 0.05
-- `staege`: > 0.25
-- `riggi_down`: < 0.025
-- `migayawa_down`: < 0.03
+```{r, fig.height=10}
+cluster_density_plot(all_info_df, auc_columns, "singler_lumped", "AUC")
+```
 
-Intriguingly we see high expression of `hallmark_EMT` and the down gene sets (genes that are repressed by `EWS-FLI`) in fibroblasts, which also happen to have high expression of the `EWS-FLI` low targets identifed by Wrenn et al.
-That paper identifies `EWS-FLI` low cells as cancer associated fibroblasts.
-Let's label these cells accordingly. 
-If cells have `in_geneset=TRUE` for the `hallmark_EMT`, `wrenn-nt5e-genes`, and `migayawa_down` they will be classified as `ews_low_cafs`.
-
-- `wrenn-nt5e-genes`: > 0.1
-- `hallmark_EMT`: > 0.05
-- `migayawa_down`: > 0.03
-- `riggi_down`: > 0.025
+The heatmap below shows the AUC values for all cells and all gene sets and the final refined cell type annotations.
 
 ```{r}
+# sort by final cell types 
 all_info_df <- all_info_df |> 
-  dplyr::mutate(
-    final_annotation = dplyr::case_when(
-      `auc_wrenn-nt5e-genes` > 0.1 & auc_miyagawa_down > 0.03 ~ "ews_low_caf",
-      `auc_aynaud-ews-targets` > 0.025 | 
-        consensus_annotation %in% c("Unknown", "muscle cell", "smooth muscle cell") ~ "ews_high",
-      .default = consensus_annotation
-    ),
-    final_lumped = final_annotation |>
-          forcats::fct_lump_n(8, other_level = "All remaining cell types", ties.method = "first") |>
-          forcats::fct_infreq() |>
-          forcats::fct_relevel("All remaining cell types", after = Inf)
-  )
+  dplyr::arrange(consensus_lumped)
 
-head(all_info_df)
-table(all_info_df$final_annotation)
-```
-
-The heatmap below shows the AUC values for all cells and all gene sets. 
-The annotations shown are the `SingleR` cell types and clusters.
-
-```{r}
-all_info_df <- all_info_df |> 
-  dplyr::arrange(final_annotation)
-
-annotated_exp_heatmap(
+single_annotation_heatmap(
   all_info_df, 
@@
   cell_type_column = "consensus_lumped", 
-  cluster_column = "final_lumped", 
   legend_title = "AUC"
@@
 
-Below we look at the mean expression of all genes in each of the custom gene sets we have defined both on a UMAP and in a density plot.
-
-```{r, fig.height=10, fig.width=10}
-# format expression data for UMAP and density plot 
-exp_df <- all_info_df |> 
-  tidyr::pivot_longer(ends_with("_mean"), names_to = "geneset", values_to = "mean") |>
-  dplyr::mutate(tumor_cell_classification = dplyr::if_else(singler_lumped == "tumor", "tumor", "other"))
-
-expression_umap(exp_df, mean, geneset)
-```
-
-Here we look at the mean expression for each gene set across all clusters.
+Below we look at the mean expression of all genes in each of the custom gene sets we have across the consensus cell types. 
+First using a density plot and then using a heatmap.
 
@@
 mean_exp_columns <- colnames(all_info_df)[which(endsWith(colnames(all_info_df), "_mean"))]
-cluster_density_plot(all_info_df, mean_exp_columns, annotation_column = "final_lumped", "mean gene expression")
+cluster_density_plot(all_info_df, mean_exp_columns, annotation_column = "consensus_lumped", "mean gene expression")
+```
+```{r}
+single_annotation_heatmap(
+  all_info_df, 
+  exp_columns = mean_exp_columns, 
+  cell_type_column = "consensus_lumped", 
+  legend_title = "AUC"
+)
 ```
 
-What are the other cell types not present here? 
+### Conclusions based on workflow results
+
+Looking at these results, it looks like things labeled as "tumor" by `SingleR` and "Unknown" in the consensus cell types have high AUC values for the `EWS-FLI1` upregulated gene sets. 
+We also see low expression of `EWS-FLI1` repressed targets in these cells. 
+It looks like the cells that are called as muscle cells in both cell types have high expression of `EWS-FLI1` targets. 
+This makes me think those are actually tumor cells that are mis-classified.
+
+Intriguingly we see high expression of `hallmark_EMT` and the `EWS-FLI1` repressed targets in fibroblasts, which also happen to have high expression of the `EWS-FLI1` low targets identifed by Wrenn _et al._ (`wrenn-nt5e-genes`).
+That paper identifies `EWS-FLI1` low cells as cancer associated fibroblasts.
+It's also important to note that many of the genes that define the `EWS-FLI1` low state are also high in fibroblasts, so distinguishing them may be difficult. 
+We also see that the chondrocytes identified in `SingleR` have high expression of this gene set.
+
+Another key observation is that we see separation between the AUC values observed in tumor/ other cells in the density plots, but these values are much lower than the AUC threshold automatically determined by `AUCell`.
+I think because of that we want to pick gene sets where we see very clear separation in AUC values between cell types to aid in finalizing our assignments.
+
+The density plots also show increased expression in the immune markers in immune cell populations and endothelial cell markers in the endothelial cells. 
+There may be a small group of cells that have high expression of the proliferative markers. 
+Perhaps we can label any cells that are tumor cells and show proliferative markers as `Tumor EWS-high proliferative`.
+
+## Refining cell type annotations
+
+Here we will combine annotations from `SingleR` and consensus cell types with help from the `AUCell` output. 
+We'll use the following criteria:
+
+- `Tumor EWS-low CAF`: High expression of `wrenn-nt5e-genes` (> 0.1 AUC) and high expression of repressed targets using `miyagawa_down` (> 0.03 AUC)
+- `Tumor EWS-high`: High expression of `aynaud-ews-targets` (>0.025 AUC) and identified as tumor with `SingleR` and identified as either unknown, muscle cell, or smooth muscle cell in the consensus cell types
+- `Tumor EWS-high proliferative`: Cells that meet the requirements for `Tumor EWS-high` and have expression of proliferative markers defined as mean > 0
+- All other cell types will be based on the `consensus_annotation`
+
+```{r}
+# set desired order for plotting final annotations 
+cell_type_order <- c(
+  "tumor EWS-high",
+  "tumor EWS-high proliferative",
+  "tumor EWS-low CAF",
+  "fibroblast",
+  "endothelial cell",
+  "macrophage",
+  "mature T cell",
+  "memory T cell",
+  "Unknown",
+  "All remaining cell types"
+)
+
+# define "final.final" cell types 
+all_info_df <- all_info_df |> 
+  dplyr::mutate(
+    final_annotation = dplyr::case_when(
+      `auc_wrenn-nt5e-genes` > 0.1 & auc_miyagawa_down > 0.03 ~ "tumor EWS-low CAF",
+      `auc_aynaud-ews-targets` > 0.025 | 
+        consensus_annotation %in% c("Unknown", "muscle cell", "smooth muscle cell") &
+        singler_lumped == "tumor" ~ "tumor EWS-high",
+      .default = consensus_annotation
+    ),
+    final_annotation = dplyr::if_else(
+      final_annotation == "tumor EWS-high" & proliferative_mean > 0,
+      "tumor EWS-high proliferative",
+      final_annotation
+    ),
+    # lump together for easier plotting
+    final_lumped = final_annotation |>
+      forcats::fct_lump_n(9, other_level = "All remaining cell types", ties.method = "first") |>
+      forcats::fct_infreq() |>
+      forcats::fct_relevel(cell_type_order)
+  )
+
+table(all_info_df$final_annotation)
+```
+For plotting purposes, we won't show any of the cell types that only have a handful of cells. 
+Let's see what cell types we won't be looking at if we only plot the top 9 cell types.
+
 ```{r}
@@
 
-Now let's make a dot plot! Or a heatmap only looking at key genes! 
-In these plots, lets group all T cells together.
+## Validation of cell type assignments
+
+In the following section we'll show some plots to help validate that we are content with our cell type assignments. 
+These plots will look specifically at expression of a set of key marker genes that we expect to up in the assigned cell types. 
+These markers are found in `references/combined-markers.tsv` and include a smaller set of markers for each of the cell types we have identified.
 
 ```{r}
-markers_df <- readr::read_tsv(file.path(module_base, "references", "combined-markers.tsv"))
-genes <- markers_df |>
+validation_markers_df <- readr::read_tsv(validation_markers_file)
+
+# pull out list of genes 
+genes <- validation_markers_df |>
   dplyr::pull(ensembl_gene_id)
 
+# get individual gene counts for all marker genes 
 gene_cts <- logcounts(sce[genes, ]) |>
@@
 
-rowData(sce)[genes, ] |> 
-  as.data.frame() |> 
-  dplyr::select(ends_with(".mean")) |> 
-  tibble::rownames_to_column("ensembl_gene_id") |> 
-  tidyr::pivot_longer(!ensembl_gene_id, names_to = "library_id", values_to = "mean")
-
-celltypes_df <- endsWith()celltypes_df <- all_info_df |> 
+# get all unique cell types from the final lumped group 
+celltypes_df <- all_info_df |> 
   dplyr::select(barcodes, final_lumped)
 
+# create a df that has gene expression column and column indicating whether or not the gene is detected in that cell 
 genes_df <- gene_cts |> 
@@
   dplyr::left_join(celltypes_df, by = c("barcodes")) |> 
+  dplyr::left_join(validation_markers_df, by = c("ensembl_gene_id")) |> 
+  dplyr::rowwise() |> 
   dplyr::mutate(
-    percent_expression = 
+    detected = gene_exp > 0 # column indicating if gene is present or not
   )
 
-head(genes_df)
+# get total number of cells per final annotation group 
+total_cells_df <- genes_df |> 
+  dplyr::select(barcodes, final_lumped) |> 
+  unique() |> 
+  dplyr::count(final_lumped, name = "total_cells")
+
+# get total number of cells each gene is detected in per group 
+# and mean gene expression per group 
+group_stats_df <- genes_df |> 
+  dplyr::group_by(final_lumped, ensembl_gene_id) |>
+  dplyr::summarize(
+    detected_count = sum(detected),
+    mean_exp = mean(gene_exp)
+  )
+
+gene_summary_df <- genes_df |> 
+  # add total cells
+  dplyr::left_join(total_cells_df, by = c("final_lumped")) |> 
+  # add per gene stats
+  dplyr::left_join(group_stats_df, by = c("ensembl_gene_id", "final_lumped")) |> 
+  dplyr::select(gene_symbol, final_lumped, cell_type, total_cells, detected_count, mean_exp) |> 
+  unique() |>
+  dplyr::rowwise() |> 
+  dplyr::mutate(
+    # get total percent
+    percent_exp = (detected_count/total_cells) * 100,
+    # order genes based on cell type they indicate
+    gene_symbol = factor(gene_symbol, levels = validation_markers_df$gene_symbol),
+    final_lumped = forcats::fct_relevel(final_lumped, cell_type_order)
+
+  ) 
+```
+
+```{r, fig.height=7}
+# filter out low expressed genes
+dotplot_df <- gene_summary_df |> 
+  dplyr::filter(mean_exp > 0, percent_exp > 10)
+
+dotplot <- ggplot(dotplot_df, aes(y = forcats::fct_rev(final_lumped), x = gene_symbol, color = mean_exp, size = percent_exp)) +
+  geom_point() +
+  scale_color_viridis_c(option = "magma") +
+  theme(
+    axis.text.x = element_text(angle = 90, hjust = 0.5)
+  ) +
+  labs(
+    x = "",
+    y = ""
+  )
+
+color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = cell_type)) + 
+  geom_tile() + 
+  scale_fill_brewer(palette = 'Set1') +
+  ggmap::theme_nothing() +
+  theme(legend.position = "bottom") +
+  labs(fill = "")
+
+dotplot + color_bar +
+  patchwork::plot_layout(ncol = 1, heights = c(4, 0.1)) 
+```
+
+A few notes from this plot:
+
+- Generally tumor markers are present throughout, although they are highest in tumor cells. 
+- Tumor EWS low CAFs show pretty strong expression of the `EWS-low` markers. 
+These markers are also present in fibroblasts, but to a lesser degree. 
+Additionally, `TNC` (which has been published to be important in the metastatic phenotype in Ewing cells) is specific to the low cells and not found in the fibroblasts, which makes me more confident that those are indeed tumor cells.
+- Endothelial cells show strong expression of `PECAM1` and `VWF`, while that is not seen in most of the other cells. 
+- All immune cell types show `PTPRC` as expected with macrophages also showing `MRC1`. 
+The only concern I see is that the `mature T cell` does not express either of the `CD3` genes, but the memory T cells do? 
+I'm not totally sure what to do there since that label is a consensus label so perhaps that gene is not as well covered in these samples?
+
+Let's make a heatmap version of the same plot.
+
+```{r, fig.width=7}
+# note that we can't use the same heatmap function as before since we are looking at cell types as rows rather than gene sets
+# we also want to specify the annotation name
+
+# first make a mtx to use for the heatmap with rows as cell types and genes as columns 
+heatmap_mtx <- gene_summary_df |> 
+  dplyr::select(gene_symbol, final_lumped, mean_exp) |> 
+  tidyr::pivot_wider(
+    names_from = gene_symbol,
+    values_from = mean_exp
+  ) |> 
+  tibble::column_to_rownames("final_lumped") |>
+  as.matrix()
+# make sure cell types are present in the right order
+heatmap_mtx <- heatmap_mtx[cell_type_order,]
+
+# get annotation colors for each of the marker gene types 
+marker_gene_categories <- unique(validation_markers_df$cell_type)
+num_categories <- length(marker_gene_categories)
+category_colors <- palette.colors(palette = "Dark2") |>
+  head(n = num_categories) |>
+  purrr::set_names(marker_gene_categories)
+
+# create annotation for heatmap
+annotation <- ComplexHeatmap::columnAnnotation(
+  marker_gene_category = validation_markers_df$cell_type,
+  col = list(
+    marker_gene_category = category_colors
+  )
+)
+
+ComplexHeatmap::Heatmap(
+    heatmap_mtx,
+    # set the color scale based on min and max values
+    col = circlize::colorRamp2(seq(min(heatmap_mtx), max(heatmap_mtx), length = 2), colors = c("white", "#00274C")),
+    border = TRUE,
+    ## Row parameters
+    cluster_rows = FALSE,
+    row_title = "",
+    row_title_side = "left",
+    row_names_side = "left",
+    row_dend_side = "right",
+    row_names_gp = grid::gpar(fontsize = 10),
+    ## Column parameters
+    cluster_columns = FALSE,
+    show_column_names = TRUE,
+    column_names_gp = grid::gpar(fontsize = 8),
+    top_annotation = annotation,
+    heatmap_legend_param = list(
+      title = "Mean expression"
+    )
+  )
+```
+
+And finally we'll look at our annotations on a UMAP! 
+Because, why not.
+
+```{r}
+ggplot(all_info_df, aes(x = UMAP1, y = UMAP2, color = final_lumped)) +
+  geom_point(alpha = 0.5, size = 0.01) +
+  # set color for all remaining cell types since there are more colors than in the palette
+  scale_color_manual(values = c(palette.colors(palette = "Set1"), "grey65", "grey90")) +
+  labs(color = "") +
+  guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))
+```
+
+## Export cell types
+
+Now that we have our cell types let's export them to a TSV to save for future use!
+
+```{r}
+annotation_df <- all_info_df |> 
+  dplyr::mutate(
+    # remove library id from barcodes since we have a library id column already 
+    barcodes = stringr::word(barcodes, -1, sep = "-"),
+    # assign ontology ID using consensus ontology ID
+    # anything without an ID is either unknown or tumor 
+    final_ontology = dplyr::if_else(
+      final_annotation == consensus_annotation,
+      consensus_ontology,
+      final_annotation
+    )
+  ) |> 
+  dplyr::select(
+    barcodes,
+    library_id,
+    sample_id,
+    sample_type,
+    singler_ontology,
+    singler_annotation,
+    consensus_annotation,
+    consensus_ontology,
+    final_annotation,
+    final_ontology
+  )
+
+readr::write_tsv(annotation_df, output_file)
 ```

### 2025-02-05 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Update 07_combined_annotation_across_samples_exploration.Rmd to pass the cli test run

@@
 params:
-  predicted.celltype.threshold: 0.85
+  predicted.celltype.threshold: 0.75
   cnv_threshold_low: 0
@@
         # proportion of cnv per chromosome
-        proportion_cnv_chr1 = srat$proportion_cnv_chr1,
-        proportion_cnv_chr2 = srat$proportion_cnv_chr2,
-        proportion_cnv_chr3 = srat$proportion_cnv_chr3,
-        proportion_cnv_chr4 = srat$proportion_cnv_chr4,
-        proportion_cnv_chr5 = srat$proportion_cnv_chr5,
-        proportion_cnv_chr6 = srat$proportion_cnv_chr6,
-        proportion_cnv_chr7 = srat$proportion_cnv_chr7,
-        proportion_cnv_chr8 = srat$proportion_cnv_chr8,
-        proportion_cnv_chr9 = srat$proportion_cnv_chr9,
-        proportion_cnv_chr10 = srat$proportion_cnv_chr10,
-        proportion_cnv_chr11 = srat$proportion_cnv_chr11,
-        proportion_cnv_chr12 = srat$proportion_cnv_chr12,
-        proportion_cnv_chr13 = srat$proportion_cnv_chr13,
-        proportion_cnv_chr14 = srat$proportion_cnv_chr14,
-        proportion_cnv_chr15 = srat$proportion_cnv_chr15,
-        proportion_cnv_chr16 = srat$proportion_cnv_chr16,
-        proportion_cnv_chr17 = srat$proportion_cnv_chr17,
-        proportion_cnv_chr18 = srat$proportion_cnv_chr18,
-        proportion_cnv_chr19 = srat$proportion_cnv_chr19,
-        proportion_cnv_chr20 = srat$proportion_cnv_chr20,
-        proportion_cnv_chr21 = srat$proportion_cnv_chr21,
-        proportion_cnv_chr22 = srat$proportion_cnv_chr22,
+        proportion_cnv_chr1p = srat$proportion_cnv_chr1p,
+        proportion_cnv_chr2p = srat$proportion_cnv_chr2p,
+        proportion_cnv_chr3p = srat$proportion_cnv_chr3p,
+        proportion_cnv_chr4p = srat$proportion_cnv_chr4p,
+        proportion_cnv_chr5p = srat$proportion_cnv_chr5p,
+        proportion_cnv_chr6p = srat$proportion_cnv_chr6p,
+        proportion_cnv_chr7p = srat$proportion_cnv_chr7p,
+        proportion_cnv_chr8p = srat$proportion_cnv_chr8p,
+        proportion_cnv_chr9p = srat$proportion_cnv_chr9p,
+        proportion_cnv_chr10p = srat$proportion_cnv_chr10p,
+        proportion_cnv_chr11p = srat$proportion_cnv_chr11p,
+        proportion_cnv_chr12p = srat$proportion_cnv_chr12p,
 
-        # cnv global estimation per chromosome
-        has_cnv_chr1 = srat$has_cnv_chr1,
-        has_cnv_chr2 = srat$has_cnv_chr2,
-        has_cnv_chr3 = srat$has_cnv_chr3,
-        has_cnv_chr4 = srat$has_cnv_chr4,
-        has_cnv_chr5 = srat$has_cnv_chr5,
-        has_cnv_chr6 = srat$has_cnv_chr6,
-        has_cnv_chr7 = srat$has_cnv_chr7,
-        has_cnv_chr8 = srat$has_cnv_chr8,
-        has_cnv_chr9 = srat$has_cnv_chr9,
-        has_cnv_chr10 = srat$has_cnv_chr10,
-        has_cnv_chr11 = srat$has_cnv_chr11,
-        has_cnv_chr12 = srat$has_cnv_chr12,
-        has_cnv_chr13 = srat$has_cnv_chr13,
-        has_cnv_chr14 = srat$has_cnv_chr14,
-        has_cnv_chr15 = srat$has_cnv_chr15,
-        has_cnv_chr16 = srat$has_cnv_chr16,
-        has_cnv_chr17 = srat$has_cnv_chr17,
-        has_cnv_chr18 = srat$has_cnv_chr18,
-        has_cnv_chr19 = srat$has_cnv_chr19,
-        has_cnv_chr20 = srat$has_cnv_chr20,
-        has_cnv_chr21 = srat$has_cnv_chr21,
-        has_cnv_chr22 = srat$has_cnv_chr22
+        proportion_cnv_chr16p = srat$proportion_cnv_chr16p,
+        proportion_cnv_chr17p = srat$proportion_cnv_chr17p,
+        proportion_cnv_chr18p = srat$proportion_cnv_chr18p,
+        proportion_cnv_chr19p = srat$proportion_cnv_chr19p,
+        proportion_cnv_chr20p = srat$proportion_cnv_chr20p,
+
+        proportion_cnv_chr1q = srat$proportion_cnv_chr1q,
+        proportion_cnv_chr2q = srat$proportion_cnv_chr2q,
+        proportion_cnv_chr3q = srat$proportion_cnv_chr3q,
+        proportion_cnv_chr4q = srat$proportion_cnv_chr4q,
+        proportion_cnv_chr5q = srat$proportion_cnv_chr5q,
+        proportion_cnv_chr6q = srat$proportion_cnv_chr6q,
+        proportion_cnv_chr7q = srat$proportion_cnv_chr7q,
+        proportion_cnv_chr8q = srat$proportion_cnv_chr8q,
+        proportion_cnv_chr9q = srat$proportion_cnv_chr9q,
+        proportion_cnv_chr10q = srat$proportion_cnv_chr10q,
+        proportion_cnv_chr11q = srat$proportion_cnv_chr11q,
+        proportion_cnv_chr12q = srat$proportion_cnv_chr12q,
+        proportion_cnv_chr13q = srat$proportion_cnv_chr13q,
+        proportion_cnv_chr14q = srat$proportion_cnv_chr14q,
+        proportion_cnv_chr15q = srat$proportion_cnv_chr15q,
+        proportion_cnv_chr16q = srat$proportion_cnv_chr16q,
+        proportion_cnv_chr17q = srat$proportion_cnv_chr17q,
+        proportion_cnv_chr18q = srat$proportion_cnv_chr18q,
+        proportion_cnv_chr19q = srat$proportion_cnv_chr19q,
+        proportion_cnv_chr20q = srat$proportion_cnv_chr20q,
+        proportion_cnv_chr21q = srat$proportion_cnv_chr21q,
+        proportion_cnv_chr22q = srat$proportion_cnv_chr22q,
+
+        # loss global estimation per chromosome
+        has_loss_chr1p = srat$has_loss_chr1p,
+        has_loss_chr2p = srat$has_loss_chr2p,
+        has_loss_chr3p = srat$has_loss_chr3p,
+        has_loss_chr4p = srat$has_loss_chr4p,
+        has_loss_chr5p = srat$has_loss_chr5p,
+        has_loss_chr6p = srat$has_loss_chr6p,
+        has_loss_chr7p = srat$has_loss_chr7p,
+        has_loss_chr8p = srat$has_loss_chr8p,
+        has_loss_chr9p = srat$has_loss_chr9p,
+        has_loss_chr10p = srat$has_loss_chr10p,
+        has_loss_chr11p = srat$has_loss_chr11p,
+        has_loss_chr12p = srat$has_loss_chr12p,
+
+        has_loss_chr16p = srat$has_loss_chr16p,
+        has_loss_chr17p = srat$has_loss_chr17p,
+        has_loss_chr18p = srat$has_loss_chr18p,
+        has_loss_chr19p = srat$has_loss_chr19p,
+        has_loss_chr20p = srat$has_loss_chr20p,
+
+        has_loss_chr1q = srat$has_loss_chr1q,
+        has_loss_chr2q = srat$has_loss_chr2q,
+        has_loss_chr3q = srat$has_loss_chr3q,
+        has_loss_chr4q = srat$has_loss_chr4q,
+        has_loss_chr5q = srat$has_loss_chr5q,
+        has_loss_chr6q = srat$has_loss_chr6q,
+        has_loss_chr7q = srat$has_loss_chr7q,
+        has_loss_chr8q = srat$has_loss_chr8q,
+        has_loss_chr9q = srat$has_loss_chr9q,
+        has_loss_chr10q = srat$has_loss_chr10q,
+        has_loss_chr11q = srat$has_loss_chr11q,
+        has_loss_chr12q = srat$has_loss_chr12q,
+        has_loss_chr13q = srat$has_loss_chr13q,
+        has_loss_chr14q = srat$has_loss_chr14q,
+        has_loss_chr15q = srat$has_loss_chr15q,
+        has_loss_chr16q = srat$has_loss_chr16q,
+        has_loss_chr17q = srat$has_loss_chr17q,
+        has_loss_chr18q = srat$has_loss_chr18q,
+        has_loss_chr19q = srat$has_loss_chr19q,
+        has_loss_chr20q = srat$has_loss_chr20q,
+        has_loss_chr21q = srat$has_loss_chr21q,
+        has_loss_chr22q = srat$has_loss_chr22q,
+
+        # dupli global estimation per chromosome
+        has_dupli_chr1p = srat$has_dupli_chr1p,
+        has_dupli_chr2p = srat$has_dupli_chr2p,
+        has_dupli_chr3p = srat$has_dupli_chr3p,
+        has_dupli_chr4p = srat$has_dupli_chr4p,
+        has_dupli_chr5p = srat$has_dupli_chr5p,
+        has_dupli_chr6p = srat$has_dupli_chr6p,
+        has_dupli_chr7p = srat$has_dupli_chr7p,
+        has_dupli_chr8p = srat$has_dupli_chr8p,
+        has_dupli_chr9p = srat$has_dupli_chr9p,
+        has_dupli_chr10p = srat$has_dupli_chr10p,
+        has_dupli_chr11p = srat$has_dupli_chr11p,
+        has_dupli_chr12p = srat$has_dupli_chr12p,
+
+        has_dupli_chr16p = srat$has_dupli_chr16p,
+        has_dupli_chr17p = srat$has_dupli_chr17p,
+        has_dupli_chr18p = srat$has_dupli_chr18p,
+        has_dupli_chr19p = srat$has_dupli_chr19p,
+        has_dupli_chr20p = srat$has_dupli_chr20p,
+
+        has_dupli_chr1q = srat$has_dupli_chr1q,
+        has_dupli_chr2q = srat$has_dupli_chr2q,
+        has_dupli_chr3q = srat$has_dupli_chr3q,
+        has_dupli_chr4q = srat$has_dupli_chr4q,
+        has_dupli_chr5q = srat$has_dupli_chr5q,
+        has_dupli_chr6q = srat$has_dupli_chr6q,
+        has_dupli_chr7q = srat$has_dupli_chr7q,
+        has_dupli_chr8q = srat$has_dupli_chr8q,
+        has_dupli_chr9q = srat$has_dupli_chr9q,
+        has_dupli_chr10q = srat$has_dupli_chr10q,
+        has_dupli_chr11q = srat$has_dupli_chr11q,
+        has_dupli_chr12q = srat$has_dupli_chr12q,
+        has_dupli_chr13q = srat$has_dupli_chr13q,
+        has_dupli_chr14q = srat$has_dupli_chr14q,
+        has_dupli_chr15q = srat$has_dupli_chr15q,
+        has_dupli_chr16q = srat$has_dupli_chr16q,
+        has_dupli_chr17q = srat$has_dupli_chr17q,
+        has_dupli_chr18q = srat$has_dupli_chr18q,
+        has_dupli_chr19q = srat$has_dupli_chr19q,
+        has_dupli_chr20q = srat$has_dupli_chr20q,
+        has_dupli_chr21q = srat$has_dupli_chr21q,
+        has_dupli_chr22q = srat$has_dupli_chr22q
       ) |>
@@
 cell_type_df <- cell_type_df |>
-  mutate(has_cnv_score = rowSums(cell_type_df[, grepl("has_cnv_chr", colnames(cell_type_df))])) |>
+  mutate(has_cnv_score = rowSums(cell_type_df[, grepl("has_loss_chr|has_dupli_chr", colnames(cell_type_df))])) |>
   mutate(has_cnv_score = case_when(
@@
 
-- _normal_ cells can be observe in all four compartments (`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV
+- _normal_ cells can be observe in all five compartments (`normal`,`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV
 We only allow a bit of flexibility in terms of CNV profile for immune and endothelium cells that have a high predicted score.
+Please note that in `06_infercnv.R`, we renamed all `immune` and `endothelium` cells with a `fetal_kidney_predicted.compartment.score` > `r params$predicted.celltype.threshold` as `normal`.
 Indeed, we know that false positive CNV can be observed in a cell type specific manner.
@@
 
-    compartment %in% c("endothelium", "immune") &
+    compartment %in% c("normal", "immune", "endothelium") &
       cell_type.score > params$predicted.celltype.threshold ~ "normal",
@@
 
-    cell_type_df$compartment %in% c("endothelium") &
-      (cell_type_df$compartment.score > params$predicted.celltype.threshold) ~ "endothelium",
-    cell_type_df$compartment %in% c("immune") &
-      (cell_type_df$compartment.score > params$predicted.celltype.threshold) ~ "immune",
+    cell_type_df$compartment %in% c("normal") &
+      cell_type_df$cell_type %in% c("endothelial cell") ~ "endothelium",
+    cell_type_df$compartment %in% c("normal") &
+      cell_type_df$cell_type %in% c("B cell",
+                                    "CD4-positive, alpha-beta T cell",
+                                    "conventional dendritic cell",
+                                    "lymphocyte",
+                                    "macrophage",
+                                    "mast cell",
+                                    "monocyte",
+                                    "natural killer cell",
+                                    "neutrophil",
+                                    "plasmacytoid dendritic cell") ~ "immune",
 
@@
 ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
-  geom_point(size = 0.2, shape = 19) +
+  geom_point(size = 1, shape = 19) +
   facet_wrap(facets = ~sample_id, ncol = 5) +
@@
 ggplot(tmp, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
-  geom_point(size = 0.2, shape = 19) +
+  geom_point(size = 1, shape = 19) +
   facet_wrap(facets = ~sample_id, ncol = 5) +
@@
 ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
-  geom_point(size = 0.2, shape = 19) +
+  geom_point(size = 1, shape = 19) +
   facet_wrap(facets = ~sample_id, ncol = 5) +
@@
 
+#### Loss on long arms
+
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
 for (i in 1:22) {
-  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_cnv_chr", i)))
+  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "q")))
+}
+```
+#### gains on long arms
+
+```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
+for (i in 1:22) {
+  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "q")))
+}
+```
+
+#### loss on short arms
+
+```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
+for (i in c(1:12, 16:20)) {
+  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "p")))
+}
+```
+
+#### gains on short arms
+
+```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
+for (i in c(1:12, 16:20)) {
+  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "p")))
 }
@@
 annotations_table <- cell_type_df |>
-  dplyr::select(
+  select(
     cell_barcode = barcode,
@@
 
-- Combining label transfer and CNV inference we have produced draft annotations for 35/40 Wilms tumor samples in SCPCP000006.
-We decided not to annotate samples for which the `infercnv` couldn't be run with a normal reference, as we don't trust the results in that case and want to avoid misclassification (`SCPCS000177`, `SCPCS000180`, `SCPCS000181`, `SCPCS000190`, `SCPCS000197`).
+- Combining label transfer and CNV inference we have produced draft annotations for all 40 Wilms tumor samples in SCPCP000006
 

### 2025-02-05 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
make tables prettier

@@
 # marker genes to be used for validating assignments 
-validation_markers_file <- file.path(module_base, "references", "combined-markers.tsv")
+validation_markers_file <- file.path(module_base, "references", "combined-validation-markers.tsv")
 ```
@@
 
-table(all_info_df$final_annotation)
+all_info_df |> 
+  dplyr::count(final_annotation) |> 
+  dplyr::arrange(desc(n))
 ```
@@
   dplyr::filter(final_annotation %in% missing_celltypes)
-table(missing_only$final_annotation)
+
+missing_only |> 
+  dplyr::count(final_annotation) |> 
+  dplyr::arrange(desc(n))
 ```

### 2025-02-06 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd
create setup functions

@@
 )
-```
-
-## Functions
-
-```{r}
-# function to read in project data frames with all cells in a project
-# output is a summarized table with total cells per sample, total cells per annotation, and number of cell types 
-summarize_celltypes <- function(file, id){
-
-  # read in data
-  df <- readr::read_tsv(file)
-
-  # get total cell count and number of assigned cell types per library
-  total_cells_df <- df |> 
-    dplyr::group_by(library_id) |> 
-    dplyr::summarize(
-      total_cells_per_library = length(library_id),
-      num_celltypes = length(unique(consensus_annotation))
-    )
-
-  summary_df <- df |> 
-    dplyr::group_by(library_id, sample_type, consensus_annotation, consensus_ontology) |> 
-    dplyr::summarize(total_cells_per_annotation = length(consensus_annotation)) |>
-    dplyr::left_join(total_cells_df, by = "library_id") |> 
-    dplyr::mutate(
-      # add percentage 
-      percent_cells_annotation = round((total_cells_per_annotation / total_cells_per_library) * 100, 2)
-    ) |> 
-    dplyr::ungroup()
-
-  return(summary_df)
-
-}
 ```
@@
 results_files <- results_files[project_ids]
+```
+
+```{r}
+# source summarize_celltypes() function
+setup_functions <- file.path(module_base, "exploratory-notebooks", "utils", "setup-functions.R")
+source(setup_functions)
 ```

### 2025-02-06 analyses/cell-type-consensus/exploratory-notebooks/utils/setup-functions.R

@@ -1,0 +1,28 @@
+# function to read in project data frames with all cells in a project
+# output is a summarized table with total cells per sample, total cells per annotation, and number of cell types 
+summarize_celltypes <- function(file, id){
+  
+  # read in data
+  df <- readr::read_tsv(file) 
+  
+  # get total cell count and number of assigned cell types per library
+  total_cells_df <- df |> 
+    dplyr::group_by(library_id) |> 
+    dplyr::summarize(
+      total_cells_per_library = length(library_id),
+      num_celltypes = length(unique(consensus_annotation))
+    )
+  
+  summary_df <- df |> 
+    dplyr::group_by(library_id, sample_type, consensus_annotation, consensus_ontology) |> 
+    dplyr::summarize(total_cells_per_annotation = length(consensus_annotation)) |>
+    dplyr::left_join(total_cells_df, by = "library_id") |> 
+    dplyr::mutate(
+      # add percentage 
+      percent_cells_annotation = round((total_cells_per_annotation / total_cells_per_library) * 100, 2)
+    ) |> 
+    dplyr::ungroup()
+  
+  return(summary_df)
+  
+}

### 2025-02-06 analyses/cell-type-consensus/exploratory-notebooks/utils/03-osteosarcoma-consensus-celltypes.Rmd
initiate osteo notebook

---
title: "Consensus cell types across osteosarcoma samples"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: show
---

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)
```

## Data setup

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)
module_base <- file.path(repository_base, "analyses", "cell-type-consensus")

# results directory with cell-type-consensus 
results_dir <- file.path(module_base, "results", "cell-type-consensus")

# data directory where project metadata files live
data_dir <- file.path(repository_base, "data", "current")
```

```{r}
# list all results files 
results_files <- list.files(results_dir, pattern = "_consensus-cell-types\\.tsv.\\gz$", full.names = TRUE)

# get project ids from file list  
project_ids <- stringr::str_remove(basename(results_files), "_consensus-cell-types.tsv.gz")
names(results_files) <- project_ids

# remove cell line projects from file list
osteo_project_ids <- c("SCPCP000017", "SCPCP000018", "SCPCP000023")
results_files <- results_files[osteo_project_ids]

# list metadata files 
metadata_files <- list.files(data_dir, pattern = "single_cell_metadata\\.tsv$", full.names = TRUE, recursive = TRUE)
metadata_ids <- stringr::word(dirname(metadata_files), -1, sep = "/")
names(metadata_files) <- metadata_ids
osteo_metadata_files <- metadata_files[osteo_project_ids]

# immune cell types 
consensus_immune_file <- file.path(module_base, "references", "consensus-immune-cell-types.tsv")
```

```{r}
# source summarize_celltypes() function
setup_functions <- file.path(module_base, "exploratory-notebooks", "utils", "setup-functions.R")
source(setup_functions)
```

```{r, message=FALSE}
# read in metadata
all_metadata_df <- osteo_metadata_files |>
  purrr::map(readr::read_tsv) |>
  dplyr::bind_rows() |>
  # select columns that might be useful 
  dplyr::select(
    project_id = scpca_project_id,
    library_id = scpca_library_id,
    disease_timing,
    age,
    sex,
    tissue_location,
    primary_or_metastasis,
    seq_unit
  )

# immune consensus cell types 
consensus_immune_types <- readr::read_tsv(consensus_immune_file) |> 
  dplyr::pull(consensus_annotation)

# read in results and prep data frame for plotting 
all_results_df <- results_files |> 
  purrr::imap(summarize_celltypes) |> 
  dplyr::bind_rows(.id = "project_id") |> 
  # join with sample metadata
  dplyr::left_join(all_metadata_df, by = c("project_id", "library_id")) |>
  dplyr::filter(sample_type != "patient-derived xenograft") |> 
  dplyr::mutate(is_immune = consensus_annotation %in% consensus_immune_types)

head(all_results_df)
```

```{r}
all_results_df |>
  dplyr::count(consensus_annotation) |> 
  dplyr::arrange(desc(n))
```

```{r}
all_results_df <- all_results_df |> 
    dplyr::mutate(
      # get most frequently observed cell types across libraries in that project 
      top_celltypes = forcats::fct_lump_n(consensus_annotation, 10, other_level = "All remaining cell types", ties.method = "first") |> 
        # sort by frequency 
        forcats::fct_infreq() |> 
        # make sure all remaining and unknown are last, use this to assign colors in specific order
        forcats::fct_relevel("All remaining cell types", "Unknown", after = Inf)
    )

# get all unique cell types ordered by frequency 
unique_celltypes <- all_results_df |> 
  dplyr::filter(!top_celltypes %in% c("All remaining cell types", "Unknown")) |> 
  dplyr::pull(top_celltypes) |> 
  unique() |>
  sort() |> 
  as.character()

# get color palette
colors <- c(
  palette.colors(palette = "alphabet")[1:9],
  "grey60", 
  "grey95"
)
names(colors) <- c(unique_celltypes, "All remaining cell types", "Unknown")

ggplot(all_results_df) + 
  aes(
    x = library_id, 
    y = percent_cells_annotation, 
    fill = top_celltypes
  ) +
  geom_col() + 
  facet_wrap(vars(project_id), scales ="free") +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(values = colors, name = "cell type") +
  theme(axis.text.x = element_blank())
```

```{r}
# t cell types
t_cell_types <- consensus_immune_types[stringr::str_detect(consensus_immune_types, "T")]
b_cell_types <- consensus_immune_types[stringr::str_detect(consensus_immune_types, "B")]
myeloid_cell_types <- c(
  "macrophage",
  "myeloid leukocyte",
  "neutrophil",
  "granulocyte",
  "monocyte",
  "erythroid lineage cell",
  "megakaryocyte",
  "eosinophil",
  "dendritic cell"
)

# look at immune cell types vs. unknown vs. other 
all_results_df <- all_results_df |> 
  dplyr::mutate(
    immune_category = dplyr::case_when(
      is_immune ~ "immune",
      consensus_annotation == "Unknown" ~ "unknown",
      .default = "other"
    ),
    broad_category = dplyr::case_when(
      consensus_annotation %in% t_cell_types ~ "T cell",
      consensus_annotation %in% b_cell_types ~ "B cell",
      consensus_annotation %in% myeloid_cell_types ~ "myeloid",
      consensus_annotation == "Unknown" ~ "unknown",
      .default = "non-immune"
    )
  )
```

```{r}
library_levels <- pct_immune_df <- all_results_df |> 
  dplyr::filter(immune_category == "immune") |> 
  dplyr::group_by(library_id) |> 
  dplyr::summarize(
    percent_immune = sum(total_cells_per_annotation)/total_cells_per_library
  ) |>
  unique() |> 
  dplyr::arrange(desc(percent_immune)) |> 
  dplyr::pull(library_id)

all_results_df <- all_results_df |> 
  dplyr::mutate(
    library_id = forcats::fct_relevel(library_id, library_levels)
  )

ggplot(all_results_df) + 
  aes(
    x = library_id, 
    y = percent_cells_annotation, 
    fill = immune_category
  ) +
  geom_col() + 
  # split samples based on sample type, patient tissue or pdx 
  #facet_wrap(vars(seq_unit), scales ="free") +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(values = c("red", "grey60", "grey95"), name = "cell type") +
  theme(axis.text.x = element_blank())
```

```{r}
# define colors for broad categories 
celtype_order <- c("myeloid", "B cell", "T cell", "non-immune", "unknown")

# get color palette
colors <- c(
  palette.colors(palette = "Dark2")[1:3],
  "grey60", 
  "grey95"
)
names(colors) <- celtype_order

ggplot(all_results_df) + 
  aes(
    x = library_id, 
    y = percent_cells_annotation, 
    fill = broad_category
  ) +
  geom_col() + 
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(values = colors, name = "cell type") +
  theme(axis.text.x = element_blank())
```

```{r}
updated_metadata_df <- all_results_df |> 
  dplyr::mutate(
    primary_or_metastasis = dplyr::if_else(!is.na(primary_or_metastasis), primary_or_metastasis, disease_timing)
  ) |> 
  dplyr::filter(primary_or_metastasis %in% c("Primary", "Metastasis"))

ggplot(updated_metadata_df) + 
  aes(
    x = library_id, 
    y = percent_cells_annotation, 
    fill = immune_category
  ) +
  geom_col() + 
  # split samples based on sample type, patient tissue or pdx 
  facet_wrap(vars(primary_or_metastasis), scales ="free") +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(values = c("red", "grey60", "grey95"), name = "cell type") +
  theme(axis.text.x = element_blank())

ggplot(updated_metadata_df, aes(x = primary_or_metastasis, y = percent_immune)) +
  ggforce::geom_sina() +
  stat_summary(fun.y=median, geom="crossbar" , color = "red", linewidth = 0.2) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        plot.margin = margin(10,10,10,10)) +
  labs(
    x = "", 
    y = "Percent of cells annotated as Immune"
  )
```

Do the same thing but with T/B/myeloid percentage? 
Will also need to update code to make sure immune percentage is actually in the df.

Then filter for initial diagnosis vs. recurrence and do the same thing

All samples together sorted by immune population 
facte by project id - what's different about osteo project 23? 
all samples broken up by primary vs. mets, stacked bar chart and sina plot 
all samples broken up by disease timing 
why is project 23 different? 
what about just T cell infiltrate? and macrophage?

### 2025-02-06 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
reorganize

@@
 fs::dir_create(results_dir)
-output_file <- file.path(results_dir, glue::glue("SCPCP000015_celltype-annotations.tsv.gz"))
+output_file <- file.path(results_dir, "SCPCP000015_celltype-annotations.tsv.gz")
 ```
@@
 source(plotting_functions)
+
+# source jaccard functions plot_jaccard()
+jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
+source(jaccard_functions)
 ```
@@
 all_info_df <- all_results_df |> 
-  dplyr::left_join(gene_exp_df, by = "barcodes")
+  dplyr::left_join(gene_exp_df, by = "barcodes") |> 
+  dplyr::mutate(
+    singler_annotation = dplyr::if_else(is.na(singler_annotation), "unknown", singler_annotation)
+  )
 ```
 
-## Results summary
+## Combine classifications from `SingleR` and consensus cell types
 
-First, we will just summarize the results from the various inputs and workflows. 
-Then we will look at those results and assign any tumor cells and refine annotations as needed.
+The first thing we will do is compare cell type annotations obtained by each method, `SingleR` with tumor cells as a reference and consensus annotations.
 
-### `aucell-singler-annotation` assignments
+Let's see how similar these annotations are and see if we can create a combined annotation. 
+Note that `SingleR` will label tumor cells but the consensus cell types only label normal cells. 
+Consensus cell types are observed when cell types from `SingleR` and `CellAssign` (using only normal tissue references) share a common ancestor. 
+If no consensus is found, the cells are labeled with "Unknown".
 
-The below UMAP shows the top cell types assigned by `SingleR` in the `aucell-singler-annotation.sh` workflow. 
-The top 7 cell types are shown and all other cell types are grouped together as "All remaining cell types".
+Any cells that are unable to be labeled via consensus cell types are labeled as "Unknown" and I expect these will line up with cells labeled as tumor cells by `SingleR`.
 
-```{r, fig.height = 5}
-plot_faceted_umap(all_info_df, singler_lumped, legend_title = "SingleR cell types") +
-  theme(strip.text = element_text(size = 8))
+```{r, fig.height=10}
+# first get jaccard index between consensus and singler
+jaccard_mtx <- make_jaccard_matrix(
+  all_info_df,
+  "consensus_annotation",
+  "singler_annotation"
+)
+
+ComplexHeatmap::Heatmap(
+  t(jaccard_mtx),
+  col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
+  border = TRUE,
+  ## Row parameters
+  cluster_rows = TRUE,
+  row_title = "SingleR with tumor reference",
+  row_title_gp = grid::gpar(fontsize = 12),
+  row_title_side = "left",
+  row_names_side = "left",
+  row_dend_side = "right",
+  row_names_gp = grid::gpar(fontsize = 10),
+  ## Column parameters
+  cluster_columns = TRUE,
+  column_title = "Consensus",
+  column_title_gp = grid::gpar(fontsize = 12),
+  column_names_side = "bottom",
+  column_names_gp = grid::gpar(fontsize = 10),
+  column_names_rot = 90,
+  ## Legend parameters
+  heatmap_legend_param = list(
+    title = "Jaccard index",
+    direction = "vertical",
+    legend_width = unit(1.5, "in")
+  ))
 ```
 
-### Consensus cell type assignments
+It looks like generally these two cell type annotations are in agreement. 
+We also see that most of the "Unknown" cells as denoted by consensus annotations are in fact lining up with tumor cells.
+It looks like some tumor cells labeled by `SingleR` are labeled as "smooth muscle cells" in the consensus cell types, so that's something to keep in mind.
 
-The below UMAP shows the top cell types for which consensus cell types were assigned.
-Consensus cell types are observed when cell types from `SingleR` and `CellAssign` share a common ancestor. 
-If no consensus is found, the cells are labeled with "Unknown". 
-We anticipate that in many cases, the "Unknown" cells will be tumor cells.
-The top 7 cell types are shown and all other cell types are grouped together as "All remaining cell types".
+Let's create a combined annotation that labels cells as tumor if the cell type label is tumor in `SingleR` and Unknown in consensus labels. 
+All other labels will be taken from the consensus cell types.
 
-```{r, fig.height=5}
-plot_faceted_umap(all_info_df, consensus_lumped, legend_title = "Consensus cell types") +
-  theme(strip.text = element_text(size = 8))
+```{r}
+# classify based on combined consensus and singler annotation 
+all_info_df <- all_info_df |> 
+  dplyr::mutate(combined_annotation = dplyr::if_else(
+    consensus_annotation == "Unknown" & singler_annotation == "tumor",
+    "tumor",
+    consensus_annotation
+  ),
+  # lump for plotting later 
+  combined_lumped = combined_annotation |>
+      forcats::fct_lump_n(9, other_level = "All remaining cell types", ties.method = "first") |>
+      forcats::fct_infreq()
+  )
+
+all_info_df |> 
+  dplyr::count(combined_annotation) |> 
+  dplyr::arrange(desc(n))
 ```
 
-### `AUCell` results
+## Use `AUCell` and mean gene set scores to refine tumor cells and identify tumor cell states
 
-The below plots show the AUC values determined by `AUCell` and output from `run-aucell-ews-signatures.sh`. 
-The first plot shows the individual AUC values on the UMAP.
+We used `AUCell` with gene sets that we expect to be up and down regulated in tumor cells. 
+We can look at the AUC values for these gene sets across all the cell types and see if there are any cells that should be classified as tumor that are not. 
+Additionally, we have gene sets that can help us distinguish EWS high and EWS low tumor cells.
 
-For context:
+First we'll look at the mean AUC value for each gene set across all cell types.
+
+Below is some information for the gene sets:
 
@@
 ```{r}
-# get the individual thresholds determined by AUCell for each msigdb geneset
-# we want this so we can show the threshold on the density plots 
-auc_threshold_df <- all_info_df |> 
-  tidyr::pivot_longer(starts_with("threshold_auc_"), names_to = "geneset", values_to = "threshold") |> 
-  dplyr::mutate(
-    geneset = stringr::str_remove(geneset, "threshold_auc_")
-  ) |> 
-  dplyr::select(barcodes, geneset, threshold)
-
-# reformat auc data for density plots and UMAPs showing AUC values 
+# reformat auc data for plots
 auc_df <- all_info_df |> 
-  tidyr::pivot_longer(starts_with("auc_"), names_to = "geneset", values_to = "auc_value") |> 
+  tidyr::pivot_longer(starts_with("auc_"), names_to = "geneset", values_to = "auc") |> 
   dplyr::mutate(
     geneset = stringr::str_remove(geneset, "auc_")
+  )
+
+# create mtx with mean auc for heatmap 
+auc_mtx <- auc_df |> 
+  dplyr::group_by(combined_annotation, geneset) |> 
+  dplyr::summarize(
+    mean_auc = mean(auc)
   ) |> 
-  dplyr::select(barcodes, UMAP1, UMAP2, geneset, auc_value) |> 
-  dplyr::left_join(auc_threshold_df, by = c("barcodes", "geneset")) |> 
-  dplyr::mutate(
-    in_geneset =  auc_value > threshold
+  tidyr::pivot_wider(names_from = geneset, 
+                     values_from = mean_auc) |> 
+  tidyr::drop_na() |> 
+  tibble::column_to_rownames("combined_annotation") |>
+  as.matrix()
+
+ComplexHeatmap::Heatmap(
+    auc_mtx,
+    # set the color scale based on min and max values
+    col = circlize::colorRamp2(seq(min(auc_mtx), max(auc_mtx), length = 2), colors = c("white", "#00274C")),
+    border = TRUE,
+    ## Row parameters
+    cluster_rows = TRUE,
+    row_title = "Combined cell types",
+    row_title_side = "left",
+    row_names_side = "left",
+    row_dend_side = "right",
+    row_names_gp = grid::gpar(fontsize = 10),
+    ## Column parameters
+    column_title = "MSigDB gene set",
+    cluster_columns = TRUE,
+    show_column_names = TRUE,
+    column_names_gp = grid::gpar(fontsize = 8),
+    heatmap_legend_param = list(
+      title = "Mean expression"
     )
+  )
+
 ```
 
-```{r, fig.height=10, fig.width=10}
-expression_umap(auc_df, auc_value, geneset)
+- Here it looks like the tumor cells have high expression of `EWS-FLI1` upregulated gene sets and low expression of `EWS-FLI1` repressed targets. 
+This suggests these cells are `EWS-high`. 
+- This same pattern is seen in cells labeled as "smooth muscle cell" and "stromal cell". 
+There may even be some slight expression of `EWS-FLI1` upregulated gene sets in "muscle cell", "Unknown", and some of the T cell populations. 
+Some of these cells are probably mistakenly labeled and should be tumor cells that are "EWS-high".
+- Fibroblasts and chondroctyes have low expression of `EWS-FLI1` upregulated gene sets and high expression of `EWS-FLI1` repressed targets. 
+Many of the genes in the `EWS-FLI1` repressed targets are also genes upregulated in fibroblasts so I'm sure we have a mix of tumor and fibroblasts in there.
+
+Because we have some normal cell types that are showing gene signatures expected to be specific to tumor cells, we should use the AUC values to help pull out any remaining cells that should be tumor cells.
+We can also use the up and down gene sets to distinguish EWS-high and EWS-low tumor cells.
+
+Let's pick some gene sets where we see variation across cell types that we can use to define cutoffs:
+
+- EWS-high gene sets: `aynaud-ews-targets` and `staege`
+- EWS-low gene sets: `wrenn-nt5e-genes` and `hallmark_EMT`
+
+```{r}
+genesets_of_interest = c(
+  "auc_aynaud-ews-targets",
+  "auc_staege",
+  "auc_wrenn-nt5e-genes",
+  "auc_hallmark_EMT"
+)
+
+cluster_density_plot(all_info_df, genesets_of_interest, "combined_lumped", "AUC")
 ```
 
-The below plot shows the distribution of AUC values for each gene set colored based on if the AUC value is above the threshold determined by `AUCell`, indicated with a dotted line.
+We can use these plots to help define cutoffs of AUC values to use for re-labeling any normal cells that should be tumor cells. 
+We expect that the endothelial cells and macrophages have low AUC values so we can use these cell types as the background. 
+Any AUC values above that background should be classified as tumor.
 
-```{r, fig.height=5}
+With this in mind, let's use these cutoffs to define tumor cells:
 
-ggplot(auc_df, aes(x = auc_value, color = in_geneset, fill = in_geneset)) +
-  geom_density(alpha = 0.5, bw = 0.01) +
-  facet_wrap(vars(geneset)) +
-  ggplot2::geom_vline(data = auc_df,
-                      mapping = aes(xintercept = threshold),
-                      lty = 2) +
-  theme(
-      aspect.ratio = 1,
-      strip.background = element_rect(fill = "transparent", linewidth = 0.5),
-      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
+- `aynaud-ews-targets` > 0.025 AUC
+- `staege` > 0.025 AUC
+- `wrenn-nt5e-genes` > 0.1 AUC
+- `hallmark_EMT` > 0.05 AUC
+
+Let's re-plot the density plots adding in a line for the AUC cutoffs we have chosen.
+
+```{r}
+cutoffs = c(
+  "auc_aynaud-ews-targets" = 0.025,
+  "auc_staege" = 0.025,
+  "auc_wrenn-nt5e-genes" = 0.1,
+  "auc_hallmark_EMT" = 0.05
 )
+
+cutoffs |>
+    purrr::imap(\(cutoff, column){
+      plot_density(
+        all_info_df,
+        column,
+        "combined_lumped"
+      ) +
+        theme(text = element_text(size = 8)) +
+        geom_vline(xintercept = cutoff)
+    }) |>
+    patchwork::wrap_plots(ncol = 2)
 ```
 
-Here we look at the AUC values for each gene set across all consensus cell types.
+I don't think these cutoffs are perfect, but I think they are reasonable. 
+Any cells that are to the left of these cutoffs will be classified as tumor.
 
-```{r, fig.height=10}
-auc_columns <- colnames(all_info_df)[which(startsWith(colnames(all_info_df), "auc_"))]
-cluster_density_plot(all_info_df, auc_columns, "consensus_lumped", "AUC")
+We can also use these to further stratify tumor cells into EWS high and EWS low classes.
+
+- `tumor EWS-low`: High expression of `wrenn-nt5e-genes` (> 0.1 AUC) and EMT markers `hallmark_EMT` (> 0.05 AUC)
+- `tumor EWS-high`: High expression of `aynaud-ews-targets` (>0.025 AUC) and `staege` (>0.025 AUC)
+
+```{r}
+# define "final.final" cell types 
+all_info_df <- all_info_df |> 
+  dplyr::mutate(
+    final_annotation = dplyr::case_when(
+      `auc_wrenn-nt5e-genes` > 0.1 & auc_hallmark_EMT > 0.05 ~ "tumor EWS-low",
+      `auc_aynaud-ews-targets` > 0.025 & auc_staege > 0.025 ~ "tumor EWS-high",
+      .default = consensus_annotation
+    ),
+    # lump together for easier plotting
+    final_lumped = final_annotation |>
+      forcats::fct_lump_n(9, other_level = "All remaining cell types", ties.method = "first") |>
+      forcats::fct_infreq() |>
+      forcats::fct_relevel("All remaining cell types", after = Inf)
+  )
+
+all_info_df |> 
+  dplyr::count(final_annotation) |> 
+  dplyr::arrange(desc(n))
 ```
 
-Here we look at the AUC values for each gene set across all `SingleR` cell types.
+Let's remake the density plot to see if we have pulled out the tumor cells from the normal cell types like muscle cells.
 
-```{r, fig.height=10}
-cluster_density_plot(all_info_df, auc_columns, "singler_lumped", "AUC")
+```{r}
+cluster_density_plot(all_info_df, genesets_of_interest, "final_lumped", "AUC")
 ```
 
-The heatmap below shows the AUC values for all cells and all gene sets and the final refined cell type annotations.
+It looks like generally the `tumor EWS-high` cells show high expression of both upregulated gene sets and `tumor EWS-low` cells show high expression of both downregulated gene sets, so I feel good about this!
+
+I will also note that there are some muscle cells that show slight expression of `aynaud-ews-targets`, but show little to no expression of `staege` targets. 
+I'm still a little worried these are actually tumor cells, but because they only show expression of one of the genesets, we are probably okay to keep them as is noting that the expression profile of muscle cells is very similar to those of tumor cells, this being sarcoma.
+
+## Identifying proliferative cells
+
+The last tumor cell state that we want to identify is if any cells can be considered proliferative using the set of proliferative markers in `tumor-cell-state-markers.tsv`. 
+Let's look at the mean expression of these markers across all cells. 
+Since we are plotting the proliferative markers, let's also look at our other custom marker gene sets.
 
 ```{r}
-# sort by final cell types 
+# get columns with mean expression of custom gene sets 
+mean_exp_columns <- colnames(all_info_df)[which(endsWith(colnames(all_info_df), "_mean"))]
+# make sure order is defined by cell type order
 all_info_df <- all_info_df |> 
-  dplyr::arrange(consensus_lumped)
+  dplyr::arrange(final_lumped)
 
@@
   all_info_df, 
-  exp_columns = auc_columns, 
-  cell_type_column = "consensus_lumped", 
+  exp_columns = mean_exp_columns, 
+  cell_type_column = "final_lumped", 
   legend_title = "AUC"
@@
 
-### Mean expression of custom gene sets
-
-Below we look at the mean expression of all genes in each of the custom gene sets we have across the consensus cell types. 
-First using a density plot and then using a heatmap.
-
-```{r, fig.height=7, warning=FALSE}
-mean_exp_columns <- colnames(all_info_df)[which(endsWith(colnames(all_info_df), "_mean"))]
-cluster_density_plot(all_info_df, mean_exp_columns, annotation_column = "consensus_lumped", "mean gene expression")
-```
-```{r}
-single_annotation_heatmap(
-  all_info_df, 
-  exp_columns = mean_exp_columns, 
-  cell_type_column = "consensus_lumped", 
-  legend_title = "AUC"
-)
-```
-
-### Conclusions based on workflow results
-
-Looking at these results, it looks like things labeled as "tumor" by `SingleR` and "Unknown" in the consensus cell types have high AUC values for the `EWS-FLI1` upregulated gene sets. 
-We also see low expression of `EWS-FLI1` repressed targets in these cells. 
-It looks like the cells that are called as muscle cells in both cell types have high expression of `EWS-FLI1` targets. 
-This makes me think those are actually tumor cells that are mis-classified.
-
-Intriguingly we see high expression of `hallmark_EMT` and the `EWS-FLI1` repressed targets in fibroblasts, which also happen to have high expression of the `EWS-FLI1` low targets identifed by Wrenn _et al._ (`wrenn-nt5e-genes`).
-That paper identifies `EWS-FLI1` low cells as cancer associated fibroblasts.
-It's also important to note that many of the genes that define the `EWS-FLI1` low state are also high in fibroblasts, so distinguishing them may be difficult. 
-We also see that the chondrocytes identified in `SingleR` have high expression of this gene set.
-
-Another key observation is that we see separation between the AUC values observed in tumor/ other cells in the density plots, but these values are much lower than the AUC threshold automatically determined by `AUCell`.
-I think because of that we want to pick gene sets where we see very clear separation in AUC values between cell types to aid in finalizing our assignments.
-
-The density plots also show increased expression in the immune markers in immune cell populations and endothelial cell markers in the endothelial cells. 
-There may be a small group of cells that have high expression of the proliferative markers. 
-Perhaps we can label any cells that are tumor cells and show proliferative markers as `Tumor EWS-high proliferative`.
-
-## Refining cell type annotations
-
-Here we will combine annotations from `SingleR` and consensus cell types with help from the `AUCell` output. 
-We'll use the following criteria:
-
-- `Tumor EWS-low CAF`: High expression of `wrenn-nt5e-genes` (> 0.1 AUC) and high expression of repressed targets using `miyagawa_down` (> 0.03 AUC)
-- `Tumor EWS-high`: High expression of `aynaud-ews-targets` (>0.025 AUC) and identified as tumor with `SingleR` and identified as either unknown, muscle cell, or smooth muscle cell in the consensus cell types
-- `Tumor EWS-high proliferative`: Cells that meet the requirements for `Tumor EWS-high` and have expression of proliferative markers defined as mean > 0
-- All other cell types will be based on the `consensus_annotation`
+It looks like we do have a set of cells that are `tumor EWS-high` and also have high expression of proliferative markers. 
+Because of this, let's label any cells that are `tumor EWS-high` and have mean expression of proliferative markers > 0 as `tumor EWS-high proliferative`.
 
 ```{r}
-# set desired order for plotting final annotations 
+# set desired order for plotting final annotations later
 cell_type_order <- c(
@@
   "tumor EWS-high proliferative",
-  "tumor EWS-low CAF",
+  "tumor EWS-low",
+  "muscle cell", 
+  "smooth muscle cell",
   "fibroblast",
@@
   "mature T cell",
-  "memory T cell",
   "Unknown",
@@
 
-# define "final.final" cell types 
 all_info_df <- all_info_df |> 
   dplyr::mutate(
-    final_annotation = dplyr::case_when(
-      `auc_wrenn-nt5e-genes` > 0.1 & auc_miyagawa_down > 0.03 ~ "tumor EWS-low CAF",
-      `auc_aynaud-ews-targets` > 0.025 | 
-        consensus_annotation %in% c("Unknown", "muscle cell", "smooth muscle cell") &
-        singler_lumped == "tumor" ~ "tumor EWS-high",
-      .default = consensus_annotation
-    ),
     final_annotation = dplyr::if_else(
@@
     ), 
-    # lump together for easier plotting
     final_lumped = final_annotation |>
-      forcats::fct_lump_n(9, other_level = "All remaining cell types", ties.method = "first") |>
+      forcats::fct_lump_n(10, other_level = "All remaining cell types", ties.method = "first") |>
       forcats::fct_infreq() |>
@@
 ```
-For plotting purposes, we won't show any of the cell types that only have a handful of cells. 
-Let's see what cell types we won't be looking at if we only plot the top 9 cell types.
 
-```{r}
-lumped_celltypes <- unique(all_info_df$final_lumped)
-all_celltypes <- unique(all_info_df$final_annotation)
-missing_celltypes <- setdiff(all_celltypes, lumped_celltypes)
-
-missing_only <- all_info_df |> 
-  dplyr::filter(final_annotation %in% missing_celltypes)
-
-missing_only |> 
-  dplyr::count(final_annotation) |> 
-  dplyr::arrange(desc(n))
-```
-
-I think we can leave these few cells alone and just lump them as "All remaining cell types" for plots as we have been doing.
+For plotting purposes, we will only show the top 10 cell types with fibroblast being the last cell type.
 
@@
 ```{r}
+# read in file with validation markers 
 validation_markers_df <- readr::read_tsv(validation_markers_file)
@@
 - Generally tumor markers are present throughout, although they are highest in tumor cells. 
-- Tumor EWS low CAFs show pretty strong expression of the `EWS-low` markers. 
+- EWS-high proliferative cells show strong expression of proliferation markers. 
+- Tumor EWS low cells show pretty strong expression of the `EWS-low` markers. 
 These markers are also present in fibroblasts, but to a lesser degree. 
@@
 - Endothelial cells show strong expression of `PECAM1` and `VWF`, while that is not seen in most of the other cells. 
-- All immune cell types show `PTPRC` as expected with macrophages also showing `MRC1`. 
-The only concern I see is that the `mature T cell` does not express either of the `CD3` genes, but the memory T cells do? 
-I'm not totally sure what to do there since that label is a consensus label so perhaps that gene is not as well covered in these samples?
+- All immune cell types show `PTPRC` as expected with macrophages also showing `MRC1` and T cells showing expression of `CD3G`
 
@@
   # set color for all remaining cell types since there are more colors than in the palette
-  scale_color_manual(values = c(palette.colors(palette = "Set1"), "grey65", "grey90")) +
+  scale_color_manual(values = c(palette.colors(palette = "Dark2"), "black", "grey65", "grey90")) +
   labs(color = "") +
   guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))
+```
+
+```{r, fig.height=7}
+plot_faceted_umap(all_info_df, final_lumped, legend_title = "Final cell type annotations") +
+  theme(strip.text = element_text(size = 6))
 ```

### 2025-02-06 analyses/cell-type-consensus/exploratory-notebooks/03-osteosarcoma-consensus-celltypes.Rmd
notebook for osteo samples

---
title: "Consensus cell types across osteosarcoma samples"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

This notebook summarizes the consensus cell types identified in all osteosarcoma samples that are part of ScPCA. 
This includes all samples from `SCPCP000017`, `SCPCP000018`, and `SCPCP000023`.

We are interested in looking at any differences in cell composition across samples, with a particular interest in looking at the immune cell populations.

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)
```

## Functions

```{r}
# barchart with or without faceting
# each bar is a stacked barchart using the fill_color
# faceting is only done if a facet_variable is provided
stacked_barchart <- function(
    df, 
    fill_color, 
    facet_variable = NULL,
    colors
){

  barchart <- ggplot(df) + 
    aes(
      x = library_id, 
      y = percent_cells_annotation, 
      fill = !!sym(fill_color)
    ) +
    geom_col() + 
    scale_y_continuous(expand = c(0,0)) +
    scale_fill_manual(values = colors) +
    theme(axis.text.x = element_blank()) +
    labs(
      fill= "cell type"
    )

  if(!is.null(facet_variable)){
    barchart <- barchart +
      facet_wrap(vars(!!sym(facet_variable)), scales ="free")
  }

  return(barchart)
}

# sina plot looking at immune percentage on the y-axis 
sina_plot <- function(df, plot_column){

  ggplot(immune_plot_df, aes(x = !!sym(plot_column), y = percent_immune)) +
    ggforce::geom_sina() +
    stat_summary(fun.y=median, geom="crossbar" , color = "red", linewidth = 0.2) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          plot.margin = margin(10,10,10,10)) +
    labs(
      x = "", 
      y = "Percent of cells annotated as Immune",
      title = plot_column
    )

}
```

## Data setup

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)
module_base <- file.path(repository_base, "analyses", "cell-type-consensus")

# results directory with cell-type-consensus 
results_dir <- file.path(module_base, "results", "cell-type-consensus")

# data directory where project metadata files live
data_dir <- file.path(repository_base, "data", "current")
```

```{r}
# list all results files 
results_files <- list.files(results_dir, pattern = "_consensus-cell-types\\.tsv.\\gz$", full.names = TRUE)

# get project ids from file list and assign as names
project_ids <- stringr::str_remove(basename(results_files), "_consensus-cell-types.tsv.gz")
names(results_files) <- project_ids

# grab only osteo projects
osteo_project_ids <- c("SCPCP000017", "SCPCP000018", "SCPCP000023")
results_files <- results_files[osteo_project_ids]

# list project metadata files and grab only osteo metadata 
metadata_files <- list.files(data_dir, pattern = "single_cell_metadata\\.tsv$", full.names = TRUE, recursive = TRUE)
metadata_ids <- stringr::word(dirname(metadata_files), -1, sep = "/")
names(metadata_files) <- metadata_ids
osteo_metadata_files <- metadata_files[osteo_project_ids]

# list of all immune cell types 
consensus_immune_file <- file.path(module_base, "references", "consensus-immune-cell-types.tsv")
```

```{r}
# source summarize_celltypes() function
setup_functions <- file.path(module_base, "exploratory-notebooks", "utils", "setup-functions.R")
source(setup_functions)
```

```{r, message=FALSE}
# get immune cell types 
immune_df <- readr::read_tsv(consensus_immune_file)
immune_types <- immune_df$consensus_annotation

# read in metadata
all_metadata_df <- osteo_metadata_files |>
  purrr::map(readr::read_tsv) |>
  dplyr::bind_rows() |>
  # select columns that might be useful 
  dplyr::select(
    project_id = scpca_project_id,
    library_id = scpca_library_id,
    disease_timing,
    age,
    sex,
    tissue_location,
    primary_or_metastasis,
    seq_unit
  )

# read in results and prep data frame for plotting 
all_results_df <- results_files |> 
  purrr::imap(summarize_celltypes) |> 
  dplyr::bind_rows(.id = "project_id") |> 
  # join with sample metadata
  dplyr::left_join(all_metadata_df, by = c("project_id", "library_id")) |>
  # remove pdx samples 
  dplyr::filter(sample_type != "patient-derived xenograft")
```

```{r}
# assign a color scheme for all bar charts

# get all possible cell types 
# there are exactly 25 of them so will use alphabet with two greys for unknown/all remaining 
unique_celltypes <- all_results_df |> 
  dplyr::filter(percent_cells_annotation > 0.1, # we're only ever going to plot cell types that are greater than 1%
                !consensus_annotation == "Unknown") |> 
  dplyr::pull(consensus_annotation) |> 
  unique() |>
  as.character()

# define colors for all cell types
all_celltype_colors <- c(
  palette.colors(palette = "alphabet")[1:25],
  "grey60", # all remaining
  "grey95", # unknown
  "grey60" # non-immune, use the same as all remaining since they are never in the same plot
)
names(all_celltype_colors) <- c(unique_celltypes, "All remaining cell types", "Unknown", "non-immune")

# define a three color scheme for immune, other, and unknown
three_color_scheme <- c(
  "immune" = "navy",
  "non-immune" = "grey60",
  "Unknown" = "grey95"
)
```

## Composition of top cell types

Here we look at the top cell types identified across all samples. 
We have already looked at this in `02-explore-consensus-results.Rmd`, but here we will plot all samples from all three osteo projects ordered by the total percent of cells labeled with a consensus label.

```{r}
# add column of "top cell types" for easier plotting 
all_results_df <- all_results_df |> 
  dplyr::mutate(
    # get most frequently observed cell types across libraries in that project 
    top_celltypes = forcats::fct_lump_n(consensus_annotation, 10, other_level = "All remaining cell types", ties.method = "first") |> 
      # sort by frequency 
      forcats::fct_infreq() |> 
      # make sure all remaining and unknown are last, use this to assign colors in specific order
      forcats::fct_relevel("All remaining cell types", "Unknown", after = Inf),
  )
```

```{r}
# get a vector of library ids ordered by total percentage annotated
library_levels <- all_results_df |> 
  dplyr::filter(consensus_annotation != "Unknown") |> 
  dplyr::group_by(library_id) |> 
  dplyr::summarize(
    total_percent_annotated = sum(total_cells_per_annotation)/total_cells_per_library
  ) |>
  unique() |> 
  dplyr::arrange(desc(total_percent_annotated)) |> 
  dplyr::pull(library_id)

# append any libraries that have 0 cells annotated
non_annotated_libs <- setdiff(unique(all_results_df$library_id), library_levels)
library_levels <- c(library_levels, non_annotated_libs)

# reorder by total percentage annotated 
total_order_df <- all_results_df |> 
  dplyr::mutate(
    library_id = forcats::fct_relevel(library_id, library_levels)
  )
```

```{r}
stacked_barchart(total_order_df, fill_color = "top_celltypes", colors = all_celltype_colors)
```

It looks like there's definitely some variation between distributions of cell types within the osteo samples. 
Generally, we see most annotated cells are smooth muscle cells and endothelial cells. 
There also appears to be some samples that have macrophages and/or T cell populations. 
We also see a handful of samples that don't have any cells that are annotated.

## Immune cell populations

Let's look specifically at immune cell populations. 
To do this we will lump all immune cells as "immune", all non-immune as "non-immune", and all unknown as "unknown". 
Here we will sort by total percentage of immune cells.

```{r}
# look at immune cell types vs. unknown vs. other 
all_results_df <- all_results_df |> 
  #dplyr::left_join(immune_df, by = c("consensus_annotation", "consensus_ontology")) |> 
  dplyr::mutate(
    # first get a column that is just immune, unknown, or other
    immune_category = dplyr::case_when(
      consensus_annotation %in% immune_types ~ "immune",
      consensus_annotation == "Unknown" ~ "Unknown",
      .default = "non-immune"
    )
  )
```

```{r}
# now sort just by immune percentage
immune_pct_df <- all_results_df |> 
  dplyr::filter(immune_category == "immune") |> 
  dplyr::group_by(library_id) |> 
  dplyr::summarize(
    percent_immune = sum(total_cells_per_annotation)/total_cells_per_library
  ) |>
  dplyr::select(library_id, percent_immune) |> 
  unique()

library_levels <- immune_pct_df |> 
  dplyr::arrange(desc(percent_immune)) |> 
  dplyr::pull(library_id)

# append any libraries that have 0 immune cells annotated
non_annotated_libs <- setdiff(unique(all_results_df$library_id), library_levels)
library_levels <- c(library_levels, non_annotated_libs)

immune_results_df <- all_results_df |> 
  dplyr::left_join(immune_pct_df, by = c("library_id")) |> 
  dplyr::mutate(
    percent_immune = dplyr::if_else(is.na(percent_immune), 0, percent_immune),
    library_id = forcats::fct_relevel(library_id, library_levels)
  )

stacked_barchart(immune_results_df, fill_color = "immune_category", colors = three_color_scheme)
```
Let's see how the percentage of immune cells compares across projects.

```{r}
stacked_barchart(immune_results_df, fill_color = "immune_category", facet_variable = "project_id", colors = three_color_scheme)
```

When looking at all samples together we do see variation in immune cells classified and it appears that `SCPCP000017` and `SCPCP000018` have more cells classified in general and have more cells classified as immune. 
It does appear that libraries that have more immune cell composition also have a higher percentage of non-immune cell types which could very well be a technical artifact and related to sample prep.

Just for visualization purposes, let's classify these as "hot" and "cold", where "hot" tumors have an immune composition > 5%.

```{r}
immune_results_df <- immune_results_df |> 
  dplyr::mutate(
    # classify hot/cold based on immune percent > 5 %
    hot_cold = dplyr::if_else(percent_immune > .05, "hot", "cold")
  )
```

```{r}
stacked_barchart(immune_results_df, fill_color = "immune_category", facet_variable = "hot_cold", colors = three_color_scheme)
```

Our threshold is pretty arbitrary, but I think there are definitely some "cold" tumors where we see little to no immune involvement.

Below we can look at all immune cell types vs. all non-immune.

```{r}
immune_results_df <- immune_results_df |> 
  dplyr::mutate(
    is_immune = dplyr::if_else(
      immune_category == "immune" | top_celltypes == "Unknown",
      consensus_annotation,
      "non-immune"
    )
  )

stacked_barchart(immune_results_df, fill_color = "is_immune", colors = all_celltype_colors)
```

Generally, if immune cells are present they tend to be macrophages. 
There's also a lot of T cell populations, which makes sense since it has been previously noted that osteosarcoma has a high percentage of T cell infiltrate (https://doi.org/10.18632/oncotarget.19071).

## Is there any relationship between immune cell percentage and clinical metadata?

One thing I was curious about was whether or not there are any differences in the presence of immune cells based on clinical metadata such as primary vs. metastasis, initial diagnosis vs. recurrence, and tissue location (bone vs. soft tissue). 
I also expect there might be some differences between single-cell and single-nuclei samples.

Below I look at the immune cell percentage stratified by each of the mentioned metadata. 
This is just a preliminary look and if we really want to dig into this further we should consider looking at algorithms for comparing cell type composition in single-cell RNA-seq data, but that is outside the scope of this notebook.

```{r}
# do some data wrangling on the metadata columns we want to look at 
immune_results_df <- immune_results_df |> 
  dplyr::mutate(
    primary_or_metastasis = dplyr::if_else(!is.na(primary_or_metastasis), primary_or_metastasis, disease_timing),
    disease_timing_mod = dplyr::if_else(disease_timing %in% c("Initial diagnosis", "Recurrence"), disease_timing, "other"),
    tissue_location_mod = dplyr::case_when(
      stringr::str_detect(tissue_location, "Bone|femur|Humerus|Tibia") ~ "Bone",
      #stringr::str_detect(tissue_location, "femur") ~ "Bone",
      stringr::str_detect(tissue_location, "lung|Lung|Chest") ~ "lung or chest",
      .default = "other"
    )
  )
```

```{r, fig.height = 10}
# get a list of metadata we care about 
metadata_categories <- c(
  "primary_or_metastasis",
  "disease_timing_mod",
  "tissue_location_mod",
  "seq_unit"
)

# make stacked bar plots looking at only immune and non immune 
metadata_categories |> 
  purrr::map(\(category){

    stacked_barchart(
      immune_results_df,
      fill_color = "immune_category",
      facet_variable = category,
      colors = three_color_scheme
    )
}) |>
  patchwork::wrap_plots(ncol = 1, guides = "collect")
```

```{r, fig.height=10}
# now look at all top celltypes 
metadata_categories |> 
  purrr::map(\(category){

    stacked_barchart(
      immune_results_df,
      fill_color = "top_celltypes",
      facet_variable = category,
      colors = all_celltype_colors
    )
}) |>
  patchwork::wrap_plots(ncol = 1, guides = "collect")
```

Below are sina plots that compare the total immune percentage for all samples in a given category.

```{r}
immune_plot_df <- immune_results_df |> 
  # just look at immune population across different diagnoses 
  dplyr::select(library_id, primary_or_metastasis, disease_timing_mod, tissue_location_mod, percent_immune, seq_unit) |> 
  unique()
```

```{r}
metadata_categories |> 
  purrr::map(\(category){
    sina_plot(immune_plot_df, category)
  })
```

It looks like there might be higher immune infiltrate in "bone" samples, but again I think if we want to make any conclusions we need to look at software that helps correct for technical artifacts. 
There may also be a difference in single-cell and single-nuclei, but there are much fewer single-cell samples so it's hard to say.

## Session info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-02-06 analyses/cell-type-consensus/exploratory-notebooks/03-osteosarcoma-consensus-celltypes.Rmd
re-render

@@
     # get most frequently observed cell types across libraries in that project 
-    top_celltypes = forcats::fct_lump_n(consensus_annotation, 10, other_level = "All remaining cell types", ties.method = "first") |> 
+    top_celltypes = forcats::fct_lump_n(consensus_annotation, 15, other_level = "All remaining cell types", ties.method = "first") |> 
       # sort by frequency 
@@
 We also see a handful of samples that don't have any cells that are annotated.
+
+```{r}
+stacked_barchart(total_order_df, fill_color = "top_celltypes", facet_variable = "project_id", colors = all_celltype_colors)
+```
+
+It looks like both total number of cells that are classified and composition of those cells is project dependent. 
+This makes sense since sample prep is probably different across labs.
 

### 2025-02-07 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
Apply suggestions from code review
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
   "sce file does not exist" = file.exists(sce_file),
-  "aucell results file does not exist" = file.exists(aucell_results_file)
+  "aucell results file does not exist" = file.exists(aucell_results_file), 
+  "at least one singler file is missing" = all(file.exists(singler_results_files),
+  "at least one consensus file is missing" = all(file.exists(consensus_results_files),
 )
@@
 ```{r}
-genesets_of_interest = c(
+genesets_of_interest <- c(
   "auc_aynaud-ews-targets",
@@
     x = "",
-    y = ""
+    y = "Final cell type annotation"
   )
@@
 - EWS-high proliferative cells show strong expression of proliferation markers. 
-- Tumor EWS low cells show pretty strong expression of the `EWS-low` markers. 
+- Tumor EWS-low cells show pretty strong expression of the `EWS-low` markers. 
 These markers are also present in fibroblasts, but to a lesser degree. 

### 2025-02-07 analyses/cell-type-ewings/template_notebooks/utils/plotting-functions.R

@@ -143,7 +143,7 @@ annotated_exp_heatmap <- function(
 #' Values in the heatmap correspond to an expression value or AUC
 #' Only one annotation will be included indicating cell type
 #'
-#' @param df Data frame containing exp_columns, cell_type_column, and cluster_column
+#' @param df Data frame containing all columns in exp_columns and cell_type_column
 #' @param exp_columns Vector of column names that contain values to be shown in the heatmap
 #' @param cell_type_column Column indicating cell types for each cell in the df
 #' @param legend_title Title to use for heatmap legend 

### 2025-02-07 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
refine plots based on review

@@
 
-- Annotations obtained by running `SingleR` with tumor cells as a reference output by `aucell-singler-annotation.sh`. 
-- Consensus cell type annotations output from the `cell-type-consensus` module. 
+- Annotations obtained by running `SingleR` with tumor cells as a reference output by [`aucell-singler-annotation.sh`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/aucell-singler-annotation.sh).
+- Consensus cell type annotations output by running [`assign-consensus-celltypes.sh` in `cell-type-consensus`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-consensus/assign-consensus-celltypes.sh).
 - AUC values as calculated by `AUCell` for a set of Ewing sarcoma specific gene sets in MSigDB output by `run-aucell-ews-signatures.sh`.
@@
   "aucell results file does not exist" = file.exists(aucell_results_file), 
-  "at least one singler file is missing" = all(file.exists(singler_results_files),
-  "at least one consensus file is missing" = all(file.exists(consensus_results_files),
+  "at least one singler file is missing" = all(file.exists(singler_results_files)),
+  "at least one consensus file is missing" = all(file.exists(consensus_results_files))
 )
@@
 
-The first thing we will do is compare cell type annotations obtained by each method, `SingleR` with tumor cells as a reference and consensus annotations.
+The first thing we will do is compare cell type annotations obtained by each method, `SingleR` with tumor cells as a reference (output by [`aucell-singler-annotation.sh`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/aucell-singler-annotation.sh)) and consensus annotations (output from [`assign-consensus-celltypes.sh` in `cell-type-consensus`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-consensus/assign-consensus-celltypes.sh).
 
@@
 
-```{r, fig.height=10}
+```{r, fig.height=5}
+# get cell types that have at least 50 cells
+singler_to_keep <- all_info_df |> 
+  dplyr::count(singler_annotation) |>
+  dplyr::filter(n >= 50) |> 
+  dplyr::pull(singler_annotation)
+
 # first get jaccard index between consensus and singler
-jaccard_mtx <- make_jaccard_matrix(
-  all_info_df,
+jaccard_mtx <- all_info_df |> 
+  dplyr::filter(singler_annotation %in% singler_to_keep) |> 
+  make_jaccard_matrix(
     "consensus_annotation",
@@
   row_dend_side = "right",
-  row_names_gp = grid::gpar(fontsize = 10),
+  row_names_gp = grid::gpar(fontsize = 12),
   ## Column parameters
@@
   column_names_side = "bottom",
-  column_names_gp = grid::gpar(fontsize = 10),
+  column_names_gp = grid::gpar(fontsize = 12),
   column_names_rot = 90,
@@
     forcats::fct_lump_n(9, other_level = "All remaining cell types", ties.method = "first") |>
-      forcats::fct_infreq()
+    forcats::fct_infreq() |> 
+    forcats::fct_relevel("All remaining cell types", after = Inf)
   )
@@
 
-- `aynaud-ews-targets` > 0.025 AUC
-- `staege` > 0.025 AUC
+- `aynaud-ews-targets` > 0.04 AUC
+- `staege` > 0.01 AUC
 - `wrenn-nt5e-genes` > 0.1 AUC
@@
 cutoffs = c(
-  "auc_aynaud-ews-targets" = 0.025,
-  "auc_staege" = 0.025,
+  "auc_aynaud-ews-targets" = 0.04,
+  "auc_staege" = 0.01,
   "auc_wrenn-nt5e-genes" = 0.1,
@@
 I don't think these cutoffs are perfect, but I think they are reasonable. 
-Any cells that are to the left of these cutoffs will be classified as tumor.
+Any cells that are to the right of these cutoffs will be classified as tumor.
 
@@
 - `tumor EWS-low`: High expression of `wrenn-nt5e-genes` (> 0.1 AUC) and EMT markers `hallmark_EMT` (> 0.05 AUC)
-- `tumor EWS-high`: High expression of `aynaud-ews-targets` (>0.025 AUC) and `staege` (>0.025 AUC)
+- `tumor EWS-high`: High expression of `aynaud-ews-targets` (>0.04 AUC) and `staege` (>0.01 AUC)
 
 ```{r}
+# top cell type order 
+density_plot_order <- c(
+  "tumor EWS-high",
+  "tumor EWS-low", 
+  "Unknown",
+  "smooth muscle cell",
+  "endothelial cell",
+  "macrophage",
+  "fibroblast",
+  "muscle cell",
+  "mature T cell",
+  "All remaining cell types"
+)
+
 # define "final.final" cell types 
@@
       `auc_wrenn-nt5e-genes` > 0.1 & auc_hallmark_EMT > 0.05 ~ "tumor EWS-low",
-      `auc_aynaud-ews-targets` > 0.025 & auc_staege > 0.025 ~ "tumor EWS-high",
+      `auc_aynaud-ews-targets` > 0.04 & auc_staege > 0.01 ~ "tumor EWS-high",
       .default = consensus_annotation
@@
       forcats::fct_infreq() |>
-      forcats::fct_relevel("All remaining cell types", after = Inf)
+      forcats::fct_relevel(density_plot_order)
   )
@@
 
+# first get tally for how many libraries each cell type is present in 
+total_libraries <- all_info_df |> 
+  dplyr::select(final_annotation, library_id) |> 
+  unique() |> # get rid of duplicate cells
+  dplyr::group_by(final_annotation) |> 
+  dplyr::summarize(
+    total_libraries = length(library_id)) |> 
+  dplyr::arrange(desc(total_libraries))
+
+# print out both total cells and total libraries 
 all_info_df |> 
-  dplyr::count(final_annotation) |>
-  dplyr::arrange(desc(n))
+  dplyr::count(final_annotation, name = "total_cells") |>
+  dplyr::left_join(total_libraries, by = "final_annotation") |> 
+  dplyr::arrange(desc(total_cells))
 ```
 
-For plotting purposes, we will only show the top 10 cell types with fibroblast being the last cell type.
+For plotting purposes, we will only show the top 10 cell types with muscle cell being the last cell type. 
+It looks like all other cell types are only present in 1-3 libraries. 
+For any cells where there's only 1 cell we may want to confirm that those cells express the expected targets, but that is outside the scope of this notebook.
 
@@
 
-```{r, fig.height=7}
+```{r, fig.height=7, fig.width=10}
 # filter out low expressed genes
 dotplot_df <- gene_summary_df |> 
-  dplyr::filter(mean_exp > 0, percent_exp > 10)
+  dplyr::filter(mean_exp > 0, percent_exp > 10) |>
+  dplyr::arrange(final_lumped) |> 
+  dplyr::mutate(y_label = as.factor(glue::glue("{final_lumped} ({total_cells})")))
 
-dotplot <- ggplot(dotplot_df, aes(y = forcats::fct_rev(final_lumped), x = gene_symbol, color = mean_exp, size = percent_exp)) +
+dotplot <- ggplot(dotplot_df, aes(y = forcats::fct_rev(y_label), x = gene_symbol, color = mean_exp, size = percent_exp)) +
   geom_point() +
-  scale_color_viridis_c(option = "magma") +
+  scale_color_viridis_c(option = "magma", limits = c(0,2.5), oob = scales::squish) +
   theme(
@@
 
+# specify legend order 
+color_order <- dotplot_df |> 
+  dplyr::pull(cell_type) |> 
+  unique()
+
 color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = cell_type)) + 
   geom_tile() + 
-  scale_fill_brewer(palette = 'Set1') +
+  scale_fill_brewer(palette = 'Set1', breaks = color_order) +
   ggmap::theme_nothing() +
@@
 These markers are also present in fibroblasts, but to a lesser degree. 
-Additionally, `TNC` (which has been published to be important in the metastatic phenotype in Ewing cells) is specific to the low cells and not found in the fibroblasts, which makes me more confident that those are indeed tumor cells.
+Additionally, `TNC`, which has shown to be secreted by Ewing cells and important in the metastatic phenotype of Ewing sarcoma (https://doi.org/10.1016/j.neo.2019.08.007), is specific to the low cells and not found in the fibroblasts, which makes me more confident that those are indeed tumor cells.
 - Endothelial cells show strong expression of `PECAM1` and `VWF`, while that is not seen in most of the other cells. 
@@
 
-```{r, fig.width=7}
+```{r, fig.width=10}
 # note that we can't use the same heatmap function as before since we are looking at cell types as rows rather than gene sets
@@
 annotation <- ComplexHeatmap::columnAnnotation(
-  marker_gene_category = validation_markers_df$cell_type,
+  category = validation_markers_df$cell_type,
   col = list(
-    marker_gene_category = category_colors
+    category = category_colors
+  ),
+  annotation_legend_param = list(
+      title = "Marker gene category",
+      labels = unique(validation_markers_df$cell_type)
     )

### 2025-02-07 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
fix some legend titles

@@
     heatmap_legend_param = list(
-      title = "Mean expression"
+      title = "Mean AUC"
     )
@@
   cell_type_column = "final_lumped", 
-  legend_title = "AUC"
+  legend_title = "Mean expression"
 )

### 2025-02-10 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
07_annotations_v2

@@
 library(DT)
+library(ggpubr)
 ```
@@
 
-- the proportion of estimated CNV per chromosome, `{i}` in 1 to 22 `proportion_cnv_chr{i}`
+- the CNV `{dupli|loss}` estimation per chromosome {i} and arm `{p|q}`: 
+`has_{dupli|loss}_chr{i}{p|q}`, in 1 to 22
 
@@
 
-```{r path_to_query}
+We will also use:
+
+- the samples metadata or clinical data reported in `r file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")`
+
+- the cytoband file to get the length of each chromosome arm (for plotting only)
+
+```{r input}
+# sample metadata
 sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
@@
 
+# the cytoband file that will be used to extract the size of each chromosome arm for plotting only
+arm_order_file <- file.path(module_base, "results", "references", "hg38_cytoBand.txt")
+
+```
+
+```{r path_to_query}
+# Extract from each Seurat object information that will be used
 sample_ids <- metadata |>
@@
         THY1 = FetchData(object = srat, vars = "ENSG00000154096", layer = "counts"),
-
-        # proportion of cnv per chromosome
-        proportion_cnv_chr1p = srat$proportion_cnv_chr1p,
-        proportion_cnv_chr2p = srat$proportion_cnv_chr2p,
-        proportion_cnv_chr3p = srat$proportion_cnv_chr3p,
-        proportion_cnv_chr4p = srat$proportion_cnv_chr4p,
-        proportion_cnv_chr5p = srat$proportion_cnv_chr5p,
-        proportion_cnv_chr6p = srat$proportion_cnv_chr6p,
-        proportion_cnv_chr7p = srat$proportion_cnv_chr7p,
-        proportion_cnv_chr8p = srat$proportion_cnv_chr8p,
-        proportion_cnv_chr9p = srat$proportion_cnv_chr9p,
-        proportion_cnv_chr10p = srat$proportion_cnv_chr10p,
-        proportion_cnv_chr11p = srat$proportion_cnv_chr11p,
-        proportion_cnv_chr12p = srat$proportion_cnv_chr12p,
-
-        proportion_cnv_chr16p = srat$proportion_cnv_chr16p,
-        proportion_cnv_chr17p = srat$proportion_cnv_chr17p,
-        proportion_cnv_chr18p = srat$proportion_cnv_chr18p,
-        proportion_cnv_chr19p = srat$proportion_cnv_chr19p,
-        proportion_cnv_chr20p = srat$proportion_cnv_chr20p,
-
-        proportion_cnv_chr1q = srat$proportion_cnv_chr1q,
-        proportion_cnv_chr2q = srat$proportion_cnv_chr2q,
-        proportion_cnv_chr3q = srat$proportion_cnv_chr3q,
-        proportion_cnv_chr4q = srat$proportion_cnv_chr4q,
-        proportion_cnv_chr5q = srat$proportion_cnv_chr5q,
-        proportion_cnv_chr6q = srat$proportion_cnv_chr6q,
-        proportion_cnv_chr7q = srat$proportion_cnv_chr7q,
-        proportion_cnv_chr8q = srat$proportion_cnv_chr8q,
-        proportion_cnv_chr9q = srat$proportion_cnv_chr9q,
-        proportion_cnv_chr10q = srat$proportion_cnv_chr10q,
-        proportion_cnv_chr11q = srat$proportion_cnv_chr11q,
-        proportion_cnv_chr12q = srat$proportion_cnv_chr12q,
-        proportion_cnv_chr13q = srat$proportion_cnv_chr13q,
-        proportion_cnv_chr14q = srat$proportion_cnv_chr14q,
-        proportion_cnv_chr15q = srat$proportion_cnv_chr15q,
-        proportion_cnv_chr16q = srat$proportion_cnv_chr16q,
-        proportion_cnv_chr17q = srat$proportion_cnv_chr17q,
-        proportion_cnv_chr18q = srat$proportion_cnv_chr18q,
-        proportion_cnv_chr19q = srat$proportion_cnv_chr19q,
-        proportion_cnv_chr20q = srat$proportion_cnv_chr20q,
-        proportion_cnv_chr21q = srat$proportion_cnv_chr21q,
-        proportion_cnv_chr22q = srat$proportion_cnv_chr22q,
 
@@
 
+```{r }
+# Add clinical metadata
+cell_type_df <- cell_type_df |>
+  left_join( metadata, by= c("sample_id" = "scpca_sample_id"))
+
+```
+
 ## Functions
@@
 do_Feature_mean <- function(df, group.by, feature) {
-  df <- df %>%
-    group_by(sample_id, !!sym(group.by)) %>%
+  df <- df |>
+    group_by(sample_id, !!sym(group.by)) |>
     summarise(m = mean(!!sym(feature)))
@@
 }
+```
+
+#### do_BoxPlot_mean
+
+`do_BoxPlot_mean` shows boxplot of the percentage of cells with CNV in a specific chromosome arm per clinical relevant group (metadata)
+
+- `df` is the name of the table containing metadata and feature expression (counts) per cells
+- `group.by` is the name of the metadata to group the cells
+- `comparison` is a list of factors in `group.by` for which we will compare the means using an unpaired Wilcoxon test
+
+```{r fig.width=20, fig.height=15, out.width='100%', results='asis'}
+
+do_BoxPlot_mean <- function(df, group.by, comparison){
+  tmp <- df |>
+    select(!!sym(group.by), sample_id, contains("has_loss_chr"), contains("has_dupli_chr")) |>
+    group_by(!!sym(group.by), sample_id) |>
+    summarise_all(.funs = mean) |>
+    pivot_longer(cols = -c(sample_id,!!sym(group.by)) ) |>
+    rename("chromosome_arm" = "name", "percentage" = "value") |>
+    mutate(chromosome_arm = factor(chromosome_arm, levels = c(has_dupli_chr, has_loss_chr)))
+
+  p <- ggplot(tmp, aes(x = !!sym(group.by), y = percentage, fill = !!sym(group.by))) +
+  geom_boxplot()+
+  scale_fill_brewer(palette="PuRd") +
+  geom_dotplot(binaxis='y', stackdir='center') +
+  facet_wrap(~chromosome_arm, nrow = 5 ) +
+  theme_classic()  +
+  stat_compare_means( method = "wilcox.test", paired = FALSE,  hide.ns = TRUE, label = "p.signif",  bracket.size = 0.5,tip.length = 0.01 ,comparison = list(comparison), size =8, label.y = 0.9)
+
+  return(p)
+}
+
 ```
@@
 
-- _normal_ cells can be observe in all five compartments (`normal`,`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV
+- _normal_ cells can be observe in all five compartments (`normal`,`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV.
 We only allow a bit of flexibility in terms of CNV profile for immune and endothelium cells that have a high predicted score.
+Indeed, we know that false positive CNV can be observed in a cell type specific manner.
 Please note that in `06_infercnv.R`, we renamed all `immune` and `endothelium` cells with a `fetal_kidney_predicted.compartment.score` > `r params$predicted.celltype.threshold` as `normal`.
-Indeed, we know that false positive CNV can be observed in a cell type specific manner.
 
@@
     compartment %in% c("fetal_nephron", "stroma") & 
-      compartment.score > params$predicted.celltype.threshold & 
       has_cnv_score == "no CNV" ~ "normal",
@@
 Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` _cancer_ cells, `r table(cell_type_df$first.level_annotation)["normal"]` _normal_ cells.
-Only `r table(cell_type_df$first.level_annotation)["unknown"]`cells remain _unknown_
+`r table(cell_type_df$first.level_annotation)["unknown"]`cells remain _unknown_
 
@@
 
-DT::datatable(table(cell_type_df$first.level_annotation, cell_type_df$sample_id))
-
 ```
@@
 We know that these _mesenchymal_ cells are cells from the cap mesenchyme that are not expected to be in a mature kidney.
-These blastema cells should express higher _CITED1_.
+These blastema cells should express higher _CITED1_ and/or _NCAM1_.
 
@@
     cell_type_df$compartment %in% c("fetal_nephron") &
-      cell_type_df$compartment.score > params$predicted.celltype.threshold &
       cell_type_df$has_cnv_score == "no CNV" ~ "kidney",
@@
     cell_type_df$compartment %in% c("stroma") &
-      cell_type_df$compartment.score > params$predicted.celltype.threshold &
       cell_type_df$has_cnv_score == "no CNV" ~ "normal stroma",
@@
 ```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
-tmp <- cell_type_df[cell_type_df$second.level_annotation != "unknown",]
-ggplot(tmp, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
-  geom_point(size = 1, shape = 19) +
-  facet_wrap(facets = ~sample_id, ncol = 5) +
-  theme_bw() +
-  theme(text = element_text(size = 22))
-```
-
-#### `Second.level_annotation` of cancer cells only - `umap` reduction
-
-```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
-ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
+cell_type_df_sub <- cell_type_df[cell_type_df$second.level_annotation != "unknown",]
+ggplot(cell_type_df_sub, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
   geom_point(size = 1, shape = 19) +
@@
 
-We look for each `second.level_annotation` and `sample_id`, the proportion of cells that has CNV in each of the chromosome.
+We look for each `second.level_annotation` and `sample_id` at the proportion of cells that has CNV in each of the chromosome.
+
+These heatmaps allow us to:
+
+- validate that mostly tumor cells do present high percentages of cells with CNV (by definition)
+
+- identify chromosome and cell types that might be more sensitive to _FALSE_ positive CNV, such as the _chr6p_ which is a hub for immune genes
+
+- get the repartition per sample of the CNV
 
@@
 for (i in 1:22) {
-  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "q")))
+  print(do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "q")))
 }
 ```
+
 #### gains on long arms
@@
 for (i in 1:22) {
-  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "q")))
+  print(do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "q")))
 }
@@
 for (i in c(1:12, 16:20)) {
-  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "p")))
+  print(do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "p")))
 }
@@
 for (i in c(1:12, 16:20)) {
-  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "p")))
+  print(do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "p")))
 }
@@
 
+### Mean CNV profile
+
+While the above heatmaps can be informative, we have to acknowledge that there are numerous and it is difficult to get a global picture of the CNV profile of the entire dataset.
+
+Here we aim to plot a mean CNV profile across the 35 annotated samples as presented in Figure 1A of [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+This represent the mean SNP Array profile of 64 pediatric kidney cancer (including 60 Wilms tumors).
+
+Here we plot for each chromosome arm (x-axis) the number of samples having a `dupli` (y-axis > 0, in red) or a `loss` (y-axis < 0, in blue).
+
+For a sample, a CNV is defined as `clonal` (opaque) if detected in more than 70% of the cells;
+and `subclonal` (transparent) when detected in 20 to 70% of the cells.
+
+```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
+
+# Here we define the vector of ordered chromosomes arms that will be used for plotting
+
+has_dupli_chr <- c("has_dupli_chr1p",                                      
+                   "has_dupli_chr1q",
+
+                   "has_dupli_chr2p",
+                   "has_dupli_chr2q",
+
+                   "has_dupli_chr3p",
+                   "has_dupli_chr3q",
+
+                   "has_dupli_chr4p",
+                   "has_dupli_chr4q",
+
+                   "has_dupli_chr5p",
+                   "has_dupli_chr5q",
+
+                   "has_dupli_chr6p",
+                   "has_dupli_chr6q",
+
+                   "has_dupli_chr7p",
+                   "has_dupli_chr7q",
+
+                   "has_dupli_chr8p",
+                   "has_dupli_chr8q",
+
+                   "has_dupli_chr9p",
+                   "has_dupli_chr9q",
+
+                   "has_dupli_chr10p",
+                   "has_dupli_chr10q",
+
+                   "has_dupli_chr11p",
+                   "has_dupli_chr11q",
+
+                   "has_dupli_chr12p",
+                   "has_dupli_chr12q",
+
+                   "has_dupli_chr13p",
+                   "has_dupli_chr13q",
+
+                   "has_dupli_chr14p",
+                   "has_dupli_chr14q",
+
+                   "has_dupli_chr15p",
+                   "has_dupli_chr15q",
+
+                   "has_dupli_chr16p",
+                   "has_dupli_chr16q",
+
+                   "has_dupli_chr17p",
+                   "has_dupli_chr17q",
+
+                   "has_dupli_chr18p",
+                   "has_dupli_chr18q",
+
+                   "has_dupli_chr19p",
+                   "has_dupli_chr19q",
+
+                   "has_dupli_chr20p",
+                   "has_dupli_chr20q",
+
+                   "has_dupli_chr21p",
+                   "has_dupli_chr21q",
+
+                   "has_dupli_chr22p",
+                   "has_dupli_chr22q"
+                   )
+
+#################################################################################
+
+# define the vector of ordered chromosomes arms##################################
+has_loss_chr <-  c("has_loss_chr1p",                                        
+                   "has_loss_chr1q",
+
+                   "has_loss_chr2p",
+                   "has_loss_chr2q",
+
+                   "has_loss_chr3p",
+                   "has_loss_chr3q",
+
+                   "has_loss_chr4p",
+                   "has_loss_chr4q",
+
+                   "has_loss_chr5p",
+                   "has_loss_chr5q",
+
+                   "has_loss_chr6p",
+                   "has_loss_chr6q",
+
+                   "has_loss_chr7p",
+                   "has_loss_chr7q",
+
+                   "has_loss_chr8p",
+                   "has_loss_chr8q",
+
+                   "has_loss_chr9p",
+                   "has_loss_chr9q",
+
+                   "has_loss_chr10p",
+                   "has_loss_chr10q",
+
+                   "has_loss_chr11p",
+                   "has_loss_chr11q",
+
+                   "has_loss_chr12p",
+                   "has_loss_chr12q",
+
+                   "has_loss_chr13p",
+                   "has_loss_chr13q",
+
+                   "has_loss_chr14p",
+                   "has_loss_chr14q",
+
+                   "has_loss_chr15p",
+                   "has_loss_chr15q",
+
+                   "has_loss_chr16p",
+                   "has_loss_chr16q",
+
+                   "has_loss_chr17p",
+                   "has_loss_chr17q",
+
+                   "has_loss_chr18p",
+                   "has_loss_chr18q",
+
+                   "has_loss_chr19p",
+                   "has_loss_chr19q",
+
+                   "has_loss_chr20p",
+                   "has_loss_chr20q",
+
+                   "has_loss_chr21p",
+                   "has_loss_chr21q",
+
+                   "has_loss_chr22p",
+                   "has_loss_chr22q"
+                   )
+
+#################################################################################
+# Here we define the size of each chromosome arm that will be used for plotting of the mean CNV profile
+
+  # Load cytoBand file
+  cytoBand <- read.table(gzfile(arm_order_file), header = FALSE, sep = "\t", stringsAsFactors = FALSE)
+  colnames(cytoBand) <- c("chrom", "start", "end", "band", "stain")
+
+  # Extract chromosome arm information
+  cytoBand <- cytoBand |>
+    mutate(arm = substr(band, 1, 1))  # Extract 'p' or 'q' from the band column
+
+  # Define arm boundaries
+  chromosome_arms <- cytoBand |>
+    group_by(chrom, arm) |>
+    summarize(
+      Start = min(start),
+      End = max(end),
+      .groups = "drop"
+    ) |>
+    mutate(Size = End - Start,
+           dupli = paste0("has_dupli_", chrom, arm),
+           loss = paste0("has_loss_", chrom, arm)) |> 
+  filter(arm != "")
+
+  chromosome_arms <- chromosome_arms |>
+    mutate(Size = Size / sum(Size))
+
+```
+
+```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
+
+# prepare the table for the number of sample with a (sub)clonal loss
+
+CNV_profile_loss <- cell_type_df_sub |>
+  select(sample_id, contains("has_loss_chr")) |>
+  group_by(sample_id) |>
+  summarise_all(.funs = mean) |>
+  pivot_longer(cols = -sample_id) |>
+  rename("chromosome_arm" = "name", "percentage" = "value") |>
+  left_join( chromosome_arms, by = c("chromosome_arm"="loss")) |>
+  mutate(chromosome_arm = factor(chromosome_arm, levels = has_loss_chr)) |>
+  mutate(absolute_70 = case_when((percentage > 0.7) ~ -1,
+         .default = 0),
+         absolute_10 = case_when((percentage > 0.1) ~ -1,
+         .default = 0))
+
+# prepare the table for the number of sample with a (sub)clonal gain
+
+CNV_profile_gain <- cell_type_df_sub |>
+  select(sample_id, contains("has_dupli_chr")) |>
+  group_by(sample_id) |>
+  summarise_all(.funs = mean) |>
+  pivot_longer(cols = -sample_id) |>
+  rename("chromosome_arm" = "name", "percentage" = "value") |>
+  left_join( chromosome_arms, by = c("chromosome_arm"="dupli")) |>
+  mutate(chromosome_arm = factor(chromosome_arm, levels = has_dupli_chr)) |>
+  mutate(absolute_70 = case_when((percentage > 0.7) ~ 1,
+         .default = 0),
+         absolute_10 = case_when((percentage > 0.1) ~ 1,
+         .default = 0))
+
+# plot
+
+loss_profile <- ggplot(CNV_profile_loss, aes(x = chromosome_arm, y = absolute_10, width = 20 * Size)) +
+  geom_bar(stat = "identity",  fill = "#41b6c4", alpha = 0.4) +
+  geom_bar(aes(x = chromosome_arm, y = absolute_70, width = 20 * Size), stat = "identity",  fill = "#41b6c4") +
+  scale_x_discrete(labels = gsub("has_loss_", "", levels(droplevels(CNV_profile_loss$chromosome_arm))))  +
+  theme_classic() +
+  ylab("Number of occurrence") +
+  xlab("chromosom arms") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12), text = element_text(size = 15))
+
+gain_profile <- ggplot(CNV_profile_gain, aes(chromosome_arm, y = absolute_10, width = 20* Size)) +
+  geom_bar(stat = "identity",  fill = "#ce1256", alpha = 0.4) +
+  geom_bar(aes(x = chromosome_arm, y = absolute_70, width = 20 * Size), stat = "identity",  fill = "#ce1256") +
+  theme_classic() +
+  theme(axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), 
+        text = element_text(size = 15)) +
+  ylab(" ")
+
+# combined plot
+wrap_plots(gain_profile, loss_profile, ncol = 1)
+
+```
+
+Comparing with the profile in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1), we identified important CNV known to play a role in Wilms tumor, including:
+
+- _chr8q_, _chr12_, _chr13_, _chr18q_ gains
+
+- _chr11q_, _chr16_, _chr17q_ loss
+
+However, we spotted some differences:
+
+- we couldn't detect any _chr4q_ loss as described in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+
+- we identified 2 losses in _chr5_ and _chr6_ that do not show up in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
+These can be `infercnv` induced _FALSE_positive CNV.
+
+Our major concern is the pattern of the _chr1_. 
+We do see a tendency of _chr1p_ loss and _chr1q_ gain but it is not as pronounced as in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
+_chr1q_ gain is however one of the most promising predictive marker of Wilms tumor outcome ([Nelson et al, Curr Opin Pediatr (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/)).
+It is however important to notice that the profile shown in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1) derived from SIOP samples that have been pre-treated with chemotherapy, while samples in the dataset are enriched in `upfront resection` samples that haven't been pre-treated, according to the COG protocol.
+So far, we do not know how the pre-operative chemotherapy impact the clonal selection of specific CNV.
+
+### Mean of the percentage of cells with CNV in a specific chromosome arm per clinical relevant group
+
+#### `vital_status`
+
+As described in [Nelson et al, Curr Opin Pediatr (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/), 
+[Phelps et al, Children (2018)](https://pmc.ncbi.nlm.nih.gov/articles/PMC6262554/) and 
+[Zheng et al, Front. Oncol (2023)](https://www.frontiersin.org/journals/oncology/articles/10.3389/fonc.2023.1137346/full), 
+LOH at _chr1p_, _chr11p_ and _chr16q_ as well as _chr1q_ gain are prognostic factors.
+
+Here we aim to see if we identify CNV associated with poorer prognosis.
+
+```{r fig.height=15, fig.width=20, message=FALSE, warning=FALSE, out.width='100%', results='asis'}
+
+do_BoxPlot_mean(df = cell_type_df_sub, group.by = "vital_status", comparison = c("Alive", "Expired") )
+
+```
+
+In our study, we validate the association of _chr1q_ gain, _chr11q_ loss and _chr14q_ loss with poorer prognosis.
+
+##### `treatment`
+
+As we don't know the effect of the treatment on the selection of cells with CNV, it is difficult to compare our CNV profile with the one in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+We those check the repartition of CNV in samples that have been pre-treated (i.e `Resection post chemotherapy`) with samples that have been removed surgically before any treatment (i.e `Upfront resection`).
+
+```{r fig.height=15, fig.width=20, message=FALSE, warning=FALSE, out.width='100%', results='asis'}
+
+do_BoxPlot_mean(df = cell_type_df_sub, group.by = "treatment", comparison = c("Upfront resection", "Resection post chemotherapy") )
+
+```
+
+Our analysis suggest that there is indeed an enrichment in cells with CNV after chemotherapy, 
+including a tendency of more _chr1q_ gain samples after chemotherapy. 
+This could explain some of the differences observed between our mean CNV profile and the one published in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+
 ### Validation of second level annotation using marker genes
+
+Finally, we aim to validate the `second level annotation` using marker genes.
 
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000081237")
+do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = "ENSG00000081237")
 ```
+
+Immune cells are well annotated based on _PTPRC_ expression.
 
@@
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000110799")
+do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = "ENSG00000110799")
 ```
 
-#### Stroma, `Vimentin` expression
+Endothelial cells are well annotated based on _VWF_ expression.
+
+#### Stroma
+
+(*) To the best of our knowledges, there is no Wilms tumor specific marker of stroma cells.
+They should express in some extend similar markers as normal stromal cells including _VIM_, _THY1_ and _COL6A3_.
+
+##### `Vimentin` expression
 
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000026025")
+do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = "ENSG00000026025")
 ```
 
-#### Stroma, `COL6A3` expression
+_VIM_ do not seem to be neither specific nor universal.
+
+##### `COL6A3` expression
 
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000163359")
+do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = "ENSG00000163359")
 ```
 
-#### Stroma, `THY1` expression
+_COL6A3_ expression is higher in Wilms tumor stroma cells and expressed across a wider range of samples.
+
+##### `THY1` expression
 
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000154096")
+do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = "ENSG00000154096")
 ```
 
-#### Blastema, `CITED1` expression
+_THY1_ do not seem to be neither specific nor universal.
+
+#### Blastema
+
+(*) To the best of our knowledges, there is no Wilms tumor specific and universel (i.e. expressed by all cells in every sample) marker of blastema cells.
+We expect however blastema cells to express higher levels of stemness markers such as _CITED1_, _SIX1/2_, _NCAM1_.
+
+##### `CITED1` expression
 
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000125931")
+do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = "ENSG00000125931")
 ```
 
-#### Blastema, `NCAM1` expression
+Blastema cells are enriched in _CITED1_+ cells, but some samples do not express it at all.
+
+##### `NCAM1` expression
 
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000149294")
+do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = "ENSG00000149294")
 ```
 
-#### stemness marker (blastema and primitive epithelium), `SIX2` expression
+_NCAM1_ seems to be more cancer specific and broadly expressed among samples.
+
+##### `SIX2` expression
 
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000170577")
+do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = "ENSG00000170577")
 ```
 
-#### Epithelium, `CDH1` expression
+#### Epithelium
+
+##### `CDH1` expression
 
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000039068")
+do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = "ENSG00000039068")
 ```
 
-#### Epithelium, `PODXL` expression
+As expected, epithelial cells express higher expression of _CDH1_.
+
+##### `PODXL` expression
 
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000128567")
+do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = "ENSG00000128567")
 ```
@@
 
-- Combining label transfer and CNV inference we have produced draft annotations for all 40 Wilms tumor samples in SCPCP000006
+- Combining label transfer and CNV inference we have produced draft annotations for 35/40 Wilms tumor samples in SCPCP000006.
+We decided not to annotate samples for which the `infercnv` couldn't be run with a normal reference, as we don't trust the results in that case and want to avoid misclassification (`SCPCS000177`, `SCPCS000180`, `SCPCS000181`, `SCPCS000190`, `SCPCS000197`).
+
+- The mean CNV profile of the Wilms tumor cohort seems reasonable in regards to the one published in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
+We identify some commonalities but also some divergence. 
+They can be due to technical limitation of our CNV inference but can also reflect some specificities of this cohort (e.g. pre/post treatment samples).
+It would be really useful to compare for some sample the `infercnv` result to a ground truth such as SNP Array (to be requested).
 

### 2025-02-10 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
Apply suggestions from code review
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
 
-This notebook looks at cell type annotations and gene set expression across all samples in `SCPCP00015` and assigns "final" annotations. 
+This notebook looks at cell type annotations and gene set expression across all samples in `SCPCP000015` and assigns "final" annotations. 
 To do this we are using the merged, but not batch-corrected, object containing the gene expression data for all samples.
@@
 - Consensus cell type annotations output by running [`assign-consensus-celltypes.sh` in `cell-type-consensus`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-consensus/assign-consensus-celltypes.sh).
-- AUC values as calculated by `AUCell` for a set of Ewing sarcoma specific gene sets in MSigDB output by `run-aucell-ews-signatures.sh`.
+- AUC values as calculated by `AUCell` for a set of Ewing sarcoma specific gene sets in MSigDB output by [`run-aucell-ews-signatures.sh`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/run-aucell-ews-signatures.sh).
 
@@
 - The `wrenn-nt5e-genes` is a list of genes found to be upregulated in `EWS-FLI1` low tumor cells/ cancer associated fibroblasts.
-- The `gobp_ECM` and `hallmark_EMT` are gene sets that are hypothesized to be upregulated in `EWS-FLI1` low tumor cells/ cancer associated fibroblasts.
+- The `gobp_ECM` and `hallmark_EMT` are gene sets that are hypothesized to be upregulated in `EWS-FLI1` low tumor cells/cancer associated fibroblasts.
 
@@
 
-The last tumor cell state that we want to identify is if any cells can be considered proliferative using the set of proliferative markers in `tumor-cell-state-markers.tsv`. 
+The last tumor cell state that we want to identify is if any cells can be considered proliferative using the set of proliferative markers in [`tumor-cell-state-markers.tsv`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/references/tumor-cell-state-markers.tsv).
 Let's look at the mean expression of these markers across all cells. 
@@
 These plots will look specifically at expression of a set of key marker genes that we expect to up in the assigned cell types. 
-These markers are found in `references/combined-markers.tsv` and include a smaller set of markers for each of the cell types we have identified.
+These markers are found in [`references/combined-validation-markers.tsv`](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-ewings/references/combined-validation-markers.tsv) and include a smaller set of markers for each of the cell types we have identified.
 

### 2025-02-10 analyses/cell-type-consensus/exploratory-notebooks/03-osteosarcoma-consensus-celltypes.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
     barchart <- barchart +
-      facet_wrap(vars(!!sym(facet_variable)), scales ="free")
+      facet_wrap(vars(!!sym(facet_variable)), scales ="free_x")
   }
@@
 metadata_files <- list.files(data_dir, pattern = "single_cell_metadata\\.tsv$", full.names = TRUE, recursive = TRUE)
-metadata_ids <- stringr::word(dirname(metadata_files), -1, sep = "/")
+metadata_ids <- fs::path_file(dirname(metadata_files))
 names(metadata_files) <- metadata_ids
@@
   dplyr::summarize(
-    total_percent_annotated = sum(total_cells_per_annotation)/total_cells_per_library
+    total_percent_annotated = sum(total_cells_per_annotation)/unique(total_cells_per_library)
   ) |>
-  unique() |> 
   dplyr::arrange(desc(total_percent_annotated)) |> 
@@
 
-It looks like there's definitely some variation between distributions of cell types within the osteo samples. 
+It looks like there's definitely some variation in the distributions of cell types among the osteo samples. 
 Generally, we see most annotated cells are smooth muscle cells and endothelial cells. 
@@
 all_results_df <- all_results_df |> 
-  #dplyr::left_join(immune_df, by = c("consensus_annotation", "consensus_ontology")) |> 
   dplyr::mutate(
@@
   dplyr::filter(immune_category == "immune") |> 
+  dplyr::mutate(annotation_percent_immune = total_cells_per_annotation/total_cells_per_library)
   dplyr::group_by(library_id) |> 
   dplyr::summarize(
-    percent_immune = sum(total_cells_per_annotation)/total_cells_per_library
-  ) |>
-  dplyr::select(library_id, percent_immune) |> 
-  unique()
+    percent_immune = sum(annotation_percent_immune)
+  )
 
@@
 }) |>
-  patchwork::wrap_plots(ncol = 1, guides = "collect")
+  patchwork::wrap_plots(ncol = 1, guides = "collect") + 
+    patchwork::plot_annotation(tag_levels = list(metadata_categories))
+
 ```

### 2025-02-10 analyses/cell-type-consensus/exploratory-notebooks/utils/setup-functions.R

@@ -6,12 +6,13 @@ summarize_celltypes <- function(file, id){
   df <- readr::read_tsv(file) 
   
   # get total cell count and number of assigned cell types per library
-  total_cells_df <- df |> 
+  df <- df |> 
     dplyr::group_by(library_id) |> 
-    dplyr::summarize(
-      total_cells_per_library = length(library_id),
+    dplyr::mutate(
+      total_cells_per_library = dplyr::n(),
       num_celltypes = length(unique(consensus_annotation))
-    )
+    ) |>
+    dplyr::ungroup()
   
   summary_df <- df |> 
     dplyr::group_by(library_id, sample_type, consensus_annotation, consensus_ontology) |> 

### 2025-02-10 analyses/cell-type-consensus/exploratory-notebooks/03-osteosarcoma-consensus-celltypes.Rmd
clean up based on review comments and remove hot/cold

@@
     df, 
-    fill_color, 
-    facet_variable = NULL,
-    colors
+    fill_column,
+    celltype_colors, # named vector where names match the values in fill_column
+    facet_variable = NULL
 ){
@@
       y = percent_cells_annotation, 
-      fill = !!sym(fill_color)
+      fill = !!sym(fill_column)
     ) +
@@
     scale_y_continuous(expand = c(0,0)) +
-    scale_fill_manual(values = colors) +
+    scale_fill_manual(values = celltype_colors) +
     theme(axis.text.x = element_blank()) +
@@
 
-  ggplot(immune_plot_df, aes(x = !!sym(plot_column), y = percent_immune)) +
+  ggplot(df, aes(x = !!sym(plot_column), y = percent_immune)) +
     ggforce::geom_sina() +
@@
 ```{r}
-stacked_barchart(total_order_df, fill_color = "top_celltypes", colors = all_celltype_colors)
+stacked_barchart(total_order_df, fill_column = "top_celltypes", celltype_colors = all_celltype_colors)
 ```
@@
 ```{r}
-stacked_barchart(total_order_df, fill_color = "top_celltypes", facet_variable = "project_id", colors = all_celltype_colors)
+stacked_barchart(total_order_df, fill_column = "top_celltypes", facet_variable = "project_id", celltype_colors = all_celltype_colors)
 ```
@@
   dplyr::filter(immune_category == "immune") |> 
-  dplyr::mutate(annotation_percent_immune = total_cells_per_annotation/total_cells_per_library)
+  dplyr::mutate(annotation_percent_immune = total_cells_per_annotation/total_cells_per_library) |>
   dplyr::group_by(library_id) |> 
@@
 
-stacked_barchart(immune_results_df, fill_color = "immune_category", colors = three_color_scheme)
+stacked_barchart(immune_results_df, fill_column = "immune_category", celltype_colors = three_color_scheme)
 ```
@@
 ```{r}
-stacked_barchart(immune_results_df, fill_color = "immune_category", facet_variable = "project_id", colors = three_color_scheme)
+stacked_barchart(immune_results_df, fill_column = "immune_category", facet_variable = "project_id", celltype_colors = three_color_scheme)
 ```
@@
 It does appear that libraries that have more immune cell composition also have a higher percentage of non-immune cell types which could very well be a technical artifact and related to sample prep.
-
-Just for visualization purposes, let's classify these as "hot" and "cold", where "hot" tumors have an immune composition > 5%.
-
-```{r}
-immune_results_df <- immune_results_df |> 
-  dplyr::mutate(
-    # classify hot/cold based on immune percent > 5 %
-    hot_cold = dplyr::if_else(percent_immune > .05, "hot", "cold")
-  )
-```
-
-```{r}
-stacked_barchart(immune_results_df, fill_color = "immune_category", facet_variable = "hot_cold", colors = three_color_scheme)
-```
-
-Our threshold is pretty arbitrary, but I think there are definitely some "cold" tumors where we see little to no immune involvement.
 
@@
 
-stacked_barchart(immune_results_df, fill_color = "is_immune", colors = all_celltype_colors)
+stacked_barchart(immune_results_df, fill_column = "is_immune", celltype_colors = all_celltype_colors)
 ```
@@
 # do some data wrangling on the metadata columns we want to look at 
-immune_results_df <- immune_results_df |> 
+immune_metadata_df <- immune_results_df |> 
   dplyr::mutate(
-    primary_or_metastasis = dplyr::if_else(!is.na(primary_or_metastasis), primary_or_metastasis, disease_timing),
-    disease_timing_mod = dplyr::if_else(disease_timing %in% c("Initial diagnosis", "Recurrence"), disease_timing, "other"),
+    primary_or_metastasis = dplyr::if_else((is.na(primary_or_metastasis) &
+                                             disease_timing != "Recurrence"), disease_timing, primary_or_metastasis) |>
+      forcats::fct_relevel(c("Primary", "Metastasis")),
+    disease_timing_mod = dplyr::if_else(disease_timing %in% c("Initial diagnosis", "Recurrence"), disease_timing, "other") |> 
+      forcats::fct_relevel("other", after = Inf),
     tissue_location_mod = dplyr::case_when(
-      stringr::str_detect(tissue_location, "Bone|femur|Humerus|Tibia") ~ "Bone",
+      stringr::str_detect(tissue_location, "Bone|femur|Humerus|Tibia") ~ "bone",
       #stringr::str_detect(tissue_location, "femur") ~ "Bone",
@@
 
+    df <- immune_metadata_df |> 
+      dplyr::select(library_id, percent_cells_annotation, immune_category, {{category}}) |> 
+      tidyr::drop_na() #remove any samples that are missing that metadata
+
     stacked_barchart(
-      immune_results_df,
-      fill_color = "immune_category",
+      df,
+      fill_column = "immune_category",
       facet_variable = category,
-      colors = three_color_scheme
+      celltype_colors = three_color_scheme
     )
@@
 
+    df <- immune_metadata_df |> 
+      dplyr::select(library_id, percent_cells_annotation, top_celltypes, {{category}}) |> 
+      tidyr::drop_na() #remove any samples that are missing that metadata
+
     stacked_barchart(
-      immune_results_df,
-      fill_color = "top_celltypes",
+      df,
+      fill_column = "top_celltypes",
       facet_variable = category,
-      colors = all_celltype_colors
+      celltype_colors = all_celltype_colors
     )
@@
 ```{r}
-immune_plot_df <- immune_results_df |> 
+immune_plot_df <- immune_metadata_df |> 
   # just look at immune population across different diagnoses 
@@
   purrr::map(\(category){
-    sina_plot(immune_plot_df, category)
+
+    df <- immune_plot_df |> 
+      dplyr::select(library_id, percent_immune, {{category}}) |> 
+      tidyr::drop_na() #remove any samples that are missing that metadata
+
+    sina_plot(df, category)
   })

### 2025-02-10 analyses/cell-type-consensus/exploratory-notebooks/utils/setup-functions.R

@@ -16,8 +16,8 @@ summarize_celltypes <- function(file, id){
   
   summary_df <- df |> 
     dplyr::group_by(library_id, sample_type, consensus_annotation, consensus_ontology) |> 
-    dplyr::summarize(total_cells_per_annotation = length(consensus_annotation)) |>
-    dplyr::left_join(total_cells_df, by = "library_id") |> 
+    dplyr::summarize(total_cells_per_annotation = dplyr::n(),
+                     total_cells_per_library = unique(total_cells_per_library)) |>
     dplyr::mutate(
       # add percentage 
       percent_cells_annotation = round((total_cells_per_annotation / total_cells_per_library) * 100, 2)

### 2025-02-11 analyses/cell-type-consensus/exploratory-notebooks/03-osteosarcoma-consensus-celltypes.Rmd

@@
       celltype_colors = three_color_scheme
-    )
+    ) +
+      labs(title = category)
 }) |>
-  patchwork::wrap_plots(ncol = 1, guides = "collect") + 
-    patchwork::plot_annotation(tag_levels = list(metadata_categories))
+  patchwork::wrap_plots(ncol = 1, guides = "collect")
 
@@
       celltype_colors = all_celltype_colors
-    )
+    ) +
+      labs(title = category)
   }) |>

### 2025-02-11 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd

@@
 - Consensus cell type annotations output by running [`assign-consensus-celltypes.sh` in `cell-type-consensus`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-consensus/assign-consensus-celltypes.sh).
-- AUC values as calculated by `AUCell` for a set of Ewing sarcoma specific gene sets in MSigDB output by [`run-aucell-ews-signatures.sh`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/run-aucell-ews-signatures.sh).
+- AUC values as calculated by `AUCell` for a set of Ewing sarcoma specific gene sets from MSigDB output by [`run-aucell-ews-signatures.sh`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/run-aucell-ews-signatures.sh).
+- Custom gene sets from [`visser-all-marker-genes.tsv`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/references/visser-all-marker-genes.tsv), [`tumor-cell-state-markers.tsv`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/references/tumor-cell-state-markers.tsv), and [`combined-validation-markers.tsv`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/references/combined-validation-markers.tsv).
 
@@
 Any cells that are unable to be labeled via consensus cell types are labeled as "Unknown" and I expect these will line up with cells labeled as tumor cells by `SingleR`.  
+Note that only cell types from `SingleR` with > 50 cells are shown in the heatmap.
 
@@
 # get columns with mean expression of custom gene sets 
-mean_exp_columns <- colnames(all_info_df)[which(endsWith(colnames(all_info_df), "_mean"))]
+mean_exp_columns <- colnames(all_info_df) |> 
+  stringr::str_subset("_mean$")
+
 # make sure order is defined by cell type order
@@
   unique() |> # get rid of duplicate cells
-  dplyr::group_by(final_annotation) |> 
-  dplyr::summarize(
-    total_libraries = length(library_id)) |> 
-  dplyr::arrange(desc(total_libraries))
+  dplyr::add_count(final_annotation, name = "total_libraries") |>
+  dplyr::select(final_annotation, total_libraries)
 
@@
   dplyr::left_join(total_libraries, by = "final_annotation") |> 
-  dplyr::arrange(desc(total_cells))
+  dplyr::arrange(desc(total_cells)) |>
+  unique()
 ```
@@
 
+Note that only genes with mean expression > 0 and expressed in at least 10% of cells are included in the dotplot. 
+All genes are included in the heatmap.
+
 ```{r}
@@
 gene_cts <- logcounts(sce[genes, ]) |>
-  as.matrix() |>
   t() |> 
@@
 ```
+
+The mean expression in this plot is capped at 2.5 for coloring purposes, but some values may be higher.
 

### 2025-02-11 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
remove unused template and move misplaced utils functions

@@
 # source in setup functions prep_results()
-setup_functions <- file.path(module_base, "template_notebooks", "utils", "setup-functions.R")
+setup_functions <- file.path(module_base, "scripts", "utils", "setup-functions.R")
 source(setup_functions)
@@
 # expression_umap(), cluster_density_plot(), and annotated_exp_heatmap()
-plotting_functions <- file.path(module_base, "template_notebooks", "utils", "plotting-functions.R")
+plotting_functions <- file.path(module_base, "scripts", "utils", "plotting-functions.R")
 source(plotting_functions)

### 2025-02-11 analyses/cell-type-consensus/exploratory-notebooks/03-osteosarcoma-consensus-celltypes.Rmd
move plotting functions to utils

@@
 )
-```
-
-## Functions
-
-```{r}
-# barchart with or without faceting
-# each bar is a stacked barchart using the fill_color
-# faceting is only done if a facet_variable is provided
-stacked_barchart <- function(
-    df, 
-    fill_column,
-    celltype_colors, # named vector where names match the values in fill_column
-    facet_variable = NULL
-){
-
-  barchart <- ggplot(df) + 
-    aes(
-      x = library_id, 
-      y = percent_cells_annotation, 
-      fill = !!sym(fill_column)
-    ) +
-    geom_col() + 
-    scale_y_continuous(expand = c(0,0)) +
-    scale_fill_manual(values = celltype_colors) +
-    theme(axis.text.x = element_blank()) +
-    labs(
-      fill= "cell type"
-    )
-
-  if(!is.null(facet_variable)){
-    barchart <- barchart +
-      facet_wrap(vars(!!sym(facet_variable)), scales ="free_x")
-  }
-
-  return(barchart)
-}
-
-# sina plot looking at immune percentage on the y-axis 
-sina_plot <- function(df, plot_column){
-
-  ggplot(df, aes(x = !!sym(plot_column), y = percent_immune)) +
-    ggforce::geom_sina() +
-    stat_summary(fun.y=median, geom="crossbar" , color = "red", linewidth = 0.2) +
-    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
-          plot.margin = margin(10,10,10,10)) +
-    labs(
-      x = "", 
-      y = "Percent of cells annotated as Immune",
-      title = plot_column
-    )
-
-}
 ```
@@
 source(setup_functions)
+
+# source stacked_barchart() and sina_plot() functions
+plotting_functions <- file.path(module_base, "exploratory-notebooks", "utils", "plotting-functions.R")
+source(plotting_functions)
 ```

### 2025-02-11 analyses/cell-type-consensus/exploratory-notebooks/04-brain-and-cns-consensus-celltypes.Rmd
notebook looking at brain/ cns samples

---
title: "Consensus cell types across samples from brain and CNS tumors"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

This notebook summarizes the consensus cell types identified in all samples from brain and CNS tumors that are part of ScPCA. 
This includes all samples from `SCPCP000001`, `SCPCP000002`, `SCPCP000010`, `SCPCP000021`, and all non-multiplexed samples from `SCPCP000009`.

We are interested in looking at any differences in cell composition across samples, with a particular interest in looking at the immune cell populations.

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)
```

## Data setup

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)
module_base <- file.path(repository_base, "analyses", "cell-type-consensus")

# results directory with cell-type-consensus 
results_dir <- file.path(module_base, "results", "cell-type-consensus")

# data directory where project metadata files live
data_dir <- file.path(repository_base, "data", "current")
```

```{r}
# list all results files 
results_files <- list.files(results_dir, pattern = "_consensus-cell-types\\.tsv.\\gz$", full.names = TRUE)

# get project ids from file list and assign as names
project_ids <- stringr::str_remove(basename(results_files), "_consensus-cell-types.tsv.gz")
names(results_files) <- project_ids

# grab only brain/cns projects
brain_project_ids <- c("SCPCP000001", "SCPCP000002", "SCPCP000010", "SCPCP000021", "SCPCP000009")
results_files <- results_files[brain_project_ids]

# list project metadata files and grab only osteo metadata 
metadata_files <- list.files(data_dir, pattern = "single_cell_metadata\\.tsv$", full.names = TRUE, recursive = TRUE)
metadata_ids <- fs::path_file(dirname(metadata_files))
names(metadata_files) <- metadata_ids
brain_metadata_files <- metadata_files[brain_project_ids]

# list of all immune cell types 
consensus_immune_file <- file.path(module_base, "references", "consensus-immune-cell-types.tsv")
```

```{r}
# source summarize_celltypes() function
setup_functions <- file.path(module_base, "exploratory-notebooks", "utils", "setup-functions.R")
source(setup_functions)

# source stacked_barchart() and sina_plot() functions
plotting_functions <- file.path(module_base, "exploratory-notebooks", "utils", "plotting-functions.R")
source(plotting_functions)
```

```{r, message=FALSE}
# get immune cell types 
immune_df <- readr::read_tsv(consensus_immune_file)
immune_types <- immune_df$consensus_annotation

# read in metadata
all_metadata_df <- brain_metadata_files |>
  purrr::map(\(file) readr::read_tsv(file, col_types = list(.default = "c"))) |>
  dplyr::bind_rows() |>
  # select columns that might be useful 
  dplyr::select(
    project_id = scpca_project_id,
    library_id = scpca_library_id,
    diagnosis,
    subdiagnosis,
    disease_timing,
    age,
    sex,
    tissue_location,
    seq_unit
  )

# define multiplexed libraries from SCPCP000009
non_multiplexed_libs <- c("SCPCL000531", "SCPCL000569", "SCPCL000570", "SCPCL000571")
multiplexed_libs <- all_metadata_df |> 
  dplyr::filter(project_id == "SCPCP000009",
                !(library_id %in% non_multiplexed_libs)) |> 
  dplyr::pull(library_id) |>
  unique()

# read in results and prep data frame for plotting 
all_results_df <- results_files |> 
  purrr::imap(summarize_celltypes) |> 
  dplyr::bind_rows(.id = "project_id") |> 
  # remove multiplexed libraries 
  dplyr::filter(!(library_id %in% multiplexed_libs)) |> 
  # join with sample metadata
  dplyr::left_join(all_metadata_df, by = c("project_id", "library_id"))
```

```{r}
# assign a color scheme for all bar charts

# get all possible cell types 
# there are exactly 26 of them so we will use alphabet with two greys for unknown/all remaining 
unique_celltypes <- all_results_df |> 
  dplyr::filter(percent_cells_annotation > 0.1, # we're only ever going to plot cell types that are greater than 1%
                !consensus_annotation == "Unknown") |> 
  dplyr::pull(consensus_annotation) |> 
  unique() |>
  as.character()

# define colors for all cell types
all_celltype_colors <- c(
  palette.colors(palette = "alphabet"),
  "grey60", # all remaining
  "grey95", # unknown
  "grey60" # non-immune, use the same as all remaining since they are never in the same plot
)
names(all_celltype_colors) <- c(unique_celltypes, "All remaining cell types", "Unknown", "non-immune")

# define a three color scheme for immune, other, and unknown
three_color_scheme <- c(
  "immune" = "navy",
  "non-immune" = "grey60",
  "Unknown" = "grey95"
)
```

## Composition of top cell types

Here we look at the top cell types identified across all samples. 
We have already looked at this in `02-explore-consensus-results.Rmd`, but here we will plot all samples from all brain and CNS projects ordered by the total percent of cells labeled with a consensus label.

```{r}
# add column of "top cell types" for easier plotting 
all_results_df <- all_results_df |> 
  dplyr::mutate(
    # get most frequently observed cell types across libraries in that project 
    top_celltypes = forcats::fct_lump_n(consensus_annotation, 15, other_level = "All remaining cell types", ties.method = "first") |> 
      # sort by frequency 
      forcats::fct_infreq() |> 
      # make sure all remaining and unknown are last, use this to assign colors in specific order
      forcats::fct_relevel("All remaining cell types", "Unknown", after = Inf),
  )
```

```{r}
# get a vector of library ids ordered by total percentage annotated
library_levels <- all_results_df |> 
  dplyr::filter(consensus_annotation != "Unknown") |> 
  dplyr::group_by(library_id) |> 
  dplyr::summarize(
    total_percent_annotated = sum(total_cells_per_annotation)/unique(total_cells_per_library)
  ) |>
  dplyr::arrange(desc(total_percent_annotated)) |> 
  dplyr::pull(library_id)

# append any libraries that have 0 cells annotated
non_annotated_libs <- setdiff(unique(all_results_df$library_id), library_levels)
library_levels <- c(library_levels, non_annotated_libs)

# reorder by total percentage annotated 
all_results_df <- all_results_df |> 
  dplyr::mutate(
    library_id = forcats::fct_relevel(library_id, library_levels)
  ) |>
  unique()
```

```{r}
stacked_barchart(all_results_df, fill_column = "top_celltypes", celltype_colors = all_celltype_colors)
```

Below we show the same data but faceted by project.

```{r}
stacked_barchart(all_results_df, fill_column = "top_celltypes", facet_variable = "project_id", celltype_colors = all_celltype_colors)
```

- Macrophages seem to be the dominant cell type across most samples. 
- There appear to be some samples that have a higher proportion of T cells, but this is not project specific. 
- As with the osteosarcoma samples there is some general heterogeneity both in the number of assigned cells and the cell types across all samples.

We hypothesize that there will be differences in the cell types observed based on the specific type of brain or CNS tumor. 
Let's look at what tumor types are represented here and then use that to assign diagnosis groups so we can separate samples by diagnosis in the later plots.

```{r}
all_results_df |>
  dplyr::select(library_id, diagnosis) |> 
  unique() |> 
  dplyr::count(diagnosis, name = "number_of_samples") |>
  dplyr::arrange(desc(number_of_samples))
```

Based on this let's make the following groups:

- Low-grade glioma: Any samples labeled with "Low-grade glioma" or are part of `SCPCP000002` which are categorized as LGG tumors. 
- High-grade glioma: Any samples labeled with "High-grade glioma" or are part of `SCPCP000001` which are categorized as HGG tumors. 
- Glioblastoma
- Pilocytic astrocytoma
- Ganglioglioma 
- ATRT 
- Other: All samples that do not fall into the above categories.

We also have one "Non-cancerous" sample that we will keep separate.

```{r}
# assign diagnosis groups to use for faceting plots 
all_results_df <- all_results_df |> 
  dplyr::mutate(
    grouped_diagnosis = dplyr::case_when(
      diagnosis %in% c("Low-grade glioma", "Glioblastoma", "Pilocytic astrocytoma", "Ganglioglioma", "High-grade glioma", "Atypical teratoid rhabdoid tumor", "Non-cancerous") ~ diagnosis,
      project_id == "SCPCP000001" ~ "High-grade glioma",
      project_id == "SCPCP000002" ~ "Low-grade glioma",
      .default =  "other"
    ) |>
      forcats::fct_relevel(c("other", "Non-cancerous"), after = Inf)
  )
```

Before we show the plots, let's look at what samples are being grouped as "other".

```{r}
all_results_df |> 
  dplyr::filter(grouped_diagnosis == "other") |> 
  dplyr::select(library_id, project_id, diagnosis, grouped_diagnosis) |> 
  unique()
```

```{r}
stacked_barchart(all_results_df, fill_column = "top_celltypes", facet_variable = "grouped_diagnosis", celltype_colors = all_celltype_colors) +
  theme(strip.text = element_text(size = 6)) 
```

It looks like we do some qualitative differences between diagnosis groups! 
In particular, Glioblastoma seems to have the least amount of macrophages and favors T cells instead. 
ATRT also looks to generally have a lower amount of classified cells compared to other diagnoses.

Also a side note that there are a lot of "Unknown" cells in the "Non-cancerous" sample. 
I would have expected cell types there to be classified since we don't have tumor, but this also might be a limit of the reference, we don't have a lot of brain specific cell types outside of neuron and astrocyte.

## Immune cell populations

Let's look specifically at immune cell populations. 
To do this we will lump all immune cells as "immune", all non-immune as "non-immune", and all unknown as "unknown". 
Here we will sort by total percentage of immune cells.

```{r}
# look at immune cell types vs. unknown vs. other 
all_results_df <- all_results_df |> 
  dplyr::mutate(
    # first get a column that is just immune, unknown, or other
    immune_category = dplyr::case_when(
      consensus_annotation %in% immune_types ~ "immune",
      consensus_annotation == "Unknown" ~ "Unknown",
      .default = "non-immune"
    ),
    library_id = as.character(library_id) #remove existing levels 
  )
```

```{r}
# now sort just by immune percentage
immune_pct_df <- all_results_df |> 
  dplyr::filter(immune_category == "immune") |> 
  dplyr::mutate(annotation_percent_immune = total_cells_per_annotation/total_cells_per_library) |>
  dplyr::group_by(library_id) |> 
  dplyr::summarize(
    percent_immune = sum(annotation_percent_immune)
  )

library_levels <- immune_pct_df |> 
  dplyr::arrange(desc(percent_immune)) |> 
  dplyr::pull(library_id)

# append any libraries that have 0 immune cells annotated
non_annotated_libs <- setdiff(unique(all_results_df$library_id), library_levels)
library_levels <- c(library_levels, non_annotated_libs)

immune_results_df <- all_results_df |> 
  dplyr::left_join(immune_pct_df, by = c("library_id")) |> 
  dplyr::mutate(
    percent_immune = dplyr::if_else(is.na(percent_immune), 0, percent_immune),
    library_id = forcats::fct_relevel(library_id, library_levels)
  )

stacked_barchart(immune_results_df, fill_column = "immune_category", celltype_colors = three_color_scheme)
```
Let's see how the percentage of immune cells compares across projects and diagnosis.

```{r}
stacked_barchart(immune_results_df, fill_column = "immune_category", facet_variable = "project_id", celltype_colors = three_color_scheme)
```

```{r}
stacked_barchart(immune_results_df, fill_column = "immune_category", facet_variable = "grouped_diagnosis", celltype_colors = three_color_scheme)
```

This isn't super interesting to me since most classified cells are immune cells and there doesn't seem to be any major differences across disease types outside of ATRT having a lower percentage of classified cells than other projects.

## Macrophage vs. T cell populations

It looks like there may be some differences between T cell and macrophage infiltrate based on the tumor type. 
Here we will look specifically at all T cells and anything in the macrophage/monocyte lineage. 
I'm including any cell types that are in our list of possible cell types and a child of [mononuclear phagocyte](https://www.ebi.ac.uk/ols4/ontologies/cl/classes/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FCL_0000113?lang=en) in the macro/mono group. 
We should also note that "macrophage" here probably refers to microglia instead.

```{r}
# t cell subtypes
t_cells <- unique_celltypes |> 
  stringr::str_subset("T")

# macro/mono lineage 
macro_mono_cells <- c(
  "macrophage", 
  "monocyte",
  "mononuclear phagocyte", 
  "dendritic cell"
)

# create broad category for T cell vs. macro/mono 
immune_results_df <- immune_results_df |> 
  dplyr::mutate(
    immune_broad_category = dplyr::case_when(
      consensus_annotation %in% t_cells ~ "T cell", 
      consensus_annotation %in% macro_mono_cells ~ "Macro/mono",
      consensus_annotation == "Unknown" ~ "Unknown",
      .default = "All remaining cell types"
    )
  )

# define colors 
immune_category_colors <- c(
  "T cell" = all_celltype_colors[["T cell"]],
  "Macro/mono" = all_celltype_colors[["macrophage"]],
  "All remaining cell types" = "grey60",
  "Unknown" = "grey90"
)
```

```{r}
stacked_barchart(immune_results_df, fill_column = "immune_broad_category", facet_variable = "grouped_diagnosis", celltype_colors = immune_category_colors) +
  theme(strip.text = element_text(size = 6)) 
```

- As expected based on the previous plot, glioblastoma appears to have greater T cell infiltrate than macrophages across all samples. 
- There appears to be a handful of samples in the "High-grade glioma" group that also have high T cell infiltrate. 
Notably, glioblastoma is a type of HGG. 
- Generally, the T cell population seems lowest in LGG. 
Ganglioglioma also has fewer samples with high percentage of T cell infiltrate and this is also a LGG. 
- It looks like about 50% of the pilocytic astrocytomas have more T cell infiltrate than macrophages.

Finally, let's group by either HGG or LGG.

```{r}
# assign diagnosis groups to use for faceting plots 
hgg_lgg_df <- immune_results_df |> 
  dplyr::mutate(
    hgg_vs_lgg = dplyr::case_when(
      grouped_diagnosis %in% c("Low-grade glioma", "Pilocytic astrocytoma", "Ganglioglioma") ~ "Low-grade glioma", 
      grouped_diagnosis %in% c("High-grade glioma", "Atypical teratoid rhabdoid tumor", "Glioblastoma") ~ "High-grade glioma",
      grouped_diagnosis == "Non-cancerous" ~ "Non-cancerous",
      .default =  "other"
    ) 
  ) |> 
  dplyr::filter(
    hgg_vs_lgg %in% c("High-grade glioma", "Low-grade glioma")
  )
```

```{r}
stacked_barchart(hgg_lgg_df, fill_column = "immune_broad_category", facet_variable = "hgg_vs_lgg", celltype_colors = immune_category_colors)
```

Consistent with the previous findings it looks like HGG tend to have higher T cell infiltrate while LGG tend to have higher macrophage infiltrate.

## Session info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-02-12 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
remove `ggpubr` dependencies and `stat_compare_mean()`

@@
 library(DT)
-library(ggpubr)
 ```
@@
   facet_wrap(~chromosome_arm, nrow = 5 ) +
-  theme_classic()  +
-  stat_compare_means( method = "wilcox.test", paired = FALSE,  hide.ns = TRUE, label = "p.signif",  bracket.size = 0.5,tip.length = 0.01 ,comparison = list(comparison), size =8, label.y = 0.9)
+  theme_classic()
 

### 2025-02-12 analyses/cell-type-consensus/exploratory-notebooks/04-brain-and-cns-consensus-celltypes.Rmd
adjust HGG and myeloid groupings

@@
 - Low-grade glioma: Any samples labeled with "Low-grade glioma" or are part of `SCPCP000002` which are categorized as LGG tumors. 
-- High-grade glioma: Any samples labeled with "High-grade glioma" or are part of `SCPCP000001` which are categorized as HGG tumors. 
+- High-grade glioma: Any samples labeled with "High-grade glioma", "Diffuse midline glioma", or are part of `SCPCP000001` which are categorized as HGG tumors. 
 - Glioblastoma
@@
       diagnosis %in% c("Low-grade glioma", "Glioblastoma", "Pilocytic astrocytoma", "Ganglioglioma", "High-grade glioma", "Atypical teratoid rhabdoid tumor", "Non-cancerous") ~ diagnosis,
-      project_id == "SCPCP000001" ~ "High-grade glioma",
+      project_id == "SCPCP000001" | diagnosis == "Diffuse midline glioma" ~ "High-grade glioma",
       project_id == "SCPCP000002" ~ "Low-grade glioma",
@@
 t_cells <- unique_celltypes |> 
-  stringr::str_subset("T")
+  stringr::str_subset("T cell")
 
 # macro/mono lineage 
-macro_mono_cells <- c(
+myeloid_cells <- c(
   "macrophage", 
@@
   "mononuclear phagocyte", 
-  "dendritic cell"
+  "myeloid leukocyte",
+  "neutrophil"
 )
@@
       consensus_annotation %in% t_cells ~ "T cell", 
-      consensus_annotation %in% macro_mono_cells ~ "Macro/mono",
+      consensus_annotation %in% myeloid_cells ~ "Myeloid",
       consensus_annotation == "Unknown" ~ "Unknown",
@@
   "T cell" = all_celltype_colors[["T cell"]],
-  "Macro/mono" = all_celltype_colors[["macrophage"]],
+  "Myeloid" = all_celltype_colors[["macrophage"]],
   "All remaining cell types" = "grey60",

### 2025-02-12 analyses/cell-type-consensus/exploratory-notebooks/04-brain-and-cns-consensus-celltypes.Rmd
Apply suggestions from code review
Co-authored-by: Jaclyn Taroni <19534205+jaclyn-taroni@users.noreply.github.com>
some rewording

@@
 
-## Macrophage vs. T cell populations
+## Myeloid vs. T cell populations
 
-It looks like there may be some differences between T cell and macrophage infiltrate based on the tumor type. 
-Here we will look specifically at all T cells and anything in the macrophage/monocyte lineage. 
-I'm including any cell types that are in our list of possible cell types and a child of [mononuclear phagocyte](https://www.ebi.ac.uk/ols4/ontologies/cl/classes/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FCL_0000113?lang=en) in the macro/mono group.
+It looks like there may be some differences between the presence of T cells and macrophages based on the tumor type. 
+Here we will look specifically at all T cells and anything in the myeloid lineage, which includes macrophages, monocytes, mononuclear phagocytes, myeloid leukoctyes, and neutrophils. 
 We should also note that "macrophage" here probably refers to microglia instead.
@@
 
-- As expected based on the previous plot, glioblastoma appears to have greater T cell infiltrate than macrophages across all samples. 
+- As expected based on the previous plot, glioblastoma appears to have a higher proportion of T cells when compared to myleoid cells across all samples. 
 - There appears to be a handful of samples in the "High-grade glioma" group that also have high T cell infiltrate. 
@@
 Ganglioglioma also has fewer samples with high percentage of T cell infiltrate and this is also a LGG. 
-- It looks like about 50% of the pilocytic astrocytomas have more T cell infiltrate than macrophages.
+- It looks like about 50% of the pilocytic astrocytomas have more T cells than myeloid cells.
 
@@
 
-Consistent with the previous findings it looks like HGG tend to have higher T cell infiltrate while LGG tend to have higher macrophage infiltrate.
+Consistent with the previous findings it looks like HGG tend to have a higher proportion of T cells while LGG tend to have higher myeloid cell proportions.
 

### 2025-02-17 analyses/cell-type-consensus/exploratory-notebooks/03-osteosarcoma-consensus-celltypes.Rmd
consensus spelling fixes

@@
 Here we look at the top cell types identified across all samples. 
-We have already looked at this in `02-explore-consensus-results.Rmd`, but here we will plot all samples from all three osteo projects ordered by the total percent of cells labeled with a consensus label.
+We have already looked at this in `02-explore-consensus-results.Rmd`, but here we will plot all samples from all three osteosarcoma projects ordered by the total percent of cells labeled with a consensus label.
 
@@
 
-It looks like there's definitely some variation in the distributions of cell types among the osteo samples. 
+It looks like there's definitely some variation in the distributions of cell types among the osteosarcoma samples. 
 Generally, we see most annotated cells are smooth muscle cells and endothelial cells. 

### 2025-02-17 analyses/cell-type-consensus/exploratory-notebooks/04-brain-and-cns-consensus-celltypes.Rmd
consensus spelling fixes

@@
 
-- As expected based on the previous plot, glioblastoma appears to have a higher proportion of T cells when compared to myleoid cells across all samples. 
+- As expected based on the previous plot, glioblastoma appears to have a higher proportion of T cells when compared to myeloid cells across all samples. 
 - There appears to be a handful of samples in the "High-grade glioma" group that also have high T cell infiltrate. 

### 2025-02-18 analyses/cell-type-consensus/exploratory-notebooks/04-brain-and-cns-consensus-celltypes.Rmd

@@
 It looks like there may be some differences between the presence of T cells and macrophages based on the tumor type. 
-Here we will look specifically at all T cells and anything in the myeloid lineage, which includes macrophages, monocytes, mononuclear phagocytes, myeloid leukoctyes, and neutrophils. 
+Here we will look specifically at all T cells and anything in the myeloid lineage, which includes macrophages, monocytes, mononuclear phagocytes, myeloid leukocytes, and neutrophils. 
 We should also note that "macrophage" here probably refers to microglia instead.

### 2025-02-18 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd

@@
 
-- Any gene sets labeled with `up` represent `EWS-FLI1` target genes that we expect to be upregulated in `EWS-FLI1` high tumor cells. 
-- Any gene sets labeled with `down` represent `EWS-FLI1` repressed genes that we expect to be downregulated in `EWS-FLI1` high tumor cells and upregulated in `EWS-FLI1` low tumor cells. 
-- The `aynaud-ews-targets` is a list of genes found to be upregulated in `EWS-FLI1` high tumor cells. 
-- The `wrenn-nt5e-genes` is a list of genes found to be upregulated in `EWS-FLI1` low tumor cells/ cancer associated fibroblasts.
-- The `gobp_ECM` and `hallmark_EMT` are gene sets that are hypothesized to be upregulated in `EWS-FLI1` low tumor cells/cancer associated fibroblasts.
+- Any gene sets labeled with `up` represent `EWS-FLI1` target genes that we expect to be up regulated in `EWS-FLI1` high tumor cells. 
+- Any gene sets labeled with `down` represent `EWS-FLI1` repressed genes that we expect to be down regulated in `EWS-FLI1` high tumor cells and up regulated in `EWS-FLI1` low tumor cells. 
+- The `aynaud-ews-targets` is a list of genes found to be up regulated in `EWS-FLI1` high tumor cells. 
+- The `wrenn-nt5e-genes` is a list of genes found to be up regulated in `EWS-FLI1` low tumor cells/ cancer associated fibroblasts.
+- The `gobp_ECM` and `hallmark_EMT` are gene sets that are hypothesized to be up regulated in `EWS-FLI1` low tumor cells/cancer associated fibroblasts.
 
@@
 
-- Here it looks like the tumor cells have high expression of `EWS-FLI1` upregulated gene sets and low expression of `EWS-FLI1` repressed targets. 
+- Here it looks like the tumor cells have high expression of `EWS-FLI1` up regulated gene sets and low expression of `EWS-FLI1` repressed targets. 
 This suggests these cells are `EWS-high`. 
 - This same pattern is seen in cells labeled as "smooth muscle cell" and "stromal cell". 
-There may even be some slight expression of `EWS-FLI1` upregulated gene sets in "muscle cell", "Unknown", and some of the T cell populations. 
+There may even be some slight expression of `EWS-FLI1` up regulated gene sets in "muscle cell", "Unknown", and some of the T cell populations. 
 Some of these cells are probably mistakenly labeled and should be tumor cells that are "EWS-high".
-- Fibroblasts and chondroctyes have low expression of `EWS-FLI1` upregulated gene sets and high expression of `EWS-FLI1` repressed targets. 
-Many of the genes in the `EWS-FLI1` repressed targets are also genes upregulated in fibroblasts so I'm sure we have a mix of tumor and fibroblasts in there.
+- Fibroblasts and chondrocytes have low expression of `EWS-FLI1` up regulated gene sets and high expression of `EWS-FLI1` repressed targets. 
+Many of the genes in the `EWS-FLI1` repressed targets are also genes up regulated in fibroblasts so I'm sure we have a mix of tumor and fibroblasts in there.
 
@@
 
-It looks like generally the `tumor EWS-high` cells show high expression of both upregulated gene sets and `tumor EWS-low` cells show high expression of both downregulated gene sets, so I feel good about this!
+It looks like generally the `tumor EWS-high` cells show high expression of both up regulated gene sets and `tumor EWS-low` cells show high expression of both down regulated gene sets, so I feel good about this!
 
 I will also note that there are some muscle cells that show slight expression of `aynaud-ews-targets`, but show little to no expression of `staege` targets. 
-I'm still a little worried these are actually tumor cells, but because they only show expression of one of the genesets, we are probably okay to keep them as is noting that the expression profile of muscle cells is very similar to those of tumor cells, this being sarcoma.
+I'm still a little worried these are actually tumor cells, but because they only show expression of one of the gene sets, we are probably okay to keep them as is noting that the expression profile of muscle cells is very similar to those of tumor cells, this being sarcoma.
 
@@
 
-Note that only genes with mean expression > 0 and expressed in at least 10% of cells are included in the dotplot. 
+Note that only genes with mean expression > 0 and expressed in at least 10% of cells are included in the dot plot. 
 All genes are included in the heatmap.

### 2025-02-19 analyses/cell-type-consensus/exploratory-notebooks/05-marker-gene-validation.Rmd
initial exploration notebook for validation plots

---
title: "Validation of marker gene expression"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  project_id: "SCPCP000001"
---

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)
```

## Data setup

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)
module_base <- file.path(repository_base, "analyses", "cell-type-consensus")

# results directory with cell-type-consensus 
results_dir <- file.path(module_base, "results", "cell-type-consensus", params$project_id)

# data directory where project metadata files live
data_dir <- file.path(repository_base, "data", "current")
```

```{r}
# get consensus results 
consensus_results_files <- list.files(results_dir, pattern = "_consensus-cell-type-assignments\\.tsv.\\gz$", recursive = TRUE, full.names = TRUE)

# get gene expression files 
gene_exp_files <- list.files(results_dir, pattern = "_marker-gene-expression\\.tsv.\\gz$", recursive = TRUE, full.names = TRUE)

# list of all validation markers
validation_markers_file <- file.path(module_base, "references", "validation-markers.tsv")

# all validation groups mapped to consensus labels 
validation_groups_file <- file.path(module_base, "references", "consensus-validation-groups.tsv")
```

```{r}
# source summarize_celltypes() function
setup_functions <- file.path(module_base, "exploratory-notebooks", "utils", "setup-functions.R")
source(setup_functions)
```

```{r, message=FALSE}
# read in consensus and gene expression results 
consensus_df <- consensus_results_files |> 
  purrr::map(readr::read_tsv) |>
  dplyr::bind_rows()

gene_exp_df <- gene_exp_files |> 
  purrr::map(readr::read_tsv) |>
  dplyr::bind_rows()

# read in validation markers 
markers_df <- readr::read_tsv(validation_markers_file) |> 
  dplyr::mutate(
    plot_label = glue::glue("{gene_symbol}-{validation_group_annotation}")
  )

validation_groups_df <- readr::read_tsv(validation_groups_file) |> 
  # rename final assigned group to avoid conflicts when merging in marker gene expression 
  # we want to separate the marker gene group from the actual cell type annotation
  dplyr::select(consensus_annotation, assigned_group = validation_group_annotation)
```

```{r}
# combine into one data frame 
combined_df <- consensus_df |> 
  dplyr::left_join(validation_groups_df, by = c("consensus_annotation")) |> 
  dplyr::left_join(gene_exp_df, by = c("barcodes", "library_id")) |> 
  # account for the same gene being present in multiple cell types 
  dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    detected = logcounts > 0 # column indicating if gene is present or not
  )

# make a data frame that just has the unique genes 
unique_gene_df <- combined_df |> 
  dplyr::filter(gene_observed_count == 1)
```

```{r}
# get total number of cells per final annotation group 
total_cells_df <- unique_gene_df |> 
  dplyr::select(library_id, barcodes, assigned_group) |> 
  unique() |> 
  dplyr::count(assigned_group, name = "total_cells")

# get total number of cells each gene is detected in per group 
# and mean gene expression per group 
group_stats_df <- unique_gene_df |> 
  dplyr::group_by(assigned_group, ensembl_gene_id) |>
  dplyr::summarize(
    detected_count = sum(detected),
    mean_exp = mean(logcounts)
  )

gene_summary_df <- group_stats_df |> 
  # add total cells
  dplyr::left_join(total_cells_df, by = c("assigned_group")) |> 
  dplyr::filter(total_cells > 50) |> 
  # add marker gene groups
  dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # get total percent
    percent_exp = (detected_count/total_cells) * 100,
    # account for NA/unknowns and set axes order
    assigned_group = dplyr::if_else(is.na(assigned_group), "unknown", assigned_group) |> 
      factor(levels =  c(unique(markers_df$validation_group_annotation), "unknown"))
  ) 
```

```{r}
# filter out low expressed genes
dotplot_df <- gene_summary_df |> 
  dplyr::filter(mean_exp > 0, percent_exp > 5) |>
  dplyr::arrange(assigned_group) |> 
  dplyr::mutate(y_label = as.factor(glue::glue("{assigned_group} ({total_cells})")))

# pull out cell types to specify legend and x-axis order
celltype_groups <- dotplot_df |> 
  dplyr::pull(assigned_group) |> 
  unique() |>
  as.character()

# specify x axis order
marker_gene_order <- markers_df |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups,
                gene_symbol %in% dotplot_df$gene_symbol) |> 
  dplyr::pull(gene_symbol)

# remove marker genes that aren't present in final annotations and set x axis order 
dotplot_df <- dotplot_df |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups) |> 
  dplyr::mutate(
    gene_symbol = factor(gene_symbol, levels = marker_gene_order)
  )

dotplot <- ggplot(dotplot_df, aes(y = forcats::fct_rev(y_label), x = gene_symbol, color = mean_exp, size = percent_exp)) +
  geom_point() +
  scale_color_viridis_c(option = "magma") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "",
    y = "Cell type annotation"
  )

color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = validation_group_annotation)) + 
  geom_tile() + 
  scale_fill_brewer(palette = "Dark2", breaks = celltype_groups) +
  #scale_fill_manual(values = color_palette, breaks = celltype_groups) +
  ggmap::theme_nothing() +
  theme(legend.position = "bottom") +
  labs(fill = "")

dotplot + color_bar +
  patchwork::plot_layout(ncol = 1, heights = c(4, 0.1))

```

```{r}
# note that we can't use the same heatmap function as before since we are looking at cell types as rows rather than gene sets
# we also want to specify the annotation name

# get annotation colors for each of the marker gene types 
filtered_markers_df <- markers_df |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups,
                gene_symbol %in% gene_summary_df$gene_symbol)

# first make a mtx to use for the heatmap with rows as cell types and genes as columns 
heatmap_mtx <- gene_summary_df |> 
  dplyr::select(gene_symbol, assigned_group, mean_exp) |> 
  dplyr::filter(gene_symbol %in% filtered_markers_df$gene_symbol) |> 
  tidyr::pivot_wider(
    names_from = gene_symbol,
    values_from = mean_exp
  ) |> 
  tibble::column_to_rownames("assigned_group") |>
  as.matrix()
# make sure cell types are present in the right order
heatmap_mtx <- heatmap_mtx[celltype_groups, filtered_markers_df$gene_symbol]

marker_gene_categories <- unique(filtered_markers_df$validation_group_annotation)
num_categories <- length(marker_gene_categories)
category_colors <- palette.colors(palette = "Dark2") |>
  head(n = num_categories) |>
  purrr::set_names(marker_gene_categories)

# create annotation for heatmap
annotation <- ComplexHeatmap::columnAnnotation(
  category = filtered_markers_df$validation_group_annotation,
  col = list(
    category = category_colors
  ),
  annotation_legend_param = list(
      title = "Marker gene category"
    )
)

ComplexHeatmap::Heatmap(
    heatmap_mtx,
    # set the color scale based on min and max values
    col = circlize::colorRamp2(seq(min(heatmap_mtx), max(heatmap_mtx), length = 2), colors = c("white", "#00274C")),
    border = TRUE,
    ## Row parameters
    cluster_rows = FALSE,
    row_title = "",
    row_title_side = "left",
    row_names_side = "left",
    row_dend_side = "right",
    row_names_gp = grid::gpar(fontsize = 10),
    ## Column parameters
    cluster_columns = FALSE,
    show_column_names = TRUE,
    column_names_gp = grid::gpar(fontsize = 8),
    top_annotation = annotation,
    heatmap_legend_param = list(
      title = "Mean expression"
    )
  )
```

Repeat the plots above but use mean expression of all marker genes across all cells

```{r}
# get total number of cells each gene is detected in per group 
# and mean gene expression per group 
group_stats_df <- combined_df |> 
  dplyr::group_by(assigned_group, ensembl_gene_id) |>
  dplyr::summarize(
    detected_count = sum(detected),
    mean_exp = mean(logcounts),
    .groups = "drop"
  ) |> 
  # add in validation group for marker genes
  dplyr::left_join(markers_df, by = c("ensembl_gene_id")) |>
  dplyr::group_by(assigned_group, validation_group_annotation) |> 
  dplyr::mutate(
    group_mean_exp = mean(mean_exp),
    group_detected_count = mean(detected_count),
    .groups = "drop"
  ) |> 
  dplyr::select(assigned_group, validation_group_annotation, group_mean_exp, group_detected_count) |>
  unique()

gene_summary_df <- group_stats_df |> 
  # add total cells
  dplyr::left_join(total_cells_df, by = c("assigned_group")) |> 
  dplyr::filter(total_cells > 50) |> 
  # add per gene stats
  #dplyr::left_join(group_stats_df, by = c("ensembl_gene_id", "assigned_group")) |> 
  #dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # get total percent
    percent_exp = (group_detected_count/total_cells) * 100,
    # account for NA/unknowns and set axes order
    assigned_group = dplyr::if_else(is.na(assigned_group), "unknown", assigned_group) |> 
      factor(levels =  c(unique(markers_df$validation_group_annotation), "unknown"))
  ) 
```

```{r}
# filter out low expressed genes
dotplot_df <- gene_summary_df |> 
  dplyr::filter(group_mean_exp > 0, percent_exp > 5) |>
  dplyr::arrange(assigned_group) |> 
  dplyr::mutate(y_label = as.factor(glue::glue("{assigned_group} ({total_cells})")))

# pull out cell types to specify legend and x-axis order
celltype_groups <- dotplot_df |> 
  dplyr::pull(assigned_group) |> 
  unique() |>
  as.character()

# remove marker genes that aren't present in final annotations and set x axis order 
dotplot_df <- dotplot_df |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups) |>
  dplyr::mutate(
    validation_group_annotation = factor(validation_group_annotation, levels = celltype_groups)
  )

dotplot <- ggplot(dotplot_df, aes(y = forcats::fct_rev(y_label), x = validation_group_annotation, color = group_mean_exp, size = percent_exp)) +
  geom_point() +
  scale_color_viridis_c(option = "magma") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "Marker gene group",
    y = "Cell type annotation"
  )

color_bar <- ggplot(dotplot_df, aes(x = validation_group_annotation, y = 1, fill = validation_group_annotation)) + 
  geom_tile() + 
  scale_fill_brewer(palette = "Dark2", breaks = celltype_groups) +
  #scale_fill_manual(values = color_palette, breaks = celltype_groups) +
  ggmap::theme_nothing() +
  theme(legend.position = "bottom") +
  labs(fill = "")

dotplot + color_bar +
  patchwork::plot_layout(ncol = 1, heights = c(4, 0.1))

```

```{r}
# get annotation colors for each of the marker gene types 
filtered_markers_df <- markers_df |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups) |> 
  dplyr::select(validation_group_annotation) |> 
  unique()

# first make a mtx to use for the heatmap with rows as cell types and genes as columns 
heatmap_mtx <- gene_summary_df |> 
  dplyr::select(validation_group_annotation, assigned_group, group_mean_exp) |> 
  tidyr::pivot_wider(
    names_from = validation_group_annotation,
    values_from = group_mean_exp
  ) |> 
  tibble::column_to_rownames("assigned_group") |>
  as.matrix()
# make sure cell types are present in the right order
heatmap_mtx <- heatmap_mtx[celltype_groups, filtered_markers_df$validation_group_annotation]

marker_gene_categories <- unique(filtered_markers_df$validation_group_annotation)
num_categories <- length(marker_gene_categories)
category_colors <- palette.colors(palette = "Dark2") |>
  head(n = num_categories) |>
  purrr::set_names(marker_gene_categories)

# create annotation for heatmap
annotation <- ComplexHeatmap::columnAnnotation(
  category = filtered_markers_df$validation_group_annotation,
  col = list(
    category = category_colors
  ),
  annotation_legend_param = list(
      title = "Marker gene category"
    )
)

ComplexHeatmap::Heatmap(
    heatmap_mtx,
    # set the color scale based on min and max values
    col = circlize::colorRamp2(seq(min(heatmap_mtx), max(heatmap_mtx), length = 2), colors = c("white", "#00274C")),
    border = TRUE,
    ## Row parameters
    cluster_rows = FALSE,
    row_title = "",
    row_title_side = "left",
    row_names_side = "left",
    row_dend_side = "right",
    row_names_gp = grid::gpar(fontsize = 10),
    ## Column parameters
    cluster_columns = FALSE,
    show_column_names = TRUE,
    column_names_gp = grid::gpar(fontsize = 8),
    top_annotation = annotation,
    heatmap_legend_param = list(
      title = "Mean expression"
    )
  )
```

## Session info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-02-20 analyses/cell-type-consensus/exploratory-notebooks/05-marker-gene-validation.Rmd

@@
 
+This notebook looks at marker gene expression for consensus cell types in `r params$project_id`. 
+The marker genes shown in this notebook can be found in [`references/validation-markers.tsv`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-consensus/references/validation-markers.tsv).
+
+All plots will also show the final cell type annotation group based on the `validation_group_annotation` in [`references/consensus-validation-groups.tsv`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv).
+This means that cells assigned to more granular cell types will be lumped together with a broader cell type, e.g., memory T cell and naive T cell will be renamed as just T cell.
+
 ```{r packages}
@@
   library(ggplot2)
+  library(patchwork)
 })
@@
 ```{r}
+# marker_heatmap
+plotting_functions <- file.path(module_base, "exploratory-notebooks", "utils", "plotting-functions.R")
+source(plotting_functions)
+```
+
+```{r}
 # get consensus results 
@@
 
-```{r}
-# source summarize_celltypes() function
-setup_functions <- file.path(module_base, "exploratory-notebooks", "utils", "setup-functions.R")
-source(setup_functions)
-```
-
 ```{r, message=FALSE}
@@
 # read in validation markers 
-markers_df <- readr::read_tsv(validation_markers_file) |> 
-  dplyr::mutate(
-    plot_label = glue::glue("{gene_symbol}-{validation_group_annotation}")
-  )
+markers_df <- readr::read_tsv(validation_markers_file)
 
@@
   # we want to separate the marker gene group from the actual cell type annotation
-  dplyr::select(consensus_annotation, assigned_group = validation_group_annotation)
+  dplyr::select(consensus_annotation, broad_celltype_group = validation_group_annotation)
 ```
@@
 ```{r}
-# combine into one data frame 
+# combine into one data frame that has all marker genes
 combined_df <- consensus_df |> 
@@
   ) 
+```
+
+## Plot setup
+
+```{r}
+# prep for plots 
+# get total number of cells per final annotation group 
+total_cells_df <- combined_df |> 
+  dplyr::select(library_id, barcodes, broad_celltype_group) |> 
+  unique() |> 
+  dplyr::count(broad_celltype_group, name = "total_cells") 
+```
+
+```{r}
+# get total number of cells each gene is detected in per group 
+# and mean gene expression per group of each marker gene and mean expression of all marker genes 
+group_stats_df <- combined_df |> 
+  dplyr::group_by(broad_celltype_group, ensembl_gene_id) |>
+  dplyr::summarize(
+    detected_count = sum(detected),
+    mean_exp = mean(logcounts)
+  ) |> 
+  # add in validation group for marker genes
+  dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |>
+  dplyr::group_by(broad_celltype_group, validation_group_annotation) |> 
+  dplyr::mutate(
+    # calculate mean expression/detected across all markers for a specific group 
+    all_markers_mean_exp = mean(mean_exp),
+    all_markers_detected_count = mean(detected_count)
+  ) |>  # add total cells
+  dplyr::left_join(total_cells_df, by = c("broad_celltype_group")) |> 
+  # for plotting we're only going to look at any cell types with > 50 cells otherwise these plots can get wild 
+  dplyr::filter(total_cells > 50) |> 
+  dplyr::rowwise() |> 
+  dplyr::mutate(
+    # get total percent expressed
+    percent_exp = (detected_count/total_cells) * 100,
+    all_markers_percent_exp = (all_markers_detected_count/total_cells) * 100, 
+    # account for NA/unknowns and set axes order
+    broad_celltype_group = dplyr::if_else(is.na(broad_celltype_group), "unknown", broad_celltype_group) |> 
+      factor(levels =  c(unique(markers_df$validation_group_annotation), "unknown"))
+  )
 
 # make a data frame that just has the unique genes 
-unique_gene_df <- combined_df |> 
+unique_gene_df <- group_stats_df |> 
   dplyr::filter(gene_observed_count == 1)
@@
 ```{r}
-# get total number of cells per final annotation group 
-total_cells_df <- unique_gene_df |> 
-  dplyr::select(library_id, barcodes, assigned_group) |> 
+# get list of celltypes to keep and assign colors 
+celltype_groups <- group_stats_df |> 
+  dplyr::pull(broad_celltype_group) |> 
   unique() |>
-  dplyr::count(assigned_group, name = "total_cells")
+  as.character()
 
-# get total number of cells each gene is detected in per group 
-# and mean gene expression per group 
-group_stats_df <- unique_gene_df |> 
-  dplyr::group_by(assigned_group, ensembl_gene_id) |>
-  dplyr::summarize(
-    detected_count = sum(detected),
-    mean_exp = mean(logcounts)
-  )
+# assign colors to use in annotation bars 
+celltype_colors <- palette.colors(palette = "alphabet")[1:length(celltype_groups)]
+names(celltype_colors) <- celltype_groups
 
-gene_summary_df <- group_stats_df |> 
-  # add total cells
-  dplyr::left_join(total_cells_df, by = c("assigned_group")) |> 
-  dplyr::filter(total_cells > 50) |> 
-  # add marker gene groups
-  dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |> 
-  dplyr::rowwise() |> 
-  dplyr::mutate(
-    # get total percent
-    percent_exp = (detected_count/total_cells) * 100,
-    # account for NA/unknowns and set axes order
-    assigned_group = dplyr::if_else(is.na(assigned_group), "unknown", assigned_group) |> 
-      factor(levels =  c(unique(markers_df$validation_group_annotation), "unknown"))
-  ) 
+# set order for cell types 
+celltype_order <- unique(filtered_markers_df$validation_group_annotation)
 ```
@@
 ```{r}
-# filter out low expressed genes
-dotplot_df <- gene_summary_df |> 
-  dplyr::filter(mean_exp > 0, percent_exp > 5) |>
-  dplyr::arrange(assigned_group) |> 
-  dplyr::mutate(y_label = as.factor(glue::glue("{assigned_group} ({total_cells})")))
-
-# pull out cell types to specify legend and x-axis order
-celltype_groups <- dotplot_df |> 
-  dplyr::pull(assigned_group) |> 
-  unique() |>
-  as.character()
+# filter markers to those that are actually relevant 
+# set order of gene symbol based on celltype groups 
+filtered_markers_df <- markers_df |> 
+  dplyr::filter(gene_observed_count == 1,
+                validation_group_annotation %in% celltype_groups,
+                gene_symbol %in% group_stats_df$gene_symbol)
 
@@
   dplyr::pull(gene_symbol)
+```
 
+## Expression of unique marker genes
+
+First we will look at the expression of marker genes that are specific to a cell type group.  
+All marker genes shown in the following plots are only found to be a marker gene for a single cell type group. 
+Only cell type groups with at least 50 cells will be shown.
+
+Note that I did try and look at all marker genes, without specifying that they must be unique, and the plots were overwhelming with lots of genes. 
+Instead I included the a separate section looking at the mean expression across all markers for a given group.
+
+### Dot plot
+
+The dot plot shows the mean expression of each marker gene across all cells assigned to a given cell type group. 
+Only genes that have mean expression > 0 and are expressed in at least 5% of cells in at least one cell type group will be shown.
+
+```{r, fig.height=10, fig.width=15}
+# filter out low expressed genes
+dotplot_df <- unique_gene_df |> 
+  dplyr::filter(mean_exp > 0, percent_exp > 5) |> 
+  dplyr::arrange(broad_celltype_group) |> 
+  # add a label for the plot 
+  dplyr::mutate(y_label = as.factor(glue::glue("{broad_celltype_group} ({total_cells})"))) |> 
   # remove marker genes that aren't present in final annotations and set x axis order 
-dotplot_df <- dotplot_df |> 
-  dplyr::filter(validation_group_annotation %in% celltype_groups) |> 
+  dplyr::filter(gene_symbol %in% marker_gene_order) |> 
   dplyr::mutate(
-    gene_symbol = factor(gene_symbol, levels = marker_gene_order)
+    # set orders of gene symbol and validation groups 
+    gene_symbol = factor(gene_symbol, levels = marker_gene_order),
+    validation_group_annotation = factor(validation_group_annotation, levels = celltype_order)
   )
 
+# make dotplot with marker gene exp
 dotplot <- ggplot(dotplot_df, aes(y = forcats::fct_rev(y_label), x = gene_symbol, color = mean_exp, size = percent_exp)) +
@@
     x = "",
-    y = "Cell type annotation"
+    y = "Broad cell type annotation",
+    color = "Mean gene expression",
+    size = "Percent cells expressed"
   )
 
+# add annotation bar aligning marker genes with validation group 
 color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = validation_group_annotation)) + 
   geom_tile() + 
-  scale_fill_brewer(palette = "Dark2", breaks = celltype_groups) +
-  #scale_fill_manual(values = color_palette, breaks = celltype_groups) +
+  scale_fill_manual(values = plot_colors, breaks = levels(dotplot_df$validation_group_annotation)) +
   ggmap::theme_nothing() +
@@
   patchwork::plot_layout(ncol = 1, heights = c(4, 0.1)) 
-
 ```
 
+### Heatmap
+
+The heatmap shows the mean expression of each marker gene across all cells assigned to a given cell type group.
+
 ```{r}
-# note that we can't use the same heatmap function as before since we are looking at cell types as rows rather than gene sets
-# we also want to specify the annotation name
-
-# get annotation colors for each of the marker gene types 
-filtered_markers_df <- markers_df |> 
-  dplyr::filter(validation_group_annotation %in% celltype_groups,
-                gene_symbol %in% gene_summary_df$gene_symbol)
-
 # first make a mtx to use for the heatmap with rows as cell types and genes as columns 
-heatmap_mtx <- gene_summary_df |> 
-  dplyr::select(gene_symbol, assigned_group, mean_exp) |> 
+unique_heatmap_mtx <- unique_gene_df |> 
+  dplyr::ungroup() |> 
+  dplyr::select(gene_symbol, broad_celltype_group, mean_exp) |> 
   dplyr::filter(gene_symbol %in% filtered_markers_df$gene_symbol) |> 
@@
   ) |> 
-  tibble::column_to_rownames("assigned_group") |>
+  tibble::column_to_rownames("broad_celltype_group") |>
   as.matrix()
 # make sure cell types are present in the right order
-heatmap_mtx <- heatmap_mtx[celltype_groups, filtered_markers_df$gene_symbol]
-
-marker_gene_categories <- unique(filtered_markers_df$validation_group_annotation)
-num_categories <- length(marker_gene_categories)
-category_colors <- palette.colors(palette = "Dark2") |>
-  head(n = num_categories) |>
-  purrr::set_names(marker_gene_categories)
+unique_heatmap_mtx <- unique_heatmap_mtx[c(celltype_order, "unknown"), filtered_markers_df$gene_symbol]
 
@@
 annotation <- ComplexHeatmap::columnAnnotation(
-  category = filtered_markers_df$validation_group_annotation,
+  group = filtered_markers_df$validation_group_annotation,
   col = list(
-    category = category_colors
+    group = celltype_colors
   ),
   annotation_legend_param = list(
-      title = "Marker gene category"
-    )
-)
-
-ComplexHeatmap::Heatmap(
-    heatmap_mtx,
-    # set the color scale based on min and max values
-    col = circlize::colorRamp2(seq(min(heatmap_mtx), max(heatmap_mtx), length = 2), colors = c("white", "#00274C")),
-    border = TRUE,
-    ## Row parameters
-    cluster_rows = FALSE,
-    row_title = "",
-    row_title_side = "left",
-    row_names_side = "left",
-    row_dend_side = "right",
-    row_names_gp = grid::gpar(fontsize = 10),
-    ## Column parameters
-    cluster_columns = FALSE,
-    show_column_names = TRUE,
-    column_names_gp = grid::gpar(fontsize = 8),
-    top_annotation = annotation,
-    heatmap_legend_param = list(
-      title = "Mean expression"
+      title = "Validation group"
     )
@@
 
-Repeat the plots above but use mean expression of all marker genes across all cells
+```{r, fig.height=7, fig.width=10}
+marker_gene_heatmap(unique_heatmap_mtx, annotation)
+```
+
+## Mean expression across all marker genes
+
+We will make the same plots but instead of looking at the expression of individual marker genes for all cell type groups, we will plot the mean expression of all marker genes for a specific cell type group.
+
+This is calculated by first getting the mean expression of each individual marker gene across all cells in a cell type group and then taking the mean of that value across all marker genes. 
+This results in one row per assigned cell type group and one column per marker gene group.
+
+### Dot plot
 
 ```{r}
-# get total number of cells each gene is detected in per group 
-# and mean gene expression per group 
-group_stats_df <- combined_df |> 
-  dplyr::group_by(assigned_group, ensembl_gene_id) |>
-  dplyr::summarize(
-    detected_count = sum(detected),
-    mean_exp = mean(logcounts),
-    .groups = "drop"
-  ) |> 
-  # add in validation group for marker genes
-  dplyr::left_join(markers_df, by = c("ensembl_gene_id")) |>
-  dplyr::group_by(assigned_group, validation_group_annotation) |> 
-  dplyr::mutate(
-    group_mean_exp = mean(mean_exp),
-    group_detected_count = mean(detected_count),
-    .groups = "drop"
-  ) |> 
-  dplyr::select(assigned_group, validation_group_annotation, group_mean_exp, group_detected_count) |>
-  unique()
+# filter low expressed genes 
+grouped_dotplot_df <- group_stats_df |> 
+  dplyr::select(broad_celltype_group, validation_group_annotation, all_markers_mean_exp, all_markers_percent_exp, total_cells) |>
+  unique() |> 
+  dplyr::filter(all_markers_mean_exp > 0, all_markers_percent_exp > 5, validation_group_annotation %in% celltype_groups) |>
+  dplyr::arrange(broad_celltype_group) |> 
+  # make sure labels are correct 
+  dplyr::mutate(y_label = as.factor(glue::glue("{broad_celltype_group} ({total_cells})")),
+                validation_group_annotation = factor(validation_group_annotation, levels = celltype_order))
 
-gene_summary_df <- group_stats_df |> 
-  # add total cells
-  dplyr::left_join(total_cells_df, by = c("assigned_group")) |> 
-  dplyr::filter(total_cells > 50) |> 
-  # add per gene stats
-  #dplyr::left_join(group_stats_df, by = c("ensembl_gene_id", "assigned_group")) |> 
-  #dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |> 
-  dplyr::rowwise() |> 
-  dplyr::mutate(
-    # get total percent
-    percent_exp = (group_detected_count/total_cells) * 100,
-    # account for NA/unknowns and set axes order
-    assigned_group = dplyr::if_else(is.na(assigned_group), "unknown", assigned_group) |> 
-      factor(levels =  c(unique(markers_df$validation_group_annotation), "unknown"))
-  ) 
-```
-
-```{r}
-# filter out low expressed genes
-dotplot_df <- gene_summary_df |> 
-  dplyr::filter(group_mean_exp > 0, percent_exp > 5) |>
-  dplyr::arrange(assigned_group) |> 
-  dplyr::mutate(y_label = as.factor(glue::glue("{assigned_group} ({total_cells})")))
-
-# pull out cell types to specify legend and x-axis order
-celltype_groups <- dotplot_df |> 
-  dplyr::pull(assigned_group) |> 
-  unique() |>
-  as.character()
-
-# remove marker genes that aren't present in final annotations and set x axis order 
-dotplot_df <- dotplot_df |> 
-  dplyr::filter(validation_group_annotation %in% celltype_groups) |>
-  dplyr::mutate(
-    validation_group_annotation = factor(validation_group_annotation, levels = celltype_groups)
-  )
-
-dotplot <- ggplot(dotplot_df, aes(y = forcats::fct_rev(y_label), x = validation_group_annotation, color = group_mean_exp, size = percent_exp)) +
+ggplot(grouped_dotplot_df, aes(y = forcats::fct_rev(y_label), x = validation_group_annotation, color = all_markers_mean_exp, size = all_markers_percent_exp)) +
   geom_point() +
@@
     x = "Marker gene group",
-    y = "Cell type annotation"
+    y = "Broad cell type annotation", 
+    color = "Mean gene expression", 
+    size = "Percent cells expressed"
   )
-
-color_bar <- ggplot(dotplot_df, aes(x = validation_group_annotation, y = 1, fill = validation_group_annotation)) + 
-  geom_tile() + 
-  scale_fill_brewer(palette = "Dark2", breaks = celltype_groups) +
-  #scale_fill_manual(values = color_palette, breaks = celltype_groups) +
-  ggmap::theme_nothing() +
-  theme(legend.position = "bottom") +
-  labs(fill = "")
-
-dotplot + color_bar +
-  patchwork::plot_layout(ncol = 1, heights = c(4, 0.1))
 
@@
 
+### Heatmap
+
 ```{r}
-# get annotation colors for each of the marker gene types 
-filtered_markers_df <- markers_df |> 
-  dplyr::filter(validation_group_annotation %in% celltype_groups) |> 
-  dplyr::select(validation_group_annotation) |> 
-  unique()
-
-# first make a mtx to use for the heatmap with rows as cell types and genes as columns 
-heatmap_mtx <- gene_summary_df |> 
-  dplyr::select(validation_group_annotation, assigned_group, group_mean_exp) |> 
-  tidyr::pivot_wider(
-    names_from = validation_group_annotation,
-    values_from = group_mean_exp
-  ) |> 
-  tibble::column_to_rownames("assigned_group") |>
-  as.matrix()
-# make sure cell types are present in the right order
-heatmap_mtx <- heatmap_mtx[celltype_groups, filtered_markers_df$validation_group_annotation]
-
-marker_gene_categories <- unique(filtered_markers_df$validation_group_annotation)
-num_categories <- length(marker_gene_categories)
-category_colors <- palette.colors(palette = "Dark2") |>
-  head(n = num_categories) |>
-  purrr::set_names(marker_gene_categories)
-
-# create annotation for heatmap
-annotation <- ComplexHeatmap::columnAnnotation(
-  category = filtered_markers_df$validation_group_annotation,
+# get annotation 
+grouped_annotation <- ComplexHeatmap::columnAnnotation(
+  group = unique(filtered_markers_df$validation_group_annotation),
   col = list(
-    category = category_colors
+    group = celltype_colors
   ),
   annotation_legend_param = list(
-      title = "Marker gene category"
+      title = "Validation group"
     )
@@
 
-ComplexHeatmap::Heatmap(
-    heatmap_mtx,
-    # set the color scale based on min and max values
-    col = circlize::colorRamp2(seq(min(heatmap_mtx), max(heatmap_mtx), length = 2), colors = c("white", "#00274C")),
-    border = TRUE,
-    ## Row parameters
-    cluster_rows = FALSE,
-    row_title = "",
-    row_title_side = "left",
-    row_names_side = "left",
-    row_dend_side = "right",
-    row_names_gp = grid::gpar(fontsize = 10),
-    ## Column parameters
-    cluster_columns = FALSE,
-    show_column_names = TRUE,
-    column_names_gp = grid::gpar(fontsize = 8),
-    top_annotation = annotation,
-    heatmap_legend_param = list(
-      title = "Mean expression"
-    )
-  )
+# first make a mtx to use for the heatmap with rows as cell types and marker gene groups as columns 
+grouped_heatmap_mtx <- group_stats_df |> 
+  dplyr::select(validation_group_annotation, broad_celltype_group, all_markers_mean_exp) |> 
+  unique() |>
+  dplyr::filter(validation_group_annotation %in% celltype_order) |> 
+  tidyr::pivot_wider(
+    names_from = validation_group_annotation,
+    values_from = all_markers_mean_exp
+  ) |> 
+  tibble::column_to_rownames("broad_celltype_group") |>
+  as.matrix()
 ```
+
+```{r}
+# make sure cell types are present in the right order
+grouped_heatmap_mtx <- grouped_heatmap_mtx[c(celltype_order, "unknown"), celltype_order]
+```
+
+```{r, fig.height=7, fig.width=10}
+marker_gene_heatmap(grouped_heatmap_mtx, grouped_annotation)
+```
+
+## Conclusions
+
+- Regardless of looking at individual genes or the mean expression of all genes, we see similar patterns where generally the cells assigned to a given cell type show expression of the appropriate marker genes when compared to other cells.
+- There seems to be pretty clear distinction of T cells, B cells, and NK cells. 
+- Monocytes, macrophages, and myeloid cells are generally hard to distinguish from each other and all share similar marker genes. 
+Perhaps we could collapse these into myeloid, but I do see expression of monocyte specific genes in monocytes only and maybe some macrophage specific genes in macrophages only, so it's probably okay to keep separate.
+I think that is more obvious in the heatmaps than in the dot plots. 
+- Neurons and astrocytes both show expression of both neuron and astrocyte marker genes. 
+It does look like there are a set of genes that are more highly expressed in the astrocyte population though, so I think we can keep those gene sets separate. 
+- Here the unknown cells don't show strong expression of any specific set of genes, but it does look like they show more expression of neuron/astrocyte markers than any of the other marker genes in the plot.
+This makes sense since we are looking at brain tumors and I expect similarity between tumor cells and neural cell types.
 

### 2025-02-20 analyses/cell-type-consensus/exploratory-notebooks/04-brain-and-cns-consensus-celltypes.Rmd
re-render brain notebook

@@
 # list all results files 
-results_files <- list.files(results_dir, pattern = "_consensus-cell-types\\.tsv.\\gz$", full.names = TRUE)
-
-# get project ids from file list and assign as names
-project_ids <- stringr::str_remove(basename(results_files), "_consensus-cell-types.tsv.gz")
-names(results_files) <- project_ids
-
-# grab only brain/cns projects
 brain_project_ids <- c("SCPCP000001", "SCPCP000002", "SCPCP000010", "SCPCP000021", "SCPCP000009")
-results_files <- results_files[brain_project_ids]
+brain_results_dirs <- file.path(results_dir, brain_project_ids)
+results_files <- brain_results_dirs |> 
+  list.files(pattern = "_processed_consensus-cell-types\\.tsv.\\gz$", recursive = TRUE, full.names = TRUE)
 
@@
   purrr::imap(summarize_celltypes) |> 
-  dplyr::bind_rows(.id = "project_id") |> 
+  dplyr::bind_rows() |> 
   # remove multiplexed libraries 
@@
 unique_celltypes <- all_results_df |> 
+  # order by frequency otherwise neurons will be light grey 
+  dplyr::arrange(desc(total_cells_per_annotation)) |> 
   dplyr::filter(percent_cells_annotation > 0.1, # we're only ever going to plot cell types that are greater than 1%
@@
   palette.colors(palette = "alphabet"),
+  "black", # extra cell type 
   "grey60", # all remaining

### 2025-02-20 analyses/cell-type-consensus/exploratory-notebooks/04-brain-and-cns-consensus-celltypes.Rmd
remove warning messages

@@
 
-```{r, message=FALSE}
+```{r, message=FALSE,warning=FALSE}
 # get immune cell types 

### 2025-02-24 analyses/cell-type-consensus/exploratory-notebooks/05-marker-gene-validation.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
     # account for NA/unknowns and set axes order
-    broad_celltype_group = dplyr::if_else(is.na(broad_celltype_group), "unknown", broad_celltype_group) |> 
+    broad_celltype_group = tidyr::replace_na(broad_celltype_group, "unknown") |> 
       factor(levels =  c(unique(markers_df$validation_group_annotation), "unknown"))

### 2025-02-25 analyses/cell-type-consensus/exploratory-notebooks/05-marker-gene-validation.Rmd

@@
 # combine into one data frame that has all marker genes
+# one row for each cell/marker gene combination (total rows= total cells x total marker genes)
+# includes original consensus annotation, braod cell type group, marker gene id, logcounts, validation group annotation, information regarding expression of marker gene in CellMarker tissues 
 combined_df <- consensus_df |> 
+  # add in broad cell type group which is used for plotting
+  # groups similar cell types together 
   dplyr::left_join(validation_groups_df, by = c("consensus_annotation")) |> 
   dplyr::left_join(gene_exp_df, by = c("barcodes", "library_id")) |> 
+  # add marker gene information (associated validation group annotation, gene observed count, percent tissues)
   # account for the same gene being present in multiple cell types 
@@
 ```{r}
-# get total number of cells each gene is detected in per group 
-# and mean gene expression per group of each marker gene and mean expression of all marker genes 
+# table with one row per unique broad cell type/ marker gene combination 
+# first all cells in with the same broad_celltype_group (determined based on consensus_annotation) are grouped together
+# then get the mean gene expression and total percentage of cells that express each marker gene across all cells in that group
+# do this for every possible marker gene/ validation group assignment 
+# second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
 group_stats_df <- combined_df |> 
+  # for each assigned cell type/marker gene combo get total detected and mean expression
   dplyr::group_by(broad_celltype_group, ensembl_gene_id) |>
@@
   # add in validation group for marker genes
+  # this includes all possible marker genes and all possible validation group assignments 
   dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |>
+  # now get the mean expression/ mean percentage across all marker genes for a given validation group
+  # here the broad_celltype_group is the final assigned annotation for that group of cells 
+  # the validation_group_annotation refers to the cell type that marker gene is associated with 
   dplyr::group_by(broad_celltype_group, validation_group_annotation) |> 
@@
 # make a data frame that just has the unique genes 
+# unique genes are those that are only observed in one cell type based on observance in Cell Marker 
+# this was determined when creating the marker gene table from Cell Marker 
 unique_gene_df <- group_stats_df |> 
@@
 names(celltype_colors) <- celltype_groups
-
-# set order for cell types 
-celltype_order <- unique(filtered_markers_df$validation_group_annotation)
-```
-
-```{r}
-# filter markers to those that are actually relevant 
-# set order of gene symbol based on celltype groups 
-filtered_markers_df <- markers_df |> 
-  dplyr::filter(gene_observed_count == 1,
-                validation_group_annotation %in% celltype_groups,
-                gene_symbol %in% group_stats_df$gene_symbol)
-
-# specify x axis order
-marker_gene_order <- markers_df |> 
-  dplyr::filter(validation_group_annotation %in% celltype_groups,
-                gene_symbol %in% dotplot_df$gene_symbol) |> 
-  dplyr::pull(gene_symbol)
 ```
@@
 Only genes that have mean expression > 0 and are expressed in at least 5% of cells in at least one cell type group will be shown.
+
+```{r}
+# filter markers to those that are actually relevant 
+# we will only plot the marker genes for cell types that are part of the assigned broad validation group for this project
+# we don't care about plotting marker genes for cell types that aren't present here 
+# note that we will use this for both the dotplot and the heatmap
+filtered_markers_df <- markers_df |> 
+  dplyr::filter(validation_group_annotation %in% celltype_groups,
+                gene_symbol %in% group_stats_df$gene_symbol)
+
+# now only keep unique markers for the dotplot 
+unique_markers_df <- filtered_markers_df |> 
+  dplyr::filter(gene_observed_count == 1)
+
+# specify x axis order for dotplot
+marker_gene_order <- unique_markers_df |> 
+  dplyr::pull(gene_symbol)
+
+# set order for cell types 
+celltype_order <- unique(unique_markers_df$validation_group_annotation)
+```
 
@@
   geom_tile() + 
-  scale_fill_manual(values = plot_colors, breaks = levels(dotplot_df$validation_group_annotation)) +
+  scale_fill_manual(values = celltype_colors, breaks = levels(dotplot_df$validation_group_annotation)) +
   ggmap::theme_nothing() +
@@
 The heatmap shows the mean expression of each marker gene across all cells assigned to a given cell type group. 
+In contrast to the dot plot, we will show all marker genes for a given cell type group, not just the uniquely expressed genes.
+
+```{r}
+# add column to uniquely label gene symbols since some genes are present across multiple cell types 
+# add this here since we will use the order in this data frame to set the order of the heatmap columns
+heatmap_markers_df <- filtered_markers_df |> 
+  dplyr::mutate(
+    unique_symbol = glue::glue("{gene_symbol}-{validation_group_annotation}")
+  ) 
+```
 
@@
 # first make a mtx to use for the heatmap with rows as cell types and genes as columns 
-unique_heatmap_mtx <- unique_gene_df |> 
+heatmap_mtx <- group_stats_df |> 
   dplyr::ungroup() |> 
-  dplyr::select(gene_symbol, broad_celltype_group, mean_exp) |> 
-  dplyr::filter(gene_symbol %in% filtered_markers_df$gene_symbol) |> 
+  # add unique label for gene symbol 
+  dplyr::mutate(
+    unique_symbol = glue::glue("{gene_symbol}-{validation_group_annotation}")
+  ) |> 
+  dplyr::select(unique_symbol, broad_celltype_group, mean_exp) |> 
+  dplyr::filter(unique_symbol %in% heatmap_markers_df$unique_symbol) |> 
   tidyr::pivot_wider(
-    names_from = gene_symbol,
+    names_from = unique_symbol,
     values_from = mean_exp
@@
 # make sure cell types are present in the right order
-unique_heatmap_mtx <- unique_heatmap_mtx[c(celltype_order, "unknown"), filtered_markers_df$gene_symbol]
+heatmap_mtx <- heatmap_mtx[c(celltype_order, "unknown"), heatmap_markers_df$unique_symbol]
 
@@
 annotation <- ComplexHeatmap::columnAnnotation(
-  group = filtered_markers_df$validation_group_annotation,
+  group = heatmap_markers_df$validation_group_annotation,
   col = list(
@@
 ```{r, fig.height=7, fig.width=10}
-marker_gene_heatmap(unique_heatmap_mtx, annotation)
+marker_gene_heatmap(heatmap_mtx, annotation)
 ```
@@
 
-We will make the same plots but instead of looking at the expression of individual marker genes for all cell type groups, we will plot the mean expression of all marker genes for a specific cell type group.
+We will make the same dot plot but instead of looking at the expression of individual marker genes for all cell type groups, we will plot the mean expression of all marker genes for a specific cell type group.
 
@@
 This results in one row per assigned cell type group and one column per marker gene group.
-
-### Dot plot
 
@@
 
-### Heatmap
-
-```{r}
-# get annotation 
-grouped_annotation <- ComplexHeatmap::columnAnnotation(
-  group = unique(filtered_markers_df$validation_group_annotation),
-  col = list(
-    group = celltype_colors
-  ),
-  annotation_legend_param = list(
-      title = "Validation group"
-    )
-)
-
-# first make a mtx to use for the heatmap with rows as cell types and marker gene groups as columns 
-grouped_heatmap_mtx <- group_stats_df |> 
-  dplyr::select(validation_group_annotation, broad_celltype_group, all_markers_mean_exp) |> 
-  unique() |>
-  dplyr::filter(validation_group_annotation %in% celltype_order) |> 
-  tidyr::pivot_wider(
-    names_from = validation_group_annotation,
-    values_from = all_markers_mean_exp
-  ) |> 
-  tibble::column_to_rownames("broad_celltype_group") |>
-  as.matrix()
-```
-
-```{r}
-# make sure cell types are present in the right order
-grouped_heatmap_mtx <- grouped_heatmap_mtx[c(celltype_order, "unknown"), celltype_order]
-```
-
-```{r, fig.height=7, fig.width=10}
-marker_gene_heatmap(grouped_heatmap_mtx, grouped_annotation)
-```
-
 ## Conclusions

### 2025-02-26 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
new template for validation reports

--- 2025-02-25 analyses/cell-type-consensus/exploratory-notebooks/05-marker-gene-validation.Rmd
@@
 output:
-  html_notebook:
+  html_document:
     toc: true
@@
 params:
-  project_id: "SCPCP000001"
+  project_id: "SCPCP000002"
 ---
@@
 
-# results directory with cell-type-consensus 
-results_dir <- file.path(module_base, "results", "cell-type-consensus", params$project_id)
-
-# data directory where project metadata files live
-data_dir <- file.path(repository_base, "data", "current")
+# path to results directory
+results_dir <- file.path(repository_base, "data", "current", "results", "cell-type-consensus", params$project_id)
 ```
@@
 # get consensus results 
-consensus_results_files <- list.files(results_dir, pattern = "_consensus-cell-type-assignments\\.tsv.\\gz$", recursive = TRUE, full.names = TRUE)
+consensus_results_files <- list.files(results_dir, pattern = "_processed_consensus-cell-types\\.tsv\\.gz$", recursive = TRUE, full.names = TRUE)
 
 # get gene expression files 
-gene_exp_files <- list.files(results_dir, pattern = "_marker-gene-expression\\.tsv.\\gz$", recursive = TRUE, full.names = TRUE)
+gene_exp_files <- list.files(results_dir, pattern = "_processed_marker-gene-expression\\.tsv\\.gz$", recursive = TRUE, full.names = TRUE)
 
+# check that consensus results exist
+stopifnot(
+  "No cell type results found in data/current/results/cell-type-consensus" = length(consensus_results_files) > 0,
+  "No gene expression files found in data/current/results/cell-type-consensus" = length(gene_exp_files) > 0
+)
+```
+
+```{r}
 # list of all validation markers
@@
   )
+
+# remove the large data frames 
+rm(consensus_df)
+rm(gene_exp_df)
 ```
@@
 
+# no longer need this and it takes up space 
+rm(combined_df)
+
 # make a data frame that just has the unique genes 
@@
 Only cell type groups with at least 50 cells will be shown.
-
-Note that I did try and look at all marker genes, without specifying that they must be unique, and the plots were overwhelming with lots of genes. 
-Instead I included the a separate section looking at the mean expression across all markers for a given group.
-
-### Dot plot
 
@@
 
-### Heatmap
-
-The heatmap shows the mean expression of each marker gene across all cells assigned to a given cell type group. 
-In contrast to the dot plot, we will show all marker genes for a given cell type group, not just the uniquely expressed genes.
-
-```{r}
-# add column to uniquely label gene symbols since some genes are present across multiple cell types 
-# add this here since we will use the order in this data frame to set the order of the heatmap columns
-heatmap_markers_df <- filtered_markers_df |> 
-  dplyr::mutate(
-    unique_symbol = glue::glue("{gene_symbol}-{validation_group_annotation}")
-  ) 
-```
-
-```{r}
-# first make a mtx to use for the heatmap with rows as cell types and genes as columns 
-heatmap_mtx <- group_stats_df |> 
-  dplyr::ungroup() |> 
-  # add unique label for gene symbol 
-  dplyr::mutate(
-    unique_symbol = glue::glue("{gene_symbol}-{validation_group_annotation}")
-  ) |> 
-  dplyr::select(unique_symbol, broad_celltype_group, mean_exp) |> 
-  dplyr::filter(unique_symbol %in% heatmap_markers_df$unique_symbol) |> 
-  tidyr::pivot_wider(
-    names_from = unique_symbol,
-    values_from = mean_exp
-  ) |> 
-  tibble::column_to_rownames("broad_celltype_group") |>
-  as.matrix()
-# make sure cell types are present in the right order
-heatmap_mtx <- heatmap_mtx[c(celltype_order, "unknown"), heatmap_markers_df$unique_symbol]
-
-# create annotation for heatmap
-annotation <- ComplexHeatmap::columnAnnotation(
-  group = heatmap_markers_df$validation_group_annotation,
-  col = list(
-    group = celltype_colors
-  ),
-  annotation_legend_param = list(
-      title = "Validation group"
-    )
-)
-```
-
-```{r, fig.height=7, fig.width=10}
-marker_gene_heatmap(heatmap_mtx, annotation)
-```
-
 ## Mean expression across all marker genes
@@
 
-## Conclusions
-
-- Regardless of looking at individual genes or the mean expression of all genes, we see similar patterns where generally the cells assigned to a given cell type show expression of the appropriate marker genes when compared to other cells.
-- There seems to be pretty clear distinction of T cells, B cells, and NK cells. 
-- Monocytes, macrophages, and myeloid cells are generally hard to distinguish from each other and all share similar marker genes. 
-Perhaps we could collapse these into myeloid, but I do see expression of monocyte specific genes in monocytes only and maybe some macrophage specific genes in macrophages only, so it's probably okay to keep separate.
-I think that is more obvious in the heatmaps than in the dot plots. 
-- Neurons and astrocytes both show expression of both neuron and astrocyte marker genes. 
-It does look like there are a set of genes that are more highly expressed in the astrocyte population though, so I think we can keep those gene sets separate. 
-- Here the unknown cells don't show strong expression of any specific set of genes, but it does look like they show more expression of neuron/astrocyte markers than any of the other marker genes in the plot.
-This makes sense since we are looking at brain tumors and I expect similarity between tumor cells and neural cell types.
-
 ## Session info

### 2025-02-26 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
keep HPC markers even though they aren't unique

@@
 unique_gene_df <- group_stats_df |> 
-  dplyr::filter(gene_observed_count == 1)
+  dplyr::filter(gene_observed_count == 1 | validation_group_annotation == "hematopoietic precursor cell")
 ```

### 2025-02-26 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
make more memory efficient and speedier

--- 2025-02-26 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
@@
 params:
-  project_id: "SCPCP000002"
+  project_id: "SCPCP000001"
 ---
@@
   library(patchwork)
+  library(data.table)
 })
@@
 # read in consensus and gene expression results 
-consensus_df <- consensus_results_files |> 
-  purrr::map(readr::read_tsv) |>
-  dplyr::bind_rows()
+# use vroom and data.table to try and be more memory-efficient
+consensus_dt <- consensus_results_files |> 
+  purrr::map(vroom::vroom) |> 
+  data.table::rbindlist(fill = TRUE)
 
-gene_exp_df <- gene_exp_files |> 
-  purrr::map(readr::read_tsv) |>
-  dplyr::bind_rows()
+gene_exp_dt <- gene_exp_files |> 
+  purrr::map(vroom::vroom) |> 
+  data.table::rbindlist(fill = TRUE)
 
-# read in validation markers 
-markers_df <- readr::read_tsv(validation_markers_file)
+# read in validation markers as data.tables
+markers_dt <- fread(validation_markers_file)
 
-validation_groups_df <- readr::read_tsv(validation_groups_file) |> 
+validation_groups_dt <- fread(validation_groups_file) |> 
   # rename final assigned group to avoid conflicts when merging in marker gene expression 
@@
 # includes original consensus annotation, braod cell type group, marker gene id, logcounts, validation group annotation, information regarding expression of marker gene in CellMarker tissues 
-combined_df <- consensus_df |> 
+consensus_dt <- consensus_dt |> 
   # add in broad cell type group which is used for plotting
   # groups similar cell types together 
-  dplyr::left_join(validation_groups_df, by = c("consensus_annotation")) |> 
-  dplyr::left_join(gene_exp_df, by = c("barcodes", "library_id")) |> 
+  merge(validation_groups_dt, by = "consensus_annotation", all.x = TRUE) |> 
+  merge(gene_exp_dt, by = c("barcodes", "library_id"), all.x = TRUE) |> 
   # add marker gene information (associated validation group annotation, gene observed count, percent tissues)
   # account for the same gene being present in multiple cell types 
-  dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |> 
-  dplyr::rowwise() |> 
-  dplyr::mutate(
-    detected = logcounts > 0 # column indicating if gene is present or not
-  )
+  merge(markers_dt, by = "ensembl_gene_id", all.x = TRUE, allow.cartesian = TRUE)
 
-# remove the large data frames 
-rm(consensus_df)
-rm(gene_exp_df)
+# add a column indicating if gene is present or not
+consensus_dt[, detected := logcounts > 0]
+
+# remove the extra data table 
+rm(gene_exp_dt)
 ```
@@
 # get total number of cells per final annotation group 
-total_cells_df <- combined_df |> 
+total_cells_df <- consensus_dt |> 
   dplyr::select(library_id, barcodes, broad_celltype_group) |> 
@@
 # second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
-group_stats_df <- combined_df |> 
+group_stats_df <- consensus_dt |> 
   # for each assigned cell type/marker gene combo get total detected and mean expression
@@
   # this includes all possible marker genes and all possible validation group assignments 
-  dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |>
+  dplyr::left_join(markers_dt, by = c("ensembl_gene_id"), relationship = "many-to-many") |>
   # now get the mean expression/ mean percentage across all marker genes for a given validation group
@@
     broad_celltype_group = tidyr::replace_na(broad_celltype_group, "unknown") |> 
-      factor(levels =  c(unique(markers_df$validation_group_annotation), "unknown"))
+      factor(levels =  c(unique(markers_dt$validation_group_annotation), "unknown"))
   )
@@
 # no longer need this and it takes up space 
-rm(combined_df)
+rm(consensus_dt)
 
@@
 unique_gene_df <- group_stats_df |> 
-  dplyr::filter(gene_observed_count == 1)
+  # keep all 6 HPC genes 
+  dplyr::filter(gene_observed_count == 1 | validation_group_annotation == "hematopoietic precursor cell")
 ```
@@
 # note that we will use this for both the dotplot and the heatmap
-filtered_markers_df <- markers_df |> 
+filtered_markers_df <- markers_dt |> 
   dplyr::filter(validation_group_annotation %in% celltype_groups,

### 2025-02-26 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
one more HPC filtering

@@
 unique_markers_df <- filtered_markers_df |> 
-  dplyr::filter(gene_observed_count == 1)
+  dplyr::filter(gene_observed_count == 1 | validation_group_annotation == "hematopoietic precursor cell")
 

### 2025-02-26 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
make note about keeping HPCs

@@
 First we will look at the expression of marker genes that are specific to a cell type group.  
-All marker genes shown in the following plots are only found to be a marker gene for a single cell type group. 
+All marker genes shown in the following plots are only found to be a marker gene for a single cell type group with the exception of marker genes for hematopoietic precursor cells, as none of those genes are unique to a single cell type and may also be found in other immune cells.
 Only cell type groups with at least 50 cells will be shown.

### 2025-02-26 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
account for cell line samples

@@
 params:
-  project_id: "SCPCP000001"
+  project_id: "SCPCP000020"
 ---
@@
 
-```{r}
+```{r, results='asis'}
+# check for if the project is all cell lines or not 
+is_cell_line <- all(consensus_dt$sample_type == "cell line")
+
+if(is_cell_line){
+  glue::glue("
+    <div class=\"alert alert-info\">
+
+    All samples in this project are from cell lines. 
+    No cell typing was performed, so no plots will be shown.
+
+    </div>
+  ")
+}
+```
+
+```{r, eval=!is_cell_line}
 # combine into one data frame that has all marker genes
@@
 
-```{r}
+```{r, eval=!is_cell_line}
 # prep for plots 
@@
 
-```{r}
+```{r, eval=!is_cell_line}
 # table with one row per unique broad cell type/ marker gene combination 
@@
 
-```{r}
+```{r, eval=!is_cell_line}
 # get list of celltypes to keep and assign colors 
@@
 
-```{r}
+```{r, eval=!is_cell_line}
 # filter markers to those that are actually relevant 
@@
 
-```{r, fig.height=10, fig.width=15}
+```{r, eval=!is_cell_line, fig.height=10, fig.width=15}
 # filter out low expressed genes
@@
 
-```{r}
+```{r, eval=!is_cell_line}
 # filter low expressed genes 

### 2025-02-26 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
make sure percent is between 0 and 100 and print out diagnoses

@@
 params:
-  project_id: "SCPCP000020"
+  project_id: "SCPCP000001"
 ---
@@
 
-```{r}
+```{r, results='asis', message=FALSE}
+# print out diagnoses for project 
+project_diagnoses_file <- file.path(module_base, "sample-info", "project-diagnoses.tsv")
+diagnosis <- readr::read_tsv(project_diagnoses_file) |> 
+  dplyr::filter(project_id == params$project_id) |> 
+  dplyr::pull(diagnosis)
+
+glue::glue("
+    <div class=\"alert alert-info\">
+
+    This project contains {diagnosis} samples.
+
+    </div>
+  ")
+```
+
+```{r, warning=FALSE}
 # get consensus results 
@@
 
-```{r, message=FALSE}
+```{r, message=FALSE, warning=FALSE}
 # read in consensus and gene expression results 
@@
 
-```{r, eval=!is_cell_line}
+```{r, eval=!is_cell_line, message=FALSE}
 # combine into one data frame that has all marker genes
@@
   # for each assigned cell type/marker gene combo get total detected and mean expression
-  dplyr::group_by(broad_celltype_group, ensembl_gene_id) |>
+  # group by both broad group and validation group to account for genes that are expressed in more than one cell type
+  dplyr::group_by(broad_celltype_group, ensembl_gene_id, validation_group_annotation) |>
   dplyr::summarize(
@@
   # this includes all possible marker genes and all possible validation group assignments 
-  dplyr::left_join(markers_dt, by = c("ensembl_gene_id"), relationship = "many-to-many") |>
+  dplyr::left_join(markers_dt, by = c("ensembl_gene_id", "validation_group_annotation"), relationship = "many-to-many") |>
   # now get the mean expression/ mean percentage across all marker genes for a given validation group

### 2025-02-27 analyses/cell-type-wilms-tumor-06/07_combined_annotation_across_samples_exploration.Rmd
notebook after review

--- 2025-02-12 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
@@
 
-_Where `cnv.thr` and `pred.thr` need to be discussed_
+| first level annotation | second level annotation | selection of the cells                                                                        | marker genes for validation | CNV validation                             |
+| ---------------------- | ----------------------- | --------------------------------------------------------------------------------------------- | --------------------------- | ------------------------------------------ |
+| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr`                                   | `VWF`                       | no CNV                                     |
+| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr`                                        | `PTPRC`, `CD163`, `CD68`    | no CNV                                     |
+| normal                 | kidney                  | `compartment == "fetal_nephron" & predicted.score > pred.thr & cnv_score < cnv.thr.lower`     | `CDH1`, `PODXL`, `LTL`      | no CNV                                     |
+| normal                 | stroma                  | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr.upper`            | `VIM`                       | no CNV                                     |
+| cancer                 | stroma                  | `compartment == "stroma" & cnv_score > cnv.thr.upper`                                         | `VIM`                       | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| cancer                 | blastema                | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr.upper`| `CITED1`                    | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| cancer                 | epithelial              | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr.upper`| `CDH1`                      | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| unknown                | -                       | the rest of the cells                                                                         | -                           | -                                          |
 
-| first level annotation | second level annotation | selection of the cells | marker genes for validation | CNV validation |
-| ---------------------- | ----------------------- | ---------------------- | --------------------------- | --------------- |
-| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr`  | `VWF`| no CNV |
-| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr`  | `PTPRC`, `CD163`, `CD68`| no CNV |
-| normal                 | kidney                  | `compartment == "fetal_nephron" & predicted.score > pred.thr & cnv_score < cnv.thr` | `CDH1`, `PODXL`, `LTL`| no CNV |
-| normal                 | stroma                  | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr`| `VIM`| no CNV |
-| cancer                 | stroma                  | `compartment == "stroma" & cnv_score > cnv.thr` | `VIM`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
-| cancer                 | blastema                | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr`  | `CITED1`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
-| cancer                 | epithelial              | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr`  | `CDH1`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
-| unknown                | -                       | the rest of the cells  | -  | -|
+with the defined parameters:
+
+- `pred.thr` = `r params$predicted.celltype.threshold`
+
+- `cnv.thr.lower` = `r params$cnv_threshold_low`
+
+- `cnv.thr.upper` = `r params$cnv_threshold_high`
 
@@
 
+```{r }
+# sample metadata
+sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
+metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
+```
+
 We extracted from the pre-processed and labeled `Seurat` object from `results/{sample_id}/06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds`.
@@
 
-- the predicted compartment in `fetal_kidney_predicted.cell_type` and `fetal_kidney_predicted.cell_type.score`
+- the predicted cell type in `fetal_kidney_predicted.cell_type` and `fetal_kidney_predicted.cell_type.score`
 
@@
 - the `counts` of [markers genes](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-wilms-tumor-06#the-table-celltype_metadatacsv-contains-the-following-column-and-information)
-
-We will also use:
-
-- the samples metadata or clinical data reported in `r file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")`
-
-- the cytoband file to get the length of each chromosome arm (for plotting only)
-
-```{r input}
-# sample metadata
-sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
-metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
-
-# the cytoband file that will be used to extract the size of each chromosome arm for plotting only
-arm_order_file <- file.path(module_base, "results", "references", "hg38_cytoBand.txt")
-
-```
 
@@
       data.frame(
-        # label transfer from the fetal kidney reference
-        cell_type = srat$fetal_kidney_predicted.cell_type,
-        compartment = srat$fetal_kidney_predicted.compartment,
-
-        # predicted.scores from the label transfer from the fetal kidney reference
-        cell_type.score = srat$fetal_kidney_predicted.cell_type.score,
-        compartment.score = srat$fetal_kidney_predicted.compartment.score,
+        srat@meta.data,
 
@@
         NCAM1 = FetchData(object = srat, vars = "ENSG00000149294", layer = "counts"),
-        THY1 = FetchData(object = srat, vars = "ENSG00000154096", layer = "counts"),
+        THY1 = FetchData(object = srat, vars = "ENSG00000154096", layer = "counts")
 
-        # loss global estimation per chromosome
-        has_loss_chr1p = srat$has_loss_chr1p,
-        has_loss_chr2p = srat$has_loss_chr2p,
-        has_loss_chr3p = srat$has_loss_chr3p,
-        has_loss_chr4p = srat$has_loss_chr4p,
-        has_loss_chr5p = srat$has_loss_chr5p,
-        has_loss_chr6p = srat$has_loss_chr6p,
-        has_loss_chr7p = srat$has_loss_chr7p,
-        has_loss_chr8p = srat$has_loss_chr8p,
-        has_loss_chr9p = srat$has_loss_chr9p,
-        has_loss_chr10p = srat$has_loss_chr10p,
-        has_loss_chr11p = srat$has_loss_chr11p,
-        has_loss_chr12p = srat$has_loss_chr12p,
-
-        has_loss_chr16p = srat$has_loss_chr16p,
-        has_loss_chr17p = srat$has_loss_chr17p,
-        has_loss_chr18p = srat$has_loss_chr18p,
-        has_loss_chr19p = srat$has_loss_chr19p,
-        has_loss_chr20p = srat$has_loss_chr20p,
-
-        has_loss_chr1q = srat$has_loss_chr1q,
-        has_loss_chr2q = srat$has_loss_chr2q,
-        has_loss_chr3q = srat$has_loss_chr3q,
-        has_loss_chr4q = srat$has_loss_chr4q,
-        has_loss_chr5q = srat$has_loss_chr5q,
-        has_loss_chr6q = srat$has_loss_chr6q,
-        has_loss_chr7q = srat$has_loss_chr7q,
-        has_loss_chr8q = srat$has_loss_chr8q,
-        has_loss_chr9q = srat$has_loss_chr9q,
-        has_loss_chr10q = srat$has_loss_chr10q,
-        has_loss_chr11q = srat$has_loss_chr11q,
-        has_loss_chr12q = srat$has_loss_chr12q,
-        has_loss_chr13q = srat$has_loss_chr13q,
-        has_loss_chr14q = srat$has_loss_chr14q,
-        has_loss_chr15q = srat$has_loss_chr15q,
-        has_loss_chr16q = srat$has_loss_chr16q,
-        has_loss_chr17q = srat$has_loss_chr17q,
-        has_loss_chr18q = srat$has_loss_chr18q,
-        has_loss_chr19q = srat$has_loss_chr19q,
-        has_loss_chr20q = srat$has_loss_chr20q,
-        has_loss_chr21q = srat$has_loss_chr21q,
-        has_loss_chr22q = srat$has_loss_chr22q,
-
-        # dupli global estimation per chromosome
-        has_dupli_chr1p = srat$has_dupli_chr1p,
-        has_dupli_chr2p = srat$has_dupli_chr2p,
-        has_dupli_chr3p = srat$has_dupli_chr3p,
-        has_dupli_chr4p = srat$has_dupli_chr4p,
-        has_dupli_chr5p = srat$has_dupli_chr5p,
-        has_dupli_chr6p = srat$has_dupli_chr6p,
-        has_dupli_chr7p = srat$has_dupli_chr7p,
-        has_dupli_chr8p = srat$has_dupli_chr8p,
-        has_dupli_chr9p = srat$has_dupli_chr9p,
-        has_dupli_chr10p = srat$has_dupli_chr10p,
-        has_dupli_chr11p = srat$has_dupli_chr11p,
-        has_dupli_chr12p = srat$has_dupli_chr12p,
-
-        has_dupli_chr16p = srat$has_dupli_chr16p,
-        has_dupli_chr17p = srat$has_dupli_chr17p,
-        has_dupli_chr18p = srat$has_dupli_chr18p,
-        has_dupli_chr19p = srat$has_dupli_chr19p,
-        has_dupli_chr20p = srat$has_dupli_chr20p,
-
-        has_dupli_chr1q = srat$has_dupli_chr1q,
-        has_dupli_chr2q = srat$has_dupli_chr2q,
-        has_dupli_chr3q = srat$has_dupli_chr3q,
-        has_dupli_chr4q = srat$has_dupli_chr4q,
-        has_dupli_chr5q = srat$has_dupli_chr5q,
-        has_dupli_chr6q = srat$has_dupli_chr6q,
-        has_dupli_chr7q = srat$has_dupli_chr7q,
-        has_dupli_chr8q = srat$has_dupli_chr8q,
-        has_dupli_chr9q = srat$has_dupli_chr9q,
-        has_dupli_chr10q = srat$has_dupli_chr10q,
-        has_dupli_chr11q = srat$has_dupli_chr11q,
-        has_dupli_chr12q = srat$has_dupli_chr12q,
-        has_dupli_chr13q = srat$has_dupli_chr13q,
-        has_dupli_chr14q = srat$has_dupli_chr14q,
-        has_dupli_chr15q = srat$has_dupli_chr15q,
-        has_dupli_chr16q = srat$has_dupli_chr16q,
-        has_dupli_chr17q = srat$has_dupli_chr17q,
-        has_dupli_chr18q = srat$has_dupli_chr18q,
-        has_dupli_chr19q = srat$has_dupli_chr19q,
-        has_dupli_chr20q = srat$has_dupli_chr20q,
-        has_dupli_chr21q = srat$has_dupli_chr21q,
-        has_dupli_chr22q = srat$has_dupli_chr22q
       ) |>
@@
     }
+
   ) |>
@@
 
+We will also use:
+
+- the samples metadata or clinical data reported in `r file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")`
+
 ```{r }
@@
   left_join( metadata, by= c("sample_id" = "scpca_sample_id"))
+
+```
+
+- the cytoband file to get the length of each chromosome arm (for plotting only)
+
+```{r input}
+# the cytoband file that will be used to extract the size of each chromosome arm for plotting only
+arm_order_file <- file.path(module_base, "results", "references", "hg38_cytoBand.txt")
+# Load cytoBand file
+cytoBand <- read.table(gzfile(arm_order_file), header = FALSE, sep = "\t", stringsAsFactors = FALSE)
+colnames(cytoBand) <- c("chrom", "start", "end", "band", "stain")
+# Here we define the size of each chromosome arm that will be used for plotting of the mean CNV profile
+  # Extract chromosome arm information
+  cytoBand <- cytoBand |>
+    mutate(arm = substr(band, 1, 1))  # Extract 'p' or 'q' from the band column
+
+  # Define arm boundaries
+  chromosome_arms <- cytoBand |>
+    group_by(chrom, arm) |>
+    summarize(
+      Start = min(start),
+      End = max(end),
+      .groups = "drop"
+    ) |>
+    mutate(Size = End - Start,
+           dupli = paste0("has_dupli_", chrom, arm),
+           loss = paste0("has_loss_", chrom, arm),
+           chrom_arm = glue::glue("{chrom}{arm}")) |> 
+    mutate(Size = Size / sum(Size)) |>
+    # Define chromosome arm order
+    mutate(chrom_arm = factor(chrom_arm, levels = c(paste0("chr", rep(1:22, each = 2), c("p", "q")),
+                                                  "chrXp", "chrXq", "chrYp", "chrYq"))) |>
+    # Sort genes by Chromosome arm and Start position
+    arrange(chrom_arm, Start)  |> 
+    filter(arm != "")
 
@@
 ```{r fig.width=20, fig.height=15, out.width='100%', results='asis'}
-
 do_BoxPlot_mean <- function(df, group.by, comparison){
@@
 
-We simply checked for each chromosome if the cell `has_cnv_chr`.
-Would the cell have less than `r params$cnv_threshold_low` chromosome with CNV, the global `has_cnv_score` will be FALSE.
-Would the cell have more than `r params$cnv_threshold_high`  chromosome with CNV, the global `has_cnv_score` will be TRUE.
+We simply checked for each chromosome the cell `cnv_chr`.
+Would the cell have less than `r params$cnv_threshold_low` chromosome with CNV, the global `cnv_score` will be FALSE.
+Would the cell have more than `r params$cnv_threshold_high`  chromosome with CNV, the global `cnv_score` will be TRUE.
 The default is set to `unknown`.
@@
 cell_type_df <- cell_type_df |>
-  mutate(has_cnv_score = rowSums(cell_type_df[, grepl("has_loss_chr|has_dupli_chr", colnames(cell_type_df))])) |>
-  mutate(has_cnv_score = case_when(
-    has_cnv_score > params$cnv_threshold_high ~ "CNV",
-    has_cnv_score <= params$cnv_threshold_low ~ "no CNV",
+  mutate(cnv_score = rowSums(cell_type_df[, grepl("has_loss_chr|has_dupli_chr", colnames(cell_type_df))], na.rm = TRUE)) |>
+  mutate(cnv_score = case_when(
+    cnv_score > params$cnv_threshold_high ~ "CNV",
+    cnv_score <= params$cnv_threshold_low ~ "no CNV",
     .default = "unknown"
@@
     # assign normal/cancer based on condition
-    compartment %in% c("fetal_nephron", "stroma") & 
-      has_cnv_score == "no CNV" ~ "normal",
+    fetal_kidney_predicted.compartment %in% c("fetal_nephron", "stroma") & 
+    cnv_score == "no CNV" ~ "normal",
 
-    compartment %in% c("normal", "immune", "endothelium") &
-      cell_type.score > params$predicted.celltype.threshold ~ "normal",
+    fetal_kidney_predicted.compartment %in% c("normal", "immune", "endothelium") &
+    fetal_kidney_predicted.compartment.score > params$predicted.celltype.threshold ~ "normal",
 
-    compartment %in% c("fetal_nephron", "stroma") & 
-      has_cnv_score == "CNV" ~ "cancer",
+    fetal_kidney_predicted.compartment %in% c("fetal_nephron", "stroma") & 
+    cnv_score == "CNV" ~ "cancer",
 
@@
 Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` _cancer_ cells, `r table(cell_type_df$first.level_annotation)["normal"]` _normal_ cells.
-`r table(cell_type_df$first.level_annotation)["unknown"]`cells remain _unknown_
+`r table(cell_type_df$first.level_annotation)["unknown"]` cells remain _unknown_
 
 ```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
-ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = first.level_annotation), shape = 19, size = 1) +
-  geom_point() +
+ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = first.level_annotation)) +
+  geom_point(size = 0.5, shape = 19, alpha = 0.5) +
   facet_wrap(facets = ~sample_id, ncol = 5) +
@@
 
-DT::datatable(table(cell_type_df$first.level_annotation, cell_type_df$sample_id))
+cell_type_df |>
+  dplyr::count(first.level_annotation, sample_id) |>
+  DT::datatable()
 
@@
 These five samples do not have enough immune and endothelium cell to be used as a normal reference in `scripts/06_infercnv.R` and we previously decided to run `scripts/06_infercnv.R` without any reference for these samples.
-In that case, the mean expression profile over the sample is taken as the normal reference, biaising the inference of CNV.
+In that case, the mean expression profile over the sample is taken as the normal reference, biasing the inference of CNV.
 
@@
 
-cell_type_df$first.level_annotation[cell_type_df$sample_id %in% c("SCPCS000177", "SCPCS000180", "SCPCS000181", "SCPCS000190", "SCPCS000197")] <- "unknown"
+cell_type_df$first.level_annotation[cell_type_df$sample_id %in% none_reference_samples] <- "unknown"
 
@@
     # assign normal cells based on condition
-    cell_type_df$compartment %in% c("fetal_nephron") &
-      cell_type_df$has_cnv_score == "no CNV" ~ "kidney",
+      cell_type_df$fetal_kidney_predicted.compartment   == c("fetal_nephron") &
+      cell_type_df$cnv_score                            == "no CNV" ~ "kidney",
 
-    cell_type_df$compartment %in% c("stroma") &
-      cell_type_df$has_cnv_score == "no CNV" ~ "normal stroma",
+      cell_type_df$fetal_kidney_predicted.compartment   == c("stroma") &
+      cell_type_df$cnv_score                            == "no CNV" ~ "normal stroma",
 
-    cell_type_df$compartment %in% c("normal") &
-      cell_type_df$cell_type %in% c("endothelial cell") ~ "endothelium",
-    cell_type_df$compartment %in% c("normal") &
-      cell_type_df$cell_type %in% c("B cell",
+      cell_type_df$fetal_kidney_predicted.compartment   == c("normal") &
+      cell_type_df$fetal_kidney_predicted.cell_type     == c("endothelial cell") ~ "endothelium",
+
+      cell_type_df$fetal_kidney_predicted.compartment   == c("normal") &
+      cell_type_df$fetal_kidney_predicted.cell_type    %in% c("B cell",
                                     "CD4-positive, alpha-beta T cell",
@@
     # assign cancer cells based on condition
-    cell_type_df$compartment %in% c("stroma") &
-      cell_type_df$has_cnv_score == "CNV" ~ "cancer stroma",
-    cell_type_df$compartment %in% c("fetal_nephron") &
-      cell_type_df$has_cnv_score == "CNV" &
-      cell_type_df$cell_type == "mesenchymal cell" ~ "blastema",
-    cell_type_df$compartment %in% c("fetal_nephron") &
-      cell_type_df$has_cnv_score == "CNV" &
-      cell_type_df$cell_type != "mesenchymal cell" ~ "cancer epithelium",
+      cell_type_df$fetal_kidney_predicted.compartment    == c("stroma") &
+      cell_type_df$cnv_score                             == "CNV" ~ "cancer stroma",
+
+      cell_type_df$fetal_kidney_predicted.compartment    == c("fetal_nephron") &
+      cell_type_df$cnv_score                             == "CNV" &
+      cell_type_df$fetal_kidney_predicted.cell_type      == "mesenchymal cell" ~ "blastema",
+
+      cell_type_df$fetal_kidney_predicted.compartment    == c("fetal_nephron") &
+      cell_type_df$cnv_score                             == "CNV" &
+      cell_type_df$fetal_kidney_predicted.cell_type      != "mesenchymal cell" ~ "cancer epithelium",
+
+      sample_id                                         %in% none_reference_samples ~ "unknown",
+
     .default = "unknown"
@@
 
-cell_type_df$second.level_annotation[cell_type_df$sample_id %in% c("SCPCS000177", "SCPCS000180", "SCPCS000181", "SCPCS000190", "SCPCS000197")] <- "unknown"
 ```
@@
 ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
-  geom_point(size = 1, shape = 19) +
+  geom_point(size = 0.5, shape = 19, alpha = 0.5) +
   facet_wrap(facets = ~sample_id, ncol = 5) +
@@
 ggplot(cell_type_df_sub, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
-  geom_point(size = 1, shape = 19) +
+  geom_point(size = 0.5, shape = 19, alpha = 0.5) +
   facet_wrap(facets = ~sample_id, ncol = 5) +
@@
 For a sample, a CNV is defined as `clonal` (opaque) if detected in more than 70% of the cells;
-and `subclonal` (transparent) when detected in 20 to 70% of the cells.
+and `subclonal` (transparent) when detected in at least 10% of the cells.
 
 ```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
+# Here we define the vector of ordered chromosomes arms that will be used for plotting
+has_dupli_chr <- names(cell_type_df)[stringr::str_starts(names(cell_type_df), "has_dupli_chr")]
 
-# Here we define the vector of ordered chromosomes arms that will be used for plotting
-
-has_dupli_chr <- c("has_dupli_chr1p",                                      
-                   "has_dupli_chr1q",
-
-                   "has_dupli_chr2p",
-                   "has_dupli_chr2q",
-
-                   "has_dupli_chr3p",
-                   "has_dupli_chr3q",
-
-                   "has_dupli_chr4p",
-                   "has_dupli_chr4q",
-
-                   "has_dupli_chr5p",
-                   "has_dupli_chr5q",
-
-                   "has_dupli_chr6p",
-                   "has_dupli_chr6q",
-
-                   "has_dupli_chr7p",
-                   "has_dupli_chr7q",
-
-                   "has_dupli_chr8p",
-                   "has_dupli_chr8q",
-
-                   "has_dupli_chr9p",
-                   "has_dupli_chr9q",
-
-                   "has_dupli_chr10p",
-                   "has_dupli_chr10q",
-
-                   "has_dupli_chr11p",
-                   "has_dupli_chr11q",
-
-                   "has_dupli_chr12p",
-                   "has_dupli_chr12q",
-
-                   "has_dupli_chr13p",
-                   "has_dupli_chr13q",
-
-                   "has_dupli_chr14p",
-                   "has_dupli_chr14q",
-
-                   "has_dupli_chr15p",
-                   "has_dupli_chr15q",
-
-                   "has_dupli_chr16p",
-                   "has_dupli_chr16q",
-
-                   "has_dupli_chr17p",
-                   "has_dupli_chr17q",
-
-                   "has_dupli_chr18p",
-                   "has_dupli_chr18q",
-
-                   "has_dupli_chr19p",
-                   "has_dupli_chr19q",
-
-                   "has_dupli_chr20p",
-                   "has_dupli_chr20q",
-
-                   "has_dupli_chr21p",
-                   "has_dupli_chr21q",
-
-                   "has_dupli_chr22p",
-                   "has_dupli_chr22q"
-                   )
-
-#################################################################################
-
-# define the vector of ordered chromosomes arms##################################
-has_loss_chr <-  c("has_loss_chr1p",                                        
-                   "has_loss_chr1q",
-
-                   "has_loss_chr2p",
-                   "has_loss_chr2q",
-
-                   "has_loss_chr3p",
-                   "has_loss_chr3q",
-
-                   "has_loss_chr4p",
-                   "has_loss_chr4q",
-
-                   "has_loss_chr5p",
-                   "has_loss_chr5q",
-
-                   "has_loss_chr6p",
-                   "has_loss_chr6q",
-
-                   "has_loss_chr7p",
-                   "has_loss_chr7q",
-
-                   "has_loss_chr8p",
-                   "has_loss_chr8q",
-
-                   "has_loss_chr9p",
-                   "has_loss_chr9q",
-
-                   "has_loss_chr10p",
-                   "has_loss_chr10q",
-
-                   "has_loss_chr11p",
-                   "has_loss_chr11q",
-
-                   "has_loss_chr12p",
-                   "has_loss_chr12q",
-
-                   "has_loss_chr13p",
-                   "has_loss_chr13q",
-
-                   "has_loss_chr14p",
-                   "has_loss_chr14q",
-
-                   "has_loss_chr15p",
-                   "has_loss_chr15q",
-
-                   "has_loss_chr16p",
-                   "has_loss_chr16q",
-
-                   "has_loss_chr17p",
-                   "has_loss_chr17q",
-
-                   "has_loss_chr18p",
-                   "has_loss_chr18q",
-
-                   "has_loss_chr19p",
-                   "has_loss_chr19q",
-
-                   "has_loss_chr20p",
-                   "has_loss_chr20q",
-
-                   "has_loss_chr21p",
-                   "has_loss_chr21q",
-
-                   "has_loss_chr22p",
-                   "has_loss_chr22q"
-                   )
-
-#################################################################################
-# Here we define the size of each chromosome arm that will be used for plotting of the mean CNV profile
-
-  # Load cytoBand file
-  cytoBand <- read.table(gzfile(arm_order_file), header = FALSE, sep = "\t", stringsAsFactors = FALSE)
-  colnames(cytoBand) <- c("chrom", "start", "end", "band", "stain")
-
-  # Extract chromosome arm information
-  cytoBand <- cytoBand |>
-    mutate(arm = substr(band, 1, 1))  # Extract 'p' or 'q' from the band column
-
-  # Define arm boundaries
-  chromosome_arms <- cytoBand |>
-    group_by(chrom, arm) |>
-    summarize(
-      Start = min(start),
-      End = max(end),
-      .groups = "drop"
-    ) |>
-    mutate(Size = End - Start,
-           dupli = paste0("has_dupli_", chrom, arm),
-           loss = paste0("has_loss_", chrom, arm)) |> 
-  filter(arm != "")
-
-  chromosome_arms <- chromosome_arms |>
-    mutate(Size = Size / sum(Size))
+# define the vector of ordered chromosomes arms
+has_loss_chr <- names(cell_type_df)[stringr::str_starts(names(cell_type_df), "has_loss_chr")]
 
@@
 
-```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-
+```{r fig.height=10, fig.width=20, message=FALSE, warning=FALSE, out.width='100%', results='asis'}
 # prepare the table for the number of sample with a (sub)clonal loss
@@
                     summarise_all(.funs = mean) |>
-  pivot_longer(cols = -sample_id) |>
-  rename("chromosome_arm" = "name", "percentage" = "value") |>
-  left_join( chromosome_arms, by = c("chromosome_arm"="loss")) |>
+                    pivot_longer(
+                      cols = -sample_id, 
+                      names_to = "chromosome_arm", 
+                      values_to = "percentage") |>
+                    left_join(chromosome_arms, by = c("chromosome_arm"="loss")) |>
                     mutate(chromosome_arm = factor(chromosome_arm, levels = has_loss_chr)) |>
-  mutate(absolute_70 = case_when((percentage > 0.7) ~ -1,
-         .default = 0),
-         absolute_10 = case_when((percentage > 0.1) ~ -1,
-         .default = 0))
+                    mutate(
+                      absolute_70 = ifelse(percentage > 0.7, -1, 0), 
+                      absolute_10 = ifelse(percentage > 0.1, -1, 0))
 
@@
                     summarise_all(.funs = mean) |>
-  pivot_longer(cols = -sample_id) |>
-  rename("chromosome_arm" = "name", "percentage" = "value") |>
-  left_join( chromosome_arms, by = c("chromosome_arm"="dupli")) |>
+                    pivot_longer(
+                      cols = -sample_id, 
+                      names_to = "chromosome_arm", 
+                      values_to = "percentage") |>
+                    left_join(chromosome_arms, by = c("chromosome_arm"="dupli")) |>
                     mutate(chromosome_arm = factor(chromosome_arm, levels = has_dupli_chr)) |>
-  mutate(absolute_70 = case_when((percentage > 0.7) ~ 1,
-         .default = 0),
-         absolute_10 = case_when((percentage > 0.1) ~ 1,
-         .default = 0))
+                    mutate(
+                      absolute_70 = ifelse(percentage > 0.7, 1, 0), 
+                      absolute_10 = ifelse(percentage > 0.1, 1, 0))
 
 # plot
+# Here we chose to loop on chromosome arms to then be able to scale each bar to the length of the corresponding genomic region using wrap_plot. 
+# Of note, the resulting plot is way better than the one obtained with facet_wrap or facet_grid so far.
 
-loss_profile <- ggplot(CNV_profile_loss, aes(x = chromosome_arm, y = absolute_10, width = 20 * Size)) +
+loss_profile <- list()
+for(chr in unique(CNV_profile_loss$chromosome_arm)){
+  loss_profile[[chr]] <- ggplot(CNV_profile_loss[CNV_profile_loss$chromosome_arm == chr,], 
+                                aes(x = chromosome_arm, y = absolute_10)) +
                         geom_bar(stat = "identity",  fill = "#41b6c4", alpha = 0.4) +
-  geom_bar(aes(x = chromosome_arm, y = absolute_70, width = 20 * Size), stat = "identity",  fill = "#41b6c4") +
-  scale_x_discrete(labels = gsub("has_loss_", "", levels(droplevels(CNV_profile_loss$chromosome_arm))))  +
-  theme_classic() +
-  ylab("Number of occurrence") +
-  xlab("chromosom arms") +
+                        geom_bar(aes(x = chromosome_arm, y = absolute_70), stat = "identity",  fill = "#41b6c4") +
+                        scale_x_discrete(labels = gsub("has_loss_", "", chr))  +
+                        theme_void() +
+                        ylim(c(-40,0)) +
                         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12), text = element_text(size = 15))
+}
+loss_profile <- wrap_plots(loss_profile, nrow = 1, widths = chromosome_arms$Size[chromosome_arms$loss %in% unique(CNV_profile_loss$chromosome_arm)], guides = "collect")
 
-gain_profile <- ggplot(CNV_profile_gain, aes(chromosome_arm, y = absolute_10, width = 20* Size)) +
+gain_profile <- list()
+for(chr in unique(CNV_profile_gain$chromosome_arm)){
+  gain_profile[[chr]] <- ggplot(CNV_profile_gain[CNV_profile_gain$chromosome_arm == chr,], aes(x = chromosome_arm, y = absolute_10)) +
   geom_bar(stat = "identity",  fill = "#ce1256", alpha = 0.4) +
-  geom_bar(aes(x = chromosome_arm, y = absolute_70, width = 20 * Size), stat = "identity",  fill = "#ce1256") +
-  theme_classic() +
-  theme(axis.title.x=element_blank(),
-        axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), 
-        text = element_text(size = 15)) +
-  ylab(" ")
+  geom_bar(aes(x = chromosome_arm, y = absolute_70), stat = "identity",  fill = "#ce1256") +
+  theme_void() +
+  ylim(c(0,40)) +
+  theme(axis.text.x = element_blank())
+}
+gain_profile <- wrap_plots(gain_profile, nrow = 1, widths = chromosome_arms$Size[chromosome_arms$dupli %in% unique(CNV_profile_gain$chromosome_arm)], guides = "collect")
 
 # combined plot
-wrap_plots(gain_profile, loss_profile, ncol = 1)
+wrap_plots(gain_profile, loss_profile, ncol = 1) + plot_layout( widths = CNV_profile_gain$Size)
 

### 2025-02-27 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

@@
   mutate(chromosome_arm = factor(chromosome_arm, levels = has_dupli_chr)) |>
-  mutate(absolute_70 = case_when((percentage > 0.7) ~ 1,
-         .default = 0),
-         absolute_10 = case_when((percentage > 0.1) ~ 1,
-         .default = 0))
+  mutate(
+    absolute_70 = ifelse(percentage > 0.7, 1, 0), 
+    absolute_10 = ifelse(percentage > 0.1, 1, 0)
+ )
 

### 2025-02-27 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
add back heatmap and skip 24

@@
 ```{r, results='asis'}
-# check for if the project is all cell lines or not 
-is_cell_line <- all(consensus_dt$sample_type == "cell line")
+# check for if the project is a cell line project or not 
+# explicitly skip project 24 since the current processed data doesn't have the updated combination of PDX and cell line
+is_cell_line <- all(consensus_dt$sample_type == "cell line") | params$project_id == "SCPCP000024"
 
@@
 
+The heatmap shows the mean expression of each marker gene across all cells assigned to a given cell type group. 
+In contrast to the dot plot, we will show all marker genes for a given cell type group, not just the uniquely expressed genes.
+
+```{r, eval=!is_cell_line}
+# add column to uniquely label gene symbols since some genes are present across multiple cell types 
+# add this here since we will use the order in this data frame to set the order of the heatmap columns
+heatmap_markers_df <- filtered_markers_df |> 
+  dplyr::mutate(
+    unique_symbol = glue::glue("{gene_symbol}-{validation_group_annotation}")
+  ) 
+```
+
+```{r, eval=!is_cell_line}
+# first make a mtx to use for the heatmap with rows as cell types and genes as columns 
+heatmap_mtx <- group_stats_df |> 
+  dplyr::ungroup() |> 
+  # add unique label for gene symbol 
+  dplyr::mutate(
+    unique_symbol = glue::glue("{gene_symbol}-{validation_group_annotation}")
+  ) |> 
+  dplyr::select(unique_symbol, broad_celltype_group, mean_exp) |> 
+  dplyr::filter(unique_symbol %in% heatmap_markers_df$unique_symbol) |> 
+  tidyr::pivot_wider(
+    names_from = unique_symbol,
+    values_from = mean_exp
+  ) |> 
+  tibble::column_to_rownames("broad_celltype_group") |>
+  as.matrix()
+# replace any NA with 0
+# accounts for marker genes that aren't found in all cells 
+heatmap_mtx[is.na(heatmap_mtx)] <- 0
+# make sure cell types are present in the right order
+heatmap_mtx <- heatmap_mtx[c(celltype_order, "unknown"), heatmap_markers_df$unique_symbol]
+
+# create annotation for heatmap
+annotation <- ComplexHeatmap::columnAnnotation(
+  group = heatmap_markers_df$validation_group_annotation,
+  col = list(
+    group = celltype_colors
+  ),
+  annotation_legend_param = list(
+      title = "Validation group"
+    )
+)
+```
+
+```{r, eval=!is_cell_line, fig.height=7, fig.width=10}
+marker_gene_heatmap(heatmap_mtx, annotation)
+```
+
 ## Mean expression across all marker genes

### 2025-02-27 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
use is renv project

@@
 ```{r base paths}
-# The base path for the OpenScPCA repository, found by its (hidden) .git directory
-repository_base <- rprojroot::find_root(rprojroot::is_git_root)
-module_base <- file.path(repository_base, "analyses", "cell-type-consensus")
+# The path for this module
+module_base <- rprojroot::find_root(rprojroot::is_renv_project)
 
 # path to results directory
-results_dir <- file.path(repository_base, "data", "current", "results", "cell-type-consensus", params$project_id)
+results_dir <- file.path(module_base, "..", "..", "data", "current", "results", "cell-type-consensus", params$project_id)
 ```

### 2025-02-27 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
use fread and back to dplyr

@@
 consensus_dt <- consensus_results_files |> 
-  purrr::map(vroom::vroom) |> 
-  data.table::rbindlist(fill = TRUE)
+  purrr::map(fread) |> 
+  data.table::rbindlist(fill = TRUE, use.names = TRUE)
 
 gene_exp_dt <- gene_exp_files |> 
-  purrr::map(vroom::vroom) |> 
-  data.table::rbindlist(fill = TRUE)
+  purrr::map(fread) |> 
+  data.table::rbindlist(fill = TRUE, use.names = TRUE)
 
@@
   # groups similar cell types together 
-  merge(validation_groups_dt, by = "consensus_annotation", all.x = TRUE) |> 
-  merge(gene_exp_dt, by = c("barcodes", "library_id"), all.x = TRUE) |> 
+  dplyr::left_join(validation_groups_dt, by = "consensus_annotation") |> 
+  dplyr::left_join(gene_exp_dt, by = c("barcodes", "library_id")) |> 
   # add marker gene information (associated validation group annotation, gene observed count, percent tissues)
   # account for the same gene being present in multiple cell types 
-  merge(markers_dt, by = "ensembl_gene_id", all.x = TRUE, allow.cartesian = TRUE)
-
-# add a column indicating if gene is present or not
-consensus_dt[, detected := logcounts > 0]
+  dplyr::left_join(markers_dt, by = "ensembl_gene_id", relationship = "many-to-many") |> 
+  dplyr::mutate(detected = logcounts > 0)
 
@@
   dplyr::select(library_id, barcodes, broad_celltype_group) |> 
-  unique() |> 
+  dplyr::distinct() |> 
   dplyr::count(broad_celltype_group, name = "total_cells") 

### 2025-03-07 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd

@@
 
-DT::datatable(table(cell_type_df$first.level_annotation, cell_type_df$sample_id))
+cell_type_df |>
+  dplyr::count(first.level_annotation, sample_id) |>
+  tidyr::pivot_wider(
+    names_from = first.level_annotation, 
+    values_from = n
+  )|>
+  DT::datatable()
 
@@
 ```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
-cell_type_df <- cell_type_df |>
-  mutate(second.level_annotation = case_when(
-    # assign normal cells based on condition
-    cell_type_df$compartment %in% c("fetal_nephron") &
-      cell_type_df$has_cnv_score == "no CNV" ~ "kidney",
-
-    cell_type_df$compartment %in% c("stroma") &
-      cell_type_df$has_cnv_score == "no CNV" ~ "normal stroma",
-
-    cell_type_df$compartment %in% c("normal") &
-      cell_type_df$cell_type %in% c("endothelial cell") ~ "endothelium",
-    cell_type_df$compartment %in% c("normal") &
-      cell_type_df$cell_type %in% c("B cell",
+# Define a vector of `fetal_kidney_predicted.cell_type` levels that corresponds to immune cells
+immune_celltypes <- c("B cell",
                                     "CD4-positive, alpha-beta T cell",
@@
                                     "neutrophil",
-                                    "plasmacytoid dendritic cell") ~ "immune",
+                                    "plasmacytoid dendritic cell")
+
+cell_type_df <- cell_type_df |>
+  mutate(second.level_annotation = case_when(
+    # assign normal cells based on condition
+      (sample_id                                         %in% none_reference_samples) ~ "unknown",
+
+      (cell_type_df$fetal_kidney_predicted.compartment   == c("fetal_nephron") &
+      cell_type_df$cnv_score                             == "no CNV") ~ "kidney",
+
+      (cell_type_df$fetal_kidney_predicted.compartment   == c("stroma") &
+      cell_type_df$cnv_score                             == "no CNV") ~ "normal stroma",
+
+      (cell_type_df$fetal_kidney_predicted.compartment   == c("normal") &
+      cell_type_df$fetal_kidney_predicted.cell_type      == c("endothelial cell")) ~ "endothelium",
+
+      (cell_type_df$fetal_kidney_predicted.compartment   == c("normal") &
+      cell_type_df$fetal_kidney_predicted.cell_type    %in% immune_celltypes) ~ "immune",
 
     # assign cancer cells based on condition
-    cell_type_df$compartment %in% c("stroma") &
-      cell_type_df$has_cnv_score == "CNV" ~ "cancer stroma",
-    cell_type_df$compartment %in% c("fetal_nephron") &
-      cell_type_df$has_cnv_score == "CNV" &
-      cell_type_df$cell_type == "mesenchymal cell" ~ "blastema",
-    cell_type_df$compartment %in% c("fetal_nephron") &
-      cell_type_df$has_cnv_score == "CNV" &
-      cell_type_df$cell_type != "mesenchymal cell" ~ "cancer epithelium",
+      (cell_type_df$fetal_kidney_predicted.compartment    == c("stroma") &
+      cell_type_df$cnv_score                              == "CNV") ~ "cancer stroma",
+
+      (cell_type_df$fetal_kidney_predicted.compartment    == c("fetal_nephron") &
+      cell_type_df$cnv_score                              == "CNV" &
+      cell_type_df$fetal_kidney_predicted.cell_type       == "mesenchymal cell") ~ "blastema",
+
+      (cell_type_df$fetal_kidney_predicted.compartment    == c("fetal_nephron") &
+      cell_type_df$cnv_score                              == "CNV" &
+      cell_type_df$fetal_kidney_predicted.cell_type       != "mesenchymal cell") ~ "cancer epithelium",
+
     .default = "unknown"
@@
 
-```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-for (i in 1:22) {
-  print(do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "q")))
-}
+```{r fig.width=25, fig.height=50, out.width='100%', results='asis'}
+# create a list of all plots by mapping over the vector of indices, instead of a for loop
+1:22 |>
+  purrr::map(
+    \(i) 
+      do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "q"))
+) |>
+# send into patchwork to make a single plot from the list of plots in 2 columns
+patchwork::wrap_plots(ncol = 2)
 ```
@@
 
-```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-for (i in 1:22) {
-  print(do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "q")))
-}
+```{r fig.width=25, fig.height=50, out.width='100%', results='asis'}
+# create a list of all plots by mapping over the vector of indices, instead of a for loop
+1:22 |>
+  purrr::map(
+    \(i) 
+      do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "q"))
+) |>
+# send into patchwork to make a single plot from the list of plots in 2 columns
+patchwork::wrap_plots(ncol = 2)
 ```
@@
 
-```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-for (i in c(1:12, 16:20)) {
-  print(do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "p")))
-}
+```{r fig.width=25, fig.height=50, out.width='100%', results='asis'}
+# create a list of all plots by mapping over the vector of indices, instead of a for loop
+c(1:12, 16:20) |>
+  purrr::map(
+    \(i) 
+      do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "p"))
+) |>
+# send into patchwork to make a single plot from the list of plots in 2 columns
+patchwork::wrap_plots(ncol = 2)
 ```
@@
 
-```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
-for (i in c(1:12, 16:20)) {
-  print(do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "p")))
-}
+```{r fig.width=25, fig.height=50, out.width='100%', results='asis'}
+# create a list of all plots by mapping over the vector of indices, instead of a for loop
+c(1:12, 16:20) |>
+  purrr::map(
+    \(i) 
+      do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "p"))
+) |>
+# send into patchwork to make a single plot from the list of plots in 2 columns
+patchwork::wrap_plots(ncol = 2)
 ```
@@
                     summarise_all(.funs = mean) |>
-  pivot_longer(cols = -sample_id) |>
-  rename("chromosome_arm" = "name", "percentage" = "value") |>
-  left_join( chromosome_arms, by = c("chromosome_arm"="dupli")) |>
+                    pivot_longer(
+                      cols = -sample_id, 
+                      names_to = "chromosome_arm", 
+                      values_to = "percentage") |>
+                    left_join(chromosome_arms, by = c("chromosome_arm"="dupli")) |>
                     mutate(chromosome_arm = factor(chromosome_arm, levels = has_dupli_chr)) |>
@@
                       absolute_70 = ifelse(percentage > 0.7, 1, 0), 
-    absolute_10 = ifelse(percentage > 0.1, 1, 0)
- )
+                      absolute_10 = ifelse(percentage > 0.1, 1, 0))
 
 # plot
+# Here we chose to loop on chromosome arms to then be able to scale each bar to the length of the corresponding genomic region using wrap_plot. 
+# Of note, the resulting plot is way better than the one obtained with facet_wrap or facet_grid so far.
 
-loss_profile <- ggplot(CNV_profile_loss, aes(x = chromosome_arm, y = absolute_10, width = 20 * Size)) +
+loss_profile <- list()
+for(chr in unique(CNV_profile_loss$chromosome_arm)){
+  loss_profile[[chr]] <- ggplot(CNV_profile_loss[CNV_profile_loss$chromosome_arm == chr,], 
+                                aes(x = chromosome_arm, y = absolute_10)) +
                         geom_bar(stat = "identity",  fill = "#41b6c4", alpha = 0.4) +
-  geom_bar(aes(x = chromosome_arm, y = absolute_70, width = 20 * Size), stat = "identity",  fill = "#41b6c4") +
-  scale_x_discrete(labels = gsub("has_loss_", "", levels(droplevels(CNV_profile_loss$chromosome_arm))))  +
-  theme_classic() +
-  ylab("Number of occurrence") +
-  xlab("chromosom arms") +
+                        geom_bar(aes(x = chromosome_arm, y = absolute_70), stat = "identity",  fill = "#41b6c4") +
+                        scale_x_discrete(labels = gsub("has_loss_", "", chr))  +
+                        theme_void() +
+                        ylim(c(-40,0)) +
                         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12), text = element_text(size = 15))
+}
+loss_profile <- wrap_plots(loss_profile, nrow = 1, widths = chromosome_arms$Size[chromosome_arms$loss %in% unique(CNV_profile_loss$chromosome_arm)], guides = "collect")
 
-gain_profile <- ggplot(CNV_profile_gain, aes(chromosome_arm, y = absolute_10, width = 20* Size)) +
+gain_profile <- list()
+for(chr in unique(CNV_profile_gain$chromosome_arm)){
+  gain_profile[[chr]] <- ggplot(CNV_profile_gain[CNV_profile_gain$chromosome_arm == chr,], aes(x = chromosome_arm, y = absolute_10)) +
   geom_bar(stat = "identity",  fill = "#ce1256", alpha = 0.4) +
-  geom_bar(aes(x = chromosome_arm, y = absolute_70, width = 20 * Size), stat = "identity",  fill = "#ce1256") +
-  theme_classic() +
-  theme(axis.title.x=element_blank(),
-        axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), 
-        text = element_text(size = 15)) +
-  ylab(" ")
+  geom_bar(aes(x = chromosome_arm, y = absolute_70), stat = "identity",  fill = "#ce1256") +
+  theme_void() +
+  ylim(c(0,40)) +
+  theme(axis.text.x = element_blank())
+}
+gain_profile <- wrap_plots(gain_profile, nrow = 1, widths = chromosome_arms$Size[chromosome_arms$dupli %in% unique(CNV_profile_gain$chromosome_arm)], guides = "collect")
 
 # combined plot
-wrap_plots(gain_profile, loss_profile, ncol = 1)
+wrap_plots(gain_profile, loss_profile, ncol = 1) + plot_layout( widths = CNV_profile_gain$Size)
 

### 2025-03-10 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
determine arm to plot based on testing. if testing, arm should be empty. In addition, notebook got styled by precommit

@@
         THY1 = FetchData(object = srat, vars = "ENSG00000154096", layer = "counts")
-
       ) |>
@@
     }
-
   ) |>
@@
 cell_type_df <- cell_type_df |>
-  left_join( metadata, by= c("sample_id" = "scpca_sample_id"))
-
+  left_join(metadata, by = c("sample_id" = "scpca_sample_id"))
 ```
@@
   ) |>
-    mutate(Size = End - Start,
+  mutate(
+    Size = End - Start,
     dupli = paste0("has_dupli_", chrom, arm),
     loss = paste0("has_loss_", chrom, arm),
-           chrom_arm = glue::glue("{chrom}{arm}")) |> 
+    chrom_arm = glue::glue("{chrom}{arm}")
+  ) |>
   mutate(Size = Size / sum(Size)) |>
   # Define chromosome arm order
-    mutate(chrom_arm = factor(chrom_arm, levels = c(paste0("chr", rep(1:22, each = 2), c("p", "q")),
-                                                  "chrXp", "chrXq", "chrYp", "chrYq"))) |>
+  mutate(chrom_arm = factor(chrom_arm, levels = c(
+    paste0("chr", rep(1:22, each = 2), c("p", "q")),
+    "chrXp", "chrXq", "chrYp", "chrYq"
+  ))) |>
   # Sort genes by Chromosome arm and Start position
@@
   filter(arm != "")
-
 ```
@@
 ```{r fig.width=20, fig.height=15, out.width='100%', results='asis'}
-do_BoxPlot_mean <- function(df, group.by, comparison){
+do_BoxPlot_mean <- function(df, group.by, comparison) {
   tmp <- df |>
@@
     summarise_all(.funs = mean) |>
-    pivot_longer(cols = -c(sample_id,!!sym(group.by)) ) |>
+    pivot_longer(cols = -c(sample_id, !!sym(group.by))) |>
     rename("chromosome_arm" = "name", "percentage" = "value") |>
@@
   p <- ggplot(tmp, aes(x = !!sym(group.by), y = percentage, fill = !!sym(group.by))) +
-  geom_boxplot()+
-  scale_fill_brewer(palette="PuRd") +
-  geom_dotplot(binaxis='y', stackdir='center') +
-  facet_wrap(~chromosome_arm, nrow = 5 ) +
+    geom_boxplot() +
+    scale_fill_brewer(palette = "PuRd") +
+    geom_dotplot(binaxis = "y", stackdir = "center") +
+    facet_wrap(~chromosome_arm, nrow = 5) +
     theme_classic()
@@
   ))
-
 ```
@@
       cnv_score == "no CNV" ~ "normal",
-
     fetal_kidney_predicted.compartment %in% c("normal", "immune", "endothelium") &
       fetal_kidney_predicted.compartment.score > params$predicted.celltype.threshold ~ "normal",
-
     fetal_kidney_predicted.compartment %in% c("fetal_nephron", "stroma") &
       cnv_score == "CNV" ~ "cancer",
-
     .default = "unknown"
   ))
-
 ```
@@
     values_from = n
-  )|>
+  ) |>
   DT::datatable()
-
 ```
@@
 ```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
-
 cell_type_df$first.level_annotation[cell_type_df$sample_id %in% none_reference_samples] <- "unknown"
-
 ```
@@
 # Define a vector of `fetal_kidney_predicted.cell_type` levels that corresponds to immune cells
-immune_celltypes <- c("B cell",
+immune_celltypes <- c(
+  "B cell",
   "CD4-positive, alpha-beta T cell",
@@
   "neutrophil",
-                                    "plasmacytoid dendritic cell")
+  "plasmacytoid dendritic cell"
+)
 
@@
     (sample_id %in% none_reference_samples) ~ "unknown",
-
     (cell_type_df$fetal_kidney_predicted.compartment == c("fetal_nephron") &
       cell_type_df$cnv_score == "no CNV") ~ "kidney",
-
     (cell_type_df$fetal_kidney_predicted.compartment == c("stroma") &
       cell_type_df$cnv_score == "no CNV") ~ "normal stroma",
-
     (cell_type_df$fetal_kidney_predicted.compartment == c("normal") &
       cell_type_df$fetal_kidney_predicted.cell_type == c("endothelial cell")) ~ "endothelium",
-
     (cell_type_df$fetal_kidney_predicted.compartment == c("normal") &
@@
       cell_type_df$cnv_score == "CNV") ~ "cancer stroma",
-
     (cell_type_df$fetal_kidney_predicted.compartment == c("fetal_nephron") &
@@
       cell_type_df$fetal_kidney_predicted.cell_type == "mesenchymal cell") ~ "blastema",
-
     (cell_type_df$fetal_kidney_predicted.compartment == c("fetal_nephron") &
@@
       cell_type_df$fetal_kidney_predicted.cell_type != "mesenchymal cell") ~ "cancer epithelium",
-
     .default = "unknown"
   ))
-
 ```
@@
 ```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
-cell_type_df_sub <- cell_type_df[cell_type_df$second.level_annotation != "unknown",]
+cell_type_df_sub <- cell_type_df[cell_type_df$second.level_annotation != "unknown", ]
 ggplot(cell_type_df_sub, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
@@
 # create a list of all plots by mapping over the vector of indices, instead of a for loop
+plot_arm <- ifelse(params$testing, "", "q") # only use "q" if NOT testing.
+
 1:22 |>
@@
     \(i)
-      do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "q"))
+    do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, plot_arm))
   ) |>
@@
 # create a list of all plots by mapping over the vector of indices, instead of a for loop
+plot_arm <- ifelse(params$testing, "", "q") # only use "q" if NOT testing.
 1:22 |>
@@
     \(i)
-      do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "q"))
+    do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, plot_arm))
   ) |>
@@
 # create a list of all plots by mapping over the vector of indices, instead of a for loop
+plot_arm <- ifelse(params$testing, "", "p") # only use "p" if NOT testing.
 c(1:12, 16:20) |>
@@
     \(i)
-      do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "p"))
+    do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, plot_arm))
   ) |>
@@
     \(i)
-      do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "p"))
+    do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, plot_arm))
   ) |>
@@
 has_loss_chr <- names(cell_type_df)[stringr::str_starts(names(cell_type_df), "has_loss_chr")]
-
 ```
@@
     names_to = "chromosome_arm",
-                      values_to = "percentage") |>
-                    left_join(chromosome_arms, by = c("chromosome_arm"="loss")) |>
+    values_to = "percentage"
+  ) |>
+  left_join(chromosome_arms, by = c("chromosome_arm" = "loss")) |>
   mutate(chromosome_arm = factor(chromosome_arm, levels = has_loss_chr)) |>
@@
     absolute_70 = ifelse(percentage > 0.7, -1, 0),
-                      absolute_10 = ifelse(percentage > 0.1, -1, 0))
+    absolute_10 = ifelse(percentage > 0.1, -1, 0)
+  )
 
@@
     names_to = "chromosome_arm",
-                      values_to = "percentage") |>
-                    left_join(chromosome_arms, by = c("chromosome_arm"="dupli")) |>
+    values_to = "percentage"
+  ) |>
+  left_join(chromosome_arms, by = c("chromosome_arm" = "dupli")) |>
   mutate(chromosome_arm = factor(chromosome_arm, levels = has_dupli_chr)) |>
@@
     absolute_70 = ifelse(percentage > 0.7, 1, 0),
-                      absolute_10 = ifelse(percentage > 0.1, 1, 0))
+    absolute_10 = ifelse(percentage > 0.1, 1, 0)
+  )
 
@@
 loss_profile <- list()
-for(chr in unique(CNV_profile_loss$chromosome_arm)){
-  loss_profile[[chr]] <- ggplot(CNV_profile_loss[CNV_profile_loss$chromosome_arm == chr,], 
-                                aes(x = chromosome_arm, y = absolute_10)) +
+for (chr in unique(CNV_profile_loss$chromosome_arm)) {
+  loss_profile[[chr]] <- ggplot(
+    CNV_profile_loss[CNV_profile_loss$chromosome_arm == chr, ],
+    aes(x = chromosome_arm, y = absolute_10)
+  ) +
     geom_bar(stat = "identity", fill = "#41b6c4", alpha = 0.4) +
@@
     theme_void() +
-                        ylim(c(-40,0)) +
-                        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12), text = element_text(size = 15))
+    ylim(c(-40, 0)) +
+    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 12), text = element_text(size = 15))
 }
@@
 gain_profile <- list()
-for(chr in unique(CNV_profile_gain$chromosome_arm)){
-  gain_profile[[chr]] <- ggplot(CNV_profile_gain[CNV_profile_gain$chromosome_arm == chr,], aes(x = chromosome_arm, y = absolute_10)) +
+for (chr in unique(CNV_profile_gain$chromosome_arm)) {
+  gain_profile[[chr]] <- ggplot(CNV_profile_gain[CNV_profile_gain$chromosome_arm == chr, ], aes(x = chromosome_arm, y = absolute_10)) +
     geom_bar(stat = "identity", fill = "#ce1256", alpha = 0.4) +
@@
     theme_void() +
-  ylim(c(0,40)) +
+    ylim(c(0, 40)) +
     theme(axis.text.x = element_blank())
@@
 # combined plot
-wrap_plots(gain_profile, loss_profile, ncol = 1) + plot_layout( widths = CNV_profile_gain$Size)
-
+wrap_plots(gain_profile, loss_profile, ncol = 1) + plot_layout(widths = CNV_profile_gain$Size)
 ```
@@
 ```{r fig.height=15, fig.width=20, message=FALSE, warning=FALSE, out.width='100%', results='asis'}
-
-do_BoxPlot_mean(df = cell_type_df_sub, group.by = "vital_status", comparison = c("Alive", "Expired") )
-
+do_BoxPlot_mean(df = cell_type_df_sub, group.by = "vital_status", comparison = c("Alive", "Expired"))
 ```
@@
 ```{r fig.height=15, fig.width=20, message=FALSE, warning=FALSE, out.width='100%', results='asis'}
-
-do_BoxPlot_mean(df = cell_type_df_sub, group.by = "treatment", comparison = c("Upfront resection", "Resection post chemotherapy") )
-
+do_BoxPlot_mean(df = cell_type_df_sub, group.by = "treatment", comparison = c("Upfront resection", "Resection post chemotherapy"))
 ```

### 2025-03-17 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
fix typos

@@
 
-The aim is to combine label transfer and CNV inference to annotate Wilms tumor samples in SCPCP000006.
-The proposed annotation will be based on the combination of:
+The aim is to combine label transfer and CNV inference to annotate Wilms tumor samples in SCPCP000006. The proposed annotation will be based on the combination of:
 
@@
 | first level annotation | second level annotation | selection of the cells | marker genes for validation | CNV validation |
-| ---------------------- | ----------------------- | --------------------------------------------------------------------------------------------- | --------------------------- | ------------------------------------------ |
+|----|----|----|----|----|
 | normal | endothelial | `compartment == "endothelium" & predicted.score > pred.thr` | `VWF` | no CNV |
@@
 | cancer | stroma | `compartment == "stroma" & cnv_score > cnv.thr.upper` | `VIM` | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
-| cancer                 | blastema                | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr.upper`| `CITED1`                    | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
-| cancer                 | epithelial              | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr.upper`| `CDH1`                      | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
-| unknown                | -                       | the rest of the cells                                                                         | -                           | -                                          |
+| cancer | blastema | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr.upper` | `CITED1` | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| cancer | epithelial | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr.upper` | `CDH1` | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| unknown | \- | the rest of the cells | \- | \- |
 
@@
 
-The report will be saved in the `notebook` directory.
-The final annotation file `SCPCP000006-annotations.tsv` will be saved in the `results` directory.
+The report will be saved in the `notebook` directory. The final annotation file `SCPCP000006-annotations.tsv` will be saved in the `results` directory.
 
@@
 
-We extracted from the pre-processed and labeled `Seurat` object from `results/{sample_id}/06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds`.
-the following information per cell:
+We extracted from the pre-processed and labeled `Seurat` object from `results/{sample_id}/06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds`. the following information per cell:
 
@@
 
-- the CNV `{dupli|loss}` estimation per chromosome {i} and arm `{p|q}`: 
-`has_{dupli|loss}_chr{i}{p|q}`, in 1 to 22
+-   the CNV `{dupli|loss}` estimation per chromosome {i} and arm `{p|q}`: `has_{dupli|loss}_chr{i}{p|q}`, in 1 to 22
 
@@
 
-#### do_BoxPlot_mean
+#### `do_BoxPlot_mean`
 
@@
 
-- _normal_ cells can be observe in all five compartments (`normal`,`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV.
+-   *normal* cells can be observe in all five compartments (`normal`,`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV.
 We only allow a bit of flexibility in terms of CNV profile for immune and endothelium cells that have a high predicted score.
 Indeed, we know that false positive CNV can be observed in a cell type specific manner.
-Please note that in `06_infercnv.R`, we renamed all `immune` and `endothelium` cells with a `fetal_kidney_predicted.compartment.score` > `r params$predicted.celltype.threshold` as `normal`.
+Please note that in `06_infercnv.R`, we renamed all `immune` and `endothelium` cells with a `fetal_kidney_predicted.compartment.score` \> `r params$predicted.celltype.threshold` as `normal`.
 
@@
 
-- _cancer_ cells are either from the `stroma` or `fetal nephron` compartments and must have at least few CNV
+-   *cancer* cells are either from the `stroma` or `fetal nephron` compartments and must have at least few CNV
 
@@
 
-Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` _cancer_ cells, `r table(cell_type_df$first.level_annotation)["normal"]` _normal_ cells.
-`r table(cell_type_df$first.level_annotation)["unknown"]` cells remain _unknown_
+Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` *cancer* cells, `r table(cell_type_df$first.level_annotation)["normal"]` *normal* cells.
+`r table(cell_type_df$first.level_annotation)["unknown"]` cells remain *unknown*
 
@@
 
-- _cancer stroma_: We define as _cancer stroma_ all cancer cells from the stroma compartment.
+-   *cancer stroma*: We define as *cancer stroma* all cancer cells from the stroma compartment.
 
-- _blastema_,: we defined as _blastema_ every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`.
-We know that these _mesenchymal_ cells are cells from the cap mesenchyme that are not expected to be in a mature kidney.
-These blastema cells should express higher _CITED1_ and/or _NCAM1_.
+-   *blastema*,: we defined as *blastema* every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`.
+We know that these *mesenchymal* cells are cells from the cap mesenchyme that are not expected to be in a mature kidney.
+These blastema cells should express higher `CITED1` and/or `NCAM1`.
 
-- _cancer epithelium_: we defined as _cancer epithelium_ all cancer cells that are neither stroma nor blastemal cells.
+-   *cancer epithelium*: we defined as *cancer epithelium* all cancer cells that are neither stroma nor blastemal cells.
 We expect these cells to express epithelial markers.
@@
 
-- identify chromosome and cell types that might be more sensitive to _FALSE_ positive CNV, such as the _chr6p_ which is a hub for immune genes
+-   identify chromosome and cell types that might be more sensitive to *FALSE* positive CNV, such as the *chr6p* which is a hub for immune genes
 
@@
 
-Here we aim to plot a mean CNV profile across the 35 annotated samples as presented in Figure 1A of [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+Here we aim to plot a mean CNV profile across the 35 annotated samples as presented in Figure 1A of [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
 This represent the mean SNP Array profile of 64 pediatric kidney cancer (including 60 Wilms tumors).
 
-Here we plot for each chromosome arm (x-axis) the number of samples having a `dupli` (y-axis > 0, in red) or a `loss` (y-axis < 0, in blue).
+Here we plot for each chromosome arm (x-axis) the number of samples having a `dupli` (y-axis \> 0, in red) or a `loss` (y-axis \< 0, in blue).
 
-For a sample, a CNV is defined as `clonal` (opaque) if detected in more than 70% of the cells;
-and `subclonal` (transparent) when detected in at least 10% of the cells.
+For a sample, a CNV is defined as `clonal` (opaque) if detected in more than 70% of the cells; and `subclonal` (transparent) when detected in at least 10% of the cells.
 
@@
 
-Comparing with the profile in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1), we identified important CNV known to play a role in Wilms tumor, including:
+Comparing with the profile in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1), we identified important CNV known to play a role in Wilms tumor, including:
 
-- _chr8q_, _chr12_, _chr13_, _chr18q_ gains
+-   *chr8q*, *chr12*, *chr13*, *chr18q* gains
 
-- _chr11q_, _chr16_, _chr17q_ loss
+-   *chr11q*, *chr16*, *chr17q* loss
 
@@
 
-- we couldn't detect any _chr4q_ loss as described in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+-   we couldn't detect any *chr4q* loss as described in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
 
-- we identified 2 losses in _chr5_ and _chr6_ that do not show up in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
-These can be `infercnv` induced _FALSE_positive CNV.
+-   we identified 2 losses in *chr5* and *chr6* that do not show up in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+These can be `infercnv` induced \_FALSE_positive CNV.
 
-Our major concern is the pattern of the _chr1_. 
-We do see a tendency of _chr1p_ loss and _chr1q_ gain but it is not as pronounced as in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
-_chr1q_ gain is however one of the most promising predictive marker of Wilms tumor outcome ([Nelson et al, Curr Opin Pediatr (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/)).
-It is however important to notice that the profile shown in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1) derived from SIOP samples that have been pre-treated with chemotherapy, while samples in the dataset are enriched in `upfront resection` samples that haven't been pre-treated, according to the COG protocol.
+Our major concern is the pattern of the *chr1*.
+We do see a tendency of *chr1p* loss and *chr1q* gain but it is not as pronounced as in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+*chr1q* gain is however one of the most promising predictive marker of Wilms tumor outcome ([Nelson et al, Current Opinion in Pediatrics (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/)).
+It is however important to notice that the profile shown in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1) derived from SIOP samples that have been pre-treated with chemotherapy, while samples in the dataset are enriched in `upfront resection` samples that haven't been pre-treated, according to the COG protocol.
 So far, we do not know how the pre-operative chemotherapy impact the clonal selection of specific CNV.
@@
 
-As described in [Nelson et al, Curr Opin Pediatr (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/), 
-[Phelps et al, Children (2018)](https://pmc.ncbi.nlm.nih.gov/articles/PMC6262554/) and 
-[Zheng et al, Front. Oncol (2023)](https://www.frontiersin.org/journals/oncology/articles/10.3389/fonc.2023.1137346/full), 
-LOH at _chr1p_, _chr11p_ and _chr16q_ as well as _chr1q_ gain are prognostic factors.
+As described in [Nelson et al, Current Opinions in Pediatrics (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/), [Phelps et al, Children (2018)](https://pmc.ncbi.nlm.nih.gov/articles/PMC6262554/) and [Zheng et al, Frontiers in Oncology (2023)](https://www.frontiersin.org/journals/oncology/articles/10.3389/fonc.2023.1137346/full), LOH at *chr1p*, *chr11p* and *chr16q* as well as *chr1q* gain are prognostic factors.
 
@@
 
-In our study, we validate the association of _chr1q_ gain, _chr11q_ loss and _chr14q_ loss with poorer prognosis.
+In our study, we validate the association of *chr1q* gain, *chr11q* loss and *chr14q* loss with poorer prognosis.
 
@@
 
-As we don't know the effect of the treatment on the selection of cells with CNV, it is difficult to compare our CNV profile with the one in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+As we don't know the effect of the treatment on the selection of cells with CNV, it is difficult to compare our CNV profile with the one in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
 We those check the repartition of CNV in samples that have been pre-treated (i.e `Resection post chemotherapy`) with samples that have been removed surgically before any treatment (i.e `Upfront resection`).
@@
 
-Our analysis suggest that there is indeed an enrichment in cells with CNV after chemotherapy, 
-including a tendency of more _chr1q_ gain samples after chemotherapy. 
-This could explain some of the differences observed between our mean CNV profile and the one published in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+Our analysis suggest that there is indeed an enrichment in cells with CNV after chemotherapy, including a tendency of more *chr1q* gain samples after chemotherapy.
+This could explain some of the differences observed between our mean CNV profile and the one published in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
 
@@
 
-Immune cells are well annotated based on _PTPRC_ expression.
+Immune cells are well annotated based on `PTPRC` expression.
 
@@
 
-Endothelial cells are well annotated based on _VWF_ expression.
+Endothelial cells are well annotated based on `VWF` expression.
 
@@
 
-(*) To the best of our knowledges, there is no Wilms tumor specific marker of stroma cells.
-They should express in some extend similar markers as normal stromal cells including _VIM_, _THY1_ and _COL6A3_.
+(\*) To the best of our knowledge, there is no Wilms tumor specific marker of stroma cells.
+They should express in some extend similar markers as normal stromal cells including *VIM*, *THY1* and *COL6A3*.
 
@@
 
-_VIM_ do not seem to be neither specific nor universal.
+*VIM* do not seem to be neither specific nor universal.
 
@@
 
-_COL6A3_ expression is higher in Wilms tumor stroma cells and expressed across a wider range of samples.
+*COL6A3* expression is higher in Wilms tumor stroma cells and expressed across a wider range of samples.
 
@@
 
-_THY1_ do not seem to be neither specific nor universal.
+*THY1* do not seem to be neither specific nor universal.
 
@@
 
-(*) To the best of our knowledges, there is no Wilms tumor specific and universel (i.e. expressed by all cells in every sample) marker of blastema cells.
-We expect however blastema cells to express higher levels of stemness markers such as _CITED1_, _SIX1/2_, _NCAM1_.
+(\*) To the best of our knowledge, there is no Wilms tumor specific and universal (i.e.
+expressed by all cells in every sample) marker of blastema cells.
+We expect however blastema cells to express higher levels of stemness markers such as `CITED1`, `SIX1/2`, `NCAM1`.
 
@@
 
-Blastema cells are enriched in _CITED1_+ cells, but some samples do not express it at all.
+Blastema cells are enriched in *CITED1*+ cells, but some samples do not express it at all.
 
@@
 
-_NCAM1_ seems to be more cancer specific and broadly expressed among samples.
+`NCAM1` seems to be more cancer specific and broadly expressed among samples.
 
@@
 
-As expected, epithelial cells express higher expression of _CDH1_.
+As expected, epithelial cells express higher expression of `CDH1`.
 
@@
 Confirm how many samples we have annotations for:
+
 ```{r}
@@
 
-- The mean CNV profile of the Wilms tumor cohort seems reasonable in regards to the one published in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
+-   The mean CNV profile of the Wilms tumor cohort seems reasonable in regards to the one published in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
 We identify some commonalities but also some divergence.
-They can be due to technical limitation of our CNV inference but can also reflect some specificities of this cohort (e.g. pre/post treatment samples).
+They can be due to technical limitation of our CNV inference but can also reflect some specificities of this cohort (e.g.
+pre/post treatment samples).
 It would be really useful to compare for some sample the `infercnv` result to a ground truth such as SNP Array (to be requested).
@@
 
-- In each histology (i.e. epithelial and stroma), the distinction between cancer and non cancer cell is difficult (as expected).
+-   In each histology (i.e.
+epithelial and stroma), the distinction between cancer and non cancer cell is difficult (as expected).
 In this analysis, we suggested to rely on the CNV score to assess the normality of the cell.

### 2025-03-17 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Some text cleanup and OSPL

@@
 
-In a second time, we will explore and validate the chosen annotation.
+We will then explore and validate the chosen annotation.
 
@@
 - `pred.thr` = `r params$predicted.celltype.threshold`
-
 - `cnv.thr.lower` = `r params$cnv_threshold_low`
-
 - `cnv.thr.upper` = `r params$cnv_threshold_high`
@@
 
-The report will be saved in the `notebook` directory. The final annotation file `SCPCP000006-annotations.tsv` will be saved in the `results` directory.
+The report will be saved in the `notebook` directory. 
+The final annotation file `SCPCP000006-annotations.tsv` will be saved in the `results` directory.
 
@@
 In this notebook, we are working with all of the samples in SCPCP000006.
-
 The sample metadata can be found in `sample_metadata_file` in the `data` folder.
@@
 
-We extracted from the pre-processed and labeled `Seurat` object from `results/{sample_id}/06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds`. the following information per cell:
+From the pre-processed and labeled `Seurat` object from `results/{sample_id}/06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds`, we will extract the following information per cell:
 
 - the predicted compartment in `fetal_kidney_predicted.compartment` and `fetal_kidney_predicted.compartment.score`
-
 - the predicted cell type in `fetal_kidney_predicted.cell_type` and `fetal_kidney_predicted.cell_type.score`
-
 - the sample identifier in `sample_id`
-
 - the CNV `{dupli|loss}` estimation per chromosome {i} and arm `{p|q}`: `has_{dupli|loss}_chr{i}{p|q}`, in 1 to 22
-
 - the `counts` of [markers genes](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-wilms-tumor-06#the-table-celltype_metadatacsv-contains-the-following-column-and-information)
@@
 - the samples metadata or clinical data reported in `r file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")`
+- the cytoband file to get the length of each chromosome arm (for plotting only)
 
-```{r }
+```{r metadata}
 # Add clinical metadata
@@
 ```
-
--   the cytoband file to get the length of each chromosome arm (for plotting only)
 
@@
 
-We simply checked for each chromosome the cell `cnv_chr`.
-Would the cell have less than `r params$cnv_threshold_low` chromosome with CNV, the global `cnv_score` will be FALSE.
-Would the cell have more than `r params$cnv_threshold_high` chromosome with CNV, the global `cnv_score` will be TRUE.
-The default is set to `unknown`.
+We simply checked the `cnv_chr` for each chromosome across cells.
+
+- If the cell has less than `r params$cnv_threshold_low` chromosome with CNV, the global `cnv_score` will be `FALSE`. 
+- If the cell has more than `r params$cnv_threshold_high` chromosome with CNV, the global `cnv_score` will be `TRUE`. 
+- The default is set to `unknown`.
 
@@
 
--   *normal* cells can be observe in all five compartments (`normal`,`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV.
+- *normal* cells can be observed in all five compartments (`normal`,`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV. 
+- *cancer* cells are either from the `stroma` or `fetal nephron` compartments and must have at least few CNV
+
 We only allow a bit of flexibility in terms of CNV profile for immune and endothelium cells that have a high predicted score. 
@@
 The thresholds used for the identification of CNV are also defined as notebook parameters with a minimum of `r params$cnv_threshold_low` and a maximum of `r params$cnv_threshold_high`.
-
--   *cancer* cells are either from the `stroma` or `fetal nephron` compartments and must have at least few CNV
 
@@
 Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` *cancer* cells, `r table(cell_type_df$first.level_annotation)["normal"]` *normal* cells.
-`r table(cell_type_df$first.level_annotation)["unknown"]` cells remain *unknown*
+
+`r table(cell_type_df$first.level_annotation)["unknown"]` cells remain *unknown*.
 
@@
 These five samples do not have enough immune and endothelium cell to be used as a normal reference in `scripts/06_infercnv.R` and we previously decided to run `scripts/06_infercnv.R` without any reference for these samples.
-In that case, the mean expression profile over the sample is taken as the normal reference, biasing the inference of CNV.
-
-For those samples, the annotations based on `infercnv` cannot be trust.
+Without a specified reference, the mean expression profile across the sample is taken as the normal reference, biasing the inference of CNV.
+For those samples, the annotations based on `infercnv` cannot be trusted. 
 To avoid any confusion in the following steps of the analysis, we decided to force the annotation of the entire samples to `unknown`.
@@
 - Normal cells from the `fetal nephron` compartment must be normal kidney cells.
-
 - Normal cells from the `stroma` compartment must be normal stroma cells.
-
 - Immune and endothelial cells have been already identified by label transfer.
@@
 - *cancer stroma*: We define as *cancer stroma* all cancer cells from the stroma compartment.
-
--   *blastema*,: we defined as *blastema* every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`.
+- *blastema*: We defined as *blastema* every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`. 
 We know that these *mesenchymal* cells are cells from the cap mesenchyme that are not expected to be in a mature kidney. 
 These blastema cells should express higher `CITED1` and/or `NCAM1`.
-
--   *cancer epithelium*: we defined as *cancer epithelium* all cancer cells that are neither stroma nor blastemal cells.
+- *cancer epithelium*: We defined as *cancer epithelium* all cancer cells that are neither stroma nor blastemal cells. 
 We expect these cells to express epithelial markers. 
@@
 - validate that mostly tumor cells do present high percentages of cells with CNV (by definition)
-
--   identify chromosome and cell types that might be more sensitive to *FALSE* positive CNV, such as the *chr6p* which is a hub for immune genes
-
+- identify chromosome and cell types that might be more sensitive to *FALSE* positive CNV, such as `chr6p` which is a hub for immune genes
 - get the repartition per sample of the CNV
@@
 
-While the above heatmaps can be informative, we have to acknowledge that there are numerous and it is difficult to get a global picture of the CNV profile of the entire dataset.
+While the above heatmaps can be informative, because there are so many of them it is difficult to get a global picture of the CNV profile of the entire dataset.
 
 Here we aim to plot a mean CNV profile across the 35 annotated samples as presented in Figure 1A of [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
-This represent the mean SNP Array profile of 64 pediatric kidney cancer (including 60 Wilms tumors).
+This represents the mean SNP Array profile of 64 pediatric kidney cancer (including 60 Wilms tumors).
 
@@
 - *chr8q*, *chr12*, *chr13*, *chr18q* gains
-
 - *chr11q*, *chr16*, *chr17q* loss
@@
 
--   we couldn't detect any *chr4q* loss as described in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+- We couldn't detect any *chr4q* loss as described in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
 
--   we identified 2 losses in *chr5* and *chr6* that do not show up in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
-These can be `infercnv` induced \_FALSE_positive CNV.
+- We identified 2 losses in *chr5* and *chr6* that do not show up in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+These may be `infercnv` induced false-positive CNV.
 
@@
 
-(\*) To the best of our knowledge, there is no Wilms tumor specific and universal (i.e.
-expressed by all cells in every sample) marker of blastema cells.
+(\*) To the best of our knowledge, there is no Wilms tumor specific and universal (i.e. expressed by all cells in every sample) marker of blastema cells. 
 We expect however blastema cells to express higher levels of stemness markers such as `CITED1`, `SIX1/2`, `NCAM1`.
@@
 
-Blastema cells are enriched in *CITED1*+ cells, but some samples do not express it at all.
+Blastema cells are enriched in `CITED1+` cells, but some samples do not express it at all.
 
@@
 We identify some commonalities but also some divergence. 
-They can be due to technical limitation of our CNV inference but can also reflect some specificities of this cohort (e.g.
-pre/post treatment samples).
+They can be due to technical limitation of our CNV inference but can also reflect some specificities of this cohort (e.g. pre/post treatment samples). 
 It would be really useful to compare for some sample the `infercnv` result to a ground truth such as SNP Array (to be requested).
@@
 
--   In each histology (i.e.
-epithelial and stroma), the distinction between cancer and non cancer cell is difficult (as expected).
+- In each histology (i.e. epithelial and stroma), the distinction between cancer and non cancer cell is difficult (as expected). 
 In this analysis, we suggested to rely on the CNV score to assess the normality of the cell. 

### 2025-03-17 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
clarify that chr text

@@
 - the sample identifier in `sample_id`
-- the CNV `{dupli|loss}` estimation per chromosome {i} and arm `{p|q}`: `has_{dupli|loss}_chr{i}{p|q}`, in 1 to 22
+- the CNV `{dupli|loss}` estimate per chromosome `{i}`, where `i \in {1-22}`, and arm `{p|q}`: `has_{dupli|loss}_chr{i}{p|q}`
 - the `counts` of [markers genes](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-wilms-tumor-06#the-table-celltype_metadatacsv-contains-the-following-column-and-information)

### 2025-03-17 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

--- 2025-03-17 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
@@
 
-The aim is to combine label transfer and CNV inference to annotate Wilms tumor samples in SCPCP000006. The proposed annotation will be based on the combination of:
+The aim is to combine label transfer and CNV inference to annotate Wilms tumor samples in SCPCP000006.
+The proposed annotation will be based on the combination of:
 
@@
 
-From the pre-processed and labeled `Seurat` object from `results/{sample_id}/06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds`, we will extract the following information per cell:
+From the pre-processed and labeled `Seurat` object in `results/{sample_id}/06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds`, we will extract the following per-cell information:
 
@@
 - the sample identifier in `sample_id`
-- the CNV `{dupli|loss}` estimation per chromosome {i} and arm `{p|q}`: `has_{dupli|loss}_chr{i}{p|q}`, in 1 to 22
+- the CNV `{dupli|loss}` estimate per chromosome `{i}`, where `i` is in `{1-22}`, and arm `{p|q}`: `has_{dupli|loss}_chr{i}{p|q}`
 - the `counts` of [markers genes](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-wilms-tumor-06#the-table-celltype_metadatacsv-contains-the-following-column-and-information)
@@
 Indeed, we know that false positive CNV can be observed in a cell type specific manner. 
-Please note that in `06_infercnv.R`, we renamed all `immune` and `endothelium` cells with a `fetal_kidney_predicted.compartment.score` \> `r params$predicted.celltype.threshold` as `normal`.
+Please note that in `06_infercnv.R`, we renamed all `immune` and `endothelium` cells with a `fetal_kidney_predicted.compartment.score` > `r params$predicted.celltype.threshold` as `normal`.
 
@@
 We know that these *mesenchymal* cells are cells from the cap mesenchyme that are not expected to be in a mature kidney. 
-These blastema cells should express higher `CITED1` and/or `NCAM1`.
+These blastema cells should express higher *CITED1* and/or *NCAM1*.
 - *cancer epithelium*: We defined as *cancer epithelium* all cancer cells that are neither stroma nor blastemal cells. 
@@
 
-*VIM* do not seem to be neither specific nor universal.
+*VIM* expression does not seem to be either specific or universal.
 
@@
 
-*THY1* do not seem to be neither specific nor universal.
+*THY1* expression does not seem to be either specific or universal.
 
@@
 (\*) To the best of our knowledge, there is no Wilms tumor specific and universal (i.e. expressed by all cells in every sample) marker of blastema cells. 
-We expect however blastema cells to express higher levels of stemness markers such as `CITED1`, `SIX1/2`, `NCAM1`.
+We expect however blastema cells to express higher levels of stemness markers such as *CITED1*, *SIX1/2*, *NCAM1*.
 
@@
 
-Blastema cells are enriched in `CITED1+` cells, but some samples do not express it at all.
+Blastema cells are enriched in *CITED1+* cells, but some samples do not express it at all.
 
@@
 
-`NCAM1` seems to be more cancer specific and broadly expressed among samples.
+*NCAM1* seems to be more cancer specific and broadly expressed among samples.
 
@@
 
-As expected, epithelial cells express higher expression of `CDH1`.
+As expected, epithelial cells express higher expression of *CDH1*.
 
@@
 
-- The mean CNV profile of the Wilms tumor cohort seems reasonable in regards to the one published in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
+- The mean CNV profile of the Wilms tumor cohort seems reasonable in regards to the one published in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
 We identify some commonalities but also some divergence. 

### 2025-03-17 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
 
-Here we plot for each chromosome arm (x-axis) the number of samples having a `dupli` (y-axis \> 0, in red) or a `loss` (y-axis \< 0, in blue).
+Here we plot for each chromosome arm (x-axis) the number of samples having a `dupli` (y-axis > 0, in red) or a `loss` (y-axis < 0, in blue).
 
@@
 
-Comparing with the profile in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1), we identified important CNV known to play a role in Wilms tumor, including:
+Comparing with the profile in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1), we identified important CNV known to play a role in Wilms tumor, including:
 
@@
 
-- We couldn't detect any *chr4q* loss as described in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+- We couldn't detect any *chr4q* loss as described in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
 
-- We identified 2 losses in *chr5* and *chr6* that do not show up in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+- We identified 2 losses in *chr5* and *chr6* that do not show up in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
 These may be `infercnv` induced false-positive CNV.
@@
 Our major concern is the pattern of the *chr1*. 
-We do see a tendency of *chr1p* loss and *chr1q* gain but it is not as pronounced as in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
-*chr1q* gain is however one of the most promising predictive marker of Wilms tumor outcome ([Nelson et al, Current Opinion in Pediatrics (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/)). 
-It is however important to notice that the profile shown in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1) derived from SIOP samples that have been pre-treated with chemotherapy, while samples in the dataset are enriched in `upfront resection` samples that haven't been pre-treated, according to the COG protocol.
+We do see a tendency of *chr1p* loss and *chr1q* gain but it is not as pronounced as in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+*chr1q* gain is however one of the most promising predictive marker of Wilms tumor outcome ([Nelson et al. (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/)).
+It is however important to notice that the profile shown in [Cresswell et al.
+(2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1) derived from SIOP samples that have been pre-treated with chemotherapy, while samples in the dataset are enriched in `upfront resection` samples that haven't been pre-treated, according to the COG protocol.
 So far, we do not know how the pre-operative chemotherapy impact the clonal selection of specific CNV.
@@
 
-As described in [Nelson et al, Current Opinions in Pediatrics (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/), [Phelps et al, Children (2018)](https://pmc.ncbi.nlm.nih.gov/articles/PMC6262554/) and [Zheng et al, Frontiers in Oncology (2023)](https://www.frontiersin.org/journals/oncology/articles/10.3389/fonc.2023.1137346/full), LOH at *chr1p*, *chr11p* and *chr16q* as well as *chr1q* gain are prognostic factors.
+As described in [Nelson et al.
+(2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/), [Phelps et al.
+(2018)](https://pmc.ncbi.nlm.nih.gov/articles/PMC6262554/) and [Zheng et al.
+(2023)](https://www.frontiersin.org/journals/oncology/articles/10.3389/fonc.2023.1137346/full), LOH at *chr1p*, *chr11p* and *chr16q* as well as *chr1q* gain are prognostic factors.
 
@@
 
-As we don't know the effect of the treatment on the selection of cells with CNV, it is difficult to compare our CNV profile with the one in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+As we don't know the effect of the treatment on the selection of cells with CNV, it is difficult to compare our CNV profile with the one in [Cresswell et al.
+(2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
 We those check the repartition of CNV in samples that have been pre-treated (i.e `Resection post chemotherapy`) with samples that have been removed surgically before any treatment (i.e `Upfront resection`).
@@
 Our analysis suggest that there is indeed an enrichment in cells with CNV after chemotherapy, including a tendency of more *chr1q* gain samples after chemotherapy.
-This could explain some of the differences observed between our mean CNV profile and the one published in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+This could explain some of the differences observed between our mean CNV profile and the one published in [Cresswell et al.
+(2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
 
@@
 
-Immune cells are well annotated based on `PTPRC` expression.
+Immune cells are well annotated based on *PTPRC* expression.
 
@@
 
-Endothelial cells are well annotated based on `VWF` expression.
+Endothelial cells are well annotated based on *VWF* expression.
 

### 2025-03-17 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
fix a few more spots including remaining genes

@@
 |----|----|----|----|----|
-| normal | endothelial | `compartment == "endothelium" & predicted.score > pred.thr` | `VWF` | no CNV |
-| normal | immune | `compartment == "immune" & predicted.score > pred.thr` | `PTPRC`, `CD163`, `CD68` | no CNV |
-| normal | kidney | `compartment == "fetal_nephron" & predicted.score > pred.thr & cnv_score < cnv.thr.lower` | `CDH1`, `PODXL`, `LTL` | no CNV |
-| normal | stroma | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr.upper` | `VIM` | no CNV |
-| cancer | stroma | `compartment == "stroma" & cnv_score > cnv.thr.upper` | `VIM` | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
-| cancer | blastema | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr.upper` | `CITED1` | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
-| cancer | epithelial | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr.upper` | `CDH1` | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| normal | endothelial | `compartment == "endothelium" & predicted.score > pred.thr` | *VWF* | no CNV |
+| normal | immune | `compartment == "immune" & predicted.score > pred.thr` | *PTPRC*, *CD163*, *CD68* | no CNV |
+| normal | kidney | `compartment == "fetal_nephron" & predicted.score > pred.thr & cnv_score < cnv.thr.lower` | *CDH1*, *PODXL*, *LTL* | no CNV |
+| normal | stroma | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr.upper` | *VIM* | no CNV |
+| cancer | stroma | `compartment == "stroma" & cnv_score > cnv.thr.upper` | *VIM* | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| cancer | blastema | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr.upper` | *CITED1* | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| cancer | epithelial | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr.upper` | *CDH1* | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
 | unknown | \- | the rest of the cells | \- | \- |
@@
 
-We simply checked the `cnv_chr` for each chromosome across cells.
+For each cell, we checked the level of CNV for each chromosome.
 
@@
 - validate that mostly tumor cells do present high percentages of cells with CNV (by definition)
-- identify chromosome and cell types that might be more sensitive to *FALSE* positive CNV, such as `chr6p` which is a hub for immune genes
+- identify chromosome and cell types that might be more sensitive to *FALSE* positive CNV, such as *chr6p* which is a hub for immune genes
 - get the repartition per sample of the CNV
@@
 
-#### Immune, `PTPRC` expression
+#### Immune, *PTPRC* expression
 
@@
 
-#### Endothelium, `VWF` expression
+#### Endothelium, *VWF* expression
 
@@
 
-##### `COL6A3` expression
+##### *COL6A3* expression
 
@@
 
-##### `THY1` expression
+##### **THY1* expression
 
@@
 
-##### `CITED1` expression
+##### *CITED1* expression
 
@@
 
-##### `NCAM1` expression
+##### *NCAM1* expression
 
@@
 
-##### `SIX2` expression
+##### *SIX2* expression
 
@@
 
-##### `CDH1` expression
+##### *CDH1* expression
 
@@
 
-##### `PODXL` expression
+##### *PODXL* expression
 

### 2025-03-17 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
pubs

@@
 
-Here we aim to plot a mean CNV profile across the 35 annotated samples as presented in Figure 1A of [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
+Here we aim to plot a mean CNV profile across the 35 annotated samples as presented in Figure 1A of [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
 This represents the mean SNP Array profile of 64 pediatric kidney cancer (including 60 Wilms tumors).

### 2025-03-19 analyses/infercnv-consensus-cell-type/notebook-template.Rmd
establish skeleton module

--- 2024-09-13 analyses/cell-type-dsrct/notebook-template.Rmd
@@
 ---
-title: "cell-type-dsrct Analysis"
+title: "infercnv-consensus-cell-type Analysis"
 author:
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "cell-type-dsrct")
+module_base <- file.path(repository_base, "analyses", "infercnv-consensus-cell-type")
 ```

### 2025-03-19 analyses/infercnv-consensus-cell-type/notebooks/01_ewings-consensus-cell-types.Rmd
add notebook exploring immune cell counts

---
title: "Explore immune cell types in Ewing sarcoma"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output: html_notebook
---

## Introduction

The goal of this notebook is to explore the distribution of immune consensus cell types in Ewing sarcoma samples (`SCPCP000015`).

Based on these findings, we can determine approaches to evaluate for pooling immune cells to use as a normal reference for `inferCNV`.

## Setup

```{r}
options(readr.show_col_types = FALSE)
```

### Paths

#### Base directories

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current results directory, found within the repository base directory
result_dir <- file.path(repository_base, "data", "current", "results")

# Path to consensus cell types for this project
celltype_dir <- file.path(result_dir, "cell-type-consensus", "SCPCP000015")

# Path to immune cell list
immune_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-immune-cell-types.tsv"
```

#### Input and output files

Set paths to input and output directories and files in the chunk below.

```{r paths}
# Input files
celltype_files <- list.files(
  path = celltype_dir,
  pattern = "_processed_consensus-cell-types\\.tsv\\.gz$",
  recursive = TRUE,
  full.names = TRUE
)

# Output files
```

## Analysis

```{r}
immune_celltypes <- readr::read_tsv(immune_url) |>
  dplyr::pull(consensus_annotation)

celltype_df <- celltype_files |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind() |>
  dplyr::select(library_id, barcodes, consensus_annotation) |>
  dplyr::mutate(
    category = dplyr::case_when(
      consensus_annotation == "Unknown" ~ "Unknown",
      consensus_annotation %in% immune_celltypes ~ "immune",
      .default = "other"
    )
  )
```

```{r}
category_count_df <- celltype_df |>
  dplyr::count(library_id, category) |>
  tidyr::pivot_wider(
    names_from = category,
    values_from = n,
    values_fill = 0
  )

# this way we can print the full table at once; RMarkdown will only show 10 rows at once
knitr::kable(category_count_df)
```
It seems most immune cells are concentrated in 2-3 samples, and most samples don't have any immune cells.

How many immune cells are there total?

```{r}
total_immune <- sum(category_count_df$immune)
total_immune
```

`inferCNV` authors [recommend that at least 100 cells](https://github.com/broadinstitute/infercnv/issues/188) be used for a "normal" reference, so we are above that threshold.

If we were to use all of these, what fraction is that per sample?
We'll just divide the total number of cells by this count, not taking into consideration how many immune cells are actually in that sample, to get a rough sense of how many immune cells there are relative to all.

```{r}
celltype_df |>
  dplyr::count(library_id, name = "total_cells") |>
  dplyr::mutate(fraction_immune = round(total_immune / total_cells, 2)) |>
  knitr::kable()
```

For the low-quality sample `SCPCL001111` this fraction isn't very meaningful since there are fewer cells in the sample than there are pooled immune cells.
For all other samples, the fraction ranges from 0.03 - 0.13.

Let's look a bit more at the breakdown of immune cell types:

```{r}
celltype_df |>
  dplyr::filter(category == "immune") |>
  dplyr::count(library_id, consensus_annotation) |>
  # this will arrange columns (after pivoting) in descending order of total counts
  dplyr::arrange(desc(n)) |>
  tidyr::pivot_wider(
    names_from = consensus_annotation,
    values_from = n,
    values_fill = 0
  )
```

Let's look at the total number of macrophages and all types of T cell, since these are the dominant cell types here.

```{r}
celltype_df |>
  dplyr::mutate(
    consensus_annotation = ifelse(
      stringr::str_detect(consensus_annotation, "T cell"),
      "T cell type",
      consensus_annotation
    )
  ) |>
  dplyr::filter(consensus_annotation %in% c("T cell type", "macrophage")) |>
  dplyr::count(consensus_annotation)
```
These constitute the vast majority of cells, so it would also be an option to use just these groups of cell types.
If we were to want a single immune cell type, then macrophage is the only suitable option to get more than 100 cells.

## Compare to ewings cell typing

We should also check whether any of the immune cells were classified as a tumor cell in the `cell-type-ewings` analysis module.
Ideally this will not have been the case, but it's an important thing to check!

## Conclusions

* There are sufficient immune cells to pool together to run `inferCNV` specifying them as a normal reference
* We could also consider using only `macrophage` cells, or combined `macrophage` and all `T cell` types together, if we wanted to assess any potential effect of limit the types of immune cells feeding into `inferCNV`.

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-03-20 analyses/infercnv-consensus-cell-type/exploratory-notebooks/01_ewings-consensus-cell-types.Rmd
notebooks -> exploratory-notebooks

--- 2025-03-19 analyses/infercnv-consensus-cell-type/notebooks/01_ewings-consensus-cell-types.Rmd
@@
 
-# Path to consensus cell types for this project
-celltype_dir <- file.path(result_dir, "cell-type-consensus", "SCPCP000015")
+# Path to cell-type-ewings results, which also contains consensus annotations
+ewings_dir <- file.path(result_dir, "cell-type-ewings", "SCPCP000015")
 
@@
 celltype_files <- list.files(
-  path = celltype_dir,
-  pattern = "_processed_consensus-cell-types\\.tsv\\.gz$",
+  path = ewings_dir,
+  pattern = "_ewing-celltype-assignments\\.tsv$",
   recursive = TRUE,
@@
 )
-
-# Output files
 ```
@@
   purrr::list_rbind() |>
-  dplyr::select(library_id, barcodes, consensus_annotation) |>
+  dplyr::select(library_id, barcodes, ewing_annotation, consensus_annotation) |>
   dplyr::mutate(
@@
 
+```{r}
+celltype_df |>
+  dplyr::filter(category == "immune") |>
+  dplyr::count(ewing_annotation) |>
+  dplyr::filter(stringr::str_detect(ewing_annotation, "tumor"))
+```
+
+Well, it appears there are some consensus immune cells which were classified as `tumor`. 
+Which samples are these in?
+
+```{r}
+celltype_df |>
+  dplyr::filter(
+    stringr::str_detect(ewing_annotation, "tumor"),
+    category == "immune"
+  ) |>
+  dplyr::count(library_id, ewing_annotation, consensus_annotation)
+```
+
+These results show macrophage and T cells, which are the two most prevalent categories immune cell categories.
+We might want to investigate these cells further, but at least we should probably exclude them as "normal" cells for `inferCNV` since their true status is bit unclear.
+
 ## Conclusions
@@
 * We could also consider using only `macrophage` cells, or combined `macrophage` and all `T cell` types together, if we wanted to assess any potential effect of limit the types of immune cells feeding into `inferCNV`.
+* There are 17 cells which are annotated as an immune consensus cell type (specifically, macrophage and mature T cell) which the `cell-type-ewings` module annotated as tumor cells.
+We would probably want to exclude these cells from any normal reference.
 

### 2025-03-20 analyses/infercnv-consensus-cell-type/exploratory-notebooks/01_ewings-consensus-cell-types.Rmd
notebook updates and re-render

@@
 * There are sufficient immune cells to pool together to run `inferCNV` specifying them as a normal reference
-* We could also consider using only `macrophage` cells, or combined `macrophage` and all `T cell` types together, if we wanted to assess any potential effect of limit the types of immune cells feeding into `inferCNV`.
+* We could also consider combined `macrophage` and all `T cell` types together to assess any potential effect of limiting the types of immune cells feeding into `inferCNV`
+* We should exclude library `SCPCL001111` from analysis since it has less than 100 cells.  
 * There are 17 cells which are annotated as an immune consensus cell type (specifically, macrophage and mature T cell) which the `cell-type-ewings` module annotated as tumor cells.

### 2025-03-20 analyses/infercnv-consensus-cell-type/exploratory-notebooks/01_ewings-consensus-cell-types.Rmd
Update analyses/infercnv-consensus-cell-type/exploratory-notebooks/01_ewings-consensus-cell-types.Rmd
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

@@
 These results show macrophage and T cells, which are the two most prevalent categories immune cell categories.
-We might want to investigate these cells further, but at least we should probably exclude them as "normal" cells for `inferCNV` since their true status is bit unclear.
+We might want to investigate these cells further, but at least we should probably exclude them as "normal" cells for `inferCNV` since their true status is a bit unclear.
 

### 2025-04-14 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
WIP: Building up template notebook with plots

---
title: "`r glue::glue('Initial inferCNV results for {params$library_id}')`"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

## Introduction

Here, we have run `inferCNV` with the following settings:

* With two references, `all-immune` and `subset-immune`
  * The former contains all consensus immune cells, _excluding_ any which were also identified as tumor in the `cell-type-ewings` analysis module
  * The latter is a subset of those immune cells considering only macrophage and T-cell cell types.

## Setup

```{r}
options(readr.show_col_types = FALSE)

library(SingleCellExperiment)
library(ggplot2)
library(patchwork)

# Set default ggplot theme, customized for UMAPs
theme_set(theme_bw())

#   theme(
#     plot.margin = margin(rep(20, 4)),
#     axis.text = element_blank(),
#     axis.ticks = element_blank(),
#     panel.grid = element_blank(),
#     aspect.ratio = 1
#   )
```

### Paths

#### Base directories

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# InferCNV directories
analysis_dir <- file.path(repository_base, "analyses", "infercnv-consensus-cell-type")
ref_dir <- file.path(analysis_dir, "references")
infercnv_dir <- file.path(analysis_dir, "results", "SCPCP000015", params$sample_id)
infercnv_immune_dir <- file.path(infercnv_dir, "ref-all-immune")
infercnv_subset_dir <- file.path(infercnv_dir, "ref-subset-immune")

# OpenScpCA data and results directories
data_dir <- file.path(repository_base, "data", "current")
```

#### Input and output files

Set paths to input and output directories and files in the chunk below.

```{r paths}
# cnv validations
validation_tsv <- file.path(ref_dir, "cnv-validation.tsv")

# references themselves so we know which cells belong to reference
immune_ref_sce_file <- file.path(ref_dir, "normal-references", "SCPCP000015", "ref-all-immune.rds")
subset_ref_sce_file <- file.path(ref_dir, "normal-references", "SCPCP000015", "ref-subset-immune.rds")

# input infercnv files from both references, "all-immune" and "subset-immune"
immune_hmm_tsv <- file.path(infercnv_immune_dir, glue::glue("{params$library_id}_cnv-metadata.tsv"))
immune_png <- file.path(infercnv_immune_dir, glue::glue("{params$library_id}_infercnv.png"))
subset_hmm_tsv <- file.path(infercnv_subset_dir, glue::glue("{params$library_id}_cnv-metadata.tsv"))
subset_png <- file.path(infercnv_subset_dir, glue::glue("{params$library_id}_infercnv.png"))

# cell types
ewings_tsv <- file.path(
  data_dir,
  "results",
  "cell-type-ewings",
  "SCPCP000015",
  params$sample_id,
  glue::glue("{params$library_id}_ewing-celltype-assignments.tsv")
)

# processed sce file
sce_file <- file.path(
  data_dir,
  "SCPCP000015",
  params$sample_id,
  glue::glue("{params$library_id}_processed.rds")
)
```

- we also need to read in the cnv-validations.tsv

- section 1: what are the cnvs that they detected? do this from the hmm output at least
- section 2: what is the relationship between consensus cell types and the cnvs? is there one?
- section 3: what are the clusters? how big are they? what is their relationship to the cnv scores? do we want to show them on a UMAP?

### Read in data

```{r message = FALSE}
validation_df <- readr::read_tsv(validation_tsv)

celltype_df <- readr::read_tsv(ewings_tsv) |>
  dplyr::select(barcodes, consensus_annotation, ewing_annotation)

immune_ref_barcodes <- readr::read_rds(immune_ref_sce_file) |>
  colnames() |>
  stringr::str_split_i("-", 2)
subset_ref_barcodes <- readr::read_rds(subset_ref_sce_file) |>
  colnames() |>
  stringr::str_split_i("-", 2)

# note these tsvs already have an indicator column for the reference
# the `subcluster` contains the inferCNV-calculated clusters
infercnv_df <- readr::read_tsv(immune_hmm_tsv) |>
  dplyr::bind_rows(
    readr::read_tsv(subset_hmm_tsv)
  ) |>
  dplyr::mutate(
    barcodes = stringr::str_remove(cell_id, glue::glue("{params$library_id}-")),
    # add indicator columns for whether the cell is reference or query
    is_immune_reference = barcodes %in% immune_ref_barcodes,
    is_subset_reference = barcodes %in% subset_ref_barcodes
  )

# Separate normal reference from query cells
# cnv_query_df <- infercnv_df |>
#  dplyr::filter(
#    stringr::str_starts(cell_id, params$library_id)
#  ) |>
#  dplyr::mutate(
#    barcodes = stringr::str_remove(cell_id, glue::glue("{params$library_id}-"))
#  )
# cnv_reference_df <- dplyr::anti_join(infercnv_df, cnv_query_df)

# Read SCE and create data frame with full results
sce <- readr::read_rds(sce_file)

cnv_df <- sce |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # add in hmm results
  dplyr::left_join(infercnv_df, by = "barcodes") |>
  # add in cell type results
  dplyr::left_join(celltype_df)

rm(sce)

# remove MT and GL chromosomes
chr_columns_to_keep <- which(
  !stringr::str_detect(
    names(cnv_df),
    "chrMT|chrGL000219.1|chrGL000194.1"
  )
)
cnv_df <- cnv_df[, chr_columns_to_keep]
```

## InferCNV Heatmaps

Below, we show the `inferCNV` heatmaps from the `immune-all` and `immune-subset` normal references for this library.

### Reference with all immune cells

![InferCNV heatmap with immune-all reference](`r immune_png`)

### Reference with a subset of immune cells

![InferCNV heatmap with immune-subset reference](`r immune_png`)

## Distributions of CNVs per cell

First, the total number of chromosomes with CNVs, with the goal of comparing reference to query cells.
We hope for a bimodal distribution where reference cells are near zero, and higher values for putative tumor cells in the query set.

```{r}
# grab any columns that have has_cnv
has_cnv_cols <- names(cnv_df)[stringr::str_starts(names(cnv_df), "has_cnv")]

per_cell_cnv_df <- cnv_df |>
  dplyr::mutate(total_cnv_per_cell = rowSums(dplyr::across(has_cnv_cols))) |>
  dplyr::select(
    barcodes,
    UMAP1,
    UMAP2,
    infercnv_cluster = subcluster,
    total_cnv_per_cell,
    consensus_annotation,
    ewing_annotation,
    normal_reference,
    is_immune_reference,
    is_subset_reference
  )
```

```{r fig.width = 10, fig.height = 5}
all_plot <- per_cell_cnv_df |>
  dplyr::filter(normal_reference == "all-immune") |>
  dplyr::rename(in_reference = is_immune_reference) |>
  ggplot() +
  aes(x = total_cnv_per_cell, fill = in_reference) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Immune-all reference"
  ) +
  theme(legend.position = "bottom")

subset_plot <- per_cell_cnv_df |>
  dplyr::filter(normal_reference == "subset-immune") |>
  dplyr::rename(in_reference = is_subset_reference) |>
  ggplot() +
  aes(x = total_cnv_per_cell, fill = in_reference) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Immune-subset reference"
  ) +
  theme(legend.position = "bottom")

wrap_plots(all_plot, subset_plot, guides = "collect") +
  plot_annotation(title = "Total CNVs per cell") &
  theme(legend.position = "bottom")
```
Now, let's look at this total CNV per cell but across cell types, using our two sets:

```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
consensus_violin <- ggplot(per_cell_cnv_df) +
  aes(
    x = consensus_annotation,
    y = total_cnv_per_cell,
    fill = normal_reference
  ) +
  geom_violin(scale = "width") +
  scale_fill_brewer(palette = "Pastel2") +
  labs(title = "Annotations from cell-type-consensus") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ewings_violin <- ggplot(per_cell_cnv_df) +
  aes(
    x = ewing_annotation,
    y = total_cnv_per_cell,
    fill = normal_reference
  ) +
  geom_violin(scale = "width") +
  scale_fill_brewer(palette = "Pastel2") +
  labs(title = "Annotations from cell-type-ewings") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

wrap_plots(consensus_violin, ewings_violin, ncol = 1, guides = "collect") +
  plot_annotation(title = "CNV scores across cell types") &
  theme(legend.position = "bottom")
```

## UMAPs

```{r fig.width = 10, fig.height = 5}
### UMAPs of the cell types as a guide ####
# TODO: colors could be more matched but may not be worth it?

ewings_umap <- per_cell_cnv_df |>
  # arbitrary filtering, this is just data
  dplyr::filter(normal_reference == "all-immune") |>
  ggplot() +
  aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(ewing_annotation, 8)) +
  geom_point(size = 0.1, alpha = 0.3) +
  facet_wrap(vars(normal_reference)) +
  labs(
    title = "Ewings cell type",
    color = "annotation"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))

consensus_umap <- per_cell_cnv_df |>
  # arbitrary filtering, this is just data
  dplyr::filter(normal_reference == "all-immune") |>
  ggplot() +
  aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(consensus_annotation, 8)) +
  geom_point(size = 0.1, alpha = 0.3) +
  facet_wrap(vars(normal_reference)) +
  labs(
    title = "Consensus cell type",
    color = "annotation"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))

wrap_plots(ewings_umap, consensus_umap)
```

```{r fig.width=10}
ggplot(per_cell_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
  geom_point(size = 0.1, alpha = 0.3) +
  scale_color_viridis_c() +
  facet_wrap(vars(normal_reference)) +
  labs(title = "Total CNV per cell")

cells_per_cluster <- per_cell_cnv_df |>
  dplyr::count(infercnv_cluster, normal_reference)

# How many cells per cluster?
ggplot(cells_per_cluster) +
  aes(
    x = infercnv_cluster, # keep in default order so "reference" and "unknown" are grouped
    y = n
  ) +
  geom_col() +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# How many CNV per cell per cluster?
ggplot(per_cell_cnv_df) +
  aes(x = infercnv_cluster, y = total_cnv_per_cell) +
  geom_point(size = 2) +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## CNV across chromosomes

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-04-14 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
WIP: Adding text to the template

@@
 
-Here, we have run `inferCNV` with the following settings:
+This notebook explores `inferCNV` results run on library `params$library_id` from `SCPCP000015`.
+`inferCNV` was run with two different normal references:
 
-* With two references, `all-immune` and `subset-immune`
-  * The former contains all consensus immune cells, _excluding_ any which were also identified as tumor in the `cell-type-ewings` analysis module
-  * The latter is a subset of those immune cells considering only macrophage and T-cell cell types.
+* `all-immune`: All consensus immune cells across `SCPCP000015` libraries, _excluding_ any which were also identified as tumor in the `cell-type-ewings` analysis module
+* `subset-immnune`: A subset of `all-immune`, considering only macrophage and T-cell cell types
+
+`inferCNV` was run with the following settings:
+
+* We used the `i3` HMM which has improved run times relative to their `i6` HMM
+* We used the default `analysis_mode = "subclusters"`, which allows the HMM to call CNVs at the level of clusters rather than whole samples or individual cells
+* We used the following clustering parameters:
+  * Leiden clustering run on PCA (not the expression matrix)
+  * `k=20` nearest neighbors with a `resolution` parameter of 1 and using the `modularity` objective function
 
@@
 
-#   theme(
-#     plot.margin = margin(rep(20, 4)),
-#     axis.text = element_blank(),
-#     axis.ticks = element_blank(),
-#     panel.grid = element_blank(),
-#     aspect.ratio = 1
-#   )
+umap_theme <- list(
+  coord_fixed(),
+  theme_classic(),
+  theme(
+    axis.ticks = element_blank(),
+    axis.text = element_blank()
+  )
+)
 ```
@@
 
-Set paths to input and output directories and files in the chunk below.
-
 ```{r paths}
-# cnv validations
+# TSV with expected Ewing CNVs
 validation_tsv <- file.path(ref_dir, "cnv-validation.tsv")
 
-# references themselves so we know which cells belong to reference
+# References themselves so we know which cells belong to reference
 immune_ref_sce_file <- file.path(ref_dir, "normal-references", "SCPCP000015", "ref-all-immune.rds")
@@
 
-# cell types
+# ewings and consensus cell types
 ewings_tsv <- file.path(
@@
 
-- we also need to read in the cnv-validations.tsv
-
-- section 1: what are the cnvs that they detected? do this from the hmm output at least
-- section 2: what is the relationship between consensus cell types and the cnvs? is there one?
-- section 3: what are the clusters? how big are they? what is their relationship to the cnv scores? do we want to show them on a UMAP?
-
 ### Read in data
@@
 ```{r message = FALSE}
-validation_df <- readr::read_tsv(validation_tsv)
+validation_df <- readr::read_tsv(validation_tsv) |>
+  dplyr::filter(diagnosis == "Ewing sarcoma")
 
@@
 
-# Separate normal reference from query cells
-# cnv_query_df <- infercnv_df |>
-#  dplyr::filter(
-#    stringr::str_starts(cell_id, params$library_id)
-#  ) |>
-#  dplyr::mutate(
-#    barcodes = stringr::str_remove(cell_id, glue::glue("{params$library_id}-"))
-#  )
-# cnv_reference_df <- dplyr::anti_join(infercnv_df, cnv_query_df)
-
 # Read SCE and create data frame with full results
-sce <- readr::read_rds(sce_file)
-
-cnv_df <- sce |>
+cnv_df <- readr::read_rds(sce_file) |>
   scuttle::makePerCellDF(use.dimred = "UMAP") |>
@@
   # add in cell type results
-  dplyr::left_join(celltype_df)
-
-rm(sce)
-
+  dplyr::left_join(celltype_df) |>
   # remove MT and GL chromosomes
-chr_columns_to_keep <- which(
-  !stringr::str_detect(
-    names(cnv_df),
-    "chrMT|chrGL000219.1|chrGL000194.1"
+  dplyr::select(
+    -contains("chrMT"),
+    -contains("chrGL")
   )
-)
-cnv_df <- cnv_df[, chr_columns_to_keep]
 ```
@@
 
-## Distributions of CNVs per cell
+## Distribution of CNVs per cell
 
-First, the total number of chromosomes with CNVs, with the goal of comparing reference to query cells.
-We hope for a bimodal distribution where reference cells are near zero, and higher values for putative tumor cells in the query set.
+This section contains figures showing the total number of CNVs detected per cell.
+
+### CNVs in reference vs query cells
+
+The figures below compare the number of CNVs per cell between reference and query cells. 
+Ideally, we'll see a bimodal distribution where reference cells are at/near zero, and there is a peak of higher values which may represent tumor cells.
+In the density plots below, distributions are colored based on whether the cells are in the reference or not (aka, query cells).
 
@@
 per_cell_cnv_df <- cnv_df |>
-  dplyr::mutate(total_cnv_per_cell = rowSums(dplyr::across(has_cnv_cols))) |>
-  dplyr::select(
-    barcodes,
-    UMAP1,
-    UMAP2,
-    infercnv_cluster = subcluster,
-    total_cnv_per_cell,
-    consensus_annotation,
-    ewing_annotation,
-    normal_reference,
-    is_immune_reference,
-    is_subset_reference
-  )
+  dplyr::mutate(total_cnv_per_cell = rowSums(dplyr::across(dplyr::all_of(has_cnv_cols)))) |>
+  dplyr::rename(infercnv_cluster = subcluster)
 ```
@@
 ```
-Now, let's look at this total CNV per cell but across cell types, using our two sets:
+
+### CNVs across cell types
+
+Next, we'll look at the per-cell distributions of CNVs across cell types.
+Specifically, we'll look at both the cell types from the `cell-type-consensus` module and the `cell-type-ewings` module separately.
+Ideally, we'll see that Unknown cells from `cell-type-consensus` and cells labeled as tumor in `cell-type-ewings` will have relatively higher CNV scores compared to other cell types.
 
@@
 
-```{r fig.width = 10, fig.height = 5}
-### UMAPs of the cell types as a guide ####
-# TODO: colors could be more matched but may not be worth it?
+Next, we'll visualize the total CNV counts as a UMAP.
 
-ewings_umap <- per_cell_cnv_df |>
+First, as a guide, let's look at the top 10 `cell-type-ewings` cell types.
+
+```{r}
+per_cell_cnv_df |>
   # arbitrary filtering, this is just data
@@
   facet_wrap(vars(normal_reference)) +
-  labs(
-    title = "Ewings cell type",
-    color = "annotation"
-  ) +
-  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
-
-consensus_umap <- per_cell_cnv_df |>
-  # arbitrary filtering, this is just data
-  dplyr::filter(normal_reference == "all-immune") |>
-  ggplot() +
-  aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(consensus_annotation, 8)) +
-  geom_point(size = 0.1, alpha = 0.3) +
-  facet_wrap(vars(normal_reference)) +
-  labs(
-    title = "Consensus cell type",
-    color = "annotation"
-  ) +
-  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
-
-wrap_plots(ewings_umap, consensus_umap)
+  labs(color = "annotation") +
+  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
+  umap_theme
 ```
+Next, the total CNVs per cell; ideally, higher values will overlap with regions in the UMAP containing putative tumor cells.
 
@@
   facet_wrap(vars(normal_reference)) +
-  labs(title = "Total CNV per cell")
+  labs(title = "Total CNV per cell") +
+  umap_theme
+```
 
+## Clusters
+
+Part of `inferCNV`'s algorithm involves clustering the data, and poor clusters have the potential to lead to inaccurate inferences of CNV. 
+To see whether we'd like to further tune the clustering parameters, we'll get a size of some of their properties here:
+
+How many cells are there per cluster?
+Clusters labeled `reference_` contain only normal reference cells, and clusters labeled `unknown_` contain the query cells.
+
+```{r}
 cells_per_cluster <- per_cell_cnv_df |>
@@
 
-# How many cells per cluster?
 ggplot(cells_per_cluster) +
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
 
+```{r}
 # How many CNV per cell per cluster?
@@
 
+We want to look across chrs now, specifically:
+
+```{r}
+validation_df
+```
+
+```{r}
+cnv_validation_df <- per_cell_cnv_df |>
+  dplyr::select(
+    barcodes,
+    infercnv_cluster,
+    total_cnv_per_cell,
+    normal_reference,
+    is_immune_reference,
+    is_subset_reference,
+    all_of(
+      c(
+        glue::glue("proportion_scaled_cnv_chr{1:22}"),
+        glue::glue("proportion_scaled_loss_chr{1:22}"),
+        glue::glue("proportion_scaled_dupli_chr{1:22}")
+      )
+    )
+  ) |>
+  tidyr::pivot_longer(
+    starts_with("proportion_scaled_"),
+    names_to = "cnv_type_raw",
+    values_to = "proportion"
+  ) |>
+  dplyr::mutate(cnv_type_raw = stringr::str_remove(cnv_type_raw, "^proportion_scaled_")) |>
+  tidyr::separate(cnv_type_raw, into = c("cnv_type", "chr")) |>
+  dplyr::mutate(chr = stringr::str_remove(chr, "^chr"), chr = as.numeric(chr))
+```
+
+```{r}
+cnv_validation_df |>
+  dplyr::filter(normal_reference == "subset-immune") |>
+  ggplot() +
+  aes(
+    x = proportion,
+    color = is_subset_reference
+  ) +
+  geom_density() +
+  facet_wrap(vars(chr))
+
+cnv_validation_df |>
+  dplyr::group_by(
+    normal_reference,
+    cnv_type,
+    chr
+  ) |>
+  dplyr::summarize(median_proportion = median(proportion)) |>
+  dplyr::arrange(cnv_type)
+```
+
 ## Session Info

### 2025-04-14 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
progress

@@
   ) |>
-  dplyr::mutate(
-    barcodes = stringr::str_remove(cell_id, glue::glue("{params$library_id}-")),
-    # add indicator columns for whether the cell is reference or query
-    is_immune_reference = barcodes %in% immune_ref_barcodes,
-    is_subset_reference = barcodes %in% subset_ref_barcodes
-  )
+  tidyr::separate(cell_id, into = c("library_id", "barcodes"))
 
-# Read SCE and create data frame with full results
-cnv_df <- readr::read_rds(sce_file) |>
+# Read SCE and save UMAP coordinates
+umap_df <- readr::read_rds(sce_file) |>
   scuttle::makePerCellDF(use.dimred = "UMAP") |>
-  # replace UMAP.1 with UMAP1
-  dplyr::rename_with(
-    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
-  ) |>
-  # add in hmm results
-  dplyr::left_join(infercnv_df, by = "barcodes") |>
-  # add in cell type results
-  dplyr::left_join(celltype_df) |>
-  # remove MT and GL chromosomes
   dplyr::select(
-    -contains("chrMT"),
-    -contains("chrGL")
+    barcodes,
+    UMAP1 = UMAP.1,
+    UMAP2 = UMAP.2
   )
@@
 
-![InferCNV heatmap with immune-subset reference](`r immune_png`)
+![InferCNV heatmap with immune-subset reference](`r subset_png`)
 
@@
 ```{r}
-# grab any columns that have has_cnv
-has_cnv_cols <- names(cnv_df)[stringr::str_starts(names(cnv_df), "has_cnv")]
-
-per_cell_cnv_df <- cnv_df |>
-  dplyr::mutate(total_cnv_per_cell = rowSums(dplyr::across(dplyr::all_of(has_cnv_cols)))) |>
-  dplyr::rename(infercnv_cluster = subcluster)
+# First, construct data frame of total `has_cnv` counts, with other variables needed for plots
+has_cnv_df <- infercnv_df |>
+  dplyr::select(
+    barcodes,
+    normal_reference,
+    subcluster,
+    starts_with("has_cnv_")
+  ) |>
+  tidyr::pivot_longer(
+    starts_with("has_cnv_"),
+    names_to = "chr",
+    values_to = "cnv"
+  ) |>
+  # sum all values for each cell, but separately for each run of inferCNV (aka each normal reference)
+  dplyr::group_by(barcodes, cell_id, normal_reference) |>
+  dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
+  # add indicator columns for whether the cell is reference or query
+  dplyr::mutate(
+    immune_reference_cell = barcodes %in% immune_ref_barcodes,
+    subset_reference_cell = barcodes %in% subset_ref_barcodes
+  ) |>
+  # join in other data
+  dplyr::left_join(celltype_df) |>
+  dplyr::left_join(umap_df)
 ```
@@
 ```{r fig.width = 10, fig.height = 5}
-all_plot <- per_cell_cnv_df |>
+all_plot <- has_cnv_df |>
   dplyr::filter(normal_reference == "all-immune") |>
-  dplyr::rename(in_reference = is_immune_reference) |>
   ggplot() +
-  aes(x = total_cnv_per_cell, fill = in_reference) +
+  aes(x = total_cnv_per_cell, fill = immune_reference_cell) +
   geom_density(alpha = 0.5) +
@@
 
-subset_plot <- per_cell_cnv_df |>
+subset_plot <- has_cnv_df |>
   dplyr::filter(normal_reference == "subset-immune") |>
-  dplyr::rename(in_reference = is_subset_reference) |>
   ggplot() +
-  aes(x = total_cnv_per_cell, fill = in_reference) +
+  aes(x = total_cnv_per_cell, fill = subset_reference_cell) +
   geom_density(alpha = 0.5) +
@@
 
-wrap_plots(all_plot, subset_plot, guides = "collect") +
-  plot_annotation(title = "Total CNVs per cell") &
-  theme(legend.position = "bottom")
+wrap_plots(all_plot, subset_plot) +
+  plot_annotation(title = "Total CNVs per cell")
 ```
@@
 Specifically, we'll look at both the cell types from the `cell-type-consensus` module and the `cell-type-ewings` module separately.
+
 Ideally, we'll see that Unknown cells from `cell-type-consensus` and cells labeled as tumor in `cell-type-ewings` will have relatively higher CNV scores compared to other cell types.
+In addition, comparing distributions between `all-immune` and `subset-immune` may be helpful to identify the optimal normal reference strategy.
 
 ```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
-consensus_violin <- ggplot(per_cell_cnv_df) +
+# drop NAs from reference cells from other libraries
+query_has_cnv_df <- has_cnv_df |>
+  tidyr::drop_na(ewing_annotation, consensus_annotation)
+
+consensus_violin <- ggplot(query_has_cnv_df) +
   aes(
@@
 
-ewings_violin <- ggplot(per_cell_cnv_df) +
+ewings_violin <- ggplot(query_has_cnv_df) +
   aes(
@@
 ```{r}
-per_cell_cnv_df |>
+query_has_cnv_df |>
   # arbitrary filtering, this is just data
@@
 
-```{r fig.width=10}
-ggplot(per_cell_cnv_df) +
+```{r fig.width=10, warning = FALSE}
+# add clusters into data frame
+has_cnv_df <- infercnv_df |>
+  dplyr::select(
+    barcodes,
+    normal_reference,
+    infercnv_cluster = subcluster
+  ) |>
+  dplyr::inner_join(
+    has_cnv_df,
+    by = c("barcodes", "normal_reference")
+  )
+
+# noting we expect warnings with this plot since there's no UMAP coordinates
+# for cells from other libraries in the reference
+ggplot(has_cnv_df) +
   aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
@@
 Part of `inferCNV`'s algorithm involves clustering the data, and poor clusters have the potential to lead to inaccurate inferences of CNV. 
-To see whether we'd like to further tune the clustering parameters, we'll get a size of some of their properties here:
+To see whether we'd like to further tune the clustering parameters, we'll get a sense of their size and rough composition here.
 
-How many cells are there per cluster?
+First, how many cells are there per cluster?
 Clusters labeled `reference_` contain only normal reference cells, and clusters labeled `unknown_` contain the query cells.
@@
 ```{r}
-cells_per_cluster <- per_cell_cnv_df |>
-  dplyr::count(infercnv_cluster, normal_reference)
-
-ggplot(cells_per_cluster) +
-  aes(
-    x = infercnv_cluster, # keep in default order so "reference" and "unknown" are grouped
-    y = n
-  ) +
-  geom_col() +
+ggplot(has_cnv_df) +
+  # keep in default order so "reference" and "unknown" are grouped
+  aes(x = infercnv_cluster) +
+  geom_bar() +
   facet_wrap(
@@
 ```
+
+Let's also look at the distributions of cell types in clusters.
+Is there a trend for which cell types tend to show up in which cluster?
+We'll plot this for both the consensus and ewing annotations, considering only the top 7 cell types for visual ease.
+
+```{r fig.width=12}
+ggplot(query_has_cnv_df) +
+  aes(x = infercnv_cluster, fill = forcats::fct_lump(consensus_annotation, 7)) +
+  geom_bar() +
+  scale_fill_brewer(palette = "Dark2") +
+  facet_wrap(
+    vars(normal_reference),
+    scales = "free_x"
+  ) +
+  labs(
+    title = "Consensus cell types",
+    fill = "annotation"
+  ) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+```{r fig.width=12}
+ggplot(query_has_cnv_df) +
+  aes(x = infercnv_cluster, fill = forcats::fct_lump(ewing_annotation, 7)) +
+  geom_bar() +
+  scale_fill_brewer(palette = "Set1") +
+  facet_wrap(
+    vars(normal_reference),
+    scales = "free_x"
+  ) +
+  labs(
+    title = "Ewing cell types",
+    fill = "annotation"
+  ) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+Let's also look more specifically at the `reference` clusters, where we might expect some batch effects since cells were taken from several different libraries.
+Do originating libraries cluster together?
+
+```{r fig.width=10}
+all_reference_clusters <- has_cnv_df |>
+  # bring in library_id
+  dplyr::inner_join(
+    infercnv_df |>
+      dplyr::select(barcodes, library_id) |>
+      unique()
+  ) |>
+  dplyr::filter(
+    normal_reference == "all-immune",
+    immune_reference_cell
+  )
+all_plot <- ggplot(all_reference_clusters) +
+  aes(x = infercnv_cluster, fill = library_id) +
+  geom_bar() +
+  labs("All immune reference") +
+  theme(legend.position = "bottom")
+
+subset_reference_clusters <- has_cnv_df |>
+  # bring in library_id
+  dplyr::inner_join(
+    infercnv_df |>
+      dplyr::select(barcodes, library_id) |>
+      unique()
+  ) |>
+  dplyr::filter(
+    normal_reference == "subset-immune",
+    subset_reference_cell
+  )
+subset_plot <- ggplot(subset_reference_clusters) +
+  aes(x = infercnv_cluster, fill = library_id) +
+  geom_bar() +
+  labs("Subset immune reference") +
+  theme(legend.position = "bottom")
+
+wrap_plots(all_plot, subset_plot, guides = "collect") &
+  theme(legend.position = "bottom")
+```
+
+How many CNVs were inferred per cluster?
+We expect a single value here, because with `analysis_mode = "subclusters"` (the `inferCNV` default), CNV parameters are estimated per cluster.
 
@@
 
-We want to look across chrs now, specifically:
+Now, we'll look at CNVs across chromosomes, including specific CNVs we expect to see in Ewing sarcoma:
 
@@
 ```{r}
-cnv_validation_df <- per_cell_cnv_df |>
+cnv_validation_df <- infercnv_df |>
   dplyr::select(

### 2025-04-15 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
Add a plot to directly compare references

@@
 * `all-immune`: All consensus immune cells across `SCPCP000015` libraries, _excluding_ any which were also identified as tumor in the `cell-type-ewings` analysis module
-* `subset-immnune`: A subset of `all-immune`, considering only macrophage and T-cell cell types
+* `subset-immune`: A subset of `all-immune`, considering only macrophage and T-cell cell types
 
@@
 
-* We used the `i3` HMM which has improved run times relative to their `i6` HMM
+* We used the `i3` HMM which has improved run times relative to the `i6` HMM.
 * We used the default `analysis_mode = "subclusters"`, which allows the HMM to call CNVs at the level of clusters rather than whole samples or individual cells
@@
 
-# Set default ggplot theme, customized for UMAPs
 theme_set(theme_bw())
@@
 
+To help contextualize these, here is the table of known CNVs in Ewing sarcoma we'll use for validation:
+
+```{r}
+validation_df
+```
+
 ### Reference with all immune cells
@@
   # sum all values for each cell, but separately for each run of inferCNV (aka each normal reference)
-  dplyr::group_by(barcodes, cell_id, normal_reference) |>
+  dplyr::group_by(barcodes, normal_reference) |>
   dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
@@
 
-First, as a guide, let's look at the top 10 `cell-type-ewings` cell types.
+First, as a guide to interpret the UMAP, let's color by the top 10 `cell-type-ewings` cell types.
 
@@
   umap_theme
+```
+
+## Compare references directly
+
+We can also look at CNV predictions directly between references using a point-density scatterplot, with the Spearman correlation shown:
+
+```{r}
+compare_refs <- has_cnv_df |>
+  # bring in library_id so we can only include those cells.
+  dplyr::inner_join(
+    infercnv_df |>
+      dplyr::select(barcodes, library_id) |>
+      unique()
+  ) |>
+  dplyr::filter(library_id == params$library_id) |>
+  dplyr::select(
+    barcodes, normal_reference, total_cnv_per_cell
+  ) |>
+  tidyr::pivot_wider(
+    names_from = normal_reference,
+    values_from = total_cnv_per_cell
+  ) |>
+  # color points this way rather than binning to ensure small bins aren't left out
+  dplyr::add_count(`all-immune`, `subset-immune`)
+
+# Determine max limit for axes
+max_axis <- max(has_cnv_df$total_cnv_per_cell)
+axis_breaks <- 1:max_axis
+
+# Calculate in spearman correlation coefficient to include in plot
+rho <- cor(compare_refs$`all-immune`, compare_refs$`subset-immune`, method = "spearman") |>
+  round(2)
+
+ggplot(compare_refs) +
+  aes(
+    x = `all-immune`,
+    y = `subset-immune`,
+    color = n
+  ) +
+  geom_point(size = 5) +
+  geom_abline(color = "red") +
+  scale_color_viridis_c() +
+  coord_equal() +
+  scale_x_continuous(limits = c(0, max_axis), breaks = 1:max_axis) +
+  scale_y_continuous(limits = c(0, max_axis), breaks = 1:max_axis) +
+  labs(
+    title = "Per-cell CNVs between normal references",
+    subtitle = glue::glue("rho = {rho}"),
+    color = "Number of cells"
+  ) +
+  theme(legend.position = "bottom")
 ```

### 2025-04-15 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
updates in general, and specifically in the cnv section at the end

@@
     toc_depth: 3
+    code_folding: hide
 params:
@@
 
+suppressPackageStartupMessages({
   library(SingleCellExperiment)
@@
   library(patchwork)
+})
 
@@
     normal_reference,
-    subcluster,
     starts_with("has_cnv_")
@@
   dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
+  dplyr::ungroup() |>
   # add indicator columns for whether the cell is reference or query
@@
 
-```{r}
+```{r, fig.width = 6}
 query_has_cnv_df |>
@@
 ```
+
 Next, the total CNVs per cell; ideally, higher values will overlap with regions in the UMAP containing putative tumor cells.
@@
 ```{r fig.width=10, warning = FALSE}
-# add clusters into data frame
-has_cnv_df <- infercnv_df |>
-  dplyr::select(
-    barcodes,
-    normal_reference,
-    infercnv_cluster = subcluster
-  ) |>
-  dplyr::inner_join(
-    has_cnv_df,
-    by = c("barcodes", "normal_reference")
-  )
-
 # noting we expect warnings with this plot since there's no UMAP coordinates
@@
 
-```{r}
+```{r, fig.height=6}
 compare_refs <- has_cnv_df |>
@@
 
-```{r}
-ggplot(has_cnv_df) +
+```{r, fig.width = 10}
+# bring in cluster and library information
+cluster_cnv_df <- has_cnv_df |>
+  dplyr::inner_join(
+    infercnv_df |>
+      dplyr::select(barcodes, library_id, subcluster, normal_reference) |>
+      unique(),
+    by = c("barcodes", "normal_reference")
+  )
+
+ggplot(cluster_cnv_df) +
   # keep in default order so "reference" and "unknown" are grouped
-  aes(x = infercnv_cluster) +
+  aes(x = subcluster) +
   geom_bar() +
@@
 ```{r fig.width=12}
-ggplot(query_has_cnv_df) +
-  aes(x = infercnv_cluster, fill = forcats::fct_lump(consensus_annotation, 7)) +
+cluster_cnv_df |>
+  # remove cells from other libraries
+  dplyr::filter(library_id == params$library_id) |>
+  ggplot() +
+  aes(x = subcluster, fill = forcats::fct_lump(consensus_annotation, 7)) +
   geom_bar() +
@@
 ```{r fig.width=12}
-ggplot(query_has_cnv_df) +
-  aes(x = infercnv_cluster, fill = forcats::fct_lump(ewing_annotation, 7)) +
+cluster_cnv_df |>
+  # remove cells from other libraries
+  dplyr::filter(library_id == params$library_id) |>
+  ggplot() +
+  aes(x = subcluster, fill = forcats::fct_lump(ewing_annotation, 7)) +
   geom_bar() +
@@
 ```{r fig.width=10}
-all_reference_clusters <- has_cnv_df |>
-  # bring in library_id
-  dplyr::inner_join(
-    infercnv_df |>
-      dplyr::select(barcodes, library_id) |>
-      unique()
-  ) |>
+all_reference_clusters <- cluster_cnv_df |>
   dplyr::filter(
@@
 all_plot <- ggplot(all_reference_clusters) +
-  aes(x = infercnv_cluster, fill = library_id) +
+  aes(x = subcluster, fill = library_id) +
   geom_bar() +
@@
 
-subset_reference_clusters <- has_cnv_df |>
+subset_reference_clusters <- cluster_cnv_df |>
   # bring in library_id
@@
 subset_plot <- ggplot(subset_reference_clusters) +
-  aes(x = infercnv_cluster, fill = library_id) +
+  aes(x = subcluster, fill = library_id) +
   geom_bar() +
@@
 # How many CNV per cell per cluster?
-ggplot(per_cell_cnv_df) +
-  aes(x = infercnv_cluster, y = total_cnv_per_cell) +
+ggplot(cluster_cnv_df) +
+  aes(x = subcluster, y = total_cnv_per_cell) +
   geom_point(size = 2) +
@@
 
-Now, we'll look at CNVs across chromosomes, including specific CNVs we expect to see in Ewing sarcoma:
+Now, we'll look at CNVs across chromosomes, specifically considering the `proportion_scaled_cnv_` output from the HMM.
+
+Again, these are the specific CNVs we expect to see in Ewing sarcoma:
 
@@
 ```{r}
+# First, construct data frame with proportion_scaled_ instead of has_cnv
 cnv_validation_df <- infercnv_df |>
@@
     barcodes,
-    infercnv_cluster,
-    total_cnv_per_cell,
+    library_id,
     normal_reference,
-    is_immune_reference,
-    is_subset_reference,
-    all_of(
-      c(
-        glue::glue("proportion_scaled_cnv_chr{1:22}"),
-        glue::glue("proportion_scaled_loss_chr{1:22}"),
-        glue::glue("proportion_scaled_dupli_chr{1:22}")
-      )
-    )
+    starts_with("proportion_scaled_")
   ) |>
+  # tidy up
   tidyr::pivot_longer(
@@
   ) |>
+  # more tidying
   dplyr::mutate(cnv_type_raw = stringr::str_remove(cnv_type_raw, "^proportion_scaled_")) |>
-  tidyr::separate(cnv_type_raw, into = c("cnv_type", "chr")) |>
-  dplyr::mutate(chr = stringr::str_remove(chr, "^chr"), chr = as.numeric(chr))
+  tidyr::separate(cnv_type_raw, sep = "_", into = c("cnv_type", "chr")) |>
+  # remove extra chromosomes and cells not in this library, and then can make chr numeric
+  dplyr::filter(
+    !stringr::str_starts(chr, "chrMT"),
+    !stringr::str_starts(chr, "chrGL"),
+    library_id == params$library_id
+  ) |>
+  dplyr::mutate(
+    chr = stringr::str_remove(chr, "^chr"),
+    chr = as.numeric(chr)
+  )
 ```
 
-```{r}
+We'll make density plots of the predicted CNV events across chromosomes, for each of the two normal references:
+
+```{r fig.height=8, fig.width=8}
+cnv_validation_df |>
+  dplyr::filter(normal_reference == "all-immune") |>
+  ggplot() +
+  aes(
+    x = proportion,
+    color = cnv_type
+  ) +
+  geom_density() +
+  facet_wrap(vars(chr)) +
+  ggtitle("All immune reference")
+
 cnv_validation_df |>
@@
     x = proportion,
-    color = is_subset_reference
+    color = cnv_type
   ) +
   geom_density() +
-  facet_wrap(vars(chr))
-
-cnv_validation_df |>
-  dplyr::group_by(
-    normal_reference,
-    cnv_type,
-    chr
-  ) |>
-  dplyr::summarize(median_proportion = median(proportion)) |>
-  dplyr::arrange(cnv_type)
+  facet_wrap(vars(chr)) +
+  ggtitle("Subset immune reference")
 ```

### 2025-04-15 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
clean up some redundant data frames

@@
 
-celltype_df <- readr::read_tsv(ewings_tsv) |>
-  dplyr::select(barcodes, consensus_annotation, ewing_annotation)
-
 immune_ref_barcodes <- readr::read_rds(immune_ref_sce_file) |>
@@
 
-# Read SCE and save UMAP coordinates
+# Read SCE and save UMAP coordinates and cell types
 umap_df <- readr::read_rds(sce_file) |>
@@
     UMAP2 = UMAP.2
+  ) |>
+  # TODO: We currently are bringing in cell types from separate files, but in the next
+  # data release they will be in the SCE objects
+  dplyr::inner_join(
+    readr::read_tsv(ewings_tsv) |>
+      dplyr::select(barcodes, consensus_annotation, ewing_annotation)
   )
@@
   ) |>
-  # join in other data
-  dplyr::left_join(celltype_df) |>
-  dplyr::left_join(umap_df)
+  # join in SCE information
+  dplyr::left_join(umap_df) |>
+  # finally, bring back in cluster and library information
+  dplyr::left_join(
+    infercnv_df |>
+      dplyr::select(barcodes, library_id, subcluster, normal_reference) |>
+      unique(),
+    by = c("barcodes", "normal_reference")
+  )
 ```
@@
 ```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
-# drop NAs from reference cells from other libraries
+# drop cells from other libraries
 query_has_cnv_df <- has_cnv_df |>
-  tidyr::drop_na(ewing_annotation, consensus_annotation)
+  dplyr::filter(library_id == params$library_id)
 
@@
 ```{r, fig.width = 6}
-query_has_cnv_df |>
-  # arbitrary filtering, this is just data
-  dplyr::filter(normal_reference == "all-immune") |>
-  ggplot() +
+ggplot(umap_df) +
   aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(ewing_annotation, 8)) +
   geom_point(size = 0.1, alpha = 0.3) +
-  facet_wrap(vars(normal_reference)) +
-  labs(color = "annotation") +
+  labs(color = "ewings annotation") +
   guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
@@
 compare_refs <- has_cnv_df |>
-  # bring in library_id so we can only include those cells.
-  dplyr::inner_join(
-    infercnv_df |>
-      dplyr::select(barcodes, library_id) |>
-      unique()
-  ) |>
   dplyr::filter(library_id == params$library_id) |>
@@
 ```{r, fig.width = 10}
-# bring in cluster and library information
-cluster_cnv_df <- has_cnv_df |>
-  dplyr::inner_join(
-    infercnv_df |>
-      dplyr::select(barcodes, library_id, subcluster, normal_reference) |>
-      unique(),
-    by = c("barcodes", "normal_reference")
-  )
-
-ggplot(cluster_cnv_df) +
+ggplot(has_cnv_df) +
   # keep in default order so "reference" and "unknown" are grouped
@@
 ```{r fig.width=12}
-cluster_cnv_df |>
-  # remove cells from other libraries
-  dplyr::filter(library_id == params$library_id) |>
-  ggplot() +
+ggplot(query_has_cnv_df) +
   aes(x = subcluster, fill = forcats::fct_lump(consensus_annotation, 7)) +
@@
 ```{r fig.width=12}
-cluster_cnv_df |>
-  # remove cells from other libraries
-  dplyr::filter(library_id == params$library_id) |>
-  ggplot() +
+ggplot(query_has_cnv_df) +
   aes(x = subcluster, fill = forcats::fct_lump(ewing_annotation, 7)) +
@@
 ```{r fig.width=10}
-all_reference_clusters <- cluster_cnv_df |>
+all_reference_clusters <- has_cnv_df |>
   dplyr::filter(
@@
 
-subset_reference_clusters <- cluster_cnv_df |>
+subset_reference_clusters <- has_cnv_df |>
   # bring in library_id
@@
 # How many CNV per cell per cluster?
-ggplot(cluster_cnv_df) +
+ggplot(has_cnv_df) +
   aes(x = subcluster, y = total_cnv_per_cell) +

### 2025-04-17 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
remove the scatterplot

@@
 
-* We used the `i3` HMM which has improved run times relative to the `i6` HMM.
+* We used the `i6` HMM
 * We used the default `analysis_mode = "subclusters"`, which allows the HMM to call CNVs at the level of clusters rather than whole samples or individual cells
@@
 
-## Compare references directly
-
-We can also look at CNV predictions directly between references using a point-density scatterplot, with the Spearman correlation shown:
-
-```{r, fig.height=6}
-compare_refs <- has_cnv_df |>
-  dplyr::filter(library_id == params$library_id) |>
-  dplyr::select(
-    barcodes, normal_reference, total_cnv_per_cell
-  ) |>
-  tidyr::pivot_wider(
-    names_from = normal_reference,
-    values_from = total_cnv_per_cell
-  ) |>
-  # color points this way rather than binning to ensure small bins aren't left out
-  dplyr::add_count(`all-immune`, `subset-immune`)
-
-# Determine max limit for axes
-max_axis <- max(has_cnv_df$total_cnv_per_cell)
-axis_breaks <- 1:max_axis
-
-# Calculate in spearman correlation coefficient to include in plot
-rho <- cor(compare_refs$`all-immune`, compare_refs$`subset-immune`, method = "spearman") |>
-  round(2)
-
-ggplot(compare_refs) +
-  aes(
-    x = `all-immune`,
-    y = `subset-immune`,
-    color = n
-  ) +
-  geom_point(size = 5) +
-  geom_abline(color = "red") +
-  scale_color_viridis_c() +
-  coord_equal() +
-  scale_x_continuous(limits = c(0, max_axis), breaks = 1:max_axis) +
-  scale_y_continuous(limits = c(0, max_axis), breaks = 1:max_axis) +
-  labs(
-    title = "Per-cell CNVs between normal references",
-    subtitle = glue::glue("rho = {rho}"),
-    color = "Number of cells"
-  ) +
-  theme(legend.position = "bottom")
-```
-
 ## Clusters

### 2025-04-17 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
Update cell type labeling scheme

@@
 
+# immune cell URL to check for ambiguous cell types
+immune_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-immune-cell-types.tsv"
+
 # References themselves so we know which cells belong to reference
@@
 
+immune_labels <- readr::read_tsv(immune_url) |>
+  dplyr::pull(consensus_annotation)
+
 immune_ref_barcodes <- readr::read_rds(immune_ref_sce_file) |>
@@
     UMAP2 = UMAP.2
+  )
+
+# TODO: We currently are bringing in cell types from separate files, but in the next
+# data release they will be in the SCE objects this could be part of the colData wrangling above.
+celltype_df <- readr::read_tsv(ewings_tsv) |>
+  # Label cells as:
+  # - tumor-immune: Cells labeled as tumor by cell-type-ewings, but immune by cell-type-consensus
+  # - tumor: Cells labeled as tumor by cell-type-ewings
+  # - their existing ewing_annotation if none of the above
+  dplyr::mutate(
+    celltype = dplyr::case_when(
+      ## tumor-immune
+      consensus_annotation %in% immune_labels &
+        stringr::str_starts(ewing_annotation, "tumor") ~ "tumor-immune",
+      ## tumor
+      stringr::str_starts(ewing_annotation, "tumor") ~ "tumor",
+      ## default
+      .default = ewing_annotation
+    )
   ) |>
-  # TODO: We currently are bringing in cell types from separate files, but in the next
-  # data release they will be in the SCE objects
-  dplyr::inner_join(
-    readr::read_tsv(ewings_tsv) |>
-      dplyr::select(barcodes, consensus_annotation, ewing_annotation)
-  )
+  dplyr::select(barcodes, celltype)
+
+umap_df <- umap_df |>
+  dplyr::left_join(celltype_df)
 ```
@@
 
-Ideally, we'll see that Unknown cells from `cell-type-consensus` and cells labeled as tumor in `cell-type-ewings` will have relatively higher CNV scores compared to other cell types.
-In addition, comparing distributions between `all-immune` and `subset-immune` may be helpful to identify the optimal normal reference strategy.
+Ideally, we'll see that tumor and Unknown cells from `cell-type-consensus` and cells labeled as tumor in `cell-type-ewings` will have relatively higher CNV scores compared to other cell types.
+
+In addition, it will be useful to:
+
+* See values for cells labeled `tumor-immune`, which were labeled as `tumor` by `cell-type-ewings` but as an immune cell type by `cell-type-consensus`
+* Compare distributions between `all-immune` and `subset-immune` which may support identifying the optimal normal reference strategy.
 
@@
 
-consensus_violin <- ggplot(query_has_cnv_df) +
+ggplot(query_has_cnv_df) +
   aes(
-    x = consensus_annotation,
+    x = celltype,
     y = total_cnv_per_cell,
@@
   scale_fill_brewer(palette = "Pastel2") +
-  labs(title = "Annotations from cell-type-consensus") +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-ewings_violin <- ggplot(query_has_cnv_df) +
-  aes(
-    x = ewing_annotation,
-    y = total_cnv_per_cell,
-    fill = normal_reference
-  ) +
-  geom_violin(scale = "width") +
-  scale_fill_brewer(palette = "Pastel2") +
-  labs(title = "Annotations from cell-type-ewings") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-wrap_plots(consensus_violin, ewings_violin, ncol = 1, guides = "collect") +
-  plot_annotation(title = "CNV scores across cell types") &
-  theme(legend.position = "bottom")
 ```
@@
 
-First, as a guide to interpret the UMAP, let's color by the top 10 `cell-type-ewings` cell types.
+First, as a guide to interpret the UMAP, let's color by cell types.
 
@@
 ggplot(umap_df) +
-  aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(ewing_annotation, 8)) +
+  aes(x = UMAP1, y = UMAP2, color = celltype) +
   geom_point(size = 0.1, alpha = 0.3) +
-  labs(color = "ewings annotation") +
   guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
@@
 Is there a trend for which cell types tend to show up in which cluster?
-We'll plot this for both the consensus and ewing annotations, considering only the top 7 cell types for visual ease.
 
@@
 ggplot(query_has_cnv_df) +
-  aes(x = subcluster, fill = forcats::fct_lump(consensus_annotation, 7)) +
-  geom_bar() +
-  scale_fill_brewer(palette = "Dark2") +
-  facet_wrap(
-    vars(normal_reference),
-    scales = "free_x"
-  ) +
-  labs(
-    title = "Consensus cell types",
-    fill = "annotation"
-  ) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-```
-
-```{r fig.width=12}
-ggplot(query_has_cnv_df) +
-  aes(x = subcluster, fill = forcats::fct_lump(ewing_annotation, 7)) +
+  aes(x = subcluster, fill = celltype) +
   geom_bar() +
@@
     scales = "free_x"
-  ) +
-  labs(
-    title = "Ewing cell types",
-    fill = "annotation"
   ) +

### 2025-04-18 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
small updates

@@
 * We used the default `analysis_mode = "subclusters"`, which allows the HMM to call CNVs at the level of clusters rather than whole samples or individual cells
-* We used the following clustering parameters:
-  * Leiden clustering run on PCA (not the expression matrix)
-  * `k=20` nearest neighbors with a `resolution` parameter of 1 and using the `modularity` objective function
+* We used the default `inferCNV` clustering parameters
 
@@
 ggplot(query_has_cnv_df) +
-  aes(x = subcluster, fill = celltype) +
+  aes(x = subcluster, fill = forcats::fct_lump(celltype, 8)) +
   geom_bar() +

### 2025-04-25 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
broad notebook updates for three references with default clustering params

@@
 ---
-title: "`r glue::glue('Initial inferCNV results for {params$library_id}')`"
+title: "`r glue::glue('inferCNV results for {params$library_id}')`"
 author: "Stephanie J. Spielman"
@@
 params:
+  library_id: SCPCL000822
   sample_id: SCPCS000490
-  library_id: SCPCL000822
 ---
@@
 This notebook explores `inferCNV` results run on library `params$library_id` from `SCPCP000015`.
-`inferCNV` was run with two different normal references:
+`inferCNV` was run with three different normal references:
 
-* `all-immune`: All consensus immune cells across `SCPCP000015` libraries, _excluding_ any which were also identified as tumor in the `cell-type-ewings` analysis module
-* `subset-immune`: A subset of `all-immune`, considering only macrophage and T-cell cell types
+* `endo`: All consensus endothelial cells across `SCPCP000015` libraries, _excluding_ any which were also identified as tumor in the `cell-type-ewings` analysis module
+* `immune`: All consensus immune cells across `SCPCP000015` libraries, _excluding_ any which were also identified as tumor in the `cell-type-ewings` analysis module
+* `endo-immune`: The `endo` and `immune` references combined
 
-`inferCNV` was run with the following settings:
+`inferCNV` was run with default settings, including:
 
-* We used the `i6` HMM
-* We used the default `analysis_mode = "subclusters"`, which allows the HMM to call CNVs at the level of clusters rather than whole samples or individual cells
-* We used the default `inferCNV` clustering parameters
+* The `i6` HMM
+* The setting `analysis_mode = "subclusters"`, which allows the HMM to call CNVs at the level of clusters rather than whole samples or individual cells
+* Default clustering parameters
 
@@
 ```{r base paths}
-# The base path for the OpenScPCA repository, found by its (hidden) .git directory
+# repo and analysis base dirs
 repository_base <- rprojroot::find_root(rprojroot::is_git_root)
+analysis_dir <- file.path(repository_base, "analyses", "infercnv-consensus-cell-type")
 
-# InferCNV directories
-analysis_dir <- file.path(repository_base, "analyses", "infercnv-consensus-cell-type")
-ref_dir <- file.path(analysis_dir, "references")
+# reference names, defined here to help define directories
+ref_names <- c("ref_endo", "ref_immune", "ref_endo-immune")
+
+# inferCNV directories
 infercnv_dir <- file.path(analysis_dir, "results", "SCPCP000015", params$sample_id)
-infercnv_immune_dir <- file.path(infercnv_dir, "ref-all-immune")
-infercnv_subset_dir <- file.path(infercnv_dir, "ref-subset-immune")
+infercnv_dirs <- file.path(infercnv_dir, ref_names)
+names(infercnv_dirs) <- ref_names
 
@@
 
-#### Input and output files
+#### Input files
 
@@
 # TSV with expected Ewing CNVs
-validation_tsv <- file.path(ref_dir, "cnv-validation.tsv")
+validation_tsv <- file.path(analysis_dir, "references", "cnv-validation.tsv")
 
@@
 
-# References themselves so we know which cells belong to reference
-immune_ref_sce_file <- file.path(ref_dir, "normal-references", "SCPCP000015", "ref-all-immune.rds")
-subset_ref_sce_file <- file.path(ref_dir, "normal-references", "SCPCP000015", "ref-subset-immune.rds")
+# processed sce file
+sce_file <- file.path(
+  data_dir,
+  "SCPCP000015",
+  params$sample_id,
+  glue::glue("{params$library_id}_processed.rds")
+)
 
-# input infercnv files from both references, "all-immune" and "subset-immune"
-immune_hmm_tsv <- file.path(infercnv_immune_dir, glue::glue("{params$library_id}_cnv-metadata.tsv"))
-immune_png <- file.path(infercnv_immune_dir, glue::glue("{params$library_id}_infercnv.png"))
-subset_hmm_tsv <- file.path(infercnv_subset_dir, glue::glue("{params$library_id}_cnv-metadata.tsv"))
-subset_png <- file.path(infercnv_subset_dir, glue::glue("{params$library_id}_infercnv.png"))
-
-# ewings and consensus cell types
+# consensus cell types, which we'll grab from the ewings results for now
+# TODO: We currently are bringing in cell types from separate files, but in the next
+# data release they will be in the SCE objects so we won't need this tsv
 ewings_tsv <- file.path(
@@
   glue::glue("{params$library_id}_ewing-celltype-assignments.tsv")
-)
-
-# processed sce file
-sce_file <- file.path(
-  data_dir,
-  "SCPCP000015",
-  params$sample_id,
-  glue::glue("{params$library_id}_processed.rds")
 )
@@
 
-immune_ref_barcodes <- readr::read_rds(immune_ref_sce_file) |>
-  colnames() |>
-  stringr::str_split_i("-", 2)
-subset_ref_barcodes <- readr::read_rds(subset_ref_sce_file) |>
-  colnames() |>
-  stringr::str_split_i("-", 2)
+# read infercnv files 
+infercnv_df <- infercnv_dirs |>
+  purrr::map(
+    \(input_dir) {
+      readr::read_tsv(
+        file.path(input_dir, glue::glue("{params$library_id}_cnv-metadata.tsv"))
+      ) |>
+        # indicator for reference vs query cell based on `subcluster` name
+        dplyr::mutate(cell_group = stringr::str_split_i(subcluster, "_", 1)) |>
+        # separate barcode and library
+        tidyr::separate(
+          cell_id, 
+          sep = "-", 
+          into = c("library_id", "barcodes"), 
+          remove = FALSE
+        )
+    }
+  ) |>
+  purrr::list_rbind(names_to = "normal_reference")
 
-# note these tsvs already have an indicator column for the reference
-# the `subcluster` contains the inferCNV-calculated clusters
-infercnv_df <- readr::read_tsv(immune_hmm_tsv) |>
-  dplyr::bind_rows(
-    readr::read_tsv(subset_hmm_tsv)
-  ) |>
-  tidyr::separate(cell_id, into = c("library_id", "barcodes"))
+# Extract a version of the data frame with just the infercnv metadata, not CNVs themselves
+infercnv_metadata_df <- infercnv_df |>
+  dplyr::select(
+    -starts_with("has_"), 
+    -starts_with("proportion_"),
+    -starts_with("top_")
+  )
 
@@
 
+```{r}
+# define the pngs
+infercnv_pngs <- file.path(infercnv_dirs, glue::glue("{params$library_id}_infercnv.png"))
+names(infercnv_pngs) <- ref_names
+```
+### Reference with all endothelial cells
+
+![InferCNV heatmap with endo reference](`r infercnv_pngs["endo"]`)
+
 ### Reference with all immune cells
 
-![InferCNV heatmap with immune-all reference](`r immune_png`)
+![InferCNV heatmap with immune reference](`r infercnv_pngs["immune"]`)
 
-### Reference with a subset of immune cells
+### Reference with both endothelial and immune cells
 
-![InferCNV heatmap with immune-subset reference](`r subset_png`)
+![InferCNV heatmap with end-immune reference](`r infercnv_pngs["endo-immune"]`)
 
@@
 Ideally, we'll see a bimodal distribution where reference cells are at/near zero, and there is a peak of higher values which may represent tumor cells.
-In the density plots below, distributions are colored based on whether the cells are in the reference or not (aka, query cells).
+In the density plots below, distributions are colored based on whether the cells are in the reference or not (aka, "unknown" cells).
 
@@
   dplyr::select(
+    library_id, 
     barcodes,
@@
   # sum all values for each cell, but separately for each run of inferCNV (aka each normal reference)
-  dplyr::group_by(barcodes, normal_reference) |>
+  dplyr::group_by(library_id, barcodes, normal_reference) |>
   dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
   dplyr::ungroup() |>
-  # add indicator columns for whether the cell is reference or query
-  dplyr::mutate(
-    immune_reference_cell = barcodes %in% immune_ref_barcodes,
-    subset_reference_cell = barcodes %in% subset_ref_barcodes
+  # bring back metadata
+  dplyr::inner_join(
+    infercnv_metadata_df, by = c("library_id", "barcodes", "normal_reference")
   ) |>
@@
   dplyr::left_join(umap_df) |>
-  # finally, bring back in cluster and library information
-  dplyr::left_join(
-    infercnv_df |>
-      dplyr::select(barcodes, library_id, subcluster, normal_reference) |>
-      unique(),
-    by = c("barcodes", "normal_reference")
-  )
+  # set levels
+  dplyr::mutate(normal_reference = factor(normal_reference, levels = ref_names))
 ```
@@
 ```{r fig.width = 10, fig.height = 5}
-all_plot <- has_cnv_df |>
-  dplyr::filter(normal_reference == "all-immune") |>
-  ggplot() +
-  aes(x = total_cnv_per_cell, fill = immune_reference_cell) +
+ggplot(has_cnv_df) +
+  aes(x = total_cnv_per_cell, fill = cell_group) +
   geom_density(alpha = 0.5) +
-  labs(
-    title = "Immune-all reference"
-  ) +
+  facet_wrap(vars(normal_reference)) +
   theme(legend.position = "bottom")
-
-subset_plot <- has_cnv_df |>
-  dplyr::filter(normal_reference == "subset-immune") |>
-  ggplot() +
-  aes(x = total_cnv_per_cell, fill = subset_reference_cell) +
-  geom_density(alpha = 0.5) +
-  labs(
-    title = "Immune-subset reference"
-  ) +
-  theme(legend.position = "bottom")
-
-wrap_plots(all_plot, subset_plot) +
-  plot_annotation(title = "Total CNVs per cell")
 ```
@@
 
-```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
+```{r fig.height=6, fig.width=12, message=FALSE, warning=FALSE}
 # drop cells from other libraries
@@
   ) +
-  geom_violin(scale = "width") +
+  geom_boxplot(linewidth = 0.25, outlier.size = 0.25) +
   scale_fill_brewer(palette = "Pastel2") +
@@
 
-First, how many cells are there per cluster?
-Clusters labeled `reference_` contain only normal reference cells, and clusters labeled `unknown_` contain the query cells.
+How many clusters are there, and how big are they?
 
-```{r, fig.width = 10}
-ggplot(has_cnv_df) +
-  # keep in default order so "reference" and "unknown" are grouped
-  aes(x = subcluster) +
+```{r}
+has_cnv_df |>
+  dplyr::count(subcluster) |>
+  dplyr::group_by(subcluster) |>
+  dplyr::arrange(desc(n))
+```
+
+```{r}
+summary(as.numeric(table(has_cnv_df$subcluster)))
+```
+
+Is there any relationship between cluster size and the number of CNVs?
+We would ideally not want there to be a relationship since this might indicate a bias.
+
+```{r}
+cluster_size_df <- has_cnv_df |>
+  dplyr::add_count(subcluster, name = "cluster_size") |>
+  dplyr::select(
+    total_cnv_per_cell,
+    cluster_size, 
+    normal_reference
+  ) |>
+  dplyr::distinct()
+
+ggplot(cluster_size_df) + 
+  aes(x = cluster_size, y = total_cnv_per_cell) +
+  geom_point() + 
+  facet_wrap(vars(normal_reference)) + 
+  geom_smooth(method = "lm")
+
+```
+
+Let's also look more specifically at the `reference` clusters, where we might expect some batch effects since cells were taken from several different libraries.
+Do originating libraries cluster together?
+
+```{r}
+reference_cells_df <- has_cnv_df |>
+  dplyr::filter(cell_group == "reference")
+
+ggplot(reference_cells_df) +
+  aes(x = subcluster, fill = library_id) +
   geom_bar() +
@@
   ) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-```
-
-Let's also look at the distributions of cell types in clusters.
-Is there a trend for which cell types tend to show up in which cluster?
-
-```{r fig.width=12}
-ggplot(query_has_cnv_df) +
-  aes(x = subcluster, fill = forcats::fct_lump(celltype, 8)) +
-  geom_bar() +
-  scale_fill_brewer(palette = "Set1") +
-  facet_wrap(
-    vars(normal_reference),
-    scales = "free_x"
-  ) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-```
-
-Let's also look more specifically at the `reference` clusters, where we might expect some batch effects since cells were taken from several different libraries.
-Do originating libraries cluster together?
-
-```{r fig.width=10}
-all_reference_clusters <- has_cnv_df |>
-  dplyr::filter(
-    normal_reference == "all-immune",
-    immune_reference_cell
+  theme(
+    legend.position = "bottom", 
+    axis.text.x = element_blank()
   )
-all_plot <- ggplot(all_reference_clusters) +
-  aes(x = subcluster, fill = library_id) +
-  geom_bar() +
-  labs("All immune reference") +
-  theme(legend.position = "bottom")
-
-subset_reference_clusters <- has_cnv_df |>
-  # bring in library_id
-  dplyr::inner_join(
-    infercnv_df |>
-      dplyr::select(barcodes, library_id) |>
-      unique()
-  ) |>
-  dplyr::filter(
-    normal_reference == "subset-immune",
-    subset_reference_cell
-  )
-subset_plot <- ggplot(subset_reference_clusters) +
-  aes(x = subcluster, fill = library_id) +
-  geom_bar() +
-  labs("Subset immune reference") +
-  theme(legend.position = "bottom")
-
-wrap_plots(all_plot, subset_plot, guides = "collect") &
-  theme(legend.position = "bottom")
-```
-
-How many CNVs were inferred per cluster?
-We expect a single value here, because with `analysis_mode = "subclusters"` (the `inferCNV` default), CNV parameters are estimated per cluster.
-
-```{r}
-# How many CNV per cell per cluster?
-ggplot(has_cnv_df) +
-  aes(x = subcluster, y = total_cnv_per_cell) +
-  geom_point(size = 2) +
-  facet_wrap(
-    vars(normal_reference),
-    scales = "free_x"
-  ) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 ```
@@
 # First, construct data frame with proportion_scaled_ instead of has_cnv
-cnv_validation_df <- infercnv_df |>
+prop_cnv_df <- infercnv_df |>
   dplyr::select(
+    library_id, 
     barcodes,
-    library_id,
     normal_reference,
@@
   ) |>
-  # tidy up
   tidyr::pivot_longer(
@@
   tidyr::separate(cnv_type_raw, sep = "_", into = c("cnv_type", "chr")) |>
-  # remove extra chromosomes and cells not in this library, and then can make chr numeric
+  # remove extra chromosomes and cells not in this library
   dplyr::filter(
@@
     chr = stringr::str_remove(chr, "^chr"),
-    chr = as.numeric(chr)
+    # set factor levels
+    chr = factor(chr, levels = 1:22), 
+    normal_reference = factor(normal_reference, levels = ref_names),
+    chr = forcats::fct_relevel(chr, as.character(validation_df$chromosome))
   )
@@
 
-We'll make density plots of the predicted CNV events across chromosomes, for each of the two normal references:
+We'll make density plots of the predicted CNV events across chromosomes, for each of the normal references:
+The panels will be ordered first by the chromosomes with known CNVs, and then in order for the rest of the chromosomes.
 
-```{r fig.height=8, fig.width=8}
-cnv_validation_df |>
-  dplyr::filter(normal_reference == "all-immune") |>
-  ggplot() +
+```{r fig.height=40, fig.width=8}
+ggplot(prop_cnv_df) + 
   aes(
@@
   geom_density() +
-  facet_wrap(vars(chr)) +
-  ggtitle("All immune reference")
-
-cnv_validation_df |>
-  dplyr::filter(normal_reference == "subset-immune") |>
-  ggplot() +
-  aes(
-    x = proportion,
-    color = cnv_type
+  # use wrap to get truly free scales 
+  facet_wrap(
+   vars(chr, normal_reference),
+   scales = "free", 
+   labeller = "label_both",
+   ncol = 3
   ) + 
-  geom_density() +
-  facet_wrap(vars(chr)) +
-  ggtitle("Subset immune reference")
+  theme(strip.text = element_text(size = 8))
 ```

### 2025-04-25 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd

@@
   dplyr::inner_join(
-    infercnv_metadata_df, by = c("library_id", "barcodes", "normal_reference")
+    infercnv_metadata_df,
+    by = c("library_id", "barcodes", "normal_reference")
   ) |>
@@
   geom_smooth(method = "lm")
-
 ```

### 2025-05-06 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
Add template notebook for ewings exploration of a single infercnv run with a single reference

---
title: "`r glue::glue('inferCNV results for {params$library_id} with {params$reference_name} reference')`"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  library_id: SCPCL000822
  sample_id: SCPCS000490
  reference_name: ref_endo
---

## Introduction

This notebook explores `inferCNV` results run on library `r params$library_id` from `SCPCP000015`.

```{r, results = 'asis'}
ref_text <- list(
  "ref_endo" = "endothelial",
  "ref_immune" = "immune",
  "ref_endo-immune" = "endothelial and immune"
)
glue::glue(
  "`inferCNV` was run with the following normal reference: `{params$reference_name}`.
  This reference contains all {ref_text[[params$reference_name]]}  cells across `SCPCP000015` libraries, excluding any which were also identified as tumor in the `cell-type-ewings` analysis module.
  "
)
```

## Setup

```{r}
options(readr.show_col_types = FALSE)

suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(ggplot2)
  library(patchwork)
})

theme_set(theme_bw())

umap_theme <- list(
  coord_fixed(),
  theme_classic(),
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )
)
```

### Paths

#### Base directories

```{r base paths}
# repo and analysis base dirs
repository_base <- rprojroot::find_root(rprojroot::is_git_root)
analysis_dir <- file.path(repository_base, "analyses", "infercnv-consensus-cell-type")

# OpenScpCA data and results directories
data_dir <- file.path(repository_base, "data", "current")

# inferCNV directory
infercnv_dir <- file.path(
  analysis_dir,
  "results",
  "SCPCP000015",
  params$sample_id,
  params$reference_name
)
```

#### Input files

```{r paths}
# TSV with expected Ewing CNVs
validation_tsv <- file.path(analysis_dir, "references", "cnv-validation.tsv")

# immune cell URL
immune_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-immune-cell-types.tsv"

# processed sce file
sce_file <- file.path(
  data_dir,
  "SCPCP000015",
  params$sample_id,
  glue::glue("{params$library_id}_processed.rds")
)

# ewing module cell types
ewings_tsv <- file.path(
  data_dir,
  "results",
  "cell-type-ewings",
  "SCPCP000015",
  params$sample_id,
  glue::glue("{params$library_id}_ewing-celltype-assignments.tsv")
)

# infercnv files
infercnv_tsv <- file.path(
  infercnv_dir,
  glue::glue("{params$library_id}_cnv-metadata.tsv")
)
infercnv_png <- file.path(
  infercnv_dir,
  glue::glue("{params$library_id}_infercnv.png")
)
```

### Read in data

```{r message = FALSE}
validation_df <- readr::read_tsv(validation_tsv) |>
  dplyr::filter(diagnosis == "Ewing sarcoma")

immune_labels <- readr::read_tsv(immune_url) |>
  dplyr::pull(consensus_annotation)

endo_labels <- c(
  "endothelial cell",
  "blood vessel endothelial cell",
  "microvascular endothelial cell"
)

# read infercnv results
infercnv_df <- readr::read_tsv(infercnv_tsv) |>
  # indicator for reference vs query cell based on `subcluster` name
  dplyr::mutate(cell_group = stringr::str_split_i(subcluster, "_", 1)) |>
  # separate barcode and library
  tidyr::separate(
    cell_id,
    sep = "-",
    into = c("library_id", "barcodes"),
    remove = FALSE
  )

# Extract a version of the data frame with just the infercnv metadata, not CNVs themselves
infercnv_metadata_df <- infercnv_df |>
  dplyr::select(
    -starts_with("has_"),
    -starts_with("proportion_"),
    -starts_with("top_")
  )

# Read SCE and save UMAP coordinates
umap_df <- readr::read_rds(sce_file) |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  dplyr::select(
    barcodes,
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2
  )

# Define cell types based on ewing module
celltype_df <- readr::read_tsv(ewings_tsv) |>
  # Label cells as:
  # - tumor-immune: Cells labeled as tumor by cell-type-ewings, but immune by cell-type-consensus
  # - tumor-endo: Cells labeled as tumor by cell-type-ewings, but endothelial by cell-type-consensus
  # - tumor: Cells labeled as tumor by cell-type-ewings
  # - their existing ewing_annotation if none of the above
  dplyr::mutate(
    celltype = dplyr::case_when(
      ## tumor-immune
      consensus_annotation %in% immune_labels &
        stringr::str_starts(ewing_annotation, "tumor") ~ "tumor-immune",
      ## tumor-endo
      consensus_annotation %in% endo_labels &
        stringr::str_starts(ewing_annotation, "tumor") ~ "tumor-endo",
      ## tumor
      stringr::str_starts(ewing_annotation, "tumor") ~ "tumor",
      ## default
      .default = ewing_annotation
    )
  ) |>
  dplyr::select(barcodes, celltype)

umap_df <- umap_df |>
  dplyr::left_join(celltype_df)
```

## InferCNV Heatmap

Below, we show the `inferCNV` heatmap for this library.
To help contextualize this image, here is the table of known CNVs in Ewing sarcoma used for validation:

```{r}
validation_df
```

![InferCNV heatmap](`r infercnv_png`)

## Distribution of CNVs per cell

This section contains figures showing the total number of CNVs detected per cell.

### CNVs in reference vs query cells

The figures below compare the number of CNVs per cell between reference and query cells.
Ideally, we'll see a bimodal distribution where reference cells are at/near zero, and there is a peak of higher values which may represent tumor cells.
In the density plots below, distributions are colored based on whether the cells are in the reference or not (aka, "unknown" cells).

```{r}
# First, construct data frame of total `has_cnv` counts, with other variables needed for plots
has_cnv_df <- infercnv_df |>
  dplyr::select(
    barcodes,
    starts_with("has_cnv_")
  ) |>
  tidyr::pivot_longer(
    starts_with("has_cnv_"),
    names_to = "chr",
    values_to = "cnv"
  ) |>
  # sum all values for each
  dplyr::group_by(barcodes) |>
  dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
  dplyr::ungroup() |>
  # bring back metadata
  dplyr::inner_join(
    infercnv_metadata_df,
    by = "barcodes"
  ) |>
  # join in SCE information
  dplyr::left_join(umap_df)
```

```{r fig.width = 10, fig.height = 5}
ggplot(has_cnv_df) +
  aes(x = total_cnv_per_cell, fill = cell_group) +
  geom_density(alpha = 0.5) +
  theme(legend.position = "bottom")
```

### CNVs across cell types

Next, we'll look at the per-cell distributions of CNVs across cell types, using labels derived from the `cell-type-ewings` module as follows:

* `tumor-endo`: Cells labeled as tumor by cell-type-ewings, but endothelial by `cell-type-consensus`
* `tumor-immune`: Cells labeled as tumor by cell-type-ewings, but immune by `cell-type-consensus`
* `tumor`: Cells labeled as tumor by `cell-type-ewings`
* `endo`: An endothelial cell type
* `immune`: An immune cell type
* If none of the above, their existing `cell-type-ewings` annotation

Ideally, we'll see that tumor and Unknown cells from will have relatively higher CNV scores compared to other cell types, in particular those cell types in the reference.

```{r fig.height=12, message=FALSE, warning=FALSE}
# drop cells from other libraries
query_has_cnv_df <- has_cnv_df |>
  dplyr::filter(library_id == params$library_id)

ggplot(query_has_cnv_df) +
  aes(
    y = celltype,
    x = total_cnv_per_cell,
    fill = cell_group
  ) +
  ggridges::geom_density_ridges2(alpha = 0.7, scale = 0.85)
```

## UMAPs

Next, we'll visualize the total CNV counts as a UMAP.
Below, the UMAP colored by cell type is shown on the left, and the UMAP colored by total CNV counts is shown on the right.
Ideally, unknown and tumor cells will overlap with regions in the UMAP with higher total CNV.

```{r}
```

```{r, fig.width = 12}
celltype_umap <- ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = celltype) +
  geom_point(size = 0.1, alpha = 0.3) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(title = "Cell types") +
  umap_theme

cnv_umap <- ggplot(query_has_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
  geom_point(size = 0.1, alpha = 0.3) +
  scale_color_viridis_c() +
  labs(title = "Total CNV per cell") +
  umap_theme

celltype_umap / cnv_umap
```

## Clusters

Part of `inferCNV`'s algorithm involves clustering the data, and poor clusters have the potential to lead to inaccurate inferences of CNV.
To see whether we'd like to further tune the clustering parameters, we'll get a sense of their size and rough composition here.

How many clusters are there, and how big are they?
We'll show this separately for reference vs query cells.

```{r}
cluster_count_df <- has_cnv_df |>
  dplyr::count(subcluster, cell_group)
ggplot(cluster_count_df) +
  aes(x = n, fill = cell_group) +
  geom_histogram(color = "grey20", bins = 15) +
  facet_wrap(vars(cell_group), scales = "free")
```

Is there any relationship between cluster size and the number of CNVs?
We would ideally not want there to be a relationship since this might indicate a bias.

```{r}
cluster_size_df <- has_cnv_df |>
  dplyr::add_count(subcluster, name = "cluster_size") |>
  dplyr::select(
    total_cnv_per_cell,
    cluster_size
  ) |>
  dplyr::distinct()

ggplot(cluster_size_df) +
  aes(x = cluster_size, y = total_cnv_per_cell) +
  geom_point() +
  geom_smooth(method = "lm")
```

Let's also look more specifically at the `reference` clusters, where we might expect some batch effects since cells were taken from several different libraries.
Do originating libraries cluster together?
When interpreting this plot, bear in mind that there might also be an association between cell types and libraries, and we would expect cell types to still cluster together.

```{r}
reference_cells_df <- has_cnv_df |>
  dplyr::filter(cell_group == "reference")

ggplot(reference_cells_df) +
  aes(x = subcluster, fill = library_id) +
  geom_bar() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_blank()
  )
```

## CNV across chromosomes

Now, we'll look at CNVs across chromosomes, specifically considering the `proportion_scaled_cnv_` output from the HMM.

Again, these are the specific CNVs we expect to see in Ewing sarcoma:

```{r}
validation_df
```

```{r}
# First, construct data frame with proportion_scaled_ instead of has_cnv
prop_cnv_df <- infercnv_df |>
  dplyr::select(
    library_id,
    barcodes,
    starts_with("proportion_scaled_")
  ) |>
  tidyr::pivot_longer(
    starts_with("proportion_scaled_"),
    names_to = "cnv_type_raw",
    values_to = "proportion"
  ) |>
  # more tidying
  dplyr::mutate(cnv_type_raw = stringr::str_remove(cnv_type_raw, "^proportion_scaled_")) |>
  tidyr::separate(cnv_type_raw, sep = "_", into = c("cnv_type", "chr")) |>
  # remove extra chromosomes and cells not in this library
  dplyr::filter(
    !stringr::str_starts(chr, "chrMT"),
    !stringr::str_starts(chr, "chrGL"),
    library_id == params$library_id
  ) |>
  dplyr::mutate(
    chr = stringr::str_remove(chr, "^chr"),
    # set factor levels
    chr = factor(chr, levels = 1:22),
    chr = forcats::fct_relevel(chr, as.character(validation_df$chromosome))
  )
```

We'll make density plots of the predicted CNV events across chromosomes, for each of the normal references:
The panels will be ordered first by the chromosomes with known CNVs, and then in order for the rest of the chromosomes.

```{r fig.height=12, fig.width=14}
ggplot(prop_cnv_df) +
  aes(x = proportion, color = cnv_type) +
  geom_density() +
  facet_wrap(
    vars(chr),
    scales = "free",
    ncol = 6
  ) +
  theme(strip.text = element_text(size = 8))
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-05-06 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
fix umap sizing

@@
 
-```{r}
-```
-
-```{r, fig.width = 12}
+```{r, fig.height = 10}
 celltype_umap <- ggplot(umap_df) +

### 2025-05-06 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
wording

@@
 
-We'll make density plots of the predicted CNV events across chromosomes, for each of the normal references:
+We'll make density plots of the predicted CNV events across chromosomes:
 The panels will be ordered first by the chromosomes with known CNVs, and then in order for the rest of the chromosomes.

### 2025-05-07 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
update ridge plot to remove low numbers of cells and include sorted n= labels

@@
 * `tumor`: Cells labeled as tumor by `cell-type-ewings`
-* `endo`: An endothelial cell type
-* `immune`: An immune cell type
 * If none of the above, their existing `cell-type-ewings` annotation
@@
 
-```{r fig.height=12, message=FALSE, warning=FALSE}
-# drop cells from other libraries
+```{r fig.height=8, message=FALSE, warning=FALSE}
+# drop cells from other libraries, and update data frame for plotting
 query_has_cnv_df <- has_cnv_df |>
-  dplyr::filter(library_id == params$library_id)
+  dplyr::filter(library_id == params$library_id) |>
+  dplyr::add_count(celltype) |>
+  # remove really low count cell types
+  dplyr::filter(n >= 3) |>
+  # add column with cell type counts, and arrange in order
+  dplyr::mutate(
+    celltype_n = glue::glue("{celltype} (n={n})"),
+    celltype_n = forcats::fct_infreq(celltype_n),
+    celltype_n = forcats::fct_rev(celltype_n)
+  )
 
 ggplot(query_has_cnv_df) +
-  aes(
-    y = celltype,
-    x = total_cnv_per_cell,
-    fill = cell_group
-  ) +
+  aes(x = total_cnv_per_cell, y = celltype_n) +
   ggridges::geom_density_ridges2(alpha = 0.7, scale = 0.85)

### 2025-05-07 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
update density plots to have two ridgeplots instead at different cell groupings, and begin updating cell type scheme

@@
 
-# Define cell types based on ewing module
-celltype_df <- readr::read_tsv(ewings_tsv) |>
-  # Label cells as:
-  # - tumor-immune: Cells labeled as tumor by cell-type-ewings, but immune by cell-type-consensus
-  # - tumor-endo: Cells labeled as tumor by cell-type-ewings, but endothelial by cell-type-consensus
-  # - tumor: Cells labeled as tumor by cell-type-ewings
-  # - their existing ewing_annotation if none of the above
-  dplyr::mutate(
-    celltype = dplyr::case_when(
-      ## tumor-immune
-      consensus_annotation %in% immune_labels &
-        stringr::str_starts(ewing_annotation, "tumor") ~ "tumor-immune",
-      ## tumor-endo
-      consensus_annotation %in% endo_labels &
-        stringr::str_starts(ewing_annotation, "tumor") ~ "tumor-endo",
-      ## tumor
-      stringr::str_starts(ewing_annotation, "tumor") ~ "tumor",
-      ## default
-      .default = ewing_annotation
-    )
-  ) |>
-  dplyr::select(barcodes, celltype)
-
-umap_df <- umap_df |>
-  dplyr::left_join(celltype_df)
+# Read ewings cell types
+celltype_df <- readr::read_tsv(ewings_tsv)
 ```
@@
 Ideally, we'll see a bimodal distribution where reference cells are at/near zero, and there is a peak of higher values which may represent tumor cells.
-In the density plots below, distributions are colored based on whether the cells are in the reference or not (aka, "unknown" cells).
 
@@
   # join in SCE information
-  dplyr::left_join(umap_df)
+  dplyr::left_join(celltype_df) |>
+  dplyr::left_join(umap_df) |>
+  # add cell type grouping column, with counts
+  dplyr::mutate(
+    celltype = dplyr::case_when(
+      cell_group == "reference" ~ "reference cell",
+      stringr::str_detect(ewing_annotation, "tumor") ~ "tumor cell",
+      .default = "other query cell"
+    )
+  ) |>
+  # add columns with cell type & group counts
+  dplyr::add_count(celltype, name = "celltype_count") |>
+  dplyr::add_count(cell_group, name = "cell_group_count") |>
+  dplyr::mutate(
+    celltype_n = glue::glue("{celltype} (n={celltype_count})"),
+    cell_group_n = glue::glue("{cell_group} (n={cell_group_count})")
+  )
+
+# Set the celltype_n factor order based on label, not frequency
+# We don't know the exact labels due to `(n=)`, so use grep
+has_cnv_df$celltype_n <- as.factor(has_cnv_df$celltype_n)
+present_values <- levels(has_cnv_df$celltype_n)
+has_cnv_df$celltype_n <- forcats::fct_relevel(
+  has_cnv_df$celltype_n,
+  c(
+    grep("reference", present_values, value = TRUE),
+    grep("tumor", present_values, value = TRUE),
+    grep("other", present_values, value = TRUE)
+  )
+)
 ```
 
-```{r fig.width = 10, fig.height = 5}
-ggplot(has_cnv_df) +
-  aes(x = total_cnv_per_cell, fill = cell_group) +
-  geom_density(alpha = 0.5) +
+Below, we show two ridge plots for the total detected CNVs across cell groups:
+
+* On the left, distributions are colored based on whether the cells are in the reference or not (aka, "query" cells)
+* On the right, distributions are colored based on a broad cell type category as follows:
+  * `reference cell`: Cells in the normal reference
+  * `tumor cell`: Cells labeled as tumor by the `cell-type-ewings` module
+  * `other query cell`: All other cells
+
+```{r fig.width = 12, fig.height = 5}
+left_plot <- ggplot(has_cnv_df) +
+  aes(y = forcats::fct_rev(cell_group_n), x = total_cnv_per_cell, fill = cell_group_n) +
+  ggridges::geom_density_ridges2(alpha = 0.7, scale = 0.85) +
+  labs(y = "Cell group (number of cells)") +
   theme(legend.position = "bottom")
+
+right_plot <- ggplot(has_cnv_df) +
+  aes(y = forcats::fct_rev(celltype_n), x = total_cnv_per_cell, fill = celltype_n) +
+  ggridges::geom_density_ridges2(alpha = 0.7, scale = 0.85) +
+  labs(y = "Cell type (number of cells)") +
+  theme(legend.position = "bottom") +
+  scale_fill_brewer(palette = "Dark2")
+
+left_plot + right_plot
 ```
@@
 Next, we'll look at the per-cell distributions of CNVs across cell types, using labels derived from the `cell-type-ewings` module as follows:
-
-* `tumor-endo`: Cells labeled as tumor by cell-type-ewings, but endothelial by `cell-type-consensus`
-* `tumor-immune`: Cells labeled as tumor by cell-type-ewings, but immune by `cell-type-consensus`
-* `tumor`: Cells labeled as tumor by `cell-type-ewings`
-* If none of the above, their existing `cell-type-ewings` annotation
 

### 2025-05-07 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
update UMAPs

@@
     axis.ticks = element_blank(),
-    axis.text = element_blank()
+    axis.text = element_blank(),
+    legend.position = "bottom"
   )
@@
 
-### CNVs in reference vs query cells
-
-The figures below compare the number of CNVs per cell between reference and query cells.
+The figures below compare the number of CNVs per cell between cell groupings.
 Ideally, we'll see a bimodal distribution where reference cells are at/near zero, and there is a peak of higher values which may represent tumor cells.
@@
   ggridges::geom_density_ridges2(alpha = 0.7, scale = 0.85) +
+  scale_fill_brewer(palette = "Dark2") +
   labs(y = "Cell group (number of cells)") +
@@
   labs(y = "Cell type (number of cells)") +
-  theme(legend.position = "bottom") +
-  scale_fill_brewer(palette = "Dark2")
+  theme(legend.position = "bottom")
 
 left_plot + right_plot
-```
-
-### CNVs across cell types
-
-Next, we'll look at the per-cell distributions of CNVs across cell types, using labels derived from the `cell-type-ewings` module as follows:
-
-Ideally, we'll see that tumor and Unknown cells from will have relatively higher CNV scores compared to other cell types, in particular those cell types in the reference.
-
-```{r fig.height=8, message=FALSE, warning=FALSE}
-# drop cells from other libraries, and update data frame for plotting
-query_has_cnv_df <- has_cnv_df |>
-  dplyr::filter(library_id == params$library_id) |>
-  dplyr::add_count(celltype) |>
-  # remove really low count cell types
-  dplyr::filter(n >= 3) |>
-  # add column with cell type counts, and arrange in order
-  dplyr::mutate(
-    celltype_n = glue::glue("{celltype} (n={n})"),
-    celltype_n = forcats::fct_infreq(celltype_n),
-    celltype_n = forcats::fct_rev(celltype_n)
-  )
-
-ggplot(query_has_cnv_df) +
-  aes(x = total_cnv_per_cell, y = celltype_n) +
-  ggridges::geom_density_ridges2(alpha = 0.7, scale = 0.85)
 ```
@@
 ```{r, fig.height = 10}
-celltype_umap <- ggplot(umap_df) +
-  aes(x = UMAP1, y = UMAP2, color = celltype) +
+# Filter to just cells in the library to avoid warnings
+query_has_cnv_df <- has_cnv_df |>
+  dplyr::filter(library_id == params$library_id)
+
+celltype_umap <- ggplot(query_has_cnv_df) +
+  aes(x = UMAP1, y = UMAP2, color = celltype_n) +
   geom_point(size = 0.1, alpha = 0.3) +
@@
 
-celltype_umap / cnv_umap
+celltype_umap + cnv_umap
 ```

### 2025-05-07 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
cluster plots

@@
 # Filter to just cells in the library to avoid warnings
-query_has_cnv_df <- has_cnv_df |>
+library_has_cnv_df <- has_cnv_df |>
   dplyr::filter(library_id == params$library_id)
 
-celltype_umap <- ggplot(query_has_cnv_df) +
+celltype_umap <- ggplot(library_has_cnv_df) +
   aes(x = UMAP1, y = UMAP2, color = celltype_n) +
@@
 
-cnv_umap <- ggplot(query_has_cnv_df) +
+cnv_umap <- ggplot(library_has_cnv_df) +
   aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
@@
 cluster_count_df <- has_cnv_df |>
-  dplyr::count(subcluster, cell_group)
+  dplyr::count(subcluster, cell_group_n)
 ggplot(cluster_count_df) +
-  aes(x = n, fill = cell_group) +
+  aes(x = n, fill = cell_group_n) +
   geom_histogram(color = "grey20", bins = 15) +
-  facet_wrap(vars(cell_group), scales = "free")
+  facet_wrap(vars(cell_group_n), scales = "free") +
+  labs(
+    x = "Number of cells in cluster",
+    y = "Number of clusters"
+  )
 ```
@@
 
-Let's also look more specifically at the `reference` clusters, where we might expect some batch effects since cells were taken from several different libraries.
+We'd also like to look at the distribution of cell types across clusters, specifically considering the non-reference clusters.
+For this, we'll focus on the top 10 cell types (based on the `cell-type-ewings` labels) considering all "tumor" as a single grouping.
+For visual ease, we'll only show clusters with at least 10 cells.
+
+```{r}
+query_only_df <- library_has_cnv_df |>
+  dplyr::filter(cell_group == "unknown") |>
+  # create a finer cell type column for this plot
+  dplyr::mutate(
+    celltype_fine = dplyr::case_when(
+      cell_group == "reference" ~ "reference cell",
+      stringr::str_detect(ewing_annotation, "tumor") ~ "tumor cell",
+      .default = ewing_annotation
+    )
+  ) |>
+  dplyr::add_count(celltype_fine, name = "celltype_fine_count") |>
+  dplyr::mutate(
+    celltype_fine_n = glue::glue("{celltype_fine} (n={celltype_fine_count})"),
+    # top 10 only
+    celltype_fine_n = forcats::fct_lump(celltype_fine_n, 10),
+  )
+
+# add n() to "Other" for plot clarity and update factor to frequency order
+current_levels <- levels(query_only_df$celltype_fine_n)
+total_other <- query_only_df |>
+  dplyr::filter(celltype_fine_n == "Other") |>
+  nrow()
+updated_other <- glue::glue("Other (n={total_other})")
+
+query_only_df <- query_only_df |>
+  dplyr::mutate(
+    # update other level to have counts
+    celltype_fine_n = as.character(celltype_fine_n),
+    celltype_fine_n = ifelse(celltype_fine_n == "Other", updated_other, celltype_fine_n),
+    # order based on freq
+    celltype_fine_n = as.factor(celltype_fine_n),
+    celltype_fine_n = forcats::fct_infreq(celltype_fine_n),
+    celltype_fine_n = forcats::fct_relevel(celltype_fine_n, updated_other, after = Inf)
+  ) |>
+  # exclude small clusters
+  dplyr::add_count(subcluster, name = "subcluster_size") |>
+  dplyr::filter(subcluster_size >= 10)
+
+# 11 hex colors!
+# adapted from 'Classic 10' from paletteer: https://r-graph-gallery.com/color-palette-finder
+colors <- c("#1F77B4", "#FF7F0E", "#2CA02C", "#D6272F", "#9467BD", "#8C564B", "#E377C2", "#7F7F7F", "#BCBD22", "#17CFFB", "#17CF1F")
+
+ggplot(query_only_df) +
+  aes(
+    x = forcats::fct_infreq(subcluster),
+    fill = celltype_fine_n
+  ) +
+  geom_bar() +
+  scale_fill_manual(values = colors) +
+  theme(axis.text.x = element_blank())
+```
+
+Finally, let's also look more specifically at the `reference` clusters, where we might expect some batch effects since cells were taken from several different libraries.
 Do originating libraries cluster together?

### 2025-05-07 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
density plot updates and other miscellany

@@
 
-# immune cell URL
-immune_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-immune-cell-types.tsv"
-
 # processed sce file
@@
   dplyr::filter(diagnosis == "Ewing sarcoma")
-
-immune_labels <- readr::read_tsv(immune_url) |>
-  dplyr::pull(consensus_annotation)
-
-endo_labels <- c(
-  "endothelial cell",
-  "blood vessel endothelial cell",
-  "microvascular endothelial cell"
-)
 
@@
 
-```{r, fig.height = 10}
+```{r, fig.height = 8}
 # Filter to just cells in the library to avoid warnings
@@
   aes(x = UMAP1, y = UMAP2, color = celltype_n) +
-  geom_point(size = 0.1, alpha = 0.3) +
+  geom_point(size = 0.25, alpha = 0.4) +
   guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
@@
   aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
-  geom_point(size = 0.1, alpha = 0.3) +
+  geom_point(size = 0.25, alpha = 0.4) +
   scale_color_viridis_c() +
@@
 
-```{r}
+```{r, fig.width = 8}
 cluster_count_df <- has_cnv_df |>
@@
 
-```{r}
+```{r, fig.width = 6}
 cluster_size_df <- has_cnv_df |>
@@
   geom_bar() +
+  scale_fill_brewer(palette = "Dark2") +
   theme(
@@
   ) |>
+  # set chr factor levels, with ones used for validation first
   dplyr::mutate(
     chr = stringr::str_remove(chr, "^chr"),
-    # set factor levels
     chr = factor(chr, levels = 1:22),
     chr = forcats::fct_relevel(chr, as.character(validation_df$chromosome))
+  ) |>
+  # finally, join in the cell types
+  dplyr::inner_join(
+    has_cnv_df |>
+      dplyr::select(barcodes, celltype_n)
   )
+
+prop_cnv_df
 ```
 
-We'll make density plots of the predicted CNV events across chromosomes:
+```{r}
+# separate gain and loss dfs
+loss_df <- prop_cnv_df |>
+  dplyr::filter(cnv_type == "loss")
+gain_df <- prop_cnv_df |>
+  dplyr::filter(cnv_type == "dupli")
+```
+
+We'll make density plots of the predicted CNV events across chromosomes, considering gain and loss separately.
 The panels will be ordered first by the chromosomes with known CNVs, and then in order for the rest of the chromosomes.
+Each line in the density plot represents a different cell type grouping.
 
 ```{r fig.height=12, fig.width=14}
-ggplot(prop_cnv_df) +
-  aes(x = proportion, color = cnv_type) +
+ggplot(loss_df) +
+  aes(x = proportion, color = celltype_n) +
   geom_density() +
@@
     scales = "free",
-    ncol = 6
+    ncol = 4 # this way the first row is the validation ones only
   ) +
+  ggtitle("Proportion of CNV loss across chromosomes for each cell type group") +
+  theme(strip.text = element_text(size = 8))
+```
+
+```{r fig.height=12, fig.width=14}
+ggplot(gain_df) +
+  aes(x = proportion, color = celltype_n) +
+  geom_density() +
+  facet_wrap(
+    vars(chr),
+    scales = "free",
+    ncol = 4 # this way the first row is the validation ones only
+  ) +
+  ggtitle("Proportion of CNV gain across chromosomes for each cell type group") +
   theme(strip.text = element_text(size = 8))

### 2025-05-07 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
plot styling updates

@@
     axis.ticks = element_blank(),
-    axis.text = element_blank(),
-    legend.position = "bottom"
+    axis.text = element_blank()
   )
@@
 
-```{r, fig.height = 8}
+```{r, fig.width = 12}
 # Filter to just cells in the library to avoid warnings
@@
   aes(x = UMAP1, y = UMAP2, color = celltype_n) +
-  geom_point(size = 0.25, alpha = 0.4) +
+  geom_point(size = 0.2, alpha = 0.3) +
   guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
@@
   aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
-  geom_point(size = 0.25, alpha = 0.4) +
+  geom_point(size = 0.2, alpha = 0.3) +
   scale_color_viridis_c() +
@@
 colors <- c("#1F77B4", "#FF7F0E", "#2CA02C", "#D6272F", "#9467BD", "#8C564B", "#E377C2", "#7F7F7F", "#BCBD22", "#17CFFB", "#17CF1F")
+```
 
+```{r, fig.width = 12}
 ggplot(query_only_df) +

### 2025-05-09 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd

@@
 # Extract a version of the data frame with just the infercnv metadata, not CNVs themselves
+# Note that we keep this separate for circumstances when the metadata has to be joined back in with results
 infercnv_metadata_df <- infercnv_df |>
@@
 has_cnv_df <- infercnv_df |>
-  dplyr::select(
-    barcodes,
-    starts_with("has_cnv_")
-  ) |>
   tidyr::pivot_longer(
@@
   dplyr::mutate(
-    celltype_n = glue::glue("{celltype} (n={celltype_count})"),
+    celltype = forcats::fct_relevel(celltype, c("reference cell", "tumor cell", "other query cell")),
+    # set celltype labels with (n=)
+    celltype = forcats::fct_relabel(
+      celltype,
+      # https://stackoverflow.com/questions/70510771/how-can-i-use-fct-relabel-but-maintain-original-factor-levels-or-how-can-i-set
+      ~ glue::glue("{levels(celltype)} (n = {celltype_count[match(levels(celltype), celltype)]})")
+    ),
     cell_group_n = glue::glue("{cell_group} (n={cell_group_count})")
-  )
-
-# Set the celltype_n factor order based on label, not frequency
-# We don't know the exact labels due to `(n=)`, so use grep
-has_cnv_df$celltype_n <- as.factor(has_cnv_df$celltype_n)
-present_values <- levels(has_cnv_df$celltype_n)
-has_cnv_df$celltype_n <- forcats::fct_relevel(
-  has_cnv_df$celltype_n,
-  c(
-    grep("reference", present_values, value = TRUE),
-    grep("tumor", present_values, value = TRUE),
-    grep("other", present_values, value = TRUE)
-  )
   )
@@
 right_plot <- ggplot(has_cnv_df) +
-  aes(y = forcats::fct_rev(celltype_n), x = total_cnv_per_cell, fill = celltype_n) +
+  aes(y = forcats::fct_rev(celltype), x = total_cnv_per_cell, fill = celltype) +
   ggridges::geom_density_ridges2(alpha = 0.7, scale = 0.85) +
@@
 celltype_umap <- ggplot(library_has_cnv_df) +
-  aes(x = UMAP1, y = UMAP2, color = celltype_n) +
+  aes(x = UMAP1, y = UMAP2, color = celltype) +
   geom_point(size = 0.2, alpha = 0.3) +
@@
 
-Now, we'll look at CNVs across chromosomes, specifically considering the `proportion_scaled_cnv_` output from the HMM.
+Now, we'll look at CNVs across chromosomes, specifically considering the `proportion_scaled_cnv_` output from the HMM focusing on gain and loss events.
 
@@
 prop_cnv_df <- infercnv_df |>
-  dplyr::select(
-    library_id,
-    barcodes,
-    starts_with("proportion_scaled_")
-  ) |>
   tidyr::pivot_longer(
@@
     has_cnv_df |>
-      dplyr::select(barcodes, celltype_n)
+      dplyr::select(barcodes, celltype)
   )
-
-prop_cnv_df
 ```
@@
 ggplot(loss_df) +
-  aes(x = proportion, color = celltype_n) +
+  aes(x = proportion, color = celltype) +
   geom_density() +
@@
 ggplot(gain_df) +
-  aes(x = proportion, color = celltype_n) +
+  aes(x = proportion, color = celltype) +
   geom_density() +

### 2025-05-09 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
incorporate color scheme into notebook

@@
 )
+
+# celltype palette file
+palette_file <- file.path("palettes", "SCPCP000015_palette.tsv")
 ```
@@
 
-# Read ewings cell types
+# Read ewings cell types and colors
 celltype_df <- readr::read_tsv(ewings_tsv)
+colors_df <- readr::read_tsv(palette_file)
 ```
@@
   dplyr::mutate(
-    celltype_fine = dplyr::case_when(
-      cell_group == "reference" ~ "reference cell",
-      stringr::str_detect(ewing_annotation, "tumor") ~ "tumor cell",
-      .default = ewing_annotation
+    celltype_fine = ifelse(
+      stringr::str_detect(ewing_annotation, "tumor"),
+      "tumor",
+      ewing_annotation
     )
   ) |>
-  dplyr::add_count(celltype_fine, name = "celltype_fine_count") |>
+  dplyr::add_count(celltype_fine) |>
   dplyr::mutate(
-    celltype_fine_n = glue::glue("{celltype_fine} (n={celltype_fine_count})"),
+    celltype_fine_n = glue::glue("{celltype_fine} (n={n})"),
     # top 10 only
-    celltype_fine_n = forcats::fct_lump(celltype_fine_n, 10),
+    celltype_fine_n = forcats::fct_lump(celltype_fine_n, 10)
   )
@@
 
-# 11 hex colors!
-# adapted from 'Classic 10' from paletteer: https://r-graph-gallery.com/color-palette-finder
-colors <- c("#1F77B4", "#FF7F0E", "#2CA02C", "#D6272F", "#9467BD", "#8C564B", "#E377C2", "#7F7F7F", "#BCBD22", "#17CFFB", "#17CF1F")
+# Create color palette with (n=) in labels
+color_n_df <- query_only_df |>
+  dplyr::select(celltype_fine, celltype_fine_n) |>
+  dplyr::inner_join(
+    colors_df,
+    by = c("celltype_fine" = "celltype")
+  ) |>
+  dplyr::select(celltype_fine_n, color) |>
+  unique()
+colors <- color_n_df$color
+names(colors) <- color_n_df$celltype_fine_n
+
+# Add in the other color with its label, since that won't be in the palette
+colors[updated_other] <- "#D2D2D2"
 ```

### 2025-05-09 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
make code a bit more concise

@@
       ewing_annotation
-    )
+    ),
+    # top 10 only - lump here to assist in color setup
+    celltype_fine = forcats::fct_lump(celltype_fine, 10)
   ) |>
   dplyr::add_count(celltype_fine) |>
-  dplyr::mutate(
-    celltype_fine_n = glue::glue("{celltype_fine} (n={n})"),
-    # top 10 only
-    celltype_fine_n = forcats::fct_lump(celltype_fine_n, 10)
-  )
-
-# add n() to "Other" for plot clarity and update factor to frequency order
-current_levels <- levels(query_only_df$celltype_fine_n)
-total_other <- query_only_df |>
-  dplyr::filter(celltype_fine_n == "Other") |>
-  nrow()
-updated_other <- glue::glue("Other (n={total_other})")
-
-query_only_df <- query_only_df |>
-  dplyr::mutate(
-    # update other level to have counts
-    celltype_fine_n = as.character(celltype_fine_n),
-    celltype_fine_n = ifelse(celltype_fine_n == "Other", updated_other, celltype_fine_n),
-    # order based on freq
-    celltype_fine_n = as.factor(celltype_fine_n),
-    celltype_fine_n = forcats::fct_infreq(celltype_fine_n),
-    celltype_fine_n = forcats::fct_relevel(celltype_fine_n, updated_other, after = Inf)
-  ) |>
+  dplyr::mutate(celltype_fine_n = glue::glue("{celltype_fine} (n={n})")) |>
   # exclude small clusters
@@
   dplyr::filter(subcluster_size >= 10)
+
+# Set cell type factor order by frequency, with the "Other (n=XX)" at the end
+other_level <- grep("Other", unique(query_only_df$celltype_fine_n), value = TRUE)
+query_only_df$celltype_fine_n <- forcats::fct_infreq(query_only_df$celltype_fine_n) |>
+  forcats::fct_relevel(other_level, after = Inf)
 
@@
 names(colors) <- color_n_df$celltype_fine_n
-
-# Add in the other color with its label, since that won't be in the palette
-colors[updated_other] <- "#D2D2D2"
 ```
@@
   scale_fill_manual(values = colors) +
+  labs(
+    fill = "Cell type (number of cells)",
+    x = "InferCNV subcluster"
+  ) +
   theme(axis.text.x = element_blank())

### 2025-05-12 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
prop_cnv_df fix

@@
 prop_cnv_df <- infercnv_df |>
+  # selecting here is helpful for debugging and viewing results
+  dplyr::select(
+    starts_with("proportion_scaled_"),
+    cell_id,
+    library_id,
+    barcodes,
+    subcluster
+  ) |>
   tidyr::pivot_longer(
@@
   tidyr::separate(cnv_type_raw, sep = "_", into = c("cnv_type", "chr")) |>
-  # remove extra chromosomes and cells not in this library
+  # remove extra chromosomes
   dplyr::filter(
     !stringr::str_starts(chr, "chrMT"),
-    !stringr::str_starts(chr, "chrGL"),
-    library_id == params$library_id
+    !stringr::str_starts(chr, "chrGL")
   ) |>
@@
     has_cnv_df |>
-      dplyr::select(barcodes, celltype)
+      dplyr::select(barcodes, celltype),
+    # we're joining a wide and long df, so we expect many matches
+    relationship = "many-to-many"
   )

### 2025-05-14 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
update text based on reference type

@@
 )
+if (stringr::str_detect(opts$reference_name, "internal")) {
+  reference_type_text <- glue::glue("within the `SCPCL000822` library")
+} else {
+  reference_type_text <- glue::glue("across all `SCPCP000015` libraries")
+}
+
 glue::glue(
   "`inferCNV` was run with the following normal reference: `{params$reference_name}`.
-  This reference contains all {ref_text[[params$reference_name]]}  cells across `SCPCP000015` libraries, excluding any which were also identified as tumor in the `cell-type-ewings` analysis module.
+  This reference contains all {ref_text[[params$reference_name]]} cells {reference_type_text}, excluding any which were also identified as tumor in the `cell-type-ewings` analysis module.
   "

### 2025-05-14 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
params not opts

@@
 )
-if (stringr::str_detect(opts$reference_name, "internal")) {
+if (stringr::str_detect(params$reference_name, "internal")) {
   reference_type_text <- glue::glue("within the `SCPCL000822` library")

### 2025-05-14 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
Apply suggestions from code review
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

--- 2025-05-14 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
@@
 )
-if (stringr::str_detect(opts$reference_name, "internal")) {
-  reference_type_text <- glue::glue("within the `SCPCL000822` library")
+if (stringr::str_detect(params$reference_name, "internal")) {
+  reference_type_text <- glue::glue("within the {params$library_id} library")
 } else {

### 2025-05-14 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
update param and text handling for _pooled/internal

--- 2025-05-14 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
@@
   sample_id: SCPCS000490
-  reference_name: ref_endo
+  reference_name: endo_pooled
 ---
@@
 ```{r, results = 'asis'}
+ref_celltype <- stringr::str_split(params$reference_name, "_")[1]
+ref_type <- stringr::str_split(params$reference_name, "_")[2]
+
 ref_text <- list(
-  "ref_endo" = "endothelial",
-  "ref_immune" = "immune",
-  "ref_endo-immune" = "endothelial and immune"
+  "endo" = "endothelial",
+  "immune" = "immune",
+  "endo-immune" = "endothelial and immune"
 )
-if (stringr::str_detect(opts$reference_name, "internal")) {
-  reference_type_text <- glue::glue("within the `SCPCL000822` library")
+if (ref_type == "internal") {
+  reference_type_text <- glue::glue("within the {params$library_id} library")
 } else {
@@
   "`inferCNV` was run with the following normal reference: `{params$reference_name}`.
-  This reference contains all {ref_text[[params$reference_name]]} cells {reference_type_text}, excluding any which were also identified as tumor in the `cell-type-ewings` analysis module.
+  This reference contains all {ref_celltype} cells {reference_type_text}, excluding any which were also identified as tumor in the `cell-type-ewings` analysis module.
   "

### 2025-05-14 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
Recreational bonus commit: Get rid of the chromosomes we dont need before infercnv to save some compute

@@
   tidyr::separate(cnv_type_raw, sep = "_", into = c("cnv_type", "chr")) |>
-  # remove extra chromosomes
-  dplyr::filter(
-    !stringr::str_starts(chr, "chrMT"),
-    !stringr::str_starts(chr, "chrGL")
-  ) |>
   # set chr factor levels, with ones used for validation first

### 2025-05-14 analyses/infercnv-consensus-cell-type/scripts/00-make-gene-order-file.R

@@ -54,7 +54,9 @@ gtf <- rtracklayer::import(local_gtf_file, feature.type = "gene")
 gtf_df <- gtf |>
   as.data.frame() |>
   dplyr::select(gene_id, seqnames, start, end) |>
-  dplyr::mutate(seqnames = glue::glue("chr{seqnames}"))
+  dplyr::mutate(seqnames = glue::glue("chr{seqnames}")) |>
+  # only keep chr1 - 22 and chrX and chrY
+  dplyr::filter(grepl("^chr([1-9]|1[0-9]|2[0-2]|X|Y)$", seqnames))
 
 # export gene order file
 readr::write_tsv(gtf_df, gene_order_file, col_names = FALSE)

### 2025-05-15 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
fix printing bug

@@
 ```{r, results = 'asis'}
-ref_celltype <- stringr::str_split(params$reference_name, "_")[1]
-ref_type <- stringr::str_split(params$reference_name, "_")[2]
+ref_celltype <- stringr::str_split_1(params$reference_name, "_")[1]
+ref_type <- stringr::str_split_1(params$reference_name, "_")[2]
 
@@
   "`inferCNV` was run with the following normal reference: `{params$reference_name}`.
-  This reference contains all {ref_celltype} cells {reference_type_text}, excluding any which were also identified as tumor in the `cell-type-ewings` analysis module.
+  This reference contains all {ref_text[[ref_celltype]]} cells {reference_type_text}, excluding any which were also identified as tumor in the `cell-type-ewings` analysis module.
   "

### 2025-05-28 analyses/infercnv-consensus-cell-type/exploratory-notebooks/02_ewings-reference-cnv.Rmd
add notebook

---
title: "Explore CNV distributions of reference cells: SCPCP000015"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

## Introduction

We have previously observed that the reference cell CNV distributions in pooled references may overlap with tumor CNV distributions.
This notebook aims to explore the extent of this trend for `SCPCP000015`.

## Setup

```{r}
options(readr.show_col_types = FALSE)
suppressPackageStartupMessages({
  library(ggplot2)
  library(SingleCellExperiment)
})

theme_set(theme_bw())
```

### Functions

```{r}
prepare_cnv_df <- function(infercnv_tsv) {
  infercnv_df <- readr::read_tsv(infercnv_tsv)

  infercnv_metadata_df <- infercnv_df |>
    dplyr::select(
      -starts_with("has_"),
      -starts_with("proportion_"),
      -starts_with("top_")
    ) |>
    dplyr::mutate(cell_group = stringr::str_split_i(subcluster, "_", 1))

  infercnv_df |>
    tidyr::pivot_longer(
      starts_with("has_cnv_"),
      names_to = "chr",
      values_to = "cnv"
    ) |>
    # sum all values for each
    dplyr::group_by(cell_id) |>
    dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
    dplyr::ungroup() |>
    # bring back metadata
    dplyr::inner_join(
      infercnv_metadata_df,
      by = "cell_id"
    )
}
```

### Paths

#### Base directories

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

module_dir <- file.path(repository_base, "analyses", "infercnv-consensus-cell-type")
result_dir <- file.path(module_dir, "results", "SCPCP000015") # organized as sample/reference/infercnv results
ewings_dir <- file.path(repository_base, "data", "current", "results", "cell-type-ewings", "SCPCP000015")
```

#### Input files

```{r}
# inferCNV output files
infercnv_files <- list.files(
  path = result_dir,
  pattern = "_cnv-metadata\\.tsv",
  recursive = TRUE,
  full.names = TRUE
) |>
  # remove any containing "endo_immune" or "internal"; we only want pooled endo or immune
  purrr::discard(\(x) {
    stringr::str_detect(x, "endo-immune|internal")
  }) |>
  # name by "library-reference"
  purrr::set_names(\(x) {
    library_id <- stringr::str_split_i(basename(x), "_", 1)
  })
#    ref_name <- stringr::str_remove(basename(dirname(x)), "_pooled")
#    glue::glue("{library_id}-{ref_name}")
# })

# ewing module cell types
celltype_files <- list.files(
  ewings_dir,
  pattern = "_ewing-celltype-assignments\\.tsv",
  recursive = TRUE,
  full.names = TRUE
) |>
  # name by library id
  purrr::set_names(\(x) {
    stringr::str_split_i(basename(x), "_", 1)
  })
```

### Read input files

We'll first read in the cell type annotations:
```{r}
celltype_df <- celltype_files |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind(names_to = "library_id") |>
  dplyr::mutate(cell_id = glue::glue("{library_id}-{barcodes}")) |>
  dplyr::select(cell_id, consensus_annotation, ewing_annotation)
```

Now, we'll read in the `inferCNV` results:
```{r}
infercnv_df <- infercnv_files |>
  purrr::map(prepare_cnv_df) |>
  purrr::list_rbind(names_to = "infercnv_library_id") |>
  dplyr::inner_join(celltype_df, by = "cell_id") |>
  # condense labels for cell type groupings
  dplyr::mutate(
    label = ifelse(
      stringr::str_detect(ewing_annotation, "tumor"),
      "tumor",
      cell_group
    ),
    label = forcats::fct_relevel(label, "unknown", "tumor", "reference")
  )
```

## Analysis

We'll first visualize the total CNV distributions for each inference for each grouping of cells:

- `unknown`: query cells
- `tumor`: query cells labeled "tumor" by `cell-type-ewings`
- `reference`: cells in the normal reference

```{r, fig.height = 8}
immune_df <- infercnv_df |>
  dplyr::filter(normal_reference == "immune_pooled")

endo_df <- infercnv_df |>
  dplyr::filter(normal_reference == "endo_pooled")

ggplot(immune_df) +
  aes(x = total_cnv_per_cell, y = label, fill = label) +
  ggridges::geom_density_ridges2(scale = 0.9) +
  facet_wrap(vars(infercnv_library_id)) +
  theme(legend.position = "none") +
  ggtitle("Immune pooled reference")

ggplot(endo_df) +
  aes(x = total_cnv_per_cell, y = label, fill = label) +
  ggridges::geom_density_ridges2(scale = 0.9) +
  facet_wrap(vars(infercnv_library_id)) +
  theme(legend.position = "none") +
  ggtitle("Endo pooled reference")
```

* The `endo` reference cell CNV distributions generally do not overlap with tumor cell CNV distributions.
* The `immune` reference cell CNV distributions do have a wider spread but do not _substantially_ overlap with the tumor cell CNV distributions.

Let's look more closely at the `immune` reference to see if any particular libraries might be driving a (minor) trend.
For each inference with the pooled `immune` reference, we'll find the 25\% top reference cells with the highest CNVs.
Do these cells tend to be from a particular library?

First, note how many reference cells come from each library:

```{r}
ref_immune_df <- immune_df |>
  dplyr::filter(cell_group == "reference") |>
  tidyr::separate(cell_id, into = c("reference_library_id", "barcode"), sep = "-", remove = FALSE)

ref_immune_df |>
  dplyr::select(cell_id) |>
  dplyr::distinct() |>
  tidyr::separate(cell_id, into = c("reference_library_id", "barcode"), sep = "-") |>
  dplyr::count(reference_library_id)
```

The table displayed below shows for each inference (`infercnv_library_id`), how many of top 25\% CNV cells in the pooled immune reference are from each library.
For example, of the reference cells with the highest CNVs in the `SCPCL000822` inference, 28 cells came from `SCPCL000822`, 30 cells came from `SCPCL000824`, etc.
Generally, these numbers are consistent across inferences although cells from `SCPCL000824` do seem to be more prevalent in certain samples.
Even so, there doesn't seem to be any major trend here.

```{r}
top25 <- ref_immune_df |>
  dplyr::group_by(infercnv_library_id) |>
  dplyr::slice_max(total_cnv_per_cell, prop = 0.25) |>
  dplyr::ungroup()
```

```{r}
top25 |>
  dplyr::count(infercnv_library_id, reference_library_id) |>
  tidyr::pivot_wider(
    names_from = reference_library_id,
    values_from = n,
    values_fill = 0
  )
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-05-28 analyses/infercnv-consensus-cell-type/exploratory-notebooks/02_ewings-reference-cnv.Rmd
some more text

@@
 ### Functions
+
+This function prepares a TSV of `inferCNV` results by creating a `total_cnv_per_cell` column to summarize CNVs per cell.
 
@@
 
+## Conclusions
+
+It does not seem that reference cell CNV distributions are problematically overlapping with tumor CNV distributions now that references were updated to generate these re-processed results.
+
 ## Session Info

### 2025-05-28 analyses/infercnv-consensus-cell-type/exploratory-notebooks/02_ewings-reference-cnv.Rmd
add endo-immune

@@
 
-This function prepares a TSV of `inferCNV` results by creating a `total_cnv_per_cell` column to summarize CNVs per cell.
+This function prepares a table of `inferCNV` results by creating a `total_cnv_per_cell` column to summarize CNVs per cell.
 
@@
 
+This function plots ridge plots of total CNV values across cell groupings for a given normal reference.
+
+```{r}
+plot_cnv_distributions <- function(df, reference_name) {
+  filtered_df <- infercnv_df |>
+    dplyr::filter(normal_reference == reference_name)
+
+  ggplot(filtered_df) +
+    aes(x = total_cnv_per_cell, y = label, fill = label) +
+    ggridges::geom_density_ridges2(scale = 0.9) +
+    facet_wrap(vars(infercnv_library_id)) +
+    theme(legend.position = "none") +
+    ggtitle(reference_name)
+}
+```
+
 ### Paths
@@
 ) |>
-  # remove any containing "endo_immune" or "internal"; we only want pooled endo or immune
+  # remove any containing "internal"; we only want pooled
   purrr::discard(\(x) {
-    stringr::str_detect(x, "endo-immune|internal")
+    stringr::str_detect(x, "internal")
   }) |>
@@
   })
-#    ref_name <- stringr::str_remove(basename(dirname(x)), "_pooled")
-#    glue::glue("{library_id}-{ref_name}")
-# })
 
@@
 ```{r, fig.height = 8}
-immune_df <- infercnv_df |>
-  dplyr::filter(normal_reference == "immune_pooled")
-
-endo_df <- infercnv_df |>
-  dplyr::filter(normal_reference == "endo_pooled")
-
-ggplot(immune_df) +
-  aes(x = total_cnv_per_cell, y = label, fill = label) +
-  ggridges::geom_density_ridges2(scale = 0.9) +
-  facet_wrap(vars(infercnv_library_id)) +
-  theme(legend.position = "none") +
-  ggtitle("Immune pooled reference")
-
-ggplot(endo_df) +
-  aes(x = total_cnv_per_cell, y = label, fill = label) +
-  ggridges::geom_density_ridges2(scale = 0.9) +
-  facet_wrap(vars(infercnv_library_id)) +
-  theme(legend.position = "none") +
-  ggtitle("Endo pooled reference")
+plot_cnv_distributions(infercnv_df, "endo_pooled")
+plot_cnv_distributions(infercnv_df, "immune_pooled")
+plot_cnv_distributions(infercnv_df, "endo-immune_pooled")
 ```
@@
 * The `endo` reference cell CNV distributions generally do not overlap with tumor cell CNV distributions.
-* The `immune` reference cell CNV distributions do have a wider spread but do not _substantially_ overlap with the tumor cell CNV distributions.
+* The `endo-immune` and `immune` reference cell CNV distributions do have a wider spread but do not _substantially_ overlap with the tumor cell CNV distributions.
 
@@
 ```{r}
-ref_immune_df <- immune_df |>
-  dplyr::filter(cell_group == "reference") |>
+ref_immune_df <- infercnv_df |>
+  dplyr::filter(
+    normal_reference == "immune_pooled",
+    cell_group == "reference"
+  ) |>
   tidyr::separate(cell_id, into = c("reference_library_id", "barcode"), sep = "-", remove = FALSE)

### 2025-05-29 analyses/infercnv-consensus-cell-type/exploratory-notebooks/02_ewings-reference-cnv.Rmd
Updated notebook to include barplot of library proportions, and add some additional text

@@
   We have previously observed that the reference cell CNV distributions in pooled references may overlap with tumor CNV distributions.
-This notebook aims to explore the extent of this trend for `SCPCP000015`.
+This notebook aims to explore the extent of this trend for `SCPCP000015`, as well as compare CNV distributions among cell groups for all libraries.
 
@@
 
-```{r, fig.height = 8}
+We show a plot faceted by library `inferCNV` was run on for each of the three pooled references explored.
+
+```{r, fig.width = 10, fig.height = 8, message = FALSE}
 plot_cnv_distributions(infercnv_df, "endo_pooled")
@@
 
-Let's look more closely at the `immune` reference to see if any particular libraries might be driving a (minor) trend.
-For each inference with the pooled `immune` reference, we'll find the 25\% top reference cells with the highest CNVs.
+Let's look more closely at the `immune_polled` reference to see if any particular libraries might be driving a (minor) trend.
+For each inference with the `immune_pooled` reference, we'll find the 25\% top reference cells with the highest CNVs.
 Do these cells tend to be from a particular library?
 
-First, note how many reference cells come from each library:
+The barplot below shows for each `inferCNV` inference with the `immune_pooled` reference, the proportion of reference cells contributed by each of the five given libraries.
 
-```{r}
-ref_immune_df <- infercnv_df |>
+```{r, fig.width = 8}
+top25_df <- infercnv_df |>
   dplyr::filter(
@@
   ) |>
-  tidyr::separate(cell_id, into = c("reference_library_id", "barcode"), sep = "-", remove = FALSE)
-
-ref_immune_df |>
-  dplyr::select(cell_id) |>
-  dplyr::distinct() |>
   tidyr::separate(cell_id, into = c("reference_library_id", "barcode"), sep = "-") |>
-  dplyr::count(reference_library_id)
-```
-
-The table displayed below shows for each inference (`infercnv_library_id`), how many of top 25\% CNV cells in the pooled immune reference are from each library.
-For example, of the reference cells with the highest CNVs in the `SCPCL000822` inference, 28 cells came from `SCPCL000822`, 30 cells came from `SCPCL000824`, etc.
-Generally, these numbers are consistent across inferences although cells from `SCPCL000824` do seem to be more prevalent in certain samples.
-Even so, there doesn't seem to be any major trend here.
-
-```{r}
-top25 <- ref_immune_df |>
+  # keep only the top 25% of cells
   dplyr::group_by(infercnv_library_id) |>
@@
   dplyr::ungroup()
+
+ggplot(top25_df) +
+  aes(x = infercnv_library_id, fill = reference_library_id) +
+  geom_bar(position = "fill") +
+  labs(
+    x = "Library that inferCNV was run on",
+    y = "Proportion of cells in the top 25% CNV\nfrom the given reference library"
+  ) +
+  theme(axis.text.x = element_text(angle = 30, hjust = 1))
 ```
 
-```{r}
-top25 |>
-  dplyr::count(infercnv_library_id, reference_library_id) |>
-  tidyr::pivot_wider(
-    names_from = reference_library_id,
-    values_from = n,
-    values_fill = 0
-  )
-```
+We see a generally consistent trend across inferences; cells in the top 25\% are similarly distributed across libraries that contributed to the reference. 
+No particular library seems to be biasing higher CNV values in the reference cells.
 
@@
 It does not seem that reference cell CNV distributions are problematically overlapping with tumor CNV distributions now that references were updated to generate these re-processed results.
+Overall, the `endo_pooled` reference appears to give the most distinct CNV distributions between reference and tumor cells.
 

### 2025-06-02 analyses/infercnv-consensus-cell-type/exploratory-notebooks/03_ewings-pooled-internal.Rmd
exploratory notebook for pooled vs internal

---
title: "Compare internal vs pooled references: SCPCP000015"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

## Introduction

For several samples in `SCPCP000015`, we were able to run `inferCNV` with a so-called "internal" normal reference, where normal cells were taken only from the library being analyzed and were not pooled across all libraries.
The goal of this notebook is to directly compared results from internal references with their pooled counterparts to assess if there is evidence that the pooled reference strategy improves tumor/normal classification.

## Setup

```{r, warning = FALSE}
options(readr.show_col_types = FALSE)
suppressPackageStartupMessages({
  library(ggplot2)
  library(SingleCellExperiment)
})

theme_set(theme_bw())
```

### Functions

This function prepares a table of `inferCNV` results by creating a `total_cnv_per_cell` column to summarize CNVs per cell.

```{r}
prepare_cnv_df <- function(infercnv_tsv) {
  infercnv_df <- readr::read_tsv(infercnv_tsv)

  infercnv_metadata_df <- infercnv_df |>
    dplyr::select(
      -starts_with("has_"),
      -starts_with("proportion_"),
      -starts_with("top_")
    ) |>
    dplyr::mutate(cell_group = stringr::str_split_i(subcluster, "_", 1))

  infercnv_df |>
    tidyr::pivot_longer(
      starts_with("has_cnv_"),
      names_to = "chr",
      values_to = "cnv"
    ) |>
    # sum all values for each
    dplyr::group_by(cell_id) |>
    dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
    dplyr::ungroup() |>
    # bring back metadata
    dplyr::inner_join(
      infercnv_metadata_df,
      by = "cell_id"
    )
}
```

This function plots ridge plots of total CNV values across cell groupings, faceted by reference type (pooled vs. internal) and library id, for a given reference cell grouping.

```{r}
plot_cnv_distributions <- function(reference_name, df) {
  df <- infercnv_df |>
    dplyr::filter(
      reference_cell_group == reference_name
    )
  ggplot(df) +
    aes(x = total_cnv_per_cell, y = label, fill = label) +
    ggridges::geom_density_ridges2(scale = 0.9) +
    facet_grid(
      cols = vars(library_id),
      rows = vars(reference_type)
    ) +
    theme(legend.position = "none") +
    ggtitle(
      glue::glue("Comparing {reference_name} internal vs. pooled")
    )
}
```

### Paths

#### Base directories

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

module_dir <- file.path(repository_base, "analyses", "infercnv-consensus-cell-type")
result_dir <- file.path(module_dir, "results", "SCPCP000015") # organized as sample/reference/infercnv results
ewings_dir <- file.path(repository_base, "data", "current", "results", "cell-type-ewings", "SCPCP000015")
```

#### Input files

```{r}
# inferCNV output files
infercnv_files <- list.files(
  path = result_dir,
  pattern = "_cnv-metadata\\.tsv",
  recursive = TRUE,
  full.names = TRUE
) |>
  # we only want internal here.
  # based on these internals, we'll also grab corresponding pooled
  purrr::keep(\(x) {
    stringr::str_detect(x, "internal")
  })

infercnv_files <- c(
  infercnv_files,
  stringr::str_replace(infercnv_files, "internal", "pooled")
) |>
  # name by library id
  purrr::set_names(\(x) {
    library_id <- stringr::str_split_i(basename(x), "_", 1)
  })

# ewing module cell types
celltype_files <- list.files(
  ewings_dir,
  pattern = "_ewing-celltype-assignments\\.tsv",
  recursive = TRUE,
  full.names = TRUE
) |>
  # name by library id
  purrr::set_names(\(x) {
    stringr::str_split_i(basename(x), "_", 1)
  })
```

### Read input files

We'll first read in the cell type annotations:
```{r}
celltype_df <- celltype_files |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind(names_to = "library_id") |>
  dplyr::mutate(cell_id = glue::glue("{library_id}-{barcodes}")) |>
  dplyr::select(cell_id, consensus_annotation, ewing_annotation)
```

Next we'll read in and prepare the `inferCNV` results:

```{r}
infercnv_df <- infercnv_files |>
  purrr::map(prepare_cnv_df) |>
  purrr::list_rbind(names_to = "library_id") |>
  dplyr::inner_join(celltype_df, by = "cell_id") |>
  # split normal_reference into "reference_cell_group" and "reference_type"
  tidyr::separate(
    normal_reference,
    into = c("reference_cell_group", "reference_type"),
    sep = "_",
    remove = FALSE # keep the normal_reference column around
  ) |>
  # condense labels for cell type groupings
  dplyr::mutate(
    label = ifelse(
      stringr::str_detect(ewing_annotation, "tumor"),
      "tumor",
      cell_group
    ),
    label = forcats::fct_relevel(label, "unknown", "tumor", "reference")
  )
```

## Visualization

We'll plot the total CNV distributions for each inference for each grouping of cells:

- `unknown`: query cells
- `tumor`: query cells labeled "tumor" by `cell-type-ewings`
- `reference`: cells in the normal reference

We'll plot this separately for each reference cell group (`endo`, `immune`, and `endo-immune`) for visual ease to make comparisons between `internal` and `pooled` reference versions.
Note that `SCPCL001112` was only run with an `endo` internal reference since it didn't have sufficient immune cells; this library will therefore only appear in the plots for the `endo` reference.

```{r message = FALSE, warning = FALSE, fig.height = 6}
unique(infercnv_df$reference_cell_group) |>
  purrr::map(plot_cnv_distributions, infercnv_df)
```

## Conclusions

* For all three libraries we were able to directly compare internal and pooled references for, the pooled reference appears to do either a similar or slightly a better job separating normal and tumor cells compared to the internal reference.
* There is no obvious inference where using a pooled reference appears _worse_ than using a pooled reference.
* Similar to previous observations we've made, the `endo` reference (either pooled or internal) shows improved separation between tumor and normal compared to references containing `immune` cells.
* Notably, `SCPCL001112` is a much smaller library than `SCPCL000822` and `SCPCL000824`, and we see evidence here that using a pooled reference dramatically improves separation between tumor and normal.

Overall, these comparisons suggest that using a pooled reference is a reasonable approach, in particular when information is limited within a single library.

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-06-02 analyses/infercnv-consensus-cell-type/exploratory-notebooks/02_ewings-reference-cnv.Rmd
fixed a couple formatting spots

@@
 ```{r}
-plot_cnv_distributions <- function(df, reference_name) {
-  filtered_df <- infercnv_df |>
+plot_cnv_distributions <- function(reference_name, df) {
+  filtered_df <- df |>
     dplyr::filter(normal_reference == reference_name)
@@
 ```{r, fig.width = 10, fig.height = 8, message = FALSE}
-plot_cnv_distributions(infercnv_df, "endo_pooled")
-plot_cnv_distributions(infercnv_df, "immune_pooled")
-plot_cnv_distributions(infercnv_df, "endo-immune_pooled")
+unique(infercnv_df$normal_reference) |>
+  purrr::map(plot_cnv_distributions, infercnv_df)
 ```

### 2025-06-03 analyses/infercnv-consensus-cell-type/exploratory-notebooks/03_ewings-pooled-internal.Rmd
Apply suggestions from code review
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

@@
 For several samples in `SCPCP000015`, we were able to run `inferCNV` with a so-called "internal" normal reference, where normal cells were taken only from the library being analyzed and were not pooled across all libraries.
-The goal of this notebook is to directly compared results from internal references with their pooled counterparts to assess if there is evidence that the pooled reference strategy improves tumor/normal classification.
+The goal of this notebook is to directly compare results from internal references with their pooled counterparts to assess if there is evidence that the pooled reference strategy improves tumor/normal classification.
 
@@
 * For all three libraries we were able to directly compare internal and pooled references for, the pooled reference appears to do either a similar or slightly a better job separating normal and tumor cells compared to the internal reference.
-* There is no obvious inference where using a pooled reference appears _worse_ than using a pooled reference.
+* There is no obvious inference where using a pooled reference appears _worse_ than using an internal reference.
 * Similar to previous observations we've made, the `endo` reference (either pooled or internal) shows improved separation between tumor and normal compared to references containing `immune` cells.

### 2025-06-03 analyses/infercnv-consensus-cell-type/exploratory-notebooks/03_ewings-pooled-internal.Rmd
fig size

@@
 
-```{r message = FALSE, warning = FALSE, fig.height = 6}
+```{r message = FALSE, warning = FALSE, fig.width = 8}
 unique(infercnv_df$reference_cell_group) |>

### 2025-06-04 analyses/infercnv-consensus-cell-type/exploratory-notebooks/SCPCP000004/01_nb-consensus-cell-types.Rmd
add exploratory notebook for NB normal cells to determine reference options

---
title: "Explore normal cell types in Neuroblastoma samples"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

## Introduction

The goal of this notebook is to explore the distribution of normal consensus cell types, specifically immune and endothelial, in the Neuroblastoma tumor (i.e., _not_ the PDX) samples from `SCPCP000004`.

Based on these findings, we can determine approaches to evaluate for pooling  cells to use as a normal reference for `inferCNV`.
We can also determine whether any samples can be run with an internal (non-pooled) reference as a control.

## Setup

```{r}
options(readr.show_col_types = FALSE)
library(ggplot2)
theme_set(theme_bw())
```

### Paths

#### Base directories

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")

# The current results directory, found within the repository base directory
result_dir <- file.path(repository_base, "data", "current", "results")

# Path to cell-type-consensus results
consensus_dir <- file.path(result_dir, "cell-type-consensus", "SCPCP000004")
```

#### Input and output files

```{r paths}
# Input consensus cell type files
celltype_files <- list.files(
  path = consensus_dir,
  pattern = "_processed_consensus-cell-types\\.tsv\\.gz$",
  recursive = TRUE,
  full.names = TRUE
) |>
  # set names as library id
  purrr::set_names(
    \(x) {
      stringr::str_split_i(basename(x), "_", 1)
    }
  )

# project sample metadata file
metadata_file <- file.path(data_dir, "single_cell_metadata.tsv")

# Path to all possible immune consensus cell types
immune_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-immune-cell-types.tsv"
```

We'll also define vectors for the consensus cell types to check for.

```{r}
immune_celltypes <- readr::read_tsv(immune_url) |>
  dplyr::pull(consensus_annotation)

# Vector of all possible endothelial cell types
endo_celltypes <- c(
  "endothelial cell",
  "blood vessel endothelial cell",
  "microvascular endothelial cell",
  "pericyte"
)
```

## Analysis

```{r}
# determine PDX libraries to exclude
tumor_library_ids <- readr::read_tsv(metadata_file) |>
  dplyr::filter(!is_xenograft) |>
  dplyr::pull(scpca_library_id)

# Prepare data frame of all cell types
celltype_df <- celltype_files |>
  # consider only non-PDX
  purrr::keep_at(tumor_library_ids) |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind() |>
  dplyr::select(library_id, barcodes, consensus_annotation) |>
  dplyr::mutate(
    category = dplyr::case_when(
      consensus_annotation %in% immune_celltypes ~ "immune",
      consensus_annotation %in% endo_celltypes ~ "endo",
      .default = "other"
    )
  )
```

The table below shows counts of endothelial, immune, and all other cells across libraries:

```{r}
category_count_df <- celltype_df |>
  dplyr::count(library_id, category)

# format for viewing
category_count_df |>
  tidyr::pivot_wider(
    names_from = category,
    values_from = n,
    values_fill = 0
  ) |>
  # order columns
  dplyr::select(library_id, endo, immune, other) |>
  # this way we can print the full table at once; RMarkdown will only show 10 rows at once
  knitr::kable()
```

Looks like we've got a lot of normal cells here, and plenty of libraries we can run with an internal reference. 
Since we have a lot of immune and endothelial cells on their own, we don't need to consider a combined reference for this project.

There are also some libraries with a very small number of cells which are presumably lower-quality and should not be included in any reference.

First, we'll find any libraries that have fewer than 1000 cells and exclude them.

```{r}
discard_libraries <- category_count_df |>
  dplyr::group_by(library_id) |>
  dplyr::summarize(total_cells = sum(n)) |>
  dplyr::filter(total_cells < 1000) |>
  dplyr::pull(library_id)

discard_libraries
```

How many cells would be in each reference for `endo` and `immune` groups?

```{r}
category_count_df <- category_count_df |>
  dplyr::filter(
    category != "other", # we dont need these counts anymore
    !(library_id %in% discard_libraries)
  )

total_ref_cells_df <- category_count_df |>
  dplyr::group_by(category) |>
  dplyr::summarize(total_cells = sum(n))
total_ref_cells_df
```
Well, that's quite a lot of cells!
Let's visualize which libraries these cells are spread across.
The barplot below shows the proportion of cells in the given category from each library.

```{r}
category_prop_df <- category_count_df |>
  dplyr::inner_join(total_ref_cells_df) |>
  dplyr::mutate(prop_total_ref_cells = n / total_cells)

ggplot(category_prop_df) +
  aes(x = library_id, y = prop_total_ref_cells, fill = category) +
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))
```
The plurality of endothelial cells are from `SCPCL000125`, and the plurality of immune cells are from both `SCPCL000126` and `SCPCL000127`.
These libraries dominate their respective reference cell categories, but cells are still reasonably spread across the other libraries.

Next, let's look more closely at the counts for cell types within these broad categories of `immune` and `endo`, since we may have enough data here to narrow the reference down further:

```{r}
fine_celltype_df <- celltype_df |>
  dplyr::filter(category != "other") |>
  dplyr::count(consensus_annotation, category) |>
  dplyr::arrange(category)

# kable to print the full table at once
knitr::kable(fine_celltype_df)
```

* Endothelial cells are nearly exclusively plain `endothelial cell`
* The immune cells are pretty spread out, and we could have cell-type specific references created from anything with more than 100 cells:

```{r}
fine_celltype_df |>
  dplyr::filter(category == "immune", n > 100) |>
  dplyr::arrange(desc(n))
```
Dominant cell types are T cells (across several categories of T cell) and macrophages.

Finally, let's see which libraries could have an internal reference.
Generally `inferCNV` recommends at least 100 cells in a reference, but since there are so many normal cells here we can increase this threshold since 100 is really a lower bound.
So, for this project, we'll use 250 as a threshold to find which libraries/internal references are possible:

```{r}
category_count_df |>
  dplyr::filter(n >= 250) |>
  dplyr::group_by(library_id) |>
  dplyr::mutate(reference_options = paste(category, collapse = ", ")) |>
  dplyr::select(library_id, reference_options) |>
  unique() |>
  # kable to print the full table at once
  knitr::kable()
```

Notably, we have previously observed with our analysis of `SCPCP000015` in this module that references with immune cells tend to show more total CNV overlap with tumor cells, since immune cells are funky! 
So, it's nice to see that we enough data to be able to use a robust `endo` reference without needing to rely on `immune`, should similar CNV patterns emerge here.

## Conclusions

* We can _definitely_ have `endo` and `immune` pooled references, and the references won't be _strongly_ biased towards coming from a specific library
* Since we have such a high number of cells in each reference cell grouping in the first place, a combined reference (e.g. `endo-immune`) does not seem necessary
* We could also consider a subset of immune cell types for a reference, particularly macrophage and T cell, but a full immune reference seems reasonable to start with
* We have 11 libraries which we could use an internal reference with: 3 of them with `endo`, and _all_ of them with `immune`

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-06-06 analyses/infercnv-consensus-cell-type/exploratory-notebooks/SCPCP000004/01_nb-consensus-cell-types.Rmd

@@
 consensus_dir <- file.path(result_dir, "cell-type-consensus", "SCPCP000004")
+
+# Path to reference cell type groupings
+reference_dir <- file.path(repository_base, "analyses", "infercnv-consensus-cell-type", "references")
 ```
@@
 
-# Path to all possible immune consensus cell types
-immune_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-immune-cell-types.tsv"
-```
-
-We'll also define vectors for the consensus cell types to check for.
-
-```{r}
-immune_celltypes <- readr::read_tsv(immune_url) |>
-  dplyr::pull(consensus_annotation)
-
-# Vector of all possible endothelial cell types
-endo_celltypes <- c(
-  "endothelial cell",
-  "blood vessel endothelial cell",
-  "microvascular endothelial cell",
-  "pericyte"
-)
+# reference cell type groups
+reference_file <- file.path(reference_dir, "reference-cell-groups.tsv")
 ```
@@
 
+# get cell groups
+cell_group_df <- readr::read_tsv(reference_file)
+
 # Prepare data frame of all cell types
@@
   dplyr::select(library_id, barcodes, consensus_annotation) |>
-  dplyr::mutate(
-    category = dplyr::case_when(
-      consensus_annotation %in% immune_celltypes ~ "immune",
-      consensus_annotation %in% endo_celltypes ~ "endo",
-      .default = "other"
-    )
-  )
+  dplyr::left_join(cell_group_df, by = "consensus_annotation")
 ```
 
-The table below shows counts of endothelial, immune, and all other cells across libraries:
+We'll first exclude any libraries that have fewer than 1000 cells from consideration in the reference.
 
 ```{r}
-category_count_df <- celltype_df |>
-  dplyr::count(library_id, category)
-
-# format for viewing
-category_count_df |>
-  tidyr::pivot_wider(
-    names_from = category,
-    values_from = n,
-    values_fill = 0
-  ) |>
-  # order columns
-  dplyr::select(library_id, endo, immune, other) |>
-  # this way we can print the full table at once; RMarkdown will only show 10 rows at once
-  knitr::kable()
-```
-
-Looks like we've got a lot of normal cells here, and plenty of libraries we can run with an internal reference. 
-Since we have a lot of immune and endothelial cells on their own, we don't need to consider a combined reference for this project.
-
-There are also some libraries with a very small number of cells which are presumably lower-quality and should not be included in any reference.
-
-First, we'll find any libraries that have fewer than 1000 cells and exclude them.
-
-```{r}
-discard_libraries <- category_count_df |>
-  dplyr::group_by(library_id) |>
-  dplyr::summarize(total_cells = sum(n)) |>
-  dplyr::filter(total_cells < 1000) |>
+discard_libraries <- celltype_df |>
+  dplyr::count(library_id) |>
+  dplyr::filter(n < 1000) |>
   dplyr::pull(library_id)
@@
 
-How many cells would be in each reference for `endo` and `immune` groups?
+```{r}
+celltype_df <- celltype_df |>
+  dplyr::filter(!(library_id %in% discard_libraries))
+```
+
+These cell groupings are present in the neuroblastoma samples we're considering:
 
 ```{r}
-category_count_df <- category_count_df |>
+unique(celltype_df$reference_cell_group)
+```
+
+We should remove any cell groups which might correspond to tumor cells, including `stromal`, `neuronal`, and `stem cell`.
+The `NA` values here correspond to `Unknown` cells, which we'll also remove.
+This means we'll keep groups `endothelial`, `epithelial`, and `immune`.
+
+```{r}
+exclude_groups <- c("stromal", "neuronal", "stem cell")
+celltype_df <- celltype_df |>
   dplyr::filter(
-    category != "other", # we dont need these counts anymore
-    !(library_id %in% discard_libraries)
+    !(reference_cell_group %in% exclude_groups),
+    consensus_annotation != "Unknown"
+  )
+```
+
+Let's begin by counting how many cells in each cell group we have per library
+
+```{r}
+celltype_count_df <- celltype_df |>
+  dplyr::count(library_id, reference_cell_group)
+
+# format for viewing
+celltype_count_wide_df <- celltype_count_df |>
+  tidyr::pivot_wider(
+    names_from = reference_cell_group,
+    values_from = n,
+    values_fill = 0
+  ) |>
+  # add column for total normal cells
+  dplyr::mutate(
+    all_normal = rowSums(dplyr::across(where(is.numeric)))
+  ) |>
+  # arrange by total normal cells
+  dplyr::arrange(desc(all_normal))
+
+# this way we can print the full table at once; RMarkdown will only show 10 rows at once
+knitr::kable(celltype_count_wide_df)
+```
+
+Virtually all of the normal cell types are `endothelial` and `immune`, with very few `adipocyte` and literally only one `epithelial` cell.
+
+We'll look at this data in barplot form to see, per-library, the relative proportions of each normal cell type group out of all normal cells in the library.
+
+```{r}
+celltype_prop_df <- celltype_count_wide_df |>
+  # long again, but keep all_normal column for calculating proportion
+  tidyr::pivot_longer(
+    -c(library_id, all_normal),
+    names_to = "reference_cell_group",
+    values_to = "n"
+  ) |>
+  dplyr::mutate(cell_group_prop = n / all_normal)
+
+ggplot(celltype_prop_df) +
+  aes(x = library_id, y = cell_group_prop, fill = reference_cell_group) +
+  geom_col() +
+  labs(y = "Proportion of normal cells in group") +
+  theme(
+    legend.position = "bottom",
+    axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
+  )
+```
+
+Based on these numbers, we can consider three normal reference options:
+
+* All normal cells
+* ALl immune cells
+* All endothelial cells
+
+How many cells would be in each of those pooled references?
+
+```{r}
+# update the count data frame without categories we don't need
+celltype_count_df <- celltype_count_wide_df |>
+  dplyr::select(-adipocyte, -epithelial) |>
+  tidyr::pivot_longer(
+    -library_id,
+    names_to = "reference_cell_group",
+    values_to = "n"
   )
 
-total_ref_cells_df <- category_count_df |>
-  dplyr::group_by(category) |>
+celltype_count_df |>
+  dplyr::group_by(reference_cell_group) |>
   dplyr::summarize(total_cells = sum(n))
-total_ref_cells_df
 ```
-Well, that's quite a lot of cells!
-Let's visualize which libraries these cells are spread across.
-The barplot below shows the proportion of cells in the given category from each library.
+
+Well, that's quite a lot of cells for these pooled references!
+A potential concern is that this many cells will incur prohibitive runtimes, so we may want to explore a subsampling strategy.
+Any subsampling strategy should aim to retain the relative cell type proportions across libraries, but we can explore that if runtime becomes a problem when we reach that stage.
+
+Finally, let's see which libraries could have an internal reference.
+`inferCNV` recommends at least 100 cells in a reference, so which combinations of library/reference are possible here?
 
 ```{r}
-category_prop_df <- category_count_df |>
-  dplyr::inner_join(total_ref_cells_df) |>
-  dplyr::mutate(prop_total_ref_cells = n / total_cells)
-
-ggplot(category_prop_df) +
-  aes(x = library_id, y = prop_total_ref_cells, fill = category) +
-  geom_col(position = "dodge") +
-  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))
-```
-The plurality of endothelial cells are from `SCPCL000125`, and the plurality of immune cells are from both `SCPCL000126` and `SCPCL000127`.
-These libraries dominate their respective reference cell categories, but cells are still reasonably spread across the other libraries.
-
-Next, let's look more closely at the counts for cell types within these broad categories of `immune` and `endo`, since we may have enough data here to narrow the reference down further:
-
-```{r}
-fine_celltype_df <- celltype_df |>
-  dplyr::filter(category != "other") |>
-  dplyr::count(consensus_annotation, category) |>
-  dplyr::arrange(category)
-
-# kable to print the full table at once
-knitr::kable(fine_celltype_df)
-```
-
-* Endothelial cells are nearly exclusively plain `endothelial cell`
-* The immune cells are pretty spread out, and we could have cell-type specific references created from anything with more than 100 cells:
-
-```{r}
-fine_celltype_df |>
-  dplyr::filter(category == "immune", n > 100) |>
-  dplyr::arrange(desc(n))
-```
-Dominant cell types are T cells (across several categories of T cell) and macrophages.
-
-Finally, let's see which libraries could have an internal reference.
-Generally `inferCNV` recommends at least 100 cells in a reference, but since there are so many normal cells here we can increase this threshold since 100 is really a lower bound.
-So, for this project, we'll use 250 as a threshold to find which libraries/internal references are possible:
-
-```{r}
-category_count_df |>
-  dplyr::filter(n >= 250) |>
+celltype_count_df |>
+  dplyr::filter(n >= 100) |>
   dplyr::group_by(library_id) |>
-  dplyr::mutate(reference_options = paste(category, collapse = ", ")) |>
+  dplyr::mutate(reference_options = paste(reference_cell_group, collapse = ", ")) |>
+  dplyr::ungroup() |>
   dplyr::select(library_id, reference_options) |>
   unique() |>
+  # add row counter so we can easily see how many libraries
+  dplyr::mutate(row_num = dplyr::row_number()) |>
   # kable to print the full table at once
@@
 
-Notably, we have previously observed with our analysis of `SCPCP000015` in this module that references with immune cells tend to show more total CNV overlap with tumor cells, since immune cells are funky!
-So, it's nice to see that we enough data to be able to use a robust `endo` reference without needing to rely on `immune`, should similar CNV patterns emerge here.
-
 ## Conclusions
 
-* We can _definitely_ have `endo` and `immune` pooled references, and the references won't be _strongly_ biased towards coming from a specific library
-* Since we have such a high number of cells in each reference cell grouping in the first place, a combined reference (e.g. `endo-immune`) does not seem necessary
-* We could also consider a subset of immune cell types for a reference, particularly macrophage and T cell, but a full immune reference seems reasonable to start with
-* We have 11 libraries which we could use an internal reference with: 3 of them with `endo`, and _all_ of them with `immune`
+* To start, we can consider three references: all normal, all immune, and possibly also all endothelial
+  * We may need to consider a subsetting strategy for these references if runtimes are too long since there would be >10,000 cells in the immune-containing pooled normal references
+* There are 19 libraries we can use internal references with, all with the option for all normal cells and some with the immune-only or endothelial-only options too
 

### 2025-06-09 analyses/infercnv-consensus-cell-type/exploratory-notebooks/SCPCP000004/01_nb-consensus-cell-types.Rmd
Apply suggestions from code review
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

@@
 The `NA` values here correspond to `Unknown` cells, which we'll also remove.
-This means we'll keep groups `endothelial`, `epithelial`, and `immune`.
+This means we'll keep groups `endothelial`, `epithelial`, `adipocyte`, and `immune`.
 
 ```{r}
+# First, we want to count the total number of cells to use for later plotting
+total_cells_df <- celltype_df |>
+  dplyr::count(library_id, name = "total_cells")
+
+# Now we can remove unwanted cells
 exclude_groups <- c("stromal", "neuronal", "stem cell")
@@
 
-Let's begin by counting how many cells in each cell group we have per library
+Let's begin by counting how many cells in each cell group we have per library.
 
@@
 
-We'll look at this data in barplot form to see, per-library, the relative proportions of each normal cell type group out of all normal cells in the library.
+We'll look at this data in barplot form to see, per-library, the relative proportions of each normal cell type group out of _all cells_ in the library.
+The height of each bar is the total fraction of normal cells in that library to begin with; note the Y-axis only ranges from around `[0,0.3]`.
 
@@
 celltype_prop_df <- celltype_count_wide_df |>
+  dplyr::select(-all_normal) |>
   # long again, but keep all_normal column for calculating proportion
   tidyr::pivot_longer(
-    -c(library_id, all_normal),
+    -library_id,
     names_to = "reference_cell_group",
@@
   ) |>
-  dplyr::mutate(cell_group_prop = n / all_normal)
+  dplyr::left_join(total_cells_df) |>
+  dplyr::mutate(cell_group_prop = n / total_cells)
 
@@
   geom_col() +
-  labs(y = "Proportion of normal cells in group") +
+  labs(y = "Proportion of cells in normal group") +
   theme(
@@
 * All normal cells
-* ALl immune cells
+* All immune cells
 * All endothelial cells
@@
 ```{r}
-# update the count data frame without categories we don't need
+# update the count data frame without categories we don't need.
+# note that this does not remove adipocyte or epithelial counts from the `all_normal` group; it just
+# removes them as standalone categories
 celltype_count_df <- celltype_count_wide_df |>

### 2025-06-11 analyses/cell-type-neuroblastoma-04/notebook-template.Rmd
initiate template with a barebones readme

--- 2024-11-19 analyses/cell-type-consensus/notebook-template.Rmd
@@
 ---
-title: "cell-type-consensus Analysis"
+title: "cell-type-neuroblastoma-04 Analysis"
 author:
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "cell-type-consensus")
+module_base <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
 ```

### 2025-07-02 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/01_reference-aggregation.Rmd
Add exploratory Rmd notebook and utils functions for jaccard

---
title: "`r glue::glue('Compare aggregated and non-aggregated reference: {params$library_id}')`"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  sample_id: "SCPCS000105"
  library_id: "SCPCL000122"
---

This notebook does some preliminary comparisons of `SingleR` results obtained with the NBAtlas reference, where the reference was either aggregated before training or not.
The goal is to determine whether there are substantial differences between aggregation versions.
Note that the goal of this notebook is _not_ specifically to validate annotations, but to compare whether the aggregated vs non-aggregated reference yield different or similar labels.
There is, however, a brief comparison to consensus annotations.

## Setup

```{r, warning = FALSE}
options(readr.show_col_types = FALSE)
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(SingleCellExperiment)
})

theme_set(theme_bw())

# Define color ramp for shared use in the heatmaps
heatmap_col_fun <- circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue"))
# Set heatmap padding option
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(0.6, "in"))
```

### Paths

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
results_dir <- file.path(module_dir, "results", "singler")
data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")

validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
```

### Functions

```{r}
# Source Jaccard and heatmap utilities functions:
source(file.path(module_dir, "exploratory-notebooks", "jaccard-utils.R"))

# This function calculates and returns a confusion matrix
#
# df: data frame with columns of interest
# annotation_col1: first column to compare; arg should be a string
# annotation_col2: second column to compare; arg should be a string
calculate_confusion <- function(
    df,
    annotation_col1,
    annotation_col2) {
  # determine all possible annotation values for caret
  all_levels <- union(
    unique(df[[annotation_col1]]),
    unique(df[[annotation_col2]])
  )

  caret::confusionMatrix(
    factor(df[[annotation_col1]], levels = all_levels),
    factor(df[[annotation_col2]], levels = all_levels),
    mode = "sens_spec"
  )
}

# This function calculates Jaccard similarity between two vectors and plots a heatmap
# Code is based on this scpca-nf code:
# https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_supplemental_report.rmd#L50
#
# df: data frame with columns of interest
# annotation_col1: first column to compare; arg should be a string
# annotation_col2: second column to compare; arg should be a string
# label1: label to use in heatmap for annotation_col1 axis
# label2: label to use in heatmap for annotation_col2 axis
make_jaccard_heatmap <- function(
    df,
    annotation_col1,
    annotation_col2,
    label1,
    label2) {
  # Calculate jaccard matrix
  jaccard_matrix <- make_jaccard_matrix(
    df,
    annotation_col1,
    annotation_col2
  )

  create_single_heatmap(
    jaccard_matrix,
    row_title = label2,
    column_title = label1,
    keep_legend_name = label2,
    col_fun = heatmap_col_fun,
    # additional arguments
    column_names_rot = 90
  ) |>
    ComplexHeatmap::draw(heatmap_legend_side = "bottom")
}

# This function calculates and plots delta median for two groups of barcodes: those in agree and those not
# This code is based on this scpca-nf code:
# https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_supplemental_report.rmd#L626
#
# df: SingleR data frame with `scores.` columns
# agree_barcodes: Vector of barcodes that have agreeing labels between aggregated and not
# plot_title: Title for plot
plot_delta_median <- function(df, agree_barcodes, plot_title) {
  # extract scores into matrix
  scores_mat <- df |>
    dplyr::select(starts_with("scores.")) |>
    as.matrix()
  rownames(scores_mat) <- df$barcodes

  # Create data frame for plotting with delta median and the full *non-pruned* cell labels
  delta_median_df <- tibble::tibble(
    delta_median = rowMaxs(scores_mat) - rowMedians(scores_mat),
    # Need to grab the non-pruned label for this plot
    full_labels = aggr$labels,
    # if pruned.labels are NA ==> low confidence
    # so, negate for this variable:
    confident = !is.na(df$pruned.labels)
  ) |>
    dplyr::mutate(
      confident = ifelse(confident, "High-quality", "Low-quality"),
      barcodes = df$barcodes,
      annotations_agree = barcodes %in% agree_barcodes
    )

  ggplot(delta_median_df) +
    aes(x = annotations_agree, y = delta_median, color = confident) +
    # don't jitter the height
    geom_jitter(height = 0, alpha = 0.7) +
    theme(legend.position = "bottom") +
    ggtitle(plot_title)
}
```

### Prepare input data

Read SCE object so we can get UMAP and consensus cell types:

```{r}
sce_file <- file.path(
  data_dir,
  params$sample_id,
  glue::glue("{params$library_id}_processed.rds")
)

sce_df <- readRDS(sce_file) |>
  scuttle::makePerCellDF() |>
  tibble::as_tibble() |>
  dplyr::select(
    barcodes,
    consensus_annotation = consensus_celltype_annotation,
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2
  )
```

Read cell type data frames:
```{r}
validation_df <- readr::read_tsv(validation_url) |>
  dplyr::select(consensus_annotation, validation_group_annotation)

aggr <- readr::read_rds(
  file.path(
    results_dir,
    params$sample_id,
    "aggregated",
    glue::glue("{params$library_id}_singler-annotations.rds")
  )
) |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::mutate(ref_type = "aggregated")

non_aggr <- readr::read_rds(
  file.path(
    results_dir,
    params$sample_id,
    "non-aggregated",
    glue::glue("{params$library_id}_singler-annotations.rds")
  )
) |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::mutate(ref_type = "non-aggregated")
```

Join and prepare data for use:

```{r}
singler_df <- data.frame(
  barcodes = aggr$barcodes,
  "aggregated" = aggr$pruned.labels,
  "not_aggregated" = non_aggr$pruned.labels
)

sce_df <- sce_df |>
  dplyr::left_join(singler_df, by = "barcodes") |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # set validation NA -> Unknown
  dplyr::mutate(
    validation_group_annotation =
      ifelse(is.na(validation_group_annotation), "Unknown", validation_group_annotation)
  )
```

## Aggregated vs. non-aggregated

First a quick and dirty query: What percent of annotations made with either the aggregated and non-aggregated reference versions agree?

```{r}
sum(singler_df$aggregated == singler_df$not_aggregated, na.rm = T) / nrow(singler_df)
```

Where specifically are their disagreements?
This table displays the number of cells that have the given cell label disagreements:

```{r}
singler_df |>
  dplyr::filter(aggregated != not_aggregated) |>
  dplyr::count(aggregated, not_aggregated) |>
  dplyr::arrange(desc(n))
```

### Delta median scores

We'll plot the `delta_median` scores for each `SingleR` inference (aggregated and non-aggregated) for cells where annotations agree vs. differ.
We would like to see that scores will be _lower_ for the cells where inferences between reference versions disagree.
Cells will also be colored by their reliability as determined by `SingleR`.

```{r}
agree_barcodes <- singler_df |>
  dplyr::filter(aggregated == not_aggregated) |>
  dplyr::pull(barcodes)

plot_delta_median(aggr, agree_barcodes, plot_title = "Aggregated delta_median") +
  plot_delta_median(non_aggr, agree_barcodes, plot_title = "Non-aggregated delta_median") &
  theme(legend.position = "none")
```

### Confusion matrix metrics

Here are metrics from a confusion matrix comparing these annotations:

```{r}
calculate_confusion(
  sce_df,
  "aggregated",
  "not_aggregated"
)
```

### Heatmap: Aggregated vs. non-aggregated

This heatmap compares annotations between aggregated and non-aggregated directly:

```{r, fig.height = 8}
make_jaccard_heatmap(
  sce_df,
  "aggregated",
  "not_aggregated",
  "Aggregated labels",
  "Non-aggregated labels"
)
```

## Heatmap: Aggregated vs. consensus annotations

This heatmap compares annotations between `SingleR` inference using the aggregated reference and consensus cell types.
The broader validation grouping for consensus cell types is used to keep the plot manageable.

```{r, fig.height = 10}
make_jaccard_heatmap(
  sce_df,
  "consensus_annotation",
  "aggregated",
  "OpenScPCA Consensus cell type validation group",
  "Aggregated labels"
)
```

## Heatmap: Non-aggregated vs. consensus annotations

This heatmap compares annotations between `SingleR` inference using the non-aggregated reference and consensus cell types.
The broader validation grouping for consensus cell types is used to keep the plot manageable.

```{r, fig.height = 10}
make_jaccard_heatmap(
  sce_df,
  "consensus_annotation",
  "not_aggregated",
  "OpenScPCA Consensus cell type validation group",
  "Non-aggregated labels"
)
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-07-02 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/jaccard-utils.R

@@ -1,0 +1,102 @@
+# This file contains functions to support making Jaccard similarity heatmaps
+# Code is adapted from:
+# https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_supplemental_report.rmd#L50
+
+#' Function to calculate Jaccard similarity on two vectors
+#'
+#' @param vec1 First vector
+#' @param vec2 Second vector
+#'
+#' @return Jaccard similarity between the vectors
+calc_jaccard <- function(vec1, vec2) {
+  length(intersect(vec1, vec2)) / length(union(vec1, vec2))
+}
+
+
+# Wrapper function to calculate jaccard similarity matrix for two categorical variables
+#'
+#' @param celltype_df The celltype_df data frame which must contain these columns:
+#'   `colname1`, `colname2`, and `barcodes`
+#' @param colname1 Column name, as a string, of first categorical variable of interest
+#' @param colname2 Column name, as a string, of second categorical variable of interest
+#'
+#' @return Jaccard similarity matrix for the two columns. `colname1` values will
+#'   be row names and `colname2` values will be column names in the final matrix
+make_jaccard_matrix <- function(celltype_df, colname1, colname2) {
+  # make lists of barcodes for each category, named by the category
+  id1_list <- split(celltype_df$barcodes, celltype_df[[colname1]])
+  id2_list <- split(celltype_df$barcodes, celltype_df[[colname2]])
+
+  # create the grid of comparisons
+  cross_df <- tidyr::expand_grid(id1 = names(id1_list), id2 = names(id2_list))
+
+  # calculate a single Jaccard index for each combination using split lists & ids
+  jaccard_scores <- cross_df |>
+    purrr::pmap_dbl(\(id1, id2){
+      calc_jaccard(id1_list[[id1]], id2_list[[id2]])
+    })
+
+  # add scores to the comparison grid and convert to matrix
+  jaccard_matrix <- cross_df |>
+    dplyr::mutate(jaccard = jaccard_scores) |>
+    # convert to matrix
+    tidyr::pivot_wider(
+      names_from = "id2",
+      values_from = "jaccard"
+    ) |>
+    tibble::column_to_rownames(var = "id1") |>
+    as.matrix()
+
+  return(jaccard_matrix)
+}
+
+#' Create a ComplexHeatmap from a matrix
+#'
+#' @param mat Matrix to create heatmap from.
[…]

### 2025-07-02 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/01_reference-aggregation.Rmd
plot fixes

@@
 params:
-  sample_id: "SCPCS000105"
-  library_id: "SCPCL000122"
+  sample_id: "SCPCS000109"
+  library_id: "SCPCL000126"
 ---
@@
 
-Read SCE object so we can get UMAP and consensus cell types:
+Read SCE object so we can get consensus cell types:
 
@@
 
-sce_df <- readRDS(sce_file) |>
-  scuttle::makePerCellDF() |>
+celltype_df <- readRDS(sce_file) |>
+  colData() |>
   tibble::as_tibble() |>
@@
     barcodes,
-    consensus_annotation = consensus_celltype_annotation,
-    UMAP1 = UMAP.1,
-    UMAP2 = UMAP.2
+    consensus_annotation = consensus_celltype_annotation
   )
@@
 
-sce_df <- sce_df |>
+celltype_df <- celltype_df |>
   dplyr::left_join(singler_df, by = "barcodes") |>
@@
 plot_delta_median(aggr, agree_barcodes, plot_title = "Aggregated delta_median") +
-  plot_delta_median(non_aggr, agree_barcodes, plot_title = "Non-aggregated delta_median") &
-  theme(legend.position = "none")
+  plot_delta_median(non_aggr, agree_barcodes, plot_title = "Non-aggregated delta_median") +
+  plot_layout(guides = "collect") & theme(legend.position = "bottom")
 ```
@@
 calculate_confusion(
-  sce_df,
+  celltype_df,
   "aggregated",
@@
 
-```{r, fig.height = 8}
+```{r, fig.height = 8, fig.width = 8}
 make_jaccard_heatmap(
-  sce_df,
+  celltype_df,
   "aggregated",
@@
 
-```{r, fig.height = 10}
+```{r, fig.height = 8, fig.width = 8}
 make_jaccard_heatmap(
-  sce_df,
-  "consensus_annotation",
+  celltype_df,
+  "validation_group_annotation",
   "aggregated",
@@
 
-```{r, fig.height = 10}
+```{r,fig.height = 8, fig.width = 8}
 make_jaccard_heatmap(
-  sce_df,
-  "consensus_annotation",
+  celltype_df,
+  "validation_group_annotation",
   "not_aggregated",

### 2025-07-03 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/01_reference-aggregation.Rmd
move the heatmap function over to jaccard-utils

@@
   )
-}
-
-# This function calculates Jaccard similarity between two vectors and plots a heatmap
-# Code is based on this scpca-nf code:
-# https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_supplemental_report.rmd#L50
-#
-# df: data frame with columns of interest
-# annotation_col1: first column to compare; arg should be a string
-# annotation_col2: second column to compare; arg should be a string
-# label1: label to use in heatmap for annotation_col1 axis
-# label2: label to use in heatmap for annotation_col2 axis
-make_jaccard_heatmap <- function(
-    df,
-    annotation_col1,
-    annotation_col2,
-    label1,
-    label2) {
-  # Calculate jaccard matrix
-  jaccard_matrix <- make_jaccard_matrix(
-    df,
-    annotation_col1,
-    annotation_col2
-  )
-
-  create_single_heatmap(
-    jaccard_matrix,
-    row_title = label2,
-    column_title = label1,
-    keep_legend_name = label2,
-    col_fun = heatmap_col_fun,
-    # additional arguments
-    column_names_rot = 90
-  ) |>
-    ComplexHeatmap::draw(heatmap_legend_side = "bottom")
 }

### 2025-07-03 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/01_reference-aggregation.Rmd
WIP path updates

@@
 
-celltype_df <- readRDS(sce_file) |>
-  colData() |>
-  tibble::as_tibble() |>
+sce_df <- readRDS(sce_file) |>
+  scuttle::makePerCellDF() |>
   dplyr::select(
     barcodes,
-    consensus_annotation = consensus_celltype_annotation
+    consensus_annotation = consensus_celltype_annotation,
+    UMAP1 = UMAP.1,
+    UMAP2 = UMAP.2
   )
@@
     "aggregated",
-    glue::glue("{params$library_id}_singler-annotations.rds")
+    glue::glue("{params$library_id}_singler-result.rds")
   )
@@
     params$sample_id,
-    "non-aggregated",
-    glue::glue("{params$library_id}_singler-annotations.rds")
+    "not-aggregated",
+    glue::glue("{params$library_id}_singler-result.rds")
   )
@@
   tibble::rownames_to_column("barcodes") |>
-  dplyr::mutate(ref_type = "non-aggregated")
+  dplyr::mutate(ref_type = "not-aggregated")
 ```

### 2025-07-03 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/01_reference-aggregation.Rmd
add UMAP with harmonized palette

@@
 params:
-  sample_id: "SCPCS000109"
-  library_id: "SCPCL000126"
+  sample_id: "SCPCS000105"
+  library_id: "SCPCL000122"
 ---
@@
 
+# broad consensus cell type groups
 validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
+
+# umap colors
+palette_url <- "https://raw.githubusercontent.com/AlexsLemonade/scpca-paper-figures/refs/heads/main/palettes/validation-group-palette.tsv"
 ```
@@
 
+palette_df <- readr::read_tsv(palette_url) # umap plotting later
+
 aggr <- readr::read_rds(
@@
   tibble::rownames_to_column("barcodes") |>
-  dplyr::mutate(ref_type = "not-aggregated")
+  dplyr::mutate(ref_type = "non-aggregated")
 ```
@@
 
-celltype_df <- celltype_df |>
+celltype_df <- sce_df |>
   dplyr::left_join(singler_df, by = "barcodes") |>
   dplyr::left_join(validation_df, by = "consensus_annotation") |>
-  # set validation NA -> Unknown
+  # set validation NA -> unknown
   dplyr::mutate(
     validation_group_annotation =
-      ifelse(is.na(validation_group_annotation), "Unknown", validation_group_annotation)
+      ifelse(is.na(validation_group_annotation), "unknown", validation_group_annotation)
   )
@@
 
+## UMAPs
+
+Below, we show the UMAPs colored by cell types for each of aggregated, non-aggregated, and consensus (again, using the validation groupings for plot visibility).
+
+Before plotting cell type labels were harmonized to the extent possible to that `NBAtlas` labels matched the corresponding consensus cell type labels (e.g., `Fibroblast` -> `fibroblast` or `Endothelial` -> `endothelial cell`).
+Cell types which not literally just spelling variants were left alone.
+
+The UMAP therefore has one overall legend for all possible cell types, so colors are directly comparable across panels.
+The specific cell types that are part of either refernce are printed below.
+
+```{r}
+# Recode NBAtlas cell types where possible so they match with colors
+celltype_recoded_df <- celltype_df |>
+  dplyr::select(-consensus_annotation) |>
+  # pivot longer for wrangling first
+  tidyr::pivot_longer(
+    c(validation_group_annotation, aggregated, not_aggregated),
+    names_to = "annotation_type",
+    values_to = "label"
+  ) |>
+  dplyr::mutate(
+    label_recoded = dplyr::case_when(
+      # nbatlas ~ consensus
+      label %in% c("Myeloid", "Fibroblast") ~ stringr::str_to_lower(label),
+      label == "Endothelial" ~ "endothelial cell",
+      label == "Plasma" ~ "plasma cell",
+      label == "NK cell" ~ "natural killer cell",
+      is.na(label) ~ "unknown",
+      .default = label
+    )
+  )
+
+# Recode some existing colors we don't need
+palette_df <- palette_df |>
+  dplyr::mutate(
+    label_recoded = dplyr::case_when(
+      validation_group_annotation == "hematopoietic precursor cell" ~ "Neuroendocrine",
+      validation_group_annotation == "muscle cell" ~ "RBCs",
+      validation_group_annotation == "adipocyte" ~ "Schwann",
+      validation_group_annotation == "pericyte" ~ "pDC",
+      validation_group_annotation == "innate lymphoid cell" ~ "Stromal other",
+      .default = validation_group_annotation
+    )
+  ) |>
+  dplyr::filter(label_recoded %in% unique(celltype_recoded_df$label_recoded))
+
+# Define palette
+pal <- palette_df$color
+names(pal) <- palette_df$label_recoded
+```
+
+```{r, fig.width = 12}
+ggplot(celltype_recoded_df) +
+  aes(x = UMAP1, y = UMAP2, color = label_recoded) +
+  geom_point(alpha = 0.6, size = 0.5) +
+  scale_color_manual(values = pal) +
+  facet_wrap(vars(annotation_type)) +
+  coord_equal() +
+  theme(
+    axis.ticks = element_blank(),
+    axis.text = element_blank(),
+    legend.position = "bottom"
+  ) +
+  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
+```
+Cell types that are part of `NBAtlas` include:
+
+```{r}
+celltype_recoded_df |>
+  dplyr::filter(annotation_type != "validation_group_annotation") |>
+  dplyr::pull(label_recoded) |>
+  unique()
+```
+
+Cell types that are part of consensus labels include:
+
+```{r}
+celltype_recoded_df |>
+  dplyr::filter(annotation_type == "validation_group_annotation") |>
+  dplyr::pull(label_recoded) |>
+  unique()
+```
+
 ## Session Info

### 2025-07-03 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/01_reference-aggregation.Rmd
file name plural

@@
 params:
-  sample_id: "SCPCS000105"
-  library_id: "SCPCL000122"
+  sample_id: "SCPCS000115"
+  library_id: "SCPCL000133"
 ---
@@
     "aggregated",
-    glue::glue("{params$library_id}_singler-result.rds")
+    glue::glue("{params$library_id}_singler-results.rds")
   )
@@
     "not-aggregated",
-    glue::glue("{params$library_id}_singler-result.rds")
+    glue::glue("{params$library_id}_singler-results.rds")
   )

### 2025-07-03 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd

@@
 
-# path to results directory
-results_dir <- file.path(module_base, "..", "..", "data", "current", "results", "cell-type-consensus", params$project_id)
+# path to the `2025-03-20` release results directory
+results_dir <- file.path(module_base, "..", "..", "data", "2025-03-20", "results", "cell-type-consensus", params$project_id)
 ```
@@
 stopifnot(
-  "No cell type results found in data/current/results/cell-type-consensus" = length(consensus_results_files) > 0,
-  "No gene expression files found in data/current/results/cell-type-consensus" = length(gene_exp_files) > 0
+  "No cell type results found in data/2025-03-20/results/cell-type-consensus. Please make sure you downloaded result files with the flag `--release 2025-03-20`." = length(consensus_results_files) > 0,
+  "No gene expression files found in data/2025-03-20/results/cell-type-consensus. Please make sure you downloaded result files with the flag `--release 2025-03-20`." = length(gene_exp_files) > 0
 )
@@
 # now only keep unique markers for the dotplot
+# except for hematopoietic precursor cells, where there are no unique markers
 unique_markers_df <- filtered_markers_df |>

### 2025-07-06 analyses/infercnv-osteosarcoma/notebook-template.Rmd
Created analysis module and updated analysis README.md

--- 2024-11-19 analyses/cell-type-consensus/notebook-template.Rmd
@@
 ---
-title: "cell-type-consensus Analysis"
+title: "infercnv-osteosarcoma Analysis"
 author:
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "cell-type-consensus")
+module_base <- file.path(repository_base, "analyses", "infercnv-osteosarcoma")
 ```

### 2025-07-07 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/generate_reference-aggregation.sh
Apply suggestions from code review
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

@@ -1,8 +1,8 @@
 #!/bin/bash
 #
 # Generate results and render the template exploratory notebook `01_reference-aggregation.Rmd`.
-# This analysis only considers a subset of samples chosen their range in
-# number of cells, diversity of consensus cell type annotations, and diversity in tech/sequencing units.
+# This analysis only considers a subset of samples 
+# They were chosen to have a range in number of cells, diversity of consensus cell type annotations, and diversity in tech/sequencing units.
 # SCPCS000104: nucleus, 10Xv2
 # SCPCS000105: cell, 10Xv2
 # SCPCS000108: cell, 10Xv3
@@ -38,7 +38,7 @@ for sample_id in $sample_ids; do
     for tsv_file in "${singler_result_dir}/${sample_id}"/aggregated/*annotations.tsv; do
         library_id=$(basename "$tsv_file" | sed 's/_singler-annotations.tsv$//')
 
-        # Skip the SCPCS000109 we're not considering here library
+        # Skip the library from SCPCS000109 we're not considering here
         if [[ $library_id == "SCPCL000127" ]]; then
             continue
         fi

### 2025-07-07 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/reference-aggregation.Rmd
Apply suggestions from code review
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>
define paths at the top

--- 2025-07-03 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/01_reference-aggregation.Rmd
@@
 data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")
+```
 
+```{r url paths}
 # broad consensus cell type groups
@@
 palette_url <- "https://raw.githubusercontent.com/AlexsLemonade/scpca-paper-figures/refs/heads/main/palettes/validation-group-palette.tsv"
+```
+
+```{r file paths}
+# SingleR results
+aggr_singler_file <- file.path(
+  results_dir,
+  params$sample_id,
+  "aggregated",
+  glue::glue("{params$library_id}_singler-results.rds")
+)
+
+non_aggr_singler_file <- file.path(
+  results_dir,
+  params$sample_id,
+  "not-aggregated",
+  glue::glue("{params$library_id}_singler-results.rds")
+)
+
+# SCE file
+sce_file <- file.path(
+  data_dir,
+  params$sample_id,
+  glue::glue("{params$library_id}_processed.rds")
+)
 ```
@@
 ```{r}
-sce_file <- file.path(
-  data_dir,
-  params$sample_id,
-  glue::glue("{params$library_id}_processed.rds")
-)
-
 sce_df <- readRDS(sce_file) |>
@@
 
-aggr <- readr::read_rds(
-  file.path(
-    results_dir,
-    params$sample_id,
-    "aggregated",
-    glue::glue("{params$library_id}_singler-results.rds")
-  )
-) |>
+aggr <- readr::read_rds(aggr_singler_file) |>
   as.data.frame() |>
-  tibble::rownames_to_column("barcodes") |>
-  dplyr::mutate(ref_type = "aggregated")
+  tibble::rownames_to_column("barcodes")
 
-non_aggr <- readr::read_rds(
-  file.path(
-    results_dir,
-    params$sample_id,
-    "not-aggregated",
-    glue::glue("{params$library_id}_singler-results.rds")
-  )
-) |>
+non_aggr <- readr::read_rds(non_aggr_singler_file) |>
   as.data.frame() |>
-  tibble::rownames_to_column("barcodes") |>
-  dplyr::mutate(ref_type = "non-aggregated")
+  tibble::rownames_to_column("barcodes")
 ```
@@
 
-## Heatmap: Aggregated vs. consensus annotations
+## Comparison with consensus cell type annotations
+### Heatmap: Aggregated vs. consensus annotations
 
@@
 
-## Heatmap: Non-aggregated vs. consensus annotations
+### Heatmap: Non-aggregated vs. consensus annotations
 
@@
 Before plotting cell type labels were harmonized to the extent possible to that `NBAtlas` labels matched the corresponding consensus cell type labels (e.g., `Fibroblast` -> `fibroblast` or `Endothelial` -> `endothelial cell`).
-Cell types which not literally just spelling variants were left alone.
+Cell types that were not literally just spelling variants were left alone.
 

### 2025-07-07 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/reference-aggregation.Rmd
wording about valdation groupings and stromal

@@
 ## Comparison with consensus cell type annotations
+
+In this section, we compare aggregated & non-aggregated annotations to the consensus cell type labels.
+In these comparisons, we consider the broader validation groupings for consensus cell types in order to keep plots manageable and more easily interpreted.
+
 ### Heatmap: Aggregated vs. consensus annotations
@@
 This heatmap compares annotations between `SingleR` inference using the aggregated reference and consensus cell types.
-The broader validation grouping for consensus cell types is used to keep the plot manageable.
 
@@
 This heatmap compares annotations between `SingleR` inference using the non-aggregated reference and consensus cell types.
-The broader validation grouping for consensus cell types is used to keep the plot manageable.
 
@@
 Before plotting cell type labels were harmonized to the extent possible to that `NBAtlas` labels matched the corresponding consensus cell type labels (e.g., `Fibroblast` -> `fibroblast` or `Endothelial` -> `endothelial cell`).
-Cell types that were not literally just spelling variants were left alone.
+Cell types that were not literally just spelling variants were left alone; note that `stromal` cell types between references were also not harmonized since this group doesn't always have a clear definition.
 

### 2025-07-07 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/reference-aggregation.Rmd

@@
 
-# Recode some existing colors we don't need
+# Recode some existing colors we don't need - this is done to update the existing color
+# palette to match the cell types we need in this plot
 palette_df <- palette_df |>
   dplyr::mutate(
+    # colors associated with these original validation groups will be now used to
+    # color the given the NBAtlas cell types
     label_recoded = dplyr::case_when(
@@
   ) |>
+  # get rid of any cell types that won't appear in the plot so palette_df just contains
+  # colors & cell types of interest
   dplyr::filter(label_recoded %in% unique(celltype_recoded_df$label_recoded))
 
-# Define palette
+# Define palette as a vector from all remaining colors/labels
 pal <- palette_df$color

### 2025-07-08 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/separate-tumor.Rmd
WIP: notebook comparing results

---
title: "Compare NE-combined vs. NE-separated reference"
subtitle: "`r glue::glue('Results for {params$sample_id}: {params$library_id}')`"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  sample_id: "SCPCS000109"
  library_id: "SCPCL000126"
---

This notebook does some preliminary comparisons of `SingleR` results obtained with these two versions of the NBAtlas (aggregated) reference:

* `NE-combined`: The reference has a single category for all Neuroendocrine (NE) cells
* `NE-separated`: The reference has two categories, Neuroendocrine and Neuroendocrine-tumor (NE-tumor) where the latter are NE cells which are part of the paper's "tumor zoom" subset and were explicitly classified as malignant

This notebook explores whether `Neuroendocrine` should be a single label or have the two separated categories.
On one hand, these different categories might indicate different tumor states or highly-similar normal cells. 
However, per direct correspondence with the NBAtlas authors, it is probably more likely that any different signal associated with the malignant cells vs other NE cells is driven by technical reasons such as data quality, and _not_ biological signal.

The goal of this notebook is to briefly look at the difference in cell type assignments between these two reference options to determine whether to proceed with a separated or combined NE category of cells, with the null that we should use the combined reference unless there is a compelling difference between the results.

## Setup

```{r, warning = FALSE}
options(readr.show_col_types = FALSE)

suppressWarnings({
  suppressPackageStartupMessages({
    library(ggplot2)
    library(patchwork)
    library(SingleCellExperiment)
  })
})

theme_set(theme_bw())

# Define color ramp for shared use in the heatmaps
heatmap_col_fun <- circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue"))
# Set heatmap padding option
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(0.6, "in"))
```

### Paths

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
results_dir <- file.path(module_dir, "results", "singler")
data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")
```

```{r url paths}
# broad consensus cell type groups
validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"

# umap colors
palette_url <- "https://raw.githubusercontent.com/AlexsLemonade/scpca-paper-figures/refs/heads/main/palettes/validation-group-palette.tsv"
```

```{r file paths}
# SingleR results
combined_singler_file <- file.path(
  results_dir,
  params$sample_id,
  "aggregated",
  "NE-tumor-combined",
  glue::glue("{params$library_id}_singler-results.rds")
)

separated_singler_file <- file.path(
  results_dir,
  params$sample_id,
  "aggregated",
  "NE-tumor-separated",
  glue::glue("{params$library_id}_singler-results.rds")
)

# SingleR trained models
model_separated_file <- file.path(
  module_dir,
  "scratch",
  "singler-model_nbatlas_aggregated_NE-tumor-separated.rds"
)
model_combined_file <- file.path(
  module_dir,
  "scratch",
  "singler-model_nbatlas_aggregated_NE-tumor-combined.rds"
)

# SCE file
sce_file <- file.path(
  data_dir,
  params$sample_id,
  glue::glue("{params$library_id}_processed.rds")
)
```

### Functions

```{r}
# Source Jaccard and heatmap utilities functions:
source(file.path(module_dir, "exploratory-notebooks", "jaccard-utils.R"))
```

### Prepare input data

Read SCE object so we can get consensus cell types:

```{r}
sce <- readRDS(sce_file)
sce_df <- sce |>
  scuttle::makePerCellDF() |>
  dplyr::select(
    barcodes,
    consensus_annotation = consensus_celltype_annotation,
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2
  )
```

Read cell type and result data frames:
```{r}
validation_df <- readr::read_tsv(validation_url) |>
  dplyr::select(consensus_annotation, validation_group_annotation)

palette_df <- readr::read_tsv(palette_url) # umap plotting later

singler_results_combined <- readr::read_rds(combined_singler_file)
singler_results_separated <- readr::read_rds(separated_singler_file)
```

Join and prepare data for use:

```{r}
celltype_df <- data.frame(
  barcodes = rownames(singler_results_combined),
  "singler_label_combined" = as.data.frame(singler_results_combined)$pruned.labels,
  "singler_label_separated" = as.data.frame(singler_results_separated)$pruned.labels
) |>
  dplyr::left_join(sce_df, by = "barcodes") |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # set validation NA -> unknown
  dplyr::mutate(
    validation_group_annotation =
      ifelse(is.na(validation_group_annotation), "unknown", validation_group_annotation)
  )

# remove sce_df for space
rm(sce_df)
```

Read the SingleR models and NBAtlas cell type marker genes:

```{r}
model_separated <- readRDS(model_separated_file)
model_combined <- readRDS(model_combined_file)
```

## Compare trained SingleR models

In this section, we will compare some of the marker genes identified in the trained SingleR models directly.

When training a model, SingleR performs pairwise comparisons to identify marker genes as the top 10 upregulated genes for each label compared to each other label.

### Intersecting NE marker genes in the separated model

We'll first look at the `NE-separated` model which has two categories for `Neuroendocrine` (NE) and `Neuroendocrine-tumor` (NE-tumor).
We'll pull out the marker genes for each and ask how similar they are.
In other words, we'll ask if there is overlap between:

* Marker genes upregulated in NE compared to each other cell type
* Marker genes upregulated in NE-tumor compared to each other cell type

```{r}
# markers upregulated NE compared to other cell types in the separated model
ne_markers <- model_separated$markers$full$Neuroendocrine

# markers upregulated NE-tumor compared to other cell types in the separated model
ne.tumor_markers <- model_separated$markers$full$`Neuroendocrine-tumor`
```

Below we print the intersection of these two sets of marker genes for each cell type comparison:
```{r}
separated_cell_types <- names(ne_markers)
non_neuro_cell_types <- separated_cell_types[!(stringr::str_detect(separated_cell_types, "Neuro"))]

intersecting_marker_genes <- non_neuro_cell_types |>
  purrr::set_names() |>
  purrr::map(
    \(cell_type) {
      intersect(
        ne_markers[[cell_type]],
        ne.tumor_markers[[cell_type]]
      )
    }
  )

# Print as data frame so it doesn't print double
intersecting_marker_genes |>
  purrr::map(
    \(x) data.frame(genes = paste(sort(x), collapse = ", "))
  ) |>
  purrr::list_rbind(names_to = "cell_type") |>
  # add column with gene count
  dplyr::mutate(n_genes = stringr::str_count(genes, ",") + 1) |>
  dplyr::select(cell_type, n_genes, gene_symbols = genes)
```

Most marker genes are actually the same for the NE and NE-tumor pairwise comparisons, indeed showing us that these categories are highly similar.
The main exception is `RBCs` for which there are apparently very different gene lists for NE and NE-tumor. 
Moreover, the genes tend to be the same across cell types, which suggests these are probably more universal NE markers.

### Expression of different NE marker genes in the separated model

Next, we'll pull out the _distinct_ marker genes associated with `NE` and `NE-tumor` groups that are also not represented in the combined model's `NE` marker genes.

We'll look at these genes' expression in cells that were annotated as `NE` and `NE-tumor` by the separated model.
We expect that these distributions will show some difference (since `SingleR` will find genes no matter what!) between annotated cell types `NE` and `NE-tumor`.
If the magnitude of the difference is large, it suggests that separating NE into two groups may not be unreasonable.
If the magnitude of the difference is small, it suggests that separating NE into two groups is oversplitting.

That said, note that there are likely very different numbers of cells labeled as `NE` and `NE-tumor`, which may influence how these plots are interpreted.
```{r}
celltype_df |>
  dplyr::filter(stringr::str_detect(singler_label_separated, "Neuro")) |>
  dplyr::count(singler_label_separated)
```

```{r}
combined_marker_genes <- model_combined$markers$full$Neuroendocrine |>
  purrr::reduce(c) |>
  unique()

cell_types <- names(ne_markers)
non_neuro_cell_types <- cell_types[!(stringr::str_detect(cell_types, "Neuro"))]

different_marker_genes <- non_neuro_cell_types |>
  purrr::set_names() |>
  purrr::map(
    \(cell_type) {
      # In NE but not in NE-tumor
      diff_genes_ne <- setdiff(
        ne_markers[[cell_type]],
        ne.tumor_markers[[cell_type]]
      )

      # In NE-tumor but not in NE
      diff_genes_ne.tumor <- setdiff(
        ne.tumor_markers[[cell_type]],
        ne_markers[[cell_type]]
      )

      return(
        list(
          "ne_only" = diff_genes_ne,
          "ne.tumor_only" = diff_genes_ne.tumor
        )
      )
    }
  ) |>
  # transpose so it is a list of two, "ne_only" and "ne.tumor_only",
  # each of which is a named list of marker genes for each cell type
  purrr::transpose() |>
  # collapse each nested list to remove cell types
  # keep only unique genes
  # remove any genes that are also in combined_marker_genes
  purrr::map(
    \(gene_list) {
      gene_vector <- gene_list |>
        purrr::reduce(c) |>
        unique()
      gene_vector[!(gene_vector %in% combined_marker_genes)]
    }
  )

# This final list are the genes uniquely upregulated in the NE and
# NE-tumor categories in the separated model
different_marker_genes
```

```{r, fig.width = 12}
plot_barcodes <- celltype_df |>
  dplyr::filter(stringr::str_detect(singler_label_separated, "Neuroendocrine")) |>
  dplyr::pull(barcodes)

# Convert rownames to symbols to match our gene lists
sce <- rOpenScPCA::sce_to_symbols(sce, reference = "sce")

# Add the separated cell types into colData as `singler_cell_type`
colData(sce) <- colData(sce) |>
  as.data.frame() |>
  dplyr::inner_join(
    celltype_df |>
      dplyr::select(barcodes, singler_cell_type = singler_label_separated),
    by = "barcodes"
  ) |>
  DataFrame(row.names = rownames(colData(sce)))

# plot the NE-only genes
plot1 <- scater::plotExpression(
  sce[different_marker_genes$ne_only, plot_barcodes],
  features = different_marker_genes$ne_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5
) +
  theme(legend.position = "none") +
  ggtitle("Neuroendocrine marker genes")

# plot the NE-tumor--only genes
plot2 <- scater::plotExpression(
  sce[different_marker_genes$ne.tumor_only, plot_barcodes],
  features = different_marker_genes$ne.tumor_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5
) +
  theme(legend.position = "none") +
  ggtitle("Neuroendocrine-tumor marker genes")

plot1 + plot2
```

### Compare separated to combined models

Next, we'll look at the NE marker genes between references.

For a quick summary, we'll ask: How many of the NE markers in the combined model (out of 10 total per cell type comparison!) are _also included_ in the separated model's NE (or NE-tumor) marker gene lists?

```{r}
combined_markers <- model_combined$markers$full$Neuroendocrine

combined_markers |>
  # only comparisons to other groups of cells
  purrr::discard_at("Neuroendocrine") |>
  purrr::imap(
    \(combined_marker_genes, cell_type) {
      separated_model_markers <- c(
        # upregulated in NE compared to cell_type
        ne_markers[[cell_type]],
        # upregulated in NE-tumor compared to cell-type
        ne.tumor_markers[[cell_type]]
      ) |> unique()

      # how many combined are in that separated list?
      sum(combined_marker_genes %in% separated_model_markers)
    }
  )
```

Here, we see that NE marker genes from the combined model are essentially entirely contained within the NE and/or NE-tumor categories of the separated model.

### Conclusions

Based on a brief comparison of marker genes in the SingleR models, we see that:

* In the _separated model_, marker genes that distinguish NE from other cell types and marker genes that distinguish NE-tumor from other cell types are highly similar.
  * These genes were identified in NBAtlas as NE marker genes as well
  * These genes tend to overlap with marker genes that distinguish NE from other cell types in the _combined model_
* It is additionally worth noting that SingleR inference took much longer with the separated than combined model

This minimal exploration does not suggest that there are salient differences between the two NE categories in the separated model.

## SingleR inference: Combined vs. separated

We'll begin with a heatmap directly comparing annotations between the two SingleR models:

```{r, fig.height = 8, fig.width = 8}
make_jaccard_heatmap(
  celltype_df,
  "singler_label_combined",
  "singler_label_separated",
  "Combined model (single NE category)",
  "Separated model (NE and NE-tumor)"
)
```

In the next section, we'll look at the numbers a bit more closely.

### Cells labeled as Neuroendocrine or Neuroendocrine-tumor

We'll first directly compare cells that were labeled as a `Neuroendocrine` cell type with either model.
Each row shows counts of cells that had given pairs corresponding labels from inference with either model.

```{r}
celltype_df |>
  dplyr::filter(
    singler_label_combined == "Neuroendocrine" | stringr::str_detect(singler_label_separated, "Neuroendocrine")
  ) |>
  dplyr::count(singler_label_combined, singler_label_separated) |>
  dplyr::arrange(desc(n))
```

### Cells with different labels between models

Next, we'll print a table that directly shows the counts of cells with different labels as inferred with either model.

```{r}
celltype_df |>
  dplyr::filter(singler_label_combined != singler_label_separated) |>
  dplyr::count(singler_label_combined, singler_label_separated) |>
  dplyr::arrange(desc(n))
```

### Unlabeled cells

We can also ask whether each model was able to label cells successfully, or whether more cells were ambiguous (aka `pruned.label` is `NA`) in one model or the other.

How many cells were `NA` in both models?

```{r}
na_cells <- celltype_df |>
  dplyr::filter(is.na(singler_label_combined), is.na(singler_label_separated)) |>
  dplyr::pull(barcodes)

length(na_cells)
```

#### Unlabeled cells in the separated model

Excluding cells which neither model labeled...

* How many cells did the separated model uniquely fail to label?
* What did the combined model label those cells?

```{r}
# if na in separated, what was it in combined?
unlabeled_in_separated <- singler_results_separated |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::filter(is.na(pruned.labels), !(barcodes %in% na_cells)) |>
  dplyr::pull(barcodes)
length(unlabeled_in_separated)
```

```{r}
celltype_df |>
  dplyr::filter(barcodes %in% unlabeled_in_separated) |>
  dplyr::count(singler_label_combined) |>
  dplyr::arrange(desc(n))
```

#### Unlabeled cells in the combined model

Excluding cells which neither model labeled...

* How many cells did the combined model uniquely fail to label?
* What did the separated model label those cells?

```{r}
# if na in combined, what was it in separated?
unlabeled_in_combined <- singler_results_combined |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::filter(is.na(pruned.labels), !(barcodes %in% na_cells)) |>
  dplyr::pull(barcodes)
length(unlabeled_in_combined)
```

```{r}
celltype_df |>
  dplyr::filter(barcodes %in% unlabeled_in_combined) |>
  dplyr::count(singler_label_separated) |>
  dplyr::arrange(desc(n))
```

### Score comparisons

In the separated reference, for cells that are labeled either `NE` or `NE-tumor` (considering the `pruned.labels` specifically), which label has the _second highest score_?
We expect that the second highest score will usually be the other `NE` label if these cell types are indeed highly similar.

Note that there were this many total NE and NE-tumor cells identified by the separated model:
```{r}
separated_ne_df <- singler_results_separated |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  # keep only cells labeled with one of the NE categories
  dplyr::filter(stringr::str_detect(pruned.labels, "Neuroendocrine"))
nrow(separated_ne_df)
```

The table below shows counts for how many times the given label is the second-highest score:

```{r}
SingleR::plotDeltaDistribution(
  singler_results_separated,
  show = "delta.next",
  size = 0.5,
  ncol = 4
)
```

```{r}
# tidy the scores into long format
scores_df <- singler_results_separated$scores |>
  as.data.frame() |>
  dplyr::mutate(barcodes = rownames(singler_results_separated)) |>
  tidyr::pivot_longer(
    -barcodes,
    names_to = "score_label",
    values_to = "score"
  )

# This dataframe grabs the labels associated with delta.next, aka the label for the 2nd highest score
next_label_df <- scores_df |>
  dplyr::group_by(barcodes) |>
  # use -score to go in descending order and then keep only the SECOND rank
  dplyr::mutate(score_rank = rank(-score)) |>
  dplyr::filter(score_rank == 2) |>
  dplyr::ungroup() |>
  dplyr::select(barcodes, delta.next_label = score_label)

# This dataframe grabs the actual scores associated with the assigned pruned.labels
full_scores_df <- scores_df |>
  # bring in the real labels and delta.next scores
  dplyr::inner_join(
    data.frame(
      barcodes = rownames(singler_results_separated),
      singler_label = singler_results_separated$pruned.labels,
      delta.next = singler_results_separated$delta.next
    ),
    by = "barcodes"
  ) |>
  # the score whose label matches the assigned label is the score we want
  dplyr::filter(score_label == singler_label) |>
  dplyr::select(barcodes, singler_label_score = score, singler_label, delta.next) |>
  # bring in the delta.next_label
  dplyr::inner_join(
    next_label_df,
    by = "barcodes"
  ) |>
  unique() |>
  # Finally, add a column for the delta.next / singler_label_score
  dplyr::mutate(delta.next_rel_difference = (delta.next / singler_label_score))
full_scores_df |>
  dplyr::filter(delta.next_fraction > 1)

ggplot(full_scores_df) +
  aes(x = singler_label, y = delta.next_fraction) +
  geom_boxplot() +
  theme(axis.text = element_text(angle = 45))
```

```{r}
# [, "Neuroendocrine"]

next_label_df

next_label_barcodes <- next_label_df$barcodes

singler_results_separated |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::select(barcodes, pruned.labels, delta.next) |>
  ggplot() +
  aes(x = pruned.labels, y = delta.next) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90))

dplyr::add_count(delta.next_label)
# format for printing
dplyr::mutate(second_highest_delta_label = stringr::str_remove(cell_type, "scores\\.")) |>
  dplyr::arrange(desc(n)) |>
  dplyr::select(
    assigned_label = pruned.labels,
    second_highest_delta_label,
  )
```

Next we'll look at the score heatmaps to see the overall support for labels as inferred with both models.

#### Separated model SingleR score heatmap

```{r, fig.width = 10}
SingleR::plotScoreHeatmap(singler_results_separated)
```

#### Combined model SingleR score heatmap

```{r, fig.width = 10}
SingleR::plotScoreHeatmap(singler_results_combined)
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-07-08 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/reference-aggregation.Rmd

--- 2025-07-07 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/reference-aggregation.Rmd
@@
 
-# Recode some existing colors we don't need
+# For a set of consensus cell types that don't appear in samples this notebook explores
+#  we'll grab their colors to use as NBAtlas labels.
+# We'll do this by recoding cell consensus types associated with a set of colors to use, and
+#  finally discard cell type labels which won't appear in the plot.
 palette_df <- palette_df |>
   dplyr::mutate(
+    # colors associated with these original validation groups will be now used to
+    # color the given the NBAtlas cell types
     label_recoded = dplyr::case_when(

### 2025-07-09 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/separate-tumor.Rmd
towards prod

@@
 In this section, we will compare some of the marker genes identified in the trained SingleR models directly.
-
-When training a model, SingleR performs pairwise comparisons to identify marker genes as the top 10 upregulated genes for each label compared to each other label.
+When training a model as we have done here, SingleR performs pairwise comparisons to identify marker genes as the top 10 upregulated genes for each label compared to each other label.
 
@@
 
-* Marker genes upregulated in NE compared to each other cell type
-* Marker genes upregulated in NE-tumor compared to each other cell type
+* Marker genes upregulated in `NE` compared to each other cell type
+* Marker genes upregulated in `NE-tumor` compared to each other cell type
 
@@
 If the magnitude of the difference is small, it suggests that separating NE into two groups is oversplitting.
+
+In the plot, each point represents gene expression in a given cell, and the black diamonds are the mean expression for that gene.
 
@@
 
-```{r, fig.width = 12}
+```{r, message=F}
 plot_barcodes <- celltype_df |>
@@
   DataFrame(row.names = rownames(colData(sce)))
+```
 
+```{r, fig.width = 12, message=F, warning=F}
 # plot the NE-only genes
@@
   colour_by = "singler_cell_type",
-  point_size = 0.5
+  point_size = 0.5,
 ) +
+  stat_summary(shape = 18) +
   theme(legend.position = "none") +
@@
 ) +
+  stat_summary(shape = 18) +
   theme(legend.position = "none") +
@@
 
-Next, we'll look at the NE marker genes between references.
+Next, we'll look at the NE marker genes between references and ask for a quick summary:
 
-For a quick summary, we'll ask: How many of the NE markers in the combined model (out of 10 total per cell type comparison!) are _also included_ in the separated model's NE (or NE-tumor) marker gene lists?
+How many of the `NE` markers in the combined model (again, out of 10 total per cell type comparison!) are _also included_ in the separated model's `NE` (or `NE-tumor`) marker gene lists?
 
@@
     }
-  )
+  ) |>
+  as.data.frame() |>
+  t()
 ```
 
-Here, we see that NE marker genes from the combined model are essentially entirely contained within the NE and/or NE-tumor categories of the separated model.
+Here, we see that `NE` marker genes from the combined model are essentially entirely contained within the `NE` and/or `NE-tumor` categories of the separated model.
 
@@
 * It is additionally worth noting that SingleR inference took much longer with the separated than combined model
-
-This minimal exploration does not suggest that there are salient differences between the two NE categories in the separated model.
 
@@
 
-In the next section, we'll look at the numbers a bit more closely.
+In the subsections below, we'll look at the numbers behind the heatmap a bit more closely.
 
@@
 
-### Score comparisons
-
-In the separated reference, for cells that are labeled either `NE` or `NE-tumor` (considering the `pruned.labels` specifically), which label has the _second highest score_?
-We expect that the second highest score will usually be the other `NE` label if these cell types are indeed highly similar.
-
-Note that there were this many total NE and NE-tumor cells identified by the separated model:
-```{r}
-separated_ne_df <- singler_results_separated |>
-  as.data.frame() |>
-  tibble::rownames_to_column("barcodes") |>
-  # keep only cells labeled with one of the NE categories
-  dplyr::filter(stringr::str_detect(pruned.labels, "Neuroendocrine"))
-nrow(separated_ne_df)
-```
-
-The table below shows counts for how many times the given label is the second-highest score:
-
-```{r}
-SingleR::plotDeltaDistribution(
-  singler_results_separated,
-  show = "delta.next",
-  size = 0.5,
-  ncol = 4
-)
-```
-
-```{r}
-# tidy the scores into long format
-scores_df <- singler_results_separated$scores |>
-  as.data.frame() |>
-  dplyr::mutate(barcodes = rownames(singler_results_separated)) |>
-  tidyr::pivot_longer(
-    -barcodes,
-    names_to = "score_label",
-    values_to = "score"
-  )
-
-# This dataframe grabs the labels associated with delta.next, aka the label for the 2nd highest score
-next_label_df <- scores_df |>
-  dplyr::group_by(barcodes) |>
-  # use -score to go in descending order and then keep only the SECOND rank
-  dplyr::mutate(score_rank = rank(-score)) |>
-  dplyr::filter(score_rank == 2) |>
-  dplyr::ungroup() |>
-  dplyr::select(barcodes, delta.next_label = score_label)
-
-# This dataframe grabs the actual scores associated with the assigned pruned.labels
-full_scores_df <- scores_df |>
-  # bring in the real labels and delta.next scores
-  dplyr::inner_join(
-    data.frame(
-      barcodes = rownames(singler_results_separated),
-      singler_label = singler_results_separated$pruned.labels,
-      delta.next = singler_results_separated$delta.next
-    ),
-    by = "barcodes"
-  ) |>
-  # the score whose label matches the assigned label is the score we want
-  dplyr::filter(score_label == singler_label) |>
-  dplyr::select(barcodes, singler_label_score = score, singler_label, delta.next) |>
-  # bring in the delta.next_label
-  dplyr::inner_join(
-    next_label_df,
-    by = "barcodes"
-  ) |>
-  unique() |>
-  # Finally, add a column for the delta.next / singler_label_score
-  dplyr::mutate(delta.next_rel_difference = (delta.next / singler_label_score))
-full_scores_df |>
-  dplyr::filter(delta.next_fraction > 1)
-
-ggplot(full_scores_df) +
-  aes(x = singler_label, y = delta.next_fraction) +
-  geom_boxplot() +
-  theme(axis.text = element_text(angle = 45))
-```
-
-```{r}
-# [, "Neuroendocrine"]
-
-next_label_df
-
-next_label_barcodes <- next_label_df$barcodes
-
-singler_results_separated |>
-  as.data.frame() |>
-  tibble::rownames_to_column("barcodes") |>
-  dplyr::select(barcodes, pruned.labels, delta.next) |>
-  ggplot() +
-  aes(x = pruned.labels, y = delta.next) +
-  geom_boxplot() +
-  theme(axis.text.x = element_text(angle = 90))
-
-dplyr::add_count(delta.next_label)
-# format for printing
-dplyr::mutate(second_highest_delta_label = stringr::str_remove(cell_type, "scores\\.")) |>
-  dplyr::arrange(desc(n)) |>
-  dplyr::select(
-    assigned_label = pruned.labels,
-    second_highest_delta_label,
-  )
-```
-
+#### Score heatmaps
 Next we'll look at the score heatmaps to see the overall support for labels as inferred with both models.
 
-#### Separated model SingleR score heatmap
+##### Separated model SingleR score heatmap
 
@@
 
-#### Combined model SingleR score heatmap
+##### Combined model SingleR score heatmap
 
@@
 SingleR::plotScoreHeatmap(singler_results_combined)
+```
+
+#### Delta median distributions
+
+Finally, we'll look at the distribution of delta.median scores in the separated model specifically for the `NE` categories.
+
+```{r, fig.width = 6}
+SingleR::plotDeltaDistribution(
+  singler_results_separated,
+  labels.use = c("Neuroendocrine", "Neuroendocrine-tumor"),
+  size = 1
+)
 ```

### 2025-07-09 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/separate-tumor.Rmd
Apply suggestions from code review
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

@@
 separated_cell_types <- names(ne_markers)
-non_neuro_cell_types <- separated_cell_types[!(stringr::str_detect(separated_cell_types, "Neuro"))]
+non_neuro_cell_types <- separated_cell_types[!(stringr::str_detect(separated_cell_types, "Neuroendocrine"))]
 
@@
 celltype_df |>
-  dplyr::filter(stringr::str_detect(singler_label_separated, "Neuro")) |>
+  dplyr::filter(stringr::str_detect(singler_label_separated, "Neuroendocrine")) |>
   dplyr::count(singler_label_separated)
@@
 ```{r}
-combined_marker_genes <- model_combined$markers$full$Neuroendocrine |>
+ne_marker_genes <- model_separated$markers$full$Neuroendocrine |>
   purrr::reduce(c) |>
@@
 
+ ne.tumor_marker_genes <- model_separated$markers$full$`Neuroendocrine-tumor` |>
+  purrr::reduce(c) |>
+  unique()
+
+ne_only <- setdiff(ne_marker_genes, ne.tumor_marker_genes)
+ne.tumor_only <- setdiff(ne.tumor_marker_genes, ne_marker_genes) 
 cell_types <- names(ne_markers)
@@
 celltype_df |>
-  dplyr::filter(
-    singler_label_combined == "Neuroendocrine" | stringr::str_detect(singler_label_separated, "Neuroendocrine")
-  ) |>
-  dplyr::count(singler_label_combined, singler_label_separated) |>
-  dplyr::arrange(desc(n))
-```
-
-### Cells with different labels between models
-
-Next, we'll print a table that directly shows the counts of cells with different labels as inferred with either model.
-
-```{r}
-celltype_df |>
-  dplyr::filter(singler_label_combined != singler_label_separated) |>
   dplyr::count(singler_label_combined, singler_label_separated) |>

### 2025-07-09 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/separate-tumor.Rmd
remove leftover bits from other notebook and reorg so singler results are first

@@
 
-```{r url paths}
-# broad consensus cell type groups
-validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
-
-# umap colors
-palette_url <- "https://raw.githubusercontent.com/AlexsLemonade/scpca-paper-figures/refs/heads/main/palettes/validation-group-palette.tsv"
-```
-
 ```{r file paths}
@@
 
-Read SCE object so we can get consensus cell types:
+Read SCE object and define data frame of UMAP coordinates.
 
@@
 sce <- readRDS(sce_file)
-sce_df <- sce |>
-  scuttle::makePerCellDF() |>
-  dplyr::select(
-    barcodes,
-    consensus_annotation = consensus_celltype_annotation,
-    UMAP1 = UMAP.1,
-    UMAP2 = UMAP.2
-  )
+
+umap_df <- reducedDim(sce, "UMAP") |>
+  as.data.frame() |>
+  tibble::rownames_to_column("barcodes")
 ```
 
-Read cell type and result data frames:
+Read SingleR result objects:
+
 ```{r}
-validation_df <- readr::read_tsv(validation_url) |>
-  dplyr::select(consensus_annotation, validation_group_annotation)
-
-palette_df <- readr::read_tsv(palette_url) # umap plotting later
-
 singler_results_combined <- readr::read_rds(combined_singler_file)
@@
 ```{r}
-celltype_df <- data.frame(
+singler_df <- data.frame(
   barcodes = rownames(singler_results_combined),
@@
 ) |>
-  dplyr::left_join(sce_df, by = "barcodes") |>
-  dplyr::left_join(validation_df, by = "consensus_annotation") |>
-  # set validation NA -> unknown
-  dplyr::mutate(
-    validation_group_annotation =
-      ifelse(is.na(validation_group_annotation), "unknown", validation_group_annotation)
-  )
-
-# remove sce_df for space
-rm(sce_df)
+  dplyr::left_join(umap_df, by = "barcodes")
 ```
@@
 model_combined <- readRDS(model_combined_file)
+```
+
+## SingleR inference: Combined vs. separated
+
+We'll begin with a heatmap directly comparing annotations between the two SingleR models:
+
+```{r, fig.height = 8, fig.width = 8}
+make_jaccard_heatmap(
+  singler_df,
+  "singler_label_combined",
+  "singler_label_separated",
+  "Combined model (single NE category)",
+  "Separated model (NE and NE-tumor)"
+)
+```
+
+In the subsections below, we'll look at the numbers behind the heatmap a bit more closely.
+
+### Cells labeled as Neuroendocrine or Neuroendocrine-tumor
+
+We'll first directly compare cells that were labeled as a `Neuroendocrine` cell type with either model.
+Each row shows counts of cells that had given pairs corresponding labels from inference with either model.
+
+```{r}
+singler_df |>
+  dplyr::count(singler_label_combined, singler_label_separated) |>
+  dplyr::arrange(desc(n))
+```
+
+### Unlabeled cells
+
+We can also ask whether each model was able to label cells successfully, or whether more cells were ambiguous (aka `pruned.label` is `NA`) in one model or the other.
+
+How many cells were `NA` in both models?
+
+```{r}
+na_cells <- singler_df |>
+  dplyr::filter(is.na(singler_label_combined), is.na(singler_label_separated)) |>
+  dplyr::pull(barcodes)
+
+length(na_cells)
+```
+
+#### Unlabeled cells in the separated model
+
+Excluding cells which neither model labeled...
+
+* How many cells did the separated model uniquely fail to label?
+* What did the combined model label those cells?
+
+```{r}
+# if na in separated, what was it in combined?
+unlabeled_in_separated <- singler_results_separated |>
+  as.data.frame() |>
+  tibble::rownames_to_column("barcodes") |>
+  dplyr::filter(is.na(pruned.labels), !(barcodes %in% na_cells)) |>
+  dplyr::pull(barcodes)
+length(unlabeled_in_separated)
+```
+
+```{r}
+singler_df |>
+  dplyr::filter(barcodes %in% unlabeled_in_separated) |>
+  dplyr::count(singler_label_combined) |>
+  dplyr::arrange(desc(n))
+```
+
+#### Unlabeled cells in the combined model
+
+Excluding cells which neither model labeled...
+
+* How many cells did the combined model uniquely fail to label?
+* What did the separated model label those cells?
+
+```{r}
+# if na in combined, what was it in separated?
+unlabeled_in_combined <- singler_results_combined |>
+  as.data.frame() |>
+  tibble::rownames_to_column("barcodes") |>
+  dplyr::filter(is.na(pruned.labels), !(barcodes %in% na_cells)) |>
+  dplyr::pull(barcodes)
+length(unlabeled_in_combined)
+```
+
+```{r}
+singler_df |>
+  dplyr::filter(barcodes %in% unlabeled_in_combined) |>
+  dplyr::count(singler_label_separated) |>
+  dplyr::arrange(desc(n))
+```
+
+#### Score heatmaps
+Next we'll look at the score heatmaps to see the overall support for labels as inferred with both models.
+
+##### Separated model SingleR score heatmap
+
+```{r, fig.width = 10}
+SingleR::plotScoreHeatmap(singler_results_separated)
+```
+
+##### Combined model SingleR score heatmap
+
+```{r, fig.width = 10}
+SingleR::plotScoreHeatmap(singler_results_combined)
+```
+
+#### Delta median distributions
+
+Finally, we'll look at the distribution of delta.median scores in the separated model specifically for the `NE` categories.
+
+```{r, fig.width = 6}
+SingleR::plotDeltaDistribution(
+  singler_results_separated,
+  labels.use = c("Neuroendocrine", "Neuroendocrine-tumor"),
+  size = 1
+)
 ```
@@
 ```{r}
-celltype_df |>
+singler_df |>
   dplyr::filter(stringr::str_detect(singler_label_separated, "Neuroendocrine")) |>
@@
 ```{r, message=F}
-plot_barcodes <- celltype_df |>
+plot_barcodes <- singler_df |>
   dplyr::filter(stringr::str_detect(singler_label_separated, "Neuroendocrine")) |>
@@
   dplyr::inner_join(
-    celltype_df |>
+    singler_df |>
       dplyr::select(barcodes, singler_cell_type = singler_label_separated),
@@
 
-## SingleR inference: Combined vs. separated
-
-We'll begin with a heatmap directly comparing annotations between the two SingleR models:
-
-```{r, fig.height = 8, fig.width = 8}
-make_jaccard_heatmap(
-  celltype_df,
-  "singler_label_combined",
-  "singler_label_separated",
-  "Combined model (single NE category)",
-  "Separated model (NE and NE-tumor)"
-)
-```
-
-In the subsections below, we'll look at the numbers behind the heatmap a bit more closely.
-
-### Cells labeled as Neuroendocrine or Neuroendocrine-tumor
-
-We'll first directly compare cells that were labeled as a `Neuroendocrine` cell type with either model.
-Each row shows counts of cells that had given pairs corresponding labels from inference with either model.
-
-```{r}
-celltype_df |>
-  dplyr::count(singler_label_combined, singler_label_separated) |>
-  dplyr::arrange(desc(n))
-```
-
-### Unlabeled cells
-
-We can also ask whether each model was able to label cells successfully, or whether more cells were ambiguous (aka `pruned.label` is `NA`) in one model or the other.
-
-How many cells were `NA` in both models?
-
-```{r}
-na_cells <- celltype_df |>
-  dplyr::filter(is.na(singler_label_combined), is.na(singler_label_separated)) |>
-  dplyr::pull(barcodes)
-
-length(na_cells)
-```
-
-#### Unlabeled cells in the separated model
-
-Excluding cells which neither model labeled...
-
-* How many cells did the separated model uniquely fail to label?
-* What did the combined model label those cells?
-
-```{r}
-# if na in separated, what was it in combined?
-unlabeled_in_separated <- singler_results_separated |>
-  as.data.frame() |>
-  tibble::rownames_to_column("barcodes") |>
-  dplyr::filter(is.na(pruned.labels), !(barcodes %in% na_cells)) |>
-  dplyr::pull(barcodes)
-length(unlabeled_in_separated)
-```
-
-```{r}
-celltype_df |>
-  dplyr::filter(barcodes %in% unlabeled_in_separated) |>
-  dplyr::count(singler_label_combined) |>
-  dplyr::arrange(desc(n))
-```
-
-#### Unlabeled cells in the combined model
-
-Excluding cells which neither model labeled...
-
-* How many cells did the combined model uniquely fail to label?
-* What did the separated model label those cells?
-
-```{r}
-# if na in combined, what was it in separated?
-unlabeled_in_combined <- singler_results_combined |>
-  as.data.frame() |>
-  tibble::rownames_to_column("barcodes") |>
-  dplyr::filter(is.na(pruned.labels), !(barcodes %in% na_cells)) |>
-  dplyr::pull(barcodes)
-length(unlabeled_in_combined)
-```
-
-```{r}
-celltype_df |>
-  dplyr::filter(barcodes %in% unlabeled_in_combined) |>
-  dplyr::count(singler_label_separated) |>
-  dplyr::arrange(desc(n))
-```
-
-#### Score heatmaps
-Next we'll look at the score heatmaps to see the overall support for labels as inferred with both models.
-
-##### Separated model SingleR score heatmap
-
-```{r, fig.width = 10}
-SingleR::plotScoreHeatmap(singler_results_separated)
-```
-
-##### Combined model SingleR score heatmap
-
-```{r, fig.width = 10}
-SingleR::plotScoreHeatmap(singler_results_combined)
-```
-
-#### Delta median distributions
-
-Finally, we'll look at the distribution of delta.median scores in the separated model specifically for the `NE` categories.
-
-```{r, fig.width = 6}
-SingleR::plotDeltaDistribution(
-  singler_results_separated,
-  labels.use = c("Neuroendocrine", "Neuroendocrine-tumor"),
-  size = 1
-)
-```
-
 ## Session Info

### 2025-07-09 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/separate-tumor.Rmd
UMAP and update to expression plot

@@
 
-We'll begin with a heatmap directly comparing annotations between the two SingleR models:
+### UMAP
+
+```{r}
+# colors adapted from:
+# https://github.com/AlexsLemonade/scpca-paper-figures/blob/a457408323495b6ecee0c63ccdd74b41e88ab703/palettes/validation-group-palette.tsv
+nbatlas_colors <- c(
+  "B cell" = "#AA0DFE",
+  "Endothelial" = "#FC1CBF",
+  "Fibroblast" = "#B10DA1",
+  "Myeloid" = "#90AD1C",
+  "Neuroendocrine" = "#FEAF16",
+  "Neuroendocrine-tumor" = "#B17A0F",
+  "NK cell" = "#FBE426",
+  "pDC" = "#16FF32",
+  "Plasma" = "#782AB6",
+  "RBCs" = "#F6222E",
+  "Schwann" = "#1CFFCE",
+  "Stromal other" = "#C4451C",
+  "T cell" = "#3283FE"
+)
+singler_long_df <- singler_df |>
+  tidyr::pivot_longer(
+    contains("singler_label"),
+    names_to = "model",
+    values_to = "singler_label"
+  ) |>
+  dplyr::mutate(model = stringr::str_remove(model, "singler_label_"))
+```
+
+First, we'll show all cell types:
+```{r}
+# remove NAs before plotting
+ggplot(tidyr::drop_na(singler_long_df)) +
+  aes(x = UMAP1, y = UMAP2, color = singler_label) +
+  geom_point(size = 0.5, alpha = 0.5) +
+  scale_color_manual(values = nbatlas_colors) +
+  facet_wrap(vars(model)) +
+  theme(legend.position = "bottom") +
+  guides(color = guide_legend(override.aes = list(size = 1, alpha = 1)))
+```
+
+Next, we'll show this faceted only for the `NE` cell types
+```{r}
+singler_ne_df <- singler_long_df |>
+  dplyr::filter(stringr::str_detect(singler_label, "Neuroendocrine"))
+
+# remove NAs before plotting
+ggplot(singler_ne_df) +
+  aes(x = UMAP1, y = UMAP2, color = singler_label) +
+  geom_point(size = 0.75, alpha = 0.5) +
+  scale_color_manual(values = nbatlas_colors) +
+  facet_grid(
+    rows = vars(model),
+    cols = vars(singler_label)
+  ) +
+  theme(legend.position = "none")
+```
+
+### Jaccard similarity heatmap
+
+Next we'll show a heatmap directly comparing annotations between the two SingleR models colored by Jaccard similarity.
 
@@
 
-### Cells labeled as Neuroendocrine or Neuroendocrine-tumor
+#### All labels
 
-We'll first directly compare cells that were labeled as a `Neuroendocrine` cell type with either model.
-Each row shows counts of cells that had given pairs corresponding labels from inference with either model.
+The follow table shows relationship between model annotations:
 
 ```{r}
-singler_df |>
-  dplyr::count(singler_label_combined, singler_label_separated) |>
-  dplyr::arrange(desc(n))
+table(
+  singler_df$singler_label_combined,
+  singler_df$singler_label_separated
+)
 ```
 
-### Unlabeled cells
+#### Unlabeled cells
 
@@
 
-Next, we'll pull out the _distinct_ marker genes associated with `NE` and `NE-tumor` groups that are also not represented in the combined model's `NE` marker genes.
+Next, we'll pull out the _distinct_ marker genes associated with `NE` and `NE-tumor` groups.
 
@@
 ```{r}
+# Define the marker genes
 ne_marker_genes <- model_separated$markers$full$Neuroendocrine |>
@@
 ne.tumor_only <- setdiff(ne.tumor_marker_genes, ne_marker_genes)
-cell_types <- names(ne_markers)
-non_neuro_cell_types <- cell_types[!(stringr::str_detect(cell_types, "Neuro"))]
-
-different_marker_genes <- non_neuro_cell_types |>
-  purrr::set_names() |>
-  purrr::map(
-    \(cell_type) {
-      # In NE but not in NE-tumor
-      diff_genes_ne <- setdiff(
-        ne_markers[[cell_type]],
-        ne.tumor_markers[[cell_type]]
-      )
-
-      # In NE-tumor but not in NE
-      diff_genes_ne.tumor <- setdiff(
-        ne.tumor_markers[[cell_type]],
-        ne_markers[[cell_type]]
-      )
-
-      return(
-        list(
-          "ne_only" = diff_genes_ne,
-          "ne.tumor_only" = diff_genes_ne.tumor
-        )
-      )
-    }
-  ) |>
-  # transpose so it is a list of two, "ne_only" and "ne.tumor_only",
-  # each of which is a named list of marker genes for each cell type
-  purrr::transpose() |>
-  # collapse each nested list to remove cell types
-  # keep only unique genes
-  # remove any genes that are also in combined_marker_genes
-  purrr::map(
-    \(gene_list) {
-      gene_vector <- gene_list |>
-        purrr::reduce(c) |>
-        unique()
-      gene_vector[!(gene_vector %in% combined_marker_genes)]
-    }
-  )
-
-# This final list are the genes uniquely upregulated in the NE and
-# NE-tumor categories in the separated model
-different_marker_genes
 ```
@@
 ```{r, message=F}
-plot_barcodes <- singler_df |>
+neuro_barcodes <- singler_df |>
   dplyr::filter(stringr::str_detect(singler_label_separated, "Neuroendocrine")) |>
@@
 
-```{r, fig.width = 12, message=F, warning=F}
+```{r, fig.width = 14, message=F, warning=F}
 # plot the NE-only genes
 plot1 <- scater::plotExpression(
-  sce[different_marker_genes$ne_only, plot_barcodes],
-  features = different_marker_genes$ne_only,
+  sce[ne_only, neuro_barcodes],
+  features = ne_only,
   x = "singler_cell_type",
@@
   point_size = 0.5,
+  ncol = 6 # facet cols
 ) +
@@
 plot2 <- scater::plotExpression(
-  sce[different_marker_genes$ne.tumor_only, plot_barcodes],
-  features = different_marker_genes$ne.tumor_only,
+  sce[ne.tumor_only, neuro_barcodes],
+  features = ne.tumor_only,
   x = "singler_cell_type",
   colour_by = "singler_cell_type",
-  point_size = 0.5
+  point_size = 0.5,
+  ncol = 6 # facet cols
 ) +
@@
 
-plot1 + plot2
+plot1 / plot2
 ```

### 2025-07-09 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/separate-tumor.Rmd
notebook cleanups at render time

@@
 
-### UMAP
+### UMAPs
 
@@
 
-First, we'll show all cell types:
+First, we'll show a UMAP of all cell types for both the combined and separated models:
+
 ```{r}
@@
 
-Next, we'll show this faceted only for the `NE` cell types
+Next, we'll show this UMAP in a faceted view for _only_ the `NE` cell types.
+This will help us see to what extent `NE` and `NE-tumor` cells are overlapping in the separated model, and how they are laid out compared to the combined model.
+
 ```{r}
@@
 
-The follow table shows relationship between model annotations:
+The follow table shows relationship between model annotations.
+The separated model is shown in rows, and the combined model in columns.
+Each value is the count of cells with the given combination of labels.
 
@@
 table(
-  singler_df$singler_label_combined,
-  singler_df$singler_label_separated
+  singler_df$singler_label_separated,
+  singler_df$singler_label_combined
 )
@@
 # Define the marker genes
-ne_marker_genes <- model_separated$markers$full$Neuroendocrine |>
+ne_marker_genes <- ne_markers |>
   purrr::reduce(c) |>
@@
 
-ne.tumor_marker_genes <- model_separated$markers$full$`Neuroendocrine-tumor` |>
+ne.tumor_marker_genes <- ne.tumor_markers |>
   purrr::reduce(c) |>
@@
 
-```{r, fig.width = 14, message=F, warning=F}
+```{r, fig.width = 12, fig.height = 14, message=F, warning=F}
 # plot the NE-only genes
@@
   stat_summary(shape = 18) +
-  theme(legend.position = "none") +
+  theme(
+    legend.position = "none",
+    axis.text.x = element_text(angle = 45, hjust = 1)
+  ) +
   ggtitle("Neuroendocrine marker genes")
@@
   stat_summary(shape = 18) +
-  theme(legend.position = "none") +
+  theme(
+    legend.position = "none",
+    axis.text.x = element_text(angle = 45, hjust = 1)
+  ) +
   ggtitle("Neuroendocrine-tumor marker genes")
@@
 
-How many of the `NE` markers in the combined model (again, out of 10 total per cell type comparison!) are _also included_ in the separated model's `NE` (or `NE-tumor`) marker gene lists?
+How many of the `NE` markers in the combined model (again, out of 10 total per cell type comparison!) are _also included_ in the separated model's `NE` or `NE-tumor` marker gene lists?
+The table below shows the counts for how many of the combined markers appear in the given separated model's marker gene lists.
 
@@
     \(combined_marker_genes, cell_type) {
-      separated_model_markers <- c(
-        # upregulated in NE compared to cell_type
-        ne_markers[[cell_type]],
-        # upregulated in NE-tumor compared to cell-type
-        ne.tumor_markers[[cell_type]]
-      ) |> unique()
-
-      # how many combined are in that separated list?
-      sum(combined_marker_genes %in% separated_model_markers)
+      data.frame(
+        n_markers_in_NE = sum(combined_marker_genes %in% ne_markers[[cell_type]]),
+        n_markers_in_NE_tumor = sum(combined_marker_genes %in% ne.tumor_markers[[cell_type]])
+      )
     }
   ) |>
-  as.data.frame() |>
-  t()
+  purrr::list_rbind(names_to = "cell_type")
 ```

### 2025-07-09 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/explore-singler-results.Rmd
Very WIP: palette and notebook to explore singler on merged

---
title: "Explore the SingleR results"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

## Setup

```{r, warning = FALSE}
options(readr.show_col_types = FALSE)
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(SingleCellExperiment)
})

theme_set(theme_bw())

# Define color ramp for shared use in the heatmaps
heatmap_col_fun <- circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue"))
# Set heatmap padding option
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(0.6, "in"))
```

### Paths

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
results_dir <- file.path(module_dir, "results", "singler") # sample_id/aggregated/NE-tumor-combined/
data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")
```

```{r url paths}
# broad consensus cell type groups
validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
```

```{r file paths}
# SingleR files
singler_files <- list.files(
  path = results_dir,
  pattern = "_singler-annotations\\.tsv",
  recursive = TRUE,
  full.names = TRUE
) |>
  # keep only those nested in the directories `aggregated/NE-tumor-combined/`
  purrr::keep(
    \(x) {
      stringr::str_detect(x, "/aggregated/NE-tumor-combined/")
    }
  ) |>
  # add names as library id
  purrr::set_names(
    \(x) {
      stringr::str_remove(basename(x), "_singler-annotations.tsv")
    }
  )

# merged SCE file
# OBTAINED FROM PORTAL DIRECTLY on 2025-07-09
sce_file <- file.path(
  data_dir,
  "SCPCP000004_merged.rds"
)
```

### Functions

```{r}
# Source Jaccard and heatmap utilities functions:
source(file.path(module_dir, "exploratory-notebooks", "jaccard-utils.R"))

# This function calculates and plots delta median for two groups of barcodes: those in agree and those not
# This code is based on this scpca-nf code:
# https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_supplemental_report.rmd#L626
#
# df: SingleR data frame with `scores.` columns
# agree_barcodes: Vector of barcodes that have agreeing labels between aggregated and not
# plot_title: Title for plot
plot_delta_median <- function(df, agree_barcodes, plot_title) {
  # extract scores into matrix
  scores_mat <- df |>
    dplyr::select(starts_with("scores.")) |>
    as.matrix()
  rownames(scores_mat) <- df$barcodes

  # Create data frame for plotting with delta median and the full *non-pruned* cell labels
  delta_median_df <- tibble::tibble(
    delta_median = rowMaxs(scores_mat) - rowMedians(scores_mat),
    # Need to grab the non-pruned label for this plot
    full_labels = aggr$labels,
    # if pruned.labels are NA ==> low confidence
    # so, negate for this variable:
    confident = !is.na(df$pruned.labels)
  ) |>
    dplyr::mutate(
      confident = ifelse(confident, "High-quality", "Low-quality"),
      barcodes = df$barcodes,
      annotations_agree = barcodes %in% agree_barcodes
    )

  ggplot(delta_median_df) +
    aes(x = annotations_agree, y = delta_median, color = confident) +
    # don't jitter the height
    geom_jitter(height = 0, alpha = 0.7) +
    theme(legend.position = "bottom") +
    ggtitle(plot_title)
}
```

### Prepare input data

Read SCE object to get consensus cell types and UMAP coordinates:

```{r}
merged_sce <- readRDS(sce_file)
# remove assays we don't need immediately for space
assay(merged_sce, "spliced") <- NULL
assay(merged_sce, "counts") <- NULL
gc()
```

Read cell type data frames:
```{r}
singler_df <- singler_files |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind(names_to = "library_id") |>
  dplyr::select(-delta.next, -labels) |>
  dplyr::mutate(
    # add cell id column for unique rows & joining
    cell_id = glue::glue("{library_id}-{barcodes}"),
    # recode NA -> "unknown_singler"
    pruned.labels = ifelse(
      is.na(pruned.labels),
      "unknown_singler",
      pruned.labels
    )
  )

validation_df <- readr::read_tsv(validation_url) |>
  dplyr::select(consensus_annotation, validation_group_annotation)
```

Join and prepare data for use:

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("barcodes", "sample_id", "library_id", "consensus_celltype_annotation"),
  use.dimred = c("UMAP")
) |>
  # there are repeated barcodes so we need to keep cell_id around
  tibble::rownames_to_column("cell_id") |>
  dplyr::rename(
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2,
    consensus_annotation = consensus_celltype_annotation
  ) |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # recode NAs to "unknown_consensus" and remove full consensus labels
  dplyr::mutate(validation_group_annotation = ifelse(
    is.na(validation_group_annotation),
    "unknown_consensus",
    validation_group_annotation
  )) |>
  dplyr::select(-consensus_annotation) |>
  dplyr::left_join(singler_df, by = c("cell_id", "barcodes", "library_id"))
```

```{r}
unify_celltypes <- function(df, label_column, recoded_column) {
  df |>
    dplyr::mutate(
      {{ recoded_column }} := dplyr::case_when(
        # nbatlas ~ consensus
        {{ label_column }} %in% c("Myeloid", "Fibroblast") ~ stringr::str_to_lower(label),
        {{ label_column }} == "Endothelial" ~ "endothelial cell",
        {{ label_column }} == "Plasma" ~ "plasma cell",
        {{ label_column }} == "NK cell" ~ "natural killer cell",
        # make the difference stromal labels more clearly distinguishable
        {{ label_column }} == "stromal cell" ~ "stromal cell (consensus)",
        {{ label_column }} == "Stromal other" ~ "Stromal other (NBAtlas)",
        .default = {{ label_column }}
      )
    )
}

# Recode NBAtlas cell types where possible so they match with colors
celltype_recoded_df <- celltype_df |>
  # rename to make annotation sources more clear
  dplyr::rename(
    "consensus_validation" = validation_group_annotation,
    "singler" = pruned.labels
  ) |>
  # pivot longer for wrangling
  tidyr::pivot_longer(
    c(consensus_validation, singler),
    names_to = "annotation_type",
    values_to = "label"
  ) |>
  unify_celltypes(label, label_recoded) |>
  dplyr::select(-label)

palette_df <- readr::read_tsv("../unified-cell-type-palette.tsv")
setdiff(unique(celltype_recoded_df$label_recoded), palette_df$unified_label)
```

```{r, fig.height = 10, fig.width = 10} 
# Pivot wider for heatmap functions
celltype_recoded_wide_df <- celltype_recoded_df |>
  tidyr::pivot_wider(
    names_from = annotation_type,
    values_from = label_recoded
  )

heatmap_col_fun <- circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue"))

make_jaccard_heatmap(
  celltype_recoded_wide_df,
  "consensus_validation",
  "singler",
  "Consensus validation label",
  "SingleR NBAtlas label"
)
```

```{r, fig.width = 12, fig.height = 8}
celltype_pal <- palette_df$hex_color
names(celltype_pal) <- palette_df$unified_label

ggplot(celltype_recoded_df) +
  aes(x = UMAP1, y = UMAP2, color = label_recoded) +
  geom_point(size = 0.25, alpha = 0.5) +
  scale_color_manual(
    values = celltype_pal,
    name = "Harmonized cell types"
  ) +
  facet_wrap(vars(annotation_type)) +
  coord_equal() +
  theme(
    legend.position = "bottom",
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
```

Let's facet just the SingleR annotations:

```{r}
# Adapted from scpca-nf
# https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_qc.rmd#L134-L221

#' Create a faceted UMAP panel where each panel has only one cell type colored
#'
#' @param umap_df Data frame with UMAP1 and UMAP2 columns
#' @param annotation_column Column containing broad cell type annotation
#' @param celltype_colors Named vector of colors to use for each broader validation group
#' @param point_size Point size. Default is 1
#'
#' @return ggplot object containing a faceted UMAP where each cell type is a facet.
#'   In each panel, the cell type of interest is colored red and all other cells are grey.
faceted_umap <- function(umap_df,
                         annotation_column,
                         celltype_colors,
                         point_size = 1) {
  # color by the annotation column but only color one cell type at a time
  faceted_umap <- ggplot(
    umap_df,
    aes(x = UMAP1, y = UMAP2, color = {{ annotation_column }})
  ) +
    # set points for all "other" points
    geom_point(
      data = dplyr::select(
        umap_df, -{{ annotation_column }}
      ),
      color = "gray90",
      alpha = 0.5,
      size = point_size
    ) +
    scale_color_manual(values = celltype_colors) +
    # set points for desired cell type
    geom_point(size = point_size, alpha = 0.5) +
    facet_wrap(
      vars({{ annotation_column }}),
      ncol = 5
    ) +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    theme_bw() +
    theme(
      legend.position = "none",
      aspect.ratio = 1,
      strip.background = element_rect(fill = "transparent")
    )

  return(faceted_umap)
}
```

```{r fig.width = 20}
singler_recoded_df <- celltype_recoded_df |>
  dplyr::filter(annotation_type == "singler")

faceted_umap(
  singler_recoded_df,
  label_recoded,
  celltype_pal
)
```

For cells that directly match up with consensus:

```{r}
direct_celltype_matches <- c(
  "B cell",
  "T cell",
  "myeloid",
  "fibroblast",
  "endothelial cell",
  "plasma cell",
  "natural killer cell",
  "unknown_consensus" # _for this plot_, we'll recode the Neuroendocrine --> "unknown consensus" for the purposes of comparison
)
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-07-09 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/reference-aggregation.Rmd

@@
   "aggregated",
+  "NE-tumor-combined",
   glue::glue("{params$library_id}_singler-results.rds")
@@
   "not-aggregated",
+  "NE-tumor-combined",
   glue::glue("{params$library_id}_singler-results.rds")

### 2025-07-10 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/separate-tumor.Rmd
update directories

@@
 module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
-results_dir <- file.path(module_dir, "results", "singler")
+results_dir <- file.path(module_dir, "results")
+combined_results_dir <- file.path(results_dir, "singler-test_combined-tumor-ne")
+separated_results_dir <- file.path(results_dir, "singler-test_separated-tumor-ne")
 data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")
@@
 combined_singler_file <- file.path(
-  results_dir,
+  combined_results_dir,
   params$sample_id,
-  "aggregated",
-  "NE-tumor-combined",
   glue::glue("{params$library_id}_singler-results.rds")
@@
 separated_singler_file <- file.path(
-  results_dir,
+  separated_results_dir,
   params$sample_id,
-  "aggregated",
-  "NE-tumor-separated",
   glue::glue("{params$library_id}_singler-results.rds")

### 2025-07-10 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/explore-singler-results.Rmd
Continued WIP notebook development

@@
 singler_files <- list.files(
-  path = results_dir,
+  path = "/Users/sjspielman/ALSF/open-scpca/OpenScPCA-analysis/analyses/cell-type-neuroblastoma-04/results/singler-test_aggregated",
   pattern = "_singler-annotations\\.tsv",
@@
   full.names = TRUE
-) |>
-  # keep only those nested in the directories `aggregated/NE-tumor-combined/`
-  purrr::keep(
-    \(x) {
-      stringr::str_detect(x, "/aggregated/NE-tumor-combined/")
-    }
 ) |>
@@
 merged_sce <- readRDS(sce_file)
-# remove assays we don't need immediately for space
+# immediately remove assays we don't need for space
 assay(merged_sce, "spliced") <- NULL
@@
   dplyr::select(consensus_annotation, validation_group_annotation)
+
+####### TEMPORARY!
+merged_sce <- merged_sce[, singler_df$cell_id]
+merged_sce
 ```
@@
 ```{r}
-unify_celltypes <- function(df, label_column, recoded_column) {
+harmonize_celltypes <- function(df, label_column, recoded_column) {
   df |>
@@
   ) |>
-  unify_celltypes(label, label_recoded) |>
+  harmonize_celltypes(label, label_recoded) |>
   dplyr::select(-label)
 
-palette_df <- readr::read_tsv("../unified-cell-type-palette.tsv")
-setdiff(unique(celltype_recoded_df$label_recoded), palette_df$unified_label)
+palette_df <- readr::read_tsv("../harmonized-cell-type-palette.tsv")
+celltype_pal <- palette_df$hex_color
+names(celltype_pal) <- palette_df$harmonized_label
 ```
@@
 ```{r, fig.width = 12, fig.height = 8}
-celltype_pal <- palette_df$hex_color
-names(celltype_pal) <- palette_df$unified_label
 
@@
 #' @param celltype_colors Named vector of colors to use for each broader validation group
-#' @param point_size Point size. Default is 1
 #'
@@
                          annotation_column,
+                         facet_column,
                          celltype_colors,
-                         point_size = 1) {
+                         facet_ncol = 4, 
+                         single_celltype = NULL) {
+
+  if (!(is.null(single_celltype))) {
+    first_df <- umap_df |>
+      dplyr::filter({{annotation_column}} == single_celltype)
+    plot_title <- single_celltype
+  } else {
+    first_df <- umap_df
+    plot_title <- ""
+  }
+
   # color by the annotation column but only color one cell type at a time
   faceted_umap <- ggplot(
-    umap_df,
+    first_df,
     aes(x = UMAP1, y = UMAP2, color = {{ annotation_column }})
@@
       alpha = 0.5,
-      size = point_size
+      size = 1
     ) +
@@
     # set points for desired cell type
-    geom_point(size = point_size, alpha = 0.5) +
+    geom_point(size = 1, alpha = 0.5) +
     facet_wrap(
-      vars({{ annotation_column }}),
-      ncol = 5
+      vars({{ facet_column }}),
+      ncol = facet_ncol
     ) +
-    # remove axis numbers and background grid
-    scale_x_continuous(labels = NULL, breaks = NULL) +
-    scale_y_continuous(labels = NULL, breaks = NULL) +
+    coord_equal() +
     theme_bw() +
@@
       legend.position = "none",
-      aspect.ratio = 1,
-      strip.background = element_rect(fill = "transparent")
-    )
+      axis.ticks = element_blank(), 
+      axis.text = element_blank()
+    ) +
+    ggtitle(plot_title)
 
@@
 
-```{r fig.width = 20}
-singler_recoded_df <- celltype_recoded_df |>
-  dplyr::filter(annotation_type == "singler")
-
+```{r fig.width = 14}
+celltype_recoded_df |>
+  dplyr::filter(annotation_type == "singler") |>
   faceted_umap(
-  singler_recoded_df,
-  label_recoded,
-  celltype_pal
+    annotation_column = label_recoded,
+    facet_column = label_recoded,
+    celltype_colors = celltype_pal
   )
@@
 
-```{r}
+```{r, fig.width = 30, fig.asp = 1}
 direct_celltype_matches <- c(
@@
   "natural killer cell",
-  "unknown_consensus" # _for this plot_, we'll recode the Neuroendocrine --> "unknown consensus" for the purposes of comparison
+  # we'll use this label to be able to directly compare unknown_consensus to Neuroendocrine
+  "putative-tumor" 
 )
+
+celltype_facet_df <- celltype_recoded_df |>
+  # recode so Neuroendocrine matches with unknown_consensus in the plot
+  dplyr::mutate(
+    label_recoded = ifelse(
+      label_recoded %in% c("Neuroendocrine", "unknown_consensus"),
+      "putative-tumor",
+      label_recoded
+    )
+  ) |>
+  dplyr::filter(label_recoded %in% direct_celltype_matches)
+
+unique(celltype_facet_df$label_recoded) |>
+  purrr::map(
+    \(cell_type) {
+      faceted_umap(
+        celltype_facet_df,
+        annotation_column = label_recoded, 
+        celltype_colors = celltype_pal,
+        facet_column = annotation_type,
+        facet_ncol = 2, 
+        single_celltype = cell_type
+      )
+  }) |>
+  patchwork::wrap_plots(ncol = 1)
+
 ```
+
+## Dotplot time
+
+This is copied from Ally's code and is NOT YET ADAPTED TO THIS NOTEBOOK.
+```{r}
+  # prep for plots
+  # get total number of cells per final annotation group
+  total_cells_df <- consensus_df |>
+    dplyr::select(unique_id, barcodes, broad_celltype_group) |>
+    dplyr::distinct() |>
+    dplyr::count(broad_celltype_group, name = "total_cells")
+
+  # table with one row per unique broad cell type/ marker gene combination
+  # first all cells in with the same broad_celltype_group (determined based on consensus_annotation) are grouped together
+  # then get the mean gene expression and total percentage of cells that express each marker gene across all cells in that group
+  # do this for every possible marker gene/ validation group assignment
+  # second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
+  group_stats_df <- consensus_df |>
+    # for each assigned cell type/marker gene combo get total detected and mean expression
+    # group by both broad group and validation group to account for genes that are expressed in more than one cell type
+    dplyr::summarize(
+      .by = c("broad_celltype_group", "ensembl_gene_id", "validation_group_annotation"),
+      detected_count = sum(detected),
+      mean_exp = mean(logcounts)
+    ) |>
+    # add in validation group for marker genes
+    # this includes all possible marker genes and all possible validation group assignments
+    dplyr::left_join(markers_df, by = c("ensembl_gene_id", "validation_group_annotation"), relationship = "many-to-many") |> # add total cells
+    dplyr::left_join(total_cells_df, by = c("broad_celltype_group")) |>
+    # only show cells that have above the minimum total number
+    dplyr::filter(total_cells > min_cells) |>
+    dplyr::mutate(
+      # get total percent expressed
+      percent_exp = (detected_count / total_cells) * 100,
+      # account for NA/unknowns and set axes order
+      broad_celltype_group = factor(broad_celltype_group, levels = c(names(celltype_colors)))
+    )
+
+  # get list of celltypes to keep and assign colors
+  celltype_groups <- group_stats_df |>
+    dplyr::pull(broad_celltype_group) |>
+    unique() |>
+    as.character()
+
+  # filter markers to those that are actually relevant
+  # we will only plot the marker genes for cell types that are part of the assigned broad validation group for this group of samples
+  # we don't care about plotting marker genes for cell types that aren't present here
+  filtered_markers_df <- markers_df |>
+    dplyr::filter(
+      validation_group_annotation %in% celltype_groups,
+      gene_symbol %in% group_stats_df$gene_symbol
+    ) |>
+    # ensure order matches the order of the legend
+    dplyr::mutate(
+      validation_group_annotation = factor(validation_group_annotation, levels = names(celltype_colors))
+    ) |>
+    dplyr::arrange(validation_group_annotation)
+
+  # specify x axis order for dotplot
+  marker_gene_order <- filtered_markers_df |>
+    dplyr::pull(gene_symbol)
+
+  # set order for cell types
+  celltype_order <- unique(filtered_markers_df$validation_group_annotation)
+
+  # filter out low expressed genes
+  dotplot_df <- group_stats_df |>
+    dplyr::filter(mean_exp > 0, percent_exp > 10) |>
+    dplyr::arrange(broad_celltype_group) |>
+    # add a label for the plot
+    dplyr::mutate(y_label = glue::glue("{broad_celltype_group} ({total_cells})")) |>
+    # remove marker genes that aren't present in final annotations and set x axis order
+    dplyr::filter(gene_symbol %in% marker_gene_order) |>
+    dplyr::mutate(
+      # set orders of gene symbol and validation groups
+      y_label = factor(y_label, levels = rev(unique(y_label))),
+      gene_symbol = factor(gene_symbol, levels = marker_gene_order),
+      validation_group_annotation = factor(validation_group_annotation, levels = celltype_order)
+    )
+
+  # make dotplot with marker gene exp
+  dotplot <- ggplot(dotplot_df, aes(y = y_label, x = gene_symbol, color = mean_exp, size = percent_exp)) +
+    geom_point() +
+    scale_color_viridis_c(option = "magma") +
+    facet_grid(cols = vars(validation_group_annotation), scales = "free", space = "free") +
+    theme_classic() +
+    theme(
+      strip.background = element_rect(fill = "transparent", color = NA),
+      strip.placement = "outside",
+      strip.text.x = element_blank(),
+      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
+      axis.ticks.x = element_blank(),
+      text = element_text(size = 14),
+      panel.spacing = unit(0.5, "lines") # adjust spacing and match with annotation bar
+    ) +
+    labs(
+      x = "",
+      y = "Broad cell type annotation",
+      color = "Mean gene expression",
+      size = "Percent cells expressed"
+    )
+
+  # add annotation bar aligning marker genes with validation group
+  color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = validation_group_annotation)) +
+    geom_tile() +
+    facet_grid(cols = vars(validation_group_annotation), scales = "free", space = "free") +
+    scale_fill_manual(values = celltype_colors, breaks = levels(dotplot_df$validation_group_annotation)) +
+    ggmap::theme_nothing() +
+    theme(
+      strip.background = element_rect(fill = "transparent", color = NA),
+      strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 12),
+      strip.placement = "outside",
+      legend.position = "none",
+      panel.spacing = unit(0.5, "lines"),
+      strip.clip = "off"
+    ) +
+    labs(fill = "")
+
+  combined_plot <- color_bar / dotplot +
+    patchwork::plot_layout(ncol = 1, heights = c(0.1, 4))
 

### 2025-07-10 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/filter-genes.Rmd
Exploratory notebook for filter genes testing

---
title: "Explore gene filtering strategy"
subtitle: "`r glue::glue('Results for {params$sample_id}: {params$library_id}')`"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  sample_id: "SCPCS000109"
  library_id: "SCPCL000127"
---

The goal of this notebook is to assess `SingleR` results inferred with two different models constructed from NBAtlas:

- A model where ribosomal and mitochondrial genes were filtered out of the reference before building the model
  - This means those genes will not come up as marker genes for cell type annotation
- A models where no genes are filtered out
  - There means there is potential for those genes to come up as marker genes for cell type annotation

This notebook compares inferences made with either model to determine whether this filtering makes an empirical difference.
It also compares marker genes in the underlying models directly to get a sense of how much the filtering influenced model training.

## Setup

```{r, warning = FALSE}
options(readr.show_col_types = FALSE)

suppressWarnings({
  suppressPackageStartupMessages({
    library(ggplot2)
    library(patchwork)
    library(SingleCellExperiment)
  })
})

theme_set(theme_bw())

# Define color ramp for shared use in the heatmaps
heatmap_col_fun <- circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue"))
# Set heatmap padding option
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(0.6, "in"))
```

### Paths

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
results_dir <- file.path(module_dir, "results")
filtered_results_dir <- file.path(results_dir, "singler-test_filter-genes-TRUE")
unfiltered_results_dir <- file.path(results_dir, "singler-test_filter-genes-FALSE")
data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")
```

```{r file paths}
# SingleR results
filtered_singler_file <- file.path(
  filtered_results_dir,
  params$sample_id,
  glue::glue("{params$library_id}_singler-results.rds")
)

unfiltered_singler_file <- file.path(
  unfiltered_results_dir,
  params$sample_id,
  glue::glue("{params$library_id}_singler-results.rds")
)

# SCE file
sce_file <- file.path(
  data_dir,
  params$sample_id,
  glue::glue("{params$library_id}_processed.rds")
)

# broad consensus cell type groups
validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
```

### Functions

```{r}
# Source Jaccard and heatmap utilities functions:
source(file.path(module_dir, "exploratory-notebooks", "jaccard-utils.R"))
```

### Prepare input data

Read SCE object and define data frame of UMAP coordinates.

```{r}
validation_df <- readr::read_tsv(validation_url) |>
  dplyr::select(consensus_annotation, validation_group_annotation)

sce <- readRDS(sce_file)

sce_df <- sce |>
  scuttle::makePerCellDF(
    use.coldata = c("barcodes", "consensus_celltype_annotation"),
    use.dimred = c("UMAP")
  ) |>
  dplyr::rename(
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2,
    consensus_annotation = consensus_celltype_annotation
  )
```

Read SingleR result objects:

```{r}
singler_results_filtered <- readr::read_rds(filtered_singler_file)
singler_results_unfiltered <- readr::read_rds(unfiltered_singler_file)
```

Read SingleR models:

```{r}
model_filtered <- readRDS(
  file.path(
    module_dir,
    "scratch",
    "singler-model_nbatlas_filter-genes-TRUE.rds"
  )
)
model_unfiltered <- readRDS(
  file.path(
    module_dir,
    "scratch",
    "singler-model_nbatlas_filter-genes-FALSE.rds"
  )
)
```

Join and prepare data for use:

```{r}
singler_df <- data.frame(
  barcodes = rownames(singler_results_filtered),
  "singler_label_filtered" = as.data.frame(singler_results_filtered)$pruned.labels,
  "singler_label_unfiltered" = as.data.frame(singler_results_unfiltered)$pruned.labels
) |>
  dplyr::left_join(sce_df, by = "barcodes") |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  dplyr::mutate(
    validation_group_annotation = ifelse(
      is.na(validation_group_annotation),
      "unknown",
      validation_group_annotation
    )
  )
```

## Filtered vs unfiltered models

### UMAP

We'll display their annotations side-by-side as a UMAP.
Cells labeled `NA` are those that SingleR could not confidently label.

```{r}
# colors adapted from:
# https://github.com/AlexsLemonade/scpca-paper-figures/blob/a457408323495b6ecee0c63ccdd74b41e88ab703/palettes/validation-group-palette.tsv
nbatlas_colors <- c(
  "B cell" = "#AA0DFE",
  "Endothelial" = "#FC1CBF",
  "Fibroblast" = "#B10DA1",
  "Myeloid" = "#90AD1C",
  "Neuroendocrine" = "#FEAF16",
  "Neuroendocrine-tumor" = "#B17A0F",
  "NK cell" = "#FBE426",
  "pDC" = "#16FF32",
  "Plasma" = "#782AB6",
  "RBCs" = "#F6222E",
  "Schwann" = "#1CFFCE",
  "Stromal other" = "#C4451C",
  "T cell" = "#3283FE"
)
singler_long_df <- singler_df |>
  tidyr::pivot_longer(
    contains("singler_label"),
    names_to = "model",
    values_to = "singler_label"
  ) |>
  dplyr::mutate(model = stringr::str_remove(model, "singler_label_"))

ggplot(singler_long_df) +
  aes(x = UMAP1, y = UMAP2, color = singler_label) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_manual(values = nbatlas_colors) +
  facet_wrap(vars(model)) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(override.aes = list(size = 1, alpha = 1)))
```

### Heatmap

We'll display the relationship between annotations as a heatmap colored by Jaccard similarity:

```{r, fig.height = 8, fig.width = 8}
make_jaccard_heatmap(
  singler_df,
  "singler_label_filtered",
  "singler_label_unfiltered",
  "Gene filtering ON",
  "Gene filtering OFF"
)
```

### Confusion matrix

Below we display the confusion matrix and associated statistics comparing these annotations.
Annotations from the unfiltered genes model are shown as "Reference," and annotations from the filtered genes model are shown as "Prediction."

Note that some cell type names have been abbreviated in this output to make it more clearly visible:

* Endothelial --> Endo
* Fibroblasts --> Fibro
* Neuroendocrine --> NE

```{r}
singler_long_df <- singler_long_df |>
  # shorten labels
  dplyr::mutate(
    short_label = dplyr::case_when(
      singler_label == "Endothelial" ~ "Endo",
      singler_label == "Fibroblast" ~ "Fibro",
      singler_label == "Neuroendocrine" ~ "NE",
      .default = singler_label
    )
  )

caret::confusionMatrix(
  factor(singler_long_df$short_label[singler_long_df$model == "filtered"]), # data
  factor(singler_long_df$short_label[singler_long_df$model == "unfiltered"]), # reference
  mode = "sens_spec"
)
```

### Unlabeled cells

We can also ask whether each model was able to label cells successfully, or whether more cells were ambiguous (aka `pruned.label` is `NA`) in one model or the other.

How many cells were `NA` in both models?

```{r}
na_cells <- singler_df |>
  dplyr::filter(is.na(singler_label_filtered), is.na(singler_label_unfiltered)) |>
  dplyr::pull(barcodes)

length(na_cells)
```

#### Unlabeled cells in the filtered genes model

Excluding cells which neither model labeled...

* How many cells did the filtered genes model uniquely fail to label?
* What did the unfiltered genes model label those cells?

```{r}
# if na in separated, what was it in combined?
unlabeled_in_filtered <- singler_results_filtered |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::filter(is.na(pruned.labels), !(barcodes %in% na_cells)) |>
  dplyr::pull(barcodes)
length(unlabeled_in_filtered)
```

```{r}
singler_df |>
  dplyr::filter(barcodes %in% unlabeled_in_filtered) |>
  dplyr::count(singler_label_unfiltered) |>
  dplyr::arrange(desc(n))
```

#### Unlabeled cells in the unfiltered model

Excluding cells which neither model labeled...

* How many cells did the unfiltered model uniquely fail to label?
* What did the filtered model label those cells?

```{r}
# if na in unfiltered, what was it in separated?
unlabeled_in_unfiltered <- singler_results_unfiltered |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::filter(is.na(pruned.labels), !(barcodes %in% na_cells)) |>
  dplyr::pull(barcodes)
length(unlabeled_in_unfiltered)
```

```{r}
singler_df |>
  dplyr::filter(barcodes %in% unlabeled_in_unfiltered) |>
  dplyr::count(singler_label_filtered) |>
  dplyr::arrange(desc(n))
```

## Comparison with consensus cell type annotations

In this section, we compare filtered & unfiltered annotations to the consensus cell type labels.
In these comparisons, we consider the broader validation groupings for consensus cell types in order to keep plots manageable and more easily interpreted.

### Heatmap: Filtered vs. consensus annotations

This heatmap compares annotations between `SingleR` inference using the filtered reference and consensus cell types.

```{r, fig.height = 8, fig.width = 8}
make_jaccard_heatmap(
  singler_df,
  "validation_group_annotation",
  "singler_label_filtered",
  "OpenScPCA Consensus cell type validation group",
  "Gene filtering ON"
)
```

### Heatmap: Unfiltered vs. consensus annotations

This heatmap compares annotations between `SingleR` inference using the unfiltered reference and consensus cell types.

```{r, fig.height = 8, fig.width = 8}
make_jaccard_heatmap(
  singler_df,
  "validation_group_annotation",
  "singler_label_unfiltered",
  "OpenScPCA Consensus cell type validation group",
  "Gene filtering OFF"
)
```

## Compare trained SingleR models

In this section, we will compare the marker genes between filtered and unfiltered models.
Output from this section will be same in all rendered notebooks since it is data-agnostic and focuses only on the trained SingleR models.

```{r}
filtered_markers <- model_filtered$markers$unique
unfiltered_markers <- model_unfiltered$markers$unique
```

First, are there any mitochondrial or ribosomal marker genes in the unfiltered model, and if so what are they?

```{r}
ribo_mito <- union(
  grep("^MT-", unfiltered_markers, value = TRUE),
  grep("^RP[SL]", unfiltered_markers, value = TRUE)
)
length(ribo_mito)
ribo_mito
```
There are quite a few ribosomal genes, but no mitochondrial genes.

Which cell types do these marker genes come up for?

```{r}
model_unfiltered$markers$full |>
  purrr::map(
    \(inner_list) {
      purrr::reduce(inner_list, c)
    }
  ) |>
  purrr::map(
    \(all_marker_genes) {
      if (any(ribo_mito %in% all_marker_genes)) {
        intersect(ribo_mito, all_marker_genes)
      }
    }
  ) |>
  # remove all the NULLs
  purrr::compact()
```

These marker genes are _only_ associated with the following cell types, and of these predominantly B-cells and T-cells.

* B cells
* Fibroblasts
* Natural killer cells
* Plasmacytoid dendritic cells (pDCs)
* T cells

What genes do these models have in common, out of the total number of marker genes in each model?

* Total filtered markers:
```{r}
length(filtered_markers)
```

* Total unfiltered markers:
```{r}
length(unfiltered_markers)
```

* Number of intersecting markers:
```{r}
intersect(
  filtered_markers,
  unfiltered_markers
) |> length()
```
* Non-intersecting markers _aside from_ ribosomal genes:
```{r}
diff_markers <- c(
  setdiff(filtered_markers, unfiltered_markers),
  setdiff(unfiltered_markers, filtered_markers)
)
diff_markers <- diff_markers[!(diff_markers %in% ribo_mito)]
diff_markers
```

There are nine different markers here. 
So, now that the ribosomal genes weren't present to serve as markers, these genes were apparently picked up.

```{r}
any(diff_markers %in% filtered_markers)
any(diff_markers %in% unfiltered_markers)
```

They are _all_ part of the `filtered_markers` reference.
Which cell type do those come up for?

```{r}
model_filtered$markers$full |>
  purrr::map(
    \(inner_list) {
      purrr::reduce(inner_list, c)
    }
  ) |>
  purrr::map(
    \(all_marker_genes) {
      if (any(diff_markers %in% all_marker_genes)) {
        intersect(diff_markers, all_marker_genes)
      }
    }
  ) |> # remove all the NULLs
  purrr::compact()
```

They are in B-cells, Fibroblasts, and T-cells, all of which were cell types with ribosomal marker genes coming up in the unfiltered model.

### Conclusions from marker gene comparisons

* Filtering genes leads to ribosomal genes no longer being selected as marker genes in the trained model
* Ribosomal genes were most commonly observed as B-cell and T-cell marker genes
* Those ribosomal genes aside, the two models identified essentially the same overall marker genes

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-07-10 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/reference-aggregation.Rmd
move jaccard-utils.R into scripts. update docs and path in notebooks to match

--- 2025-07-08 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/reference-aggregation.Rmd
@@
 module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
-results_dir <- file.path(module_dir, "results", "singler")
+results_dir <- file.path(module_dir, "results")
+aggr_results_dir <- file.path(results_dir, "singler-test_aggregated")
+non_aggr_results_dir <- file.path(results_dir, "singler-test_not-aggregated")
 data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")
@@
 aggr_singler_file <- file.path(
-  results_dir,
+  aggr_results_dir,
   params$sample_id,
-  "aggregated",
   glue::glue("{params$library_id}_singler-results.rds")
@@
 non_aggr_singler_file <- file.path(
-  results_dir,
+  non_aggr_results_dir,
   params$sample_id,
-  "not-aggregated",
   glue::glue("{params$library_id}_singler-results.rds")
@@
 # Source Jaccard and heatmap utilities functions:
-source(file.path(module_dir, "exploratory-notebooks", "jaccard-utils.R"))
+source(file.path(module_dir, "scripts", "utils", "jaccard-utils.R"))
 

### 2025-07-10 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/explore-singler-results.Rmd
WIP continues: UMAPs are cleaned up and we have a baseline dotplot

--- 2025-07-09 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/explore-singler-results.Rmd
@@
 module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
-results_dir <- file.path(module_dir, "results", "singler") # sample_id/aggregated/NE-tumor-combined/
+results_dir <- file.path(module_dir, "results", "singler-filtered-genes")
 data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")
-```
-
-```{r url paths}
-# broad consensus cell type groups
-validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
 ```
@@
 )
+
+# broad consensus cell type groups
+validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
+
+# palette file
+palette_file <- file.path(
+  module_dir,
+  "palettes",
+  "harmonized-cell-type-palette.tsv"
+)
 ```
@@
 # Source Jaccard and heatmap utilities functions:
-source(file.path(module_dir, "exploratory-notebooks", "jaccard-utils.R"))
+source(file.path(module_dir, "scripts", "utils", "jaccard-utils.R"))
+```
 
-# This function calculates and plots delta median for two groups of barcodes: those in agree and those not
-# This code is based on this scpca-nf code:
-# https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_supplemental_report.rmd#L626
-#
-# df: SingleR data frame with `scores.` columns
-# agree_barcodes: Vector of barcodes that have agreeing labels between aggregated and not
-# plot_title: Title for plot
-plot_delta_median <- function(df, agree_barcodes, plot_title) {
-  # extract scores into matrix
-  scores_mat <- df |>
-    dplyr::select(starts_with("scores.")) |>
-    as.matrix()
-  rownames(scores_mat) <- df$barcodes
+```{r}
+harmonize_celltypes <- function(df, label_column, recoded_column) {
+  df |>
+    dplyr::mutate(
+      {{ recoded_column }} := dplyr::case_when(
+        # nbatlas ~ consensus
+        {{ label_column }} %in% c("Myeloid", "Fibroblast") ~ stringr::str_to_lower({{ label_column }}),
+        {{ label_column }} == "Endothelial" ~ "endothelial cell",
+        {{ label_column }} == "Plasma" ~ "plasma cell",
+        {{ label_column }} == "NK cell" ~ "natural killer cell",
+        # make the difference stromal labels more clearly distinguishable
+        {{ label_column }} == "stromal cell" ~ "stromal cell (consensus)",
+        {{ label_column }} == "Stromal other" ~ "Stromal other (NBAtlas)",
+        .default = {{ label_column }}
+      )
+    )
+}
+```
 
-  # Create data frame for plotting with delta median and the full *non-pruned* cell labels
-  delta_median_df <- tibble::tibble(
-    delta_median = rowMaxs(scores_mat) - rowMedians(scores_mat),
-    # Need to grab the non-pruned label for this plot
-    full_labels = aggr$labels,
-    # if pruned.labels are NA ==> low confidence
-    # so, negate for this variable:
-    confident = !is.na(df$pruned.labels)
-  ) |>
-    dplyr::mutate(
-      confident = ifelse(confident, "High-quality", "Low-quality"),
-      barcodes = df$barcodes,
-      annotations_agree = barcodes %in% agree_barcodes
+```{r}
+# Adapted from scpca-nf
+# https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_qc.rmd#L134-L221
+
+#' Create a faceted UMAP panel where each panel has only one cell type colored
+#'
+#' @param umap_df Data frame with UMAP1 and UMAP2 columns
+#' @param annotation_column Column containing broad cell type annotation
+#' @param celltype_colors Named vector of colors to use for each broader validation group
+#' @param facet_type Whether to use facet_wrap or facet_grid
+#' @param annotation_type_column Additional column to use if facet_type is "grid"
+#'
+#' @return ggplot object containing a faceted UMAP where each cell type is a facet.
+#'   In each panel, the cell type of interest is colored red and all other cells are grey.
+faceted_umap <- function(umap_df,
+                         annotation_column,
+                         celltype_colors,
+                         facet_type = c("wrap", "grid"),
+                         annotation_type_column = NULL) {
+  facet_type <- match.arg(facet_type)
+
+  # color by the annotation column but only color one cell type at a time
+  faceted_umap <- ggplot(
+    umap_df,
+    aes(x = UMAP1, y = UMAP2, color = {{ annotation_column }})
+  ) +
+    # set points for all "other" points
+    geom_point(
+      data = dplyr::select(
+        umap_df, -{{ annotation_column }}
+      ),
+      color = "gray90",
+      alpha = 0.5,
+      size = 0.5
+    ) +
+    scale_color_manual(values = celltype_colors) +
+    # set points for desired cell type
+    geom_point(size = 0.5, alpha = 0.5) +
+    coord_equal() +
+    theme_bw() +
+    theme(
+      legend.position = "none",
+      axis.ticks = element_blank(),
+      axis.text = element_blank()
     )
 
-  ggplot(delta_median_df) +
-    aes(x = annotations_agree, y = delta_median, color = confident) +
-    # don't jitter the height
-    geom_jitter(height = 0, alpha = 0.7) +
-    theme(legend.position = "bottom") +
-    ggtitle(plot_title)
+  # add faceting as specified
+  if (facet_type == "wrap") {
+    faceted_umap <- faceted_umap +
+      facet_wrap(
+        vars({{ annotation_column }}),
+        ncol = 4
+      )
+  } else {
+    faceted_umap <- faceted_umap +
+      facet_grid(
+        rows = vars({{ annotation_column }}),
+        cols = vars({{ annotation_type_column }})
+      )
+  }
+  return(faceted_umap)
 }
@@
 gc()
+
+# RUH ROH!!!
+# https://github.com/AlexsLemonade/scpcaTools/issues/298
+merged_sce$cell_id <- colnames(merged_sce)
 ```
@@
   dplyr::select(consensus_annotation, validation_group_annotation)
+
+# set up palette for umaps
+palette_df <- readr::read_tsv(palette_file)
+celltype_pal <- palette_df$hex_color
+names(celltype_pal) <- palette_df$harmonized_label
 ```
@@
 ```{r}
-unify_celltypes <- function(df, label_column, recoded_column) {
-  df |>
-    dplyr::mutate(
-      {{ recoded_column }} := dplyr::case_when(
-        # nbatlas ~ consensus
-        {{ label_column }} %in% c("Myeloid", "Fibroblast") ~ stringr::str_to_lower(label),
-        {{ label_column }} == "Endothelial" ~ "endothelial cell",
-        {{ label_column }} == "Plasma" ~ "plasma cell",
-        {{ label_column }} == "NK cell" ~ "natural killer cell",
-        # make the difference stromal labels more clearly distinguishable
-        {{ label_column }} == "stromal cell" ~ "stromal cell (consensus)",
-        {{ label_column }} == "Stromal other" ~ "Stromal other (NBAtlas)",
-        .default = {{ label_column }}
-      )
-    )
-}
-
 # Recode NBAtlas cell types where possible so they match with colors
@@
   ) |>
-  unify_celltypes(label, label_recoded) |>
+  harmonize_celltypes(label, label_recoded) |>
   dplyr::select(-label)
-
-palette_df <- readr::read_tsv("../unified-cell-type-palette.tsv")
-setdiff(unique(celltype_recoded_df$label_recoded), palette_df$unified_label)
 ```
@@
 
-```{r, fig.width = 12, fig.height = 8}
-celltype_pal <- palette_df$hex_color
-names(celltype_pal) <- palette_df$unified_label
-
+```{r, fig.width = 10}
 ggplot(celltype_recoded_df) +
@@
 
-```{r}
-# Adapted from scpca-nf
-# https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_qc.rmd#L134-L221
-
-#' Create a faceted UMAP panel where each panel has only one cell type colored
-#'
-#' @param umap_df Data frame with UMAP1 and UMAP2 columns
-#' @param annotation_column Column containing broad cell type annotation
-#' @param celltype_colors Named vector of colors to use for each broader validation group
-#' @param point_size Point size. Default is 1
-#'
-#' @return ggplot object containing a faceted UMAP where each cell type is a facet.
-#'   In each panel, the cell type of interest is colored red and all other cells are grey.
-faceted_umap <- function(umap_df,
-                         annotation_column,
-                         celltype_colors,
-                         point_size = 1) {
-  # color by the annotation column but only color one cell type at a time
-  faceted_umap <- ggplot(
-    umap_df,
-    aes(x = UMAP1, y = UMAP2, color = {{ annotation_column }})
-  ) +
-    # set points for all "other" points
-    geom_point(
-      data = dplyr::select(
-        umap_df, -{{ annotation_column }}
-      ),
-      color = "gray90",
-      alpha = 0.5,
-      size = point_size
-    ) +
-    scale_color_manual(values = celltype_colors) +
-    # set points for desired cell type
-    geom_point(size = point_size, alpha = 0.5) +
-    facet_wrap(
-      vars({{ annotation_column }}),
-      ncol = 5
-    ) +
-    # remove axis numbers and background grid
-    scale_x_continuous(labels = NULL, breaks = NULL) +
-    scale_y_continuous(labels = NULL, breaks = NULL) +
-    theme_bw() +
-    theme(
-      legend.position = "none",
-      aspect.ratio = 1,
-      strip.background = element_rect(fill = "transparent")
-    )
-
-  return(faceted_umap)
-}
-```
-
-```{r fig.width = 20}
-singler_recoded_df <- celltype_recoded_df |>
-  dplyr::filter(annotation_type == "singler")
-
+```{r fig.width = 16, fig.height = 16}
+celltype_recoded_df |>
+  dplyr::filter(annotation_type == "singler") |>
   faceted_umap(
-  singler_recoded_df,
-  label_recoded,
-  celltype_pal
+    annotation_column = label_recoded,
+    celltype_colors = celltype_pal
   )
@@
 
-```{r}
+```{r, fig.width=8, fig.height=32}
 direct_celltype_matches <- c(
@@
   "natural killer cell",
-  "unknown_consensus" # _for this plot_, we'll recode the Neuroendocrine --> "unknown consensus" for the purposes of comparison
+  # we'll use this label to be able to directly compare unknown_consensus to Neuroendocrine
+  "putative-tumor"
 )
+
+celltype_facet_df <- celltype_recoded_df |>
+  # recode so Neuroendocrine matches with unknown_consensus in the plot
+  dplyr::mutate(
+    label_recoded = ifelse(
+      label_recoded %in% c("Neuroendocrine", "unknown_consensus"),
+      "putative-tumor",
+      label_recoded
+    )
+  ) |>
+  dplyr::filter(label_recoded %in% direct_celltype_matches)
+
+faceted_umap(
+  celltype_facet_df,
+  annotation_column = label_recoded,
+  celltype_colors = celltype_pal,
+  facet_type = "grid",
+  annotation_type_column = annotation_type
+)
+```
+
+## Dotplot time
+
+```{r}
+# get top 7 upregulated marker genes per cell type
+nbatlas_markers_df <- readr::read_tsv("../references/nbatlas-marker-genes.tsv") |>
+  dplyr::filter(direction == "up") |>
+  dplyr::group_by(NBAtlas_label) |>
+  dplyr::mutate(rank_lfc = rank(-avg_log2FC)) |>
+  dplyr::ungroup() |>
+  dplyr::filter(rank_lfc <= 5) |>
+  harmonize_celltypes(NBAtlas_label, label_recoded) |>
+  dplyr::select(label_recoded, gene_symbol, ensembl_gene_id, avg_log2FC)
+nbatlas_markers_df
+
+all_markers
+```
+
+```{r}
+# extract gene expression for marker genes from SCE
+# list of all marker genes
+all_markers <- nbatlas_markers_df |>
+  dplyr::pull(ensembl_gene_id) |>
+  unique()
+
+# we only care about if that gene is expressed otherwise we won't waste memory and include it
+gene_sums <- rowData(merged_sce) |>
+  as.data.frame() |>
+  dplyr::select(contains("detected")) |>
+  as.matrix() |>
+  rowSums()
+expressed_genes <- names(gene_sums)[gene_sums > 0]
+
+# get markers that are expressed
+expressed_markers <- intersect(all_markers, expressed_genes)
+expressed_markers
+```
+
+```{r}
+# get logcounts from merged_sce for expressed genes
+gene_exp_df <- scuttle::makePerCellDF(
+  merged_sce,
+  features = expressed_markers,
+  assay.type = "logcounts",
+  use.coldata = "cell_id",
+  use.dimred = FALSE
+) |>
+  tidyr::pivot_longer(starts_with("ENSG"), names_to = "ensembl_gene_id", values_to = "logcounts") |>
+  dplyr::mutate(detected = logcounts > 0)
+
+# Join with cell type results and marker gene info
+all_info_df <- celltype_recoded_df |>
+  dplyr::filter(annotation_type == "singler") |>
+  dplyr::select(cell_id, label_recoded) |>
+  dplyr::left_join(gene_exp_df, by = "cell_id") |>
+  # account for the same gene being present in multiple cell types
+  dplyr::left_join(nbatlas_markers_df, by = c("label_recoded", "ensembl_gene_id"), relationship = "many-to-many")
+```
+
+```{r}
+marker_gene_dotplot(
+  all_info_df,
+  markers_df = nbatlas_markers_df,
+  celltype_colors = celltype_pal
+)
+```
+
+CODE WORKS, NEEDS HELP THOUGH!
+what I need to do when I come back to this update my names to match allys code
+- broad_celltype_group is her singler
+- validation_group_annotation is her nbatlas marker gene df label
+
+Then i can use the same code to make a consensus dotplot.
+
+```{r}
+min_cells <- 0
+# prep for plots
+# get total number of cells per final annotation group
+total_cells_df <- all_info_df |>
+  dplyr::select(cell_id, label_recoded) |>
+  dplyr::distinct() |>
+  dplyr::count(label_recoded, name = "total_cells")
+
+# table with one row per unique broad cell type/ marker gene combination
+# first all cells in with the same broad_celltype_group (determined based on consensus_annotation) are grouped together
+# then get the mean gene expression and total percentage of cells that express each marker gene across all cells in that group
+# do this for every possible marker gene/ validation group assignment
+# second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
+group_stats_df <- all_info_df |>
+  # for each assigned cell type/marker gene combo get total detected and mean expression
+  # group by both broad group and validation group to account for genes that are expressed in more than one cell type
+  dplyr::summarize(
+    .by = c("label_recoded", "ensembl_gene_id"),
+    detected_count = sum(detected),
+    mean_exp = mean(logcounts)
+  ) |>
+  # add in validation group for marker genes
+  # this includes all possible marker genes and all possible validation group assignments
+  dplyr::left_join(nbatlas_markers_df, by = c("ensembl_gene_id", "label_recoded"), relationship = "many-to-many") |> # add total cells
+  dplyr::left_join(total_cells_df, by = c("label_recoded")) |>
+  # only show cells that have above the minimum total number
+  dplyr::filter(total_cells > min_cells) |>
+  dplyr::mutate(
+    # get total percent expressed
+    percent_exp = (detected_count / total_cells) * 100,
+    # account for NA/unknowns and set axes order
+    broad_celltype_group = factor(label_recoded, levels = c(names(celltype_pal)))
+  )
+
+# specify x axis order for dotplot
+marker_gene_order <- nbatlas_markers_df |>
+  dplyr::pull(gene_symbol) |>
+  unique()
+
+# set order for cell types
+celltype_order <- unique(nbatlas_markers_df$label_recoded)
+
+# filter out low expressed genes
+dotplot_df <- group_stats_df |>
+  dplyr::filter(mean_exp > 0, percent_exp > 10) |>
+  dplyr::arrange(label_recoded) |>
+  # add a label for the plot
+  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})")) |>
+  # remove marker genes that aren't present in final annotations and set x axis order
+  dplyr::filter(gene_symbol %in% marker_gene_order) |>
+  dplyr::mutate(
+    # set orders of gene symbol and validation groups
+    y_label = factor(y_label, levels = rev(unique(y_label))),
+    gene_symbol = factor(gene_symbol, levels = marker_gene_order),
+    label_recoded = factor(label_recoded, levels = celltype_order)
+  )
+
+# make dotplot with marker gene exp
+dotplot <- ggplot(dotplot_df, aes(y = y_label, x = gene_symbol, color = mean_exp, size = percent_exp)) +
+  geom_point() +
+  scale_color_viridis_c(option = "magma") +
+  facet_grid(cols = vars(label_recoded), scales = "free", space = "free") +
+  theme_classic() +
+  theme(
+    strip.background = element_rect(fill = "transparent", color = NA),
+    strip.placement = "outside",
+    strip.text.x = element_blank(),
+    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
+    axis.ticks.x = element_blank(),
+    text = element_text(size = 14),
+    panel.spacing = unit(0.5, "lines") # adjust spacing and match with annotation bar
+  ) +
+  labs(
+    x = "",
+    y = "label_recoded",
+    color = "Mean gene expression",
+    size = "Percent cells expressed"
+  )
+
+# add annotation bar aligning marker genes with validation group
+color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = label_recoded)) +
+  geom_tile() +
+  facet_grid(cols = vars(label_recoded), scales = "free", space = "free") +
+  scale_fill_manual(values = celltype_pal, breaks = levels(dotplot_df$label_recoded)) +
+  ggmap::theme_nothing() +
+  theme(
+    strip.background = element_rect(fill = "transparent", color = NA),
+    strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 12),
+    strip.placement = "outside",
+    legend.position = "none",
+    panel.spacing = unit(0.5, "lines"),
+    strip.clip = "off"
+  ) +
+  labs(fill = "")
+
+combined_plot <- color_bar / dotplot +
+  patchwork::plot_layout(ncol = 1, heights = c(0.1, 4))
+```
+
+```{r fig.width = 20}
+combined_plot
 ```

### 2025-07-10 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/explore-singler-results.Rmd
results path

@@
 module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
-results_dir <- file.path(module_dir, "results", "singler-filtered-genes")
+results_dir <- file.path(module_dir, "results", "singler")
 data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")

### 2025-07-10 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/explore-singler-results.Rmd
WIP: dotplot customization 2

@@
 ```{r}
-# get top 7 upregulated marker genes per cell type
+# get top 10??? upregulated marker genes per cell type
 nbatlas_markers_df <- readr::read_tsv("../references/nbatlas-marker-genes.tsv") |>
@@
   dplyr::ungroup() |>
-  dplyr::filter(rank_lfc <= 5) |>
-  harmonize_celltypes(NBAtlas_label, label_recoded) |>
-  dplyr::select(label_recoded, gene_symbol, ensembl_gene_id, avg_log2FC)
-nbatlas_markers_df
-
-all_markers
+  dplyr::filter(rank_lfc <= 10) |>
+  harmonize_celltypes(NBAtlas_label, NBAtlas_label_recoded) |>
+  dplyr::select(NBAtlas_label_recoded, gene_symbol, ensembl_gene_id, avg_log2FC)
 ```
@@
 expressed_markers <- intersect(all_markers, expressed_genes)
-expressed_markers
 ```
@@
 all_info_df <- celltype_recoded_df |>
-  dplyr::filter(annotation_type == "singler") |>
+  dplyr::filter(
+    annotation_type == "singler",
+    # we don't consider the NAs here
+    label_recoded != "singler_unknown"
+  ) |>
   dplyr::select(cell_id, label_recoded) |>
@@
   # account for the same gene being present in multiple cell types
-  dplyr::left_join(nbatlas_markers_df, by = c("label_recoded", "ensembl_gene_id"), relationship = "many-to-many")
+  dplyr::left_join(nbatlas_markers_df, by = "ensembl_gene_id", relationship = "many-to-many")
 ```
@@
 ```{r}
-marker_gene_dotplot(
-  all_info_df,
-  markers_df = nbatlas_markers_df,
-  celltype_colors = celltype_pal
-)
-```
+min_cells <- 0
 
-CODE WORKS, NEEDS HELP THOUGH!
-what I need to do when I come back to this update my names to match allys code
-- broad_celltype_group is her singler
-- validation_group_annotation is her nbatlas marker gene df label
-
-Then i can use the same code to make a consensus dotplot.
-
-```{r}
-min_cells <- 0
 # prep for plots
@@
   dplyr::distinct() |>
-  dplyr::count(label_recoded, name = "total_cells")
+  dplyr::count(label_recoded, name = "total_cells") |>
+  dplyr::arrange(desc(total_cells)) |>
+  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})"))
+
+# y axis - singler
+y_label_order <- rev(total_cells_df$y_label)
+
+# bar order - NBAtlas
+nbatlas_order <- total_cells_df$label_recoded
+
+# x axis marker genes - nbatlas
+nbatlas_markers_df <- nbatlas_markers_df |>
+  dplyr::filter(ensembl_gene_id %in% expressed_markers) |>
+  dplyr::mutate(
+    NBAtlas_label_recoded = factor(NBAtlas_label_recoded, levels = nbatlas_order)
+  ) |>
+  dplyr::arrange(NBAtlas_label_recoded)
+
+marker_gene_order <- nbatlas_markers_df |>
+  dplyr::pull(gene_symbol) |>
+  unique()
 
@@
   # group by both broad group and validation group to account for genes that are expressed in more than one cell type
+  # remove genes that aren't present in final annotations
+  dplyr::filter(gene_symbol %in% marker_gene_order) |>
+  dplyr::group_by(label_recoded, NBAtlas_label_recoded, ensembl_gene_id) |>
   dplyr::summarize(
-    .by = c("label_recoded", "ensembl_gene_id"),
     detected_count = sum(detected),
@@
   ) |>
+  dplyr::ungroup() |>
+  # add total cells
+  dplyr::left_join(total_cells_df, by = "label_recoded") |>
+  # get total percent expressed
+  dplyr::mutate(percent_exp = (detected_count / total_cells) * 100) |>
+  # remove lowly expressed marker genes and any that aren't present in final annotations
+  dplyr::filter(
+    mean_exp > 0,
+    percent_exp > 10,
+    # only show cells that have above the minimum total number
+    total_cells > min_cells
+  ) |>
   # add in validation group for marker genes
   # this includes all possible marker genes and all possible validation group assignments
-  dplyr::left_join(nbatlas_markers_df, by = c("ensembl_gene_id", "label_recoded"), relationship = "many-to-many") |> # add total cells
-  dplyr::left_join(total_cells_df, by = c("label_recoded")) |>
-  # only show cells that have above the minimum total number
-  dplyr::filter(total_cells > min_cells) |>
-  dplyr::mutate(
-    # get total percent expressed
-    percent_exp = (detected_count / total_cells) * 100,
-    # account for NA/unknowns and set axes order
-    broad_celltype_group = factor(label_recoded, levels = c(names(celltype_pal)))
+  dplyr::left_join(
+    nbatlas_markers_df,
+    by = c("ensembl_gene_id", "NBAtlas_label_recoded"),
+    relationship = "many-to-many"
   )
 
-# specify x axis order for dotplot
-marker_gene_order <- nbatlas_markers_df |>
-  dplyr::pull(gene_symbol) |>
-  unique()
-
-# set order for cell types
-celltype_order <- unique(nbatlas_markers_df$label_recoded)
-
-# filter out low expressed genes
 dotplot_df <- group_stats_df |>
-  dplyr::filter(mean_exp > 0, percent_exp > 10) |>
-  dplyr::arrange(label_recoded) |>
-  # add a label for the plot
-  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})")) |>
-  # remove marker genes that aren't present in final annotations and set x axis order
-  dplyr::filter(gene_symbol %in% marker_gene_order) |>
   dplyr::mutate(
     # set orders of gene symbol and validation groups
-    y_label = factor(y_label, levels = rev(unique(y_label))),
+    y_label = factor(y_label, y_label_order),
     gene_symbol = factor(gene_symbol, levels = marker_gene_order),
-    label_recoded = factor(label_recoded, levels = celltype_order)
+    NBAtlas_label_recoded = factor(NBAtlas_label_recoded, levels = nbatlas_order)
   )
+```
 
+```{r}
 # make dotplot with marker gene exp
@@
   scale_color_viridis_c(option = "magma") +
-  facet_grid(cols = vars(label_recoded), scales = "free", space = "free") +
+  facet_grid(cols = vars(NBAtlas_label_recoded), scales = "free", space = "free") +
   theme_classic() +
@@
   labs(
-    x = "",
-    y = "label_recoded",
+    x = "NBAtlas cell type marker genes",
+    y = "SingleR label",
     color = "Mean gene expression",
@@
 # add annotation bar aligning marker genes with validation group
-color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = label_recoded)) +
+color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = NBAtlas_label_recoded)) +
   geom_tile() +
-  facet_grid(cols = vars(label_recoded), scales = "free", space = "free") +
-  scale_fill_manual(values = celltype_pal, breaks = levels(dotplot_df$label_recoded)) +
+  facet_grid(cols = vars(NBAtlas_label_recoded), scales = "free", space = "free") +
+  scale_fill_manual(values = celltype_pal, breaks = levels(dotplot_df$NBAtlas_label_recoded)) +
   ggmap::theme_nothing() +

### 2025-07-11 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/explore-singler-results.Rmd
get dotplot code into function and fill out that section, plus some overall organization and text

@@
 module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
+ref_dir <- file.path(module_dir, "references")
 results_dir <- file.path(module_dir, "results", "singler")
@@
 )
+
+# marker genes for validation
+consensus_markers_file <- file.path(ref_dir, "consensus-marker-genes.tsv")
+nbatlas_markers_file <- file.path(ref_dir, "nbatlas-marker-genes.tsv")
 ```
@@
 ```{r}
-# Source Jaccard and heatmap utilities functions:
+# Source Jaccard and heatmap utilities functions
 source(file.path(module_dir, "scripts", "utils", "jaccard-utils.R"))
@@
 ```{r}
+#' Function to update NBAtlas labels
+#' This function matches NBAtlas labels to consensus labels where appropriate to facilitate plots
+#'
+#' @param df Data frame with columns to recode
+#' @param label_column Name of column with current labels
+#' @param recoded_column Name of new column to create with harmonized labels
+#'
+#' @returns Data frame with new column as specified containing recoded labels
 harmonize_celltypes <- function(df, label_column, recoded_column) {
@@
         {{ label_column }} == "NK cell" ~ "natural killer cell",
-        # make the difference stromal labels more clearly distinguishable
+        # also make the different stromal labels more clearly distinguishable
         {{ label_column }} == "stromal cell" ~ "stromal cell (consensus)",
@@
 ```{r}
-# Adapted from scpca-nf
-# https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_qc.rmd#L134-L221
-
 #' Create a faceted UMAP panel where each panel has only one cell type colored
+#' Adapted from scpca-nf:
+#' https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_qc.rmd#L134-L221
 #'
@@
 
+```{r}
+#' Prepare and create a marker gene expression dotplot
+#' Adapted from corresponding code in this ews-nf report:
+#' https://github.com/AlexsLemonade/ews-nf/blob/bd38f2bd5ae581ea7dcbd98c5ef717afb9015766/templates/summary-report.Rmd
+#'
+#' @param merged_sce Merged SCE object to plot with a `cell_id` column in the colData
+#' @param markers_df Data frame of marker genes for validation.
+#' This function expects columns `marker_gene_label` (cell types), `ensembl_gene_id`, and `gene_symbol`
+#' @param singler_df Data frame of singler annotations to plot.
+#' This function expects columns called `label_recoded` (cell types) and `cell_id`.
+#' The `cell_id` values should match the `cell_id` colData in the merged_sce
+#' @param total_cells_df Data frame of cell counts and plot order.
+#' This function expects columns called `label_recoded` (cell types), `y_label` (cell types with (total cells) as factor for plot order), and `total_cells`
+#' @param expressed_genes Vector of genes that are expressed in the merged_sce
+#' @param bar_order Vector for the annotation bar order
+#' @param min_cells Only include genes present in at least this many cells
+#'
+#' @returns Dotplot object
+generate_dotplot <- function(
+    merged_sce,
+    markers_df,
+    singler_df,
+    total_cells_df,
+    expressed_genes,
+    bar_order,
+    min_cells = 0) {
+  all_markers <- markers_df |>
+    dplyr::pull(ensembl_gene_id) |>
+    unique()
+
+  # consider only markers that are expressed
+  expressed_markers <- intersect(all_markers, expressed_genes)
+
+  # get logcounts from merged_sce for expressed genes
+  gene_exp_df <- scuttle::makePerCellDF(
+    merged_sce,
+    features = expressed_markers,
+    assay.type = "logcounts",
+    use.coldata = "cell_id",
+    use.dimred = FALSE
+  ) |>
+    tidyr::pivot_longer(starts_with("ENSG"), names_to = "ensembl_gene_id", values_to = "logcounts") |>
+    dplyr::mutate(detected = logcounts > 0)
+
+  # Join with cell type results and marker gene info
+  all_info_df <- singler_df |>
+    dplyr::left_join(gene_exp_df, by = "cell_id") |>
+    # account for the same gene being present in multiple cell types
+    dplyr::left_join(markers_df, by = "ensembl_gene_id", relationship = "many-to-many")
+
+  # define y- and x- axis orders
+
+  # y axis - singler
+  y_label_order <- rev(total_cells_df$y_label)
+
+  # x axis - marker genes. this uses the passed in bar_order
+  markers_df <- markers_df |>
+    dplyr::filter(ensembl_gene_id %in% expressed_markers) |>
+    dplyr::mutate(marker_gene_label = factor(marker_gene_label, levels = bar_order)) |>
+    dplyr::arrange(marker_gene_label)
+
+  marker_gene_order <- markers_df |>
+    dplyr::pull(gene_symbol) |>
+    unique()
+
+  group_stats_df <- all_info_df |>
+    # remove genes that aren't present in final annotations
+    dplyr::filter(gene_symbol %in% marker_gene_order) |>
+    # for each assigned cell type/marker gene combo get total detected and mean expression
+    dplyr::group_by(label_recoded, marker_gene_label, ensembl_gene_id) |>
+    dplyr::summarize(
+      detected_count = sum(detected),
+      mean_exp = mean(logcounts)
+    ) |>
+    dplyr::ungroup() |>
+    # add total cells
+    dplyr::left_join(total_cells_df, by = "label_recoded") |>
+    # get total percent expressed
+    dplyr::mutate(percent_exp = (detected_count / total_cells) * 100) |>
+    # add in validation group for marker genes
+    # this includes all possible marker genes and all possible validation group assignments
+    dplyr::left_join(
+      markers_df,
+      by = c("ensembl_gene_id", "marker_gene_label"),
+      relationship = "many-to-many"
+    )
+
+  dotplot_df <- group_stats_df |>
+    dplyr::filter(
+      # remove lowly expressed marker genes
+      mean_exp > 0,
+      percent_exp > 10,
+      # only show cells that have above the minimum total number
+      total_cells > min_cells
+    ) |>
+    # set orders of gene symbol and validation groups
+    dplyr::mutate(
+      y_label = factor(y_label, y_label_order),
+      gene_symbol = factor(gene_symbol, levels = marker_gene_order),
+      marker_gene_label = factor(marker_gene_label, levels = bar_order)
+    )
+
+  ### Make the dotplot!
+  dotplot <- ggplot(dotplot_df) +
+    aes(
+      y = y_label,
+      x = gene_symbol,
+      color = mean_exp,
+      size = percent_exp
+    ) +
+    geom_point() +
+    scale_color_viridis_c(option = "magma") +
+    facet_grid(cols = vars(marker_gene_label), scales = "free", space = "free") +
+    theme_classic() +
+    theme(
+      strip.background = element_rect(fill = "transparent", color = NA),
+      strip.placement = "outside",
+      strip.text.x = element_blank(),
+      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
+      axis.ticks.x = element_blank(),
+      text = element_text(size = 14),
+      panel.spacing = unit(0.5, "lines") # adjust spacing and match with annotation bar
+    ) +
+    labs(
+      x = "Validation marker genes",
+      y = "SingleR label",
+      color = "Mean gene expression",
+      size = "Percent cells expressed"
+    )
+
+  # add annotation bar aligning marker genes with validation group
+  color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = marker_gene_label)) +
+    geom_tile() +
+    facet_grid(cols = vars(marker_gene_label), scales = "free", space = "free") +
+    scale_fill_manual(values = celltype_pal, breaks = levels(dotplot_df$marker_gene_label)) +
+    ggmap::theme_nothing() +
+    theme(
+      strip.background = element_rect(fill = "transparent", color = NA),
+      strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 12),
+      strip.placement = "outside",
+      legend.position = "none",
+      panel.spacing = unit(0.5, "lines"),
+      strip.clip = "off"
+    ) +
+    labs(fill = "")
+
+  combined_plot <- color_bar / dotplot +
+    patchwork::plot_layout(ncol = 1, heights = c(0.1, 4))
+
+  return(combined_plot)
+}
+```
+
 ### Prepare input data
@@
 assay(merged_sce, "counts") <- NULL
-gc()
 
@@
 Read cell type data frames:
+
 ```{r}
@@
 
+# validation data frames
 validation_df <- readr::read_tsv(validation_url) |>
   dplyr::select(consensus_annotation, validation_group_annotation)
+consensus_markers_df <- readr::read_tsv(consensus_markers_file)
+nbatlas_markers_df <- readr::read_tsv(nbatlas_markers_file)
 
@@
   dplyr::left_join(singler_df, by = c("cell_id", "barcodes", "library_id"))
-```
 
-```{r}
 # Recode NBAtlas cell types where possible so they match with colors
@@
 ```
+
+## Heatmap
+
+This section compares SingleR annotations to consensus cell type annotations using a heatmap.
+The heatmap is colored by Jaccard similarity index.
 
@@
 
-```{r, fig.width = 10}
+## UMAPs
+
+This section visualizes and compares SingleR annotations to consensus cell type annotations using UMAPs.
+Note that the displayed UMAP is from a merged object that has _not been batch-corrected._
+
+Cell type labels have been harmonized between sources wherever possible.
+Note that each set of labels has its own "stromal" category which the labels distinguish.
+In addition, cells labeled `unknown_consensus` are those with no assigned consensus label, and cells labeled `unknown_singler` are those where SingleR could not confidently assign a label.
+
+### Complete UMAP
+
+First, we display the consensus and SingleR annotations for all cells.
+
+```{r, fig.width = 14, fig.height = 7}
 ggplot(celltype_recoded_df) +
@@
 
-Let's facet just the SingleR annotations:
+### SingleR annotations only
+
+Below we display a faceted UMAP to highlight the SingleR annotations.
+Light gray cells in each panel represent all other cell types.
 
@@
 
-For cells that directly match up with consensus:
+### Faceted comparison to consensus cell types
 
-```{r, fig.width=8, fig.height=32}
+Below, we display faceted UMAPs highlighting a single cell type but considering only cell types that have direct correspondence between SingleR and consensus cell types.
+This allows us to see if the normal cells that SingleR is labeling correspond well to the normal cells identified by consensus labels.
+Each row displays a cell type where the left column shows the consensus version and the right column shows the SingleR version.
+
+In addition, we include a category below `putative-tumor` to directly compare the unknown consensus labels with the `Neuroendocrine` cells labeled by SingleR. 
+While these categories are not necessarily directly comparable, they are each most likely to contain tumor cells.
+
+```{r, fig.width=7, fig.height=28}
 direct_celltype_matches <- c(
@@
 
-## Dotplot time
+## Marker gene expression dotplots
+
+In this section we'll validate annotations using two sets of marker genes:
+
+* Marker genes identified in NBAtlas corresponding to the atlas cell types
+  * This will tell us if we are picking up comparable signal in our data that plays well with NBAtlas
+* Consensus cell type marker genes
+  * This will provide an independent assessment of the SingleR normal cell type labels
 
 ```{r}
-# get top 10??? upregulated marker genes per cell type
-nbatlas_markers_df <- readr::read_tsv("../references/nbatlas-marker-genes.tsv") |>
-  dplyr::filter(direction == "up") |>
-  dplyr::group_by(NBAtlas_label) |>
-  dplyr::mutate(rank_lfc = rank(-avg_log2FC)) |>
-  dplyr::ungroup() |>
-  dplyr::filter(rank_lfc <= 10) |>
-  harmonize_celltypes(NBAtlas_label, NBAtlas_label_recoded) |>
-  dplyr::select(NBAtlas_label_recoded, gene_symbol, ensembl_gene_id, avg_log2FC)
-```
-
-```{r}
-# extract gene expression for marker genes from SCE
-# list of all marker genes
-all_markers <- nbatlas_markers_df |>
-  dplyr::pull(ensembl_gene_id) |>
-  unique()
-
+# Prepare the expressed_genes vector
 # we only care about if that gene is expressed otherwise we won't waste memory and include it
@@
 
-# get markers that are expressed
-expressed_markers <- intersect(all_markers, expressed_genes)
-```
-
-```{r}
-# get logcounts from merged_sce for expressed genes
-gene_exp_df <- scuttle::makePerCellDF(
-  merged_sce,
-  features = expressed_markers,
-  assay.type = "logcounts",
-  use.coldata = "cell_id",
-  use.dimred = FALSE
-) |>
-  tidyr::pivot_longer(starts_with("ENSG"), names_to = "ensembl_gene_id", values_to = "logcounts") |>
-  dplyr::mutate(detected = logcounts > 0)
-
-# Join with cell type results and marker gene info
-all_info_df <- celltype_recoded_df |>
+# Prepare data frame with singler labels to plot
+singler_recoded_df <- celltype_recoded_df |>
   dplyr::filter(
@@
     # we don't consider the NAs here
-    label_recoded != "singler_unknown"
+    label_recoded != "unknown_singler"
   ) |>
-  dplyr::select(cell_id, label_recoded) |>
-  dplyr::left_join(gene_exp_df, by = "cell_id") |>
-  # account for the same gene being present in multiple cell types
-  dplyr::left_join(nbatlas_markers_df, by = "ensembl_gene_id", relationship = "many-to-many")
-```
+  dplyr::select(cell_id, label_recoded)
 
-```{r}
-min_cells <- 0
-
-# prep for plots
-# get total number of cells per final annotation group
-total_cells_df <- all_info_df |>
-  dplyr::select(cell_id, label_recoded) |>
-  dplyr::distinct() |>
+# get total number of cells per final annotation group and set up y_label
+total_cells_df <- singler_recoded_df |>
   dplyr::count(label_recoded, name = "total_cells") |>
@@
 
-# y axis - singler
-y_label_order <- rev(total_cells_df$y_label)
+singler_order <- total_cells_df$y_label
+total_cells_df$y_label <- factor(total_cells_df$y_label, levels = singler_order)
+```
 
-# bar order - NBAtlas
-nbatlas_order <- total_cells_df$label_recoded
+### NBAtlas marker genes
 
-# x axis marker genes - nbatlas
+We'll show the top-seven highest logFC marker genes per cell type for validation.
+
+```{r}
+# number of marker genes to consider per validation group
+n_marker_genes <- 7
+
 nbatlas_markers_df <- nbatlas_markers_df |>
-  dplyr::filter(ensembl_gene_id %in% expressed_markers) |>
-  dplyr::mutate(
-    NBAtlas_label_recoded = factor(NBAtlas_label_recoded, levels = nbatlas_order)
-  ) |>
-  dplyr::arrange(NBAtlas_label_recoded)
+  # keep only the top `n_marker_genes`
+  dplyr::filter(direction == "up") |>
+  dplyr::group_by(NBAtlas_label) |>
+  dplyr::mutate(rank_lfc = rank(-avg_log2FC)) |>
+  dplyr::ungroup() |>
+  dplyr::filter(rank_lfc <= n_marker_genes) |>
+  harmonize_celltypes(NBAtlas_label, marker_gene_label) |>
+  dplyr::select(marker_gene_label, gene_symbol, ensembl_gene_id)
 
-marker_gene_order <- nbatlas_markers_df |>
-  dplyr::pull(gene_symbol) |>
-  unique()
+nbatlas_bar_order <- total_cells_df$label_recoded
+```
 
-# table with one row per unique broad cell type/ marker gene combination
-# first all cells in with the same broad_celltype_group (determined based on consensus_annotation) are grouped together
-# then get the mean gene expression and total percentage of cells that express each marker gene across all cells in that group
-# do this for every possible marker gene/ validation group assignment
-# second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
-group_stats_df <- all_info_df |>
-  # for each assigned cell type/marker gene combo get total detected and mean expression
-  # group by both broad group and validation group to account for genes that are expressed in more than one cell type
-  # remove genes that aren't present in final annotations
-  dplyr::filter(gene_symbol %in% marker_gene_order) |>
-  dplyr::group_by(label_recoded, NBAtlas_label_recoded, ensembl_gene_id) |>
-  dplyr::summarize(
-    detected_count = sum(detected),
-    mean_exp = mean(logcounts)
-  ) |>
-  dplyr::ungroup() |>
-  # add total cells
-  dplyr::left_join(total_cells_df, by = "label_recoded") |>
-  # get total percent expressed
-  dplyr::mutate(percent_exp = (detected_count / total_cells) * 100) |>
-  # remove lowly expressed marker genes and any that aren't present in final annotations
-  dplyr::filter(
-    mean_exp > 0,
-    percent_exp > 10,
-    # only show cells that have above the minimum total number
-    total_cells > min_cells
-  ) |>
-  # add in validation group for marker genes
-  # this includes all possible marker genes and all possible validation group assignments
-  dplyr::left_join(
+```{r, fig.width = 24, fig.height = 12}
+generate_dotplot(
+  merged_sce,
   nbatlas_markers_df,
-    by = c("ensembl_gene_id", "NBAtlas_label_recoded"),
-    relationship = "many-to-many"
-  )
-
-dotplot_df <- group_stats_df |>
-  dplyr::mutate(
-    # set orders of gene symbol and validation groups
-    y_label = factor(y_label, y_label_order),
-    gene_symbol = factor(gene_symbol, levels = marker_gene_order),
-    NBAtlas_label_recoded = factor(NBAtlas_label_recoded, levels = nbatlas_order)
+  singler_recoded_df,
+  total_cells_df,
+  expressed_genes,
+  nbatlas_bar_order
 )
@@
 
+### Consensus validation marker genes
+
 ```{r}
-# make dotplot with marker gene exp
-dotplot <- ggplot(dotplot_df, aes(y = y_label, x = gene_symbol, color = mean_exp, size = percent_exp)) +
-  geom_point() +
-  scale_color_viridis_c(option = "magma") +
-  facet_grid(cols = vars(NBAtlas_label_recoded), scales = "free", space = "free") +
-  theme_classic() +
-  theme(
-    strip.background = element_rect(fill = "transparent", color = NA),
-    strip.placement = "outside",
-    strip.text.x = element_blank(),
-    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
-    axis.ticks.x = element_blank(),
-    text = element_text(size = 14),
-    panel.spacing = unit(0.5, "lines") # adjust spacing and match with annotation bar
-  ) +
-  labs(
-    x = "NBAtlas cell type marker genes",
-    y = "SingleR label",
-    color = "Mean gene expression",
-    size = "Percent cells expressed"
+# prepare for dotplot
+
+consensus_markers_df <- consensus_markers_df |>
+  # we'll use this to help determine the bar order
+  harmonize_celltypes(NBAtlas_label, NBAtlas_label_recoded) |>
+  dplyr::select(
+    marker_gene_label = validation_group_annotation,
+    ensembl_gene_id,
+    gene_symbol,
+    NBAtlas_label_recoded
   )
 
-# add annotation bar aligning marker genes with validation group
-color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = NBAtlas_label_recoded)) +
-  geom_tile() +
-  facet_grid(cols = vars(NBAtlas_label_recoded), scales = "free", space = "free") +
-  scale_fill_manual(values = celltype_pal, breaks = levels(dotplot_df$NBAtlas_label_recoded)) +
-  ggmap::theme_nothing() +
-  theme(
-    strip.background = element_rect(fill = "transparent", color = NA),
-    strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 12),
-    strip.placement = "outside",
-    legend.position = "none",
-    panel.spacing = unit(0.5, "lines"),
-    strip.clip = "off"
-  ) +
-  labs(fill = "")
-
-combined_plot <- color_bar / dotplot +
-  patchwork::plot_layout(ncol = 1, heights = c(0.1, 4))
+# get the bar order
+consensus_bar_order <- total_cells_df |>
+  dplyr::select(label_recoded, y_label) |>
+  # inner!!!!
+  dplyr::inner_join(consensus_markers_df, by = c("label_recoded" = "NBAtlas_label_recoded")) |>
+  dplyr::arrange(y_label) |>
+  dplyr::pull(marker_gene_label) |>
+  unique()
 ```
 
-```{r fig.width = 20}
-combined_plot
+```{r, fig.width = 22, fig.height = 12}
+generate_dotplot(
+  merged_sce,
+  consensus_markers_df,
+  singler_recoded_df,
+  total_cells_df,
+  expressed_genes,
+  consensus_bar_order
+)
 ```

### 2025-07-11 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd

--- 2025-07-11 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/explore-singler-results.Rmd
@@
 ---
+
+This notebooks explores the results from running cell type annotation with `SingleR` using the NBAtlas.
+The NBAtlas reference was aggregated with `SingelR` prior to model training.
+
+In this notebook, we visualize inferred cell type annotations directly, compare them to normal consensus cell types, and validate cell type assignments with marker genes.
 
@@
 * Consensus cell type marker genes
-  * This will provide an independent assessment of the SingleR normal cell type labels
+  * This will provide an independent assessment of the reliability of SingleR normal cell type assignments
 

### 2025-07-11 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-testing/separate-tumor.Rmd

--- 2025-07-10 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/separate-tumor.Rmd
@@
   "scratch",
-  "singler-model_nbatlas_aggregated_NE-tumor-separated.rds"
+  "singler-model_nbatlas_separated.rds"
 )
@@
   "scratch",
-  "singler-model_nbatlas_aggregated_NE-tumor-combined.rds"
+  "singler-model_nbatlas_combined.rds"
 )
@@
 # Source Jaccard and heatmap utilities functions:
-source(file.path(module_dir, "exploratory-notebooks", "jaccard-utils.R"))
+source(file.path(module_dir, "scripts", "utils", "jaccard-utils.R"))
 ```

### 2025-07-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/filter-genes.Rmd

@@
 * Neuroendocrine --> NE
+* Stromal other --> stromal
 
@@
       singler_label == "Neuroendocrine" ~ "NE",
+      singler_label == "Stromal other" ~ "stromal",
       .default = singler_label

### 2025-07-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
move functions into separate utils file

@@
 source(file.path(module_dir, "scripts", "utils", "jaccard-utils.R"))
-```
 
-```{r}
-#' Function to update NBAtlas labels
-#' This function matches NBAtlas labels to consensus labels where appropriate to facilitate plots
-#'
-#' @param df Data frame with columns to recode
-#' @param label_column Name of column with current labels
-#' @param recoded_column Name of new column to create with harmonized labels
-#'
-#' @returns Data frame with new column as specified containing recoded labels
-harmonize_celltypes <- function(df, label_column, recoded_column) {
-  df |>
-    dplyr::mutate(
-      {{ recoded_column }} := dplyr::case_when(
-        # nbatlas ~ consensus
-        {{ label_column }} %in% c("Myeloid", "Fibroblast") ~ stringr::str_to_lower({{ label_column }}),
-        {{ label_column }} == "Endothelial" ~ "endothelial cell",
-        {{ label_column }} == "Plasma" ~ "plasma cell",
-        {{ label_column }} == "NK cell" ~ "natural killer cell",
-        # also make the different stromal labels more clearly distinguishable
-        {{ label_column }} == "stromal cell" ~ "stromal cell (consensus)",
-        {{ label_column }} == "Stromal other" ~ "Stromal other (NBAtlas)",
-        .default = {{ label_column }}
-      )
-    )
-}
-```
-
-```{r}
-#' Create a faceted UMAP panel where each panel has only one cell type colored
-#' Adapted from scpca-nf:
-#' https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_qc.rmd#L134-L221
-#'
-#' @param umap_df Data frame with UMAP1 and UMAP2 columns
-#' @param annotation_column Column containing broad cell type annotation
-#' @param celltype_colors Named vector of colors to use for each broader validation group
-#' @param facet_type Whether to use facet_wrap or facet_grid
-#' @param annotation_type_column Additional column to use if facet_type is "grid"
-#'
-#' @return ggplot object containing a faceted UMAP where each cell type is a facet.
-#'   In each panel, the cell type of interest is colored red and all other cells are grey.
-faceted_umap <- function(umap_df,
-                         annotation_column,
-                         celltype_colors,
-                         facet_type = c("wrap", "grid"),
-                         annotation_type_column = NULL) {
-  facet_type <- match.arg(facet_type)
-
-  # color by the annotation column but only color one cell type at a time
-  faceted_umap <- ggplot(
-    umap_df,
-    aes(x = UMAP1, y = UMAP2, color = {{ annotation_column }})
-  ) +
-    # set points for all "other" points
-    geom_point(
-      data = dplyr::select(
-        umap_df, -{{ annotation_column }}
-      ),
-      color = "gray90",
-      alpha = 0.5,
-      size = 0.5
-    ) +
-    scale_color_manual(values = celltype_colors) +
-    # set points for desired cell type
-    geom_point(size = 0.5, alpha = 0.5) +
-    coord_equal() +
-    theme_bw() +
-    theme(
-      legend.position = "none",
-      axis.ticks = element_blank(),
-      axis.text = element_blank()
-    )
-
-  # add faceting as specified
-  if (facet_type == "wrap") {
-    faceted_umap <- faceted_umap +
-      facet_wrap(
-        vars({{ annotation_column }}),
-        ncol = 4
-      )
-  } else {
-    faceted_umap <- faceted_umap +
-      facet_grid(
-        rows = vars({{ annotation_column }}),
-        cols = vars({{ annotation_type_column }})
-      )
-  }
-  return(faceted_umap)
-}
-```
-
-```{r}
-#' Prepare and create a marker gene expression dotplot
-#' Adapted from corresponding code in this ews-nf report:
-#' https://github.com/AlexsLemonade/ews-nf/blob/bd38f2bd5ae581ea7dcbd98c5ef717afb9015766/templates/summary-report.Rmd
-#'
-#' @param merged_sce Merged SCE object to plot with a `cell_id` column in the colData
-#' @param markers_df Data frame of marker genes for validation.
-#' This function expects columns `marker_gene_label` (cell types), `ensembl_gene_id`, and `gene_symbol`
-#' @param singler_df Data frame of singler annotations to plot.
-#' This function expects columns called `label_recoded` (cell types) and `cell_id`.
-#' The `cell_id` values should match the `cell_id` colData in the merged_sce
-#' @param total_cells_df Data frame of cell counts and plot order.
-#' This function expects columns called `label_recoded` (cell types), `y_label` (cell types with (total cells) as factor for plot order), and `total_cells`
-#' @param expressed_genes Vector of genes that are expressed in the merged_sce
-#' @param bar_order Vector for the annotation bar order
-#' @param min_cells Only include genes present in at least this many cells
-#'
-#' @returns Dotplot object
-generate_dotplot <- function(
-    merged_sce,
-    markers_df,
-    singler_df,
-    total_cells_df,
-    expressed_genes,
-    bar_order,
-    min_cells = 0) {
-  all_markers <- markers_df |>
-    dplyr::pull(ensembl_gene_id) |>
-    unique()
-
-  # consider only markers that are expressed
-  expressed_markers <- intersect(all_markers, expressed_genes)
-
-  # get logcounts from merged_sce for expressed genes
-  gene_exp_df <- scuttle::makePerCellDF(
-    merged_sce,
-    features = expressed_markers,
-    assay.type = "logcounts",
-    use.coldata = "cell_id",
-    use.dimred = FALSE
-  ) |>
-    tidyr::pivot_longer(starts_with("ENSG"), names_to = "ensembl_gene_id", values_to = "logcounts") |>
-    dplyr::mutate(detected = logcounts > 0)
-
-  # Join with cell type results and marker gene info
-  all_info_df <- singler_df |>
-    dplyr::left_join(gene_exp_df, by = "cell_id") |>
-    # account for the same gene being present in multiple cell types
-    dplyr::left_join(markers_df, by = "ensembl_gene_id", relationship = "many-to-many")
-
-  # define y- and x- axis orders
-
-  # y axis - singler
-  y_label_order <- rev(total_cells_df$y_label)
-
-  # x axis - marker genes. this uses the passed in bar_order
-  markers_df <- markers_df |>
-    dplyr::filter(ensembl_gene_id %in% expressed_markers) |>
-    dplyr::mutate(marker_gene_label = factor(marker_gene_label, levels = bar_order)) |>
-    dplyr::arrange(marker_gene_label)
-
-  marker_gene_order <- markers_df |>
-    dplyr::pull(gene_symbol) |>
-    unique()
-
-  group_stats_df <- all_info_df |>
-    # remove genes that aren't present in final annotations
-    dplyr::filter(gene_symbol %in% marker_gene_order) |>
-    # for each assigned cell type/marker gene combo get total detected and mean expression
-    dplyr::group_by(label_recoded, marker_gene_label, ensembl_gene_id) |>
-    dplyr::summarize(
-      detected_count = sum(detected),
-      mean_exp = mean(logcounts)
-    ) |>
-    dplyr::ungroup() |>
-    # add total cells
-    dplyr::left_join(total_cells_df, by = "label_recoded") |>
-    # get total percent expressed
-    dplyr::mutate(percent_exp = (detected_count / total_cells) * 100) |>
-    # add in validation group for marker genes
-    # this includes all possible marker genes and all possible validation group assignments
-    dplyr::left_join(
-      markers_df,
-      by = c("ensembl_gene_id", "marker_gene_label"),
-      relationship = "many-to-many"
-    )
-
-  dotplot_df <- group_stats_df |>
-    dplyr::filter(
-      # remove lowly expressed marker genes
-      mean_exp > 0,
-      percent_exp > 10,
-      # only show cells that have above the minimum total number
-      total_cells > min_cells
-    ) |>
-    # set orders of gene symbol and validation groups
-    dplyr::mutate(
-      y_label = factor(y_label, y_label_order),
-      gene_symbol = factor(gene_symbol, levels = marker_gene_order),
-      marker_gene_label = factor(marker_gene_label, levels = bar_order)
-    )
-
-  ### Make the dotplot!
-  dotplot <- ggplot(dotplot_df) +
-    aes(
-      y = y_label,
-      x = gene_symbol,
-      color = mean_exp,
-      size = percent_exp
-    ) +
-    geom_point() +
-    scale_color_viridis_c(option = "magma") +
-    facet_grid(cols = vars(marker_gene_label), scales = "free", space = "free") +
-    theme_classic() +
-    theme(
-      strip.background = element_rect(fill = "transparent", color = NA),
-      strip.placement = "outside",
-      strip.text.x = element_blank(),
-      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
-      axis.ticks.x = element_blank(),
-      text = element_text(size = 14),
-      panel.spacing = unit(0.5, "lines") # adjust spacing and match with annotation bar
-    ) +
-    labs(
-      x = "Validation marker genes",
-      y = "SingleR label",
-      color = "Mean gene expression",
-      size = "Percent cells expressed"
-    )
-
-  # add annotation bar aligning marker genes with validation group
-  color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = marker_gene_label)) +
-    geom_tile() +
-    facet_grid(cols = vars(marker_gene_label), scales = "free", space = "free") +
-    scale_fill_manual(values = celltype_pal, breaks = levels(dotplot_df$marker_gene_label)) +
-    ggmap::theme_nothing() +
-    theme(
-      strip.background = element_rect(fill = "transparent", color = NA),
-      strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 12),
-      strip.placement = "outside",
-      legend.position = "none",
-      panel.spacing = unit(0.5, "lines"),
-      strip.clip = "off"
-    ) +
-    labs(fill = "")
-
-  combined_plot <- color_bar / dotplot +
-    patchwork::plot_layout(ncol = 1, heights = c(0.1, 4))
-
-  return(combined_plot)
-}
+# Source additional utilities functions:
+# - harmonize_celltypes()
+# - faceted_umap()
+# - generate_dotplot()
+source(file.path(module_dir, "scripts", "utils", "celltype-utils.R"))
 ```

### 2025-07-14 analyses/cell-type-neuroblastoma-04/scripts/utils/celltype-utils.R

@@ -1,0 +1,248 @@
+# This file contains the following functions:
+# - harmonize_celltypes()
+# - faceted_umap()
+# - generate_dotplot()
+
+#' Function to update NBAtlas labels
+#' This function matches NBAtlas labels to consensus labels where appropriate to facilitate plots
+#'
+#' @param df Data frame with columns to recode
+#' @param label_column Name of column with current labels
+#' @param recoded_column Name of new column to create with harmonized labels
+#'
+#' @returns Data frame with new column as specified containing recoded labels
+harmonize_celltypes <- function(df, label_column, recoded_column) {
+  df |>
+    dplyr::mutate(
+      {{ recoded_column }} := dplyr::case_when(
+        # nbatlas ~ consensus
+        {{ label_column }} %in% c("Myeloid", "Fibroblast") ~ stringr::str_to_lower({{ label_column }}),
+        {{ label_column }} == "Endothelial" ~ "endothelial cell",
+        {{ label_column }} == "Plasma" ~ "plasma cell",
+        {{ label_column }} == "NK cell" ~ "natural killer cell",
+        # also make the different stromal labels more clearly distinguishable
+        {{ label_column }} == "stromal cell" ~ "stromal cell (consensus)",
+        {{ label_column }} == "Stromal other" ~ "Stromal other (NBAtlas)",
+        .default = {{ label_column }}
+      )
+    )
+}
+
+
+#' Create a faceted UMAP panel where each panel has only one cell type colored
+#' Adapted from scpca-nf:
+#' https://github.com/AlexsLemonade/scpca-nf/blob/4bb82aa635b572a62f2028dbec587fcfd2155e26/templates/qc_report/celltypes_qc.rmd#L134-L221
+#'
+#' @param umap_df Data frame with UMAP1 and UMAP2 columns
+#' @param annotation_column Column containing broad cell type annotation
+#' @param celltype_colors Named vector of colors to use for each broader validation group
+#' @param facet_type Whether to use facet_wrap or facet_grid
+#' @param annotation_type_column Additional column to use if facet_type is "grid"
+#'
+#' @return ggplot object containing a faceted UMAP where each cell type is a facet.
+#'   In each panel, the cell type of interest is colored red and all other cells are grey.
+faceted_umap <- function(umap_df,
+                         annotation_column,
+                         celltype_colors,
[…]

### 2025-07-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
fix faceted umap prep code - there are several myeloid types to collapse in consensus for this comparison

@@
   "T cell",
+  # note that we need to collapse the consensus to compare
   "myeloid",
@@
 
+# Define the different myeloid types in consensus annotations
+consensus_myeloid <- c("myeloid", "dendritic cell", "monocyte", "macrophage")
+
 celltype_facet_df <- celltype_recoded_df |>
+  dplyr::mutate(
     # recode so Neuroendocrine matches with unknown_consensus in the plot
-  dplyr::mutate(
     label_recoded = ifelse(
@@
       "putative-tumor",
+      label_recoded
+    ), 
+    # recode all myeloid types, since NBAtlas has one group but we have several categories
+    label_recoded = ifelse(
+      label_recoded %in% consensus_myeloid,
+      "myeloid",
       label_recoded

### 2025-07-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
Very WIP: expression plot prep

@@
 
+```{r}
+# this time we are not considering putative tumor
+direct_celltype_matches <- c(
+  "B cell",
+  "T cell",
+  # note that we need to collapse the consensus to compare
+  "myeloid",
+  "fibroblast",
+  "endothelial cell",
+  "plasma cell",
+  "natural killer cell", 
+  "unknown_consensus" # we do want to keep these for this analysis
+)
+
+# Define the different myeloid types in consensus annotations
+consensus_myeloid <- c("myeloid", "dendritic cell", "monocyte", "macrophage")
+
+situation_df <- celltype_recoded_df |>
+  # recode labels for comparison
+  dplyr::mutate(
+    label_recoded = ifelse(
+      label_recoded %in% consensus_myeloid,
+      "myeloid",
+      label_recoded
+  )) |>
+  dplyr::filter(label_recoded %in% direct_celltype_matches) |>
+  dplyr::select(cell_id, annotation_type, label_recoded) |>
+  # make it wider to label the situation
+  tidyr::pivot_wider(
+    names_from = "annotation_type",
+    values_from = "label_recoded"
+  ) |>
+  # remove NAs:
+  # - consensus NAs are those with other cell types not to include here
+  # - singler NAs were unclassified cells
+  tidyr::drop_na() |>
+  dplyr::mutate(
+    annotation_situation = dplyr::case_when(
+      consensus_validation == "unknown_consensus" ~ "consensus not classified", 
+      consensus_validation == singler ~ "matching annotations",
+      consensus_validation != singler ~ "different annotations"
+    )
+  ) |>
+  # make it long again
+  tidyr::pivot_longer(
+    c(consensus_validation, singler),
+    names_to = "annotation_type", 
+    values_to = "label_recoded"
+  )
+
+cells <- unique(situation_df$cell_id)
+genes <- unique(consensus_markers_df$ensembl_gene_id)
+```
+
+```{r}
+scuttle::makePerCellDF(
+    merged_sce[genes, cells],
+    features = genes,
+    assay.type = "logcounts",
+    use.coldata = "cell_id",
+    use.dimred = FALSE
+  ) |>
+    tidyr::pivot_longer(starts_with("ENSG"), names_to = "ensembl_gene_id", values_to = "logcounts") |>
+  dplyr::left_join(situation_df, by = "cell_id") |>
+  dplyr::group_by(label_recoded, marker_gene_label, ensembl_gene_id) |>
+  dplyr::summarize(
+    detected_count = sum(detected),
+    mean_exp = mean(logcounts)
+  ) |>
+  dplyr::ungroup()
+
+ggplot(situation_df) + 
+  aes(x = singler, y = )
+
+  group_stats_df <- all_info_df |>
+    # remove genes that aren't present in final annotations
+    dplyr::filter(gene_symbol %in% marker_gene_order) |>
+    # for each assigned cell type/marker gene combo get total detected and mean expression
+    dplyr::group_by(label_recoded, marker_gene_label, ensembl_gene_id) |>
+    dplyr::summarize(
+      detected_count = sum(detected),
+      mean_exp = mean(logcounts)
+    ) |>
+
+```
+
 ## Session Info

### 2025-07-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
update faceted umaps for updated cell types

@@
   "T cell",
-  # note that we need to collapse the consensus to compare
   "myeloid",
+  "macrophage",
+  "monocyte",
+  "dendritic cell",
+  "natural killer cell",
   "fibroblast",
@@
 
-# Define the different myeloid types in consensus annotations
-consensus_myeloid <- c("myeloid", "dendritic cell", "monocyte", "macrophage")
-
 celltype_facet_df <- celltype_recoded_df |>
@@
       label_recoded
-    ), 
-    # recode all myeloid types, since NBAtlas has one group but we have several categories
-    label_recoded = ifelse(
-      label_recoded %in% consensus_myeloid,
-      "myeloid",
-      label_recoded
-    )
-  ) |>
+    )) |>
   dplyr::filter(label_recoded %in% direct_celltype_matches)

### 2025-07-15 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
swap in validation url and add table of cell types and labels

@@
 # marker genes for validation
-consensus_markers_file <- file.path(ref_dir, "consensus-marker-genes.tsv")
+consensus_markers_file <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/validation-markers.tsv"
 nbatlas_markers_file <- file.path(ref_dir, "nbatlas-marker-genes.tsv")
@@
 ```
+
+## Cell type comparisons
+
+Throughout this notebook, we compare the cell type labels from `SingleR` with the consensus cell type labels.
+To facilitate this, the `SingleR` labels from `NBAtlas` were grouped together to both a) mirror (where possible) the broad consensus labels, and b) to make some visualizations more manageable.
+
+The table below shows the `NBAtlas` cell types and how they are represented in visualizations, unless otherwise stated.
+
+| Broad label shown in figures | Associated NBAtlas labels                                       |
+| ---------------------------- | --------------------------------------------------------------- |
+| `T cell`                     | `CD4+ T cell`, `CD8+ T cell`, `Treg`, `NKT cell`                |
+| `dendritic cell`             | `cDC1`, `cDC2/DC3`, `Migratory cDC`                             |
+| `natural killer cell cell`   | `Circulating NK cell`, `TOX2+/KIT+ NK cell`, `Resident NK cell` |
+| `monocyte`                   | `CD4+ T cell`, `Classical monocyte`, `Patrolling monocyte`      |
+| `myeloid`                    | `Neutrophil`                                                    |
+| `Neuroendocrine`             | `Neuroendocrine`                                                |
+| `fibroblast`                 | `Fibroblast`                                                    |
+| `macrophage`                 | `Macrophage`                                                    |
+| `B cell`                     | `B cell`                                                        |
+| `plasma cell`                | `Plasma`                                                        |
+| `endothelial`                | `Endothelial`                                                   |
+| `Stromal other (NBAtlas)`    | `Stromal other`                                                 |
+| `Schwann`                    | `Schwann`                                                       |
+| `RBCs`                       | `RBCs`                                                          |
+| `Immune cycling`             | `Immune cycling`                                                |
+| `pDC`                        | `pDC`                                                           |
+
+Cells labeled `singler_unknown` are those for which `SingleR` could not reliably assign an annotation.
 
@@
 
-We'll show the top-seven highest logFC marker genes per cell type for validation.
+We'll show the top-5 highest log2FC marker genes per cell type for validation.
 
@@
 # number of marker genes to consider per validation group
-n_marker_genes <- 7
+n_marker_genes <- 5
 

### 2025-07-15 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
got dotplots working with new marker genes and labels, including use of updated palettes

@@
 
-# palette file
-palette_file <- file.path(
+# palette files
+recoded_palette_file <- file.path(
   module_dir,
@@
   "harmonized-cell-type-palette.tsv"
+)
+
+nbatlas_palette_file <- file.path(
+  module_dir,
+  "palettes",
+  "nbatlas-marker-genes-palette.tsv"
 )
@@
 
-# set up palette for umaps
-palette_df <- readr::read_tsv(palette_file)
-celltype_pal <- palette_df$hex_color
-names(celltype_pal) <- palette_df$harmonized_label
+# set up palettes
+
+# recoded to shared colors
+recoded_palette_df <- readr::read_tsv(recoded_palette_file)
+recoded_celltype_pal <- recoded_palette_df$hex_color
+names(recoded_celltype_pal) <- recoded_palette_df$harmonized_label
+
+# only the NBAtlas labels - use for validation dotplot
+nbatlas_palette_df <- readr::read_tsv(nbatlas_palette_file)
+nbatlas_celltype_pal <- nbatlas_palette_df$hex_color
+names(nbatlas_celltype_pal) <- nbatlas_palette_df$NBAtlas_label
+
 ```
@@
   scale_color_manual(
-    values = celltype_pal,
+    values = recoded_celltype_pal,
     name = "Harmonized cell types"
@@
     annotation_column = label_recoded,
-    celltype_colors = celltype_pal
+    celltype_colors = recoded_celltype_pal
   )
@@
 While these categories are not necessarily directly comparable, they are each most likely to contain tumor cells.
+
+Also, note that the `myeloid`-labeled SingleR cells are _only_ neutrophils, but the consensus myeloid cagetory contains additonal cell types.
 
@@
   annotation_column = label_recoded,
-  celltype_colors = celltype_pal,
+  celltype_colors = recoded_celltype_pal,
   facet_type = "grid",
@@
 expressed_genes <- names(gene_sums)[gene_sums > 0]
+```
 
+### NBAtlas marker genes
+
+We'll show the top-5 highest log2FC marker genes per cell type for validation.
+
+```{r}
+# number of marker genes to consider per validation group
+n_marker_genes <- 5
+
+top_nbatlas_markers_df <- nbatlas_markers_df |>
+  # keep only the top `n_marker_genes`
+  dplyr::filter(direction == "up") |>
+  dplyr::group_by(NBAtlas_label) |>
+  dplyr::mutate(rank_lfc = rank(-avg_log2FC)) |>
+  dplyr::ungroup() |>
+  dplyr::filter(rank_lfc <= n_marker_genes) |>
+  dplyr::select(marker_gene_label = NBAtlas_label, gene_symbol, ensembl_gene_id)
+```
+
+```{r, fig.width = 24, fig.height = 12}
+# we want to plot the literal labels here
+full_celltype_df <- celltype_df |>
+  dplyr::select(
+    cell_id, 
+    # the dotplot function expects this column name
+    label_recoded = pruned.labels
+  ) |>
+  dplyr::mutate(
+    label_recoded = dplyr::case_when(
+      # collapse the dendritic cells & monocytes to match what's in the marker genes
+      stringr::str_detect(label_recoded, "cDC") ~ "cDC", 
+      stringr::str_detect(label_recoded, "monocyte") ~ "Monocyte",
+      label_recoded == "NE" ~ "Neuroendocrine",
+      .default = label_recoded
+  )) |>
+  dplyr::filter(label_recoded != "unknown_singler")
+
+# get total number of cells per final annotation group and set up y_label
+total_cells_df <- full_celltype_df |>
+  dplyr::count(label_recoded, name = "total_cells") |>
+  dplyr::arrange(desc(total_cells)) |>
+  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})"))
+
+singler_order <- total_cells_df$y_label
+total_cells_df$y_label <- factor(total_cells_df$y_label, levels = singler_order)
+nbatlas_bar_order <- total_cells_df$label_recoded
+```
+
+```{r, fig.width = 32, fig.height = 16}
+generate_dotplot(
+  merged_sce = merged_sce,
+  markers_df = top_nbatlas_markers_df,
+  singler_df = full_celltype_df,
+  total_cells_df = total_cells_df,
+  expressed_genes = expressed_genes,
+  bar_order = nbatlas_bar_order, 
+  celltype_palette = nbatlas_celltype_pal, 
+  min_cells = 0
+)
+```
+
+### Consensus validation marker genes
+
+```{r}
 # Prepare data frame with singler labels to plot
-singler_recoded_df <- celltype_recoded_df |>
-  dplyr::filter(
-    annotation_type == "singler",
+singler_recoded_df <- celltype_recoded_wide_df |>
   # we don't consider the NAs here
-    label_recoded != "unknown_singler"
-  ) |>
-  dplyr::select(cell_id, label_recoded)
+  dplyr::filter(singler != "unknown_singler") |>
+  dplyr::select(cell_id, label_recoded = singler)
 
@@
 
-### NBAtlas marker genes
-
-We'll show the top-5 highest log2FC marker genes per cell type for validation.
-
 ```{r}
-# number of marker genes to consider per validation group
-n_marker_genes <- 5
-
-nbatlas_markers_df <- nbatlas_markers_df |>
-  # keep only the top `n_marker_genes`
-  dplyr::filter(direction == "up") |>
-  dplyr::group_by(NBAtlas_label) |>
-  dplyr::mutate(rank_lfc = rank(-avg_log2FC)) |>
-  dplyr::ungroup() |>
-  dplyr::filter(rank_lfc <= n_marker_genes) |>
-  harmonize_celltypes(NBAtlas_label, marker_gene_label) |>
-  dplyr::select(marker_gene_label, gene_symbol, ensembl_gene_id)
-
-nbatlas_bar_order <- total_cells_df$label_recoded
-```
-
-```{r, fig.width = 24, fig.height = 12}
-generate_dotplot(
-  merged_sce,
-  nbatlas_markers_df,
-  singler_recoded_df,
-  total_cells_df,
-  expressed_genes,
-  nbatlas_bar_order
-)
-```
-
-### Consensus validation marker genes
-
-```{r}
-# prepare for dotplot
-
-consensus_markers_df <- consensus_markers_df |>
-  # we'll use this to help determine the bar order
-  harmonize_celltypes(NBAtlas_label, NBAtlas_label_recoded) |>
+# prepare markers for dotplot
+dotplot_consensus_markers_df <- consensus_markers_df |>
+  # consider only matching labels
+  dplyr::filter(validation_group_annotation %in% singler_recoded_df$label_recoded) |>
   dplyr::select(
@@
     ensembl_gene_id,
-    gene_symbol,
-    NBAtlas_label_recoded
+    gene_symbol
   )
@@
   # inner!!!!
-  dplyr::inner_join(consensus_markers_df, by = c("label_recoded" = "NBAtlas_label_recoded")) |>
+  dplyr::inner_join(
+    dotplot_consensus_markers_df, 
+    by = c("label_recoded" = "marker_gene_label")
+  ) |>
   dplyr::arrange(y_label) |>
-  dplyr::pull(marker_gene_label) |>
+  dplyr::pull(label_recoded) |>
   unique()
@@
 
-```{r, fig.width = 22, fig.height = 12}
+```{r, fig.width = 26, fig.height = 12}
 generate_dotplot(
-  merged_sce,
-  consensus_markers_df,
-  singler_recoded_df,
-  total_cells_df,
-  expressed_genes,
-  consensus_bar_order
+  merged_sce = merged_sce,
+  markers_df = dotplot_consensus_markers_df,
+  singler_df = singler_recoded_df,
+  total_cells_df = total_cells_df,
+  expressed_genes = expressed_genes,
+  bar_order = consensus_bar_order, 
+  celltype_palette = recoded_celltype_pal, 
+  min_cells = 0
 )

### 2025-07-15 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
add final section boxplot

@@
 
-Also, note that the `myeloid`-labeled SingleR cells are _only_ neutrophils, but the consensus myeloid cagetory contains additonal cell types.
+Also, note that the `myeloid`-labeled SingleR cells are _only_ neutrophils, but the consensus myeloid grouping contains additional cell types.
 
@@
 
+## Compare expression when SingleR and consensus annotations match
+
+Next, we look at expression of consensus validation marker genes for each normal/harmonized cell type label.
+We compare expression across groups of cells:
+
+* `matching annotations`: Cells whose consensus annotation matches the SingleR annotation
+* `different annotations`: Cells whose consensus annotation differs from the SingleR annotation
+* `consensus not classified`: Cells where the consensus annotation is unknown
+
+Each panel contains only cells that SingleR annotated with the given label, and shows only marker genes associated with that given cell type. 
+We expect that cells whose consensus and SingleR annotations agree may show higher gene expression for the given marker.
+
 ```{r}
@@
 
-situation_df <- celltype_recoded_df |>
+# Data frame with column `compare_annotations` that indicates
+# if consensus and singler (roughly) match
+compare_annotations_df <- celltype_recoded_df |>
   # recode labels for comparison
@@
   dplyr::mutate(
-    annotation_situation = dplyr::case_when(
+    compare_annotations = dplyr::case_when(
       consensus_validation == "unknown_consensus" ~ "consensus not classified", 
@@
       consensus_validation != singler ~ "different annotations"
-    )
+    ), 
+    compare_annotations = forcats::fct_relevel(compare_annotations, "matching annotations", "different annotations", "consensus not classified")
   ) |>
-  # make it long again
-  tidyr::pivot_longer(
-    c(consensus_validation, singler),
-    names_to = "annotation_type", 
-    values_to = "label_recoded"
-  )
+  # we can remove the consensus annotation now
+  dplyr::select(cell_id, singler, compare_annotations)
 
-cells <- unique(situation_df$cell_id)
-genes <- unique(consensus_markers_df$ensembl_gene_id)
+direct_markers_df <- consensus_markers_df |>
+  dplyr::filter(validation_group_annotation %in% direct_celltype_matches) |>
+  dplyr::select(marker_gene_label = validation_group_annotation, ensembl_gene_id, gene_symbol)
+cells <- unique(compare_annotations_df$cell_id)
+genes <- unique(direct_markers_df$ensembl_gene_id)
 ```
@@
 ```{r}
-scuttle::makePerCellDF(
+compare_annotations_expr_df <- scuttle::makePerCellDF(
   merged_sce[genes, cells],
@@
   tidyr::pivot_longer(starts_with("ENSG"), names_to = "ensembl_gene_id", values_to = "logcounts") |>
-  dplyr::left_join(situation_df, by = "cell_id") |>
-  dplyr::group_by(label_recoded, marker_gene_label, ensembl_gene_id) |>
-  dplyr::summarize(
-    detected_count = sum(detected),
-    mean_exp = mean(logcounts)
-  ) |>
-  dplyr::ungroup()
+  # filter to only cells with expression
+  dplyr::filter(logcounts > 0) |>
+  # join other data frames
+  dplyr::left_join(compare_annotations_df, by = "cell_id") |>
+  dplyr::left_join(direct_markers_df, by = "ensembl_gene_id", relationship = "many-to-many")
+```
 
-ggplot(situation_df) + 
-  aes(x = singler, y = )
+```{r}
+unique(compare_annotations_expr_df$singler) |>
+  purrr::map(
+    \(celltype) {
 
-  group_stats_df <- all_info_df |>
-    # remove genes that aren't present in final annotations
-    dplyr::filter(gene_symbol %in% marker_gene_order) |>
-    # for each assigned cell type/marker gene combo get total detected and mean expression
-    dplyr::group_by(label_recoded, marker_gene_label, ensembl_gene_id) |>
-    dplyr::summarize(
-      detected_count = sum(detected),
-      mean_exp = mean(logcounts)
-    ) |>
+      plot_df <- compare_annotations_expr_df |>
+        dplyr::filter(singler == celltype, marker_gene_label == celltype)
 
+      ggplot(plot_df) + 
+        aes(x = gene_symbol, y = logcounts, fill = compare_annotations) +
+        geom_boxplot(outlier.size = 0.25) + 
+        ggtitle(glue::glue("cells annotated as: {celltype}"))
+
+  }) |>
+  patchwork::wrap_plots(ncol = 1) + patchwork::plot_layout(guides = "collect")
 ```

### 2025-07-15 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
added text throughout and adjusted fig size

@@
 
-## Cell type comparisons
+## Cell types in NBAtlas
 
 Throughout this notebook, we compare the cell type labels from `SingleR` with the consensus cell type labels.
-To facilitate this, the `SingleR` labels from `NBAtlas` were grouped together to both a) mirror (where possible) the broad consensus labels, and b) to make some visualizations more manageable.
+To facilitate this, the resulting `SingleR` labels from `NBAtlas` were grouped together to both a) mirror (where possible) the broad consensus labels, and b) to make some visualizations more manageable.
 
@@
 ```
+
+We broadly see high compatibility between SingleR and consensus labels.
+In addition, SingleR mostly classifies the unknown consensus cells as Neuroendocrine, which is consistent with our expectation that these are mostly tumor cells.
 
@@
 
+There appears to be a reasonable correspondence between consensus- and SingleR-colored UMAPS for each cell type.
+We see that SingleR classified more cells compared to consensus, which we expected, and we also see that the additional cells that SingleR labeled are roughly in the same neighborhood as their corresponding consensus labels.
+
 ## Marker gene expression dotplots
@@
 
-We'll show the top-5 highest log2FC marker genes per cell type for validation.
+This dotplot shows the top-5 highest log2FC marker genes per NBAtlas cell type for validation.
+Marker genes are taken directly from the NBAtlas paper where they were identified with `Seurat::FindMarkers()`.
+
+This plot does not show the recoded SingleR labels but rather all labels that have associated marker genes; therefore, there are more cell type categories in this plot than in other plots in this report.
 
@@
 ```
+For validation, we expect to see high marker gene expression along the diagonal of the dotplot where cell type labels match with their marker genes.
+We do broadly see high expression along the diagonal, which means our labeled cells have some amount of similar signal to those in NBAtlas.
+Some sets of marker genes show high expression across multiple cell type categories in particular for closely related cell types (e.g.
+monocyte, macrophage, myeloid, dendritic, or T-cells and NK-cells), but there do not appear to be any strongly unexpected gene expression patterns.
 
 ### Consensus validation marker genes
+
+This dotplot shows expression of consensus validation marker genes across matching cell types labeled with SingleR.
 
@@
 ```
+
+Again, marker gene expression tends to be high in the matching cell type categories, with the possible exception of `B cell` which has somewhat lower gene expression of fewer genes; but, there is still some expected expression there.
+We also see some "promiscuity" where marker genes for closely-related cell types show high expression.
 
@@
 
-```{r}
+```{r, fig.width = 10, fig.height = 20}
 unique(compare_annotations_expr_df$singler) |>
@@
 
+In most cases, gene expression for cells with matching consensus/SingleR annotations is higher or about the same as is gene expression for cells with different annotations.
+
 ## Session Info

### 2025-07-16 analyses/cell-type-osteosarcoma-23/notebook-template.Rmd
Analysis module creation

--- 2025-07-06 analyses/infercnv-osteosarcoma/notebook-template.Rmd
@@
 ---
-title: "infercnv-osteosarcoma Analysis"
+title: "cell-type-osteosarcoma-23 Analysis"
 author:
@@
 # The path to this module
-module_base <- file.path(repository_base, "analyses", "infercnv-osteosarcoma")
+module_base <- file.path(repository_base, "analyses", "cell-type-osteosarcoma-23")
 ```

### 2025-07-16 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
we can use the right merge-sce path now that consensus are there

@@
 data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")
+merged_dir <- file.path(repository_base, "data", "current", "results", "merge-sce", "SCPCP000004")
 ```
@@
 # merged SCE file
-# OBTAINED FROM PORTAL DIRECTLY on 2025-07-09
 sce_file <- file.path(
-  data_dir,
+  merged_dir,
   "SCPCP000004_merged.rds"
@@
 
-# RUH ROH!!!
-# https://github.com/AlexsLemonade/scpcaTools/issues/298
+# Ensure cell_id is correct
 merged_sce$cell_id <- colnames(merged_sce)

### 2025-07-16 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
all the SingleR into backticks, and programmable table

@@
 This notebooks explores the results from running cell type annotation with `SingleR` using the NBAtlas.
-The NBAtlas reference was aggregated with `SingelR` prior to model training.
+The NBAtlas reference was aggregated with `SingleR` prior to model training.
 
@@
 sce_file <- file.path(
-  merged_dir,
+  data_dir,
   "SCPCP000004_merged.rds"
@@
   ) |> 
-  harmonize_celltypes(label, label_recoded) |>
-  dplyr::select(-label)
+  harmonize_celltypes(label, label_recoded)
 ```
@@
 The table below shows the `NBAtlas` cell types and how they are represented in visualizations, unless otherwise stated.
+Cells labeled `singler_unknown` are those for which `SingleR` could not reliably assign an annotation.
 
-| Broad label shown in figures | Associated NBAtlas labels                                       |
-| ---------------------------- | --------------------------------------------------------------- |
-| `T cell`                     | `CD4+ T cell`, `CD8+ T cell`, `Treg`, `NKT cell`                |
-| `dendritic cell`             | `cDC1`, `cDC2/DC3`, `Migratory cDC`                             |
-| `natural killer cell cell`   | `Circulating NK cell`, `TOX2+/KIT+ NK cell`, `Resident NK cell` |
-| `monocyte`                   | `CD4+ T cell`, `Classical monocyte`, `Patrolling monocyte`      |
-| `myeloid`                    | `Neutrophil`                                                    |
-| `Neuroendocrine`             | `Neuroendocrine`                                                |
-| `fibroblast`                 | `Fibroblast`                                                    |
-| `macrophage`                 | `Macrophage`                                                    |
-| `B cell`                     | `B cell`                                                        |
-| `plasma cell`                | `Plasma`                                                        |
-| `endothelial`                | `Endothelial`                                                   |
-| `Stromal other (NBAtlas)`    | `Stromal other`                                                 |
-| `Schwann`                    | `Schwann`                                                       |
-| `RBCs`                       | `RBCs`                                                          |
-| `Immune cycling`             | `Immune cycling`                                                |
-| `pDC`                        | `pDC`                                                           |
-
-Cells labeled `singler_unknown` are those for which `SingleR` could not reliably assign an annotation.
+```{r}
+celltype_recoded_df |>
+  dplyr::filter(annotation_type == "singler") |>
+  dplyr::select(label, label_recoded) |>
+  dplyr::distinct() |>
+  dplyr::group_by(label_recoded) |>
+  dplyr::summarize(all_labels = paste(label, collapse = ", ")) |>
+  dplyr::arrange(label_recoded) |>
+  dplyr::rename(
+    "Label shown in figures" = label_recoded,
+    "Underlying NBAltas label(s)" = all_labels
+  ) |>
+  knitr::kable()
+```
 
@@
 
-This section compares SingleR annotations to consensus cell type annotations using a heatmap.
+This section compares `SingleR` annotations to consensus cell type annotations using a heatmap.
 The heatmap is colored by Jaccard similarity index.
@@
 
-We broadly see high compatibility between SingleR and consensus labels.
+We broadly see high compatibility between `SingleR` and consensus labels.
 In addition, SingleR mostly classifies the unknown consensus cells as Neuroendocrine, which is consistent with our expectation that these are mostly tumor cells.
@@
 
-This section visualizes and compares SingleR annotations to consensus cell type annotations using UMAPs.
+This section visualizes and compares `SingleR` annotations to consensus cell type annotations using UMAPs.
 Note that the displayed UMAP is from a merged object that has _not been batch-corrected._
@@
 Note that each set of labels has its own "stromal" category which the labels distinguish.
-In addition, cells labeled `unknown_consensus` are those with no assigned consensus label, and cells labeled `unknown_singler` are those where SingleR could not confidently assign a label.
+In addition, cells labeled `unknown_consensus` are those with no assigned consensus label, and cells labeled `unknown_singler` are those where `SingleR` could not confidently assign a label.
 
@@
 
-First, we display the consensus and SingleR annotations for all cells.
+First, we display the consensus and `SingleR` annotations for all cells.
 
@@
 
-Below we display a faceted UMAP to highlight the SingleR annotations.
+Below we display a faceted UMAP to highlight the `SingleR` annotations.
 Light gray cells in each panel represent all other cell types.
@@
 
-Below, we display faceted UMAPs highlighting a single cell type but considering only cell types that have direct correspondence between SingleR and consensus cell types.
-This allows us to see if the normal cells that SingleR is labeling correspond well to the normal cells identified by consensus labels.
-Each row displays a cell type where the left column shows the consensus version and the right column shows the SingleR version.
+Below, we display faceted UMAPs highlighting a single cell type but considering only cell types that have direct correspondence between `SingleR` and consensus cell types.
+This allows us to see if the normal cells that `SingleR` is labeling correspond well to the normal cells identified by consensus labels.
+Each row displays a cell type where the left column shows the consensus version and the right column shows the `SingleR` version.
 
-In addition, we include a category below `putative-tumor` to directly compare the unknown consensus labels with the `Neuroendocrine` cells labeled by SingleR. 
+In addition, we include a category below `putative-tumor` to directly compare the unknown consensus labels with the `Neuroendocrine` cells labeled by `SingleR.` 
 While these categories are not necessarily directly comparable, they are each most likely to contain tumor cells.
 
-Also, note that the `myeloid`-labeled SingleR cells are _only_ neutrophils, but the consensus myeloid grouping contains additional cell types.
+Also, note that the `myeloid`-labeled `SingleR` cells are _only_ neutrophils, but the consensus myeloid grouping contains additional cell types.
 
@@
 
-There appears to be a reasonable correspondence between consensus- and SingleR-colored UMAPS for each cell type.
-We see that SingleR classified more cells compared to consensus, which we expected, and we also see that the additional cells that SingleR labeled are roughly in the same neighborhood as their corresponding consensus labels.
+There appears to be a reasonable correspondence between consensus- and `SingleR`-colored UMAPS for each cell type.
+We see that `SingleR` classified more cells compared to consensus, which we expected, and we also see that the additional cells that `SingleR` labeled are roughly in the same neighborhood as their corresponding consensus labels.
 
@@
 * Consensus cell type marker genes
-  * This will provide an independent assessment of the reliability of SingleR normal cell type assignments
+  * This will provide an independent assessment of the reliability of `SingleR` normal cell type assignments
 
@@
 
-This plot does not show the recoded SingleR labels but rather all labels that have associated marker genes; therefore, there are more cell type categories in this plot than in other plots in this report.
+This plot does not show the recoded `SingleR` labels but rather all labels that have associated marker genes; therefore, there are more cell type categories in this plot than in other plots in this report.
 
@@
 
-This dotplot shows expression of consensus validation marker genes across matching cell types labeled with SingleR.
+This dotplot shows expression of consensus validation marker genes across matching cell types labeled with `SingleR`.
 
@@
 
-* `matching annotations`: Cells whose consensus annotation matches the SingleR annotation
-* `different annotations`: Cells whose consensus annotation differs from the SingleR annotation
+* `matching annotations`: Cells whose consensus annotation matches the `SingleR` annotation
+* `different annotations`: Cells whose consensus annotation differs from the `SingleR` annotation
 * `consensus not classified`: Cells where the consensus annotation is unknown
 
-Each panel contains only cells that SingleR annotated with the given label, and shows only marker genes associated with that given cell type. 
-We expect that cells whose consensus and SingleR annotations agree may show higher gene expression for the given marker.
+Each panel contains only cells that `SingleR` annotated with the given label, and shows only marker genes associated with that given cell type. 
+We expect that cells whose consensus and `SingleR` annotations agree may show higher gene expression for the given marker.
 
@@
 
-In most cases, gene expression for cells with matching consensus/SingleR annotations is higher or about the same as is gene expression for cells with different annotations.
+In most cases, gene expression for cells with matching consensus/`SingleR` annotations is higher or about the same as is gene expression for cells with different annotations.
 

### 2025-07-16 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
reorder nbatlas dotplot

@@
   dplyr::select(-consensus_annotation) |>
-  dplyr::left_join(singler_df, by = c("cell_id", "barcodes", "library_id"))
+  dplyr::left_join(singler_df, by = c("cell_id", "barcodes", "library_id")) |>
+  # we do specifically need to change pruned.labels from NE -> Neuroendocrine
+  dplyr::mutate(
+    pruned.labels = ifelse(pruned.labels == "NE", "Neuroendocrine", pruned.labels)
+  )
 
@@
   harmonize_celltypes(label, label_recoded)
+
 ```
@@
 ```{r}
-celltype_recoded_df |>
-  dplyr::filter(annotation_type == "singler") |>
+label_map_df <- celltype_recoded_df |>
+  dplyr::filter(annotation_type == "singler", label != "unknown_singler") |>
   dplyr::select(label, label_recoded) |>
-  dplyr::distinct() |>
+  dplyr::distinct()
+
+label_map_df |>
   dplyr::group_by(label_recoded) |>
@@
 full_celltype_df <- celltype_df |>
-  dplyr::select(
-    cell_id, 
-    # the dotplot function expects this column name
-    label_recoded = pruned.labels
+  dplyr::filter(pruned.labels != "unknown_singler") |>
+  dplyr::select(cell_id, label = pruned.labels) |>
+  # collapse certain labels that are grouped in the marker genes
+  dplyr::mutate(
+   label = dplyr::case_when(
+     stringr::str_detect(label, "cDC") ~ "cDC",
+     stringr::str_detect(label, "monocyte") ~ "Monocyte",
+     .default = label
+  ))
+
+# Also update label_map_df to collapse those labels
+label_map_df <- label_map_df |>
+  dplyr::mutate(
+   label = dplyr::case_when(
+     stringr::str_detect(label, "cDC") ~ "cDC",
+     stringr::str_detect(label, "monocyte") ~ "Monocyte",
+     .default = label)
   ) |>
-  dplyr::mutate(
-    label_recoded = dplyr::case_when(
-      # collapse the dendritic cells & monocytes to match what's in the marker genes
-      stringr::str_detect(label_recoded, "cDC") ~ "cDC", 
-      stringr::str_detect(label_recoded, "monocyte") ~ "Monocyte",
-      label_recoded == "NE" ~ "Neuroendocrine",
-      .default = label_recoded
-  )) |>
-  dplyr::filter(label_recoded != "unknown_singler")
+  unique()
+
+# Prepare an order of label_recoded to help set the order for the literal underlying labels
+label_coded_factor <- celltype_recoded_df |>
+  dplyr::filter(annotation_type == "singler", label_recoded != "unknown_singler") |>
+  dplyr::count(label_recoded, name = "total_cells") |>
+  dplyr::arrange(desc(total_cells)) |>
+  dplyr::pull(label_recoded)
+label_recoded_factor <- factor(label_coded_factor, levels = label_coded_factor)
 
@@
 total_cells_df <- full_celltype_df |>
-  dplyr::count(label_recoded, name = "total_cells") |>
-  dplyr::arrange(desc(total_cells)) |>
-  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})"))
+  dplyr::count(label, name = "total_cells") |>
+  # order so cell types are grouped
+  dplyr::inner_join(label_map_df, by = "label" ) |>
+  dplyr::mutate(
+    label_recoded = as.factor(label_recoded),
+    label_recoded = forcats::fct_relevel(label_recoded, levels(label_recoded_factor))
+  ) |>
+  dplyr::arrange(label_recoded) |>
+  dplyr::mutate(y_label = glue::glue("{label} ({total_cells})")) |>
+  # rename as expected for dotplot function - label_recoded needs to be the main column
+  dplyr::select(-label_recoded) |>
+  dplyr::rename(label_recoded = label)
 
-singler_order <- total_cells_df$y_label
-total_cells_df$y_label <- factor(total_cells_df$y_label, levels = singler_order)
+total_cells_df$y_label <- factor(total_cells_df$y_label, levels =  total_cells_df$y_label)
 nbatlas_bar_order <- total_cells_df$label_recoded
@@
   markers_df = top_nbatlas_markers_df,
-  singler_df = full_celltype_df,
+  # rename this after we've done all the wrangling
+  singler_df = full_celltype_df |> dplyr::rename(label_recoded = label),
   total_cells_df = total_cells_df,

### 2025-07-16 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
Update ordering for consensus dotplot

@@
 
-total_cells_df$y_label <- factor(total_cells_df$y_label, levels =  total_cells_df$y_label)
+total_cells_df$y_label <- factor(total_cells_df$y_label, levels = rev(total_cells_df$y_label))
 nbatlas_bar_order <- total_cells_df$label_recoded
@@
 
+Importantly, these marker genes are not necessarily canonical marker genes for the normal/immune cell types present, but were empirically-determined based on the `NBAtlas` expression patterns.
+There may therefore be overlapping marker genes between categories, and the given marker genes may not be the "most" precise for the cell types if analyzed independently.
+
 ### Consensus validation marker genes
@@
 This dotplot shows expression of consensus validation marker genes across matching cell types labeled with `SingleR`.
+Harmonized `SingleR` labels with direct matches are shown on the top of the y-axis (`endothelial cell` through `plasma cell`), and the remaining `SingleR` labels are below.
 
@@
 
-# get total number of cells per final annotation group and set up y_label
+# to match up consensus and NBAtlas for ordering
+direct_celltype_matches <- c(
+  "B cell",
+  "T cell",
+  "myeloid",
+  "macrophage", 
+  "monocyte",
+  "fibroblast",
+  "dendritic cell",
+  "endothelial cell",
+  "plasma cell",
+  "natural killer cell"
+)
+
 total_cells_df <- singler_recoded_df |>
   dplyr::count(label_recoded, name = "total_cells") |>
-  dplyr::arrange(desc(total_cells)) |>
+  # labels that appear in both consensus+nbatlas should appear first
+  dplyr::mutate(in_both = label_recoded %in% direct_celltype_matches) |>
+  dplyr::group_by(in_both) |>
+  dplyr::arrange(total_cells, .by_group = TRUE) |>
+  dplyr::ungroup() |>
   dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})"))
+total_cells_df$y_label <- factor(total_cells_df$y_label, levels = total_cells_df$y_label)
 
-singler_order <- total_cells_df$y_label
-total_cells_df$y_label <- factor(total_cells_df$y_label, levels = singler_order)
+# get the annotation bar order
+consensus_bar_order <- total_cells_df |>
+  dplyr::filter(in_both) |>
+  dplyr::pull(label_recoded) |>
+  rev()
 ```
@@
   )
-
-# get the bar order
-consensus_bar_order <- total_cells_df |>
-  dplyr::select(label_recoded, y_label) |>
-  # inner!!!!
-  dplyr::inner_join(
-    dotplot_consensus_markers_df, 
-    by = c("label_recoded" = "marker_gene_label")
-  ) |>
-  dplyr::arrange(y_label) |>
-  dplyr::pull(label_recoded) |>
-  unique()
 ```

### 2025-07-16 analyses/cell-type-neuroblastoma-04/scripts/utils/celltype-utils.R

@@ -163,7 +163,7 @@ generate_dotplot <- function(
   # define y- and x- axis orders
 
   # y axis - singler
-  y_label_order <- rev(total_cells_df$y_label)
+  y_label_order <- levels(total_cells_df$y_label)
 
   # x axis - marker genes. this uses the passed in bar_order
   markers_df <- markers_df |>

### 2025-07-16 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
add final plot and helper function for subsetting nbatlas marker genes

@@
 # Source additional utilities functions:
+# - select_nbatlas_markers()
 # - harmonize_celltypes()
@@
     "Underlying NBAltas label(s)" = all_labels
-  ) |>
-  knitr::kable()
+  )
 ```
@@
 
-```{r, fig.height = 10, fig.width = 10} 
+```{r, fig.height = 8, fig.width = 8} 
 # Pivot wider for heatmap functions
 celltype_recoded_wide_df <- celltype_recoded_df |>
+  dplyr::select(-label) |>
   tidyr::pivot_wider(
@@
 ```{r}
-# number of marker genes to consider per validation group
-n_marker_genes <- 5
-
-top_nbatlas_markers_df <- nbatlas_markers_df |>
-  # keep only the top `n_marker_genes`
-  dplyr::filter(direction == "up") |>
-  dplyr::group_by(NBAtlas_label) |>
-  dplyr::mutate(rank_lfc = rank(-avg_log2FC)) |>
-  dplyr::ungroup() |>
-  dplyr::filter(rank_lfc <= n_marker_genes) |>
-  dplyr::select(marker_gene_label = NBAtlas_label, gene_symbol, ensembl_gene_id)
+# data frame of top 5 upregulated genes per cell type
+top_nbatlas_markers_df <- subset_nbatlas_markers(nbatlas_markers_df, 5, "up")
 ```
@@
 Next, we look at expression of consensus validation marker genes for each normal/harmonized cell type label.
-We compare expression across groups of cells:
+We're specifically interested in the marker gene expression for cells that were classified by `SingleR` but have an unknown consensus cell type.
+We assume the unknown consensus cells may be more likely to be tumor because they could not be robustly classified otherwise, but since some of those cells were called "normal" by `SingleR` we'd like to get a sense of whether they really are normal.
 
-* `matching annotations`: Cells whose consensus annotation matches the `SingleR` annotation
-* `different annotations`: Cells whose consensus annotation differs from the `SingleR` annotation
-* `consensus not classified`: Cells where the consensus annotation is unknown
+We have two groups:
 
-Each panel contains only cells that `SingleR` annotated with the given label, and shows only marker genes associated with that given cell type. 
-We expect that cells whose consensus and `SingleR` annotations agree may show higher gene expression for the given marker.
+* `SingleR` cells classified as `Neuroendocrine`; these are putatively tumor cells
+  * We expect that these will have higher `Neuroendocrine` marker gene expression and lower normal gene expression
+* `SingleR` cells not classified as `Neuroendocrine`; these are more likely to be normal cells
+  * We expect that these will have higher normal marker gene expression
+
+For this, we'll consider the average gene expression among the top 20 NBAtlas marker genes for each cell type.
 
 ```{r}
-# this time we are not considering putative tumor
-direct_celltype_matches <- c(
-  "B cell",
-  "T cell",
-  # note that we need to collapse the consensus to compare
-  "myeloid",
-  "fibroblast",
-  "endothelial cell",
-  "plasma cell",
-  "natural killer cell", 
-  "unknown_consensus" # we do want to keep these for this analysis
-)
+# data frame of top 20 upregulated genes per cell type
+top_nbatlas_markers_df <- subset_nbatlas_markers(nbatlas_markers_df, 20, "up")
 
-# Define the different myeloid types in consensus annotations
-consensus_myeloid <- c("myeloid", "dendritic cell", "monocyte", "macrophage")
+unknown_consensus_ids <- celltype_recoded_wide_df |>
+  dplyr::filter(consensus_validation == "unknown_consensus") |>
+  dplyr::pull(cell_id)
 
-# Data frame with column `compare_annotations` that indicates
-# if consensus and singler (roughly) match
-compare_annotations_df <- celltype_recoded_df |>
-  # recode labels for comparison
+unknown_consensus_celltype_df <- full_celltype_df |>
+  dplyr::filter(cell_id %in% unknown_consensus_ids) |>
   dplyr::mutate(
-    label_recoded = ifelse(
-      label_recoded %in% consensus_myeloid,
-      "myeloid",
-      label_recoded
-  )) |>
-  dplyr::filter(label_recoded %in% direct_celltype_matches) |>
-  dplyr::select(cell_id, annotation_type, label_recoded) |>
-  # make it wider to label the situation
-  tidyr::pivot_wider(
-    names_from = "annotation_type",
-    values_from = "label_recoded"
+    singler = ifelse(
+      label == "Neuroendocrine", "NE (likely tumor)", "Non-NE (likely normal)")
   ) |>
-  # remove NAs:
-  # - consensus NAs are those with other cell types not to include here
-  # - singler NAs were unclassified cells
-  tidyr::drop_na() |>
-  dplyr::mutate(
-    compare_annotations = dplyr::case_when(
-      consensus_validation == "unknown_consensus" ~ "consensus not classified", 
-      consensus_validation == singler ~ "matching annotations",
-      consensus_validation != singler ~ "different annotations"
-    ), 
-    compare_annotations = forcats::fct_relevel(compare_annotations, "matching annotations", "different annotations", "consensus not classified")
-  ) |>
-  # we can remove the consensus annotation now
-  dplyr::select(cell_id, singler, compare_annotations)
+  dplyr::select(-label)
 
-direct_markers_df <- consensus_markers_df |>
-  dplyr::filter(validation_group_annotation %in% direct_celltype_matches) |>
-  dplyr::select(marker_gene_label = validation_group_annotation, ensembl_gene_id, gene_symbol)
-cells <- unique(compare_annotations_df$cell_id)
-genes <- unique(direct_markers_df$ensembl_gene_id)
-```
+goi <- unique(top_nbatlas_markers_df$ensembl_gene_id)
 
-```{r}
 compare_annotations_expr_df <- scuttle::makePerCellDF(
-  merged_sce[genes, cells],
-  features = genes,
+  merged_sce[goi, unknown_consensus_ids],
+  features = goi,
   assay.type = "logcounts",
@@
   dplyr::filter(logcounts > 0) |>
-  # join other data frames
-  dplyr::left_join(compare_annotations_df, by = "cell_id") |>
-  dplyr::left_join(direct_markers_df, by = "ensembl_gene_id", relationship = "many-to-many")
+  # join the annotations and marker gene data frame; use inner since we've filtered already
+  dplyr::inner_join(unknown_consensus_celltype_df, by = "cell_id") |>
+  dplyr::left_join(top_nbatlas_markers_df, by = "ensembl_gene_id", relationship = "many-to-many") |>
+  # average by marker_gene_label
+  dplyr::group_by(singler, marker_gene_label, ensembl_gene_id) |>
+  dplyr::summarize(mean_exp = mean(logcounts)) |>
+  dplyr::ungroup() |>
+  dplyr::mutate(marker_gene_label = factor(marker_gene_label), 
+                marker_gene_label = forcats::fct_relevel(marker_gene_label, nbatlas_bar_order))
+
 ```
 
-```{r, fig.width = 10, fig.height = 20}
-unique(compare_annotations_expr_df$singler) |>
-  purrr::map(
-    \(celltype) {
-
-      plot_df <- compare_annotations_expr_df |>
-        dplyr::filter(singler == celltype, marker_gene_label == celltype)
-
-      ggplot(plot_df) + 
-        aes(x = gene_symbol, y = logcounts, fill = compare_annotations) +
-        geom_boxplot(outlier.size = 0.25) + 
-        ggtitle(glue::glue("cells annotated as: {celltype}"))
-
-  }) |>
-  patchwork::wrap_plots(ncol = 1) + patchwork::plot_layout(guides = "collect")
+```{r, fig.width = 12, fig.height = 5}
+ggplot(compare_annotations_expr_df) + 
+  aes(x = marker_gene_label, y = mean_exp, color = singler) + 
+  ggforce::geom_sina(position = position_dodge(width = 0.7)) + 
+  labs(
+    x = "NBAtlas marker gene label", 
+    y = "Mean expression across marker genes",
+    color = "SingleR annotation"
+  ) + 
+  theme(
+    axis.text.x = element_text(angle = 45, hjust = 1)
+  )
 ```
-
-In most cases, gene expression for cells with matching consensus/`SingleR` annotations is higher or about the same as is gene expression for cells with different annotations.
+The left-most gene set on the x-axis is `Neuroendocrine`, and normal cell types follow.
+This plot suggests that the normal cells `SingleR` is picking up are likely to be normal cells: normal gene expression for non-NE cells is higher compared to NE cells, and we see the opposite trend for the `Neuroendocrine` marker genes.
 

### 2025-07-16 analyses/cell-type-neuroblastoma-04/scripts/utils/celltype-utils.R

@@ -1,7 +1,33 @@
 # This file contains the following functions:
+# - select_nbatlas_markers()
 # - harmonize_celltypes()
 # - faceted_umap()
 # - generate_dotplot()
+
+
+
+#' Helper function to prepare the NBAtlas marker data frame
+#'
+#' @param nbatlas_markers_df Unmanipulated data frame of NBAtlas markers
+#' @param n_genes Number of marker genes to keep per cell type
+#' @param directon Whether up or downregulated genes should be considered; default is "up"
+#'
+#' @returns Data frame with columns `marker_gene_label`, `gene_symbol`, and `ensembl_gene_id`
+subset_nbatlas_markers <- function(nbatlas_markers_df, n_genes, direction = c("up", "down")) {
+
+  direction <- match.arg(direction)
+
+  nbatlas_markers_df |>
+    # keep only the top `n_marker_genes`
+    dplyr::filter(direction == "up") |>
+    dplyr::group_by(NBAtlas_label) |>
+    dplyr::mutate(rank_lfc = rank(-avg_log2FC)) |>
+    dplyr::ungroup() |>
+    dplyr::filter(rank_lfc <= n_genes) |>
+    dplyr::select(marker_gene_label = NBAtlas_label, gene_symbol, ensembl_gene_id)
+}
+
+
 
 #' Function to update NBAtlas labels
 #' This function matches NBAtlas labels to consensus labels where appropriate to facilitate plots

### 2025-07-17 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd

@@
 ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(0.6, "in"))
+
+# settings
+options(
+  dplyr.summarise.inform = FALSE, 
+  readr.show_col_types = FALSE
+)
 ```
@@
 sce_file <- file.path(
-  data_dir,
+  merged_dir,
   "SCPCP000004_merged.rds"
@@
 
-## Compare expression when SingleR and consensus annotations match
+## Gene expression of unclassified consensus cells
 

### 2025-07-17 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-testing/filter-genes.Rmd

--- 2025-07-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/filter-genes.Rmd
@@
 
-* Endothelial --> Endo
-* Fibroblasts --> Fibro
-* Neuroendocrine --> NE
-* Stromal other --> stromal
+* `Endothelial` --> `Endo`
+* `Fibroblasts` --> `Fibro`
+* `Neuroendocrine` --> `NE`
+* `Stromal other` --> `stromal`
 

### 2025-07-19 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd

@@
 
-# only the NBAtlas labels - use for validation dotplot
+# only the NBAtlas labels - use for validation dot plot
 nbatlas_palette_df <- readr::read_tsv(nbatlas_palette_file)
@@
 
-## Marker gene expression dotplots
+## Marker gene expression dot plots
 
@@
 
-This dotplot shows the top-5 highest log2FC marker genes per NBAtlas cell type for validation.
+This dot plot shows the top-5 highest `log2FC` marker genes per NBAtlas cell type for validation.
 Marker genes are taken directly from the NBAtlas paper where they were identified with `Seurat::FindMarkers()`.
@@
 ```
-For validation, we expect to see high marker gene expression along the diagonal of the dotplot where cell type labels match with their marker genes.
+For validation, we expect to see high marker gene expression along the diagonal of the dot plot where cell type labels match with their marker genes.
 We do broadly see high expression along the diagonal, which means our labeled cells have some amount of similar signal to those in NBAtlas.
@@
 
-This dotplot shows expression of consensus validation marker genes across matching cell types labeled with `SingleR`.
+This dot plot shows expression of consensus validation marker genes across matching cell types labeled with `SingleR`.
 Harmonized `SingleR` labels with direct matches are shown on the top of the y-axis (`endothelial cell` through `plasma cell`), and the remaining `SingleR` labels are below.

### 2025-07-24 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
remove the really low-quality libraries, and a gitleaks spot to override

@@
 assay(merged_sce, "counts") <- NULL
+reducedDim(merged_sce, "PCA") <- NULL
 
-# Ensure cell_id is correct
-merged_sce$cell_id <- colnames(merged_sce)
+# Define libraries to remove
+# See: https://github.com/AlexsLemonade/OpenScPCA-analysis/issues/1234#issuecomment-3113966395
+remove_libraries <- c("SCPCL000124", "SCPCL001058")
+keep_cells <- grep(
+  paste(remove_libraries, collapse="|"), 
+  colnames(merged_sce), 
+  invert = TRUE,
+  value = TRUE
+)
+
+merged_sce <- merged_sce[, keep_cells]
+rm(keep_cells)
 ```
@@
 singler_df <- singler_files |>
+  # remove libraries
+  purrr::discard_at(remove_libraries) |>
   purrr::map(readr::read_tsv) |>

### 2025-07-24 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
add brief section looking at smaller libraries

@@
   )
-
-heatmap_col_fun <- circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue"))
 
@@
 
+## Inferences from libraries with lower cell counts
+
+In this section, we briefly look at annotation results for libraries with fewer than 1000 cells which may be lower quality than other libraries.
+
+Those libraries are:
+
+```{r}
+libraries_low_cells <- celltype_recoded_df |>
+  dplyr::filter(annotation_type == "singler") |>
+  dplyr::select(-annotation_type) |>
+  dplyr::count(library_id, name = "total_cells") |>
+  dplyr::filter(total_cells < 1000) 
+libraries_low_cells
+```
+
+### Heatmap
+
+This section shows a heatmap comparing annotations from these libraries only to the consensus annotations.
+
+```{r, fig.width = 8, fig.height = 8}
+make_jaccard_heatmap(
+  celltype_recoded_wide_df |> dplyr::filter(library_id %in% libraries_low_cells$library_id),
+  "consensus_validation",
+  "singler",
+  "Consensus validation label",
+  "SingleR NBAtlas label"
+)
+```
+
+The heatmap comparison to consensus cell types for these libraries looks reasonable; corresponding cell types tend to match up.
+Note that because this heatmap includes fewer libraries, there are fewer cell types to display overall.
+
+### Percent of unclassified cells
+
+Next, we'll ask whether the percent of unclassified cells (aka, `unknown_singler`) is much higher in these libraries compared to larger libraries.
+
+```{r}
+unknown_singler_df <- celltype_recoded_wide_df |>
+  dplyr::group_by(library_id) |>
+  dplyr::summarize(unknown_cells= sum(singler == "unknown_singler"))
+
+celltype_recoded_wide_df |>
+  dplyr::add_count(library_id, name = "total_cells") |>
+  dplyr::inner_join(unknown_singler_df, by = "library_id") |>
+  dplyr::select(library_id, total_cells, unknown_cells) |>
+  dplyr::distinct() |>
+  dplyr::mutate(percent_singler_unclassified = round(unknown_cells/total_cells, 2)) |>
+  dplyr::arrange(total_cells)
+```
+
+Most cells in the relatively smaller libraries are classified, and the percentage of unclassified cells (when non-0) is fully in line with the larger libraries.
+
 ## Session Info

### 2025-07-24 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
add barplots and PDX section

@@
 
-## Inferences from libraries with lower cell counts
+## Inferences from smaller vs. larger libraries
 
@@
 ```
+
+### Distirbution of cell types
+
+```{r, fig.width = 10, fig.height = 4}
+celltype_recoded_wide_df |>
+  dplyr::mutate(
+    library_size = ifelse(library_id %in% libraries_low_cells$library_id, "<1000 cells", ">1000 cells")
+  ) |>
+  ggplot() + 
+  aes(x = library_id, fill = singler) + 
+  geom_bar(position = "fill") + 
+  scale_fill_manual(values = recoded_celltype_pal) + 
+  facet_wrap(vars(library_size), scales = "free_x") + 
+  scale_y_continuous(expand = c(0,0)) +
+  theme(
+    legend.position = "bottom",
+    axis.text.x = element_blank()
+  )
+```
+Although there are only four libraries with <1000 cells so it is challenging to make definitive conclusions, it does not seem like the overall cell type distributions are very different between smaller and larger libraries.
 
@@
 
-Next, we'll ask whether the percent of unclassified cells (aka, `unknown_singler`) is much higher in these libraries compared to larger libraries.
+Next, we'll ask whether the percent of unclassified cells (aka, `unknown_singler`) can be explained by library size.
+Below, we plot the relationship between total cells and fraction of cells that SingleR failed to confidently classify.
 
@@
 
-celltype_recoded_wide_df |>
+percent_unclassified_df <- celltype_recoded_wide_df |>
   dplyr::add_count(library_id, name = "total_cells") |>
@@
   dplyr::distinct() |>
-  dplyr::mutate(percent_singler_unclassified = round(unknown_cells/total_cells, 2)) |>
-  dplyr::arrange(total_cells)
+  dplyr::mutate(frac_singler_unclassified = unknown_cells/total_cells)
+
+ggplot(percent_unclassified_df) + 
+  aes(x = total_cells, y = frac_singler_unclassified) + 
+  geom_point() + 
+  geom_smooth(method = "lm")
 ```
 
-Most cells in the relatively smaller libraries are classified, and the percentage of unclassified cells (when non-0) is fully in line with the larger libraries.
+There doesn't appear to be a relationship here, and further the percent of unclassified cells are all under 7\% with most below around 4\%.
+
+## Inferences from PDX libraries
+
+In this section, we briefly look at annotation results for libraries from PDX samples, which may show distinct trends.
+Here is the breakdown of the number of PDX vs tissue libraries:
+
+```{r}
+# This was being developed while this bug was present, so we have to check the type first
+# https://github.com/AlexsLemonade/scpcaTools/issues/302
+pdx_class <- class(colData(merged_sce)$is_xenograft)
+if (pdx_class == "character") {
+  pdx_val <- "TRUE"
+} else {
+  pdx_val <- TRUE
+}
+
+pdx_libraries <- colData(merged_sce) |>
+  as.data.frame() |>
+  dplyr::filter(is_xenograft == pdx_val) |>
+  dplyr::pull(library_id) |>
+  unique()
+
+celltype_recoded_wide_df <- celltype_recoded_wide_df |>
+  dplyr::mutate(sample_type = ifelse(library_id %in% pdx_libraries, "pdx", "tissue"))
+
+celltype_recoded_wide_df |>
+  dplyr::select(library_id, sample_type) |>
+  dplyr::distinct() |>
+  dplyr::count(sample_type)
+```
+
+### Distribution of cell types
+
+We'll first look at the cell type distributions for each library between sample types:
+
+```{r, fig.width = 14, fig.height = 6}
+ggplot(celltype_recoded_wide_df) + 
+  aes(x = library_id, fill = singler) + 
+  geom_bar(position = "fill") + 
+  scale_fill_manual(values = recoded_celltype_pal) + 
+  facet_wrap(vars(sample_type), scales = "free_x") + 
+  scale_y_continuous(expand = c(0,0)) +
+  theme(
+    legend.position = "bottom",
+    axis.text.x = element_blank()
+  )
+```
+
+PDX samples have, not unexpectedly, much less cell type diversity overall with most cells being tumor or similar and immune cells rarely present.
+This is consistent with expectations in general for the biology of PDX's.
+
+### Heatmap
+
+This section shows a heatmap compaing SingleR annotations to consensus cell types, for PDX samples only.
+
+```{r, fig.width = 8, fig.height = 8}
+make_jaccard_heatmap(
+  celltype_recoded_wide_df |> dplyr::filter(sample_type == "pdx"),
+  "consensus_validation",
+  "singler",
+  "Consensus validation label",
+  "SingleR NBAtlas label"
+)
+```
+This heatmap comparison seems reasonable, again taking into consideration that there are relatively fewer cell types in the PDX samples, and corresponding cell types tend to match up.
 

### 2025-07-24 analyses/cell-type-ewings/template_notebooks/auc-singler-workflow/02-singler-results.Rmd
sort order and make sure at least 2 levels to compare

@@
 # these should be tumor normal but sometimes that's not the case
-singler_cell_types <- unique(caret_df$singler_tumor_normal) |> as.character()
-aucell_cell_types <- unique(caret_df$aucell_annotation) |> as.character()
+singler_cell_types <- unique(caret_df$singler_tumor_normal) |> as.character() |> sort()
+aucell_cell_types <- unique(caret_df$aucell_annotation) |> as.character() |> sort()
 
 # only calculate confusion matrix if all cell types are the same
-if (all(singler_cell_types == aucell_cell_types)) {
+if (all(singler_cell_types == aucell_cell_types) & length(singler_cell_types) > 1) {
   caret::confusionMatrix(

### 2025-07-25 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
Update analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

@@
 ```{r, fig.width = 8, fig.height = 8}
+low_cells_df <- celltype_recoded_wide_df |> 
+  dplyr::filter(library_id %in% libraries_low_cells$library_id)
+
 make_jaccard_heatmap(
-  celltype_recoded_wide_df |> dplyr::filter(library_id %in% libraries_low_cells$library_id),
+  low_cells_df,
   "consensus_validation",

### 2025-07-28 analyses/cell-type-ewings/template_notebooks/auc-singler-workflow/02-singler-results.Rmd

--- 2024-12-13 analyses/cell-type-ewings/template_notebooks/auc-singler-workflow/02-singler-results.Rmd
@@
 # only calculate confusion matrix if all cell types are the same
-if (all(singler_cell_types == aucell_cell_types)) {
+if (setequal(singler_cell_types, aucell_cell_types) & length(singler_cell_types) > 1) {
   caret::confusionMatrix(

### 2025-07-31 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd

@@
 # list all results files 
-results_files <- list.files(results_dir, pattern = "_consensus-cell-types\\.tsv.\\gz$", full.names = TRUE)
+results_files <- list.files(results_dir, pattern = "_processed_consensus-cell-types\\.tsv.\\gz$", recursive = TRUE, full.names = TRUE)
 
-# get project ids from file list  
-project_ids <- stringr::str_remove(basename(results_files), "_consensus-cell-types.tsv.gz")
-names(results_files) <- project_ids
-
-# remove cell line projects from file list
+# define cell line projects to remove
 cell_line_projects <- c("SCPCP000020", "SCPCP000024")
-project_ids <- setdiff(project_ids, cell_line_projects) # remove cell line projects
-results_files <- results_files[project_ids]
 ```
@@
 
-```{r, message=FALSE}
+```{r, message=FALSE, warning=FALSE}
 # read in diagnoses
@@
   purrr::imap(summarize_celltypes) |> 
-  dplyr::bind_rows(.id = "project_id") |> 
+  dplyr::bind_rows() |> 
+  # remove cell line projects
+  dplyr::filter(!project_id %in% cell_line_projects) |> 
   # add in diagnoses 
@@
   palette.colors(palette = "alphabet"),
-  "black", # 1 extra since alphabet is 26 and we have 27, this will be plasma cell which shows up once 
+  "lightyellow", # plasma cell 
+  "black", # glial cell
   "grey60", 

### 2025-07-31 analyses/cell-type-consensus/exploratory-notebooks/03-osteosarcoma-consensus-celltypes.Rmd

@@
 ```{r}
-# list all results files 
-results_files <- list.files(results_dir, pattern = "_consensus-cell-types\\.tsv.\\gz$", full.names = TRUE)
-
-# get project ids from file list and assign as names
-project_ids <- stringr::str_remove(basename(results_files), "_consensus-cell-types.tsv.gz")
-names(results_files) <- project_ids
-
 # grab only osteo projects
 osteo_project_ids <- c("SCPCP000017", "SCPCP000018", "SCPCP000023")
-results_files <- results_files[osteo_project_ids]
+osteo_results_dirs <- file.path(results_dir, osteo_project_ids)
+# list all results files 
+results_files <- osteo_results_dirs |> 
+  list.files(pattern = "_processed_consensus-cell-types\\.tsv.\\gz$", recursive = TRUE, full.names = TRUE)
 
@@
 
-```{r, message=FALSE}
+```{r, message=FALSE, warning=FALSE}
 # get immune cell types 
@@
   purrr::imap(summarize_celltypes) |> 
-  dplyr::bind_rows(.id = "project_id") |> 
+  dplyr::bind_rows() |> 
   # join with sample metadata

### 2025-07-31 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd

--- 2025-02-27 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
@@
 # now only keep unique markers for the dotplot
+# except for hematopoietic precursor cells, where there are no unique markers
 unique_markers_df <- filtered_markers_df |> 

### 2025-07-31 analyses/infercnv-consensus-cell-type/exploratory-notebooks/SCPCP000004/01_nb-consensus-cell-types.Rmd

@@
 options(readr.show_col_types = FALSE)
+suppressPackageStartupMessages({
   library(ggplot2)
+  library(SingleCellExperiment)
+})
 theme_set(theme_bw())
@@
 
-# The current results directory, found within the repository base directory
-result_dir <- file.path(repository_base, "data", "current", "results")
-
-# Path to cell-type-consensus results
-consensus_dir <- file.path(result_dir, "cell-type-consensus", "SCPCP000004")
-
 # Path to reference cell type groupings
@@
 ```{r paths}
-# Input consensus cell type files
-celltype_files <- list.files(
-  path = consensus_dir,
-  pattern = "_processed_consensus-cell-types\\.tsv\\.gz$",
+# Input sce files to get consensus cell types
+sce_files <- list.files(
+  path = data_dir,
+  pattern = "_processed\\.rds$",
   recursive = TRUE,
@@
 # Prepare data frame of all cell types
-celltype_df <- celltype_files |>
+celltype_df <- sce_files |>
   # consider only non-PDX
   purrr::keep_at(tumor_library_ids) |>
-  purrr::map(readr::read_tsv) |>
-  purrr::list_rbind() |>
-  dplyr::select(library_id, barcodes, consensus_annotation) |>
+  purrr::map(
+    \(sce_file) {
+      readRDS(sce_file) |>
+        colData() |>
+        as.data.frame() |>
+        dplyr::select(
+          barcodes, 
+          consensus_annotation = consensus_celltype_annotation
+        )
+  }) |>
+  purrr::list_rbind(names_to = "library_id") |>
   dplyr::left_join(cell_group_df, by = "consensus_annotation")

### 2025-07-31 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd

@@
 # repo and analysis base dirs
-repository_base <- rprojroot::find_root(rprojroot::is_git_root)
+module_base <- rprojroot::find_root(rprojroot::is_renv_project)
+repository_base <- file.path(module_base, "..", "..") # we can't use `rprojroot::is_git_root` in Docker, so back up to OpenScPCA base
 analysis_dir <- file.path(repository_base, "analyses", "infercnv-consensus-cell-type")

### 2025-08-13 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
add notebook comparing scanvi and singler

---
title: "Compare scANVI and SingleR annotations"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

We have used `SingleR` and `scANVI/scArches` to perform cell type annotation on `SCPCP000004` samples using the `NBAtlas` reference.
The goal of this notebook is to compare these two sets of inferences to determine if they generally agree.

## Setup

```{r, warning = FALSE}
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(SingleCellExperiment)
})

theme_set(theme_bw())

# Define color ramp for shared use in the heatmaps
heatmap_col_fun <- circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue"))
# Set heatmap padding option
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(0.6, "in"))

# settings
options(
  dplyr.summarise.inform = FALSE, 
  readr.show_col_types = FALSE
)
```

### Paths

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

merged_dir <- file.path(repository_base, "data", "current", "results", "merge-sce", "SCPCP000004")
module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
ref_dir <- file.path(module_dir, "references")
results_dir <- file.path(module_dir, "results")
singler_dir <- file.path(results_dir, "singler")
scanvi_dir <- file.path(results_dir, "scanvi")
```

```{r file paths}
# merged SCE file
sce_file <- file.path(
  merged_dir,
  "SCPCP000004_merged.rds"
)

# SingleR files
singler_files <- list.files(
  path = singler_dir,
  pattern = "_singler-annotations\\.tsv",
  recursive = TRUE,
  full.names = TRUE
) |>
  # add names as library id
  purrr::set_names(
    \(x) {
      stringr::str_remove(basename(x), "_singler-annotations.tsv")
    }
  )

# scanvi predictions
scanvi_file <- file.path(
  scanvi_dir, 
  "scanvi-predictions.tsv"
)

# broad consensus cell type groups
validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"

# palette files
recoded_palette_file <- file.path(
  module_dir,
  "palettes",
  "harmonized-cell-type-palette.tsv"
)
```

### Functions

```{r}
# Source Jaccard and heatmap utilities functions
source(file.path(module_dir, "scripts", "utils", "jaccard-utils.R"))

# Source additional utilities functions:
# - select_nbatlas_markers()
# - harmonize_celltypes()
# - faceted_umap()
# - generate_dotplot()
source(file.path(module_dir, "scripts", "utils", "celltype-utils.R"))
```

### Read and prepare input data

Read merged SCE object:

```{r}
merged_sce <- readRDS(sce_file)
# immediately remove assays we don't need for space
assay(merged_sce, "spliced") <- NULL
assay(merged_sce, "counts") <- NULL
reducedDim(merged_sce, "PCA") <- NULL
```

Read SingleR results:

```{r}
singler_df <- singler_files |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind(names_to = "library_id") |>
  dplyr::mutate(
    # add cell id column for unique rows & joining
    cell_id = glue::glue("{library_id}-{barcodes}"),
    # recode NA -> "unknown_singler"
    singler = ifelse(is.na(pruned.labels),"unknown_singler", pruned.labels)
  ) |>
  dplyr::select(cell_id, singler, singler_delta_next = delta.next)
```

Read and prepare scanvi results:

```{r}
scanvi_full_df <- readr::read_tsv(
  file.path(scanvi_dir, "scanvi_predictions.tsv")
) |>
  dplyr::select(
    cell_id, 
    scanvi = scanvi_prediction, 
    starts_with("pp_")
  ) |>
  tidyr::pivot_longer(
    starts_with("pp_"), 
    names_to = "posterior_celltype", 
    values_to = "posterior"
  ) |>
  dplyr::mutate(posterior_celltype = stringr::str_remove(posterior_celltype, "^pp_"))

# create dedicated column for the posterior associated with the label
scanvi_df <- scanvi_full_df |>
  dplyr::filter(scanvi == posterior_celltype) |>
  dplyr::select(cell_id, scanvi, scanvi_posterior = posterior)
```

Read and prepare additional helper files for viz:

```{r}
# validation groups data frame
validation_df <- readr::read_tsv(validation_url) |>
  dplyr::select(consensus_annotation, validation_group_annotation)

# set up palettes
# recoded to shared colors
recoded_palette_df <- readr::read_tsv(recoded_palette_file)
recoded_celltype_pal <- recoded_palette_df$hex_color
names(recoded_celltype_pal) <- recoded_palette_df$harmonized_label
```

## scANVI posterior probabilities

Before comparing to `SingleR`, we should get a sense of reliability for the `scANVI` results.
`scANVI` returns a cell-specific posterior probability for each label, so we'll begin by looking at these distributions.

First, what is the distribution of posterior probabilities for assigned labels?
We'll look at this on a per-library basis to see if any libraries may have less confident annotations in general.

```{r fig.height=12, fig.width=15}
scanvi_df |>
  tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE) |>
  dplyr::add_count(library_id) |>
  dplyr::mutate(library_id = glue::glue("{library_id} (N={n})")) |>
  ############ into ggplot ################
  ggplot() + 
  aes(x = scanvi_posterior) + 
  # ugly but _very visible_ color
  geom_histogram(bins = 100, fill = "magenta", color = "magenta") + 
  facet_wrap(vars(library_id), scales = "free_y", ncol = 7)
```

The posterior probabilities here are all _exceptionally_ high, consistently across all libraries, which tells us that on the whole, `scANVI` is confident in assignments.

Let's instead look at these probabilities on a per-cell type basis:

```{r fig.height=8, fig.width=12}
ggplot(scanvi_df) + 
  aes(x = scanvi_posterior) + 
  # ugly but _very visible_ color
  geom_histogram(bins = 100, fill = "magenta", color = "magenta") + 
  facet_wrap(vars(scanvi), scales = "free_y", ncol = 5) 
```

While annotation reliability is about the same on a per-library basis, it seems that certain cell types may be much easier to classify than others which may be a factor of how well represented they are in the `NBAtlas` reference.
We'll look at the median value for each as well as the number of cells that were annotated for each label:

```{r}
scanvi_df |>
  dplyr::group_by(scanvi) |>
  dplyr::summarize(
    median_pp = median(scanvi_posterior), 
    total_cells = dplyr::n()
  ) |>
  dplyr::arrange(median_pp)
```

There are very few cells in general associated with the cell types that have very low median posterior probabilities (`NKT cell` and `Migratory cDC`), which is comforting.

A `0.9` threshold for the posterior probability here seems reasonable given how high these values are in general. 
We'll therefore recode cells with posterior probabilities below 0.9 to `unknown_scanvi`.

```{r}
pp_threshold <- 0.9
scanvi_df <- scanvi_df |>
  dplyr::mutate(scanvi = ifelse(
    scanvi_posterior < pp_threshold, 
    "unknown_scanvi",
    scanvi
  ))
```

## Compare labels

We'll create a single data frame with UMAP coordinates and all cell types (consensus, `SingleR`, and `scANVI`):

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("sample_id", "consensus_celltype_annotation"),
  use.dimred = c("UMAP")
) |>
  # there ARE repeated barcodes so we need to keep cell_id around
  tibble::rownames_to_column("cell_id") |>
  dplyr::rename(
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2,
    consensus_annotation = consensus_celltype_annotation
  ) |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # recode consensus NAs to "unknown_consensus"
  dplyr::mutate(
    validation_group_annotation = ifelse(
      is.na(validation_group_annotation), "unknown_consensus", validation_group_annotation
    )
  ) |>
  dplyr::select(-consensus_annotation) |>
  # merge in the annotations
  dplyr::left_join(singler_df, by = "cell_id") |>
  dplyr::left_join(scanvi_df, by = "cell_id") |>
  # make it tidy
  tidyr::pivot_longer(
    c(validation_group_annotation, singler, scanvi), 
    names_to = "celltype_method", 
    values_to = "label"
  ) |>
  harmonize_celltypes(label, label_recoded) 
```

## UMAP

This section displays the UMAP for the _merged object_ (not integrated) with cell type labels harmonized for comparison.
We show a panel for each cell type method.

```{r, fig.width = 12, fig.height = 8}
ggplot(celltype_df) +
  aes(x = UMAP1, y = UMAP2, color = label_recoded) +
  geom_point(alpha = 0.5, size = 0.2) +
  scale_color_manual(values = recoded_celltype_pal) +
  facet_wrap(vars(celltype_method)) +
  coord_equal() +
  theme(
    legend.position = "bottom", 
    axis.text = element_blank(), 
    axis.ticks = element_blank()
  ) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
```

## Heatmap

This section shows a stacked heatmap comparing the `SingleR` and `scANVI` annotations to the consensus validation groups.

```{r, fig.height = 13, fig.width = 12} 
# make a wide version for the heatmap
celltype_wide_df <- celltype_df |>
  dplyr::select(-label) |>
  tidyr::pivot_wider(
    names_from = celltype_method,
    values_from = label_recoded
  )

list(
  "SingleR annotations" = make_jaccard_matrix(celltype_wide_df, "validation_group_annotation", "singler"),
  "scANVI annotations" = make_jaccard_matrix(celltype_wide_df, "validation_group_annotation", "scanvi")
) |>
  purrr::imap(
    \(jaccard_mat, name) create_single_heatmap(
        mat = jaccard_mat,
        row_title = name,
        column_title = "Consensus validation groups",
        keep_legend_name = "SingleR annotations" # arbitrary
    )
  ) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(heatmap_legend_side = "right")
```

We'll also show some statistics from a confusion matrix comparison between `SingleR` and `scANVI`:

```{r}
# ensure factors are compatible for caret
all_levels <- c(
  celltype_wide_df$scanvi, 
  celltype_wide_df$singler
) |> unique()

caret::confusionMatrix(
  factor(celltype_wide_df$scanvi, levels = all_levels),  
  factor(celltype_wide_df$singler, levels = all_levels)
)$overall

```

## Cross-library comparison

Across libraries, what fraction of `SingleR` and `scANVI` labels exactly agree?
Note that this analysis does not consider similar labels (e.g. two types of T cells) but only does a quick look at identical labels.
We'll also highlight the PDX samples here to see if there was any associated bias.

```{r, fig.width = 10, fig.height = 4}
frac_agree_df <- celltype_wide_df |>
  tidyr::separate(cell_id, into = c("library_id", "barcode")) |>
  dplyr::group_by(library_id) |>
  dplyr::summarize(
    frac_agree = sum(scanvi == singler)/dplyr::n(), 
    library_size = dplyr::n()
  )
frac_agree_df <- colData(merged_sce) |> 
  as.data.frame() |> 
  dplyr::select(library_id, is_xenograft) |> 
  unique() |>
  dplyr::right_join(frac_agree_df, by = "library_id")

p1 <- ggplot(frac_agree_df) + 
  aes(x = frac_agree) + 
  geom_histogram() + 
  xlab("Fraction of SingleR and scANVI cells\nthat agree, per library")

p2 <- ggplot(frac_agree_df) + 
  aes(x = library_size, y = frac_agree, color = is_xenograft) + 
  geom_point() +
  labs(
    x = "Library size (n cells)",
    y = "Fraction of SingleR and scANVI\ncells that agree"
  ) +
  theme(legend.position = "bottom")

p1 + p2

```

## Conclusions

Overall `SingleR` and `scANVI` appear to have made highly similar predictions for the vast majority of libraries.
These cell types are also fairly similar to our existing consensus cell types.
Therefore, we can use agreement between `SingleR` and `scANVI` to derive final annotations for this analysis module.

```{r}
sessionInfo()
```

### 2025-08-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
update posterior section and land at 0.75

@@
 )
+nbatlas_palette_file <- file.path(
+  module_dir,
+  "palettes",
+  "nbatlas-plus-cell-type-palette.tsv"
+)
 ```
@@
 # set up palettes
-# recoded to shared colors
 recoded_palette_df <- readr::read_tsv(recoded_palette_file)
@@
 names(recoded_celltype_pal) <- recoded_palette_df$harmonized_label
+
+nbatlas_palette_df <- readr::read_tsv(nbatlas_palette_file)
+nbatlas_celltype_pal <- nbatlas_palette_df$hex_color
+names(nbatlas_celltype_pal) <- nbatlas_palette_df$NBAtlas_label
 ```
@@
 First, what is the distribution of posterior probabilities for assigned labels?
-We'll look at this on a per-library basis to see if any libraries may have less confident annotations in general.
+We'll look at the minimum, maximum, and median posterior probabilities across libraries in a table.
 
-```{r fig.height=12, fig.width=15}
+```{r}
 scanvi_df |>
   tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE) |>
-  dplyr::add_count(library_id) |>
-  dplyr::mutate(library_id = glue::glue("{library_id} (N={n})")) |>
-  ############ into ggplot ################
-  ggplot() + 
-  aes(x = scanvi_posterior) + 
-  # ugly but _very visible_ color
-  geom_histogram(bins = 100, fill = "magenta", color = "magenta") + 
-  facet_wrap(vars(library_id), scales = "free_y", ncol = 7)
+  dplyr::group_by(library_id) |>
+  dplyr::summarize(
+    library_size = dplyr::n(), 
+    min_posterior = min(scanvi_posterior), 
+    median_posterior = median(scanvi_posterior), 
+    max_posterior = max(scanvi_posterior)
+  ) |>
+  dplyr::arrange(library_size)
 ```
 
-The posterior probabilities here are all _exceptionally_ high, consistently across all libraries, which tells us that on the whole, `scANVI` is confident in assignments.
+Across libraries, the median posterior probability is always exceptionally high at >=0.99, which suggests `scANVI` confidence is roughly the same across libraries.
 
-Let's instead look at these probabilities on a per-cell type basis:
+Let's instead look at these probabilities on a per-cell type basis, this time with a ridge plot including a line for the median value for each distribution:
 
-```{r fig.height=8, fig.width=12}
-ggplot(scanvi_df) + 
-  aes(x = scanvi_posterior) + 
-  # ugly but _very visible_ color
-  geom_histogram(bins = 100, fill = "magenta", color = "magenta") + 
-  facet_wrap(vars(scanvi), scales = "free_y", ncol = 5) 
+```{r fig.height=8, fig.width=8}
+scanvi_df |>
+  dplyr::group_by(scanvi) |>
+  dplyr::mutate(scanvi = glue::glue("{scanvi} (n={dplyr::n()})")) |>
+  ########### into ggplot ############
+  ggplot() + 
+    aes(y = forcats::fct_infreq(scanvi), x = scanvi_posterior) + 
+    # show the median
+    ggridges::stat_density_ridges(quantile_lines = TRUE, quantiles = 2, alpha = 0.7) +
+    labs(x = "posterior probability", y = "scANVI label (number of cells)") 
 ```
@@
 While annotation reliability is about the same on a per-library basis, it seems that certain cell types may be much easier to classify than others which may be a factor of how well represented they are in the `NBAtlas` reference.
-We'll look at the median value for each as well as the number of cells that were annotated for each label:
+There are very few cells in general associated with the cell types that have very low median posterior probabilities (`NKT cell`, `Migratory cDC`, `Circulating NK cell`).
+
+A `0.75` threshold seems like it may be a good middle-ground for preserving labels while removing low-confidence annotations. 
+What effect will this have?
+
+```{r}
+sum(scanvi_df$scanvi_posterior < 0.75) / nrow(scanvi_df)
+```
+
+This will throw out a bit under 2% of cells:
 
@@
 scanvi_df |>
-  dplyr::group_by(scanvi) |>
-  dplyr::summarize(
-    median_pp = median(scanvi_posterior), 
-    total_cells = dplyr::n()
-  ) |>
-  dplyr::arrange(median_pp)
+  dplyr::filter(scanvi_posterior < 0.75) |>
+  dplyr::count(scanvi) |>
+  dplyr::arrange(desc(n))
 ```
 
-There are very few cells in general associated with the cell types that have very low median posterior probabilities (`NKT cell` and `Migratory cDC`), which is comforting.
-
-A `0.9` threshold for the posterior probability here seems reasonable given how high these values are in general. 
-We'll therefore recode cells with posterior probabilities below 0.9 to `unknown_scanvi`.
+We'll therefore recode cells with posterior probabilities below 0.75 to `unknown_scanvi`.
 
 ```{r}
-pp_threshold <- 0.9
+pp_threshold <- 0.75
 scanvi_df <- scanvi_df |>
@@
 
-We'll create a single data frame with UMAP coordinates and all cell types (consensus, `SingleR`, and `scANVI`):
+We'll create a single data frame with UMAP coordinates and all cell types (consensus, `SingleR`, and `scANVI`).
 

### 2025-08-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
restore histogram and show ridge plot with vline

@@
 First, what is the distribution of posterior probabilities for assigned labels?
-We'll look at the minimum, maximum, and median posterior probabilities across libraries in a table.
+We'll make a histogram below using a very visible color to ensure the plot is legible, where each panel is a library, and libraries are ordered by size.
 
-```{r}
+```{r fig.height=12, fig.width=15}
 scanvi_df |>
   tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE) |>
-  dplyr::group_by(library_id) |>
-  dplyr::summarize(
-    library_size = dplyr::n(), 
-    min_posterior = min(scanvi_posterior), 
-    median_posterior = median(scanvi_posterior), 
-    max_posterior = max(scanvi_posterior)
+  dplyr::add_count(library_id) |>
+  dplyr::mutate(
+    library_id = glue::glue("{library_id} (N={n})"),
+    library_id = factor(library_id), 
+    library_id = forcats::fct_infreq(library_id), 
+    library_id = forcats::fct_rev(library_id)
   ) |>
-  dplyr::arrange(library_size)
+  ############ into ggplot ################
+  ggplot() + 
+  aes(x = scanvi_posterior) + 
+  # ugly but _very visible_ color
+  geom_histogram(bins = 100, fill = "magenta", color = "magenta") + 
+  facet_wrap(vars(library_id), scales = "free_y", ncol = 7)
 ```
 
-Across libraries, the median posterior probability is always exceptionally high at >=0.99, which suggests `scANVI` confidence is roughly the same across libraries.
+Across libraries, the median posterior probability is always exceptionally high at which suggests `scANVI` confidence is roughly the same across libraries.
 
@@
 ```{r fig.height=8, fig.width=8}
-scanvi_df |>
+scanvi_ridgeplot <- scanvi_df |>
   dplyr::group_by(scanvi) |>
@@
     labs(x = "posterior probability", y = "scANVI label (number of cells)") 
+scanvi_ridgeplot
 ```
@@
 
-A `0.75` threshold seems like it may be a good middle-ground for preserving labels while removing low-confidence annotations. 
-What effect will this have?
+A `0.75` threshold seems like it may be a good middle-ground for preserving labels while removing low-confidence annotations:
+
+```{r fig.height=8, fig.width=8}
+scanvi_ridgeplot +
+  geom_vline(xintercept = 0.75, color = "firebrick")
+```
+
+How many cells will this remove?
 
@@
 
-This will throw out a bit under 2% of cells:
+What are those cell types?
 

### 2025-08-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
update umaps, heatmaps, and confusion matrix to ensure we have direct comparisons also

@@
 theme_set(theme_bw())
+set.seed(2025) # used for some viz
 
@@
 )
+
+# label mapping
+nbatlas_label_map_file <- file.path(ref_dir, "nbatlas-label-map.tsv")
 ```
@@
     # recode NA -> "unknown_singler"
-    singler = ifelse(is.na(pruned.labels),"unknown_singler", pruned.labels)
+    singler = ifelse(is.na(pruned.labels),"Unknown", pruned.labels)
   ) |>
@@
 
+# label mapping
+nbatlas_label_map <- readr::read_tsv(nbatlas_label_map_file)
+
 # set up palettes
@@
 names(recoded_celltype_pal) <- recoded_palette_df$harmonized_label
+# for this notebook, we want just a single Unknown
+recoded_celltype_pal["Unknown"] <- recoded_celltype_pal["unknown_singler"] 
+recoded_celltype_pal["unknown_singler"] <- NULL
+recoded_celltype_pal["unknown_consensus"] <- NULL
 
@@
 
-We'll therefore recode cells with posterior probabilities below 0.75 to `unknown_scanvi`.
+We'll therefore recode cells with posterior probabilities below 0.75 to `Unknown`.
 
@@
     scanvi_posterior < pp_threshold, 
-    "unknown_scanvi",
+    "Unknown",
     scanvi
@@
 We'll create a single data frame with UMAP coordinates and all cell types (consensus, `SingleR`, and `scANVI`). 
+We'll do this at two levels of organization:
+
+- One with harmonized labels to be able to compare to consensus
+- One with original NBAtlas labels to just compare `SingleR` to `scANVI`
 
@@
   merged_sce,
-  use.coldata = c("sample_id", "consensus_celltype_annotation"),
+  use.coldata = c("sample_id", "is_xenograft", "consensus_celltype_annotation"),
   use.dimred = c("UMAP")
@@
   dplyr::left_join(validation_df, by = "consensus_annotation") |>
-  # recode consensus NAs to "unknown_consensus"
+  # recode consensus NAs to "Unknown"
   dplyr::mutate(
     validation_group_annotation = ifelse(
-      is.na(validation_group_annotation), "unknown_consensus", validation_group_annotation
+      is.na(validation_group_annotation), "Unknown", validation_group_annotation
     )
@@
   ) |>
-  harmonize_celltypes(label, label_recoded) 
+  harmonize_celltypes(label, label_recoded) |>
+  dplyr::mutate(label = ifelse(label == "NE", "Neuroendocrine", label))
 ```
@@
 
-This section displays the UMAP for the _merged object_ (not integrated) with cell type labels harmonized for comparison.
+This section displays the UMAP for the _merged object_ (not integrated) with cell type labels across methods.
 We show a panel for each cell type method.
+
+### SingleR vs scANVI
+
+This set of UMAPs shows annotation results from `SingleR` and `scANVI`.
+
+Since there are many cell types in this plot, it has been colored so that cell types in the same "family" each share a color:
+
+* T cells: `T cell`, `Treg`, `CD4+ T cell`, `CD8+ T cell`
+* Natural killer cells: `NK cell`, `Circulating NK cell`, `Resident NK cell`, `TOX2+/KIT+ NK cell`
+* Conventional dendritic cells: `cDC`, `cDC1`, `cDC2/DC3`, `Migratory cDC`
+* Monocytes: `Monocyte`, `Classical monocyte`, `Patrolling monocyte`
+
+```{r, fig.width = 12, fig.height = 8}
+nbatlas_t_celltypes <- nbatlas_label_map |>
+  dplyr::filter(NBAtlas_label == "T cell") |>
+  dplyr::pull(NBAtlas_label_immune_zoom)
+
+nbatlas_nk_celltypes <- nbatlas_label_map |>
+  dplyr::filter(NBAtlas_label == "NK cell") |>
+  dplyr::pull(NBAtlas_label_immune_zoom)
+
+nbatlas_myeloid_celltypes <- nbatlas_label_map |>
+  dplyr::filter(NBAtlas_label == "Myeloid") |>
+  dplyr::pull(NBAtlas_label_immune_zoom)
+
+# Note that cDC and Monocyte are not direct labels in the disseminated NBAtlas object,
+# but these groupings are used in the paper's marker gene analysis.
+nbatlas_cdc_celltypes <- nbatlas_label_map |>
+  dplyr::filter(NBAtlas_label == "cDC") |>
+  dplyr::pull(NBAtlas_label_immune_zoom)
+
+nbatlas_monocyte_celltypes <- nbatlas_label_map |>
+  dplyr::filter(NBAtlas_label == "Monocyte") |>
+  dplyr::pull(NBAtlas_label_immune_zoom)
+
+umap_direct_df <- celltype_df |>
+  dplyr::filter(celltype_method != "validation_group_annotation") |>
+  # shuffle points to ensure order they are plotted is random
+  dplyr::sample_frac(1) |> 
+  # order celltypes so that families are grouped together for a cleaner legend
+  # note this will give warnings if any of those cells aren't in the annotation, but that's fine
+  dplyr::mutate(
+    label = factor(label), 
+    label = forcats::fct_relevel(
+      label, 
+      "T cell", nbatlas_t_celltypes, 
+      "NK cell", nbatlas_nk_celltypes,
+      "cDC", nbatlas_cdc_celltypes, 
+      "Monocyte", nbatlas_monocyte_celltypes
+    )
+  )
+
+ggplot(umap_direct_df) +
+  aes(x = UMAP1, y = UMAP2, color = label) +
+  geom_point(alpha = 0.5, size = 0.2) +
+  scale_color_manual(values = nbatlas_celltype_pal) +
+  facet_wrap(vars(celltype_method)) +
+  coord_equal() + 
+  theme(
+    legend.position = "bottom", 
+    axis.text = element_blank(), 
+    axis.ticks = element_blank()
+  ) +
+  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
+```
+
+Cell type annotations look broadly consistent between these methods, although `SingleR` appears to assign more heterogeneity within the "clumps" of cells (which likely correspond to individual libraries in this merged object).
+
+### Harmonized labels for consensus comparison
+
+This set of UMAPs harmonizes cell type labels into broader groupings to enable a comparison with consensus annotations.
+Note that the `Unknown` cell type refers to cells uncategorized by `SingleR` or `scANVI`.
 
@@
 ```
+We see broad compatibility among all three approaches, and the Unknown consensus indeed appear to be mostly tumor cells (here, Neuroendocrine).
+This is reassuring that we have some robust cell type identifications here!
 
@@
 
-This section shows a stacked heatmap comparing the `SingleR` and `scANVI` annotations to the consensus validation groups.
+This section shows heatmaps comparing annotations to one another.
 
-```{r, fig.height = 13, fig.width = 12} 
+### SingleR vs scANVI
+
+This heatmap directly compares `SingleR` and `scANVI` annotations.
+
+```{r, fig.height = 11, fig.width = 11} 
 # make a wide version for the heatmap
 celltype_wide_df <- celltype_df |>
-  dplyr::select(-label) |>
+  dplyr::filter(celltype_method != "validation_group_annotation") |>
+  dplyr::select(-label_recoded) |>
   tidyr::pivot_wider(
     names_from = celltype_method,
-    values_from = label_recoded
+    values_from = label
   )
-
-list(
-  "SingleR annotations" = make_jaccard_matrix(celltype_wide_df, "validation_group_annotation", "singler"),
-  "scANVI annotations" = make_jaccard_matrix(celltype_wide_df, "validation_group_annotation", "scanvi")
-) |>
-  purrr::imap(
-    \(jaccard_mat, name) create_single_heatmap(
-        mat = jaccard_mat,
-        row_title = name,
-        column_title = "Consensus validation groups",
-        keep_legend_name = "SingleR annotations" # arbitrary
+make_jaccard_heatmap(
+  celltype_wide_df,
+  "scanvi",
+  "singler",
+  "scANVI label",
+  "SingleR label"
 )
-  ) |>
-  # concatenate vertically into HeatmapList object
-  purrr::reduce(ComplexHeatmap::`%v%`) |>
-  ComplexHeatmap::draw(heatmap_legend_side = "right")
 ```
 
-We'll also show some statistics from a confusion matrix comparison between `SingleR` and `scANVI`:
+This heatmap shows excellent correspondence between `SingleR` and `scANVI` annotations with most of the weight along the diagonal.
+
+We'll nail this down a bit further with some confusion matrix metrics:
 
@@
 ```
+
+### Harmonized labels for consensus comparison
+
+This stacked heatmap compares all three methods to one another, using broader cell type groupings to enable a more direct comparison with consensus annotations.
+
+```{r, fig.height = 13, fig.width = 12} 
+# make a wide version for the heatmap
+celltype_recoded_wide_df <- celltype_df |>
+  dplyr::select(-label) |>
+  tidyr::pivot_wider(
+    names_from = celltype_method,
+    values_from = label_recoded
+  )
+
+list(
+  "SingleR annotations" = make_jaccard_matrix(celltype_recoded_wide_df, "validation_group_annotation", "singler"),
+  "scANVI annotations" = make_jaccard_matrix(celltype_recoded_wide_df, "validation_group_annotation", "scanvi")
+) |>
+  purrr::imap(
+    \(jaccard_mat, name) create_single_heatmap(
+        mat = jaccard_mat,
+        row_title = name,
+        column_title = "Consensus validation groups",
+        keep_legend_name = "SingleR annotations" # arbitrary for labeling
+    )
+  ) |>
+  # concatenate vertically into HeatmapList object
+  purrr::reduce(ComplexHeatmap::`%v%`) |>
+  ComplexHeatmap::draw(heatmap_legend_side = "right")
+```
+
+Again, we see excellent correspondence among all three methods.
 

### 2025-08-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
update section where we look at overall agreement

@@
   "palettes",
-  "nbatlas-plus-cell-type-palette.tsv"
+  "nbatlas-cell-type-palette.tsv"
 )
@@
 scanvi_df |>
-  tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE) |>
+  tidyr::separate(cell_id, into = c("library_id", "barcode")) |>
   dplyr::add_count(library_id) |>
@@
   harmonize_celltypes(label, label_recoded) |>
-  dplyr::mutate(label = ifelse(label == "NE", "Neuroendocrine", label))
+  dplyr::mutate(label = ifelse(label == "NE", "Neuroendocrine", label)) |>
+  tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE)
 ```
@@
   dplyr::filter(NBAtlas_label == "Monocyte") |>
+  dplyr::pull(NBAtlas_label_immune_zoom)
+
+nbatlas_myeloid_celltypes <- nbatlas_label_map |>
+  dplyr::filter(NBAtlas_label == "Myeloid") |>
   dplyr::pull(NBAtlas_label_immune_zoom)
@@
   )
+
 make_jaccard_heatmap(
@@
 
-```{r, fig.width = 10, fig.height = 4}
-frac_agree_df <- celltype_wide_df |>
-  tidyr::separate(cell_id, into = c("library_id", "barcode")) |>
+> I think it would be helpful to get a table in here that shows you % of each library that has exact agreement between SingleR and scANVI, % of each library where cells are in the same family (e.g., the same cell is labeled as two different T cell types), and % of each library that disagrees and is not in the same family.
+
+In this case, we consider these families specifically:
+
+* T cells: `T cell`, `Treg`, `CD4+ T cell`, `CD8+ T cell`
+* Natural killer cells: `NK cell`, `Circulating NK cell`, `Resident NK cell`, `TOX2+/KIT+ NK cell`
+* Conventional dendritic cells (cDC): `cDC`, `cDC1`, `cDC2/DC3`, `Migratory cDC`
+* Monocytes: `Monocyte`, `Classical monocyte`, `Patrolling monocyte`
+* Myeloid: `Neutrophil` and `Macrophage`
+
+```{r}
+# celltype_wide_df uses the original labels
+identical_agree_df <- celltype_wide_df |>
   dplyr::group_by(library_id) |>
   dplyr::summarize(
-    frac_agree = sum(scanvi == singler)/dplyr::n(), 
+    frac_identical = sum(scanvi == singler)/dplyr::n(), 
     library_size = dplyr::n()
   )
-frac_agree_df <- colData(merged_sce) |> 
-  as.data.frame() |> 
+
+# create a wide data frame with "family" labels where possible
+celltype_family_df <- celltype_df |>
+  dplyr::filter(celltype_method != "validation_group_annotation") |>
+  # create label_family grouping
+  dplyr::mutate(
+    label_family = dplyr::case_when(
+      label %in% nbatlas_t_celltypes ~ "T cell",
+      label %in% nbatlas_nk_celltypes ~ "NK cell",
+      label %in% nbatlas_cdc_celltypes ~ "cDC",
+      label %in% nbatlas_monocyte_celltypes ~ "Monocyte",
+      label %in% nbatlas_myeloid_celltypes ~ "Myeloid", 
+      .default = label
+    )
+  ) |>
+  dplyr::select(-label_recoded, -label) |>
+  tidyr::pivot_wider(
+    names_from = celltype_method,
+    values_from = label_family
+  )
+
+family_agree_df <- celltype_family_df |>
+  dplyr::group_by(library_id) |>
+  dplyr::summarize(frac_same_family = sum(scanvi == singler)/dplyr::n())
+
+# if families disagree, then there is no agreement at all
+disagree_df <- celltype_family_df |>
+  dplyr::group_by(library_id) |>
+  dplyr::summarize(frac_disagree = sum(scanvi != singler)/dplyr::n())
+```
+
+```{r}
+# combine into a single table for display & viz
+frac_df <- identical_agree_df |>
+  dplyr::left_join(family_agree_df) |>
+  dplyr::left_join(disagree_df) |>
+  # round to 4 decimals
+  dplyr::mutate(dplyr::across(contains("frac_"), round, 4)) |>
+  dplyr::select(
+    library_id, 
+    library_size, 
+    frac_identical, 
+    frac_same_family, 
+    frac_disagree
+  ) |>
+  dplyr::arrange(library_size)
+
+# use DT for improved exploration ability
+DT::datatable(frac_df, rownames = FALSE)
+```
+
+```{r}
+# add PDX column for plotting
+frac_df <- celltype_df |> 
   dplyr::select(library_id, is_xenograft) |>
   unique() |>
-  dplyr::right_join(frac_agree_df, by = "library_id")
+  dplyr::right_join(frac_df)
 
-p1 <- ggplot(frac_agree_df) + 
-  aes(x = frac_agree) + 
-  geom_histogram() + 
-  xlab("Fraction of SingleR and scANVI cells\nthat agree, per library")
+ggplot(frac_df) + 
+  aes(x = library_size, y = frac_same_family) + 
+  geom_point(aes(color = is_xenograft)) + 
+  geom_smooth(method = "lm") +
+  labs(
+    x = "Library size",
+    y = "Fraction of SingleR/scANVI labels from same family"
+  ) +
+  # stop going above 1
+  scale_y_continuous(limits = c(0, 1))
+```
 
-p2 <- ggplot(frac_agree_df) + 
-  aes(x = library_size, y = frac_agree, color = is_xenograft) + 
-  geom_point() +
-  labs(
-    x = "Library size (n cells)",
-    y = "Fraction of SingleR and scANVI\ncells that agree"
-  ) +
-  theme(legend.position = "bottom")
+Most cells have agreeing annotations in the majority of libraries.
+There appears to be a slight trend (not signficant) that larger libraries have more agreement between methods, but this also seems driven by the libraries with low correspondence.
 
-p1 + p2
+### Libraries with high levels of disagreement
 
-```
+There are four libraries (2 PDX and 2 tissue) which show less than 50% agreement, so we should look at these a bit more closely.
+Notably, these do _not_ correspond to the four smallest libraries.
 

### 2025-08-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
add section comparing bad libraries

@@
 
+data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")
 merged_dir <- file.path(repository_base, "data", "current", "results", "merge-sce", "SCPCP000004")
@@
 nbatlas_label_map_file <- file.path(ref_dir, "nbatlas-label-map.tsv")
+
+# sample metadata
+sample_metadata_file <- file.path(data_dir, "single_cell_metadata.tsv")
 ```
@@
 names(nbatlas_celltype_pal) <- nbatlas_palette_df$NBAtlas_label
+
+# sample metadata
+sample_metadata <- readr::read_tsv(sample_metadata_file)
 ```
@@
 Across libraries, what fraction of `SingleR` and `scANVI` labels exactly agree?
-Note that this analysis does not consider similar labels (e.g. two types of T cells) but only does a quick look at identical labels.
-We'll also highlight the PDX samples here to see if there was any associated bias.
+For this, we'll compare among:
 
-> I think it would be helpful to get a table in here that shows you % of each library that has exact agreement between SingleR and scANVI, % of each library where cells are in the same family (e.g., the same cell is labeled as two different T cell types), and % of each library that disagrees and is not in the same family.
-
-In this case, we consider these families specifically:
-
+* Fraction of identical labels
+* Fraction of labels that agree at least on the level of their cell type "family," where we define these families based on groups of NBAtlas annotations:
   * T cells: `T cell`, `Treg`, `CD4+ T cell`, `CD8+ T cell`
@@
   * Myeloid: `Neutrophil` and `Macrophage`
+* Fraction of labels that entirely disagree
+
+The table below shows these fractions for each library along with the library size.
+We also visualize the relationship between library size and family-based agreement.
 
@@
 
-```{r}
+```{r, message = FALSE}
 # combine into a single table for display & viz
@@
   # round to 4 decimals
-  dplyr::mutate(dplyr::across(contains("frac_"), round, 4)) |>
+  dplyr::mutate(dplyr::across(contains("frac_"), \(x) round(x, 4))) |>
   dplyr::select(
@@
 
-```{r}
+```{r, message = FALSE, fig.width = 5, fig.height = 3}
 # add PDX column for plotting
@@
     x = "Library size",
-    y = "Fraction of SingleR/scANVI labels from same family"
+    y = "Fraction of SingleR/scANVI labels\nfrom the same family"
   ) +
@@
 Most cells have agreeing annotations in the majority of libraries.
-There appears to be a slight trend (not signficant) that larger libraries have more agreement between methods, but this also seems driven by the libraries with low correspondence.
+Considering closely-related cells in the same family usually provides a couple more cells to be annotated, but not a large percentage.
+There appears to be a slight trend (not significant) that larger libraries have more agreement between methods, but this also seems driven by the libraries with low correspondence.
 
@@
 Notably, these do _not_ correspond to the four smallest libraries.
+
+Here is the sample metadata for these four samples:
+
+```{r}
+disagree_ids <- frac_df |>
+  dplyr::filter(frac_disagree >= 0.5) |>
+  dplyr::pull(library_id)
+
+disagree_metadata <- sample_metadata |>
+  dplyr::filter(scpca_library_id %in% disagree_ids)
+```
+
+Looking closely here, we can see that these samples were in fact derived from two participants where one sample is a xenograft and the other is tissue.
+
+```{r}
+disagree_metadata |>
+  dplyr::select(scpca_library_id, sample_type, participant_id) |>
+  dplyr::arrange(participant_id)
+```
+
+These participant IDs are not associated with any other samples:
+
+```{r}
+sample_metadata |>
+  dplyr::filter(participant_id %in% disagree_metadata$participant_id) |>
+  dplyr::filter(!(scpca_library_id) %in% disagree_ids)
+```
+
+It's possible that patient-specific biology explains the discrepancies we see for these samples' cell types.
+
+We'll look more closely at their disagreements as well with heatmaps:
+
+```{r, fig.height = 8, fig.width = 8} 
+disagree_ids |>
+  # walk to only print the heatmaps once
+  purrr::walk(
+    \(id) {
+      library_df <- celltype_wide_df |>
+        dplyr::filter(library_id == id)
+
+     frac_disagree <- frac_df |>
+       dplyr::filter(library_id == id) |>
+       dplyr::pull(frac_disagree)
+
+     heatmap_title <- glue::glue("{id} (fraction disagree = {frac_disagree})")
+
+      make_jaccard_heatmap(
+        library_df,
+        "scanvi",
+        "singler",
+        glue::glue("{heatmap_title}\n\nscANVI label"),
+        "SingleR label"
+      )
+  })
+```
+
+For these libraries, the main source of disagreement is `scANVI` applying more homogeneous labeling of `Neuroendocrine` compared to `SingleR` which is assigning more varied cell types.
+Note that the two libraries where `scANVI` is _only_ detecting `Neuroendocrine` cells are the PDX libraries, for which we expect lower cell type diversity in general.
+It's not unlikely that a lot of what `SingleR` has labeled `Stromal other` or `Schwann` may be tumor.
 

### 2025-08-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
add final touches and render

@@
 recoded_celltype_pal["Unknown"] <- recoded_celltype_pal["unknown_singler"]
-recoded_celltype_pal["unknown_singler"] <- NULL
-recoded_celltype_pal["unknown_consensus"] <- NULL
 
@@
 
-How many cells will this remove?
+What fraction of cells will this remove?
 
@@
 ```
+Notably, none are Neuroendocrine (the tumor type).
 
-We'll therefore recode cells with posterior probabilities below 0.75 to `Unknown`.
+We'll apply the 0.75 threshold below by recoding these cells to `Unknown`.
 
@@
   tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE)
+
+# can remove now for space
+rm(merged_sce)
 ```
@@
 ```{r, fig.width = 12, fig.height = 8}
+# define families using the provided map
 nbatlas_t_celltypes <- nbatlas_label_map |>
@@
   dplyr::filter(NBAtlas_label == "NK cell") |>
-  dplyr::pull(NBAtlas_label_immune_zoom)
-
-nbatlas_myeloid_celltypes <- nbatlas_label_map |>
-  dplyr::filter(NBAtlas_label == "Myeloid") |>
   dplyr::pull(NBAtlas_label_immune_zoom)
@@
 
+# we'll use these later but not for this plot
 nbatlas_myeloid_celltypes <- nbatlas_label_map |>
@@
 
-```{r, fig.height = 11, fig.width = 11} 
+```{r, fig.height = 9, fig.width = 9} 
 # make a wide version for the heatmap
@@
 
-```{r}
+```{r, message=FALSE}
 # ensure factors are compatible for caret

### 2025-08-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
Update analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>

@@
 
-Across libraries, the median posterior probability is always exceptionally high at which suggests `scANVI` confidence is roughly the same across libraries.
+Across libraries, the median posterior probability is always exceptionally high which suggests `scANVI` confidence is roughly the same across libraries.
 

### 2025-08-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd

@@
 
-A `0.75` threshold seems like it may be a good middle-ground for preserving labels while removing low-confidence annotations:
-
-```{r fig.height=8, fig.width=8}
-scanvi_ridgeplot +
-  geom_vline(xintercept = 0.75, color = "firebrick")
-```
-
+A `0.75` threshold seems like it may be a good middle-ground for preserving labels while removing low-confidence annotations.
 What fraction of cells will this remove?

### 2025-08-14 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
add annotation notebook

---
title: "Derive final annotations based on NBAtlas reference"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  scanvi_pp_threshold: 0.75    
  annotations_tsv: "../results/SCPCP000004-annotations.tsv"
---

In this analysis module, we have annotated cell types in `SCPCP000004` samples using two methods: `SingleR` and `scANVI/scArches`, both using the `NBAtlas` dataset as a reference.
In this notebook, we derive the final annotations using information from the `SingleR` labels, the `scANVI/scArches` labels, and existing consensus cell type annotations.
All final annotations use labels present in the `NBAtlas` dataset.

We use the following approach to determine the final annotation.

First, we prioritize inferences with `SingleR` and `scANVI/scArches`.
If these methods returned the exact same label, we assign this label.

Next, we compare labels across all three methods: if _two out of three_ methods agree, we assign the agreeing label.
Note that this requires mapping consensus cell types to NBAtlas labels, which is only possible for these cell types: endothelial cells, fibroblasts, and immune cell types except for the `NBAtlas` labels `RBCs` and `pDC` which do not have corresponding consensus labels.

In this approach, there are two ways that cell type annotations can agree, and we treat them as follows:

* Cell type annotations can have directly corresponding labels. For example:
  * If the consensus cell type is `endothelial cell` and at least one of the `NBAtlas`-derived annotations is `Endothelial`, we assign the label `Endothelial`.
* Cell type annotations can differ but be in the same "family" of cell types. For example:
  * If the consensus cell type is `mature T cell` and at least one of the `NBAtlas`-derived annotations is `CD4+ T cell`, we assign the broader label `T cell`.
  * If one `NBAtlas`-derived annotation is `Patrolling monocyte` and the other is `Classical monocyte`, we assign the broader label `Monocyte`.

Finally, to the extent possible, we assign CL ontology IDs to the final annotations before export.

## Setup

```{r, warning = FALSE}
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(SingleCellExperiment)
})

theme_set(
  theme_bw() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(), 
    legend.position = "bottom", 
    aspect.ratio = 1
  )
)

# settings
options(
  dplyr.summarise.inform = FALSE, 
  readr.show_col_types = FALSE
)
```

### Functions

The function `get_unified_label()` defined in the chunk below determines a final label based on correspondence between at least two out of three labels, out of `SingleR`, `scANVI/scArches`, and the consensus label.

```{r}
#' Get unified cell type label
#'
#' @param annotation_df Data frame to annotate with columns `consensus`, `singler`, `scanvi`, and `cell_type_assignment`
#' @param consensus_options Vector of consensus cell types that match the final label, used for `consensus`
#' @param nbatlas_options Vector of NBAtlas cell types that match the final label, used for `singler` and `scanvi`
#' @param final_label Final label to apply if at least two methods have labels present in their respective vectors
#'
#' @returns Updated annotation_df with `cell_type_assignment` filled in with the `final_label` value where appropriate
get_unified_label <- function(
    annotation_df, 
    consensus_options, 
    nbatlas_options, 
    final_label) {

  annotation_df |>
    dplyr::mutate(
      # count how many annotations are in agreement
      n = consensus %in% consensus_options +
        singler %in% nbatlas_options + 
        scanvi %in% nbatlas_options, 
      # if >=2 agree, use this annotation
      cell_type_assignment = ifelse(
        is.na(cell_type_assignment) & n >= 2, 
        final_label, 
        cell_type_assignment
    )) |>
    dplyr::select(-n)
}
```

### Paths

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")
merged_dir <- file.path(repository_base, "data", "current", "results", "merge-sce", "SCPCP000004")
module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
ref_dir <- file.path(module_dir, "references")
results_dir <- file.path(module_dir, "results")
singler_dir <- file.path(results_dir, "singler")
scanvi_dir <- file.path(results_dir, "scanvi")
```

```{r file paths}
# merged SCE file
sce_file <- file.path(
  merged_dir,
  "SCPCP000004_merged.rds"
)

# SingleR files
singler_files <- list.files(
  path = singler_dir,
  pattern = "_singler-annotations\\.tsv",
  recursive = TRUE,
  full.names = TRUE
) |>
  # add names as library id
  purrr::set_names(
    \(x) {
      stringr::str_remove(basename(x), "_singler-annotations.tsv")
    }
  )

# scanvi predictions
scanvi_file <- file.path(
  scanvi_dir, 
  "scanvi-predictions.tsv"
)

# palette file
palette_file <- file.path(
  module_dir,
  "palettes",
  "submission-palette.tsv"
)

# label map file
nbatlas_label_map_file <- file.path(
  ref_dir, 
  "nbatlas-label-map.tsv"
)

# label ontologies file
nbatlas_ontology_file <- file.path(
  ref_dir, 
  "nbatlas-ontology-ids.tsv"
)

validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
```

### Read and prepare input data

Read merged SCE object:

```{r}
merged_sce <- readRDS(sce_file)
# immediately remove assays we don't need for space
assay(merged_sce, "spliced") <- NULL
assay(merged_sce, "counts") <- NULL
reducedDim(merged_sce, "PCA") <- NULL
```

Read SingleR results:

```{r}
singler_df <- singler_files |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind(names_to = "library_id") |>
  dplyr::mutate(
    # add cell id column for unique rows & joining
    cell_id = glue::glue("{library_id}-{barcodes}"),
    # recode NA -> "Unknown"
    singler = ifelse(is.na(pruned.labels), "Unknown", pruned.labels)
  ) |>
  dplyr::select(cell_id, singler)
```

Read scANVI results:

```{r}
scanvi_df <- readr::read_tsv(
  file.path(scanvi_dir, "scanvi_predictions.tsv")
) |>
  dplyr::select(
    cell_id, 
    scanvi = scanvi_prediction, 
    starts_with("pp_")
  ) |>
  tidyr::pivot_longer(
    starts_with("pp_"), 
    names_to = "posterior_celltype", 
    values_to = "posterior"
  ) |>
  dplyr::mutate(posterior_celltype = stringr::str_remove(posterior_celltype, "^pp_")) |>
  dplyr::filter(scanvi == posterior_celltype) |>
  # recode to Unknown if below the threshold
  dplyr::mutate(scanvi = ifelse(posterior < params$scanvi_pp_threshold, "Unknown", scanvi)) |>
  dplyr::select(cell_id, scanvi)
```

Read additional helper files:

```{r}
validation_df <- readr::read_tsv(validation_url)

palette_df <- readr::read_tsv(palette_file)
celltype_pal <- palette_df$hex_color
names(celltype_pal) <- palette_df$NBAtlas_label

nbatlas_label_map <- readr::read_tsv(nbatlas_label_map_file)
nbatlas_ontology_ids <- readr::read_tsv(nbatlas_ontology_file)
```

Combine annotations into a single data frame:

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("sample_id", "is_xenograft", "consensus_celltype_annotation")
) |>
  tibble::rownames_to_column("cell_id") |>
  tidyr::separate(cell_id, into = c("library_id", "cell_barcode"), remove = FALSE) |>
  # keep UMAP for eventual viz
  dplyr::rename(
    consensus = consensus_celltype_annotation, 
    UMAP1 = UMAP.1, 
    UMAP2 = UMAP.2
  ) |>
  dplyr::left_join(singler_df, by = "cell_id") |>
  dplyr::left_join(scanvi_df, by = "cell_id")

# Pull out metadata we'll want later
sample_metadata <- celltype_df |>
  dplyr::select(sample_id, library_id, is_xenograft) |>
  unique()

# We can remove the merged_sce now
rm(merged_sce)
```

Define closely-related groups of cell types that will be used for resolving annotations:

```{r}
#### Vectors of closely-related NBAtlas immune cell types
nbatlas_t_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "T cell") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

nbatlas_nk_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "NK cell") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

nbatlas_myeloid_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "Myeloid") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

# Note that cDC and Monocyte are not direct labels in the disseminated NBAtlas object,
# but these groupings are used in the paper's marker gene analysis.
nbatlas_cdc_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "cDC") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

nbatlas_monocyte_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "Monocyte") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

#### Vectors of closely-related consensus cell types
consensus_t_celltypes <- validation_df |> 
  dplyr::filter(validation_group_annotation == "T cell") |>
  dplyr::pull(consensus_annotation)

consensus_endo_celltypes <- validation_df |> 
  dplyr::filter(validation_group_annotation == "endothelial cell") |>
  dplyr::pull(consensus_annotation)

consensus_myeloid_celltypes <- validation_df |> 
  dplyr::filter(
    # consider only groups corresponding to what NBAtlas breaks up 
    # the Myeloid group into in the tumor zoom
    validation_group_annotation %in% c("myeloid", "monocyte", "macrophage")
  )|>
  dplyr::pull(consensus_annotation)
```

## Annotate

In this section, we'll derive the final annotations.

First, we'll label any cells that have exactly matching `scANVI` and `SingleR` labels.

```{r}
annotation_df <- celltype_df |>
  dplyr::select(cell_id, consensus, singler, scanvi) |>
  dplyr::mutate(
    cell_type_assignment = ifelse(
      scanvi == singler, scanvi, NA
    )
  )
```

Next, we'll get unified labels for any cells where at least two of three methods agree.

```{r}
annotation_df <- annotation_df |>
  get_unified_label("fibroblast", "Fibroblast", "Fibroblast") |>
  get_unified_label("macrophage", "Macrophage", "Macrophage") |>
  get_unified_label("neutrophil", "Neutrophil", "Neutrophil") |>
  get_unified_label("plasma cell", "Plasma", "Plasma") |>
  get_unified_label(consensus_endo_celltypes, "Endothelial", "Endothelial") |>
  get_unified_label("natural killer cell", nbatlas_nk_celltypes, "NK cell") |>
  get_unified_label("monocyte", nbatlas_monocyte_celltypes, "Monocyte") |>
  get_unified_label("dendritic cell", nbatlas_cdc_celltypes, "cDC") |>
  get_unified_label(consensus_t_celltypes, nbatlas_t_celltypes, "T cell") |>
  # This must be last since it is the broadest category
  get_unified_label(consensus_myeloid_celltypes, nbatlas_myeloid_celltypes, "Myeloid")
```

## Explore final annotations

What fraction of cells have been annotated?

```{r}
sum(!is.na(annotation_df$cell_type_assignment)) / nrow(annotation_df)
```

What fraction of cells have been annotated per library?

```{r, fig.width = 8, fig.height = 4}
frac_labeled_df <- annotation_df |>
  tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE) |>
  dplyr::group_by(library_id) |>
  dplyr::summarize(
    frac_labeled = sum(!is.na(cell_type_assignment))/dplyr::n(), 
    library_size = dplyr::n()
  ) |>
  # get PDX logical for coloring
  dplyr::inner_join(sample_metadata, by = "library_id")

p1 <- ggplot(frac_labeled_df) + 
  aes(x = frac_labeled) + 
  geom_histogram(bins = 40) +
  labs(x = "Fraction of library labeled") +
  theme_bw()

p2 <- ggplot(frac_labeled_df) + 
  aes(x = library_size, y = frac_labeled, color = is_xenograft) + 
  geom_point() + 
  labs(
    x = "Total cells in library",
    y = "Fraction of library labeled"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

p1 + p2
```

### UMAPs

In this section, we show the UMAP from the merged (not batch-corrected!) `SCPCP000004` object colored by cell type annotations and other potentially relevant metadata.

```{r}
# Finalize the annotations
submission_df <- annotation_df |>
  # add ontology ids
  dplyr::left_join(
    nbatlas_ontology_ids, 
    by = c("cell_type_assignment" = "NBAtlas_label")
  ) |>
  # recode labels
  dplyr::mutate(
    cell_type_assignment = dplyr::case_when(
      is.na(cell_type_assignment) ~ "Unknown", 
      cell_type_assignment == "NE" ~ "Neuroendocrine/tumor",
      .default = cell_type_assignment
   ),
    tumor_cell_classification = dplyr::case_when(
      cell_type_assignment == "Unknown" ~ "Unknown", 
      cell_type_assignment == "Neuroendocrine/tumor" ~ "tumor", 
      .default = "normal"
    )
  ) |>
  tidyr::separate(cell_id, into = c("library_id", "cell_barcode"), remove = FALSE) |>
  dplyr::select(
    cell_id, library_id, cell_barcode, cell_type_assignment, tumor_cell_classification, CL_ontology_id
  )
```

```{r}
# prepare data frame for plotting
umap_df <- celltype_df |>
  dplyr::select(cell_id, sample_id, library_id, is_xenograft, UMAP1, UMAP2) |>
  dplyr::left_join(submission_df, by = "cell_id") |>
  # group "same family" cell types in the factor for cleaner legend
  # note this will give warnings if any of those cells aren't in the final annotation, but that's fine
  dplyr::mutate(
    cell_type_assignment = factor(cell_type_assignment), 
    cell_type_assignment = forcats::fct_relevel(
      cell_type_assignment, 
      "T cell", nbatlas_t_celltypes, 
      "NK cell", nbatlas_nk_celltypes,
      "cDC", nbatlas_cdc_celltypes, 
      "Monocyte", nbatlas_monocyte_celltypes
    )
  )
```

#### Colored by cell type assignment

Note that in the UMAP below, cell types in the same "family" each share a color:

* T cells: `T cell`, `Treg`, `CD4+ T cell`, `CD8+ T cell`
* Natural killer cells: `NK cell`, `Circulating NK cell`, `Resident NK cell`, `TOX2+/KIT+ NK cell`
* Conventional dendritic cells: `cDC`, `cDC1`, `cDC2/DC3`, `Migratory cDC`
* Monocytes: `Monocyte`, `Classical monocyte`, `Patrolling monocyte`

```{r fig.width = 10, fig.height = 10}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = cell_type_assignment) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_manual(values = celltype_pal) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1), ncol=4))
```

#### Colored by tumor classification

```{r fig.width = 8, fig.height = 8}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = tumor_cell_classification) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_brewer(palette = "Set2") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
```

#### Colored by PDX status

```{r fig.width = 8, fig.height = 8}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = is_xenograft) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_brewer(palette = "Set1") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
```

## Export

```{r}
readr::write_tsv(
  submission_df,
  params$annotations_tsv
)
```

## Session Info

```{r}
sessionInfo()
```

### 2025-08-14 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
path

@@
   scanvi_pp_threshold: 0.75    
-  annotations_tsv: "../results/SCPCP000004-annotations.tsv"
+  annotations_tsv: "results/SCPCP000004-annotations.tsv"
 ---

### 2025-08-14 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
dont use git for paths since i dont want to update docker

--- 2025-08-14 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
@@
 
-We use the following approach to determine the final annotation.
+We use the following approach to determine the final annotation:
 
@@
 
-Next, we compare labels across all three methods: if _two out of three_ methods agree, we assign the agreeing label.
-Note that this requires mapping consensus cell types to NBAtlas labels, which is only possible for these cell types: endothelial cells, fibroblasts, and immune cell types except for the `NBAtlas` labels `RBCs` and `pDC` which do not have corresponding consensus labels.
+Next, we compare labels across all three methods: if only _two out of three_ methods agree, we assign the agreeing label.
+Note that this requires mapping consensus cell types to NBAtlas labels, which is only possible for these cell types: endothelial cells, fibroblasts, B cells, and immune cell types except for the `NBAtlas` labels `RBCs` and `pDC` which do not have corresponding consensus labels.
 
@@
 ```{r base paths}
-# The base path for the OpenScPCA repository, found by its (hidden) .git directory
-repository_base <- rprojroot::find_root(rprojroot::is_git_root)
+module_dir <- rprojroot::find_root(rprojroot::is_renv_project)
+repository_base <- file.path(module_dir, "..", "..")
 
@@
 
-```{r fig.width = 10, fig.height = 10}
+```{r fig.width = 8, fig.height = 8}
 ggplot(umap_df) +
@@
   scale_color_manual(values = celltype_pal) +
-  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1), ncol=4))
+  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
 ```

### 2025-08-14 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
restore some changes i accidentally got rid of

--- 2025-08-14 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
@@
 
-Next, we compare labels across all three methods: if _two out of three_ methods agree, we assign the agreeing label.
+Next, we compare labels across all three methods: if at least _two out of three_ methods agree, we assign the agreeing label.
 Note that this requires mapping consensus cell types to NBAtlas labels, which is only possible for these cell types: endothelial cells, fibroblasts, and immune cell types except for the `NBAtlas` labels `RBCs` and `pDC` which do not have corresponding consensus labels.
@@
   scale_color_manual(values = celltype_pal) +
-  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1), ncol=4))
+guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
 ```

### 2025-08-14 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
dont overwrite your own edit

@@
 merged_dir <- file.path(repository_base, "data", "current", "results", "merge-sce", "SCPCP000004")
-module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
 ref_dir <- file.path(module_dir, "references")

### 2025-08-14 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
aha, wrong palette file name
Ok, I had a palette bug

@@
   "palettes",
-  "submission-palette.tsv"
+  "nbatlas-cell-type-palette.tsv"
 )
@@
   )
+# update palette label as well
+names(celltype_pal)[names(celltype_pal) == "Neuroendocrine"] <- "Neuroendocrine/tumor"
 ```

### 2025-08-15 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd

@@
 
-### Distirbution of cell types
+### Distribution of cell types
 
@@
 PDX samples have, not unexpectedly, much less cell type diversity overall with most cells being tumor or similar and immune cells rarely present.
-This is consistent with expectations in general for the biology of PDX's.
+This is consistent with expectations in general for the biology of PDX samples.
 
@@
 
-This section shows a heatmap compaing SingleR annotations to consensus cell types, for PDX samples only.
+This section shows a heatmap comparing SingleR annotations to consensus cell types, for PDX samples only.
 

### 2025-08-18 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
WIP responding to reviews; not expected to render

@@
 
+```{r}
+# Source additional utilities functions:
+# - subset_nbatlas_markers()
+# - harmonize_celltypes()
+# - faceted_umap()
+# - generate_dotplot()
+source(file.path(module_dir, "scripts", "utils", "celltype-utils.R"))
+```
+
 ```{r file paths}
@@
 
-validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
+# marker genes from NBAtlas
+nbatlas_markers_file <- file.path(
+  ref_dir, 
+  "nbatlas-marker-genes.tsv"
+)
 ```
@@
 ```{r}
-validation_df <- readr::read_tsv(validation_url)
-
 palette_df <- readr::read_tsv(palette_file)
@@
 
-nbatlas_label_map <- readr::read_tsv(nbatlas_label_map_file)
-nbatlas_ontology_ids <- readr::read_tsv(nbatlas_ontology_file)
+label_map_df <- readr::read_tsv(nbatlas_label_map_file)
+ontology_df <- readr::read_tsv(nbatlas_ontology_file)
+
+nbatlas_markers_df <- readr::read_tsv(nbatlas_markers_file)
 ```
@@
   merged_sce,
-  use.coldata = c("sample_id", "is_xenograft", "consensus_celltype_annotation")
+  use.coldata = c("sample_id", "is_xenograft", "consensus_celltype_annotation", "consensus_celltype_ontology")
 ) |>
@@
     consensus = consensus_celltype_annotation, 
+    consensus_ontology = consensus_celltype_ontology,
     UMAP1 = UMAP.1, 
@@
 
-Define closely-related groups of cell types that will be used for resolving annotations:
-
-```{r}
-#### Vectors of closely-related NBAtlas immune cell types
-nbatlas_t_celltypes <- nbatlas_label_map |>
-  dplyr::filter(NBAtlas_label == "T cell") |>
-  dplyr::pull(NBAtlas_label_immune_zoom)
-
-nbatlas_nk_celltypes <- nbatlas_label_map |>
-  dplyr::filter(NBAtlas_label == "NK cell") |>
-  dplyr::pull(NBAtlas_label_immune_zoom)
-
-nbatlas_myeloid_celltypes <- nbatlas_label_map |>
-  dplyr::filter(NBAtlas_label == "Myeloid") |>
-  dplyr::pull(NBAtlas_label_immune_zoom)
-
-# Note that cDC and Monocyte are not direct labels in the disseminated NBAtlas object,
-# but these groupings are used in the paper's marker gene analysis.
-nbatlas_cdc_celltypes <- nbatlas_label_map |>
-  dplyr::filter(NBAtlas_label == "cDC") |>
-  dplyr::pull(NBAtlas_label_immune_zoom)
-
-nbatlas_monocyte_celltypes <- nbatlas_label_map |>
-  dplyr::filter(NBAtlas_label == "Monocyte") |>
-  dplyr::pull(NBAtlas_label_immune_zoom)
-
-#### Vectors of closely-related consensus cell types
-consensus_t_celltypes <- validation_df |> 
-  dplyr::filter(validation_group_annotation == "T cell") |>
-  dplyr::pull(consensus_annotation)
-
-consensus_endo_celltypes <- validation_df |> 
-  dplyr::filter(validation_group_annotation == "endothelial cell") |>
-  dplyr::pull(consensus_annotation)
-
-consensus_myeloid_celltypes <- validation_df |> 
-  dplyr::filter(
-    # consider only groups corresponding to what NBAtlas breaks up 
-    # the Myeloid group into in the tumor zoom
-    validation_group_annotation %in% c("myeloid", "monocyte", "macrophage")
-  )|>
-  dplyr::pull(consensus_annotation)
-```
-
 ## Annotate
@@
 
-First, we'll label any cells that have exactly matching `scANVI` and `SingleR` labels.
+```{r}
+ontology_slim_df <- ontology_df |>
+  dplyr::select(-CL_annotation)
+
+prep_for_annotation <- function(
+    df, 
+    annot_type, 
+    ontology_df, 
+    label_map_df) {
+
+ df |>
+    dplyr::select(all_of(c("cell_id", annot_type))) |>
+    dplyr::rename(label = annot_type) |>
+    # Join the family labels
+    dplyr::left_join(label_map_df, by = c("label" = "NBAtlas_label_immune_zoom")) |>
+    dplyr::rename(family_label = NBAtlas_label) |>
+    # Ensure "Unknown" has a family label instead of NA for future-proofing
+    dplyr::mutate(family_label = ifelse(is.na(family_label), "Unknown", family_label)) |>
+    # Join in the ontology labels to match the **celltype** ontologies (label column)
+    dplyr::left_join(ontology_slim_df, by = c("label" = "NBAtlas_label")) |>
+    dplyr::rename(label_ontology = CL_ontology_id) |>
+    # Join in the ontology labels to match the **family** ontologies (family_label column)
+    dplyr::left_join(ontology_slim_df, by = c("family_label" = "NBAtlas_label")) |>
+    dplyr::rename(family_ontology = CL_ontology_id) |>
+    # We don't need the labels now that we have ontologies
+    dplyr::select(-label, -family_label) |>
+    # rename columns to start with `annot_type`
+    dplyr::rename_with(\(x) {paste(annot_type, x, sep = "_")}, -cell_id)
+}
+
+singler_annotation_df <- prep_for_annotation(celltype_df, "singler", ontology_slim_df, label_map_df)
+scanvi_annotation_df <- prep_for_annotation(celltype_df, "scanvi", ontology_slim_df, label_map_df)
+
+annotation_df <- celltype_df |>
+  dplyr::select(cell_id, consensus_ontology) |>
+  dplyr::left_join(singler_annotation_df, by = "cell_id") |>
+  dplyr::left_join(scanvi_annotation_df, by = "cell_id") |>
+  dplyr::select(cell_id, consensus_ontology, starts_with("singler"), starts_with("scanvi"))
+```
 
 ```{r}
-annotation_df <- celltype_df |>
-  dplyr::select(cell_id, consensus, singler, scanvi) |>
+annotation_df <- annotation_df |> 
   dplyr::mutate(
-    cell_type_assignment = ifelse(
-      scanvi == singler, scanvi, NA
+    CL_ontology_id = dplyr::case_when(
+      # Use directly matching label
+      scanvi_label_ontology == singler_label_ontology ~ scanvi_label_ontology, 
+      # Check for family agreement between SingleR/scANVI
+      scanvi_family_ontology == singler_family_ontology ~ scanvi_family_ontology,
+      # Now use agreement with consensus to assign a label: first by label, and then by family
+      singler_label_ontology == consensus_ontology ~ singler_label_ontology,
+      scanvi_label_ontology == consensus_ontology ~ scanvi_label_ontology,
+      singler_family_ontology == consensus_ontology ~ singler_family_ontology,
+      scanvi_family_ontology == consensus_ontology ~ scanvi_family_ontology,
+      # Everything else is NA
+      .default = NA_character_
     )
-  )
+  ) |>
+  # keep only the final ontologies
+  dplyr::select(cell_id, CL_ontology_id)
+
+missing_cell_id <- annotation_df |>dplyr::filter(is.na(CL_ontology_id)) |>
+  dplyr::pull(cell_id)
+
+sjs_version <- readr::read_tsv(params$annotations_tsv)
+
+stephyes_allyno <- sjs_version |>
+  dplyr::filter(cell_id %in% missing_cell_id) |>
+  dplyr::filter(!is.na(CL_ontology_id)) |>
+  dplyr::pull(cell_id)
+
+celltype_df |>
+  dplyr::select(cell_id, singler, scanvi, consensus) |>
+  dplyr::filter(cell_id %in% stephyes_allyno)
 ```
@@
 ```{r}
-sum(!is.na(annotation_df$cell_type_assignment)) / nrow(annotation_df)
+sum(!is.na(annotation_df$CL_ontology_id)) / nrow(annotation_df)
 ```
@@
 
+#### Faceted by cell type assignment
+
+This UMAP is the same as the previous one, except each facet highlights one specific cell type.
+
+```{r fig.width = 16, fig.height = 16}
+umap_df |>
+  faceted_umap(
+    annotation_column = cell_type_assignment,
+    celltype_colors = celltype_pal
+  )
+```
+
 #### Colored by tumor classification
@@
 
+### Dot plot
+
+This section shows dot plots of marker gene expression for the final annotations.
+Specifically, we show the top 5 highest `log2FC` marker genes per NBAtlas cell type for validation.
+Marker genes that are present in three or more cell types are not included.
+Marker genes are taken directly from the NBAtlas paper where they were identified with `Seurat::FindMarkers()`.
+
+```{r}
+# data frame of top 5 upregulated genes per cell type
+top_nbatlas_markers_df <- subset_nbatlas_markers(nbatlas_markers_df, 5, "up") |>
+  # only markers in no more than 2 categories
+  # this is a good middle ground to ensure there are at least 2 markers per cell type
+  dplyr::add_count(ensembl_gene_id) |>
+  dplyr::filter(n < 3) |>
+  dplyr::select(-n) 
+```
+
+```{r, fig.width = 24, fig.height = 12}
+
+# get total number of cells per final annotation group and set up y_label
+total_cells_df <- full_celltype_df |>
+  dplyr::count(label, name = "total_cells") |>
+  # order so cell types are grouped
+  dplyr::inner_join(label_map_df, by = "label" ) |>
+  dplyr::mutate(
+    label_recoded = as.factor(label_recoded),
+    label_recoded = forcats::fct_relevel(label_recoded, levels(label_recoded_factor))
+  ) |>
+  dplyr::arrange(label_recoded) |>
+  dplyr::mutate(y_label = glue::glue("{label} ({total_cells})")) |>
+  # rename as expected for dotplot function - label_recoded needs to be the main column
+  dplyr::select(-label_recoded) |>
+  dplyr::rename(label_recoded = label)
+
+total_cells_df$y_label <- factor(total_cells_df$y_label, levels = rev(total_cells_df$y_label))
+nbatlas_bar_order <- total_cells_df$label_recoded
+```
+
+```{r, fig.width = 32, fig.height = 16}
+generate_dotplot(
+  merged_sce = merged_sce,
+  markers_df = top_nbatlas_markers_df,
+  # rename this after we've done all the wrangling
+  singler_df = full_celltype_df |> dplyr::rename(label_recoded = label),
+  total_cells_df = total_cells_df,
+  expressed_genes = expressed_genes,
+  bar_order = nbatlas_bar_order, 
+  celltype_palette = nbatlas_celltype_pal, 
+  min_cells = 0
+)
+```
+
 ## Export

### 2025-08-18 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
WIP: add dotplot code, some of which is temporary

@@
 theme_set(
-  theme_bw() +
-  theme(
-    axis.text = element_blank(), 
-    axis.ticks = element_blank(), 
-    legend.position = "bottom", 
-    aspect.ratio = 1
-  )
+  theme_bw()
 )
@@
   unique()
-
-# We can remove the merged_sce now
-rm(merged_sce)
 ```
@@
 ```{r fig.width = 16, fig.height = 16}
+umap_theme <- list(
+  theme(
+    axis.text = element_blank(), 
+    axis.ticks = element_blank(), 
+    legend.position = "bottom", 
+    aspect.ratio = 1
+  )
+)
+
 umap_df |>
@@
     celltype_colors = celltype_pal
-  )
+  ) + 
+  umap_theme
 ```
@@
   scale_color_brewer(palette = "Set2") +
-  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
+  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
+  umap_theme
 ```
@@
   scale_color_brewer(palette = "Set1") +
-  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
+  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
+  umap_theme
 ```
@@
 Specifically, we show the top 5 highest `log2FC` marker genes per NBAtlas cell type for validation.
-Marker genes that are present in three or more cell types are not included.
 Marker genes are taken directly from the NBAtlas paper where they were identified with `Seurat::FindMarkers()`.
 
+Marker genes that are present in three or more cell types are not included.
+In addition, the upper limit for the color scale has been censored to a maximum of 6.
+
 ```{r}
+# Temporary to develop code for next section
+annotation_df <- readr::read_tsv(params$annotations_tsv) |>
+  dplyr::filter(cell_type_assignment != "Unknown") |>
+  # recode cell types to match what NBAtlas uses for their marker genes
+  dplyr::mutate(
+    dotplot_celltype = dplyr::case_when(
+      stringr::str_detect(cell_type_assignment, "monocyte") ~ "Monocyte", 
+      stringr::str_detect(cell_type_assignment, "cDC") ~ "cDC", 
+      stringr::str_detect(cell_type_assignment, "Neuroendocrine") ~ "Neuroendocrine", 
+      .default = cell_type_assignment
+    )
+  )
+
 # data frame of top 5 upregulated genes per cell type
@@
 
-```{r, fig.width = 24, fig.height = 12}
-
+```{r}
 # get total number of cells per final annotation group and set up y_label
-total_cells_df <- full_celltype_df |>
-  dplyr::count(label, name = "total_cells") |>
-  # order so cell types are grouped
-  dplyr::inner_join(label_map_df, by = "label" ) |>
-  dplyr::mutate(
-    label_recoded = as.factor(label_recoded),
-    label_recoded = forcats::fct_relevel(label_recoded, levels(label_recoded_factor))
-  ) |>
-  dplyr::arrange(label_recoded) |>
-  dplyr::mutate(y_label = glue::glue("{label} ({total_cells})")) |>
-  # rename as expected for dotplot function - label_recoded needs to be the main column
-  dplyr::select(-label_recoded) |>
-  dplyr::rename(label_recoded = label)
+total_cells_df <- annotation_df |>
+  dplyr::count(dotplot_celltype, name = "total_cells") |>
+  # remove infrequent cell types
+  dplyr::filter(total_cells > 10) |>
+  dplyr::mutate(y_label = glue::glue("{dotplot_celltype} ({total_cells})")) |>
+  # order by count, but put Unknown at the end
+  dplyr::arrange(desc(total_cells)) |> 
+  # amazing new dplyr trick; only this row is TRUE, so it gets moved to the last row
+  dplyr::arrange(dotplot_celltype == "Unknown") |>
+  # the dotplot function expects this column name
+  dplyr::rename(label_recoded = dotplot_celltype)
 
@@
 nbatlas_bar_order <- total_cells_df$label_recoded
+```
+
+```{r}
+# Prepare the expressed_genes vector
+# we only care about if that gene is expressed otherwise we won't waste memory and include it
+gene_sums <- rowData(merged_sce) |>
+  as.data.frame() |>
+  dplyr::select(contains("detected")) |>
+  as.matrix() |>
+  rowSums()
+expressed_genes <- names(gene_sums)[gene_sums > 0]
 ```
@@
   # rename this after we've done all the wrangling
-  singler_df = full_celltype_df |> dplyr::rename(label_recoded = label),
+  celltype_df = annotation_df |> dplyr::rename(label_recoded = dotplot_celltype),
   total_cells_df = total_cells_df,
@@
   bar_order = nbatlas_bar_order, 
-  celltype_palette = nbatlas_celltype_pal, 
+  celltype_palette = celltype_pal, 
   min_cells = 0
-)
+) +
+  theme(legend.position = "bottom")
 ```

### 2025-08-18 analyses/cell-type-consensus/exploratory-notebooks/06-update-consensus-scimilarity.Rmd
notebook for exploring consensus cell type pairs

---
title: "Summary of cell type ontologies in reference files"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
    df_print: "kable"
---

This notebook explores incorporating annotations from the `SCimilarity` foundational model to the consensus cell type reference. 
The full list of cell type annotations possible for `SCimilarity` can be found in the [`cell-type-scimilarity` module](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-scimilarity/references/scimilarity-mapped-ontologies.tsv).

For the updated consensus cell types, we plan to assign a consensus cell type label if 2 of the 3 cell type annotations agree.

- First, all possible combinations of the three annotations are identified. 
- Then we take all pairs for each combination and find the latest common ancestor (LCA) using the [`ontoProc::findCommonAncestors()` function](https://rdrr.io/bioc/ontoProc/man/findCommonAncestors.html).
- The LCA will then be assigned to the consensus cell type label for that _pair_ if it meets a set of rules that we previously defined (total descendants <= 170 with some exceptions). 
We will look at all cell type labels that are kept and discarded using our previous rules and identify if we need to make updates to the exceptions. 
- Finally we will map the LCA for each possible pair back to the combinations of all three cell types. 
If one possible LCA is identified, that will be used as the consensus cell type label. 
If there are multiple possible LCAs, we will choose the most specific term as the label.

## Setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_bw()
)
```

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The path to this module
ref_dir <- file.path(repository_base, "analyses", "cell-type-consensus", "references")

# path to ref files
panglao_file <- file.path(ref_dir, "panglao-cell-type-ontologies.tsv")
blueprint_file <- file.path(ref_dir, "blueprint-mapped-ontologies.tsv")
scimilarity_file <- file.path(repository_base, "analyses", "cell-type-scimilarity", "references", "scimilarity-mapped-ontologies.tsv")
```

```{r}
# grab obo file
cl_ont <- ontologyIndex::get_ontology("http://purl.obolibrary.org/obo/cl/releases/2024-09-26/cl-basic.obo")

# set up the graph to use for assigning LCA terms 
parent_terms <- cl_ont$parents
cl_graph <- igraph::make_graph(rbind(unlist(parent_terms), rep(names(parent_terms), lengths(parent_terms))))
```

```{r, message=FALSE}
# read in reference files
panglao_df <- readr::read_tsv(panglao_file) |>
  # rename columns to have panglao in them for easy joining later
  dplyr::select(
    panglao_ontology = "ontology_id",
    panglao_annotation = "human_readable_value",
    original_panglao_name = "panglao_cell_type" # keep original name since some map to the same ontology ID
  ) |> 
  # remove any cell types that don't have ontologies 
  tidyr::drop_na()

# read in blueprint data 
blueprint_df <- readr::read_tsv(blueprint_file)

# read in scimilarity 
scimilarity_df <- readr::read_tsv(scimilarity_file) |> 
  dplyr::select(
    scimilarity_ontology = "scimilarity_celltype_ontology",
    scimilarity_annotation = "human_readable_value",
    original_scimilarity_name = "scimilarity_celltype_annotation"
  )
```

```{r, message=FALSE}
# get total descendants for each term in CL  
# turn cl_ont into data frame with one row per term
cl_df <- data.frame(
  cl_ontology = cl_ont$id,
  cl_annotation = cl_ont$name
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    descendants = list(ontologyIndex::get_descendants(cl_ont, cl_ontology, exclude_roots = TRUE)),
    total_descendants = length(descendants)
  )
```

## Determine LCA for all possible pairs

In this section we will first create a table of all possible combinations of the three cell type annotations (`SingleR`, `CellAssign`, and `SCimilarity`). 
For each combination, we will get all possible pairs and compute the LCA.

```{r, warning=FALSE, message=FALSE}
# get a data frame with all combinations of panglao, blueprint, and scimilarity terms
# one row for each combination
all_ref_df <- expand.grid(
  panglao_ontology = panglao_df$panglao_ontology,
  blueprint_ontology = blueprint_df$blueprint_ontology,
  scimilarity_ontology = scimilarity_df$scimilarity_ontology
) |> 
  # remove any potentially redundant combinations based on ontology Ids 
  dplyr::distinct() |> 
  # create a column with all possible pairs between the three tools 
  dplyr::mutate(pair_1 = glue::glue("{panglao_ontology};{blueprint_ontology}"),
                pair_2 = glue::glue("{panglao_ontology};{scimilarity_ontology}"),
                pair_3 = glue::glue("{blueprint_ontology};{scimilarity_ontology}")) |> 
  tidyr::pivot_longer(cols = c(pair_1, pair_2, pair_3), names_to = "number", values_to = "pairs")

# get all possible pairs from each combination and get lca for each pair that is unique 
lca_pairs_df <- all_ref_df |> 
  dplyr::select(pairs) |> 
  dplyr::distinct() |> 
  tidyr::separate(pairs, into = c("id1", "id2"), sep = ";") |> 
  dplyr::rowwise() |>
  dplyr::mutate(
    # least common shared ancestor
    lca = list(rownames(ontoProc::findCommonAncestors(id1, id2, g = cl_graph)))
  )
```

```{r, warning=FALSE}
# add lca and total number of descendants to data frame
# expand to have one row per unique combo + unique lca
# remember sometimes there are multiple LCA that will get returned 
lca_df <- lca_pairs_df |>
  dplyr::mutate(
    total_lca = length(lca), # get total number for filtering later 
    lca = paste0(lca, collapse = ",") # make it easier to split the lca terms
  ) |>
  # split each lca term into its own column
  tidyr::separate(lca, into = c("lca_1", "lca_2", "lca_3"), sep = ",") |>
  # transpose so that instead of lca being in a column there is one row per lca 
  tidyr::pivot_longer(
    cols = dplyr::starts_with("lca"),
    names_to = "lca_number",
    values_to = "lca"
  ) |>
  tidyr::drop_na() |>
  dplyr::select(-lca_number) |> # don't need this column 
  # account for any cases where the ontology IDs are exact matches
  # r complains about doing this earlier since the lca column holds lists until now
  dplyr::mutate(lca = dplyr::if_else(id1 == id2, id1, lca)) |> 
  # join in information for each of the lca terms including name and number of descendants
  dplyr::left_join(cl_df, by = c("lca" = "cl_ontology")) |> 
  # add in legible names for id1/id2 pairs 
  dplyr::mutate(
    id1_annotation = cl_ont$name[id1],
    id2_annotation = cl_ont$name[id2]
  )

```

### Retained celltypes using existing rules

We previously excluded any LCA that had > 170 descendants or belonged to one of these cell types: "bone cell", "lining cell", "blood cell", "progenitor cell", "supporting cell".

Let's see what cell types we would keep if we kept these rules and see if any of them seem too "broad".

```{r}
# vector of cell types that we want to exclude 
celltypes_to_exclude <- c("bone cell", "lining cell", "blood cell", "progenitor cell", "supporting cell")

# get only celltypes that remain after implementing the previous rules 
lca_df |>
  dplyr::filter(total_descendants <= 170 & !(cl_annotation %in% celltypes_to_exclude)) |>
  dplyr::select(cl_annotation, total_descendants) |>
  dplyr::arrange(desc(total_descendants)) |>
  unique() |> 
  # print out in a readable table
  DT::datatable(rownames = FALSE)
```

There are a few cell types we might want to explore further that might seem too broad. 
Let's print out a table that look at the possible matches that give rise to those cell types below.

```{r}
celltypes_to_explore <- c(
  "peptide hormone secreting cell",
  "embryonic cell (metazoa)",
  "biogenic amine secreting cell", 
  "protein secreting cell",
  "antibody secreting cell",
  "extracellular matrix secreting cell",
  "surfactant secreting cell"
)

# for all cell types we want to look at print out a table of the pairs that give rise to that cell type
celltypes_to_explore |> 
  purrr::map(
    \(celltype){
      lca_df |> 
        dplyr::filter(cl_annotation == {{celltype}}) |> 
        dplyr::select(id1_annotation, id2_annotation, total_lca, cl_annotation) |> 
        unique() |> 
        DT::datatable(rownames = FALSE)
    }
  )
```

Looking at these tables, I might get rid of "biogenic amine secreting cell", "protein secreting cell", and "extracellular matrix secreting cell". 
These cell types arise because of matches from cells that have similar functions but are very different. 
For example, platelets and mast cells are pretty different but give the LCA term "biogenic amine secreting cell".

I would keep the following:

- Matches that result in "peptide hormone secreting cell" all appear to be from the pancreas. 
- Matches that result in "embryonic cell (metazoa)" are all developmental cell types that are found in the embryo and mirror some pediatric cancer cell types. 
I think having that information could be useful. 
- "Antibody secreting cell" refers to "A lymphocyte of B lineage" according to [Cell Ontology](https://www.ebi.ac.uk/ols4/ontologies/cl/classes/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FCL_0000946) and I think knowing that a cell is some sort of B cell is helpful.
- "Surfactant secreting cell" is specific to the lung and both cell types that appear in matches are lung cell types.

One thing to note is that most of these have a total LCA count > 1 meaning there is at least one other match for those cell types. 
Maybe we should consider removing anything with more than 1 LCA which is what we were doing previously when we only were matching Blueprint and Panglao.

```{r}
# updated celltypes to exclude 
celltypes_to_exclude <- c(celltypes_to_exclude, "biogenic amine secreting cell", "protein secreting cell", "extracellular matrix secreting cell")

# create a df with just lca that meet the specified criteria so far 
filtered_celltype_df <- lca_df |>
  dplyr::filter(total_descendants <= 170  & !cl_annotation %in% celltypes_to_exclude)

# print out all matches when total lca > 1
filtered_celltype_df |> 
  dplyr::filter(total_lca > 1) |> 
  dplyr::select(cl_annotation) |> 
  unique() |> 
  DT::datatable(rownames = FALSE)
```

Previously when we looked at the LCA for pairs with moret than 1 match, we had seen only very broad cell types. 
That doesn't seem to be the case here as we see things like "memory T cell".

Perhaps instead of just removing the pairs with multiple LCA, we first filter out the terms that don't meet our criteria. 
Then if multiple LCA remain we find the LCA of all the remaining LCA's and if that new LCA meets our criteria we use that instead.

```{r, warning=FALSE, message=FALSE}
# get the new LCA number after removing the non-specific terms 
filtered_celltype_df <- filtered_celltype_df |> 
  dplyr::group_by(id1, id2) |> 
  dplyr::mutate(new_lca_pair = length(lca))

# now pull out any pairs where only one of the LCA remains 
# find the LCA between the remaining ones and then filter based on total descendants
# are there any that are remaining that still have multiple LCA? 
new_lca_df <- filtered_celltype_df |>
  dplyr::filter(new_lca_pair > 1) |>
  dplyr::group_by(id1, id2) |> 
  dplyr::summarize(
    updated_lca = list(unique(rownames(ontoProc::findCommonAncestors(lca, g = cl_graph))))
    ) |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    new_total_lca = length(updated_lca),
    updated_lca = paste0(updated_lca, collapse = ","),
    id1_annotation = cl_ont$name[id1],
    id2_annotation = cl_ont$name[id2]
  ) |> 
  # split each lca term into its own column
  tidyr::separate(updated_lca, into = c("updated_lca_1", "updated_lca_2", "updated_lca_3"), sep = ",") |>
  # transpose so that instead of lca being in a column there is one row per lca 
  tidyr::pivot_longer(
    cols = dplyr::starts_with("updated_lca"),
    names_to = "lca_number",
    values_to = "updated_lca"
  ) |>
  tidyr::drop_na() |>
  dplyr::select(-lca_number) |>
   dplyr::left_join(cl_df, by = c("updated_lca" = "cl_ontology"))
```

First for any pairs that have LCA > 1, what are they? 
```{r}
new_lca_df |> 
  dplyr::filter(new_total_lca > 1) |> 
  dplyr::select(id1_annotation, id2_annotation, cl_annotation, total_descendants)
```

I think here we just want to keep "kidney tubule cell" if one of the LCA's includes that term. 
In that case, there are three possible LCA but "kidney tubule cell" is the most specific (fewest descendants).

In the other cases, we should keep any possible LCA that meet the total descendants criteria and remove all others.

Now let's look at the remaining matches that give rise to 1 LCA and see what pairs pass the 170 total descendants threshold.

```{r}
new_lca_df |> 
  dplyr::filter(total_descendants <= 170 & new_total_lca == 1) |> 
  dplyr::select(id1_annotation, id2_annotation, cl_annotation)
```

These all seem like reasonable assignments. 
Let's create a dataframe with the pairs and final matches for these so we can update the whole assignment table later.

```{r}
# firt get a dataframe of the matches we will keep that had multiple LCA and their final annotations
multi_lca_keep_df <- new_lca_df |> 
  dplyr::filter(total_descendants <= 170 & cl_annotation != "epithelial cell of nephron" | cl_annotation == "kidney tubule cell") |> 
  dplyr::select(id1, id2, id1_annotation, id2_annotation, lca = updated_lca, cl_annotation, total_descendants) |> 
  unique()
```

### Discarded cell types

Let's look at what possible cell types would be discarded if we used these rules. 
Previously we were keeping "neuron" and "epithelial cell" even though they had a larger number of descendants. 
Are there any other cell types that appear that we should include in that list?

```{r}
lca_df |>
  dplyr::filter(total_descendants > 170) |>
  dplyr::select(cl_annotation, total_descendants) |>
  dplyr::arrange(desc(total_descendants)) |>
  unique() |> 
  DT::datatable(rownames = FALSE)
```

We should probably also include "columnar/cuboidal epithelial cell" and "endo-epithelial cell". 
Let's look at the matches that led to those LCA to see if there is any reason we should not include them.

```{r}
lca_df |> 
  dplyr::filter(cl_annotation == "columnar/cuboidal epithelial cell") |> 
  dplyr::select(id1_annotation, id2_annotation, cl_annotation) |> 
  unique() |> 
  DT::datatable(rownames = FALSE)
```

```{r}
lca_df |> 
  dplyr::filter(cl_annotation == "endo-epithelial cell") |> 
  dplyr::select(id1_annotation, id2_annotation, cl_annotation) |> 
  unique() |> 
  DT::datatable(rownames = FALSE)
```

I think we should add these two terms to our list of cell types to keep alongside "neuron" and "epithelial cell" when the total descendants is higher than 170.

## Updated consensus cell type pairs

Using the above exploration, we should update our rules for labeling the pairs with their LCA to fit one of the following criteria:

- A single LCA term with <= 170 total descendants
- LCA term that is not one of: "bone cell", "lining cell", "blood cell", "progenitor cell", "supporting cell", "biogenic amine secreting cell", "protein secreting cell", or "extracellular matrix secreting cell"
- LCA term that is "neuron" or "epithelial cell"
- A secondary LCA term (LCA computed all LCAs returned from the initial pairing) that is a single LCA value with <= 170 total descendants or one of "epithelial cell", "kidney tubule cell", or "hematopoietic precursor cell"

```{r}
# now filter to remove the multi LCA matches and add back in the updated assignments for them 
filtered_lca_df <- lca_df |> 
  # first set criteria to keep cell types, > 170, celltypes of interest 
  dplyr::filter(total_descendants <= 170 & !(cl_annotation %in% celltypes_to_exclude) 
                | (cl_annotation == "neuron"
                   # only keep epithelial if not keratinocyte
                | (cl_annotation == "epithelial cell" & id1_annotation != "keratinocyte" & id2_annotation != "keratinocyte"))) |> 
  dplyr::group_by(id1, id2) |> 
  # after filtering, see how many LCAs are left
  dplyr::mutate(new_lca_pair = length(lca)) |> 
   # everything with more than 1 lca gets removed with the exception of HSCs
  dplyr::filter(new_lca_pair <=1 | cl_annotation == "hematopoietic precursor cell") |> 
  dplyr::select(id1, id2, id1_annotation, id2_annotation, lca, cl_annotation, total_descendants) |> 
  # add in pairs with multi LCA and their final match 
  dplyr::bind_rows(multi_lca_keep_df) |> 
  unique()
```

Let's make sure we only have unique combinations:

```{r}
dup <- filtered_lca_df |> 
  dplyr::select(id1, id2) |> 
  duplicated()

dim(filtered_lca_df[dup, ])[1]
```

## Obtaining the final consensus cell type

Now that we have the LCA to use for all possible pairs, we need to combine them back with the possible combinations of all three annotation methods to obtain the final consensus cell type. 
For each combination there should be three possible LCA.

- If they are all unknown and no LCA was assigned, then no consensus cell type will be assigned. 
- If there is one unique LCA then that will be used as the consensus cell type. 
- If there are more than one unique LCA, then we will need to decide how to assign the consensus cell type.

```{r}
# combine lca with all combinations 
combined_ref_df <- all_ref_df |> 
  # split pairs to join with final lca assignment
  tidyr::separate(col = "pairs", into = c("id1", "id2"), sep = ";") |> 
  dplyr::left_join(filtered_lca_df, by = c("id1", "id2")) |> 
  # get rid of any combinations that don't have any possible matches 
  tidyr::drop_na()
```

Let's see how often we have a single unique match. 
If there's only one match we'll use that, otherwise we need to figure out how to make the final assignment.

```{r}
consensus_ref_df <- combined_ref_df |> 
  dplyr::group_by(panglao_ontology, blueprint_ontology, scimilarity_ontology) |> 
  dplyr::summarize(
    total_matches = length(unique(lca)),
    consensus_celltype_ontology = dplyr::case_when(
      total_matches == 1 ~ lca,
      total_matches > 1 ~ paste0(unique(lca), collapse = ",")
    )
  )

table(consensus_ref_df$total_matches)
```

TODO: Figure out how to assign the LCA for ~ 16000 combinations that result in multiple LCAs

## Session info

```{r}
sessionInfo()
```

### 2025-08-18 analyses/cell-type-consensus/exploratory-notebooks/06-update-consensus-scimilarity.Rmd
include extra epithelial cells

@@
         unique() |> 
-        DT::datatable(rownames = FALSE)
+        kable()
     }
@@
   dplyr::filter(total_descendants <= 170 & !(cl_annotation %in% celltypes_to_exclude) 
-                | (cl_annotation == "neuron"
+                | cl_annotation %in% c("neuron","columnar/cuboidal epithelial cell", "endo-epithelial cell")
                    # only keep epithelial if not keratinocyte
-                | (cl_annotation == "epithelial cell" & id1_annotation != "keratinocyte" & id2_annotation != "keratinocyte"))) |> 
+                | (cl_annotation == "epithelial cell" & id1_annotation != "keratinocyte" & id2_annotation != "keratinocyte")) |> 
   dplyr::group_by(id1, id2) |> 

### 2025-08-18 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
Fix some spelling while I'm here for #1276
substantially simplify code using updated family TSV

@@
 # Source additional utilities functions:
-# - select_nbatlas_markers()
+# - subset_nbatlas_markers()
 # - harmonize_celltypes()
@@
 
-# label mapping
-nbatlas_label_map <- readr::read_tsv(nbatlas_label_map_file)
-
 # set up palettes
@@
 sample_metadata <- readr::read_tsv(sample_metadata_file)
+
+# label mapping
+nbatlas_label_map <- readr::read_tsv(nbatlas_label_map_file)
 ```
@@
   tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE)
-
-# can remove now for space
-rm(merged_sce)
 ```
@@
 ```{r, fig.width = 12, fig.height = 8}
-# define families using the provided map
-nbatlas_t_celltypes <- nbatlas_label_map |>
-  dplyr::filter(NBAtlas_label == "T cell") |>
-  dplyr::pull(NBAtlas_label_immune_zoom)
-
-nbatlas_nk_celltypes <- nbatlas_label_map |>
-  dplyr::filter(NBAtlas_label == "NK cell") |>
-  dplyr::pull(NBAtlas_label_immune_zoom)
-
-# Note that cDC and Monocyte are not direct labels in the disseminated NBAtlas object,
-# but these groupings are used in the paper's marker gene analysis.
-nbatlas_cdc_celltypes <- nbatlas_label_map |>
-  dplyr::filter(NBAtlas_label == "cDC") |>
-  dplyr::pull(NBAtlas_label_immune_zoom)
-
-nbatlas_monocyte_celltypes <- nbatlas_label_map |>
-  dplyr::filter(NBAtlas_label == "Monocyte") |>
-  dplyr::pull(NBAtlas_label_immune_zoom)
-
-# we'll use these later but not for this plot
-nbatlas_myeloid_celltypes <- nbatlas_label_map |>
-  dplyr::filter(NBAtlas_label == "Myeloid") |>
-  dplyr::pull(NBAtlas_label_immune_zoom)
+label_order <- nbatlas_label_map |>
+  dplyr::add_count(NBAtlas_family) |>
+  dplyr::arrange(desc(n), NBAtlas_family) |>
+  dplyr::pull(NBAtlas_label)
 
@@
   # note this will give warnings if any of those cells aren't in the annotation, but that's fine
-  dplyr::mutate(
-    label = factor(label), 
-    label = forcats::fct_relevel(
-      label, 
-      "T cell", nbatlas_t_celltypes, 
-      "NK cell", nbatlas_nk_celltypes,
-      "cDC", nbatlas_cdc_celltypes, 
-      "Monocyte", nbatlas_monocyte_celltypes
-    )
-  )
+  dplyr::mutate(label = factor(label, levels = label_order))
 
@@
   * Monocytes: `Monocyte`, `Classical monocyte`, `Patrolling monocyte`
-  * Myeloid: `Neutrophil` and `Macrophage`
 * Fraction of labels that entirely disagree
@@
   dplyr::filter(celltype_method != "validation_group_annotation") |>
-  # create label_family grouping
-  dplyr::mutate(
-    label_family = dplyr::case_when(
-      label %in% nbatlas_t_celltypes ~ "T cell",
-      label %in% nbatlas_nk_celltypes ~ "NK cell",
-      label %in% nbatlas_cdc_celltypes ~ "cDC",
-      label %in% nbatlas_monocyte_celltypes ~ "Monocyte",
-      label %in% nbatlas_myeloid_celltypes ~ "Myeloid", 
-      .default = label
-    )
+  dplyr::left_join(
+    nbatlas_label_map, by = c("label" = "NBAtlas_label")
   ) |>
-  dplyr::select(-label_recoded, -label) |>
+  dplyr::select(cell_id, library_id, celltype_method, label) |>
   tidyr::pivot_wider(
     names_from = celltype_method,
-    values_from = label_family
+    values_from = label
   )

### 2025-08-19 analyses/cell-type-consensus/exploratory-notebooks/06-update-consensus-scimilarity.Rmd
fix some rendering

@@
         dplyr::select(id1_annotation, id2_annotation, total_lca, cl_annotation) |> 
-        unique() |> 
-        kable()
+        unique()
     }

### 2025-08-19 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
a few more cleanups

@@
 theme_set(theme_bw())
+umap_theme <- list(
+  theme(
+    aspect.ratio = 1,
+    legend.position = "bottom", 
+    axis.text = element_blank(), 
+    axis.ticks = element_blank()
+  ) 
+)
 set.seed(2025) # used for some viz
@@
 )
+
 ```
@@
 
-Since there are many cell types in this plot, it has been colored so that cell types in the same "family" each share a color:
+Since there are many cell types in this plot, it has been colored so that all cell types in a given "family" each share a color, where we consider these family groupings:
 
-* T cells: `T cell`, `Treg`, `CD4+ T cell`, `CD8+ T cell`
-* Natural killer cells: `NK cell`, `Circulating NK cell`, `Resident NK cell`, `TOX2+/KIT+ NK cell`
-* Conventional dendritic cells: `cDC`, `cDC1`, `cDC2/DC3`, `Migratory cDC`
-* Monocytes: `Monocyte`, `Classical monocyte`, `Patrolling monocyte`
+* T cells
+* Natural killer cells
+* Conventional dendritic cells
+* Monocytes
 
@@
   dplyr::pull(NBAtlas_label)
+label_order <- c(label_order, "Unknown")
 
@@
   facet_wrap(vars(celltype_method)) +
-  coord_equal() + 
-  theme(
-    legend.position = "bottom", 
-    axis.text = element_blank(), 
-    axis.ticks = element_blank()
-  ) +
+  umap_theme +
   guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
@@
   facet_wrap(vars(celltype_method)) +
-  coord_equal() +
-  theme(
-    legend.position = "bottom", 
-    axis.text = element_blank(), 
-    axis.ticks = element_blank()
-  ) +
+  umap_theme +
   guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))

### 2025-08-19 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
add final annotation notebook, massively refreshed from prior versions

@@
   scanvi_pp_threshold: 0.75    
-  annotations_tsv: "results/SCPCP000004-annotations.tsv"
+  annotations_tsv: "../results/SCPCP000004-annotations.tsv"
 ---
 
-In this analysis module, we have annotated cell types in `SCPCP000004` samples using two methods: `SingleR` and `scANVI/scArches`, both using the `NBAtlas` dataset as a reference.
+In this analysis module, we have annotated cell types in `SCPCP000004` samples using two methods: `SingleR` and `scANVI/scArches`, both using the NBAtlas dataset as a reference.
 In this notebook, we derive the final annotations using information from the `SingleR` labels, the `scANVI/scArches` labels, and existing consensus cell type annotations.
-All final annotations use labels present in the `NBAtlas` dataset.
+All final annotations use labels present in the NBAtlas dataset.
 
@@
 Next, we compare labels across all three methods: if at least _two out of three_ methods agree, we assign the agreeing label.
-Note that this requires mapping consensus cell types to NBAtlas labels, which is only possible for these cell types: endothelial cells, fibroblasts, and immune cell types except for the `NBAtlas` labels `RBCs` and `pDC` which do not have corresponding consensus labels.
+Note that this requires mapping consensus cell types to NBAtlas labels, which is only possible for these cell types: endothelial cells, fibroblasts, and immune cell types except for the NBAtlas labels `RBCs` and `pDC` which do not have corresponding consensus labels.
 
@@
 * Cell type annotations can have directly corresponding labels. For example:
-  * If the consensus cell type is `endothelial cell` and at least one of the `NBAtlas`-derived annotations is `Endothelial`, we assign the label `Endothelial`.
+  * If the consensus cell type is `endothelial cell` and at least one of the NBAtlas-derived annotations is `Endothelial`, we assign the label `Endothelial`.
 * Cell type annotations can differ but be in the same "family" of cell types. For example:
-  * If the consensus cell type is `mature T cell` and at least one of the `NBAtlas`-derived annotations is `CD4+ T cell`, we assign the broader label `T cell`.
-  * If one `NBAtlas`-derived annotation is `Patrolling monocyte` and the other is `Classical monocyte`, we assign the broader label `Monocyte`.
+  * If the consensus cell type is `mature T cell` and at least one of the NBAtlas-derived annotations is `CD4+ T cell`, we assign the broader label `T cell`.
+  * If one NBAtlas-derived annotation is `Patrolling monocyte` and the other is `Classical monocyte`, we assign the broader label `Monocyte`.
 
-Finally, to the extent possible, we assign CL ontology IDs to the final annotations before export.
+We use CL ontology ids to determine agreement throughout. 
+For more information about how these ontology ids were determined for NBAtlas labels, see `../references/README.md`.
+
+For the final exported TSV file, we further designate any cells labeled as `Neuroendocrine` as `tumor`, following how this label is interpreted in NBAtlas.
 
@@
 )
+umap_theme <- list(
+  theme(
+    axis.text = element_blank(), 
+    axis.ticks = element_blank(), 
+    legend.position = "bottom", 
+    aspect.ratio = 1
+  )
+)
 
@@
 )
-```
-
-### Functions
-
-The function `get_unified_label()` defined in the chunk below determines a final label based on correspondence between at least two out of three labels, out of `SingleR`, `scANVI/scArches`, and the consensus label.
-
-```{r}
-#' Get unified cell type label
-#'
-#' @param annotation_df Data frame to annotate with columns `consensus`, `singler`, `scanvi`, and `cell_type_assignment`
-#' @param consensus_options Vector of consensus cell types that match the final label, used for `consensus`
-#' @param nbatlas_options Vector of NBAtlas cell types that match the final label, used for `singler` and `scanvi`
-#' @param final_label Final label to apply if at least two methods have labels present in their respective vectors
-#'
-#' @returns Updated annotation_df with `cell_type_assignment` filled in with the `final_label` value where appropriate
-get_unified_label <- function(
-    annotation_df, 
-    consensus_options, 
-    nbatlas_options, 
-    final_label) {
-
-  annotation_df |>
-    dplyr::mutate(
-      # count how many annotations are in agreement
-      n = consensus %in% consensus_options +
-        singler %in% nbatlas_options + 
-        scanvi %in% nbatlas_options, 
-      # if >=2 agree, use this annotation
-      cell_type_assignment = ifelse(
-        is.na(cell_type_assignment) & n >= 2, 
-        final_label, 
-        cell_type_assignment
-    )) |>
-    dplyr::select(-n)
-}
 ```
@@
 scanvi_dir <- file.path(results_dir, "scanvi")
-```
-
-```{r}
-# Source additional utilities functions:
-# - subset_nbatlas_markers()
-# - harmonize_celltypes()
-# - faceted_umap()
-# - generate_dotplot()
-source(file.path(module_dir, "scripts", "utils", "celltype-utils.R"))
 ```
@@
 )
+
+# validation groups URL, used to obtain consensus "family" ontologies (aka broad validation group ontologies)
+validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
+```
+
+### Functions
+
+```{r}
+# Source utilities functions:
+# - subset_nbatlas_markers()
+# - faceted_umap()
+# - generate_dotplot()
+source(file.path(module_dir, "scripts", "utils", "celltype-utils.R"))
+```
+
+```{r}
+
+#' Prepare data frame of scANVI or SingleR labels for annotation
+#'
+#' @param df Data frame to prepare
+#' @param annot_type "singler" or "scanvi"
+#' @param ontology_df Data frame of ontology ids
+#' @param label_map_df Data frame mapping labels to family labels
+#'
+#' @returns Wide data frame with ontologies and family labels for the given annotation type
+prep_for_annotation <- function(
+    df, 
+    annot_type, 
+    ontology_df, 
+    label_map_df) {
+
+  df |>
+    dplyr::select(all_of(c("cell_id", annot_type))) |>
+    dplyr::rename(label = annot_type) |>
+    ####### Join in the family labels
+    dplyr::left_join(label_map_df, by = c("label" = "NBAtlas_label")) |>
+    dplyr::rename(family = NBAtlas_family) |>
+    ######### Obtain LABEL ontologies
+    dplyr::left_join(ontology_slim_df, by = c("label" = "NBAtlas_label")) |>
+    dplyr::rename(label_ontology = CL_ontology_id) |>
+    ######## Obtain FAMILY ontologies
+    dplyr::left_join(ontology_slim_df, by = c("family" = "NBAtlas_label")) |>
+    dplyr::rename(family_ontology = CL_ontology_id) |>
+    # rename columns to start with `annot_type`
+    dplyr::rename_with(\(x) {paste(annot_type, x, sep = "_")}, -cell_id) 
+}
 ```
@@
 
-Read SingleR results:
+Read `SingleR` results:
 
@@
     cell_id = glue::glue("{library_id}-{barcodes}"),
-    # recode NA -> "Unknown"
-    singler = ifelse(is.na(pruned.labels), "Unknown", pruned.labels)
+    # recode NA -> "Unknown" and NE -> "Neuroendocrine"
+    singler = dplyr::case_when(
+      is.na(pruned.labels) ~ "Unknown", 
+      pruned.labels == "NE" ~ "Neuroendocrine",
+      .default = pruned.labels)
   ) |>
@@
 
-Read scANVI results:
+Read `scANVI` results:
 
@@
   dplyr::filter(scanvi == posterior_celltype) |>
-  # recode to Unknown if below the threshold
-  dplyr::mutate(scanvi = ifelse(posterior < params$scanvi_pp_threshold, "Unknown", scanvi)) |>
+  # recode to Unknown if below the threshold, and NE -> Neuroendocrine
+  dplyr::mutate(scanvi = dplyr::case_when(
+    posterior < params$scanvi_pp_threshold ~ "Unknown", 
+    scanvi == "NE" ~ "Neuroendocrine", 
+    .default = scanvi)) |>
   dplyr::select(cell_id, scanvi)
@@
 nbatlas_markers_df <- readr::read_tsv(nbatlas_markers_file)
+
+validation_df <- readr::read_tsv(validation_url) |>
+  dplyr::rename(
+    consensus = consensus_annotation,
+    consensus_family_label = validation_group_annotation,
+    consensus_family_ontology = validation_group_ontology 
+  )
 ```
@@
 
-In this section, we'll derive the final annotations.
+Firt we prepare the data frame for annotation.
 
-```{r}
+```{r, warning = FALSE}
 ontology_slim_df <- ontology_df |>
@@
 
-prep_for_annotation <- function(
-    df, 
-    annot_type, 
-    ontology_df, 
-    label_map_df) {
-
- df |>
-    dplyr::select(all_of(c("cell_id", annot_type))) |>
-    dplyr::rename(label = annot_type) |>
-    # Join the family labels
-    dplyr::left_join(label_map_df, by = c("label" = "NBAtlas_label_immune_zoom")) |>
-    dplyr::rename(family_label = NBAtlas_label) |>
-    # Ensure "Unknown" has a family label instead of NA for future-proofing
-    dplyr::mutate(family_label = ifelse(is.na(family_label), "Unknown", family_label)) |>
-    # Join in the ontology labels to match the **celltype** ontologies (label column)
-    dplyr::left_join(ontology_slim_df, by = c("label" = "NBAtlas_label")) |>
-    dplyr::rename(label_ontology = CL_ontology_id) |>
-    # Join in the ontology labels to match the **family** ontologies (family_label column)
-    dplyr::left_join(ontology_slim_df, by = c("family_label" = "NBAtlas_label")) |>
-    dplyr::rename(family_ontology = CL_ontology_id) |>
-    # We don't need the labels now that we have ontologies
-    dplyr::select(-label, -family_label) |>
-    # rename columns to start with `annot_type`
-    dplyr::rename_with(\(x) {paste(annot_type, x, sep = "_")}, -cell_id)
-}
-
-singler_annotation_df <- prep_for_annotation(celltype_df, "singler", ontology_slim_df, label_map_df)
-scanvi_annotation_df <- prep_for_annotation(celltype_df, "scanvi", ontology_slim_df, label_map_df)
+singler_annotation_df <- prep_for_annotation(
+  celltype_df, 
+  "singler", 
+  ontology_slim_df, 
+  label_map_df
+)
+scanvi_annotation_df <- prep_for_annotation(
+  celltype_df, 
+  "scanvi", 
+  ontology_slim_df, 
+  label_map_df
+)
 
 annotation_df <- celltype_df |>
-  dplyr::select(cell_id, consensus_ontology) |>
+  dplyr::select(cell_id, consensus, consensus_ontology) |>
   dplyr::left_join(singler_annotation_df, by = "cell_id") |>
   dplyr::left_join(scanvi_annotation_df, by = "cell_id") |>
-  dplyr::select(cell_id, consensus_ontology, starts_with("singler"), starts_with("scanvi"))
+  dplyr::left_join(validation_df, by = c("consensus", "consensus_ontology")) |>
+  dplyr::select(cell_id, starts_with("consensus"), starts_with("singler"), starts_with("scanvi"))
 ```
 
+In the code below, we compare ontology ids to derive final annotations.
+However, based on those comparisons, we assign a final _label_, not final ontology id, and then add the final ontology ids back into those annotations.
+This is because we have several `NA` ontology ids, and we would not be able to unambiguously map their labels in if we assigned ontologies.
+
 ```{r}
-annotation_df <- annotation_df |> 
+final_annotation_df <- annotation_df |> 
   dplyr::mutate(
-    CL_ontology_id = dplyr::case_when(
-      # Use directly matching label
-      scanvi_label_ontology == singler_label_ontology ~ scanvi_label_ontology, 
-      # Check for family agreement between SingleR/scANVI
-      scanvi_family_ontology == singler_family_ontology ~ scanvi_family_ontology,
-      # Now use agreement with consensus to assign a label: first by label, and then by family
-      singler_label_ontology == consensus_ontology ~ singler_label_ontology,
-      scanvi_label_ontology == consensus_ontology ~ scanvi_label_ontology,
-      singler_family_ontology == consensus_ontology ~ singler_family_ontology,
-      scanvi_family_ontology == consensus_ontology ~ scanvi_family_ontology,
-      # Everything else is NA
-      .default = NA_character_
+    # For all comparisons, make sure we aren't comparing NA to NA and getting TRUE
+    final_label = dplyr::case_when(
+      ########### Check for exact match between SingleR/scANVI
+      !is.na(scanvi_label_ontology) & singler_label_ontology == scanvi_label_ontology ~ scanvi_label, 
+      ########### Check for family match between SingleR/scANVI
+      !is.na(scanvi_family_ontology) & singler_family_ontology == scanvi_family_ontology ~ scanvi_family,
+      ########## Now use agreement with consensus to assign a label: first by label, and then by family
+      !is.na(consensus_ontology) & singler_label_ontology  == consensus_ontology ~ singler_label,
+      !is.na(consensus_ontology) & scanvi_label_ontology   == consensus_ontology ~ scanvi_label,
+      !is.na(consensus_family_ontology) & singler_family_ontology == consensus_family_ontology ~ singler_family,
+      !is.na(consensus_family_ontology) & scanvi_family_ontology  == consensus_family_ontology ~ scanvi_family,
+      # Everything else is Unknown
+      .default = "Unknown"
     )
   ) |>
-  # keep only the final ontologies
-  dplyr::select(cell_id, CL_ontology_id)
-
-missing_cell_id <- annotation_df |>dplyr::filter(is.na(CL_ontology_id)) |>
-  dplyr::pull(cell_id)
-
-sjs_version <- readr::read_tsv(params$annotations_tsv)
-
-stephyes_allyno <- sjs_version |>
-  dplyr::filter(cell_id %in% missing_cell_id) |>
-  dplyr::filter(!is.na(CL_ontology_id)) |>
-  dplyr::pull(cell_id)
-
-celltype_df |>
-  dplyr::select(cell_id, singler, scanvi, consensus) |>
-  dplyr::filter(cell_id %in% stephyes_allyno)
-```
-
-Next, we'll get unified labels for any cells where at least two of three methods agree.
-
-```{r}
-annotation_df <- annotation_df |>
-  get_unified_label("fibroblast", "Fibroblast", "Fibroblast") |>
-  get_unified_label("macrophage", "Macrophage", "Macrophage") |>
-  get_unified_label("neutrophil", "Neutrophil", "Neutrophil") |>
-  get_unified_label("plasma cell", "Plasma", "Plasma") |>
-  get_unified_label(consensus_endo_celltypes, "Endothelial", "Endothelial") |>
-  get_unified_label("natural killer cell", nbatlas_nk_celltypes, "NK cell") |>
-  get_unified_label("monocyte", nbatlas_monocyte_celltypes, "Monocyte") |>
-  get_unified_label("dendritic cell", nbatlas_cdc_celltypes, "cDC") |>
-  get_unified_label(consensus_t_celltypes, nbatlas_t_celltypes, "T cell") |>
-  # This must be last since it is the broadest category
-  get_unified_label(consensus_myeloid_celltypes, nbatlas_myeloid_celltypes, "Myeloid")
+  # Now, we can bring map the final ontologies into the data frame
+  dplyr::inner_join(
+    ontology_df |> dplyr::rename(final_label = NBAtlas_label, final_id = CL_ontology_id), 
+    by = "final_label"
+  ) |>
+  # add tumor column for visualization
+  dplyr::mutate(cell_class = dplyr::case_when(
+    final_label == "Neuroendocrine" ~ "tumor", 
+    final_label == "Unknown" ~ "unknown", 
+    .default = "normal"
+  ))
 ```
@@
 ```{r}
-sum(!is.na(annotation_df$CL_ontology_id)) / nrow(annotation_df)
+sum(!is.na(final_annotation_df$final_id)) / nrow(final_annotation_df)
 ```
@@
 ```{r, fig.width = 8, fig.height = 4}
-frac_labeled_df <- annotation_df |>
+frac_labeled_df <- final_annotation_df |>
   tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE) |>
@@
   dplyr::summarize(
-    frac_labeled = sum(!is.na(cell_type_assignment))/dplyr::n(), 
+    frac_labeled = sum(!is.na(final_id))/dplyr::n(), 
     library_size = dplyr::n()
@@
 ```{r}
-# Finalize the annotations
-submission_df <- annotation_df |>
-  # add ontology ids
-  dplyr::left_join(
-    nbatlas_ontology_ids, 
-    by = c("cell_type_assignment" = "NBAtlas_label")
-  ) |>
-  # recode labels
-  dplyr::mutate(
-    cell_type_assignment = dplyr::case_when(
-      is.na(cell_type_assignment) ~ "Unknown", 
-      cell_type_assignment == "NE" ~ "Neuroendocrine/tumor",
-      .default = cell_type_assignment
-   ),
-    tumor_cell_classification = dplyr::case_when(
-      cell_type_assignment == "Unknown" ~ "Unknown", 
-      cell_type_assignment == "Neuroendocrine/tumor" ~ "tumor", 
-      .default = "normal"
-    )
-  ) |>
-  tidyr::separate(cell_id, into = c("library_id", "cell_barcode"), remove = FALSE) |>
-  dplyr::select(
-    cell_id, library_id, cell_barcode, cell_type_assignment, tumor_cell_classification, CL_ontology_id
-  )
-# update palette label as well
-names(celltype_pal)[names(celltype_pal) == "Neuroendocrine"] <- "Neuroendocrine/tumor"
-```
+label_order <- label_map_df |>
+  dplyr::add_count(NBAtlas_family) |>
+  dplyr::arrange(desc(n), NBAtlas_family) |>
+  dplyr::filter(NBAtlas_label %in% final_annotation_df$final_label) |>
+  dplyr::pull(NBAtlas_label)
+label_order <- c(label_order, "Unknown")
 
-```{r}
 # prepare data frame for plotting
@@
   dplyr::select(cell_id, sample_id, library_id, is_xenograft, UMAP1, UMAP2) |>
-  dplyr::left_join(submission_df, by = "cell_id") |>
-  # group "same family" cell types in the factor for cleaner legend
-  # note this will give warnings if any of those cells aren't in the final annotation, but that's fine
-  dplyr::mutate(
-    cell_type_assignment = factor(cell_type_assignment), 
-    cell_type_assignment = forcats::fct_relevel(
-      cell_type_assignment, 
-      "T cell", nbatlas_t_celltypes, 
-      "NK cell", nbatlas_nk_celltypes,
-      "cDC", nbatlas_cdc_celltypes, 
-      "Monocyte", nbatlas_monocyte_celltypes
-    )
-  )
+  dplyr::left_join(final_annotation_df, by = "cell_id") |>
+  dplyr::mutate(final_label = factor(final_label, levels = label_order))
 ```
@@
 
-Note that in the UMAP below, cell types in the same "family" each share a color:
+Note that in the UMAP below, cell types in these same "family" each share a color:
 
-* T cells: `T cell`, `Treg`, `CD4+ T cell`, `CD8+ T cell`
-* Natural killer cells: `NK cell`, `Circulating NK cell`, `Resident NK cell`, `TOX2+/KIT+ NK cell`
-* Conventional dendritic cells: `cDC`, `cDC1`, `cDC2/DC3`, `Migratory cDC`
-* Monocytes: `Monocyte`, `Classical monocyte`, `Patrolling monocyte`
+* T cells
+* Natural killer cells
+* Conventional dendritic cells
+* Monocytes
 
@@
 ggplot(umap_df) +
-  aes(x = UMAP1, y = UMAP2, color = cell_type_assignment) +
+  aes(x = UMAP1, y = UMAP2, color = final_label) +
   geom_point(alpha = 0.2, size = 0.3) +
   scale_color_manual(values = celltype_pal) +
-guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
+  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
+  umap_theme +
+  theme(legend.position = "bottom")
 ```
@@
 This UMAP is the same as the previous one, except each facet highlights one specific cell type.
+In this plot, only cell types with at least 10 cells are shown.
 
 ```{r fig.width = 16, fig.height = 16}
-umap_theme <- list(
-  theme(
-    axis.text = element_blank(), 
-    axis.ticks = element_blank(), 
-    legend.position = "bottom", 
-    aspect.ratio = 1
-  )
-)
-
 umap_df |>
+  dplyr::add_count(final_label) |>
+  dplyr::filter(n > 10) |>
   faceted_umap(
-    annotation_column = cell_type_assignment,
-    celltype_colors = celltype_pal
+    annotation_column = final_label,
+    celltype_colors = celltype_pal, 
+    facet_wrap_cols = 6
   ) + 
-  umap_theme
+    umap_theme +
+    theme(legend.position = "none")
 ```
@@
 ggplot(umap_df) +
-  aes(x = UMAP1, y = UMAP2, color = tumor_cell_classification) +
+  aes(x = UMAP1, y = UMAP2, color = cell_class) +
   geom_point(alpha = 0.2, size = 0.3) +
@@
 Marker genes that are present in three or more cell types are not included.
-In addition, the upper limit for the color scale has been censored to a maximum of 6.
+In addition, the upper limit for the color scale has been censored to a maximum of 6 to ensure visibility of expression levels across all cell types.
 
 ```{r}
-# Temporary to develop code for next section
-annotation_df <- readr::read_tsv(params$annotations_tsv) |>
-  dplyr::filter(cell_type_assignment != "Unknown") |>
-  # recode cell types to match what NBAtlas uses for their marker genes
+input_dotplot_df <- final_annotation_df |>
+  dplyr::filter(final_label != "Unknown") |>
+  # recode cell types to match what is present in NBAtlas marker genes
   dplyr::mutate(
-    dotplot_celltype = dplyr::case_when(
-      stringr::str_detect(cell_type_assignment, "monocyte") ~ "Monocyte", 
-      stringr::str_detect(cell_type_assignment, "cDC") ~ "cDC", 
-      stringr::str_detect(cell_type_assignment, "Neuroendocrine") ~ "Neuroendocrine", 
-      .default = cell_type_assignment
-    )
-  )
+    label_recoded = dplyr::case_when(
+      stringr::str_detect(final_label, "monocyte") ~ "Monocyte", 
+      stringr::str_detect(final_label, "cDC") ~ "cDC",
+      .default = final_label
+  )) |>
+  dplyr::select(cell_id, label_recoded)
 
@@
 # get total number of cells per final annotation group and set up y_label
-total_cells_df <- annotation_df |>
-  dplyr::count(dotplot_celltype, name = "total_cells") |>
-  # remove infrequent cell types
-  dplyr::filter(total_cells > 10) |>
-  dplyr::mutate(y_label = glue::glue("{dotplot_celltype} ({total_cells})")) |>
-  # order by count, but put Unknown at the end
-  dplyr::arrange(desc(total_cells)) |> 
+total_cells_df <- input_dotplot_df |>
+  dplyr::count(label_recoded, name = "total_cells") |>
+  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})")) |>
+  dplyr::inner_join(label_map_df, by = c("label_recoded" = "NBAtlas_label")) |>
+  dplyr::group_by(NBAtlas_family) |>
+  dplyr::mutate(family_total_cells = sum(total_cells)) |>
+  # order by family count, but put Unknown at the end
+  dplyr::arrange(desc(family_total_cells)) |>
   # amazing new dplyr trick; only this row is TRUE, so it gets moved to the last row
-  dplyr::arrange(dotplot_celltype == "Unknown") |>
-  # the dotplot function expects this column name
-  dplyr::rename(label_recoded = dotplot_celltype)
+  dplyr::arrange(label_recoded == "Unknown")
 
@@
   markers_df = top_nbatlas_markers_df,
-  # rename this after we've done all the wrangling
-  celltype_df = annotation_df |> dplyr::rename(label_recoded = dotplot_celltype),
+  celltype_df = input_dotplot_df,
   total_cells_df = total_cells_df,
@@
   celltype_palette = celltype_pal, 
-  min_cells = 0
+  min_cells = 10 # only show cell types with at least 10 cells
 ) +
@@
 
+Export the final table with submission-ready column names.
+
 ```{r}
+final_annotation_df |>
+  dplyr::select(
+    cell_id, 
+    cell_type_assignment = final_label, 
+    tumor_cell_classification = cell_class,
+    CL_ontology_id = final_id, 
+    CL_annotation
+  ) |>
   readr::write_tsv(
-  submission_df,
     params$annotations_tsv

### 2025-08-19 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
add testing flag for CI since dotplot fails, and update notebook text to match README

@@
   scanvi_pp_threshold: 0.75    
-  annotations_tsv: "../results/SCPCP000004-annotations.tsv"
+  annotations_tsv: "results/SCPCP000004-annotations.tsv"
+  testing: FALSE
 ---
@@
 We use the following approach to determine the final annotation.
-
-First, we prioritize inferences with `SingleR` and `scANVI/scArches`.
-If these methods returned the exact same label, we assign this label.
-
-Next, we compare labels across all three methods: if at least _two out of three_ methods agree, we assign the agreeing label.
-Note that this requires mapping consensus cell types to NBAtlas labels, which is only possible for these cell types: endothelial cells, fibroblasts, and immune cell types except for the NBAtlas labels `RBCs` and `pDC` which do not have corresponding consensus labels.
-
-In this approach, there are two ways that cell type annotations can agree, and we treat them as follows:
-
-* Cell type annotations can have directly corresponding labels. For example:
-  * If the consensus cell type is `endothelial cell` and at least one of the NBAtlas-derived annotations is `Endothelial`, we assign the label `Endothelial`.
-* Cell type annotations can differ but be in the same "family" of cell types. For example:
-  * If the consensus cell type is `mature T cell` and at least one of the NBAtlas-derived annotations is `CD4+ T cell`, we assign the broader label `T cell`.
-  * If one NBAtlas-derived annotation is `Patrolling monocyte` and the other is `Classical monocyte`, we assign the broader label `Monocyte`.
-
-We use CL ontology ids to determine agreement throughout. 
+Throughout, we use CL ontology ids to perform comparisons. 
 For more information about how these ontology ids were determined for NBAtlas labels, see `../references/README.md`.
 
-For the final exported TSV file, we further designate any cells labeled as `Neuroendocrine` as `tumor`, following how this label is interpreted in NBAtlas.
+1. If `SingleR` and `scANVI/scArches` labels exactly agree, we assign that label
+2. If broad label families (e.g., `T cell` would be the broad family for `CD4+ T cell`) for `SingleR` and `scANVI/scArches` labels agree, we assign the broad family label.
+  * See [`references/nbatlas-label-map.tsv`](references/nbatlas-label-map.tsv) and [`references/README.md`](references/README.md) for how labels were mapped to families.
+3. If `SingleR` and `scANVI/scArches` labels disagree, we then compare to the consensus cell type; if at least one inference agrees with the consensus cell type, we assign the inferred label.
+  * We perform this first at the label level and then at the broad family level.
+4. Finally, following the conclusions of Bonine et al. (2024), we designate any cell labeled as `Neuroendocrine` as a tumor cell.
 
@@
 
-```{r, fig.width = 32, fig.height = 16}
+```{r, fig.width = 32, fig.height = 16, eval = !params$testing}
 generate_dotplot(

### 2025-08-19 analyses/cell-type-neuroblastoma-04/run-analysis.sh

@@ -260,5 +260,6 @@ python ${script_dir}/03c_run-scanvi-label-transfer.py \
 Rscript -e "rmarkdown::render('final-annotation.Rmd',
   params = list(
       scanvi_pp_threshold = 0.75,
-      annotations_tsv = '${final_annotations_tsv}'
+      annotations_tsv = '${final_annotations_tsv}',
+      testing = ${testing}
 ))"

### 2025-08-19 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd

@@
   ) |>
-  # Now, we can bring map the final ontologies into the data frame
+  # Now, we can bring the final ontologies into the data frame
   dplyr::inner_join(
@@
 
-```{r fig.width = 8, fig.height = 8}
+```{r fig.width = 7, fig.height = 7}
 ggplot(umap_df) +
@@
 Marker genes that are present in three or more cell types are not included.
+Only cell types with at least 10 detected cells are shown, and cells without an assigned label are not shown.
 In addition, the upper limit for the color scale has been censored to a maximum of 6 to ensure visibility of expression levels across all cell types.
@@
   dplyr::filter(final_label != "Unknown") |>
+  # keep only cells that appear > 10 times
+  dplyr::add_count(final_label) |>
+  dplyr::filter(n > 10) |>
   # recode cell types to match what is present in NBAtlas marker genes
@@
 
-```{r, fig.width = 32, fig.height = 16, eval = !params$testing}
+```{r, fig.width = 24, fig.height = 16}
+# placing this in eval in chunk settings makes the chunk
+# ignore width and height, so we use an if
+if (!(params$testing)) {
   generate_dotplot(
@@
     celltype_palette = celltype_pal,
-  min_cells = 10 # only show cell types with at least 10 cells
+    text_size = 22,
+    bar_text_size = 20
   ) +
     theme(legend.position = "bottom")
+}
+
 ```

### 2025-08-20 analyses/cell-type-consensus/exploratory-notebooks/06-update-consensus-scimilarity.Rmd

@@
 
-Previously when we looked at the LCA for pairs with moret than 1 match, we had seen only very broad cell types. 
+Previously, when we looked at the LCA for pairs with more than 1 match, we had seen only very broad cell types. 
 That doesn't seem to be the case here as we see things like "memory T cell".
@@
 Perhaps instead of just removing the pairs with multiple LCA, we first filter out the terms that don't meet our criteria. 
-Then if multiple LCA remain we find the LCA of all the remaining LCA's and if that new LCA meets our criteria we use that instead.
+Then, if multiple LCAs remain, we find the LCA of all the remaining LCAs, and if that new LCA meets our criteria, we use that instead.
 
@@
 ```{r}
-# firt get a dataframe of the matches we will keep that had multiple LCA and their final annotations
+# first get a data frame of the matches we will keep that had multiple LCA and their final annotations
 multi_lca_keep_df <- new_lca_df |> 

### 2025-08-21 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
add notebook to look at counts of reference/query

---
title: "Diagnoses and cell types"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output: html_notebook
params:
  n_reference_threshold: 100
---

The goal of this notebook is to look at cell type distributions across samples to determine how many samples could be run through `inferCNV`.
Specifically, we are interested in seeing the relative numbers of cells intended as `inferCNV` reference vs. query.
Cell types were determined to be reference or query based on the sample's diagnosis, as mapped out in `../references/broad-diagnosis-map.tsv`.
We consider a threshold of at least `r params$n_reference_threshold` reference cells needed to run `inferCNV`.

## Setup

```{r, warning = FALSE}
# settings
options(
  dplyr.summarise.inform = FALSE, 
  readr.show_col_types = FALSE
)
```

### Paths

Define directories:

```{r}
module_dir <- rprojroot::find_root(rprojroot::is_renv_project)
repository_base <- file.path(module_dir, "..", "..")

ref_dir <- file.path(module_dir, "references")
data_dir <- file.path(repository_base, "data", "current")
consensus_dir <- file.path(repository_base, "data", "current", "results", "cell-type-consensus")
consensus_module_dir <- file.path(repository_base, "analyses", "cell-type-consensus")
```

Define paths to input files to read in:

```{r}
# map broad to submitted diagnoses
diagnosis_map_file <- file.path(ref_dir, "broad-diagnosis-map.tsv")

# map broad diagnoses to cell type groups
diagnosis_celltype_map_file <- file.path(ref_dir, "diagnosis-celltype-groups.tsv")

# map cell type groups to constituent cell types
validation_file <- file.path(
  consensus_module_dir, 
  "references", 
  "consensus-validation-groups.tsv"
)

# metadata files, for getting sample diagnoses
metadata_files <- list.files(
  path = data_dir,
  pattern = "single_cell_metadata\\.tsv$",
  recursive = TRUE,
  full.names = TRUE
)

# TSVs of consensus cell types
consensus_files <- list.files(
  path = consensus_dir,
  pattern = "_processed_consensus-cell-types\\.tsv\\.gz$",
  recursive = TRUE,
  full.names = TRUE
) |>
  # set names to filter on later so we don't read files we don't need
  purrr::set_names(\(x) {
    stringr::str_split_i(basename(x), "_", 1)
  })
```

### Read input data

```{r}
diagnosis_map <- readr::read_tsv(diagnosis_map_file)
diagnosis_celltype_map <- readr::read_tsv(diagnosis_celltype_map_file)
validation_group_df <- readr::read_tsv(validation_url)
```

```{r}
sample_diagnosis_df <- metadata_files |>
  purrr::map(
    \(file) {
      readr::read_tsv(file) |>
        # don't consider multiplexed or cell line
        dplyr::filter(!is_multiplexed, !is_cell_line) |>
        dplyr::select(contains("scpca"), diagnosis)
  }) |>
  dplyr::bind_rows() |>
  # use inner_join to remove normals
  dplyr::inner_join(diagnosis_map, by = c("diagnosis" = "submitted_diagnosis")) |>
  dplyr::select(-ontology_id, -human_readable_value)
```

```{r}
consensus_df <- consensus_files |>
  # only keep samples in sample_diagnosis_df
  purrr::keep_at(sample_diagnosis_df$scpca_library_id) |>
  purrr::imap(
    \(file, id) {

      # bump guess_max to kill problems() warnings
      readr::read_tsv(file, guess_max = 15000) |>
       dplyr::select(
         scpca_project_id = project_id,
         scpca_sample_id = sample_id,
         scpca_library_id = library_id,
         barcodes,
         consensus_annotation,
         consensus_ontology
       )
  }) |>
  dplyr::bind_rows()
```

### Prepare data for analysis

Here, we map over each diagnosis group to count the number of reference and query cells for each sample in that group.
We then combine into a single data frame of reference/query cell counts per sample.

```{r}
sample_counts_df <- unique(sample_diagnosis_df$diagnosis_group) |>
  purrr::map(
    \(broad_diagnosis) {

      celltype_groups <- diagnosis_celltype_map |>
        dplyr::filter(diagnosis_group == broad_diagnosis) |>
        tidyr::separate_rows(celltype_groups, sep = ", ") |>
        dplyr::pull(celltype_groups)

      reference_celltypes <- validation_group_df |>
        dplyr::filter(validation_group_annotation %in% celltype_groups) |>
        dplyr::pull(consensus_annotation)

      library_ids <- sample_diagnosis_df |>
        dplyr::filter(diagnosis_group == broad_diagnosis) |>
        dplyr::pull(scpca_library_id)

      count_df <- consensus_df |>
        dplyr::filter(scpca_library_id %in% library_ids) |>
        dplyr::mutate(
          cell_class = ifelse(
            consensus_annotation %in% reference_celltypes, 
            "n_reference", 
            "n_query"
        )) |>
        dplyr::count(scpca_library_id, cell_class) |>
        tidyr::pivot_wider(
          names_from = cell_class, 
          values_from = n, 
          values_fill = 0
        ) |>
        dplyr::inner_join(sample_diagnosis_df, by = "scpca_library_id") |>
        dplyr::select(
          scpca_project_id, 
          scpca_library_id, 
          diagnosis_group, 
          diagnosis, 
          # there may not be any n_reference, so this is more flexible
          contains("n_")
        )

      # Create n_reference if it doesn't exist
      if (!("n_reference" %in% names(count_df))) {
        count_df$n_reference <- 0
      }

      return(count_df)
  }) |>
  dplyr::bind_rows()
```

## Analysis

Using a threshold of needing at least `r params$n_reference_threshold` reference cells, how many samples could be run through `inferCNV`?

```{r}
usable_df <- sample_counts_df |>
  dplyr::group_by(diagnosis_group) |>
  dplyr::summarize(
    n_usable = sum(n_reference >= params$n_reference_threshold),
    total_group_samples = dplyr::n(), 
    frac_usable = round(n_usable / total_group_samples, 3) 
  ) |>
  dplyr::arrange(desc(total_group_samples))

usable_df
```
All of our broad diagnosis groups are represented, which is nice to see.

In total, we can run this fraction of ScPCA samples:

```{r}
sum(sample_counts_df$n_reference >= params$n_reference_threshold) / nrow(sample_counts_df)
```

Of the samples with sufficient reference cells, do any have _fewer_ query than reference cells?

```{r}
sample_counts_df |>
  # consider usable only
  dplyr::filter(n_reference >= params$n_reference_threshold) |>
  dplyr::filter(n_query < params$n_reference_threshold)
```

There is one project about 7x as many reference as query cells; let's look more closely:

```{r}
consensus_df |>
  dplyr::filter(scpca_library_id == "SCPCL000021") |>
  dplyr::count(consensus_annotation) |>
  dplyr::arrange(n)
```

This sample indeed is dominated by immune cell types.
This sample may just have very low tumor purity, but I think it's still fine to run through `inferCNV`.

## Conclusions

Currently, _before_ incorporating `SCimilarity` into our consensus cell types, we are able to run about 44% of samples through `inferCNV`. 
All broad diagnosis groups are represented.

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-08-21 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
text updates

@@
 diagnosis_celltype_map <- readr::read_tsv(diagnosis_celltype_map_file)
-validation_group_df <- readr::read_tsv(validation_url)
+validation_group_df <- readr::read_tsv(validation_file)
 ```
@@
 
-In total, we can run this fraction of ScPCA samples:
+Which diagnosis groups have _no samples_ with sufficient reference cells?
+
+```{r}
+usable_df |>
+  dplyr::filter(frac_usable == 0)
+```
+It's not too surprising to see `Mixed phenotype acute leukemia` here since most immune cell types can't be included in the reference, and the other groups comprise very small numbers of samples.
+
+In total, this fraction of ScPCA samples have sufficient reference cells to run through `inferCNV`:
 
@@
 
-This sample indeed is dominated by immune cell types.
-This sample may just have very low tumor purity, but I think it's still fine to run through `inferCNV`.
+This sample indeed is dominated by immune cell types with very low tumor purity, but it still seems to have enough query cells to not be a problem and should be ok to run.
 
@@
 Currently, _before_ incorporating `SCimilarity` into our consensus cell types, we are able to run about 44% of samples through `inferCNV`. 
-All broad diagnosis groups are represented.
+Most broad diagnosis groups are represented, except `Mixed phenotype acute leukemia` and a handful of other solid tumors.
 

### 2025-08-21 analyses/cell-type-consensus/exploratory-notebooks/06-update-consensus-scimilarity.Rmd
figure out how to deal with multiple lcas

@@
 
+### Handling multiple LCA matches
+
+First, let's see what cell types have multiple matches.
+
 ```{r}
@@
 # create a df with just lca that meet the specified criteria so far 
-filtered_celltype_df <- lca_df |>
+filtered_lca_df <- lca_df |>
   dplyr::filter(total_descendants <= 170  & !cl_annotation %in% celltypes_to_exclude)
@@
 # print out all matches when total lca > 1
-filtered_celltype_df |> 
+filtered_lca_df |> 
   dplyr::filter(total_lca > 1) |> 
@@
 
+#### Re-calculating the LCA
+
 Perhaps instead of just removing the pairs with multiple LCA, we first filter out the terms that don't meet our criteria. 
@@
 # get the new LCA number after removing the non-specific terms 
-filtered_celltype_df <- filtered_celltype_df |> 
+filtered_lca_df <- filtered_lca_df |> 
   dplyr::group_by(id1, id2) |> 
   dplyr::mutate(new_lca_pair = length(lca))
+```
 
+```{r, warning=FALSE, message=FALSE}
 # now pull out any pairs where only one of the LCA remains 
@@
 # are there any that are remaining that still have multiple LCA? 
-new_lca_df <- filtered_celltype_df |>
+new_lca_df <- filtered_lca_df |>
   dplyr::filter(new_lca_pair > 1) |>
@@
 
+#### Choosing a single LCA
+
+Alternatively, we could identify some systematic approach to choose one of the possible LCAs. 
+Let's first look at some of the pairs and see if using the total descendants would be appropriate:
+
+```{r}
+filtered_lca_df |> 
+  dplyr::filter(new_lca_pair > 1) |> 
+  dplyr::ungroup() |> 
+  dplyr::select(id1_annotation, id2_annotation, cl_annotation, total_descendants) |> 
+  DT::datatable(rownames = FALSE)
+```
+
+In looking at these pairs, for the most part we see one term that is obviously "more specific" than the other terms.
+
+- A lot of matches give both "memory T cell" and "mature alpha-beta T cell". 
+In this case, "memory T cell" is more specific and has fewer descendants. 
+- We see "peptide hormone secreting cell" and another more specific term like "pancreatic endocrine cell" or "enteroendocrine cell" with fewer descendants. 
+- "glandular epithelial cell" and "epithelial cell of pancreas" also show up, where "epithelial cell of pancreas" has fewer descendants.
+
+The exception to this is "mononuclear phagocyte" and "myeloid leukocyte" which to me are both equally vague. 
+"myeloid leukocyte" does have a fewer number of descendants though so if we use that rule, I do think it would be reasonable to choose "myeloid leukocyte".
+
+Generally, I think we could choose the term with the fewest descendants as the appropriate LCA, which would be consistent with using the number of descendants in the first place to set our original cutoffs.
+This approach seems more straight forward than recalculating the LCA as we did in the previous section.
+
+```{r}
+filtered_lca_df <- filtered_lca_df |> 
+  dplyr::group_by(id1, id2) |> 
+  # get the cl_annotation with the minimum value in total_descendants column 
+  # discard all other rows that aren't the minimum value 
+  dplyr::slice_min(total_descendants, with_ties = FALSE) |> 
+  dplyr::ungroup()
+```
+
 ### Discarded cell types
@@
 - LCA term that is "neuron" or "epithelial cell"
-- A secondary LCA term (LCA computed all LCAs returned from the initial pairing) that is a single LCA value with <= 170 total descendants or one of "epithelial cell", "kidney tubule cell", or "hematopoietic precursor cell"
+- The LCA term with the fewest descendants when multiple LCA terms are found that meet the above criteria
 
 ```{r}
-# now filter to remove the multi LCA matches and add back in the updated assignments for them 
+# create a table with all pairs and their LCA
 filtered_lca_df <- lca_df |> 
@@
   dplyr::group_by(id1, id2) |> 
-  # after filtering, see how many LCAs are left
-  dplyr::mutate(new_lca_pair = length(lca)) |> 
-   # everything with more than 1 lca gets removed with the exception of HSCs
-  dplyr::filter(new_lca_pair <=1 | cl_annotation == "hematopoietic precursor cell") |> 
-  dplyr::select(id1, id2, id1_annotation, id2_annotation, lca, cl_annotation, total_descendants) |> 
-  # add in pairs with multi LCA and their final match 
-  dplyr::bind_rows(multi_lca_keep_df) |> 
-  unique()
+  # get the cl_annotation with the minimum value in total_descendants column 
+  # discard all other rows that aren't the minimum value 
+  dplyr::slice_min(total_descendants, with_ties = FALSE) |> 
+  dplyr::ungroup()
 ```
@@
 dim(filtered_lca_df[dup, ])[1]
+```
+
+Now what cell types are we retaining:
+
+```{r}
+filtered_lca_df |> 
+  dplyr::select(cl_annotation, total_descendants) |> 
+  dplyr::arrange(desc(total_descendants)) |> 
+  dplyr::distinct() |> 
+  DT::datatable(rownames = FALSE)
 ```
@@
   dplyr::summarize(
-    total_matches = length(unique(lca)),
-    consensus_celltype_ontology = dplyr::case_when(
-      total_matches == 1 ~ lca,
-      total_matches > 1 ~ paste0(unique(lca), collapse = ",")
+    total_matches = length(unique(cl_annotation)),
+    consensus_celltype_annotation = dplyr::case_when(
+      total_matches == 1 ~ cl_annotation,
+      total_matches > 1 ~ paste0(unique(cl_annotation), collapse = ";")
+    ),
+    # aggregate total_descendants for the unique annotations only
+    total_descendants = paste0(
+      unique(total_descendants[match(unique(cl_annotation), cl_annotation)]),
+      collapse = ";"
     )
-  )
+
+  ) |> 
+  dplyr::distinct()
 
@@
 
-TODO: Figure out how to assign the LCA for ~ 16000 combinations that result in multiple LCAs
+```{r}
+multi_lca_df <- consensus_ref_df |> 
+  dplyr::ungroup() |> 
+  dplyr::filter(total_matches > 1) |> 
+  dplyr::select(consensus_celltype_annotation, total_descendants) |> 
+  dplyr::distinct()
+
+multi_lca_df |> 
+  DT::datatable(rownames = FALSE)
+
+```
+
+Generally speaking it looks like there are terms that are of the same lineage which makes sense. 
+I think we could just stick with our previous guidelines of choosing the label with the fewest descendants or is the most specific term.
+
+The only problem we might have is when terms have an equivalent number of descendants. 
+Let's see how many cases that applies too.
+
+```{r}
+# first get a dataframe of the minimum descendants, keeping all ties 
+final_df <- combined_ref_df |> 
+  # first get rid of any cases where more than one pair gives the same lca term 
+  dplyr::select(panglao_ontology, blueprint_ontology, scimilarity_ontology, lca, cl_annotation, total_descendants) |> 
+  dplyr::distinct() |> 
+  # now select the lca term with the minimum total descendants 
+  dplyr::group_by(panglao_ontology, blueprint_ontology, scimilarity_ontology) |> 
+  dplyr::slice_min(total_descendants)
+```
+
+```{r}
+# are there any cases where we have ties? 
+dup <- final_df |> 
+  dplyr::select(panglao_ontology, blueprint_ontology, scimilarity_ontology) |> 
+  duplicated()
+
+final_df[dup, ] |> 
+  DT::datatable(rownames = FALSE)
+```
+
+It looks like there are occassions where the lca term is both `hematopoietic precursor cell` and `stem cell`. 
+These terms both have the same number of total descendants. 
+Manually looking at the Cell Ontology, they also both only have 2 ancestors. 
+I think I would favor choosing `hematopoietic precursor cell` over `stem cell` since that specifically refers to a cell in the blood lineage. 
+Also glancing at the terms from blueprint here, they all appear to be from the HSC lineage. 
+Let's double check that.
+
+```{r}
+final_df[dup, ] |> 
+  dplyr::pull(blueprint_ontology) |> 
+  as.character() |> 
+  unique()
+```
+
+These terms correspond to `megakaryocyte-erythroid progenitor cell`,  `hematopoietic stem cell`, `hematopoietic multipotent progenitor cell`, `common lymphoid progenitor`, `granulocyte monoctye progenitor cell`, and `common myeloid progenitor`.
+
+For this case, we should choose `hematopoietic precursor cell` over `stem cell`.
+
+## Conclusions
+
+Combining all the information from this notebook, we should use the following set of rules to create the consensus cell type reference:
+
+Identify the LCA for all possible pairs of annotations and keep the LCA if the following criteria are met:
+
+- LCA has <= 170 total descendants or belongs to `neuron`, `epithelial cell`, `columnar/cuboidal epithelial cell`, or `endo-epithelial cell`
+- LCA term that is not one of: "bone cell", "lining cell", "blood cell", "progenitor cell", "supporting cell", "biogenic amine secreting cell", "protein secreting cell", or "extracellular matrix secreting cell"
+- If multiple LCAs are found that meet the above criteria, the LCA term with the fewest descendants is retained
+
+For all possible combinations of the three annotations, if at least one pair has an LCA that meet the above criteria, a consensus cell type is assigned using the LCA with the fewest descendants. 
+The only exception to this rule is if the annotations return both `hematopoietic precursor cell` and `stem cell` as possible LCAs.
+In this case, `hematopoietic precursor cell` is assigned as the consensus cell type.
 

### 2025-08-21 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
add barplots of cell group proportions per library

@@
 ```{r, warning = FALSE}
+library(ggplot2)
+theme_set(theme_bw())
+
 # settings
@@
 
+# validation palette
+palette_url <- "https://raw.githubusercontent.com/AlexsLemonade/scpca-paper-figures/refs/heads/main/palettes/validation-group-palette.tsv"
+
 # metadata files, for getting sample diagnoses
@@
 validation_group_df <- readr::read_tsv(validation_file)
+palette_df <- readr::read_tsv(palette_url)
+celltype_pal <- palette_df$color |>
+  purrr::set_names(palette_df$validation_group_annotation)
+celltype_pal["query"] <- celltype_pal["unknown"]
 ```
@@
 
-Here, we map over each diagnosis group to count the number of reference and query cells for each sample in that group.
-We then combine into a single data frame of reference/query cell counts per sample.
+Here, we map over each diagnosis group to label each cell as `query` or `reference`.
+Then, we count the number of each category.
 
 ```{r}
-sample_counts_df <- unique(sample_diagnosis_df$diagnosis_group) |>
+query_reference_df <- unique(sample_diagnosis_df$diagnosis_group) |>
   purrr::map(
@@
 
-      count_df <- consensus_df |>
+      consensus_df |>
         dplyr::filter(scpca_library_id %in% library_ids) |>
@@
             consensus_annotation %in% reference_celltypes, 
-            "n_reference", 
-            "n_query"
-        )) |>
+            "reference", 
+            "query"
+        ))
+  }) |>
+  dplyr::bind_rows()
+```
+
+```{r}
+sample_counts_df <- query_reference_df |>
   dplyr::count(scpca_library_id, cell_class) |>
@@
     diagnosis, 
-          # there may not be any n_reference, so this is more flexible
-          contains("n_")
+    query, 
+    reference
   )
-
-      # Create n_reference if it doesn't exist
-      if (!("n_reference" %in% names(count_df))) {
-        count_df$n_reference <- 0
-      }
-
-      return(count_df)
-  }) |>
-  dplyr::bind_rows()
 ```
@@
   dplyr::summarize(
-    n_usable = sum(n_reference >= params$n_reference_threshold),
+    n_usable = sum(reference >= params$n_reference_threshold),
     total_group_samples = dplyr::n(), 
@@
   ) |>
-  dplyr::arrange(desc(total_group_samples))
+  dplyr::arrange(desc(total_group_samples)) |>
+  dplyr::select(diagnosis_group, total_group_samples, n_usable, frac_usable)
 
@@
 ```{r}
-sum(sample_counts_df$n_reference >= params$n_reference_threshold) / nrow(sample_counts_df)
+sum(sample_counts_df$reference >= params$n_reference_threshold) / nrow(sample_counts_df)
 ```
@@
   # consider usable only
-  dplyr::filter(n_reference >= params$n_reference_threshold) |>
-  dplyr::filter(n_query < params$n_reference_threshold)
+  dplyr::filter(reference >= params$n_reference_threshold) |>
+  dplyr::filter(query < params$n_reference_threshold)
 ```
@@
 
+## Plots
+
+In this section, we plot the relative contributions of each cell type group in the samples we can run. 
+We show the relative proportions of consensus cell type groups in the _reference_, and we show all other query cells grouped together in gray.
+The x-axis is library ids, and the y-axis is the proportion of each cell type group.
+
+```{r}
+usable_libraries <- sample_counts_df |>
+  dplyr::filter(reference >= params$n_reference_threshold) |>
+  dplyr::pull(scpca_library_id)
+
+plot_df <- query_reference_df |>
+  dplyr::filter(scpca_library_id %in% usable_libraries) |>
+  dplyr::left_join(validation_group_df, by = c("consensus_ontology", "consensus_annotation")) |>
+  dplyr::select(
+    scpca_project_id, 
+    scpca_library_id, 
+    barcodes, 
+    cell_class, 
+    validation_group_annotation
+  ) |>
+  dplyr::mutate(
+    label = ifelse(
+      cell_class == "query", 
+      "query",
+      validation_group_annotation
+    ),
+    # move "query" first
+    label = forcats::fct_relevel(label, "query", after = 0)
+  )
+```
+
+```{r fig.width = 16, fig.height = 14}
+ggplot(plot_df) + 
+  aes(x = scpca_library_id, fill = label) + 
+  geom_bar(position = "fill") +
+  scale_fill_manual(values = celltype_pal) + 
+  facet_wrap(
+    vars(scpca_project_id), 
+    scales = "free_x", 
+    ncol = 4
+  ) + 
+  ylab("proportion") +
+  theme(
+    axis.text.x = element_blank(), 
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 14),
+    strip.text = element_text(size = 13),
+    legend.text = element_text(size = 12),
+    legend.position = "bottom"
+  )
+```
+
 ## Conclusions
@@
 Most broad diagnosis groups are represented, except `Mixed phenotype acute leukemia` and a handful of other solid tumors.
+All eligible projects except `SCPCP000011` (Retinoblastoma samples) are represented as well.
 

### 2025-08-21 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
Update analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
Co-authored-by: Ally Hawkins <54039191+allyhawkins@users.noreply.github.com>
keep multiplexed, and save intermediate dataframe for math

@@
 ```{r}
-sample_diagnosis_df <- metadata_files |>
+all_sample_diagnosis_df <- metadata_files |>
   purrr::map(
@@
       readr::read_tsv(file) |>
-        # don't consider multiplexed or cell line
-        dplyr::filter(!is_multiplexed, !is_cell_line) |>
-        dplyr::select(contains("scpca"), diagnosis)
+        dplyr::select(contains("scpca"), diagnosis, is_cell_line, is_multiplexed)
   }) |>
@@
   dplyr::select(-ontology_id, -human_readable_value)
+
+total_samples <- nrow(all_sample_diagnosis_df) # needed for later math
+
+sample_diagnosis_df <- all_sample_diagnosis_df |>
+  # don't consider cell line (or multiplexed?)
+  # for now at least, we are retaining multiplexed
+  dplyr::filter(!is_cell_line)
 ```
@@
 ```{r}
-sum(sample_counts_df$reference >= params$n_reference_threshold) / nrow(sample_counts_df)
+sum(sample_counts_df$reference >= params$n_reference_threshold) / total_samples
 ```
@@
 
-There is one project about 7x as many reference as query cells; let's look more closely:
+There is one library with about 7x as many reference as query cells; let's look more closely:
 

### 2025-08-21 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
fix counting to handle multiplexed samples, and update language throughout to libraries which is actually what we are counting

@@
 
-The goal of this notebook is to look at cell type distributions across samples to determine how many samples could be run through `inferCNV`.
+The goal of this notebook is to look at cell type distributions across libraries to determine how many could be run through `inferCNV`.
 Specifically, we are interested in seeing the relative numbers of cells intended as `inferCNV` reference vs. query.
@@
 ```{r}
-all_sample_diagnosis_df <- metadata_files |>
+ all_library_diagnosis_df <- metadata_files |>
   purrr::map(
     \(file) {
-      readr::read_tsv(file) |>
-        dplyr::select(contains("scpca"), diagnosis, is_cell_line, is_multiplexed)
+      df <- readr::read_tsv(file) |>
+        dplyr::select(scpca_project_id, scpca_library_id, diagnosis, is_cell_line, is_multiplexed) |>
+        # have to join here because of multiplexed; inner_join also removes normals
+        dplyr::inner_join(diagnosis_map, by = c("diagnosis" = "submitted_diagnosis")) |>
+        dplyr::select(-diagnosis, -ontology_id, -human_readable_value) |>
+        dplyr::distinct()
   }) |>
-  dplyr::bind_rows() |>
-  # use inner_join to remove normals
-  dplyr::inner_join(diagnosis_map, by = c("diagnosis" = "submitted_diagnosis")) |>
-  dplyr::select(-ontology_id, -human_readable_value)
+  dplyr::bind_rows()
 
-total_samples <- nrow(all_sample_diagnosis_df) # needed for later math
+total_libraries <- nrow(all_library_diagnosis_df) # needed for later math
 
-sample_diagnosis_df <- all_sample_diagnosis_df |>
+library_diagnosis_df <- all_library_diagnosis_df |>
   # don't consider cell line (or multiplexed?)
@@
 consensus_df <- consensus_files |>
-  # only keep samples in sample_diagnosis_df
-  purrr::keep_at(sample_diagnosis_df$scpca_library_id) |>
+  # only keep libraries in scpca_project_id_diagnosis_df
+  purrr::keep_at(library_diagnosis_df$scpca_library_id) |>
   purrr::imap(
@@
          scpca_project_id = project_id,
-         scpca_sample_id = sample_id,
          scpca_library_id = library_id,
@@
 ```{r}
-query_reference_df <- unique(sample_diagnosis_df$diagnosis_group) |>
+query_reference_df <- unique(library_diagnosis_df$diagnosis_group) |>
   purrr::map(
@@
 
-      library_ids <- sample_diagnosis_df |>
+      library_ids <- library_diagnosis_df |>
         dplyr::filter(diagnosis_group == broad_diagnosis) |>
@@
 ```{r}
-sample_counts_df <- query_reference_df |>
+library_counts_df <- query_reference_df |>
   dplyr::count(scpca_library_id, cell_class) |>
@@
   ) |>
-  dplyr::inner_join(sample_diagnosis_df, by = "scpca_library_id") |>
+  dplyr::inner_join(library_diagnosis_df, by = "scpca_library_id") |>
   dplyr::select(
@@
     diagnosis_group, 
-    diagnosis, 
     query, 
@@
 
-Using a threshold of needing at least `r params$n_reference_threshold` reference cells, how many samples could be run through `inferCNV`?
+Using a threshold of needing at least `r params$n_reference_threshold` reference cells, how many libraries could be run through `inferCNV`?
 
 ```{r}
-usable_df <- sample_counts_df |>
+usable_df <- library_counts_df |>
   dplyr::group_by(diagnosis_group) |>
@@
     n_usable = sum(reference >= params$n_reference_threshold),
-    total_group_samples = dplyr::n(), 
-    frac_usable = round(n_usable / total_group_samples, 3) 
+    total_group_libs = dplyr::n(), 
+    frac_usable = round(n_usable / total_group_libs, 3) 
   ) |>
-  dplyr::arrange(desc(total_group_samples)) |>
-  dplyr::select(diagnosis_group, total_group_samples, n_usable, frac_usable)
+  dplyr::arrange(desc(total_group_libs)) |>
+  dplyr::select(diagnosis_group, total_group_libs, n_usable, frac_usable)
 
@@
 ```
-All of our broad diagnosis groups are represented, which is nice to see.
+Except for a handful of "other solid tumors," our broad diagnosis groups are well-represented, which is nice to see.
 
-Which diagnosis groups have _no samples_ with sufficient reference cells?
+Which diagnosis groups have _no libraries_ with sufficient reference cells?
 
@@
 ```
-It's not too surprising to see `Mixed phenotype acute leukemia` here since most immune cell types can't be included in the reference, and the other groups comprise very small numbers of samples.
+It's not too surprising to see `Mixed phenotype acute leukemia` here since most immune cell types can't be included in the reference, and the other groups comprise very small numbers of libraries.
 
-In total, this fraction of ScPCA samples have sufficient reference cells to run through `inferCNV`:
+In total, this fraction of ScPCA libraries have sufficient reference cells to run through `inferCNV`:
 
 ```{r}
-sum(sample_counts_df$reference >= params$n_reference_threshold) / total_samples
+sum(library_counts_df$reference >= params$n_reference_threshold) / total_libraries
 ```
 
-Of the samples with sufficient reference cells, do any have _fewer_ query than reference cells?
+Of the libraries with sufficient reference cells, do any have _fewer_ query than reference cells?
 
 ```{r}
-sample_counts_df |>
+library_counts_df |>
   # consider usable only
@@
 
-This sample indeed is dominated by immune cell types with very low tumor purity, but it still seems to have enough query cells to not be a problem and should be ok to run.
+This library indeed is dominated by immune cell types with very low tumor purity, but it still seems to have enough query cells to not be a problem and should be ok to run.
 
@@
 
-In this section, we plot the relative contributions of each cell type group in the samples we can run. 
+In this section, we plot the relative contributions of each cell type group in the libraries we can run. 
 We show the relative proportions of consensus cell type groups in the _reference_, and we show all other query cells grouped together in gray.
@@
 ```{r}
-usable_libraries <- sample_counts_df |>
+usable_libraries <- library_counts_df |>
   dplyr::filter(reference >= params$n_reference_threshold) |>
@@
 
-Currently, _before_ incorporating `SCimilarity` into our consensus cell types, we are able to run about 44% of samples through `inferCNV`. 
+Currently, _before_ incorporating `SCimilarity` into our consensus cell types, we are able to run about 44% of libraries through `inferCNV`. 
 Most broad diagnosis groups are represented, except `Mixed phenotype acute leukemia` and a handful of other solid tumors.

### 2025-08-25 analyses/cell-type-consensus/exploratory-notebooks/06-update-consensus-scimilarity.Rmd

@@
 
+It looks like we have ~7500 cases where there is more than one unique LCA identified. 
+What are the possible LCAs for those scenarios?
+
 ```{r}
@@
 Generally speaking it looks like there are terms that are of the same lineage which makes sense. 
-I think we could just stick with our previous guidelines of choosing the label with the fewest descendants or is the most specific term.
+I think we could just stick with our previous guidelines of choosing the label with the fewest descendants or the most specific term.
 

### 2025-08-25 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
add more NE annotations

@@
     * We perform this first at the label level and then at the broad family level.
-4. Finally, following the conclusions of Bonine et al. (2024), we designate any cell labeled as `Neuroendocrine` as a tumor cell.
+4.
+If the consensus cell type is `Unknown` and one of `SingleR` or `scANVI/scArches` labels is `Neuroendocrine` and the other label is one of `Schwann cell`, `Stromal other` or `fibroblast`, we assign the `Neuroendocrine` label
+5. Finally, following the conclusions of Bonine et al. (2024), we designate any cell labeled as `Neuroendocrine` as a tumor cell.
 
@@
 
+# Define vector of tumor-like cell types
+tumorlike_cells <- c("Schwann", "Stromal other", "Fibroblast")
+
+# pull out NE ontology for assignment code
+ne_ontology <- ontology_slim_df$CL_ontology_id[ontology_slim_df$NBAtlas_label == "Neuroendocrine"]
+
 annotation_df <- celltype_df |>
@@
       !is.na(consensus_family_ontology) & scanvi_family_ontology  == consensus_family_ontology ~ scanvi_family,
+      ########## Assign Neuroendocrine where possible; for this step, we refer to NBAtlas labels directly since
+      # the `tumorlike_cells` don't all have ontology ids
+      is.na(consensus_ontology) & singler_label == "Neuroendocrine" & scanvi_label %in% tumorlike_cells ~ "Neuroendocrine",
+      is.na(consensus_ontology) & scanvi_label == "Neuroendocrine" & singler_label %in% tumorlike_cells ~ "Neuroendocrine",
       # Everything else is Unknown

### 2025-08-25 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
remove unneeded stuff

@@
 4.
-If the consensus cell type is `Unknown` and one of `SingleR` or `scANVI/scArches` labels is `Neuroendocrine` and the other label is one of `Schwann cell`, `Stromal other` or `fibroblast`, we assign the `Neuroendocrine` label
+If the consensus cell type is `Unknown` and one of `SingleR` or `scANVI/scArches` labels is `Neuroendocrine` and the other label is one of `Schwann`, `Stromal other` or `Fibroblast`, we assign the `Neuroendocrine` label
 5. Finally, following the conclusions of Bonine et al. (2024), we designate any cell labeled as `Neuroendocrine` as a tumor cell.
@@
 tumorlike_cells <- c("Schwann", "Stromal other", "Fibroblast")
-
-# pull out NE ontology for assignment code
-ne_ontology <- ontology_slim_df$CL_ontology_id[ontology_slim_df$NBAtlas_label == "Neuroendocrine"]
 

### 2025-09-04 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
multiplexed sentence
update sep

@@
         dplyr::filter(diagnosis_group == broad_diagnosis) |>
-        tidyr::separate_rows(celltype_groups, sep = ", ") |>
+        tidyr::separate_rows(celltype_groups, sep = ",") |>
         dplyr::pull(celltype_groups)
@@
 
+Note that we are including multiplexed libraries from project `SCPCP000009` here as well; 11 libraries from that project from are eligible, most of which are multiplexed.
+These libraries would effectively be using pooled references in `inferCNV`, since we did not demultiplex libraries.
+
 ## Session Info

### 2025-09-05 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
Update analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

@@
         tidyr::separate_rows(celltype_groups, sep = ",") |>
-        dplyr::pull(celltype_groups)
+        dplyr::pull(celltype_groups) |>
+        stringr::str_trim()
 

### 2025-09-09 analyses/cell-type-scimilarity/exploratory-notebooks/01-scimilarity-results-summary.Rmd

---
title: "Explore SCimilarity cell type annotations"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

This notebook summarizes the findings from assigning cell type labels to all ScPCA samples using `SCimilarity`.

Running this notebook requires the results from the `cell-type-consensus` and `cell-type-scimilarity` module in `OpenScPCA-nf` to be saved to `OpenScPCA-analysis/data/current/results`.

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)

# Define color ramp for shared use in the heatmaps
heatmap_col_fun <- circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue"))
# Set heatmap padding option
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(0.6, "in"))

# settings
options(
  dplyr.summarise.inform = FALSE, 
  readr.show_col_types = FALSE
)

```

## Functions

```{r}
# function to read in both scimilarity and consensus results and combine into a single data frame 
prep_data <- function(library_id, scimilarity_file, consensus_file){

  scimilarity_df <- readr::read_tsv(scimilarity_file)
  consensus_df <- readr::read_tsv(consensus_file)

  annotation_df <- consensus_df |> 
    dplyr::left_join(scimilarity_df, by = c("barcodes" = "barcode")) |> 
    dplyr::mutate(
      cell_id = glue::glue("{library_id}-{barcodes}"), 
      total_cells_per_library = dplyr::n()
    )

  return(annotation_df)

}
```

```{r}
# Copied from scpca-nf template report 
#' Format DT datatables
#'
#' @param df Data frame to format
#'
#' @return DT datatable
format_datatable <- function(df, caption) {
  df |>
    DT::datatable(
      rownames = FALSE,
      extensions = "Scroller",
      caption = caption,
      options = list(
        scroller = TRUE,
        deferRender = TRUE,
        scrollX = TRUE,
        scrollY = 400,
        scrollCollapse = TRUE,
        language = list(search = "Filter:"),
        columnDefs = list(list(
          # render missing values as "NA" rather than blanks in the table
          targets = "_all",
          render = DT::JS(
            "function(data, type, row, meta) {",
            "return data === null ? 'NA' : data;",
            "}"
          )
        ))
      )
    )
}
```

## Data setup

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)
# results directory with cell-type-consensus 
results_dir <- file.path(repository_base, "data", "current", "results")
consensus_results_dir <- file.path(results_dir, "cell-type-consensus")
scimilarity_results_dir <- file.path(results_dir, "cell-type-scimilarity")

# diagnoses table used for labeling plots from cell-type-consensus module
diagnoses_file <- file.path(repository_base, "analyses", "cell-type-consensus", "sample-info", "project-diagnoses.tsv")
```

```{r}
# use the jaccard functions from cell-type-neuroblastoma-04 module
jaccard_functions <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04", "scripts", "utils", "jaccard-utils.R")
source(jaccard_functions)
```

```{r}
# list all results files 
consensus_results_files <- list.files(consensus_results_dir, pattern = "_processed_consensus-cell-types\\.tsv.\\gz$", recursive = TRUE, full.names = TRUE)
consensus_files_df <- data.frame(
  library_id = stringr::word(basename(consensus_results_files), 1, sep = "_"),
  consensus_file = consensus_results_files
)

scimilarity_results_files <- list.files(scimilarity_results_dir, pattern = "_processed_scimilarity-celltype-assignments\\.tsv.\\gz$", recursive = TRUE, full.names = TRUE)
scimilarity_files_df <- data.frame(
  library_id = stringr::word(basename(scimilarity_results_files), 1, sep = "_"),
  scimilarity_file = scimilarity_results_files
)

all_files_df <- scimilarity_files_df |> 
  dplyr::left_join(consensus_files_df, by = "library_id")

# define cell line projects to remove
cell_line_projects <- c("SCPCP000020", "SCPCP000024")
```

```{r}
# check that everything has a matching consensus file 
# we joined the consensus into the scimilarity so only libraries with scimilarity are included 
if(sum(is.na(all_files_df$consensus_file)) > 0){
  stop("Missing matching consensus cell type file for all libraries")
}
```

```{r, message=FALSE, warning=FALSE}
# read in diagnoses
diagnoses_df <- readr::read_tsv(diagnoses_file)

# read in all results as a single dataframe
results_df <- all_files_df |> 
  purrr::pmap(prep_data) |> 
  dplyr::bind_rows() |> 
  # remove cell line projects
  dplyr::filter(!project_id %in% cell_line_projects) |> 
  # add in diagnoses 
  dplyr::left_join(diagnoses_df, by = "project_id") |> 
  dplyr::mutate(
    # create a label for plotting
    project_label = glue::glue("{project_id}:{diagnosis}")
  )
```

## `SCimilarity` results

Below we will summarize the results from `SCimilarity` on all ScPCA projects.

### Table of all cell types in ScPCA

First, we just look at the cell types identified and see if any of the cells are not assigned.

```{r}
sum(is.na(results_df$scimilarity_celltype_annotation))
```

It looks like every cell gets a cell type, so that's something to keep in mind when looking at these results.

Let's see what cell types are presented in our data?

```{r}
results_df |> 
  dplyr::ungroup() |> 
  dplyr::count(scimilarity_celltype_annotation, name = "total_cells") |> 
  dplyr::arrange(desc(total_cells)) |> 
      format_datatable(caption = "All samples")
```

So it looks like we have a lot of "native cell" labels which really just means the cell is a cell and didn't align with anything else in the reference. 
This label is probably similar to our "other" label in `CellAssign`.

Also, most of our cells are progenitor cells or hematopoietic stem cells, which either means a lot of the cells are tumor cells and show stem and progenitor like properties or these are from the blood tumors which make up a large part of our atlas.
Either way, we have cell types!

### Table of cell types per project

This section shows a table of all cell types identified in each project. 
Each project will have its own table.

```{r}
# get all the unique labels
project_labels <- unique(results_df$project_label)

tables <- project_labels |> 
  purrr::map(\(label) {

    results_df |> 
      dplyr::filter(project_label == label) |> 
      dplyr::group_by(scimilarity_celltype_annotation) |> 
      dplyr::summarise(
        total_cells = dplyr::n()
      ) |> 
      dplyr::arrange(desc(total_cells)) |> 
      format_datatable(caption = label)

  })

htmltools::tagList(tables)
```

### Comparison to existing consensus cell types

In this section, we compare the top 15 `SCimilarity` cell type annotations to the consensus cell types determined using only `CellAssign` and `SingleR` annotations for all libraries in a single project. 
The rows correspond to consensus cell types and the columns correspond to `SCimilarity` annotations. 
All other cell types not in the top 10 represented cell types in a project are grouped into the "All remaining cell types" category.

```{r}
results_df <- results_df |> 
    dplyr::group_by(project_id) |> 
    dplyr::mutate(
      # get most frequently observed cell types across libraries in that project 
      scimilarity_celltype_lumped = forcats::fct_lump_n(scimilarity_celltype_annotation, 15, other_level = "All remaining cell types", ties.method = "first") |> 
        # sort by frequency 
        forcats::fct_infreq() |> 
        # make sure all remaining and unknown are last, use this to assign colors in specific order
        forcats::fct_relevel("All remaining cell types", after = Inf) |> 
        as.character()
    )
```

```{r, fig.height = 10, fig.width = 10}
project_labels |> 
  purrr::walk(\(label){

    results_df |> 
      dplyr::filter(project_label == label) |> 
      make_jaccard_heatmap(
        annotation_col1 = "scimilarity_celltype_lumped",
        annotation_col2 = "consensus_annotation",
        label1 = glue::glue("{label} \nSCimilarity"),
        label2 = "Consensus"
      )

  })
```

As expected there is some direct agreement or places where more broad consensus cell types are split up across more granular cell types in `SCimilarity`. 
A few notable examples:

- `SCPCP000001`: Cells labeled as macrophages in consensus are split between glial and microglial cells in `SCimilarity`. 
Natural killer cells match up between consensus and `SCimilarity`. 
- `SCPCP000006`: Endothelial cells match up between consensus and `SCimilarity`. 
The unknown cells in consensus correspond to mesenchymal stem cells and a variety of kidney cells. 
- `SCPCP000007`: Mature T cell in consensus is split between CD4 and CD8 T cells in `SCimilarity`.

We also see places where there is not agreement, which is to be expected.

## Conclusions

Of the 203 possible cell types in `SCimilarity`, 191 of them are represented in ScPCA samples. 
There are quite a few scenarios where consensus cell types "agree" with `SCimilarity`. 
It will be informative to look at the new consensus cell types and how they change with the addition of the `SCimilarity` annotations, but in general the addition of `SCimilarity` seems quite useful!

## Session info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-09-12 analyses/cell-type-consensus/exploratory-notebooks/07-compare-consensus.Rmd
notebook comparing consensus cell type outputs

---
title: "Compare consensus cell types with SCimilarity to consensus cell types without SCimilarity"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

This notebook compares the results from assigning consensus cell types with `SCimilarity` to the previous consensus cell types without `SCimilarity`. 
All results from the `cell-type-consensus` module in `OpenScPCA-nf` must be saved to `results` prior to rendering this notebook.

Note that any multiplexed libraries have been excluded from this notebook as `SCimilarity` is not run on those libraries when using the `OpenScPCA-nf` module.

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)

# Define color ramp for shared use in the heatmaps
heatmap_col_fun <- circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue"))
# Set heatmap padding option
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(0.6, "in"))

# settings
options(
  dplyr.summarise.inform = FALSE, 
  readr.show_col_types = FALSE
)
```

## Data setup

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)
module_base <- file.path(repository_base, "analyses", "cell-type-consensus")

# results directory with cell-type-consensus 
results_dir <- file.path(module_base, "results", "cell-type-consensus")

# diagnoses table used for labeling plots 
diagnoses_file <- file.path(module_base, "sample-info", "project-diagnoses.tsv")
```

```{r}
# use the jaccard functions from cell-type-neuroblastoma-04 module
jaccard_functions <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04", "scripts", "utils", "jaccard-utils.R")
source(jaccard_functions)
```

```{r}
# list all results files 
results_files <- list.files(results_dir, pattern = "_consensus-cell-types\\.tsv.\\gz$", recursive = TRUE, full.names = TRUE)

# define cell line projects to remove
cell_line_projects <- c("SCPCP000020", "SCPCP000024")
```

```{r, message=FALSE, warning=FALSE}
# read in diagnoses
diagnoses_df <- readr::read_tsv(diagnoses_file)

# read in results and prep data frame for plotting 
all_results_df <- results_files |> 
  purrr::map(readr::read_tsv, col_types = "c")|> 
  dplyr::bind_rows() |> 
  # remove cell line projects
  dplyr::filter(!project_id %in% cell_line_projects,
                # remove multiplexed libraries since we didn't run scimilarity on them
                !stringr::str_detect(sample_id, ";")) |> 
  # add in diagnoses 
  dplyr::left_join(diagnoses_df, by = "project_id") |> 
  dplyr::mutate(
    # create a label for plotting
    project_label = glue::glue("{project_id}:{diagnosis}")
  )
```

```{r, message=FALSE, warning=FALSE}
# Turn into a dataframe with one row per library per celltype
# Make a table for each consensus method separately and then stack by consensus cell type old/new 
# one row for old and one row for each cell type

# first add columns for total cells per library, number of new and old cell types
grouped_df <- all_results_df |> 
    dplyr::group_by(library_id) |> 
    dplyr::mutate(
      total_cells_per_library = dplyr::n(),
      num_old_celltypes = length(unique(consensus_annotation)),
      num_new_celltypes = length(unique(existing_consensus_celltype_annotation))
    ) |>
    dplyr::ungroup()

# get stats for old consensus cell types
old_consensus_df <- grouped_df |> 
    dplyr::group_by(project_id, project_label, diagnosis, library_id, sample_type, existing_consensus_celltype_annotation, existing_consensus_celltype_ontology) |> 
    dplyr::summarize(num_celltypes = unique(num_new_celltypes), # constant for each library
                     total_cells_per_annotation = dplyr::n(),
                     total_cells_per_library = unique(total_cells_per_library)) |>
    dplyr::mutate(
      # add percentage 
      percent_cells_annotation = round((total_cells_per_annotation / total_cells_per_library) * 100, 2)
    ) |> 
    dplyr::ungroup()

# stats for new consensus cell types
new_consensus_df <- grouped_df |> 
    dplyr::group_by(project_id, project_label, diagnosis, library_id, sample_type, consensus_annotation, consensus_ontology) |> 
    dplyr::summarize(num_celltypes = unique(num_old_celltypes), # constant for each library
                     total_cells_per_annotation = dplyr::n(),
                     total_cells_per_library = unique(total_cells_per_library)) |>
    dplyr::mutate(
      # add percentage 
      percent_cells_annotation = round((total_cells_per_annotation / total_cells_per_library) * 100, 2)
    ) |> 
    dplyr::ungroup()

# combine into a single df and add a column to indiciate old/new 
df_list <- list(old_consensus_df, new_consensus_df) |> 
  purrr::set_names(c("no_scimilarity", "with_scimilarity"))

all_grouped_df <- dplyr::bind_rows(df_list, .id = "consensus_type")
```

## Number of unknown cells

The plot below shows the number of cells annotated as unknown for each library using old (without `SCimilarity`) vs. new (with `SCimilarity`) consensus annotations. 
The red bar indicates the median percentage of cells.

```{r, fig.height=7}
# only get the unknown cells
unknown_only <- all_grouped_df |> 
  dplyr::filter(consensus_annotation == "Unknown" | existing_consensus_celltype_annotation == "Unknown")

# compare the distribution of unknown with and without scimilarity
ggplot(unknown_only, aes(x = consensus_type, y = percent_cells_annotation)) +
  ggforce::geom_sina(size = 0.1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        plot.margin = margin(10,10,10,10)) +
  labs(
    x = "", 
    y = "Percent of cells annotated as Unknown"
  ) +
  stat_summary(fun=median, geom="crossbar" , color = "red", linewidth = 0.2)

```

As expected, the percentage of cells annotated as unknown is lower when `SCimilarity` is incorporated, suggesting we are able to annotate more cells with a meaningful label that is not unknown.

## Number of cell types observed

Below we look at the number of cell types observed in each project for all samples.

```{r, fig.height=15, fig.width=10}
num_celltypes_df <- all_grouped_df |> 
  # add a new line for facet labels 
  dplyr::mutate(facet_label = glue::glue("{project_id}\n{diagnosis}")) |>
  dplyr::select(facet_label, library_id, num_celltypes, consensus_type) |> 
  unique()

ggplot(num_celltypes_df, 
       aes(x = consensus_type, y = num_celltypes, group = library_id)) +
  geom_point(aes(color = consensus_type)) +
  geom_line(aes(group = library_id), color = "gray60", alpha = 0.7) +
  facet_wrap(vars(facet_label), ncol = 3) +
  labs(
    x = "Consensus cell type",
    y = "Number of cell types"
  ) +
  theme_bw()

```

As expected, we see that the number of cell types identified increases with the addition of `SCimilarity`.

## Heatmaps comparing old and new consensus cell types

In this section, we compare the top 15 cell type annotations for consensus cell types with `SCimilarity` (rows) to consensus cell types without `SCimilarity` (columns). 
All other cell types not in the top 15 represented cell types in a project are grouped into the "All remaining cell types" category.

Note that in some cases, there are fewer than 15 unique cell types.

```{r, warning=FALSE}
all_results_df <- all_results_df |> 
    dplyr::group_by(project_id) |> 
    dplyr::mutate(
      # get most frequently observed cell types across libraries in that project 
      existing_consensus_celltype_lumped = forcats::fct_lump_n(existing_consensus_celltype_annotation, 15, other_level = "All remaining cell types", ties.method = "first") |> 
        # sort by frequency 
        forcats::fct_infreq() |> 
        # make sure all remaining and unknown are last, use this to assign colors in specific order
        forcats::fct_relevel("All remaining cell types", after = Inf) |> 
        as.character(), 
      # do the same thing for the new consensus cell types
      consensus_celltype_lumped = forcats::fct_lump_n(consensus_annotation, 15, other_level = "All remaining cell types", ties.method = "first") |> 
        # sort by frequency 
        forcats::fct_infreq() |> 
        # make sure all remaining and unknown are last, use this to assign colors in specific order
        forcats::fct_relevel("All remaining cell types", after = Inf) |> 
        as.character(), 
      # jaccard functions expect a cell id column
      cell_id = glue::glue("{library_id}-{barcodes}")
    )

project_labels <- unique(all_results_df$project_label)
```

```{r, fig.height = 10, fig.width = 10}
project_labels |> 
  purrr::walk(\(label){

    all_results_df |> 
      dplyr::filter(project_label == label) |> 
      make_jaccard_heatmap(
        annotation_col1 = "existing_consensus_celltype_lumped",
        annotation_col2 = "consensus_celltype_lumped",
        label1 = glue::glue("{label} \nNo Scimilarity"),
        label2 = "With SCimilarity"
      )

  })
```

In general we see a fair amount of agreement in the overall cell types are defined with a bit more granularity in the cell types assigned with `SCimilarity`.

## Distribution of new consensus cell types

Now we look at the distribution of the cell types in each sample. 
For these plots, we will pull out the top 9 cell types for each project. 
All other cells will be labeled with "All remaining cell types".

The top cell types are determined by counting how many libraries each cell type is found in within a project and taking the most frequent types.

```{r, warning=FALSE}
plot_df <- all_grouped_df |> 
  dplyr::group_by(project_id) |> 
  # remove the old results from plotting
  dplyr::select(-c(existing_consensus_celltype_annotation, existing_consensus_celltype_ontology)) |> 
  tidyr::drop_na(consensus_annotation) |> 
  dplyr::mutate(
    # get most frequently observed cell types across libraries in that project 
    top_celltypes = forcats::fct_lump_n(consensus_annotation, 9, other_level = "All remaining cell types", ties.method = "first") |> 
      # sort by frequency 
      forcats::fct_infreq() |> 
      # make sure all remaining and unknown are last, use this to assign colors in specific order
      forcats::fct_relevel("All remaining cell types", "Unknown", after = Inf)
  ) |> 
  dplyr::ungroup()

# get all unique cell types ordered by frequency 
unique_celltypes <- plot_df |> 
  dplyr::filter(!top_celltypes %in% c("All remaining cell types", "Unknown")) |> 
  dplyr::pull(top_celltypes) |> 
  unique() |>
  sort() |> 
  as.character()

# get color palette
colors <- c(
  colorRampPalette(RColorBrewer::brewer.pal(12, "Paired"))(length(unique_celltypes)),
  "grey60", # all remaining
  "grey95" # unknown
)
names(colors) <- c(unique_celltypes, "All remaining cell types", "Unknown")
```

```{r, fig.height=60, fig.width=10}
# stacked bar chart showing the distribution of the top 9 cell types for each project, including Unknown
project_labels |> 
  purrr::map(\(label){

    project_df <- plot_df |> 
      dplyr::filter(project_label == label) |> 
      dplyr::mutate(
        # relevel factors for specific project 
        top_celltypes = forcats::fct_infreq(top_celltypes) |> 
          forcats::fct_relevel("All remaining cell types", "Unknown", after = Inf)
      )

    # make a stacked bar chart with top cell types 
    ggplot(project_df) + 
      aes(
        x = library_id, 
        y = percent_cells_annotation, 
        fill = top_celltypes
      ) +
      geom_col() + 
      # split samples based on sample type, patient tissue or pdx 
      facet_wrap(vars(sample_type), scales ="free") +
      scale_y_continuous(expand = c(0,0)) +
      scale_fill_manual(values = colors, name = "cell type") +
      ggtitle(label) +
      theme(axis.text.x = element_blank())

    }) |>
  patchwork::wrap_plots(ncol = 1)
```

We are definitely labeling more cells than we were previously and in some cases we may be assigning a "normal" cell type to tumor cells. 
This is not totally surprising as many solid tumors have cells that resemble fibroblasts and muscle cells. 
We also see various progenitors and HSCs being labeled in the leukemias which reflect how leukemia cells resemble those cell types. 
Overall, I think adding in `SCimilarity` is giving us a lot more information than without it.

## Session info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```

### 2025-09-12 analyses/cell-type-consensus/exploratory-notebooks/07-compare-consensus.Rmd
old is new and new is old

@@
       total_cells_per_library = dplyr::n(),
-      num_old_celltypes = length(unique(consensus_annotation)),
-      num_new_celltypes = length(unique(existing_consensus_celltype_annotation))
+      num_old_celltypes = length(unique(existing_consensus_celltype_annotation)),
+      num_new_celltypes = length(unique(consensus_annotation))
     ) |>
@@
     dplyr::group_by(project_id, project_label, diagnosis, library_id, sample_type, existing_consensus_celltype_annotation, existing_consensus_celltype_ontology) |> 
-    dplyr::summarize(num_celltypes = unique(num_new_celltypes), # constant for each library
+    dplyr::summarize(num_celltypes = unique(num_old_celltypes), # constant for each library
                      total_cells_per_annotation = dplyr::n(),
@@
     dplyr::group_by(project_id, project_label, diagnosis, library_id, sample_type, consensus_annotation, consensus_ontology) |> 
-    dplyr::summarize(num_celltypes = unique(num_old_celltypes), # constant for each library
+    dplyr::summarize(num_celltypes = unique(num_new_celltypes), # constant for each library
                      total_cells_per_annotation = dplyr::n(),

### 2025-09-12 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
update validation notebook template

@@
   library(patchwork)
-  library(data.table)
 })
@@
 )
+
+# Use DuckDB for dplyr functions
+suppressMessages(duckplyr::methods_overwrite())
 ```
@@
 # path to results directory
-results_dir <- file.path(module_base, "..", "..", "data", "current", "results", "cell-type-consensus", params$project_id)
+results_dir <- file.path(module_base, "results", "cell-type-consensus", params$project_id)
 ```
@@
 # get consensus results 
-consensus_results_files <- list.files(results_dir, pattern = "_processed_consensus-cell-types\\.tsv\\.gz$", recursive = TRUE, full.names = TRUE)
+consensus_results_files <- list.files(results_dir, pattern = "_consensus-cell-types\\.tsv\\.gz$", recursive = TRUE, full.names = TRUE)
 
 # get gene expression files 
-gene_exp_files <- list.files(results_dir, pattern = "_processed_marker-gene-expression\\.tsv\\.gz$", recursive = TRUE, full.names = TRUE)
+gene_exp_files <- list.files(results_dir, pattern = "_marker-gene-expression\\.tsv\\.gz$", recursive = TRUE, full.names = TRUE)
 
@@
 ```{r, message=FALSE, warning=FALSE}
-# read in consensus and gene expression results 
-# use vroom and data.table to try and be more memory-efficient
-consensus_dt <- consensus_results_files |> 
-  purrr::map(fread) |> 
-  data.table::rbindlist(fill = TRUE, use.names = TRUE)
+# read in validation markers as data.tables
+markers_df <- duckplyr::read_csv_duckdb(validation_markers_file) |> 
+  duckplyr::as_duckdb_tibble()
 
-gene_exp_dt <- gene_exp_files |> 
-  purrr::map(fread) |> 
-  data.table::rbindlist(fill = TRUE, use.names = TRUE)
-
-# read in validation markers as data.tables
-markers_dt <- fread(validation_markers_file)
-
-validation_groups_dt <- fread(validation_groups_file) |> 
+validation_groups_df <- duckplyr::read_csv_duckdb(validation_groups_file) |> 
   # rename final assigned group to avoid conflicts when merging in marker gene expression 
   # we want to separate the marker gene group from the actual cell type annotation
-  dplyr::select(consensus_annotation, broad_celltype_group = validation_group_annotation)
+  dplyr::select(consensus_annotation, broad_celltype_group = validation_group_annotation) |> 
+  duckplyr::as_duckdb_tibble()
+
+# read in files directly to duckdb tables
+  # specify all varchar for consensus to avoid parsing error
+consensus_df <- duckplyr::read_csv_duckdb(consensus_results_files, options = list(sep = "\t", union_by_name = TRUE))
+
+gene_exp_df <- duckplyr::read_csv_duckdb(gene_exp_files, options = list(sep = "\t", union_by_name = TRUE))
 ```
@@
 # explicitly skip project 24 since the current processed data doesn't have the updated combination of PDX and cell line
-is_cell_line <- all(consensus_dt$sample_type == "cell line") | params$project_id == "SCPCP000024"
+is_cell_line <- all(consensus_df$sample_type == "cell line") | params$project_id == "SCPCP000024"
 
@@
 # includes original consensus annotation, braod cell type group, marker gene id, logcounts, validation group annotation, information regarding expression of marker gene in CellMarker tissues 
-consensus_dt <- consensus_dt |> 
+consensus_df <- consensus_df |> 
   # add in broad cell type group which is used for plotting
   # groups similar cell types together 
-  dplyr::left_join(validation_groups_dt, by = "consensus_annotation") |> 
-  dplyr::left_join(gene_exp_dt, by = c("barcodes", "library_id")) |> 
+  dplyr::left_join(validation_groups_df, by = "consensus_annotation") |> 
+  dplyr::left_join(gene_exp_df, by = c("barcodes", "library_id")) |>
   # add marker gene information (associated validation group annotation, gene observed count, percent tissues)
   # account for the same gene being present in multiple cell types 
-  dplyr::left_join(markers_dt, by = "ensembl_gene_id", relationship = "many-to-many") |> 
+  # we only want genes that are in both and throw out extra consensus genes 
+  dplyr::left_join(markers_df, by = "ensembl_gene_id", relationship = "many-to-many") |> 
   dplyr::mutate(detected = logcounts > 0)
-
-# remove the extra data table 
-rm(gene_exp_dt)
 ```
@@
 # get total number of cells per final annotation group 
-total_cells_df <- consensus_dt |> 
+total_cells_df <- consensus_df |> 
   dplyr::select(library_id, barcodes, broad_celltype_group) |> 
@@
 # second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
-group_stats_df <- consensus_dt |> 
+group_stats_df <- consensus_df |> 
   # for each assigned cell type/marker gene combo get total detected and mean expression
   # group by both broad group and validation group to account for genes that are expressed in more than one cell type
-  dplyr::group_by(broad_celltype_group, ensembl_gene_id, validation_group_annotation) |>
   dplyr::summarize(
+    .by = c("broad_celltype_group", "ensembl_gene_id", "validation_group_annotation"),
     detected_count = sum(detected),
@@
   # this includes all possible marker genes and all possible validation group assignments 
-  dplyr::left_join(markers_dt, by = c("ensembl_gene_id", "validation_group_annotation"), relationship = "many-to-many") |>
+  dplyr::left_join(markers_df, by = c("ensembl_gene_id", "validation_group_annotation"), relationship = "many-to-many") |>
   # now get the mean expression/ mean percentage across all marker genes for a given validation group
@@
   dplyr::filter(total_cells > 50) |> 
-  dplyr::rowwise() |> 
   dplyr::mutate(
       # get total percent expressed
-    percent_exp = (detected_count/total_cells) * 100,
+      percent_exp = (detected_count / total_cells) * 100,
       all_markers_percent_exp = (all_markers_detected_count/total_cells) * 100,
@@
       broad_celltype_group = tidyr::replace_na(broad_celltype_group, "unknown") |>
-      factor(levels =  c(unique(markers_dt$validation_group_annotation), "unknown"))
+        factor(levels = c(unique(markers_df$validation_group_annotation), "unknown"))
     )
@@
 # no longer need this and it takes up space 
-rm(consensus_dt)
+rm(consensus_df)
 
@@
 # assign colors to use in annotation bars 
-celltype_colors <- palette.colors(palette = "alphabet")[1:length(celltype_groups)]
+celltype_colors <-  colorRampPalette(RColorBrewer::brewer.pal(12, "Paired"))(length(celltype_groups))
 names(celltype_colors) <- celltype_groups
@@
 # note that we will use this for both the dotplot and the heatmap
-filtered_markers_df <- markers_dt |> 
+filtered_markers_df <- markers_df |> 
   dplyr::filter(validation_group_annotation %in% celltype_groups,
@@
 
-```{r, eval=!is_cell_line, fig.height=10, fig.width=15}
+```{r, eval=!is_cell_line, fig.height=15, fig.width=25}
 # filter out low expressed genes
@@
   geom_tile() +
+  facet_grid(cols = vars(validation_group_annotation), scales = "free", space = "free") +
   scale_fill_manual(values = celltype_colors, breaks = levels(dotplot_df$validation_group_annotation)) +
   ggmap::theme_nothing() +
-  theme(legend.position = "bottom") +
+  theme(
+    strip.background = element_rect(fill = "transparent", color = NA),
+    strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 12),
+    strip.placement = "outside",
+    legend.position = "none",
+    panel.spacing = unit(0.5, "lines"),
+    strip.clip = "off"
+  ) +
   labs(fill = "")
 
-dotplot + color_bar +
-  patchwork::plot_layout(ncol = 1, heights = c(4, 0.1)) 
+color_bar / dotplot +
+  patchwork::plot_layout(ncol = 1, heights = c(0.1, 4))
 ```

### 2025-09-15 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
define cell line projects

@@
 # check for if the project is a cell line project or not 
-# explicitly skip project 24 since the current processed data doesn't have the updated combination of PDX and cell line
-is_cell_line <- all(consensus_df$sample_type == "cell line") | params$project_id == "SCPCP000024"
+is_cell_line <- params$project_id %in% c("SCPCP000020", "SCPCP000024")
 
@@
         factor(levels = c(unique(markers_df$validation_group_annotation), "unknown"))
-    )
+    ) |> 
+  # remove any marker genes that don't have an assigned validation group
+  tidyr::drop_na()
 

### 2025-09-15 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
change back directory for CI

@@
 params:
-  project_id: "SCPCP000001"
+  project_id: "SCPCP000006"
 ---
@@
 # path to results directory
-results_dir <- file.path(module_base, "results", "cell-type-consensus", params$project_id)
+results_dir <- file.path(module_base, "..", "..", "data", "current", "results", "cell-type-consensus", params$project_id)
 ```
@@
   theme(
-    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
+    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
   ) +

### 2025-09-15 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
change default back to SCPCP000001

@@
 params:
-  project_id: "SCPCP000006"
+  project_id: "SCPCP000001"
 ---

### 2025-10-08 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
update notebook now that consensus are updated

@@
 ```{r}
-usable_df <- library_counts_df |>
+usable_counts_df <- library_counts_df |>
   dplyr::group_by(diagnosis_group) |>
@@
 
-usable_df
+usable_counts_df
 ```
@@
 ```{r}
-usable_df |>
+usable_counts_df |>
   dplyr::filter(frac_usable == 0)
@@
 ```{r}
-library_counts_df |>
-  # consider usable only
-  dplyr::filter(reference >= params$n_reference_threshold) |>
+usable_libraries_df <- library_counts_df |>
+  dplyr::filter(reference >= params$n_reference_threshold)
+
+fewer_query_df <- usable_libraries_df |>
   dplyr::filter(query < params$n_reference_threshold)
+
+fewer_query_df
 ```
 
-There is one library with about 7x as many reference as query cells; let's look more closely:
+There are five library with an order of magnitude more  7x as many reference as query cells; let's look more closely at the validation groups present to get a sense.
 
@@
 consensus_df |>
-  dplyr::filter(scpca_library_id == "SCPCL000021") |>
-  dplyr::count(consensus_annotation) |>
-  dplyr::arrange(n)
+  dplyr::filter(scpca_library_id %in% fewer_query_df$scpca_library_id) |>
+  dplyr::left_join(validation_group_df, by = c("consensus_ontology", "consensus_annotation")) |>
+  dplyr::count(scpca_library_id, validation_group_annotation) |>
+  dplyr::arrange(scpca_library_id)
 ```
 
-This library indeed is dominated by immune cell types with very low tumor purity, but it still seems to have enough query cells to not be a problem and should be ok to run.
+These librares are dominated by immune cell types with very few Unknown/tumor cells, but it's still worth running them through `inferCNV` while being aware that we _may_ experience issues with these samples.
 
-## Plots
+We should also check the eligible `SCPCP000009` libraries to see if they are multiplexed:
+
+```{r}
+library_diagnosis_df |>
+  dplyr::filter(scpca_project_id == "SCPCP000009", scpca_library_id %in% usable_libraries_df$scpca_library_id) |>
+  dplyr::count(is_multiplexed)
+```
+
+## Plot cell types
 
@@
 ```{r}
-usable_libraries <- library_counts_df |>
-  dplyr::filter(reference >= params$n_reference_threshold) |>
-  dplyr::pull(scpca_library_id)
-
 plot_df <- query_reference_df |>
-  dplyr::filter(scpca_library_id %in% usable_libraries) |>
+  dplyr::filter(scpca_library_id %in% usable_libraries_df$scpca_library_id) |>
   dplyr::left_join(validation_group_df, by = c("consensus_ontology", "consensus_annotation")) |>
@@
 
-Currently, _before_ incorporating `SCimilarity` into our consensus cell types, we are able to run about 44% of libraries through `inferCNV`. 
-Most broad diagnosis groups are represented, except `Mixed phenotype acute leukemia` and a handful of other solid tumors.
-All eligible projects except `SCPCP000011` (Retinoblastoma samples) are represented as well.
+We are able to run about 51% of libraries through `inferCNV`, based on consensus cell types derived from `SingleR`, `CellAssign`, and `SCimilarity` labels.
+Most broad diagnosis groups are represented, except `Mixed phenotype acute leukemia` (n=18) and a handful of other solid tumors (n=3 among Gastrointestinal stromal tumor and Hepatoblastoma).
+All eligible projects are represented as well.
 
-Note that we are including multiplexed libraries from project `SCPCP000009` here as well; 11 libraries from that project from are eligible, most of which are multiplexed.
+Note that we are including multiplexed libraries from project `SCPCP000009` here as well; 12 libraries from that project from are eligible, and all but of of those are multiplexed.
 These libraries would effectively be using pooled references in `inferCNV`, since we did not demultiplex libraries.

### 2025-10-08 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
phrasing

@@
 
-There are five library with an order of magnitude more  7x as many reference as query cells; let's look more closely at the validation groups present to get a sense.
+There are five library with an order of magnitude more reference than query cells; let's look more closely at the validation groups present to get a sense.
 

### 2025-10-08 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
typo

@@
 
-Note that we are including multiplexed libraries from project `SCPCP000009` here as well; 12 libraries from that project from are eligible, and all but of of those are multiplexed.
+Note that we are including multiplexed libraries from project `SCPCP000009` here as well; 12 libraries from that project from are eligible, and all but one of those are multiplexed.
 These libraries would effectively be using pooled references in `inferCNV`, since we did not demultiplex libraries.

### 2025-10-08 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
ensure we are only using most up to date consensus cells with scimilarity, meaning no assessment of multiplexed

--- 2025-10-08 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
@@
 library_diagnosis_df <- all_library_diagnosis_df |>
-  # don't consider cell line (or multiplexed?)
-  # for now at least, we are retaining multiplexed
-  dplyr::filter(!is_cell_line)
+  # don't consider cell line or multiplexed in this notebook
+  # the multiplexed samples were not processed with SCimilarity in OpenScPCA-nf
+  dplyr::filter(!is_cell_line, !is_multiplexed)
 ```
@@
       # bump guess_max to kill problems() warnings
-      readr::read_tsv(file, guess_max = 15000) |>
+      consensus_df <- readr::read_tsv(file, guess_max = 15000)
+
+      # ensure we have the most up-to-date files - aka, with SCimilarity
+      stopifnot("Missing SCimilarity in the consensus files" = "scimilarity_celltype_annotation" %in% colnames(consensus_df))
+
+      consensus_df |>
         dplyr::select(
@@
 
-We should also check the eligible `SCPCP000009` libraries to see if they are multiplexed:
-
-```{r}
-library_diagnosis_df |>
-  dplyr::filter(scpca_project_id == "SCPCP000009", scpca_library_id %in% usable_libraries_df$scpca_library_id) |>
-  dplyr::count(is_multiplexed)
-```
-
 ## Plot cell types
@@
 
-We are able to run about 51% of libraries through `inferCNV`, based on consensus cell types derived from `SingleR`, `CellAssign`, and `SCimilarity` labels.
+We are able to run about 50\% of libraries through `inferCNV`, based on consensus cell types derived from `SingleR`, `CellAssign`, and `SCimilarity` labels.
 Most broad diagnosis groups are represented, except `Mixed phenotype acute leukemia` (n=18) and a handful of other solid tumors (n=3 among Gastrointestinal stromal tumor and Hepatoblastoma).
@@
 
-Note that we are including multiplexed libraries from project `SCPCP000009` here as well; 12 libraries from that project from are eligible, and all but of of those are multiplexed.
-These libraries would effectively be using pooled references in `inferCNV`, since we did not demultiplex libraries.
+Note that it was not possible to fully assess multiplexed libraries here since they did not undergo `SCimilarity` (and hence their consensus cell types are not updated) inference in `OpenScPCA-nf`. 
+They will be included when running through `scpca-nf`, however, assuming they have sufficient reference cells; it is important to bear in mind that these libraries would effectively be using pooled references in `inferCNV`, since we did not demultiplex libraries.
 

### 2025-10-09 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd

@@
 
-There are five library with an order of magnitude more reference than query cells; let's look more closely at the validation groups present to get a sense.
+There are five libraries with an order of magnitude more reference than query cells; let's look more closely at the validation groups present to get a sense.
 

### 2025-10-16 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
spell libraries

@@
 
-These librares are dominated by immune cell types with very few Unknown/tumor cells, but it's still worth running them through `inferCNV` while being aware that we _may_ experience issues with these samples.
+These libraries are dominated by immune cell types with very few Unknown/tumor cells, but it's still worth running them through `inferCNV` while being aware that we _may_ experience issues with these samples.
 

### 2025-12-17 analyses/cell-type-ewings/template_notebooks/cnv-workflow/01-marker-genes.Rmd
manually run typos to fix a bunch of typos that are reasonable to fix

@@
 marker_plot_df <- celltype_df |>
-  dplyr::mutate(gene_expresed = dplyr::if_else(transformed_gene_expression > 0, TRUE, FALSE)) |>
-  dplyr::select(gene_symbol, gene_expresed, sum_classification)
+  dplyr::mutate(gene_expressed = dplyr::if_else(transformed_gene_expression > 0, TRUE, FALSE)) |>
+  dplyr::select(gene_symbol, gene_expressed, sum_classification)
 
-ggplot(marker_plot_df, aes(x = sum_classification, fill = gene_expresed)) +
+ggplot(marker_plot_df, aes(x = sum_classification, fill = gene_expressed)) +
   geom_bar(stat = "count") +

### 2025-12-17 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
manually run typos to fix a bunch of typos that are reasonable to fix

@@
   seed: 12345
-  padj_thershold: 0.05  # Threshold for adjusted p-value
+  padj_threshold: 0.05  # Threshold for adjusted p-value
   lfc_threshold: 1      # Threshold for log fold change
@@
 # filter the most relevant markers
-s.markers <- de_results[de_results$padj < params$padj_thershold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold, ]
+s.markers <- de_results[de_results$padj < params$padj_threshold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold, ]
 # add gene symbol for easiest interpretation of the result

### 2025-12-17 analyses/cell-type-wilms-tumor-06/supplemental_notebooks/cnv_exploration/06_infercnv_exploration.Rmd
manually run typos to fix a bunch of typos that are reasonable to fix

--- 2024-11-25 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd
@@
 ```{r}
-Do_CNV_score <- function(seurat_oject, infercnv_obj, reference_value) {
+Do_CNV_score <- function(seurat_object, infercnv_obj, reference_value) {
   score <- apply(infercnv_obj@expr.data, 2, function(x) {
@@
 
-  seurat_obj <- AddMetaData(seurat_oject, score, col.name = glue::glue("CNV-score_", reference_value))
+  seurat_obj <- AddMetaData(seurat_object, score, col.name = glue::glue("CNV-score_", reference_value))
   return(seurat_obj)

### 2025-12-17 analyses/cell-type-ewings/aucell-singler-annotation.sh

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-# This script first identfies tumor cells using AUCell for all samples in SCPCP000015
+# This script first identifies tumor cells using AUCell for all samples in SCPCP000015
 # Cells that are identified as tumor cells using AUCell are used to create a
 # combined tumor cell reference to use with SingleR
 # SingleR is then run on all samples using the combined tumor cell reference and

### 2025-12-17 analyses/cell-type-ewings/scripts/cnv-workflow/03-run-copykat.R

@@ -59,7 +59,7 @@ sce <- readr::read_rds(opt$sce_file)
 # get library id to use as prefix for all output file names
 library_id <- metadata(sce)$library_id
 
-# contstruct and create output folder if not already present
+# construct and create output folder if not already present
 fs::dir_create(opt$results_dir)
 
 # define scratch directory

### 2025-12-17 analyses/cell-type-ewings/scripts/utils/plotting-functions.R

@@ -45,7 +45,7 @@ expression_umap <- function(
 
 #' Density plot showing expression across clusters
 #'
-#' @param df Data frame with clusters for each cell and expresion values to be plotted 
+#' @param df Data frame with clusters for each cell and expression values to be plotted 
 #'   Must contain the `cluster` column 
 #' @param expression_columns Vector of columns present in the data frame that contain some
 #'   expression value to show on the x-axis of the density plot, such as AUC values or mean expression

### 2025-12-17 analyses/cell-type-wilms-tumor-06/scripts/06b_build-normal_reference.R

@@ -48,7 +48,7 @@ for (sample_id in sample_upfront_resection) {
 # merge all normal cells in one `Seurat` object
 srat_normal <- base::merge(s[[1]], y = s[2:length(s)])
 
-# we only keep as nornal cells immune and endothelial cells with a high mapping score
+# we only keep as normal cells immune and endothelial cells with a high mapping score
 srat_normal <- subset(srat_normal, subset = fetal_kidney_predicted.compartment.score > score_threshold)
 
 # save the `Seurat` object

### 2025-12-17 analyses/doublet-detection/scripts/01a_run-scdblfinder.R

@@ -61,7 +61,7 @@ run_scdblfinder <- function(sce,
   result_df <- result_df |>
     as.data.frame() |>
     tibble::rownames_to_column("barcodes") |>
-    # remove artifical doublets
+    # remove artificial doublets
     dplyr::filter(!stringr::str_starts(barcodes, "rDbl")) |>
     # keep only columns of interest
     dplyr::select(dplyr::all_of(columns_to_keep))

### 2025-12-17 analyses/infercnv-consensus-cell-type/scripts/build-normal-references.R

@@ -40,7 +40,7 @@ option_list <- list(
     type = "character",
     default = "",
     help = "Comma-separated list of reference group names to build references for.
-      Use the string 'normal' to include a combined set of normal cell types, which can be further specifed with the `normal_reference_groups` argument."
+      Use the string 'normal' to include a combined set of normal cell types, which can be further specified with the `normal_reference_groups` argument."
   ),
   make_option(
     opt_str = "--normal_reference_groups",

### 2025-12-17 analyses/metacells/scripts/run-seacells.py

@@ -66,7 +66,7 @@ def run_seacells(
         An AnnData object containing the data to run the SEACells algorithm on.
     cell_ratio : float
         The ratio of cells to metacells to use; i.e. number of cells per metacell
-        Default is 75, based on recommentations in https://github.com/dpeerlab/SEACells/blob/3462c624ffae0df6d3930490f345f00196c3503e/notebooks/SEACell_computation.ipynb
+        Default is 75, based on recommendations in https://github.com/dpeerlab/SEACells/blob/3462c624ffae0df6d3930490f345f00196c3503e/notebooks/SEACell_computation.ipynb
     min_cells : int
         The minimum number of cells for the SEACells algorithm to run,
         and the minimum number of metacells to create

### 2025-12-18 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
Spelling updates for #1276
NBAltas -> NBAtlas

@@
     "Label shown in figures" = label_recoded,
-    "Underlying NBAltas label(s)" = all_labels
+    "Underlying NBAtlas label(s)" = all_labels
   )
@@
 PDX samples have, not unexpectedly, much less cell type diversity overall with most cells being tumor or similar and immune cells rarely present.
-This is consistent with expectations in general for the biology of PDX samples.
+This is consistent with expectations in general for the biology of xenografts.
 

### 2025-12-19 analyses/cell-type-consensus/exploratory-notebooks/07-compare-consensus.Rmd

@@
 
-# combine into a single df and add a column to indiciate old/new 
+# combine into a single df and add a column to indicate old/new 
 df_list <- list(old_consensus_df, new_consensus_df) |> 

### 2025-12-19 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd

@@
 - Using `InferCNV` without a group of normal cells does not seem feasible for this particular sample since all tumor cells appear to have the same set of CNVs with little variation. 
-- We can define tumor cells by setting a cutoff for the number of CNVs that must be present in a cell, but I'm not sure how transferrable this is to other samples. 
+- We can define tumor cells by setting a cutoff for the number of CNVs that must be present in a cell, but I'm not sure how transferable this is to other samples. 
 - `InferCNV` can be helpful in looking at specific CNVs that are expected in a tumor type, something that is harder to look at in a tool like `CopyKAT`.

### 2025-12-19 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
caught one more ospl

@@
 |----|----|----|----|----|
-| normal | endothelial | `compartment == "endothelium" & predicted.score > pred.thr` | *VWF* | no CNV |
-| normal | immune | `compartment == "immune" & predicted.score > pred.thr` | *PTPRC*, *CD163*, *CD68* | no CNV |
-| normal | kidney | `compartment == "fetal_nephron" & predicted.score > pred.thr & cnv_score < cnv.thr.lower` | *CDH1*, *PODXL*, *LTL* | no CNV |
-| normal | stroma | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr.upper` | *VIM* | no CNV |
-| cancer | stroma | `compartment == "stroma" & cnv_score > cnv.thr.upper` | *VIM* | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
-| cancer | blastema | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr.upper` | *CITED1* | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
-| cancer | epithelial | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr.upper` | *CDH1* | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| normal | endothelial | `compartment == "endothelium" & predicted.score > pred_thr` | *VWF* | no CNV |
+| normal | immune | `compartment == "immune" & predicted.score > pred_thr` | *PTPRC*, *CD163*, *CD68* | no CNV |
+| normal | kidney | `compartment == "fetal_nephron" & predicted.score > pred_thr & cnv_score < cnv_threshold_low` | *CDH1*, *PODXL*, *LTL* | no CNV |
+| normal | stroma | `compartment == "stroma" & predicted.score > pred_thr & cnv_score < cnv_threshold_upper` | *VIM* | no CNV |
+| cancer | stroma | `compartment == "stroma" & cnv_score > cnv_threshold_upper` | *VIM* | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| cancer | blastema | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv_threshold_upper` | *CITED1* | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
+| cancer | epithelial | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv_threshold_upper` | *CDH1* | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
 | unknown | \- | the rest of the cells | \- | \- |
@@
 
-- `pred.thr` = `r params$predicted.celltype.threshold`
-- `cnv.thr.lower` = `r params$cnv_threshold_low`
-- `cnv.thr.upper` = `r params$cnv_threshold_high`
+- `pred_thr` = `r params$predicted.celltype.threshold`
+- `cnv_threshold_low` = `r params$cnv_threshold_low`
+- `cnv_threshold_upper` = `r params$cnv_threshold_high`
 
@@
 Our analysis suggest that there is indeed an enrichment in cells with CNV after chemotherapy, including a tendency of more *chr1q* gain samples after chemotherapy. 
-This could explain some of the differences observed between our mean CNV profile and the one published in [Cresswell et al.
-(2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
+This could explain some of the differences observed between our mean CNV profile and the one published in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
 

### 2025-12-19 analyses/hello-clusters/02_compare-clustering-parameters.Rmd
fix typo and re-render html

@@
 
-      # calculate neighbhorhood purity, starting from silhouette_df
+      # calculate neighborhood purity, starting from silhouette_df
       rOpenScPCA::calculate_purity(pca_matrix, silhouette_df)
@@
 In particular, we use faceting to help display all information in one plot, by placing nearest neighbor values on the X-axis and faceting by resolution values.
-Since silhouette width and neighhorbood purity calculations using generally similar code, we'll just show neighborhood purity here.
+Since silhouette width and neighborhood purity calculations using generally similar code, we'll just show neighborhood purity here.
 
